\begin{letter}{a}
\tbLX{a}\tbHM{} \tbPH{à} \tbPS{conj.} \tbSN \tbGE{and} \tbXV{O waɛ la a wa gaŋ.} \tbXE{S/he came and left.} \tbSN \tbGE{in order to} \tbXV{O waɛ la a na nyɛ te.} \tbXE{S/he has come in order to visit us.}
\tbLX{a}\tbHM{} \tbPH{á} \tbPS{conj.} \tbGE{but} \tbXV{A wola ka o ba wa?} \tbXE{But why didn't s/he come?} \tbVA{ana}
\tbLX{a}\tbHM{} \tbPH{à} \tbPS{pron.} \tbGE{they, them (3rd person plural weak form nonhuman)} \tbXV{Vɛŋ ka a kpɛ a zage.} \tbXE{Let them go into the pen.} \tbXV{De a moɔ eŋ a.} \tbXE{Give them the grass.}
\tbLX{a}\tbHM{} \tbPH{à} \tbPS{def. art.} \tbGE{the} \tbXV{A bie kpɛ la a die.} \tbXE{The child has gone into the room.}
\tbLX{a}\tbHM{} \tbPH{á} \tbPS{v.} \tbGE{to hate} \tbXV{Te a ba la.} \tbXE{We hate them.} \tbPD{aɛ, aara, aara, aareba, aaraa, aareba }
\tbLX{aa}\tbHM{} \tbPH{áà} \tbPS{interj.} \tbGE{expression of surprise and regret} \tbXV{ A ŋmane ŋmarɛɛ la, aa.} \tbXE{The calabash is broken; what a pity!} \tbXV{ Aa, ka N da baŋ ka lɛ N gaaɛ.} \tbXE{Oh, if I had known, I would have gone.}
\tbLX{aa}\tbHM{} \tbPH{áá} \tbPS{interj.} \tbGE{expression of regret}
\tbLX{aa}\tbHM{} \tbPH{àà} \tbPS{interj.} \tbGE{expression expressing  {``}I told you so."}
\tbLX{aae}\tbHM{} \tbPH{ááɪ́} \tbPS{interj.} \tbGE{expression of pain}
\tbLX{a-a}\tbHM{} \tbPH{á-à} \tbPS{interj.} \tbSN \tbGE{expression of disapproval} \tbXV{A-a, ta de ona.} \tbXE{Oh no, don't go that way.} \tbSN \tbGE{ why?; expression of doubt} \tbXV{A-a, fo naŋ ba gaa?} \tbXE{Well, you haven't gone yet?}
\tbLX{ãa}\tbHM{} \tbPH{ã̀ã́} \tbPS{adv.} \tbSN \tbGE{yes (answer to a call)} \tbXV{N boɔle la a bie ka o sage ka {``}ãa."} \tbXE{I called the child and he answered, {``}Yes."} \tbSN \tbGE{what? (exclamation of disbelief)} \tbXV{Ka Naa boɔle fo ka fo zagere, ãa.} \tbXE{You refuse to answere the chief's call, what?}
\tbLX{ãaa}\tbHM{} \tbPH{ã̀ã́ã̀} \tbPS{interj.} \tbGE{is that so? (expression of uncertain surprise)} \tbXV{Bayuoː A pɔgebil-pɛɛre na taa la poɔ. Ayuoː Ãaa?} \tbXE{Bayuo: That teenage girl is pregnant. Ayuo: Is that true?}
\tbLX{âaa}\tbHM{} \tbPH{ã̀ã́ã́} \tbPS{part.} \tbGE{answer in agreement} \tbXV{Fo na gaa la? âaa.} \tbXE{Will you go? Yes, I will.}
\tbLX{aa-ee}\tbHM{} \tbPH{áá-ɪ̀ɪ́} \tbPS{interj.} \tbGE{expression of lack of confidence in someone's ability to perform a task} \tbXV{Aa-ee, fo koŋ daŋ ma ta.} \tbXE{I don't think you can get there before me.}
\tbLX{aa-hii}\tbHM{} \tbPH{áá-hĩ̀ĩ̀} \tbPS{interj.} \tbGE{expression of total despair}
\tbLX{aane}\tbHM{} \tbPH{áánɪ̀} \tbPS{adj.} \tbGE{infertile (woman )} \tbOP{aama} \tbTP{aanɛɛ}
\tbLX{aaŋ}\tbHM{} \tbPH{áàŋ} \tbPS{interj.} \tbGE{well then}
\tbLX{aba}\tbHM{} \tbPH{ábà} \tbPS{interj.} \tbSN \tbGE{expression of despair over a repeated mishap} \tbXV{A bie la lee la, aba.} \tbXE{The child has fallen again, oh no!} \tbSN \tbGE{expression of satisfaction or agreement} \tbXV{Aba, lɛ ka N boɔrɔ.} \tbXE{Good, that is what I want.} \tbSN \tbGE{expression of empathy} \tbXV{Aba, mɔ kyɛ so a bie kõɔ} \tbXE{Try and bathe the child.}
\tbLX{abaa}\tbHM{} \tbPH{ábàà} \tbPS{interj.} \tbSN \tbGE{exclamation of satisfaction or commendation} \tbXV{Abaa, fo waa bie.} \tbXE{Well done, you are a good child.} \tbSN \tbGE{exclamation of despair or being distraught} \tbXV{Abaa, vɛŋ a zɔɔre lɛ.} \tbXE{Oh, stop the fight.}
\tbLX{abɛ}\tbHM{} \tbPH{àbɛ́} \tbPS{n.} \tbSN \tbGE{palm tree} \tbSN \tbGE{palm fruit}
\tbLX{abɛbiri}\tbHM{} \tbPH{àbɛ̀bírì} \tbPS{n.} \tbGE{palm fruit} \tbSG{abɛbiruu} \tbOP{abɛbie} \tbTP{abɛbiree}
\tbLX{abɛdãã}\tbHM{} \tbPH{àbɛ́dã́ã̀} \tbPS{n.} \tbGE{palm wine} \tbSG{abɛdããõ} \tbTP{abɛdaanɛɛ}
\tbLX{abɛkãã}\tbHM{} \tbPH{àbɛ́kã́ã̀} \tbPS{n.} \tbGE{palm oil} \tbTP{abɛkaanɛɛ}
\tbLX{abɛsaare}\tbHM{} \tbPH{ábɛ́sáárɪ̀} \tbPS{n.} \tbGE{palm-frond broom} \tbOP{abɛsaɛ} \tbTP{abɛsaarɛɛ}
\tbLX{abɛteɛ}\tbHM{} \tbPH{àbɛ̀tɪ̀ɛ̀} \tbPS{n.} \tbGE{palm tree} \tbOP{abɛteere} \tbTP{abɛteerɛɛ}
\tbLX{abuobo}\tbHM{} \tbPH{ábúóbò} \tbPS{interrogative pronoun} \tbGE{which ones (nonhuman)} \tbXV{Abuobo ka fo boɔrɔ?} \tbXE{Which ones do you want?}
\tbLX{Afereka}\tbHM{} \tbPH{áfɪ́rɪ́kà} \tbPS{n.} \tbGE{Africa} \tbOP{Aferekare} \tbTP{Aferekarɛɛ}
\tbLX{Aferekaneɛ}\tbHM{} \tbPH{áfíríkànɪ́ɛ́} \tbPS{n.} \tbGE{African} \tbOP{Aferekanoba}
\tbLX{afu}\tbHM{} \tbPH{àfú} \tbPS{n.} \tbGE{preface (of a book)} \tbOP{afuri} \tbTP{afuree}
\tbLX{age}\tbHM{} \tbPH{ágɪ́} \tbPS{v.} \tbSN \tbGE{to dance} \tbSN \tbGE{to jump up and down} \tbPD{agɛɛ, agra, agra, agreba, agraa, agreba } \tbSE{} \tbGE{}
\tbLX{age}\tbHM{} \tbPH{ágɪ̀} \tbPS{v.} \tbSN \tbGE{to fly} \tbSN \tbGE{erect, ejaculate (penis)} \tbSN \tbGE{to jump over} \tbSN \tbGE{to be frightened; to miss a heartbeat} \tbPD{agɛɛ, agra, agra, agreba, agraa, agreba } \tbSE{pɔlaa age} \tbGE{to be frightened}
\tbLX{agoo}\tbHM{} \tbPH{àgòò} \tbPS{v.} \tbGE{to ask for permission to join someone (knock door)} %\tbPD{-, -, -, -, -, -}
\tbLX{agoo}\tbHM{} \tbPH{àgòò} \tbPS{n.} \tbGE{a type of silky, colourful fabric} \tbOP{agoori} \tbTP{agooree}
\tbLX{agye}\tbHM{} \tbPH{ádzé} \tbPS{n.} \tbGE{bad luck} \tbOP{agyeri} \tbTP{agyeree}
\tbLX{ahâa}\tbHM{} \tbPH{àhã́ã́} \tbPS{interj.} \tbGE{that's right (expression of agreement)} \tbXV{Ahaa, o binzie la be.} \tbXE{That's right, that is its position.}
\tbLX{ahaŋ}\tbHM{} \tbPH{àháŋ} \tbPS{interj.} \tbGE{an expression that denotes {``}well?"} \tbXV{Ahaŋ a polisi la waana.} \tbXE{All is well, as the police are coming.}
\tbLX{ai}\tbHM{} \tbPH{àí} \tbPS{part.} \tbGE{no}
\tbLX{aii}\tbHM{} \tbPH{áìì} \tbPS{interj.} \tbGE{an expression of despair} \tbXV{Aii, a nɔgyɛle ŋmarɛɛ.} \tbXE{Oh no; the egg is broken!}
\tbLX{akaase}\tbHM{} \tbPH{àkáásɪ̀} \tbPS{interj.} \tbGE{expression of boastful challenge} \tbXV{Akaase, neɛ kyebe na daŋ ma kɔnoɔre.} \tbXE{I dare anyone to finish his portion of the weeding before me.}
\tbLX{akaasemanakei}\tbHM{} \tbPH{ákáásɪ̀mànà} \tbPS{interj.} \tbGE{expression of disappointment at failure} \tbXV{A bie ba toɔ tuo daga, akaasemanakei.} \tbXE{The child has failed to carry the box, what a pity.} \tbVA{akaase}
\tbLX{akai}\tbHM{} \tbPH{àkàí} \tbPS{interj.} \tbGE{expression of disgust and disagreement} \tbXV{ Akai, fo deme kyebe.} \tbXE{Too bad, you have no case.}
\tbLX{aku}\tbHM{} \tbPH{ákú} \tbPS{interj.} \tbGE{expression of despair, loss, empathy} \tbXV{A nannyigri la maaleŋ fa la n zɔmeŋɛ na, aku.} \tbXE{The thieves have robbed my friend again, how awful!} \tbVA{akuroku}
\tbLX{akuroku}\tbHM{} \tbPH{àkúrókú} \tbPS{interj.} \tbGE{emphatic form of aku} \tbVA{akuraku; aku}
\tbLX{akuu}\tbHM{} \tbPH{àkúú} \tbPS{interj.} \tbGE{expression indicating one's inability to do a task} \tbXV{Akuu, ta ko ma, N koŋ toɔ̂ a tuori tuo.} \tbXE{Ei, don't kill me; I can't carry the load.}
\tbLX{Ala}\tbHM{} \tbPH{álà} \tbPS{n.} \tbGE{God}
\tbLX{ala}\tbHM{} \tbPH{álà} \tbPS{interj.} \tbGE{I swear} \tbXV{Ala, a dɔɔ ŋa naane so a woɔ.} \tbXE{I swear, the bag does not belong to this man.}
\tbLX{alaakosebaroo}\tbHM{} \tbPH{àláákosɪ́bárʊ́ʊ́} \tbPS{interj.} \tbGE{in the name of God}
\tbLX{ale}\tbHM{} \tbPH{álɪ̀} \tbPS{v.} \tbGE{to open (ring), to stretch out} \tbXV{Ka fooŋ ba ale a fenetere, o koŋ kpɛ a fo nubiri.} \tbXE{If you don't open up the ring you can't put it on your finger.} \tbPD{alɛɛ, alla, alla, aleba, allaa}%, , -}
\tbLX{alee}\tbHM{} \tbPH{àléé} \tbPS{part.} \tbGE{expression ordering one to get out; away with you} \tbXV{A bibiiri naŋ di a bondirii baare la ka ba ba yeli ka alee, ka ba wa yi gaa sakuuri.} \tbXE{When the children had finished eating, their father ordered them to go to school at once.} %\tbPD{-, -, -, -, -, -}
\tbLX{aleɛma}\tbHM{} \tbPH{àlɪ́ɛ́mà} \tbPS{n.} \tbGE{metaphor (literature)}
\tbLX{alɛ}\tbHM{} \tbPH{álɛ̀} \tbPS{interj.} \tbGE{expression of surprise mixed with fear} \tbXV{Alɛ, fentel diglii ka fo deɛnɛ ne?} \tbXE{Whoa, you're playing with electricity?}
\tbLX{aloo}\tbHM{} \tbPH{álòó} \tbPS{interj.} \tbGE{hello}
\tbLX{aloopelee}\tbHM{} \tbPH{álóópèléè} \tbPS{n.} \tbGE{aeroplane} \tbOP{aloopelli} \tbTP{aloopellee} \tbVA{ayɔɔmapelee}
\tbLX{aloopeleeduoraa}\tbHM{} \tbPH{álóópèléédúóráà} \tbPS{n.} \tbGE{airport} \tbOP{aloopeleeduori}
\tbLX{ama}\tbHM{} \tbPH{àmá} \tbPS{conj.} \tbGE{but, however} \tbXV{ A bɔlŋmeɛrebɛ mɔɛ la ama ba ba di.} \tbXE{The football players played well, but they did not win.}
\tbLX{ama}\tbHM{} \tbPH{ámà} \tbPS{pron.} \tbGE{these (3rd person plural demonstrative nonhuman)} \tbXV{A dogere ama daaroŋ tuo.} \tbXE{These pots are expensive.}
\tbLX{ambaŋenaa}\tbHM{} \tbPH{ámbáŋɪ́náá} \tbPS{n.} \tbGE{a type of small tree with large leaves and pink fruit} \tbOP{ambaŋene} \tbTP{ambaŋenɛɛ}
\tbLX{Ameleka}\tbHM{} \tbPH{ámɪ́lɪ́kà} \tbPS{n.} \tbGE{ America} \tbOP{amelekare} \tbTP{amelekarɛɛ} \tbVA{Hameleka}
\tbLX{aminekaŋa}\tbHM{} \tbPH{àmínékàŋá} \tbPS{adv.} \tbSN \tbGE{perhaps, maybe} \tbXV{Aminekaŋa o koŋ wa pampana.} \tbXE{Perhaps s/he will not come now.} \tbSN \tbGE{sometimes} \tbVA{minekaŋa}
\tbLX{amma}\tbHM{} \tbPH{ámmá} \tbPS{conj.} \tbGE{much more than (exaggeration)} \tbXV{A bie zoro to, amma soɔŋaa.} \tbXE{The boy runs really fast, much faster than the hare.}
\tbLX{ammeseɛrɛ}\tbHM{} \tbPH{ámmɪ̀sɪ́ɛ́rɛ̀} \tbPS{n.} \tbGE{thanks} \tbXV{Fo taa la ammeseɛrɛ ne fo naŋ toŋ a toma.} \tbXE{I am thankful to you for doing the work.} \tbTP{ammeseɛrɛɛ}
\tbLX{ammine}\tbHM{} \tbPH{ámmìné} \tbPS{pron.} \tbGE{which people (3rd person plural human interrogative pronoun)}
\tbLX{ampe}\tbHM{} \tbPH{ámpé} \tbPS{n.} \tbGE{a game traditionally played by women} \tbOP{amperi} \tbTP{amperee}
\tbLX{ana}\tbHM{} \tbPH{áná} \tbPS{pron.} \tbGE{those (3rd person plural strong form nonhuman)} \tbXV{Ana ka te na wuo.} \tbXE{We'll collect those.}
\tbLX{anaare}\tbHM{} \tbPH{ànáárɪ̀} \tbPS{adj.} \tbGE{four} \tbOP{anaare} \tbTP{anaarɛɛ} \tbVA{naare}
\tbLX{ananso}\tbHM{} \tbPH{ánánsʊ́} \tbPS{conj.} \tbGE{because} \tbXV{O naŋ taa libie ananso ka o ba mɛ yiri.} \tbXE{S/he built a house because s/he has money.} \tbVA{ana la so}
\tbLX{ananso}\tbHM{} \tbPH{ánánsʊ́} \tbPS{n.} \tbGE{reason} \tbXV{A fo toma zagere gaabo ananso la wola?} \tbXE{What is your reason for refusing to go to work?}
\tbLX{anaŋ}\tbHM{} \tbPH{ánǎŋ} \tbPS{pron.} \tbGE{those ones (3rd person plural demonstrative)} \tbXV{Te na iri la anaŋ naŋ sãa te wuli a boŋkoɔrɔ.} \tbXE{We'll select the ones that are bad and show them to the seller.}
\tbLX{anaŋ}\tbHM{} \tbPH{ánáŋ} \tbPS{pron.} \tbGE{it (a problem, a situation)} \tbXV{Ka anaŋ waa lɛ, ta di a seemaa.} \tbXE{If that is the case, don't eat the food.}
\tbLX{andonɛɛ}\tbHM{} \tbPH{ándʊ́nɛ́ɛ́} \tbPS{n.} \tbSN \tbGE{world} \tbXV{Andonɛɛ waa la yɛloŋ.} \tbXE{The world is large.} \tbSN \tbGE{people of the world} \tbXV{Andonɛɛ ba baŋ ka a wono ba mine.} \tbXE{People of the world do not know that some people are suffering.}
\tbLX{ane}\tbHM{} \tbPH{ànɪ̀} \tbPS{conj.} \tbSN \tbGE{and} \tbXV{Dɛre ane Daŋyi maŋ daŋ la sakuuri gaabo.} \tbXE{Dɛre and Daŋyi always go to school early.} \tbSN \tbGE{although, despite} \tbXV{Ane ka zie e la kpeɛŋaa kyɛ noba mine naŋ ba sage dɔge seŋ lɛ puori tuubu.} \tbXE{Although times are hard, some people do not want to practise family planning.} \tbSN \tbGE{with} \tbXV{O pɔge daa la a dɔɔ ane o faŋa zaa ka o te le.} \tbXE{The woman pushed the man with all her strength and he fell over.}
\tbLX{aneazaa}\tbHM{} \tbPH{ánɪ́ázàá} \tbPS{conj.} \tbGE{despite, even though, in spite of} \tbXV{Aneazaa ka saa ba miire kyɛ kõɔ naŋ be la a baa poɔ.} \tbXE{Even though it has not been raining, there is still water in the pond.}
\tbLX{aneɛ}\tbHM{} \tbPH{ánɪ́ɛ̀} \tbPS{n.} \tbGE{effort, attempt} \tbXV{A tontonema da erɛ la aneɛ na baare a toma kyɛ ka saa ba sage.} \tbXE{The workers tried to finish the job but the rain would not allow them.}
\tbLX{anesɛ}\tbHM{} \tbPH{ánɪ́sɛ̀} \tbPS{adj.} \tbGE{welcome (an expression of goodwill to a stranger who has just arrived)} \tbXV{Anesɛ, kpeɛrɛ waana.} \tbXE{Welcome, come in!} %\tbPD{-, -, -, -, -, -}
\tbLX{angyoɔso}\tbHM{} \tbPH{ándzʊ̀ɔ̀sʊ́} \tbPS{n.} \tbGE{gratitude, thanks} \tbXV{Fo taa la angyoɔso ne a kyɔɔtaa fo naŋ ko ma.} \tbXE{Accept my thanks for the gift.} \tbOP{angyoɔsere} \tbTP{angyoɔserɛɛ}
\tbLX{anii}\tbHM{} \tbPH{àníì} \tbPS{adj.} \tbGE{eight} \tbOP{aniini} \tbTP{aniinee}
\tbLX{anlee}\tbHM{} \tbPH{ánléé} \tbPS{n.} \tbSN \tbGE{a traditional women's dance} \tbSN \tbGE{music for such a dance} \tbOP{anleeri} \tbTP{anlerri/anleree} \tbVA{an-yee}
\tbLX{annoo}\tbHM{} \tbPH{ánnʊ́ʊ́} \tbPS{pron.} \tbGE{who;  which person (interrogative pronoun)} \tbVA{aŋ la o}
\tbLX{ansaŋ}\tbHM{} \tbPH{ánsáŋ} \tbPS{conj.} \tbGE{before} \tbXV{Vɛŋ ka saa mi ansaŋ ka fo bore a kamaana.} \tbXE{Let it rain before you sow the maize.}
\tbLX{ansiŋinee}\tbHM{} \tbPH{ánsíŋínéé} \tbPS{n.} \tbSN \tbGE{a type of edible berry that turns violet when ripe} \tbSN \tbGE{the tree of this fruit} \tbSG{ansiŋinuu} \tbOP{ansiŋini} \tbVA{ansiinee}
\tbLX{ansoɔnee}\tbHM{} \tbPH{ánsʊ́ɔ́nɪ́ɪ́} \tbPS{n.} \tbGE{a woman who has recently given birth} \tbOP{ansoɔne/ansoɔmɔ} \tbTP{ansoɔnɛɛ}
\tbLX{antere}\tbHM{} \tbPH{ántèré} \tbPS{n.} \tbGE{noon; used as a greeting, good afternoon (Southern Dagaare dialect)} %\tbPD{-, -, -, -, -, -}
\tbLX{anuu}\tbHM{} \tbPH{ànúú} \tbPS{adj.} \tbGE{five} \tbOP{anuuri} \tbTP{anuuree}
\tbLX{anwoŋ}\tbHM{} \tbPH{ánwóŋ} \tbPS{n.} \tbGE{one who is great or strong (physically or spiritually)} \tbOP{annwonni} \tbTP{anwonnee}
\tbLX{aŋ}\tbHM{} \tbPH{ǎŋ} \tbPS{adj.} \tbGE{to be alerted; used to caution or persuade} \tbXV{Ka fo wa gɛrɛ a libidaana na yiri, aŋ baŋ ne baare.} \tbXE{If you are going to that rich man’s house you'd better beware of the dogs.} %\tbPD{-, -, -, -, -, -}
\tbLX{aŋ}\tbHM{} \tbPH{áŋ} \tbPS{pron.} \tbGE{who (interrogative pronoun)} \tbXV{Aŋ la ba nɔŋ bompaala?} \tbXE{Who does not like new things?}
\tbLX{aŋa}\tbHM{} \tbPH{áŋá} \tbPS{n.} \tbGE{no one (retort; implies the matter is none of the asker's business)} \tbXV{Aŋ la wa a kyɛ? Aŋa la wa.} \tbXE{Who came here? No one. (Why do you care?)}
\tbLX{aŋa}\tbHM{} \tbPH{àŋá} \tbPS{adv.} \tbGE{like} \tbXV{O kono la aŋa bibile na.} \tbXE{S/he cries like a baby.} \tbVA{ŋa}
\tbLX{aŋgaara}\tbHM{} \tbPH{áŋgáárá} \tbPS{adj.} \tbGE{having the blades burnt away, leaving the stalks (used for grass)} \tbXV{A kyeɛ zo kpɛ la a mɔ-aŋgaara poɔ.} \tbXE{The squirrel has escaped into the stalky bush.} \tbOP{aŋgaare} \tbTP{aŋgaarɛɛ}
\tbLX{aŋgeli}\tbHM{} \tbPH{áŋgélí} \tbPS{n.} \tbGE{a type of long grass used to weave sleeping mats} \tbOP{aŋgelli} \tbTP{aŋgelee}
\tbLX{aŋgoɔle}\tbHM{} \tbPH{áŋgʊ́ɔ̀lɪ̀} \tbPS{v.} \tbGE{to adorn, to decorate} \tbXV{Ba aŋgoɔle la a naa wiri.} \tbXE{They have decorated the chief's horse.} \tbPD{aŋgoɔlɛɛ, aŋgoɔlɔ, aŋgoɔlɔ, aŋgoɔleba, aŋgoɔlaa}%, , -}
\tbLX{aŋguoŋaa}\tbHM{} \tbPH{áŋgúóŋáà} \tbPS{n.} \tbGE{a type of tree with round, pink fruit} \tbOP{aŋguoni/aŋguomo} \tbTP{aŋguonee} \tbVA{pl.var. aŋguôaa}
\tbLX{aŋkaa}\tbHM{} \tbPH{àŋkàá} \tbPS{n.} \tbGE{orange (fruit)} \tbSG{aŋkaaroo} \tbOP{aŋkaare/aŋkaɛ} \tbTP{aŋkarɛɛ}
\tbLX{Aŋkara}\tbHM{} \tbPH{àŋkàrà} \tbPS{n.} \tbGE{Accra} \tbOP{Aŋkarre} \tbTP{Aŋkarrɛɛ}
\tbLX{aŋko}\tbHM{} \tbPH{áŋkò} \tbPS{n.} \tbGE{river blindness} \tbOP{aŋkori} \tbTP{aŋkoree}
\tbLX{aŋkorɔ}\tbHM{} \tbPH{áŋkʊ̀rɔ́} \tbPS{n.} \tbGE{barrel, drum} \tbOP{aŋkorre} \tbTP{aŋkorɛɛ/aŋkorrɛɛ} \tbVA{aŋkora}
\tbLX{ão}\tbHM{} \tbPH{ã̀ʊ̃̀} \tbPS{interj.} \tbGE{of course (expression of certainty)} \tbXV{A. Fo na kuli la be? B. ão.} \tbXE{A. Will you go home? B. Of course I willl.}
\tbLX{apa}\tbHM{} \tbPH{ápà} \tbPS{interj.} \tbGE{expression of unpleasant surprise} \tbXV{A soɔŋaa muri la a baa bare kyɛ zo, apa.} \tbXE{The rabbit dodged the dog and escaped, what a pity.} \tbVA{aparapa}
\tbLX{aparapa}\tbHM{} \tbPH{ápárápà} \tbPS{interj.} \tbGE{emphatic expression of despair} \tbXV{Aparapa, a soɔŋaa ŋa meŋ la zoe?} \tbXE{Oh no, this hare too has escaped?} \tbVA{apa}
\tbLX{ara}\tbHM{} \tbPH{àrá} \tbPS{adj.} \tbGE{to gather up one's courage} \tbXV{A pɔge ara kyɛ daa la a dɔɔlee eŋ tambɔgeŋ.} \tbXE{The woman gathered up her courage and pushed the man into a ditch.} %\tbPD{-, -, -, -, -, -}
\tbLX{arambiri}\tbHM{} \tbPH{árámbírí} \tbPS{n.} \tbGE{kidney} \tbXV{A dɔɔta yeli ka a baala arambiri kaŋa sãaɛ la.} \tbXE{The doctor said one of the patient's kidneys had failed.} \tbSG{arambiruu} \tbOP{arembie} \tbTP{arambiiree}
\tbLX{are}\tbHM{} \tbPH{árɪ́} \tbPS{v. pl.} \tbGE{to cut, harvest (maize)}
\tbLX{are}\tbHM{} \tbPH{árɪ́} \tbPS{v.} \tbGE{to be related through the mother's line} \tbXV{N are la Kusielee.} \tbXE{I am maternally related to the Kusielee clan.} \tbPD{arɛɛ, ara, ara, areba, araa}%, , -}
\tbLX{are}\tbHM{} \tbPH{àrɪ̀} \tbPS{v.} \tbSN \tbGE{to stand} \tbSN \tbGE{to wait} \tbXV{N baŋ ka o na are la ka N gaa te pɔge o.} \tbXE{I know s/he will wait so that I go to meet him/her.} \tbSN \tbGE{to stop} \tbXV{A wɔɔkye kyɛŋɛɛ la tasoga tasoga leɛ are.} \tbXE{The watch worked for a while and then stopped.} \tbPD{arɛɛ, ara, ara, areba, araa}%, , -}
\tbLX{are}\tbHM{} \tbPH{árɪ̀} \tbPS{v.} \tbGE{to tear off from the main body (branch or limb of an animal)} \tbXV{A maŋgo wɔŋ zue la ka a ulee kaŋa are le.} \tbXE{The mango tree has borne so many fruits that one of the branches tore off.} \tbPD{arɛɛ, ara, ara, areba, araa, areba }
\tbLX{areba}\tbHM{} \tbPH{árɪ́bá} \tbPS{n.} \tbGE{maternal uncle} \tbOP{aremine} \tbTP{arrɛɛ}
\tbLX{arekpoŋ}\tbHM{} \tbPH{árɪ́kpóŋ} \tbPS{n.} \tbGE{eldest niece} \tbOP{arekponni} \tbTP{arekponnee}
\tbLX{arelee}\tbHM{} \tbPH{árɪ́lèé} \tbPS{n.} \tbGE{nephew or niece} \tbOP{arebilii} \tbTP{arebilee/arebillee}
\tbLX{aremateke}\tbHM{} \tbPH{àrɪ́mátɪ́kɪ̀} \tbPS{n.} \tbGE{arithmetic} \tbOP{arematekere} \tbVA{kuntaŋ (Akan)}
\tbLX{arezie}\tbHM{} \tbPH{àrɪ̀zíé} \tbPS{n.} \tbSN \tbGE{standing place, spot} \tbSN \tbGE{post, position, rank} \tbXV{O arezie kpoŋ.} \tbXE{S/he holds a high position.} \tbSN \tbGE{position, stand (for or against an idea)} \tbXV{O ba wuli o arezie a nɔkpeɛne poɔ.} \tbXE{She has not indicated her position with regard to the argument.} \tbOP{areziiri} \tbTP{areziiree}
\tbLX{aroozie}\tbHM{} \tbPH{árʊ́ʊ́zíé} \tbPS{n.} \tbSN \tbGE{standing place; position} \tbSN \tbGE{adverb of place (on, under)} \tbOP{arooziiri} \tbTP{aroo\-zii\-ree}
\tbLX{aseŋ}\tbHM{} \tbPH{àsɪ́ŋ} \tbPS{adv.} \tbGE{for instance, like}
\tbLX{asibiti}\tbHM{} \tbPH{ásíbítì} \tbPS{n.} \tbGE{hospital, clinic} \tbOP{asibitiri} \tbTP{asibitiree} \tbVA{asiiti, asimeti, asebere}
\tbLX{ata}\tbHM{} \tbPH{átà} \tbPS{adj.} \tbGE{three} \tbOP{atare} \tbTP{atarɛɛ} \tbSE{ata ata} \tbGE{in threes}
\tbLX{awae}\tbHM{} \tbPH{àwáɪ̀} \tbPS{adj.} \tbGE{nine} \tbOP{awarre} \tbTP{awarɛɛ/awarrɛɛ} \tbVA{wae}\tbSE{awae awae} \tbGE{nine each}
\tbLX{awola}\tbHM{} \tbPH{àwʊ́là} \tbPS{interrogative pronoun} \tbSN \tbGE{how much (price)} \tbXV{Awola ka fo da a kparoo?} \tbXE{How much did the shirt cost you?} \tbSN \tbGE{how many (nonhuman)} \tbXV{Teere awola ka yɛ sɛle?} \tbXE{How many trees did you plant?}
\tbLX{awommu}\tbHM{} \tbPH{àwómmú} \tbPS{n.} \tbGE{hardship, suffering} \tbXV{Ka kɔŋeŋ le awommu maŋ pɔge la noba yaga.} \tbXE{When there is a famine, many people experience a lot of suffering.} \tbTP{awonnee} \tbVA{awonaa}
\tbLX{awulo}\tbHM{} \tbPH{áwúló} \tbPS{n.} \tbGE{continuous, loud, sharp cry of pain} \tbXV{A pɔge na bie naŋ kpi la lɔɔrɔ awulo.} \tbXE{The woman who lost her son is crying hysterically.} \tbTP{awulee}
\tbLX{ayi}\tbHM{} \tbPH{áyì} \tbPS{adj.} \tbGE{two} \tbOP{ayiri} \tbTP{ayiree} \tbSE{ayi ayi} \tbGE{in twos}
\tbLX{ayililili}\tbHM{} \tbPH{áyílílílí} \tbPS{interj.} \tbGE{joyous, shrill shouting in commendation (usually by women)} \tbXV{A pɔgeba ŋmɛ la kulinni, ayililili, eŋ a bawaseɛrebɛ.} \tbXE{The women made shrill noises of commendation for the dancers.} \tbVA{ayiyiyiyi}
\tbLX{ayoɔbo}\tbHM{} \tbPH{àyʊ̀ɔ̀bʊ́} \tbPS{adj.} \tbGE{six} \tbOP{ayoɔre} \tbTP{ayoɔrɛɛ}
\tbLX{Ayoree}\tbHM{} \tbPH{áyóréé} \tbPS{n.} \tbGE{name of a clan} \tbOP{Ayoruu}
\tbLX{Ayɔɔ}\tbHM{} \tbPH{áyɔ̀ɔ̀} \tbPS{n.} \tbGE{name traditionallly given to a female child born on a market day} \tbOP{Ayɔɔmine} \tbTP{Ayɔɔrɛɛ}
\tbLX{ayɔpõĩ}\tbHM{} \tbPH{àyʊ̀ɔ̀pṍĩ̀} \tbPS{adj.} \tbGE{seven} \tbOP{ayoɔ\-pon\-ni} \tbTP{ayoɔponnee}
\tbLX{Ayuo}\tbHM{} \tbPH{áyúó} \tbPS{n.} \tbGE{name traditionally given to a female child born on a funeral day} \tbOP{Ayumine} \tbTP{Ayuu\-ree}
\tbLX{Azaasoba}\tbHM{} \tbPH{àzàásʊ́bá} \tbPS{n.} \tbGE{almighty}
\end{letter}
