
\begin{letter}{b}
\tbLX{ba}\tbHM{} \tbPH{bá} \tbPS{adv.} \tbGE{not (negative particle)} \tbXV{N ba nyɛ o zenɛ.} \tbXE{I have not seen him today.}
\tbLX{ba}\tbHM{} \tbPH{bà} \tbPS{pron.} \tbSN \tbGE{they, them (3rd person plural weak form human)} \tbXV{Ba na wa la.} \tbXE{They will come.} \tbSN \tbGE{their (3rd person possessive human)} \tbXV{Te de la ba bondirii biŋ.} \tbXE{We have reserved their food.} \tbSN \tbGE{them (human)} \tbXV{Te de la a bondirii ko ba.} \tbXE{We have given them the food.}
\tbLX{ba}\tbHM{} \tbPH{bà} \tbPS{v.} \tbGE{to fix a pointed object firmly in the ground} \tbXV{Ba ba la daare gɔlle a dabɔgɔ zaa.} \tbXE{They have put up planks around the whole garden.} \tbPD{baɛ, baara, baara, bareba, baaraa}%, , -}
\tbLX{ba}\tbHM{} \tbPH{bà} \tbPS{v.} \tbGE{to take a long drink from a container} \tbXV{A saana de la a kõɔ ba lɛɛ ka a ŋmane wa te kyɛre.} \tbXE{The stranger drank deeply until the calabash was empty.} \tbPD{baɛ, baara, baara, bareba, baaraa}%, , -}
\tbLX{ba}\tbHM{} \tbPH{bá} \tbPS{v.} \tbGE{to gallop} \tbPD{baɛ, baara, baara, bareba, baaraa}%, , -}
\tbLX{ba}\tbHM{} \tbPH{bâ} \tbPS{n.} \tbSN \tbGE{father} \tbSN \tbGE{friend (between males)} \tbSN \tbGE{Mister} \tbOP{bamine} \tbTP{barɛɛ}
\tbLX{baa}\tbHM{} \tbPH{báà} \tbPS{conj.} \tbGE{or (normally used in question tag)} \tbXV{Te na gaa la, baa wolɔ?} \tbXE{We'll go, won't we?} \tbVA{bee}
\tbLX{baa}\tbHM{} \tbPH{báá} \tbPS{n.} \tbGE{dog} \tbOP{baare} \tbTP{baarɛɛ}
\tbLX{baa}\tbHM{} \tbPH{báá} \tbPS{n.} \tbGE{a small black ground ant-like creature which burrows into the ground with the tip of its abdomen} \tbOP{baare} \tbTP{baarɛɛ}
\tbLX{baa}\tbHM{} \tbPH{báá} \tbPS{n.} \tbGE{a type of children's game} \tbOP{baare} \tbTP{baarɛɛ}
\tbLX{baa}\tbHM{} \tbPH{bàá} \tbPS{n.} \tbGE{body of water} \tbOP{bare} \tbTP{barɛɛ}
\tbLX{baa}\tbHM{} \tbPH{bàà} \tbPS{v.} \tbSN \tbGE{to grow up, to flourish} \tbXV{Sẽɛ saŋa, moɔ maŋ baa la wogiri lɛ.} \tbXE{During the rainy season, grass normally grows tall.} \tbSN \tbGE{to mature} \tbXV{A pɔgelee naŋ ba baa azuiŋ o ba ta serɛ kuluu.} \tbXE{The girl has not yet matured enough, so she is not marriageable.} \tbSN \tbGE{to survive, to live} \tbXV{A baala naŋ ba kpi zenɛ o na baa la.} \tbXE{Since the patient has not died today, s/he will survive.} \tbPD{baa, baaɛ, baara, baara, baaraa, baabo}
\tbLX{baabaa}\tbHM{} \tbPH{bààbáá} \tbPS{n.} \tbGE{father (familiar term of address used by toddlers)} \tbOP{baabaamine} \tbTP{baabaarɛɛ}
\tbLX{Baabol}\tbHM{} \tbPH{báábʊ̀l} \tbPS{n.} \tbGE{Bible} \tbOP{baabola} \tbTP{baabolɛɛ/baabollɛɛ}
\tbLX{bãabõo}\tbHM{} \tbPH{bã̀ã̀bʊ̃́ʊ̃́} \tbPS{n.} \tbGE{bustard} \tbOP{bãaboomɔ/bãaboone} \tbTP{bãaboomɛɛ/bãaboonɛɛ}
\tbLX{baaebaae}\tbHM{} \tbPH{bááɪ́bááɪ́} \tbPS{interj.} \tbGE{goodbye}
\tbLX{baage}\tbHM{} \tbPH{báágɪ̀} \tbPS{n.} \tbGE{bag} \tbOP{baagere} \tbTP{bagerɛɛ} \tbVA{baagye}
\tbLX{baala}\tbHM{} \tbPH{bààlá} \tbPS{n.} \tbGE{patient, sick person} \tbOP{baalba} \tbTP{baalbɛɛ}
\tbLX{baalaa}\tbHM{} \tbPH{bààlàà} \tbPS{adj.} \tbSN \tbGE{slim, slender, thin} \tbXV{N da nyɛ la pɔgebil baalaa kaŋa a seɛre zie.} \tbXE{I saw a slender girl at the dance.} \tbSN \tbGE{narrow} \tbXV{A sori waa la baalaa lɛ.} \tbXE{The road is very narrow.} \tbOP{baale} \tbTP{baalɛɛ}
\tbLX{baale}\tbHM{} \tbPH{báálɪ̀} \tbPS{v.} \tbGE{to ache slightly (disconfort of the stomach)} \tbXV{Ka maaŋ iri ba gaa bangyeraa n poɔ maŋ baala la.} \tbXE{If I don’t go to the toilet for the day I normally have a slight stomachache.} \tbPD{baalɛɛ, baala, baala, baaleba, baalaa}%, , -}
\tbLX{baale}\tbHM{} \tbPH{báálɪ̀} \tbPS{v.} \tbGE{to separate (grains from unwanted material)} \tbXV{A pɔge baale iri la a seŋkãabie yi a pɛgere poɔ.} \tbXE{The woman separated the groundnuts from the shells.} \tbPD{baalɛɛ, baala, baala, baaleba, baalaa}%, , -}
\tbLX{baalebayiri}\tbHM{} \tbPH{bààlbáyírì} \tbPS{n.} \tbGE{hospital} \tbOP{baalebayie} \tbTP{baalebayiree}
\tbLX{baalimbo}\tbHM{} \tbPH{bààlímbò} \tbPS{n.} \tbGE{the act of supporting someone by holding them by the midsection} \tbXV{Ba kpɛ la a baala baalimbo gaa ne a asibiti.} \tbXE{They supported the patient by the midsection and took him to the hospital.} \tbOP{baalimbori} \tbTP{baalimboree} \tbVA{boolimbo}
\tbLX{baalombiri}\tbHM{} \tbPH{bààlʊ́mbírì} \tbPS{n.} \tbGE{germ, bacterium} \tbOP{baalombie} \tbTP{baalombiiree}
\tbLX{baalomboɔlaa}\tbHM{} \tbPH{báálʊ̀mbʊ̀ɔ̀làá} \tbPS{n.} \tbGE{feast organised for a friend} \tbXV{Te daare gaa la baalomboɔlaa Loraa poɔ.} \tbXE{We attended a friendship feast in Lawra.} \tbOP{baalomboɔle}
\tbLX{baalonlɔnnaa}\tbHM{} \tbPH{bààlʊ̀nlɔ́nnáá} \tbPS{n.} \tbGE{contagious disease} \tbOP{baalonlɔnnɔ} \tbTP{baalonlɔnnɛɛ}
\tbLX{baalonnooraa}\tbHM{} \tbPH{bààlʊ́nnʊ̀ʊ̀ráà} \tbPS{n.} \tbGE{recurring illness} \tbOP{baalonnoore}
\tbLX{baaloŋ}\tbHM{} \tbPH{báálʊ́ŋ} \tbPS{n.} \tbGE{friendship} \tbXV{Ba nyɔge la baaloŋ saŋa zaa.} \tbXE{They have been friends for a long time.} \tbOP{baalondeme}
\tbLX{baaloŋ}\tbHM{} \tbPH{bààlʊ́ŋ} \tbPS{n.} \tbGE{illness, sickness} \tbXV{A baala baaloŋ seɛ o la zenɛ.} \tbXE{The patient's illness has improved today.}
\tbLX{baaloŋ}\tbHM{} \tbPH{bààlʊ̀ŋ} \tbPS{adj.} \tbSN \tbGE{calm} \tbXV{Zenɛ, a mane kõɔ̃ e la baaloŋ.} \tbXE{Today, the sea is calm.} \tbSN \tbGE{gentle, gently} \tbXV{A doɔ baaloŋ na wa la kyɛ.} \tbXE{That gentle man came here.} \tbXV{A doɔ wa la baaloŋ.} \tbXE{The man is gentle.} \tbXV{De a bileɛ baaloŋ.} \tbXE{Take the baby gently.} \tbSN \tbGE{soft (objects, sounds)} \tbXV{A kapuro ŋa e la baaloŋ yelmeŋɛ.} \tbXE{This pillow is really soft.} \tbSN \tbGE{smooth (powdery substances)} \tbXV{Bɛnzɔŋ maŋ waa la baaloŋ.} \tbXE{Bean flour is usually smooth.}
\tbLX{baalugiri}\tbHM{} \tbPH{bààlùgìrí} \tbPS{n.} \tbGE{armpit} \tbSG{baalugiruu} \tbOP{baalugo} \tbTP{baalugiree} \tbVA{bagelugri}
\tbLX{baaluu}\tbHM{} \tbPH{bààlúú} \tbPS{n.} \tbSN \tbGE{gallon measure} \tbSN \tbGE{balloon} \tbSG{baaluuruu} \tbOP{baaluuri} \tbTP{baaluuree}
\tbLX{baa-ne-ŋmaaŋa}\tbHM{} \tbPH{bàá-nɪ́-ŋmááŋà} \tbPS{n.} \tbGE{Ursa Major (constellation); also known as Big Dipper, Plough, Great Bear}
\tbLX{baapɛmpɛle}\tbHM{} \tbPH{bàápɛ̀mpɛ́lɛ̀} \tbPS{n.} \tbGE{lily pads (pl.)} \tbSG{baapɛmpɛloo} \tbOP{baapɛmpɛlɛ} \tbTP{baapɛmpɛlɛɛ/baapɛmpɛllɛɛ}
\tbLX{baaraa}\tbHM{} \tbPH{bááráá} \tbPS{n.} \tbGE{end, last} \tbXV{A baaraa zaa te na nyɛ la a nannyige.} \tbXE{At the end of it all, we will find the thief.}
\tbLX{baaraakyi}\tbHM{} \tbPH{báárááʧì} \tbPS{n.} \tbGE{bridge} \tbOP{baaraakyiri} \tbTP{baaraakyiree}
\tbLX{baare}\tbHM{} \tbPH{báárɪ̀} \tbPS{v.} \tbSN \tbGE{to finish, to end, to complete} \tbXV{A tontonema baare la a toma.} \tbXE{The workers have finished the work.} \tbSN \tbGE{to be used up} \tbXV{A baaloŋ vɛŋɛɛ la ka a dɔɔ bɛroŋ na baare la o zaa.} \tbXE{Sickness has caused that man to lose weight.} \tbPD{baarɛɛ, baara, baara, baareba, baaraa}%, , -}
\tbLX{baasaabiri}\tbHM{} \tbPH{bààsààbírí} \tbPS{n.} \tbSN \tbGE{a seed from a type of tree} \tbSN \tbGE{a type of large bead made from this seed} \tbSG{baasaabiruu} \tbOP{baasaabie} \tbTP{baasaabiiree}
\tbLX{Baasaale}\tbHM{} \tbPH{bàásààlɪ̀} \tbPS{n.} \tbGE{Baasare (ethnic group)} \tbSG{baasaaloo}
\tbLX{baasakuuri}\tbHM{} \tbPH{báásàkúúrì} \tbPS{n.} \tbGE{bicycle} \tbOP{baasakue} \tbTP{baasakuuree}
\tbLX{baataŋaa}\tbHM{} \tbPH{báátáŋáá} \tbPS{n.} \tbGE{a type of tree that resembles the shea nut tree} \tbOP{baataŋema} \tbTP{baataŋemɛɛ/baataanɛɛ}
\tbLX{babammo}\tbHM{} \tbPH{bábámmʊ́} \tbPS{n.} \tbGE{ignorance}
\tbLX{babaŋena}\tbHM{} \tbPH{bábáŋɪ́ná} \tbPS{n.} \tbGE{one who is innocent} \tbOP{babaŋenema} \tbTP{babaŋenemɛɛ}
\tbLX{babaŋyeli}\tbHM{} \tbPH{bábàŋyélí} \tbPS{n.} \tbGE{innocence} \tbVA{dakumo}
\tbLX{babẽe}\tbHM{} \tbPH{bàbɪ̃́ɪ̃́} \tbPS{n.} \tbGE{broth from dog meat} \tbTP{babeenɛɛ}
\tbLX{babeɛ}\tbHM{} \tbPH{bábɪ́ɛ́} \tbPS{n.} \tbGE{a friend that cannot be trusted} \tbOP{babeere} \tbTP{babeerɛɛ}
\tbLX{babeɛraa}\tbHM{} \tbPH{bàbɪ̀ɛ́ráà} \tbPS{n.} \tbGE{bridge} \tbOP{babeɛre} \tbTP{babeɛrɛɛ} \tbVA{baaraakyi, takoriko}
\tbLX{babere}\tbHM{} \tbPH{bàbɪ̀rɪ́} \tbPS{n.} \tbGE{the breast of a dog} \tbOP{baberɛ} \tbTP{baberɛɛ}
\tbLX{babini}\tbHM{} \tbPH{bàbíní} \tbPS{n.} \tbGE{dog faeces} \tbOP{babinee} \tbTP{babinnee}
\tbLX{babiŋ}\tbHM{} \tbPH{bábíŋ} \tbPS{n.} \tbGE{a type of tree} \tbOP{babinni} \tbTP{babinnee}
\tbLX{babiri}\tbHM{} \tbPH{bábírì} \tbPS{n.} \tbGE{bosom friend} \tbOP{babie} \tbTP{babiiree} \tbVA{batuo}
\tbLX{babogi}\tbHM{} \tbPH{bàbógí} \tbPS{n.} \tbGE{the deepest part of a body of water} \tbOP{babogri} \tbTP{babogree}
\tbLX{babullaa}\tbHM{} \tbPH{bàbùlláà} \tbPS{n.} \tbGE{spring, pond where water is drawn in small amounts (usually during dry season)} \tbOP{babulli}
\tbLX{badaa}\tbHM{} \tbPH{bàdáá} \tbPS{n.} \tbGE{male dog} \tbOP{badaare} \tbTP{badaarɛɛ}
\tbLX{badaa-de-nyoɔre}\tbHM{} \tbPH{bàdáádíɲɛʊ́ɔ́rɪ́} \tbPS{n.} \tbGE{the last day of a festival} \tbTP{badaadenyoɔrɛɛ} \tbVA{baraade\-nyoɔ\-re}
\tbLX{badalane}\tbHM{} \tbPH{bàdàlánɪ́} \tbPS{n.} \tbGE{uncastrated male dog} \tbOP{badalama} \tbTP{badalamɛɛ}
\tbLX{badalee}\tbHM{} \tbPH{bàdàléé} \tbPS{n.} \tbGE{male puppy} \tbOP{badabilii} \tbTP{ba\-da\-bi\-lee/badabillee}
\tbLX{badavaraa}\tbHM{} \tbPH{bàdàváràà} \tbPS{n.} \tbGE{castrated dog} \tbOP{badavarre} \tbTP{badavarɛɛ/badavarrɛɛ}
\tbLX{badɛre}\tbHM{} \tbPH{bádɛ́rɪ́} \tbPS{n.} \tbSN \tbGE{spider} \tbSN \tbGE{cunning person} \tbOP{badɛrre} \tbTP{badɛrɛɛ}
\tbLX{badie}\tbHM{} \tbPH{bàdìé} \tbPS{n.} \tbGE{the deepest part of a body of water} \tbOP{baderi} \tbTP{baderee}
\tbLX{badoge}\tbHM{} \tbPH{bàdʊ̀gɪ̀} \tbPS{n.} \tbGE{a plant with a sweet smell, used to season sauce} \tbOP{badogre} \tbTP{badogrɛɛ}
\tbLX{badɔgerɔ}\tbHM{} \tbPH{bàdɔ̀grɔ́} \tbPS{n.} \tbGE{female dog that has recently given birth} \tbOP{badɔgere} \tbTP{badɔgerɛɛ} \tbVA{badɔgera}
\tbLX{bafoɔlaa}\tbHM{} \tbPH{bàfʊ́ɔ́láá} \tbPS{n.} \tbSN \tbGE{valley} \tbSN \tbGE{creek} \tbOP{bafoɔle} \tbTP{baroɔlɛɛ}
\tbLX{baga}\tbHM{} \tbPH{bágà} \tbPS{n.} \tbGE{a triangular garment worn by men as underwear} \tbOP{bagere} \tbTP{bagerɛɛ} \tbVA{lere, para}
\tbLX{baga}\tbHM{} \tbPH{bàgà} \tbPS{n.} \tbGE{not caring, unconcerned} \tbXV{Ka onaŋ ba wa a pɔgebeɛloŋ kyɛ n baga kyebe.} \tbXE{I don't care if s/he doesn't attend the wedding.}
\tbLX{baganaa}\tbHM{} \tbPH{bàgànàà} \tbPS{n.} \tbGE{dog leash} \tbOP{baganne} \tbTP{baganɛɛ/bagannɛɛ}
\tbLX{bagane}\tbHM{} \tbPH{bàgánɪ̀} \tbPS{n.} \tbGE{dog skin, dog hide} \tbOP{bagama} \tbTP{bagannɛɛ}
\tbLX{bagaŋ}\tbHM{} \tbPH{bágáŋ} \tbPS{n.} \tbGE{a mad dog} \tbOP{baganne} \tbTP{baganɛɛ/bagannɛɛ} \tbVA{bagao}
\tbLX{bagao}\tbHM{} \tbPH{bàgáʊ̀} \tbPS{n.} \tbGE{mad dog} \tbOP{bagagre} \tbTP{bagagrɛɛ} \tbVA{bagaŋ}
\tbLX{bageba}\tbHM{} \tbPH{bágɪ́bá} \tbPS{n.} \tbGE{a type of sacred cult which initiates boys and girls into adulthood} \tbTP{bagebɛɛ} \tbVA{bɔgebɔ}
\tbLX{bageboge}\tbHM{} \tbPH{bàgɪ̀bʊ̀gɪ̀} \tbPS{n.} \tbGE{the act of divining} \tbXV{Kiritabiiri kyiiruŋ la bageboge.} \tbXE{It is a taboo for Christians to participate  in divining.} \tbVA{bɔgeboge}
\tbLX{bageboge-dagoli}\tbHM{} \tbPH{bàgɪ̀bʊ̀g-dàgólí} \tbPS{n.} \tbGE{divining stick} \tbOP{bageboge-dagolo} \tbTP{bageboge-dagolee} \tbVA{bɔgeboge-dagoli}
\tbLX{bageboge-karaa}\tbHM{} \tbPH{bàgɪ̀bʊ̀g-kàràá} \tbPS{n.} \tbGE{chief diviner} \tbOP{bagebogekarre} \tbTP{bagebogekarɛɛ} \tbVA{bɔgeboge-karaa}
\tbLX{bageboge-woɔ}\tbHM{} \tbPH{bàgɪ̀bʊ̀gɪ̀-wʊ́ɔ́} \tbPS{n.} \tbGE{divining bag} \tbOP{bageboge-woore} \tbTP{bageboge-woorɛɛ} \tbVA{bɔgeboge-woɔ}
\tbLX{bagere}\tbHM{} \tbPH{bàgrɪ́} \tbPS{n.} \tbGE{sacrifice for gods} \tbOP{baga} \tbTP{bagɛɛ} \tbVA{bɔgere}
\tbLX{bagukyɔfeɛle}\tbHM{} \tbPH{bágùʧɔ̀nfɪ́ɛ́lɪ́} \tbPS{n.} \tbGE{very early morning (usually cold)}
\tbLX{baguo}\tbHM{} \tbPH{bágúó} \tbPS{n.} \tbGE{morning} \tbOP{ba\-guu\-ri} \tbTP{baguuree} \tbVA{beguo}
\tbLX{bagupi}\tbHM{} \tbPH{bàgùpí} \tbPS{n.} \tbGE{early morning} \tbOP{baguipiri} \tbTP{bagupiree} \tbVA{bebipi}
\tbLX{bagupipi}\tbHM{} \tbPH{bágùpípí} \tbPS{n.} \tbGE{very early morning} \tbOP{bagupipiri} \tbTP{bagupipiree} \tbVA{bagupi}
\tbLX{bagbɛre}\tbHM{} \tbPH{bàgbɛ́rɪ̀} \tbPS{n.} \tbGE{leg of a dog} \tbOP{bagbɛɛ} \tbTP{bagbɛɛrɛɛ}
\tbLX{bagbɛre}\tbHM{} \tbPH{bàgbɛ́rɪ̀} \tbPS{n.} \tbGE{valley} \tbOP{ba\-gbɛɛ} \tbTP{bagbɛɛrɛɛ}
\tbLX{bagboŋ}\tbHM{} \tbPH{bágbòŋ} \tbPS{n.} \tbGE{fool} \tbOP{ba\-gbon\-ni} \tbTP{bagbonnee}
\tbLX{bagbori}\tbHM{} \tbPH{bàgbòrí} \tbPS{n.} \tbGE{trough for a dog} \tbOP{bagboe} \tbTP{bagboree}
\tbLX{bagyeɛnɛ}\tbHM{} \tbPH{bàdzɪ̀ɛ̀nɛ́} \tbPS{n.} \tbGE{a type of children's dance} \tbTP{bagyeɛnɛɛ}
\tbLX{bagyoɔ}\tbHM{} \tbPH{bàdzʊ́ɔ́} \tbPS{n.} \tbSN \tbGE{squirrel-like rodent (believed to be very sexually active)} \tbSN \tbGE{sex maniac (male)} \tbVA{bagyoa, bazoɔ}
\tbLX{bagyɔroo}\tbHM{} \tbPH{bàdzɔ̀rʊ́ʊ́} \tbPS{n.} \tbGE{a thick, fleshy, poisonous creeping plant}
\tbLX{ba-irime}\tbHM{} \tbPH{bá-írímé} \tbPS{n.} \tbGE{a type of wild mouse found along river valleys which has tender skin} \tbOP{ba-irri} \tbTP{ba-iree}
\tbLX{bakana}\tbHM{} \tbPH{bàkáná} \tbPS{n.} \tbGE{broth from dog meat, made with flour} \tbTP{bakanɛɛ}
\tbLX{bakore}\tbHM{} \tbPH{bákʊ́rɪ́} \tbPS{n.} \tbGE{trusted friend} \tbOP{bakorɔ/bakoremine} \tbTP{bakorɛɛ/bakorrɛɛ}
\tbLX{bakɔ}\tbHM{} \tbPH{bàkɔ̀} \tbPS{n.} \tbGE{a type of yam (early maturing)} \tbOP{bakɔɛ} \tbTP{bakɔrɛɛ}
\tbLX{bakpoŋ}\tbHM{} \tbPH{bàkpóŋ} \tbPS{n.} \tbSN \tbGE{large stream, river} \tbSN \tbGE{(capitalized) the name of a tributary of the Black Volta River in the Upper West Region} \tbOP{bakponni} \tbTP{bakponnee}
\tbLX{balaa}\tbHM{} \tbPH{báláà} \tbPS{n.} \tbGE{waistband (pair of shorts)} \tbOP{balle} \tbTP{balɛɛ}
\tbLX{balambaaraa}\tbHM{} \tbPH{bálámbááráá} \tbPS{n.} \tbGE{peg used for tethering animals} \tbOP{balambaare} \tbTP{balambaarɛɛ} \tbVA{balambiri}
\tbLX{balambiri}\tbHM{} \tbPH{bálámbírí} \tbPS{n.} \tbGE{peg used for tethering animals} \tbOP{balambie} \tbTP{balambiiree} \tbVA{balambaaraa}
\tbLX{balantamm}\tbHM{} \tbPH{bàlàntàmm} \tbPS{adv.} \tbGE{large and heavy} \tbXV{Wɔɔ waa la kpoŋi balantamm.} \tbXE{The elephant is enormously large.}
\tbLX{balante}\tbHM{} \tbPH{bálántɪ́} \tbPS{n.} \tbGE{belt} \tbSG{ba\-lan\-tee\-rɛɛ} \tbOP{ba\-lan\-te\-re} \tbTP{ba\-lan\-te\-re} \tbVA{bɛlɛnte}
\tbLX{bale}\tbHM{} \tbPH{bàlɪ̀} \tbPS{v.} \tbSN \tbGE{to be tired} \tbXV{O toŋ la a toma te zuo lɛnso ka o bale.} \tbXE{He has overworked himself; that's why he is tired.} \tbSN \tbGE{to be fed up} \tbXV{Bebiri zaa ka o maŋ wa sɔre libie, maa eŋ bale la o yɛlɛ.} \tbXE{He comes to beg for money every day; I'm fed up with him.} \tbSN \tbGE{to be weak} \tbXV{Ka neɛŋ kore o maŋ bale la.} \tbXE{One becomes weak when one is old.} \tbSN \tbGE{to be soft} \tbXV{A gane koe la, yɛ mere kõɔ ka o bale.} \tbXE{The animal skin is dry and hard, so sprinkle water on it to make it soft.} \tbPD{balɛɛ, balla, balla, baleba, balaa, baleba }
\tbLX{bale}\tbHM{} \tbPH{bálɪ́} \tbPS{n.} \tbGE{tribe, ethnic group} \tbXV{Dagaaba bale kaŋa yuori la Kowɛrɛɛ.} \tbXE{Kowɛrɛɛ is one of the tribes among the Dagaaba.} \tbOP{balmine} \tbTP{balɛɛ}
\tbLX{bale}\tbHM{} \tbPH{bálɪ́} \tbPS{v.} \tbSN \tbGE{avoid, evade} \tbXV{O daare to ma lɛnso ka o maŋ wa nyɛ ma bale.} \tbXE{She insulted me the last time, and that's why she avoids me when she sees me.} \tbSN \tbGE{divert} \tbXV{Ba naŋ da beɛrɛ a baa ba bale la a sori.} \tbXE{When they were constucting the bridge, they diverted the road.} \tbPD{balɛɛ, bala, bala, baleba, balaa, baleba }
\tbLX{bale}\tbHM{} \tbPH{bálɪ́} \tbPS{adj.} \tbGE{of a kind} \tbXV{A toma ŋa bale ba kye kyɛ.} \tbXE{We do not have his kind of work here.}
\tbLX{balee}\tbHM{} \tbPH{bàléé} \tbPS{n.} \tbGE{brook, stream} \tbOP{babilii} \tbTP{babillee}
\tbLX{balee}\tbHM{} \tbPH{bàléé} \tbPS{n.} \tbGE{puppy} \tbOP{babilii} \tbTP{babillee}
\tbLX{balee}\tbHM{} \tbPH{bàlɪ̀ɪ̀} \tbPS{adj.} \tbGE{of a dubious and dangerous nature} \tbXV{Pɔge koŋ sage toŋ tombalee ŋa tɔ.} \tbXE{A woman will not do this sort of shady thing.}
\tbLX{baleŋkpɔgre}\tbHM{} \tbPH{bàlɪ́ŋkpɔ̀grì} \tbPS{n.} \tbGE{tick} \tbOP{baleŋkpɔgɔ} \tbTP{baleŋkpɔgrɛɛ}
\tbLX{balɔre}\tbHM{} \tbPH{bálɔ́rɪ́} \tbPS{n.} \tbSN \tbGE{ugly person or thing} \tbSN \tbGE{one without strength} \tbOP{balɔɛ} \tbTP{balɔrɛɛ}
\tbLX{balɔrroŋ}\tbHM{} \tbPH{bálɔ̀rrʊ́ŋ} \tbPS{n.} \tbSN \tbGE{not being strong} \tbSN \tbGE{act of being ugly}
\tbLX{bam}\tbHM{} \tbPH{bǎm} \tbPS{n.} \tbGE{a type of large yam} \tbOP{bama} \tbTP{bamɛɛ}
\tbLX{bama}\tbHM{} \tbPH{bámà} \tbPS{pron.} \tbGE{these (3rd person plural demonstrative human)} \tbXV{A noba bama la maŋ toŋ a toma wiẽu.} \tbXE{These are the people who normally do the work quickly.} \tbXV{Bibiiri bama la ka ba iri.} \tbXE{These are the children that have been chosen.}
\tbLX{bambaala}\tbHM{} \tbPH{bàmbààlá} \tbPS{n.} \tbGE{unassuming person; poor person} \tbOP{bambaaleba} \tbTP{bambaalebɛɛ}
\tbLX{bambaalaa}\tbHM{} \tbPH{bámbááláá} \tbPS{n.} \tbGE{small fish, fingerling} \tbOP{bambaale} \tbTP{bambaalɛɛ}
\tbLX{bambalaaraa}\tbHM{} \tbPH{bámbálááráá} \tbPS{n.} \tbGE{swallow (bird)} \tbOP{bambalaare} \tbTP{bambalaarɛɛ}
\tbLX{bambaŋ}\tbHM{} \tbPH{bámbáŋ} \tbPS{adv.} \tbGE{hot, steamy} \tbXV{A zee, re ba taa diibu, a toloŋ bambaŋ.} \tbXE{The soup is too hot; it can't be eaten.} \tbXV{Tentoloŋ saŋa a ŋmenaa maŋ toloŋ bambaŋ.} \tbXE{During the hot season, the sun is always extremely hot.}
\tbLX{bambiŋi}\tbHM{} \tbPH{bàmbìŋ} \tbPS{n.} \tbGE{an arrangement made in advance by two parties}
\tbLX{bambugo}\tbHM{} \tbPH{bámbúgó} \tbPS{n.} \tbGE{confused person; fool} \tbOP{bambugri} \tbTP{ bambugree}
\tbLX{bambuŋi}\tbHM{} \tbPH{bàmbúŋ} \tbPS{n.} \tbGE{salp (fish of the class Thaliacea)} \tbOP{babunni} \tbTP{bambunnee}
\tbLX{bambuŋi}\tbHM{} \tbPH{bàmbúŋ} \tbPS{n.} \tbGE{locally made gunpowder} \tbOP{babunni} \tbTP{bambunnee}
\tbLX{bamm}\tbHM{} \tbPH{bàmm} \tbPS{adv.} \tbGE{big and heavy} \tbXV{A nannyige zu mɔrɛɛ bamm lɛ.} \tbXE{The thief's head is badly swollen.}
\tbLX{bammo}\tbHM{} \tbPH{bàmmʊ́} \tbPS{n.} \tbGE{intelligence, wit} \tbXV{Kɔkɔlaa taa la bammo gaŋ baa.} \tbXE{The cat is more intelligent than the dog.}
\tbLX{bamoɔ}\tbHM{} \tbPH{bàmʊ́ɔ́} \tbPS{n.} \tbGE{a type of tall, strong grass found around bodies of water, whose dry stalks are used for weaving} \tbSG{bamoɔnoo} \tbOP{bamoore} \tbTP{bamoorɛɛ}
\tbLX{bandaa}\tbHM{} \tbPH{bándáá} \tbPS{n.} \tbGE{lizard (male)} \tbOP{bandaare} \tbTP{bandarɛɛ/bandaarɛɛ} \tbVA{baŋa, bɔndaa}
\tbLX{bandaakpogilo}\tbHM{} \tbPH{bàndáákpóglò} \tbPS{n.} \tbGE{the act of hitting the opponent's armpits with one's knees to weaken the opponent (a wrestling move)} \tbXV{Bayuo to la Dɛre ka Dɛre nyɔge o lɔɔ tɔ bandaakpogilo.} \tbXE{Bayuo insulted Dɛre, and Dɛre threw Bayuo down and used his knees to hit his armpits.} \tbTP{bandaakpogilee}
\tbLX{bandakarekpele}\tbHM{} \tbPH{bàndàkárkpɪ́lɪ́} \tbPS{n.} \tbGE{Agama agama lizard; also known as red-headed rock agama} \tbOP{bandakarekpelɛ} \tbTP{bandakarekpelɛɛ} \tbVA{karekpele}
\tbLX{bandakarema}\tbHM{} \tbPH{bàndàkàrɪ̀mà} \tbPS{n.} \tbGE{large male lizard} \tbTP{bandakaremɛɛ}
\tbLX{bandanyene}\tbHM{} \tbPH{bàndàɲɪ́nɪ́} \tbPS{n.} \tbGE{tooth of a lizard} \tbOP{bandanyemɛ} \tbTP{bandanyennɛɛ/bandanyemɛɛ}
\tbLX{bandapoleɛne}\tbHM{} \tbPH{bàndàpʊ́lɪ́ɛ́nɪ́} \tbPS{n.} \tbGE{female lizard} \tbOP{bandapoleɛmɛ} \tbTP{bandapoleɛmɛɛ/bandapoleɛnɛɛ} \tbVA{poleɛne}
\tbLX{bandazeɛraa}\tbHM{} \tbPH{bàndàzɪ̀ɛ́ráá} \tbPS{n.} \tbGE{kite; small bird of prey (family Accipitridae)} \tbOP{bandazeɛre} \tbTP{bandazeɛrɛɛ}
\tbLX{bandazoore}\tbHM{} \tbPH{bàndàzʊ́ʊ́rɪ́} \tbPS{n.} \tbGE{tail of a lizard} \tbOP{bandazoɛ} \tbTP{bandazoorɛɛ}
\tbLX{banfane}\tbHM{} \tbPH{bànfánɪ́} \tbPS{n.} \tbGE{a bangle that is believed to have medicinal value and may be worn to cure a disease} \tbOP{banfama} \tbTP{banfanɛɛ}
\tbLX{bangye}\tbHM{} \tbPH{bándzɪ́} \tbPS{n.} \tbGE{bank} \tbOP{bangyere} \tbTP{bangyerɛɛ}
\tbLX{bangyeraa}\tbHM{} \tbPH{bándzɪ́ráá} \tbPS{n.} \tbSN \tbGE{latrine; washroom} \tbXV{Pampana ka fooŋ mɛ yiri a seŋ ka bagyeraa bebe.} \tbXE{Modern houses are supposed to have built-in washrooms.} \tbSN \tbGE{feces} \tbXV{A bie nyɛ la bini meele a sɛŋ.} \tbXE{The child has soiled the mat with feces.} \tbOP{bangyerre} \tbTP{bangyerɛɛ/bangyerrɛɛ}
\tbLX{ba-nimie-anaare}\tbHM{} \tbPH{bà-nímíè-ànàárɪ̀} \tbPS{n.} \tbGE{a dog, usually with big spots above the eyes that look like a second pair of eyes, believed to have supernatural powers} \tbOP{ba-nimie-anaare} \tbTP{ba-nimie-anaarɛɛ}
\tbLX{bantakperee}\tbHM{} \tbPH{bàntákpérèé} \tbPS{n.} \tbGE{a type of spotted lizard (believed to be poisonous)} \tbOP{bantakperri} \tbTP{bantakperree}
\tbLX{bantoɔ}\tbHM{} \tbPH{bántʊ́ɔ́} \tbPS{n.} \tbGE{rectal bulb syringe} \tbOP{bantoore} \tbTP{bantoorɛɛ} \tbVA{bɔntoɔ}
\tbLX{banweebaa}\tbHM{} \tbPH{bànwɪ́ɪ́báá} \tbPS{n.} \tbGE{carnivorous bird (bigger than a kite but strong and fast)} \tbOP{banweere} \tbTP{banweerɛɛ}
\tbLX{banzeɛ}\tbHM{} \tbPH{bànzɪ̀ɛ̀} \tbPS{n.} \tbGE{copper bangle} \tbOP{banzeere} \tbTP{banzeerɛɛ}
\tbLX{banzeɛraa}\tbHM{} \tbPH{bǎnzɪ́ɛ́ráá} \tbPS{n.} \tbGE{bird that eats lizards} \tbOP{banzeɛre} \tbTP{banzeɛrɛɛ}
\tbLX{banzɔlɔ}\tbHM{} \tbPH{bànzɔ̀lɔ̀} \tbPS{n.} \tbGE{chain} \tbOP{banzɔlle} \tbTP{banzɔllɛɛ} \tbVA{bɔnzɔlɔ}
\tbLX{banyaa}\tbHM{} \tbPH{bàɲáà} \tbPS{n.} \tbGE{chest of a dog} \tbOP{banyaɛ/banyaane} \tbTP{banyaanɛɛ}
\tbLX{banyagere}\tbHM{} \tbPH{bàɲágrɪ́} \tbPS{n.} \tbGE{intestine of a dog} \tbOP{banyaga} \tbTP{banyagɛɛ/banyagerɛɛ}
\tbLX{banyaŋaa}\tbHM{} \tbPH{bàɲáŋáá} \tbPS{n.} \tbGE{female dog} \tbOP{banyaŋene/banyanne} \tbTP{banyaŋenɛɛ/banyannɛɛ}
\tbLX{Banyene}\tbHM{} \tbPH{bàɲɪ̀nɪ̀} \tbPS{n.} \tbGE{homeland of the Banyenɛɛ clan}
\tbLX{Banyenɛɛ}\tbHM{} \tbPH{bàɲɪ́nɛ́ɛ́} \tbPS{n.} \tbGE{name of a clan} \tbSG{banyenoo}
\tbLX{banyɛtɔgle}\tbHM{} \tbPH{báɲɛtɔ́glɪ́} \tbPS{n.} \tbGE{imitation} \tbXV{Bibiiri taa la banyɛtɔgle, azuiŋ neŋkponni seŋ ka ba taa ba menne velaa.} \tbXE{Children like to immitate, so adults should behave well.}
\tbLX{baŋa}\tbHM{} \tbPH{báŋá} \tbPS{n.} \tbGE{bangle} \tbOP{banne} \tbTP{bannɛɛ} \tbVA{bɔŋɔ}
\tbLX{baŋa}\tbHM{} \tbPH{bàŋá} \tbPS{n.} \tbGE{lizard} \tbOP{banne} \tbTP{bannɛɛ} \tbVA{bandaa}
\tbLX{baŋe}\tbHM{} \tbPH{bàŋ} \tbPS{v.} \tbSN \tbGE{to know (something or someone)} \tbXV{N baŋe la a pɔge ŋa serɛ, o yuori la Dabuo.} \tbXE{I know this woman's husband; he is called Dabuo.} \tbSN \tbGE{to be aware of} \tbXV{N baŋe ne la o waabo.} \tbXE{I know he's coming.} \tbPD{baŋɛɛ, banna, banna, banema, bannaa, banema } \tbVA{bɔŋe}
\tbLX{baŋena}\tbHM{} \tbPH{bàŋɪ́ná} \tbPS{n.} \tbSN \tbGE{area between the shoulder and neck} \tbSN \tbGE{trickster} \tbOP{baŋema/banema} \tbTP{baŋemɛɛ/banemɛɛ} \tbVA{bɔŋenɔ}
\tbLX{baŋenaa}\tbHM{} \tbPH{báŋɪ́náá} \tbPS{n.} \tbGE{a type of tree that has medicinal properties} \tbOP{baŋene} \tbTP{baŋenɛɛ} \tbVA{bɔŋenaa}
\tbLX{baŋene}\tbHM{} \tbPH{bàŋɪ̀nɪ́} \tbPS{n.} \tbGE{cleverness} \tbXV{Nannyige maŋ taa la baŋene, ka o nyɔgebo maŋ e tuo.} \tbXE{A thief usually has such cleverness that it is difficult to catch him.} \tbVA{bɔŋene}
\tbLX{baŋgori}\tbHM{} \tbPH{báŋgórí} \tbPS{n.} \tbGE{high-heeled shoes} \tbSG{baŋgoruu} \tbOP{baŋgorri} \tbTP{baŋgoree/baŋgorree}
\tbLX{baŋkaroŋ}\tbHM{} \tbPH{bàŋkàrʊ́ŋ} \tbPS{n.} \tbGE{a type of brown bird with long, thin legs} \tbOP{baŋkarre} \tbTP{baŋkarɛɛ/baŋkarrɛɛ}
\tbLX{baŋkoroŋ}\tbHM{} \tbPH{bàŋkʊ́rʊ́ŋ} \tbPS{n.} \tbGE{foreknowledge}
\tbLX{baŋŋmeɛle}\tbHM{} \tbPH{bàŋŋmɪ́ɛ́lɪ́} \tbPS{n.} \tbGE{twisted bangle, often with a spiritual or medicinal significance} \tbOP{ baŋŋmeɛlɛ} \tbTP{baŋŋmeɛlɛɛ}
\tbLX{ba-ɔge}\tbHM{} \tbPH{bà-ɔ̀gɪ́} \tbPS{n.} \tbGE{a dog's nest, usually for a female that has recently given birth} \tbOP{ba-ɔgre} \tbTP{ba-ɔgɛɛ/ba-ɔgrɛɛ}
\tbLX{bapare}\tbHM{} \tbPH{bápárɪ́} \tbPS{n.} \tbGE{close friend} \tbOP{baparemine} \tbTP{baparɛɛ} \tbVA{batuo}
\tbLX{bapare}\tbHM{} \tbPH{bàpárɪ́} \tbPS{n.} \tbGE{end of a valley} \tbOP{baparɛɛ}
\tbLX{bapiiraa}\tbHM{} \tbPH{bàpííráá} \tbPS{n.} \tbGE{a puppy earmarked for someone} \tbOP{bapiiri} \tbTP{bapiiree}
\tbLX{bapimbaare}\tbHM{} \tbPH{bápímbáárɪ̀} \tbPS{n.} \tbGE{a small part of the stomach of a ruminant that has many compartments} \tbOP{bapimbaare} \tbTP{bapimbaarɛɛ}
\tbLX{bara}\tbHM{} \tbPH{bárá} \tbPS{n.} \tbGE{boastful threat, often considered inappropriate or rude} \tbXV{Dakoraa naŋ baŋ ka dɔɔ zaa kye a yiri poɔ la so ka o wa fare a bara ama taaba.} \tbXE{Knowing that all the men had left the house, Dakoraa went to make boastful threats.} \tbTP{barɛɛ}
\tbLX{baraa}\tbHM{} \tbPH{báráá} \tbPS{n.} \tbGE{not making any effort; helpless}
\tbLX{bare}\tbHM{} \tbPH{bàrɪ̀} \tbPS{v.} \tbSN \tbGE{to leave, to release, to let go} \tbXV{Ba bare la sakubiiri.} \tbXE{They have let the schoolchildren go for vacation.} \tbSN \tbGE{to discontinue, to abstain, to cease, to stop} \tbXV{A baala bare la a dɔɔta gaabo.} \tbXE{The patient has stayed away from the hospital.} \tbSN \tbGE{to divorce (initiated by husband)} \tbXV{A dɔɔ gaa bare la o pɔge.} \tbXE{The man has divorced his wife.} \tbSN \tbGE{not having power} \tbXV{A dɔɔ ŋa eŋ barɛɛ la te zuo, o koŋ tõɔ a dɛre do.} \tbXE{This man is so useless that he cannot climb the ladder.} \tbPD{barɛɛ, bara, bara, bareba, baraa, bareba } \tbSE{bare gyɛle} \tbGE{to be careless; to not pay attention} \tbSE{bare yɛŋe} \tbGE{to feel free, to relax}
\tbLX{bare}\tbHM{} \tbPH{bàrɪ̀} \tbPS{v.} \tbSN \tbGE{to cultivate seeds} \tbXV{A pɔge bare la kommie a dabɔge poɔ.} \tbXE{The woman has cultivated tomato seeds in the garden.} \tbSN \tbGE{to offer firstfruits to the gods} \tbXV{Bɔgebogereba maŋ bare la bompaala kyɛ baŋ di.} \tbXE{Diviners normally offer firstfruits to their gods before eating.} \tbPD{barɛɛ, bara, bara, bareba, baraa, bareba }

\tbLX{barekanneɛbaa}\tbHM{} \tbPH{bàrɪ̀kánnɪ́ɛ́báá} \tbPS{n.} \tbGE{local spirits (alcohol)} \tbOP{barekanneɛbaare} \tbTP{barekanneɛbaarɛɛ}

\tbLX{barre}\tbHM{} \tbPH{bárrɪ́} \tbPS{v.} \tbGE{to haggle} \tbPD{barrɛɛ, barra, barra, bareba, barraa}%, , -}
\tbLX{ba-sakɔre}\tbHM{} \tbPH{bà-sakɔrɪ́} \tbPS{n.} \tbGE{lazy and useless dog}
\tbLX{basasale}\tbHM{} \tbPH{bàsàsàlɪ́} \tbPS{n.} \tbGE{slippery waterway} \tbTP{basasalɛɛ}
\tbLX{ba-soɔmee}\tbHM{} \tbPH{bà-sʊ́ɔ́mɪ́ɪ́} \tbPS{n.} \tbGE{weak and sickly dog} \tbVA{ba-sakɔre}
\tbLX{bataabo}\tbHM{} \tbPH{bátáábʊ́} \tbPS{n.} \tbGE{the state of not having anything}
\tbLX{batara}\tbHM{} \tbPH{bátárá} \tbPS{n.} \tbGE{poor or needy person} \tbOP{batareba} \tbTP{batarebɛɛ}
\tbLX{batuo}\tbHM{} \tbPH{bátúó} \tbPS{n.} \tbGE{bosom friend} \tbOP{batuuri} \tbTP{batuuree} \tbVA{bakore bapare, bameŋɛ, babiri}
\tbLX{bavilime}\tbHM{} \tbPH{bàvílímé} \tbPS{n.} \tbGE{lake} \tbOP{bavilimeri} \tbTP{bavilimee}
\tbLX{bawaa}\tbHM{} \tbPH{báwáá} \tbPS{n.} \tbGE{a type of dance} \tbOP{bawaare} \tbTP{bawaarɛɛ}
\tbLX{Bayaayiree}\tbHM{} \tbPH{báyààyíréé} \tbPS{n.} \tbGE{people of Bayaayiri} \tbSG{bayaayiruu}
\tbLX{Bayaayiri}\tbHM{} \tbPH{báyààyírí} \tbPS{n.} \tbSN \tbGE{a settlement in the Upper West Region} \tbSN \tbGE{ancestors' dwelling; next world}
\tbLX{Bayuo}\tbHM{} \tbPH{báyúó} \tbPS{n.} \tbGE{name traditionally given to a male child born on a funeral day} \tbTP{bayumine}
\tbLX{bazẽe}\tbHM{} \tbPH{bàzɪ̃́ɪ̃̀} \tbPS{n.} \tbGE{blood of a dog} \tbTP{bazeenɛɛ}
\tbLX{bazele}\tbHM{} \tbPH{bàzɪ̀lɪ́} \tbPS{n.} \tbGE{tongue of a dog} \tbOP{bazelɛ} \tbTP{bazelɛɛ}
\tbLX{bazoɔ}\tbHM{} \tbPH{bàzʊ́ɔ̀} \tbPS{n.} \tbGE{a type of wild fox} \tbTP{bazoorɛɛ} \tbVA{bazoa, bagyoɔ}
\tbLX{bazoraa}\tbHM{} \tbPH{bàzóráà} \tbPS{n.} \tbGE{fast dog} \tbOP{bazorri} \tbTP{bazorree}
\tbLX{bazugiri}\tbHM{} \tbPH{bázúgrí} \tbPS{n.} \tbGE{a type of wild mouse that has thick fur} \tbOP{bazugo} \tbTP{bazugee/bazugiree}
\tbLX{be}\tbHM{} \tbPH{bě} \tbPS{v.} \tbGE{to exist, to be there} \tbXV{A baala be la asibiti poɔ.} \tbXE{The patient is in the hospital.} \tbPD{bee, bere, bere, beribe, beraa, beribe } \tbSE{be nimiri} \tbGE{to be awake, to be alert}
\tbLX{be}\tbHM{} \tbPH{bě} \tbPS{loc.} \tbGE{there (place)} \tbXV{N ba nyɛ fo be.} \tbXE{I did not see you there.}
\tbLX{be}\tbHM{} \tbPH{bɪ̀} \tbPS{v.} \tbSN \tbGE{to become boiled or cooked} \tbXV{A nɛne ba be, o varema.} \tbXE{The meat is not well boiled, it is hard.} \tbSN \tbGE{to be smooth (flour)} \tbXV{A mansinni ɛre la a bɛnzɔŋ ka a be.} \tbXE{The machine has ground the bean flour very smooth.} \tbSN \tbGE{to mature, to ripen (fruit, millet)} \tbXV{Ka taŋaaŋ wa be o maŋ pɔre le la.} \tbXE{A ripe shea fruit normally falls off the tree.} \tbSN \tbGE{to ferment (alcohol)} \tbXV{A dãa naŋ ba be azuiŋ a ba bugiro.} \tbXE{The pito is not yet fermented, so it is not alcholic.} \tbSN \tbGE{to fortify (spiritually)} \tbXV{Pɔlebilii yaga maŋ boɔrɔ ka ba tegere ba menne a be.} \tbXE{Today, most young men want to fortify themselves through spritual powers.} \tbPD{beɛ,  berɛ, beerɛ, berebɛ, beraa, berebɛ }
\tbLX{be}\tbHM{} \tbPH{bɪ́} \tbPS{conj.} \tbGE{or} \tbXV{Fo na gaa la be?} \tbXE{Will you go or not?} \tbVA{bee}
\tbLX{bebe}\tbHM{} \tbPH{bèbé} \tbPS{adj.} \tbGE{present, available} \tbXV{Noba yaga da be be la kyɛ ka Naa wa.} \tbXE{There were many people present when the Chief came.}
\tbLX{bebelaa}\tbHM{} \tbPH{bɪ́bɪ́láá} \tbPS{n.} \tbGE{termite hill} \tbOP{bebelle} \tbTP{bebellɛɛ} \tbVA{belaa}
\tbLX{bebelebiri}\tbHM{} \tbPH{bɪ́bɪ̀lɪ̀bírì} \tbPS{n.} \tbGE{white ant} \tbSG{bebelebiruu} \tbOP{bebelebie} \tbTP{bebelebiiree}
\tbLX{beberaa}\tbHM{} \tbPH{bébéráà} \tbPS{n.} \tbGE{a pessimist} \tbOP{beberri} \tbVA{beberee}
\tbLX{beberee}\tbHM{} \tbPH{bébéréè} \tbPS{n.} \tbGE{one who is malicious} \tbOP{beberri} \tbTP{beberree} \tbVA{bebberaa}
\tbLX{beberiyɛlɛ}\tbHM{} \tbPH{bébèrìyɛ́lɛ́} \tbPS{n.} \tbGE{malicious talk} \tbSG{beberiyeli}
\tbLX{beberuŋ}\tbHM{} \tbPH{bébérúŋ} \tbPS{n.} \tbGE{the act of causing trouble}
\tbLX{bebibaaraa}\tbHM{} \tbPH{bèbíbááráá} \tbPS{n.} \tbGE{last day} \tbOP{bebibaare} \tbTP{bebibaarɛɛ}
\tbLX{bebibalaa}\tbHM{} \tbPH{bèbíbàlàà} \tbPS{n.} \tbGE{particular day}
\tbLX{bebidɛŋe}\tbHM{} \tbPH{bèbídɛ́ŋ} \tbPS{n.} \tbGE{first day} \tbOP{bebidɛnne} \tbTP{bebidɛnnɛɛ} \tbVA{bebidɛndɛŋ}
\tbLX{bebifaa}\tbHM{} \tbPH{bèbífáá} \tbPS{n.} \tbGE{sabbath; sacred day} \tbOP{bebifaare} \tbTP{bebifaarɛɛ}
\tbLX{bebigbuli}\tbHM{} \tbPH{bèbígbùlì} \tbPS{n.} \tbGE{whole day} \tbOP{bebigbulo} \tbTP{bebigbulee}
\tbLX{bebikore}\tbHM{} \tbPH{bèbíkʊ́rɪ́} \tbPS{n.} \tbGE{a former day, a day in the past} \tbOP{bebikorɔ/bebikora} \tbTP{bebikorɛɛ}
\tbLX{bebipaalaa}\tbHM{} \tbPH{bèbípááláá} \tbPS{n.} \tbGE{new day; different day} \tbOP{bebipaala} \tbTP{bebipaalɛɛ}
\tbLX{bebiri}\tbHM{} \tbPH{bèbírì} \tbPS{n.} \tbGE{date, day} \tbOP{bebie} \tbTP{bebiiri/bebiiree}
\tbLX{bebisaana}\tbHM{} \tbPH{bèbísááná} \tbPS{n.} \tbGE{strange day} \tbOP{bebisaama}
\tbLX{bebisoŋ}\tbHM{} \tbPH{bèbísʊ́ŋ} \tbPS{n.} \tbGE{good day, holy day} \tbOP{bebisonne} \tbTP{bebisonnɛɛ}
\tbLX{bebitɔɔre}\tbHM{} \tbPH{bèbítɔ̀ɔ̀rɪ́} \tbPS{n.} \tbGE{distant date} \tbOP{bebitɔbɔ} \tbTP{bebitɔɔrɛɛ}
\tbLX{bebivuo}\tbHM{} \tbPH{bèbívúó} \tbPS{n.} \tbGE{free day} \tbOP{bebivori} \tbTP{bebivoree}
\tbLX{bebiyeni}\tbHM{} \tbPH{bèbíyéní} \tbPS{n.} \tbGE{one day, a single day} \tbOP{bebiyaga} \tbVA{bebiyenaa}
\tbLX{bee}\tbHM{} \tbPH{bèé} \tbPS{adj.} \tbGE{to exist; to be present} \tbXV{O bee be.} \tbXE{S/he is there.} %\tbPD{-, -, -, -, -, - }
    \tbVA{be; be la}
\tbLX{bee}\tbHM{} \tbPH{bɪ́ɪ́} \tbPS{conj.} \tbGE{or} \tbXV{O gaaɛ la bee o ba gaa?} \tbXE{Has s/he gone or not?} \tbVA{be, beɛ, baa}
\tbLX{bẽe}\tbHM{} \tbPH{bɪ̃́ɪ̃́} \tbPS{n.} \tbSN \tbGE{concentrated extraction from something (tasty, interesting)} \tbSN \tbGE{broth from meat} \tbTP{beenɛɛ}
\tbLX{beele}\tbHM{} \tbPH{bɪ́ɪ́lɪ̀} \tbPS{v.} \tbGE{to envy (with evil intent)} \tbPD{beelɛɛ, beelɛ, beelɛ, beelebɛ, beelaa, beelebɛ }
\tbLX{beele}\tbHM{} \tbPH{bɪ́ɪ́lɪ̀} \tbPS{v.} \tbGE{to move slightly away from a spot} \tbPD{beelɛɛ, beelɛ, beelɛ, beelebɛ, beelaa, beelebɛ }
\tbLX{beene}\tbHM{} \tbPH{bɪ́ɪ́nɪ́} \tbPS{n.} \tbSN \tbGE{strong person} \tbSN \tbGE{one with supernatural power} \tbOP{beemɛ} \tbTP{beemɛɛ}
\tbLX{beene}\tbHM{} \tbPH{bɪ́ɪ́nɪ́} \tbPS{adj.} \tbSN \tbGE{well-cooked, boiled} \tbSN \tbGE{ripe (fruit)} \tbOP{beemɛ} \tbTP{beenɛɛ}
\tbLX{beene}\tbHM{} \tbPH{bɪ̀ɪ̀nɪ̀} \tbPS{v.} \tbSN \tbGE{to move majestically} \tbSN \tbGE{to roam round leisurely} \tbPD{beenɛɛ, beenɛ, beenɛ, beenemɛ, beenaa}%, , -}
\tbLX{beere}\tbHM{} \tbPH{bɪ̀ɪ̀rɪ̀} \tbPS{v.} \tbSN \tbGE{to kill, to finish off, to beat mercilessly} \tbSN \tbGE{to boil pito in the final stage of production} \tbPD{beerɛɛ, beerɛ, beerɛ, beerebɛ, beeraa}%, , -}
\tbLX{beɛ}\tbHM{} \tbPH{bɪ̀ɛ̀} \tbPS{v.} \tbSN \tbGE{to move out of the way (allowing another to pass)} \tbSN \tbGE{to divert} \tbSN \tbGE{to pass someone by, to overtake} \tbPD{beɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa}%, , -}
\tbLX{beɛ}\tbHM{} \tbPH{bɪ̀ɛ̀} \tbPS{v.} \tbGE{to slice (meat)} \tbXV{A nɛŋŋmaara beɛ la a nɛne eŋ a doge poɔ.} \tbXE{The butcher sliced the meat and put it into the pot.} \tbPD{beɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, - } \tbVA{beɛle}
\tbLX{beɛ}\tbHM{} \tbPH{bíɛ́} \tbPS{adj.} \tbGE{funny; strange} \tbOP{beere} \tbTP{beerɛɛ} \tbVA{beɛbeɛ}
\tbLX{bẽɛ}\tbHM{} \tbPH{bɪ̃́ɛ̃̀} \tbPS{n.} \tbGE{beer} \tbSG{beɛnoo} \tbTP{beɛnɛɛ}
\tbLX{beɛbeɛ}\tbHM{} \tbPH{bɪ́ɛ́bɪ́ɛ́} \tbPS{adj.} \tbGE{strange} \tbVA{beɛ}
\tbLX{beɛlaa}\tbHM{} \tbPH{bɪ́ɛ́láá} \tbPS{adj.} \tbGE{small, little (quantity, time)} \tbOP{beɛlɛ}
\tbLX{beɛle}\tbHM{} \tbPH{bɪ̀ɛ̀lɪ̀} \tbPS{v.} \tbGE{to accompany} \tbPD{beɛlɛɛ, beɛlɛ, beɛlɛ, beɛlebɛ, beɛlaa, beɛlebɛ } \tbVA{beɛre}
\tbLX{beɛle}\tbHM{} \tbPH{bɪ̀ɛ̀lɪ̀} \tbPS{v.} \tbGE{to cut in strips  (meat, hide)} \tbPD{beɛlɛɛ, beɛlɛ, beɛlɛ, beɛlebɛ, beɛlaa, beɛlebɛ } \tbVA{beɛre}
\tbLX{beɛlɛbeɛlɛ}\tbHM{} \tbPH{bɪ́ɛ́lɛ́bɪ́ɛ́lɛ́} \tbPS{adv.} \tbGE{slowly; bit by bit}
\tbLX{bẽɛo}\tbHM{} \tbPH{bɪ̃̀ɛ̃̀ʊ̃́} \tbPS{n.} \tbGE{multitude, crowd} \tbOP{beɛnɛɛ}
\tbLX{beɛre}\tbHM{} \tbPH{bɪ́ɛ́rɪ̀} \tbPS{n.} \tbGE{older brother} \tbOP{beɛremine} \tbTP{beɛrɛɛ}
\tbLX{beɛre}\tbHM{} \tbPH{bɪ́ɛ́rɪ́} \tbPS{v.} \tbGE{to follow up} \tbPD{beɛrɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, beɛrebɛ }
\tbLX{beɛre}\tbHM{} \tbPH{bɪ̀ɛ̀rɪ̀} \tbPS{v.} \tbGE{to cut into strips} \tbPD{beɛrɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, beɛrebeɛraa } \tbVA{beɛle}
\tbLX{bege}\tbHM{} \tbPH{bɪ̀gɪ̀} \tbPS{v.} \tbGE{to stop up, to close a passage} \tbPD{begɛɛ, begerɛ, begeraa, begerebɛ, begeraa, begerebɛ }
\tbLX{bege}\tbHM{} \tbPH{bɪ̀gɪ́} \tbPS{n.} \tbGE{law} \tbOP{begɛ} \tbTP{begɛɛ} \tbVA{merɛ (Akan)}
\tbLX{begebinne}\tbHM{} \tbPH{bɪ̀gɪ̀bínné} \tbPS{n.} \tbSN \tbGE{lawmaker} \tbSN \tbGE{member of parliament} \tbOP{begebinime} \tbTP{begebinimee}
\tbLX{begebiŋi-duoraa}\tbHM{} \tbPH{bɪ̀gɪ̀bíŋ-duoraa} \tbPS{n.} \tbGE{parliament house} \tbOP{begebiŋi-duori} \tbTP{begebiŋi-duoree}
\tbLX{begere}\tbHM{} \tbPH{bɪ̀gɪ̀rɪ́} \tbPS{n.} \tbGE{wing} \tbOP{begɛ} \tbTP{begerɛɛ}
\tbLX{belaa}\tbHM{} \tbPH{bɪ́láá} \tbPS{n.} \tbGE{anthill} \tbOP{belle} \tbTP{bellɛɛ} \tbVA{bebelaa}
\tbLX{belaa}\tbHM{} \tbPH{bìláà} \tbPS{pron.} \tbGE{what (interrogative pronoun)} \tbXV{Belaa ka fo erɛ?} \tbXE{What are you doing?}
\tbLX{belaa}\tbHM{} \tbPH{bɪ́láá} \tbPS{adv.} \tbSN \tbGE{small (quantity), few} \tbXV{A laa poɔ bondirii waa la belaa lɛ.} \tbXE{The amount of food is very small.} \tbXV{Zenɛ a tontonema waa la belaa.} \tbXE{Today the workers are few.} \tbSN \tbGE{short while (time)} \tbXV{Vɛŋ ka belaa ka fo gaa.} \tbXE{Wait for a short while before you go.}
\tbLX{bele}\tbHM{} \tbPH{bɪ́lɪ́} \tbPS{v.} \tbGE{to do extraneous work} \tbPD{belɛɛ, belɛ, belɛ, belbɛ, belaa}%, , -}
\tbLX{bele}\tbHM{} \tbPH{bɪ́lɪ́} \tbPS{n.} \tbGE{yeast} \tbOP{belɛ} \tbTP{belɛɛ}
\tbLX{belefele}\tbHM{} \tbPH{bɪ̀lɪ̀fɪ̀lɪ̀} \tbPS{adv.} \tbGE{big and fat (person)}
\tbLX{benee}\tbHM{} \tbPH{bènéé} \tbPS{adv.} \tbGE{that place}
\tbLX{benne}\tbHM{} \tbPH{bɪ̀nnɪ̀} \tbPS{v.} \tbGE{to intrude} \tbPD{bennɛɛ, bennɛ, bennɛ, benemɛ, bennaa, benemɛ }
\tbLX{bennɛ}\tbHM{} \tbPH{bɪ́nnɛ́} \tbPS{n.} \tbGE{intruder (especially to meals)} \tbOP{benemɛ} \tbTP{benemɛɛ}
\tbLX{benni}\tbHM{} \tbPH{bénnì} \tbPS{v.} \tbGE{to doubt} \tbPD{bennee, benne, benne, benime, bennaa}%, -, -}
    \tbVA{bonne}
\tbLX{bentere}\tbHM{} \tbPH{bɪ̀ntɪ́rɪ̀} \tbPS{n.} \tbGE{a large gourd drum} \tbOP{benteɛ} \tbTP{benterɛɛ} \tbVA{bontere}
\tbLX{beo}\tbHM{} \tbPH{bɪ̀ʊ̀} \tbPS{adv.} \tbGE{big and stagnant}
\tbLX{bere}\tbHM{} \tbPH{bɪ́rɪ́} \tbPS{n.} \tbSN \tbGE{a type of small plant whose leaves and flowers are used as vegetables} \tbSN \tbGE{the leaves of this plant} \tbTP{berrɛɛ} \tbVA{berenɛmɛ}
\tbLX{bere}\tbHM{} \tbPH{bɪ̀rɪ́} \tbPS{n.} \tbSN \tbGE{breast} \tbOP{berɛ} \tbTP{berɛɛ}
\tbLX{bere}\tbHM{} \tbPH{bɪ̀rɪ̀} \tbPS{v.} \tbGE{protect by supernatural means} \tbPD{berɛɛ, berɛ, berɛ, berebɛ, beraa, beremɛ }
\tbLX{bere}\tbHM{} \tbPH{bɪ̀rɪ̀} \tbPS{adj.} \tbGE{powerful (supernaturally)}
\tbLX{berebaaraa}\tbHM{} \tbPH{bɪ̀rɪ̀bááráá} \tbPS{n.} \tbGE{last born (of humans and other mammals)} \tbOP{berebaare}
\tbLX{berekaroo}\tbHM{} \tbPH{bɪ̀rìkàrʊ́ʊ́} \tbPS{n.} \tbSN \tbGE{a type of disease of the breast} \tbSN \tbGE{a type of charm believed to inflict this disease} \tbOP{berekarre} \tbTP{berekarɛɛ}
\tbLX{berekuuri}\tbHM{} \tbPH{bìrɪ̀kúúrí} \tbPS{n.} \tbGE{thick centre of the breast} \tbSG{berekuuree} \tbOP{berekubo}
\tbLX{berekpulo}\tbHM{} \tbPH{bɪ̀rɪ̀kpúló} \tbPS{n.} \tbGE{hard, firm breasts (of a young woman)} \tbTP{berkpulee}
\tbLX{beremɛ}\tbHM{} \tbPH{bɪ́rɪ́mɛ́} \tbPS{n.} \tbGE{one who possesses supernatural powers} \tbTP{beremɛɛ}
\tbLX{berenɛmɛ}\tbHM{} \tbPH{bɪ̀rɪ̀nɛ́mɛ́} \tbPS{n.} \tbGE{the flower of the bere plant} \tbSG{berenɛnoo} \tbTP{berenɛmɛɛ} \tbVA{bere}
\tbLX{berenoɔ}\tbHM{} \tbPH{bɪ̀rɪ̀nʊ́ɔ̀} \tbPS{n.} \tbGE{a bird used in rituals to induce breast milk in a nursing mother} \tbOP{berenoore} \tbTP{berenoorɛɛ}
\tbLX{berenoɔre}\tbHM{} \tbPH{bɪ̀rɪ̀nʊ́ɔ́rɪ̀} \tbPS{n.} \tbGE{nipple (of a breast)} \tbOP{berenɔɛ} \tbTP{berenɔrɛɛ}
\tbLX{berese}\tbHM{} \tbPH{bɪ̀rɪ̀sɪ̀} \tbPS{adv.} \tbGE{extremely large and fat}
\tbLX{berewoɔ}\tbHM{} \tbPH{bɪ̀rɪ̀wʊ́ɔ́} \tbPS{n.} \tbGE{slang term for female breast} \tbOP{berewoore} \tbTP{berewoorɛɛ}
\tbLX{bereyɛre}\tbHM{} \tbPH{bɪ̀rɪ̀yɛ́rɪ́} \tbPS{n.} \tbGE{a type of wild plant that is used as a vegetable} \tbTP{bereyɛrɛɛ}
\tbLX{berɛ}\tbHM{} \tbPH{bìrɛ́} \tbPS{v.} \tbSN \tbGE{to wish} \tbXV{O berɛ ŋa zie da koŋ nyaa.} \tbXE{S/he wished day would never come.} \tbXV{O berɛ ona la da na gaa a tigiri.} \tbXE{S/he wished s/he would be the one to attend the feast.} \tbSN \tbGE{take by surprise} \tbXV{O berɛ ka saana la wa kpɛ. } \tbXE{Before he realized it, a stranger arrived.} %\tbPD{-, -, -, -, -, -}
\tbLX{berondɔndɔre}\tbHM{} \tbPH{bɪ̀rʊ́ndɔ̀ndɔ̀rɪ́} \tbPS{n.} \tbGE{curdled milk} \tbTP{berondɔndɔrɛɛ}
\tbLX{beroŋ}\tbHM{} \tbPH{bɪ̀rʊ́ŋ} \tbPS{n.} \tbGE{milk}
\tbLX{beroŋkãa}\tbHM{} \tbPH{bɪ́rʊ́ŋkã́ã̀} \tbPS{n.} \tbGE{fat from fresh milk; butter} \tbSG{beroŋkaamo} \tbTP{beroŋkaanɛɛ}
\tbLX{berroŋ}\tbHM{} \tbPH{bɪ̀rrʊ́ŋ} \tbPS{n.} \tbGE{the state of being supernaturally powerful}
\tbLX{bɛ}\tbHM{} \tbPH{bɛ̀} \tbPS{v.} \tbGE{to be sick, to be ill} \tbXV{Ka saaŋ poɔ a bie o na bɛ la.} \tbXE{If the child is out in the downpour he will get sick.} \tbPD{bɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, beɛrebɛ }
\tbLX{bɛ}\tbHM{} \tbPH{bɛ̀} \tbPS{v.} \tbGE{to seal, to close a hole with a muddy substance} \tbXV{Ka fooŋ ba bɛ a peɛ voɔre ka saaŋ wa miire a die na kyire la.} \tbXE{If you do not patch the hole in the roof, the room will leak when there is rain.} \tbPD{bɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, beɛrebɛ }
\tbLX{bɛ}\tbHM{} \tbPH{bɛ́} \tbPS{pron.} \tbGE{where (interrogative pronoun)} \tbXV{A libie bɛ?} \tbXE{Where is the money?} \tbVA{bɛɛ}
\tbLX{bɛbɛgere}\tbHM{} \tbPH{bɛ̀bɛ̌grɪ̀} \tbPS{n.} \tbGE{mud} \tbOP{bɛbɛgɛ} \tbTP{bɛbɛgerɛɛ}
\tbLX{bɛɛ}\tbHM{} \tbPH{bɛ́ɛ̀} \tbPS{pron.} \tbGE{where (interrogative pronoun), emphatic form} \tbXV{A libie bɛɛ?} \tbXE{Where the hell is the money?} \tbVA{bɛ}
\tbLX{bɛɛɛ}\tbHM{} \tbPH{bɛ̀ɛ́ɛ̀} \tbPS{n.} \tbGE{sheep} \tbOP{bɛɛre} \tbTP{bɛɛrɛɛ}
\tbLX{bɛɛne}\tbHM{} \tbPH{bɛ́ɛ́nɪ́} \tbPS{n.} \tbGE{hemp seeds}
\tbLX{bɛge}\tbHM{} \tbPH{bɛ̀gɪ̀} \tbPS{v.} \tbSN \tbGE{to embank} \tbSN \tbGE{to block with a dam} \tbPD{bɛgɛɛ, bɛgerɛ, bɛgerɛ, bɛgerebɛ, bɛgeraa, bɛgerebɛ }
\tbLX{bɛgelaa}\tbHM{} \tbPH{bɛ̀gɪ̀làá} \tbPS{n.} \tbGE{a shell of a particular type of seed, used to scrape inside of cooking utensils} \tbOP{bɛgele} \tbTP{bɛgelɛɛ}
\tbLX{bɛlɛmbɛgere}\tbHM{} \tbPH{bɛ̀lɛ̀mbɛ́grɪ́} \tbPS{n.} \tbGE{nightengale} \tbOP{bɛlɛmbɛgɛ} \tbTP{bɛlɛmbɛgerɛɛ} \tbVA{bɛlembɛgre}
\tbLX{bɛlɛnte}\tbHM{} \tbPH{bɛ́lɛ́ntɪ̀} \tbPS{n.} \tbGE{belt} \tbOP{bɛlɛntere} \tbTP{bɛlɛnteerɛɛ}
\tbLX{bɛlle}\tbHM{} \tbPH{bɛ̀llɪ̀} \tbPS{v.} \tbGE{to deceive} \tbPD{bɛllɛɛ, bɛllɛ, bɛllɛ, bɛlbɛ, bɛllaa}%, , -}
\tbLX{bɛlle}\tbHM{} \tbPH{bɛ̀llɪ̀} \tbPS{v.} \tbGE{to coax (a baby)} \tbPD{bɛllɛɛ, bɛllɛ, bɛllɛ, bɛlbɛ, bɛllaa}%, , -}
\tbLX{bɛlle}\tbHM{} \tbPH{bɛ́llɪ̀} \tbPS{v.} \tbGE{to look at closely, to inspect} \tbPD{bɛllɛɛ, bɛllɛ, bɛllɛ, bɛlbɛ, bɛllaa}%, , -}
\tbLX{bɛllebɛllɛ}\tbHM{} \tbPH{bɛ̀lbɛ́llɛ́} \tbPS{n.} \tbGE{bean flour shaped into slabs and steamed} \tbTP{bɛllebɛllɛɛ} \tbVA{tumpanne}
\tbLX{bɛmbɛŋ}\tbHM{} \tbPH{bɛ́mbɛ́ŋ} \tbPS{adv.} \tbGE{extremely black} \tbXV{A peroo e la sɔgelaa bɛmbɛŋ.} \tbXE{The sheep is extremely black.}
\tbLX{bɛmbɛŋ}\tbHM{} \tbPH{bɛ̀mbɛ̌ŋ} \tbPS{v.} \tbGE{to move in a shaky manner} \tbPD{bɛmbɛŋɛɛ, bɛmbɛnɛ, bɛmbɛnɛ, bɛmbɛnemɛ, bɛmbɛnaa}%, , -}
\tbLX{bɛmbiri}\tbHM{} \tbPH{bɛ́mbírì} \tbPS{n.} \tbGE{bean seed} \tbOP{bɛmbie} \tbTP{bɛmbiiree}
\tbLX{bɛmbugo}\tbHM{} \tbPH{bɛ́mbùgó} \tbPS{n.} \tbGE{barn for storing beans} \tbOP{bɛmbugri} \tbTP{bɛmbugree}
\tbLX{bɛmbuli}\tbHM{} \tbPH{bɛ́mbúlí} \tbPS{n.} \tbGE{bean sprout} \tbOP{bɛmbulo} \tbTP{bɛmbullee}
\tbLX{bɛmpaala}\tbHM{} \tbPH{bɛ́mpáálà} \tbPS{n.} \tbGE{new, fresh beans} \tbTP{bɛmpaalɛɛ}
\tbLX{bɛmpelaa}\tbHM{} \tbPH{bɛ́mpɪ́láá} \tbPS{n.} \tbGE{white beans} \tbOP{bɛmpeɛle} \tbTP{bɛmpeɛlɛɛ}
\tbLX{bɛmpɛɛre}\tbHM{} \tbPH{bɛ́mpɛ̀ɛ̀rɪ́} \tbPS{n.} \tbGE{unripened bean seed} \tbOP{bɛmpɛbɛ} \tbTP{bɛmpɛbɛɛ}
\tbLX{bɛmpɛge}\tbHM{} \tbPH{bɛ́mpɛ́gɪ́} \tbPS{n.} \tbGE{bean husk} \tbOP{bɛmpɛgre/bɛmpɛgɛ} \tbTP{bɛmpɛgrɛɛ}
\tbLX{bɛndaa}\tbHM{} \tbPH{bɛ́ndàà} \tbPS{n.} \tbGE{bean stalk} \tbOP{bɛndaare} \tbTP{bɛndarɛɛ/bɛndaarɛɛ}
\tbLX{bɛndaa}\tbHM{} \tbPH{bɛ́ndáá} \tbPS{n.} \tbGE{a cooked bean dish (usually not seasoned)} \tbOP{bɛndaare} \tbTP{bɛndarɛɛ/bɛndaarɛɛ}
\tbLX{bɛnwɔne}\tbHM{} \tbPH{bɛ́nwɔ́nɪ́} \tbPS{n.} \tbGE{the fruit of a bean plant} \tbOP{bɛnwɔmɔ} \tbTP{bɛnwɔnnɛɛ}
\tbLX{bɛnzɔŋe}\tbHM{} \tbPH{bɛ́nzɔ́ŋ} \tbPS{n.} \tbGE{bean flour} \tbOP{bɛnzɔnne} \tbTP{bɛnzɔnnɛɛ}
\tbLX{bɛnzɔɔre}\tbHM{} \tbPH{bɛ́nzɔ̀ɔ̀rɪ́} \tbPS{n.} \tbGE{tender (seedless) fruit of a bean plant} \tbOP{bɛnzɔbɔ} \tbTP{bɛnzɔrɛɛ}
\tbLX{bɛŋe}\tbHM{} \tbPH{bɛ̂ŋ} \tbPS{v.} \tbSN \tbGE{to sift, to separate grains from unwanted material} \tbSN \tbGE{to skim (usually by tapping water with the back of the hand or a container to push away debris)} \tbSN \tbGE{to move unsteadily, to sway} \tbPD{bɛŋɛɛ, bɛnnɛ, bɛnnɛ, bɛnemɛ, bɛnnaa, bɛnemɛ } \tbVA{pɛŋe}
\tbLX{bɛŋe}\tbHM{} \tbPH{bɛ̀ŋ} \tbPS{v.} \tbGE{to pretend to invite someone to share something (but not whole-heartedly)} \tbPD{bɛŋɛɛ, bɛnnɛ, bɛnnɛ, bɛnnaa, bɛnemɛ}%, , -}
\tbLX{bɛŋɛ}\tbHM{} \tbPH{bɛ́ŋɛ̀} \tbPS{n.} \tbGE{beans} \tbSG{bɛmmo} \tbTP{bɛnnɛɛ}
\tbLX{bɛŋgbɛre}\tbHM{} \tbPH{bɛ̀ŋgbɛ̀rɪ́} \tbPS{n.} \tbGE{a variety of beans} \tbOP{bɛŋgbɛrre} \tbTP{bɛŋgbɛrɛɛ}
\tbLX{bɛŋkorɔ}\tbHM{} \tbPH{bɛ́ŋkʊ́rɔ́} \tbPS{n.} \tbGE{old beans} \tbTP{bɛŋkorɛɛ} \tbVA{bɛŋkora}
\tbLX{bɛŋkuoŋaa}\tbHM{} \tbPH{bɛ́ŋkúóŋáá} \tbPS{n.} \tbGE{dry bean} \tbOP{bɛŋkuoni} \tbTP{bɛŋkuonee}
\tbLX{bɛraa}\tbHM{} \tbPH{bɛ́ráá} \tbPS{n.} \tbSN \tbGE{trap, snare} \tbSN \tbGE{bait (fishhook)} \tbSN \tbGE{poison} \tbOP{bɛrre} \tbTP{bɛrɛɛ}
\tbLX{bɛraa}\tbHM{} \tbPH{bɛ̀ràá} \tbPS{n.} \tbGE{razor} \tbOP{bɛrre} \tbTP{bɛrrɛɛ}
\tbLX{bɛre}\tbHM{} \tbPH{bɛ̀rɪ̀} \tbPS{v.} \tbSN \tbGE{to set (a trap)} \tbSN \tbGE{to poison} \tbXV{Dabẽɛ deme maŋ boɔrɔ ka ba bɛre ba dendɔmɔ ko.} \tbXE{Cowards always try to kill their enemies by poisoning them.} \tbPD{bɛrɛɛ, bɛrɛ, bɛrɛ, bɛrebɛ, bɛraa, bɛrebɛ }
\tbLX{bɛre}\tbHM{} \tbPH{bɛ̀rɪ̀} \tbPS{v.} \tbGE{to ponder, to wonder} \tbXV{O da di la a bondirii tasoga leɛ zeŋe bɛre.} \tbXE{After eating for some time, she sat back and thought.} \tbPD{bɛrɛɛ, bɛrɛ, bɛrɛ, bɛrebɛ, bɛraa, bɛrebɛ }
\tbLX{bɛre}\tbHM{} \tbPH{bɛ́rɪ̀} \tbPS{v.} \tbSN \tbGE{to exclude, to separate from the herd} \tbXV{A naa\-kyee\-nɛ bɛre la naadɔgerɔ na bare.} \tbXE{The cattle herder has separated the cow that had recently calved from the herd.} \tbSN \tbGE{to detach or cut off (a large piece)} \tbXV{A naŋkpaana bɛre la a naabo gbɛre koɔre.} \tbXE{The butcher has cut off and sold the thigh of the cow.} \tbPD{bɛrɛɛ, bɛrɛ, bɛrɛ, bɛrebɛ, bɛraa, bɛrebɛ }
\tbLX{bɛren-uri}\tbHM{} \tbPH{bɛ̀rɪ̀n-ùrí} \tbPS{n.} \tbGE{pollen from the kenaf plant (a fibre-producing plant similar to hibiscus cannabinus) which causes itching} \tbTP{bɛren-uree} \tbVA{bɛren-uri}
\tbLX{bɛroŋ}\tbHM{} \tbPH{bɛ́rʊ́ŋ} \tbPS{adj.} \tbGE{fat, thick} \tbOP{bɛrɛ} \tbTP{bɛrɛɛ/bɛrrɛɛ}
\tbLX{bibile}\tbHM{} \tbPH{bìbílé} \tbPS{n.} \tbGE{small child} \tbOP{bibilii}
\tbLX{bibile-deɛnaa}\tbHM{} \tbPH{bìbílédɪ́ɛ́náá} \tbPS{n.} \tbGE{toy} \tbOP{bibiledeɛne} \tbTP{bibiledeɛnɛɛ}
\tbLX{bibilpɔɔ}\tbHM{} \tbPH{bìbìlpɔ̀ɔ́} \tbPS{n.} \tbGE{young child} \tbOP{bibilpɔɔre} \tbTP{bibilpɔɔrɛɛ}
\tbLX{bibore}\tbHM{} \tbPH{bíbʊ̀rɪ̀} \tbPS{n.} \tbGE{the act of playing chess}
\tbLX{bidɛndɛŋe}\tbHM{} \tbPH{bìdɛ̀ndɛ́ŋ} \tbPS{n.} \tbGE{firstborn (human)} \tbOP{bidɛndɛnne} \tbTP{bidɛndɛnnɛɛ} \tbVA{bidɛŋ}
\tbLX{bidɛŋe}\tbHM{} \tbPH{bìdɛ́ŋ} \tbPS{n.} \tbGE{firstborn (human)} \tbOP{bidɛnne} \tbTP{bidɛnnɛɛ} \tbVA{bidɛndɛŋ}
\tbLX{bidɔŋe}\tbHM{} \tbPH{bìdɔ̂ŋ} \tbPS{n.} \tbGE{sibling (clan level)} \tbOP{bidɔnne} \tbTP{bidɔnnɛɛ}
\tbLX{bidɔɔ}\tbHM{} \tbPH{bìdɔ́ɔ́} \tbPS{n.} \tbGE{son} \tbOP{bidɔbɔ} \tbTP{bidɔɔrɛɛ}
\tbLX{bie}\tbHM{} \tbPH{bíé} \tbPS{n.} \tbGE{child} \tbOP{biiri} \tbTP{biiree}
\tbLX{bie}\tbHM{} \tbPH{bíé} \tbPS{n.} \tbGE{day, date} \tbXV{Fo sori na yiibu bie la dabuo?} \tbXE{What date have you set for your journey?} \tbOP{biiri} \tbTP{biiree}
\tbLX{bie}\tbHM{} \tbPH{bíè} \tbPS{n.} \tbGE{a type of game played using marbles} \tbOP{biree/biiree}
\tbLX{bie}\tbHM{} \tbPH{bíé} \tbPS{adv.} \tbGE{tomorrow} \tbVA{bieo}
\tbLX{bie}\tbHM{} \tbPH{bìè} \tbPS{v.} \tbGE{to come to pass} \tbXV{A pɔlbile na de la pɔge, a bebiri la naŋ ba bie.} \tbXE{The young man will marry, it just has not happened yet.} \tbPD{-, biere, biere, bieribe, bieraa, bieribe }
\tbLX{biebie}\tbHM{} \tbPH{bíébíé} \tbPS{adv.} \tbGE{daily; every day}
\tbLX{bieo}\tbHM{} \tbPH{bíèó} \tbPS{adv.} \tbGE{tomorrow} \tbXV{Bieo, ba na wa la.} \tbXE{Tomorrow, they will come.} \tbVA{bie}
\tbLX{bieri}\tbHM{} \tbPH{bíérí} \tbPS{n.} \tbGE{stinginess, unwillingness to share or open up}
\tbLX{bifaa}\tbHM{} \tbPH{bìfáá} \tbPS{n.} \tbGE{bad child} \tbOP{bifaare} \tbTP{bifaarɛɛ}
\tbLX{bifaa}\tbHM{} \tbPH{bífáá} \tbPS{n.} \tbGE{bad pebble} \tbOP{bifaare} \tbTP{bifaarɛɛ}
\tbLX{bifẽe}\tbHM{} \tbPH{bífɪ̃́ɪ̃́} \tbPS{n.} \tbGE{small quantity of pebbles} \tbOP{bifeene} \tbTP{bifeenɛɛ}
\tbLX{bifõɔo}\tbHM{} \tbPH{bífʊ̃́ɔ̃́ʊ̃́} \tbPS{n.} \tbGE{light pebble} \tbOP{bifoɔne} \tbTP{bifoɔnɛɛ}
\tbLX{bigaalaa}\tbHM{} \tbPH{bìgáálàá} \tbPS{n.} \tbGE{any bedding for a child} \tbOP{bigaale}
\tbLX{bigaale}\tbHM{} \tbPH{bìgààlɪ̀} \tbPS{n.} \tbGE{postnatal care} \tbOP{bigaale}
\tbLX{bigane}\tbHM{} \tbPH{bìgánɪ̀} \tbPS{n.} \tbGE{leather spread for a baby cot} \tbOP{bigama} \tbTP{bigamɛɛ}
\tbLX{bigeŋ}\tbHM{} \tbPH{bìgèŋ} \tbPS{adv.} \tbGE{all over (area)}
\tbLX{bigeŋ}\tbHM{} \tbPH{bìgèŋ} \tbPS{adv.} \tbGE{big heavy sound (fall of a big wall, gunshot)}
\tbLX{bigileŋ}\tbHM{} \tbPH{bìgìlèŋ} \tbPS{adj.} \tbGE{large and heavy} \tbXV{A kuuri waa la bigileŋ lɛ.} \tbXE{The stone is large and heavy.}
\tbLX{bigiri}\tbHM{} \tbPH{bígírí} \tbPS{n.} \tbSN \tbGE{fibre} \tbSN \tbGE{kenaf; a plant that produces fibre} \tbOP{bige} \tbTP{bigee}
\tbLX{bigiri}\tbHM{} \tbPH{bígírí} \tbPS{adj.} \tbGE{fibre-like} \tbXV{A maŋgo deɛ waa la bigiri.} \tbXE{The mango is fibrous.}
\tbLX{bigiri}\tbHM{} \tbPH{bìgìrì} \tbPS{v.} \tbGE{to describe} \tbPD{bigiree, bigire, bigire, bigiribe, bigiraa, bigiribe }
\tbLX{biiree}\tbHM{} \tbPH{bììréé} \tbPS{n.} \tbGE{sand} \tbOP{biiri} \tbTP{biiree}
\tbLX{biiridalempoɔ}\tbHM{} \tbPH{bììrìdàlɪ̀mpʊ̀ɔ̀} \tbPS{n.} \tbGE{desert} \tbOP{biiridalempoore} \tbTP{bii\-ridalemporɛɛ} \tbVA{biirikpoŋ}
\tbLX{bii\-rikpoŋi}\tbHM{} \tbPH{bíírìkpôŋ} \tbPS{n.} \tbGE{desert} \tbOP{bii\-rikponni} \tbTP{bii\-rikponnee} \tbVA{biiridalempoɔ}
\tbLX{bikaa}\tbHM{} \tbPH{bìkáà} \tbPS{n.} \tbGE{the act of caring for children}
\tbLX{bikaale}\tbHM{} \tbPH{bíkààlɪ̀} \tbPS{n.} \tbGE{the act of gathering pebbles}
\tbLX{bikaara}\tbHM{} \tbPH{bìkáárá} \tbPS{n.} \tbGE{caretaker of a child} \tbOP{bikaareba} \tbTP{bikarebɛɛ}
\tbLX{biko}\tbHM{} \tbPH{bìkʊ́} \tbPS{n.} \tbGE{the act of killing children}
\tbLX{bikoŋkonaa}\tbHM{} \tbPH{bìkóŋkónáá} \tbPS{n.} \tbGE{a child that cries a lot} \tbOP{bikoŋkonni} \tbTP{bikoŋkonnee}
\tbLX{bikoɔraa}\tbHM{} \tbPH{bìkʊ́ɔ́ráá} \tbPS{n.} \tbGE{industrious person} \tbOP{bikoɔre}
\tbLX{bikore}\tbHM{} \tbPH{bìkʊ́rɪ́} \tbPS{n.} \tbGE{old chap, child} \tbOP{bikorɔ/bikoreba} \tbTP{bikorɛɛ}
\tbLX{bikũu}\tbHM{} \tbPH{bìkṹũ̀} \tbPS{n.} \tbGE{baby cot} \tbOP{bikuuni} \tbTP{bikuunee}
\tbLX{bikũu}\tbHM{} \tbPH{bìkṹṹ} \tbPS{n.} \tbGE{dead child} \tbOP{bikuu\-ni} \tbTP{bikuunee}
\tbLX{bikpeebɛ}\tbHM{} \tbPH{bìkpɪ́ɪ́bɛ́} \tbPS{n.} \tbGE{orphan} \tbOP{bikpeere/bikpeemine} \tbTP{bikpeerɛɛ}
\tbLX{bikyelaa}\tbHM{} \tbPH{bìʧɪ́láà} \tbPS{n.} \tbGE{smart child} \tbOP{bikelle} \tbTP{bikyelɛɛ}
\tbLX{bikyɛnaa}\tbHM{} \tbPH{bìʧɛ́náá} \tbPS{n.} \tbGE{toddler} \tbOP{bikyɛnne/bikyɛnemɛ} \tbTP{bikyɛnemɛɛ}
\tbLX{bildɛŋe}\tbHM{} \tbPH{bìldɛ́ŋ} \tbPS{n.} \tbGE{firstborn (nonhuman)} \tbOP{bildɛnne} \tbTP{bildɛnnɛɛ}
\tbLX{bile}\tbHM{} \tbPH{bílé} \tbPS{adj.} \tbSN \tbGE{small} \tbXV{A kparoo waa la bile; o koŋ kpɛ a bie.} \tbXE{The shirt is small; it won't fit the child.} \tbSN \tbGE{younger} \tbOP{bilii}
\tbLX{bile}\tbHM{} \tbPH{bílé} \tbPS{n.} \tbGE{offspring (nonhuman)} \tbOP{bilii} \tbTP{bilee}
\tbLX{bileɛ}\tbHM{} \tbPH{bìlɪ́ɛ́} \tbPS{n.} \tbGE{young baby} \tbOP{bileɛre} \tbTP{bileɛrɛɛ} \tbVA{bilɛɛ}
\tbLX{bilembileŋ}\tbHM{} \tbPH{bílémbíléŋ} \tbPS{adj.} \tbGE{round and smooth}
\tbLX{bilembileŋ}\tbHM{} \tbPH{bìlèmbíléŋ} \tbPS{n.} \tbGE{caterpillar track, continous track} \tbOP{bilembilenni} \tbTP{bilembilennee}
\tbLX{bilembileŋ}\tbHM{} \tbPH{bìlèmbíléŋ} \tbPS{adj.} \tbGE{large and heavy, with weight distributed unevenly}
\tbLX{bilentimm}\tbHM{} \tbPH{bìlèntìmm} \tbPS{adj.} \tbGE{bloated and large}
\tbLX{bilɛɛ}\tbHM{} \tbPH{bìlìɛ̀} \tbPS{n.} \tbGE{baby (newly born)} \tbOP{bilɛɛre} \tbTP{bilɛɛrɛɛ} \tbVA{bileɛ}
\tbLX{bilibala}\tbHM{} \tbPH{bìlìbálá} \tbPS{adv.} \tbGE{grabbing on to a huge object by throwing both arms around it} \tbXV{O le nyɔge la a bombɛroŋ na lɛ bilibala.} \tbXE{S/he firmly grabbed the fat one.}
\tbLX{billi}\tbHM{} \tbPH{bìllì} \tbPS{v.} \tbGE{to roll (a heavy object along the ground)} \tbPD{billee, bille, bille, bilibe, billaa}%}%, , -}
\tbLX{billi}\tbHM{} \tbPH{bíllí} \tbPS{v.} \tbGE{to roll between the palms; to mould} \tbPD{billee, bille, bille, bilibe, billaa}%}%, , -}
\tbLX{bimaraa}\tbHM{} \tbPH{bìmáráá} \tbPS{n.} \tbGE{baby sling} \tbOP{bimarre} \tbVA{maraa}
\tbLX{bimbiŋ}\tbHM{} \tbPH{bímbíŋ} \tbPS{adv.} \tbGE{extremely dark (rain clouds)} \tbXV{A saa koe la bimbiŋ lɛ.} \tbXE{The rain clouds have gotten awfully thick.}
\tbLX{bimbiŋe}\tbHM{} \tbPH{bìmbìŋé} \tbPS{n.} \tbGE{pen, poultry shed} \tbOP{bimbinni} \tbTP{bimbinnee} \tbVA{nɔzage}
\tbLX{bimeŋɛ}\tbHM{} \tbPH{bìmɪ́ŋɛ́} \tbPS{n.} \tbGE{good child} \tbOP{bimenne} \tbTP{bimennɛɛ}
\tbLX{bimm}\tbHM{} \tbPH{bìmm} \tbPS{adv.} \tbGE{large and solid} \tbXV{A bugo e la kpoŋ bimm lɛ.} \tbXE{The granary is big and solid.}
%\tbLX{bimm}\tbHM{} \tbPH{bimm} \tbPS{adj.} \tbGE{solid} \tbXV{A taabo daa waa la bimm lɛ.} \tbXE{The timber log is solid.}
\tbLX{bimpeɛle}\tbHM{} \tbPH{bìmpɪ́ɛ́lɪ́} \tbPS{n.} \tbSN \tbGE{white faeces} \tbSN \tbGE{fear (used with nyɛ...)} \tbXV{Ba toŋ yeli ka n pɔge ma beɛrɛ la ka o deɛ yɔ nyeɛrɛ bimpeɛle.} \tbXE{A message was sent to inform my wife that her mother was sick and she is in utter distress and fear.} \tbTP{bimpeɛlɛɛ}
\tbLX{bimporɔ}\tbHM{} \tbPH{bìmpʊ́rɔ́} \tbPS{n.} \tbGE{flatulence} \tbOP{bimporre} \tbTP{bimporɛɛ/bimporrɛɛ} \tbVA{bimpora}
\tbLX{bin-eɛraa}\tbHM{} \tbPH{bìn-ɪ́ɛ́ráá} \tbPS{n.} \tbGE{cloth used for cleaning faeces (of a baby)} \tbOP{bin-eɛre} \tbTP{bin-eɛrɛɛ}
\tbLX{bini}\tbHM{} \tbPH{bíní} \tbPS{n.} \tbGE{excrement}  \tbTP{binee}
\tbLX{binkyoo}\tbHM{} \tbPH{bìnʧóó} \tbPS{n.} \tbGE{open place full of defecation} \tbOP{binkyoori} \tbTP{binkyooree}
\tbLX{binnyeɛbo}\tbHM{} \tbPH{bìnɲɛɪ́ɛ́bʊ́} \tbPS{n.} \tbGE{act of misbehavior}
\tbLX{binɔnaa}\tbHM{} \tbPH{bìnɔ̀náá} \tbPS{n.} \tbGE{beloved child; favorite} \tbOP{binɔnne}
\tbLX{binɔŋe}\tbHM{} \tbPH{bìnɔ̀ŋ} \tbPS{n.} \tbGE{the act of loving children; the act of pampering} \tbOP{binɔnema}
\tbLX{bintanyeɛrɛ}\tbHM{} \tbPH{bìntàɲɪ́ɛ́rɛ́} \tbPS{n.} \tbGE{a type of diarrhoea} \tbOP{bintanyeɛre} \tbTP{bintanyeɛrɛɛ}
\tbLX{binzẽe}\tbHM{} \tbPH{bìnzɪ̃́ɪ́} \tbPS{n.} \tbGE{dysentery} \tbTP{binzeenɛɛ}
\tbLX{biŋi}\tbHM{} \tbPH{bìŋ} \tbPS{v.} \tbSN \tbGE{to put down, to place, to set out} \tbSN \tbGE{to store (up)} \tbPD{biŋee, binne, binne, binime, binnaa, bi\-ni\-me }
\tbLX{biŋibiŋi}\tbHM{} \tbPH{bìŋìbíŋí} \tbPS{v.} \tbGE{to shift unsteadily, teeter} \tbXV{A tanzuŋ waa la bile, ka a doge biŋibiŋine.} \tbXE{The pot holder is small and so the pot is moving unsteadily.} \tbPD{biŋibiŋee, biŋibiŋine, biŋibiŋine, biŋibiŋi\-ni\-me, biŋibiŋinaa, biŋibiŋinime }
\tbLX{biŋkõɔ}\tbHM{} \tbPH{bǐŋkʊ̃́ɔ̃́} \tbPS{n.} \tbGE{watery faeces, diarrhoea} \tbTP{biŋkoɔnɛɛ}
\tbLX{biŋkɔnnyeɛrɛ}\tbHM{} \tbPH{bìŋkɔ́nɲɪ́ɛ́rɛ́} \tbPS{n.} \tbGE{one suffering from diarrhoea} \tbOP{biŋkɔnnyeɛrebɛ} \tbTP{biŋkɔnnyeɛrebɛɛ}
\tbLX{bipeɛ}\tbHM{} \tbPH{bìpɪ́ɛ́} \tbPS{n.} \tbGE{baby cot} \tbOP{bipɛre} \tbTP{bipɛrɛɛ} \tbVA{bikũu}
\tbLX{bipoɔnaa}\tbHM{} \tbPH{bìpʊ́ɔ́náá} \tbPS{n.} \tbGE{pampered child} \tbOP{bipoɔne} \tbTP{bipoɔnɛɛ}
\tbLX{bipɔlbile}\tbHM{} \tbPH{bìpɔ́lbílé} \tbPS{n.} \tbGE{a lad; young man}
\tbLX{bipɔlee}\tbHM{} \tbPH{bìpɔ́lɪ́ɪ́} \tbPS{n.} \tbGE{young man, lad} \tbOP{bipɔlɔ} \tbTP{bipɔlɛɛ/bipɔllɛɛ} \tbVA{bipɔlee, bipɔlbile}
\tbLX{bipɔllee}\tbHM{} \tbPH{bìpɔ́lléé} \tbPS{n.} \tbGE{young man} \tbOP{bipɔlɔ} \tbTP{bipɔlɛɛ} \tbVA{bipɔlee, bipɔlbile}
\tbLX{bipuulee}\tbHM{} \tbPH{bìpúúléé} \tbPS{n.} \tbGE{young woman} \tbOP{bipuuli}
\tbLX{biraa}\tbHM{} \tbPH{bíráá} \tbPS{n.} \tbGE{one who stammers} \tbOP{biribe} \tbTP{biribee}
\tbLX{biri}\tbHM{} \tbPH{bírí} \tbPS{v.} \tbGE{cause debris to drop (down from a height)} \tbXV{Gyaŋ\-gbon\-ni biri la tɛne a dasɛrɛɛ poɔ eŋe a kõɔ poɔ.} \tbXE{The mice broke debris from the ceiling into the water.} \tbPD{biree, bire, bire, biribe, biraa, biribe }
\tbLX{biri}\tbHM{} \tbPH{bírì} \tbPS{v.} \tbGE{to stutter, to stammer} \tbPD{biree, bire, bire, biribe, biraa}%}%, , -}
\tbLX{biri}\tbHM{} \tbPH{bìrì} \tbPS{v.} \tbGE{to pass by narrowly, to slip past} \tbXV{O deɛ biri la a taŋa kyɛ le.} \tbXE{S/he just passed the hill a little and fell.} \tbPD{biree, bire, bire, biribe, biraa, biribe }
\tbLX{biri}\tbHM{} \tbPH{bírì} \tbPS{n.} \tbGE{chaff from malt} \tbOP{biree}
\tbLX{biri}\tbHM{} \tbPH{bírì} \tbPS{n.} \tbGE{seed, nut} \tbOP{bie} \tbTP{bii\-ree}
\tbLX{biri}\tbHM{} \tbPH{bírì} \tbPS{n.} \tbGE{something of worth} \tbXV{Te zenɛ toma ŋa eŋ yi la biri.} \tbXE{Our work today has been fruitful.}
\tbLX{biribiri}\tbHM{} \tbPH{bìrìbìrì} \tbPS{adv.} \tbGE{rushing in large numbers e.g. in a stampede}
\tbLX{biribiri}\tbHM{} \tbPH{bíríbírí} \tbPS{adv.} \tbSN \tbGE{completely, totally} \tbXV{O zagerɛɛ biribiri ka o koŋ gaa.} \tbXE{He has completely refused to go.} \tbSN \tbGE{completely black}
\tbLX{biribiri}\tbHM{} \tbPH{bìrìbírí} \tbPS{v.} \tbGE{to slip, to not fit in (joint)} \tbXV{A sori na taa la kusiwara te zuo ka fo maŋ kyɛnɛ ka fo gbɛɛ bibire.} \tbXE{That road has too much gravel, so your feet slip around as you walk on it.} \tbPD{biribiree, biribire, biribire, biribiribe, biribiraa}%}%, , -}
\tbLX{birikyi}\tbHM{} \tbPH{bìríʧì} \tbPS{n.} \tbSN \tbGE{brake} \tbSN \tbGE{brick} \tbOP{birikyiri} \tbTP{birikyiree} \tbVA{birisi}
\tbLX{birikyilaa}\tbHM{} \tbPH{bìríʧìláá} \tbPS{n.} \tbGE{mould (for bricks)} \tbOP{birikyilaa} \tbTP{birikyilaarɛɛ}
\tbLX{birikyisaabo}\tbHM{} \tbPH{bìríʧìsáábʊ́} \tbPS{n.} \tbGE{brake pad} \tbTP{birikyisaarɛɛ}
\tbLX{birisi}\tbHM{} \tbPH{bìrísì} \tbPS{n.} \tbGE{brick} \tbSG{birisiruu} \tbOP{birisiri} \tbTP{birisiree} \tbVA{birikyi}
\tbLX{biriyokpekperaa}\tbHM{} \tbPH{bìríyǒkpɪ́kpìráà} \tbPS{n.} \tbGE{wild duck-like water bird} \tbOP{biriyokpekperre} \tbTP{biri\-yo\-kpe\-kpe\-rɛɛ}
\tbLX{biriyonaameɛo}\tbHM{} \tbPH{bíríyònààmɪ̀ɛ́ʊ} \tbPS{n.} \tbGE{very large duck} \tbOP{biriyonaameɛne} \tbTP{biriyonaameɛnɛɛ}
\tbLX{biriyuo}\tbHM{} \tbPH{bìríyúó} \tbPS{n.} \tbGE{duck} \tbSG{biriyoruu} \tbOP{biriyori/biriyoe} \tbTP{biri\-yo\-ree} \tbVA{biriyori}
\tbLX{bisaa}\tbHM{} \tbPH{bìsàá} \tbPS{n.} \tbGE{father (human)} \tbOP{bisaamine} \tbTP{bisaarɛɛ}
\tbLX{bisoŋ}\tbHM{} \tbPH{bìsʊ́ŋ} \tbPS{n.} \tbGE{good child, well-behaved child} \tbOP{bisonne} \tbTP{bisonnɛɛ}
\tbLX{bitoroo}\tbHM{} \tbPH{bìtòróò} \tbPS{n.} \tbGE{virgin} \tbOP{bitorri} \tbTP{bitoree}
\tbLX{bitɔgelɔ}\tbHM{} \tbPH{bìtɔ́glɔ̀} \tbPS{n.} \tbGE{midwife} \tbOP{bitɔgelebɔ/bitɔgeleba} \tbTP{bitɔgelebɛɛ} \tbVA{bitɔgela}
\tbLX{bituuro}\tbHM{} \tbPH{bìtúúró} \tbPS{n.} \tbGE{baby adopter, nanny} \tbOP{bituuribo} \tbTP{bi\-tuu\-ri\-bee}
\tbLX{biweɛmɛ}\tbHM{} \tbPH{bìwɪ́ɛ́mɛ́} \tbPS{n.} \tbGE{strayed or wayward child} \tbOP{biweɛne} \tbTP{biweɛnɛɛ}
\tbLX{biweɛraa}\tbHM{} \tbPH{bìwɪ́ɛ́ráá} \tbPS{n.} \tbGE{spoilt child} \tbOP{biweɛre}
\tbLX{biwolaa}\tbHM{} \tbPH{bìwʊ́láá} \tbPS{n.} \tbSN \tbGE{problem child} \tbSN \tbGE{beloved child} \tbOP{biwolle} \tbTP{biwolɛɛ}
\tbLX{biyaalaa}\tbHM{} \tbPH{bìyááláá} \tbPS{n.} \tbGE{babysitter} \tbOP{biyaalba} \tbTP{biyaalbɛɛ}
\tbLX{biyaale}\tbHM{} \tbPH{bìyáálɪ́} \tbPS{n.} \tbGE{the act of babysitting}
\tbLX{biyaaleyieluŋ}\tbHM{} \tbPH{bìyáálɪ̀yíélúŋ} \tbPS{n.} \tbGE{lullaby} \tbOP{biyaaleyiele} \tbVA{biyagleyieluŋ}
\tbLX{biyoɔraa}\tbHM{} \tbPH{bìyʊ́ɔ́ráá} \tbPS{n.} \tbGE{well-travelled person; one who roams a lot} \tbOP{biyoɔre} \tbTP{biyoɔrɛɛ}
\tbLX{biyopore}\tbHM{} \tbPH{bìyópʊ̀rɪ̀} \tbPS{n.} \tbGE{naming ceremony (baby)}
\tbLX{biyuori}\tbHM{} \tbPH{bìyúórí} \tbPS{n.} \tbGE{popular person} \tbOP{biyoe} \tbTP{biyoree}
\tbLX{bizɛgɛ}\tbHM{} \tbPH{bìzɛ̀gɛ́} \tbPS{n.} \tbGE{young baby} \tbOP{bizɛgre} \tbTP{bizɛgɛɛ}
\tbLX{bizɔnzɔɔ}\tbHM{} \tbPH{bìzɔ̀nzɔ́ɔ́} \tbPS{n.} \tbGE{young baby} \tbOP{bizɔnzɔɔbɔ} \tbTP{bizɔnzɔɔrɛɛ}
\tbLX{bo}\tbHM{} \tbPH{bʊ́} \tbPS{v.} \tbGE{to gather momentum (rain seen coming from afar in full scale)} \tbPD{boɛ, boorɔ, boorɔ, booreba, booraa, booreba }
\tbLX{bo}\tbHM{} \tbPH{bʊ̀} \tbPS{v.} \tbSN \tbGE{to confer, to deliberate, to plan together} \tbXV{Ba na bo la sɛre kyɛ gaa a kɔɔto.} \tbXE{They will confer before they go to court.} \tbSN \tbGE{to suppose} \tbXV{Fo bo ka boŋ ka a saana na taa wa ne?} \tbXE{What do you suppose the stranger will bring along?} \tbPD{boɛ, boorɔ, boorɔ, booreba, booraa, booreba }
\tbLX{bobɛllɛ}\tbHM{} \tbPH{bʊ̀bɛ́llɛ́} \tbPS{n.} \tbGE{a type of carnivorous bird} \tbOP{bobɛlle}
\tbLX{boboo}\tbHM{} \tbPH{bʊ̀bʊ́ʊ́} \tbPS{adj.} \tbGE{extremely voluminous}
\tbLX{boboruu}\tbHM{} \tbPH{bóbórúú} \tbPS{n.} \tbGE{a type of termite} \tbOP{boborri/boboree} \tbTP{boborree}
\tbLX{bobɔge}\tbHM{} \tbPH{bʊ̀bɔ̀gɪ́} \tbPS{n.} \tbGE{huge carnivorous bird} \tbOP{bobɔgre} \tbTP{bobɔgrɛɛ}
\tbLX{bodaa}\tbHM{} \tbPH{bʊ̀dáá} \tbPS{n.} \tbGE{male goat} \tbOP{bodaare} \tbTP{bodaarɛɛ/bodaɛ}
\tbLX{bodaa-ansiinee}\tbHM{} \tbPH{bʊ̀dáá-ánsíínéé} \tbPS{n.} \tbGE{a type of tree with edible fruit} \tbOP{bodaa-ansiini} \tbVA{boraa-ansiinee}
\tbLX{bodaanonnoonaa}\tbHM{} \tbPH{bʊ̀dáánʊ́nnʊ́ʊ́náá} \tbPS{n.} \tbGE{a type of tree with edible fruit} \tbOP{bodaanonnoone} \tbTP{bodaanonnoonɛɛ} \tbVA{bo\-non\-noo\-naa}
\tbLX{bodaasensɔɔre}\tbHM{} \tbPH{bʊ̀dáásɪ̌nsɔ́ɔ́rɪ́} \tbPS{n.} \tbGE{wild tigernut (not edible)} \tbOP{bodaasensɔbɔ} \tbTP{bodaasensɔɔrɛɛ}
\tbLX{bodalane}\tbHM{} \tbPH{bʊ́dàlánɪ́} \tbPS{n.} \tbGE{male goat (uncastrated)} \tbOP{bodalama} \tbTP{bodalamɛɛ}
\tbLX{bodalee}\tbHM{} \tbPH{bʊ̀dàléé} \tbPS{n.} \tbGE{ young male goat} \tbOP{bodabilii} \tbTP{bodabilee}
\tbLX{bodateɛne}\tbHM{} \tbPH{bʊ̀dàtɪ́ɛ́nɪ́} \tbPS{n.} \tbGE{male goat with long hair on the chin; bearded male goat}
\tbLX{bodavaraa}\tbHM{} \tbPH{bʊ̀dàváràà} \tbPS{n.} \tbGE{castrated male goat} \tbOP{bodavarre} \tbTP{bodavarɛɛ/bodavarrɛɛ}
\tbLX{bogere}\tbHM{} \tbPH{bʊ́grɪ̀} \tbPS{v.} \tbGE{to break into little bits or pulp} \tbPD{bogerɛɛ, bogerɔ, bogerɔ, bogerebɔ, bogeraa, bogerebɔ }
\tbLX{bogi}\tbHM{} \tbPH{bògí} \tbPS{n.} \tbSN \tbGE{hole} \tbSN \tbGE{grave, tomb} \tbSN \tbGE{pit} \tbSN \tbGE{ditch} \tbSN \tbGE{limit} \tbOP{bogiri} \tbTP{bogree}
\tbLX{bogigbɛre-katakye}\tbHM{} \tbPH{bògìgbɛre-kátákyè} \tbPS{n.} \tbGE{subway, underground train} \tbOP{bogigbɛre-katakyeri} \tbTP{bogigbɛre-katakyeree}
\tbLX{bogikantoo}\tbHM{} \tbPH{bògíkántóó} \tbPS{n.} \tbGE{a type of rodent (not edible)} \tbOP{bogikantori} \tbTP{bogikantoree}
\tbLX{bogisulee}\tbHM{} \tbPH{bògsùlèè} \tbPS{n.} \tbGE{round, deep tomb (like a well)} \tbOP{bogisulli} \tbTP{bogisullee}
\tbLX{bogiti}\tbHM{} \tbPH{bógítì} \tbPS{n.} \tbGE{bucket} \tbOP{bogitiri} \tbTP{bogitiree}
\tbLX{bogi-ũu}\tbHM{} \tbPH{bògì-ũ̀ũ̀} \tbPS{n.} \tbGE{the act of burying a human corpse}
\tbLX{bogi-ũu-faare}\tbHM{} \tbPH{bògì-ũ̀ũ̀-fáárɪ́} \tbPS{n.} \tbGE{xylophone music signaling the completion of the burial of the dead} \tbTP{bogi-ũu-faarɛɛ}
\tbLX{bogi-ũu-libie}\tbHM{} \tbPH{bògì-ũ̀ũ̀-líbíè} \tbPS{n.} \tbGE{fee for those who bury  the dead} \tbTP{bogi-ũu-libiiree}
\tbLX{bogi-ũu-naazoɔ}\tbHM{} \tbPH{bògì-ũ̀ũ̀-náázʊ́ɔ́} \tbPS{n.} \tbGE{chief undertaker (one who buries the dead)} \tbOP{bogi-ũu-naazoore} \tbTP{bogi-ũu-naazoorɛɛ}
\tbLX{bogi-ũuno}\tbHM{} \tbPH{bògì-ũ̀ũ̀nó} \tbPS{n.} \tbGE{one who buries the dead} \tbOP{bogi-ũunimo} \tbTP{bogi-ũunimee}
\tbLX{bogi-ũu-saabo}\tbHM{} \tbPH{bògì-ũ̀ũ̀-sáábʊ́} \tbPS{n.} \tbGE{ritual TZ for those who bury the dead} \tbTP{bogi-ũu-saarɛɛ}
\tbLX{bogi-ũu-sawɔlɔ}\tbHM{} \tbPH{bògì-ũ̀ũ̀-sáwɔ́lɔ́} \tbPS{n.} \tbGE{ritual dish for those who bury the dead} \tbTP{bogi-ũu-sawɔlɛɛ}
\tbLX{bogi-ũu-zanna}\tbHM{} \tbPH{bògì-ũ̀ũ̀-zànná} \tbPS{n.} \tbGE{amateur undertaker} \tbOP{bogi-ũu-zanema} \tbTP{bogi-ũu-zanemɛɛ}
\tbLX{bogivuo}\tbHM{} \tbPH{bògìvùò} \tbPS{n.} \tbGE{uncovered tomb} \tbOP{bogivori} \tbTP{bogivoree}
\tbLX{bogiwɛraa}\tbHM{} \tbPH{bògwɛ́ráà} \tbPS{n.} \tbGE{oblong tomb} \tbOP{bogiwɛrre}
\tbLX{bogizeɛ}\tbHM{} \tbPH{bògzɪ̀ɛ̀} \tbPS{n.} \tbGE{grave, deep pit} \tbOP{bogizeere} \tbTP{bogizeerɛɛ}
\tbLX{bogɔbogɔ}\tbHM{} \tbPH{bʊ́gɔ́bʊ́gɔ́} \tbPS{adv.} \tbGE{ground into pulp} \tbVA{bogaboga}
\tbLX{boguolo}\tbHM{} \tbPH{bʊ̀gúóló} \tbPS{n.} \tbGE{goats given to one to raise in return for some} \tbTP{boguolee}
\tbLX{bog-uuno}\tbHM{} \tbPH{bòg-úúnó} \tbPS{n.} \tbGE{one who buries the dead; undertaker} \tbOP{bog-uunimo} \tbTP{bog-uunimee}
\tbLX{bokolɔ}\tbHM{} \tbPH{bʊ̀kʊ̀lɔ̀} \tbPS{n.} \tbGE{tasteless powdered stuff, tobacco} \tbOP{bokolle} \tbTP{bokolɛɛ} \tbVA{bokola}
\tbLX{bokontawɛ}\tbHM{} \tbPH{bʊ̀kòntáwɛ́} \tbPS{n.} \tbGE{a daily marketplace} \tbOP{bokontawɛre} \tbTP{bokontawɛrɛɛ}
\tbLX{bokoŋ}\tbHM{} \tbPH{bʊ̀kóŋ} \tbPS{n.} \tbGE{lean or thin goat} \tbOP{bokonni} \tbTP{bokonnee}
\tbLX{bokore}\tbHM{} \tbPH{bʊ̀kʊ́rɪ́} \tbPS{n.} \tbGE{old goat} \tbOP{bokora} \tbTP{bokorɛɛ}
\tbLX{bokpɛ}\tbHM{} \tbPH{bʊ̀kpɛ̀} \tbPS{v.} \tbGE{to sympathise with} \tbPD{bokpɛɛ, boorɔkpeɛrɛ, bokpeɛrɛ, bokpeɛrebɛ, bokpeɛraa}%}%, , -}
\tbLX{bolaakaatoo}\tbHM{} \tbPH{bʊ̀lààkààtʊ̀ʊ̀} \tbPS{adv.} \tbGE{talking continuously, noisy}
\tbLX{bolaare}\tbHM{} \tbPH{bʊ̀láárɪ́} \tbPS{v.} \tbGE{to burst strongly} \tbXV{A lɔɔre kɔbɔ da bolaarɛɛ la.} \tbXE{The lorry tyre blasted with a loud noise.} \tbPD{bolaarɛɛ, bolaara, bolaara, bolaareba, bolaaraa, bolaareba } \tbVA{belaare}
\tbLX{bole}\tbHM{} \tbPH{bʊ́lɪ́} \tbPS{v.} \tbGE{to smear all over} \tbPD{bolɛɛ, bola, bola, bolba, bolaa}%}%, , -}
\tbLX{bolebole}\tbHM{} \tbPH{bʊ̀lbʊ́lɪ́} \tbPS{v.} \tbGE{to speak unintelligibly, to mumble} \tbPD{bolebolɛɛ, bolebolɔ, bolebolɔ, bolebolebɔ, bolebolaa, bolebolebɔ }
\tbLX{bolee}\tbHM{} \tbPH{bʊ̀léé} \tbPS{n.} \tbGE{kid (young goat)} \tbOP{bobili} \tbTP{bobillee}
\tbLX{bolenaa}\tbHM{} \tbPH{bʊ̀lénáá} \tbPS{n.} \tbSN \tbGE{tethered goat} \tbSN \tbGE{one restricted from moving} \tbOP{bolenni} \tbTP{bolennee}
\tbLX{bo-leŋi}\tbHM{} \tbPH{bʊ̀-léŋ} \tbPS{n.} \tbGE{the act of tethering goats}
\tbLX{boli}\tbHM{} \tbPH{bólí} \tbPS{v.} \tbGE{to be foolish} \tbPD{bolee, bolo, bolo, bolibo, bolaa, bolibo }
\tbLX{boli}\tbHM{} \tbPH{bòlí} \tbPS{adj.} \tbGE{not adulterated, pure e.g. honey or oil}
\tbLX{bolle}\tbHM{} \tbPH{bʊ̀llɪ̀} \tbPS{v.} \tbGE{to murmur; to talk inaudibly} \tbPD{bollɛɛ, bolla, bolla, boleba, bollaa, boleba }
\tbLX{bolɔ}\tbHM{} \tbPH{bʊ̀lɔ́} \tbPS{n.} \tbGE{a type of vegetable} \tbVA{bola}
\tbLX{bolɔbolɔ}\tbHM{} \tbPH{bʊ́lɔ́bʊ́lɔ́} \tbPS{adv.} \tbGE{lukewarm  (water)} \tbVA{bolabola}
\tbLX{boluu}\tbHM{} \tbPH{bòlúú} \tbPS{pron.} \tbGE{what (interrogative pronoun)}
\tbLX{bombaala}\tbHM{} \tbPH{bʊ́mbààlá} \tbPS{n.} \tbGE{a sick thing} \tbOP{bombaaleba} \tbTP{bombaalɛɛ}
\tbLX{bombaalaa}\tbHM{} \tbPH{bʊ́mbààlàà} \tbPS{n.} \tbGE{a slim or narrow thing} \tbOP{bombaale}
\tbLX{bombaaraa}\tbHM{} \tbPH{bʊ́mbààráà} \tbPS{n.} \tbGE{something that can grow} \tbOP{bombaare}
\tbLX{bombaaraa}\tbHM{} \tbPH{bʊ́mbááráá} \tbPS{n.} \tbGE{the last one} \tbOP{bombaare}
\tbLX{bombalaa}\tbHM{} \tbPH{bʊ́mbáláá} \tbPS{n.} \tbGE{something which is avoided} \tbOP{bomballe}
\tbLX{bombalaa}\tbHM{} \tbPH{bʊ́mbàlàà} \tbPS{n.} \tbGE{kind of thing (referring to something huge or strange)} \tbXV{Yeŋ ka fo ta pɔge a bombalaa ŋa?} \tbXE{Where did you get this kind of thing?}
\tbLX{bombaltakyelmɛ}\tbHM{} \tbPH{bòmbàltáʧɪ́mɛ́} \tbPS{n.} \tbGE{things of all sorts} \tbTP{bombaltakyelmɛɛ}
\tbLX{bombannaa}\tbHM{} \tbPH{bʊ́mbànnáà} \tbPS{n.} \tbGE{something which is known} \tbOP{bombanne} \tbVA{bombaŋenaa}
\tbLX{bombaŋe}\tbHM{} \tbPH{bʊ́mbàŋ} \tbPS{n.} \tbSN \tbGE{the act of knowing} \tbSN \tbGE{knowledge}
\tbLX{bombare}\tbHM{} \tbPH{bʊ̌mbàrɪ̀} \tbPS{n.} \tbSN \tbGE{the act of releasing; the act of setting animals free} \tbSN \tbGE{the act of offering firstfruits to gods} \tbOP{bombarre}
\tbLX{bombeene}\tbHM{} \tbPH{bʊ́mbɪ́ɪ́nɪ́} \tbPS{n.} \tbGE{something that is cooked or ripe} \tbOP{bombeemɛ} \tbTP{bombeenɛɛ}
\tbLX{bombeɛ}\tbHM{} \tbPH{bʊ́mbɪ́ɛ́} \tbPS{n.} \tbGE{something strange or mysterious} \tbOP{bombeere} \tbTP{bombeerɛɛ}
\tbLX{bombeɛraa}\tbHM{} \tbPH{bʊ́mbɪ̀ɛ́ráà} \tbPS{n.} \tbSN \tbGE{sliced strips (of leather)} \tbSN \tbGE{that which is patched or to be patched (as in car tyre)} \tbOP{bombeɛre} \tbTP{bombɛrɛɛ}
\tbLX{bombeɛre}\tbHM{} \tbPH{bʊ́mbɪ̀ɛ̀rɪ́} \tbPS{n.} \tbGE{boils} \tbOP{bombɛbɛ} \tbTP{bombɛrɛɛ}
\tbLX{bombiiri}\tbHM{} \tbPH{bʊ́mbíírí} \tbPS{n.} \tbGE{very young ones; unhatched embryos in eggs} \tbTP{bombiiree}
\tbLX{bombile}\tbHM{} \tbPH{bòmbílé} \tbPS{n.} \tbGE{ young donkey; foal} \tbOP{bombilii} \tbTP{bombillee}
\tbLX{bombile}\tbHM{} \tbPH{bʊ́mbílé} \tbPS{n.} \tbGE{small thing} \tbOP{bombilii}
\tbLX{bombiri}\tbHM{} \tbPH{bʊ́mbírì} \tbPS{n.} \tbGE{seed; cereal} \tbSG{bombiruu} \tbOP{bombie} \tbTP{bombiiree}
\tbLX{bombo}\tbHM{} \tbPH{bǒmbò} \tbPS{n.} \tbSN \tbGE{a small plant with a jellylike tuber} \tbSN \tbGE{the tuber or flour made from the tuber of this plant} \tbSN \tbGE{fried dish made from the flour of this plant} \tbSG{bomboruu (for plant or tuber)} \tbOP{bombori} \tbTP{bomboree}
\tbLX{bombolaa}\tbHM{} \tbPH{bʊ́mbʊ́láá} \tbPS{n.} \tbGE{a thing that is not of the same species; a strange thing} \tbOP{bombolle} \tbTP{bombollɛɛ}
\tbLX{bombone}\tbHM{} \tbPH{bʊ́mbʊ́nɪ̀} \tbPS{n.} \tbGE{a useless thing} \tbOP{bomboma} \tbTP{bombonnɛɛ}
\tbLX{bombonoone}\tbHM{} \tbPH{bǒmbònʊ̀ʊ̀nɪ̀} \tbPS{n.} \tbGE{a type of tree} \tbOP{bombonooma} \tbTP{ bombonoonɛɛ}
\tbLX{bomboŋɔ}\tbHM{} \tbPH{bʊ́mbʊ̀ŋɔ́} \tbPS{n.} \tbGE{something that cannot be classified} \tbOP{bombonne} \tbTP{bombonnɛɛ}
\tbLX{bombooree}\tbHM{} \tbPH{bʊ́mbʊ́ʊ́rɪ́} \tbPS{n.} \tbSN \tbGE{something that is kept for posterity} \tbSN \tbGE{seed for sowing} \tbOP{bomboora} \tbTP{bomboorɛɛ}
\tbLX{bomboorɔ}\tbHM{} \tbPH{bʊ́mbʊ́ʊ́rɔ́} \tbPS{n.} \tbGE{seed for planting} \tbSG{bombooroo} \tbTP{bomboorɛɛ} \tbVA{bomboora}
\tbLX{bomboɔraa}\tbHM{} \tbPH{bʊ́mbʊ́ɔ́ráá} \tbPS{n.} \tbGE{something that is sought; something that is wanted or liked} \tbOP{bomboɔre} \tbTP{bomboɔrɛɛ}
\tbLX{bomboɔre}\tbHM{} \tbPH{bʊ́mbʊ́ɔ́rɪ́} \tbPS{n.} \tbSN \tbGE{things one wishes to have} \tbSN \tbGE{wealth}
\tbLX{bomboraa}\tbHM{} \tbPH{bʊ́mbʊ̀ráà} \tbPS{n.} \tbGE{something that is sown; a crop} \tbOP{bomborre} \tbTP{ bomborrɛɛ}
\tbLX{bombore}\tbHM{} \tbPH{bʊ́mbʊ̀rɪ̀} \tbPS{n.} \tbGE{the act of sowing}
\tbLX{bombo sawɔlɔ}\tbHM{} \tbPH{bòmbò sáwɔ́lɔ́} \tbPS{n.} \tbGE{fried duff from bombo (flour)} \tbTP{bombo sawɔlɛɛ} \tbVA{bombo sawala}
\tbLX{bombɔ}\tbHM{} \tbPH{bʊ́mbɔ́} \tbPS{n.} \tbGE{quest for wealth}
\tbLX{bombɔge}\tbHM{} \tbPH{bʊ́mbɔ́gɪ́} \tbPS{n.} \tbGE{the act of grinding into fine pulp} \tbOP{bombɔgre} \tbTP{bombɔgrɛɛ}
\tbLX{bombɔgɔ}\tbHM{} \tbPH{bʊ́mbɔ́gɔ́} \tbPS{n.} \tbGE{the forearm or leg of an animal} \tbOP{bombɔgerɔ} \tbTP{bombɔgerɛɛ} \tbVA{bombɔge}
\tbLX{bombugilaa}\tbHM{} \tbPH{bʊ́mbúglàá} \tbPS{n.} \tbGE{something that is nursed} \tbOP{bombugili}
\tbLX{bombulaa}\tbHM{} \tbPH{bʊ́mbúláá} \tbPS{n.} \tbGE{something that can sprout from a seed} \tbOP{bombulli} \tbTP{bombullee}
\tbLX{bombuli}\tbHM{} \tbPH{bʊ́mbúlí} \tbPS{n.} \tbGE{shoot of a new plant} \tbOP{bombulo} \tbTP{bombullee}
\tbLX{bombuuraa}\tbHM{} \tbPH{bʊ́mbúúràá} \tbPS{n.} \tbSN \tbGE{a measure} \tbSN \tbGE{something that is respected} \tbOP{bombuuri}
\tbLX{bommaalaa}\tbHM{} \tbPH{bʊ́mmààláà} \tbPS{n.} \tbGE{something that is made} \tbOP{bommaale}
\tbLX{bommaale}\tbHM{} \tbPH{bʊ́mmààlɪ̀} \tbPS{n.} \tbGE{the act of making} \tbOP{bommaaleba}
\tbLX{bommaane}\tbHM{} \tbPH{bʊ́mmàànɪ́} \tbPS{n.} \tbSN \tbGE{an egg that is immature} \tbSN \tbGE{something that is tiny} \tbOP{bommaama}
\tbLX{bommaaroŋ}\tbHM{} \tbPH{bʊ́mmáárʊ́ŋ} \tbPS{n.} \tbGE{a wet, cold, or fresh thing} \tbOP{bommaara}
\tbLX{bommannaa}\tbHM{} \tbPH{bʊ́mmánnáà} \tbPS{n.} \tbGE{a measure; something that is measured}
\tbLX{bompaala}\tbHM{} \tbPH{bʊ́mpáálà} \tbPS{n.} \tbSN \tbGE{new or fresh things} \tbSN \tbGE{firstfruits of crops} \tbOP{bompaala} \tbTP{bompalɛɛ}
\tbLX{bompaare}\tbHM{} \tbPH{bʊ́mpáárɪ̀} \tbPS{n.} \tbGE{the vulva of any animal} \tbOP{bompɛɛrɛɛ}
\tbLX{bompaare}\tbHM{} \tbPH{bǒmpáárɪ̀} \tbPS{n.} \tbGE{the vulva of a donkey} \tbOP{bompaɛ} \tbTP{ bompɛɛrɛɛ}
\tbLX{bompaare}\tbHM{} \tbPH{bǒmpáárɪ̀} \tbPS{n.} \tbGE{foot rot} \tbOP{bompaɛ} \tbTP{ bompɛɛrɛɛ}
\tbLX{bompare}\tbHM{} \tbPH{bʊ́mpárɪ̀} \tbPS{n.} \tbSN \tbGE{residue} \tbSN \tbGE{bottom} \tbOP{bomparɛɛ}
\tbLX{bompeelaa}\tbHM{} \tbPH{bʊ́mpɪ́ɪ́láá} \tbPS{n.} \tbGE{a very young animal} \tbOP{bompeele} \tbTP{bompeelɛɛ}
\tbLX{bompeelaa}\tbHM{} \tbPH{bʊ́mpɪ́ɪ́láá} \tbPS{n.} \tbGE{a piece torn from a piece of cloth} \tbOP{bompeele} \tbTP{bompeelɛɛ}
\tbLX{bompege}\tbHM{} \tbPH{bʊ́mpɪ̀gɪ̀} \tbPS{n.} \tbGE{the act of decorating, improving or adding worth to something}
\tbLX{bompegelaa}\tbHM{} \tbPH{bʊ́mpɪ́gláà} \tbPS{n.} \tbGE{something that is moving and must be intercepted} \tbOP{bompegele} \tbTP{bompegelɛɛ}
\tbLX{bompegeraa}\tbHM{} \tbPH{bʊ́mpɪ́gráà} \tbPS{n.} \tbGE{something that is adorned} \tbOP{bompegere} \tbTP{bompegerɛɛ}
\tbLX{bompegɛ}\tbHM{} \tbPH{bʊ́mpɪ́gɛ́} \tbPS{n.} \tbGE{a piece of a broken container (calabash)} \tbOP{bompegerre} \tbTP{ bompegerɛɛ}
\tbLX{bompelaa}\tbHM{} \tbPH{bʊ́mpɪ̀làá} \tbPS{n.} \tbGE{a white thing} \tbOP{bompeɛle} \tbTP{bompeɛlɛɛ}
\tbLX{bompere}\tbHM{} \tbPH{bʊ́mpɪ̀rɪ̀} \tbPS{n.} \tbGE{the act of dehusking (removing the dry outer cover of a grain)}
\tbLX{bompere}\tbHM{} \tbPH{bʊ́pɪ́rɪ́} \tbPS{n.} \tbGE{the act of scratching (as by fowl)}
\tbLX{bompɛɛre}\tbHM{} \tbPH{bʊ́mpɛ̀ɛ̀rɪ́} \tbPS{n.} \tbGE{an immature thing} \tbOP{bompɛbɛ}
\tbLX{bompɛge}\tbHM{} \tbPH{bʊ́mpɛ́gɪ̀} \tbPS{n.} \tbGE{the act of washing the shell or bark of something}
\tbLX{bompɛgelaa}\tbHM{} \tbPH{bʊ́mpɛ́gláà} \tbPS{n.} \tbGE{something that is carried by holding it to one's side} \tbOP{bompɛgele} \tbTP{bompɛgelɛɛ}
\tbLX{bompɛgeraa}\tbHM{} \tbPH{bʊ́pɛ́gráá} \tbPS{n.} \tbSN \tbGE{something that has been washed} \tbSN \tbGE{something for washing} \tbOP{bompɛgere} \tbTP{bompɛgerɛɛ}
\tbLX{bompɛlaa}\tbHM{} \tbPH{bʊ́mpɛ́láá} \tbPS{n.} \tbGE{something that can be peeled} \tbOP{bompɛlle} \tbTP{ bompɛllɛɛ}
\tbLX{bompɛle}\tbHM{} \tbPH{bʊ́mpɛ́lɪ́} \tbPS{n.} \tbGE{the act of peeling}
\tbLX{bompɛle}\tbHM{} \tbPH{bʊḿpɛ̀lì} \tbPS{n.} \tbGE{peel} \tbOP{bompɛlɛ} \tbTP{bompɛllɛɛ}
\tbLX{bompɛnnaa}\tbHM{} \tbPH{bʊ́mpɛ̀nnáá} \tbPS{n.} \tbGE{something that is carved} \tbOP{bompɛnne} \tbTP{bompɛnnɛɛ}
\tbLX{bompɛnne}\tbHM{} \tbPH{bʊ́mpɛ̀nnɪ̀} \tbPS{n.} \tbGE{the act of carving}
\tbLX{bompi}\tbHM{} \tbPH{bʊ́mpí} \tbPS{n.} \tbSN \tbGE{the process of becoming distended or bloated} \tbSN \tbGE{claims in advance for something (see pi)}
\tbLX{bompigiraa}\tbHM{} \tbPH{bʊ́mpǐgráà} \tbPS{n.} \tbGE{something that is to be watched or ought to be watched} \tbOP{bompigiri}
\tbLX{bompiiraa}\tbHM{} \tbPH{bʊ́mpììráà} \tbPS{n.} \tbSN \tbGE{a lost thing that is found; an unexpected find} \tbSN \tbGE{an easy find} \tbOP{bompiiri}
\tbLX{bompiiraa}\tbHM{} \tbPH{bʊ́mpííráá} \tbPS{n.} \tbGE{a distended thing} \tbOP{bompiiri}
\tbLX{bompiiraa}\tbHM{} \tbPH{bʊ́mpííráá} \tbPS{n.} \tbGE{something that has been earmarked for someone} \tbOP{bompiiri}
\tbLX{bompiire}\tbHM{} \tbPH{bʊ́mpííré} \tbPS{n.} \tbGE{one who finds a lost item} \tbOP{bompiiribe} \tbTP{bompiiribee}
\tbLX{bompollaa}\tbHM{} \tbPH{bʊ́mpʊ́lláá} \tbPS{n.} \tbGE{something on which an oath is taken} \tbOP{bompolle}
\tbLX{bompoɔ}\tbHM{} \tbPH{bʊ́mpʊ́ɔ́} \tbPS{n.} \tbGE{a pregnant thing (nonhuman)} \tbOP{bompoore} \tbTP{ bompoorɛɛ}
\tbLX{bompoɔraa}\tbHM{} \tbPH{bʊ́mpʊ́ɔ́ráá} \tbPS{n.} \tbGE{something by which one swears} \tbOP{bompoɔre} \tbTP{ bompoɔrɛɛ}
\tbLX{bomporaa}\tbHM{} \tbPH{bʊ́mpʊ́ráá} \tbPS{n.} \tbGE{something that is poured} \tbOP{bomporre} \tbTP{ bomporrɛɛ}
\tbLX{bompore}\tbHM{} \tbPH{bʊ́mpʊ̀rɪ̀} \tbPS{n.} \tbGE{a situation of plentiful things} \tbOP{bomporre}
\tbLX{bompore}\tbHM{} \tbPH{bʊ́mpʊ́rɪ́} \tbPS{n.} \tbGE{something that is small and short; a little thing} \tbOP{bompora} \tbTP{bomporɛɛ}
\tbLX{bompore}\tbHM{} \tbPH{bʊ́mpʊ́rɪ̀} \tbPS{n.} \tbGE{the act of pouring}
\tbLX{bompɔ}\tbHM{} \tbPH{bʊ́mpɔ́} \tbPS{n.} \tbGE{the act of swearing or taking an oath on something}
\tbLX{bompɔge}\tbHM{} \tbPH{bʊ́mpɔ́gɪ́} \tbPS{n.} \tbGE{a female thing} \tbOP{bompɔgeba} \tbTP{bompɔgrɛɛ}
\tbLX{bompɔge}\tbHM{} \tbPH{bʊ́mpɔ̀gɪ̀} \tbPS{n.} \tbGE{the act of confining animals in enclosures}
\tbLX{bompɔgeraa}\tbHM{} \tbPH{bʊ́mpɔ̀gráà} \tbPS{n.} \tbGE{something that is confined in an enclosure} \tbOP{bompɔgere}
\tbLX{bompɔle}\tbHM{} \tbPH{bʊ́mpɔ́lɪ̀} \tbPS{n.} \tbGE{a young thing} \tbOP{bompɔlɔ} \tbTP{ bompɔllɛɛ}
\tbLX{bompɔre}\tbHM{} \tbPH{bʊ́mpɔ́rɪ́} \tbPS{n.} \tbGE{the act of plucking something; (fruits or groundnuts)}
\tbLX{bompɔrɔ}\tbHM{} \tbPH{bʊ́mpɔ́rɔ̀} \tbPS{n.} \tbGE{harvester} \tbOP{bompɔreba} \tbTP{bompɔrebɛɛ}
\tbLX{bompugo}\tbHM{} \tbPH{bʊ́mpúgó} \tbPS{n.} \tbGE{an ash-coloured thing} \tbOP{bompugri} \tbTP{ bompugree}
\tbLX{bompulonpuloŋ}\tbHM{} \tbPH{bʊ́mpùlònpúlóŋ} \tbPS{n.} \tbGE{various things mixed up}
\tbLX{bompuuraa}\tbHM{} \tbPH{bʊ́mpùùráà} \tbPS{n.} \tbGE{something that has flowered} \tbOP{bom\-puu\-ri} \tbTP{bompuuree}
\tbLX{bompuuraa}\tbHM{} \tbPH{bʊ́mpùùráà} \tbPS{n.} \tbGE{something that is or can be praised} \tbOP{bompuuri} \tbTP{bompuuree}
\tbLX{bompuuri}\tbHM{} \tbPH{bʊ́mpúúrí} \tbPS{n.} \tbGE{flowers of things} \tbTP{bompuuree}
\tbLX{bonaare}\tbHM{} \tbPH{bʊ̀náárɪ̀} \tbPS{adj.} \tbGE{four times}
\tbLX{bon-agraa}\tbHM{} \tbPH{bʊ́ŋágráà} \tbPS{n.} \tbSN \tbGE{winged creature} \tbSN \tbGE{ insect} \tbOP{boŋ-agre}
\tbLX{bonda}\tbHM{} \tbPH{bʊ́ndà} \tbPS{n.} \tbGE{the act of buying}
\tbLX{bondaa}\tbHM{} \tbPH{bòndáá} \tbPS{n.} \tbGE{male donkey} \tbOP{bondaare} \tbTP{bondarɛɛ}
\tbLX{bondaa}\tbHM{} \tbPH{bʊ́ndáá} \tbPS{n.} \tbGE{a male thing} \tbOP{bondaare} \tbTP{bondarɛɛ}
\tbLX{bondaa}\tbHM{} \tbPH{bʊ́ndáà} \tbPS{n.} \tbGE{the act of pushing} \tbOP{bonda}
\tbLX{bondaana}\tbHM{} \tbPH{bʊ́ndááná} \tbPS{n.} \tbGE{rich person; one who has wealth}  \tbOP{bondaanema/bondaama} \tbTP{ bondaamɛɛ/bondaanɛɛ}
\tbLX{bondaaraa}\tbHM{} \tbPH{bʊ́ndààráà} \tbPS{n.} \tbGE{something that is bought; something that is meant for sale} \tbOP{bondaare}
\tbLX{bondaga}\tbHM{} \tbPH{bòndàgà} \tbPS{n.} \tbGE{carefree} \tbOP{bondagre} \tbTP{ bondagrɛɛ}
\tbLX{bonde}\tbHM{} \tbPH{bʊ́ndɪ́} \tbPS{n.} \tbSN \tbGE{the act of taking} \tbSN \tbGE{the act of harvesting}
\tbLX{bondegere}\tbHM{} \tbPH{bʊ́ndɪ̀grɪ} \tbPS{n.} \tbGE{a dwarflike thing} \tbOP{bondegɛ}
\tbLX{bondemannewullaa}\tbHM{} \tbPH{bʊ́ndɪ́mánnɪ́wúlláá} \tbPS{n.} \tbGE{example} \tbOP{bondemannewulli}
\tbLX{bondɛgere}\tbHM{} \tbPH{bʊ́ndɛ̌grɪ̀} \tbPS{n.} \tbGE{a dirty thing} \tbOP{bondɛgɛ}
\tbLX{bondi}\tbHM{} \tbPH{bʊ́ndì} \tbPS{n.} \tbGE{the act of eating}
\tbLX{bondikaama}\tbHM{} \tbPH{bʊ̀ndìkáámá} \tbPS{n.} \tbGE{delicious and assorted food; luncheon} \tbTP{bondikaanɛɛ}
\tbLX{bondikaraa}\tbHM{} \tbPH{bʊ́ndìkáráá} \tbPS{n.} \tbGE{raw food; food that is not well cooked} \tbOP{bondikare} \tbTP{bondikarɛɛ}
\tbLX{bondi-kpaa}\tbHM{} \tbPH{bóndì-kpàà} \tbPS{n.} \tbGE{leftover food} \tbOP{bondi-kpaare} \tbTP{bondi-kpaarɛɛ}
\tbLX{bondimaala}\tbHM{} \tbPH{bʊ́ndímáálá} \tbPS{n.} \tbGE{cook, one who prepares food}
\tbLX{bondiraa}\tbHM{} \tbPH{bʊ́ndìráà} \tbPS{n.} \tbSN \tbGE{food; something that is edible} \tbSN \tbGE{a sharp thing (blade)} \tbOP{bondirri} \tbTP{bondirree}
\tbLX{bondire}\tbHM{} \tbPH{bʊ́ndíré} \tbPS{n.} \tbSN \tbGE{one who eats} \tbSN \tbGE{rightful owner} \tbOP{bondiribe} \tbTP{bondiribee}
\tbLX{bondirii}\tbHM{} \tbPH{bʊ́ndìríì} \tbPS{n.} \tbGE{food, things to eat} \tbTP{bondirree}
\tbLX{bondi-sãa} \tbHM{} \tbPH{bʊ́ndì-sã̀ã̀} \tbPS{n.} \tbGE{the act of wasting food}
\tbLX{bondizie}\tbHM{} \tbPH{bʊ́ndìzíé} \tbPS{n.} \tbGE{restaurant} \tbOP{bondiziiri} \tbTP{bondiziiree}
\tbLX{bondogelaa}\tbHM{} \tbPH{bʊ́ndʊ́gláá} \tbPS{n.} \tbGE{something that is used for cooking} \tbOP{bondogele}
\tbLX{bondogeraa}\tbHM{} \tbPH{bʊ́ndʊ́gráá} \tbPS{n.} \tbGE{a boiled thing} \tbOP{bondogere}
\tbLX{bondonaa}\tbHM{} \tbPH{bʊ́ndʊ̀náà} \tbPS{n.} \tbSN \tbGE{something that bites} \tbSN \tbGE{insect, bug} \tbOP{bondonne} \tbTP{bondonnɛɛ} \tbVA{\mbox{kyibe}}
\tbLX{bondori}\tbHM{} \tbPH{bʊ́ndórí} \tbPS{n.} \tbGE{ heap} \tbOP{bondoree} \tbTP{bondorree}
\tbLX{bondɔgelaa}\tbHM{} \tbPH{bʊ́ndɔ́gláá} \tbPS{n.} \tbGE{an extra thing} \tbOP{bondɔgele} \tbLX{bondɔgeraa}\tbHM{} \tbPH{bʊ́ndɔ̀gèràà} \tbPS{n.} \tbGE{hat which has been born} \tbOP{bondɔgere}
\tbLX{bondɔgerɔ}\tbHM{} \tbPH{bʊ́ndɔ̀gèrɔ̀} \tbPS{n.} \tbGE{an animal that has recently littered} \tbOP{bondɔgere} \tbTP{bondɔgerɛɛ}
\tbLX{bone}\tbHM{} \tbPH{bʊ́nɪ̀} \tbPS{n.} \tbGE{thing} \tbOP{boma} \tbTP{bonnɛɛ}
\tbLX{bone}\tbHM{} \tbPH{bʊ́nɪ́} \tbPS{adj.} \tbGE{worthless or useless} \tbOP{boma}
\tbLX{bon-ennaa}\tbHM{} \tbPH{bʊ́n-ɪ́nnáà} \tbPS{n.} \tbGE{container} \tbOP{bon-enne} \tbTP{bon-ennɛɛ}
\tbLX{bonfa}\tbHM{} \tbPH{bʊ́nfá} \tbPS{n.} \tbGE{the act of robbing, robbery}
\tbLX{bonfaa}\tbHM{} \tbPH{bʊ́nfáá} \tbPS{n.} \tbGE{a bad or evil thing} \tbOP{bonfaare} \tbTP{ bonfaarɛɛ}
\tbLX{bonfaa}\tbHM{} \tbPH{bʊ́nfáà} \tbPS{n.} \tbGE{robbery}
\tbLX{bonfaara}\tbHM{} \tbPH{bʊ́nfáárá} \tbPS{n.} \tbGE{robber, thief} \tbOP{bonfareba} \tbTP{ bonfabarɛɛ}
\tbLX{bonfaaraa}\tbHM{} \tbPH{bʊ́nfááráá} \tbPS{n.} \tbGE{something that is robbed} \tbOP{bonfaare} \tbTP{ bonfaarɛɛ}
\tbLX{bonfane}\tbHM{} \tbPH{bʊ́nfànɪ́} \tbPS{n.} \tbGE{a rough thing} \tbOP{bonfama} \tbTP{ bonfannɛɛ}
\tbLX{bonfarafara}\tbHM{} \tbPH{bʊ́nfàràfàrà} \tbPS{n.} \tbGE{a rough, scaly thing}
\tbLX{bonfẽe} \tbHM{} \tbPH{bʊ́nfɪ̃́ɪ̃́} \tbPS{n.} \tbGE{a small quantity} \tbOP{bonfeene} \tbTP{bonfeenɛɛ}
\tbLX{bonfeeloo}\tbHM{} \tbPH{bʊ́fɪ́ɪ́lʊ́ʊ́} \tbPS{n.} \tbGE{an opening that is small and tight} \tbOP{bonfeelɛ} \tbTP{ bonfeelɛɛ}
\tbLX{bonfegɛ}\tbHM{} \tbPH{bʊ́nfɪ̀gɛ́} \tbPS{n.} \tbGE{narrow space} \tbVA{bonfogeloo}
\tbLX{bonfogeloo}\tbHM{} \tbPH{bʊ́nfʊ́glʊ́ʊ́} \tbPS{n.} \tbGE{a narrow thing} \tbOP{bofogela} \tbTP{ bonfogelɛɛ}
\tbLX{bonfogiluu}\tbHM{} \tbPH{bʊ́nfóglúú} \tbPS{n.} \tbGE{a seedless pod} \tbOP{bonfogilo} \tbTP{ bonfogilee}
\tbLX{bonfole}\tbHM{} \tbPH{bʊ́nfʊ̀lɪ́} \tbPS{n.} \tbGE{an immature thing} \tbOP{bonfola} \tbTP{ bonfolɛɛ}
\tbLX{bonfolo}\tbHM{} \tbPH{bʊ́nfòló} \tbPS{n.} \tbGE{powerful spiritual beings} \tbOP{bonfolo} \tbTP{ bonfolee}
\tbLX{bonfõɔo}\tbHM{} \tbPH{bʊ́nfʊ̃́ɔ̃́ʊ̃́} \tbPS{n.} \tbGE{a light thing} \tbOP{bonfoɔne} \tbTP{bonfoɔnɛɛ}
\tbLX{bonfɔraa}\tbHM{} \tbPH{bʊ́nfɔ́ráá} \tbPS{n.} \tbGE{a narrow thing} \tbOP{bonfɔrre} \tbTP{bonfɔrɛɛ}
\tbLX{bonfuraa}\tbHM{} \tbPH{bʊ́nfùráà} \tbPS{n.} \tbGE{something that is used to plug an opening} \tbOP{bonfurri} \tbTP{ bonfurree}
\tbLX{bonfuuraa}\tbHM{} \tbPH{bʊ́nfùùráá} \tbPS{n.} \tbSN \tbGE{something that is cool} \tbSN \tbGE{something that threatens} \tbOP{bonfuuri} \tbTP{bonfuuree}
\tbLX{bonii}\tbHM{} \tbPH{bʊ̀níì} \tbPS{adj.} \tbGE{eight times}
\tbLX{bonkyaaraa}\tbHM{} \tbPH{bʊ́nʧááráá} \tbPS{n.} \tbGE{gift} \tbOP{bonkyaare} \tbTP{bonkyaarɛɛ}
\tbLX{bonkyẽɛ̃}\tbHM{} \tbPH{bʊ́nʧɪ̃́ɛ̃́} \tbPS{n.} \tbGE{a torn thing} \tbOP{bonkyeɛne/bonkyeɛnɛ} \tbTP{ bonkyeɛnɛɛ}
\tbLX{bonkyeɛraa}\tbHM{} \tbPH{bʊ́nʧɪ́ɛ́ráá} \tbPS{n.} \tbSN \tbGE{something that is contributed} \tbSN \tbGE{something that is cut} \tbSN \tbGE{something used for cutting} \tbOP{bonkyeɛre} \tbTP{ bonkyeɛrɛɛ}
\tbLX{bonkyɛlee}\tbHM{} \tbPH{bʊ́nʧɛ́lɪ́ɪ́} \tbPS{n.} \tbGE{a half thing} \tbOP{bonkyɛlle}
\tbLX{bonlaane}\tbHM{} \tbPH{bʊ́nláánɪ́} \tbPS{n.} \tbGE{a worthless thing} \tbOP{bonlaama} \tbTP{ bonlaanɛɛ}
\tbLX{bonlane}\tbHM{} \tbPH{bʊ́nlánɪ́} \tbPS{n.} \tbGE{uncastrated animal} \tbOP{bonlama} \tbTP{ bonlamɛɛ}
\tbLX{bonlee}\tbHM{} \tbPH{bʊ́nléé} \tbPS{n.} \tbGE{a small thing e.g. object/nonhuman} \tbOP{bombilii} \tbVA{bombile}
\tbLX{bonlee}\tbHM{} \tbPH{bònléé} \tbPS{n.} \tbGE{colt, young donkey} \tbOP{bombilii}
\tbLX{bonnaakale}\tbHM{} \tbPH{bǒnnààkálɪ́} \tbPS{n.} \tbGE{the act of rolling over and over. (donkey having a sun bath)} \tbTP{bonnakalɛɛ}
\tbLX{bonnaŋ}\tbHM{} \tbPH{bʊ́nnáŋ} \tbPS{n.} \tbGE{scarcity} \tbOP{bonnanne} \tbTP{ bonnannnɛɛ}
\tbLX{bonne}\tbHM{} \tbPH{bʊ̀nnɪ̀, bʊ́nnɪ́} \tbPS{v.} \tbGE{to doubt; to speculate, to guess} \tbPD{bonnɛɛ, bonnɔ, bonnɔ, bonema, bonnaa, bonema }
\tbLX{bonneɛ}\tbHM{} \tbPH{bʊ́nnɪ́ɛ́} \tbPS{n.} \tbSN \tbGE{the act of trampling on things} \tbSN \tbGE{the act of ironing clothes} \tbOP{bonneɛne} \tbTP{bonneɛnɛɛ}
\tbLX{bonni}\tbHM{} \tbPH{bónní} \tbPS{v.} \tbSN \tbGE{to burn (pain)} \tbSN \tbGE{to burn with fire} \tbPD{bonnee, bonno, bonno, bonnaa, bonimo}%}%, , -}
\tbLX{bonnimiri}\tbHM{} \tbPH{bʊ́nnímírì} \tbPS{n.} \tbGE{something that is very valuable} \tbOP{bonnimie} \tbTP{ bonimiiree}
\tbLX{bonnoɔ}\tbHM{} \tbPH{bʊ́nnʊ̀ɔ̀} \tbPS{n.} \tbGE{a sweet thing} \tbOP{bonnoore} \tbTP{ bonnoorɛɛ}
\tbLX{bonnyaa}\tbHM{} \tbPH{bʊ́nɲáà} \tbPS{n.} \tbSN \tbGE{chest of;} \tbSN \tbGE{most important part of} \tbOP{bonnyaane} \tbTP{bonnyaanɛɛ}
\tbLX{bonnyaanaa}\tbHM{} \tbPH{bʊ́nɲáànáà} \tbPS{n.} \tbGE{a clever or wise animal} \tbOP{bonnyaane}
\tbLX{bonnyaara}\tbHM{} \tbPH{bʊ́nɲàárɪ̀} \tbPS{n.} \tbGE{coarseness (flour)} \tbTP{bonnyaarɛɛ}
\tbLX{bonnyagere}\tbHM{} \tbPH{bʊ́nɲágɪ́rɪ́} \tbPS{n.} \tbGE{intestines of nonhuman creatures} \tbOP{bonnyaga} \tbTP{bonnyagerɛɛ}
\tbLX{bonnyaŋaa}\tbHM{} \tbPH{bʊ́nɲáŋáá} \tbPS{n.} \tbGE{female} \tbOP{bonnyaŋene/bonnyaane} \tbTP{bonnyaanɛɛ}
\tbLX{bonnyaŋaa}\tbHM{} \tbPH{bònɲàŋáà} \tbPS{n.} \tbGE{female donkey} \tbOP{boŋnyaŋene} \tbTP{ boŋnyaŋenɛɛ}
\tbLX{bonnyuu}\tbHM{} \tbPH{bʊ́nɲúú} \tbPS{n.} \tbSN \tbGE{something that smells} \tbSN \tbGE{smelling medicinal lotion} \tbOP{bonnyuuri} \tbTP{bonyuu\-ree}
\tbLX{bonnyuu}\tbHM{} \tbPH{bʊ́nɲúú} \tbPS{n.} \tbGE{the neck of something} \tbOP{bonnyuuri} \tbTP{bonyuu\-ree}
\tbLX{bonnyuuraa}\tbHM{} \tbPH{bʊ́nɲúúráá} \tbPS{n.} \tbSN \tbGE{something that is drunk; a drink} \tbSN \tbGE{something that is used for drinking} \tbOP{bonnyuuri}
\tbLX{bononnoonaa}\tbHM{} \tbPH{bʊ̀nʊ́nnʊ́ʊ́náá} \tbPS{n.} \tbGE{a type of tree with juicy fruits} \tbVA{bodaanonnoonaa}
\tbLX{bonoone}\tbHM{} \tbPH{bʊ̀nʊ̀ʊ̀nɪ̀} \tbPS{n.} \tbGE{a type of tree with fleshy edible fruits; medicinal} \tbOP{bonooma} \tbTP{ bonoonɛɛ}
\tbLX{bonɔbonɔ}\tbHM{} \tbPH{bʊ̀nɔ̀bʊ̀nɔ̀} \tbPS{adv.} \tbGE{oily-looking} \tbXV{A zeɛre zu deɛ e la bonɔbonɔ lɛ.} \tbXE{The surface of the soup looks oily.} \tbVA{bonabona}
\tbLX{bonsãa}\tbHM{} \tbPH{bʊ́nsã̀ã̀} \tbPS{n.} \tbGE{the act of being wasteful, destructive}
\tbLX{bonsaalaa}\tbHM{} \tbPH{bʊ́nsààlàà} \tbPS{n.} \tbGE{something that is smooth} \tbOP{bonsaale} \tbTP{ bonsaalɛɛ}
\tbLX{bonsaaloŋ}\tbHM{} \tbPH{bʊ́nsààlʊ́ŋ} \tbPS{n.} \tbGE{a slippery thing} \tbOP{bonsaale} \tbTP{bonsaalɛɛ}
\tbLX{bonsaana}\tbHM{} \tbPH{bʊ́nsáánà} \tbPS{n.} \tbGE{a new or strange thing} \tbOP{bonsaama} \tbTP{ bonsaanɛɛ}
\tbLX{bonsaana}\tbHM{} \tbPH{bʊ́nsááná} \tbPS{n.} \tbGE{one who destroys} \tbOP{bonsaanema}
\tbLX{bonsaanaa}\tbHM{} \tbPH{bʊ́nsáánáá} \tbPS{n.} \tbGE{a spoilt thing} \tbOP{bonsaane} \tbTP{ bonsaanɛɛ}
\tbLX{bonsammo}\tbHM{} \tbPH{bʊ́nsàmmʊ́} \tbPS{n.} \tbGE{unshelled nut} \tbOP{bonsama} \tbTP{bonsamɛɛ}
\tbLX{bonsaraa}\tbHM{} \tbPH{bʊ́nsáràà} \tbPS{n.} \tbGE{a young female} \tbOP{bonsarre} \tbTP{ bonsarɛɛ}
\tbLX{bonseenaa}\tbHM{} \tbPH{bʊ́nsɪ́ɪ́náà} \tbPS{n.} \tbGE{something that is touched} \tbOP{boseene} \tbTP{ bonseenɛɛ}
\tbLX{bonseɛraa}\tbHM{} \tbPH{bʊ́nsɪ́ɛ́ráá} \tbPS{n.} \tbSN \tbGE{something that is worn on the waist} \tbSN \tbGE{something that is sewn} \tbOP{bonseɛre} \tbTP{ bonseɛrɛɛ}
\tbLX{bonseɛraa}\tbHM{} \tbPH{bʊ́nsɪ́ɛ́ráá} \tbPS{n.} \tbGE{that which is better than} \tbOP{bonseɛre} \tbTP{ bonseɛrɛɛ}
\tbLX{bonseɛraa}\tbHM{} \tbPH{bʊ́nsɪ́ɛ́ráá} \tbPS{n.} \tbGE{something that is roasted} \tbOP{bonseɛre} \tbTP{bonseɛrɛɛ}
\tbLX{bonsennaa}\tbHM{} \tbPH{bʊ̀nsɪ́nnáà} \tbPS{n.} \tbGE{something that is the right size} \tbOP{bonsenne} \tbTP{bonsennɛɛ}
\tbLX{bonserɛ}\tbHM{} \tbPH{bʊ́nsɪ̀rɛ́} \tbPS{n.} \tbGE{a valuable thing} \tbOP{bonserre} \tbTP{ bonserɛɛ/bonserrɛɛ}
\tbLX{bonsɛgeraa}\tbHM{} \tbPH{bʊ́nsɛ́gráá} \tbPS{n.} \tbGE{something that is written} \tbOP{bonsɛgere} \tbTP{ bonsɛgerɛɛ}
\tbLX{bonsɛgeraa}\tbHM{} \tbPH{bʊ́nsɛ̀gráà} \tbPS{n.} \tbGE{something that is used to make another thing sturdy} \tbOP{bonsɛgere}
\tbLX{bonsigiraa}\tbHM{} \tbPH{bʊ́nsìgráà} \tbPS{n.} \tbSN \tbGE{something whose price is reduced} \tbSN \tbGE{something that is lowered (height)} \tbOP{bonsigiri} \tbTP{bonsigiree}
\tbLX{bonso}\tbHM{} \tbPH{bʊ́nsʊ́} \tbPS{n.} \tbGE{the act of owning}
\tbLX{bonso}\tbHM{} \tbPH{bónsó} \tbPS{pron.} \tbGE{why (interrogative pronoun)} \tbXV{Bonso ka fo ba gaa a daa?} \tbXE{Why have you not gone to the market?}
\tbLX{bonso}\tbHM{} \tbPH{bónsó} \tbPS{conj.} \tbGE{because} \tbXV{Ba ba gaa a sori bonso a lɔɔre sãaɛ la.} \tbXE{They could not go on the trip because the car broke down.}
\tbLX{bonsoŋ}\tbHM{} \tbPH{bʊ́nsʊ́ŋ} \tbPS{n.} \tbGE{a good or blessed thing} \tbOP{bonsonne} \tbTP{bonsonnɛɛ}
\tbLX{bonsooraa}\tbHM{} \tbPH{bʊ́nsʊ́ʊ́ráá} \tbPS{n.} \tbGE{that which is owned; a belonging} \tbOP{bonsoore} \tbTP{bonsoorɛɛ}
\tbLX{bonsooraa}\tbHM{} \tbPH{bʊ́nsʊ́ʊ́ráà} \tbPS{n.} \tbGE{something that is asked for or wanted}
\tbLX{bonsɔgelaa}\tbHM{} \tbPH{bʊ́nsɔ́gláà} \tbPS{n.} \tbSN \tbGE{a black thing} \tbSN \tbGE{a hidden thing} \tbOP{bonsɔgelɔ} \tbTP{ bonsɔgelɛɛ}
\tbLX{bonsɔgele}\tbHM{} \tbPH{bʊ́nsɔ́glɪ́} \tbPS{n.} \tbGE{the act of hiding things}
\tbLX{bonsɔraa}\tbHM{} \tbPH{bʊ́nsɔ̀ráà} \tbPS{n.} \tbGE{something that is begged for} \tbOP{bonsɔrre} \tbTP{bonsɔrɛɛ}
\tbLX{bonsɔraa}\tbHM{} \tbPH{bʊ́sɔ́ráá} \tbPS{n.} \tbGE{something that has been counted} \tbXV{A pɔge iri la a bonsɔraa kaŋa.} \tbXE{The woman took one of the already counted ones.} \tbOP{bonsɔrre}
\tbLX{bonsu}\tbHM{} \tbPH{bʊ́nsù} \tbPS{n.} \tbGE{the act of wearing (clothes)}
\tbLX{bonsu}\tbHM{} \tbPH{bʊ́nsù} \tbPS{n.} \tbGE{the act of pushing things into an opening}
\tbLX{bonsugo}\tbHM{} \tbPH{bʊ́nsúgò} \tbPS{n.} \tbGE{things in great numbers} \tbOP{bonsugri} \tbTP{ bonsugree}
\tbLX{bonsuuraa}\tbHM{} \tbPH{bʊ́nsúùráà} \tbPS{n.} \tbGE{garment} \tbOP{bonsuuri} \tbTP{ bonsuuree}
\tbLX{bontaa}\tbHM{} \tbPH{bʊ́ntáà} \tbPS{n.} \tbGE{the act of owning}
\tbLX{bontage}\tbHM{} \tbPH{bʊ́ntàgɪ́} \tbPS{n.} \tbGE{a thick thing} \tbOP{bontagre} \tbTP{ bontagrɛɛ}
\tbLX{bontage}\tbHM{} \tbPH{bʊ́ntágɪ̀} \tbPS{n.} \tbGE{the act of pulling things}
\tbLX{bontale}\tbHM{} \tbPH{bʊ́ntálɪ̀} \tbPS{n.} \tbGE{a lobe of  (kola)} \tbOP{bontala} \tbTP{bontalɛɛ}
\tbLX{bontaŋa}\tbHM{} \tbPH{bòtáŋá} \tbPS{n.} \tbGE{camel} \tbOP{bontanne} \tbTP{bontannɛɛ}
\tbLX{bontara}\tbHM{} \tbPH{bʊ́ntárá} \tbPS{n.} \tbGE{rich man; one who owns many things} \tbOP{bontareba} \tbTP{bontarebɛɛ}
\tbLX{bonteɛ}\tbHM{} \tbPH{bʊ́ntɪ̀ɛ̀} \tbPS{n.} \tbSN \tbGE{the act of shooting} \tbSN \tbGE{the act of supporting things}
\tbLX{bonteɛraa}\tbHM{} \tbPH{bʊ́ntɪ̀ɛ̀ráà} \tbPS{n.} \tbGE{something that supports something else} \tbOP{bonteɛre} \tbTP{bonteɛrɛɛ}
\tbLX{bonteɛraa}\tbHM{} \tbPH{bʊ́ntɪ̀ɛ̀ráà} \tbPS{n.} \tbGE{a boil} \tbOP{bonteɛre} \tbTP{bonteɛrɛɛ}
\tbLX{bonteɛraa}\tbHM{} \tbPH{bʊ́ntɪ̀ɛ̀ráà} \tbPS{n.} \tbGE{something that is shot at} \tbOP{bonteɛre} \tbTP{bonteɛrɛɛ}
\tbLX{bonteɛrɛ}\tbHM{} \tbPH{bʊ́ntɪ́ɛ́rɛ́} \tbPS{n.} \tbGE{shooter} \tbOP{bonteɛrebɛ} \tbTP{bonteɛrebɛɛ}
\tbLX{bontegeroŋ}\tbHM{} \tbPH{bʊ́ntɪ́grʊ́ŋ} \tbPS{n.} \tbGE{a heavy thing} \tbOP{ bontegerɛ} \tbTP{bontegerɛɛ}
\tbLX{bonteraa}\tbHM{} \tbPH{bʊ́ntɪ́ráà} \tbPS{n.} \tbGE{something that is given out or offered} \tbOP{bonterre}
\tbLX{bontere}\tbHM{} \tbPH{bʊ́ntɪ́rɪ́} \tbPS{n.} \tbGE{the act of giving out things}
\tbLX{bontere}\tbHM{} \tbPH{bʊ̀ntɪ́rɪ} \tbPS{n.} \tbGE{a big drum made of animal skin and a gourd} \tbVA{bentere}
\tbLX{bontɛge}\tbHM{} \tbPH{bʊ́tɛ́gɪ̀} \tbPS{n.} \tbGE{the act of exchanging}
\tbLX{bontɛgeraa}\tbHM{} \tbPH{bʊ́ntɛ̀gráà} \tbPS{n.} \tbGE{something that is traded for} \tbOP{bontɛgere} \tbTP{bontɛgerɛɛ}
\tbLX{bontɛgerɛ}\tbHM{} \tbPH{bʊ́ntɛ́grɛ́} \tbPS{n.} \tbGE{one who exchanges} \tbOP{bontɛgerebɛ}
\tbLX{bonto}\tbHM{} \tbPH{bʊ́ntʊ́} \tbPS{n.} \tbGE{act of insulting of things}
\tbLX{bonto}\tbHM{} \tbPH{bòntó} \tbPS{n.} \tbGE{type of cigar made by the Dagaabe} \tbOP{bontori} \tbTP{bontoree}
\tbLX{bontobo}\tbHM{} \tbPH{bʊ́ntòbó} \tbPS{n.} \tbSN \tbGE{a type of plant; medicinal} \tbSN \tbGE{the immediate tender leaves of a plant (bean leaves as vegetable for soup)} \tbSN \tbGE{an animal with prominent ears}
\tbLX{bontoɔlaa}\tbHM{} \tbPH{bʊ́ntʊ́ɔ́láá} \tbPS{n.} \tbGE{that which is sent by a messenger} \tbOP{bontoɔle/bontoɔlɔ} \tbTP{bontoɔlɛɛ}
\tbLX{bontɔraa}\tbHM{} \tbPH{bʊ́ntɔ̀ráà} \tbPS{n.} \tbGE{something that is shared} \tbOP{bontɔrre}
\tbLX{bontɔre}\tbHM{} \tbPH{bʊ́ntɔ̀rɪ́} \tbPS{n.} \tbGE{a worthy thing} \tbOP{bontɔɛ} \tbTP{bontɔrɛɛ}
\tbLX{bontɔre}\tbHM{} \tbPH{bʊ́ntɔ̀rɪ̀} \tbPS{n.} \tbGE{the act of giving out or sharing to as many as possible}
\tbLX{bontuo}\tbHM{} \tbPH{bʊ́ntúó} \tbPS{n.} \tbSN \tbGE{a bitter thing} \tbSN \tbGE{a poisonous or dangerous thing} \tbOP{bontuuri} \tbTP{bonturee}
\tbLX{bonuu}\tbHM{} \tbPH{bʊ̀núú} \tbPS{adj.} \tbGE{five times} \tbOP{bonuuri} \tbTP{bonuuree}
\tbLX{bonvaane}\tbHM{} \tbPH{bʊ́nvàànɪ́} \tbPS{n.} \tbGE{a shrivelled thing} \tbOP{bonvaama} \tbTP{ bonvaa\-nɛɛ}
\tbLX{bonvaraa}\tbHM{} \tbPH{bʊ́nváràà} \tbPS{n.} \tbGE{castrated animal} \tbOP{bonvarre} \tbTP{ bonvarrɛɛ}
\tbLX{bonvelaa}\tbHM{} \tbPH{bʊ́nvɪ̀làá} \tbPS{n.} \tbGE{a good thing} \tbOP{bonveɛle} \tbTP{ bonveɛlɛɛ}
\tbLX{bonveraa}\tbHM{} \tbPH{bʊ́nvɪ̀ráà} \tbPS{n.} \tbSN \tbGE{an implement used in throwing stones (catapult)} \tbSN \tbGE{something that has been shot at by means of a catapult} \tbOP{bonverre}
\tbLX{bonveraa}\tbHM{} \tbPH{bʊ́nvɪ̀ráà} \tbPS{n.} \tbGE{a shift of a bone} \tbOP{bonverre}
\tbLX{bonvere}\tbHM{} \tbPH{bʊ́nvɪ̀rɪ̀} \tbPS{n.} \tbGE{the act of spraining (body part)}
\tbLX{bonvere}\tbHM{} \tbPH{bʊ́nvɪ̀rɪ̀} \tbPS{n.} \tbGE{the act of shooting with a catapult}
\tbLX{bonvooraa}\tbHM{} \tbPH{bʊ́nvʊ́ʊ́ráá} \tbPS{n.} \tbGE{something that breathes} \tbOP{bonvoore} \tbTP{ bonvoorɛɛ}
\tbLX{bonvoɔloŋ}\tbHM{} \tbPH{bʊ́nvʊ̀ɔ̀lʊ̌ŋ} \tbPS{n.} \tbGE{a slimy thing} \tbOP{bonvoɔle} \tbTP{ bonvoɔlɛɛ}
\tbLX{bonvore}\tbHM{} \tbPH{bʊ́nvʊ́rɪ̀} \tbPS{n.} \tbGE{a living thing} \tbOP{bonvoɛ/bonvoba} \tbTP{bonvorɛɛ}
\tbLX{bonvuo}\tbHM{} \tbPH{bʊ́nvùó} \tbPS{n.} \tbGE{an open thing; an uncovered enclosure or container} \tbOP{bonvori} \tbTP{ bonvoree}
\tbLX{bonvuuraa}\tbHM{} \tbPH{bʊ́nvúúráá} \tbPS{n.} \tbSN \tbGE{a thing that crawls (e.g. creature without legs)} \tbSN \tbGE{a thing that is dragged along the ground} \tbOP{bonvuuri} \tbTP{ bonvuuree}
\tbLX{bonweɛ}\tbHM{} \tbPH{bʊ́nwɪ́ɛ́} \tbPS{n.} \tbGE{a common thing} \tbOP{bonweere} \tbTP{bonweerɛɛ}
\tbLX{bon-yaga}\tbHM{} \tbPH{bʊ́ŋ-yágà} \tbPS{n.} \tbGE{many things} \tbTP{bon-yagrɛɛ} \tbVA{boŋyaga}
\tbLX{bon-yeɛle}\tbHM{} \tbPH{bʊ́n-yɪ́ɛ́lɪ́} \tbPS{n.} \tbGE{that which is winnowed or to be winnowed}
\tbLX{bon-yɛɛ}\tbHM{} \tbPH{bʊn-yɛɛ} \tbPS{n.} \tbSN \tbGE{unimportant things} \tbSN \tbGE{wild crops} \tbSG{bon-yɛroo} \tbTP{bon-yɛrɛɛ} \tbVA{boŋyɛɛ}
\tbLX{bon-yɛraa}\tbHM{} \tbPH{bʊ́n-yɛ̀ráà} \tbPS{n.} \tbSN \tbGE{something that is hung around the neck} \tbSN \tbGE{garment} \tbOP{bon-yɛrre} \tbTP{bon-yɛrrɛɛ}
\tbLX{bon-yoɔraa}\tbHM{} \tbPH{bʊ́n-yʊ́ɔ́ráá} \tbPS{n.} \tbSN \tbGE{something that roams or wanders about} \tbSN \tbGE{rheumatism} \tbOP{bon-yoɔre} \tbTP{ bon-yoɔrɛɛ}
\tbLX{bon-yɔmɔ}\tbHM{} \tbPH{bʊ́n-yɔ̀mɔ́} \tbPS{n.} \tbGE{plentiful and free thing(s)} \tbTP{ bon-yɔmɛɛ}
\tbLX{bon-yuo}\tbHM{} \tbPH{bʊ́n-yúó} \tbPS{n.} \tbGE{different one} \tbXV{A bie yage la  a kparoo na bare kyɛ de bonyuo.} \tbXE{The boy took off that shirt and took a different one.} \tbOP{bon-yuobo} \tbVA{boŋyuo}
\tbLX{bon-yuori}\tbHM{} \tbPH{bún-yúórì} \tbPS{n.} \tbGE{a popular or prominent thing} \tbOP{bon-yoe} \tbTP{ bon-yoree}
\tbLX{bonzaa}\tbHM{} \tbPH{bʊ̂nzàá} \tbPS{n.} \tbSN \tbGE{anything, everything} \tbXV{Fo na nyɛ la bonzaa fo naŋ boɔrɔ.} \tbXE{You will get anything that you want.} \tbSN \tbGE{everyone} \tbXV{Bonzaa naŋ wa kyɛ taa la puoraa.} \tbXE{Thanks to everyone who has come here.}
\tbLX{bonzagelaa}\tbHM{} \tbPH{bʊ́nzágláá} \tbPS{n.} \tbGE{an empty thing} \tbOP{bonzagele}
\tbLX{bonzeɛ}\tbHM{} \tbPH{bʊ́nzɪ̀ɛ̀} \tbPS{n.} \tbGE{a red thing} \tbOP{bonzeere} \tbTP{bonzeerɛɛ}
\tbLX{bonzoolɔ}\tbHM{} \tbPH{bʊ́nzʊ́ʊ́lɔ́} \tbPS{n.} \tbGE{a thing with hairy ends (corncob)} \tbTP{bon\-zoo\-lɛɛ} \tbVA{bonzoola}
\tbLX{bonyanlori}\tbHM{} \tbPH{bʊ̀ɲánlórí} \tbPS{n.} \tbGE{very large female goat} \tbTP{bonyanloree}
\tbLX{bonyaŋaa}\tbHM{} \tbPH{bʊ̀nɲáŋáá} \tbPS{n.} \tbGE{female mother goat} \tbOP{bonyaŋene} \tbTP{bonyaŋenɛɛ}
\tbLX{bonyɛ}\tbHM{} \tbPH{bʊ̀ɲɛ́} \tbPS{n.} \tbGE{the act of having wicked ideas} \tbXV{A dɔɔ suuri da iri ne la o pɔge kyɛ o da ba de bonyɛ tuuro ne o.} \tbXE{The man was annoyed with his wife but he did not have any wicked ideas towards her.}
\tbLX{bonyɛ}\tbHM{} \tbPH{bʊ̀ɲɛ́} \tbPS{adj.} \tbGE{having wicked ideas}
\tbLX{boŋ}\tbHM{} \tbPH{bóŋ} \tbPS{pron.} \tbGE{what (interrogative pronoun)} \tbXV{Boŋ ka fo na e zenɛ?} \tbXE{What will you do today?}
\tbLX{boŋemaŋ}\tbHM{} \tbPH{bʊ́ŋɪ̀máŋ} \tbPS{adj.} \tbGE{hazy}
\tbLX{boŋganaa}\tbHM{} \tbPH{bʊŋganaa} \tbPS{n.} \tbGE{bed, something that is slept on} \tbOP{boŋganne} \tbTP{ boŋganɛɛ}
\tbLX{boŋgoraa}\tbHM{} \tbPH{bʊ́ŋgʊ̀ráà} \tbPS{n.} \tbGE{something that is abundant} \tbOP{boŋgorre} \tbTP{boŋgorɛɛ}
\tbLX{boŋgori}\tbHM{} \tbPH{bʊ́ŋgòrí} \tbPS{n.} \tbGE{half-ripe fruit} \tbOP{boŋgoe} \tbTP{ boŋgoree}
\tbLX{boŋgori}\tbHM{} \tbPH{bóŋgórí} \tbPS{n.} \tbGE{shoes  (pl.)} \tbTP{boŋgoree}
\tbLX{boŋgu}\tbHM{} \tbPH{bʊ́ŋgù} \tbPS{n.} \tbGE{the act of watching over something (crops from animals)}
\tbLX{boŋguolaa}\tbHM{} \tbPH{bʊ́ŋgúóláá} \tbPS{n.} \tbSN \tbGE{something that is raised or reared} \tbSN \tbGE{an animal given to someone to be raised} \tbOP{boŋguoli} \tbTP{boŋguolee}
\tbLX{boŋguoli}\tbHM{} \tbPH{bʊ́ŋgúólí} \tbPS{n.} \tbGE{domestic animals and birds (poultry and animal husbandry)} \tbOP{boŋguolo}
\tbLX{boŋguoli}\tbHM{} \tbPH{bʊ̌ŋgúólì} \tbPS{n.} \tbGE{the rearing of animals}
\tbLX{boŋguolo}\tbHM{} \tbPH{bʊ́ŋgúólò} \tbPS{n.} \tbGE{things that are reared (domesticated animals and birds)} \tbVA{doŋguolo}
\tbLX{boŋguraa}\tbHM{} \tbPH{bʊ́ŋgùráà} \tbPS{n.} \tbGE{something jumped upon} \tbOP{boŋgurri}
\tbLX{boŋguuro}\tbHM{} \tbPH{bʊ́ŋgúúro} \tbPS{n.} \tbGE{one who watches over things} \tbOP{boŋguuribo} \tbTP{boŋguuribee}
\tbLX{boŋgbuli}\tbHM{} \tbPH{bʊ́ŋgbùlí} \tbPS{n.} \tbSN \tbGE{a whole thing} \tbSN \tbGE{tumour} \tbOP{boŋgbulo} \tbTP{boŋgbulee}
\tbLX{boŋkaa}\tbHM{} \tbPH{bʊ́ŋkáá} \tbPS{n.} \tbGE{invocation of spirits} \tbOP{boŋkaareba} \tbTP{boŋkaarebɛɛ} \tbVA{boŋkaale}
\tbLX{boŋkaa}\tbHM{} \tbPH{bʊ́ŋkàà} \tbPS{n.} \tbSN \tbGE{the act of taking care of something} \tbSN \tbGE{the act of staring at something}
\tbLX{boŋkãa}\tbHM{} \tbPH{bʊ́ŋkã́ã̀} \tbPS{n.} \tbSN \tbGE{oil or fat from an animal} \tbSN \tbGE{something oily or full of fat} \tbSG{boŋkaanoo} \tbTP{boŋkaanɛɛ}
\tbLX{boŋkaara}\tbHM{} \tbPH{bʊ́ŋkáára} \tbPS{n.} \tbGE{one who watches over something} \tbOP{boŋkaareba} \tbTP{boŋkaarebɛɛ}
\tbLX{boŋkaŋa}\tbHM{} \tbPH{bʊ́ŋkáŋá} \tbPS{n.} \tbGE{something} \tbOP{bommine} \tbVA{boŋkaŋa}
\tbLX{boŋkapala}\tbHM{} \tbPH{bʊ́ŋkpálá} \tbPS{n.} \tbGE{a type of plant} \tbTP{boŋkapalɛɛ}
\tbLX{boŋkaraa}\tbHM{} \tbPH{bʊ́ŋkáráá} \tbPS{n.} \tbGE{a raw, uncooked, or unripe thing} \tbOP{boŋkare} \tbTP{ boŋkarɛɛ}
\tbLX{boŋkoŋi}\tbHM{} \tbPH{bʊ́ŋkǒŋ} \tbPS{n.} \tbGE{a lean thing} \tbOP{boŋkonni} \tbTP{boŋkonnee}
\tbLX{boŋkõɔ}\tbHM{} \tbPH{bʊ́ŋkʊ̃́ɔ̃́} \tbPS{n.} \tbGE{a watery thing} \tbTP{ boŋkoɔnɛɛ}
\tbLX{boŋkoɔraa}\tbHM{} \tbPH{bʊ́ŋkʊ́ɔ́ráá} \tbPS{n.} \tbSN \tbGE{crop; something that is cultivated} \tbSN \tbGE{something used for ploughing or farming} \tbOP{boŋkoɔre} \tbTP{boŋkoɔrɛɛ}
\tbLX{boŋkoɔraa}\tbHM{} \tbPH{bʊ́ŋkʊ́ɔ́ráà} \tbPS{n.} \tbGE{something meant for sale (goods)} \tbOP{boŋkoɔre} \tbTP{boŋkoɔrɛɛ}
\tbLX{boŋkorabinzie}\tbHM{} \tbPH{bʊ́ŋkʊ́rábìnzíé} \tbPS{n.} \tbGE{museum} \tbOP{boŋkorabinziiri}
\tbLX{boŋkore}\tbHM{} \tbPH{bʊ́ŋkʊ́rɪ́} \tbPS{n.} \tbGE{old thing} \tbOP{boŋkorɔ} \tbTP{boŋkorɛɛ}
\tbLX{boŋkuoŋaa}\tbHM{} \tbPH{bʊ́ŋkúóŋáá} \tbPS{n.} \tbGE{a dry thing} \tbOP{boŋkuoni} \tbTP{ boŋkuonee}
\tbLX{boŋkũu}\tbHM{} \tbPH{bʊ́ŋkṹṹ} \tbPS{n.} \tbGE{a dead thing} \tbOP{bokuuni} \tbTP{ boŋkuunee}
\tbLX{boŋkpɔgɔ}\tbHM{} \tbPH{bʊ́ŋkpɔ́gɔ̀} \tbPS{n.} \tbSN \tbGE{lumps} \tbSN \tbGE{a type of disease whose symptoms are swellings and boils} \tbTP{boŋkpɔgerɛɛ} \tbVA{boŋkpɔga/boŋkpaga}
\tbLX{Boŋŋo}\tbHM{} \tbPH{bóŋŋò} \tbPS{n.} \tbGE{a traditional festival of Jirapa  traditional area}
\tbLX{boŋŋmaa}\tbHM{} \tbPH{bʊ́ŋŋmáá} \tbPS{n.} \tbGE{a piece of something} \tbOP{boŋŋmaara} \tbTP{ boŋŋmaarɛɛ}
\tbLX{boŋŋmaa}\tbHM{} \tbPH{bʊ́ŋŋmàà} \tbPS{n.} \tbGE{the practice of cutting things}
\tbLX{boŋŋmaa}\tbHM{} \tbPH{bʊ́ŋŋmàà} \tbPS{n.} \tbGE{the confining of domestic animals that have been ranging freely (usually at the beginning of the farming season)}
\tbLX{boŋŋmare}\tbHM{} \tbPH{bʊ́ŋŋmárɪ́} \tbPS{n.} \tbGE{the act of breaking things} \tbOP{boŋŋmarre} \tbTP{ boŋŋmarɛɛ}
\tbLX{boŋŋmare}\tbHM{} \tbPH{bʊ́ŋŋmàrɪ́} \tbPS{n.} \tbGE{something that is spoilt (food)} \tbTP{boŋŋmarɛɛ}
\tbLX{boŋŋmeɛraa}\tbHM{} \tbPH{bʊ́ŋŋmɪ́ɛ́ráà} \tbPS{n.} \tbGE{musical instrument} \tbOP{boŋŋmeɛre}
\tbLX{boŋo}\tbHM{} \tbPH{bòŋó} \tbPS{n.} \tbGE{donkey} \tbOP{bonni} \tbTP{ bonnee}
\tbLX{boŋɔ}\tbHM{} \tbPH{bʊ̀ŋɔ́} \tbPS{adv.} \tbGE{done anyhow} \tbVA{boŋa}
\tbLX{boo}\tbHM{} \tbPH{bóò} \tbPS{pron.} \tbGE{which one; which of them (interrogative pronoun)} \tbVA{buori}
\tbLX{boo}\tbHM{} \tbPH{bʊ̀ʊ̀} \tbPS{adj.} \tbGE{voluminous e.g. a dress}
\tbLX{bõo}\tbHM{} \tbPH{bʊ̃̀ʊ̃̀} \tbPS{adv.} \tbGE{oily-looking (top of sauce)}
\tbLX{boolimbo}\tbHM{} \tbPH{bòòlímbò} \tbPS{n.} \tbGE{taking one by the shoulders and chest (patient) (typically used with kpɛ)} \tbOP{boolimbori} \tbTP{boolimboree}
\tbLX{boore}\tbHM{} \tbPH{bʊ̀ʊ̀rɪ̀} \tbPS{v.} \tbGE{to attack suddenly and fiercely, to rush on} \tbPD{boorɛɛ, boorɔ, boorɔ, boorebɔ, booraa, boorebɔ }
\tbLX{boore}\tbHM{} \tbPH{bʊ́ʊ́rɪ́} \tbPS{n.} \tbGE{of same stalk, species, descendant} \tbOP{boora} \tbTP{boorɛɛ} \tbVA{booree}
\tbLX{boore}\tbHM{} \tbPH{bʊ̀ʊ̀rɪ́} \tbPS{n.} \tbGE{agreement, plan} \tbTP{boorɛɛ}
\tbLX{boorɔ}\tbHM{} \tbPH{bʊ́ʊ́rɔ́} \tbPS{n.} \tbSN \tbGE{off same type, family} \tbSN \tbGE{seed for planting} \tbSG{booroo} \tbVA{boora}
\tbLX{boɔ}\tbHM{} \tbPH{bʊ́ɔ́} \tbPS{n.} \tbGE{goat} \tbOP{boore} \tbTP{boorɛɛ}
\tbLX{bõɔ}\tbHM{} \tbPH{bʊ̃̀ɔ̃̀} \tbPS{v.} \tbGE{to scoop up a large amount of something} \tbXV{ O bõɔ la a kãa poli o eŋɛ zaa.} \tbXE{She scooped up a large amount of lotion and rubbed it all over her body.} \tbPD{bõɔɛ, boɔnɔ, boɔnɔ, boɔnema, boɔnaa}%}%, , -}
\tbLX{bõɔ}\tbHM{} \tbPH{bʊ̃̀ɔ̃̀} \tbPS{v.} \tbGE{snub} \tbXV{A dɔɔ suuri iree la ka o maŋ deɛ bõɔ neɛ te lɔɔ.} \tbXE{The man is so annoyed that he snubs anybody he comes across.} \tbPD{bõɔɛ, boɔnɔ, boɔnɔ, boɔnema, boɔnaa}%}%, , -}
\tbLX{boɔbo}\tbHM{} \tbPH{bʊ́ɔ́bʊ̀} \tbPS{n.} \tbSN \tbGE{something that is preferred} \tbSN \tbGE{the act of liking or searching for something}
\tbLX{boɔlaa}\tbHM{} \tbPH{bʊ̀ɔ̀làá} \tbPS{n.} \tbSN \tbGE{a call} \tbSN \tbGE{an invitation}
\tbLX{boɔle}\tbHM{} \tbPH{bʊ̀ɔ̀lɪ̀} \tbPS{v.} \tbSN \tbGE{to call} \tbSN \tbGE{to invite} \tbPD{boɔlɛɛ, boɔlɔ, boɔlɔ, boɔleba, boɔlaa, boɔleba }
\tbLX{boɔraa}\tbHM{} \tbPH{bʊ̀ɔ̀ráà} \tbPS{n.} \tbSN \tbGE{sack} \tbSN \tbGE{one hundred pounds} \tbOP{boɔre} \tbTP{boɔrɛɛ}
\tbLX{boɔre}\tbHM{} \tbPH{bʊ̀ɔ̀rɪ́} \tbPS{v.} \tbGE{to smear with thick paste} \tbPD{boɔrɛɛ, boɔrɔ, boɔrɔ, boɔrebɔ, boɔraa, boɔrebɔ }
\tbLX{boɔrɔ}\tbHM{} \tbPH{bʊ́ɔ́rɔ̀} \tbPS{v.} \tbSN \tbGE{to like, to love; to desire, to want} \tbSN \tbGE{to need} \tbPD{bɔɛ, boɔrɔ, boɔrɔ, boɔreba, boɔraa}%}%, , -}
\tbLX{boɔso}\tbHM{} \tbPH{bʊ́ɔ́sʊ́} \tbPS{n.} \tbGE{bus} \tbOP{boɔsere} \tbTP{ boɔserɛɛ}
\tbLX{bopeɛ}\tbHM{} \tbPH{bʊ̀pɪ́ɛ́} \tbPS{n.} \tbGE{traditional storey building} \tbOP{bopeere} \tbTP{bopeerɛɛ} \tbVA{peɛzudie}
\tbLX{boraa}\tbHM{} \tbPH{bʊ́ráá} \tbPS{n.} \tbGE{a stick used in sowing grains} \tbOP{borre} \tbTP{borrɛɛ} \tbVA{boredaa}
\tbLX{bore}\tbHM{} \tbPH{bʊ̀rɪ̀} \tbPS{v.} \tbGE{to break part of a lump off in big chunks} \tbPD{borɛɛ, bora, bora, boreba, boraa}%}%, , -}
\tbLX{bore}\tbHM{} \tbPH{bʊ̀rɪ̀} \tbPS{v.} \tbGE{to sow, to plant} \tbPD{borɛɛ, bora, bora, boreba, boraa, boreba }
\tbLX{boredaa}\tbHM{} \tbPH{bʊ̀rɪ̀dàá} \tbPS{n.} \tbGE{a stick for sowing} \tbOP{boredaare} \tbTP{ boredarɛɛ/boredaarɛɛ}
\tbLX{borenyɛ}\tbHM{} \tbPH{bʊ́rɪ́ɲɛ̀} \tbPS{n.} \tbGE{Christmas} \tbOP{borenyɛre} \tbTP{ borenyɛrɛɛ}
\tbLX{boreŋmaakpantoli}\tbHM{} \tbPH{bʊ̀rɪ̀ŋmàáakpántólí} \tbPS{n.} \tbGE{a type of large, beautiful bird with a sweet voice} \tbOP{boreŋmaakpantolo} \tbTP{boreŋmaakpantolee}
\tbLX{boreŋmaao}\tbHM{} \tbPH{bʊ̀rɪ̀ŋmààʊ́} \tbPS{n.} \tbGE{a large bird with a strong sweet voice} \tbOP{boreŋmaane} \tbTP{ boreŋmaanɛɛ}
\tbLX{boreŋmanlee}\tbHM{} \tbPH{bʊ̀rɪ̀ŋmánléé} \tbPS{n.} \tbGE{calabash for sowing} \tbOP{boreŋmambili} \tbTP{boreŋmambilee}
\tbLX{boribara}\tbHM{} \tbPH{bòrìbàrà} \tbPS{adv.} \tbGE{done poorly}
\tbLX{boribori}\tbHM{} \tbPH{bòrìbórí} \tbPS{v.} \tbGE{to carry out poorly (cleaning, talk)} \tbPD{boriboree, boriboro, boriboro, boriboribo, boriboraa}%}%, , -}
\tbLX{boroboro}\tbHM{} \tbPH{bòròbòrò} \tbPS{n.} \tbGE{bread} \tbOP{boroborri} \tbTP{boroboree}
\tbLX{boroborozɔŋ}\tbHM{} \tbPH{bòròbòròzɔ̂ŋ} \tbPS{n.} \tbGE{bread flour} \tbTP{boroborozɔnnɛɛ}
\tbLX{boroto}\tbHM{} \tbPH{bòròtò} \tbPS{adj.} \tbGE{ big and round}
\tbLX{boroto}\tbHM{} \tbPH{bʊ̀rʊ̀tʊ̀} \tbPS{adv.} \tbGE{fall flat} \tbXV{A bie zoro gɛrɛ te ŋmɛ gbɛre a gaa te boroto o naŋ le.} \tbXE{As the boy ran along he stumbled on a stone and crashed to the ground.}
\tbLX{borɔbɛ}\tbHM{} \tbPH{bʊ̀rɔ̀bɛ́} \tbPS{n.} \tbGE{pineapple fruit or plant} \tbOP{borɔbɛre} \tbTP{borɔbɛrɛɛ}
\tbLX{borɔborɔ}\tbHM{} \tbPH{bʊ̀rɔ̀bʊ̀rɔ̀} \tbPS{adv.} \tbGE{dull colour}
\tbLX{borɔso}\tbHM{} \tbPH{bʊ̀rɔ́sʊ̀} \tbPS{n.} \tbGE{ brush} \tbOP{borɔsore} \tbTP{ borɔsorɛɛ}
\tbLX{bosaraa}\tbHM{} \tbPH{bʊ̀sáràà} \tbPS{n.} \tbGE{young female goat} \tbOP{bosarre} \tbTP{bosarrɛɛ}
\tbLX{bosere}\tbHM{} \tbPH{bʊ̀sɪ́rɪ̀} \tbPS{n.} \tbGE{a necklace made of huge beads} \tbOP{boseɛ/boserre} \tbTP{boserɛɛ}
\tbLX{bovammo}\tbHM{} \tbPH{bʊ̀vàmmʊ́} \tbPS{n.} \tbGE{a type of bow, usually not made from bamboo} \tbOP{bovama} \tbTP{ bovannɛɛ}
\tbLX{bovaraa}\tbHM{} \tbPH{bʊ̀váràá} \tbPS{n.} \tbGE{castrated goat} \tbOP{bovarre} \tbTP{ bovarrɛɛ}
\tbLX{bowola}\tbHM{} \tbPH{bʊ́wʊ́lá} \tbPS{adv.} \tbGE{how many times}
\tbLX{boyaa}\tbHM{} \tbPH{bóyáá} \tbPS{interj.} \tbGE{expression of surprise} \tbVA{boyaaŋ}
\tbLX{boyaaŋ}\tbHM{} \tbPH{bòyááŋ} \tbPS{interj.} \tbGE{expression of surprise}
\tbLX{boyoɔbo}\tbHM{} \tbPH{bʊ̀yʊ̀ɔ́bʊ́} \tbPS{adj.} \tbGE{six times}
\tbLX{boyɔpõi}\tbHM{} \tbPH{bʊ̀yɔ̀pṍĩ́} \tbPS{adj.} \tbGE{seven times}
\tbLX{bozeɛ}\tbHM{} \tbPH{bʊ̀zɪ́ɛ́} \tbPS{n.} \tbGE{dog} \tbOP{bozeere} \tbTP{bozeerɛɛ}
\tbLX{bɔ}\tbHM{} \tbPH{bɔ́} \tbPS{v.} \tbSN \tbGE{to search; to look for} \tbXV{N bie bɔrɛɛ la, yɛ soŋe, ma ka N bɔ.} \tbXE{My child is lost; help me look for her.} \tbSN \tbGE{to court, to woo} \tbXV{Dagaaba maŋ bɔ la pɔge kyɛ de.} \tbXE{The Dagaaba usually court a woman before marriage.} \tbSN \tbGE{to engage in sexual intercourse, to make love} \tbXV{Ka fooŋ bɔ neɛ pɔge ba na ŋmɛ fo la sane yaga.} \tbXE{If you make love to a married woman, you will be heavily fined.} \tbPD{bɔɛ, boɔrɔ, boɔrɔ, boɔrebɔ, boɔraa, boɔrebɔ }
\tbLX{bɔ}\tbHM{} \tbPH{bɔ̀} \tbPS{v.} \tbGE{to clean a very dirty place or thing} \tbPD{bɔɛ, boɔrɔ, boɔrɔ, boɔrebɔ, boɔraa, boɔrebɔ }
\tbLX{bɔbɔe}\tbHM{} \tbPH{bɔ́bɔ́ɪ̀} \tbPS{n.} \tbGE{duck} \tbOP{bɔbɔɛ} \tbTP{bɔbɔrɛɛ}
\tbLX{bɔbɔlaa}\tbHM{} \tbPH{bɔ́bɔ́láá} \tbPS{n.} \tbGE{a wild animal like a dog} \tbOP{bɔbɔlle} \tbTP{bɔbɔlɛɛ}
\tbLX{bɔe}\tbHM{} \tbPH{bɔ̀ɪ́} \tbPS{n.} \tbGE{servant} \tbOP{bɔemine}
\tbLX{bɔ̃ee}\tbHM{} \tbPH{bɔ̃̀ɪ̃̀ɪ̃̀} \tbPS{adv.} \tbGE{calmly, slowly} \tbXV{O yeli la ayɛlɛ bɔ̃ee lɛ.} \tbXE{She spoke gently.}
\tbLX{bɔge}\tbHM{} \tbPH{bɔ́gɪ́} \tbPS{v.} \tbGE{to grind into smooth flour} \tbPD{bɔgɛɛ, bɔgerɔ, bɔgerɔ, bɔgerebɔ, bɔgeraa, bɔgerebɔ } \tbVA{bage}
\tbLX{bɔge}\tbHM{} \tbPH{bɔ́gɪ́} \tbPS{v.} \tbSN \tbGE{to hit with many arrows; to sting or bite many times (insects)} \tbSN \tbGE{to apply poisonous lotion to arrows} \tbPD{bɔgɛɛ, bɔgerɔ, bɔgerɔ, bɔgerebɔ, bɔgeraa, bɔgerebɔ } \tbVA{bage}
\tbLX{bɔge}\tbHM{} \tbPH{bɔ́gɪ́} \tbPS{v.} \tbGE{to put a fence around something} \tbPD{bɔgɛɛ, bɔgerɔ, bɔgerɔ, bɔgerebɔ, bɔgeraa, bɔgerebɔ } \tbVA{bage}
\tbLX{bɔge}\tbHM{} \tbPH{bɔ́gɪ́} \tbPS{n.} \tbSN \tbGE{shoulder} \tbSN \tbGE{forelimb of an animal} \tbOP{bɔgere} \tbTP{bɔgerɛɛ} \tbVA{bɔga/baga}
\tbLX{bɔgebie}\tbHM{} \tbPH{bɔ̀gìbíé} \tbPS{n.} \tbGE{misfortune} \tbOP{bɔgebiiri} \tbTP{bɔgebiiree} \tbVA{bɔɔbie/bagebie}
\tbLX{bɔgebiesoba}\tbHM{} \tbPH{bɔ̀gɪ̀bíésʊ́bá} \tbPS{n.} \tbGE{unfortunate person; unlucky person} \tbOP{ bɔgebiedeme} \tbVA{bagebiesoba}
\tbLX{bɔgebogerɔ}\tbHM{} \tbPH{bɔ̀gɪ̀bʊ́grɔ́} \tbPS{n.} \tbGE{sorcerer, soothsayer} \tbOP{bɔgebogerebɔ} \tbTP{bɔgebogerebɛɛ} \tbVA{bagebogera}
\tbLX{bɔgedie}\tbHM{} \tbPH{bɔ̀gɪ̀dìè} \tbPS{n.} \tbGE{room for divining} \tbOP{bɔgederi} \tbTP{bɔgederee}
\tbLX{bɔgele}\tbHM{} \tbPH{bɔ̀glɪ̀} \tbPS{v.} \tbSN \tbGE{to stalk (hunting)} \tbXV{Naŋkpaana zaa maŋ bɔgele la wɛdoŋɔ a pɔge o ŋmɛ.} \tbXE{A hunter will normally stalk a wild animal to get close enough to shoot.} \tbSN \tbGE{to feel, to move one's hand over} \tbXV{A zɔŋɔ bɔgele la a gado zu te nyɛ a o kparoo.} \tbXE{The blind man felt along the bed and found his shirt.} \tbPD{bɔgelɛɛ, bɔgelɔ, bɔgelɔ, bɔgeleba, bɔgelaa}%}%, , -}
\tbLX{bɔgemaale}\tbHM{} \tbPH{bɔ̀gɪ̀mààlɪ̀} \tbPS{n.} \tbGE{the act of making sacrifice} \tbOP{bɔgemaaleba} \tbTP{bɔgemaalbɛɛ} \tbVA{bagemaale}
\tbLX{bɔgeŋmampegere}\tbHM{} \tbPH{bɔ̀gɪ̀ŋmǎmpɪ́gɪ́rɪ́} \tbPS{n.} \tbGE{shoulder} \tbVA{bageŋmampere}
\tbLX{bɔgeŋmampere}\tbHM{} \tbPH{bɔ̀gɪ̀ŋmàmpɪ́rɪ́} \tbPS{n.} \tbGE{shoulder blade} \tbOP{bɔgeŋmamperre} \tbTP{bɔgeŋmamperɛɛ} \tbVA{bɔgeŋmampegere}
\tbLX{bɔgeŋmane}\tbHM{} \tbPH{bɔ̀gɪ̀ŋmánɪ́} \tbPS{n.} \tbGE{forearm} \tbOP{bɔgeŋmama} \tbTP{bɔgeŋmamɛɛ} \tbVA{bageŋmane}
\tbLX{bɔgere}\tbHM{} \tbPH{bɔ̌grɪ́} \tbPS{n.} \tbGE{part, portion of a whole} \tbTP{bɔgerɛɛ} \tbVA{bɔgre}
\tbLX{bɔgere}\tbHM{} \tbPH{bɔ̀gɪ̀rɪ́} \tbPS{n.} \tbSN \tbGE{fetish, object believed to embody a spirit} \tbSN \tbGE{sacrifice, offering to a divinity} \tbOP{bɔgɔ/bɔga} \tbTP{bɔgerɛɛ/bɔgɛɛ} \tbVA{bagere}
\tbLX{bɔgesaa}\tbHM{} \tbPH{bɔ̀gsàà} \tbPS{n.} \tbGE{counsellor for a person being initiated into the bagre cult} \tbOP{bɔgesaamine} \tbTP{bɔgesaarɛɛ} \tbVA{bagesaa}
\tbLX{bɔgesoŋ}\tbHM{} \tbPH{bɔ̀gɪ̀sʊ̂ŋ} \tbPS{n.} \tbGE{luck} \tbOP{bɔgesonne} \tbTP{bɔgesonnɛɛ} \tbVA{bagesoŋ}
\tbLX{bɔgetaŋkpolo}\tbHM{} \tbPH{bɔ̀gɪ̀táŋkpóló} \tbPS{n.} \tbGE{cult members, those who have gone through the initiation process for a cult} \tbOP{bɔgetaŋkpolli} \tbTP{bɔgetaŋkpolee} \tbVA{bagetaŋkpolo}
\tbLX{bɔgɔ}\tbHM{} \tbPH{bɔ́gɔ̀} \tbPS{n.} \tbGE{track, path (usually left by passing through grassy bush)} \tbOP{bɔgere} \tbTP{bɔgerɛɛ} \tbVA{bɔga}
\tbLX{bɔgɔ}\tbHM{} \tbPH{bɔ́gɔ́} \tbPS{n.} \tbGE{shoulder} \tbOP{bɔgere} \tbTP{bɔgerɛɛ} \tbVA{bɔge/bɔga/bage}
\tbLX{bɔlaa}\tbHM{} \tbPH{bɔ̀làà} \tbPS{adv.} \tbGE{ long, round and smooth}
\tbLX{bɔle}\tbHM{} \tbPH{bɔ́lɪ́} \tbPS{n.} \tbGE{ball} \tbOP{bɔlɔ} \tbTP{bɔlɛɛ/bɔllɛɛ}
\tbLX{bɔle}\tbHM{} \tbPH{bɔ̀lɪ̀} \tbPS{n.} \tbGE{field rat} \tbSG{bɔloo} \tbOP{bɔlɔ} \tbTP{bɔllɛɛ}
\tbLX{bɔlembɔ}\tbHM{} \tbPH{bɔ̀lɪ̀mbɔ́} \tbPS{n.} \tbGE{bottle} \tbSG{bɔlembɔroo} \tbOP{ bɔlembɔre} \tbTP{bɔlembɔrɛɛ} \tbVA{pilintoɔ}
\tbLX{bɔll}\tbHM{} \tbPH{bɔ̀ll} \tbPS{adj.} \tbGE{long and roundish}
\tbLX{bɔlle}\tbHM{} \tbPH{bɔ̀llɪ̀} \tbPS{v.} \tbSN \tbGE{to move all over an object (insects)} \tbSN \tbGE{to smear liquid on a surface in an untidy way} \tbPD{bɔllɛɛ, bɔllɔ, bɔllɔ, bɔlbɔ, bɔllaa}%}%, , -}
\tbLX{bɔlɔmbɔlɔŋ}\tbHM{} \tbPH{bɔ́lɔ́mbɔ́lɔ́ŋ} \tbPS{adj.} \tbGE{smooth and shiny}
\tbLX{bɔmbɔrezeɛ}\tbHM{} \tbPH{bɔ̀mbɔ̀rzɪ̀ɛ́} \tbPS{n.} \tbGE{a type of snake} \tbOP{bɔmbɔrezeere} \tbTP{bɔmbɔrezeerɛɛ}
\tbLX{bɔmbɔroŋ}\tbHM{} \tbPH{bɔ̌mbɔ̀rʊ́ŋ} \tbPS{n.} \tbGE{a type of snake} \tbOP{bɔmbɔrre} \tbTP{bɔmbɔrɛɛ}
\tbLX{bɔmpɔrɔ}\tbHM{} \tbPH{bɔ̀mpɔ́rɔ́} \tbPS{n.} \tbGE{mumps} \tbOP{bɔmpɔrre} \tbTP{ bɔmpɔrɛɛ}
\tbLX{bɔndaa}\tbHM{} \tbPH{bɔ̀ndáá} \tbPS{n.} \tbGE{lizard (variation of ‘bandaa’(individual) and ‘baŋa’(dialectal))} \tbOP{bɔndaare} \tbTP{bɔndarɛɛ} \tbVA{bandaa}
\tbLX{bɔŋenaa}\tbHM{} \tbPH{bɔ́ŋɪ́náá} \tbPS{n.} \tbGE{type of tree, medicinal} \tbOP{bɔŋene} \tbTP{bɔŋenɛɛ}
\tbLX{bɔŋenɔ}\tbHM{} \tbPH{bɔ̀ŋɪ́nɔ́} \tbPS{n.} \tbGE{witty person} \tbOP{bɔŋenema} \tbTP{bɔŋenemɛɛ}
\tbLX{bɔŋɔ}\tbHM{} \tbPH{bɔ̀ŋɔ́} \tbPS{n.} \tbGE{lizard} \tbOP{bɔnne} \tbTP{bɔnnɛɛ} \tbVA{baŋa/bɔŋa}
\tbLX{bɔŋɔ}\tbHM{} \tbPH{bɔ́ŋɔ́} \tbPS{n.} \tbGE{bangle} \tbOP{bɔnne} \tbTP{bɔnnɛɛ} \tbVA{baŋa}
\tbLX{bɔɔ}\tbHM{} \tbPH{bɔ̀ɔ̀} \tbPS{v.} \tbGE{to dress very well} \tbPD{bɔɔɛ, bɔɔrɔ, bɔɔrɔ, bɔɔrebɔ, bɔɔraa, bɔɔrebɔ }
\tbLX{bɔɔderoo}\tbHM{} \tbPH{bɔ̀ɔ̀dɪ́rʊ́ʊ́} \tbPS{n.} \tbGE{plantain} \tbSG{bɔɔderoo} \tbOP{bɔɔdeɛ} \tbTP{bɔɔdeerɛɛ}
\tbLX{bɔɔdoa}\tbHM{} \tbPH{bɔ̀ɔ̀dʊ́áá} \tbPS{n.} \tbGE{towel} \tbOP{bɔɔdoore} \tbTP{bɔɔdoorɛɛ}
\tbLX{bɔɔre}\tbHM{} \tbPH{bɔ̀ɔ̀rɪ́} \tbPS{n.} \tbGE{sacrifice, deity} \tbOP{bɔbɔ} \tbVA{bɔgre}
\tbLX{bɔre}\tbHM{} \tbPH{bɔ́rɪ̀} \tbPS{v.} \tbGE{to break off a large chunk (from a bigger mass)} \tbPD{bɔrɛɛ, bɔrɔ, bɔrɔ, bɔreba, bɔraa, bɔreba }
\tbLX{bɔre}\tbHM{} \tbPH{bɔ̀rɪ̀} \tbPS{v.} \tbGE{to blossom, to flourish (plants)} \tbPD{bɔrɛɛ, bɔrɔ, bɔrɔ, bɔrebɔ, bɔraa}%}%, , -}
\tbLX{bɔre}\tbHM{} \tbPH{bɔ̀rɪ̀} \tbPS{v.} \tbGE{to get lost, to go astray; to make a mistake} \tbPD{bɔrɛɛ, bɔrɔ, bɔrɔ, bɔrebɔ, bɔraa}%}%, , -}
\tbLX{bɔre}\tbHM{} \tbPH{bɔ́rɪ́} \tbPS{n.} \tbGE{puff adder; a thick large poisonous snake} \tbOP{bɔrre} \tbTP{bɔrɛɛ} \tbVA{bɔsere}
\tbLX{bɔreferoo}\tbHM{} \tbPH{bɔ̀rɪ́fɪ́rʊ́ʊ́} \tbPS{n.} \tbGE{pawpaw} \tbOP{bɔreferre} \tbTP{bɔreferrɛɛ}
\tbLX{Bɔrefɔ}\tbHM{} \tbPH{bɔ̀rɪ̀fɔ́} \tbPS{n.} \tbGE{English} \tbOP{bɔrefɔre} \tbTP{ bɔrefɔrɛɛ} \tbVA{bɔrɔ/naasaaleŋ}
\tbLX{bu}\tbHM{} \tbPH{bù} \tbPS{v.} \tbSN \tbGE{to get wet} \tbXV{Saa poɔ la a tontonema ka ba bonsuuri zaa bu.} \tbXE{The rain beat down on the workers and soaked all their clothes.} \tbSN \tbGE{to mix dry powder with liquid or oil (mortar)} \tbXV{A demeɛrɛ bu la tɛne na mɛ a dankyini.} \tbXE{The mason has prepared mortar to build the wall with.} \tbPD{bue, buu\-ro, buu\-ro, buuribo, buuraa, buuribo }
\tbLX{bu}\tbHM{} \tbPH{bú} \tbPS{v.} \tbGE{to measure (liquid), to calculate} \tbXV{Ba bu la kãã bɔlembɔre ata eŋ a doge poɔ.} \tbXE{The have measured out three bottles of oil into the pot.} \tbPD{bue, buuro, buuro, buuribo, buuraa, buuribo }
\tbLX{bu}\tbHM{} \tbPH{bú} \tbPS{v.} \tbSN \tbGE{to ponder, to reflect} \tbXV{N bu la a yɛlɛ n zu poɔ lɛɛ kyɛ N naŋ ba baŋ a pare.} \tbXE{I've pondered over the issue for some time now, but I can't seem to come to terms with it.} \tbSN \tbGE{to respect , to value} \tbXV{Ka fooŋ ba bu ma N meŋ koŋ bu fo.} \tbXE{If you don't respect me, I wont respect you either.} \tbPD{bue, buu\-ro, buu\-ro, buuribo, buuraa, buuribo }
\tbLX{bufuli}\tbHM{} \tbPH{bùfúlí} \tbPS{n.} \tbGE{a blood-sucking maggot-like worm, usually found under floors} \tbOP{bufulo} \tbTP{bufulee}
\tbLX{bugagu}\tbHM{} \tbPH{bùgàgú} \tbPS{n.} \tbGE{a type of tree, medicinal} \tbOP{bugaguri} \tbTP{bugaguree}
\tbLX{bugi}\tbHM{} \tbPH{bùgì} \tbPS{v.} \tbSN \tbGE{to be confused, to be perplexed} \tbSN \tbGE{to be drunk} \tbXV{O yãã nyu la dãã bugi.} \tbXE{As usual, he has been drinking and is drunk.} \tbPD{bugee, bugiro, bugiro, bugiribo, bugiraa}%}%, , -}
\tbLX{bugi}\tbHM{} \tbPH{búgí} \tbPS{v.} \tbGE{to pound into powder} \tbXV{Ba bugi a kyooma na tɔ kãã.} \tbXE{The shea nuts have been pounded into powder in order to prepare shea butter.} \tbPD{bugee, bugro, bugro, bugribo, bugraa}%}%, , -}
\tbLX{bugi}\tbHM{} \tbPH{búgí} \tbPS{v.} \tbGE{to make fresh mounds (farming)} \tbXV{A koɔraa bugi la o weɛ yuoni ŋa azuiŋ o bɛŋɛ wɔŋɛɛ la soŋ.} \tbXE{The farmer made fresh mounds; consequently, he had a good yield of beans.} \tbPD{bugee, bugro, bugro, bugribo, bugraa}%}%, , -}
\tbLX{bugikpara}\tbHM{} \tbPH{bùgìkpárá} \tbPS{n.} \tbGE{a moulded granary with two apartments} \tbOP{bugkparre} \tbTP{bugkparɛɛ}
\tbLX{bugili}\tbHM{} \tbPH{búgílì} \tbPS{v.} \tbGE{to tend (seed)} \tbPD{bugilee, bugilo, bugilo, buglibo, bugilaa, - } \tbVA{buuli/buoli}
\tbLX{bugiliŋe}\tbHM{} \tbPH{bùglíŋé} \tbPS{n.} \tbGE{lid for a barn or granary} \tbOP{bugilinni} \tbTP{bugilinnee}
\tbLX{bugimoɔ}\tbHM{} \tbPH{bùgìmʊ́ɔ́} \tbPS{n.} \tbGE{a type of grass, the straw of which is mixed with clay to build a granary} \tbSG{bugimoɔnoo} \tbTP{bugimoɔnɛɛ}
\tbLX{bugipelaa}\tbHM{} \tbPH{bùgìpɪ̀làá} \tbPS{n.} \tbGE{a small granary} \tbOP{bugipeɛle} \tbTP{ bugipeɛlɛɛ}
\tbLX{bugita}\tbHM{} \tbPH{búgítà} \tbPS{n.} \tbGE{kettle} \tbOP{bugitare} \tbTP{ bugitarɛɛ}
\tbLX{bugo}\tbHM{} \tbPH{bùgó} \tbPS{n.} \tbGE{granary} \tbOP{bugri} \tbTP{bugree}
\tbLX{bukulo} \tbPH{búkúló} \tbPS{adv.} \tbGE{perish completely} \tbXV{ A vũũ di la  yiri zaa bukulo.} \tbXE{Fire has burnt the house completely.}
\tbLX{bulee}\tbHM{} \tbPH{búléé} \tbPS{n.} \tbGE{well} \tbOP{bulli} \tbTP{ bullee}
\tbLX{buli}\tbHM{} \tbPH{búlí} \tbPS{v.} \tbSN \tbGE{to germinate, to sprout, to grow out from} \tbXV{Ba bore la kyi kyɛ a naŋ ba buli sɛre.} \tbXE{They have sown millet, but it has not germinated yet.} \tbSN \tbGE{to spring up, to ooze (water)} \tbXV{Te tu la a bulee ka o sigi kyɛ o naŋ ba buli kõɔ.} \tbXE{We have dug the well quite deep, but water has not started to flow.} \tbPD{bulee, bulo, bulo, bulibo, bulaa, bulibo }
\tbLX{buli}\tbHM{} \tbPH{búlí} \tbPS{n.} \tbGE{sprout} \tbOP{bulo}
\tbLX{buliba}\tbHM{} \tbPH{búlíbà} \tbPS{v.} \tbGE{to run fast} \tbPD{bulibaɛ, bulobaara, bulibaara, bulibaareba, bulibaaraa}%}%, , -}
\tbLX{bulinimiri}\tbHM{} \tbPH{bùlnímírì} \tbPS{n.} \tbGE{the point in a well from which water flows out} \tbOP{bulinimie} \tbTP{bulinimiiree}
\tbLX{bulli}\tbHM{} \tbPH{bùllì} \tbPS{v.} \tbGE{to collect bits of water flowing from a well or hole (e.g. during draught)} \tbPD{bullee, bullo, bullo, bulibo, bullaa, bulibo }
\tbLX{bulo}\tbHM{} \tbPH{bùlò} \tbPS{n.} \tbGE{soup from seed or nuts (beans, groundnuts)} \tbTP{bulee/bullee}
\tbLX{bulugiri}\tbHM{} \tbPH{bùlùgrí} \tbPS{n.} \tbGE{a container made from a gourd} \tbOP{bulugo} \tbTP{bulugiree} \tbVA{buluuri}
\tbLX{buluuri}\tbHM{} \tbPH{bùlùùrí} \tbPS{n.} \tbGE{a fat goat} \tbTP{buluuree.} \tbVA{bulugiri}
\tbLX{bumbunturi}\tbHM{} \tbPH{bùmbùntùrì} \tbPS{n.} \tbGE{a type of tree with edible fruits} \tbOP{bumbunturri} \tbTP{bumbunturee}
\tbLX{buŋi}\tbHM{} \tbPH{búŋ} \tbPS{v.} \tbGE{to get mouldy (seed)} \tbPD{buŋee, buŋino, buŋino, buŋinimo, buŋinaa, buŋinimo }
\tbLX{buo}\tbHM{} \tbPH{búò} \tbPS{pron.} \tbGE{which one (interrogative pronoun)} \tbVA{buori}
\tbLX{buoli}\tbHM{} \tbPH{búólì} \tbPS{v.} \tbGE{to nurse (a seedling)} \tbPD{buolee, buolo, buolo, buolibo, buolaa, buolibo } \tbVA{buuli/bugili}
\tbLX{buori}\tbHM{} \tbPH{bùòrì} \tbPS{v.} \tbGE{to hurt} \tbPD{buoree, buoro, buoro, buoribo, buoraa, buoribo }
\tbLX{buori}\tbHM{} \tbPH{bùórì} \tbPS{pron.} \tbGE{which (interrogative pronoun)} \tbVA{buo/booː abuo/abobo}
\tbLX{buri}\tbHM{} \tbPH{bùrì} \tbPS{v.} \tbGE{to soften by moistening} \tbXV{O buri la a gane a na tõɔ̃sɛ a woɔ.} \tbXE{He softened the leather so he could sew the bag.} \tbPD{buree, buro, buro, buribo, buraa, buribo }
\tbLX{buri}\tbHM{} \tbPH{bùrì} \tbPS{v.} \tbGE{to gain weight} \tbXV{Beri ayi ŋa a dɔɔ ŋmeene na leɛ buri la o zaa.} \tbXE{Lately, that lean man has been putting on weight.} \tbPD{buree, buro, buro, buribo, buraa, buribo }
\tbLX{buri}\tbHM{} \tbPH{búrí} \tbPS{v.} \tbGE{to immerse a container in liquid or powder to fill it} \tbPD{buree, buro, buro, buribo, buraa, buro }
\tbLX{buriburi}\tbHM{} \tbPH{bùrìbúrí} \tbPS{v.} \tbGE{to be confused} \tbPD{buriburee, buriburo, buriburo, buriburibo, buriburaa, buriburibo }
\tbLX{buriburi}\tbHM{} \tbPH{búríbúrí} \tbPS{n.} \tbGE{state of confusion}
\tbLX{burimo}\tbHM{} \tbPH{bùrìmó} \tbPS{n.} \tbGE{a type of wild animal the size of a zebra} \tbOP{burri} \tbTP{burree}
\tbLX{buro}\tbHM{} \tbPH{bùró} \tbPS{adj.} \tbGE{in dry powdered form (flour)} \tbXV{O nɔŋ la zɔmburo mogebo.} \tbXE{He likes eating unsoaked flour.} \tbOP{burri} \tbTP{burree}
\tbLX{buroni}\tbHM{} \tbPH{bùròní} \tbPS{n.} \tbGE{white man} \tbOP{buroniri} \tbTP{buroniree}
\tbLX{buroŋ-buroŋ}\tbHM{} \tbPH{búróŋ-búróŋ} \tbPS{adv.} \tbGE{done haphazardly (without order)} \tbXV{A yɛlɛ deɛ waa la buromburoŋ lɛ.} \tbXE{The case is just in a confused state.} \tbVA{buromburoŋ}
\tbLX{burri}\tbHM{} \tbPH{búrrì} \tbPS{v.} \tbGE{to fool, to outwit} \tbPD{burree, burro, burro, buribo, burraa, burro }
\tbLX{buseli}\tbHM{} \tbPH{búsélí} \tbPS{n.} \tbGE{very heavy object} \tbOP{busele} \tbTP{ buselee}
\tbLX{buulaa}\tbHM{} \tbPH{bùùláá} \tbPS{n.} \tbGE{playground} \tbOP{buuli} \tbTP{buulee}
\tbLX{buuli}\tbHM{} \tbPH{bùùlí} \tbPS{n.} \tbGE{an insect that makes honey in the ground} \tbOP{buulo} \tbTP{buulee} \tbVA{wirikuoluŋ}
\tbLX{buuli}\tbHM{} \tbPH{búúlí} \tbPS{n.} \tbGE{bells worn on the wrist to play xylophone} \tbOP{buulo} \tbTP{buulee}
\tbLX{buuli}\tbHM{} \tbPH{búúlì} \tbPS{v.} \tbGE{nurse (seed)} \tbXV{O buu\-li la kommie a dabɔge poɔ.} \tbXE{She has nursed tomoto seed in the garden.} \tbPD{buulee, buulo, buulo, buulibo, buulaa, buulibo } \tbVA{buoli/bugili}
\tbLX{buuli}\tbHM{} \tbPH{bùùlì} \tbPS{v.} \tbGE{to boil (first stage in making porridge paste; saabo)} \tbPD{buu\-lee, buulo, buulo, buulibo, buulaa, buulo }
\tbLX{buuli}\tbHM{} \tbPH{bùùlì} \tbPS{v.} \tbGE{to murmur, to grumble} \tbXV{Ka o poɔŋ sãã o maŋ buuli la yɛlɛ yaga.} \tbXE{When she is sad, she grumbles a lot.} \tbPD{buulee, buulo, buu\-lo, buulibo, buulaa, buulo }
\tbLX{buulo}\tbHM{} \tbPH{búúló} \tbPS{n.} \tbGE{something that complements or adorns another thing} \tbXV{Nɔtoro e la dɔɔ noɔre buu\-lo.} \tbXE{A mustache complements a man's mouth.}
\tbLX{buulunwala}\tbHM{} \tbPH{bùùlùnwálá} \tbPS{n.} \tbGE{light porridge} \tbTP{buulunwalɛɛ}
\tbLX{buulunwoɔlɔ}\tbHM{} \tbPH{bùùlùnwʊ́ɔ́lɔ́} \tbPS{n.} \tbGE{porridge used in the preparation of TZ} \tbTP{buulunwoɔlɛɛ} \tbVA{buulunwoɔla}
\tbLX{buuluŋ}\tbHM{} \tbPH{búúlúŋ} \tbPS{n.} \tbGE{porridge} \tbTP{buu\-lu\-nee}
\tbLX{buuluŋ}\tbHM{} \tbPH{bùúlúŋ} \tbPS{n.} \tbGE{fleshy portion (meat)} \tbOP{buulo} \tbTP{buulee}
\end{letter}
