
\begin{letter}{e}
\tbLX{e}\tbHM{} \tbPH{ɪ̀} \tbPS{v.} \tbSN \tbGE{to act, to do} \tbXV{Te na e o la velaa.} \tbXE{We shall do him good.} \tbSN \tbGE{to let} \tbXV{E ka te gaa pampana.} \tbXE{Let us go now.} \tbSN \tbGE{to fetch} \tbXV{A dɔɔ e la a kyi ko a pɔge.} \tbXE{The man has fetched  millet for the woman.} \tbSN \tbGE{to be} \tbXV{A bie e la kpeɛŋaa.} \tbXE{The child is strong.} \tbPD{eɛɛ, erɛ, erɛ, erebɛ, eraa, erebɛ }
\tbLX{eele}\tbHM{} \tbPH{ɪ́ɪ́lɪ́} \tbPS{n.} \tbSN \tbGE{horn} \tbSN \tbGE{antenna} \tbOP{eelɛ} \tbTP{eelɛɛ}
\tbLX{eɛ}\tbHM{} \tbPH{ɪ̀ɛ̀} \tbPS{v.} \tbSN \tbGE{to pinch} \tbSN \tbGE{to hint} \tbXV{Ba eɛ la a yɛlɛ fẽẽ ka N woŋ.} \tbXE{They have given me a hint of the situation.} \tbPD{eɛɛ, eɛrɛ, eɛrɛ, eɛrebɛ, eɛraa, eɛrebɛ }
\tbLX{ẽɛ}\tbHM{} \tbPH{ɪ̃́ɛ̃́} \tbPS{n.} \tbGE{} \tbSG{eɛnoo} \tbOP{ẽɛ} \tbTP{enɛɛ} \tbVA{eɛne}
\tbLX{ẽɛ}\tbHM{} \tbPH{ɪ̃́ɛ̃̀} \tbPS{v.} \tbSN \tbGE{to mark the foundation of a building} \tbSN \tbGE{to cut a very small piece} \tbPD{eɛɛ, eɛnɛ, eɛnɛ, eɛnemɛ, eɛnaa, eɛnemɛ }
\tbLX{eɛ eŋe}\tbHM{} \tbPH{ɪ̀ɛ̀ ɪ̀ŋ} \tbPS{v.} \tbGE{to flee, to run away} \tbXV{A bie zu pɔre la maŋgo ka a daana waana ka o eɛ eŋ.} \tbXE{The boy stole a mango and when he saw the owner coming he took to his heels.} \tbPD{eɛ eŋɛɛ, eɛrɛ ennɛ, eɛ ennɛ, eɛ e\-ne\-mɛ, eɛ ennaa}%, , -}
\tbLX{eɛne}\tbHM{} \tbPH{ɪ̀ɛ̀nɪ̀} \tbPS{n.} \tbGE{hippopotamus} \tbOP{eɛmɛ} \tbTP{eɛnɛɛ}
\tbLX{eɛne}\tbHM{} \tbPH{ɪ́ɛ́nɪ́} \tbPS{n.} \tbGE{claw} \tbSG{eɛnoo} \tbOP{eɛmɛ} \tbTP{eɛnɛɛ}
\tbLX{eɛne}\tbHM{} \tbPH{ɪ̀ɛ̌nɪ́} \tbPS{n.} \tbGE{a type of deer} \tbOP{eɛmɛ} \tbTP{eɛmɛɛr}
\tbLX{eɛne}\tbHM{} \tbPH{ɪ̀ɛ́nɪ́} \tbPS{adj.} \tbGE{tiny, e.g. fingernail} \tbOP{eɛmɛ} \tbTP{eɛmɛɛ}
\tbLX{eɛne}\tbHM{} \tbPH{ɪ̀ɛ̀nɪ̀} \tbPS{v.} \tbGE{to claw e.g. a cat climbing a wall} %\tbPD{-, -, -, -, -, -}
\tbLX{ẽɛo}\tbHM{} \tbPH{ɪ̃̀ɛ̃̀ʊ̃́} \tbPS{n.} \tbGE{type of deer} \tbOP{eɛne} \tbTP{eɛnɛɛ}
\tbLX{eɛraa}\tbHM{} \tbPH{ɪ́ɛ̀ráá} \tbPS{n.} \tbGE{hook, e.g. fishhook} \tbOP{eɛre} \tbTP{eɛrɛɛ}
\tbLX{eɛre}\tbHM{} \tbPH{ɪ́ɛ́rɪ̀} \tbPS{v.} \tbGE{to wipe away e.g. dirt, tears} \tbPD{eɛrɛɛ, eɛrɛ, eɛrɛ, eɛrebɛ, eɛraa, eɛrebɛ }
\tbLX{eɛre}\tbHM{} \tbPH{ɪ̀ɛ̀rɪ̀} \tbPS{v.} \tbGE{to claw at} \tbPD{eɛrɛɛ, eɛrɛ, eɛrɛ, eɛrebɛ, eɛraa, eɛrebɛ }
\tbLX{eɛre-miri}\tbHM{} \tbPH{ɪ̀ɛ̀rɪ̀-mírì} \tbPS{n.} \tbGE{fishing line} \tbOP{eɛre-mie} \tbTP{eɛre-miiree}
\tbLX{e-la}\tbHM{} \tbPH{ɪ̀-là} \tbPS{v.} \tbGE{to tease} \tbPD{e-laɛ, erɛ laara, e-laara, e-laareba, e-laaraa, e-lareba }
\tbLX{e-laara}\tbHM{} \tbPH{ɪ̀-láárá} \tbPS{n.} \tbGE{one who teases} \tbOP{e-lareba} \tbTP{e-larebɛɛ}
\tbLX{embaare}\tbHM{} \tbPH{ɪ̀mbáárɪ́} \tbPS{n.} \tbGE{laziness} \tbTP{embarɛɛ} \tbVA{entuo}
\tbLX{emii}\tbHM{} \tbPH{ɪ́míì} \tbPS{n.} \tbGE{act of desperation; impatience} \tbTP{emmiinee}
\tbLX{emmaa}\tbHM{} \tbPH{ɪ̀máà} \tbPS{n.} \tbGE{the condition of being in a good state}
\tbLX{emmaaroŋ}\tbHM{} \tbPH{ɪ̀mmáárʊ́ŋ} \tbPS{n.} \tbGE{health} \tbOP{emmaara} \tbTP{emmaarɛɛ}
\tbLX{emmii}\tbHM{} \tbPH{ɪ̀mmìì} \tbPS{n.} \tbGE{a body that is in a painful condition} \tbOP{emmiime} \tbTP{emmiinee}
\tbLX{emmiiruŋ}\tbHM{} \tbPH{ɪ̀mmììrûŋ} \tbPS{n.} \tbSN \tbGE{a body in pain} \tbSN \tbGE{stunted body} \tbTP{em\-mii\-ree}
\tbLX{emmiiruŋ}\tbHM{} \tbPH{ɪ̀mmììrûŋ} \tbPS{adj.} \tbGE{stunted}
\tbLX{Emɔ}\tbHM{} \tbPH{ɪ́mɔ̀} \tbPS{n.} \tbSN \tbGE{the name of a clan in Southern Dagao} \tbSN \tbGE{appellation of this group} \tbSG{emoɔloo} \tbOP{emoɔlɔ} \tbTP{emoɔlɛɛ}
\tbLX{empoɔne}\tbHM{} \tbPH{ɪ́mpʊ́ɔ́nɪ́} \tbPS{n.} \tbGE{ocean} \tbOP{mpoɔnɛɛ} \tbVA{po}
\tbLX{endaa}\tbHM{} \tbPH{ɪ̀ndàá} \tbPS{n.} \tbGE{frame (of a person's body)} \tbOP{endaare} \tbTP{endarɛɛ/endaarɛɛ}
\tbLX{endaa waaloŋ}\tbHM{} \tbPH{ɪ̌ndàá-wàálʊ̂ŋ} \tbPS{n.} \tbGE{skin colour, appearance of the body}
\tbLX{enfuomaala}\tbHM{} \tbPH{ɪ̀nfùòmáálá} \tbPS{n.} \tbGE{one who makes pictures} \tbOP{enfuomaaleba} \tbTP{enfuomaalebɛɛ}
\tbLX{enfuomaale}\tbHM{} \tbPH{ɪ̀nfùòmàálɪ̀} \tbPS{n.} \tbGE{the act of making a picture (draw)}
\tbLX{enfuoni}\tbHM{} \tbPH{ɪ̀nfúóní} \tbPS{n.} \tbGE{image, photograph, picture} \tbOP{enfuomo} \tbTP{enfuonee}
\tbLX{enfuori}\tbHM{} \tbPH{ɪ̀nfùòrì} \tbPS{n.} \tbGE{the act of bewitching; the act of casting a spell}
\tbLX{enne}\tbHM{} \tbPH{ɪ̀nnɪ́} \tbPS{n.} \tbGE{act of respect} \tbTP{ennɛɛ} \tbVA{eŋene}
\tbLX{ennoɔ}\tbHM{} \tbPH{ɪ̀nnʊ̀ɔ́} \tbPS{n.} \tbGE{happiness} \tbOP{ennoore} \tbTP{ennoorɛɛ}
\tbLX{ensɔgelaa}\tbHM{} \tbPH{ɪ́nsɔ́gláá} \tbPS{n.} \tbGE{black skin} \tbOP{ensɔglɔ}
\tbLX{entaa}\tbHM{} \tbPH{ɪ̀ntáá} \tbPS{n.} \tbGE{mutual respect}
\tbLX{entage}\tbHM{} \tbPH{ɪ̌ntágɪ́} \tbPS{n.} \tbGE{muscular body} \tbOP{entagre} \tbTP{entagrɛɛ}
\tbLX{entole}\tbHM{} \tbPH{ɪ́ntʊ́lɪ̀} \tbPS{n.} \tbGE{fever, rise in body temperature} \tbOP{entolɔ/entola} \tbTP{entolɛɛ}
\tbLX{entoloŋ}\tbHM{} \tbPH{ɪ́ntʊ́lʊ̂ŋ} \tbPS{n.} \tbGE{fever} \tbOP{entola} \tbTP{entollɛɛ}
\tbLX{entukpalaa}\tbHM{} \tbPH{ɪ̂ntùkpàláà} \tbPS{n.} \tbGE{extreme laziness} \tbOP{entukpalle} \tbTP{en\-tu\-kpa\-lɛɛ}
\tbLX{entuo}\tbHM{} \tbPH{ɪ̀ntúó} \tbPS{n.} \tbGE{laziness} \tbOP{entuuri} \tbTP{entuuree} \tbVA{embaare}
\tbLX{enzeɛ}\tbHM{} \tbPH{ǐnzɪ̀ɛ́} \tbPS{n.} \tbSN \tbGE{light skin} \tbSN \tbGE{leprosy} \tbOP{enzeere} \tbTP{enzeerɛɛ}
\tbLX{eŋ}\tbHM{} \tbPH{ɪ́ŋ} \tbPS{part.} \tbGE{as} \tbXV{Maa eŋ koŋ gaa.} \tbXE{As for me I won't go.}
\tbLX{eŋe}\tbHM{} \tbPH{ɪ̀ŋ} \tbPS{v.} \tbSN \tbGE{to put in} \tbXV{o ɔŋ la kõɔ eŋe a doge poɔ.} \tbXE{S/he has put water into the pot.} \tbSN \tbGE{to put on (shoes, ring, spectacles)} \tbSN \tbGE{to respect} \tbXV{A seŋe ka bibile maŋ eŋe neŋkpoŋ.} \tbXE{The young should respect the elderly.}   \tbPD{eŋɛɛ, ennɛ, ennɛ, enemɛ, ennaa}%, , -}
\tbSE{eŋe daa} \tbGE{to set a date}
\tbSE{eŋe daa} \tbGE{to beat}
\tbLX{eŋene} \tbPH{ɪ́ŋɪ́nɪ́} \tbPS{n.} \tbGE{main hole leading to a rodent's living room in its burrow}  \tbOP{enne} \tbTP{eŋenɛɛ}
\tbLX{eŋene} \tbPH{ɪ́ŋɪ́nɪ́} \tbPS{n.} \tbGE{respect, regard for, honour} \tbXV{A teŋɛ deme zaa taa la eŋene kora a naa.} \tbXE{All the people in town have respect for the chief.} \tbTP{ennɛɛ} \tbVA{enne}
\tbLX{eŋɛ}\tbHM{} \tbPH{ɪ́ŋɛ́} \tbPS{pron.} \tbGE{what, who} \tbXV{Boŋ eŋɛ ka fo wa?} \tbXE{What have you come for?}
\tbLX{eŋɛ}\tbHM{} \tbPH{ɪ́ŋɛ́} \tbPS{n.} \tbGE{body (surface of)} \tbOP{enne} \tbTP{ennɛɛ}
\tbLX{eŋgane}\tbHM{} \tbPH{ɪ̀ŋgánɪ̀} \tbPS{n.} \tbGE{skin (human body)} \tbSG{eŋganoo} \tbOP{eŋgama} \tbTP{eŋgannɛɛ}
\tbLX{Eŋgelɛ}\tbHM{} \tbPH{ɪ́ŋgɪ̀lɛ̀} \tbPS{n.} \tbGE{England} \tbTP{eŋgelɛɛ}
\tbLX{Eŋgilisi}\tbHM{} \tbPH{ɪ́ŋgìlìsì} \tbPS{n.} \tbGE{English} \tbOP{ɪŋlisiri} \tbTP{ɪŋgilisiree}
\tbLX{eŋkɔbɔ}\tbHM{} \tbPH{ɪ̀ŋkɔ́bɔ́} \tbPS{n.} \tbGE{skeleton} \tbVA{eŋkɔba}
\tbLX{eŋkpaŋkpoli}\tbHM{} \tbPH{ɪ́ŋkpáŋkpòlì} \tbPS{n.} \tbGE{naked body} \tbVA{eŋkpoŋkpoli, eŋkpoli}
\tbLX{eŋkpoli}\tbHM{} \tbPH{ɪ̀ŋkpòlɪ́} \tbPS{n.} \tbGE{naked body} \tbOP{eŋkpolo} \tbTP{eŋkpolee} \tbVA{eŋkpoŋkpoli, eŋkpaŋkpoli}
\tbLX{eŋkpoŋkpoli}\tbHM{} \tbPH{ɪ̀ŋkpóŋkpólì} \tbPS{n.} \tbGE{naked body} \tbOP{eŋkpoŋkpolo} \tbTP{eŋkpoŋkpolee} \tbVA{eŋkpoli, eŋ\-kpaŋ\-kpoli}
\tbLX{eŋ noɔre}\tbHM{} \tbPH{ɪ̀ŋ nʊ́ɔ́rɪ̀} \tbPS{v.} \tbGE{to promise} %\tbPD{-, -, -, -, -, -}
\tbLX{eŋ poɔ}\tbHM{} \tbPH{ɪ̀ŋ pʊ́ɔ́} \tbPS{v.} \tbGE{to empregnate} %\tbPD{-, -, -, -, -, -}
\tbLX{eŋyoɔre}\tbHM{} \tbPH{ɪ̀ŋyʊ̀ɔ̀rɪ̀} \tbPS{adj.} \tbGE{that which does not produce excitement} \tbVA{enyoɔre}
\tbLX{eŋ zẽɛ}\tbHM{} \tbPH{ɪ̀ŋ zɪ̃̀ɛ̃} \tbPS{v.} \tbGE{to punish severely} \tbPD{eŋɛɛ zẽɛ̃, ennɛ zẽɛ̃}%, , -, -, -, -}
\tbLX{eroŋ}\tbHM{} \tbPH{ɪ̀rʊ̂ŋ} \tbPS{n.} \tbGE{conduct} \tbOP{erre} \tbTP{errɛɛ}
\tbLX{eroŋ}\tbHM{} \tbPH{ɪ̀rʊ̂ŋ} \tbPS{n.} \tbGE{verb} \tbOP{erre} \tbTP{errɛɛ}
\tbLX{Etore}\tbHM{} \tbPH{ítʊ́rɪ̀} \tbPS{n.} \tbGE{Easter} \tbOP{itorɛɛ}
\tbLX{etɔ}\tbHM{} \tbPH{ɪ́tɔ̀} \tbPS{n.} \tbGE{the name of a clan in Southern Dagao} \tbOP{etoɔlɔ} \tbTP{etoɔlɛɛ}
\tbLX{ewa}\tbHM{} \tbPH{ɪ́wà} \tbPS{n.} \tbSN \tbGE{a kinship group} \tbSN \tbGE{appellation of this group} \tbSG{ewaaloo} \tbOP{ewaala} \tbTP{ewaalɛɛ}
\tbLX{ewɛ}\tbHM{} \tbPH{ɪ́wɛ̀} \tbPS{n.} \tbGE{the name of a clan in Southern Dagao} \tbOP{eweɛlɛ} \tbTP{eweɛlɛɛ}
\end{letter}
