
\begin{letter}{g}
\tbLX{ga}\tbHM{} \tbPH{gà} \tbPS{v.} \tbGE{to eat TZ without soup} \tbPD{gaɛ, gaara, gaara, garba, gaaraa, gaareba }
\tbLX{ga}\tbHM{} \tbPH{gá} \tbPS{v.} \tbGE{to run wildly or in a frenzy e.g. cow stung by tsetse fly} \tbPD{gagaɛ, gaara, gaara, gareba, gaaraa, gareba }
\tbLX{gaa}\tbHM{} \tbPH{gáà} \tbPS{n.} \tbGE{ebony (tree or fruit)} \tbOP{gaare} \tbTP{garɛɛ/gaarɛɛ}
\tbLX{gaa}\tbHM{} \tbPH{gàà} \tbPS{v.} \tbGE{to leave} \tbPD{gaaɛ, gɛrɛ, gɛrɛ, gɛrebɛ, gɛraa, gɛrebɛ }
\tbLX{gaa}\tbHM{} \tbPH{gáà} \tbPS{v.} \tbGE{to scoop pulp e.g. soil} \tbPD{gaaɛ, gaara, gaara, gaareba, gaaraa, gaareba }
\tbLX{gãa}\tbHM{} \tbPH{gã̀ã̀} \tbPS{adv.} \tbGE{outstretched and long} \tbXV{A sɔrsɔrɔ tẽɛ la o nu gãa ka nembaaloŋ erɛ a noba ka ba korɔ o libie.} \tbXE{The beggar stretched out his hand ernestly and the onlookers were moved and began offering him money.}
\tbLX{gaabu}\tbHM{} \tbPH{gáábú} \tbPS{n.} \tbGE{onion} \tbOP{gaaburi} \tbTP{gaaburee/gaaburree} \tbVA{zam\-mo}
\tbLX{gaafara}\tbHM{} \tbPH{gááfárà} \tbPS{n.} \tbGE{excuse, sorry} \tbSG{gaafaroo} \tbOP{gaafarre} \tbTP{gaafarɛɛ}
\tbLX{gaale}\tbHM{} \tbPH{gáálɪ̀} \tbPS{v.} \tbSN \tbGE{to place flat on the ground, to put to bed} \tbSN \tbGE{to attend prenatal or postnatal care} \tbPD{gaalɛɛ, gaala, gaala, gaalba, gaalaa, gaaleba }
\tbLX{Gaana}\tbHM{} \tbPH{gáánà} \tbPS{n.} \tbGE{Ghana} \tbOP{gaanarɪ} \tbTP{gaanɛɛ}
\tbLX{gaane}\tbHM{} \tbPH{gàànɪ̀} \tbPS{v.} \tbSN \tbGE{to stride; to take strides of equal length with energy and confidence} \tbSN \tbGE{to wear by passing a strip between the buttocks (triangular underwear)} \tbXV{Neŋkponni mine taa la parre a maŋ de gaane yelmaale daare.} \tbXE{Some elderly men have triangular underwear which they wear on festive occasions.} \tbPD{gaanɛɛ, gaana, gaana, gaanema, gaanema, gaangaana }
\tbLX{gaanoo}\tbHM{} \tbPH{gaanoo} \tbPS{n.} \tbGE{the act of striding} \tbVA{gaane}
\tbLX{gabaa}\tbHM{} \tbPH{gàbàà} \tbPS{adv.} \tbGE{open and extensive}
\tbLX{gabadayaa}\tbHM{} \tbPH{gàbádáyáá} \tbPS{adj.} \tbGE{straightaway} \tbVA{gaba}
\tbLX{gabadii}\tbHM{} \tbPH{gàbàdìì} \tbPS{n.} \tbGE{a type of woolen cloth used to make suits}
\tbLX{gabaŋgabaŋ}\tbHM{} \tbPH{gàbàŋgàbàŋ} \tbPS{adv.} \tbGE{anyhow}
\tbLX{gadaa}\tbHM{} \tbPH{gàdàà} \tbPS{n.} \tbSN \tbGE{ebony stick} \tbSN \tbGE{sign of extradition} \tbXV{A tendaana ba la nɔgraa gadaa.} \tbXE{The landlord has ordered the sojourner to leave.}
\tbLX{gado}\tbHM{} \tbPH{gádó} \tbPS{n.} \tbGE{bed} \tbOP{gadori} \tbTP{gadoree/gadorree} \tbVA{godo}
\tbLX{gagaraa}\tbHM{} \tbPH{gágáráà} \tbPS{n.} \tbSN \tbGE{traitor} \tbSN \tbGE{One who denies the truth and treads on falsehood, inconsistent} \tbOP{gagarre} \tbTP{gagarɛɛ} \tbVA{minaafigi}
\tbLX{gakogili}\tbHM{} \tbPH{gákóglíí} \tbPS{n.} \tbGE{dried berry of the ebony tree} \tbSG{gakogiluu} \tbOP{gakogilo} \tbTP{gakogilee}
\tbLX{galaa}\tbHM{} \tbPH{gàláà} \tbPS{n.} \tbGE{fool} \tbOP{galle} \tbTP{gallɛɛ}
\tbLX{galbie}\tbHM{} \tbPH{gálbíé} \tbPS{n.} \tbGE{a great person} \tbOP{galbiiri} \tbTP{galbiiree}
\tbLX{galbie}\tbHM{} \tbPH{gálbíé} \tbPS{n.} \tbGE{a child born after more than 9 months of pregnancy} \tbOP{galbiiri} \tbTP{galbiiree}
\tbLX{gale}\tbHM{} \tbPH{gálɪ́} \tbPS{v.} \tbGE{to speak in anger; to reproach} \tbPD{galɛɛ, gala, gala, galeba, galaa, galeba }
\tbLX{gale}\tbHM{} \tbPH{gàlɪ̀} \tbPS{v.} \tbGE{to miss the regular time} \tbXV{O daŋe ba gale toma waabo bebiyeni gba.} \tbXE{S/he never missed going to work even for a day.} \tbPD{galɛɛ, gala, gala, galba, galaa, galba } \tbVA{galle}
\tbLX{galegale}\tbHM{} \tbPH{gálgál} \tbPS{adv.} \tbGE{of quick fast steps (run)}
\tbLX{galegale}\tbHM{} \tbPH{gálɪ́gálɪ́} \tbPS{n.} \tbGE{dog}
\tbLX{galema}\tbHM{} \tbPH{gálɪ́má} \tbPS{adj.} \tbGE{strange} \tbOP{galemalema} \tbVA{galee}
\tbLX{galeŋgaa}\tbHM{} \tbPH{gàlɪ́ŋgáà} \tbPS{n.} \tbGE{crow} \tbOP{galeŋgaare} \tbTP{galeŋgaarɛɛ}
\tbLX{galeŋgoɔnaa}\tbHM{} \tbPH{gàlɪ̀ŋgʊ́ɔ́náá} \tbPS{n.} \tbGE{heron} \tbOP{galeŋgoɔne} \tbTP{galeŋgoɔnɛɛ} \tbVA{gɔleŋgoɔnaa}
\tbLX{galiŋkoori}\tbHM{} \tbPH{gàlíŋkóórì} \tbPS{n.} \tbGE{millipede} \tbSG{galiŋkooruu} \tbOP{galiŋkoro} \tbTP{galiŋkoree}
\tbLX{galle}\tbHM{} \tbPH{gàllɪ̀} \tbPS{v.} \tbGE{to pass, to miss} \tbXV{Ka pɔgeŋ wa galle o pɛgebo kaapɔge poɔ la ka o taa.} \tbXE{If a woman misses her period, then she may be pregnant.} \tbPD{gallɛɛ, galla, galla, galeba, gallaa}%, , -}
\tbLX{galle}\tbHM{} \tbPH{gàllɪ̀} \tbPS{v.} \tbGE{manoeuvre with skill and expertise} \tbXV{Te karema maŋ galle la a kɔkɔre ka noɔ kpɛ fo.} \tbXE{Our teacher manipulates the language such that you enjoy his talk.} \tbPD{gallɛɛ, galla, galla, galeba, gallaa}%, , -}
\tbLX{Galyuoni}\tbHM{} \tbPH{gàlyúóní} \tbPS{n.} \tbGE{name traditionally given to a child born after 9 months of pregnancy} \tbVA{Yuoni}
\tbLX{gambanna}\tbHM{} \tbPH{gámbánná} \tbPS{n.} \tbGE{one who is literate} \tbOP{gambanema} \tbTP{gambannɛɛ}
\tbLX{gammiri}\tbHM{} \tbPH{gámmírì} \tbPS{n.} \tbGE{woven twine string worn by women around the waist} \tbSG{gammiruu} \tbOP{gammie} \tbTP{gammiiree}
\tbLX{gammisɔgelaa}\tbHM{} \tbPH{gámmísɔ́gláà} \tbPS{n.} \tbGE{woven leather waist string} \tbOP{gammisɔgelɔ}
\tbLX{gammizoolɔ}\tbHM{} \tbPH{gámmízóólɔ̀} \tbPS{n.} \tbGE{waist strings for adolescent girl} \tbOP{gammizoolɛɛ} \tbVA{gammizoola}
\tbLX{gampɛle}\tbHM{} \tbPH{gámpɛ́lɪ̀} \tbPS{n.} \tbSN \tbGE{sheet of paper} \tbSN \tbGE{page of a book} \tbOP{ gampɛlɛ}
\tbLX{gamporrizie}\tbHM{} \tbPH{gámpórrìzíé} \tbPS{n.} \tbGE{post office} \tbOP{gamporriziiri} \tbTP{gamporriziiree}
\tbLX{ganaa}\tbHM{} \tbPH{gànàá} \tbPS{n.} \tbSN \tbGE{strip of leather} \tbSN \tbGE{twine for chief mourners of a deceased person} \tbOP{ganne} \tbTP{gannɛɛ}
\tbLX{ganaaganaa}\tbHM{} \tbPH{gánáágánáá} \tbPS{adv.} \tbGE{small and stretched like a string, thin}
\tbLX{ganaaganaa}\tbHM{} \tbPH{gànààgànàà} \tbPS{adv.} \tbGE{large and stretched} \tbXV{A dɔbɔ ti la taa ganaaganaa, ka ba faara faana.} \tbXE{The men hooked each other tightly and fiercely such that people were finding it difficult to separate them.}
\tbLX{gandaa}\tbHM{} \tbPH{gàndáá} \tbPS{n.} \tbSN \tbGE{hero, great person, protagonist} \tbSN \tbGE{hero in literature} \tbOP{gandaare} \tbTP{gandaarɛɛ}
\tbLX{gandaa}\tbHM{} \tbPH{gàndáá} \tbPS{n.} \tbGE{placenta} \tbOP{gandaare} \tbTP{gandaarɛɛ}
\tbLX{gane}\tbHM{} \tbPH{gánɪ̀} \tbPS{n.} \tbSN \tbGE{book} \tbSN \tbGE{hide, skin} \tbOP{gama} \tbTP{gannɛɛ/gamɛɛ}
\tbLX{gangyinni}\tbHM{} \tbPH{gàndzínní} \tbPS{n.} \tbGE{the act of doing things in excess e.g. eating} \tbTP{gangyinnee} \tbVA{gangyinnuŋ}
\tbLX{gangyinnuŋ}\tbHM{} \tbPH{gàndzínnúŋ} \tbPS{n.} \tbGE{the act of doing things in excess} \tbVA{gangyinnuu, gangyinni}
\tbLX{ganlaa}\tbHM{} \tbPH{gánláá} \tbPS{n.} \tbGE{a type of dance} \tbOP{ganlaare} \tbTP{ganlaarɛɛ}
\tbLX{ganlɔgɔ}\tbHM{} \tbPH{gànlɔ̀gɔ̀} \tbPS{n.} \tbGE{small leather bag worn on the shoulder} \tbOP{ganlɔgre} \tbTP{ganlɔgrɛɛ}
\tbLX{gansɛgerɛ}\tbHM{} \tbPH{gánsɛ́grɛ̀} \tbPS{n.} \tbGE{writer, letter writer} \tbOP{gansegerebɛ} \tbTP{ganɛgerebɛɛ} \tbVA{gaŋŋmaara}
\tbLX{gansɔraa}\tbHM{} \tbPH{gánsɔ́ráá} \tbPS{n.} \tbGE{a book for reading} \tbOP{gansɔrre} \tbTP{gansɔrrɛɛ}
\tbLX{gantoɔle}\tbHM{} \tbPH{gántʊ́ɔ́lɪ̀} \tbPS{n.} \tbGE{mail} \tbSG{gantoɔlaa} \tbTP{gantoɔlɛɛ}
\tbLX{gaŋa}\tbHM{} \tbPH{gàŋá} \tbPS{n.} \tbGE{An enclosure for livestock e.g. cattle} \tbOP{ganne} \tbTP{gannɛɛ} \tbVA{gaŋe}
\tbLX{gaŋe}\tbHM{} \tbPH{gáŋ} \tbPS{adj.} \tbGE{mad or insane (dogs)} \tbXV{Bagaŋe zaameŋ doŋe la n baa.} \tbXE{A mad dog bit my dog yesterday.}
\tbLX{gaŋe}\tbHM{} \tbPH{gâŋ} \tbPS{v.} \tbSN \tbGE{to lie down} \tbSN \tbGE{to sleep} \tbPD{gaŋɛɛ, gana, gana, ganema, ganaa, ganema }
\tbLX{gaŋe}\tbHM{} \tbPH{gàŋɪ̀} \tbPS{v.} \tbSN \tbGE{to step over} \tbXV{A baa age gaŋ la a bogi.} \tbXE{The dog jumped over the hole.} \tbSN \tbGE{to overtake,} \tbXV{Ba lɔɔre zo wa gaŋ la te deni na bare.} \tbXE{Their lorry overtook ours.} \tbSN \tbGE{to be more than, to exceed} \tbXV{A kparoo waa la kpoŋ gaŋ a bie.} \tbXE{The shirt is bigger than the child.} \tbPD{gaŋɛɛ, gaŋena/gana, gana, gaŋenema/ganema, gaŋenaa/ganaa, gaŋenema/ganema }
\tbLX{gaŋene} \tbPH{gàŋɪ̀nɪ́} \tbPS{n.} \tbGE{next one} \tbXV{N daga maŋ biŋ la a gaŋene die na poɔ.} \tbXE{My box is kept in the room after the first one.}
\tbLX{gaŋene} \tbPH{gàŋɪ̀nɪ́} \tbPS{n.} \tbGE{that which is too much}
\tbLX{gaŋgaa}\tbHM{} \tbPH{gàŋgàá} \tbPS{n.} \tbGE{drum (general)} \tbOP{gaŋgarɪ} \tbTP{gaŋgarɛɛ}
\tbLX{gaŋgaanaa}\tbHM{} \tbPH{gǎŋgàànáà} \tbPS{n.} \tbSN \tbGE{reed} \tbSN \tbGE{vine} \tbOP{gaŋgaane} \tbTP{gaŋgaanɛɛ}
\tbLX{gaŋgadaa}\tbHM{} \tbPH{gàŋgàdáá} \tbPS{n.} \tbGE{small drum with a quieter sound} \tbOP{gaŋgadaare} \tbTP{gaŋgadaarɛɛ}
\tbLX{gaŋgadalee}\tbHM{} \tbPH{gàŋgàdàléé} \tbPS{n.} \tbGE{drumstick} \tbOP{gaŋgadabilii} \tbVA{gaŋga\-da\-bi\-li}
\tbLX{gaŋgakpoŋ}\tbHM{} \tbPH{gàŋgàkpóŋ} \tbPS{n.} \tbGE{bass drum} \tbOP{gaŋgakponni} \tbTP{gaŋgakponnee}
\tbLX{gaŋgalee}\tbHM{} \tbPH{gàŋgàléé} \tbPS{n.} \tbGE{small drum; alto drum} \tbOP{gaŋgabilii} \tbTP{gaŋgabilee}
\tbLX{gaŋganyaŋaa}\tbHM{} \tbPH{gàŋgàɲáŋáá} \tbPS{n.} \tbGE{drum with a deep, loud sound} \tbOP{gaŋganyaŋene} \tbTP{gaŋganyaŋenɛɛ}
\tbLX{gaŋkanna}\tbHM{} \tbPH{gáŋkánná} \tbPS{n.} \tbGE{reader, letter or book reader} \tbOP{gaŋkanema} \tbTP{gaŋkanemɛɛ}
\tbLX{gaŋkoŋi}\tbHM{} \tbPH{gàŋkǒŋ} \tbPS{n.} \tbGE{funeral string/strip given without any accompanying funeral donation} \tbOP{gaŋkonni} \tbTP{gaŋkonnee}
\tbLX{gaŋkoŋi}\tbHM{} \tbPH{gáŋkóŋ} \tbPS{n.} \tbGE{dry and crumpled animal hide} \tbOP{gaŋkonni} \tbTP{gaŋkonnee}
\tbLX{gaŋkuoŋaa}\tbHM{} \tbPH{gàŋkúóŋáá} \tbPS{n.} \tbGE{dried strip of leather} \tbOP{gaŋkuoni} \tbTP{gaŋkuonee}
\tbLX{gaŋŋmaara}\tbHM{} \tbPH{gáŋŋmáárá} \tbPS{n.} \tbGE{writer} \tbOP{gaŋŋmaareba} \tbTP{gaŋŋmaarebɛɛ} \tbVA{gansɛgrɛ}
\tbLX{gara}\tbHM{} \tbPH{gàrá} \tbPS{n.} \tbGE{tapeworm} \tbOP{garre} \tbTP{garɛɛ/garrɛɛ}
\tbLX{gara}\tbHM{} \tbPH{gárá} \tbPS{n.} \tbGE{hobble (rope for holding two legs of an animal)} \tbOP{garre} \tbTP{garɛɛ/garrɛɛ} \tbVA{gbara}
\tbLX{garaa}\tbHM{} \tbPH{gáráá} \tbPS{adj.} \tbGE{small and pointed}
\tbLX{garaa}\tbHM{} \tbPH{gàráà} \tbPS{n.} \tbGE{ridge} \tbOP{garre} \tbTP{garrɛɛ}
\tbLX{gare}\tbHM{} \tbPH{gárɪ́} \tbPS{v.} \tbSN \tbGE{to stumble, to trip over} \tbSN \tbGE{to tie or hold legs together} \tbPD{garɛɛ, gara, gara, gareba, garaa, gareba }
\tbLX{gare}\tbHM{} \tbPH{gàrɪ̀} \tbPS{v.} \tbGE{to place across, to close (road)} \tbXV{A dɔɔ de la daa gare a sori.} \tbXE{The man has placed a plank across the road.} \tbPD{garɛɛ, gara, gara, gareba, garaa, gareba }
\tbLX{gare}\tbHM{} \tbPH{gárɪ́} \tbPS{v.} \tbGE{go bad, sour (food)} \tbPD{garɛɛ, gara}%, , -, -, -, -}
\tbLX{garegyerɛ}\tbHM{} \tbPH{gárɪ́yɪ̀rɛ} \tbPS{n.} \tbGE{a shrub with milky sap, poisonous} \tbSG{garegyeroo} \tbTP{gargyerɛɛ}
\tbLX{Garegyerɛbaa}\tbHM{} \tbPH{gàrɪ́dzìrɛ́bàá} \tbPS{n.} \tbGE{One of the main district headquarters of the Dagaare speaking area whose dialeɔt is often referred to as the ‘central dialect’} \tbVA{Gyerɛbaa}
\tbLX{garepegele}\tbHM{} \tbPH{gàrɪ̀pɪ̀glɪ̀} \tbPS{v.} \tbGE{to crisscross} \tbXV{A teɛ le la garepegele a sori.} \tbXE{The tree fell and crossed/bocked the road.} \tbPD{garepegelɛɛ, garapegelɛ, garapegelɛ, garepegelebɛ, garepegelaa, garepeglebɛ }
\tbLX{garewaa}\tbHM{} \tbPH{gárɪ́wáá} \tbPS{n.} \tbGE{four-gallon tin} \tbOP{garewaare} \tbTP{garewaarɛɛ}
\tbLX{garoŋ}\tbHM{} \tbPH{gárʊ́ŋ} \tbPS{n.} \tbGE{sour-tasting substance}
\tbLX{garre}\tbHM{} \tbPH{gárrɪ́} \tbPS{v.} \tbGE{to deny} \tbPD{garrɛɛ, garra, garra, gareba, garaa, gareba }
\tbLX{gasalaa}\tbHM{} \tbPH{gàsàlàà} \tbPS{adj.} \tbGE{gaping}
\tbLX{gegili}\tbHM{} \tbPH{gèglì} \tbPS{v.} \tbGE{be half-closed} \tbPD{gegilee, gegile, gegile, gegilibe, gegilaa, gegilibe } \tbVA{ogili}
\tbLX{geli}\tbHM{} \tbPH{gélì} \tbPS{v.} \tbGE{to lift over something heavy lying flat on the ground} \tbXV{Ba geli la a kuri gbori tiili a dankyini poɔ.} \tbXE{They have lifted over the iron gate unto the wall.} \tbPD{gelee, gelle, gelle, gelibe, -, gelibe }
\tbLX{geli}\tbHM{} \tbPH{gélì} \tbPS{v.} \tbGE{to burn badly (food)} \tbPD{gelee, gelle, gelle, gelibe, -, gelibe }
\tbLX{geli}\tbHM{} \tbPH{gélì} \tbPS{v.} \tbGE{to be mad, to behave abnormally or foolishly} \tbPD{gelee, gelle, gelle, gelibe, -, gelibe }
\tbLX{gelime}\tbHM{} \tbPH{gélímè} \tbPS{n.} \tbGE{insanity}
\tbLX{gemee}\tbHM{} \tbPH{géméé} \tbPS{adj.} \tbGE{mad, insane}
\tbLX{genegene}\tbHM{} \tbPH{gènègènè} \tbPS{adv.} \tbSN \tbGE{tinted (colour)} \tbXV{A boɔ e la zeɛ genegene.} \tbXE{The goat is tinted brown.} \tbSN \tbGE{not appearing clearly} \tbXV{Zie naŋ da wa nyaana genegene la ka o wa.} \tbXE{S/he came when dawn had not clearly set in.}
\tbLX{geŋgeŋ}\tbHM{} \tbPH{géŋgéŋ} \tbPS{adj.} \tbGE{firmly fixed}
\tbLX{geŋi}\tbHM{} \tbPH{gêŋ} \tbPS{v.} \tbGE{to stretch out to reach e.g. to look over a wall} \tbPD{geŋee, geŋine/genne, geŋine/genne, geŋinime/genime, geŋinaa/gennaa, genime }
\tbLX{geŋi}\tbHM{} \tbPH{gêŋ} \tbPS{v.} \tbGE{to be mad, to be angry} \tbPD{geŋee, geŋine/genne, geŋine/genne, geŋinime/genime, geŋinaa/gennaa, genime }
\tbLX{geregere}\tbHM{} \tbPH{gèrègèrè} \tbPS{adv.} \tbGE{moderately hard e.g. half-cooked yam}
\tbLX{gɛbɛɛ}\tbHM{} \tbPH{gɛ̀bɛ́ɛ́} \tbPS{n.} \tbGE{fool, lunatic}
\tbLX{gieri}\tbHM{} \tbPH{gíérì} \tbPS{v.} \tbGE{to belch} \tbPD{gieree, giere, giere, gieribe, gieraa, gieribe }
\tbLX{go}\tbHM{} \tbPH{gó} \tbPS{v.} \tbGE{to stoop, to bend down} \tbPD{goe, guoro, guoro, guoribo, goraa, guoribo }
\tbLX{go}\tbHM{} \tbPH{gò} \tbPS{v.} \tbGE{watch over} \tbPD{goe, goro, goro, goribo, goraa, - } \tbVA{gu}
\tbLX{gobaa}\tbHM{} \tbPH{gòbàà} \tbPS{n.} \tbGE{shame, disgrace}
\tbLX{gogo}\tbHM{} \tbPH{gógó} \tbPS{n.} \tbSN \tbGE{clock} \tbSN \tbGE{hour} \tbOP{gogori} \tbTP{gogoree}
\tbLX{gogobiri}\tbHM{} \tbPH{gògòbírí} \tbPS{n.} \tbGE{zebu (bull with a humped back)} \tbOP{gogobie} \tbTP{gogobiiree}
\tbLX{gogoo}\tbHM{} \tbPH{gògóó} \tbPS{adj.} \tbGE{standing out with long appendages e.g. pointing}
\tbLX{gogoɔ}\tbHM{} \tbPH{gʊ̀gʊ̀ɔ́} \tbPS{n.} \tbGE{sawfish} \tbOP{gogoɔre} \tbTP{gogoɔrɛɛ}
\tbLX{golebɛlɛ}\tbHM{} \tbPH{gʊ̀lɪ̀bɛ̀lɛ̀} \tbPS{adj.} \tbGE{lethargic and unsteady (very sick person)}
\tbLX{goligoli}\tbHM{} \tbPH{gòlgólí} \tbPS{v.} \tbGE{to scribble} \tbPD{goligolee, goligolo, goligolo, golgolibo, golgolaa, golgolibo }
\tbLX{gollaa}\tbHM{} \tbPH{gólláà} \tbPS{n.} \tbGE{a game in which players hop on one leg} \tbOP{golli} \tbTP{gollee}
\tbLX{golle}\tbHM{} \tbPH{gʊ̀llɪ̀} \tbPS{v.} \tbGE{to mix up any which way (clothes)} \tbPD{gollɛɛ, gollɔ/golla, gollɔ/golla, golba, gollaa}%, , -}
\tbLX{golli}\tbHM{} \tbPH{góllì} \tbPS{v.} \tbGE{to hop on one leg} \tbPD{gollee, gollo, gollo, golibo, gollaa, golibo }
\tbLX{goloŋgoloŋ}\tbHM{} \tbPH{gòlòŋgólóŋ} \tbPS{adv.} \tbGE{long and twisted}
\tbLX{golɔgolɔ}\tbHM{} \tbPH{gʊ́lɔ́gʊ́lɔ́} \tbPS{adj.} \tbGE{soft and wavy (hair)}
\tbLX{golɔnsomm}\tbHM{} \tbPH{gʊ̀lɔ̀nsʊ̀mm} \tbPS{adv.} \tbGE{roundish and big}
\tbLX{golɔnsomm}\tbHM{} \tbPH{gʊ́lɔ́nsʊ́mm} \tbPS{adv.} \tbGE{roundish and small}
\tbLX{gombɔle}\tbHM{} \tbPH{gòmbɔ́lɪ́} \tbPS{n.} \tbGE{kapok fruit} \tbOP{gombɔlɔ} \tbTP{gombɔlɛɛ}
\tbLX{gomɔgomɔ}\tbHM{} \tbPH{gʊ̀mɔ̀gʊ̀mɔ̀} \tbPS{adv.} \tbGE{not straightforward} \tbVA{gomagoma}
\tbLX{gondoŋ}\tbHM{} \tbPH{gòndóŋ} \tbPS{n.} \tbGE{skirt} \tbOP{gon\-don\-ni} \tbTP{gondonnee}
\tbLX{gone}\tbHM{} \tbPH{gʊ́nɪ́} \tbPS{n.} \tbGE{structural flaw} \tbOP{gonne} \tbTP{gonɛɛ/gonnɛɛ}
\tbLX{gonnaabo}\tbHM{} \tbPH{gónnáábʊ́} \tbPS{n.} \tbGE{plant louse} \tbOP{gonnii} \tbTP{gonniinee}
\tbLX{gonni}\tbHM{} \tbPH{gònní} \tbPS{n.} \tbGE{cotton} \tbSG{gonnuu} \tbOP{} \tbTP{gonnee}
\tbLX{gonseɛ}\tbHM{} \tbPH{gònsɪ̀ɛ́} \tbPS{n.} \tbGE{fall; season of intermitent wind}
\tbLX{goŋo}\tbHM{} \tbPH{gòŋó} \tbPS{n.} \tbGE{cotton tree} \tbOP{gonni} \tbTP{gonnee}
\tbLX{goŋmane}\tbHM{} \tbPH{góŋmánɪ́} \tbPS{n.} \tbGE{one hundred kola nuts} \tbOP{goŋmama} \tbTP{goŋmannɛɛ}
\tbLX{goole}\tbHM{} \tbPH{gʊ̀ʊ̀lɪ̀} \tbPS{v.} \tbGE{to mix, to stir (thick paste)} \tbPD{goolɛɛ, goolɔ, goolɔ, gooleba, goolaa, gooleba }
\tbLX{goone}\tbHM{} \tbPH{gʊ̀ʊ̀nì} \tbPS{v.} \tbGE{to walk majestically with an up-and-down body movement} \tbXV{A daanedann da maŋe la a naa niŋŋ kyɛ goone iri.} \tbXE{The praise singer would walk majestically with up-and-down-movement of the body in front of the King.} \tbPD{goonɛɛ, goonɔ, goonɔ, goonema, goonaa, goonema }
\tbLX{goɔ}\tbHM{} \tbPH{gʊ́ɔ̀} \tbPS{n.} \tbGE{thorn} \tbOP{goore} \tbTP{goorɛɛ}
\tbLX{goɔ}\tbHM{} \tbPH{gʊ̀ɔ̀} \tbPS{v.} \tbGE{to bend at the trunk} \tbXV{Pɔgeba maŋ goɔ la kaabo kyɛ ŋma.} \tbXE{Women usually hold down guinea corn to cut the head.} \tbPD{goɔɛ, goɔrɔ, goɔrɔ, goɔrbɔ, goɔraa, goɔrbɔ }
\tbLX{goɔ}\tbHM{} \tbPH{gʊ̀ɔ̀} \tbPS{v.} \tbGE{to prevent from progressing; to cause to return or refrain from something} \tbPD{goɔɛ, goɔrɔ, goɔrɔ, goɔrbɔ, goɔraa, goɔrbɔ }
\tbLX{goɔ}\tbHM{} \tbPH{gʊ̀ɔ̀} \tbPS{v.} \tbGE{to bite (snake)} \tbPD{goɔɛ, goɔrɔ, goɔrɔ, goɔrbɔ, goɔraa, goɔrbɔ }
\tbLX{gõɔ}\tbHM{} \tbPH{gʊ̃́ɔ̃́} \tbPS{n.} \tbSN \tbGE{kiln (place for baking clay items to make them hard)} \tbSN \tbGE{arena, grouping point (ko-gõɔ)} \tbOP{goɔne} \tbTP{goɔnɛɛ}
\tbLX{gõɔo}\tbHM{} \tbPH{gʊ̃́ɔ̃̀ʊ̃́} \tbPS{n.} \tbGE{sleep} \tbOP{goɔnɛɛ}
\tbLX{goɔraa}\tbHM{} \tbPH{gʊ̀ɔ̀ràá} \tbPS{n.} \tbGE{a type of small tree with medicinal properties} \tbOP{goɔre} \tbTP{goɔrɛɛ}
\tbLX{goɔre}\tbHM{} \tbPH{gʊ́ɔ́rɪ́} \tbPS{v.} \tbGE{to walk or run half stooping} \tbPD{goɔrɛɛ, goɔrɔ, goɔrɔ, goɔreba, goɔraa, goɔreba }
\tbLX{goɔre}\tbHM{} \tbPH{gʊ̀ɔ̀rɪ̀} \tbPS{v.} \tbSN \tbGE{to go round and round} \tbSN \tbGE{to splash mashed ashes round the outer wall of a house (ritual to rid the house of activities of witchraft)} \tbPD{goɔrɛɛ, goɔrɔ, goɔrɔ, goɔreba, goɔraa}%, , -}
\tbLX{gora}\tbHM{} \tbPH{gʊ́rà} \tbPS{n.} \tbSN \tbGE{hunchback} \tbSN \tbGE{disease of the back} \tbOP{gorre} \tbTP{gorrɛɛ} \tbVA{gorɔ}
\tbLX{goraa}\tbHM{} \tbPH{gòràà} \tbPS{adj.} \tbGE{curved}
\tbLX{goraa}\tbHM{} \tbPH{gòràà} \tbPS{adj.} \tbGE{ashamed}
\tbLX{gore}\tbHM{} \tbPH{gʊ̀rɪ̀} \tbPS{v.} \tbSN \tbGE{to be abundant} \tbSN \tbGE{to be overburdened with weight} \tbPD{gorɛɛ, gora, gora, goreba, goraa}%, , -}
\tbLX{gore}\tbHM{} \tbPH{gʊ́rɪ́} \tbPS{v.} \tbGE{to pluck feathers or fur} \tbPD{gorɛɛ, gora, gora, gorba, goraa}%, , -}
\tbLX{gorgore}\tbHM{} \tbPH{gʊ̀rgʊ́rɪ́} \tbPS{v.} \tbGE{to ruffle} \tbPD{gorgorɛɛ, gorgora, gorgora, gorgorba, gorgoraa, gorgorba }
\tbLX{gori}\tbHM{} \tbPH{górí} \tbPS{n.} \tbGE{dowry, bride price} \tbOP{gorri} \tbTP{goree}
\tbLX{gori}\tbHM{} \tbPH{gòrì} \tbPS{v.} \tbGE{to bend an object (into a sharp curve)} \tbPD{goree, goro, goro, gorbo, goraa, gorbo }
\tbLX{gori}\tbHM{} \tbPH{górí} \tbPS{n.} \tbGE{big pile of; load} \tbOP{goe} \tbTP{goree}
\tbLX{gori-gori-gori}\tbHM{} \tbPH{gór-gór-gór} \tbPS{adv.} \tbGE{motionless}
\tbLX{gozambuuli}\tbHM{} \tbPH{gòzàmbúúlí} \tbPS{n.} \tbGE{the fruit of a type of thorn tree} \tbOP{gozambuulo} \tbTP{gozambuulee} \tbVA{gɔzambuuli}
\tbLX{gozaŋaa}\tbHM{} \tbPH{gózàŋàà} \tbPS{n.} \tbGE{a type of thorn tree with yellow fruit} \tbOP{gozanne} \tbTP{gozannɛɛ} \tbVA{gɔzaŋaa}
\tbLX{gɔbaa}\tbHM{} \tbPH{gɔ̀báà} \tbPS{n.} \tbSN \tbGE{skillful person; one full of dexterity} \tbSN \tbGE{artist (music, visual arts)} \tbOP{gɔɔre} \tbTP{gɔɔrɛɛ}
\tbLX{gɔbenɛ}\tbHM{} \tbPH{gɔ́bɪ̀nɛ̀} \tbPS{n.} \tbGE{governor, government} \tbOP{gɔbenne} \tbTP{gɔbenɛɛ}
\tbLX{gɔbenɛnte}\tbHM{} \tbPH{gɔ́bɪ́nɛ́ntɪ̀} \tbPS{n.} \tbGE{government} \tbOP{gɔbenɛntere} \tbTP{gɔbenɛnterɛɛ}
\tbLX{gɔfuuraa}\tbHM{} \tbPH{gɔ́fúúráá} \tbPS{n.} \tbGE{a type of tree with medicinal properties} \tbOP{gɔ\-fuu\-ri} \tbTP{gɔfuuree}
\tbLX{gɔge}\tbHM{} \tbPH{gɔ́gɪ́} \tbPS{v.} \tbGE{to nod} \tbPD{gɔgɛɛ, gɔgerɔ, gɔgerɔ, gɔgerebɔ, gɔgraa, gɔgerebɔ }
\tbLX{gɔgɔ}\tbHM{} \tbPH{gɔ́gɔ̀} \tbPS{n.} \tbSN \tbGE{a type of tree whose bark is used in tanning} \tbSN \tbGE{to type of black bird with a long tail} \tbOP{gɔgɔre} \tbTP{gɔgɔrɛɛ}
\tbLX{gɔgɔ}\tbHM{} \tbPH{gɔ̀gɔ́} \tbPS{v.} \tbGE{to bend halfway; to stoop} \tbPD{gɔgɔɛ, gɔgɔɔrɔ, gɔgɔɔrɔ, gɔgɔrebɔ, gɔgɔraa, gɔgɔrebɔ }
\tbLX{gɔllaa}\tbHM{} \tbPH{gɔ́lláá} \tbPS{n.} \tbGE{a type of small, fatty field mouse} \tbOP{gɔlle} \tbTP{gɔllɛɛ} \tbVA{mɔɔtere}
\tbLX{gɔlle}\tbHM{} \tbPH{gɔ̀llɪ̀} \tbPS{v.} \tbSN \tbGE{to go round, to detour} \tbSN \tbGE{to surround, to encircle} \tbPD{gɔllɛɛ, gɔllɔ, gɔllɔ, gɔlebɔ, gɔllaa, gɔlebɔ }
\tbLX{gɔlɔnteebo}\tbHM{} \tbPH{gɔ́lɔ́ntɪ̀ɪ̀bʊ́} \tbPS{n.} \tbGE{chameleon} \tbOP{gɔlɔnteere} \tbTP{gɔlɔnteerɛɛ} \tbVA{gɔmɔ/gɔmɔteebo}
\tbLX{gɔmberaa}\tbHM{} \tbPH{gɔ́mbíráá} \tbPS{n.} \tbGE{state of sleeplessness} \tbOP{gɔmberre} \tbTP{gɔmberɛɛ}
\tbLX{gɔmbulo}\tbHM{} \tbPH{gɔ̀mbùlò} \tbPS{n.} \tbGE{unique sound in poetry, sound that is pleasing to the ear}
\tbLX{gɔmɔ}\tbHM{} \tbPH{gɔ́mɔ̀} \tbPS{n.} \tbGE{noise, disturbance} \tbOP{gɔnne} \tbTP{gɔnnɛɛ} \tbVA{gɔnne}
\tbLX{gɔnda}\tbHM{} \tbPH{gɔ̌ndà} \tbPS{n.} \tbGE{pawpaw, papaya} \tbSG{gɔndaroo} \tbOP{gɔndare} \tbTP{gɔndarɛɛ}
\tbLX{gɔnne}\tbHM{} \tbPH{gɔ́nnɪ̀} \tbPS{n.} \tbGE{noise} \tbOP{gɔmɔ} \tbTP{gɔnnɛɛ}
\tbLX{gɔntɛre}\tbHM{} \tbPH{gɔ̀ntɛ́rɪ́} \tbPS{v.} \tbGE{to talk incessantly} \tbPD{gɔntɛrɛɛ, gɔnɔtɛrɛ, gɔnɔtɛrɛ, gɔntɛrebɛ}%, , -, -}
\tbLX{gɔntɛre}\tbHM{} \tbPH{gɔ̀ntɛ́rɪ́} \tbPS{v.} \tbGE{to burst out, especially when angry} \tbPD{gɔntɛrɛɛ, gɔnɔtɛrɛ, gɔnɔtɛrɛ, gɔntɛrebɛ}%, , -, -}
\tbLX{gɔnzuuro}\tbHM{} \tbPH{gɔ́nzúúrò} \tbPS{n.} \tbGE{sleepiness} \tbXV{O taa la gɔnzuuro.} \tbXE{He/she is feeling sleepy.} \tbTP{gɔnzuuree}
\tbLX{gɔŋe}\tbHM{} \tbPH{gɔ́ŋ} \tbPS{adj.} \tbGE{crooked, bent}
\tbLX{gɔŋe}\tbHM{} \tbPH{gɔ̀ŋ} \tbPS{v.} \tbGE{to make crooked, to incline toward} \tbPD{gɔŋɛɛ, gɔnnɔ, gɔnnɔ, gɔnema, gɔnnaa, gɔnema }
\tbLX{gɔŋe}\tbHM{} \tbPH{gɔ́ŋ} \tbPS{v.} \tbGE{to make noise, to quarrel} \tbPD{gɔŋɛɛ, gɔnnɔ, gɔnɔ, gɔnemɔ, gɔnnaa, gɔnemɔ }
\tbLX{gɔŋ-eŋ}\tbHM{} \tbPH{gɔ̀ŋ-ɪ̀ŋ} \tbPS{v.} \tbGE{to redirect or state to suit one's purpose} \tbXV{A yɛlɛ ba waa lɛ a pɔge gɔŋ-eŋɛɛ la.} \tbXE{That is not the true state of the matter; the woman has tried to add something to make it look like that.} \tbPD{gɔŋ-eŋɛɛ, gɔnnɔ-ennɛ, gɔŋ-ennɛ, gɔŋ-ennaa, gɔŋ-enemɛ}%, , -}
\tbLX{gɔŋgɔlɔŋ}\tbHM{} \tbPH{gɔ̀ŋgɔ̀lɔ̀ŋ} \tbPS{adj.} \tbGE{long and winding}
\tbLX{gɔŋgɔŋe}\tbHM{} \tbPH{gɔ̀ŋgɔ́ŋ} \tbPS{v.} \tbGE{to walk wearily and unsteadily; to trudge} \tbPD{gɔŋgɔŋɛɛ, gɔŋgɔnɔ, gɔŋgɔnɔ, gɔŋgɔnema, gɔŋgɔnaa, gɔŋgɔnema }
\tbLX{gɔŋgbire}\tbHM{} \tbPH{gɔ́ŋgbírè} \tbPS{n.} \tbGE{one who is asleep} \tbOP{gɔŋgbiribe} \tbTP{gɔŋ\-gbi\-ri\-bee}
\tbLX{gɔŋɔ}\tbHM{} \tbPH{gɔ́ŋɔ́} \tbPS{n.} \tbGE{type of guinea corn} \tbTP{gɔnnɛɛ}
\tbLX{gɔɔ}\tbHM{} \tbPH{gɔ̀ɔ́} \tbPS{n.} \tbGE{left hand, left side} \tbOP{gɔɔe} \tbTP{gɔɔrɛɛ}
\tbLX{gɔɔ}\tbHM{} \tbPH{gɔ̀ɔ́} \tbPS{adj.} \tbGE{left-handed} \tbOP{gɔɔe} \tbTP{gɔɔrɛɛ}
\tbLX{gɔɔ}\tbHM{} \tbPH{gɔ̀ɔ̀} \tbPS{v.} \tbGE{to go round and round} \tbPD{gɔɔɛ, gɔɔrɔ, gɔɔrɔ, gɔɔrebɔ, gɔɔraa, gɔɔrebɔ } \tbVA{goɔre}
\tbLX{gɔɔrombiri}\tbHM{} \tbPH{gɔ́ɔ́rʊ́mbírí} \tbPS{n.} \tbGE{literary device (lit.)} \tbSG{gɔɔrombiruu} \tbOP{gɔɔrombie} \tbTP{gɔɔrombiree}
\tbLX{gɔpelaa}\tbHM{} \tbPH{gɔ̀pɪ̀làá} \tbPS{n.} \tbGE{a type of tree with medicinal properties} \tbOP{gɔpeɛle} \tbTP{gɔpeɛlɛɛ}
\tbLX{gɔre}\tbHM{} \tbPH{gɔ̀rɪ̀} \tbPS{v.} \tbSN \tbGE{to bend} \tbSN \tbGE{to mould a hoe into a curved shape} \tbPD{gɔrɛɛ, gɔrɔ, gɔrɔ, gɔrebɔ, gɔraa, gɔrebɔ }
\tbLX{gɔre}\tbHM{} \tbPH{gɔ́rɪ́} \tbPS{n.} \tbGE{scythe} \tbOP{gɔrre} \tbTP{gɔrɛɛ}
\tbLX{gɔsɔgelaa}\tbHM{} \tbPH{gɔ́sɔ́glàá} \tbPS{n.} \tbGE{a type of thorn tree} \tbOP{gɔsɔgelɔ} \tbTP{gɔsɔgelɛɛ}
\tbLX{gɔteɛ}\tbHM{} \tbPH{gɔ̂tɪ̀ɛ́} \tbPS{n.} \tbGE{a tree with thorns} \tbOP{gɔteere} \tbTP{gɔteerɛɛ}
\tbLX{gɔzeɛ}\tbHM{} \tbPH{gɔ̀zɪ̀ɛ́} \tbPS{n.} \tbGE{a type of tree with medicinal properties}
\tbLX{gu}\tbHM{} \tbPH{gù} \tbPS{v.} \tbSN \tbGE{to watch out for} \tbSN \tbGE{to look after} \tbSN \tbGE{to protect, to defend} \tbSN \tbGE{to wait for (someone),} \tbSN \tbGE{to waylay (someone)} \tbPD{gue, guuro, guu\-ro, guribo, guuraa, guribo } \tbVA{go}\tbSE{gu zɔɔ} \tbGE{to ambush}
\tbLX{gufu}\tbHM{} \tbPH{gúfù} \tbPS{n.} \tbGE{interpreter} \tbOP{gufuri} \tbTP{gufuree/gufurree}
\tbLX{Gumo}\tbHM{} \tbPH{gúmó} \tbPS{n.} \tbGE{a town in the Sisaalaland noted for soothsaying and divining} \tbTP{gumee}
\tbLX{gunni}\tbHM{} \tbPH{gúnní} \tbPS{v.} \tbGE{to stalk (hunter tracking game)} \tbPD{gunnee, gunno, gunno, gunimo, gunnaa}%, , -}
\tbLX{gunni}\tbHM{} \tbPH{gùnnì} \tbPS{v.} \tbGE{to pick tiny leaves from plants} \tbPD{gunnee, gunno, gunno, gunimo, gunnaa}%, , -}
\tbLX{guo}\tbHM{} \tbPH{gùò} \tbPS{n.} \tbGE{heavy club with knobbed end} \tbOP{gori} \tbTP{goree} \tbVA{daguo}
\tbLX{guoli}\tbHM{} \tbPH{gùòlì} \tbPS{v.} \tbGE{to coil (rope)} \tbPD{guolee, guolo, guolo, guolibo, guolaa, guolibo }
\tbLX{guoli}\tbHM{} \tbPH{gúólì} \tbPS{v.} \tbSN \tbGE{to bring up (child)} \tbSN \tbGE{to raise (animals)} \tbPD{guolee, guolo, guolo, guolibo, guolaa, guolibo }
\tbLX{guoli}\tbHM{} \tbPH{gúólí} \tbPS{n.} \tbGE{pot rest; pad on which pots are placed} \tbOP{guolo} \tbTP{guolee}
\tbLX{guori}\tbHM{} \tbPH{gúórí} \tbPS{n.} \tbGE{kola nuts} \tbSG{guoruu} \tbOP{goe} \tbTP{goree}
\tbLX{guri}\tbHM{} \tbPH{gùri} \tbPS{v.} \tbSN \tbGE{to throw one's body onto another (wrestling, catching a fowl)} \tbSN \tbGE{to fold up (leg)} \tbPD{guree, guro, guro, guribo, guraa}%, , -}
\tbSE{guri pɔge}\tbHM{}  \tbGE{to rape} %\tbPD{-, -, -, -, -, -}
%\tbLX{gurri}\tbHM{} \tbPH{gúrrí} \tbPS{n.} \tbGE{mushroom} \tbSG{gurruu} \tbTP{gurree}
\tbLX{guruŋ}\tbHM{} \tbPH{gùrúŋ} \tbPS{n.} \tbGE{mushroom} \tbSG{gurruu} \tbOP{gurri} \tbTP{gurree}
\tbLX{guu}\tbHM{} \tbPH{gúú} \tbPS{v.} \tbGE{to be unable, to fail} \tbPD{guue, guuro, guuro, guuribo, guuraa, guuribo }
\tbLX{guu}\tbHM{} \tbPH{gùú} \tbPS{adj.} \tbGE{blunt, not sharp} \tbOP{guuri} \tbTP{guuree}
\tbLX{guu}\tbHM{} \tbPH{gùù} \tbPS{v.} \tbGE{to be blunt} \tbPD{guue, guuro, guuro, guurbo, guuraa, guurbo }
\tbLX{guuli}\tbHM{} \tbPH{gúúlì} \tbPS{v.} \tbGE{to delay someone} \tbPD{guu\-lee, guulo, guulo, guulibo, guulaa, guulibo } \tbVA{gbuuli}
\tbLX{guuli}\tbHM{} \tbPH{gúúlì} \tbPS{v.} \tbGE{to be not yet healed of a sprain} \tbPD{guulee, guulo, guulo, guulibo, guulaa, guulibo } \tbVA{gbuuli}
\tbLX{guuli}\tbHM{} \tbPH{gúúlì} \tbPS{v.} \tbGE{to delay someone} \tbPD{guulee, guulo, guulo, guulibo, guulaa, guulibo } \tbVA{gbuuli}
\tbLX{guuli}\tbHM{} \tbPH{gúúlí} \tbPS{n.} \tbGE{knot} \tbOP{guulo} \tbTP{guulee}
%\tbLX{gu zɔɔ}\tbHM{} \tbPH{gù zɔ̀ɔ́} \tbPS{v.} \tbGE{to ambush} \tbPD{gue  zɔɔ, guuro zɔɔ, zɔɔ guuro}%, , -, -, -}
\end{letter}
