\begin{letter}{u}
\tbLX{ugi}\tbHM{} \tbPH{úgì} \tbPS{v.} \tbGE{to expand e.g. cooked rice} \tbPD{ugee, ugro, ugro, ugribo, ugraa}%, , -}
\tbLX{ulee}\tbHM{} \tbPH{ùléè} \tbPS{n.} \tbGE{tree branch} \tbOP{ulli} \tbTP{ullee}
\tbLX{uli}\tbHM{} \tbPH{úlí} \tbPS{n.} \tbGE{fog, mist} \tbOP{ulli} \tbTP{ullee}
\tbLX{uli}\tbHM{} \tbPH{ùlì} \tbPS{v.} \tbSN \tbGE{to be dusty} \tbSN \tbGE{to become moldy e.g. food} \tbXV{Ka boroboroŋ wa kɔɔre o maŋ uli la.} \tbXE{When bread is kept for long it becomes moldy.} \tbPD{ulee, ullo, ullo, ulibo, ullaa}%, , -}
\tbLX{ulimo}\tbHM{} \tbPH{úlmó} \tbPS{adj.} \tbGE{grey}
\tbLX{unni}\tbHM{} \tbPH{ùnnì} \tbPS{v.} \tbSN \tbGE{unburying something buried} \tbXV{Polisiri yeli ka ba ba baŋ a dɔɔ kũũ, pare a gaa te unni a kũũ.} \tbXE{The police said they did not know the cause of the man's death, so they went and exhumed the corpse.} \tbSN \tbGE{to bring back past memories} \tbXV{Ka Dɛre ne o pɔgeŋ wa zɔɔrɔ zaa a pɔge maŋ unni la yelikorɔ.} \tbXE{Each time Dɛre and his wife quarrel, the wife will always recount past events.} \tbPD{unnee, unno, unno, unimo, unnaa}%, , -}
\tbLX{unni}\tbHM{} \tbPH{únnì} \tbPS{v.} \tbGE{to remind} \tbXV{N deɛmɛ zãã wa unni ma la a koɔbo.} \tbXE{My in-law reminded me about the day of the group farming.} \tbPD{unnee, unno, unno, unimo, unnaa}%, , -}
\tbLX{uoni}\tbHM{} \tbPH{úònì} \tbPS{n.} \tbGE{dry season} \tbOP{uomo} \tbTP{uonee}
\tbLX{uori}\tbHM{} \tbPH{úórì} \tbPS{v.} \tbGE{to warm oneself (by fire, sun)} \tbPD{uoree, uoro, uoro, uoribo, uoraa}%, , -}
\tbLX{uori}\tbHM{} \tbPH{úórì} \tbPS{v.} \tbGE{to mourn} \tbPD{uoree, uoro, uoro, uoribo, uoraa}%, , -}
\tbLX{uori}\tbHM{} \tbPH{úórì} \tbPS{v.} \tbGE{to peer (shading one's face with the hand)} \tbPD{uoree, uoro, uoro, uoribo, uoraa}%, , -}
\tbLX{uori}\tbHM{} \tbPH{ùòrì} \tbPS{v.} \tbSN \tbGE{to bellow (bull)} \tbSN \tbGE{to groan; a sound uttered in pain or grief} \tbPD{uoree, uoro, uoro, uoribo, uoraa}%, , -}
\tbLX{uri}\tbHM{} \tbPH{ùrí} \tbPS{n.} \tbSN \tbGE{itchy stuff} \tbSN \tbGE{chaff}  \tbTP{urree}
\tbLX{uri}\tbHM{} \tbPH{ùrì} \tbPS{v.} \tbSN \tbGE{to take unaware} \tbSN \tbGE{to be shocked} \tbXV{Maa ne a baala naŋ yeli yɛlɛ ba kɔɔre, o kũũ uri ma la.} \tbXE{I spoke with the patient not long ago; I'm shocked at his sudden death.} \tbPD{uree, uro, uro, uribo, uraa}%, , -}
\tbLX{uri}\tbHM{} \tbPH{úrì} \tbPS{v.} \tbGE{to make burn e.g. fire} %\tbPD{-, -, -, -, -, -}
\tbLX{uri}\tbHM{} \tbPH{ùrí} \tbPS{n.} \tbGE{hair-like projections on the straw of the grass family and other plants that cause itchiness of the skin}
\tbLX{uri}\tbHM{} \tbPH{úrí} \tbPS{v.} \tbGE{to scrub a very dirty object e.g. cooking pot} \tbPD{uree, uro, uro, uribo, uraa}%, , -}
\tbLX{ũu}\tbHM{} \tbPH{ṹṹ} \tbPS{v.} \tbGE{to close tightly (fist, eyes)} \tbPD{ũue, uuno, uuno, uuimo, uunaa}%, , -}
\tbLX{ũu}\tbHM{} \tbPH{ũ̀ũ̀} \tbPS{v.} \tbGE{to bury} \tbPD{ũue, uuno, uuno, uuimo, uunaa}%, , -}
\tbLX{uuni}\tbHM{} \tbPH{úúnì} \tbPS{v.} \tbSN \tbGE{to crouch, to hide} \tbXV{Ba digi la a bie ka o zo te uuni tuu poɔ.} \tbXE{As they chased the boy, he ran and hid in a thicket.} \tbSN \tbGE{to incubate, to brood, to sit on (a fowl on its eggs)} \tbXV{Noɔ maŋ uuni la daare ata kyɛ wɛge.} \tbXE{A fowl sits on its eggs before they hatch.} \tbPD{uunee, uuno, uuno, uunimo, uunaa}%, , -}
\tbLX{uuni}\tbHM{} \tbPH{úúnì} \tbPS{v.} \tbGE{to form pus e.g. a swelling} \tbXV{Goɔ daare kyɔge la n nubiri ka o mɔre eŋ kõɔ ka ba wɛre.} \tbXE{A thorn pierced my finger and it got swollen, formed pus, and was operated upon.} \tbPD{uunee, uuno, uuno, uunimo, uunaa}%, , -}
\tbLX{uuri}\tbHM{} \tbPH{úúrí} \tbPS{v.} \tbGE{to hoot, to call out e.g. lost person} \tbPD{uuree, uuro, uuro, uuribo, uuraa}%, , -}
\tbLX{uuri}\tbHM{} \tbPH{ùùrì} \tbPS{v.} \tbSN \tbGE{to smoke out} \tbSN \tbGE{to warm a place with smoke} \tbPD{uuree, uuro, uuro, uuribo, uuraa}%, , -}
\tbLX{uuruŋ}\tbHM{} \tbPH{úúrúŋ} \tbPS{n.} \tbGE{dust} \tbTP{uuurunnee}
\end{letter}
