\begin{letter}{v}
\tbLX{va}\tbHM{} \tbPH{và} \tbPS{v.} \tbSN \tbGE{to whip} \tbSN \tbGE{to thresh (millet)} \tbPD{vaɛ, -, vaara, vareba, vaaraa}%, , -}
\tbLX{va}\tbHM{} \tbPH{vá} \tbPS{v.} \tbGE{snatch; to run directly and suddenly at someone seizing him; used with `-de’ (va-de) } %\tbPD{-, -, -, -, -, -}
\tbLX{va}\tbHM{} \tbPH{và} \tbPS{v.} \tbGE{to graze a target; miss narrowly} \tbPD{vaɛ, -, vaara, vareba, vaaraa}%, , -}
\tbLX{va}\tbHM{} \tbPH{vá} \tbPS{v.} \tbGE{to rush, to dash} \tbXV{A dɔbɔ naŋ ŋmeɛrɛ la ka kaŋa wa va de o tɔ gaa lɔɔ.} \tbXE{As the men argued, one suddenly rushed on the other and threw him down.} \tbPD{vaɛ, vaara, vaara, vareba, vaaraa}%, , -}
\tbLX{vaa}\tbHM{} \tbPH{váà} \tbPS{n.} \tbGE{leaf} \tbSG{vabo} \tbOP{vaare} \tbTP{vaarɛɛ}
\tbLX{vaa}\tbHM{} \tbPH{vàà} \tbPS{v.} \tbGE{collect or take everything from a container} \tbPD{vaaɛ, vaara, vaara, vaareba, vaaraa}%, , -}
\tbLX{vaa}\tbHM{} \tbPH{vàà} \tbPS{v.} \tbGE{to draw e.g. bowstring} \tbPD{vaaɛ, vaara, vaara, vaareba, vaaraa}%, , -}
\tbLX{vaa}\tbHM{} \tbPH{vàà} \tbPS{v.} \tbGE{to lift an opponent's leg (wrestling)} \tbPD{vaaɛ, vaara, vaara, vaareba, vaaraa}%, , -}
\tbLX{vaagaviige}\tbHM{} \tbPH{vààgàvíígè} \tbPS{v.} \tbGE{I command you (to die)} %\tbPD{-, -, -, -, -, -}
\tbLX{vaa-iri}\tbHM{} \tbPH{vàà-ìrì} \tbPS{v.} \tbGE{to stand up quickly} \tbPD{vaa-iree, vaara-ire, vaa-ire, vaa-iribe}%, , -, -}
\tbLX{vaale}\tbHM{} \tbPH{vààlɪ̀} \tbPS{v.} \tbGE{to gather e.g. rubbish, garbage} \tbPD{vaalɛɛ,vaala, vaala, vaaleba,vaalebɛɛ vaaloo}
\tbLX{vaalee}\tbHM{} \tbPH{váálɪ́ɪ́} \tbPS{n.} \tbGE{speck} \tbSG{vaaloo} \tbOP{vaala} \tbTP{vaalɛɛ} \tbVA{vaale}
\tbLX{vaali}\tbHM{} \tbPH{vààlì} \tbPS{adv.} \tbGE{quickly, in a fast and sudden manner, e.g.~in standing up} %\tbPD{-, -, -, -, -, -}
\tbLX{vaane}\tbHM{} \tbPH{vàànɪ̀} \tbPS{v.} \tbSN \tbGE{to shrink, contract} \tbSN \tbGE{to wrinkle} \tbPD{vaanɛɛ, vaana, vaana, vaaneba, vaanaa}%, , -}
\tbLX{vaane}\tbHM{} \tbPH{váánɪ́} \tbPS{adj.} \tbGE{shrunken}
\tbLX{vãao}\tbHM{} \tbPH{vã̀ã̀ʊ̃̀} \tbPS{adv.} \tbGE{courageously}
\tbLX{vaare}\tbHM{} \tbPH{vààrɪ̀} \tbPS{v.} \tbGE{to wrap oneself, usually with a cloth} \tbPD{vaarɛɛ, vaara, vaara, vaareba, vaaraa}%, , -}
\tbLX{vaare}\tbHM{} \tbPH{váárɪ̀} \tbPS{v.} \tbGE{to tug, to pull strongly and suddenly towards oneself} \tbXV{A bie da nyɔge la a pedaa miri kyɛ ka a peroo vaare o lɔɔ.} \tbXE{The boy held onto the rope around the ram's neck, but it tugged him to the ground.} \tbPD{vaarɛɛ, vaara, vaara, vaareba, vaaraa}%, , -}
\tbLX{vaare}\tbHM{} \tbPH{váárɪ̀} \tbPS{v.} \tbGE{to rinse} \tbXV{Fanfane fuuri naŋ be la laare poɔ a seŋ ka vaare.} \tbXE{There are still soapsuds in the bowls, so they need to be rinsed.} \tbPD{vaarɛɛ, vaara, vaara, vaareba, vaaraa}%, , -}
\tbLX{vabɔre}\tbHM{} \tbPH{vábɔ́rɪ̀} \tbPS{n.} \tbGE{spring} \tbTP{vabɔrrɛɛ}
\tbLX{vãee}\tbHM{} \tbPH{vã̀ɪ̃́ɪ̃̀} \tbPS{adv.} \tbGE{aromatically}
\tbLX{vage}\tbHM{} \tbPH{vágɪ́} \tbPS{v.} \tbGE{sounding like a bird pecking at a soft object (onomatopoeia)} \tbPD{vagɛɛ, vagra, vagra, vagreba, vagraa}%, , -}
\tbLX{vage}\tbHM{} \tbPH{vàgɪ̀} \tbPS{adv.} \tbGE{sound of pecking}
\tbLX{vagere}\tbHM{} \tbPH{vàgɪ̀rɪ̀} \tbPS{adv.} \tbGE{pecking}
\tbLX{vakogilo}\tbHM{} \tbPH{vákóglo} \tbPS{n.} \tbGE{dried leaves (that once were worn by women)} \tbSG{vakogiluu}
\tbLX{vakuoŋaa}\tbHM{} \tbPH{vákúóŋáá} \tbPS{n.} \tbGE{dry leaf} \tbOP{vakuoni}
\tbLX{vakuoŋaa}\tbHM{} \tbPH{vákúóŋáá} \tbPS{n.} \tbGE{bad luck} \tbOP{vakuoni}
\tbLX{valenvõɔ}\tbHM{} \tbPH{válɪ́nvʊ̃́ɔ̃̀} \tbPS{n.} \tbGE{mason wasp, potter wasp} \tbSG{valenvoɔnoo} \tbOP{valenvoɔne} \tbTP{valenvoɔnɛɛ} \tbVA{valenvõa}
\tbLX{valere}\tbHM{} \tbPH{vàlɪ́rɪ́} \tbPS{v.} \tbGE{to intertwine e.g. weaving} %\tbPD{-, -, -, -, -, -}
\tbLX{valimmu}\tbHM{} \tbPH{vàlímmú} \tbPS{n.} \tbGE{a type of small tree with soft large leaves which women used to wear to cover parts of their private parts}  \tbOP{valinni} \tbTP{valinnee}
\tbLX{valinvaa}\tbHM{} \tbPH{vàlìnv́áà} \tbPS{n.} \tbSN \tbGE{leaf of the palm tree} \tbSN \tbGE{the leaf of the valimmu tree} \tbOP{valinvaare} \tbTP{va\-lin\-vaa\-rɛɛ}
\tbLX{valinvuuluŋ}\tbHM{} \tbPH{v̀à\-lìn\-vúú\-lúŋ} \tbPS{n.} \tbGE{soot} \tbOP{valinvuulo/valinvuuli} \tbTP{va\-lin\-vuu\-lee} \tbVA{valanvuuluŋ, vuuluŋ}
\tbLX{valogema}\tbHM{} \tbPH{válʊ́gɪ́mà} \tbPS{n.} \tbGE{green, fresh, and tender leaves}
\tbLX{vama}\tbHM{} \tbPH{vàmà} \tbPS{adv.} \tbGE{effortlessly}
\tbLX{vamaaroŋ}\tbHM{} \tbPH{vámáárʊ́ŋ} \tbPS{n.} \tbGE{green leaf} \tbOP{vamaara} \tbTP{vamaarɛɛ}
\tbLX{vanne}\tbHM{} \tbPH{vánnɪ̀} \tbPS{v.} \tbGE{to stretch, to grow longer} \tbPD{vannɛɛ, vanna, vanna, vannaa}%, , -, -}
\tbLX{vapeɛlegboro}\tbHM{} \tbPH{vápíɛ́lgbóró} \tbPS{n.} \tbGE{a type of tree with medicinal properties}
\tbLX{vapɛge}\tbHM{} \tbPH{vápɛ́gɪ́} \tbPS{n.} \tbGE{fallen leaf} \tbOP{vapɛgre} \tbTP{vapgrɛɛ}
\tbLX{vapielee}\tbHM{} \tbPH{vápìèlèè} \tbPS{n.} \tbGE{a type of tree with medicinal properties} \tbOP{vapieli}
\tbLX{vapɔrɔ}\tbHM{} \tbPH{vápɔ̀rɔ̀} \tbPS{n.} \tbGE{fresh green leaves} \tbTP{vapɔrrɛɛ}
\tbLX{vare}\tbHM{} \tbPH{vàrɪ̀} \tbPS{v.} \tbGE{to wrinkle} \tbSE{vare niŋe} \tbGE{make a face; usually in disgust or anger}
\tbPD{varɛɛ, vara, vara, vareba, varaa, vareba }
\tbLX{vare}\tbHM{} \tbPH{vàrɪ̀} \tbPS{v.} \tbGE{to make laundry not well washed e.g. clothes} \tbPD{varɛɛ, vara, vara, vareba, varaa, vareba }
\tbLX{vare}\tbHM{} \tbPH{vàrɪ̀} \tbPS{v.} \tbGE{to make the layout of a mound} \tbPD{varɛɛ, vara, vara, vareba, varaa, vareba }
\tbLX{vare}\tbHM{} \tbPH{várɪ́} \tbPS{v.} \tbGE{to jump, to leap, to hop} \tbPD{varoo, vara, vara, varibo, varaa, variba }
\tbLX{vare}\tbHM{} \tbPH{várɪ́} \tbPS{v.} \tbGE{to castrate} \tbPD{varoo, vara, vara, varibo, varaa, variba }
\tbLX{varekpara}\tbHM{} \tbPH{vàrìkpàrà} \tbPS{adv.} \tbGE{long and loose e.g. twine}
\tbLX{varekpenɛ}\tbHM{} \tbPH{várɪ́kpɪ́nɛ́} \tbPS{n.} \tbGE{thick paste from beans} \tbSG{varekpenoo} \tbTP{varekpennɛɛ}
\tbLX{varema}\tbHM{} \tbPH{várɪ́má} \tbPS{adj.} \tbSN \tbGE{tough e.g. meat, rope} \tbSN \tbGE{difficult} \tbXV{A pɔgelee na nɔŋe la laare kyɛ o boɔbo yɛlɛ varema.} \tbXE{That girl likes smiling but she is very difficult when it comes to dating her.}

\tbLX{varevare}\tbHM{} \tbPH{vàrɪ̀várɪ́} \tbPS{v.} \tbGE{to heckle} \tbPD{varevarɛɛ, varevara, varevara, varevareba, varevaraa}%, , -}
\tbLX{varevare}\tbHM{} \tbPH{vàrɪ̀várɪ́} \tbPS{v.} \tbGE{to pull along} \tbPD{varevarɛɛ, varevara, varevara, varevareba, varevaraa}%, , -}
\tbLX{varoo}\tbHM{} \tbPH{várʊ̀ʊ̀} \tbPS{adj.} \tbGE{tough e.g. meat} \tbVA{varema}
\tbLX{ve}\tbHM{} \tbPH{vɪ́} \tbPS{v.} \tbGE{to knit} \tbPD{veɛɛ, verɛ, verɛ, verebɛ, veraa}%, , -}
\tbLX{vebo}\tbHM{} \tbPH{vɪ̀bʊ̀} \tbPS{adj.} \tbGE{striking directly e.g. arrow}
\tbLX{vedeɛ}\tbHM{} \tbPH{vɪ̀dɪ̀ɛ́} \tbPS{n.} \tbGE{sugarcane} \tbOP{vederre} \tbTP{vedeerɛɛ}
\tbLX{veene}\tbHM{} \tbPH{vɪ̀ɪ̀nɪ̀} \tbPS{n.} \tbGE{a type of small bird with very lean legs (short form of daveene)} \tbOP{veemɛ} \tbTP{veenɛɛ} \tbVA{daveene}
\tbLX{veene}\tbHM{} \tbPH{vɪ́ɪ́nɪ́} \tbPS{v.} \tbGE{to tie tightly} \tbPD{veenɛɛ, veenɛ, veenɛ, veenemɛ, veenaa}%, , -}
\tbLX{veɛle}\tbHM{} \tbPH{vɪ̀ɛ̀lɪ̀} \tbPS{v.} \tbGE{to be beautiful} \tbPD{veɛlɛɛ, veɛlɛ, veɛlɛ, veɛle}%, , -, -}
\tbLX{veɛlɛ}\tbHM{} \tbPH{vɪ̀ɛ̀lɛ́} \tbPS{adj.} \tbGE{beautiful}
\tbLX{veɛlɛyeŋe}\tbHM{} \tbPH{vɪ̀ɛ̀lɛ́yéŋé} \tbPS{n.} \tbGE{one with physical beauty only}
\tbLX{veɛloŋ}\tbHM{} \tbPH{vɪ̀ɛ́lʊ́ŋ} \tbPS{n.} \tbSN \tbGE{beauty} \tbSN \tbGE{goodness} \tbSN \tbGE{light}
\tbLX{veɛre}\tbHM{} \tbPH{vɪ̀ɛ̀rɪ̀} \tbPS{v.} \tbGE{to push aside growth e.g. in searching for something in an overgrown bush} \tbPD{veɛrɛɛ, veɛrɛ, veɛrɛ, veɛrebɛ, veɛraa}%, , -}
\tbLX{veɛre}\tbHM{} \tbPH{vɪ́ɛ́rɪ̀} \tbPS{v.} \tbGE{to fly low and slowly} \tbPD{veɛrɛɛ, veɛrɛ, veɛrɛ, veɛrebɛ, veɛraa}%, , -}
\tbLX{veɛre}\tbHM{} \tbPH{vɪ́ɛ́rɪ̀} \tbPS{v.} \tbGE{to evade e.g. a verbal attack} \tbPD{veɛrɛɛ, veɛrɛ, veɛrɛ, veɛrebɛ, veɛraa}%, , -}
\tbLX{veɛrenyɔge}\tbHM{} \tbPH{vɪ̀ɛ̀rɪ̀ɲɔ́gɪ́} \tbPS{n.} \tbGE{rodent}
\tbLX{velaa}\tbHM{} \tbPH{vɪ̀làà} \tbPS{adj.} \tbSN \tbGE{good, nice} \tbSN \tbGE{beautiful} \tbOP{veɛle}
\tbLX{vell}\tbHM{} \tbPH{vɪ̀ll} \tbPS{adj.} \tbGE{full to the brim}
\tbLX{velle}\tbHM{} \tbPH{vɪ̀llɪ̀} \tbPS{v.} \tbGE{to hold the legs tightly in such a way as to make the person lose balance} \tbXV{A pɔge velle la a dɔɔ gbɛɛ ka o le.} \tbXE{The woman held the man’s legs so tightly that he lost balance and fell.} \tbPD{vellɛɛ, vellɛ, vellɛ, velbɛ, vellaa}%, , -}
\tbLX{velɔɔ}\tbHM{} \tbPH{vɪ̀lɔ̀ɔ̀} \tbPS{adj.} \tbGE{turning instantly}
\tbLX{venenne}\tbHM{} \tbPH{vɪ̀nɪ̀nnɪ̀} \tbPS{adj.} \tbGE{elastic}
\tbLX{venne}\tbHM{} \tbPH{vɪ́nnɪ́} \tbPS{v.} \tbGE{to throw off (wrestling)} \tbPD{vennɛɛ, vennɛ, vennɛ, venemɛ, vennaa}%, , -}
\tbLX{venoo}\tbHM{} \tbPH{vɪ́nʊ̀ʊ̀} \tbPS{n.} \tbGE{wine}
\tbLX{venɔɔ}\tbHM{} \tbPH{vɪ̀nɔ̀ɔ̀} \tbPS{adj.} \tbGE{thick and sticky e.g. grease}
\tbLX{venveŋ}\tbHM{} \tbPH{vɪ́nvɪ́ŋ} \tbPS{n.} \tbGE{climax}
\tbLX{venveŋ}\tbHM{} \tbPH{vɪ́nvɪ́ŋ} \tbPS{adv.} \tbGE{at the maximum} \tbXV{A daa naŋ da w ate kpɛ venveŋ saŋa la ka ba da moɔle a zuyɔɔ yɛlɛ.} \tbXE{It was at the climax of the market activities that they announced the payment of taxes.}
\tbLX{vere}\tbHM{} \tbPH{vɪ̀rɪ̀} \tbPS{v.} \tbGE{to shoot with a catapult} \tbPD{verɛɛ, verɛ, verɛ, verebɛ, veraa, verebɛ }
\tbLX{vere}\tbHM{} \tbPH{vɪ̀rɪ̀} \tbPS{v.} \tbSN \tbGE{to sprain} \tbSN \tbGE{to dislocate (joint)} \tbPD{verɛɛ, verɛ, verɛ, verebɛ, veraa, verebɛ }
\tbLX{vere}\tbHM{} \tbPH{vɪ̀rɪ̀} \tbPS{v.} \tbGE{to make an effort, to strain} \tbPD{verɛɛ, verɛ, verɛ, verebɛ, veraa, verebɛ }
\tbLX{vere}\tbHM{} \tbPH{vɪ́rɪ́} \tbPS{v.} \tbGE{to fence e.g. garden} \tbPD{verɛɛ, verɛ, verɛ, verebɛ, veraa}%, , -}
\tbLX{verevere}\tbHM{} \tbPH{vɪ̀rɪ̀vɪ́rɪ́} \tbPS{v.} \tbGE{to pull very hard on e.g. tug of war} \tbPD{vereverɛɛ, vereverɛ, vereverɛ, vereverebɛ, vereveraa}%, , -}
\tbLX{verɛnverɛŋ}\tbHM{} \tbPH{vɪ̀rɛ̀nvɪ́rɛ́ŋ} \tbPS{adj.} \tbGE{found everywhere}
\tbLX{verɛverɛ}\tbHM{} \tbPH{vɪ̀rɛ̀vɪ̀rɛ̀} \tbPS{adj.} \tbGE{scattered all over}
\tbLX{vɛ̃ɛ}\tbHM{} \tbPH{vɛ̃̀ɛ̃̀} \tbPS{adv.} \tbGE{less sunny, dusky}
\tbLX{vɛle}\tbHM{} \tbPH{vɛ̀lɪ̀} \tbPS{v.} \tbGE{to become clear e.g. day, mind} \tbPD{vɛlɛɛ, vɛllɛ, vɛllɛ, vɛlebɛ, vɛllaa}%, , -}
\tbLX{vɛlevɛle}\tbHM{} \tbPH{vɛ̀lvɛ̀l} \tbPS{adj.} \tbGE{very long and and wavy}
\tbLX{vɛmm}\tbHM{} \tbPH{vɛ́mm} \tbPS{adv.} \tbGE{smoothly}
\tbLX{vɛnɛvɛnɛ}\tbHM{} \tbPH{vɛ̀nɛ̀vɛ̀nɛ̀} \tbPS{adj.} \tbSN \tbGE{sticky e.g. honey} \tbSN \tbGE{oily}
\tbLX{vɛnlɛmm}\tbHM{} \tbPH{vɛ́nlɛ́mm} \tbPS{adj.} \tbGE{slim and beautiful}
\tbLX{vɛŋe}\tbHM{} \tbPH{vɛ́ŋ} \tbPS{v.} \tbSN \tbGE{to leave} \tbSN \tbGE{to let go} \tbSN \tbGE{to stop} \tbPD{vɛŋɛɛ, vɛnɛ, vɛnɛ, vɛnemɛ, vɛnaa}%, , -}
\tbLX{vɛre}\tbHM{} \tbPH{vɛ̀rɪ̀} \tbPS{v.} \tbGE{to prepare mortar} \tbPD{vɛrɛɛ, vɛrɛ, vɛrɛ, vɛrebɛ, vɛraa}%, , -}
\tbLX{vɛre}\tbHM{} \tbPH{vɛ́rɪ́} \tbPS{v.} \tbGE{to scatter thick liquid e.g. mortar} \tbPD{vɛrɛɛ, vɛrɛ, vɛrɛ, vɛrebɛ, vɛraa}%, , -}
\tbLX{vɛre}\tbHM{} \tbPH{vɛ́rɪ́} \tbPS{n.} \tbGE{stony farmland} \tbOP{vɛrɛɛ}
\tbLX{vɛrɛɛ}\tbHM{} \tbPH{vɛ̀rɛ̀ɛ̀} \tbPS{adj.} \tbGE{lying low e.g. dew-laden grass}
\tbLX{vɛvɛ}\tbHM{} \tbPH{vɛ́vɛ́} \tbPS{adj.} \tbGE{glidingly e.g. a bird}
\tbLX{vɛvɛre}\tbHM{} \tbPH{vɛ̀vɛ̀rɪ́} \tbPS{n.} \tbGE{mud, marsh} \tbTP{vɛvɛrɛɛ}
\tbLX{vi}\tbHM{} \tbPH{vî} \tbPS{n.} \tbGE{shame, disgrace} \tbOP{viri} \tbTP{viree}
\tbLX{vigi}\tbHM{} \tbPH{vígì} \tbPS{n.} \tbGE{owl} \tbOP{vigiri} \tbTP{vigree} \tbVA{vig, vib}
\tbLX{vigi}\tbHM{} \tbPH{vígì} \tbPS{v.} \tbGE{to throw out in a whole (solid waste)} \tbPD{vigee, vigre, vigre, vigribe, vigraa}%, , -}
\tbLX{viiri}\tbHM{} \tbPH{vììrì} \tbPS{v.} \tbSN \tbGE{to turn around, to rotate} \tbSN \tbGE{to visit a place repeatedly} \tbPD{viiree, viire, viire, viiribe, viiraa, viiribe }
\tbLX{vilenkyimm}\tbHM{} \tbPH{vílénkyímm} \tbPS{adj.} \tbGE{twisted}
\tbLX{vilenva}\tbHM{} \tbPH{vílénvà} \tbPS{n.} \tbGE{going out of sight} \tbXV{A karema yeli ka o erɛ vilenva ka a karembiiri yaarɛɛ.} \tbXE{No sooner had the teacher left than the students dispersed.}
\tbLX{vilenvileŋ}\tbHM{} \tbPH{vílénvíléŋ} \tbPS{adv.} \tbGE{round, circular}
\tbLX{vili}\tbHM{} \tbPH{vílí} \tbPS{v.} \tbGE{to entwine, to tie, to wrap around} \tbPD{vilee, vile, vile, vilbe, vilaqa}%, , -}
\tbLX{vilime}\tbHM{} \tbPH{vílímé} \tbPS{adj.} \tbGE{circular} \tbVA{vilenvileŋ}
\tbLX{virenvireŋ}\tbHM{} \tbPH{vírénvíréŋ} \tbPS{adv.} \tbGE{totally e.g. refuse}
\tbLX{viretaŋgbelli}\tbHM{} \tbPH{vìrèntáŋgbéllí} \tbPS{n.} \tbGE{your limit, first and last (colloquialism)}
\tbLX{vitaŋkpulee}\tbHM{} \tbPH{vítáŋkpúléé} \tbPS{n.} \tbGE{a type of small owl} \tbOP{vitaŋkpulli}
\tbLX{vo}\tbHM{} \tbPH{vò} \tbPS{v.} \tbGE{to make less crowded} \tbPD{voe, voro, voro}%, , -, -, -}
\tbLX{vogɛɛ}\tbHM{} \tbPH{vʊ̀gɛ̀ɛ̀} \tbPS{adv.} \tbGE{voluminous e.g. pair of trousers}
\tbLX{vokyɔ}\tbHM{} \tbPH{vʊ̀ʧɔ̀} \tbPS{n.} \tbSN \tbGE{a type of tree with long fruit} \tbSN \tbGE{the fruit of this tree (used in kneading)} \tbOP{vokyɔre} \tbTP{vokyɔrɛɛ}
\tbLX{volavola}\tbHM{} \tbPH{vʊ̀làvʊ̀là} \tbPS{adv.} \tbGE{slimly} \tbVA{volɔvolɔ}
\tbLX{volɔvolɔ}\tbHM{} \tbPH{vʊ̀lɔ̀vʊ̀lɔ̀} \tbPS{adv.} \tbGE{slimly} \tbVA{volavola}
\tbLX{vonne}\tbHM{} \tbPH{vʊ̀nnɪ̀} \tbPS{v.} \tbGE{to throw far away} \tbPD{vonnɛɛ, vonna, vonna, vonema, vonnaa}%, , -}
\tbLX{voŋe}\tbHM{} \tbPH{vʊ́ŋ} \tbPS{v.} \tbGE{to stay put and refuse to move} \tbPD{voŋɛɛ, vonɔ, vonɔ, vonema, vonaa}%, , -}
\tbLX{võo}\tbHM{} \tbPH{vʊ̃́ʊ̃̀} \tbPS{v.} \tbGE{to mix, to disturb e.g. water} \tbPD{voõɛ, voonɔ, voonɔ, voonema, voonaa,voonoo} \tbVA{voŋe}
\tbLX{voone}\tbHM{} \tbPH{vʊ̀ʊ̀nɪ̀} \tbPS{v.} \tbGE{to sound, to vibrate} \tbPD{voonɛɛ, voona/voonɔ, voona/voonɔ}%, , -, -, -}
\tbLX{vooraa}\tbHM{} \tbPH{vʊ́ʊ́ráá} \tbPS{n.} \tbGE{a type of lumpy paste made from bean flour} \tbOP{voore}
\tbLX{voore}\tbHM{} \tbPH{vʊ̀ʊ̀rɪ̀} \tbPS{v.} \tbGE{echo, resound (noise)} \tbPD{voorɛɛ, voora, voora, vooreba, vooraa}%, , -}
\tbLX{voore}\tbHM{} \tbPH{vʊ́ʊ́rɪ́} \tbPS{v.} \tbSN \tbGE{to breathe, to take a breath} \tbSN \tbGE{to be relieved, to be free, to rest} \tbXV{A saama kulee la ka N naa voore.} \tbXE{The guets have gone and now I'm relieved.} \tbPD{voorɛɛ, voora, voora, vooreba, vooraa, vooreba } \tbSE{voore fõõ} \tbGE{to take a deep breath, to sigh}
\tbLX{Vooronsoŋ}\tbHM{} \tbPH{vʊ́ʊ́rʊ́nsʊ́ŋ} \tbPS{n.} \tbGE{Holy Spirit}
\tbLX{vooroŋ}\tbHM{} \tbPH{vʊ́ʊ́rʊ́ŋ} \tbPS{n.} \tbSN \tbGE{life} \tbSN \tbGE{breathed air, breath} \tbOP{voora} \tbTP{voorɛɛ}
\tbLX{voowoɔ}\tbHM{} \tbPH{vʊ́ʊ́wʊ́ɔ́} \tbPS{n.} \tbGE{womb} \tbOP{voowoore} \tbTP{voowoorɛɛ}
\tbLX{voɔ}\tbHM{} \tbPH{vʊ́ɔ̀} \tbPS{v.} \tbGE{to pull out, to uproot} \tbPD{voɔɛ, voɔrɔ, voɔrɔ, voɔreba, vɔraa}%, , -}
\tbLX{voɔbale}\tbHM{} \tbPH{vʊ́ɔ́bàlɪ̀} \tbPS{n.} \tbGE{arrow with many hooks}
\tbLX{voɔloŋ}\tbHM{} \tbPH{vʊ̀ɔ̀lʊ́ŋ} \tbPS{n.} \tbGE{slimy sap (from a type of tree) used to settle sediments in liquids}
\tbLX{voɔloŋ}\tbHM{} \tbPH{vʊ̀ɔ̀lʊ́ŋ} \tbPS{adj.} \tbGE{slimy}
\tbLX{voɔre}\tbHM{} \tbPH{vʊ́ɔ́rɪ́} \tbPS{n.} \tbSN \tbGE{opening, hole (through an object)} \tbSN \tbGE{a small door, usually at the back of a compound house (comparable to an emergency exit)} \tbSN \tbGE{small space} \tbXV{N daare ba nyɛ voɔre gaa a kuori.} \tbXE{I did not get the chance to attend the funeral.} \tbOP{vɔ̃ɛ} \tbTP{vɔrɛɛ}
\tbLX{voɔre}\tbHM{} \tbPH{vʊ̀ɔ̀rɪ̀} \tbPS{v.} \tbGE{to be worn out, to be spent, to be tired and weak} \tbPD{voɔrɛɛ, voɔrɔ, voɔrɔ, voɔreba, voɔraa}%, , -}
\tbLX{voɔre}\tbHM{} \tbPH{vʊ̀ɔ̀rɪ̀} \tbPS{v.} \tbGE{to get scorched e.g. plants during drought} \tbPD{voɔrɛɛ, voɔrɔ, voɔrɔ, voɔreba, voɔraa}%, , -}
\tbLX{voɔre}\tbHM{} \tbPH{vʊ̀ɔ̀rɪ̀} \tbPS{v.} \tbGE{to smear} \tbPD{voɔrɛɛ, voɔrɔ, voɔrɔ, voɔreba, voɔraa}%, , -}
\tbLX{voraa}\tbHM{} \tbPH{vʊ́ráá} \tbPS{n.} \tbGE{a hand drill; see kyaleere}
\tbLX{vore}\tbHM{} \tbPH{vʊ́rɪ́} \tbPS{v.} \tbGE{to make a hole through, to pierce e.g. ears, lips} \tbPD{vorɛɛ, vorɔ, vorɔ, voreba, voraa, voreba }
\tbLX{vore}\tbHM{} \tbPH{vʊ́rɪ́} \tbPS{v.} \tbGE{to weave} \tbPD{vorɛɛ, vorɔ, vorɔ, voreba, voraa, voreba }
\tbLX{vore}\tbHM{} \tbPH{vʊ́rɪ̀} \tbPS{n.} \tbGE{foetus} \tbOP{voɛ} \tbTP{vorɛɛ}
\tbLX{vore}\tbHM{} \tbPH{vʊ́rɪ̀} \tbPS{adj.} \tbGE{alive} \tbOP{voɛ} \tbTP{vorɛɛ}
\tbLX{vori}\tbHM{} \tbPH{vórì} \tbPS{v.} \tbGE{to rot, to spoil completely} \tbPD{voree, voro, voro, voribo, voraa}%, , -}
\tbLX{vorikporo}\tbHM{} \tbPH{vóríkpóró} \tbPS{adv.} \tbGE{watery manner}
\tbLX{vorovoro}\tbHM{} \tbPH{vòròvòrò} \tbPS{adj.} \tbGE{watery}
\tbLX{vɔ}\tbHM{} \tbPH{vɔ́} \tbPS{v.} \tbGE{to perform a type of ritual for a dead male} \tbPD{vɔɛ, voɔrɔ, voɔrɔ, voɔreba, voɔraa}%, , -}
\tbLX{vɔga}\tbHM{} \tbPH{vɔ́gà} \tbPS{n.} \tbGE{track, trail} \tbVA{vɔgɔ}
\tbLX{vɔgaa}\tbHM{} \tbPH{vɔ̀gáá} \tbPS{n.} \tbGE{silk cotton tree} \tbOP{vɔgre} \tbTP{vɔgrɛɛ}
\tbLX{vɔge}\tbHM{} \tbPH{vɔ̀gɪ̀} \tbPS{v.} \tbSN \tbGE{to take off} \tbSN \tbGE{to uncover} \tbPD{vɔgɛɛ, vɔgrɔ, vɔgrɔ, vɔgreba, vɔgraa}%, , -}
\tbLX{vɔge}\tbHM{} \tbPH{vɔ̀gɪ̀} \tbPS{v.} \tbSN \tbGE{to scald} \tbSN \tbGE{to get burnt from hot substance} \tbPD{vɔgɛɛ, vɔgrɔ, vɔgrɔ, vɔgreba, vɔgraa}%, , -}
\tbLX{vɔgele}\tbHM{} \tbPH{vɔ̀glɪ̀} \tbPS{v.} \tbSN \tbGE{to cover} \tbSN \tbGE{to put on top e.g. hat} \tbSN \tbGE{to overturn e.g. container} \tbPD{vɔglɛɛ, vɔglɔ, vɔglɔ, vɔgleba, vɔglaa, vɔgleba }
\tbLX{vɔgɔ}\tbHM{} \tbPH{vɔ́gɔ̀} \tbPS{n.} \tbGE{track, trail} \tbOP{vɔgre} \tbTP{vɔgrɛɛ} \tbVA{vɔga}
\tbLX{vɔlaa}\tbHM{} \tbPH{vɔ́láà} \tbPS{n.} \tbGE{great desire for}
\tbLX{vɔle}\tbHM{} \tbPH{vɔ́lɪ̀} \tbPS{n.} \tbGE{great desire for meat; carnivorousness} \tbOP{vɔlɛɛ} \tbTP{vɔllɛɛ}
\tbLX{vɔle}\tbHM{} \tbPH{vɔ́lɪ́} \tbPS{v.} \tbGE{to swallow} \tbPD{vɔlɛɛ, vɔlɔ, vɔlɔ, vɔlebɔ, vɔlaa, vɔlebɔ }
\tbLX{vɔll}\tbHM{} \tbPH{vɔ̀ll} \tbPS{adj.} \tbGE{beautifully; usually in the state of growing up}
\tbLX{vɔrekpɔrɔ}\tbHM{} \tbPH{vɔ̀rɪ̀kpɔ̀rɔ̀} \tbPS{adv.} \tbSN \tbGE{wet and unwieldy} \tbXV{Saseɛ fue la ka saa meŋ mi kõɔ lɔre a sagere ka a zie e lɛ vɔrekpɔrɔ.} \tbXE{The rain which followed the rainstorm has soaked the garbage in the yard making it wet and disorderly.} \tbSN \tbGE{sound of fall by a long winding thing e.g. snake, rope (onomatopoeia)}
\tbLX{vu}\tbHM{} \tbPH{vú} \tbPS{v.} \tbSN \tbGE{to crawl e.g. child} \tbSN \tbGE{to drag along the ground} \tbPD{vue, vuuro, vuuro, vuribo, vuuraa, vuribo }
\tbLX{vubiri}\tbHM{} \tbPH{vúbìrí} \tbPS{n.} \tbGE{mound for yam plant} \tbOP{vubie} \tbTP{vubiiree}
\tbLX{vuduori}\tbHM{} \tbPH{vúdúórì} \tbPS{n.} \tbGE{the act of making yam mounds}
\tbLX{vuduoro}\tbHM{} \tbPH{vúdúóró} \tbPS{n.} \tbGE{one who makes yam mounds} \tbOP{vuduoribo}
\tbLX{vulaŋkaŋkpogili}\tbHM{} \tbPH{vùlàŋkǎŋkpóglí} \tbPS{n.} \tbGE{final thing} \tbOP{vulaŋkaŋkpolo} \tbTP{vulaŋkaŋkpolee}
\tbLX{vuloŋkpogili}\tbHM{} \tbPH{vùlòŋkpóglí} \tbPS{n.} \tbGE{beetle} \tbOP{vuloŋkpogilo} \tbTP{vuloŋkpogilee} \tbVA{vulaŋkpogli}
\tbLX{vulotokpolo}\tbHM{} \tbPH{vùlòtókpóló} \tbPS{n.}  \tbSN \tbGE{piece of work meant to be one’s work for the day} \tbSN  \tbGE{stanza (literature)} \tbOP{vulotokpolli} \tbTP{vulotokpolee}
\tbLX{vunni}\tbHM{} \tbPH{vùnnì} \tbPS{v.} \tbGE{to throw far away} \tbPD{vunnee, vunno, vunno, vunimo, vunnaa}%, , -}
\tbLX{vuo}\tbHM{} \tbPH{vùó} \tbPS{n.} \tbSN \tbGE{space,  passage} \tbSN \tbGE{chance, time} \tbSN \tbGE{occasion}
\tbLX{vuo}\tbHM{} \tbPH{vùó} \tbPS{adj.} \tbGE{hollow} \tbOP{vori} \tbTP{voree}
\tbLX{vuoli}\tbHM{} \tbPH{vúólì} \tbPS{v.} \tbSN \tbGE{to bark (dog)} \tbSN \tbGE{to hoot (to put to shame)} \tbPD{vuolee, vuolo, vuolo, vuolibo, vuolaa, - } \tbVA{wuoli}
\tbSE{vuoli kaa} \tbGE{to peer at}
\tbLX{vutaŋkpulee}\tbHM{} \tbPH{vútàŋkpùlèè} \tbPS{n.} \tbGE{gate pillar} \tbOP{vutaŋkpulli} \tbTP{vutaŋkpullee}
\tbLX{vuu}\tbHM{} \tbPH{vúù} \tbPS{v.} \tbGE{to drag} \tbXV{Ba vuu la a kamaana boɔraa boŋyeni na.} \tbXE{They dragged that one sack of maize.} \tbPD{vuue, vuuro, vuuro, vuuribo, vuuraa}%, , -}
\tbLX{vũu}\tbHM{} \tbPH{vũ̀ṹ} \tbPS{n.} \tbGE{fire} \tbSG{vuunuu} \tbTP{vuunee}
\tbLX{vũudindigili}\tbHM{} \tbPH{vũ̀ṹdíndíglí} \tbPS{n.} \tbSN \tbGE{electricity} \tbSN \tbGE{electric light} \tbVA{vũudigili}
\tbLX{vũudiraa}\tbHM{} \tbPH{vũ̀ũ̀dìráà} \tbPS{n.} \tbGE{flame} \tbOP{vũudirri} \tbTP{vũudiree}
\tbLX{vũudirri}\tbHM{} \tbPH{vũ̀ũ̀dìrrì} \tbPS{n.} \tbGE{a stream of burning}
\tbLX{vũugyɛle}\tbHM{} \tbPH{vũ̀ũ̀gyɛ́lɪ́} \tbPS{n.} \tbGE{lightbulb} \tbOP{vũugyɛlɛ} \tbTP{vũugyɛlɛɛ} \tbVA{vũũnimiri}
\tbLX{vuuli}\tbHM{} \tbPH{vúúlí} \tbPS{v.} \tbGE{to singe} \tbPD{vuulee, vuulo, vuulo, vuulibo, vuulaa}%, , -}
\tbLX{vũulunsaa}\tbHM{} \tbPH{vũ̀ũ̀lúnsáà} \tbPS{n.} \tbGE{first rain after the dry season}
\tbLX{vuuluŋ}\tbHM{} \tbPH{vúúlúŋ} \tbPS{n.} \tbGE{an animal believed to have supernatural power and usually attack people in the night} \tbVA{valinvuuluŋ}
\tbLX{vuuluŋ}\tbHM{} \tbPH{vúúlúŋ} \tbPS{n.} \tbGE{short form of valinvuuluŋ (soot)} \tbVA{valinvuuluŋ}
\tbLX{vuunimiri}\tbHM{} \tbPH{vùùnímírì} \tbPS{n.} \tbGE{lightbulb} \tbOP{vuunimie} \tbTP{vuunimiiree} \tbVA{vũugyɛle}
\tbLX{vuuri}\tbHM{} \tbPH{vùùrì} \tbPS{v.} \tbGE{to feign; to pretend or attempt to do something (which doesn't happen)} \tbPD{vuuree, vuuro, vuuro, vuuribo, vuuraa, vuuribo }
\tbLX{vuuri}\tbHM{} \tbPH{vúúrì} \tbPS{n.} \tbGE{small oar-like stick for stirring porridge} \tbOP{vue} \tbTP{vuuree}
\tbLX{vuuri}\tbHM{} \tbPH{vúúrì} \tbPS{n.} \tbGE{yam farm} \tbOP{vue} \tbTP{vuuree}
\tbLX{vũusaalbiri}\tbHM{} \tbPH{vũ̀ũ̀sáálbírí} \tbPS{n.} \tbGE{burning charcoal} \tbVA{vũusaale}
\tbLX{vũusaale}\tbHM{} \tbPH{vũ̀ũ̀sáálɪ́} \tbPS{n.} \tbGE{burning coal} \tbOP{vũusaala} \tbTP{vũusaaloo} \tbVA{vũusaalbiri}
\tbLX{vũuvaale}\tbHM{} \tbPH{vũ̀úváálɪ́} \tbPS{n.} \tbGE{a type of bird which hovers around bushes that are being burnt to catch fleeing insects} \tbOP{vũuvaala}
\tbLX{vũuyɛlɛ}\tbHM{} \tbPH{vũ̀ũ̀gyɛ́lɛ́} \tbPS{n.} \tbGE{lightbulb} \tbOP{vũũgyɛlɛ} \tbTP{vũũgyɛlɛɛ}
\tbLX{vũuzeɛle}\tbHM{} \tbPH{vũ̀ũ̀zɪ́ɛ́lɪ́} \tbPS{n.} \tbGE{flying burnt particles from a fire e.g. grass} \tbSG{vũuzeɛloo} \tbOP{vũuzeɛlɛ} \tbTP{vũuzeɛlɛɛ}
\tbLX{vũuzele}\tbHM{} \tbPH{vũ̀ũ̀zɪ̀lɪ́} \tbPS{n.} \tbGE{flame, tongue of fire} \tbOP{vũuzelɛ} \tbTP{vũuzelɛɛ}
\end{letter}
