\chapter{Stimuli for perception test}
	\label{app.stimuli}
	\subsection*{\textsc{nurse}}
	\label{sec:nurse}
	
	\begin{minipage}[t]{0.49\textwidth}
		\subsubsection*{Final}
			\begin{enumerate}\footnotesize
				\item People in that town almost never went to ch\uline{ur}ch.
				\item It was best to hide fast when it was his t\uline{ur}n.
				\item The old cat still had quite a nice, soft f\uline{ur}.
				\item There are hundreds of nice names for a g\uline{ir}l.
				\item He always managed to ruin his new sh\uline{ir}t.
				\item She wasn't sure this was the right w\uline{or}d.
			\end{enumerate}
	\end{minipage}
	\begin{minipage}[t]{0.49\textwidth}
		\subsubsection*{Medial}
			\begin{enumerate}\footnotesize
				\item In that town ch\uline{ur}ch was not popular with people.
				\item When it was his t\uline{ur}n we often hid extra fast.
				\item The cat's nice, soft f\uline{ur} now appeared rather sordid.
				\item Mary was a nice g\uline{ir}l, but not a smart one.
				\item He wanted that sh\uline{ir}t but couldn't pay for it.
				\item She knew the w\uline{or}d but pretended not to.
			\end{enumerate}
	\end{minipage}
			
	\subsection*{happ\textsc{y}}
	\label{sec:happy}

	\begin{minipage}[t]{0.49\textwidth}
		\subsubsection*{Final}
			\begin{enumerate}\footnotesize
				\item At the end of the day we all want to be happ\uline{y}.
				\item He had always known that she was laz\uline{y}.
				\item She had never thought it important to be prett\uline{y}.
				\item There are lots of new hotels in the cit\uline{y}.
				\item For two years, they had tried for a bab\uline{y}.
				\item No one in our street could be called sting\uline{y}.
			\end{enumerate}
	\end{minipage}
	\begin{minipage}[t]{0.49\textwidth}
		\subsubsection*{Medial}
			\begin{enumerate}\footnotesize
				\item We all want to be happ\uline{y}, but not all in the same way.
				\item He knew she was laz\uline{y}, and mean as well.
				\item She had never been prett\uline{y} until about age fourteen.
				\item Most hotels in the cit\uline{y} are now located downtown.
				\item Two toddlers and a bab\uline{y} required all their attention.
				\item No one there was sting\uline{y}, at least not in public.
			\end{enumerate}
	\end{minipage}

	\subsection*{Lenition of /k/}
	\label{sec:k}

	\begin{minipage}[t]{0.49\textwidth}
		\subsubsection*{Final}
			\begin{enumerate}\footnotesize
				\item She's just someone you have got to li\uline{ke}.
				\item The film's not half as good as the boo\uline{k}.
				\item Bill had never even been part of the pa\uline{ck}.
				\item He's good, but I wouldn't call him a ha\uline{ck}er.
				\item They used to meet quite often to play snoo\uline{k}er.
				\item The way he moved reminded us of a chi\uline{ck}en.
			\end{enumerate}
	\end{minipage}
	\begin{minipage}[t]{0.49\textwidth}
		\subsubsection*{Medial}
			\begin{enumerate}\footnotesize
				\item She's just someone you li\uline{ke}, regardless of unpleasant habits.
				\item I quite enjoyed the boo\uline{k}, but the film's terrible.
				\item Well, there was the pa\uline{ck}, but that was different.
				\item He's not quite a ha\uline{ck}er, at least not yet.
				\item They used to play snoo\uline{k}er in their favourite pub.
				\item We had both ordered chi\uline{ck}en and got beef instead.
			\end{enumerate}
	\end{minipage}

	\subsection*{Velar nasal plus}
	\label{sec:ng}
	
		\begin{minipage}[t]{0.49\textwidth}
			\subsubsection*{Final}
			\begin{enumerate}\footnotesize
				\item As a child Heather wanted to be a si\uline{ng}er.
				\item Please put your coat on the free ha\uline{ng}er.
				\item I bet that speech will be rather lo\uline{ng}ish.
				\item This is by far my wife's favourite so\uline{ng}.
				\item None of us was a member of a ga\uline{ng}.
				\item At the time I was still quite you\uline{ng}.
			\end{enumerate}
		\end{minipage}
		\begin{minipage}[t]{0.49\textwidth}
			\subsubsection*{Medial}
			\begin{enumerate}\footnotesize
				\item Often, a famous rock si\uline{ng}er lives on the edge.
				\item Put it on the ha\uline{ng}er to your right, please.
				\item It'll be rather lo\uline{ng}ish, so don't fall asleep.
				\item She adores this so\uline{ng} because of its rhythm.
				\item It was a notorious ga\uline{ng} but she wasn't afraid.
				\item Jake appeared quite you\uline{ng}, which often annoyed him.
			\end{enumerate}
		\end{minipage}