\addchap{Acknowledgments}
\begin{refsection}

This dissertation would not have been possible without the guidance and assistance of several people who in one way or another contributed and extended their valuable time and expertise in the groundwork and accomplishment of this study.

My utmost appreciation to Professor Hubert Devonish, without whose excellent supervision, this would not have been possible. Professor Devonish has challenged my abilities at all levels, thereby allowing me to surpass thresholds that I never thought it possible to surpass.

I am deeply grateful to Professor Ewart Thomas of Stanford University, who made himself available to aid me with most, if not all, my statistical analysis related matters. I would like to express my gratitude for his expertise and the patience he showed to me, a non-statistician.

Many thanks, to Dr. Karen Carpenter for her altruistic and dependable support as the coordinator of what was a very effective writing group. Her guidance and support, especially at times when what laid ahead seemed overwhelming, will not be forgotten.

I would like to thank Lubova Dubilina, who spent countless hours with me, helping to format my work. Thank you for your assistance, patience and resolve to see me complete what I started.

Last but by no means least, my sincerest gratitude to my loving and supportive parents, Augustus Sherriah, and Verna Sherriah, my best friend and brother, Courtney Smith, my dearest friends and colleagues, Audene Henry-Harvey, Daidrah Smith, Kedisha Williams, Yanique Wallace, Tashieka Burris and the countless others who supported me in more ways than one, throughout this journey.

\printbibliography[heading=subbibliography]
\end{refsection}

