\section{A car accident}\label{sec:CarCrash}
\subsection{Metadata}

\wg\begin{tabular}{lL{.67\textwidth}}
File-name:					& aaz20130911{\_}02\\
Archive-link:				& \url{http://catalog.paradisec.org.au/collections/OE1/items/aaz20130911_02}\\
Original Name:			& aaz-20130911-2-DominggusBani-HenkiOra- CeritaOtoJato\\
Language:						& Amarasi [aaz] \\
Dialect:						& Kotos; Koro{\Q}oto hamlet \\
Location:						& Nekmese{\Q}, Amarasi Selatan, Kupang \\
Date:								&	11/09/2013\\
Speaker(s):					& Dominggus Bani (D), Heronimus Bani (R), Henki Ora (H), Sefnat Bois, and occasional others\\
Recorded by: 				& Heronimus Bani\\
Transcribed by:			& Heronimus Bani\\
Interlinear by:			& Owen Edwards \\
Indonesian/Kupang:	& Heronimus Bani\\
Free English by:		& Owen Edwards\\
Genre:							& conversation\\
Summary:						& conversation about a car accident\\
%Length:		& 1.43\\
\end{tabular}

\subsection{Notes}
This text is a conversation about a recent car crash.
As is to be expected from natural free-flowing conversation,
there are many instances in which more than one person is speaking at once.
Given this, it was not possible for the transcriber (Heronimus Bani)
to transcribe every voice at every point in the recording.
I have listened through the entire text several times and edited where necessary.
Where there is doubt over the exact transcription,
I have deferred to the original.

The three dominant participants are Dominggus Bani
(D), Heronimus Bani (R) and Henki Ora (H).
Names of other participants
are given in full before their contributions.
When a speaker makes multiple consecutive contributions,
only the first contribution is marked.
The recording begins after the conversation has begun
and the topic of conversation has been established.

\newpage
\subsection{The text}
\begin{exe}
\renewcommand{\N}{\tsc{3gen}}

\ex[R:]{\ve{Onai te ma saa{\Q}, naa,}}
\exi{0.00}[]{\glll	{onai =t}, {onai =t}, =ma, =ma, saaʔ \sf{naa} \\
									{onai =te} {onai =te} =ma =ma saaʔ \sf{naa}\\
									and.then and.then =and =and what well \\
						\glt	`and then, and then, and, and what, well'}

\ex[]{\ve{kedalaman ma{\Q}boik{\Q} ee, keefn ii mnanu{\Q}.}}
\exi{0.03}[]{\glll	\sf{kedalaman} maʔboikʔ=ee keefn=ii mnanuʔ \\
									\sf{kedalaman} maʔbokiʔ=ee kefan=ii mnanuʔ \\
									deep suspended={\ee} gap={\ii} deep \\
						\glt	`(it was) deep (the car) was suspended, the gap was deep.'}

\ex[D:]{\ve{Re{\Q} natoon ii nak am, pas anritu{\Q} neu re{\Q} mnaun{\Q} ii jeen am,
				ka tahini mnaun{\Q} ii basik ate, ka tahiin, neor hit toka ma, es he tahiin.}}
%\exi{0.05}[]{\begin{xlist}
\exi{0.05}{\begin{xlist}
	\ex{\glll	reʔ na-toon=ii n-aka =m, \\
						reʔ na-tona=ii n-aka =ma\\
						{\req} \na-tell={\ii} \n-say =and\\
			\glt	`That's what they said,'}
	
	\ex{\glll	\sf{pas} a|n-rituʔ n-eu reʔ mnaunʔ=ii\j=ena =m, \\
						\sf{pas} {\a}n-rituʔ n-eu reʔ mnanuʔ=ii=ena =ma \\
						exact \a\n-roll \n-{\eu} {\reqt} deep={\ii}={\een} =and \\
			\glt	`and it rolled exactly into the deep space'}
	
	\ex{\glll	ka= ta-hini mnaunʔ=ii basik =at, \\
						ka= ta-hini mnanuʔ=ii basik =te \\
						{\ka}= \ta-know depth={\ii} how.much ={\te} \\
			\glt	`we don't know how deep it was'}
	
	\ex{\glll	ka= ta-hiin, neor hiit t-oka =m es he ta-hiin\\
						ka= ta-hini nero hiit t-oka =ma es he ta-hini \\
						{\ka}= {\ta}-know not {\hiit} {\t-\ok\U} =and one {\he} {\ta}-know \\
			\glt	`we don't know, we weren't with (them) to know' \txrf{0.10}}
\end{xlist}}

\ex[]{\ve{Mnanu{\Q}, oot goe, nak sin na{\Q} nateut oto.}}
\exi{0.12}[]{
	\glll	mnanuʔ ootgw=ee n-ak siin naʔ na-teut oto\\
				mnanuʔ oto=ee n-ak sini naʔ na-tetu oto \\
				deep car={\ee} \n-say {\siin} then \nat-upright car \\
	\glt	`(It was) deep, the car, they said they then stood the car upright'}

\ex[R:]{\ve{Sekau es neki?}}
\exi{0.15}[]{\glll
	sekau ees n-eki\\
	sekau esa n-eki\\
	who {\esc} \n-bring{\U} \\
\glt `Who was the one driving?'}

\ex{Sefnat Bois: \ve{Cuma mana fa te, nmouf goen ate,}}\vspace{-4pt}
\exi{0.16 }[]{\glll
	\sf{suma} nehh, mana =fa =te n-moofgw=ena =te\\
	\sf{suma} {} mana =fa =te n-mofu=ena =te \\
	only {} like.that ={\fa} ={\te} \n-fall={\een} ={\te} \\
\glt `Only, umm, when (it was) like that, when it fell,'}

\ex[R:]{\ve{Rem ee naah mes,}}
\exi{0.18}[]{\glll
	reem=ee na-ah mes \\
	reem=ee na-ah mes\\
	brakes={\ee} \na-eat but\\
\glt `the brakes failed? but{\ldots}'}

\ex[H:]{\ve{Rem ee naah. Semantara n{\Q}antareek.}}
\exi{0.19}[]{\glll
	reem=ee na-ah \sf{sementara} n-\sf{ʔantareek} \\
	reem=ee na-ah \sf{sementara} n-\sf{ʔantareek} \\
	brakes={\ee} \na-eat during \n-backing \\
\glt `The brakes failed, while they were backing.' }

\ex[R:]{\ve{Ohh, semantara n{\Q}antareek.}}
\exi{0.21}[]{\glll
	ohh, \sf{sementara} n-\sf{ʔantareek} \\
	{} \sf{sementara} n-\sf{ʔantareek} \\
	oh during \n-backing \\
\glt `Oh, while they were backing.'}

\ex[H:]{\ve{Jadi, in ka nakeo fa mnaun he--}}
\exi{0.23}[]{\glll
	\sf{{\j}adi} iin ka= nauhh ka= na-keo =fa mnaun he--\\
	\sf{{\j}adi} ini ka= {} ka= na-keo =fa mnanuʔ {}\\
	so {\iin} {\ka}= umm {\ka}= \na-be.aware ={\fa} deep{\M} {}\\
\glt `So, he wasn't, wasn't aware (it was) deep'}

\newpage
\ex[]{\ve{Posisi n{\Q}antareek in ka bisa nbi fa nee, saap ma{\Q}bake{\Q}.}}
\exi{0.25}[]{\begin{xlist}
	\ex{\glll	\sf{posisi} n-\sf{ʔantareek} iin ka= bisa n-bi =fa nee\\
						\sf{posisi} n-\sf{ʔantareek} ini ka= bisa n-bi =fa nee\\
						posisi \n-backing {\iin} {\ka}= able \n-{\bi} ={\fa} {\nee}\\
			\glt	`His position was backing, he couldn't get there '}
	\ex{\glll	saap maʔbakeʔ, \\
						saap maʔbakeʔ \\
						because narrow \\
			\glt	`because it was narrow.'}
\end{xlist}}

\ex[]{\ve{Bait in he naim bare hena{\Q} n{\Q}antareek ate, bisa.}}
\exi{0.28}[]{\glll
	bait iin he n-aim bare henaʔ n-\sf{ʔantareek} =at, bisa. \\
	bait ini he n-ami bare henaʔ n-\sf{ʔantareek} =te bisa \\
	actually {\iin} {\he} \n-look.for place {\he} \n-backing ={\te} able \\
\glt `Actually if he had looked for a place to back, he could have'}

\ex[R:]{\ve{In nareen on ma n{\Q}antareek anbi n--}}
\exi{0.31}[]{\glll
	iin na-reen=oo-n =ma n-\sf{ʔantareek} a|n-bi n-- \\
	ini na-rena=oo-n =ma n-\sf{ʔantareek} {\a}n-bi {}\\
	{\iin} \na-force={\oo=\N} =and \n-backing \a\n-{\bi} {}\\
\glt `He forced himself, and went back into it, he was in{\ldots}'}

\ex[H:]{\ve{Nabara ma{\Q}bake{\Q}.}}
\exi{0.32}[]{\glll
	na-bara maʔbakeʔ \\
	na-bara maʔbakeʔ \\
	\na-forever{\U} narrow \\
\glt `He was stuck in the narrow (place)' \txrf{0.32}}

\ex{Sefnat Bois: \ve{In he nbibi.}}\vspace{-4pt}
\exi{0.34}[]{\glll
	iin he n-bibi \\
	ini he n-bibi \\
	{\iin} {\he} \n-shrink {\U} \\
\glt `He would've wanted to shrink (the car)'}

\ex[D:]{\ve{Nak, oot gui nasnii, mak, am, nakamaf am,}}
\exi{0.35}[]{\glll	n-ak, ootgw=ii na-snii m-ak, =am, na-kamaf =am\\
										n-ak oto=ii na-snii m-ak =ma na-kamaf =ma\\
										\n-say car={\ii} \na-slope \m-say and \na-what's.it =and\\
							\glt	`he said, the car was sloping, you think, and what's it and'}
\ex[]{\ve{nasnii, ntaikobi nkoon, na{\Q} natetu.}}
\exi{0.38}[]{\glll	na-snii n-taikobi n-koon, naʔ na-tetu\\
										na-snii n-taikobi n-kono naʔ na-tetu\\
										\na-slope \n-fall \n-keep.on then \nat-upright{\U}\\
							\glt	`it was sloping, fell over, kept on, and only then he got the car upright' }

\ex[H:]{\ve{Onai ma, srutun re{\Q} ia, in nmouf goen.}}
\exi{0.40}[]{\glll
	{onai =ma} srutun reʔ ia, iin n-moofgw=een \\
	{onai =ma} srutun reʔ ia ini n-mofu=ena \\
	and.so suddenly {\req} {\ia} {\iin} \n-fall={\een} \\
\glt `and suddenly like this, it fell down'}

\ex{Sam Ora: \ve{Oh, mak oot gui in nmese nnao kuun.}}\vspace{-4pt}
\exi{0.42}[]{\glll
	ohh, m-ak, ootgw=ii iin n-mese n-nao kuu-n \\
	{} m-ak oto=ii ini n-mese n-nao kuu-n \\
	oh \m-say car={\ii} {\iin} \n-alone \n-go alone-{\N} \\
\glt `Oh, you think the car went by itself.'}

\ex[R:]{\ve{Mak, sofir ii nmouf goen?}}
\exi{0.43}[]{\glll
	m-ak ahh, sofiir=ii n-moofgw=een\\
	m-ak {} sofiir=ii n-mofu=ena\\
	\m-say {} driver={\ii} \n-fall={\een} \\
\glt `Do you think the driver fell?' }

\ex{Stef Ora: \ve{Tua.}}\vspace{-4pt}
\exi{0.45}[]{\gll
	tua \\
%	tua \\
	{\tua} \\
\glt `yes'}

\ex[R:]{\ve{Tuan?}}
\exi{0.46}[]{\gll
	tua-n \\
	owner-{\N} \\
\glt `(did you say) its owner?'}

\ex[H:]{\ve{Onaim, in nmeo te, oot gui in nmese ntaikob-koib.}}
\exi{0.47}[]{\glll
	{onai =m} iin n-meo =t, ootgw=ii iin n-mese n-taikob{\tl}koib \\
	{onai =m} ini n-meo =te oto=ii ini n-mese n-taikob{\tl}kobi \\
	and.so {\iin} \n-see ={\te} car={\ii} {\iin} \n-alone \n-{\prd}fall \\
\glt `And so when he saw it, the car fell down by itself'}

\newpage
\ex[D:]{\ve{Onai te, oirf ii nok aanh ii sin nbin belakang.}}
\exi{0.52}[]{\glll
	{onai =te} oir-f=ii n-ok aanh=ii siin n-bi=n a|\sf{blaka\ng} \\
	{onai =te} ori-f=ii n-oka anah=ii sini n-bi=n {\a}\sf{blaka\ng} \\
	and.then ySi-\f={\ii} {\n-\ok} child={\ii} {\siin} \n-\bi={\einV} {\a}back \\
\glt `and his younger brother with his child were in the back (of the car)'}

\ex[R:]{\ve{Orif Joni.}}
\exi{0.52}[]{\glll
	ori-f Joni. \\
	ori-f Joni \\
	ySi-{\F} Joni \\
\glt `the younger brother was Johnny.' \txrf{0.52}}

\ex[D:]{\ve{Tuan ii nnaben ate oni{\Q} maineun{\Q} een ate,}}
\exi{0.52}[]{\glll
	tua-n=ii, n-naben =at oniʔ maineunʔ=ena =te. \\
	tua-n=ii n-naben =te oniʔ mainenuʔ=ena =te \\
	owner-{\N}={\ii} \n-feel ={\te} maybe wide.length={\een} ={\te} \\
\glt `The owner, maybe he felt as though there was enough space'}

\ex[]{\ve{Tuan ii nnaben ate mnaun{\Q} een, ro in nrete npoi kuun.}}
\exi{0.55}{\begin{xlist}

	\ex{\glll
		tua-n=ii n-naben =at mnaunʔ=een\\
		tua-n=ii n-naben =te mnanuʔ=ena\\
		owner-{\N}={\ii} \n-feel ={\te} deep={\een}  \\
	\glt `The owner felt it was (too) deep,'}
	
	\ex{\glll
		ro iin n-rete n-poi kuu-n. \\
		ro ini n-rete n-poi kuu-n \\
		must {\iin} \n-jump \a\n-exit alone-{\N} \\
	\glt `he had to jump out by himself'}
	
\end{xlist}}

\ex[R:]{\ve{Aina, in nasaeb ba{\Q}-ba{\Q}uk atoin{\Q} ein?}}
\exi{1.00}[]{\glll
	aina, iin na-sae-b baʔ{\tl}baʔuk atoinʔ=ein\\
	aina ini na-sae-b {\prd}baʔuk atoniʔ=eni \\
	mother {\iin} \na-go.up-{\b} prd-several man={\ein} \\
\glt `Oh my, how many people was he carrying?'}

\ex[D:]{\ve{Molak am, muhiin he,}}
\exi{1.01}[]{\glll
	\sf{molak} =am mu-hiin he \\
	\sf{molak} =ma mu-hini he \\
	log and \muu-know {\he} \\
\glt `(he was carrying) logs, and you know...'}

\ex[R:]{\ve{Ma{\Q}fena{\Q}.}}
\exi{1.04}[]{\glll
	maʔfenaʔ \\
	maʔfenaʔ \\
	heavy \\
\glt `heavy'}

\ex[H:]{\ve{In nak fe{\Q} nasaeban naan tuka{\Q} bo{\Q} esa?}}
\exi{1.05}[]{\glll
	iin n-ak feʔ na-sae-ba=n naan tukaʔ boʔ=esa\\
	ini n-ak feʔ na-sae-ba=n naan tukaʔ boʔ=esa \\
	{\iin} \n-say still \nat-go.up-{\b}={\einV} {\naan} slice ten=one{\U} \\
\glt `he said, he was carrying ten of them, right?'}

\ex[R:]{\ve{Tuka{\Q} bo{\Q} es, mes mainenu{\Q}!}}
\exi{1.09}[]{\glll
	tukaʔ boʔ ees, mes mainenuʔ\\
	tukaʔ boʔ esa mes mainenuʔ\\
	slice ten one but wide.length\\
\glt `Ten of them. But that's too much!'}

\ex[H:]{\ve{Onai te, nak posisi n{\Q}antareek in nasaeba{\Q} nteni{\Q}.}}
\exi{1.10}[]{\glll
	{onai =t} n-ak \sf{posisi} n-\sf{ʔantareek} iin na-sae-baʔ n-teniʔ, \\
	{onai =te} n-ak \sf{posisi} n-\sf{ʔantareek} ini na-sae-{\b} n-teniʔ \\
	and.then \n-say position \n-backing={\ii} {\iin} \nat-go.up-{\b} \n-again \\
\glt `And then he said he was backing, he was carrying more'}

\ex[R:]{\ve{He nteni{\Q}.}}
\exi{1.13}[]{\glll
	he n-teniʔ \\
	he n-teniʔ \\
	{\he} \n-again \\
\glt `He wanted more.'}

\ex[D:]{\ve{Tasaeba{\Q} molak on re{\Q} nee ja te, ma{\Q}fena{\Q}.}}
\exi{1.13}[]{\glll
	ta-sae-baʔ \sf{molak} on reʔ nee\j=aa =t maʔfenaʔ \\
	ta-sae-baʔ \sf{molak} on reʔ nee=aa =te maʔfenaʔ \\
	\tg-go.up-{\b} log like {\reqt} {\nee=\aa} ={\te} heavy \\
\glt `carrying logs like that, it's heavy'}

\ex{Rehuel Nakmofa: \ve{Ma{\Q}fena{\Q}, papan re{\Q},}}\vspace{-4pt}
\exi{1.15}[]{\gll
	maʔfenaʔ, papan reʔ \\
	heavy plank {\req}\\
\glt `heavy, planks which {\ldots}'}

\ex[D:]{\ve{Papan, fe{\Q} papan noo nautn es, ma{\Q}kafa{\Q} fe{\Q}.}}
\exi{1.17}[]{\glll
	mahh papan, feʔ {noo nautn--,} papan noo nautn=ees, \hspace{20mm} maʔkafaʔ feʔ \\
	{} papan feʔ noo papan noo natun=ees {} maʔkafaʔ feʔ \\
	{} plank still {\manaq} plank {\manaq} hundred=one {} light still \\
\glt `Umm, planks, still a hundred, a hundred planks is still light!'}

\ex[H:]{\ve{Onai te, hi misaah miit noo nautn es!}}
\exi{1.21}[]{\glll
	{onai =t} naʔ hii mi-saah m-iit noo nautn=ees \\
	{onai =te} naʔ hii mi-saha m-ita noo natun=ees \\
	and.then {} {\hii} \mi-carry \m-try {\manaq} hundred=one \\
\glt `Well then, why don't you try and carry a hundred planks?'}

\exi{1.23}{\emph{[laughter]}}

\ex[D:]{\ve{Aah, hit tareta{\Q} nok oot goe ma hi ta{\Q}uab,}}
\exi{1.24}[]{\glll
	aah, hiit ta-retaʔ n-ok ootgw=ee =ma hiit ta-ʔuab {\ldots}\\
	{} hiit ta-retaʔ n-oka oto=ee =ma hiit ta-uaba \\
	ah {\hiit} {\ta}-story {\n-\ok} car={\ee} =ma {\hiit} {\ta}-speak \\
\glt `Ah yes! But we're talking about the car! And we're talking {\ldots}'}

\exi{1.26}{\emph{[laughter]}}

\ex[H:]{\ve{Au {\Q}ak, hi misoba{\Q} noo nautn es.}}
\exi{1.28}[]{\glll
	au ʔ-ak hii m-sobaʔ noo nautn=ees \\
	au ʔ-ak hii m-sobaʔ noo natun=ees \\
	{\au} \q-say {\hii} \m-try {\manaq} hundred=one \\
\glt `I said, you try (and carry) a hundred of them'}

\ex[D:]{\ve{Sonde, no nautn es ate, oot gui ma{\Q}kaaf{\Q} ii, naena te, mainenu{\Q}.}}
\exi{1.30}[]{\glll
	\sf{sonde}, noo nautn=esa =t ootgw=ii maʔkaafʔ=ii \hspace{20mm} n-aena =t, mainenuʔ \\
	\sf{sonde} noo natun=esa =te oto=ii maʔkafaʔ=ii {} n-aena =te mainenuʔ \\
	not {\manaq} hundred=one{\U} ={\te} car={\ii} light={\ii} {} \n-run{\U} ={\te} excessive \\
\glt `No, a hundred of them, (in) the car is light, (it) goes quickly, too much'}

\newpage
\ex[R:]{\ve{Onaim, ameent ee neu haa nai{\Q} Firgo.}}
\exi{1.33}[]{\glll
	{onai =m}, mhh, a-meen-t=ee n-eu =ha naiʔ Firgo\\
	{onai =m} {} a-mena-t=ee n-eu =ha naiʔ Firgo \\
	and.so {} {\at}-sick-{\at}={\ee} \n-{\eu} =only {\naiq} Firgo \\
\glt `And so, umm, the only one injured is Firgo.'}

\ex{Rehuel Nakmofa: \ve{Firgo nmees.}}\vspace{-4pt}
\exi{1.35}[]{\glll
	Firgo n-mees \\
	Firgo n-mese \\
	Firgo \n-alone \\
\glt `Just Firgo.'}

\ex[R:]{\ve{On dusun, tak, asaunt ee, nua sin oirf ii ka saa{\Q}.}}
\exi{1.36}{\begin{xlist}

		\ex{\glll
			on nehh, \sf{dusun}, ehh, t-ak a-saun-t=ee\\
			on {} \sf{dusun} {} t-ak a-sanu-t=ee \\
			{\on} {} county {} {\t-\ak} {\at}-descend-{\at}={\ee} \\
		\glt `Like, umm, the county (head), the one who fell down,'}

		\ex{\glll
			nua siin oir-f=ii ka= saaʔ \\
			nua sini ori-f=ii ka= saaʔ \\
			two {\siin} ySi-{\F=\ii} {\ka}= what \\
		\glt `nothing happened to those two kids.'}

\end{xlist}}

\ex{Rehuel Nakmofa: \ve{Nak ka saa{\Q} fa.}}\vspace{-4pt}
\exi{1.40}[]{\glll
	n-ak, ka= saaʔ =fa\\
	n-ak ka= saaʔ =fa \\
	\n-say {\ka}= what ={\fa} \\
\glt `they said nothing happened (to them)'}

\ex{Adi Bani: \ve{Nok keun{\Q} aa te ean{\Q} ee nasoin.}}\vspace{-4pt}
\exi{1.41}[]{\glll
	n-ok keunʔ =at, eanʔ=ee na-soin \\
	n-oka kenuʔ =te enoʔ=ee na-soni \\
	{\n-\ok} fortune ={\te} door={\ee} \na-open \\
\glt `It's fortunate, the door opened'}

\ex[R:]{\ve{Neu reko.}}
\exi{1.43}[]{\glll
	neu reko\\
	neu reko \\
	already good \\
\glt `Well, good.'}

\end{exe}