
\section{Moo{\Q}hitu{\Q}}\label{sec:MooHit}

\subsection{Metadata}
\wg\begin{tabular}{lL{.67\textwidth}}
File-name:					& aaz20120715{\_}04\\
Archive-link:				& \url{http://catalog.paradisec.org.au/collections/OE1/items/aaz20120715_04}\\
Original Name:			& aaz-20120715-4-Nekmese-KusnawiBani-2\\
Language:						& Amarasi [aaz] \\
Dialect:						& Kotos; Koro{\Q}oto hamlet \\
Location:						& Nekmese{\Q}, Amarasi Selatan, Kupang \\
Date:								& 15/07/2015 \\
Speaker(s):					& Taniel Feni, Kusnawi Bani, Heronimus Bani\\
Recorded by: 				& Daniel Kaufman, Heronimus Bani\\
Transcribed by:			& Yedida Ora\\
Interlinear by:			& Owen Edwards \\
Indonesian/Kupang:	& Yedida Ora\\
Free English by:		& Owen Edwards\\
Genre:							& folk-tale\\
Summary:						& Story of Moo{\Q}hitu{\Q}, a mythical snake who created the world\\
Video:							& \url{https://www.youtube.com/watch?v=Z_2D9WhYuuM&list=PLcXFPx-z7B0q_2Ns3iYHigEY77DG4kXSU&index=15}\\
\end{tabular}

\subsection{Notes}
The original recording contains a number of separate stories.
I present here only the first story:
the story of Moo{\Q}hitu{\Q}.
This may be a conflation of two separate myths.
The first is a creation myth about how Moo{\Q}hitu{\Q}, a snake-like being,
separates the sky, land, and sea.
The second myth is about how a python copulated with women
and is, perhaps, an explanation for the origin of men.
In Timorese thought the human world cannot exist without
women, who are the source of life.
This leaves unexplained the origin of men.\footnote{
		Thanks goes to James Fox for this explanation.}

The information in the myth is incredibly dense
in parts and certain information is left unexplained and/or
assumed to be known by the hearers.
Footnotes provide additional explanations
as well as possible alternate readings.

\subsection{The text}
\begin{exe}

%\ex{\ve{Neno naa re{\Q} paha{\gap}{\Q}piin ii,}}
%\exi{0.00}{\glll	neno naa reʔ paha{\gap}ʔpiin-n=ii\\
%									neno naa reʔ paha{\gap}ʔpina-n=ii\\
%									day {\naa} {\reqt} land{\gap}below-{\N} ={\ii}\\
%						\glt	`At that time the world{\ldots}'}

%\ex[R:]{\ve{mee, mee, mee,}
%				\glt `Hey, wait.'}

%\ex[T:]{\ve{Oi, oi oi oi}
%				\glt `Oh.'}

%\ex[R:]{\ve{\sf{Yep.}}
%			\glt `Yep (start).'}

\ex{\ve{Neno naa pah-a{\Q}pinan ia, ankobub on bare mese{\Q}.}}
\exi{0.05}{\glll	neno naa paha{\gap}ʔpina-n ia a|n-kobub on bare meseʔ \\
									neno naa paha{\gap}ʔpina-n ia {\a}n-kobub on bare meseʔ \\
									day {\naa} land{\gap}below-{\N} {\ia}	\a\n-piled.up {\on} place one \\
						\glt	`At that time this world was all piled up in one place' }

\ex{\ve{Ka nmui{\Q} fa mainuan.}}
\exi{0.10}{\glll	ka= n-muiʔ =fa maanuan\\
									ka= n-muʔi  =fa mainuan  \\
									{\ka}= \n-exist ={\fa} open \\
						\glt	`there was no openness/space.' }

\ex{\ve{Ka nmui{\Q} fa oe.}}
\exi{0.13}{\glll	ka= n-muiʔ =fa oe\\
									ka= n-muʔi  =fa oe\\
									{\ka}= \n-exist ={\fa} water\\
						\glt	`There was no water.' }

\ex{\label{ex:0.14}\ve{{\Q}Aa{\Q} oe ji nmees, ka tiit fa auf meto{\Q}.}}\footnote{
				Line \qf{ex:0.14}: The meaning of initial phonetic [ʔaːʔ] is currently unclear.
				It may be from the root \ve{{\rt}aʔa} `ritual speech, poetic speech' and
				could, perhaps, mean something like `I am telling it according to tradition'.
				In some other varieties of Meto \ve{{\rt}aʔa} simply means `speak, talk'.}
\exi{0.14}{\glll	ʔ-aaʔ oo\j=ii n-mees ka= tiit =fa auf metoʔ\\
									ʔ-aʔa oe=ii n-mese ka= tita =fa afu metoʔ\\
									\q-speak water={\ii} \n-alone {\ka}= exist ={\fa} ground{\M} dry\\
						\glt	`I say, there was only water, there was no dry ground.'}

\ex{\label{ex:0.17}\ve{Afu ma neno nmanaa{\Q}.}}
\exi{0.17}[K:]{\glll	afu =m neno n-ma-naaʔ\\
											afu =ma neno n-ma-naʔa \\
											ground =and sky \n-\mak-hold \\
								\glt	`The ground and sky held on to one another.' \footnote{
													Line \qf{ex:0.17} is spoken by Kusnawi Bani.}}

\newpage
\ex{\ve{Afu ma neno nmanaa{\Q}. Meis{\Q}ookn ii nnaa{\Q}.}}
\exi{0.18}[T:]{\glll	afu =m neno n-ma-naaʔ meisʔookn=ii n-naaʔ\\
											afu =ma neno n-ma-naʔa  meisʔokan=ii n-naʔa \\
											ground =and sky \n-\mak-hold dark={\ii} \n-hold \\
								\glt	`The ground and sky held on to one another, darkness held (fast).' }

\ex{\label{ex:0.20}\ve{Tapi re{\Q} kauna{\Q} ia, in nmoni nbi oe ji naan ii.}}
\exi{0.20}{\begin{xlist}
	\ex{\glll	\sf{tapi} ahh reʔ kaunaʔ ia \\
						\sf{tapi} {} reʔ kaunaʔ ia \\
						but {} {\req} snake {\ia} \\}
	\ex{\glll	iin n-moni n-bi oo\j=ee naan-n=ii, \\
						ini n-moni n-bi oe=ee nana-n=ii \\
						{\iin} \n-live \n-loc water={\ee} inside-{\N}={\ii} \\
			\glt	`but as for this snake, he was living inside the water,'}\footnote{
								Line \qf{ex:0.20}:	The snake is Moo{\Q}hitu{\Q}.}
\end{xlist}}

\ex{\ve{Noki-noki te, in naskeke nfena nhake ma,}}
\exi{0.24}{\gll	noki-noki =te iin na-skeke n-fena n-hake =ma\\
									eventually ={\te} {\iin} \na-suddenly \n-rise \n-stand{\U} =and\\
						\glt	`after a while, he suddenly stood up and' }

\ex{\ve{in nfeen es mee te neon gui natsiri{\Q}, natsiri{\Q}, sampai in ntea re{\Q} aat neno nee msa{\Q}, in natuin ee ma,}}\label{ex:0.27}
\exi{0.27}{\begin{xlist}
	\ex{\glll	iin, iin n-feen es mee =t nee{\ng}gw=ii na-tsiriʔ na-tsiriʔ \\
						ini ini n-fena  es mee =te neno=ii na-tsiriʔ na-tsiriʔ \\
						{\iin} {\iin} \n-rise {\et} where ={\te} sky={\ii} \n-spread \n-spread \\
			\glt	`as he went up to somewhere, the sky spread and spread (upwards)'\footnote{
							Lines \qf{ex:0.27} and \qf{ex:0.34} explain how Moo{\Q}hitu{\Q}
							pushed up the sky, thus separating it from the water.}}

	\ex{\glll	\sf{sampe} iin n-tea reʔ aat neno nee msaʔ\\
						\sf{sampe} ini n-tea reʔ ata neno nee msaʔ \\
						until {\iin} \n-up.to {\req} up sky {\nee} also \\
			\glt `until when he arrived at (the place) where the top of the sky also is,'}

	\ex{\glll	iin na-tuin=ee =ma\\
						ini na-tuin=ee =ma\\
						{\iin} \n-follow={\eeV} =and\\
			\glt `he followed it and,'}

\end{xlist}}

\newpage
\ex{\ve{Anhake {\Q}roo-roo te, es naa neon goe na{\Q} anmana{\Q}a ma,}}\label{ex:0.34}
\exi{0.34}{\begin{xlist}
	\ex{\glll	a|n-hake ʔro{\tl}roo =t \\
						n-hake ʔro{\tl}roo =te \\
						\n-stand {\prd}far ={\te} \\}
	\ex{\glll	es naa nee{\ng}gw=ee naʔ a|n-ma-naʔa =ma	\\
						es naa neno=ee naʔ {\a}n-ma-naʔa =ma	\\
						{\et} {\naa} sky={\ee} then \a\n-\mak-hold{\U} =and\\
			\glt	`when he had stood up for a long time, at that place
							only then the sky held fast (in relation to him) and,' }
\end{xlist}}

\ex{\ve{na{\Q} nsanu nfani kre{\Q}o-kre{\Q}o ma nfani nbi in baran.}}
\exi{0.37}{\begin{xlist}
	\ex{\glll	naʔ n-sanu n-fani kreʔo{\tl}kreʔo =ma\\
						naʔ n-sanu n-fani kreʔo{\tl}kreʔo =ma\\
						then \n-descend \n-return {\frd}bit =and\\}
	\ex{\glll	n-fani n-bi iin bara-n\\
						n-fani n-bi ini bara-n\\
						\n-return \n-{\bi} {\iin} place-{\N}\\
			\glt	`then (he) went back down bit by bit and returned to his place'}
\end{xlist}}

\ex{\ve{Nfani nbi in baarn ii. In baarn ee et oe je nanan.}}
\exi{0.40}{\begin{xlist}
	\ex{\glll	n-fani n-bi iin baar-n=ii \\
						n-fani n-bi ini bara-n=ii \\
						\n-return \n-{\bi} {\iin} place-\N={\ii}  \\}
	\ex{\glll	iin baar-n=ee et oo\j=ee nana-n\\
						ini bara-n=ee et oe=ee nana-n\\
						{\iin} place-\N={\ee} {\et} water={\ee} inside-{\N}\\
			\glt	`(He) went back to his place. His place was inside the water.'}
\end{xlist}}

\ex{\ve{Nbi-bi oe je naan ee onai te, anmo{\Q}e ma npoi jeen anbi meto{\Q}.}}
\exi{0.43}{\begin{xlist}
	\ex{\glll	n-bi{\tl}bi oo\j=ee naan-n=ee {onai =te}\\
						n-ni{\tl}bi oe=ee nana-n=ee {onai =te}\\
						\n-{\prd}loc water={\ee} inside\N={\ee} then\\
			\glt	`after he had been in the water for a while then,'}
	\ex{\glll	a|nmoʔe =ma npoo\j=ena n-bi metoʔ\\
						{\a}n-moʔe =ma n-poi=ena n-bi metoʔ \\
						{\a}\n-do{\U} and \n-exit={\een} \n-{\bi} dry \\
			\glt \lh{\a}`(he) made (dry land) and went out onto dry land,'}
\end{xlist}}

\ex{\ve{Npoi nbi meot{\Q} ee onai te, in ka nmui{\Q} fa bare he natua ma,}}\label{ex:0.47}
\exi{0.47}{\begin{xlist}
	\ex{\glll	n-poi n-bi meotʔ=ee {onai =te}\\
						n-poi n-bi metoʔ=ee {onai =te}\\
						\n-exit \n-{\bi} dry={\ee} then\\}
	\ex{\glll	iin ka= n-muiʔ =fa bare he na-tua =m\\
						ini ka= n-muʔi =fa bare he na-tua =ma\\
						{\iin} {\ka}= \n-exist ={\fa} place {\he} \na-live =and\\
			\glt	`having gone out onto the dry land, he didn't have a place to live and,' }
\end{xlist}}

\ex{\ve{he natua te, baer mainuan.}}
\exi{0.51}{\glll	he na-tua =te he-- baer mainuan\\
									he na-tua =te {} bare mainuan\\
									{\he} \n-live top {} place{\M} open\\
						\glt	`he would (have to) live in an open place,' }

\ex{\ve{Natua te, baer ko{\Q}u.}}
\exi{0.53}{\glll	na-tua =te baer koʔu\\
									na-tua =te bare koʔu \\
									\na-live ={\te} place{\M} big \\
						\glt	`live in a big place,' }

\ex{\ve{Akhirnya, naim naan baer jes am namaika{\Q} nbi Smara{\Q} tunan.}}\label{ex:0.57}
\exi{0.57}{\begin{xlist}
	\ex{\glll	\sf{ahirɲa} ahh n-aim naan baar\j=esa =m namaikaʔ an--,  \\
						\sf{ahirɲa} {} n-ami naan bare=esa =ma na-maikaʔ {} \\
						in.the.end {} \n-look.for {\naan} place={\es} =and \na-stay {}  \\
			\glt `in the end, (he) looked there for a place and settled,'}
	\ex{\glll	na-maikaʔ n-bi Smaraʔ tunan\\
						na-maikaʔ n-bi Smaraʔ tuna-n\\
						\na-stay \n-{\bi} Smara{\Q} top-{\N}\\
			\glt `(he) settled on top of Smara{\Q}.' (a headland on the southern coast)}
\end{xlist}}

\ex{\ve{Namaika{\Q} nbi Smara{\Q} tuun ee ma,}}
\exi{1.01}{\glll	na-maikaʔ n-bi Smaraʔ tuun-n=ee =ma\\
									na-maikaʔ n-bi Smaraʔ tuna-n=ee =ma\\
									\n-stay \n-{\bi} Smara{\Q} top-\N={\ee} =and\\
						\glt	`settled on top of Smara{\Q} and' }

\newpage
\ex{\ve{In re{\Q} fee mnais unu{\Q} ma nai{\Q} unu{\Q} nnao nakbatun anbi tasi.}}
\exi{1.05}{\begin{xlist}
	\ex{\glll	iin reʔ fee mnais unuʔ =ma naiʔ unuʔ \\
						ini reʔ fee mnasiʔ unuʔ =ma naʔi unuʔ \\
						3sg {\req} wife old{\M} past and grandfather{\M} past \\
			\glt	`he (was) where old women of past times and old men of past times'}
	\ex{\glll	n-nao na-kbatu=n a|n-bi tasi\\
						n-nao na-kbatu=n {\a}n-bi tasi\\
						\n-go \n-shell={\einV} {\a}\n-{\bi} sea\\
			\glt	`went and collected shells by the sea,'}
\end{xlist}}

\ex{\label{ex:1.07}\ve{Ntea uab reu{\Q}f ii jena ma.}}
\exi{1.07}{\glll	n-tea uab reuʔf=ii\j=ena =ma ahh\\
					n-tea uaba reʔuf=ii=ena =ma {}\\
					\n-arrive speech{\M} bad={\ii}={\een} =and {}\\
		\glt	`he went there (to do things which are) bad to talk about.'\footnote{
								Line \qf{ex:1.07} is obscure.
								It probably foreshadows that the actions
								Moo{\Q}hitu{\Q} is about to carry out are bad to talk about.	
								Just after this line Kusnawi Bani says one or two inaudible words.}}

\ex{\ve{In fee je msa{\Q} nua sin huma{\Q} mese{\Q} tapi bifee je bifee biasa.}}\label{ex:1.10}
\exi{1.10}{\begin{xlist}
	\ex{\glll	iin fee\j=ee msaʔ nua sin humaʔ meseʔ \\
						ini fee=ee msaʔ nua sin humaʔ meseʔ \\
						{\iin} wife={\ee} also two {\siin} kind one \\}
	\ex{\glll	\sf{tapi} bifee\j=ee bifee \sf{biasa}\\
						\sf{tapi} bifee=ee bifee \sf{biasa}\\
						but woman={\ee} woman normal\\
		\glt `he and his wife were the same, but the woman was a normal woman'\footnote{
						Line (\ref{ex:1.10}a):
						The wife of Moo{\Q}hitu{\Q} has not been introduced before.
						The reference to her being a normal woman is probably a contrast with
						the fact that Moo{\Q}hitu{\Q} is a snake-like being.}}
\end{xlist}}

\ex{\ve{Cuma atoin{\Q} ein ee nteek ee te nak: Moo{\Q}hitu{\Q}.}}\label{ex:1.14}
\exi{1.14}{\glll	\sf{suma} atoinʔ=ein=ee n-teek=ee =te n-ak: Mooʔhituʔ. \\
									\sf{suma} atoniʔ=eni=ee n-teka=ee =te n-ak Mooʔhituʔ \\
									only man=\ein={\ee}	\n-call={\eeV} ={\te} {\n-\ak} Moo{\Q}hitu{\Q} \\
						\glt	`Only the men called him Moo{\Q}hitu{\Q}.'\footnote{
										Line (\ref{ex:1.14}): that only the men call him Moo{\Q}hitu{\Q} is
										probably a reference to his phallic shape and/or nature.}}

\newpage
\ex{\ve{Moo{\Q}hitu{\Q} re{\Q} naan, in kauna{\Q}.}}
\exi{1.16}{\glll	Mooʔhituʔ reʔ naan iin kaunaʔ\\
									Mooʔhituʔ reʔ naan ini kaunaʔ \\
									Moo{\Q}hitu{\Q} {\req} {\naan} {\iin} snake \\
						\glt `That Moo{\Q}hitu{\Q} was/is a snake.' }

\ex{\ve{Kauna{\Q}, mes huum atoni{\Q} on re{\Q} hit.}}
\exi{1.19}{\glll	kaunaʔ mes huum atoniʔ on reʔ hiit\\
									kaunaʔ mes humaʔ atoniʔ on reʔ hiti\\
									snake but face{\M} man like {\reqt} {\hiit}\\
						\glt `(He was) a snake but (he had) a human face/form like us.' }

\ex{\ve{Cuma in kaan ee es re{\Q} nai{\Q} Moo{\Q}hitu{\Q}.}}
\exi{1.22}{\glll	\sf{ʧuma} iin kaan-n=ee eseʔ {} naiʔ Mooʔhituʔ\\
									\sf{ʧuma} ini kana-n=ee esa reʔ naiʔ Mooʔhituʔ \\
									only {\iin} name\N={\ee} {\esc} {\req} {\naiq} Moo{\Q}hitu{\Q} \\
						\glt	`It was only his name which was Moo{\Q}hitu{\Q}.' }

\ex{\ve{In nfena nhake te, mo{\Q}ok hitu, mes ho muhiin he moo{\Q}k es ate, he mnaun{\Q} ii ba{\Q}uk.}}
\exi{1.24}{\begin{xlist}
\ex{\glll	iin n-fena n-hake =t moʔok hitu\\
					ini n-fena n-hake =te moʔok hitu \\
					{\iin} \n-rise \n-stand{\U} ={\te} section seven{\U}\\
		\glt `If he stood up (there would be) seven sections,'\footnote{
						Line \qf{ex:1.24}: an explanation of the name Moo{\Q}hitu{\Q}.
						It is from the root \ve{moʔok}
						`section of something long; joints of a finger, nodes of bamboo'
						and \ve{hitu} `seven'.}}\label{ex:1.24}
\ex{\glll	mes hoo mu-hiin he mooʔk=esa =t he mnaunʔ=ii baʔuk\\
					mes hoo mu-hini he moʔok=esa =te he mnanuʔ=ii baʔuk \\
					but {\hoo} \muu-know {\he} section=one ={\te} {\he} long={\ii} several \\
		\glt `but if you know (the length of) one section, it would be very long.'\footnote{
						Line \qf{ex:1.28}: Moo{\Q}hitu{\Q} is so long, that it is hard to know
						how long even a single section of him would be.}}\label{ex:1.28}
\end{xlist}}

\ex{\ve{Akhirnya in nhake nbi Smara{\Q} tuun ee te, bifee ngguin nakbatun nbi nahen nee kboa{\Q} ko{\Q}u.}}
\exi{1.30}{\begin{xlist}
	\ex{\glll	\sf{ahirɲa} iin n-hake n-bi Smaraʔ tuun-n=ee =te  \\
						\sf{ahirɲa} ini n-hake n-bi Smaraʔ tuna-n=ee =te  \\
						in.the.end {\iin} \n-stand \n-{\bi} Smara{\Q} top-\N={\ee} ={\te}  \\
			\glt	`In the end while he was standing on top of Smara{\Q},'}
	\ex{\glll	bifee={\ng}gwein na-kbatu=n n-bi nahen nee kboaʔ koʔu\\
						bifee=eni na-kbatu=n n-bi nahe-n nee kboʔes koʔu \\
						woman={\ein} \na-shell-{\einV} \n-{\bi} down-{\N} {\nee} clump{\M} big \\
			\glt	`the women were collecting sea shells down there in a big clump.'}
\end{xlist}}

\ex{\ve{In naim ranan huma{\Q}-huma{\Q} akhirnya,}}
\exi{1.34}{\glll	iin n-aim ranan humaʔ{\tl}humaʔ \sf{ahirɲa}\\
									ini n-ami  ranan humaʔ{\tl}humaʔ \sf{ahirɲa}\\
									{\iin} \n-look.for road {\frd}kind in.the.end\\
						\glt	`he was looking for various ways, and in the end,' }

\ex{\ve{permisi, ma re{\Q} in nnao npeo{\Q} afu, nmoe{\Q} on umeek ji ma,}}
\exi{1.37}{\begin{xlist}
	\ex{\glll	\sf{parmisi} =m reʔ iin nahh hihh \\
						\sf{parmisi} =ma reʔ ini {} {} \\
						excuse.me =and {\req} {\iin} {} {}  \\
			\glt	`excuse me, and it was where he, uhh'\footnote{
							Line \qf{ex:1.37a}: The narrator uses \it{permisi}
							to signal that he is about to talk of sexual matters.}}\label{ex:1.37a}
	\ex{\glll	iin n-nao n-peoʔ afu n-moaʔ on umeek\j=ii =ma\\
						ini n-nao n-peʔo  afu n-moʔe on umeke=ii =ma\\
						{\iin} \n-go \n-go.by ground \n-do like wolf.snake={\ii} =and\\
			\glt	`he went along the ground he was doing it like the wolf snake,'\footnote{
							Line \qf{ex:1.37b}: \ve{umeke} = \it{Lycodon sp.},
							a kind of non-poisonous red snake}}\label{ex:1.37b}
\end{xlist}}

\ex{\ve{in tuan ii nbi ata {\Q}toe{\Q}f ee tuun ee te, in aon ee es anaot ma,}}
\exi{1.42}{\begin{xlist}
	\ex{\glll	iin, iin tua-n=ii n-bi ata ʔtoeʔf=ee \hspace{20mm} tuun-n=ee =t \\
						ini ini  tua-n=ii n-bi ata ʔtoʔef=ee {} tuna-n=ee =te \\
						{\iin} {\iin} self-\N={\ii} \n-{\bi} up mountain={\ee} {} top-\N={\ee} ={\te} \\
			\glt `while his self was up on top of the mountain,'}
	\ex{\glll	iin ao-n=ee ees a-nao-t =ma\\
						ini ao-n=ee ees a-nao-t =ma\\
						{\iin} body-\N={\ee} {\esc} {\at}-go-{\at} =and\\
			\glt `his body (was the) one which went and'}
\end{xlist}}

\newpage
\ex{\ve{in nkoin re{\Q} bifee ngguin nbi tasi.}}
\exi{1.46}{\glll	iin n-koin reʔ bifee={\ng}gwein n-bi tasi\\
									ini n-koni  reʔ bifee=eni n-bi tasi \\
									{\iin} \n-copulate {\reqt} woman={\ein} \n-{\bi} sea \\
					\glt `he copulated with those women at the sea.'}

\ex{\ve{Ka nakeon fa.}}
\exi{1.48}{\glll	ka= na-keo=n =fa\\
									ka= na-keo=n =fa\\
									{\ka}= \n-aware={\einV} ={\fa}\\
						\glt `They weren't aware of it.' }

\ex{\ve{In a{\Q}maen ii es anpeo{\Q} afu. In nmoe{\Q} jon on kaun{\Q} ii ma, nnonok anpeo{\Q} auf gui ma,}}
\exi{1.49}{\begin{xlist}
	\ex{\glll	iin, iin, ina ʔ-mae-n=ii esa n-peoʔ afu \\
						ini ini ina ʔ-mae-n=ii esa n-peʔo  afu \\
						{\iin} {\iin} {\iin} {\qq}-shame-\N={\ii} {\esc} \n-go.by ground  \\
			\glt `his private part was the one which went along the ground.'}
	\ex{\glll	iin n-mooʔ\j=oo-n on kaunʔ=ii =ma \\
						ini n-moʔe=oo-n on kaunaʔ=ii =ma \\
						{\iin} \n-do={\oo-\N} like snake={\ii} =and  \\
			\glt `he made himself like a snake and,'}
	\ex{\glll	n-nonok a|n-peoʔ aafgw=ii =ma\\
						n-nonok {\a}n-peʔo afu=ii =ma\\
						\n-crawl \a\n-go.by ground={\ii} =and\\
			\glt `crawled along the ground and,'}
\end{xlist}}

\ex{\ve{nnaob antama ma, in nkoin re{\Q} bifee ngguin. Sin nakeon fa.}}
\exi{1.55}{\begin{xlist}
	\ex{\glll	n-nao-b a|n-tama =m, iin n-koin reʔ bifee={\ng}gwein\\
						n-nao-b {\a}n-tama =ma ini n-koni  reʔ bifee=eni \\
						\n-go-{\b} \a\n-enter{\U} =and {\iin} \n-copulate {\reqt} woman={\ein} \\
			\glt `(he) made (it) go (and) penetrate and he copulated with the women'}
	\ex{\glll	siin ka= na-keo=n =fa\\
						sini ka= na-keo=n	=fa \\
						{\siin} {\ka}= \na-aware={\einV}	={\fa} \\
			\glt	`they weren't aware of it'}
\end{xlist}}

%\ex{\ve{Nbi tais je ma,}}
%\exi{1.58}{\glll	n-bi taas\j=ee =m\\
%									n-bi tasi=ee =ma\\
%									\n-{\bi} sea={\ee} =and\\
%						\glt `He was at the sea and {\ldots}' }

\end{exe}
