\addchap{Appendix A: Interlinear texts}
{
Below are three glossed and interlinearised texts, all of which are part of the PLRN text corpus (\sectref{sec:1.6.2}). The first text is a children’s story, written during a writer’s workshop in 1984. The second text is a dramatic retelling of a traditional story by Luis Avaka Paoa (‘Papa Kiko’), a renowned storyteller; a very short version of the same story was published by \citet{Blixen1974}. Number three is a description of a fishing trip, composed as part of a schoolbook containing stories about traditional activities on Rapa Nui.

\section*{\textit{Te tātane taŋata} – The devilman (R215)}\label{sec:a.1}
\addcontentsline{toc}{section}{\textit{Te tātane taŋata} – The devilman (R215)}

by Virginia Haoa Cardinali (\citealt{HaoaCardinali1984}; \citealt[Vol. 3:118–119]{WeberWeber1990Mai})

\bigskip\gll
\textbf{\textup{01}} Ko ahiahi pō {\ꞌ}ā. \textbf{\textup{02}} Te ŋā poki nei e rua: e tahi ko Kihi te {\ꞌ}īŋoa, e rima ō{\ꞌ}ona matahiti, poki teatea, ritorito, he tau nō; te rua poki ko {\ꞌ}Atera te {\ꞌ}īŋoa, e hitu matahiti, me{\ꞌ}e rakerake a vērā. \textbf{\textup{03}} Mahana tāpati {\ꞌ}i te pō, he haka rivariva e te rāua māmā mo ma{\ꞌ}u i te kai ki te rāua koro. \textbf{\textup{04}} Te aŋa iŋa {\ꞌ}i te kona motore mo haka pura i te mōrī paurō te mahana {\ꞌ}i te pō. ~\textbf{\textup{05}} Ko ha{\ꞌ}a{\ꞌ}ī {\ꞌ}ā tū kai era e nua {\ꞌ}i roto i te pani e tahi, {\ꞌ}ai ka viri rō hai pānio teatea.\\ 
~ \textsc{prf} evening night \textsc{cont} ~  \textsc{art} \textsc{pl} child \textsc{prox} \textsc{num} two \textsc{num} one \textsc{prom} Kihi \textsc{art} name \textsc{num} five \textsc{poss.3sg.o} year child white clear \textsc{ntr} pretty just \textsc{art} two child \textsc{prom} Atera \textsc{art} name \textsc{num} seven year thing bad:\textsc{red} \textsc{prop} poor\_thing ~ day Sunday at \textsc{art} night \textsc{ntr} \textsc{caus} good:\textsc{red} \textsc{ag} \textsc{art} \textsc{3pl} mother for carry \textsc{acc} \textsc{art} food to \textsc{art} \textsc{3pl} Dad ~  \textsc{art} work \textsc{nmlz} at \textsc{art} place engine for \textsc{caus} shine \textsc{acc} \textsc{art} light every \textsc{art} day at \textsc{art} night ~ \textsc{prf} fill \textsc{cont} \textsc{dem} food \textsc{dist} \textsc{ag} Mum at inside at \textsc{art} pan \textsc{num} one \textsc{subs} \textsc{cntg} wrap \textsc{emph} with towel white\\ 
 
\medskip\glt
\textbf{\textup{01}} It was evening. ~\textbf{\textup{02}} There were two children: one was called Kihi, she was five years old, a fair child, light-skinned, just pretty; the other child was called Atera, seven years old, the poor one was ugly. ~\textbf{\textup{03}} On Sunday night, their mother made preparations to take food to their father. ~\textbf{\textup{04}} He worked at the electrical power plant every day at night. ~\textbf{05} Mother had put the food in a pan and wrapped it in a white towel.
%\todo[inline]{It may be good to add blank lines before the free translation. What do you think? One way would be to use glll and keep the third line empty}
~

\bigskip\gll
\textbf{\textup{06}} He hahari i te pū{\ꞌ}oko o tū ŋā poki era, he haka uru i te paratoa, {\ꞌ}ai ka va{\ꞌ}ai rō tū pū{\ꞌ}ahu kai era ki te poki {\ꞌ}atariki era. ~\textbf{\textup{07}} He kī ia e nua: \textbf{\textup{08}} —Ka ma{\ꞌ}u hiohio te kai ena mā koro. ~\textbf{\textup{09}} E ma{\ꞌ}u hiohio {\ꞌ}i roto i tu{\ꞌ}u rima; ta{\ꞌ}e mo haka pakō tā{\ꞌ}ue. ~\textbf{\textup{10}} E tahi rima ena ō{\ꞌ}ou ka ma{\ꞌ}u ena i te kai, e tahi i te rima o tu{\ꞌ}u taina; ararua nō {\ꞌ}ina ko haka pako{\ꞌ}o ki te hora hope{\ꞌ}a.\\
~ \textsc{ntr} comb \textsc{acc} \textsc{art} head of \textsc{dem} \textsc{pl} child \textsc{dist} \textsc{ntr} \textsc{caus} dress \textsc{acc} \textsc{art} jacket \textsc{subs} \textsc{cntg} give \textsc{emph} \textsc{dem} bundle food \textsc{dist} to \textsc{art} child firstborn \textsc{dist} ~ \textsc{ntr} say then \textsc{ag} Mum ~ ~~~~\textsc{imp} carry strong \textsc{art} food \textsc{med} for\textsc{.a} Dad ~ \textsc{exh} carry strong at inside at \textsc{poss.2sg.o} hand \textsc{conneg} for \textsc{caus} loose by\_chance ~ \textsc{num} one hand \textsc{med} \textsc{poss.2sg.o} \textsc{cntg} carry \textsc{med} \textsc{acc} \textsc{art} food \textsc{num} one \textsc{acc} \textsc{art} hand of \textsc{poss.2sg.o} sibling the\_two just \textsc{neg} \textsc{neg.ipfv} \textsc{caus} loose to \textsc{art} time last\\
 
\medskip\glt
\textbf{\textup{06}} She combed the children’s hair, put on their jacket and gave the bundle of food to the oldest one. ~\textbf{\textup{07}} Then mother said, “Hold the food for Dad firmly. ~\textbf{\textup{09}} Hold it tight in your hands; you must not let go of it. ~\textbf{\textup{10}} With one hand carry the food, with the other hold on to you sister’s hand; don’t ever let go of each other.”

~

\bigskip\gll
\textbf{\textup{11}} He e{\ꞌ}a ia tū ŋā poki era a te vāeŋa o te ara he ha{\ꞌ}ere he iri. \textbf{\textup{12}} {\ꞌ}Ina e tahi vānaŋa rere ararua; ko momou {\ꞌ}ā {\ꞌ}i te ri{\ꞌ}ari{\ꞌ}a {\ꞌ}i te pōhāhā. \textbf{\textup{13}} {\ꞌ}E {\ꞌ}i te ri{\ꞌ}ari{\ꞌ}a {\ꞌ}i tū pōhāhā era, he pa{\ꞌ}ahia te rima ararua. \textbf{\textup{14}} He ha{\ꞌ}amata te paŋaha{\ꞌ}a o te rāua va{\ꞌ}e, pē ira {\ꞌ}ā tū kai era mā koro.\\
~ \textsc{ntr} go\_out then \textsc{dem} \textsc{art} child \textsc{dist} by \textsc{art} middle of \textsc{art} road \textsc{ntr} walk \textsc{ntr} ascend ~ \textsc{neg} \textsc{num} one word fly the\_two \textsc{prf} \textsc{pl}:quiet \textsc{cont} at \textsc{art} fear at art dark  ~ and at \textsc{art} fear at \textsc{dem} dark \textsc{dist} \textsc{ntr} sweat \textsc{art} hand the\_two  ~ \textsc{ntr} begin \textsc{art} heavy of \textsc{art} \textsc{3pl} foot like \textsc{ana} \textsc{ident} \textsc{dem} food \textsc{dist} for\textsc{.a} Dad\\

\medskip\glt
\textbf{\textup{11}} The children went out by the middle of the road and walked up. ~\textbf{\textup{12}} The two didn’t say anything; they were silent, because they were afraid in the dark. ~\textbf{\textup{13}} And because they were afraid in the dark, both had sweaty hands. ~\textbf{\textup{14}} Their feet started to feel heavy, and the food for Dad felt heavy as well.


~

\bigskip\gll
\textbf{\textup{15}} Ko hinihini {\ꞌ}ā te hora e iri era, he {\ꞌ}ui ia e {\ꞌ}Atera ki a Kihi: \textbf{\textup{16}} —¿E ko haŋa {\ꞌ}ō rō koe mo ma{\ꞌ}u i te kai nei mā koro? \textbf{\textup{17}} Ka ma{\ꞌ}u {\ꞌ}iti{\ꞌ}iti koe mo haka ora {\ꞌ}iti{\ꞌ}iti o tō{\ꞌ}oku rima. \textbf{\textup{18}} Terā ka kī e Kihi: ~\textbf{\textup{19}} ––¡Ko aha {\ꞌ}ā i a koe! ~\textbf{\textup{20}} ¿Ko haŋa {\ꞌ}ana {\ꞌ}ō pēaha koe mo pako{\ꞌ}o tō{\ꞌ}oku rima? ~\textbf{\textup{21}} He momou haka{\ꞌ}ou ararua, {\ꞌ}ai ka iri nō.\\
~ \textsc{prf} delay:\textsc{red} \textsc{cont} \textsc{art} time \textsc{ipfv} ascend \textsc{dist} \textsc{ntr} ask then \textsc{ag} Atera to \textsc{prop} Kihi ~ ~~~~\textsc{ipfv} \textsc{neg.ipfv} want really \textsc{emph} \textsc{2sg} for carry \textsc{acc} \textsc{art} food \textsc{prox} for\textsc{.a} Dad  ~ \textsc{imp} carry little \textsc{2sg} for \textsc{caus} rest little of \textsc{poss.1sg.o} hand ~ then \textsc{cntg} say \textsc{ag} Kihi ~ ~~~~~\textsc{prf} what \textsc{cont} at \textsc{prop} \textsc{2sg} ~ ~\textsc{prf} want \textsc{cont} really perhaps \textsc{2sg} for loose \textsc{poss.1sg.o} hand ~ \textsc{ntr} \textsc{pl}:quiet again the\_two \textsc{subs} \textsc{cntg} ascend just\\

\medskip\glt
\textbf{\textup{15}} When they had walked for a while, Atera asked Kihi, ~\textbf{\textup{16}} “Don’t you want to carry the food for Dad? ~\textbf{\textup{17}} Carry it for a little while, so my hand can rest a little.” ~\textbf{\textup{18}} Kihi said ~\textbf{\textup{19~~}}“What are you thinking! ~\textbf{\textup{20}} Do you really want to let go of my hand?” ~
\textbf{\textup{21}} The two were silent again, while they kept going up.


~

\bigskip\gll
\textbf{\textup{22}} I roaroa haka{\ꞌ}ou era te hora, he kī haka{\ꞌ}ou e {\ꞌ}Atera: \textbf{\textup{23}} —¡Mo ta{\ꞌ}e haŋa ō{\ꞌ}ou mo ma{\ꞌ}u i te kai nei mā koro, he haka pakō e au tu{\ꞌ}u rima nei! \textbf{\textup{24}} He oho ia a Kihi mo taŋi, {\ꞌ}ai ka kī rō: ~\textbf{\textup{25}} —¡E te taina riva ō{\ꞌ}oku ē, {\ꞌ}ina koe ko haka pakō i tō{\ꞌ}oku rima {\ꞌ}o ri{\ꞌ}ari{\ꞌ}a rō au {\ꞌ}i te tātane! \textbf{\textup{26}} Ki hāhine nō tāua mo tu{\ꞌ}u ana ma{\ꞌ}u iho e au te kai.\\
~ \textsc{pfv} long again \textsc{dist} \textsc{art} time \textsc{ntr} say again \textsc{ag} Atera ~ ~~~~if \textsc{conneg} want \textsc{poss.2sg.o} for carry \textsc{acc} \textsc{art} food \textsc{prox} for\textsc{.a} Dad \textsc{ntr} \textsc{caus} loose \textsc{ag} \textsc{1sg} \textsc{poss.2sg.o} hand \textsc{prox} ~ \textsc{ntr} go then \textsc{prop} Kihi for cry \textsc{subs} \textsc{cntg} say \textsc{emph} ~ ~~~~\textsc{voc} \textsc{art} sibling good \textsc{poss.1sg.o} \textsc{voc} \textsc{neg} \textsc{2sg} \textsc{neg.ipfv} \textsc{caus} loose \textsc{acc} \textsc{art} hand lest fear \textsc{emph} \textsc{1sg} at \textsc{art} devil  ~ when near just \textsc{1du.incl} for arrive \textsc{irr} carry just\_then \textsc{ag} \textsc{1sg} \textsc{art} food\\

\medskip\glt
\textbf{\textup{22}} After a long time, Atera said again, ~\textbf{\textup{23}} “If you don’t want to carry the food for Dad, I will let go of your hand!”~\textbf{\textup{24}} Kihi was about to cry and said, ~\textbf{\textup{25}} “My dear sister, don’t let go of my hand, or else I will be afraid of the devil! ~\textbf{\textup{26}} When we are almost there, then I will carry the food.”


~

\bigskip\gll
\textbf{\textup{27}} He kī ia e {\ꞌ}Atera: ~\textbf{\textup{28}} —¡Ko aha {\ꞌ}ā koe i vānaŋa mai ai i te vānaŋa o te tātane! ~\textbf{\textup{29}} ¿Hoki ko tike{\ꞌ}a {\ꞌ}ā e koe te tātane ra{\ꞌ}e? \textbf{\textup{30}} He pāhono mai ia e Kihi: ~\textbf{\textup{31}} —Te parauti{\ꞌ}a, kai tike{\ꞌ}a {\ꞌ}ā e au te tātane ra{\ꞌ}e. \textbf{\textup{32}} O tētahi ŋā poki {\ꞌ}ō tā{\ꞌ}au i {\ꞌ}a{\ꞌ}amu mai pē nei ē: e ai rō {\ꞌ}ā te tātane {\ꞌ}e {\ꞌ}i te pō e e{\ꞌ}a e ha{\ꞌ}ere nei.\\
~ \textsc{ntr} say then \textsc{ag} Atera ~ ~~~~\textsc{prf} what \textsc{cont} \textsc{2sg} \textsc{pfv} speak hither \textsc{pvp} \textsc{acc} \textsc{art} word of \textsc{art} devil ~ ~~\textsc{pq} \textsc{prf} see \textsc{cont} \textsc{ag} \textsc{2sg} \textsc{art} devil first  ~ \textsc{ntr} answer hither then \textsc{ag} Kihi ~ ~~~\textsc{art} truth \textsc{neg.pfv} see \textsc{cont} \textsc{ag} \textsc{1sg} \textsc{art} devil first ~ of other \textsc{pl} child really \textsc{poss.2sg.a} \textsc{pfv} tell hither like \textsc{prox} thus \textsc{ipfv} exist \textsc{emph} \textsc{cont} \textsc{art} devil and at \textsc{art} night \textsc{ipfv} go\_out \textsc{ipfv} walk \textsc{prox}\\

\medskip\glt
\textbf{\textup{27}} Atera said, ~\textbf{\textup{28}} “What are you talking about the devil! ~\textbf{\textup{29}} Have you ever seen a devil?” ~\textbf{\textup{30}} Kihi replied, ~\textbf{\textup{31}} “The truth is, I have never seen a devil. ~\textbf{\textup{32}} Other children have told that devils exist and that they go out and walk around at night.”

~

\bigskip\gll
\textbf{\textup{33}} Ka topa tū vānaŋa era {\ꞌ}a Kihi, {\ꞌ}ī rāua ka u{\ꞌ}i atu ena e noho nō mai {\ꞌ}ā te tātane e tahi {\ꞌ}i ruŋa i te {\ꞌ}āua. ~\textbf{\textup{34}} A ruŋa i pe{\ꞌ}e ai ararua ko kikikikiu atu {\ꞌ}ana ia. ~\textbf{\textup{35}} E tu{\ꞌ}u nō {\ꞌ}ā te re{\ꞌ}o ki te rāua hare era {\ꞌ}e pē ira {\ꞌ}ā ki te kona aŋa era o koro. \textbf{\textup{36}} He momoko mai te taŋata ta{\ꞌ}ato{\ꞌ}a mai roto i te rāua hare. \textbf{\textup{37}} Pē ira {\ꞌ}ā a nua ararua ko koro ko tahuti mai {\ꞌ}ā i ŋaro{\ꞌ}a era te re{\ꞌ}o o tū ŋā rikiriki era o rāua. ~\textbf{\textup{38}} I tu{\ꞌ}u era ki tū kona era o tū ŋā poki era ko te kikikikiu haŋa {\ꞌ}ā, e puru rō {\ꞌ}ā te {\ꞌ}āriŋa ararua hai tū paratoa era o rāua.\\
~ \textsc{cntg} happen \textsc{dem} word \textsc{dist} of\textsc{.a} Kihi \textsc{imm} \textsc{3pl} \textsc{cntg} look away \textsc{med} \textsc{ipfv} sit just hither \textsc{cont} \textsc{art} devil \textsc{num} one at above at \textsc{art} fence ~ by above \textsc{pfv} jump\_up \textsc{pvp} the\_two \textsc{prf} \textsc{red}:shriek away \textsc{cont} then ~ \textsc{ipfv} arrive just \textsc{cont} \textsc{art} voice to \textsc{art} \textsc{3pl} house \textsc{dist} and like \textsc{ana} \textsc{ident} to \textsc{art} place work \textsc{dist} of Dad  ~ \textsc{ntr} \textsc{pl}:rush hither \textsc{art} man all from inside at \textsc{art} \textsc{3pl} house  ~ like \textsc{ana} \textsc{ident} \textsc{prop} Mum the\_two \textsc{prom} Dad \textsc{prf} run hither \textsc{cont} \textsc{pfv} perceive \textsc{dist} \textsc{art} voice of \textsc{dem} \textsc{pl} small.\textsc{pl} \textsc{dist} of \textsc{3pl} ~ \textsc{pfv} arrive \textsc{dist} to \textsc{dem} place \textsc{dist} of \textsc{dem} \textsc{pl} child \textsc{dist} \textsc{prom} \textsc{art} \textsc{red}:shriek \textsc{nmlz} \textsc{ident} \textsc{ipfv} close \textsc{emph} \textsc{cont} \textsc{art} face the\_two \textsc{ins} \textsc{dem} jacket \textsc{dist} of \textsc{3pl}\\

\medskip\glt
\textbf{\textup{33}} Just when Kihi had said that, they saw a devil that was sitting on a fence. ~\textbf{\textup{35}} They both jumped up and yelled. ~ ~\textbf{\textup{35}} The sound could be heard at their house and also at the place where Dad worked. ~\textbf{\textup{36}} All the people rushed out from their houses. ~\textbf{\textup{37}} Mum and Dad came running when they heard the voice of their little ones. ~\textbf{\textup{38}} When they came to the place where their children were, the two were still yelling, covering their face with their jackets. 


~

\bigskip\gll
\textbf{\textup{39}} He haka hāhine atu ia a nua ararua ko koro, {\ꞌ}ai ka kī rō: \textbf{\textup{40}} —Ka momou, e koā vovo ē. ~\textbf{\textup{41}} Ta{\ꞌ}e he tātane te me{\ꞌ}e era o ruŋa i te {\ꞌ}āua. ~\textbf{\textup{42}} He mautini piro ko kakaro {\ꞌ}ā te roto e Hao Kūmā. \textbf{\textup{43}} {\ꞌ}Ai ka aŋa rō te mata, te haha, {\ꞌ}ai ka pu{\ꞌ}a rō hai paratoa {\ꞌ}e he to{\ꞌ}o mai he haka eke ki ruŋa o te {\ꞌ}āua mo haka ri{\ꞌ}ari{\ꞌ}a o te hua{\ꞌ}ai ha{\ꞌ}ere pō.\\
~ \textsc{ntr} \textsc{caus} near away then \textsc{prop} Mum the\_two \textsc{prom} Dad \textsc{subs} \textsc{cntg} say \textsc{emph} ~ ~~~\textsc{imp} \textsc{pl}:quiet \textsc{voc} \textsc{coll} dear\_girl \textsc{voc} ~ \textsc{conneg} \textsc{pred} devil \textsc{art} thing \textsc{dist} of above at \textsc{art} fence ~ \textsc{pred} pumpkin rotten \textsc{prf} hollow\_out \textsc{cont} \textsc{art} inside \textsc{ag} Hao Kuma ~ \textsc{subs} \textsc{cntg} make \textsc{emph} \textsc{art} eye \textsc{art} mouth \textsc{subs} \textsc{cntg} cover \textsc{emph} with jacket and \textsc{ntr} take hither \textsc{ntr} \textsc{caus} go\_up to above of \textsc{art} fence for \textsc{caus} fear of \textsc{art} family walk night\\

\medskip\glt
\textbf{\textup{39}} Mum and Dad came near and said, ~\textbf{\textup{40}} “Be quiet, dear girls. ~\textbf{\textup{41}} That thing on the fence is not a devil. \textbf{\textup{42}} It’s a rotten pumpkin which Hao Kuma has hollowed out. ~\textbf{\textup{43}} He made eyes and a mouth, then he covered it with a jacket and put it on top of the fence to scare the people who walk by at night.”

\newpage
\section*{\textit{Tikitiki {\ꞌ}a {\ꞌ}Ataraŋa} – Tikitiki a Ataranga (R352)}\label{sec:a.2}
\addcontentsline{toc}{section}{\textit{Tikitiki {\ꞌ}a {\ꞌ}Ataraŋa} – Tikitiki a Ataranga (R352)}

by Luis Avaka Paoa (\citealt[Vol. 4:93–98]{WeberWeber1990Mai})

~

\bigskip\gll
\textbf{\textup{01}} Hora nei he {\ꞌ}a{\ꞌ}amu atu au i te {\ꞌ}a{\ꞌ}amu era o Tikitiki {\ꞌ}a {\ꞌ}Ataraŋa. \textbf{\textup{02}} A Tikitiki {\ꞌ}a {\ꞌ}Ataraŋa, kai aŋiaŋi mai e au he aha tō{\ꞌ}ona mata he aha tō{\ꞌ}ona hakaara, he Tūpāhotu {\ꞌ}o he Miru. ~\textbf{\textup{03}} {\ꞌ}Ina kai aŋiaŋi rivariva mai e au. \textbf{\textup{04}} Te me{\ꞌ}e ō{\ꞌ}ona i aŋiaŋi mai e au, he nua ō{\ꞌ}ona te {\ꞌ}īŋoa ko Nuahine {\ꞌ}a Raŋi Kotekote. ~\textbf{\textup{05}} I poreko era te poki nei, he hāŋai e te rū{\ꞌ}au nei ararua ko tā{\ꞌ}ana kenu. ~\textbf{\textup{06}} He nuinui, he mate te koro. ~\textbf{\textup{07}} I mate era te koro o te poki nei, he toe he rū{\ꞌ}au nō nei, he noho he hāŋai ē... ~\textbf{\textup{08}} he nuinui.\\
~ time \textsc{prox} \textsc{ntr} tell away \textsc{1sg} \textsc{acc} \textsc{art} story \textsc{dist} of Tikitiki a Ataranga  ~ \textsc{prop} Tikitiki a Ataranga \textsc{neg.pfv} certain hither \textsc{ag} \textsc{1sg} \textsc{pred} what \textsc{poss.3sg.o} tribe \textsc{pred} what \textsc{poss.3sg.o} descendance \textsc{pred} Tupahotu of \textsc{pred} Miru ~ \textsc{neg} \textsc{neg.pfv} certain good:\textsc{red} hither \textsc{ag} \textsc{1sg} ~ \textsc{art} thing \textsc{poss.3sg.o} \textsc{pfv} certain hither \textsc{ag} \textsc{1sg} \textsc{pred} Mum \textsc{poss.3sg.o} \textsc{art} name \textsc{prom} Nuahine a Rangi Kotekote ~ \textsc{pfv} born \textsc{dist} \textsc{art} child \textsc{prox} \textsc{ntr} raise \textsc{ag} \textsc{art} old\_woman \textsc{prox} the\_two \textsc{prom} \textsc{poss.3sg.a} husband ~ \textsc{ntr} big:\textsc{red} \textsc{ntr} die \textsc{art} Dad ~ \textsc{pfv} die \textsc{dist} \textsc{art} Dad of \textsc{art} child \textsc{prox} \textsc{ntr} remain \textsc{pred} old\_woman just \textsc{prox} \textsc{ntr} stay \textsc{ntr} raise on\_and\_on ~ \textsc{ntr} big:\textsc{red}\\

\medskip\glt
\textbf{\textup{01}} Now I’m going to tell the story of Tikitiki a Ataranga. ~\textbf{\textup{02}} Concerning Tiki a Ataranga, I don’t know what his tribe was or his descendance, whether Tupahotu of Miru. ~\textbf{\textup{03}} I don’t know it well. ~\textbf{\textup{04}} The thing about him I do know, is that his mother’s name was Nuahine a Rangi Kotekote. ~\textbf{\textup{05}} When this child was born, this old woman brought it up with her husband. ~\textbf{\textup{06}} When he was bigger, his father died. ~\textbf{\textup{07}} When the child’s father had died, just the old woman was left; she continued to raise the child ~\textbf{\textup{08}} until he was grown up.


~

\bigskip\gll
\textbf{\textup{09}} He moe tou poki era ko Tikitiki {\ꞌ}a {\ꞌ}Ataraŋa ki tā{\ꞌ}ana vi{\ꞌ}e Tūpāhotu. \textbf{\textup{10}} He poreko e toru poki: e rua poki tane, e tahi poki vahine. \textbf{\textup{11}} He hāŋai i tū ŋā poki era he nunui.\\
~ \textsc{ntr} lie\_down \textsc{dem} child \textsc{dist} \textsc{prom} Tikitiki a Ataranga to \textsc{poss.3sg.a} woman Tupahotu ~ \textsc{ntr} born \textsc{num} three child \textsc{num} two child male \textsc{num} one child female  ~ \textsc{ntr} raise \textsc{acc} \textsc{dem} \textsc{pl} child \textsc{dist} \textsc{ntr} \textsc{pl}:big\\

\medskip\glt
\textbf{\textup{09}} The boy Tikitiki a Ataranga married a Tupahotu wife. ~\textbf{\textup{10}} Three children were born: two boys and one girl. ~\textbf{\textup{11}} They raised the children and they grew up.


~

\bigskip\gll
\textbf{\textup{12}} I hāŋai era i nunui era, te aŋa o tū rū{\ꞌ}au māmā era he kā i te {\ꞌ}umu paurō te mahana. ~\textbf{\textup{13}} E kā era i tou {\ꞌ}umu era paurō te mahana, {\ꞌ}ina he {\ꞌ}ō{\ꞌ}otu te {\ꞌ}umu e pō rō era. \textbf{\textup{14}} E pō rō {\ꞌ}ā e {\ꞌ}ō{\ꞌ}otu era te {\ꞌ}umu e ma{\ꞌ}oa era, {\ꞌ}ai ka kakai ananake ko te ŋā poki. \textbf{\textup{15}} He u{\ꞌ}i pa{\ꞌ}i i te me{\ꞌ}e era {\ꞌ}ina {\ꞌ}ō he kai {\ꞌ}i te {\ꞌ}ōtea {\ꞌ}ā. ~\textbf{\textup{16}} E pō rō {\ꞌ}ana {\ꞌ}ō e kai era, e mōkirokiro rō {\ꞌ}ā. ~\textbf{\textup{17}} He {\ꞌ}aroha pa{\ꞌ}i he kī: \textbf{\textup{18}} —¡Kai riva {\ꞌ}ō te kai nei o tātou, e repa ē! \textbf{\textup{19}} {\ꞌ}I te {\ꞌ}ao nui {\ꞌ}ō e kai nei te ŋā poki, e ha{\ꞌ}uru rō {\ꞌ}ō te ŋā poki.\\
~ \textsc{pfv} raise \textsc{dist} \textsc{pfv} \textsc{pl}:big \textsc{dist} \textsc{art} work of \textsc{dem} old\_woman mother \textsc{dist} \textsc{pred} kindle \textsc{acc} \textsc{art} earth\_oven every \textsc{art} day ~ \textsc{ipfv} kindle \textsc{dist} \textsc{acc} \textsc{dem} earth\_oven \textsc{dist} every \textsc{art} day \textsc{neg} \textsc{ntr} cooked \textsc{art} earth\_oven \textsc{ipfv} night \textsc{emph} \textsc{dist}  ~ \textsc{ipfv} night \textsc{emph} \textsc{cont} \textsc{ipfv} cooked \textsc{dist} \textsc{art} earth\_oven \textsc{ipfv} open\_earth\_oven \textsc{dist} \textsc{subs} \textsc{cntg} \textsc{pl}:eat together \textsc{prom} \textsc{art} \textsc{pl} child ~ \textsc{ntr} look in\_fact \textsc{acc} \textsc{art} thing \textsc{dist} \textsc{neg} really \textsc{ntr} eat at \textsc{art} daytime \textsc{ident} ~ \textsc{ipfv} night \textsc{emph} \textsc{cont} really \textsc{ipfv} eat \textsc{dist} \textsc{ipfv} nightfall \textsc{emph} \textsc{cont} ~ \textsc{ntr} pity in\_fact \textsc{ntr} say ~ ~~~~\textsc{neg.pfv} good really \textsc{art} eat \textsc{prox} of \textsc{1pl.incl} \textsc{voc} young\_man \textsc{voc}  ~ in \textsc{art} day big really \textsc{ipfv} eat \textsc{prox} \textsc{art} \textsc{pl} child \textsc{ipfv} sleep \textsc{emph} really \textsc{art} \textsc{pl} child\\

\medskip\glt
\textbf{\textup{12}} When they had raised them and they had grown up, what the old mother did was cooking food in the earth oven every day. ~\textbf{\textup{13}} When she cooked food in the oven every day, the food was not cooked before it got dark. ~\textbf{\textup{14}} When it was dark, the food would be done and she would open the earth oven; then she would eat with the children. ~\textbf{\textup{15}} So she saw that they did not eat in the daylight. ~\textbf{\textup{16}} It was night when they ate, it was dark. ~\textbf{\textup{17}} So she took pity and said, ~\textbf{\textup{18}} “The way we eat is not right, my son! ~\textbf{\textup{19}} The children eat at midnight, when they are asleep.”


~

\bigskip\gll
\textbf{\textup{20}} He oti {\ꞌ}ana he kī e tou poki era o tou rū{\ꞌ}au era ko Tikitiki {\ꞌ}a {\ꞌ}Ataraŋa: \textbf{\textup{21}} —¿Ma{\ꞌ}a {\ꞌ}ā rō koe, e nua ē? \textbf{\textup{22}} He kī tou rū{\ꞌ}au era: ~\textbf{\textup{23}} —{\ꞌ}E, ¿he aha? \textbf{\textup{24}} —A au he e{\ꞌ}a a au he oho. ~\textbf{\textup{25}} He oho au ki ŋā hare he no{\ꞌ}ino{\ꞌ}i hai kona mahute, hauhau mo hiro o te taura mo tāea e au i te ra{\ꞌ}ā mo here. \textbf{\textup{26}} ¿{\ꞌ}E ku tano {\ꞌ}ā, e nua ē? \textbf{\textup{27}} —e kī era ki tō{\ꞌ}ona nua nei. \textbf{\textup{28}} He kī: ~\textbf{\textup{29}} —{\ꞌ}\=Eē, ku tano {\ꞌ}ā.\\
~ \textsc{ntr} finish \textsc{cont} \textsc{ntr} say \textsc{ag} \textsc{dem} child \textsc{dist} of \textsc{dem} old\_woman \textsc{dist} \textsc{prom} Tikitiki a Ataranga ~ ~~~~know \textsc{cont} \textsc{emph} \textsc{2sg} \textsc{voc} Mum \textsc{voc}  ~ \textsc{ntr} say \textsc{dem} old\_woman \textsc{dist} ~ ~~~and ~\textsc{pred} what ~ ~~~\textsc{prop} \textsc{1sg} \textsc{ntr} go\_out \textsc{prop} \textsc{1sg} \textsc{ntr} go ~ \textsc{ntr} go \textsc{1sg} to \textsc{pl} house \textsc{pl} request:\textsc{red} \textsc{ins} place mulberry kind\_of\_tree for braid of \textsc{art} rope for catch\_with\_lasso \textsc{ag} \textsc{1sg} \textsc{acc} \textsc{art} sun for tie ~ ~and \textsc{prf} correct \textsc{cont} \textsc{voc} Mum \textsc{voc} ~ ~~~\textsc{ipfv} say \textsc{dist} to \textsc{poss.3sg.o} Mum \textsc{prox} ~ \textsc{ntr} say ~ ~~~yes \textsc{prf} correct \textsc{cont}\\

\medskip\glt
\textbf{\textup{20}} Then the son of the old woman, Tikitiki a Ataranga, said, ~\textbf{\textup{21}} “You know what, Mum?” ~\textbf{\textup{22}} The old woman said, ~\textbf{\textup{23}} “Well, what?” \textbf{\textup{24}} “I will go out. ~\textbf{\textup{25}} I will go to the houses and ask for mulberry and hauhau fibers to braid a rope to catch the sun with a lasso. ~\textbf{\textup{26}} Is that okay, Mum?” ~\textbf{\textup{27}} he said to his mother. ~\textbf{\textup{28}} She said, ~\textbf{\textup{29}} “Yes, that’s okay.”

\newpage 
\gll
\textbf{\textup{30}} I e{\ꞌ}a era {\ꞌ}i te {\ꞌ}ao era {\ꞌ}ā o te rua ra{\ꞌ}ā, he oho he tu{\ꞌ}u he pa{\ꞌ}o mai i tā{\ꞌ}ana mahute i tā{\ꞌ}ana hauhau, he haha{\ꞌ}o ki roto te vai he haka pā mo haka marere mo haka piro haka kōpiro {\ꞌ}i roto i te vai mo te taura mo hiri. \textbf{\textup{31}} He oho ki te hare o tō{\ꞌ}ona ŋaruhoa e noho era ananake. ~\textbf{\textup{32}} {\ꞌ}E i tupu{\ꞌ}aki {\ꞌ}ā {\ꞌ}i tō{\ꞌ}ona hare, he oho he tu{\ꞌ}u he no{\ꞌ}ino{\ꞌ}i hai mahute hai hauhau. \textbf{\textup{33}} He rova{\ꞌ}a mai. \textbf{34} He ma{\ꞌ}u he oho mai he tu{\ꞌ}u, he pu{\ꞌ}a he haka kōpiro mo te taura mo hiri.\\
~ \textsc{pfv} go\_out \textsc{dist} at \textsc{art} dawn \textsc{dist} \textsc{ident} of \textsc{art} two day \textsc{ntr} go \textsc{ntr} arrive \textsc{ntr} chop hither \textsc{acc} \textsc{poss.3sg.a} mulberry \textsc{acc} \textsc{poss.3sg.a} kind\_of\_tree \textsc{ntr} insert to inside \textsc{art} water \textsc{ntr} \textsc{caus} double for \textsc{caus} fall\_apart for \textsc{caus} rotten \textsc{caus} ferment at inside at \textsc{art} water for \textsc{art} rope for braid ~ \textsc{ntr} go to \textsc{art} house of \textsc{poss.3sg.o} friend \textsc{ipfv} stay \textsc{dist} together ~ and \textsc{pfv} near \textsc{cont} at \textsc{poss.3sg.o} house \textsc{ntr} go \textsc{ntr} arrive \textsc{ntr} request:\textsc{red} \textsc{ins} mulberry \textsc{ins} kind\_of\_tree ~ \textsc{ntr} obtain hither ~ \textsc{ntr} carry \textsc{ntr} go hither \textsc{ntr} arrive \textsc{ntr} cover \textsc{ntr} \textsc{caus} ferment for \textsc{art} rope for braid\\

\medskip\glt
\textbf{\textup{30}} When he had gone out at dawn the next day, he went and chopped mulberry and hauhau trees; he put the fibres in the water and folded them so they would fall apart and rot and ferment in the water, to braid a rope from. ~\textbf{\textup{31}} He went to the houses of his friends who lived together with him. ~\textbf{\textup{32}} And when he was near their houses, he went and asked for mulberry and hauhau. ~\textbf{\textup{33}} He got them. ~\textbf{\textup{34}} He carried them away and covered them (with water) to ferment to braid a rope from.


~

\bigskip\gll
\textbf{\textup{35}} I oti era tū me{\ꞌ}e era, he to{\ꞌ}o koro{\ꞌ}iti mai tū taura era i tou hau era, he hiro i te taura. ~\textbf{\textup{36}} He hiro ka hiro era ē... ~\textbf{\textup{37}} ka kumi ka kumi. \textbf{\textup{38}} He to{\ꞌ}o mai i tū taura era he aŋa {\ꞌ}ā ka oti rō. \textbf{\textup{39}} I oti era tū taura era, he oho mai he kī ki tū rū{\ꞌ}au māmā era: \textbf{\textup{40}} —\=Apō {\ꞌ}i te {\ꞌ}ao era {\ꞌ}ā, e oŋa iho nō mai {\ꞌ}ā te ra{\ꞌ}ā, he rere au he oho he tu{\ꞌ}u he tāea i te ŋao o te ra{\ꞌ}ā, he here a ruŋa i te mā{\ꞌ}ea. \textbf{\textup{41}} Haka rito koe, e nua ē, mo kā i to tātou {\ꞌ}umu āpō. \textbf{\textup{42}} He kī tū rū{\ꞌ}au era: ~\textbf{\textup{43}} —Ku tano {\ꞌ}ā.\\
~ \textsc{pfv} finish \textsc{dist} \textsc{dem} thing \textsc{dist} \textsc{ntr} take slowly hither \textsc{dem} rope \textsc{dist} \textsc{acc} \textsc{dem} cord \textsc{dist} \textsc{ntr} braid \textsc{acc} \textsc{art} rope ~ \textsc{ntr} braid \textsc{cntg} braid \textsc{dist} on\_and\_on ~ \textsc{cntg} long \textsc{cntg} long ~ \textsc{ntr} take hither \textsc{acc} \textsc{dem} rope \textsc{dist} \textsc{ntr} make until \textsc{cntg} finish \textsc{emph}  ~ \textsc{pfv} finish \textsc{dist} \textsc{dem} rope \textsc{dist} \textsc{ntr} go hither \textsc{ntr} say to \textsc{dem} old\_woman mother \textsc{dist} ~ ~~tomorrow at \textsc{art} dawn \textsc{dist} \textsc{ident} \textsc{ipfv} appear just\_then just hither \textsc{cont} \textsc{art} sun \textsc{ntr} jump \textsc{1sg} \textsc{ntr} go \textsc{ntr} arrive \textsc{ntr} catch\_with\_lasso \textsc{acc} \textsc{art} neck of \textsc{art} sun \textsc{ntr} tie by above at \textsc{art} stone ~ \textsc{caus} ready \textsc{2sg} \textsc{voc} Mum \textsc{voc} for kindle \textsc{acc} \textsc{art}:of \textsc{1pl.incl} earth\_oven tomorrow ~ \textsc{ntr} say \textsc{dem} old\_woman \textsc{dist} ~ ~~~\textsc{prf} correct \textsc{cont}\\
% \medskip
\newpage 
\glt
\textbf{\textup{35}} When that was done, he took the fibres bit by bit and braided a rope. ~\textbf{\textup{36}} He braided and braided ~\textbf{\textup{37}} until the rope was very long. ~\textbf{\textup{38}} He took the rope and kept working until it was finished. ~\textbf{\textup{39}} When the rope was finished, he went and said to his old mother, ~\textbf{\textup{40}} “Tomorrow early in the morning, just when the sun appears, I will jump and catch the neck of the sun with a lasso and tie it to a rock. ~\textbf{\textup{41}} Get ready, Mum, to cook our earth oven tomorrow.” ~\textbf{\textup{42}} The old woman said, ~\textbf{\textup{43}} “That’s okay.”


~

\bigskip\gll
\textbf{\textup{44}} Ka moe ka {\ꞌ}ara, ka moe ka {\ꞌ}ara ē.... ~\textbf{\textup{45}} {\ꞌ}I te {\ꞌ}ao era {\ꞌ}ā i {\ꞌ}ara rō ai, he tiaki he haka tau he noho he u{\ꞌ}i ki te ra{\ꞌ}ā. \textbf{\textup{46}} I oho mai era mo hāhine mai mo e{\ꞌ}a mai o te ra{\ꞌ}ā, he rere he oho. \textbf{\textup{47}} Ka rere atu ka oho atu, ka oŋa mai te ra{\ꞌ}ā, he haka eke hai taura ki te ŋao, he here hiohio ta{\ꞌ}a ika. ~\textbf{\textup{48}} He totoi he oho mai i te taura, he tu{\ꞌ}u mai he here ki ruŋa ki te puku mā{\ꞌ}ea ena, {\ꞌ}ā ka ŋōŋō rō ka ŋōŋō rō, {\ꞌ}ā ka harara rō. \textbf{\textup{49}} He here, {\ꞌ}ā ka hiohio rō, {\ꞌ}ina kai hōrou haka{\ꞌ}ou mai tū ra{\ꞌ}ā era. \textbf{\textup{50}} He iri mai tū ra{\ꞌ}ā era koro{\ꞌ}iti nō koro{\ꞌ}iti nō i iri mai ai.\\
~ \textsc{cntg} lie\_down \textsc{cntg} wake\_up \textsc{cntg} lie\_down \textsc{cntg} wake\_up on\_and\_on ~ at \textsc{art} dawn \textsc{dist} \textsc{ident} \textsc{pfv} wake\_up \textsc{emph} \textsc{pvp} \textsc{ntr} wait \textsc{ntr} \textsc{caus} hang \textsc{ntr} stay \textsc{ntr} look to \textsc{art} sun  ~ \textsc{pfv} go hither \textsc{dist} for near hither for go\_out hither of \textsc{art} sun \textsc{ntr} jump \textsc{ntr} go  ~ \textsc{cntg} jump away \textsc{cntg} go away \textsc{cntg} appear hither \textsc{art} sun \textsc{ntr} \textsc{caus} go\_up \textsc{ins} rope to \textsc{art} neck \textsc{ntr} tie strong \textsc{poss.2sg.a} victim ~ \textsc{ntr} drag \textsc{ntr} go hither \textsc{acc} \textsc{art} rope \textsc{ntr} arrive hither \textsc{ntr} tie to above to \textsc{art} boulder stone \textsc{med} until \textsc{cntg} firm \textsc{emph} \textsc{cntg} firm \textsc{emph} until \textsc{cntg} stiff \textsc{emph} ~ \textsc{ntr} tie until \textsc{cntg} strong \textsc{emph} \textsc{neg} \textsc{neg.pfv} quickly again hither \textsc{dem} sun \textsc{dist}  ~ \textsc{ntr} ascend hither \textsc{dem} sun \textsc{dist} slowly just slowly just \textsc{pfv} ascend hither \textsc{pvp}\\

\medskip\glt
\textbf{\textup{44}} He slept and woke up, he slept and woke up... ~\textbf{\textup{45}} Early in the morning he woke up and stayed on the lookout, keeping an eye on the sun. ~\textbf{\textup{46}} When the sun was close to rising, he jumped up. ~\textbf{\textup{47}} Just when he jumped, the sun appeared; he put the rope to the neck and tied his victim firmly. ~\textbf{\textup{48}} He dragged the rope and tied it to a stone boulder, so it was very firmly tied and the rope was taut. ~\textbf{\textup{49}} He tied it firmly and the sun did not go quickly any more. ~\textbf{\textup{50}} The sun came up; slowly, slowly it came up.


~

\bigskip\gll
\textbf{\textup{51}} He e{\ꞌ}a mai tou rū{\ꞌ}au era {\ꞌ}i tou hora {\ꞌ}ā e ki tā{\ꞌ}ana {\ꞌ}umu, he amo ka ma{\ꞌ}itaki rō. ~\textbf{\textup{52}} He penapena i te {\ꞌ}umu pae, he tutu. \textbf{\textup{53}} He oho ki tā{\ꞌ}ana kūmā he keri mai he tata, ki tā{\ꞌ}ana moa he unu. \textbf{\textup{54}} Pōpōrā e ke{\ꞌ}oke{\ꞌ}o mai era te rima o tou rū{\ꞌ}au era. ~\textbf{\textup{55}} He to{\ꞌ}o mai he kā i tou {\ꞌ}umu era he {\ꞌ}ō{\ꞌ}otu. \textbf{\textup{56}} He ta{\ꞌ}o i te kūmā, i te taro, i te {\ꞌ}uhi, i te moa. \textbf{\textup{57}} Ko turu {\ꞌ}ā tū taŋata era ki tai ko hī mai {\ꞌ}ā i te ika, e here {\ꞌ}ā i te kōreha. ~\textbf{\textup{58}} Tanu te {\ꞌ}umu ra{\ꞌ}e nei o te rū{\ꞌ}au nei, ko tata mai {\ꞌ}ā tū taŋata era mo tou ika era.\\
~ \textsc{ntr} go\_out hither \textsc{dem} old\_woman \textsc{dist} at \textsc{dem} time \textsc{ident} quickly to \textsc{poss.3sg.a} earth\_oven \textsc{ntr} wipe \textsc{cntg} clean \textsc{emph} ~ \textsc{ntr} arrange\_firewood \textsc{acc} \textsc{art} earth\_oven hewn\_stones \textsc{ntr} set\_fire ~ \textsc{ntr} go to \textsc{poss.3sg.a} sweet\_potato \textsc{ntr} dig hither \textsc{ntr} wash to \textsc{poss.3sg.a} chicken \textsc{ntr} pluck ~ quickly \textsc{ipfv} hurry hither \textsc{dist} \textsc{art} hand of \textsc{dem} old\_woman \textsc{dist} ~ \textsc{ntr} take hither \textsc{ntr} kindle \textsc{acc} \textsc{dem} earth\_oven \textsc{dist} \textsc{ntr} cooked  ~ \textsc{ntr} cook \textsc{acc} \textsc{art} sweet\_potato \textsc{acc} \textsc{art} taro \textsc{acc} \textsc{art} yam \textsc{acc} \textsc{art} chicken ~ \textsc{prf} go\_down \textsc{cont} \textsc{dem} man \textsc{dist} to sea \textsc{prf} to\_fish hither \textsc{cont} \textsc{acc} \textsc{art} fish \textsc{ipfv} tie \textsc{cont} \textsc{acc} \textsc{art} eel ~ bury \textsc{art} earth\_oven first \textsc{prox} of \textsc{art} old\_woman \textsc{prox} \textsc{prf} approach hither \textsc{cont} \textsc{dem} man \textsc{dist} for \textsc{dem} fish \textsc{dist}\\

\medskip\glt
\textbf{\textup{51}} At that time the old women went quickly to her earth oven and wiped it clean. ~\textbf{\textup{52}} She arranged the firewood over the earth oven made of hewn stones, and kindled it. \textbf{\textup{53~~}}She went and dug up sweet potatoes and washed them, and plucked chickens. ~\textbf{\textup{54}} The hands of the old woman moved quickly. ~\textbf{\textup{55}} She lighted the earth oven and got the food cooked. ~\textbf{\textup{56}} She cooked sweet potatoes, taro, yam and chicken. ~\textbf{\textup{57}} The man had gone down to sea to catch fish with hook and line and to catch eels with a trap. \textbf{\textup{58~~}}The old woman covered the first earth oven when the man had come back with his fish.


~

\bigskip\gll
\textbf{\textup{59}} Puhi haka{\ꞌ}ou {\ꞌ}ā ka rua {\ꞌ}umu. ~\textbf{\textup{60}} Ko {\ꞌ}ō{\ꞌ}otu haka{\ꞌ}ou mai {\ꞌ}ā te ta{\ꞌ}o, penapena haka{\ꞌ}ou mai {\ꞌ}ā ka toru {\ꞌ}umu. \textbf{\textup{61}} He e{\ꞌ}a mai tou rū{\ꞌ}au era ki tou {\ꞌ}umu ra{\ꞌ}e era, he mātaki mai. \textbf{\textup{62}} He raŋi mai ki te ŋā poki: \textbf{\textup{63}} —¡Ka oho mai! \textbf{\textup{64}} He oho atu tou vi{\ꞌ}e hunoŋa era, tou ŋā poki era he takataka, he ma{\ꞌ}oa tou {\ꞌ}umu era.\\
~ blow again \textsc{cont} \textsc{cntg} two earth\_oven ~ \textsc{prf} cooked again hither \textsc{cont} \textsc{art} cook arrange\_firewood again hither \textsc{cont} \textsc{cntg} three earth\_oven  ~ \textsc{ntr} go\_out hither \textsc{dem} old\_woman \textsc{dist} to \textsc{dem} earth\_oven first \textsc{dist} \textsc{ntr} open hither  ~ \textsc{ntr} call hither to \textsc{art} \textsc{pl} child ~ ~~~~~\textsc{imp} go hither  ~ \textsc{ntr} go away \textsc{dem} woman child\_in\_law \textsc{dist} \textsc{dem} \textsc{pl} child \textsc{dist} \textsc{ntr} gather:\textsc{red} \textsc{ntr} open\_earth\_oven \textsc{dem} earth\_oven \textsc{dist}\\

\medskip\glt
\textbf{\textup{59}} She lighted another earth oven, the second one. ~\textbf{\textup{60}} When the food was cooked again, she arranged firewood again for the third earth oven. ~\textbf{\textup{61}} The old woman went to the first earth oven and opened it. ~\textbf{\textup{62}} She cried to the children, ~\textbf{\textup{63}} “Come!” ~\textbf{\textup{64}} Her daughter-in-law and the children gathered and she opened the earth oven.


~

\bigskip\gll
\textbf{\textup{65}} He hoa te kai ki ruŋa, he kī: ~\textbf{\textup{66}} —Pē nei hoki ta{\ꞌ}a me{\ꞌ}e. \textbf{\textup{67}} {\ꞌ}\=I tātou ka rovā kai. ~\textbf{\textup{68}} Ka kai iho tātou {\ꞌ}i te hora mā{\ꞌ}eha {\ꞌ}i te {\ꞌ}ōtea. ~\textbf{\textup{69}} Kai riva hoki ta{\ꞌ}a me{\ꞌ}e era tātou e kakai era {\ꞌ}i te pō {\ꞌ}i te {\ꞌ}ao nui. ~\textbf{\textup{70}} {\ꞌ}Ina he mā{\ꞌ}eha mo u{\ꞌ}i iŋa i te kai. \textbf{\textup{71}} E hā{\ꞌ}uru rō era te ŋā poki, {\ꞌ}ina he kakai. \textbf{\textup{72}} ¡{\ꞌ}\=A, me{\ꞌ}e {\ꞌ}aroha! ~\textbf{\textup{73}} {\ꞌ}O ira {\ꞌ}ī tātou ka to{\ꞌ}o rō mai ka kai rō atu. \textbf{\textup{74}} He kai mai i tou {\ꞌ}umu era, he oti. \textbf{\textup{75}} He ma{\ꞌ}oa haka{\ꞌ}ou mai ka rua {\ꞌ}umu {\ꞌ}i ruŋa {\ꞌ}ā, he kai haka{\ꞌ}ou mai. ~\textbf{\textup{76}} He ta{\ꞌ}o haka{\ꞌ}ou mai i te {\ꞌ}umu. \textbf{\textup{77}} I ta{\ꞌ}o i oho nō i ta{\ꞌ}o i oho nō.\\
~ \textsc{ntr} throw \textsc{art} food to above \textsc{ntr} say ~ ~~~like \textsc{prox} indeed \textsc{poss.2sg.a} thing  ~ \textsc{imm} \textsc{1pl.incl} \textsc{cntg} obtain food ~ \textsc{cntg} eat just\_now \textsc{1pl.incl} at \textsc{art} time light at \textsc{art} daytime ~ \textsc{neg.pfv} good indeed \textsc{poss.2sg.a} thing \textsc{dist} \textsc{1pl.incl} \textsc{ipfv} \textsc{pl}:eat \textsc{dist} at \textsc{art} night at \textsc{art} dusk big ~ \textsc{neg} \textsc{pred} light for see \textsc{nmlz} \textsc{acc} \textsc{art} food  ~ \textsc{ipfv} \textsc{pl}:sleep \textsc{emph} \textsc{dist} \textsc{art} \textsc{pl} child \textsc{neg} \textsc{ntr} \textsc{pl}:eat  ~ ~~ah thing pity ~ because\_of \textsc{ana} \textsc{imm} \textsc{1pl.incl} \textsc{cntg} take \textsc{emph} hither \textsc{cntg} eat \textsc{emph} away ~ \textsc{ntr} eat hither \textsc{acc} \textsc{dem} earth\_oven \textsc{dist} \textsc{ntr} finish  ~ \textsc{ntr} open\_earth\_oven again hither \textsc{cntg} two earth\_oven at above \textsc{ident} \textsc{ntr} eat again hither ~ \textsc{ntr} cook again hither \textsc{acc} \textsc{art} earth\_oven  ~ \textsc{pfv} cook \textsc{pfv} go just \textsc{pfv} cook \textsc{pfv} go just\\

\medskip\glt
\textbf{\textup{65}} She took the food out and said, ~\textbf{\textup{66}} “Here, look at this. ~\textbf{\textup{67}} Now we have something to eat. ~\textbf{\textup{68}} We will eat in the daytime, when it is light. ~\textbf{\textup{69}} It was not good what we did, eating in the dark at midnight. ~\textbf{\textup{70}} There was no light to see the food. \textbf{\textup{71~~}}The children were asleep and did not eat. ~\textbf{\textup{72}} Ah, what a pity! ~\textbf{\textup{73}} So we will go ahead now and eat.” ~\textbf{\textup{74}} They ate the food from the earth oven and finished. ~\textbf{\textup{75}} Then they opened the second earth oven and ate again. ~\textbf{\textup{76}} They cooked another earth oven. ~\textbf{\textup{77}} They just kept cooking and cooking.


~

\bigskip\gll
\textbf{\textup{78}} I ahiahi era, he tike{\ꞌ}a e te taŋata ko here {\ꞌ}ā i te ra{\ꞌ}ā. \textbf{\textup{79}} He oho mai he tu{\ꞌ}u mai he haka takataka {\ꞌ}i tū hare era o tū taŋata era. ~\textbf{\textup{80}} He taū te tau{\ꞌ}a, he rakerake te tau{\ꞌ}a he noho he kakai. \textbf{\textup{81}} Kakai i te tau{\ꞌ}a nei, ka rakerake rō {\ꞌ}i te hare era o tū taŋata era. \textbf{\textup{82}} He kī: ~\textbf{\textup{83}} —¿He aha koe i to{\ꞌ}o mai ai i here ai i te ra{\ꞌ}ā?\\
~ \textsc{pfv} evening \textsc{dist} \textsc{ntr} see \textsc{ag} \textsc{art} man \textsc{prf} tie \textsc{cont} \textsc{acc} \textsc{art} sun  ~ \textsc{ntr} go hither \textsc{ntr} arrive hither \textsc{ntr} \textsc{caus} gather:\textsc{red} at \textsc{dem} house \textsc{dist} of \textsc{dem} man \textsc{dist} ~ \textsc{ntr} to\_fight \textsc{art} fight \textsc{ntr} bad:\textsc{red} \textsc{art} fight \textsc{ntr} stay \textsc{ntr} quarrel  ~ quarrel \textsc{acc} \textsc{art} fight \textsc{prox} \textsc{cntg} bad:\textsc{red} \textsc{emph} at \textsc{art} house \textsc{dist} of \textsc{dem} man \textsc{dist}  ~ \textsc{ntr} say ~ ~~~~~\textsc{pred}~ what \textsc{2sg} \textsc{pfv} take hither \textsc{pvp} \textsc{pfv} tie \textsc{pvp} \textsc{acc} \textsc{art} sun\\

\medskip\glt
\textbf{\textup{78}} In the evening, the people saw that the sun had been tied. ~\textbf{\textup{79}} They came and gathered at the house of that man. ~\textbf{\textup{80}} A fight broke out, there was an ugly quarrel. \textbf{\textup{81~~}}There was a bad quarrel at the house of that man. ~\textbf{\textup{82}} They said, ~\textbf{\textup{83}} “Why did you take the sun and tie it up?”


~

\bigskip\gll
\textbf{\textup{84}} He kī e tū taŋata era: ~\textbf{\textup{85}} —Ka koa ho{\ꞌ}i kōrua. \textbf{\textup{86}} ¿Kai koa {\ꞌ}ō kōrua {\ꞌ}i te me{\ꞌ}e rivariva? ~\textbf{\textup{87}} Ku to{\ꞌ}o mai {\ꞌ}ā a au i te ra{\ꞌ}ā ku here {\ꞌ}ā mo haka mā{\ꞌ}eha mo tātou mo kakai. \textbf{\textup{88}} {\ꞌ}O ira ho{\ꞌ}i au i here ena. \textbf{\textup{89}} Nā ka u{\ꞌ}i rā kōrua, ka hia {\ꞌ}umu nei {\ꞌ}ā{\ꞌ}aku e kā atu ena. \textbf{\textup{90}} He kā te {\ꞌ}umu nei he ma{\ꞌ}oa he kai. ~\textbf{\textup{91}} He kā haka{\ꞌ}ou he ma{\ꞌ}oa he kai. ~\textbf{\textup{92}} {\ꞌ}Ai e kā haka{\ꞌ}ou mai era.\\
~ \textsc{ntr} say \textsc{ag} \textsc{dem} man \textsc{dist} ~ ~~~\textsc{imp} happy indeed \textsc{2pl}  ~ ~\textsc{neg.pfv} happy really \textsc{2pl} at \textsc{art} thing good ~ \textsc{prf} take hither \textsc{cont} \textsc{prop} \textsc{1sg} \textsc{acc} \textsc{art} sun \textsc{pfv} tie \textsc{cont} for \textsc{caus} light for \textsc{1pl.incl} for \textsc{pl}:eat  ~ because\_of \textsc{ana} indeed \textsc{1sg} \textsc{pfv} tie \textsc{med} ~ \textsc{med} \textsc{imp} look \textsc{intens} \textsc{2pl} \textsc{cntg} how\_many earth\_oven \textsc{prox} \textsc{poss.1sg.a} \textsc{ipfv} kindle away \textsc{med}  ~ \textsc{ntr} kindle \textsc{art} earth\_oven \textsc{prox} \textsc{ntr} open\_earth\_oven \textsc{ntr} eat ~ \textsc{ntr} kindle again \textsc{ntr} open\_earth\_oven \textsc{ntr} eat ~ \textsc{subs} \textsc{ipfv} kindle again hither \textsc{dist}\\

\medskip\glt
\textbf{\textup{84}} The man said, ~\textbf{\textup{85}} “Be happy! ~\textbf{\textup{86}} Aren’t you happy with a good thing? ~ ~\textbf{\textup{87}} I have taken the sun and tied it up to give us light to eat. ~\textbf{\textup{88}} That’s why I tied it up. ~\textbf{\textup{89}} There, look how many earth ovens I’ve been cooking. ~\textbf{\textup{90}} We lighted this earth oven, opened it and ate. ~\textbf{\textup{91}} We lighted another, opened it and ate. ~\textbf{\textup{92}} Then we lighted another.”


~

\bigskip\gll
\textbf{\textup{93}} He ture ka rakerake rō te tau{\ꞌ}a. ~\textbf{\textup{94}} He rere mai e te taŋata he horehore i tou taura era. ~\textbf{\textup{95}} He momotu he haka poko{\ꞌ}o. \textbf{\textup{96}} I haka poko{\ꞌ}o atu era i te ra{\ꞌ}ā era, he nini rō {\ꞌ}ai tū ra{\ꞌ}ā era he oho he ŋaro. ~\textbf{\textup{97}} He raŋi e te rū{\ꞌ}au nei: ~\textbf{\textup{98}} —E pa{\ꞌ}o rō kōrua i te ra{\ꞌ}ā here e tā{\ꞌ}aku poki. \textbf{\textup{99}} Rohirohi e tā{\ꞌ}aku poki i iri ai i here mai ai mo haka mā{\ꞌ}eha o tātou. \textbf{\textup{100}} {\ꞌ}E {\ꞌ}ī au ka raŋi rō atu: e to{\ꞌ}o rō au i ta{\ꞌ}a me{\ꞌ}e hukahuka ena mai ruŋa i te henua. ~\textbf{\textup{101}} {\ꞌ}\=I ka to{\ꞌ}o rō.\\
~ \textsc{ntr} scold \textsc{cntg} bad:\textsc{red} \textsc{emph} \textsc{art} fight ~ \textsc{ntr} jump hither \textsc{ag} \textsc{art} man \textsc{ntr} cut:\textsc{red} \textsc{acc} \textsc{dem} rope \textsc{dist} ~ \textsc{ntr} break\_completely \textsc{ntr} \textsc{caus} loose  ~ \textsc{pfv} \textsc{caus} loose away \textsc{dist} \textsc{acc} \textsc{art} sun \textsc{dist} \textsc{ntr} spin \textsc{emph} \textsc{subs} \textsc{dem} sun \textsc{dist} \textsc{ntr} go \textsc{ntr} disappear ~ \textsc{ntr} call \textsc{ag} \textsc{art} old\_woman \textsc{prox} ~ ~~~\textsc{ipfv} chop \textsc{emph} \textsc{2pl} \textsc{acc} \textsc{art} sun tie \textsc{ag} \textsc{poss.1sg.a} child ~ tired \textsc{ag} \textsc{poss.1sg.a} child \textsc{pfv} ascend \textsc{pvp} \textsc{pfv} tie hither \textsc{pvp} for \textsc{caus} light of \textsc{1pl.incl} ~ and \textsc{imm} \textsc{1sg} \textsc{cntg} call \textsc{emph} away \textsc{ipfv} take \textsc{emph} \textsc{1sg} \textsc{acc} \textsc{poss.2sg.} thing firewood \textsc{med} from above at \textsc{art} land  ~ \textsc{imm} \textsc{cntg} take \textsc{emph}\\

\medskip\glt
\textbf{\textup{93}} They made reproaches and quarreled bitterly. ~\textbf{\textup{94}} The people jumped up and cut the rope. ~\textbf{\textup{95}} It snapped and the sun was loose. ~\textbf{\textup{96}} When the sun was loose, it turned around and disappeared. ~\textbf{\textup{97}} The old woman cried out, ~\textbf{\textup{98}} “You’re chopping the rope with which my son tied the sun. ~\textbf{\textup{99}} My sun wore himself out to go up and tie the sun to give us light. \textbf{\textup{100}} Now I’ll tell you: I will take your firewood from the earth. ~\textbf{\textup{101}} I’ll take it straightaway.”


~
\newpage
\gll
\textbf{\textup{102}} He rere mai te rū{\ꞌ}au nei ki tā{\ꞌ}ana {\ꞌ}umu pae, ki tā{\ꞌ}ana hukahuka, ki tā{\ꞌ}ana kai, ki tā{\ꞌ}ana me{\ꞌ}e ta{\ꞌ}ato{\ꞌ}a. \textbf{\textup{103}} He {\ꞌ}apa tahi ko te ŋā poki, ko te hare, ko te me{\ꞌ}e ta{\ꞌ}ato{\ꞌ}a. \textbf{\textup{104}} He rere he oho rō atu {\ꞌ}ai, ko te iri iŋa {\ꞌ}ā ko te oho oŋa {\ꞌ}ā, ē... ~\textbf{\textup{105}} he ŋaro ki roto ki te māhina, he oti rō {\ꞌ}ai. \textbf{\textup{106}} Ki roto i te māhina i ŋaro ai te rū{\ꞌ}au nei. \textbf{\textup{107}} Te {\ꞌ}īŋoa o tou rū{\ꞌ}au era ko Nuahine {\ꞌ}a Raŋi Kotekote. \textbf{\textup{108}} He ŋaro rō atu {\ꞌ}ai ki roto ki te māhina.\\
~ \textsc{ntr} jump hither \textsc{art} old\_woman \textsc{prox} to \textsc{poss.3sg.a} earth\_oven hewn\_stones to \textsc{poss.3sg.a} firewood to \textsc{poss.3sg.a} food to \textsc{poss.3sg.a} thing all  ~ \textsc{ntr} carry all \textsc{prom} \textsc{art} \textsc{pl} child \textsc{prom} \textsc{art} house \textsc{prom} \textsc{art} thing all  ~ \textsc{ntr} fly \textsc{ntr} go \textsc{emph} away \textsc{subs} \textsc{prom} \textsc{art} ascend \textsc{nmlz} \textsc{ident} \textsc{prom} \textsc{art} go \textsc{nmlz} \textsc{ident} on\_and\_on ~ \textsc{ntr} disappear to inside to \textsc{art} moon \textsc{ntr} finish \textsc{emph} \textsc{subs}  ~ to inside at \textsc{art} moon \textsc{pfv} disappear \textsc{pvp} \textsc{art} old\_woman \textsc{prox}  ~ \textsc{art} name of \textsc{dem} old\_woman \textsc{dist} \textsc{prom} Nuahine a Rangi Kotekote  ~ \textsc{ntr} disappear \textsc{emph} away \textsc{subs} to inside to \textsc{art} moon\\

\medskip\glt
\textbf{\textup{102}} The old woman jumped to her stone earth oven, to her firewood, to her food, to all her things. ~\textbf{\textup{103}} She carried all the children, the house, everything. ~\textbf{\textup{104}} She flew away, going up, going away... ~\textbf{\textup{105}} and disappeared into the moon, and that was it. \textbf{\textup{106}} Inside the moon the old woman disappeared. ~\textbf{\textup{107}} The name of the old woman was Nuahine a Rangi Kotekote. ~\textbf{\textup{108}} She disappeared into the moon.


~

\largerpage
\medskip\gll
\textbf{\textup{109}} He kī te taŋata: ~\textbf{\textup{110}} —¡{\ꞌ}\=A, ko tere {\ꞌ}ā te rū{\ꞌ}au! \textbf{\textup{111}} I {\ꞌ}ōtea mai era ki te popohaŋa he u{\ꞌ}i te taŋata ku oŋe {\ꞌ}ā, {\ꞌ}ina he hukahuka. \textbf{\textup{112}} Ku oŋe {\ꞌ}ā te kai, ko pakapaka {\ꞌ}ā ku mei {\ꞌ}ā. \textbf{\textup{113}} He kī: \textbf{\textup{114}} —¡{\ꞌ}\=A! ~\textbf{\textup{115}} O tou rū{\ꞌ}au era {\ꞌ}ā te kai, e topa era te mau, te hukahuka e ai era. ~\textbf{\textup{116}} Ku kore {\ꞌ}ā te hukahuka mo tunu o te kai. ~\textbf{\textup{117}} Ku pakapaka {\ꞌ}ā te henua. ~\textbf{\textup{118}} Ku oŋe {\ꞌ}ā tātou. \textbf{\textup{119}} He oti mau {\ꞌ}ā. ~\textbf{\textup{120}} ¿Ka aha rā ia? \textbf{\textup{121}} Ku tere {\ꞌ}ā te rū{\ꞌ}au nei.\\
~ \textsc{ntr} say \textsc{art} man ~ ~~~~~ah \textsc{prf} run \textsc{cont} \textsc{art} old\_woman  ~ \textsc{pfv} daylight hither \textsc{dist} to \textsc{art} dawn \textsc{ntr} look \textsc{art} man \textsc{prf} shortage \textsc{cont} \textsc{neg} \textsc{pred} firewood ~ \textsc{prf} shortage \textsc{cont} \textsc{art} food \textsc{prf} dry:\textsc{red} \textsc{cont} \textsc{prf} wither \textsc{cont}  ~ \textsc{ntr} say ~ ~~~~ah ~ of \textsc{dem} old\_woman \textsc{dist} \textsc{ident} \textsc{art} food \textsc{ipfv} happen \textsc{dist} \textsc{art} abundance \textsc{art} firewood \textsc{ipfv} exist \textsc{dist} ~ \textsc{prf} lack \textsc{cont} \textsc{art} firewood for cook of \textsc{art} food ~ \textsc{prf} dry:\textsc{red} \textsc{cont} \textsc{art} land ~ \textsc{prf} shortage \textsc{cont} \textsc{1pl.incl}  ~ \textsc{ntr} finish really \textsc{cont} ~ ~\textsc{cntg} what \textsc{intens} then  ~ \textsc{prf} run \textsc{cont} \textsc{art} old\_woman \textsc{prox}\\

\medskip\glt
\textbf{\textup{109}} The people said, ~\textbf{\textup{110}} “Ah, the old woman has run off!” ~\textbf{\textup{111}} At dawn the next day the people saw that there was shortage, there was no firewood. ~\textbf{\textup{112}} The food was scarce, it was dry and had withered. ~\textbf{\textup{113}} They said, ~\textbf{\textup{114}} “Ah! ~\textbf{\textup{115}} It was because of the woman that there was food, that there was abundance, that there was firewood. \textbf{\textup{116}} Now there is no firewood to cook food. ~\textbf{\textup{117}} The land is dry. ~\textbf{\textup{118}} We are in need. \textbf{\textup{119}} That’s how it is. ~\textbf{\textup{120}} What can we do? ~\textbf{\textup{121}} The old woman has run off.”


\newpage 
\gll
\textbf{\textup{122}} Kai {\ꞌ}ite atu ho{\ꞌ}i e te taŋata pē nei ē: tou rū{\ꞌ}au era {\ꞌ}ō e ai rō {\ꞌ}ā tō{\ꞌ}ona mana; rū{\ꞌ}au mana. \textbf{\textup{123}} Pē ira {\ꞌ}ā tū taŋata era. \textbf{\textup{124}} {\ꞌ}O ira ho{\ꞌ}i i iri era i here era i te ra{\ꞌ}ā era. ~\textbf{\textup{125}} E ai rō {\ꞌ}ā to rāua mana.\\
~ \textsc{neg.pfv} know away indeed \textsc{ag} \textsc{art} man like \textsc{prox} thus \textsc{dem} old\_woman \textsc{dist} really \textsc{ipfv} exist \textsc{emph} \textsc{cont} \textsc{poss.3sg.o} power old\_woman power  ~ like \textsc{ana} \textsc{ident} \textsc{dem} man \textsc{dist} ~ because\_of \textsc{ana} indeed \textsc{pfv} ascend \textsc{dist} \textsc{pfv} tie \textsc{dist} \textsc{acc} art sun \textsc{dist} ~ \textsc{ipfv} exist \textsc{emph} \textsc{cont} \textsc{art}:of \textsc{3pl} power\\

\medskip\glt
\textbf{\textup{122}} The people did not know that the old woman had mana (supernatural power); it was a woman with power. ~\textbf{\textup{123}} The same was true for the man. ~\textbf{\textup{124}} That’s why he went up and tied the son. ~\textbf{\textup{125}} They had supernatural power.


~

\bigskip\gll
\textbf{\textup{126}} He oti mau {\ꞌ}ā. ~\textbf{\textup{127}} Ku ŋaro ku oti {\ꞌ}ā te rū{\ꞌ}au nei ki roto ki te māhina. ~\textbf{\textup{128}} {\ꞌ}E ku noho {\ꞌ}ā te taŋata o te henua he taŋi rō {\ꞌ}ai mo tou rū{\ꞌ}au era {\ꞌ}i tū me{\ꞌ}e era ko ŋaro {\ꞌ}ā. \textbf{\textup{129}} He oti mau rō {\ꞌ}ai te {\ꞌ}a{\ꞌ}amu nei o Tikitiki {\ꞌ}a {\ꞌ}Ataraŋa.\\
~ \textsc{ntr} finish really \textsc{cont} ~ \textsc{prf} disappear \textsc{prf} finish \textsc{cont} \textsc{art} old\_woman \textsc{prox} to inside to \textsc{art} moon ~ and \textsc{prf} stay \textsc{cont} \textsc{art} man of \textsc{art} land \textsc{ntr} cry \textsc{emph} \textsc{subs} for \textsc{dem} old\_woman \textsc{dist} at \textsc{dem} thing \textsc{dist} \textsc{prf} disappear \textsc{cont}  ~ \textsc{ntr} finish really \textsc{emph} \textsc{subs} \textsc{art} story \textsc{prox} of Tikitiki a Ataranga\\

\medskip\glt
\textbf{\textup{126}} That’s all. ~\textbf{\textup{127}} The old woman had disappeared into the moon. ~\textbf{\textup{128}} And the people of the land stayed and cried for the woman, because she had disappeared. ~\textbf{\textup{129}} The story of Tikitiki a Ataranga is finished.

\newpage
\section*{\textit{He oho iŋa o te nu{\ꞌ}u hoko rua rama} – The trip of two people who went torch fishing (R357)} \label{sec:a.3}
\addcontentsline{toc}{section}{\textit{He oho iŋa o te nu{\ꞌ}u hoko rua rama} – The trip of two people who went torch fishing (R357)}
(\citealt[Vol. 5:22–23]{WeberWeber1990Mai})

~

\bigskip\gll
\textbf{\textup{01}} Ko te {\ꞌ}āva{\ꞌ}e era o te evinio {\ꞌ}e ko te mahana maha ia, he e{\ꞌ}a te nu{\ꞌ}u hoko rua, he pu{\ꞌ}a i te hoi, he aŋa i te rama, he haka rito i te rāua me{\ꞌ}e, he eke te hoi, he oho ararua a te pā{\ꞌ}eŋa ena o Vaihū. \textbf{\textup{02}} I tu{\ꞌ}u era ki {\ꞌ}Akahaŋa, he noho he tiaki ki te tai hahati mo rama.\\
~ \textsc{prom} \textsc{art} month \textsc{dist} of \textsc{art} Lent and \textsc{prom} \textsc{art} day four then \textsc{ntr} go\_out \textsc{art} people \textsc{num.pers} two \textsc{ntr} cover \textsc{acc} \textsc{art} horse \textsc{ntr} make \textsc{acc} \textsc{art} torch \textsc{ntr} \textsc{caus} ready \textsc{acc} \textsc{art} \textsc{3pl} thing \textsc{ntr} go\_up \textsc{art} horse \textsc{ntr} go the\_two by \textsc{art} side \textsc{med} of Vaihu ~ \textsc{pfv} arrive \textsc{dist} to Akahanga \textsc{ntr} stay \textsc{ntr} wait to \textsc{art} sea break for fish\_with\_torch\\

\medskip\glt
\textbf{\textup{01}} In the month of Lent on a Thursday, two people went out; they saddled their horses, made torches and prepared their things; they mounted their horses and went together to the side of Vaihu. ~\textbf{\textup{02}} When they arrived at Akahanga, they stayed and waited for the rising tide (suitable) for torch fishing.


~

\bigskip\gll
\textbf{\textup{03}} Ko rote {\ꞌ}ahuru mā ho{\ꞌ}e o te pō ia, he e{\ꞌ}a ararua, he tutu i te rama, he topa he rama {\ꞌ}i rote haŋa era o {\ꞌ}Akahaŋa. \textbf{\textup{04}} E rama nō {\ꞌ}ā, i hoa rō atu ai te {\ꞌ}ua ki te kāve{\ꞌ}uve{\ꞌ}u o te tua{\ꞌ}ivi. \textbf{\textup{05}} Kai rova{\ꞌ}a hia e tahi me{\ꞌ}e. ~\textbf{\textup{06}} He e{\ꞌ}a ararua, he tētere ki rote {\ꞌ}ana, he nonoho koia ko mamae {\ꞌ}i tū me{\ꞌ}e ta{\ꞌ}e rova{\ꞌ}a era.\\
~ \textsc{prom} inside\_\textsc{art} ten plus one of \textsc{art} night then \textsc{ntr} go\_out the\_two \textsc{ntr} set\_fire \textsc{acc} \textsc{art} torch \textsc{ntr} descend \textsc{ntr} fish\_with\_torch at inside\_the bay \textsc{dist} of Akahanga  ~ \textsc{ipfv} fish\_with\_torch just \textsc{cont} \textsc{pfv} throw \textsc{emph} away then \textsc{art} rain to \textsc{art} ache of \textsc{art} back ~ \textsc{neg.pfv} obtain yet \textsc{num} one thing ~ \textsc{ntr} go\_out the\_two \textsc{ntr} \textsc{pl}:run to inside\_the cave \textsc{ntr} \textsc{pl}:stay with \textsc{prom} pain at \textsc{dem} thing \textsc{conneg} obtain \textsc{dist}\\

\medskip\glt
\textbf{\textup{03}} Around eleven o’clock at night the two went out, lighted their torches, went down and went torch fishing in the bay of Akahanga. ~\textbf{\textup{04}} While they were fishing, suddenly the rain came down, lashing their backs. ~\textbf{\textup{05}} They didn’t catch anything. ~\textbf{\textup{06}} The two left and fled into a cave, sad because they hadn’t caught anything.


~

\bigskip\gll
\textbf{\textup{07}} I nonoho era, he papaŋaha{\ꞌ}a {\ꞌ}i te ha{\ꞌ}uru. ~\textbf{\textup{08}} E ha{\ꞌ}uru nō {\ꞌ}ā, he tu{\ꞌ}u atu hoko rua haka{\ꞌ}ou nu{\ꞌ}u mai te puhi iŋa mo te evinio. \textbf{\textup{09}} He hipa atu he haka {\ꞌ}ara i tū ŋā ha{\ꞌ}uru era. ~\textbf{\textup{10}} I {\ꞌ}a{\ꞌ}ara mai era, he eke te hoi, he oho mai mai {\ꞌ}Akahaŋa, he tu{\ꞌ}u ki Motu o Pope. ~\textbf{\textup{11}} He topa haka{\ꞌ}ou ararua {\ꞌ}i tū kona era he tutu te rama.\\
~ \textsc{pfv} \textsc{pl}:stay \textsc{dist} \textsc{ntr} \textsc{pl}:heavy at \textsc{art} sleep ~ \textsc{ipfv} sleep just \textsc{cont} \textsc{ntr} arrive away \textsc{num.pers} two again people from \textsc{art} fish\_at\_night \textsc{nmlz} for \textsc{art} Lent  ~ \textsc{ntr} pass\_by away \textsc{ntr} \textsc{caus} wake\_up \textsc{acc} \textsc{dem} \textsc{pl} sleep \textsc{dist} ~ \textsc{pfv} \textsc{pl}:wake\_up hither \textsc{dist} \textsc{ntr} go\_up \textsc{art} horse \textsc{ntr} go hither from Akahanga \textsc{ntr} arrive to Motu o Pope ~ \textsc{ntr} descend again the\_two at \textsc{dem} place \textsc{dist} \textsc{ntr} set\_fire \textsc{art} torch\\

\medskip\glt
\textbf{\textup{07}} While they stayed there, they fell asleep. ~\textbf{\textup{08}} While they were sleeping, two other people arrived, who had been fishing at night for Lent. ~\textbf{\textup{09}} They came by and woke the sleepers up. ~\textbf{\textup{10}} Once awake, they mounted their horses, left Akahanga and went to Motu o Pope. ~\textbf{\textup{11}} There the two descended again and lighted their torches.


~

\bigskip\gll
\textbf{\textup{12}} I mā{\ꞌ}eha atu era tū rama era, {\ꞌ}ī ka u{\ꞌ}i atu ena ko te {\ꞌ}ura ku noho {\ꞌ}ā {\ꞌ}i tū kona era. ~\textbf{\textup{13}} He topa ararua he {\ꞌ}aru{\ꞌ}aru mai i tū ŋā {\ꞌ}ura era mai tū kona era. ~\textbf{\textup{14}} He rova{\ꞌ}a e ho{\ꞌ}e {\ꞌ}ahuru mā piti {\ꞌ}ura. ~\textbf{\textup{15}} Ku koa {\ꞌ}ā ko ruŋa i te hoi, he {\ꞌ}ehu he oho rō mai {\ꞌ}ai ki Haŋa Roa {\ꞌ}i tū hora era.\\
~ \textsc{pfv} light away \textsc{dist} \textsc{dem} torch \textsc{dist} \textsc{imm} \textsc{cntg} look away \textsc{med} \textsc{prom} \textsc{art} lobster \textsc{prf} stay \textsc{cont} at \textsc{dem} place \textsc{dist} ~ \textsc{ntr} descend the\_two \textsc{ntr} grab:\textsc{red} hither \textsc{acc} \textsc{dem} \textsc{pl} lobster \textsc{dist} from \textsc{dem} place \textsc{dist} ~ \textsc{ntr} obtain \textsc{num} one ten plus two lobster ~ \textsc{prf} happy \textsc{cont} \textsc{prom} above at \textsc{art} horse \textsc{ntr} twilight \textsc{ntr} go \textsc{emph} hither \textsc{subs} to Hanga Roa at \textsc{dem} time \textsc{dist}\\

\medskip\glt
\textbf{\textup{12}} When their torches gave light, they saw lobsters sitting in that place. ~\textbf{\textup{13}} The two went down and grabbed the lobsters from there. ~\textbf{\textup{14}} They got twelve lobsters. ~\textbf{\textup{15}} Happily they climbed their horses, and in the early morning light they went to Hanga Roa.


~

\bigskip\gll
\textbf{\textup{16}} Tū oho era ho{\ꞌ}i i oho ai e oho {\ꞌ}ā mo rova{\ꞌ}a {\ꞌ}ura mo e{\ꞌ}a ki ruŋa ki te vaka ki te nanue hī {\ꞌ}i te motu mo kai o te evinio. \textbf{\textup{17}} {\ꞌ}I te me{\ꞌ}e era, te taŋata o nei {\ꞌ}i rā hora {\ꞌ}ina he kai i te kiko {\ꞌ}i te ta{\ꞌ}ato{\ꞌ}a mahana pae o te evinio.\\
~ \textsc{dem} go \textsc{dist} indeed \textsc{pfv} go \textsc{pvp} \textsc{ipfv} go \textsc{cont} for obtain lobster for go\_out to above to \textsc{art} boat to \textsc{art} kind\_of\_fish to\_fish at \textsc{art} islet for eat for \textsc{art} Lent  ~ at \textsc{art} thing \textsc{dist} \textsc{art} person of \textsc{prox} at \textsc{dist} time \textsc{neg} \textsc{ntr} eat \textsc{acc} \textsc{art} meat at \textsc{art} all day five of \textsc{art} Lent\\

\medskip\glt
\textbf{\textup{16}} This trip happened when they went to catch lobsters (as bait), to go out by boat to fish for nanue at the islets, to eat during Lent. ~\textbf{\textup{17}} Because at that time the people here did not eat meat on Fridays during Lent.
}
