\chapter{Prepared script for all interviews of JAMPRO informants}\label{appendix1}
104 informants were interviewed using this script. 82 of these interviews were recorded and transcribed phonetically for later sociolinguistic analysis.

\begin{enumerate}
	\item  Was the informant told of the purpose of the interview? 13 (yes)
\end{enumerate}

\section{Personal data collected}
\begin{enumerate}
\item Sex: 82 (female), 22 (male)
\item Age: 35 (20--29), 33 (30--39), 16 (40--49), 20 (50--65)
\item Residence: 65 (Portmore), 4 (Franklin Town), 21 (Liguanea), 9 (Red Hills), 4 (Stony Hills), 1 (no data)
\item Race: 53 (black), 35 (brown), 1 (Chinese), 1 (Indian), 14 (no data noted)
\item Years at \isi{JAMPRO}: 30 (10+), 38 (5+), 28 (–5), 8 (new)
\item Level of Education: 4 (primary), 10 (secondary), 43 (post-secondary), 22 (tertiary), 24 (graduate)
\item Salary Scale: 7 (ancillary), 31 (secretarial), 42 (professional), 10 (director), 14 (\isi{senior management})
\item Spouse’s Occupation: 1 (“cleaner”), 7 (“artisan”), 23 (“teacher”), 11 (“doctor”), 2 (business), 60 (none)
\item Parent’s Occupation: 17 (“cleaner”), 39 (“artisan”), 30 (“teacher”), 8 (“doctor”), 6 (business) 4 (no data)
\item  Floor: 32 (1), 17 (2), 14 (3), 28 (4), 10 (5), 3 (overseas posting)
\item  Job Description: 24 (frontline - local clients), 39 (frontline - local \& foreign clients), 41 (no clients)
\item  Employment History: 26 (never promoted), 37 (one promotion), 30 (multiple), 11 (no data)
\item  Transport to Work: 30 (bus), 50 (own car), 24 (lift)
\item  Organizational Section: 24 (\isi{JNIP}), 15 (\isi{JNEC}), 3 (\isi{JIDC}), 1 (JNIC), 61 (none)
\end{enumerate}


\section{Data on patterns of workplace interaction}
\begin{enumerate}
\item Which members of staff do you lunch with?
\item Which members of staff live in your residential community?
\item Are any members of staff related to you?
\item Which members of staff do you consider friends?
\item Which members of staff do you see outside of work?
\item Are you active on the Staff Association (\isi{JSA}, hereafter)?
\item Do you hold any (elected) position on the \isi{JSA}?
\end{enumerate}

\section{Data on working at JAMPRO}
\begin{enumerate}
\item What type of person does well and moves ahead in \isi{JAMPRO}?
\item Why do you think staffers feel as they do about the \isi{JSA}?
\item Who is responsible for your promotion?
\item Do you see yourself here in 5 or 10 years? 36 (yes), 50 (no), 18 (don’t know)
\item How important do you think these are in determining who does well here:
\item \isi{gender} (36.5\%)     b) colour (16\%)     c) class (40\%)      d) education (87.5\%)     e) other (15\%) (These percentages are calculated out of a possible 104 mentions of a particular social factor.)
\item Which levels in \isi{JAMPRO} do you think have authority and constitute top management?
\end{enumerate}
