%!TEX root = ../main.tex

\appendix
\addchap{Sample text: Nzürna trikasi}

\section*{\emph{Nzürna trikasi}}\label{nzurna}

This text belongs to a genre called \emph{nzürna trikasi} `\emph{nzürna} stories'. I translate \emph{nzürna} with `devil', `spirit' or `witch'. Although all of these translations fall short of a full description, the \emph{nzürna} character has some resemblance to witches in a western context. They are malevolent beings, usually old women, who live in the forest. They have long eyebrows and sharp fingernails, with which they disembowel people to devour them. They can change their appearance to look like a human being. They can summon and control animals, especially the centipede (\emph{züm}). They often trick people who foolishly walk alone in the forest. \footnote{All texts in the appendix have been edited in the following way: (i) I corrected mistakes that came up during the transcription (ii) I have removed overly long speech pauses and (iii) I have lumped together some annotations and split others. All editorial changes were kept to a minimum. The unedited versions can be found in the archive.}

Although \emph{nzürna} stories belong to a particular place, they are very much public stories. The \emph{nzürna} character is often joked about openly. For example, one may call a person or a dog ``\emph{nzürna}'', when it roams around in the dark. Although there are many local variations of the \emph{nzürna} theme, we can identify some recurring elements. First, the \emph{nzürna} often lives in a tree, usually a \emph{wäsi} tree. Secondly, most stories involve some innocent person who is killed and eaten. Third, relatives and friends of the victim take revenge by burning the \emph{nzürna}.

The following \emph{nzürna} story belongs to the hamlet of Firra. The narrator is Maraga Kwozi. He was born in Firra, but he told me the story in Morehead. This \emph{nzürna} story deviates in two points. First, the \emph{nzürna} character lives together with a husband, and they have children. Secondly, the \emph{nzürna} character lived in harmony with the people of Firra up until she kills and eats a visitor.\\

\noindent
This text can be accessed under: \href{https://zenodo.org/record/1294666}{https://zenodo.org/record/1294666}

\clearpage
\begin{exe}
\exi{1}
	\emph{zaföwä ... fthé kabe keke kwot tüfr thfrärm}\\
	\gll zafe=wä (.) fthé kabe keke kwot tüfr thf\stem{rä}rm\\
	before={\Emph} (.) when man {\Neg} properly plenty \Stpl:\Sbj:\Pst:\Dur/be\\
	\trans `Long time ago, that is when there were not many people'
\exi{2}
	\emph{thwamnzrm zane kafar baden thé z kabe enrera}\\
	\gll thwa\stem{m}nzrm zane kafar bad=en thé z kabe en\stem{rä}ra\\
	\Stpl:\Sbj:\Pst:\Dur/dwell \Dem:{\Prox} big ground={\Loc} when {\Iam} man \Stpl:\Sbj:\Pst:\Ipfv:\Venit/be\\
	\trans `and they were living here on this land. That is the time when people came.'
\exi{3}
	\emph{nä kabe thfamnzrm ... mogarkamen ... kar nima rä ... mogarkam}\\
	\gll nä kabe thfa\stem{m}nzrm (.) mogarkam=en (.) kar nima \stem{rä} mogarkam\\
	{\Indf} man \Stpl:\Sbj:\Pst:\Dur/dwell (.) mogarkam={\Loc} (.) village {like.this} \Tsg.\F:\Nonpast:\Ipfv/be mogarkam\\
	\trans `Some people lived in Mogarkam ... There is a village there ... Mogarkam.'
\exi{4}
	\emph{okay, nä thfamnzrm firran}\\
	\gll okay nä thfa\stem{m}nzrm firra=n\\
	okay {\Indf} \Stpl:\Sbj:\Pst:\Dur/dwell firra={\Loc}\\
	\trans `Okay, others lived in Firra.'
\exi{5}
	\emph{okay, nä fä fefe thwamnzrm mänwä kar bramöwä erä}\\
	\gll okay nä fä fefe thwa\stem{m}nzrm mä=wä kar bramöwä e\stem{rä}\\
	okay {\Indf} {\Dist} really \Stpl:\Sbj:\Pst:\Dur/dwell where={\Emph} place all \Stpl:\Sbj:\Nonpast:\Ipfv/be\\
	\trans `Okay, others lived right there, where all the villages (and hamlets) are.'
\exi{6}
	\emph{firra mrmren ... mane zfrärm ... nzürna ŋare bobo zwamnzrm}\\
	\gll firra mrmr=en (.) mane zf\stem{rä}rm (.) nzürna ŋare bobo zwa\stem{m}nzrm\\
	firra inside={\Loc} (.) which \Tsg.\F:\Sbj:\Pst:\Dur/be (.) nzürna woman \Med.{\All} \Tsg.\F:\Sbj:\Pst:\Dur/dwell\\
	\trans `As for Firra, a nzürna woman lived in the village.'
\exi{7}
	\emph{nzürna ŋare nafafisrwä thfrnm}\\
	\gll nzürna ŋare nafa-fis=r=wä thf\stem{rn}m\\
	nzürna woman \Third.\Poss-husband=\Assoc.\Du={\Emph} \Stdu:\Sbj:\Pst:\Dur/be\\
	\trans `The nzürna woman was with her husband.'
\exi{8}
	\emph{nafafis yf nagawa ... tnztnz kabe sfrärm}\\
	\gll nafa-fis yf nagawa (.) tnz-tnz kabe sf\stem{rä}rm\\
	\Third.\Poss-husband name nagawa (.) \Redup-short man \Tsg.\Masc:\Sbj:\Pst:\Dur/be\\
	\trans `His name (was) Nagawa ... He was a short guy.'
\exi{9}
	\emph{nafane ŋare ... nzürna ŋare fof yf mane zfrärm zafo ... nafrr thwamrnm}\\
	\gll nafane ŋare (.) nzürna ŋare fof yf mane zf\stem{rä}rm zafo (.) nafrr thwam\stem{rn}m\\
	\Tsg.{\Poss} woman (.) nzürna woman {\Emph} name which \Tsg.\F:\Sbj:\Pst:\Dur/be zafo (.) \Tdu.{\Assoc} \Stdu:\Sbj:\Pst:\Dur/be\\
	\trans `His wife ... the nzürna woman whose name was Zafo ... He lived with her.'
\exi{10}
	\emph{wati, mä fefe thwamrnm wäsü ... nafanme mnz zfrärm}\\
	\gll wati, mä fefe thwam\stem{rn}m wäsü (.) nafanme mnz zf\stem{rä}rm\\
	then where really \Stdu:\Sbj:\Pst:\Dur/dwell wäsü (.) \Tnsg.{\Poss} house \Tsg.\F:\Sbj:\Pst:\Dur/be\\
	\trans `Where they really lived (was) the wäsü tree (Ficus elastica) ... it was their house.'
\exi{11}
	\emph{wäsü kafar sukogrm ... ane yfön thuthkrnm}\\
	\gll wäsü kafar su\stem{kogr}m (.) ane yfö=n thu\stem{thkr}nm\\
	wäsü big \Tsg.\Masc:\Sbj:\Pst:\Dur:\Stat/stand (.) {\Dem} hole={\Loc} \Stdu:\Sbj:\Pst:\Dur:\Stat/be.inside\\
	\trans `There was a big wäsü tree standing ... They were inside that hole.'
\exi{12}
	\emph{boba mnz nafanme zfrärm mä thwamrnm}\\
	\gll boba mnz nafanme zf\stem{rä}rm mä thwa\stem{m}rnm\\
	\Med.{\Abl} house \Tnsg.{\Poss} \Tsg.\F:\Sbj:\Pst:\Dur/be where \Stdu:\Sbj:\Pst:\Dur/dwell\\
	\trans `This was their house, where they were living'
\exi{13}
	\emph{firra kar mrmren kabe thwamnzrm fobo}\\
	\gll firra kar mrmr=en kabe thwa\stem{m}nzrm fobo\\
	firra village inside={\Loc} man \Stpl:\Sbj:\Pst:\Dur/dwell \Dist.{\All}\\
	\trans `People were living over there in the village of Firra.'
\exi{14}
	\emph{kabe fthé kwarfakunzrmth fthé thfyakm}\\
	\gll kabe fthé kwa\stem{rfaku}nzrmth fthé thf\stem{yak}m\\
	man when \Stpl:\Sbj:\Pst:\Dur/sprinkle when \Stpl:\Sbj:\Pst:\Dur/walk\\
	\trans `When the people spread out, when they went ...'
\exi{15}
	\emph{nima ŋarake znfo o fiyafr o ... nima efothen ... etfthmöwä fthé thfyakm}\\
	\gll nima ŋarake zn=fo o fiyaf=r o (.) nima efoth=en (.) etfth=me=wä fthé thf\stem{yak}m\\
	{like.this} fence place={\Loc} or hunting={\Purp} or (.) {like.this} day={\Loc} (.) sleep=\Ins={\Emph} when \Stpl:\Sbj:\Pst:\Dur/walk\\
	\trans `like this to the garden place or hunting or during the day ... or when they went overnight'
\exi{16}
	\emph{ane nzürna ŋare ausi fof kwänzinzr ... fi zwanyakm}\\
	\gll ane nzürna ŋare ausi fof kwän\stem{zinzr} (.) fi zwan\stem{yak}m\\
	{\Dem} spirit woman {old woman} {\Emph} \Stsg:\Sbj:\Iter:\Venit/replace (.) \Third.{\Abs} \Tsg.\F:\Sbj:\Pst:\Dur:\Venit/walk\\
	\trans `that nzürna woman, that old woman always took over the place ... she came.'
\exi{17}
	\emph{gatha kar fthé thumarwrm ... gathagathame thfnakwrmth mnz gatha kar.}\\
	\gll {gatha kar} fthé thu\stem{mar}wrm (.) {gathagatha}=me thf\stem{nak}wrmth mnz {gatha kar}\\
	rubbish when \Stsg:\Sbj>\Stpl:\Obj:\Pst:\Dur/see (.) bad={\Ins} \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Dur/put.down house rubbish\\
	\trans `When she saw the rubbish ... they had carelessly put down the rubbish in the house'
\exi{18}
	\emph{dödö thfefaf ane zurenwrmo mnz fath thwafiyokwrm}\\
	\gll dödö thfe\stem{faf} ane zu\stem{ren}wrmo mnz fath thwa\stem{fiyok}wrm\\
	broom \Stsg:\Sbj>\Stpl:\Obj:\Iter/hold {\Dem} \Sg:\Sbj>\Tsg.\F:\Pst:\Dur:\Andat/sweep house clear.place \Sg:\Sbj>\Stpl:\Io:\Pst:\Dur/make\\
	\trans `She always grabbed the broom, swept the house and cleaned it for them.'
\exi{19}
	\emph{nafanme kkauna monme gathagathame thfnakwrmth kwot namäme thfanakwrm}\\
	\gll nafanme k-kauna mon=me gathagatha=me thf\stem{nak}wrmth kwot namä=me thfa\stem{nak}wrm\\
	\Tnsg.{\Poss} \Redup-stuff how={\Ins} bad={\Ins} \Stpl:\Sbj>\Stpl:\Io:\Pst:\Dur/put.down properly good={\Ins} \Stsg:\Sbj>\Stpl:\Obj:\Pst:\Dur/put.down\\
	\trans `How they had dropped their things carelessly, she was sorting them properly.'
\exi{20}
	\emph{mnzen thwarakthkwrmo ... mni tnztnz rä ... kwanbrigwrm nafanemäwä mnzfo}\\
	\gll mnz=en thwa\stem{rakthk}wrmo (.) mni tnz-tnz \stem{rä} (.) kwan\stem{brig}wrm nafaneme=wä mnz=fo\\
	house={\Loc} \Sg:\Sbj>\Stpl:\Io:\Pst:\Dur:\Andat/put.on.top (.) firewood \Redup-short \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be (.) \Stsg:\Sbj:\Pst:\Dur/return \Tnsg.\Poss=Emph{} house=\All\\
	\trans `She put their (things) back in the house ... for example the small pices of firewood ... She brought them back to their house.'
\exi{21}
	\emph{fthé we thwanyakm thwänthor ... ane mnz woga fthé swänthor}\\
	\gll fthé we thwan\stem{yak}m thwän\stem{thor} (.) ane mnz woga fthé swän\stem{thor}\\
	when also \Stpl:\Sbj:\Pst:\Dur:\Venit/walk \Stpl:\Sbj:\Iter:\Venit/arrive (.) {\Dem} house man when \Tsg.\Masc:\Sbj:\Iter:\Venit/arrive\\
	\trans `When they were coming back, each time they arrived ... each time when the houseowner arrived'
\exi{22}
	\emph{``oh zane ŋare z nzwänyak mnz fath zf nzürenwro zrä.''}\\
	\gll oh zane ŋare z nz=wän\stem{yak} mnz fath zf nz=w\stem{ren}wro z=\stem{rä}\\
	oh \Dem:{\Prox} woman {\Iam} \Immpst=\Tsg.\F:\Nonpast:\Ipfv:\Venit/walk house {clear place} {\Imm} \Immpst=\Sg:\Sbj>\Tsg.\F:\Nonpast:\Ipfv:\Andat/sweep \Prox=\Tsg.\F:\Sbj:\Nonpast:\Ipfv/be\\
	\trans `(he said) ``Oh, this woman already came. She has swept the houseyard just now.'''
\exi{23}
	\emph{zafe kabe miyatha thfrärm nafanme rzarsi monme zfrärm ... ane kar woganzo}\\
	\gll zafe kabe miyatha thf\stem{rä}rm nafanme rzar-si mon=me zf\stem{rä}rm (.) ane kar woga=nzo\\
	before man knowledgeable \Stpl:\Sbj:\Pst:\Dur/be \Tnsg.{\Poss} tie-{\Nmlz} how={\Ins} \Tsg.\F:\Sbj:\Pst:\Dur/be (.) {\Dem} village man={\Only}\\
	\trans `Before, the people people knew about their ties (to her) ... only those village people.'
\exi{24}
	\emph{nafä fi monme nzürna ŋareyä kwarzarwrmth}\\
	\gll nafä fi mon=me nzürna ŋare=ä kwa\stem{rzar}wrmth\\
	\Tpl.{\Assoc} \Third.{\Abs} how={\Ins} spirit woman=\Assoc.{\Pl} \Stpl:\Sbj:\Pst:\Dur/tie\\
	\trans `how they were behaving towards the \emph{nzürna} woman.'
\exi{25}
	\emph{keke thufnzrm ane karma kabe naf ... bänema fi nar wogathatha bäne thfrärm nima}\\
	\gll keke thu\stem{fn}nzrm ane kar=ma kabe naf (.) bäne=ma fi nar woga=thatha bäne thf\stem{rä}rm nima\\
	{\Neg} \Stsg:\Sbj>\Stpl:\Obj:\Pst:\Dur/hit {\Dem} village={\Char} man \Tsg.{\Erg} (.) \Dem:\Med={\Char} \Third.{\Abs} friend man=\Simil{} \Dem:\Med{} \Stpl:\Sbj:\Pst:\Dur/be {like.this}\\
	\trans `She did not attack the people from that village ... because they were like friends ...'
\exi{26}
	\emph{miyatha thfrärm ... nafane nagayé thfrärm naf thwamonegwrm ... kabe fefe}\\
	\gll miyatha thf\stem{rä}rm (.) nafane nagayé thf\stem{rä}rm naf thwa\stem{moneg}wrm (.) kabe fefe\\
	knowledgeable \Stpl:\Sbj:\Pst:\Dur/be (.) \Tnsg.{\Poss} children \Stpl:\Sbj:\Pst:\Dur/be \Tsg.{\Erg} \Stsg:\Sbj>\Stpl:\Io:\Pst:\Dur/look.after (.) man really\\
	\trans `They knew about this. They were her children. She looked after them ... really the people.'
\exi{27}
	\emph{wati, nä kayé ... mogarkamma kabe nima sfyakm firrafo}\\
	\gll wati nä kayé (.) mogarkam=ma kabe nima sf\stem{yak}m firra=fo\\
	then {\Indf} yesterday (.) mogarkam={\Char} man {like.this} \Tsg.\Masc:\Sbj:\Pst:\Dur/walk firra={\All}\\
	\trans `Well, one day, a man from Mogarkam walked this way to Firra.'
\exi{28}
	\emph{wati, fi mane yara nama zokwasi woga yara}\\
	\gll wati fi mane ya\stem{r}a nama zokwasi woga ya\stem{r}a\\
	then \Third.{\Abs} which \Tsg.\Masc:\Sbj:\Pst:\Ipfv/be nama language man \Tsg.\Masc:\Sbj:\Pst:\Ipfv/be\\
	\trans `As for this one, he was a speaker of Nama.'
\exi{29}
	\emph{firran mane thwamnzrm mema zokwasi woga yara ... fthé thwamnzrm kabe}\\
	\gll firra=n mane thwa\stem{m}nzrm mema zokwasi woga ya\stem{r}a (.) fthé thwa\stem{m}nzrm kabe\\
	firra={\Loc} which \Stpl:\Sbj:\Pst:\Dur/dwell mema language man \Tsg.\Masc:\Pst:\Ipfv/be (.) when \Stpl:\Sbj:\Pst:\Dur/dwell man\\
	\trans `As for the ones who lived in Firra, they were speaker of Mema. That'a when the people lived in Firra.'
\exi{30}
	\emph{wati, fi mane yaka e ``krara krara krara''}\\
	\gll wati fi mane \stem{yak}a e 3x{krara}\\
	then \Third.{\Abs} which \Tsg.\Masc:\Sbj:\Pst:\Ipfv/walk until 3x{sound.of.coockatoo}\\
	\trans `Well, when he walked. ``krara krara krara'''
\exi{31}
	\emph{firra sathora fof with fren fof ``krara krara krara''}\\
	\gll firra sa\stem{thor}a fof with fr=en fof 3x{krara}\\
	firra \Tsg.\Masc:\Sbj:\Pst:\Pfv{} {\Emph} banana stem={\Loc} {\Emph} 3x{sound.of.coockatoo}\\
	\trans `He arrived in Firra (and he went) between the banana trees ``krara krara krara'''
\exi{32}
	\emph{fi zära yakme we senis zära ... kabe wokuthé zäkora nima kabe}\\
	\gll fi zä\stem{r}a yakme we senis zä\stem{r}a (.) kabe woku-thé zä\stem{kor}a nima kabe\\
	\Third.{\Abs} \Stsg:\Sbj:\Pst:\Pfv/do quickly also change \Stsg:\Sbj:\Pst:\Pfv/do (.) man skin-{\Adlzr} \Stsg:\Sbj:\Pst:\Pfv/become {like.this} man\\
	\trans `He quickly changed. He became like a human being ... like a man.'
\exi{33}
	\emph{ane si thäbu zanfr ra zane thfrärm ... ofe ŋarerath ... zäwthefa ... kabe zäkora}\\
	\gll ane si thäbu zanfr ra zane thf\stem{rä}rm (.) ofe ŋa\stem{rä}rath (.) zä\stem{wthef}a (.) kabe zä\stem{kor}a\\
	{\Dem} eye hair long what \Dem:{\Prox} \Stpl:\Sbj:\Pst:\Dur/be (.) absence \Stpl:\Sbj:\Pst:\Ipfv/do (.) \Stsg:\Sbj:\Pst:\Pfv/change (.) man \Stsg:\Sbj:\Pst:\Pfv/become\\
	\trans `These long eyebrows and whatever else there was, it disappeared. He changed. He became a human.'
\largerpage	
\exi{34}
	\emph{watik ŋare nima zräzigrm ``awe nzone moba nzranyak?''}\\
	\gll watik ŋare nima zrä\stem{zigrm} awe nzone moba nzran\stem{yak}\\
	then woman {\Quot} \Tsg.\F:\Sbj:\Irr:\Pfv/look.around come \Fsg.{\Poss} where.{\Abl} \Ssg:\Sbj:\Irr:\Ipfv:\Venit/walk\\\
	\trans `Well, the woman was looking around and said ``Come my friend, where do you come from?''{}'
\newpage 
\exi{35}
	\emph{naf we komnzo zära nima ``oh zane ausinzo zf zagathifth''}\\
	\gll naf we komnzo zä\stem{r}a nima oh zane ausi=nzo zf za\stem{gathif}th\\
	\Tsg.{\Erg} also only \Stsg:\Sbj:\Pst:\Pfv/do {\Quot} oh \Dem:{\Prox} {old woman}={\Only} {\Imm} \Stpl:\Sbj>\Tsg.\F:\Obj:\Rpst:\Pfv/leave\\\
	\trans `He was also thinking ``Oh, they have left only this old woman behind.'''
\exi{36}
	\emph{``kabe matak erä nima z bramöwä kwafarkwrth nima erä ŋarsfo'' ... ``awow''}\\
	\gll kabe matak e\stem{rä} nima z bramöwä kwa\stem{fark}wrth nima e\stem{rä} ŋars=fo (.) awow\\
	man nothing \Stpl:\Sbj:\Nonpast:\Ipfv/be {like.this} {\Iam} all \Stpl:\Sbj:\Pst:\Ipfv/set.off {like.this} \Stpl:\Sbj:\Nonpast:\Ipfv/be river={\All} (.) okay\\
	\trans ` ``Nobody is here. All the people left this way to the river.'' ... ``Okay'''
\exi{37}
	\emph{yamenzo srathams ... kramath with tayo yanrkunzr ... yarithr}\\
	\gll yame=nzo sra\stem{thams} (.) kra\stem{math} with tayo yan\stem{rku}nzr (.) ya\stem{ri}thr\\
	mat={\Only} \Stsg:\Sbj>\Tsg.\Masc:\Io:\Irr:\Pfv/spread.out (.) \Stsg:\Sbj:\Irr:\Pfv/run banana ripe \Stsg:\Sbj>\Tsg.\Masc:\Io:\Nonpast:\Ipfv/knock.down (.) \Stsg:\Sbj>\Tsg.\Masc:\Io:\Nonpast:\Ipfv/give\\
	\trans `She spread (the mat) for him, ran and knocked down some ripe banana for him ... and gave them to him.'
\exi{38}
	\emph{kafar famä zäkora nima ``nzone dagonma zane zf yé. z nzyanyak''}\\
	\gll kafar fam=ä zä\stem{kor}a nima nzone dagon=ma zane zf \stem{yé} z nz=yan\stem{yak}\\
	big thought={\Assoc} \Stsg:\Sbj:\Pst:\Pfv/become {\Quot} \Fsg.{\Poss} food={\Char} \Dem:{\Prox} {\Imm} \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/be {\Iam} \Immpst=\Tsg.\Masc:\Sbj:\Nonpast:\Ipfv:\Venit/walk\\\
	\trans `She had big thoughts ``This one here is my dinner. He came already.'''
\exi{39}
	\emph{bänema yrgfakwa nima ``zane karma keke yé. moba zane nm nzyanyak?''}\\
	\gll bäne=ma y\stem{rgfak}wa nima zane kar=ma keke \stem{yé} moba zane nm nz=yanyak\\
	\Dem:\Med={\Char} \Stsg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Ipfv/recognise {\Quot} \Dem:{\Prox} village={\Char} {\Neg} \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/be where.{\Abl} \Dem:{\Prox} maybe  \Immpst=\Tsg.\Masc:\Sbj:\Nonpast:\Ipfv:\Venit/walk\\
	\trans `because she realised ``He is not from this village. Where might he have come from?'''
\exi{40}
	\emph{garamgaram srethkäf ``kwa ŋabrigwr? efoth byé!''}\\
	\gll {garamgaram} sre\stem{thkäf} kwa ŋa\stem{brig}wr efoth b=\stem{yé}\\
	{sweet.talk} \Stsg:\Sbj>\Tsg.\Masc:\Obj:\Irr:\Pfv/start {\Fut} \Stsg:\Sbj:\Nonpast:\Ipfv/return sun \Med=\Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/be\\
	\trans `She started sweet-talking him ``Will you return today? The sun setting already?'''
\exi{41}
	\emph{``keke, zä zf kwa worugr. kwa fof thrämonesé kayé fthé thräthor.''}\\
	\gll keke zä zf kwa wo\stem{rug}r kwa fof thrä\stem{mones}é kayé fthé thrä\stem{thor}\\
	{\Neg} {\Prox} {\Imm} {\Fut} \Fsg:\Sbj:\Nonpast:\Ipfv/sleep {\Fut} {\Emph} \Fsg:\Sbj>\Stpl:\Obj:\Irr:\Pfv/wait tomorrow when \Stpl:\Sbj:\Irr:\Pfv/arrive\\
	\trans ` ``No, I will sleep right here. I will wait until they return tomorrow.'''
\exi{42}
	\emph{zbär ... faf ŋathamsakrnth ... etfth kramnzerth}\\
	\gll zbär (.) faf ŋa\stem{thamsak}rnth (.) etfth kra\stem{mnzer}th\\
	night (.) place \Stdu:\Sbj:\Nonpast:\Ipfv/spread.out (.) sleep \Stdu:\Sbj:\Irr:\Pfv/fall.asleep\\
	\trans `The night came and they spread the mats. They fell asleep.'
\exi{43}
	\emph{etfth kwosi krämnzer ... ausi nzürna ŋare krebnaf ``züm züm züm züm''}\\
	\gll etfth kwosi krä\stem{mnzer} (.) ausi nzürna ŋare kre\stem{bnaf} 4x(züm)\\
	sleep dead \Stsg:\Sbj:\Irr:\Pfv/fall.asleep (.) {old woman} spirit woman \Stsg:\Sbj:\Irr:\Pfv/wake.up 4x(centipede)\\
	\trans `He was fast asleep! The nzürna woman woke up (and called out) ``centipedes! centipedes! centipedes! centipedes!'''
\exi{44}
	\emph{subnazrm fof ... sain swarithrm ... wati}\\
	\gll su\stem{bnaz}rm fof (.) sain swa\stem{ri}thrm (.) wati\\
	\Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Dur/wake.up {\Emph} (.) sign \Sg:\Sbj>\Tsg.\Masc:\Io:\Pst:\Dur/wake.up (.) enough\\
	\trans `She was really waking him up, giving him a sign ... but no.'
\exi{45}
	\emph{keke zethäkna ane}\\
	\gll keke ze\stem{thäkn}a ane\\
	{\Neg} \Sg:\Sbj:\Pst:\Pfv/shake {\Dem}\\
	\trans `That guy did not move.'
\exi{46}
	\emph{yaka zanrnzo srewakuth.}\\
	\gll yaka zan=r=nzo sre\stem{wakuth}\\
	yam.stick hit=\Purp={\Only} \Sg:\Sbj>\Tsg.\Masc:\Obj:\Irr:\Pfv/pick.up\\
	\trans `She picked up the yam stick to kill him.'
\exi{47}
	\emph{di fof safrnza kwosi.}\\
	\gll di fof sa\stem{frnz}a kwosi\\
	{back.of.head} {\Emph} \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Pfv/uproot dead\\\
	\trans `She whacked him on the head and killed him.'
\exi{48}
	\emph{kwot yanatha fä fof ... bramöwä.}\\
	\gll kwot ya\stem{na}tha fä fof (.) bramöwä\\
	properly \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Ipfv/eat {\Dist} {\Emph} (.) all\\
	\trans `She ate him there ... all of him.'
\exi{49}
	\emph{sabtha wthnzo ezänzr. füni komnzo bikogr firran.}\\
	\gll sa\stem{bth}a wth=nzo e\stem{zä}nzr füni komnzo b=y\stem{kogr} firra=n\\
	\Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Pfv/finish intestines={\Only} \Stsg:\Sbj>\Stpl:\Obj:\Nonpast:\Ipfv/carry füni  still \Med=\Tsg.\Masc:\Sbj:\Nonpast:\Stat/stand firra={\Loc}\\
	\trans `She finished him and carried away only the intestines. The füni tree still stands in Firra.'
\exi{50}
	\emph{wämne ... yf füni yé ... firran bä ykogr.}\\
	\gll wämne (.) yf füni \stem{yé} (.) firra=n bä y\stem{kogr}\\
	tree (.) name füni \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/be (.) firra={\Loc} \Med{} \Tsg.\Masc:\Sbj:\Nonpast:\Stat/stand\\
	\trans `The name of the tree is füni. It stands there in Firra'
\exi{51}
	{\footnotesize{Maraga addresses me directly now.}}\\
	\emph{nä kayé fthé boba gnyako nima kwa ymarwr ... ane kafar wämne.}\\
	\gll nä kayé fthé boba gn\stem{yak}o nima kwa y\stem{mar}wr (.) ane kafar wämne\\
	{\Indf} yesterday when \Med.{\All} \Ssg:\Sbj:\Imp:\Ipfv:\Andat/go {like.this} {\Fut} \Stsg:\Sbj>\Tsg.\Masc:\Nonpast:\Ipfv/see (.) {\Dem} big tree\\
	\trans `When you go there some day, you will see it ... that big tree.'
\exi{52}
	{\footnotesize{Marua, who sits in the back, tells him that I have been to Firra in the previous week.}}\\
	\emph{z nyakako? ... zba mothfa mane ykogr füni.}\\
	\gll z n\stem{yak}ako (.) zba moth=fa mane y\stem{kogr} füni\\
	{\Iam} \Ssg:\Sbj:\Pst:\Ipfv:\Andat/walk (.) \Prox.{\Abl} path={\Abl} which \Tsg.\Masc:\Sbj:\Nonpast:\Stat/stand füni\\
	\trans `You already went? It is here on the road, where the füni tree stands.'
\exi{53}
	\emph{ane bafen ... yakan dganzo saräsa.}\\
	\gll ane baf=en (.) yaka=n dga=nzo sa\stem{räs}a\\
	{\Dem} \Recog={\Loc} (.) yamstick={\Loc} bifurcation={\Only} \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Pfv/erect\\
	\trans `At that place ... on the yamstick ... she rammed it in the ground.'
\newpage 	
\exi{54}
	\emph{wth fobo fof thämira ... ane kabeane wth. fi zäbrimako ... zäthbako mnzen.}\\
	\gll wth fobo fof thä\stem{mir}a (.) ane kabe=ane wth fi zä\stem{brim}ako (.) zä\stem{thb}ako mnz=en\\
	intestines \Dist.{\All} {\Emph} \Sg:\Sbj>\Stpl:\Obj:\Pst:\Pfv/hang (.) {\Dem} man={\Poss} intestines \Third.{\Abs} \Sg:\Sbj:\Pst:\Pfv:\Andat/return (.) \Sg:\Sbj:\Pst:\Pfv:\Andat/enter house=\Loc\\
	\trans `She hanged the intestines up there ... that guy's intestines! Then she went back. She went inside the house.'
\exi{55}
	\emph{nafafis oromanf zräses fof ``be ranzo änfiyokwr, ah? ... bä moba nrä? ... mä nznrugr?''}\\
	\gll nafa-fis oroman=f zrä\stem{ses} fof be ra=nzo än\stem{fiyok}wr ah (.) bä moba n\stem{rä} (.) mä nzn\stem{rugr}\\
	\Third.\Poss-husband {old.man}=\Erg.{\Sg} \Stsg:\Sbj>\Tsg.\F:\Obj:\Irr:\Pfv/ask {\Emph} \Ssg.{\Erg} what={\Only} \Stsg:\Sbj>\Stpl:\Obj:\Nonpast:\Ipfv:\Venit/make ah (.) \Second.{\Abs} where.{\Abl} \Ssg:\Sbj:\Nonpast:\Ipfv/be (.) where \Ssg:\Sbj:\Rpst:\Ipfv/sleep\\
	\trans `Her husband, the old man, asked her ``Hey, just what have you been up to? Where are you coming from? Where have you slept?'''
\exi{56}
	\emph{``mä kwa! bä fof zämnzeré ... zbärma.''}\\
	\gll mä kwa bä fof zä\stem{mnzer}é (.) zbär=ma\\
	where {\Fut} \Med{} {\Emph} \Fsg:\Sbj:\Rpst:\Pfv/fall.asleep (.) night=\Char\\
	\trans `She replied ``Where do you think? I slept there because it got night.'''
\exi{57}
	\emph{``nagayaneme znsän zwäfonz. ane gathagathame kkauna mane egathikwroth.''}\\
	\gll nagayé=aneme znsä=n zwä\stem{fonz} ane gathagatha=me k-kauna mane e\stem{gathik}wroth\\
	children=\Poss.{\Nsg} work={\Loc} \Fsg:\Sbj:\Rpst:\Pfv/be.caught.by.nightfall {\Dem} bad={\Ins} \Redup-thing which \Pl:\Sbj>\Stpl:\Obj:\Nonpast:\Ipfv:\Andat/leave\\
	\trans ```I was caught by nightfall while working for the children sorting those things which they leave scattered around.'''
\exi{58}
	\emph{nafafis miyamr.}\\
	\gll nafa-fis miyamr\\
	\Third.\Poss-husband ignorance\\
	\trans `Her husband was clueless.'
\exi{59}
	\emph{fi thé enthorakwa ... mnz kabe fof ... nima thäzigrthma}\\
	\gll fi fthé en\stem{thorak}wa (.) mnz kabe fof (.) nima thä\stem{zigrthm}a\\
	\Third.{\Abs} when \Stpl:\Sbj:\Pst:\Ipfv:\Venit/arrive (.) house people {\Emph} (.) {\Quot} \Stpl:\Sbj:\Pst:\Pfv/look\\
	\trans `At that time the house owners returned to the village. They looked around and said,'
\exi{60}
	\emph{``nä tmatm ffé nzŋawänzr ... manema kabe zä naf nzyanathr?''}\\
	\gll nä tmatm fefe nz=ŋa\stem{wä}nzr (.) mane=ma kabe zä naf nz=ya\stem{na}thr\\
	{\Indf} event real \Immpst=\Stsg:\Sbj:\Nonpast:\Ipfv/happen (.) which={\Char} man {\Prox} \Tsg.{\Erg} \Immpst=\Stsg:\Sbj>\Tsg.\Masc:\Obj:\Nonpast:\Ipfv/eat\\
	\trans ` ``Something terrible has just happened. From which village was the man who she ate here?'''
\exi{61}
	\emph{äniyaka zbär zf zukwinzrmth zfkonzrmth.}\\
	\gll än\stem{yak}a zbär zf zu\stem{kwi}nzrmth zf\stem{ko}nzrmth\\
	\Stpl:\Sbj:\Pst:\Ipfv:\Venit/walk night {\Imm} \Stpl:\Sbj>\Tsg.\F:\Obj:\Pst:\Dur/argue \Stpl:\Sbj>\Tsg.\F:\Obj:\Pst:\Dur/tell\\
	\trans `In the night, they came right here and they cursed her and told her.'
\exi{62}
	\emph{zäbrimath ``mon kwa wäfiyokwre? bänema kabe z nzirärkwr ... z nzyanathr.''}\\
	\gll zä\stem{brim}ath mon kwa wä\stem{fiyok}wre bäne=ma kabe z nz=y\stem{rärk}wr (.) z nzyanathr\\
	\Stpl:\Sbj:\Pst:\Pfv/return how {\Fut} \Fpl:\Sbj>:\Tsg.\F:\Obj:\Nonpast:\Ipfv/make \Recog={\Char} man {\Iam} \Immpst=\Stsg:\Sbj>\Tsg.\Masc:\Obj:\Nonpast:\Ipfv/mess.up (.) {\Iam} \Immpst=\Stsg:\Sbj>\Tsg.\Masc:\Obj:\Nonpast:\Ipfv/eat\\
	\trans `They returned and said ``What are we going to do with her? She already messed up this man. She already ate him.'''
\exi{63}
	\emph{wati bthan kabe thfrärm ... kabe firran mane thwamnzrm.}\\
	\gll wati bthan kabe thf\stem{rä}rm (.) kabe firra=n mane thwa\stem{m}nzrm\\
	then magic man \Stpl:\Sbj:\Pst:\Dur/be (.) men firra={\Loc} who \Stpl:\Sbj:\Pst:\Dur/be\\
	\trans `Okay, there were sorcerers living in Firra.'
\exi{64}
	\emph{wati, ttmatm äfiyokwrth. ttmatm zwafiyokwrmth}\\
	\gll wati t-tmatm ä\stem{fiyok}wrth t-tmatm zwa\stem{fiyok}wrmth\\
	then \Redup-action \Stpl:\Sbj>\Stpl:\Obj:\Nonpast:\Ipfv/make \Redup-action \Stpl:\Sbj>\Tsg.\F:\Pst:\Dur/make\\
	\trans `Okay, they make their magic things. They were doing this to her.'
\exi{65}
	\emph{fam wäfiyokwrth näbinzo ``mnime n zräföfe.'' ... nima tmatmr rä}\\
	\gll fam wä\stem{fiyok}wrth näbi=nzo mni=me n zrä\stem{föf}e (.) nima tmatm=r \stem{rä}\\
	thought \Stpl:\Sbj>\Tsg.\F:\Nonpast:\Ipfv/make one={\Only} fire={\Ins} {\Imn} \Fpl:\Sbj>\Tsg.\F:\Obj:\Irr:\Pfv/burn (.) {like.this} action={\Purp} \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be\\
	\trans `They came up with a plan ``We will burn her with fire.'' This was the plan.'	
\exi{66}
	\emph{wati, bthan tmatmme nafane fam zwarmänwrmth.}\\
	\gll wati bthan tmatm=me nafane fam zwa\stem{rmän}wrmth\\
	then magic action={\Ins} \Tsg.{\Poss} though \Stpl:\Sbj>\Tsg.\F:\Io:\Pst:\Dur/close\\
	\trans `Okay, they were distracting her mind with magic.'
\exi{67}
	\emph{wärmänwath e ane bafen keke kwamätrakwrm bänema fam z zürmänth.}\\
	\gll wä\stem{rmän}wath e ane baf=en keke kwa\stem{mätrak}wrm bäne=ma fam z zü\stem{rmän}th\\
	\Stpl:\Sbj>\Tsg:\Obj:\Io:\Nonpast:\Ipfv/close until {\Dem} \Recog={\Loc} {\Neg} \Sg:\Sbj:\Pst:\Dur/come.out \Recog={\Loc} thought {\Iam} \Stpl:\Sbj>\Tsg.\F:\Io:\Iter/close\\
	\trans `They were distracting her so that she did not come out of that place, because her thoughts were always distracted.'
\exi{68}
	\emph{wtri we z zära nima ``z zwemarth ane yam fiyoksin.''}\\
	\gll wtri we z zä\stem{r}a nima z zwe\stem{mar}th ane yam fiyok-si=n\\
	fear also {\Iam} \Tsg.\F:\Sbj:\Pst:\Ipfv/be {\Quot} {\Iam} \Stpl:\Sbj>\Fsg:\Obj:\Rpst:\Pfv/see {\Dem} event make-\Nmlz=\Loc\\
	\trans `She was also afraid and thought ``They already know that I have done this.'''
\exi{69}
	\emph{mni wthomonwrth yfö mä zfrärm.}\\
	\gll mni w\stem{thomon}wrth yfö mä zf\stem{rä}rm\\
	fire \Stpl:\Sbj>\Tsg.\F:\Nonpast:\Ipfv/pile.up.fire hole where \Tsg.\F:\Sbj:\Pst:\Dur/be\\
	\trans `They piled up the fire wood where the entrance was.'
\exi{70}
	\emph{mni wthomonwath a zräföfth.}\\
	\gll mni w\stem{thomon}wath a zrä\stem{föf}th\\
	fire \Stpl:\Sbj>\Tsg.\F:\Pst:\Ipfv/pile.up.fire until \Stpl:\Sbj>\Tsg.\F:\Obj:\Irr:\Pfv/burn\\
	\trans `They piled it up and they set it on fire.'
\exi{71}
	\emph{fi yame yrsifnzo zukonzrm boba wämne yfön fof.}\\
	\gll fi yame yr-si=f=nzo zu\stem{ko}nzrm boba wämne yfö=n fof\\
	\Third.{\Abs} mat weave-\Nmlz=\Erg={\Only} \Sg:\Sbj>\Tsg.\F:\Obj:\Pst:\Dur/become \Med.{\Abl} tree hole={\Loc} {\Emph}\\
	\trans `She was preoccupied with weaving the mat there in the tree hole.'
\exi{72}
	\emph{nafafis bana krebnaf krekaris ``u'' mni u kwan fof.}\\
	\gll nafa-fis bana kre\stem{bnaf} kre\stem{karis} u mni {u.kwan} fof\\
	\Third.\Poss-husband poor \Stsg:\Sbj:\Irr:\Pfv/wake.up \Stsg:\Sbj:\Irr:\Pfv/hear u fire {sound.of.strong.wind} \Emph\\
	\trans `Her husband woke up and heard it:``uh'' the sound of the fire.'
\exi{73}
	\emph{kafar wäsü sukogrm mrab fren.}\\
	\gll kafar wäsü su\stem{kogr}m mrab fr=en\\
	big {tree.species} \Tsg.\Masc:\Sbj:\Pst:\Dur/stand bamboo grove=\Loc\\
	\trans `The big wäsü tree was standing there in a bamboo grove.'
\exi{74}
	\emph{waisamen ... waisamen ane kar yf rä mobo zwamnzrm ... mrab fr thden.}\\
	\gll waisamen (.) waisamen ane kar yf \stem{rä} mobo zwa\stem{m}nzrm (.) mrab fr thd=en\\
	waisamen (.) waisamen {\Dem} place name \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be where.{\All} \Tsg.\F:\Sbj:\Pst:\Dur/dwell (.) bamboo grove middle={\Loc}\\
	\trans `Waisamen ... Waisamen is the name of that place where she was living ... in the middle of a bamboo grove.'
\exi{75}
	\emph{krär ... nafafis zräs ``be ranzo kayé thwanfiyokwr?''}\\
	\gll krä\stem{r} (.) nafa-fis zrä\stem{s} be ra=nzo kayé thwan\stem{fiyok}wr\\
	\Stsg:\Sbj:\Irr:\Pfv/do (.) \Third.\Poss-husband \Stsg:\Sbj>\Tsg.\F:\Irr:\Pfv/ask \Ssg.{\Erg} what={\Only} yesterday \Stsg:\Sbj>\Stpl:\Obj:\Rpst:\Ipfv:\Venit/make\\
	\trans `He got up. Her husband asked ``Just what have you done to them yesterday?'''
\exi{76}
	\emph{``ra kwa thanfiyokwé?'' }\\
	\gll ra kwa than\stem{fiyok}wé\\
	what {\Fut} \Fsg:\Sbj>\Stpl:\Obj:\Rpst:\Ipfv:\Venit/make\\
	\trans `She replied ``What do you think I have done?'''
\exi{77}
	\emph{``ra kwan we rä ah?''}\\
	\gll ra kwan we \stem{rä} ah\\
	what sound also \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be ah\\
	\trans `He asked ``and what is that sound, eh?'''
\exi{78}
	\emph{``nzukar banafa borbor bana sathor kma borbor u kwan zfrärm zufsgwrm fof.''}\\
	\gll nzu-kar bana=fa borbor bana sa\stem{thor} kma borbor {u kwan} zf\stem{rä}rm zu\stem{fsg}wrm fof\\
	\Fsg.\Poss-place poor={\Abl} thunderstorm poor \Tsg.\Masc:\Rpst:\Pfv/arrive {\Pot} thunderstorm {roaring.sound} \Tsg.\F:\Sbj:\Pst:\Dur/be \Sg:\Sbj>\Tsg.\F:\Obj:\Pst:\Dur/blow \Emph\\
	\trans `She replied ``The thunderstorm is coming from my poor village. It must be the sound of the storm blowing.'''
\newpage 	
\exi{79}
	\emph{mni kwarsirm. mni komnzo zöfthé zethkäfa.}\\
	\gll mni kwa\stem{rsi}rm mni komnzo zöfthé ze\stem{thkäf}a\\
	fire \Sg:\Sbj:\Pst:\Dur/burn fire just before \Sg:\Sbj:\Pst:\Pfv/start\\
	\trans `It was the fire burning. The fire which has just started to burn.'
\exi{80}
	\emph{nima sräzigrm nafafis bana fof zänmätra fof mni zbo z zamara.}\\
	\gll nima srä\stem{zigrm} nafa-fis bana fof zän\stem{mätr}a fof mni zbo z za\stem{mar}a\\
	{like.this} \Tsg.\Masc:\Sbj:\Irr:\Pfv/look \Third.\Poss-husband poor {\Emph} \Sg:\Sbj:\Pst:\Pfv/exit {\Emph} fire \Prox.{\All} {\Iam} \Sg:\Sbj>\Tsg.\F:\Obj:\Pst:\Pfv/see\\
	\trans `He looked around. Her poor husband stepped outside and saw the fire close by.' 	
\exi{81}
	\emph{rürä fthé zagathifa nima sathfärako.}\\
	\gll rürä fthé za\stem{gathif}a nima sa\stem{thfär}ako\\
	alone when \Sg:\Sbj>\Tsg.\F:\Obj:\Pst:\Pfv/leave {like.this} \Tsg.\Masc:\Sbj:\Pst:\Pfv:\Andat/jump\\
	\trans `That was when he left her alone. He jumped out.'
\exi{82}
	\emph{nima fi fthé sathfärako yakäsü ... trtha zuthorakwrm.}\\
	\gll nima fi fthé sa\stem{thfär}ako yak=ä=sü (.) trtha zu\stem{thorak}wrm\\
	{like.this} \Third.{\Abs} when \Tsg.\Masc:\Sbj:\Pst:\Pfv:\Andat/jump run=\Assoc=\Etc{} (.) life \Sg:\Sbj>\Tsg.\F:\Obj:\Pst:\Dur/search\\
	\trans `When he jumped out in a rush, he was running for his life.'
\exi{83}
	\emph{foba näbi fthé zanmatha fof.}\\
	\gll foba näbi fthé zan\stem{math}a fof\\
	\Dist.{\Abl} one when \Sg:\Sbj:\Pst:\Pfv:\Venit/ran \Emph\\
	\trans `From there, he ran for good.'
\exi{84}
	\emph{emoth fäth nima ämnzr oten.}\\
	\gll emoth fäth nima ä\stem{m}nzr ote=n\\
	girl \Dim{} {like.this} \Stpl:\Sbj:\Nonpast:\Ipfv/dwell ote={\Loc}\\
	\trans `The daughters are living there in Ote.'
\exi{85}
	\emph{komnzo zena bobo rä. ane kar we nä fof rä trikasi kar fof.}\\
	\gll komnzo zena bobo \stem{rä} ane kar we nä fof \stem{rä} trik-si kar fof\\
	still today \Med.{\Abl} \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be {\Dem} village also {\Indf} {\Emph} \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be tell-{\Nmlz} village {\Emph}\\
	\trans `This place is still there and there is also a story about that place.'
\exi{86}
	\emph{dödö fr rä kafar dödö fr zbo thden rä.}\\
	\gll dödö fr \stem{rä} kafar dödö fr zbo thd=en \stem{rä}\\
	{tree.species} grove \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be big {tree.species} grove \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be middle={\Loc} \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be\\
	\trans `The is a dödö grove (Melaleuca sp), a big dödö grove about halfway.'
\exi{87}
	\emph{näbüwä thé zanmatha ote. emoth fäthnm thrätrif}\\
	\gll näbi-wä fthé zan\stem{math}a ote emoth fäth=nm thrä\stem{trif}\\
	one={\Emph} when \Sg:\Sbj:\Pst:\Pfv:\Venit/run ote girl \Dim=\Dat.{\Nsg} \Stsg:\Sbj>\Stpl:\Io:\Irr:\Pfv/tell\\
	\trans `Then he ran off for good to Ote. He told the girls: '
\exi{88}
	\emph{``beŋame ausi ... bzaföfth ... nafanemäwä!''}\\
	\gll be-ŋame ausi (.) b=za\stem{föf}th (.) nafane=ma=wä\\
	\Snsg.\Poss-mother {old woman} (.) \Med=\Stpl:\Sbj>\Tsg.\F:\Obj:\Pst:\Pfv/burn (.) \Tsg.\Poss=\Char={\Emph}\\
	\trans ` ``Your mother, the old woman ... they burned her there! It was all her own fault!'''
\exi{89}
	\emph{``kafar yam zwafiyokwr ... kabe nä z swanathr!''}\\
	\gll kafar yam zwa\stem{fiyok}wr (.) kabe nä z swa\stem{na}thr\\
	big event \Stsg:\Sbj>\Tsg.\F:\Obj:\Rpst:\Ipfv/make (.) man {\Indf} {\Iam} \Stsg:\Sbj>\Tsg.\Masc:\Obj:\Rpst:\Ipfv/eat\\
	\trans ` ``She has made a big mistake. She ate a man!'''
\exi{90}
	\emph{fi nimanzo fefe yarako.}\\
	\gll fi nima=nzo fefe ya\stem{r}ako\\
	\Third.{\Abs} {like.this}={\Only} really \Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Andat/be\\
	\trans `He had left just like this.'
\exi{91}
	\emph{fi nima mni zewaräfa ... ŋarsira}\\
	\gll fi nima mni ze\stem{waräf}a (.) ŋa\stem{rsir}a\\
	but {like.this} fire \Sg:\Sbj:\Pst:\Pfv/burn.down (.) \Sg:\Sbj:\Pst:\Ipfv/burn\\
	\trans `But the fire burned down ... it was burning'
\exi{92}
	\emph{ŋarsira kma zräzigrm ``moba kwa krämätré? moba?''}\\
	\gll ŋa\stem{rsir}a kma zrä\stem{zigrm} moba kwa krä\stem{mätr}é moba\\
	\Sg:\Sbj:\Pst:\Ipfv/burn {\Pot} \Tsg.\F:\Sbj:\Irr:\Pfv/burn where.{\Abl} {\Fut} \Fsg:\Sbj:\Irr:\Pfv/exit where.{\Abl}\\
	\trans `It was burning and she tried to escape and said ``Where will I get out? Where?'''
\exi{93}
	\emph{näbi fefe zaföfath ŋarsira eee kwot zäbtha.}\\
	\gll näbi fefe za\stem{föf}ath ŋa\stem{rsir}a eee kwot zä\stem{bth}a\\
	one really \Stpl:\Sbj>\Tsg.\F:\Pst:\Pfv/burn.down \Sg:\Sbj:\Pst:\Ipfv/burn until properly \Sg:\Sbj:\Pst:\Pfv/finish\\
	\trans `They really burned her for good. The fire burned until she was finished.'
\exi{94}
	\emph{brbrnzo fof n zäthba bafen ... ymden fof.}\\
	\gll brbr=nzo fof n zä\stem{thb}a baf=en (.) ymd=en fof\\
	spirit={\Only} {\Emph} {\Imn} \Sg:\Sbj:\Pst:\Pfv/enter \Recog={\Loc} (.) bird={\Loc} \Emph\\
	\trans `Only her spirit was about to enter that bird.'
\exi{95}
	\emph{``kuka kuka'' fä mane wänor ``kuka kuka''}\\
	\gll kuka kuka fä mane wä\stem{nor} kuka kuka\\
	kuka kuka {\Dist} who \Tsg.\F:\Sbj:\Nonpast:\Ipfv/shout kuka kuka\\
	\trans ` ``kuka kuka'' the one that shouts out ``kuka kuka'' over there.'
\exi{96}
	\emph{krärth ane bthan woga ane kuka kuka zrämgth krätr.}\\
	\gll krä\stem{r}th ane bthan woga ane kuka kuka zrä\stem{mg}th krä\stem{tr}\\
	\Stpl:\Sbj:\Irr:\Pfv/do {\Dem} magic man {\Dem} kuka kuka \Stpl:\Sbj>\Tsg.\F:\Irr:\Pfv/shoot \Stsg:\Sbj:\Irr:\Pfv/fall\\
	\trans `They got up. The sorcerers shot that \emph{kuka kuka} bird and it fell down.'
\exi{97}
	\emph{wati, fefe zaföfath ane fobo fä zäbtha.}\\
	\gll wati fefe za\stem{föf}ath ane fobo fä zä\stem{bth}a\\
	then really \Stpl:\Sbj>\Tsg.\F:\Pst:\Pfv/burn.down {\Dem} \Dist.{\All} {\Dist} \Sg:\Sbj:\Pst:\Pfv/finish\\
	\trans `They burned the bird  over there until that it was finished.'
\exi{98}
	\emph{ane thrma mni fthé zäbtha wati nagawa ŋabrigwa sir}\\
	\gll  ane thrma mni fthé zä\stem{bth}a wati nagawa ŋa\stem{brig}wa si=r\\
	{\Dem} after fire when \Sg:\Sbj:\Pst:\Pfv/finish then nagawa \Sg:\Sbj:\Pst:\Ipfv/return eye=\Purp\\
	\trans `After the fire had finished, Nagawa went back to see.'
\exi{99}
	\emph{``komnzo rä o z kwarsir mnin?''}\\
	\gll komnzo \stem{rä} o z kwa\stem{rsir} mni=n\\
	sill \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be or {\Iam} \Sg:\Sbj:\Rpst:\Ipfv/burn fire=\Loc\\\
	\trans `He asked himself ``Is she still alive or did she burn in the fire?'''
\exi{100}
	\emph{ŋabrigwa bobomr we waisam wäsü thé sanmara ``watik fi nafazfthenwä''}\\
	\gll ŋa\stem{brig}wa bobomr we waisam (.) wäsü thé san\stem{mar}a watik fi nafa-zfth=en=wä\\
	\Sg:\Sbj:\Pst:\Ipfv/return until also waisam (.) wäsü when \Sg:\Sbj:\Pst:\Pfv:\Venit/see then \Third.{\Abs} \Third.\Poss-fault=\Loc={\Emph}\\
	\trans `He walked until Waisam. When he saw the wäsü tree he said ``Well, it was her own fault.'''
\exi{101}
	\emph{yanzo bobo yanora ... nafaŋareanema ... wati, fi näbi zäbrima.}\\
	\gll ya=nzo bobo ya\stem{nor}a (.) nafa-ŋare=ane=ma (.) wati fi näbi zä\stem{brim}a\\
	cry={\Only} \Med.{\All} \Tsg.\Masc:\Sbj:\Pst:\Ipfv/shout (.) \Third.\Poss-woman=\Poss={\Char} (.) then \Third.{\Abs} one \Sg:\Sbj:\Pst:\Pfv/return\\
	\trans `He was crying badly for his wife. Then he returned for good.'
\exi{102}
	\emph{zbo yamnzr ane woga oten. emoth fäthä ämnzr.}\\
	\gll zbo ya\stem{m}nzr ane woga ote=n emoth fäth=ä ä\stem{m}nzr\\
	\Prox.{\All} \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/dwell {\Dem} man ote=/Loc{} girl \Dim=\Assoc.{\Pl} \Stpl:\Sbj:\Nonpast:\Ipfv/dwell\\
	\trans `That man lives here in Ote. He lives together with his daughters.'
\exi{103}
	\emph{watik kabeyé komnzo fä nomai sumarwre ymarwre fthé}\\
	\gll watik kabe=yé komnzo fä nomai su\stem{mar}wrth y\stem{mar}wre fthé\\
	then man=\Erg.{\Nsg} still {\Dist} always \Stpl:\Sbj>\Tsg.\Masc:\Obj:\Rpst:\Ipfv/see \Fnsg:\Sbj>\Tsg.\Masc:\Obj:\Rpst:\Ipfv/see when\\
	\trans `The people still see him there, we see him when...'
\exi{104}
	\emph{fä ŋaritakwr nima firrafo yak we nima ŋabrigwr}\\
	\gll fä ŋa\stem{ritak}wr nima firra=fo \stem{yak} we nima ŋa\stem{brig}wr\\
	{\Dist} \Stsg:\Sbj:\Nonpast:\Ipfv/cross {like.this} firra={\All} \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/return\\
	\trans `he goes across the river and when he goes to Firra and also when he returns that same way again.'
\exi{105}
	\emph{tnz fäth ane kabe yé}\\
	\gll tnz fäth ane kabe \stem{yé}\\
	short \Dim{} {\Dem} man \Tsg.\Masc:\Nonpast:\Ipfv/be\\
	\trans `He is a short guy.'
\exi{106}
	\emph{ane nzürna ŋareane ... zokwasi nimame fof rä fof}\\
	\gll ane nzürna ŋare=ane (.) zokwasi nima=me fof \stem{rä} fof\\
	{\Dem} nzürna woman={\Poss} (.) speech like.this={\Ins} {\Emph} \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be {\Emph}\\
	\trans `This nzürna woman's story really happened like this.'
\exi{107}
	\emph{mane bobo firran zwamnzrm.}\\
	\gll mane bobo firra=n zwa\stem{m}nzrm\\
	who \Med.{\All} firra={\Loc} \Tsg.\F:\Sbj:\Pst:\Dur/dwell\\
	\trans `The nzürne who was living there in Firra.'
\newpage 	
\exi{108}
	\emph{tüfr yam nä ffé thwafiyokwrm}\\
	\gll tüfr yam nä ffé thwa\stem{fiyok}wrm\\
	many event {\Indf} really \Sg:\Sbj>\Stpl:\Obj:\Pst:\Dur/make\\
	\trans `She did many things, '
\exi{109}
	\emph{fi fathfa ane fof wäfiyokwa}\\
	\gll fi fath=fa ane fof wä\stem{fiyok}wa\\
	but clear.place={\Abl} {\Dem} {\Emph} \Sg:\Sbj>\Tsg.\F:\Obj:\Pst:\Ipfv/make\\
	\trans `but she did this one thing in public.'
\exi{110}
	\emph{nä karma kabe mane yanatha mogarkamma.}\\
	\gll nä kar=ma kabe mane ya\stem{na}tha mogarkam=ma.\\
	{\Indf} village={\Char} man who \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Ipfv/eat mogarkam=\Char\\
	\trans `Eating this man from another village from Mogarkam.'
\exi{111}
	\emph{nafane zokwasi ... ane trikasi fobonzo wythk fof brä ... ane nzürna ŋareanema.}\\
	\gll nafane zokwasi (.) ane trik-si fobo=nzo w\stem{ythk} fof b=\stem{rä} (.) ane nzürna ŋare=ane=ma\\
	\Tsg.{\Poss} words (.) {\Dem} tell-{\Nmlz} \Dist.\All={\Only} \Tsg.\F:\Sbj:\Nonpast:\Ipfv/come.to.end {\Emph} \Med=\Tsg.\F:\Sbj:\Nonpast:\Ipfv/be (.) {\Dem} nzürna woman=\Poss.\Sg={\Char}\\
	\trans `Her story story finishes there, the one about that nzürna woman.'
\exi{112}
	\emph{watik, fobo fof zräkoré.}\\
	\gll watik fobo fof zrä\stem{kor}é\\
	then \Dist.{\All} {\Emph} \Fsg:\Sbj>\Tsg.\F:\Obj:\Irr:\Pfv/speak\\
	\trans `Well, I have told it from there.'
\exi{113}
	\emph{nä karen nima nä buné bänema ...}\\
	\gll nä kar=en nima {nä.bun=é} bäne=ma (.)\\
	{\Indf} village={\Loc} like.this \Indf=\Erg.{\Nsg} \Recog={\Char}\\
	\trans `In other villages, others can tell'
\exi{114}
	\emph{nä nzürna ŋare zokwasi trikasi bä räro fi ane kar woga mane erä fi ane miyatha erä.}\\
	\gll nä nzürna ŋare zokwasi trik-si bä \stem{rä}ro fi ane kar woga mane e\stem{rä} fi ane miyatha e\stem{rä}\\
	{\Indf} nzürna woman words tell-{\Nmlz} \Med{} \Tsg.\F:\Sbj:\Nonpast:\Ipfv:\Andat/be but {\Dem} village man who \Stpl:\Sbj:\Nonpast:\Ipfv/be \Third.{\Abs} {\Dem} knowledgeable \Stpl:\Sbj:\Nonpast:\Ipfv/be\\
	\trans `other nzürna stories there. But it is those other village people who know about these.'
\newpage 	
\exi{115}
	\emph{nzefe nzüwäbragwé nima ni miyatha nrä.}\\
	\gll nze=wä nz=wä\stem{brag}wé nima ni miyatha n\stem{rä}\\
	\Fsg.\Erg={\Emph} \Immpst=\Fsg:\Sbj>\Tsg.\F:\Obj:\Nonpast:\Ipfv/follow like.this {\Fnsg} knowledgeable \Fpl:\Sbj:\Nonpast:\Ipfv/be\\
	\trans `I just followed the story as we know it.'
\exi{116}
	\emph{nzekaren ane yam kwafiyokwrm nzenme ŋafyé mä thwamnzrm.}\\
	\gll nze-kar=en ane yam kwa\stem{fiyok}wrm nzenme ŋafe=yé mä thwa\stem{m}nzrm\\
	\Fnsg.\Poss-village={\Loc} {\Dem} event \Sg:\Sbj:\Pst:\Dur/make \Fnsg.{\Poss} father=\Abs.{\Nsg} where \Stpl:\Sbj:\Pst:\Dur/dwell\\
	\trans `She was doing this in our village, where our fathers were living'
\exi{117}
	\emph{ŋafyé we nzenm natrikwath.}\\
	\gll ŋafe=yé we nzenm na\stem{trik}wath\\
	father=\Erg.{\Nsg} also \Fnsg.{\Dat} \Stpl:\Sbj>\Fpl:\Io:\Pst:\Ipfv/tell\\
	\trans `and our fathers also told us about it.'
\exi{118}
	\emph{nima zbo zf zakoré ... fof zäbthé.}\\
	\gll nima zbo zf za\stem{kor}é (.) fof zä\stem{bth}é.\\
	like.this \Prox.{\All} {\Imm} \Fsg:\Sbj>\Tsg.\F:\Obj:\Rpst:\Pfv/speak (.) {\Emph} \Fsg:\Sbj>\Tsg.\F:\Obj:\Rpst:\Pfv/finish\\
	\trans `I said it like this and I finished it.'	\Corpus{tci20120901-01}{}
\end{exe}%nzürna ngare