%!TEX root = ../main.tex

\appendix
\addchap{Sample text: Kwafar}

\section*{\emph{Kwafar}}\label{kwafar}

The following text was prompted by my question ``Where did the yams come from?'' It was told by Abia Bai. The text should be seen as a compendium rather than a single storyline. It was recorded as my fieldtrip in 2013 came to an end. During the previous weeks, I had talked with Abia many times about different topics and he promised to tell me these stories properly in the context of a recording session. 

The text can be cut into three storylines, which I have been told independently by others. The first part is the \emph{Kwafar} myth. \emph{Kwafar} is a place off the coast between the island of New Guinea and the Australian continent. According to the story, there was a large \emph{wäsi} tree at \emph{kwafar} and the people of different tribes and languages lived together in this tree. Eventually, the tree burned down and the people started spreading out from there. Many clans of the Morehead district have an apical ancestor who came from \emph{kwafar}. One of the many myths located at \emph{Kwafar} involves two brothers, who went hunting in the area. The brothers came across a mysterious being which devours the bodies of those people who have died in the fire. The two brothers try to shoot the creature, but only the older brother is successful. As his arrow pierces the creature, a flood of water bursts out of the wound. In recent versions of the myth, the younger brother is said to be white like Europeans. He owns a shotgun instead of a bow. He runs south towards what is now Australia. The older brother runs north. He stops the flood by beating the water with branches of \emph{dödö} (Melaleuca sp). At this point, Abia transitions into the second part. This is the story of \emph{Mathkwi}, the apical ancestor of his clan. This story involves many small episodes about the route that \emph{Mathkwi} took and all the things he carried and brought along. The third part is about customs and traditions around yam cultivation and a particular magic stone which Abia's father used to own.\\

\noindent
This text can be accessed under: \href{https://zenodo.org/record/1292876}{https://zenodo.org/record/1292876}

\clearpage
\begin{exe}
\exi{1} 
	\emph{moba zrathkäfe?}\\
	\gll moba zra\stem{thkäf}e\\ 
	where.{\Abl} \Fdu:\Sbj:\Irr:\Pfv/start\\
	\trans `Where do we start?'
\exi{2}   
	{\footnotesize{CD speaking:}}\\
	\emph{wawa moba enrera?}\\
	\gll wawa moba en\stem{rä}ra\\ 
	yam where.{\Abl} \Stpl:\Sbj:\Pst:\Ipfv:\Venit/be\\
	\trans `Where did the yams come from?'
\exi{3}
	\emph{okay, kwa zöbthé zrathkäfe nimame trikasi fof ... kwafar ...}\\
	\gll okay kwa zöbthé zra\stem{thkäf}e nima=me trik-si fof (.) kwafar (.)\\ 
	okay {\Fut} first \Fdu:\Sbj:\Irr:\Pfv/start like.this={\Ins} tell-{\Nmlz} {\Emph} (.) kwafar (.)\\
	\trans `Okay, first we will start the story really ... with Kwafar.'
\exi{4} 
	\emph{nimame fof nzranyan e zbo zrabthe}\\
	\gll nima=me fof nzran\stem{yan} e zbo zra\stem{bth}e\\ 
	like.this={\Ins} {\Emph} \Fdu:\Sbj:\Irr:\Ipfv:\Venit/walk until \Prox.{\All} \Fdu:\Sbj:\Irr:\Pfv/finish\\\
	\trans `We will go like this until we finish the story here:'
\exi{5} 
	\emph{ra nzigfu enfathwath}\\
	\gll ra nzigfu en\stem{fath}wath\\ 
	what magic.stone \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Ipfv:\Venit/hold\\
	\trans `about what magic rain stones were they holding'
\exi{6} 
	\emph{ra fofosa nzigfu enfathwath?}\\
	\gll ra fofosa nzigfu enfathwath\\ 
	what heart magic.stone \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Ipfv:\Venit/hold\\
	\trans `and what magic yam stones were they holding.'
\exi{7} 
	\emph{watik zbo zf zrabthe aneme fof.}\\
	\gll watik zbo zf zra\stem{bth}e ane=me fof\\ 
	then \Prox.{\All} {\Imm} \Fdu:\Sbj:\Irr:\Pfv/finish \Dem={\Ins} {\Emph}\\
	\trans `We will finish with this topic right here.'
\exi{8} 
	\emph{trikasi näbi kwa wänyak.}\\
	\gll trik-si näbi kwa wän\stem{yak}\\ 
	tell-{\Nmlz} one {\Fut} \Tsg.\F:\Sbj:\Nonpast:\Ipfv:\Venit/walk\\
	\trans `In this way, it will come as one story.'
\exi{9} 
	\emph{zrethkäfé?}\\
	\gll zre\stem{thkäf}é\\ 
	\Fsg:\Sbj>\Tsg.\F:\Obj:\Irr:\Pfv/start\\
	\trans `Should I start?'
\newpage 	
\exi{10} 
	\emph{okay, zane mane rä ... zane trikasi ... ŋafyf bäyf mane ŋatrikwa}\\
	\gll okay zane mane \stem{rä} (.) zane trik-si (.) ŋafe=f bäi=f mane ŋa\stem{trik}wa\\ 
	okay \Dem:{\Prox} who \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be (.) \Dem:{\Prox} tell-{\Nmlz} (.) father=\Erg.{\Sg} bäi=\Erg.{\Sg} who \Sg:\Sbj:\Pst:\Ipfv/tell\\
	\trans `As for this one, this story, it was father Bäi who told it.'
\exi{11} 
	\emph{nzenm natrikwa ... watik ane trikasi fof zena ŋaritakwr.}\\
	\gll nzenm na\stem{trik}wa (.) watik ane trik-si fof zena ŋa\stem{ritak}wr\\ 
	\Fnsg.{\Dat} \Sg:\Sbj>\Fpl:\Io:\Pst:\Ipfv/tell (.) then {\Dem} tell-{\Nmlz} {\Emph} today \Stsg:\Sbj:\Nonpast:\Ipfv/cross.over\\
	\trans `He told it to us and today it will pass on.'
\exi{12} 
	\emph{trikasi mane rä kwafarma rä.}\\
	\gll trik-si mane \stem{rä} kwafar=ma \stem{rä}\\ 
	tell-{\Nmlz} which \Tsg.\F:\Nonpast:\Ipfv/be kwafar={\Char} \Tsg.\F:\Nonpast:\Ipfv/be\\
	\trans `This story is about Kwafar.'
\exi{13} 
	\emph{``kwafar'' ŋafyf nima fof kwatrikwrm ``kwafar mane rera thden rera''}\\
	\gll kwafar ŋafe=f nima fof kwa\stem{trik}wrm kwafar mane \stem{rä}ra thd=en \stem{rä}ra\\ 
	kwafar father=\Erg.{\Sg} {\Quot} {\Emph} \Sg:\Sbj:\Pst:\Dur/tell kwafar which \Tsg.\F:\Sbj:\Pst:\Ipfv/be middle={\Loc} \Tsg.\F:\Sbj:\Pst:\Ipfv/be\\
	\trans ` ``Kwafar'', father was telling us, ``Kwafar was in the middle.'''
\exi{14} 
	\emph{zane zena mane bad mane wythk.}\\
	\gll zane zena mane bad mane w\stem{ythk}\\ 
	\Dem:{\Prox} today which ground which \Tsg.\F:\Sbj:\Nonpast:\Ipfv/come.to.end\\
	\trans `Here, where this land ends today,'
\exi{15} 
	\emph{mazo mä ŋakonzr a australiane bad mä wythk.}\\
	\gll mazo mä ŋa\stem{ko}nzr a australia=ane bad mä w\stem{ythk}\\ 
	ocean where \Stsg:\Sbj:\Nonpast:\Ipfv/become until australia=\Poss.{\Sg} ground where \Tsg.\F:\Sbj:\Nonpast:\Ipfv/come.to.end\\
	\trans `where the ocean begins until where the Australian continent ends:'
\exi{16} 
	\emph{fä mä fi zfrärm ane kwafar fof ... kabe mä kwamosinzrmth.}\\
	\gll fä mä fi zf\stem{rä}rm ane kwafar fof (.) kabe mä kwa\stem{mosi}nzrmth \\ 
	{\Dist} where \Third.{\Abs} \Tsg.\F:\Sbj:\Pst:\Dur/be {\Dem} kwafar {\Emph} (.) people where \Stpl:\Sbj:\Pst:\Dur/gather\\
	\trans `that's where Kwafar was located and where the people were gathering.'
\newpage 	
\exi{17} 
	\emph{wäsi warfo thfrugrm.}\\
	\gll wäsi warfo thf\stem{rugr}m\\ 
	wäsi above \Stpl:\Sbj:\Pst:\Dur/sleep\\
	\trans `People were sleeping on top of the wäsi tree (Ficus elastica).'
\exi{18} 
	\emph{wäsi bäne ykonzrth nä bä bikogro ... zärkarä.}\\
	\gll wäsi bäne y\stem{ko}nzrth nä bä b=y\stem{kogr}o (.) zär=karä\\ 
	wäsi \Dem:\Med{} \Stpl:\Sbj>\Tsg.\Masc:\Obj:\Nonpast:\Ipfv/speak {\Indf} \Med{} \Med=\Tsg.\Masc:\Nonpast:\Ipfv:\Andat/stand (.) shade={\Prop}\\
	\trans `They call this one wäsi. There is one standing over there ... the one with shade.'
\exi{19} 
	\emph{kabe fä fof thwamnzrm fof.}\\
	\gll kabe fä fof thwa\stem{m}nzrm fof\\ 
	people {\Dist} {\Emph} \Stpl:\Sbj:\Pst:\Dur/dwell {\Emph}\\
	\trans `The people were living there.'
\exi{20} 
	\emph{zokwasi ffrümenzo ... nä zfthen thwamnzrm nä thden thwamnzrm nä kerker thwamnzrm.}\\
	\gll zokwasi f-frü=me=nzo (.) nä zfth=en thwa\stem{m}nzrm nä thd=en thwa\stem{m}nzrm nä kerker thwa\stem{m}nzrm\\ 
	language \Redup-single=\Ins={\Only} (.) {\Indf} base={\Loc} \Stpl:\Sbj:\Pst:\Dur/dwell {\Indf} middle={\Loc} \Stpl:\Sbj:\Pst:\Dur/dwell {\Indf} tail \Stpl:\Sbj:\Pst:\Dur/dwell\\
	\trans `They spoke different languages ... some people were living at the base, some people were living in the middle and some people were living up in the branches.'
\exi{21} 
	\emph{watik zokwasi ane ffrümenzo kwanafrmth.}\\
	\gll watik zokwasi ane f-frü=me=nzo kwa\stem{na}frmth\\ 
	then language {\Dem} \Redup-single=\Ins={\Only} \Stpl:\Sbj:\Pst:\Dur/talk\\
	\trans `Well, they were speaking different languages.'
\exi{22} 
	\emph{nä kayé wäsi ane zäföfa fof ... zästha fof.}\\
	\gll nä kayé wäsi ane zä\stem{föf}a fof (.) zä\stem{sth}a fof\\ 
	{\Indf} yesterday wäsi {\Dem} \Sg:\Sbj:\Pst:\Ipfv/burn {\Emph} (.) \Sg:\Sbj:\Pst:\Pfv/set.alight {\Emph}\\
	\trans `One day that wäsi tree burned down. It really went up in flames.'
\exi{23} 
	\emph{nä kabe nima kwakwikwrmth}\\
	\gll nä kabe nima kwa\stem{kwi}kwrmth\\ 
	{\Indf} people like.this \Stpl:\Sbj:\Pst:\Dur/run\\
	\trans `Some people ran away this way.'
\exi{24} 
	\emph{nä kabe nima mnin kwarsirwrmth}\\
	\gll nä kabe nima mni=n kwa\stem{rsir}wrmth\\ 
	{\Indf} people like.this fire={\Loc} \Stpl:\Sbj:\Pst:\Dur/burn\\
	\trans `Some people burned in the fire.'
\exi{25} 
	\emph{watik wäsi ane kwot ŋarsira ... zäbtha}\\
	\gll watik wäsi ane kwot ŋa\stem{rsir}a (.) zä\stem{bth}a\\ 
	then wäsi {\Dem} properly \Sg:\Sbj:\Pst:\Ipfv/burn (.) \Sg:\Sbj:\Pst:\Pfv/finish\\\
	\trans `That wäsi tree burned completely. It finished.'
\exi{26} 
	\emph{kabe bä mane thwägrm warfo nä mrmr ... fi we nimäwä kwarsirwrmth}\\
	\gll kabe bä mane th\stem{wägr}m warfo nä mrmr (.) fi we nima=wä kwa\stem{rsir}wrmth\\ 
	people \Med{} who \Stpl:\Sbj:\Pst:\Dur/be.on.top above {\Indf} inside (.) \Third.{\Abs} also like.this={\Emph} \Stpl:\Sbj:\Pst:\Dur/burn\\
	\trans `The people who lived on top and some who lived inside, they burned.'
\exi{27} 
	\emph{watik ... ezi ... kabe ane frümenzo tnägsi zethkäfath ... bä frümenzo thwamnzrm.}\\
	\gll watik (.) ezi (.) kabe ane frü=me=nzo tnäg-si zä\stem{thkäf}ath (.) bä frü=me=nzo thwa\stem{m}nzrm\\ 
	then (.) morning (.) people {\Dem} single=\Ins={\Only} lose-{\Nmlz} \Stpl:\Sbj:\Pst:\Pfv/start (.) \Med{} single=\Ins={\Only} \Stpl:\Sbj:\Pst:\Dur/dwell\\
	\trans `Then, in the morning, the people began to scatter all over the place. They were living by themselves.'
\exi{28} 
	\emph{watik, mni fthé ŋarsira ... kar ane bramöwä ŋarsira fof ... thgathg zfrärm ... fath thefath fath}\\
	\gll watik mni fthé ŋa\stem{rsir}a (.) kar ane bramöwä ŋa\stem{rsir}a fof (.) thgathg zf\stem{rä}rm (.) fath thefath fath\\ 
	then fire when \Sg:\Sbj:\Pst:\Ipfv/burn (.) place {\Dem} all \Sg:\Sbj:\Pst:\Ipfv/burn {\Emph} (.) scorched.place \Tsg.\F:\Sbj:\Pst:\Dur/be (.) clear.place burned.place clear.place\\
	\trans `When the fire burned, it burned really the whole area. It became a scorched landscape, a clear place.'
\exi{29} 
	\emph{watik menzmenz ane fof yabun kafar ... thgathg bänemr ane fof zenfara ...}\\
	\gll watik menz-menz ane fof yabun kafar (.) thgathg bänemr ane fof zen\stem{far}a (.)\\ 
	then \Redup-story.man {\Dem} {\Emph} fat big (.) burned.place \Recog.{\Purp} {\Dem} {\Emph} \Sg:\Sbj:\Pst:\Pfv/set.off (.)\\
	\trans `Well, that big, fat creature went to the burned place to get and eat those ones ...'
\exi{30} 
	\emph{kabe mane thfthnm kwosi.}\\
	\gll kabe mane thf\stem{thn}m kwosi\\ 
	people who \Stpl:\Sbj:\Pst:\Dur/lie.down dead\\
	\trans `the people who were lying around dead.'
\exi{31} 
	\emph{watik ... gwamf ŋatha thäsa ... ezi ... ane ... thefath thgathgen fof ... yaser}\\
	\gll watik (.) gwam=f ŋatha thä\stem{s}a (.) ezi (.) ane (.) thefath thgathg=en fof (.) yase=r\\ 
	then (.) gwam=\Erg.{\Sg} dog \Sg:\Sbj>\Stpl:\Obj:\Pst:\Ipfv/call.for (.) morning (.) {\Dem} (.) burned.place scorched.place={\Loc} {\Emph} (.) meat=\Purp \\
	\trans `Well, Gwam called for the dogs for hunting, in the morning at that scorched place.'
\exi{32} 
	\emph{watik ŋatha anenzo fof sathkäfa.}\\
	\gll watik ŋatha ane=nzo fof sa\stem{thkäf}a\\ 
	then dog \Dem={\Only} {\Emph} \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Pfv/start\\
	\trans `Well, he started hunting with that one dog only.'
\exi{33} 
	\emph{ŋatha ane swaruthrm gwam mon nima yarera}\\
	\gll ŋatha ane swa\stem{ru}thrm gwam mon nima ya\stem{rä}ra\\ 
	dog {\Dem} \Sg:\Sbj>\Tsg.\Masc:\Io:\Pst:\Dur/bark gwam how like.this \Sg:\Sbj>\Tsg.\Masc:\Io:\Pst:\Ipfv/do\\
	\trans `The dog was barking at the creature and Gwam noticed it.'
\exi{34} 
	\emph{eda erna kabe kafar yf mane thfrnm ... nafangthrwä gwam ... muri}\\
	\gll eda e\stem{rn}a kabe kafar yf mane thf\stem{rn}m (.) nafa-ngth=r=wä gwam (.) muri\\ 
	two \Stdu:\Sbj:\Pst:\Ipfv/be man big name who \Stdu:\Sbj:\Pst:\Dur/be (.) \Third.\Poss-younger.sibling=\Assoc.\Du={\Emph} gwam (.) muri\\
	\trans `They were two men who had well-known names Gwam with his small brother Muri.'
\exi{35} 
	\emph{gwam yara nafanane ... muri nafangth}\\
	\gll gwam ya\stem{r}a nafa-nane (.) muri nafa-ngth\\ 
	gwam \Tsg.\Masc:\Sbj:\Pst:\Ipfv/be \Third.\Poss-older.sibling (.) muri \Third.\Poss-younger.sibling\\
	\trans `Gwam was his older brother and Muri the younger brother.'
\exi{36} 
	\emph{wati gwamf ane fof ezi ŋatha thäsa thgathgen e}\\
	\gll wati gwam=f ane fof ezi ŋatha thä\stem{s}a thgathg=en e\\ 
	then gwam=\Erg.{\Sg} {\Dem} {\Emph} morning dog \Sg:\Sbj>\Stpl:\Obj:\Pst:\Pfv/call.out burned.place until\\
	\trans `Well, that Gwam was calling out for the dogs in that burned place.'
\exi{37} 
	\emph{anenzo fof ŋatha yayamgwa ... yayamgwa.}\\
	\gll ane=nzo fof ŋatha ya\stem{yamg}wa (.) ya\stem{yamg}wa\\ 
	\Dem={\Only} {\Emph} dog \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Ipfv/shock (.) \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Ipfv/shock\\
	\trans `That creature shocked the dog, it shocked him.'
\newpage 	
\exi{38} 
	\emph{ane menznzo fof kabe maf änatha fof}\\
	\gll ane menz=nzo fof kabe maf ä\stem{na}tha fof\\ 
	{\Dem} story.man={\Only} {\Emph} people who.\Erg.{\Sg} \Sg:\Sbj>\Stpl:\Obj:\Pst:\Ipfv/eat {\Emph}\\
	\trans `that creature which ate the people.'
\exi{39} 
	\emph{fewakaf kwosi thwanathrm}\\
	\gll fewa=kaf kwosi thwa\stem{na}thrm\\ 
	stench={\Prop} dead \Sg:\Sbj>\Stpl:\Obj:\Pst:\Dur/eat\\
	\trans `It was eating the rotten corpses.'
\exi{40} 
	\emph{murif zagr ymarwa fof ... maf yé? gwamf!}\\
	\gll muri=f zagr y\stem{mar}wa fof (.) maf \stem{yé} gwam=f\\ 
	muri=\Erg.{\Sg} far \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Ipfv/see {\Emph} (.) who.\Erg.{\Sg} \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/be gwam=\Erg.\Sg\\
	\trans `Muri was seeing him from a distance ... Who's that? Gwam! (not Muri)'
\exi{41} 
	\emph{``ra bäne yé?'' nima n samara ... o ``ra menzmenz yé?''}\\
	\gll ra bäne \stem{yé} nima n sa\stem{mar}a (.) o ra menz-menz \stem{yé}\\ 
	what \Dem:\Med{} \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/be like.this {\Imn} \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Pfv/see (.) or what \Redup-story.man \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/be\\
	\trans ` ``What is this?'' he was about to see it ``What creature is this?'''
\exi{42} 
	\emph{kabe nrma fi fobo fof ŋagathikwa fof ... ane menzmenz}\\
	\gll kabe nr=ma fi fobo fof ŋa\stem{gathik}wa fof (.) ane menz-menz\\ 
	people stomach={\Char} \Third.{\Abs} \Dist.{\All} {\Emph} \Sg:\Sbj:\Pst:\Ipfv/stop {\Emph} (.) {\Dem} \Redup-story.man\\
	\trans `Because its stomach was full with people, it stopped there ... that creature.'
\exi{43} 
	\emph{kabe ane zenthkäfath yak.}\\
	\gll kabe ane zen\stem{thkäf}ath yak\\ 
	man {\Dem} \Stpl:\Sbj:\Pst:\Ipfv:\Venit/start running\\
	\trans `The people (survivors) started running here.'
\exi{44} 
	\emph{ŋatha mane kwaruthrm tifr ... yf ŋatha yara ane tifr}\\
	\gll ŋatha mane kwa\stem{ru}thrm tifr (.) yf ŋatha ya\stem{r}a ane tifr\\ 
	dog who \Sg:\Sbj:\Pst:\Dur/bark tifr (.) name dog \Tsg.\Masc:\Sbj:\Pst:\Ipfv/be {\Dem} tifr\\
	\trans `As for the barking dog, it was Tifr. The dog's name was Tifr.'
	
\newpage 	
\exi{45} 
	\emph{wati sathkäfath.}\\
	\gll wati sa\stem{thkäf}ath\\ 
	then \Stpl:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Ipfv/start\\
	\trans `Then, they started engaging the creature.'
\exi{46} 
	\emph{kabeyé ane dunzi kma sfruthrmth ... keke}\\
	\gll kabe=yé ane dunzi kma sf\stem{ru}thrmth (.) keke\\ 
	man=\Erg.{\Nsg} {\Dem} arrow {\Pot} \Stpl:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Dur/shoot (.) \Neg\\
	\trans `The people were trying to shoot arrows at the creature without success.'
\exi{47} 
	\emph{gwamf nafangth sräkor ``muri! zba känrit nzuzawe! nzefé biruthro.''}\\
	\gll gwam=f nafa-ngth srä\stem{kor} muri zba kän\stem{rit} nzu-zawe nze=wä b=y\stem{ru}thro\\ 
	gwam=\Erg.{\Sg} \Third.\Poss-younger.sibling \Stsg:\Sbj>\Tsg.\Masc:\Obj:\Irr:\Pfv/speak muri \Prox.{\Abl} \Ssg:\Sbj:\Imp:\Pfv/cross.over \Fsg.\Poss-side \Fsg.\Erg={\Emph} \Med=\Sg:\Sbj>\Tsg.\Masc:\Obj:\Nonpast:\Ipfv:\Andat/shoot\\
	\trans `Gwam said to his small brother: ``Muri! Come over to my side! I will shoot it there.'''
\exi{48} 
	\emph{naf nima ``samg! bänema nä buné fof yruthrth byé keke kwosi yathizr.''}\\
	\gll naf nima sa\stem{mg} bäne=ma {nä bun=é} fof y\stem{ru}thrth b=\stem{yé} keke kwosi ya\stem{thiz}r\\ 
	\Tsg.{\Erg} {\Quot} \Ssg:\Sbj>\Tsg.\Masc:\Obj:\Imp:\Pfv/shoot \Recog={\Char} \Indf=\Erg.{\Nsg} {\Emph} \Stpl:\Sbj>\Tsg.\Masc:\Obj:\Nonpast:\Ipfv/shoot \Med=\Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/be {\Neg} dead \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/die\\
	\trans `He replied: ``Shoot it! Because others are shooting and it is not dying.'''
\exi{49} 
	\emph{naf nima: ``keke fi miyamr erä fofosa mä rä. nze komnzo zimarwé fof.''}\\
	\gll naf nima keke fi miyamr e\stem{rä} fofosa mä \stem{rä} nze komnzo z=y\stem{mar}wé fof\\ 
	\Tsg.{\Erg} {\Quot} {\Neg} \Third.{\Abs} ignorant \Stpl:\Sbj:\Nonpast:\Ipfv/be heart where \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be \Fsg.{\Erg} only \Prox=\Fsg:\Sbj>\Tsg.\Masc:\Obj:\Nonpast:\Ipfv/see {\Emph}\\
	\trans `He replied: ``No, they do not know where the heart is. Only I can see it here.'''
\exi{50} 
	\emph{zirkn thfrnm. nä bun kwanafrm. nä bun kwanafrm.}\\
	\gll zirkn thf\stem{rn}m {nä bun} kwa\stem{na}frm {nä bun} kwa\stem{na}frm\\ 
	persistent \Stdu:\Sbj:\Pst:\Dur/be {\Indf} \Sg:\Sbj:\Pst:\Dur/talk {\Indf} \Sg:\Sbj:\Pst:\Dur/talk\\\
	\trans `They were going back and forth. One was talking and then the other was talking.'
\newpage 	
\exi{51} 
	\emph{watik ``ngth biruthé!''}\\
	\gll watik ngth b=y\stem{ru}thé\\ 
	then younger.sibling \Med=\Fsg:\Sbj>\Tsg.\Masc:\Obj:\Nonpast:\Ipfv/shoot\\
	\trans `Well, (Gwam said) ``Brother, I shoot it now!'''
\exi{52} 
	\emph{``famkaräsü gnräré!'' ... nafananaf ane fof}\\
	\gll fam=karä=sü gn\stem{rä}ré (.) nafa-nana=f ane fof\\ 
	thought=\Prop=\Etc{} \Ssg:\Sbj:\Imp:\Ipfv/be (.) \Third.\Poss-older.brother=\Erg.{\Sg} {\Dem} {\Emph}\\
	\trans `(Muri said) ``You must watch out!'' His big brother (was the one who shot).'
\exi{53} 
	\emph{trikasi nima rä}\\
	\gll trik-si nima \stem{rä}\\ 
	tell-{\Nmlz} like.this \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be\\
	\trans `The story is like this:'
\exi{54} 
	\emph{nafangth kma markai näbikarä sfrärm}\\
	\gll nafa-ngth kma markai näbi=karä sf\stem{rä}rm\\ 
	\Third.\Poss-younger.sibling {\Pot} white.man bow={\Prop} \Tsg.\Masc:\Sbj:\Pst:\Dur/be\\
	\trans `His small brother must have had a shotgun.'
\exi{55} 
	\emph{watik nafangth mane yara naf keke samga ... nafananafnzo}\\
	\gll watik nafa-ngth mane ya\stem{r}a naf keke sa\stem{mg}a (.) nafa-nana=f=nzo\\ 
	then \Third.\Poss-younger.sibling who \Tsg.\Masc:\Sbj:\Pst:\Ipfv/be \Tsg.{\Erg} {\Neg} \Sg:\Sbj>\Tsg.\Masc:\Pst:\Pfv/shoot (.) \Third.\Poss-older.brother=\Erg.\Sg=\Only\\
	\trans `But his small brother did not hit the creature, only his older brother.'
\exi{56} 
	\emph{näbi ŋathunza ... zf sfthnm}\\
	\gll näbi ŋa\stem{thu}nza (.) zf sf\stem{thn}m\\ 
	bow \Sg:\Sbj:\Pst:\Ipfv/fold (.) {\Imm} \Tsg.\Masc:\Sbj:\Pst:\Dur/lie.down\\
	\trans `He drew the bow while the creature was laying down here,'
\exi{57} 
	\emph{yo kwan ... fof sargosira fofosa fefen}\\
	\gll {yo.kwan} (.) fof sa\stem{rgosi}ra fofosa ffe=n\\ 
	sound.of.arrow (.) {\Emph} \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Ipfv/penetrate heart real={\Loc}\\
	\trans `Wham! The arrow poked right through to the heart.'
\exi{58} 
	\emph{no fof zärfetha}\\
	\gll no fof zä\stem{rfeth}a\\ 
	water {\Emph} \Sg:\Sbj:\Pst:\Pfv/burst\\
	\trans `Water bursted out.'
\exi{59} 
	\emph{no ane zamatha}\\
	\gll no ane za\stem{math}a\\ 
	water {\Dem} \Sg:\Sbj:\Pst:\Pfv/run\\
	\trans `That water was gushing out.'
\newpage 	
\exi{60} 
	\emph{wati no mane kwakwirm fof}\\
	\gll wati no mane kwa\stem{kwir}m fof\\ 
	then water which \Sg:\Sbj:\Pst:\Dur/run {\Emph}\\
	\trans `Well, the water that was flowing'
\exi{61} 
	\emph{wäsi zrminz mä ŋanrsira fof ... mni mä ŋanrsira}\\
	\gll wäsi zrminz mä ŋan\stem{rsir}a fof (.) mni mä ŋan\stem{rsir}a\\ 
	wäsi root where \Sg:\Sbj:\Pst:\Ipfv:\Venit/burn {\Emph} (.) fire where \Sg:\Sbj:\Pst:\Ipfv:\Venit/burn\\
	\trans `to the place where the wäsi roots had burned, where the fire had burned.'
\exi{62} 
	\emph{no fä kwanthorthrm fof ... ane zrminz fof}\\
	\gll no fä kwan\stem{thor}thrm fof (.) ane zrminz fof\\ 
	water {\Dist} \Sg:\Sbj:\Pst:\Dur:\Venit/enter {\Emph} (.) {\Dem} root {\Emph}\\
	\trans `The water went inside there ... into those roots.'
\exi{63} 
	\emph{nof nä nima thärkothmako. nä nima thänkothma nzezawe.}\\
	\gll no=f nä nima thä\stem{kothm}ako nä nima thän\stem{kothm}a nze-zawe\\ 
	water={\Erg} {\Indf} like.this \Sg:\Sbj>\Stpl:\Obj:\Pst:\Pfv:\Andat/chase {\Indf} like.this \Sg:\Sbj>\Stpl:\Obj:\Pst:\Pfv:\Venit/chase \Fnsg.\Poss-side\\
	\trans `The water chased some people that way and it chased others here to our side.'
\exi{64} 
	\emph{gwamane nima zenmathath ... muriane nima.}\\
	\gll gwam=ane nima zen\stem{math}ath (.) muri=ane nima\\ 
	gwam=\Poss.{\Sg} like.this \Stpl:\Sbj:\Pst:\Pfv:\Venit/run (.) muri=\Poss.{\Sg} like.this\\
	\trans `Gwam's people ran this way. Muri's people ran that way (towards Australia).'
\exi{65} 
	\emph{mane ŋankwirwath zentnäthath}\\
	\gll mane ŋankwirwath zentnäthath\\ 
	who \Stpl:\Sbj:\Pst:\Ipfv:\Venit/run \Stpl:\Sbj:\Pst:\Pfv:\Venit/scatter\\
	\trans `Those who came running started to go in different directions.'
\exi{66} 
	\emph{nä enrera bawi.}\\
	\gll nä en\stem{rä}ra bawi\\ 
	{\Indf} \Sg:\Sbj:\Pst:\Ipfv:\Venit/be bawi\\
	\trans `Some came to Bawi.'
\exi{67} 
	\emph{wartha nima bämnzr wartha a kondomarin ... smärki.}\\
	\gll wartha nima b=ä\stem{m}nzr wartha a kondomarin (.) smärki\\ 
	wartha like.this \Med=\Stpl:\Sbj:\Nonpast:\Ipfv/dwell wartha and marind (.) smärki\\
	\trans `Like the Wartha people living there. The Wartha and Marind ... and the Smärki.'
	\newpage
\exi{68} 
	\emph{nafanme ... foba fof ŋankwira fof}\\
	\gll nafanme (.) foba fof ŋan\stem{kwi}ra fof\\ 
	\Tnsg.{\Poss} (.) \Dist.{\Abl} {\Emph} \Sg:\Sbj:\Pst:\Ipfv:\Venit/run {\Emph}\\
	\trans `Their ancestor was coming really from there.'
\exi{69} 
	\emph{fi foba fof ŋankwirwath ... bawi}\\
	\gll fi foba fof ŋan\stem{kwi}rwath (.) bawi\\ 
	\Third.{\Abs} \Dist.{\Abl} {\Emph} \Stpl:\Sbj:\Pst:\Ipfv:\Venit/run (.) bawi\\
	\trans `They came from Bawi.'
\exi{70} 
	\emph{watik gwamf fä fof mni ... bäne zafrafa fof no.}\\
	\gll watik gwam=f fä fof mni (.) bäne za\stem{fraf}a fof no\\ 
	then gwam=\Erg.{\Sg} {\Dist} {\Emph} fire (.) \Recog.{\Abs} \Sg:\Sbj>\Tsg.\F:\Obj:\Pst:\Pfv/extinguish {\Emph} water\\
	\trans `Okay, Gwam extinguished the fire there ... I mean the water.'
\exi{71} 
	\emph{dödöme zakwra.}\\
	\gll dödö=me za\stem{kwr}a\\ 
	dödö={\Ins} \Sg:\Sbj>\Tsg.\F:\Obj:\Pst:\Pfv/hit\\
	\trans `He hit the water with the dödö plant (Melaleuca sp).'
\exi{72} 
	\emph{watik no fä fof zäkora ... keke kwa nof zanmäyofa.}\\
	\gll watik no fä fof zä\stem{kor}a (.) keke kwa no=f zan\stem{mäyof}a\\ 
	then water {\Dist} {\Emph} \Sg:\Sbj:\Pst:\Ipfv/become (.) {\Neg} {\Fut} water={\Erg} \Sg:\Sbj>\Tsg.\F:\Obj:\Pst:\Pfv:\Venit/continue\\\
	\trans `Okay, the water stopped there. It did not continue to come our way.'
\exi{73} 
	\emph{fobo fof no ŋagathikwa fof.}\\
	\gll fobo fof no ŋa\stem{gathik}wa fof\\ 
	\Dist.{\All} {\Emph} water \Sg:\Sbj:\Pst:\Ipfv/stop \Emph\\
	\trans `The flood stopped there.'
\exi{74} 
	\emph{watik fi mane enrera e ... zwari ... wartha fof.}\\
	\gll watik fi mane en\stem{rä}ra e (.) zwari (.) wartha fof\\ 
	then \Third.{\Abs} who \Sg:\Sbj:\Pst:\Ipfv:\Venit/be until (.) zwari (.) wartha {\Emph}\\
	\trans `Those who came until Zwari (= Bawi) were really the Wartha people.'
\exi{75} 
	\emph{watik fä fof zwarin zämsath.}\\
	\gll watik fä fof zwari=n zä\stem{ms}ath\\ 
	then {\Dist} {\Emph} zwari={\Loc} \Stpl:\Sbj:\Pst:\Pfv/dwell\\
	\trans `They settled there in Zwari.'
\exi{76} 
	\emph{zokwasi fthé emarwath ffrümenzo ... watik kondomarin nima feräro.}\\
	\gll zokwasi fthé e\stem{mar}wath f-frü=me=nzo (.) watik kondomarin nima f=e\stem{rä}ro\\ 
	language when \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Ipfv/see \Redup-single=\Ins={\Only} (.) then marind like.this \Dist=\Stpl:\Sbj:\Pst:\Ipfv:\Andat/be\\
	\trans `When they saw that people spoke different languages, then the Marind moved on that way.' 	
\exi{77} 
	\emph{zena boba wazi fi berä merauken.}\\
	\gll zena boba wazi fi b=e\stem{rä} merauke=n\\ 
	today \Med.{\Abl} side \Third.{\Abs} \Med=\Stpl:\Sbj:\Nonpast:\Ipfv/be merauke=\Loc\\
	\trans `Today, they are on the other side, there in Merauke.'
\exi{78} 
	\emph{nä mane erera zwarifa ŋafrezath thoro.}\\
	\gll nä mane e\stem{rä}ra zwari=fa ŋa\stem{frez}ath thoro\\ 
	{\Indf} who \Stpl:\Sbj:\Pst:\Ipfv/be zwari={\Abl} \Stpl:\Sbj:\Pst:\Ipfv/come.up.from.river thoro\\
	\trans `As for others, they came up from Zwari to Thoro.'
\exi{79} 
	\emph{watik thoron fä fthé zemarath we nimäwä fof ... zokwasi ffrümenzo.}\\
	\gll watik thoro=n fä fthé ze\stem{mar}ath we nima=wä fof (.) zokwasi f-frü=me=nzo\\ 
	then thoro={\Loc} {\Dist} when \Stpl:\Sbj:\Pst:\Ipfv/see also like.this={\Emph} {\Emph} (.) language \Redup-single=\Ins=\Only\\
	\trans `When they looked at themselves in Thoro, it was the same thing again ... different languages.'
\exi{80} 
	\emph{watik foba zethkäfath nimame kwasogwrmth.}\\
	\gll watik foba ze\stem{thkäf}ath nima=me kwa\stem{sog}wrmth\\ 
	then \Dist.{\Abl} \Stpl:\Sbj:\Pst:\Ipfv/start like.this \Stpl:\Sbj:\Pst:\Dur/climb\\
	\trans `Then they began walking from there. They came up this way.'
\exi{81} 
	\emph{okay, nä mane enrera bäne ... zwari ... zwarifa e bäne ... tamgakar.}\\
	\gll okay nä mane en\stem{rä}ra bäne (.) zwari (.) zwari=fa e bäne . tamgakar\\ 
	okay {\Indf} who \Stpl:\Sbj:\Pst:\Ipfv:\Venit/be \Recog.{\Abs} (.) zwari (.) zwari={\Abl} until \Recog.{\Abs} (.) tamgakar\\
	\trans `Okay, some came until Zwari. From Zwari until Tamgakar.'
\exi{82} 
	\emph{nima bä ämnzr safs}\\
	\gll nima bä ä\stem{m}nzr safs\\ 
	like.this \Med{} \Stpl:\Sbj:\Nonpast:\Ipfv/dwell safs\\
	\trans `like the ones who live there in Safs.'
\exi{83} 
	\emph{wati fi fä fof thfyakm.}\\
	\gll wati fi fä fof thf\stem{yak}m\\ 
	then \Third.{\Abs} {\Dist} {\Emph} \Stpl:\Sbj:\Pst:\Dur/walk\\
	\trans `Okay, this is how they were going.'
\newpage 
\exi{84} 
	\emph{nzenme mane yanra ... mä ŋankwirwath komo fä ŋanfrezath ... komo.}\\
	\gll nzenme mane yan\stem{r}a (.) mä ŋan\stem{kwir}wath komo fä ŋan\stem{frez}ath (.) komo\\ 
	\Fnsg./Poss{} who \Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Venit/be (.) where \Stpl:\Sbj:\Pst:\Ipfv:\Venit/run komo {\Dist} \Stpl:\Sbj:\Pst:\Ipfv:\Venit/come.up.from.water (.) komo\\\
	\trans `As for our ancestor, he was running to Komo. He came up there in Komo.'
\exi{85} 
	\emph{nzenme mayawama kabe nä fä thägathizath.}\\
	\gll nzenme mayawa=ma kabe nä fä thä\stem{gathiz}ath\\ 
	\Fnsg.{\Poss} mayawa={\Char} man {\Indf} {\Dist} \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Pfv/leave\\
	\trans `Our Mayawa man left some people there.'
\exi{86} 
	\emph{we foba ... thden nä kwot we mayawama kabe fof.}\\
	\gll we foba (.) thd=en nä kwot we mayawa=ma kabe fof\\ 
	also \Dist.{\Abl} (.) middle={\Loc} {\Indf} properly also mayawa={\Char} people \Emph\\
	\trans `and again ... halfway he left some more Mayawa people again.'
\exi{87} 
	\emph{foba ... baguma kabe ... zena mifnen zämnzr.}\\
	\gll foba (.) bagu=ma kabe (.) zena mifne=n z=ä\stem{m}nzr\\ 
	\Dist.{\Abl} (.) bagu={\Char} people (.) today mibini={\Loc} \Prox=\Stpl:\Sbj:\Nonpast:\Ipfv/dwell\\
	\trans `He left some Bagu people there. They live in Mibini today.'
\exi{88} 
	\emph{sagara fä thägathinzath. okay fi nima erera ... mogarkam.}\\
	\gll sagara fä thä\stem{gathinz}ath okay fi nima e\stem{rä}ra (.) mogarkam\\ 
	sagara {\Dist} \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Ipfv/leave okay \Third.{\Abs} like.this \Stpl:\Sbj:\Pst:\Ipfv/be (.) mogarkam\\
	\trans `They left some Sagara people there. They used to live in Mogarkam.'
\exi{89} 
	\emph{nä mane erera nima erera bäne ... drdr ... nä sagara fof.}\\
	\gll nä mane e\stem{rä}ra nima e\stem{rä}ra bäne (.) drdr (.) nä sagara fof\\ 
	{\Indf} who \Stpl:\Sbj:\Pst:\Ipfv/be like.this \Stpl:\Sbj:\Pst:\Ipfv/be \Recog.{\Abs} (.) derideri (.) {\Indf} sagara \Emph\\
	\trans `Others were there in Derideri ... another Sagara clan.'
\exi{90} 
	\emph{bagu mane enrera bäne ... mäta.}\\
	\gll bagu mane en\stem{rä}ra bäne (.) mäta\\ 
	bagu who \Stpl:\Sbj:\Pst:\Ipfv:\Venit/be \Recog.{\Abs} (.) mäta\\
	\trans `The Bagus who were coming, they went to Mäta.'
\newpage 
\exi{91} 
	\emph{sagara mane enrera garaita.}\\
	\gll sagara mane en\stem{rä}ra garaita\\ 
	sagara who \Stpl:\Sbj:\Pst:\Ipfv:\Venit/be garaita\\
	\trans `and the Sagaras continued to Garaita.'
\exi{92} 
	\emph{mayawa ni zbo zf nnrera.}\\
	\gll mayawa ni zbo zf nn\stem{rä}ra\\ 
	mayawa {\Fnsg} \Prox.{\All} {\Imm} \Fpl:\Sbj:\Pst:\Ipfv:\Venit/be\\
	\trans `We Mayawas came right here.'
\exi{93} 
	\emph{okay nzenme bada ... mrzarane bada mane yanra ... fi fof yanra bäne ... mathkwi.}\\
	\gll okay nzenme bada (.) mrzar=ane bada mane yan\stem{r}a (.) fi fof yan\stem{r}a bäne (.) mathkwi\\ 
	okay \Fnsg.{\Poss} ancestor (.) mrzar=\Poss.{\Sg} ancestor who \Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Venit/be (.) \Third.{\Abs} {\Emph} \Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Venit/be \Recog.{\Abs} (.) mathkwi\\
	\trans `Okay, our ancestor, the Mrzar clan's ancestor who came was Mathkwi.'
\exi{94} 
	\emph{yf ane yanra mathkwi!}\\
	\gll yf ane yan\stem{r}a mathkwi\\ 
	name {\Dem} \Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Venit/be mathkwi\\
	\trans `That was the name, Mathkwi!'
\exi{95} 
	\emph{mathkwif ane enfathwa ... wawa fofosa.}\\
	\gll mathkwi=f ane en\stem{fath}wa (.) wawa fofosa\\ 
	mathkwi=\Erg.{\Sg} {\Dem} \Sg:\Sbj>\Stpl:\Obj:\Pst:\Ipfv:\Venit/hold (.) yam heart\\
	\trans `Mathkwi was holding those magic yam stones.'
\exi{96} 
	\emph{naf ane ynfathwa fof.}\\
	\gll naf ane yn\stem{fath}wa fof.\\ 
	\Tsg.{\Erg} {\Dem} \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Ipfv:\Venit/hold \Emph\\
	\trans `He was holding that one.'
\exi{97} 
	\emph{wati näbi ane komnzo fofosa yara wawama ... nasi ... duga ... biskar ... dagon nä berä fof}\\
	\gll wati näbi ane komnzo fofosa ya\stem{r}a wawa=ma (.) nasi (.) duga (.) biskar (.) dagon nä b=e\stem{rä} fof\\ 
	then one {\Dem} only heart \Tsg.\Masc:\Sbj:\Pst:\Ipfv/be yam={\Char} (.) long.yam (.) taro (.) cassava (.) food {\Indf} \Med=\Stpl:\Sbj:\Nonpast:\Ipfv/be {\Emph}\\
	\trans `Okay, there was just one stone for yams, long yams, taro, cassava and the other food there.'
\exi{99} 
	\emph{watik fi anekarä fof yanra fof.}\\
	\gll watik fi ane=karä fof yan\stem{r}a fof\\ 
	then \Third.{\Abs} \Dem={\Prop} {\Emph} \Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Venit/be \Emph\\
	\trans `He came with this one.'
\exi{100} 
	\emph{mane yanyaka e ... wm bä ythn ... zabrta.}\\
	\gll mane yan\stem{yak}a e (.) wm bä y\stem{thn} (.) zabrta\\ 
	who \Tsg.\Masc:\Sbj:\Pst:\Ipfv/walk until (.) stone \Med{} \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/lie.down (.) zabrta\\
	\trans `As he came to the place where the stone is lying, there at Zabrta,'
\exi{101} 
	\emph{fä fof ŋanritakwath fof.}\\
	\gll fä fof ŋan\stem{ritak}wath fof\\ 
	{\Dist} {\Emph} \Stpl:\Sbj:\Pst:\Ipfv:\Venit/cross {\Emph}\\\
	\trans `they crossed the river.'
\exi{102} 
	\emph{kwanritakwrmth trkren.}\\
	\gll kwan\stem{ritak}wrmth trkr=en\\ 
	\Stpl:\Sbj:\Pst:\Dur:\Venit/cross flood=\Loc\\
	\trans `They were going across during the rainy season.'
\exi{103} 
	\emph{watik, nima n fam zära ``garaita zawe? keke, nä kabe foba z sfyak.''}\\
	\gll watik nima n fam zä\stem{r}a garaita zawe keke nä kabe foba z sf\stem{yak}\\ 
	then {\Quot} {\Imn} though \Sg:\Sbj:\Pst:\Ipfv/do garaita side {\Neg} {\Indf} man \Dist.{\Abl} {\Iam} \Tsg.\Masc:\Sbj:\Pst:\Ipfv/walk\\
	\trans `He was thinking: ``Should I go to Garaita? No, another man went this way already.'''
\exi{104} 
	\emph{watik, nima zethkäfa fi ... safs.}\\
	\gll watik nima ze\stem{thkäf}a fi (.) safs\\ 
	then like.this \Sg:\Sbj:\Pst:\Pfv/start \Third.{\Abs} (.) safs\\
	\trans `Then, he started coming this way towards Safs.'
\exi{105} 
	\emph{nimame ane zethkäfa mothr mane yanra e ... akrimogo.}\\
	\gll nima=me ane ze\stem{thkäf}a moth=r mane yan\stem{r}a e (.) akrimogo\\ 
	like.this={\Ins} {\Dem} \Sg:\Sbj:\Pst:\Pfv/start walking={\Purp} who \Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Venit/be until (.) akrimogo\\
	\trans `He started to walk like this, he walked until Akrimogo.'
\exi{106} 
	\emph{yam fä fof thremar fof.}\\
	\gll yam fä fof thre\stem{mar} fof\\ 
	footprint {\Dist} {\Emph} \Stsg:\Sbj>\Stpl:\Obj:\Irr:\Pfv/see {\Emph}\\
	\trans `He saw footprints there.'
\exi{107} 
	\emph{``oh, nä nima z eräro.''}\\
	\gll oh nä nima z e\stem{rä}ro\\ 
	oh {\Indf} like.this {\Iam} \Stpl:\Sbj:\Nonpast:\Ipfv:\Andat/be\\
	\trans `He said ``Oh, others were walking along here already.'''
  \newpage 
\exi{108} 
	\emph{watik, keräfi foba fof zäzira fof e ... kar yf rä ymnz.}\\
	\gll watik keräfi foba fof zä\stem{zi}ra fof e (.) kar yf \stem{rä} ymnz\\ 
	then blackpalm \Dist.{\Abl} {\Emph} \Sg:\Sbj:\Pst:\Pfv/throw {\Emph} until (.) place name \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be ymnz\\
	\trans `Then he shot an arrow from there until ... the name of the place is Ymnz.'
\exi{109} 
	\emph{watik, fobo fof ``oh, kabe bä yé ... watik, nimame wyak.''}\\
	\gll watik fobo fof oh kabe bä \stem{yé} (.) watik nima=me w\stem{yak}\\ 
	then \Dist.{\All} {\Emph} oh man \Med{} \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/be (.) then like.this={\Ins} \Fsg:\Sbj:\Nonpast:\Ipfv/walk\\
	\trans `Okay, there he said: ``Oh, there is a man there! Okay, I will go that path then.'''
\exi{110} 
	\emph{watik foba fof akrimogofa zenfara fof.}\\
	\gll watik foba fof akrimogo=fa zen\stem{far}a fof\\ 
	then \Dist.{\Abl} {\Emph} akrimogo={\Abl} \Sg:\Sbj:\Pst:\Pfv/set.off {\Emph}\\
	\trans `Okay, he set off from there, from Akrimogo.'
\exi{111} 
	\emph{akrimogo ... foba näbi yanyaka. karane yf rä füsari.}\\
	\gll akrimogo (.) foba näbi yan\stem{yak} kar=ane yf \stem{rä} füsari\\ 
	akrimogo (.) \Dist.{\Abl} one \Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Venit/walk place=\Poss.{\Sg} name \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be füsäri\\
	\trans `From Akrimogo he was coming straight to ... the name of that place is Füsari.'
\exi{112} 
	\emph{füsärifa ... rarafü kar ... rarafü karfa ... kafrir fä ttfön zänrita e ... bäne ... zofok.}\\
	\gll füsäri=fa (.) rarafü kar (.) rarafü kar=fa (.) kafrir fä ttfö=n zän\stem{rit}a e (.) bäne (.) zofok\\ 
	füsäri={\Abl} (.) rarafü place (.) rarafü place={\Abl} (.) kafrir {\Dist} creek={\Loc} \Sg:\Sbj:\Pst:\Pfv:\Venit/cross until (.) \Recog.{\Abs} (.) zofok\\
	\trans `From Füsari to Rarafü, from Rarafü to Kafrir, he crossed the creek, until he arrived at Zofok.'
\exi{113} 
	\emph{zofok fä yamthiza.}\\
	\gll zofok fä ya\stem{mthi}za\\ 
	zofok {\Dist} \Tsg.\Masc:\Sbj:\Pst:\Ipfv/rest\\
	\trans `He rested there in Zofok.'
\exi{114} 
	\emph{nabi komnzo bekogr.}\\
	\gll nabi komnzo b=e\stem{kogr}\\ 
	bamboo still \Med=\Stpl:\Sbj:\Nonpast:\Stat/be.standing \\\
	\trans `The bamboos are still standing there.'
\exi{115} 
	\emph{nabi ŋatr fä fof zurärm zwafrmnzrm.}\\
	\gll nabi ŋatr fä fof zu\stem{rä}rm zwa\stem{frm}nzrm\\ 
	bow bowstring {\Dist} {\Emph} \Sg:\Sbj>\Tsg.\F:\Obj:\Pst:\Dur/do \Sg:\Sbj>\Tsg.\F:\Obj:\Pst:\Dur/prepare\\
	\trans `He was fixing his bowstring there.'
\exi{116} 
	\emph{zurzirakwa fof.}\\
	\gll zu\stem{rzirak}wa fof\\ 
	\Sg:\Sbj>\Tsg.\F:\Obj:\Pst:\Ipfv/tie \Emph\\
	\trans `He tied it properly.'
\exi{117} 
	\emph{bäne yanatha ... nasi nömä. nasi nömä yanatha.}\\
	\gll bäne ya\stem{na}tha (.) nasi nömä nasi nömä ya\stem{na}tha\\ 
	\Recog.{\Abs} \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Ipfv/eat (.) long.yam yamcake long.yam yamcake \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Ipfv/eat\\
	\trans `He ate that yamcake (from long yams). He ate the yamcake.'
\exi{118} 
	\emph{rfarrfar futhfuth mane erera ... watik, wmr ane fof ŋakwthenzath fof.}\\
	\gll rfar-rfar futh-futh mane e\stem{rä}ra (.) watik wm=r ane fof ŋa\stem{kwthe}nzath fof\\ 
	\Redup-crumb{} \Redup-scraps which \Sg:\Sbj:\Pst:\Ipfv/be (.) then stone={\Purp} {\Dem} {\Emph} \Stpl:\Sbj:\Pst:\Ipfv/change {\Emph}\\
	\trans `As for those crumbs, those scraps, they turned into stones.'
\exi{119} 
	\emph{zäkwtherath ... watik komnzo berästhgr.}\\
	\gll zä\stem{kwther}ath (.) watik komnzo b=e\stem{räs}thgr\\ 
	\Stpl:\Sbj:\Pst:\Pfv/change (.) then still \Med=\Stpl:\Sbj:\Nonpast:\Stat/be.erected\\
	\trans `They changed into stones. They are still sticking out there.'
\exi{120} 
	\emph{wm mane yé ... ynfathwa fof no nzigfu ... watik ane fof yräza fof ... zofok kar.}\\
	\gll wm mane \stem{yé} (.) yn\stem{fath}wa fof no nzigfu (.) watik ane fof y\stem{rä}za fof (.) zofok kar\\ 
	stone which \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/be (.) \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Ipfv:\Venit/hold {\Emph} water magic.stone (.) then {\Dem} {\Emph} \Sg:\Sbj>\Tsg.\Masc:\Obj:\Nonpast:\Ipfv/erect {\Emph} (.) zofok place\\
	\trans `As for the stone that he held, the rain magic stone, he pushed it in the ground at Zofok.'
\exi{121} 
	\emph{watik foba yanyaka bäne ... misa zfth. mäbri misa zfth yrn.}\\
	\gll watik foba yan\stem{yak}a bäne (.) {misa.zfth} mäbri {misa.zfth} yrn\\ 
	then \Dist.{\Abl} \Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Venit/walk \Recog.{\Abs} (.) {misa.zfth} mäbri {misa.zfth} yrn\\
	\trans `Then he went to Misa Zfth from there. Mäbri, Misa Zfth and Yrn.'
\newpage 
\exi{122} 
	\emph{fä zänrsöftha fof ... yanyaka benzü zfth.}\\
	\gll fä zän\stem{rsöfth}a fof (.) yan\stem{yak}a {benzü.zfth}\\ 
	{\Dist} \Sg:\Sbj:\Pst:\Pfv:\Venit/descend {\Emph} (.) \Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Venit/walk {benzü.zfth}\\
	\trans `There he came down and walked to Benzü Zfth.'
\exi{123} 
	\emph{foba fof ymd threnkaris fof ... afa kfokfo ythama}\\
	\gll foba fof ymd thren\stem{karis} fof (.) {afa.kfokfo} ythama\\ 
	\Dist.{\Abl} {\Emph} bird \Stsg:\Sbj>\Stpl:\Obj:\Irr:\Pfv:\Venit/hear {\Emph} (.) {afa kfokfo} ythama\\
	\trans `From there he heard those birds ... \emph{afa kfokfo} and \emph{ythama}.'\footnote{These are: \emph{afa kfokfo} `Hooded Butcherbird' and \emph{ythama} `Raggiana Bird-of-Paradise'.}
\exi{124} 
	\emph{fam zära ``kar bä rä. ah, kar töna fobo fof wyak fof.''}\\
	\gll fam zä\stem{r}a kar bä \stem{rä} ah kar töna fobo fof wo\stem{yak} fof\\ 
	thought \Sg:\Sbj:\Pst:\Pfv/do place \Med{} \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be ah place high.ground \Dist.{\All} {\Emph} \Fsg:\Sbj:\Nonpast:\Ipfv/walk \Emph\\
	\trans `He thought ``There is a place there. Ah, I will go there to the high ground.'''
\exi{125} 
	\emph{yanyaka fä fof zänrita fof rä kukwrb fr zra ... mnzär fr neba.}\\
	\gll yan\stem{yak}a fä fof zän\stem{rit}a fof \stem{rä} {kukwrb fr} zra (.) {mnzär fr} neba\\ 
	\Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Venit/walk {\Dist} {\Emph} \Sg:\Sbj:\Pst:\Pfv:\Venit/cross.over {\Emph} \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be {kukwrb fr} swamp (.) {mnzär fr} opposite\\
	\trans `He walked and crossed the river at Kukwrb Fr swamp opposite from Mnzär Fr.'
\exi{126} 
	\emph{wati fä fof yanyaka fof ... mä swanyakm ... mä zänfrefa ... nömä futhfuth ... fä fof ... ŋantnägwath.}\\
	\gll wati fä fof yan\stem{yak}a fof (.) mä swan\stem{yak}m (.) mä zän\stem{fref}a (.) nömä futh-futh (.) fä fof (.) ŋan\stem{tnäg}wath\\ 
	then {\Dist} {\Emph} \Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Venit/walk {\Emph} (.) where \Tsg.\Masc:\Sbj:\Pst:\Dur:\Venit/walk (.) where \Sg:\Sbj:\Pst:\Pfv:\Venit/come.up.from.river (.) yamcake \Redup-scraps (.) {\Dist} {\Emph} (.) \Stpl:\Sbj:\Pst:\Ipfv:\Venit/lose\\\
	\trans `Well, he walked there. and where he was walking, where he came up, he dropped those crumbs from the yamcake.'
\exi{127} 
	\emph{mane yanra e zrä zöfäthak bä brä brä ... zafe ŋazi fr ... nä fof ethn berä}\\
	\gll mane yan\stem{r}a e z=\stem{rä} zöfäthak bä b=\stem{rä} b=\stem{rä} (.) zafe ŋazi fr (.) nä fof e\stem{thn} b=e\stem{rä}\\ 
	who \Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Venit/be until \Prox=\Tsg.\F:\Sbj:\Nonpast:\Ipfv/be zöfäthak \Med{} \Med=\Tsg.\F:\Sbj:\Nonpast:\Ipfv/be \Med=\Tsg.\F:\Sbj:\Nonpast:\Ipfv/be (.) old coconut stem (.) {\Indf} {\Emph} \Stpl:\Sbj:\Nonpast:\Stat/be.lying \Med=\Stpl:\Sbj:\Nonpast:\Ipfv/be\\
	\trans `He walked up right here Zöfäthak, over there by the old coconut trees. There are some stones lying down there.'
\exi{128} 
	\emph{watik nä fä fof ŋantnägwath fof}\\
	\gll watik nä fä fof ŋan\stem{tnäg}wath fof\\ 
	then {\Indf} {\Dist} {\Emph} \Stpl:\Sbj:\Pst:\Ipfv:\Venit/lose {\Emph}\\
	\trans `Well, some more crumbs were dropped there.'
\exi{129} 
	\emph{fä fof sakuka ``oh, zane zf zunthorakwé.''}\\
	\gll fä fof sa\stem{kuk}a oh zane zf zun\stem{thorak}wé\\ 
	{\Dist} {\Emph} \Tsg.\Masc:\Sbj:\Pst:\Pfv/stand oh \Dem:{\Prox} {\Imm} \Fsg:\Sbj>\Tsg.\F:\Obj:\Rpst:\Ipfv:\Venit/search\\
	\trans `There he stood and said ``Oh, this is what I was looking for.'''
\exi{130} 
	\emph{fz zamara afa kfokfo zakarisa ... bäne zakarisa ythama}\\
	\gll fz za\stem{mar}a {afa.kfokfo} za\stem{karis}a (.) bäne za\stem{karis}a ythama\\ 
	forest \Sg:\Sbj>\Tsg.\F:\Obj:\Pst:\Pfv/see butcherbird \Sg:\Sbj:\Pst:\Pfv/hear (.) \Recog.{\Abs} \Sg:\Sbj:\Pst:\Pfv/hear bird.of.paradise\\
	\trans `He saw the forest, he heard the butcherbird and the bird of paradise.'
\exi{131} 
	\emph{watik krenafth ``nima wyak. zbo kar rä farem kar.''}\\
	\gll watik kre\stem{nafth} nima wo\stem{yak} zbo kar \stem{rä} farem kar\\ 
	then \Stsg:\Sbj:\Irr:\Pfv/speak like.this \Fsg:\Sbj:\Nonpast:\Ipfv/walk \Prox.{\All} place \Tsg.\F:\Sbj:\Nonpast:\Ipfv/be farem place\\
	\trans `Then he said ``I will go this way. There is a place here, Faremkar.'''
\exi{132} 
	\emph{watik, fthé yaka bobo, foba krekaris ``oh, füthan nä zbo kabe yamnzr.''}\\
	\gll watik fthé \stem{yak}a bobo foba kre\stem{karis} oh fütha=n nä zbo kabe ya\stem{m}nzr\\ 
	then when \Tsg.\Masc:\Sbj:\Pst:\Ipfv/walk \Med.{\All} \Dist.{\Abl} \Stsg:\Sbj:\Irr:\Pfv/hear oh fütha={\Loc} {\Indf} \Prox.{\All} man \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/dwell\\
	\trans `When he walked there, he heard someone from over there ``Oh, a man lives here in Fütha.'''
\exi{133} 
	\emph{we foba krekaris ``oh, faremkaren kabe yé.''}\\
	\gll we foba kre\stem{karis} oh faremkar=en kabe \stem{yé}\\ 
	also \Dist.{\Abl} \Stsg:\Sbj:\Irr:\Pfv/hear oh faremkar={\Loc} man \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv/be\\
	\trans `He also heard someone from over there ``Oh, a man lives here in Faremkar.'''
\exi{134} 
	\emph{watik yako.}\\
	\gll watik \stem{yak}o\\ 
	then \Tsg.\Masc:\Sbj:\Nonpast:\Ipfv:\Andat/walk\\
	\trans `Then he walked away.'
\exi{135} 
	\emph{faremaneme kabe z sathora.}\\
	\gll farem=aneme kabe z sa\stem{thor}a\\ 
	farem=\Poss.{\Nsg} man {\Iam} \Tsg.\Masc:\Sbj:\Pst:\Pfv/arrive\\
	\trans `The Farem clan's man had already arrived,'
\exi{136} 
	\emph{bafane bada fof ... fatamaane.}\\
	\gll baf=ane bada fof (.) fatama=ane\\ 
	\Recog=\Poss.{\Sg} ancestor {\Emph} (.) fatama=\Poss.{\Sg}\\
	\trans `that one's ancestor, Fatama's ancestor.'
\exi{137} 
	\emph{farem thden watik foba fof sräkor ``foba fof bä fä fof gnamnzé! ey, fisor bthanen käms!''}\\
	\gll farem thd=en watik foba fof srä\stem{kor} foba fof bä fä fof gna\stem{m}nzé ey {fisor bthan=en} kä\stem{m}s\\ 
	farem middle={\Loc} then \Dist.{\Abl} {\Emph} \Stsg:\Sbj>\Tsg.\Masc:\Obj:\Irr:\Pfv/speak \Dist.{\Abl} {\Emph} \Second.{\Abs} {\Dist} {\Emph} \Ssg:\Sbj:\Imp:\Stat/dwell hey {fisor bthan={\Loc}} \Ssg:\Sbj:\Imp:\Pfv/dwell\\
	\trans `In the middle of Farem, he told him from there: ``You stay right there! Hey, you settle at Fisor Bthan!'''
\exi{138} 
	\emph{wati we nä sräthoro ... bäne ... wazu.}\\
	\gll wati we nä srä\stem{thor}o (.) bäne (.) wazu\\ 
	then also {\Indf} \Tsg.\Masc:\Sbj:\Irr:\Pfv:\Andat/arrive (.) \Recog.{\Abs} (.) wazu\\
	\trans `Then, another one arrived ... that one ... Wazu.'
\exi{139} 
	\emph{fä fof sräkor ``watik, foba fof käms wazufa!''}\\
	\gll fä fof srä\stem{kor} watik foba fof kä\stem{m}s wazu=fa\\ 
	{\Dist} {\Emph} \Stsg:\Sbj>\Tsg.\Masc:\Obj:\Irr:\Pfv/speak then \Dist.{\Abl} {\Emph} \Ssg:\Sbj:\Imp:\Pfv/dwell wazu={\Abl}\\\
	\trans `He told him there: ``Okay, you settle there at Wazu!'''
\exi{140} 
	\emph{watik fthé zamara katan fäth ane zfrärm}\\
	\gll watik fthé za\stem{mar}a katan fäth ane zf\stem{rä}rm\\ 
	then when \Sg:\Sbj>\Tsg.\F:\Obj:\Pst:\Pfv/see small \Dim{} {\Dem} \Tsg.\F:\Sbj:\Pst:\Dur/be\\
	\trans `He looked at the place. It was a small patch.'
\exi{141} 
	\emph{``kwa nzä zä zf kwramnzr? nima ŋabrigwé.''}\\
	\gll kwa nzä zä zf kwra\stem{m}nzr nima ŋa\stem{brig}wé\\ 
	{\Fut} \Fsg.{\Abs} {\Prox} {\Imm} \Fsg:\Sbj:\Irr:\Ipfv/dwell like.this \Fsg:\Sbj:\Nonpast:\Ipfv/return\\
	\trans ` ``Will I stay right here? I will go back this way.'''
\exi{142} 
	\emph{boba fthmäsü zänbrima watik ... nasi nömä ane fof tnägsi thenthkäfa rrfar ... futhfuth.}\\
	\gll boba fthmäsü zän\stem{brim}a watik (.) nasi nömä ane fof tnäg-si then\stem{thkäf}a r-rfar (.) futh-futh\\ 
	\Med.{\Abl} while \Sg:\Sbj:\Pst:\Pfv:\Venit/return then (.) long.yam yamcake {\Dem} {\Emph} lose-{\Nmlz} \Sg:\Sbj>\Stpl:\Obj:\Pst:\Pfv:\Venit/start \Redup-crumb (.) \Redup-scrap\\
	\trans `While he was coming back from there, he started dropping this yamcake crumbs ... the scraps.'
\exi{143} 
	\emph{zä zf e zane zf zethno zerä}\\
	\gll zä zf e zane zf z=e\stem{thn}o z=e\stem{rä}\\ 
	{\Prox} {\Imm} until \Dem:{\Prox} {\Imm} \Prox=\Stpl:\Sbj:\Nonpast:\Stat:\Andat/be.lying \Prox=\Stpl:\Sbj:\Nonpast:\Ipfv/be\\
	\trans `Right here to these these stones right here.'
\exi{144} 
	\emph{fä fof ane futhfuth thuntnägwrm}\\
	\gll fä fof ane futh-futh thun\stem{tnäg}wrm\\ 
	{\Dist} {\Emph} {\Dem} \Redup-scrap \Sg:\Sbj>\Stpl:\Obj:\Pst:\Dur:\Venit/lose\\
	\trans `Over there he was dropping the scraps.'
\exi{145} 
	\emph{nä bä enthn e nima zyaro e masu.}\\
	\gll nä bä en\stem{thn} e nima z=ya\stem{r}o e masu\\ 
	{\Indf} \Med{} \Stpl:\Sbj:\Nonpast:\Stat:\Venit/be.lying until like.this \Prox=\Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Andat/be until masu\\
	\trans `There are some stones lying there. and then he walked that way until Masu.'
\exi{146} 
	\emph{foba fof sathora ``nzukar, zä zf ämnzr!''}\\
	\gll foba fof sa\stem{thor}a nzu-kar zä zf ä\stem{m}nzr\\ 
	\Dist.{\Abl} {\Emph} \Tsg.\Masc:\Sbj:\Pst:\Pfv/arrive \Fsg.\Poss-place {\Prox} {\Imm} \Stpl:\Sbj:\Nonpast:\Ipfv/dwell\\
	\trans `He arrived over there and said ``This is my place. My people will live right here!'''
\exi{147} 
	\emph{watik menz kar ane fof zräkorth bäne ... yari.}\\
	\gll watik menz kar ane fof zrä\stem{kor}th bäne (.) yari\\ 
	then myth place {\Dem} {\Emph} \Stpl:\Sbj>\Tsg.\F:\Obj:\Irr:\Pfv/call \Recog.{\Abs} (.) yari\\
	\trans `Well, they call this story place Yari.'
\newpage 
\exi{148} 
	\emph{yari sathora fof.}\\
	\gll yari sa\stem{thor}a fof\\ 
	yari \Tsg.\Masc:\Sbj:\Pst:\Pfv/arrive {\Emph}\\
	\trans `He had arrived at Yari.'
\exi{149} 
	\emph{watik fä fof ... no nzigfukarä fi fof sathora fof.}\\
	\gll watik fä fof (.) no nzigfu=karä fi fof sa\stem{thor}a fof\\ 
	then {\Dist} {\Emph} (.) rain rain.stone={\Prop} \Third.{\Abs} {\Emph} \Tsg.\Masc:\Sbj:\Pst:\Pfv:\Venit/arrive {\Emph}\\
	\trans `Over there ... he arrived with the magic rain stone.'
\exi{150} 
	\emph{fi mane yanra nzigfu nä fofosa yfathwa fof nasi, wawa, duga, fiskar ... ranzo fä dagon eräro.}\\
	\gll fi mane yan\stem{r}a nzigfu nä fofosa y\stem{fath}wa fof nasi wawa duga fiskar (.) ra=nzo fä dagon e\stem{rä}ro\\ 
	\Third.{\Abs} who \Tsg.\Masc:\Sbj:\Pst:\Ipfv:\Venit/be rain.stone {\Indf} heart \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Ipfv/hold {\Emph} long.yam yam taro cassava (.) what={\Only} {\Dist} food \Stpl:\Sbj:\Nonpast:\Ipfv:\Andat/be\\
	\trans `As he came, he had this rain stone and another stone ... for long yams, yams, taro and cassava ... all the crops.'
\exi{151} 
	\emph{anekaräsü swamnzrm fof.}\\
	\gll ane=karä=sü swa\stem{m}nzrm fof\\ 
	\Dem=\Prop=\Etc{} \Tsg.\Masc:\Sbj:\Pst:\Dur/dwell {\Emph}\\
	\trans `He was staying with these ones.'
\exi{152} 
	\emph{fthé wawa thuworthrmth.}\\
	\gll fthé wawa thu\stem{wor}thrmth\\ 
	when yam \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Dur/plant\\
	\trans `Whenever the people were planting yams,'
\exi{153} 
	\emph{watik sfrärm e wawa taga kwot thkarthé kwafiyokwrmth.}\\
	\gll watik sf\stem{rä}rm e wawa taga kwot thkarthé kwa\stem{fiyok}wrmth\\ 
	then \Tsg.\Masc:\Sbj:\Pst:\Dur/be until yam leaf properly hard \Stpl:\Sbj:\Pst:\Dur/make\\
	\trans `he was there until the yam leaves were becoming dry.'
\largerpage
\exi{154} 
	\emph{watik fthé fof wawa taga nä thurtnwrm ... nasi taga ... kemar taga ... taga bäne bera biskar duga.}\\
	\gll watik fthé fof wawa taga nä thu\stem{rtn}wrm (.) nasi taga (.) kemar taga (.) taga bäne b=e\stem{r}a biskar duga\\ 
	then when {\Emph} yam leaf {\Indf} \Sg:\Sbj>\Stpl:\Obj:\Pst:\Dur/pull.off (.) long.yam leaf (.) kemar leaf (.) leaf \Recog.{\Abs} \Med=\Stpl:\Sbj:\Pst:\Ipfv/be cassava taro\\
	\trans `That was when he pulled of some yam leaves, long yam leaves, kemar (type of yam) leaves and those leaves there ... cassava and taro.'
\exi{155} 
	\emph{watik nzigfu mrmr foba sfrärm ... ane tagame sumyuknwrm.}\\
	\gll watik nzigfu mrmr foba sf\stem{rä}rm (.) ane taga=me su\stem{myuk}nwrm\\ 
	then magic.stone inside \Dist.{\Abl} \Tsg.\Masc:\Sbj:\Pst:\Dur/be (.) {\Dem} leaf={\Ins} \Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Dur/wrap\\
	\trans `The magic stone was there inside. He was wrapping it with these leaves.'
\exi{156} 
	\emph{surdiknwrm ... watik wawa zfthen swäzin}\\
	\gll su\stem{rdikn}wrm (.) watik wawa zfth=en swä\stem{zin}\\ 
	\Sg:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Dur/tie.around (.) then yam base={\Loc} \Stsg:\Sbj>\Tsg.\Masc:\Obj:\Iter/put.down\\
	\trans `He tied them around. Then he used to put it down to the yams.'
\exi{157} 
	\emph{sfthnm e wawa fthé thwemar nima thkarthé zäkorth.}\\
	\gll sf\stem{thn}m e wawa fthé thwe\stem{mar} nima thkarthé zä\stem{kor}th\\ 
	\Tsg.\Masc:\Sbj:\Pst:\Dur/be.lying until yam when \Stsg:\Sbj:>\Stpl:\Obj:\Iter/see like.this hard \Stpl:\Sbj:\Pst:\Pfv/become\\
	\trans `The stone was lying there until he saw that the leaves became dry.'
\exi{158} 
	\emph{watik ausiausi thukonzrm ``käthfe kabe!''}\\
	\gll watik ausi-ausi thu\stem{ko}nzrm kä\stem{thf}e kabe\\ 
	then \Redup-old.woman \Sg:\Sbj>\Stpl:\Obj:\Pst:\Dur/say \Spl:\Sbj:\Imp:\Pfv/walk people\\
	\trans `Then he said to the women: ``People, go!'''
\exi{159} 
	\emph{ausiausi thfyakm ŋanz ffrümenzoma ...}\\
	\gll ausi-ausi thf\stem{yak}m ŋanz f-frü=me=nzo=ma (.)\\ 
	\Redup-old.woman \Stpl:\Sbj:\Pst:\Dur/walk garden.row \Redup-single=\Ins=\Only={\Char} (.)\\
	\trans `The women went and took from each patch ...'
\exi{160} 
	\emph{wawa ane ... ebar fr wawa ebar fr kafar}\\
	\gll wawa ane (.) ebar fr wawa ebar fr kafar\\ 
	yam {\Dem} (.) head stem yam head stem big\\
	\trans `those yams ... the best yams and big yams.'
\exi{161} 
	\emph{watik nä ŋanzma wawa näbi nä ŋanzma nä ŋanzma nä ŋanzma nimanzo watik}\\
	\gll watik nä ŋanz=ma wawa näbi nä ŋanz=ma nä ŋanz=ma nä ŋanz=ma nima=nzo watik\\ 
	then {\Indf} garden.row yam one {\Indf} garden.row {\Indf} garden.row {\Indf} garden.row like.this={\Only} then\\
	\trans `one yam from one patch, from another patch, from another patch, from another patch ... just like this.'
\exi{162} 
	\emph{mnime thufränzrmth ... watik thufthakwrmth foba ... karome thurzathrmth.}\\
	\gll mni=me thu\stem{frä}nzrmth (.) watik thu\stem{fthak}wrmth foba (.) karo=me thu\stem{rza}thrmth\\ 
	fire={\Ins} \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Dur/singe (.) then \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Dur/take.out.of.fire \Dist.{\Abl} (.) oven={\Ins} \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Dur/cook.in.oven\\
	\trans `They were burning the hair off the yams. Then they took them out of the fire and cooked them in the ground oven.'
\exi{163} 
	\emph{zizi ane fof thfzänzrmth bobo ... far mä suräzrmth.}\\
	\gll zizi ane fof thf\stem{zä}nzrmth bobo (.) far mä su\stem{räz}rmth\\ 
	afternoon {\Dem} {\Emph} \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Dur/carry \Med.{\All} (.) post where \Stpl:\Sbj>\Tsg.\Masc:\Obj:\Pst:\Dur/erect\\
	\trans `In the afternoon, they carried these yams where they had planted a post.'
\exi{164} 
	\emph{mathkwi o karawa o kukuma o ote ...}\\
	\gll mathkwi o karawa o kukuma o ote (.)\\ 
	mathkwi or karawa or kukuma or ote (.)\\
	\trans `Mathkwi or Karawa or Kukuma or Ote ...'
\exi{165} 
	\emph{watik ane far fof sfrästhgrm wawa fobo fof thunakwrm.}\\
	\gll watik ane far fof sf\stem{räs}thgrm wawa fobo fof thu\stem{nak}wrm\\ 
	then {\Dem} post {\Emph} \Tsg.\Masc:\Sbj:\Pst:\Dur/be.erected yam \Dist.{\All} {\Emph} \Sg:\Sbj>\Stpl:\Obj:\Pst:\Dur/put.down\\
	\trans `Well, that post was standing there and he put the yams down on its base.'
\exi{166} 
	\emph{fobo ffé n wawa kwanäbünzrmth ... kwosi kwakonzrmth bänemr e ... tayo tfotfo.}\\
	\gll fobo ffé n wawa kwa\stem{näbü}nzrmth (.) kwosi kwa\stem{ko}nzrmth bänemr e (.) tayo tfotfo\\ 
	\Dist.{\All} really {\Imn} yam \Stpl:\Sbj:\Pst:\Dur/decompose (.) rotten \Stpl:\Sbj:\Pst:\Dur/become \Recog.{\Purp} until (.) ripe almost\\
	\trans `When those yams were about to fall apart, about to become rotten, the yams (in the ground) were almost ready.'
\exi{167} 
	\emph{tayo wawa fthé kwakonzrmth ...}\\
	\gll tayo wawa fthé kwa\stem{ko}nzrmth (.)\\ 
	ripe yam when \Stpl:\Sbj:\Pst:\Dur/become (.)\\
	\trans `When the yams were getting ready ...'
\exi{168} 
	\emph{rfnaksir bobo zarfa thfrärm ... mäta garaita}\\
	\gll rfnak-si=r bobo zarfa thf\stem{rä}rm (.) mäta garaita\\ 
	taste-\Nmlz={\Purp} \Med.{\All} ear \Stpl:\Sbj:\Pst:\Dur/be (.) mäta garaita\\
	\trans `they heard from Mata and Garaita that they started to taste them.'
\newpage 
\exi{169} 
	\emph{nafa fthé kwänrfnth ... ``ayo!'' wrwr fof zefaro ... swänrifthth}\\
	\gll nafa fthé kwän\stem{rfn}th (.) ayo wrwr fof ze\stem{far}o (.) swän\stem{rifth}th\\ 
	\Tnsg.{\Erg} when \Stpl:\Sbj:\Iter/taste (.) watch.out eastwind {\Emph} \Sg:\Sbj:\Pst:\Pfv:\Andat/set.off (.) \Stpl:\Sbj>\Tsg.\Masc:\Obj:\Iter/send\\
	\trans `When tasting the yams, they always shouted ``Watch out!!'' and the eastwind blew the message here.'
\exi{170} 
	\emph{watik we masu karé kwekaristh ``oh, nafa z zärfnth.''}\\
	\gll watik we masu kar=é kwe\stem{karis}th oh nafa z zä\stem{rfn}th\\ 
	then also masu place=\Erg.{\Nsg} \Stpl:\Sbj:\Iter/hear oh \Tnsg.{\Erg} {\Iam} \Stpl:\Sbj:\Rpst:\Pfv/taste\\
	\trans `The Masu people used to hear this and said ``Oh, they have started tasting the yams already.'''
\exi{171} 
	\emph{we kwot we näbikakme we nä wawa thfrärmth katan o kafar ... thuwoknzrmth ... watik kwarzathrmth.}\\
	\gll we kwot we näbi-kak=me we nä wawa thf\stem{rä}rmth katan o kafar (.) thu\stem{wok}nzrmth (.) watik kwa\stem{rza}thrmth\\ 
	also properly also one-\Distr={\Ins} also {\Indf} yam \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Dur/do small or big (.) \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Dur/choose (.) then \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Dur/cook.in.oven\\
	\trans `Now they also took some yams, one by one, big or small. They choose some and cooked them in the oven.'
\exi{172} 
	\emph{tawar ane thfrärmth ŋazi thurwrmth ... kwot thufathwrmth kobakob}\\
	\gll tawar ane thf\stem{rä}rmth ŋazi thu\stem{rw}rmth (.) kwot thu\stem{fath}wrmth kobakob\\ 
	yam.pulp {\Dem} \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Dur/do coconut \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Dur/scrape (.) properly \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Dur/hold round.object\\
	\trans `They took out the yam pulp and mixed it with scraped coconut and then they formed little round balls out of the dough.'
\exi{173} 
	\emph{kwarfnakwrmth watik nima kwanrzrmth.}\\
	\gll kwa\stem{rfnak}wrmth watik nima kwan\stem{rz}rmth\\ 
	\Sg:\Sbj:\Pst:\Dur/taste then like.this \Sg:\Sbj:\Pst:\Dur:\Venit/throw\\
	\trans `Then they tasted the yams and they threw their arms up this way.'
\exi{174} 
	\emph{fatr nima thwafiyokwrmth ``ayo! farem benm fräro!'' ... nima fof}\\
	\gll fatr nima thwa\stem{fiyok}wrmth ayo farem benm f=\stem{rä}ro (.) nima fof\\ 
	upper.arm like.this \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Dur/make watch.out farem \Snsg.{\Dat} \Dist=\Tsg.\F:\Sbj:\Nonpast:\Ipfv:\Andat/be (.) like.this {\Emph}\\
	\trans `They threw the arms up like this and shouted ``Watch out! Farem people, this is for you!'' like this.'
\exi{175} 
	\emph{watik ane fthé kwärit ane tmatm}\\
	\gll watik ane fthé kwä\stem{rit} ane tmatm\\ 
	then {\Dem} when \Stsg:\Sbj:\Iter/pass.by {\Dem} event\\
	\trans `Each time when that ritual had passed'
\exi{176} 
	\emph{rfnaksi tmatm thumarwrmth e ... rrr kwan fthé bäne kwäkorth ... tayo kwot thuwäkwrm}\\
	\gll rfnak-si tmatm thu\stem{mar}wrmth e (.) {rrr kwan} fthé bäne kwä\stem{kor}th (.) tayo kwot thu\stem{wäk}wrm\\ 
	taste-{\Nmlz} event \Stpl:\Sbj>\Stpl:\Obj:\Pst:\Dur/see until (.) rustling.sound when \Recog.{\Abs} \Stpl:\Sbj:\Iter/become (.) ripe properly \Stpl:\Sbj:\Pst:\Dur/ripen\\
	\trans `when they saw the tasting rituals, when the yam leaves were rustling, when the yams were ready ...'
\exi{177} 
	\emph{watik fthé fof yaka swefafth.}\\
	\gll watik fthé fof yaka swe\stem{faf}th\\ 
	then when {\Emph} digging.stick \Stpl:\Sbj>\Tsg.\Masc:\Obj:\Iter/hold\\
	\trans `that was when they picked up the digging stick and began to harvest.'
\exi{178} 
	\emph{anenzo fof ... ane tmatm kwaritakwrm e zbo bäidbo ... bäi kafar zäkora}\\
	\gll ane=nzo fof (.) ane tmatm kwa\stem{ritak}wrm e zbo bäi=dbo (.) bäi kafar zä\stem{kor}a\\ 
	\Dem={\Only} {\Emph} (.) {\Dem} event \Sg:\Sbj:\Pst:\Dur/cross.over until \Prox.{\All} bäi=\All.{\Sg} (.) bäi big \Sg:\Sbj:\Pst:\Pfv/become\\
	\trans `That was it. That ritual was passed on to Bäi. Bäi had become a big man.'
\exi{179} 
	\emph{nafaŋafyf ... nafane ŋafyf ane fof sara fof ... foba fof otef.}\\
	\gll nafa-ŋafe=f (.) nafane ŋafe=f ane fof sa\stem{r}a fof (.) foba fof ote=f\\ 
	\Third.\Poss-father=\Erg.{\Sg} (.) \Tsg.{\Poss} father=\Erg.{\Sg} {\Dem} {\Emph} \Sg:\Sbj>\Tsg.\Masc:\Io:\Pst:\Pfv/give {\Emph} (.) \Dist.{\Abl} {\Emph} ote=\Erg.\Sg\\
	\trans `His father and his father passed on this tradition ... all the way from Ote.'
\largerpage
\exi{180} 
	\emph{watik naf we ane fof thwamonegwrm no bäne ... no nzigfu a fofosa frä ... dagon fofosa fof.}\\
	\gll watik naf we ane fof thwa\stem{moneg}wrm no bäne (.) no nzigfu a fofosa f=\stem{rä} (.) dagon fofosa fof\\ 
	then \Tsg.{\Erg} also {\Dem} {\Emph} \Sg:\Sbj>\Stpl:\Obj:\Pst:\Dur/look.after rain \Recog.{\Abs} (.) rain rain.stone and heart \Dist=\Tsg.\F:\Sbj:\Nonpast:\Ipfv/be (.) food heart {\Emph}\\
	\trans `He also looked after that rain magic stone and the other stone there ... the magic food stone.'
\exi{181} 
	\emph{foba e ni kafar ŋankonzake.}\\
	\gll foba e ni kafar ŋan\stem{ko}nzake\\ 
	\Dist.{\Abl} until {\Fnsg} big \Fpl:\Sbj:\Nonpast:\Ipfv:\Venit/become\\
	\trans `Later we grew up.' 
\exi{182} 
	\emph{nzesinenwä ane fof komnzo thfrnm ane eda ... eda rokar fof.}\\
	\gll nze-si=en=wä ane fof komnzo thf\stem{rn}m ane eda (.) eda rokar fof\\ 
	\Fnsg.\Poss-eye=\Loc={\Emph} {\Dem} {\Emph} still \Stdu:\Sbj:\Pst:\Dur/be {\Dem} two (.) two things \Emph\\
	\trans `We still saw those two with our own eyes ... those two stones.'
\exi{183} 
	\emph{e nama masun ane yam tmatm z zwabrgwre fof.}\\
	\gll e nama masu=n ane yam tmatm z zwa\stem{brg}wre fof\\ 
	until recently masu={\Loc} {\Dem} custom event {\Iam} \Fpl:\Sbj>\Tsg.\F:\Obj:\Rpst:\Ipfv/follow \Emph\\
	\trans `We have followed this tradition until recently in Masu.'
\exi{184} 
	\emph{e watik foba zänbrimake zena mänwä zä namnzr zf ... znrä.}\\
	\gll e watik foba zän\stem{brim}ake zena mä=wä zä na\stem{m}nzr zf (.) z=n\stem{rä}\\ 
	until then \Dist.{\Abl} \Fpl:\Sbj:\Rpst:\Pfv/return today where={\Emph} {\Prox} \Fpl:\Sbj:\Nonpast:\Ipfv/dwell {\Imm} (.) \Prox=\Fpl:\Sbj:\Nonpast:\Ipfv/be\\
	\trans `Then we returned from there to where we are living now ... right here.'
\exi{185} 
	\emph{watik fi fthmäsü kwik ... kwosi yara ... greg täwdben ane thfrärm.}\\
	\gll watik fi fthmäsü kwik (.) kwosi ya\stem{r}a (.) greg täw=dben ane thf\stem{rä}rm\\ 
	then \Third.{\Abs} meanwhile sick (.) dead \Tsg.\Masc:\Sbj:\Pst:\Ipfv/be (.) greg father=\Loc.{\Sg} {\Dem} \Stpl:\Sbj:\Pst:\Dur/be\\\
	\trans `In the meanwhile father had become sick and died. Those stones were with Greg's father.'
\exi{186} 
	\emph{ane bäne ... nzigfu thfrnm edawä.}\\
	\gll ane bäne (.) nzigfu thf\stem{rn}m eda=wä\\ 
	{\Dem} \Recog.{\Abs} (.) magic.stone \Stdu:\Sbj:\Pst:\Dur/be two={\Emph}\\
	\trans `Those were magic stones ... those two.'
\exi{187} 
	\emph{watik nzenme ŋafe fthmäsü kwosi yara.}\\
	\gll watik nzenme ŋafe fthmäsü kwosi ya\stem{r}a\\ 
	then \Fnsg.{\Poss} father meanwhile dead \Tsg.\Masc:\Sbj:\Pst:\Ipfv/be\\
	\trans `Well, our father died in the meantime ...'
\exi{188} 
	\emph{watik foba ni miyamr nrä mafadben zena ethn.}\\
	\gll watik foba ni miyamr n\stem{rä} mafa=dben zena e\stem{thn}\\ 
	then \Dist.{\Abl} {\Fnsg} ignorant \Fpl:\Sbj:\Nonpast:\Ipfv/be who=\Loc.{\Sg} today \Stpl:\Sbj:\Nonpast:\Ipfv/be.lying\\
	\trans `and since then we do not know with whom these (magic stones) are now.' 	
\exi{189} 
	\emph{z thrifthmath fof}\\
	\gll z th\stem{rifthm}ath fof\\ 
	{\Iam} \Stpl:\Sbj>\Stdu:\Obj:\Pst:\Pfv/hide fof\\
	\trans `They might have hidden them.'
\exi{190} 
	\emph{watik ane bäne mane rera ... ane trikasi mane nŋatrikwé fof ... ŋafynm badafa ane fof ŋanritakwa fof}\\
	\gll watik ane bäne mane \stem{rä}ra (.) ane trik-si mane n=ŋa\stem{trik}wé fof (.) ŋafe=nm bada=fa ane fof ŋan\stem{ritak}wa fof\\ 
	then {\Dem} \Recog.{\Abs} which \Tsg.\F:\Sbj:\Pst:\Ipfv/be (.) {\Dem} tell-{\Nmlz} which \Immpst=\Fsg:\Sbj:\Nonpast:\Ipfv/tell {\Emph} (.) father=\Dat.{\Nsg} ancestor={\Abl} {\Dem} {\Emph} \Sg:\Sbj:\Pst:\Ipfv/pass {\Emph}\\
	\trans `The story, which I have just told, was passed on from the ancestor to the fathers.'
\exi{191} 
	\emph{bada aki kwark benrera fof ... zath kwark enrera e ŋafydbo we nzedbo fof n zänrita nima}\\
	\gll bada aki kwark b=en\stem{rä}ra fof (.) zath kwark en\stem{rä}ra e ŋafe=dbo we nzedbo fof n zän\stem{rit}a nima\\ 
	ancestor grandfather deceased \Med=\Stpl:\Sbj:\Pst:\Ipfv:\Venit/be {\Emph} (.) grandfather deceased \Stpl:\Sbj:\Pst:\Ipfv:\Venit/be until father=\All.{\Sg} also \Fnsg.{\All} {\Emph} {\Iam} \Sg:\Sbj:\Pst:\Pfv/pass like.this\\
	\trans `From the ancestors to the late grandfathers until it came to the fathers. It was about to pass to us...'
\exi{192} 
	\emph{watik maf keke wäbragwr ane}\\
	\gll watik maf keke wä\stem{brag}wr ane\\ 
	then who.\Erg.{\Sg} {\Neg} \Stsg:\Sbj>\Tsg.\F:\Obj:\Nonpast:\Ipfv/follow {\Dem}\\\
	\trans `but nobody follows (these traditions) anymore.'\Corpus{tci20131013-01}{}
\end{exe}%kwafar