%!TEX root = ../main.tex

\addchap{Acknowledgments}

This grammar of Komnzo started out as my PhD project at the Australian National University in Canberra. Since 2016, there have been many additions and revisions to this grammar, but the majority of the contents are the same as in the final version of my dissertation. These changes are the result of the comments given by reviewers, editors and proof-readers as well as the ever-increasing knowledge I receive from my Komnzo speaking friends.

This book would not have been possible without the support of the Farem people who took on the task of teaching me their language. I am deeply grateful to them for welcoming me in their community, for feeding me and keeping me safe at all times, for the patience they have had with my probing questions, and above all for sharing their language and culture with me. It is impossible to acknowledge everyone who assisted in teaching me Komnzo, for every exchange provided a contribution to my knowledge. Amongst my indigenous teachers were: Abia Bai, Nakre Abia, Daure Kaumb, Riley Abia, \textsuperscript{\dag}Marua Bai, Lucy Abia, Sékri Karémbu, Janet Abia, Steven Karémbu, Caspar Mokai, Karo Abia, Kaumb Bai, Moses Abia and Albert Mokai.

My principal supervisor Nicholas Evans first suggested the Tonda languages as an area of research. Nick's enthusiasm and challenging criticism has helped me to sharpen my analysis and description of Komnzo. His open-mindedness about fieldwork and his holistic approach to language documentation made me see fascinating details of the language. I have greatly appreciated the contributions of I Wayan Arka, Andrew Pawley, Mark Ellison and Mark Donohue, who took over supervision at various periods over the years. I thank Ulrike Mosel, Ger Reesink and Martin Haspelmath for reviewing the dissertation. Their positive as well as their challenging comments have greatly improved this grammar. I am indebted to the administrative staff at the School of Culture, History and Language who have helped me navigating the bureaucracy. I would like to thank Jo Bushby, Penelope Judd, and Stephen Meatheringham. I am grateful to Kay Dancey at cartographic unit for the linguistic map of the Morehead district (see Figure \ref{fig-01-SNGmap}).

There are a number of people who have helped me with specific knowledge and advice concerning virtually all aspects of carrying out research in the Morehead district. I would like to thank Mary Ayres who invited me to her house in Philadelphia and shared her fieldnotes from Rouku. Paul O'Rear and Risto Sarsa have answered many questions and requests about Rouku and Yokwa. Thanks also to Garrick Hitchcock and Kevin Murphy for sharing their knowledge about and experience in working in this part of the country. I also thank Jeff Siegel for his help both in Morehead and in other places where we have met. I want to thank Cezar Fernandez and the staff of New Century Hotel in Daru, who continue to make my brief stays in this town a pleasant experience, and Douglas Dawi, Andrew Little, Peter Paradi, and Charlie Subam for transporting me and my equipment safely between Daru and the Morehead district.

Over the years, I have received financial support from the DOBES project of the Volkswagen Foundation, the Stephen and Helen Wurm Bequest at ANU, The Max Planck Institute for Evolutionary Anthropology and the ACR Centre of Excellence for the Dynamics of Language at ANU. I thank all these institutions for making language documentation and description possible. I thank especially Vera Szöllösi-Brenig for organising the DOBES project, as well as Paul Trijlsbeek, Han Sloetjes and Alexander König for providing technical training and assistance.

I thank the staff at Language Science Press, especially Sebastian Nordhoff and Martin Haspelmath. Moreover, I am grateful to the numerous volunteers who spent their valuable time during the publication process of this book. I embrace you in the spirit of true open-access publishing.

To my family and friends who have provided me with moral and practical support over the years, I am deeply grateful. Too many years and too many people have passed by to thank everyone, but particular thanks go to Darja Hoenigman, Aung Si, Charlotte van Tongeren, Sebastien Lacrampe, Penny Johnson, Gary Kildea, Matthew Carroll, Beth Evans, Alena Witzlack-Makarevich and Sonja Riesberg. Lastly, I want to thank my parents Regina and Jörg, my brother Matthias and his family for their continued support over the last years.