\addchap{Acknowledgements}
%\chapter*{Acknowledgements}

In writing this book I have benefited greatly from communication with Balthasar Bickel, Claire Bowern, Rob 
Boyd, Morten Christiansen, Jeremy Collins, Grev Corbett, Stephen Cowley, Sonia Cristofaro, Bill Croft, Jennifer Culbertson, Dan Dediu, Mark Dingemanse, Daniel Dor, Grant Evans\dag, Nick Evans, Bill Foley, Bill Hanks, Martin Haspelmath, Larry Hyman, Jennifer Johnson-Hanks, Simon Kirby, Wolfgang Klein, Chris Knight, Paul Kockelman, Michael Lempert, Steve Levinson, Elena Lieven, Hugo Mercier, Pieter Muysken, Csaba Pl\'{e}h, Joanna R\k{a}czaszek-Leonardi, Keren Rice, Peter Richerson, Se\'{a}n Roberts, Giovanni Rossi, Wendy Sandler, Jack Sidnell, Chris Sinha, Hedvig Skirg\aa{}rd, Kenny Smith, Dan Sperber, Sune Vork Steffensen, Monica Tamariz, Jordan Zlatev, and Chip Zuckerman. I thank participants at the conference 
\textit{Naturalistic Approaches to Culture} (Balatonvilagos 2011), the 
conference \textit{Social Origins of Language} (London 2011), the 
conference \textit{Language, Culture, and Mind V} (Lisbon 2012), the workshop 
\textit{Rethinking Meaning} (Bologna 2012), the \textit{Minerva-Gentner Symposium on 
Emergent Languages and Cultural Evolution} (Nijmegen 2013), and the retreat \textit{Dependencies among Systems of Language} (Ch\^{a}teau de la Poste 2014) for comments, reactions, and inspiration. 

For troubleshooting with \LaTeX\ I am grateful to Sebastian Nordhoff and Se\'{a}n Roberts. This work is supported by the European Research Council 
(grant 240853 \textit{Human Sociality and Systems of Language Use}, 2010--2014), and 
the Max Planck Institute for Psycholinguistics, Nijmegen. Chapters 2--5 are thoroughly revised versions of previously published works: as chapters in 
\textit{Social Origins of Language} (OUP, \citeyear{DorEtAl2014}, edited by D. Dor, C. Knight, and J. Lewis), and \textit{The Cambridge Handbook of Linguistic Anthropology} 
(CUP, \citeyear{EnfieldEtAl2014handbookla}, edited by N. J. Enfield, P. Kockelman, and J. Sidnell). Further, certain parts draw on sections of Enfield (\citeyear{enfield_transmission_2008}, \citeyear{enfield_relationship_2013}, 
chap. 11).

I dedicate this book to Grant Evans (1948--2014): historian, sociologist, and anthropologist of Southeast Asia. In our conversations over nearly 20 years, Grant always challenged my natural tendency to focus on items. He never stopped pushing me to acknowledge the causal reality of socio-cultural systems. His intellectual engagement has been one of the main motivations for me to confront the item/system problem that is at the heart of this book.