\addchap{Acknowledgments}\largerpage
This grammar could not have been written without the support of many people and institutions.

First of all I am indebted to the Sanzhi community, in particular to my main assistant and consultant Gadzhimurad Gadzhimuradov, his wife Fatimat (\sqt{Batichaj}) and his entire family. His passion for Sanzhi culture, history and language, his extraordinary patience in dealing with my endless questions and his linguistic intuitions made the work on this grammar possible and in fact a wonderful experience. He and his family hosted me during the entire project and allowed me to fully concentrate on my work while being their guest in Druzhba.

The grammar is one of the results of the language documentation project \textit{Documenting Dargi languages in Dagestan -- Shiri and Sanzhi} (2012--2019). The project was supported by many other members of the Sanzhi community in addition to Gadzhimurad, in particular by Isakadi Bakhmudov and Asabali Gadzhimuradov. A big \textit{barkalla} goes to all of you!

The project has received financial support through a grant from the Volkswagen Foundation (DobeS program) to Diana Forker (Grant Number 86 357). This generous support is gratefully acknowledged. I am particularly thankful for the help of Vera Szöllösi (Volkswagen Foundation) and for the administrative support of Geoffrey Haig (University of Bamberg).

The project has been conducted by an international team of researchers, first of all Rasul Mutalov, whose idea it was to document endangered Dargwa languages and to apply for a grant from the Volkswagen Foundation. I also wish to thank my other two fellow project members, Oleg Belyaev and Iwona Kaliszewska, as well as the student assistants André Müller, Teresa Klemm, Felix Anker, and Nils Schiborr for scientific, technical and practical support at all stages of the project.

The grammar has profited from many discussions with researchers, colleagues and friends, especially the very active ``Caucasian community'' scattered around the globe, but united by the interest and the fascination for the languages of the Caucasus. In particular I want to thank the reviewers who kindly provided many comments, corrections and useful suggestions: Gilles Authier, Natasha Bogomolova, Ilya Chechuro, Denis Creissels, Misha Daniel, Nina Dobrushina, Dima Ganenkov, Linda Konnerth, Yura Lander, Timur Maisak, Zarina Molochieva, George Moroz, Jérémy Pasquereau, Stefan Schnell, Nina Sumbatova, and Jussi Ylikoski. I am also grateful to Sebastian Nordhoff, Felix Kopecky and the proof readers from Language Science Press for making it possible to publish high quality books that are accessible to everybody. Finally I am lucky to have the best family support I can think of.
\\\\
Tanzt eure Revolution!
