\chapter{Phonology}\label{cpt:phonology}
    \addtocontents{toc}{\protect\enlargethispage{\baselineskip}}

Sanzhi phonology is typical for East Caucasian languages with its relative large consonant inventory (\refsec{sec:Consonant inventory}) and medium vowel inventory (\refsec{sec:Vowel inventory}). Other topics covered in this chapter are the \isi{syllable structure} (\refsec{sec:Syllable and word structure}), \isi{pharyngealization} (\refsec{sec:Pharyngealization}), \isi{stress} (\refsec{sec:Word stress}), and phonological and morphophonological alternations (\refsec{sec:Phonological and morphophonological alternations}). 


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Consonant inventory}
\label{sec:Consonant inventory}

\reftab{tab:The consonant inventory of Sanzhi Dargwa} displays the consonant inventory for Sanzhi. The table gives the phonemic value of the \isi{consonants} and  displays the orthographic representation used in this grammar in italics (see also page xvii for the Cyrillic \isi{orthography}). The three series of stops are, in the order given in the table: voiceless non-\isi{ejective}, voiced, and voiceless \isi{ejective}. The two series of fricatives are voiceless and voiced. All velars and uvulars also occur in labialized form. All voiceless non-\isi{ejective} stops and fricatives (except for the pharyngeal/epiglottal and the glottal sounds) also occur as geminates (i.e. tense).

\begin{table}\footnotesize
	\caption{The consonant inventory of Sanzhi Dargwa\label{tab:The consonant inventory of Sanzhi Dargwa}}
	\begin{tabularx}{1\textwidth}{Q c@{ }c@{ }c     c@{ }c@{ }c  c@{ }c c c@{ }c@{ }c c@{ }c c c}
		\lsptoprule
		& 	\multicolumn{3}{c}{\rotatebox{90}{bilabial}}
		& 	\multicolumn{3}{c}{\rotatebox{90}{alveolar}}
		& 	\multicolumn{2}{c}{\rotatebox{90}{postalv}}
		& 	\rotatebox{90}{palatal}
		& 	\multicolumn{3}{c}{\rotatebox{90}{velar}}
		& 	\multicolumn{2}{c}{\rotatebox{90}{uvular}}
		& 	\rotatebox{90}{\parbox{1.5cm}{pharyngeal/\linebreak epiglottal}}
		&	\rotatebox{90}{glottal}\\\midrule
			stop		& /p/ 	& /b/ 	& /pʼ/ 	& /t/ 	& /d/ 	& /tʼ/	& {} 	& {} 	& {} 	& /k/ 	& /ɡ/ 	& /kʼ/ 	& /q/ 	& /qʼ/ 	& /ʡ/ 	& /ʔ/\\
			{}		& \tit{p} & \tit{b} & \tit{pʼ} & \tit{t} & \tit{d} & \tit{tʼ} & {} & {} & {} & \tit{k} & \tit{g} & \tit{kʼ} & \tit{q} & \tit{qʼ} & \tit{ʡ} & \tit{ʔ}\\
			{}		& {}	& {} 	& {} 	& {} 	& {} 	& {}	& {} 	& {} 	& {} 	& /kʷ/ 	& /ɡʷ/ & /kʼʷ/	& /qʷ/ & /qʼʷ/	& {} 	& {}\\
			{}		& {} 	& {} 	& {} 	& {} 	& {} 	& {} 	& {} 	& {} 	& {} 	& \tit{kʷ} & \tit{gʷ}	& \tit{kʼʷ} & \tit{qʷ} & \tit{qʼʷ} & {} & {}\\
			{}		& /pː/ 	& {} 	& {} 	& /tː/	& {} 	& {} 	& {} 	& {} 	& {} 	& /kː/ 	& {} 	& {} 	& /qː/ 	& {} 	& {} 	& {}\\
			{}		& \tit{pː } 	& {} & {} & \tit{tː }	& {} & {} & {} 	& {} 	& {} 	& \tit{kː } & {} & {} 	& \tit{qː } & {} & {} 	& {}\\
			{}		& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& /kːʷ/	& {}	& {}	& /qːʷ/	& {}	& {}	& {}\\
			{}		& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& \tit{kːʷ} & {} & {}	& \tit{qːʷ} & {}	& {}	& {}\\\midrule

			fricative	& {}	& {}	& {}	& /s/ 	& /z/	& {}	& /ʃ/	& /ʒ/	& {}	& {}	& /x/	& {}	& /χ/	& /ʁ/	& /ħ/	& /h/\\
			{}		& {}	& {}	& {}	& \tit{s}	& \tit{z}	& {}	& \tit{š}	&  \tit{ž}	& {}	& {}	& \tit{x}	& {}	& \tit{χ}	& \tit{ʁ}	& \tit{ħ}	& \tit{h}\\
			{}		& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& /xʷ/	& {}	& /χʷ/ & /ʁʷ/	& {}	& {}\\
			{}		& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& \tit{xʷ}	& {}	& \tit{χʷ} 	& \tit{ʁʷ}	& {}	& {}\\
			{}		& {}	& {}	& {}	& /sː/	& {}	& {}	& /ʃː/	& /ʃː/	& {}	& {}	& /xː/	& {}	& /χː/	& {}	& {}	& {}\\
			{}		& {}	& {}	& {}	& \tit{sː}& {}	& {}	& \tit{šː}	& \tit{šː} & {}	& {}	& \tit{xː} & {}	& \tit{χː} & {}	& {}	& {}\\
			{}		& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& /χːʷ/	& {}	& {}	& {}\\
			{}		& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& \tit{χːʷ} & {} & {}	& {}\\\midrule

			affricate	& {}	& {}	& {}	& /t͡s/ & /t͡sʼ/	& {}	& /t͡ʃ/	& /t͡ʃʼ/ & {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}\\
			{}		& {}	& {}	& {}	& \tit{c} & \tit{cʼ} & {} & \tit{č} & \tit{čʼ}  & {} & {} & {} & {}	& {}	& {}	& {}	& {}\\
			{}		& {}	& {}	& {}	& /t͡sː/	& {}	& {}	& /t͡ʃː/	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}\\
			{}		& {}	& {}	& {}	& \tit{cː} & {} & {}	& \tit{čː}	& {} & {} & {}	& {}	& {}	& {}	& {}	& {}	& {}\\\midrule

			nasal		& {}	& /m/	& {}	& {}	& /n/	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}\\
			{}		& {}	& \tit{m} & {} & {}	& \tit{n} & {} & {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}\\\midrule

			liquid		& {}	& {}	& {}	& /r/	& /l/	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}\\
			{}		& {}	& {}	& {}	&  \tit{r}	& \tit{l} & {} & {} & {} & {}	& {}	& {}	& {}	& {}	& {}	& {}	& {}\\\midrule

			semivowel		& {}	& /w/	& {}	& {}	& {}	& {}	& {}	& {}	& /j/	& {}	& {}	& {}	& {}	& {}	& {}	& {}\\
					& {}	& \tit{w} & {} & {}	& {}	& {}	& {}	& {}	& \tit{j} & {} & {}	& {}	& {}	& {}	& {}	& {}\\
		\lspbottomrule
	\end{tabularx}
\end{table}

The uvular stops /q/ and /qʷ/ have strong friction that makes them sound almost like affricates /q͡χ/ and /q͡χʷ/. The friction is absent from the \isi{ejective} /q'/ and the geminates /qː/ and /qːʷ/.

The phonemic glottal stop is found in the noun \tit{beʔe} \sqt{blood} and at the end of some words, for instance in the root-final position of two verbs \tit{ha-ʔ-} (\tsc{pfv})\slash\tit{h-erʔ-} (\tsc{ipfv}) \sqt{say} and \tit{b-erʔ-} (\tsc{pfv})\slash\tit{b-uʔ-} (\tsc{ipfv}) \sqt{rot} and the numeral \tit{kːaʔ-al} \sqt{eight}. Except for \tit{beʔe} \sqt{blood}, only loan words and names contain the glottal stop in root-medial position (e.g. \tit{daʔim} \sqt{continuation}, in the male name \tit{žaˁbraˁʔil}).

A non-phonemic glottal stop, which is not written, occurs before word-initial non-pharyngealized vowels, e.g. \tit{aba} [ʔaba] \sqt{mother}, including vowel-initial words in compounds, for example \tit{ca-ibil} [t͡saʔibıl] \sqt{first} (one\tsc{-ord}), or occasionally at other morpheme boundaries of inflected words, for example, \tit{a-uk-un}  \sqt{not eating} (\tsc{neg-}eat\tsc{.ipfv-icvb}) can be pronounced [aʔukʊn] or [aʊ̯kʊn].

The semivowel /w/ is realized as a voiced labiodental fricative [v] or as a labial-velar approximant [w]. 

In addition to the segments listed in \reftab{tab:The consonant inventory of Sanzhi Dargwa}, the voiceless labiodental fricative /f/ is attested in the \isi{ideophone} \tit{uf b-ik'ʷ-ij} \sqt{blow} (whew \tsc{hpl-}say\tsc{.ipfv-inf}) and in loan words, mostly from Russian, e.g. \textit{forel} \sqt{trout}. In older loans it had been replaced with /p/, e.g. \textit{purma} \sqt{uniform} (< \textit{forma}).

All plain \isi{consonants} occur in word-initial, word-medial, and word-final position. Geminates are never found in syllable-final position. Three labialized \isi{consonants} (/q'ʷ/, /χʷ/, /ʁʷ/) are also not attested in syllable-final position. \reftab{tab:Distribution of consonants@A} shows the \isi{distribution of consonants} by means of example words. The table contains a \isi{number} of morphologically complex words for which the relevant sound happens to occur at the end of the root, but within the stem because the root is followed by suffixes (the root is given in boldface). 

\begin{table}
	\caption{Distribution of consonants}
	\label{tab:Distribution of consonants@A}
	\footnotesize
	\begin{tabularx}{1\textwidth}[]{%
		>{\raggedright\arraybackslash\itshape}p{10pt}
		>{\raggedright\arraybackslash\hangindent=0.5em\itshape}X
		>{\raggedright\arraybackslash\hangindent=0.5em\itshape}X
		>{\raggedright\arraybackslash\hangindent=0.5em\itshape}X}
		\lsptoprule
 			{}	&	\upshape initial			&	\upshape medial			&	\upshape final\\
		\midrule
			p	&	puq'a \sqtx{nest}			&	qupi \sqtx{hoe}			&	t'up \sqtx{finger}\\
			b	&	bec' \sqtx{wolf}			&	heba \sqtx{then}			&	urχːab \sqtx{mill}\\
			p’	&	p'aq' \sqtx{shake off}		&	q'aˁp'i \sqtx{shutter}		&	lap' \sqtx{wave}\\
			pː	&	pːiħaˁla \sqtx{feather}		&	k'apːur \sqtx{leaf}			&	\tmd\\
			t	&	tum \sqtx{hill}			&	kːaˁta \sqtx{cat}			&	it \sqtx{that}\\
			d	&	du \sqtx{1\tsc{sg}}			&	juldaš \sqtx{friend}			&	ca-d \sqtx{is} \upshape(\tsc{cop}-\tsc{n})\\
			t’	&	t'up \sqtx{finger}			&	kːat'i \sqtx{scarf}			&	t'ult' \sqtx{bread}\\
			tː	&	tːaˁm \sqtx{trap}			&	tːutːu \sqtx{beak}			&	\tmd\\	   
			k	&	kabc \sqtx{skin, fell}		&	dukala \sqtx{apron}		&	dek \sqtx{dung}\\
			g	&	gurmedi \sqtx{type of kerchief}	&	zigar \sqtx{hurry}			&	dig \sqtx{meat}\\
			k’	&	k'apːur \sqtx{leaf}			&	nik'a \sqtx{little, small}		&	hek' \sqtx{this/that (up)}\\
			kʷ	&	kʷač'a \sqtx{paw}			&	mikʷa \sqtx{fingernail}		&	nekʷ \sqtx{straw}\\
			gʷ	&	gʷargʷal \sqtx{onion}		&	targʷa \sqtx{weasel}		&	mergʷ \sqtx{lair, den}\\
			kʼʷ	&	k'ʷel \sqtx{two}			&	r-\textbf{ik'ʷ}-ij\footnote{The relevant roots of morphologically complex words are given in boldface. In these words, the respective sound occurs at the end of the root, but within the stem because the root is followed by suffixes.} \sqtx{say} \upshape(\tsc{f-}say\tsc{.ipfv-inf})	&	erk'ʷ \sqtx{river}\\
			kː	&	kːaˁta \sqtx{cat}			&	kːalkːi \sqtx{tree}			&	\tmd\\
			kːʷ	&	kːʷacːa \sqtx{mare}			&	\textbf{akːʷ}-ar \sqtx{without} \upshape(\tsc{cop.neg-prs}) &	\tmd\\
			q	&	qaˁr \sqtx{pear}			&	b-aqil \sqtx{much}			&	qːaq \sqtx{back}\\
			q’	&	q'aˁp'i \sqtx{shutter}		&	puq'a \sqtx{nest}			&	aq' \sqtx{flock}\\
			qʷ	&	qʷesːa \sqtx{ashes}			&	ha-\textbf{lqʷ}-an  \sqtx{the climbing one} \upshape(\tsc{up}-direct\tsc{.ipfv-ptcp}) & daˁrqʷ \sqtx{barn}\\
			qʼʷ	&	q'ʷaˁl \sqtx{cow}			&	b-\textbf{elq'ʷ}-ij \sqtx{break} \upshape(\tsc{n-}break\tsc{.pfv-inf})		&	\tmd\\
			qː	&	qːap \sqtx{sack}			&	qːuˁlqːuˁ \sqtx{scythe}		&	\tmd\\
			qːʷ	&	qːʷaz \sqtx{goose}			&	\textbf{miriqːʷ}-e  \sqtx{worms} \upshape(worm\tsc{-pl})		&	\tmd\\
			s	&	sala \sqtx{in front, before}	&	qusmuk \sqtx{cupboard}		&	dus \sqtx{year}\\
			z	&	zija \sqtx{horsefly}			&	zize \sqtx{strawberry}		&	keruz \sqtx{slope}\\
			sː	&	sːika \sqtx{bear}			&	musːa \sqtx{place}			&	\tmd\\
			š	&	šal \sqtx{direction, side}		&	haniša \sqtx{summer}		&	juldaš \sqtx{friend}\\
			ž	&	žergʷa \sqtx{wasp}			&	ižal \sqtx{today}			&	hež \sqtx{this}\\
			šː	&	šːi \sqtx{village}			&	dešːa \sqtx{ancient}		&	\tmd\\
			x	&	xujal \sqtx{five}			&	xurxe \sqtx{sobber}		&	c'erx \sqtx{fat}\\
			xʷ	&	xʷit' \sqtx{whistle} (\isit{ideophone})	&	\textbf{ixʷ}-le \sqtx{early}			&	dirixʷ \sqtx{fog}\\
			xː	&	xːamxːa \sqtx{foam}			&	dirxːa \sqtx{stick}			&	\tmd\\
			χ	&	χat'a \sqtx{bowl}			&	alχni \sqtx{saw}			&	maχ \sqtx{barrow}\\
			ʁ	&	ʁajal \sqtx{twenty}			&	pːurʁum \sqtx{carriage}		&	qːabaʁ \sqtx{pumpkin}\\
			χʷ	&	χʷal-le \sqtx{much, a lot}		&	b-\textbf{iχʷ}-ij  \sqtx{be, become} \upshape(\tsc{n-}be\tsc{.pfv-inf})	&	\tmd\\
			ʁʷ	&	ʁʷab \sqtx{ploughshare}		&	aʁʷal \sqtx{four}			&	\tmd\\
			χː	&	χːula \sqtx{big, tall}			&	duχːu \sqtx{clever}			&	\tmd\\
			χːʷ	&	χːʷe \sqtx{dog}			&	ha-d-\textbf{erχːʷ}-ij \sqtx{fulfill} \upshape(\tsc{up}\tsc{-n-}fulfill\tsc{.pfv-inf}) &	\tmd\\
			c	&	ca \sqtx{one}			&	q'aca \sqtx{he-goat}		&	kabc \sqtx{skin, fell}\\
			c’	&	c'il \sqtx{then}			&	imc'a \sqtx{superflous}		&	bec' \sqtx{wolf}\\
			cː	&	cːab \sqtx{sky}			&	kːancːa \sqtx{step}			&	\tmd\\
			č	&	čina-b \sqtx{where\tsc{-n}}	&	ʡaˁči \sqtx{work}			&	deč \sqtx{drinking}\\
			čʼ	&	č'an \sqtx{wind, storm}		&	kʷač'a \sqtx{paw}			&	ʡaˁmč' \sqtx{May it peel off!} \upshape(peel\tsc{.pfv.opt})\\
			čː	&	čːaˁʡaˁl \sqtx{tomorrow, morning} &	ečːa \sqtx{she-goat}		&	\tmd\\
	\end{tabularx}
\end{table}
\begin{table}
	%\caption{Distribution of consonants (cont.)}
	%\label{tab:Distribution of consonants@B}
	\footnotesize
	\begin{tabularx}{1\textwidth}[]{%
		>{\raggedright\arraybackslash\itshape}p{10pt}
		>{\raggedright\arraybackslash\hangindent=0.5em\itshape}X
		>{\raggedright\arraybackslash\hangindent=0.5em\itshape}X
		>{\raggedright\arraybackslash\hangindent=0.5em\itshape}X}
		\midrule
 			{}	&	\upshape initial			&	\upshape medial			&	\upshape final\\
		\midrule
			ʔ	&	aba \sqtx{mother}			&	beʔe \sqtx{blood}			&	b-aʔ \sqtx{begin}\\
			ħ	&	ħaˁšak \sqtx{pot}			&	pːiħaˁla \sqtx{feather}		&	ʡaˁħ \sqtx{good}\\
			ʡ	&	ʡaˁbal \sqtx{three}			&	čːaˁʡaˁl \sqtx{tomorrow, morning}&	daˁʡ \sqtx{face}\\
			h	&	hel \sqtx{that}			&	buhem \sqtx{bundle}		&	b-ah \sqtx{owner}\\
			m	&	mikʷa \sqtx{fingernail}		&	gurmendi \sqtx{type of kerchief}	&	t'em \sqtx{smell}\\
			n	&	nekʷ \sqtx{straw}			&	haniša \sqtx{summer}		&	arin \sqtx{too much}\\
			r	&	rucːi \sqtx{sister}			&	rursːi \sqtx{girl, daughter}		&	q'ar \sqtx{herbs}\\
			l	&	lazun \sqtx{dough}			&	ʡuˁla \sqtx{wheel}			&	hel \sqtx{that}\\
			w	&	weral \sqtx{seven}			&	gawhar \sqtx{pupil}		&	alaw \sqtx{around}\\
			j	&	jangi \sqtx{new}			&	zija \sqtx{horsefly}			&	hej \sqtx{this}\\
		\lspbottomrule
	\end{tabularx}
\end{table}

Final voiced stops do not undergo devoicing. Final voiceless non-\isi{ejective} stops (/p/, /t/, /k/, /q/) are post-aspirated. Stops in final position are released. They are also released when a homorganic consonant follows, e.g. \tit{urek-c'al} \sqt{sixty}, \tit{ħaˁžat-ce} \sqt{necessary} (need\tsc{-dd.sg}), \tit{c'elt-ne} \sqt{gravestone\tsc{-pl}}, \tit{le-d=nu} (exist-\tsc{npl=prt}). If the voiceless stops /t/, /k/, or the voiceless affricate /t͡s/ occur at morpheme boundaries and are followed by homorganic \isi{consonants}, all \isi{consonants} are fully pronounced and released \refex{ex:gemination A phon}. Neither /t/ nor /k/ nor /t͡s/ become geminates under the described conditions, although gemination is otherwise a frequent process that applies across morpheme boundaries (\refsec{ssec:Gemination and degemination}). However, the \isi{ejective} stop /k’/ can turn into a plain stop as shown in the examples in \refex{ex:ikka phon}.

\begin{exe}
	\ex	\label{ex:gemination A phon}
	\begin{xlist}
		\ex	\tit{b-uˁc-ce}\slash\tit{b-uˁc-te} \sqt{thick} (\tsc{n-}thick\tsc{-dd.sg}\slash\tsc{n-}thick\tsc{-dd.pl})	\label{ex:bucce phon}
		\ex	\tit{tunt-ce}\slash\tit{tunt-te} \sqt{daring}	\label{ex:tuntce phon}
		\ex	\tit{ik'-ka} / \tit{hek'-ka} \sqt{from that} (\tsc{dem.up}\tsc{-abl}) (alternatively \tit{ik-ka} / \tit{hek-ka})
		\label{ex:ikka phon}
	\end{xlist}
\end{exe}

All velar and uvular \isi{consonants} occur in plain and labialized forms. The labialized velars and uvulars can be followed by all vowels except /u/. Labialization is mostly found with syllable-initial \isi{consonants}, but as \reftab{tab:Distribution of consonants@A} shows, there are also words with labialized \isi{consonants} in final position. In most words, \isi{labialization} is restricted to one consonant per root, but there are a \isi{number} of words with two labialized \isi{consonants}, e.g. \tit{gʷagʷa} \sqt{flower}, \tit{gʷargʷal} \sqt{onion}, and \tit{xʷixʷit'} \sqt{pipe}. In addition to \isi{labialization} in roots, deletion of the vowel /u/ triggers \isi{labialization} of the preceding consonant or following consonant (\refsec{ssec:Labialization and delabialization}). Labialized \isi{consonants} are mostly found in \isi{nouns}, numerals, \isi{adjectives}, adverbs, and verbs and also attested in a few \is{particle}particles, but not in pronouns or suffixes. Labialization is absent from Standard Dargwa and therefore speakers who have been trained in the Standard Dargwa \isi{orthography} do not write them in Sanzhi, although they pronounce them. Younger speakers often replace labialized \isi{consonants} by plain \isi{consonants} and change a preceding or following \tit{a} to \tit{o} (in speech and writing). Minimal pairs for some labialized \isi{consonants} are given in \refex{ex:labialization phon@A}.

\begin{exe}
	\ex	\label{ex:labialization phon@A}
	\begin{xlist}
		\TabPositions{14em}
		\ex	\tit{d-elq'-ij} (\tsc{pfv}) \sqt{grind}	\tab \tit{d-elq'ʷ-ij} (\tsc{pfv}) \sqt{break}\label{ex:delqij phon}
		
		\ex	\tit{b-iχ-ij} (\tsc{pfv}) \sqt{tie, fasten}	\tab \tit{b-iχʷ-ij} (\tsc{pfv}) \sqt{be, become, be able}	\label{ex:bixwij phon}
		\ex	\tit{akri} \sqt{Akri} (place name)	\tab \tit{akʷri} \sqt{be not} (\tsc{cop.neg.msd}) \label{ex:akri phon}
		\ex	\tit{ik'-i-j} \sqt{for this / that up}	\tab \tit{ik'ʷ-ij} \sqt{say}	(say\tsc{.m.ipfv-inf}) \label{ex:ikwij phon}\\
		(\tsc{dem.up}\tsc{-obl-dat}) 
	\end{xlist}
\end{exe}

Geminates are always voiceless, non-\isi{ejective}, and unaspirated. All voiceless non-eject\-ive\is{ejective} obstruents, except for pharyngeal/epiglottal and glottal segments, occur as geminates, and even a \isi{number} of labialized \isi{consonants} are geminates. The phonemic status of geminates is proven by the minimal pairs and minimal oppositions in \refex{ex:gemination B phon@A}.\largerpage

\begin{exe}
	\ex	\label{ex:gemination B phon@A}
	\begin{xlist}
\TabPositions{14em}
		\ex	\tit{iχ-i-j} \sqt{for this / that down}	\tab \tit{iχː-ij} \sqt{guard, protect, care}	\\
		(\tsc{dem.down}\tsc{-obl-dat})
		\ex	\tit{b-uq-ij} (\tsc{pfv}) \sqt{run, go} \tab \tit{b-uqː-ij} \sqt{carry, bring}	\label{ex:buqij phon}\pagebreak
	
		\ex	\tit{bus-ij} \sqt{rain}	\tab \tit{b-usː-ij} \sqt{sleep, fall asleep} \label{ex:busij phon}
		
		\ex	\tit{b-ač-ij} (\tsc{pfv}) \sqt{smear, spread} \tab \tit{ačː-ij} \sqt{strike, hit onself} 	\label{ex:bacij phonA}
		
		\ex	\tit{b-ac-ij} (\tsc{pfv}) \sqt{plough} \tab \tit{acːi-j} \sqt{to the uncle} (uncle\tsc{-dat})	\label{ex:bacij phonB}
		\ex	\tit{het-i-j} \sqt{for that} (\tsc{dem-obl-dat}) \tab \tit{hetːi} \sqt{those}	\label{ex:hetij phonC}
		
	\end{xlist}
\end{exe}

Geminate fricatives are not always easy to identify because fricatives can be tense in emphatic pronunciation. But geminate stops and affricates are clearly audible as such, because there is a significant difference in the closure duration between singletons and geminates. Gemination can probably be analyzed as a difference between lax and tense \isi{consonants}, but the exact phonetic properties of geminates still need to be clarified by future research.

In addition to their occurrence in stems, geminates occur at morpheme boundaries (see \refsec{ssec:Gemination and degemination} below). A few sonorants can also occur as tense \isi{consonants} within roots (/n/, /m/, /l/, /r/, and /w/) and/or at morpheme boundaries, but their phonemic status needs further clarification. Only geminates of /n/, /r/, and /l/ are found in native items \refex{ex:gemination C1 phon}; the other sonorants are only found in loan words \refex{ex:gemination C2 phon}.

\begin{exe}
	\ex	\label{ex:gemination C1 phon}
	%\begin{xlist}
	\TabPositions{11em}
		\tit{t'unneq} \sqt{basket}	 \tab \tit{=malle} (emphatic \isi{particle})	\\
		\tit{-lla\slash -la} (\isi{genitive} suffix)	\tab \tit{-lle\slash -le} (adverbialzing suffix)	\\
		\tit{urra} \sqt{foreign}	
	%\end{xlist}

	\ex	\label{ex:gemination C2 phon}
	%\begin{xlist}
	\TabPositions{11em}
			\tit{Allah} \sqt{Allah}	 \tab 	\tit{amma} \sqt{but}	\\
			\tit{sːurrat} \sqt{picture}	 \tab 	\tit{Maˁħaˁmma} (male personal name)		
	%\end{xlist}
\end{exe}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Vowel inventory}\label{sec:Vowel inventory}\largerpage

Sanzhi has four plain vowels and three pharyngealized vowels, of which one (\tit{iˁ}) is very rare and whose phonemic status needs further clarification. Pharyngealized vowels and \isi{pharyngealization} are treated in \refsec{sec:Pharyngealization}. The vowels \tit{i}, \tit{e} and \tit{u} have lax and tense variants, whose distribution is not entirely clear. \reftab{tab:The vowel inventory of Sanzhi Dargwa} shows the vowel inventory with the orthographic symbols used in this grammar. 


There is a long vowel [aː], which is not phonemic, but occurs relatively frequently. It shows up only as sequences of homorganic vowels at morpheme boundaries \refex{ex:long monophthongs phon} (most often in negated verb forms), and occasionally as a stressed variant of short vowels. Long vowels mostly occur in open syllables, but can occasionally also be found in closed syllables. The negative present-tense copula-auxiliary normally has a short vowel, but when it is used as existential or \isi{locational copula}, the first vowel becomes long \refex{ex:akkuu akkuu phon}.

\begin{exe}
	\ex	\label{ex:long monophthongs phon}
	\begin{xlist}
		\ex	\tit{aːgur} < \tit{a-ag-ur} \sqt{did not go} (\tsc{neg-}go\tsc{.pfv-pret})	\label{ex:aagur phon}
		\ex	\tit{aːčːib} < \tit{a-ačː-ib} \sqt{did not get} (\tsc{neg-}get\tsc{.pfv-pret})	\label{ex:aciib phon}
		\ex	\tit{čiaːžib} < \tit{či-a-w-až-ib} \sqt{did not see} (\tsc{spr-neg-m-}see\tsc{.pfv-pret})	\label{ex:ciawazib phon}
		\ex	\tit{b-aːkːu} \sqt{does not exist} vs. \tit{akːu} \sqt{is not} (\tsc{cop.neg.prs})	\label{ex:akkuu akkuu phon}
	\end{xlist}
\end{exe}

The long high front vowel [iː] is rarely found when spatial \is{preverb}preverbs are prefixed to some verbs having [i] as stem vowel (see \refsec{ssec:Sequences of identical vowels} below for examples).\pagebreak

\begin{table}
	\caption{The vowel inventory of Sanzhi Dargwa}
	\label{tab:The vowel inventory of Sanzhi Dargwa}
	\begin{tabular}{lccc}
		\lsptoprule
			{}	&	front			&	central	&	back\\
		\midrule
			high	&	/i/	&	{}		&	/u/, /uˁ/\\
			{}	&	[ı], [i]; [ıˁ], [iˁ]	&	{}		&	[u], [ʊ]; [ʊˁ]\\
			{}	&	\tit{i; iˁ}			&	{}		&	\tit{u; uˁ}\\[2mm]

			mid	&	/e/			&	{}		&	{}\\
			{}	&	[ε], [e]			&	{}		&	{}\\
			{}	&	\tit{e}			&	{}		&	{}\\[2mm]

			low	&	{}			&	/a/; /aˁ/	&	{}\\
			{}	&	{}			&	[a]; [aˁ]	&	{}\\
			{}	&	{}			&	\tit{a; aˁ}	&	{}\\
		\lspbottomrule
	\end{tabular}
\end{table}

Sanzhi also has four diphthongs [ʊɪ̯], [aɪ̯], [εɪ̯], and [aʊ̯] that can be analyzed as consisting of two phonemes, a vowel, and a semivowel. Examples are given in \refex{ex:diphthongs phon}.

\begin{exe}
	\ex	\label{ex:diphthongs phon}
	%\begin{xlist}
	\TabPositions{20em}
		\tit{čuj} \sqt{for themselves} (\tsc{refl.pl.dat})	\tab \tit{nejg} \sqt{milk}	\\
		\tit{ʁaj} \sqt{word, talk}		\tab  \tit{alaw} \sqt{around}	\\
		\tit{caw} \sqt{is} (\tsc{cop.prs.m}); \sqt{himself} (\tsc{refl.sg.m})	
	%\end{xlist}
\end{exe}




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Syllable and word structure}
\label{sec:Syllable and word structure}

The minimal syllable consists of a single vowel. Initial vowels are always preceded by a non-phonemic glottal stop not indicated in the \isi{orthography}. The syllables in monomorphemic native words are V, VC, VCC, CV, CVC and CVCC. In other words, syllables never have complex onsets, but can have complex codas. The general \isi{syllable structure} is shown in \refex{ex:syllable structure A phon}.

\begin{exe}
	\ex	(C)V(C)(C) \label{ex:syllable structure A phon}
\end{exe}

In the onset, every consonantal phoneme can occur (see \reftab{tab:Distribution of consonants@A} above for examples), whereas in the coda not all \isi{consonants} are allowed. Note, however, that simple underived verbs have stronger restrictions because they can basically only have /l/ and /r/ in the onset as well as the pharyngeal stop /ʡ/ (in addition to \isi{gender} exponents and \isi{consonants} used in the deixis/elevation \is{preverb}preverbs, see \refsec{sec:The structure underived verbal stems} for more details on the structure of verbs). Labialized \isi{consonants} in syllable-final position are rarer than in syllable-initial position, but they are attested. Ejective \isi{consonants} are also found (\reftab{tab:Distribution of consonants@A}). By contrast, geminate (i.e. tense) \isi{consonants} are prohibited in the coda of syllables. Thus, \is{geminate consonant}geminate consonants in roots that happen to occur at the end of syllables in morphologically complex words, for instance after suffixation, are regularly shortened (see \refsec{ssec:Gemination and degemination} for examples). The nucleus consists of one vowel, which under certain circumstances can be long (\refsec{sec:Vowel inventory}). The minimal syllable (and word) consists of the nucleus only \refex{ex:syllable V phon}. Some words can be seen as containing diphthongs, but diphthongs are analyzed as a sequence of a vowel and a semivowel. The most frequent syllable type is CV \refex{ex:syllable CV phon}, but VC \refex{ex:syllable VC phon} and CVC \refex{ex:syllable CVC phon} are also relatively common. By contrast, syllables of the type V are relatively rare, and \tit{u} \sqt{you} is the only native word that consists just of a minimal syllable.

\begin{exe}
\TabPositions{11em}
	\ex	V	\label{ex:syllable V phon} \\
		\tit{u} \sqt{you}	\tsc{2sg} \tab \tit{\tbf{a}.law} \sqt{around}	
\end{exe}	

\begin{exe}
\TabPositions{11em}
	\ex	CV	\label{ex:syllable CV phon} \\
		\tit{šːi} \sqt{village}	\tab \tit{qu} \sqt{field}	\\
		\tit{χːʷe} \sqt{dog} 	\tab\tit{ʁuˁ.ra} \sqt{hare}		
		\tit{a.\textbf{ba}} \sqt{mother}	\tab \tit{du.ra.zi} \sqt{threshing floor}	
\end{exe}	

\begin{exe}
\TabPositions{11em}
	\ex	VC	\label{ex:syllable VC phon} \\
		\tit{at} \sqt{to / for you}	(\tsc{2sg.dat}) \tab \tit{\tbf{eb}.la} \sqt{in spring}	
\end{exe}

\begin{exe}
\TabPositions{11em}
	\ex	CVC	\label{ex:syllable CVC phon} \\
		\tit{dus} \sqt{year}	\tab 	\tit{ʁaj} \sqt{word, talk}	\\
		\tit{mi.\tbf{riqʷ}} \sqt{worm}	\tab 	\tit{ʡaˁ.\tbf{jar}} \sqt{dance}	\\
		\tit{\tbf{ʡaˁn}.čːi} \sqt{earth, clay}	\tab 	\tit{qaˁš.qaˁr} \sqt{scab}	
\end{exe}

As mentioned in \refsec{sec:Vowel inventory}, there are no phonemic long vowels. Long vowels occasionally show up at morpheme boundaries or when the vowels are stressed or emphasized.

The only types of superheavy syllables are VCC \refex{ex:syllable VCC phon} and CVCC \refex{ex:syllable CVCC phon}, with only sonorants (/r/, /l/, /n/, /m/, /j/) and /b/ permitted in the position of the first consonant in the coda. Up to now I found only one exceptional noun that has a fricative before the second consonant, this being \tit{q'ast} \sqt{aim, intention, plan}. This noun is a loan ultimately from Arabic (\tit{qaṣ̊d}); in Standard Dargwa its form is \tit{q'as}. The syllable-final \isi{consonants} of superheavy syllables can only be plain stops, fricatives, or affricates including \is{ejective}ejectives, geminates, and labialized stops (i.e. obstruents). Although they are mostly voiceless, there are also a few examples of voiced fricatives in the final position of (C)VCC syllables \refex{ex:syllable VCC phon}, \refex{ex:syllable CVCC phon}.

\begin{exe}
\TabPositions{13em}
	\ex	VCC	\label{ex:syllable VCC phon}\\
		\tit{ims} \sqt{moth}	\tab 		\tit{\textbf{alχ}.ni} \sqt{saw}	\\
		\tit{arʁ} \sqt{weather}	\tab 		\tit{irk} \sqt{threshing board}	
\end{exe}

\begin{exe}
\TabPositions{13em}
	\ex	CVCC	\label{ex:syllable CVCC phon}\\
		\tit{laˁbz} \sqt{mortar}	\tab 		\tit{daˁrqʷ} \sqt{barn}	\\
		\tit{nejg} \sqt{milk} 	\tab 		\tit{laˁmc'} \sqt{lightning}	\\
		\tit{kabc} \sqt{skin, fell}	\tab 		\tit{\textbf{c'ult}.mi} \sqt{plum}	\\
		\tit{jebš} \sqt{base}	\tab 		\tit{t'ult'} \sqt{bread}	\\
		\tit{b-ark} \sqt{inside}	\tab 		\tit{b-arx} \sqt{direct, straight} 	\\
		\tit{ku.\textbf{bart}} \sqt{pressed dung}	\tab 		\tit{qːuˁš.\textbf{tːunk'}} \sqt{rolling pin}	\\
		\tit{q'ast} \sqt{target, intention, idea}	
\end{exe}
 
There are no native words with syllable-initial consonant clusters. Consonant clusters in (older) loans are broken up by insertion of \isi{epenthetic vowels} either between initial consonant clusters or before them. In the first case, the vowels vary and are often identical to the following vowel, as the first three words in \refex{ex:epenthesis phon} show. In the second case, the vowel is /i/, as in the last two words:

\begin{exe}
	\ex	\label{ex:epenthesis phon}
	\TabPositions{11em}
	\tit{purust'in} \sqt{bed sheet}	\tab 	< Russian \textit{prostynja}	\\
		\tit{kːalas} \sqt{class}	\tab 	< Russian \textit{klass}	\\
		\tit{kiniga} \sqt{book}		\tab 	< Russian \textit{kniga}	\\
		\tit{ispirt} \sqt{alcohol}	\tab < Russian \textit{spirt} \\
				\tit{ispakulan} \sqt{speculator} \tab  < Russian \textit{spekuljant} 		
\end{exe}

Another possibility is to apply metathesis, though this process is very rare, for example Russian \tit{brigadir} > Sanzhi \tit{birgadir} \sqt{brigadier}.

The minimal word (i.e. free root) has the shape V (see the example in \refex{ex:syllable V phon} above). Minimal bound roots seem to consist of a single consonant and are only found among verbs. Examples are \tit{ha-ʔ-} (\tsc{pfv}) \sqt{say} and \tit{ka-xʷ-}\slash\tit{ha-xʷ-} (\tsc{pfv}) \sqt{pour, add}. These verbs obligatorily contain \is{preverb}preverbs and the vowel can be analyzed as either belonging to the \isi{preverb} (which results then in the monosegmental verb stems) or to the verbs, or two both (\tit{ha-} + \tit{aʔ-} > \tit{haːʔ-} > \tit{haʔ-}).


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Pharyngealization}
\label{sec:Pharyngealization}

The most frequent pharyngealized vowel is /aˁ/, but /uˁ/ is also relatively common,\linebreak whereas [iˁ] is restricted to very few words. The vowel /aˁ/ has phonemic status in Sanzhi as the following minimal pairs and minimal oppositions show \refex{ex:pharyngealization minimal pairs A phon@A}.

\begin{exe}
	\ex	\label{ex:pharyngealization minimal pairs A phon@A}
	\begin{xlist}
\TabPositions{13em}
		\ex	\tit{šaˁm} \sqt{candle} \tab \tit{šam} \sqt{one year old ram}
		\ex	\tit{qːaˁp} (\isi{preverb}) \sqt{twitch} \tab \tit{qːap} \sqt{sack}
		\ex	\tit{b-aˁʡ} \sqt{leaf, side, face} \tab \tit{b-aʔ} \sqt{end, beginning, edge}
		\ex	\tit{waˁħ} (or \tit{wah}) (\isi{interjection}) \tab \tit{w-ah} \sqt{owner} (masc. singular)
		\ex	\tit{ʡaˁʁʷa-l} \sqt{fat} (fat\tsc{-advz}) \tab \tit{aʁʷ-al} \sqt{four} (four\tsc{-num})
		\ex	\tit{χːaˁb} \sqt{grave} \tab \tit{urχːab} \sqt{mill}
	\end{xlist}
\end{exe}

The vowel /uˁ/ is far less frequent than /aˁ/, and thus I so far have found only one minimal pair and only a few examples of minimal oppositions in which the pharyngealized vowels only occur after uvular and pharyngeal sounds \refex{ex:pharyngealization minimal pairs B phon@A}.

\begin{exe}
	\ex	\label{ex:pharyngealization minimal pairs B phon@A}
	\begin{xlist}
	\TabPositions{15em}
		\ex	\tit{ruˁqː-uˁl} / \tit{ruˁqː-ul} \sqt{educating} \tab \tit{r-uqː-ul} \sqt{bringing}\\
		(educate\tsc{-icvb}) \tab (\tsc{f-}bring\tsc{-icvb})
		\ex	\tit{ʁuˁb-e} \sqt{potatoes} (potato\tsc{-pl}) \tab \tit{qːajʁu-be} \sqt{sorrows} (sorrow\tsc{-pl})
		\ex	\tit{ʡaˁχːuˁl} \sqt{guest}		\tab \tit{duχːu-l} \sqt{cleverly} (clever\tsc{-advz})
		\ex	\tit{ʡuˁla} \sqt{wheel}		\tab \tit{ul-la} \sqt{eye's} (eye\tsc{-gen})
	\end{xlist}
\end{exe}

There are a few words that seem to have a pharyngealized high front vowel, e.g. \tit{b-iħ-iˁb} \sqt{they fought} (\tsc{hpl-}fight\tsc{.pfv-pret}), \tit{w-irʡ-iˁb} \sqt{(they) betrayed him} (\tsc{m-}betray\tsc{.pfv-pret}), \tit{b-iˁʡ-iˁj} `steal' (\tsc{n-}steal\tsc{.pfv-inf}), \tit{čːiˁħri} (village name). However, speakers are uncertain about the presence of [iˁ] in Sanzhi words. Furthermore, I do not have any \mbox{(near-)}\linebreak minimal pairs with pharyngealized and non-pharyngealized high front vowels, and thus further research is needed.

The vast majority of pharyngealized vowels occur in the adjacency of the uvular or pharyngeal \isi{consonants} (see \reftab{tab:Distribution of consonants@A}). When pharyngealized vowels occur in roots that contain those \isi{consonants}, the vowels most frequently follow the \isi{consonants}, but can also precede them \refex{ex:uvular pharyngealphon1}. The respective \isi{consonants} are /q/, /q'/, /qː/, /χ/, /ʁ/, /χː/, /ʡ/, /ħ/ for /aˁ/ and /uˁ/, and for /aˁ/ also the labialized \isi{consonants} /qʷ/, /q'ʷ/, /χʷ/, and /ʁʷ/. The remaining uvular and pharyngeal \isi{consonants} (/qːʷ/, /χːʷ/) are in general rare and I have not found any words that contain both the \isi{consonants} and pharyngealized vowels.

\begin{exe}
	\ex	\label{ex:uvular pharyngealphon1}
\TabPositions{15em}
\tit{ʡuˁrʡ-e} \sqt{chickens} (chicken\tsc{-pl})	\tab \tit{ʡaˁnčːi} \sqt{clay, earth}	\\
\tit{q'aˁlči} \sqt{foot}	\tab \tit{ʁuˁc} \sqt{arrow}	\\
\tit{naˁq'iš} \sqt{drawing}	\tab \tit{q'ʷaˁl} \sqt{cow}	\\
\tit{daˁrqʷ} \sqt{barn}	
\end{exe}

The pharyngeal stop /ʡ/ cannot be followed by non-pharyngealized /a/ or /u/, but only by non-pharyngealized /e/ or /i/, that is */ʡa/ and */ʡu/ \refex{ex:uvular pharyngealphon1}. And the pharyngealized vowels /aˁ/ and /uˁ/ are never followed by the glottal fricative /h/, but only by the pharyngeal fricative /ħ/, that is */aˁh/, */uˁh/.

Nevertheless, pharyngeal /aˁ/ and to a lesser extent /uˁ/ can also be found in stems that do not contain uvular or pharyngeal phonemes \refex{ex:not uvular phon} (see also the first minimal pair in \refex{ex:pharyngealization minimal pairs A phon@A} above).

\begin{exe}
	\ex	\label{ex:not uvular phon}
\TabPositions{12em}
		\tit{naˁs} \sqt{dirt}	\tab 		\tit{baˁs} \sqt{argument}	\\
		\tit{laˁbz} \sqt{mortar}	\tab 		\tit{čaˁč} \sqt{haircut}	\\
		\tit{t'uˁ} \sqt{leg}	\tab 		\tit{čaˁt} \sqt{mud}	\\
		\tit{jaˁlči} \sqt{worker}	\tab 		\tit{šuˁra} \sqt{puddle}	

\end{exe}

There are a \isi{number} of words that contain two pharyngealized vowels. The vowels can be either identical (i.e. both vowels are /aˁ/ or both vowels are /uˁ/ or they are /aˁ/ and /uˁ/ in either order \refex{ex:two pharyngealized vowels phon}.

\begin{exe}
	\ex	\label{ex:two pharyngealized vowels phon}
\TabPositions{18em}
		\tit{muˁʡaˁlim} \sqt{teacher}	\tab 	\tit{daˁquˁpːe} \sqt{wounds}	\\
		\tit{qaˁjquˁjte} \sqt{jaw}		\tab 	\tit{ʡaˁχːuˁl} \sqt{guest}	\\
		\tit{q'aˁq'aˁ} \sqt{basin}		\tab 	\tit{naˁqaˁ} \sqt{oat}			\\
		\tit{ʡaˁrʡaˁ} \sqt{chicken}		\tab 	\tit{daˁrχaˁ} \sqt{evening}	\\
		\tit{qːuˁlqːuˁ-l} \sqt{by means of a/the scythe} 	\tab \tit{qːuˁnqːuˁpːe} \sqt{noses}	 \\
		(scythe\tsc{-erg}) 
\end{exe}

There is also some variation with those words that contain two pharyngealized vowels, in the sense that some speakers pharyngealize only one vowel whereas others pharyngealize both. The vowel that is optionally pharyngealized can be the first \refex{ex:pharyngealization of first vowel or both vowels} or the second vowel \refex{ex:pharyngealization of second vowel or both vowels}.

\begin{exe}
	\ex	{\isi{pharyngealization} of first vowel or both vowels}\label{ex:pharyngealization of first vowel or both vowels}
	\begin{xlist}
		\ex	\tit{puˁšːuˁk'}\slash\tit{puˁšːuk'} \sqt{blister}
		\ex	\tit{maˁlʡuˁn}\slash\tit{maˁlʡun} \sqt{snake}
		\ex	\tit{ʡuˁruˁs}\slash\tit{ʡuˁrus} \sqt{Russian}
	\end{xlist}

	\ex	{\isi{pharyngealization} of second vowel or both vowels}\label{ex:pharyngealization of second vowel or both vowels}
	\begin{xlist}
		\ex	\tit{daˁʡaˁna}\slash\tit{daʡaˁna} \sqt{secret, secretly}
		\ex	\tit{durħuˁ}\slash\tit{duˁrħuˁ} \sqt{boy, son}
	\end{xlist}
\end{exe}

I found very few words that contain only one vowel that can optionally be pharyngealized \refex{ex:optional pharyngealization vowels phon}.

\begin{exe}
	\ex	\label{ex:optional pharyngealization vowels phon}
	\begin{xlist}
		\ex	\tit{duħi}\slash\tit{duˁħi} \sqt{snow}
		\ex	\tit{zaˁnʁ}\slash\tit{zanʁ} \sqt{ring} (\isi{ideophone})
		\ex	\tit{čaˁʁir}\slash\tit{čaʁir} \sqt{wine}
	\end{xlist}
\end{exe}

There are also two derivational suffixes containing pharyngealized vowels. The suffixes are \tit{-q'aˁ} and \tit{-uˁq'}. They are not productive and derive \isi{agent} \isi{nouns} from other \isi{nouns}, infinitives, and parts of \is{compound verb}compound verbs form (\refsec{ssec:Agent nouns with -q'aˁ, -uˁq' and -kar}). These suffixes do not have allomorphs with plain vowels.

Pharyngealization is a suprasegmental feature that spreads to inflectional prefixes and suffixes, even in those words that do not contain pharyngealized vowels in the root, but uvular/pharyngeal \isi{consonants}. Only those prefixes and suffixes are affected that start with the vowels /a/ and /u/ such that in affixes only the pharyngealized vowels /aˁ/ and /uˁ/ occur, but no other pharyngealized vowels. Other affixes that contain the same vowels but start with a consonant do not have pharyngealized variants, for instance the vowel in the local \isi{participle} suffix \tit{-na} cannot be pharyngealized \refex{ex:pharyngealization na phon}. 

\begin{exe}
	\ex	\label{ex:pharyngealization na phon}
\TabPositions{12em}
	\tit{guči d-urq-aˁdi} \sqt{I gathered} \tab vs. \tit{guči d-urq-na} \sqt{the place of gathering}
\end{exe}

In the case of the \isi{negation} prefix \textit{a-} this leads to a long pharyngealized vowel (see \refsec{ssec:Sequences of identical vowels} for long vowels resulting from sequences of identical vowels):

\begin{exe}
	\ex	\label{ex:pharyngealization long vowel}
\begin{xlist}
\ex	\tit{či-aˁ-aˁħ-un} > \tit{či-aˁːħ-un} \sqt{did not fly on (something)} (\tsc{spr-neg-}fly\tsc{.pfv-pret})
\ex	\tit{aˁ-w-aˁq-ib} > \tit{aˁ-aˁq-ib} > \tit{aˁːq-ib} \sqt{did not hit} (msc.) (\tsc{neg-}hit\tsc{.pfv-pret})
	\end{xlist}
\end{exe}	


Furthermore, only affixes in immediately preceding or following syllables are affected. Pharyngealization does not spread over the entire word. For \isi{nouns} the suffixes containing pharyngealized vowels are the plural and oblique plural suffixes, as well as one suffix deriving actions \isi{nouns} (\tit{-a}; \refsec{ssec:Action and event nouns with -uti and -a}). For verbs, the suffixes can be derivational (the \isi{causative} suffix, the spatial \is{preverb}preverbs) or inflectional (\isi{negation} prefixes \tit{a-} and \tit{ma-}, various TAM suffixes). Examples are provided in \refex{ex:pharyngealization spread phon}.\pagebreak

\begin{exe}
	\ex	\label{ex:pharyngealization spread phon}
\begin{xlist}
		\ex \tit{qːuˁnqː-uˁpːe} \sqt{noses} (nose\tsc{-pl})	
		\ex \tit{baliqː-aˁ-lla} \sqt{of the fish} (fish\tsc{-obl.pl-gen})
		\ex \tit{ruˁrq-uˁl} \sqt{boiling} (boil\tsc{-icvb})	
		\ex \tit{b-iħ-aˁq-ib} \sqt{made fight} (\tsc{hpl-}fight\tsc{-pfv-caus-pret}) 
		\ex \tit{b-iʡ-uˁn} \sqt{stole} (\tsc{n-}steal\tsc{.pfv-pret})	
		\ex	\tit{b-aˁq-aˁjaˁ} \sqt{hit it!} (\tsc{n-}hit\tsc{.pfv-imp.pl})
\end{xlist}
\end{exe}

The pharyngealized articulation associated with the vowel is maintained when the vowel changes, that is when there is \isi{vowel mutation} \tit{a} > \tit{u}, as, for instance, with plural forms of some \isi{nouns} \refex{ex:vowel mutation plural phon}.

\begin{exe}
	\ex	\label{ex:vowel mutation plural phon}
	\begin{xlist}
		\ex	\tit{ʡaˁrʡaˁ} \sqt{chicken} > \tit{ʡuˁrʡ-e} \sqt{chickens} (chicken\tsc{-pl})
		\ex	\tit{q'ʷaˁl} \sqt{cow} > \tit{q'uˁl-e} \sqt{cows} (cow\tsc{-pl})
	\end{xlist}
\end{exe}

There is one verb \sqt{go} that occurs without a root vowel when prefixes are attached and with a root vowel that can be pharyngealized or plain otherwise. The suffixes used with this verb are obligatorily pharyngealized \refex{ex:go pharyngealization phonA}, \refex{ex:go pharyngealization phonB}, whereas for prefixes \isi{pharyngealization} is optional \refex{ex:go pharyngealization phonC}.

\begin{exe}
	\ex	\label{ex:go pharyngealization phon}
	\begin{xlist}
		\ex	\tit{maˁ-q'-aˁtːa} \sqt{do not go!} (\tsc{proh-}go\tsc{-proh.sg}) \label{ex:go pharyngealization phonA}
		\ex	\tit{b-uq'-aˁq-ij} \sqt{to make it go} (\tsc{n-}go\tsc{-caus-inf}) \label{ex:go pharyngealization phonB}
		\ex	\tit{saˁ-q'-aˁn} vs. \tit{sa-q'-aˁn} \sqt{going} (\tsc{hither}-go\tsc{-ptcp}) \label{ex:go pharyngealization phonC}
	\end{xlist}
\end{exe}

At least with some affixes, \isi{pharyngealization} is optional, and one can find one and the same inflected word form with and without affixes that contain pharyngealized vowels \refex{ex:optional pharyngealization phon}.

\begin{exe}
	\ex	\label{ex:optional pharyngealization phon}
	\begin{xlist}
	\TabPositions{13em}
		\ex	\tit{b-aˁħ-uˁn-ce} vs. \tit{b-aˁħ-un-ce} 		\tab \sqt{wet} (\tsc{n-}become.wet\tsc{.pfv-dd.sg})
		\ex	\tit{guči b-aˁq-aˁraj} vs. \tit{b-aˁq-araj}	\tab \sqt{gather} (gather \tsc{hpl-}assemble\tsc{-subj})
		\ex	\tit{kaˁ-q-aˁja!} vs. \tit{ka-q-aˁja!} 		\tab \sqt{drag!} (\tsc{down}-drag\tsc{.pfv-imp.pl})
	\end{xlist}
\end{exe}




Pharyngealization includes loan words, even recent borrowings from Russian \refex{ex:loan pharyngealization phon}, which are not pharyngealized in the donor language. It is even noticeable when (older) Sanzhi people speak Russian.

\begin{exe}
	\ex	\label{ex:loan pharyngealization phon}
	\TabPositions{12em}
		\tit{čaˁj} \sqt{tea}	\tab 	< Russian \textit{čaj}	 \\
		\tit{šaˁbk'a} \sqt{hat}	\tab 	< Russian 	\textit{šapka} \\
		\tit{ʡaˁšibkːa} \sqt{mistake}	\tab 	< Russian \textit{ošibka}	 \\	
		\tit{šljaˁp'a} \sqt{hat} \tab 	< Russian	\textit{šljapa} \\
		\tit{ʡaˁčkːabe} \sqt{glasses} \tab 	< Russian	\textit{očki} \\
		\tit{luˁkːujte} \sqt{lungs} \tab 	< Russian	\textit{legkie} 
\end{exe}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Word stress}
\label{sec:Word stress}

Stress is not a very prominent category in Sanzhi Dargwa. The \isi{stress} is quite weak and the \isi{stress} properties of words are very hard to determine, but they seem to be largely lexicalized. There are a few examples of minimal pairs or near minimal pairs that differ only in \isi{stress} \refex{ex:stress minimal pairs phon}.

\begin{exe}
	\ex	\label{ex:stress minimal pairs phon}
	\begin{xlist}
\TabPositions{14em}
		\ex	\tit{búk’ul} \sqt{freezing}		\tab \tit{b-uk’úl} \sqt{thin, slender}
		\ex	\tit{sːála} \sqt{wedge}		\tab \tit{salá} \sqt{in front}		
		\ex	\tit{ákːʷar} \sqt{without} (postposition) \tab \tit{akːʷár} \sqt{not being} (\isi{participle})
		\ex	\tit{hána} \sqt{cast iron} \tab \tit{haná} \sqt{now}
	\end{xlist}
\end{exe}

Some affixes attract \isi{stress}, so that the position of \isi{stress} in roots and in inflected word forms of one and the same lexeme may differ. For instance, plural suffixes of \isi{nouns} normally attract \isi{stress} \refex{ex:plural attract stress phon}.

\begin{exe}
	\ex	\label{ex:plural attract stress phon}
	\begin{xlist}
		\ex	\tit{qːap} \sqt{sack} > \tit{qːup-né} \sqt{sack\tsc{-pl}}
		\ex	\tit{kur} \sqt{pit} > \tit{kur-mé} \sqt{pit\tsc{-pl}}
	\end{xlist}
\end{exe}

The factors influencing placement of \isi{stress} require further research.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Phonological and morphophonological alternations}
\label{sec:Phonological and morphophonological alternations}

Sanzhi Dargwa has a variety of phonological and morphophonological alternations that affect vowels and \isi{consonants}. Some of the processes that target vowels result from the fact that hiatus is not allowed, and therefore the underlying word forms have to be changed. A \isi{number} of processes such as \isi{vowel deletion}, alternation in the form of enclitics\slash affixes, \isi{degemination} are syllable repair mechanisms, but others do not serve this function.

Processes affecting vowels are \isi{vowel deletion} (\refsec{ssec:Vowel deletion (vowel syncope)}), alternations in the form of enclitics\slash suffixes (\refsec{ssec:Alternations in the form of enclitics / suffixes}), \isi{glide insertion} (\refsec{ssec:Glide insertion}), \isi{glottal stop insertion} (\refsec{ssec:Glottal stop insertion}), long vowels resulting from sequences of identical vowels (\refsec{ssec:Sequences of identical vowels}), \isi{pharyngealization} and formation of diphthongs (\refsec{ssec:Other general processes affecting vowels}), and \isi{vowel mutation} (\refsec{ssec:Vowel mutation (apophony)}).

Processes affecting mainly \isi{consonants} are \isi{assimilation} (\refsec{ssec:Assimilation}), \isi{palatalization} (\refsec{ssec:Palatalization}), \isi{labialization}\slash \isi{delabialization} (\refsec{ssec:Labialization and delabialization}), gemination (in combination with devoicing) and \isi{degemination} (\refsec{ssec:Gemination and degemination}) (although \isi{labialization} and \isi{delabialization} also have an effect on vowels).


% --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- %

\subsection{Vowel deletion (vowel syncope)}
\label{ssec:Vowel deletion (vowel syncope)}

Vowel deletion (vowel syncope) is one means of avoiding two subsequent vowels at a morpheme juncture. It is mainly found with encliticized negative auxiliaries. There are three types of \isi{vowel deletion} within the domain of verbal morphology. First, sequences of identical vowels might lead to the deletion of one vowel or to \isi{vowel deletion} in combination with mutation (\refsec{ssec:Vowel mutation (apophony)}). Second, the initial vowel of the negative auxiliary is deleted when the auxiliary is used as an \isi{enclitic} \refex{ex:vowel syncope verbs phon}.

\begin{exe}
	\ex	\label{ex:vowel syncope verbs phon}
	\begin{xlist}
		\ex	\tit{biχuble + akːu} > \tit{b-iχ-ub-le=kːu} (\tsc{n-}become\tsc{.pfv-pret-cvb=}\tsc{cop.neg})			
		\ex	\tit{qːuʁace + akːu} > \tit{qːuʁa-ce=kːu} (beautiful-\tsc{dd.sg=}\tsc{cop.neg})		
		\ex	\tit{kabišːible + akːʷadi} > \tit{ka-b-išː-ib-le=kːʷadi}\newline\hspace*{1em}(\tsc{down}\tsc{-n-}put\tsc{.pfv-pret-cvb=}\tsc{cop.pst.neg.1}) 
		\ex	\tit{χʷalle + akːʷi} > \tit{χʷal-le=kːʷi} (big\tsc{-advz=}\tsc{cop.pst.neg})
	\end{xlist}
\end{exe}

Third, there is one verb `go, direct oneself, move' of which the root vowel \tit{u} is deleted when the \isi{gender} agreement is masculine singular and the verbal root is preceded by the deixis/elevation \is{preverb}preverbs or the \isi{negation} prefixes \refex{ex:habulqan phon}. These \is{preverb}preverbs and prefixes end in \tit{a}. The process is accompanied by the \isi{labialization} of the root consonant, that is, the labial feature turns from a vowel feature into a consonantal feature. See \refsec{ssec:Labialization and delabialization} below for more examples with the same verb stem.

\begin{exe}
	\ex	\tit{ha-(w)-ulq-an} > \tit{halqʷan} \sqt{the one that goes upwards}\newline\hspace*{1em}(\tsc{up-m-}direct\tsc{.ipfv-ptcp}) (compare with \tit{ha-b-ulq-an} \tsc{up-n-}direct\tsc{.ipfv-ptcp})	\label{ex:habulqan phon}
\end{exe}

As for nominal morphology, \isi{vowel deletion} that is not caused by sequences of vowels is a regular component of plural formation. The deletion of \tit{a}, \tit{i}, or \tit{u} in the final syllable of mostly disyllabic \isi{nouns} is found with the suffixes \tit{-be}, \tit{-me}, \tit{-re}, \tit{-e}, \tit{-ne}, \tit{-upːe}, \tit{-urbe}, and \tit{-ube} (\refsec{sec:nounnumber}). If the last (usually stem-final) obstruent is a geminate it undergoes \isi{degemination}. Examples are:

\begin{exe}
	\ex	\label{ex:vowel syncope nouns phon}
	\begin{xlist}
		\ex	\tit{šuša} > \tit{šuš-ne} \sqt{bottle}
		\ex	\tit{durħuˁ} > \tit{durħ-ne} \sqt{boy, son},
		\ex	\tit{daˁqaˁ} > \tit{daˁq-uˁpːe} \sqt{wound}
		\ex	\tit{rursːi} > \tit{rurs-be} \sqt{girl, daughter}
		\ex	\tit{murgul} > \tit{murgl-e} \sqt{man}
		\ex	\tit{k'apːur} > \tit{k'apr-e} \sqt{leaf}
	\end{xlist}
\end{exe}


% --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- %

\subsection{Alternations in the form of enclitics\slash suffixes}
\label{ssec:Alternations in the form of enclitics / suffixes}

There are suffixes and enclitics that have allomorphs whose use depends on the \isi{syllable structure} of the item to which the suffixes or enclitics are added. The general rule is that suffixes\slash enclitics consisting of a single consonant are attached to vowel-final words and suffixes\slash enclitics of the form CV or VC to consonant-final words. Relevant suffixes and enclitics are:

\begin{itemize}
	\item	the \isi{ergative} suffix \tit{-l}\slash\tit{-li} (\refsec{sssec:Ergative}), e.g. \textit{du-l} (\tsc{1sg-erg}) vs. \textit{kulpat-li} (family-\tsc{erg})
	\item	the \isi{adverbializer} \tit{-l}\slash\tit{-le} (\refsec{ssec:The adverbializer -le}), e.g. \textit{razi-l} \sqt{happily} vs. \textit{c'aq'-le} \sqt{strongly}
	\item	the emphatic \isi{enclitic} \tit{=n}\slash\tit{=nu} (\refsec{ssec:The enclitic =n(u)}), e.g. \textit{akːu=n} (\tsc{cop.neg.prs=prt}) vs. \textit{le-d=nu} (exist-\tsc{npl=prt})
	\item	the \isi{enclitic} for polar \isi{questions} \tit{=w}\slash\tit{=uw} (\refsec{sec:Simple polar questions and disjunctive polar questions}) \textit{arg-ul=de=w?} (go.\tsc{ipfv-icvb=2sg=q}) \sqt{Are you going?} vs. \textit{le-b=uw?} (exist-\tsc{n=q}) \sqt{Does it exist?}
\end{itemize}

The last \isi{enclitic} has another allomorph \tit{=ew} that only occurs after the \isi{imperfective converb} suffix \tit{-ul}, perhaps to avoid having two identical vowels in two adjacent syllables (although this is generally allowed).

One can argue that the underlying form of the suffixes is the one with the vowel \linebreak (\tit{-li}, \tit{-le}) and that the vowel is deleted in the appropriate contexts. However, there is no phonotactic need for such a deletion and it even goes against the general preference of open syllables in final position (\refsec{sec:Syllable and word structure}).


% --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- %

\subsection{Glide insertion}
\label{ssec:Glide insertion}

Glide insertion represents a regular form of allomorphy. It is only found with the palatal glide \textit{j} and only before certain suffixes or enclitics that start with the vowel \textit{a}. The respective suffixes and enclitics are:

\begin{itemize}
	\item	the derivational suffix used to form the numerals 2\tnd10, 20, as well as 100: \tit{-al}\slash\tit{-jal}, e.g. \tit{aʁʷ-al} \sqt{four} vs. \tit{xu-jal} \sqt{five} 
	\item	the derivational suffix \sqt{X-times} for the formation of \isi{multiplicative numerals} \tit{-na}\slash\tit{-jna}, e.g. \tit{a{\vuvfr}{\lab}-na} \sqt{four times} vs. \tit{{\eppl}a{\pha}-jna} \sqt{three times}
	\item	one of the allomorphs of the \isi{spatial case} suffix of the \tsc{loc}-series: \tit{-a}\slash\tit{-ja}, e.g. \textit{kis-n-a-b} \sqt{in the pockets} (pocket-\textsc{pl}-\textsc{obl}.\textsc{loc}-\textsc{n}) vs. \textit{tusnaqːa-ja-b} \sqt{in the prisons} (prison.\linebreak\textsc{obl.pl-loc-hpl}) 
	\item	optional marker for non-indicative verb forms that  serves as address \isi{particle} for plural addressees \tit{-a}\slash\tit{-ja}, e.g. in the \isi{imperative} plural suffix \tit{-ene-ja} (alternative variant of \tit{-ene}); the \isi{particle} \tit{ma} vs. \tit{ma-ja} `Here, take!'
	\item	the \isi{enclitic} marking content \isi{questions}: \tit{=e} after \isi{consonants} and \tit{=ja} after vowels, e.g. \tit{čina-b=e?} \sqt{Where is it?} (where\tsc{-n=q}) vs. \tit{ča=ja?} \sqt{Who?} (who\tsc{=q})
	\item	the \isi{enclitic} marking embedded \isi{questions} and forming specific \is{indefinite pronoun}indefinite pronouns: \tit{=el} after \isi{consonants} and \tit{=jal} after vowels, e.g. \tit{ča=jal} \sqt{somebody}, \tit{ce=jal}	\sqt{something} vs. \tit{čina-b=el} \sqt{somewhere} 
\end{itemize}


% --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- %

\subsection{Glottal stop insertion}
\label{ssec:Glottal stop insertion}

Another means of avoiding two adjacent vowels is the insertion of a glottal stop. This occurs when spatial \is{preverb}preverbs and \isi{negation} prefixes are added to vowel-initial verbs. In the following examples the glottal stops are given (although they are normally not written in this position).

\begin{exe}
	\ex	\label{ex:glottal stop inversion phon}
	\begin{xlist}
		\TabPositions{3em}
		\ex	\tit{iʔa:}	\tab	\tit{biʔat'un} < \tit{b-i-at'-un} `stuck into' (\tsc{n-in}-stick.\tsc{pfv-pret})
		\ex	\label{Don't take it!Phon} \tit{aʔi:} 	\tab	\tit{maʔisːit} `Don't take it!'	(alternative: \tit{majsːit})	
		\ex	\tit{uʔi:}	\tab	\tit{guʔičib} < \tit{gu-ič-ib} 	`occurred under' (\tsc{sub}-occur.\tsc{pfv.m-pret})
		\ex	\tit{uʔa:}	\tab	\tit{gu-ʔagur}	`went under' (\tsc{sub}-go.\tsc{pfv.pret})\\
		\tab \tab \tit{gu-ʔa-lik'-un} `is not listening' (\tsc{sub-neg}-listen-\tsc{icvb})
		\ex	\label{hitherGOPHON} \tit{aʔa:} 	\tab	\tit{sa-ʔargul} `is coming' (\tsc{hither}-go.\tsc{ipfv.icvb})\\ 
		\tab \tab (alternative: \tit{saːrgul})	
		\ex	\tit{iʔi:}	\tab	\tit{čiʔiʁij} \sqt{understand}	
		\ex	\label{insideADDPHON} \tit{iʔi:}	\tab	\tit{biʔiʁitːe} `you add it inside' (\tsc{n.inside}.add.\tsc{ipfv.2sg})\\
		\tab \tab (alternative: \tit{biːʁitːe})	
	\end{xlist}
\end{exe}

As can be seen in the above examples, in some cases alternative processes can be applied, namely the formation of a long vowel in the case of a sequence of two identical vowels \refex{hitherGOPHON}, \refex{insideADDPHON}, and the change from /i/ to /j/ if the first vowel is /a/ and the second is /i/ \refex{Don't take it!Phon}. The same can be observed for combinations of three vowels or \is{compound verb}compound verbs. For instance, if the verb form \tit{sa-ʔargul} given in \refex{hitherGOPHON} is negated, we get \tit{sa-a-ʔargul}. In principle, it is possible to pronounce all three vowels separately, although this is not the preferred variant in practice. Similarly, in \is{compound verb}compound verbs the first part can be pronounced as a phonologically independent word or the two vowels can fuse across word boundaries, for example \tit{dak'u uqandel} vs. \tit{dak'uːqandel} \sqt{if he appeared}. The precise conditions of \isi{glottal stop insertion} still require clarification.


% --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- %

\subsection{Sequences of identical vowels}
\label{ssec:Sequences of identical vowels}

Long vowels can be the result of a sequence of two identical vowels or of the vowel /i/ plus the semivowel /j/. The latter happens when the \isi{dative} suffix is added to nominals. The only long vowels are /aː/, /aˁː/, /iː/, and in one case after \isi{vowel mutation} /eː/. The emergence of long vowels from two identical vowels is in many cases optional, with the insertion of a glottal stop being the usual alternative. Two identical vowels at morpheme boundaries occur only with verbs, either when the \isi{negation} prefixes \tit{a-} and \tit{ma-} are used or with spatial \is{preverb}preverbs. Note that in particular with the \isi{negation} prefix it is the long vowel that carries the meaning of \isi{negation}. If the two identical vowels would be shortened, the resulting verb form would be identical to the affirmative verb form and the negative meaning would be lost. Examples of the sequence are given in \refex{ex:identical vowels A phon} and for the long pharyngealized low central vowel in \refex{ex:pharyngealization long vowel}.

\begin{exe}
	\ex	\label{ex:identical vowels A phon}
	\begin{xlist}
		\ex	\tit{a-ag-ur} > \tit{aːgur} \sqt{did not go} (\tsc{neg-}go\tsc{.pfv-pret})
		\ex	\tit{sa-arg-ul} > \tit{saːrgul} \sqt{coming here} (\tsc{hither}-go\tsc{.ipfv-icvb})\newline\hspace*{1em}(alternative: \tit{saʔargul}, \refex{hitherGOPHON})
		\ex	\tit{qum a-art-u} > \tit{qum aːrtu} \sqt{does not forget}\newline\hspace*{1em}(forget \tsc{neg-}forget\tsc{.ipfv-prs.3}) 
		\ex	\tit{sa-a-ka-b-išː-ib} > \tit{saːkabišːib} \sqt{did not put down}\newline\hspace*{1em}(\tsc{in.front-neg-down-n-}put\tsc{.pfv-pret})
		\ex	\tit{a-erč-ur} > \tit{eːrčur}\slash\tit{aʔerčur} \sqt{did not saw} (\tsc{neg-}saw\tsc{.pfv-pret})
		\ex	\tit{b-i-iʁ-itːe} > \tit{biːʁitːe} \sqt{you add it in} (\tsc{n-in}-add\tsc{.ipfv-prs.2sg})\newline\hspace*{1em}(alternative: \tit{biʔiʁitːe}, \refex{insideADDPHON})
	\end{xlist}
\end{exe}

Furthermore, the masculine singular \isi{agreement prefix} \tit{w-} can optionally be deleted when it occurs between two identical vowels (and, more generally, before /i/). This process also leads to long vowels [aː] and [iː] \refex{ex:identical vowels B phon}.

\begin{exe}
	\ex	\label{ex:identical vowels B phon}
	\begin{xlist}
		\ex	\tit{a-w-at-ur} > \tit{aːtur} \sqt{did not let him} (\tsc{neg-m-}let\tsc{.pfv-pret})\newline\hspace*{1em}(alternative: \tit{a-w-at-ur})
		\ex	\tit{ma-w-ax-utːa} > \tit{maːxutːa} \sqt{do not go!} (\tsc{proh-m-}go\tsc{-proh.sg})\newline\hspace*{1em}(alternative: \tit{ma-w-ax-utːa})
		\ex	\tit{w-i-a-w-ax-an=de} >  \tit{wiʔaːxande} \sqt{you will not go inside}\newline\hspace*{1em}(\tsc{m-}\tsc{in-neg-m-}go\tsc{-ptcp=2sg}) (alternative: \tit{w-i-a-w-ax-an=de})
		\ex	\tit{či-w-ig-ul=de} > \tit{čiːgulde} \sqt{you see him} (\tsc{spr-m-}see\tsc{.ipfv-icvb=2sg})\newline\hspace*{1em}(alternative: \tit{čiwigulde})
	\end{xlist}
\end{exe}


% --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- %

\subsection{Other general processes affecting vowels: Pharyngealization and formation of diphthongs}
\label{ssec:Other general processes affecting vowels}

Pharyngealization is a frequent process that is attested with verbal and nominal affixes containing \tit{u} or \tit{a}. The \isi{pharyngealization} feature of verbal and nominal stems spreads to the closest prefixes or suffixes, but not to the entire word. The nominal affixes that have pharyngealized allomorphs are the plural suffix \tit{-upːe}, oblique plural suffix \tit{-a} and the suffix \tit{-a} deriving actions \isi{nouns} from verbs. The verbal suffixes are a variety of derivational and inflectional suffixes. See \refsec{sec:Pharyngealization} above for more details and relevant examples.

There are two diphthongs that arise when vowels are followed by the semivowels /w/ and /j/. The diphthongs [aɪ̯] and [aʊ̯], written \tit{aj} and \tit{aw}, are found in a few roots, e.g. \tit{ʁaj} \sqt{word, talk}, \tit{alaw} \sqt{around}; more are given in \refex{ex:diphthongs phon}. They also arise during certain inflectional or derivational processes. The first diphthong is attested with verbs having \tit{i} as the root vowel and consisting only of one consonant, that is, verbs of the structure \tit{(b-)iC(ː)}. They may have or may not have \isi{gender} prefixes. When a spatial \isi{preverb} \tit{ka-}, \tit{ha-},\tit{ sa-}, or the \isi{negation} prefixes (\tit{a-}, \tit{ma-}) are added, the result is \tit{a} + \tit{i} > \tit{aj}. For verbs with \isi{gender} prefixes the process only applies when the \isi{gender} prefix \tit{w-} for masculine singular is omitted, which is always possible for verbs that have the root vowel \tit{i}. The forms with the \isi{gender} prefixes \tit{b-} or \tit{r-} that do not contain the diphthong are given in brackets.

\begin{exe}
	\ex	{\tit{a} + \tit{i} > \tit{aj}} (iC(ː) > jC(ː))\label{ex:a i aj phon}
	\begin{xlist}
		\ex	\tit{ha-(w)-icː-ij} > \tit{hajcːij} \sqt{to stand up} (\tsc{up-m-}stand\tsc{.pfv-inf}) (\tit{ha-b-icː-ij})
		\ex	\tit{a-(w)-iχʷ-ij} > \tit{ajχʷij} \sqt{to not be able} (\tsc{neg-m-}can\tsc{.pfv-inf}) (\tit{a-b-iχʷ-ij})
		\ex	\tit{ma-isː-it} > \tit{majsːit} \sqt{Do not shave!} (\tsc{neg-}shave\tsc{.ipfv-proh.sg})
	\end{xlist}
\end{exe}

This process is optional to some degree. This means that under certain circumstances that need further investigation, the two adjacent vowels \tit{a} and \tit{i} can be pronounced separately, not forming a diphthong. For instance, \tit{majsːit} can alternatively be pronounced \tit{maʔisːit} \refex{Don't take it!Phon}.

The diphthong [aʊ̯] arises when spatial \is{preverb}preverbs or \isi{negation} prefixes with the final vowel \tit{a} are added to verbs with the root vowel \tit{u}. This can be verbs with a \isi{gender} prefix (\tit{b-uC(ː)-}) that are inflected for masculine singular \isi{gender} agreement (\refsec{sec:Gender/number agreement}). The masculine singular prefix \tit{w-} is regularly dropped before verbs with the root vowel \tit{u} (e.g. \tit{uc-ib} \sqt{caught him} vs. \tit{r-uc-ib} \sqt{caught her}), and then the combination of the two subsequent vowels turns into a diphthong that will be written \tit{aw} \refex{ex:a u aw phon}. In the following examples forms with overt \is{agreement prefix}agreement prefixes \textit{b-} or \textit{d-} are given in brackets at the end of the example lines.

\begin{exe}
	\ex	{\tit{a} + \tit{u} > \tit{aw}} (uC(ː) > wC(ː)) \label{ex:a u aw phon}
	\begin{xlist}
		\ex	\tit{tːura-(w)-uq-un} > \tit{tːurawqun} \sqt{he went outside} (\tsc{outside-m-}go\tsc{.pfv-pret}) (vs. \tit{tːura b-uq-un}) 
		\ex	\tit{sa-(w)-uq-un} > \tit{sawqun} \sqt{he came} (\tsc{hither-m-}go\tsc{.pfv-pret}) (vs. \tit{sa-b-uq-un})
		\ex	\tit{gu-sa-(w)-uc-ib} > \tit{gusawcib} \sqt{kept him} (\tsc{down-hither-m-}keep\tsc{.pfv-pret}) (vs. \tit{gu-sa-b-uc-ib})
		\ex	\tit{a-(w)-uq'-idel} > \tit{awq'idel} \sqt{should I not go} (\tsc{neg-m-}go\tsc{-modq}) (vs. \tit{a-d-uq'-idel})
	\end{xlist}
\end{exe}

The same happens to verbs that do not have a \isi{gender} prefix (\tit{uC(ː)-}) when the root is preceded by prefixes ending in \textit{a} (\ref{ex:a u aw phon2}). And again there are exceptions to the rule, e.g. in \textit{sauq'ij} \sqt{go towards, go to meet} the two vowels do not form a diphthong, but are separated by a glottal stop.

\begin{exe}
	\ex	{\tit{a} + \tit{u} > \tit{aw}} (uC(ː) > wC(ː)) \label{ex:a u aw phon2}
	\begin{xlist}
		\ex	\textit{ka-utː-ij} > \textit{kawtːij} \sqt{tear off, rip off} (\tsc{down}-tear.\tsc{ipfv-inf})
		\ex \textit{ha-utː-ij} > \textit{hawtːij} \sqt{pull out, disassemble, take apart} (\tsc{up}-tear.\tsc{ipfv-inf})
		\ex \textit{ha-uχːaq-ij} > \textit{hawχːaqij} \sqt{ignite, set fire} (\tsc{up}-sparkle-\tsc{caus-inf})
	\end{xlist}
\end{exe}
% --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- %

\subsection{Vowel mutation (apophony)}
\label{ssec:Vowel mutation (apophony)}

Vowel mutation is found with inflected \isi{nouns} and verbs. In the case of \isi{nouns}, it is triggered by suffixation, in the case of verbs by prefixation.

The vowel \tit{a} in the final syllable of \isi{nouns} ending in a consonant is raised and backed when one of the plural suffixes \tit{-e}, \tit{-te}, \tit{-be}, and \tit{-re} containing close-mid vowels is added, that is, there is \isi{vowel mutation} \tit{a} > \tit{u} (including \tit{aˁ} > \tit{uˁ}). The process can be accompanied by \isi{delabialization} (\refsec{ssec:Labialization and delabialization} shows examples). There are also one instance each of \tit{e} > \tit{u} and \tit{e} > \tit{i} under the same conditions. Relevant examples are given in \refex{ex:a u e u e i phon}. See \refsec{sec:nounnumber} for more \isi{nouns}.\pagebreak

\begin{exe}
	\ex	{\tit{a} > \tit{u}; \tit{e} > \tit{u}; \tit{e} > \tit{i}}\label{ex:a u e u e i phon}
	\begin{xlist}
		\ex	\tit{qːap} > \tit{qːup-re} \sqt{sacks}
		\ex	\tit{χabar} > \tit{χabur-te} \sqt{stories, news}
		\ex	\tit{nez} > \tit{nuz-be} \sqt{lice}
		\ex	\tit{ʁez} > \tit{ʁiz-be} \sqt{hairs}
	\end{xlist}
\end{exe}

Vowel mutation with verbs occurs when the spatial \is{preverb}preverbs or \isi{negation} prefixes with the final vowel \tit{a} are prefixed. The first type of verbal \isi{vowel mutation} happens with verbs containing the stem vowel \tit{i} that are inflected for masculine singular or lack \isi{gender} \is{agreement prefix}agreement prefixes. The \isi{gender} prefix is dropped and the two vowels merge. Verb forms with overt \isi{gender} prefixes are given in brackets at the end of the example lines for comparison.

\begin{exe}
	\ex	{\tit{a} + \tit{i} > \tit{e}}\label{ex:a i e phon}
	\begin{xlist}
		\ex	\tit{sa-(w)-irʁ-an} > \tit{serʁan} \sqt{the one that comes} (\tsc{hither-m-}come\tsc{-ptcp}) (vs. \tit{sa-b-irʁ-an}) 
		\ex	\tit{a-(w)-irχʷ-ar} > \tit{erχʷar} \sqt{cannot} (\tsc{neg-m-}be.able\tsc{.ipfv-prs.3}) (vs. \tit{a-b-irχʷ-ar})
		\ex	\tit{ka-(w)-irg-an=da} > \tit{kerganda} \sqt{I will sit down} (\tsc{down-m-}be\tsc{.ipfv-ptcp=1}) (vs. \tit{ka-r-irg-an=da})
	\end{xlist}
\end{exe}

This process is optional, but again the circumstances under which alternatives are allowed need to be clarified \refex{ex:does not understand phon}.

\begin{exe}
	\ex	\label{ex:does not understand phon}
	\begin{xlist}
		\ex	\tit{han a-w-irk-u} > \tit{han awirku}\slash\tit{han.erku}\slash\tit{han.aʔirku} \sqt{does not remember\newline\hspace*{1em}him} (remember \tsc{neg-m-}occur\tsc{.ipfv-prs.3})  
		\ex	\tit{a-irʁ-ib=da} > \tit{erʁibda}\slash\tit{aʔirʁibda} \sqt{I did not understand}\newline\hspace*{1em}(\tsc{neg-}understand\tsc{.pfv-pret=1})
	\end{xlist}
\end{exe}
%

The second type of verbal \isi{vowel mutation} happens with verbs that have the stem vowel \tit{e} and lack \isi{gender} \is{agreement prefix}agreement prefixes \refex{ex:a e e phon}. Note that in the first verb given below the \isi{vowel mutation} results in a long vowel because the \isi{negation} prefix \textit{a-} assimilated to the stem vowel and this, in turn, leads to a sequence of two identical vowels, which then becomes a long vowel. This process commonly occurs when the \isi{negation} prefix is added to verbs beginning with the vowel \textit{a} because if the sequence would be shortened, the negated form would be identical to the affirmative form and \isi{negation} could not be expressed \refex{ex:identical vowels A phon}. The same logic applies to \tit{eːrčur} \refex{ex:sawPHON}.

\begin{exe}
	\ex	{\tit{a} + \tit{e} > \tit{e}}\label{ex:a e e phon}
	\begin{xlist}
		\ex	\label{ex:sawPHON} \tit{a-erč-ur} > \tit{aʔerčur}\slash\tit{eːrčur} (\tsc{neg-}saw\tsc{.pfv-pret})
		\ex	\tit{ha-erʔ-ul} > \tit{herʔul} (\tsc{up}-say.\tsc{ipfv-icvb})
	\end{xlist}
\end{exe}

Finally, the combination of spatial \is{preverb}preverbs ending with \tit{i} and a verb without a \isi{gender} prefix and \tit{a} as stem vowel or a following \isi{preverb} \tit{ha-} \sqt{upwards} also leads to \isi{vowel mutation}. In the second case, when two \is{preverb}preverbs combine, then the \isi{vowel mutation} is initiated by the disappearance of the glottal fricative. The affected \is{preverb}preverbs are \tit{či-} \sqt{on} + \tit{ha-} > \tit{če-, kʷi-} \sqt{in the hands} + \tit{ha} > \tit{kʷe-, hitːi-} \sqt{behind, after} + \tit{ha} > \tit{hitːe-, b-i} \sqt{in, inside} + \tit{ha-} > \tit{be-}. We can analyze this process as lowering of the vowel of the second \isi{preverb} \refex{ex:i a e phon}. Again the process is optional and does not occur in slow, careful speech.

\begin{exe}
	\ex	{\tit{i} + \tit{a} > \tit{e}}\label{ex:i a e phon}
	\begin{xlist}
		\ex	\tit{či-ag-ur} > \tit{čegur} \sqt{s/he went} (\tsc{spr-}go\tsc{.pfv-pret})
		\ex	\tit{či-ha-b-išː-ib} > \tit{čebišːib} \sqt{s/he put it up} (\tsc{spr-up}\tsc{-n-}put\tsc{.pfv-pret})
		\ex	\tit{kʷi-ha-b-uc-ib} > \tit{kʷebucib} \sqt{s/he kept it in the hands}\newline\hspace*{1em}(\tsc{in.hands-up-n}-keep\tsc{.pfv-pret})
	\end{xlist}
\end{exe}


% --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- %

\subsection{Assimilation}
\label{ssec:Assimilation}

Progressive \isi{assimilation} occurs with all verbal and nominal suffixes that have initial \tit{l}. The liquid assimilates to a preceding sonorant \tit{n} or \tit{r} \refex{ex:n l nn r l rr phon}. The following suffixes are affected:

\begin{itemize}
	\ex	\isi{genitive} case: \tit{-la} > \tit{-na}\slash\tit{-ra}
	\ex	\tsc{loc}-series (\isi{spatial case}): \tit{-le} > \tit{-ne}\slash\tit{-re}
	\ex	\isi{ergative} case\slash \isi{oblique stem} marker: \tit{-li} > \tit{-ni}\slash\tit{-ri}
	\ex	\isi{perfective converb}\slash \isi{adverbializer}: \tit{-le} > \tit{-ne}\slash\tit{-re}
	\ex	anteriority\slash causality converb \tit{-la} > \tit{-na}\slash\tit{-ra}
\end{itemize}

\begin{exe}
	\ex	{\tit{n} + \tit{l} > \tit{nn}; \tit{r} + \tit{l} > \tit{rr}}\label{ex:n l nn r l rr phon}
	\begin{xlist}
		\ex	\tit{cin-la} > \tit{cinna} \sqt{his\slash her} (\tsc{refl.sg.obl-gen})
		\ex	\tit{tuχtur-li} > \tit{tuχturri} (doctor\tsc{-erg})
		\ex	\tit{b-uč'-un-le} > \tit{buč'unne} \sqt{have read} (\tsc{n-}read\tsc{.pfv-pret-cvb})
	\end{xlist}
\end{exe}

With many words the process is optional, and in careful speech no \isi{assimilation} takes place.


% --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- %

\subsection{Palatalization}\label{ssec:Palatalization}\largerpage

Palatalization of velar \isi{consonants} occurs with verbs when the \isi{causative} suffix \tit{-aq} or suffixes starting with the front vowels \tit{i} \refex{ex:x s xx s phonA}, \refex{ex:g z phon}, \refex{ex:k c kk c k c phonA}, \refex{ex:k c kk c k c phonB} or \tit{e} \refex{ex:x s xx s phonB}, \refex{ex:k c kk c k c phonC} are added, or occasionally when the \isi{masdar} suffix \tit{-ni} is following \refex{ex:k c kk c k c phonD}.

\begin{exe}
	\ex	{\tit{x} > \tit{š}, \tit{xː} > \tit{šː}}\label{ex:x s xx s phon}
	\begin{xlist}
		\ex	\tit{či-ka-b-ixː-a} \sqt{put it on!} (\tsc{spr-down}\tsc{-n-}put\tsc{.pfv-imp.sg})\newline\hspace*{1em}vs. \tit{či-ka-b-išː-ij} \sqt{to put it on} (\tsc{spr-down}\tsc{-n-}put\tsc{.pfv-inf}) \label{ex:x s xx s phonA}
		\ex	\tit{b-ax-ul} \sqt{going} (\tsc{n-}go\tsc{-icvb})\newline\hspace*{1em}vs. \tit{w-aš-e!} \sqt{Go!} (\tsc{m-}go\tsc{-imp.sg}) \label{ex:x s xx s phonB}
	\end{xlist}

	\ex	{\tit{g} > \tit{ž}}\label{ex:g z phon} \\
	%\begin{xlist}
	\tit{b-ug-ul} \sqt{remaining} (\tsc{n-}stay\tsc{-icvb})\newline\hspace*{1em}vs. \tit{b-už-ib} \sqt{remained} (\tsc{n-}stay\tsc{-pret}) 
	%\end{xlist}

	\ex	{\tit{k} > \tit{č}, \tit{kː} > \tit{čː}, \tit{k'} > \tit{č'}}\label{ex:k c kk c k c phon}
	\begin{xlist}
		\ex	\tit{b-uk-ul} \sqt{gathering} (\tsc{n-}gather\tsc{-icvb})\newline\hspace*{1em}vs. \tit{b-uč-ib} \sqt{gathered it} (\tsc{n-}gather\tsc{-pret}) \label{ex:k c kk c k c phonA}
		\ex	\tit{b-ikː-a} \sqt{give it!} (\tsc{n-}give\tsc{.pfv-imp.sg})\newline\hspace*{1em}vs. \tit{b-ičː-ib} \sqt{gave it} (\tsc{n-}give\tsc{.pfv-pret}) \label{ex:k c kk c k c phonB}
		\ex	\tit{er w-erk'-araj} \sqt{in order to look at him} (look \tsc{m-}look\tsc{.pfv-subj})\newline\hspace*{1em}vs. \tit{er w-erč'-e} \sqt{Look!} (look \tsc{m-}look\tsc{.pfv-imp.sg}) \label{ex:k c kk c k c phonC}
		\ex	\tit{b-ebk'-a} \sqt{death} (\tsc{n-}die\tsc{.pfv-nmlz})\newline\hspace*{1em}vs. \tit{b-ebč'-ni} \sqt{death} (\tsc{n-}die\tsc{.pfv-msd}) \label{ex:k c kk c k c phonD}
	\end{xlist}
\end{exe}

When the \isi{masdar} suffix is added the process is optional, at least with some verbs \refex{ex:masdar palatalization phon} (although it occurs when other suffixes are added). However, with a few verbs such as \tit{er b-ik'ʷ-ni} `looking' (look \tsc{-n-}say\tsc{.ipfv-msd}) it is ungrammatical.

\begin{exe}
	\ex	\label{ex:masdar palatalization phon}
	\begin{xlist}
		\ex	\tit{ubč'-ni}\slash\tit{ubk'-ni} (die\tsc{.m.ipfv-msd}) < \tit{b-ubk'-} (\tsc{n-}die\tsc{.ipfv-}) 
		\ex	\tit{b-arč-ni}\slash\tit{b-ark-ni} (\tsc{n-}find\tsc{.pfv-msd}) < \tit{b-arkː-} (\tsc{n-}find\tsc{.pfv-})
	\end{xlist}
\end{exe}


% --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- %

\subsection{Labialization and delabialization}
\label{ssec:Labialization and delabialization}

There are two instances of \isi{labialization} of stops triggered by the round vowel \tit{u}. In the first instance, a preceding vowel is lost and the loss is compensated for by labializing the following stop (another example with the same verb has been provided in \refsec{ssec:Vowel deletion (vowel syncope)} above):

\begin{exe} \label{ex:insidephon2}
	\ex	\tit{w-i-ha-(w)-ulq-an} > \tit{wihalqʷan} \sqt{the one that goes inside}\newline\hspace*{2em}(\tsc{m-in-up-(m)}-go\tsc{.ipfv-ptcp}) (compare with \tit{b-i-ha-b-ulq-an})
\end{exe}

The second instance represents the combination of the two spatial \is{preverb}preverbs \tit{gu-} \sqt{under} and \tit{ha-} \sqt{upwards}. The glottal fricative between the two vowels is lost and the round vowel disappears, leaving the initial stop labialized, that is \tit{gu-ha-} > \tit{gʷa-} \refex{ex:She set up a phon}.

\begin{exe}
	\ex	\label{ex:She set up a phon}
	\gll	c'a	gʷa-b-iq'-un	ca-b\\
		fire	\tsc{from.under.up}\tsc{-n-}set.fire\tsc{.pfv-pret}	 \tsc{cop-n}\\
	\glt	\sqt{(She) set up a fire.}
\end{exe}

De\isi{labialization} is a more widespread and predictable process. It occurs when verbs that contain labialized stem \isi{consonants} take suffixes beginning with the round vowel \tit{u} (i.e. one of the preterite allomorphs \tit{-ub}, \tit{-ur,} or \tit{-un}):

\begin{exe}
	\ex	\label{ex:delabialization A phon}
	\begin{xlist}
		\ex	\tit{b-elk'ʷ-ij} \sqt{write} > \tit{b-elk-un} \sqt{wrote} (\tsc{n-}write\tsc{.pfv-pret})
		\ex	\tit{kaxʷ-ij} \sqt{kill} > \tit{kax-ub} \sqt{killed} (kill\tsc{.pfv-pret})
		\ex	\tit{ergʷ-ij} \sqt{sieve} > \tit{erg-ur} \sqt{sieved} (sieve\tsc{.pfv-pret})
	\end{xlist}
\end{exe}

With \isi{nouns} \isi{delabialization} occurs in the formation of the plural. When the plural suffix or the oblique plural suffix is added to \isi{nouns} that have a vowel \tit{a}\slash\tit{aˁ} in the root that undergoes \isi{vowel mutation} \tit{a}\slash\tit{aˁ} > \tit{u}\slash\tit{uˁ}, then the mutation is accompanied by \isi{delabialization} of a stop that precedes or follows the mutated vowel. Furthermore, plural suffixes containing \tit{u} also trigger \isi{delabialization} of preceding \isi{consonants} when they are added \refex{ex:plural delabialization phon}.

\begin{exe}
	\ex	\label{ex:plural delabialization phon}
\TabPositions{13em}
		\tit{daˁrqʷ} \sqt{barn} > \tit{duˁrq-be} \tab 	\tit{q'ʷaˁl} \sqt{cow} > \tit{q'uˁl-e} \\
		\tit{qːʷaz} \sqt{goose} > \tit{qːuz-re}	\tab 	\tit{mikʷa} \sqt{fingernail} > \tit{mik-upːe} \\
		\tit{χːʷe} \sqt{dog} > \tit{χː-ude}
\end{exe}

Other plural suffixes do not lead to \isi{vowel mutation}, and thus labialized \isi{consonants} are preserved, for example:

\begin{exe}
	\ex	\label{ex:no plural vowel mutation phon}
\TabPositions{13em}
		\tit{kːʷacːa} \sqt{mare} > \tit{kːʷac-ne}	\tab 	\tit{targʷa} \sqt{weasel} > \tit{targʷ-ne} \\
		\tit{gʷagʷa} \sqt{flower} > \tit{gʷagʷne}	\tab 	\tit{žilixʷa} \sqt{saddle} > \tit{žilixʷme} 
\end{exe}


% --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- %

\subsection{Gemination and degemination}
\label{ssec:Gemination and degemination}
Gemination is not a common process, whereas \isi{degemination} is frequent. There is optional gemination in combination with devoicing, which always involves at least one \isi{gender} affix. This process occurs only with the \isi{gender} affixes \tit{b} (neuter singular\slash human plural) and \tit{d} (neuter plural\slash first and second person plural). The two lax voiced \isi{consonants} become tense and devoiced when they are preceded or followed by an identical consonant. This can either be the same \isi{gender} affix or the \isi{past tense} \isi{enclitic} \tit{=de}, the attributive plural suffix \tit{-te} or occasionally when a \isi{preverb} ending in \tit{p} is used in a complex verb (there are no \is{preverb}preverbs ending in \tit{d}). Examples are given in \refex{ex:d dt t phon} and \refex{ex:bp b p phon}. In careful speech the two \isi{consonants} are pronounced individually, and no gemination and devoicing take place.

\begin{exe}
	\ex	{\tit{d} + \tit{d/t} > \tit{tː}}\label{ex:d dt t phon}
	\begin{xlist}
		\ex	\tit{či-d-d-iχ-un} > \tit{čitːiχun} \sqt{(they) tied them} (\tsc{spr-npl-npl-}tie\tsc{.pfv-pret})
		\ex	\tit{le-d=de} > \tit{letːe} \sqt{we were there} (exist\tsc{-npl=pst}) (\tit{le-b=de})
		\ex	\tit{xari-d-te} > \tit{xaritːe} \sqt{the ones down} (down-\tsc{npl-dd.pl}) (\tit{xari-b-te})
	\end{xlist}

	\ex	{\tit{b/p} + \tit{b} > \tit{pː}}\label{ex:bp b p phon}
	\begin{xlist}
		\ex	\tit{gu-b-b-iči-b} > \tit{gupːičib} \sqt{it lost} (\tsc{sub-n-n-}occur\tsc{.pfv-pret})
		\ex	\tit{χːap b-arq'-ib} > \tit{χːapːarq'ib} \sqt{grabbed it} (grab \tsc{n-}do\tsc{.pfv-pret})
	\end{xlist}
\end{exe}

Gemination does not occur when two voiceless \isi{consonants} follow each other, for example \tit{ħaˁžat-te} (need\tsc{-dd.pl}).

Furthermore, a \isi{number} of verbal suffixes such as the \isi{habitual present} suffixes contain geminates, e.g. \tit{-tːe} (\tsc{-2sg.prs}) (\refsec{sec:vis-habitualpresent}). These suffixes are probably diachronically complex in their morphology, but since they synchronically function as entire morphemes that are not further split up, they are not treated here.


Geminates are regularly degeminated when they end up in syllable-final position, because geminates in syllable-final position are prohibited (see \refsec{sec:Consonant inventory}, \refsec{sec:Syllable and word structure}). Therefore, when suffixation leads to resyllabification, then \isi{degemination} takes place, that is, tense \isi{consonants} become lax. Voicing is not affected. Within the nominal morphology we find \isi{degemination} of stops, fricatives, and affricates when the plural suffixes \tit{-be}, \tit{-ne,} and \tit{-me} are added \refex{ex:degemination be ne me phon}.

\begin{exe}
	\ex	\label{ex:degemination be ne me phon}
\TabPositions{14em} 
		\tit{rur.sːi} > \tit{rurs-be} \sqt{girl, daughter}	\tab 	\tit{cːa.cːi} > \tit{cːac-be} \sqt{thorn} \\
		\tit{c'el.tːa} > \tit{c'elt-me} \sqt{gravestone}		\tab 	\tit{e.čːa} > \tit{eč-ne} \sqt{she-goat}
\end{exe}

Similarly, a \isi{number} of \isi{nouns} have underlying geminates (stops and fricatives) in the word-final position that are only pronounced as geminates when suffixes that begin with a vowel (e.g. the plural suffixes \tit{-e} and \tit{-upːe}) are attached  \refex{ex:gemination e upee phon}. In those plural \isi{nouns} the geminates occur in syllable-initial position. By contrast, when the \isi{nouns} are used in the singular or when suffixes that start with \isi{consonants} are added (e.g. the \isi{ergative} suffix \textit{-li}), then the stops and fricatives are degeminated. More examples can be found in \refsec{sec:nounnumber}. In the examples in \refex{ex:gemination e upee phon} first the plural forms are given and then the singular forms.
  

\begin{exe}
	\ex	\label{ex:gemination e upee phon}
\TabPositions{13em} 
		 \tit{juldašːe} > \tit{juldaš} \sqt{friend}	\tab 		\tit{baliqːe} > \tit{baliq} \sqt{fish} \\
		\tit{ʡuˁrusːe} > \tit{ʡuˁrus} \sqt{Russian}	\tab 		\tit{ħaˁšukːe} > \tit{ħaˁšuk} \sqt{pot} \\
		\tit{miriqːʷe} > \tit{miriqʷ}  \sqt{worm}	\tab 		\tit{t'upːe} > \tit{t'up} \sqt{finger} \\
		\tit{qːuˁnqːuˁpːe} > \tit{qːuˁnq}  \sqt{nose}
\end{exe}



Within the verbal system, \isi{degemination} can only occur when consonant-initial suffixes are added to verbal roots that have geminated \isi{consonants}. The only relevant suffixes are the \isi{masdar} suffix \tit{-ni} (or \tit{-ri}) and the \isi{locative participle} \tit{-na}.

\begin{exe}
	\ex	\label{ex:degemination masdar locative phon}
	\begin{xlist}
		\ex	\tit{ha-qː-ij} (\tsc{up}-carry\tsc{-inf}) > \tit{haq-ni} (\isi{masdar})
		\ex	\tit{ka-b-ičː-ij} (\tsc{down-n-}cut.up\tsc{.pfv-inf}) > \tit{kabič-ni} (\isi{masdar})
		\ex	\tit{b-arcː-ij} (\tsc{n-}get.tired\tsc{.pfv-inf}) > \tit{barc-ni} (\isi{masdar})
		\ex	\tit{akːʷ-} (\tsc{cop.neg}) > \tit{akʷ-ni}\slash\tit{akʷ-ri} (\isi{masdar})
		\ex	\tit{b-učː-ij} (\tsc{n-}drink\tsc{.pfv-inf}) > \tit{buč-na} (\isi{locative participle})
	\end{xlist}
\end{exe}
