\chapter{Indicative synthetic verb forms}
\label{cpt:verbs-indicativesynthetic}

Sanzhi Dargwa has only two indicative synthetic verb forms that head independent clauses, the \isi{habitual present} (\refsec{sec:vis-habitualpresent}) and the \isi{habitual past} (\refsec{sec:vis-habitualpast}). They are formed by adding stem augmentation vowels and \isi{person agreement} markers to verbal stems that have \isi{imperfective aspect}. The stem augmentation vowels occur only with first and second person forms and are also used in \is{conditional clause}conditional clauses with synthetic verb forms (\refcpt{cpt:conditionalconcessiveclauses}). They are \tit{u} for \is{intransitive verb}intransitive verbs and \tit{i} for \is{transitive verb}transitive verbs in the \isi{habitual present}, and \tit{a} for all verbs in the \isi{habitual past} (with the exception of the verb \tit{b-aχ-} (\tsc{pfv})\slash\tit{b-alχ-} (\tsc{ipfv}) \sqt{know}, which also has \tit{a} as the stem augmentation in the \isi{habitual present}). The stem augmentation vowels are not separately glossed in the examples, but given together with the person/tense suffixes.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Habitual present}
\label{sec:vis-habitualpresent}

The \isi{habitual present} is formed by adding person suffixes to the augmented stem of imperfective verbs (\reftab{tab:habitualpresent}). The third person has always the suffix \tit{-u}, which can thus be interpreted as a person marker, although it most probably originates from the stem augmentation vowel for \is{intransitive verb}intransitive verbs. Alternatively there is the suffix \tit{-ar} for the third person (see below for a discussion). \reftab{tab:habitualpresent-examples} shows paradigms of three verbs. The \isi{intransitive verb} `say' is given in the female form for singular persons.\largerpage[2]

\begin{table}[H]
	\caption{Person suffixes for the habitual present (without stem augmentation vowels)}
	\label{tab:habitualpresent}
	\small
	\begin{tabularx}{0.40\textwidth}[]{%
		>{\centering\arraybackslash}p{10pt}
		>{\centering\arraybackslash}X
		>{\centering\arraybackslash}X}
		
		\lsptoprule
			{}	&	singular	&	plural\\
		\midrule
			1	&	\multicolumn{2}{c}{\tit{-d}}\\
			2	&	\tit{-tːe}	&	\tit{-tːa}\\
			3	&	\multicolumn{2}{c}{\tit{-u\slash -ar}}\\
		\lspbottomrule
	\end{tabularx}
\end{table}

\begin{table}[H]
	\caption{Some illustrative paradigms of the habitual present}
	\label{tab:habitualpresent-examples}
	\small
	\begin{tabularx}{1\textwidth}[]{%
		>{\centering\arraybackslash\small}p{10pt}
		>{\raggedright\arraybackslash}X
		>{\raggedright\arraybackslash}X
		>{\raggedright\arraybackslash}X
		>{\raggedright\arraybackslash}X
		>{\raggedright\arraybackslash}X
		>{\raggedright\arraybackslash}X}
		
		\lsptoprule
			{}	&	\multicolumn{2}{c}{\sqt{say}}
				&	\multicolumn{2}{c}{\sqt{do}}
				&	\multicolumn{2}{c}{\sqt{know}}\\\cmidrule(lr){2-3}\cmidrule(lr){4-5}\cmidrule(lr){6-7}

			{}	&	\multicolumn{1}{c}{singular} &	\multicolumn{1}{c}{plural}
				&	\multicolumn{1}{c}{singular} &	\multicolumn{1}{c}{plural}
				&	\multicolumn{1}{c}{singular} &	\multicolumn{1}{c}{plural}\\

		\midrule

			1	&	\tit{r-ik'-u-d}		&	\tit{d-ik'-u-d}
				&	\tit{b-irq'-i-d}	&	\tit{b-irq'-i-d}
				&	\tit{b-alχ-a-d}	&	\tit{b-alχ-a-d}\\

			2	&	\tit{r-ik'-u-tːe}	&	\tit{d-ik'-u-tːa}
				&	\tit{b-irq'-i-tːe}	&	\tit{b-irq'-i-tːa}
				&	\tit{b-alχ-a-tːe}	&	\tit{b-alχ-a-tːa}\\

			3	&	\tit{r-ik'-u}		&	\tit{b-ik'-u}
				&	\tit{b-irq'-u}		&	\tit{b-irq'-u}
				&	\tit{b-alχ-u}		&	\tit{b-alχ-u}\\
		\lspbottomrule
	\end{tabularx}
\end{table}


% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - %

\begin{description}[leftmargin=*]
\item[Semantic domains]\label{sssec:vis-semanticdomains}
\begin{enumerate}
    \item[]
	\item	habitual: used in time-less utterances that state general characteristics (of people, situations, etc.), in procedural texts, and for the description of (traditional) habits:%
	\begin{exe}
		\ex	\label{ex:We make egg khinkal}
		\gll	duq-n-a-lla	χːink'-e	d-irq'-id,	wec'al	duqu	k-ert'-id,	c'il	nejg	k-ert'-id\\
			egg-\tsc{pl}-\tsc{obl}-\tsc{gen}	khinkal-\tsc{pl}	\tsc{npl}-do\tsc{.ipfv-1.prs}	ten	egg	\tsc{down}-pour\tsc{.ipfv-1.prs}	then	milk	\tsc{down}-pour\tsc{.ipfv-1.prs}\\
		\glt	\sqt{We make egg khinkal. We pour ten eggs; then we pour milk.}

		\ex	\label{ex:I don’t forget anything}
		\gll	dam	qum.a.art-id	cik'al\\
			\tsc{1sg.dat}	forget\tsc{.ipfv.neg-1.prs}	anything\\
		\glt	\sqt{I don't forget anything.}

		\ex	\label{ex:Can you do this}
		\gll	u-l	b-arq'-ij	w-irχ-utːe=w?\\
			\tsc{2sg-erg}	\tsc{n-}do\tsc{.pfv-inf}	\tsc{m-}be.able\tsc{.ipfv-2sg.prs=q}\\
		\glt	\sqt{Can you do this?}
	\end{exe}

	\item	future and potential future: \sqt{will}/\sqt{could}/\sqt{should}, including the apodosis of conditionals:
	\begin{exe}
		\ex	\tnm{[Tell your head of administration to wait one more day.]}	\label{ex:Then I willcould be there in his office}\\
		\gll	c'il	du	hextːu-w	w-irχʷ-ud	hek'-i-la	kabinet-le-w\\
			then	\tsc{1sg}	there\tsc{.up-m}	\tsc{m-}be.\tsc{ipfv-1.prs}	\tsc{dem.up-obl-gen}	office\tsc{-loc-m}\\
		\glt	\sqt{Then I will/could be there in his office.}

		\ex	\label{ex:Shouldwill I sing my sad song or not}
		\gll	di-la	šišːim-la	dalaj	b-elč'-id=aw	a-b-elč'-id=aw?\\
			\tsc{1sg-gen}	suffering-\tsc{gen}	song	\tsc{n-}read\tsc{.ipfv-1.prs=q}	\tsc{neg-n-}read\tsc{.ipfv-1.prs=q}\\
		\glt	\sqt{Should\slash will I sing my sad song or not?} (a more literal translation is: \sqt{To sing or not to sing the song about my sufferings?})
	\end{exe}
\end{enumerate}
\end{description}

The habitual/future polysemy is common for Dagestanian languages \citep{Tatevosov2005} and also cross-linguistically well-attested \citep{Haspelmath1998}. The future reading has developed from the habitual reading, but it is only available for predicates that express transitory and accidental properties \refex{ex:Then I willcould be there in his office}, \refex{ex:I do not want to sit with you}. Predicates that denote temporally stable and essential properties that characterize their referents only express the habitual meaning \refex{ex:I don't know anything}, \refex{ex:Madina loves me}.

\begin{exe}
	\ex	\label{ex:I do not want to sit with you}
	\gll	hel	prosto,	``dam	a-b-ikː-ar'', ∅-ik'-ul ca-w,	``ašːi-cːella ka-jž-ij w-elqː-un-ne=da''	∅-ik'-ul	ca-w\\
		that	simply	\tsc{1sg.dat}	\tsc{neg-n-}want\tsc{.ipfv-3.prs}	\tsc{m-}say\tsc{.ipfv-icvb}	\tsc{cop-m}	\tsc{2pl-comit}	\tsc{down}-remain\tsc{.pfv-inf	} \tsc{m-}sate\tsc{.pfv-pret-cvb=1} \tsc{m-}say\tsc{.ipfv-icvb}	\tsc{cop-m}\\
	\glt	\sqt{He simply says, \dqt{I do not want to sit with you, I had enough of you,} he says.}

	\ex	\label{ex:I don't know anything}
	\gll	ca	cik'al	a-b-alχ-ad\\
		one	something	\tsc{neg-n-}know\tsc{.ipfv-1.prs}\\
	\glt	\sqt{I don't know anything.} (NOT: \sqt{I will\slash should not know anything.})

	\ex	\label{ex:Madina loves me}
	\gll	Madina-j	du	w-ičː-aq-id\\
		Madina\tsc{-dat}	\tsc{1sg}	\tsc{m-}want\tsc{.ipfv-caus-1.prs}\\
	\glt	\sqt{Madina loves me (masc.).} (NOT: \sqt{Madina will\slash should love me.}) (E)
\end{exe}

As can be seen in \reftab{tab:habitualpresent}, the third person has two suffixes, \tit{-u} and \tit{-ar}. The latter suffix is less frequently attested in the corpus. It is homophonous with the third person realis \isi{conditional} suffix -\textit{ar} (\refsec{sec:realisconditional}), and therefore not always easy to identify in texts. It seems that there is a slight semantic difference such that \tit{-u} can refer to single events whereas \tit{-ar} refers to habitually occurring events, but this difference is hard to detect and not always clear. For instance, \refex{ex:Don't you remember her, Bahamma} means \sqt{remember from time to time, think of}, whereas \tit{han b-irk-u} would just mean \sqt{remember (once)}. Similarly, \tit{bek' icː-u} means \sqt{the head aches (now)}, whereas \tit{bek' icː-ar} means that the head aches again and again, like when people have migraine.
By contrast, example \refex{ex:Can Hamid really throw clay into his eyes} shows an utterance, in which \tit{w-irχʷ-ar} could be replaced by \tit{w-irχ-u} without any change in meaning.  
\begin{exe}
	\ex	\label{ex:Don't you remember her, Bahamma}
	\gll	at	han	a-r-irk-ar=uw,	Baˁħaˁmma?	\\
		\tsc{2sg.dat}	remember	\tsc{neg-f-}occur\tsc{.ipfv-3.prs=q}		Bahamma\\
	\glt	\sqt{Don't you remember her, Bahamma?}
	
	\ex	\label{ex:Can Hamid really throw clay into his eyes}
	\gll	c'il	w-irχʷ-ar=uw	hati	ħaˁmid-li	cin-na	ul-b-a-cːe	lak'	d-arq'-ij	il-tːi	ʡaˁnčːi?\\
		then	\tsc{m-}be.able\tsc{.ipfv-3.prs=q}	really	Hamid\tsc{-erg}	\tsc{refl.sg-gen}	eye\tsc{-pl-obl-in}	throw	\tsc{npl}-do.\tsc{pfv-inf}	that-\tsc{pl}	clay\\
	\glt	\sqt{Can Hamid really throw clay into his eyes?}
\end{exe}

The verb \tit{b-ik'ʷ-ij} \sqt{say} is the most frequently used verb with \tit{-ar}, and it is frequently but not always translated as \isi{past tense} (\tie\ \sqt{said}) without any habitual seamntics  when it bears this suffix. I do not have an explanation for why the verb form (\textit{r/b-/d-})-\textit{ik'ʷ-ar} conveys non-\isi{habitual past} time semantics. Example \refex{ex:Zhabrail and his family do not invite you} illustrates the use of both suffixes -\textit{u} and -\textit{ar} with this verb in one sentence. 

\begin{exe}
	\ex	\label{ex:Zhabrail and his family do not invite you}
	\gll	``žaˁbraˁʔil-qal	r-aš	a-b-ik'-u=w?''	haʔ-ib=da.	``a-b-ik'-u'',	r-ik'ʷ-ar\\
		Zhabrail\tsc{-assoc}	\tsc{f-}go	\tsc{neg-n-}say\tsc{.ipfv-3.prs=q}	say\tsc{.pfv-pret=1}	\tsc{neg-n-}say\tsc{.ipfv-3.prs}	\tsc{f-}say\tsc{.ipfv-3.prs}\\
	\glt	\sqt{I said, ``Zhabrail and his family do not invite you?'' She said, ``They don't invite me.''} (lit. \sqt{Don't they say \dqt{Come!}})
\end{exe}


Negation is expressed through the prefix \tit{a-} \refex{ex:Don't you remember her, Bahamma}. Some \is{affective verb}affective verbs allow for the \isi{ergative} construction (in addition to the \isi{dative} construction) with the \isi{habitual present} (this phenomenon requires future research, but see \refsec{sec:Bivalent affective verbs} for some more examples).


% --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- %

\section{Habitual past}
\label{sec:vis-habitualpast}

The \isi{habitual past} is the past-tense counterpart of the \isi{habitual present}. It is only formed from the imperfective stem by means of a suffix \tit{-a} that is followed by person markers (first and second person). The person markers for second person are identical to the person markers used for the \isi{habitual present} such that \tit{b-aχ-} (\tsc{pfv})\slash\tit{b-alχ-} (\tsc{ipfv}) \sqt{know} has identical forms for the \isi{habitual present} and past in the second person (compare \reftab{tab:habitualpresent-examples} and \reftab{tab:habitualpast-examples}). In the third person, \tit{-a} is absent. Instead, the suffix \tit{-i} or alternatively the longer variants \tit{-iri} or, rarely, \tit{-ini} are used (\reftab{tab:habitualpast}). As an alternative to \tit{-a} plus person suffix, \tit{-i(ri)} can also be used with first and second person without any difference in meaning.

\begin{table}
	\caption{Person suffixes for the habitual past}
	\label{tab:habitualpast}
	\small
	\begin{tabularx}{0.40\textwidth}[]{%
		>{\centering\arraybackslash}p{10pt}
		>{\centering\arraybackslash}X
		>{\centering\arraybackslash}X}
		
		\lsptoprule
			{}	&	singular		&	plural\\
		\midrule
			1	&	\multicolumn{2}{c}{\tit{-di\slash -i(ri)}}\\
			2	&	\tit{-tːe\slash -i(ri)}	&	\tit{-tːa\slash -i(ri)}\\
			3	&	\multicolumn{2}{c}{\tit{-i(ri)\slash -ini}}\\
		\lspbottomrule
	\end{tabularx}
\end{table}

\begin{table}
	\caption{Some illustrative paradigms of the habitual past}
	\label{tab:habitualpast-examples}
	\small
	\begin{tabularx}{1\textwidth}[]{%
		>{\centering\arraybackslash\small}p{10pt}
		>{\raggedright\arraybackslash}X
		>{\raggedright\arraybackslash}X
		>{\raggedright\arraybackslash}X
		>{\raggedright\arraybackslash}X
		>{\raggedright\arraybackslash}X
		>{\raggedright\arraybackslash}X}
		
		\lsptoprule
			{}	&	\multicolumn{2}{c}{\sqt{say}}
				&	\multicolumn{2}{c}{\sqt{do}}
				&	\multicolumn{2}{c}{\sqt{know}}\\\cmidrule(lr){2-3}\cmidrule(lr){4-5}\cmidrule(lr){6-7}

			{}	&	\multicolumn{1}{c}{singular} &	\multicolumn{1}{c}{plural}
				&	\multicolumn{1}{c}{singular} &	\multicolumn{1}{c}{plural}
				&	\multicolumn{1}{c}{singular} &	\multicolumn{1}{c}{plural}\\

		\midrule

			1	&	\tit{r-ik'ʷ-a-di}	&	\tit{d-ik'ʷ-a-di}
				&	\tit{b-irq'-a-di}	&	\tit{b-irq'-a-di}
				&	\tit{b-alχ-a-di}	&	\tit{b-alχ-a-di}\\

			2	&	\tit{r-ik'ʷ-a-tːe}	&	\mbox{\tit{d-ik'ʷ-a-tːa}}
				&	\tit{b-irq'-a-tːe}	&	\tit{b-irq'-a-tːa}
				&	\tit{b-alχ-a-tːe}	&	\tit{b-alχ-a-tːa}\\

			3	&	\tit{r-ik'ʷ-i(ri)}	&	\tit{b-ik'ʷ-i(ri)}
				&	\tit{b-irq'-i(ri)}	&	\tit{b-irq'-i(ri)}
				&	\tit{b-alχ-i(ri)}	&	\tit{b-alχ-i(ri)}\\
		\lspbottomrule
	\end{tabularx}
\end{table}

The semantic domain is habitual situations with past time reference. The verb form is used to express habitually occurring actions in the past \refex{ex:He always told us}, employed in characterizing persons \refex{ex:Icari people ate it}, when referring to occupations, and so on. The functional range of the \isi{habitual past} also includes the expression of future-in-the-past in the protasis of past conditionals and irrealis conditionals \refex{ex:If my mother's brother would not have brought me to Chechnya}. As with the \isi{habitual present}, \isi{negation} is expressed through the prefix \tit{a-} \refex{ex:If my mother's brother would not have brought me to Chechnya} and some \is{affective verb}affective verbs additionally allow for the \isi{ergative} construction with the \isi{habitual past} (for more information see \refsec{sec:Bivalent affective verbs}).

\begin{exe}
	\ex	\label{ex:He always told us}
	\gll	har	zamana	herʔ-i	nišːi-cːe,	Sanži-le	w-ax-an=da\\
		every	time	say.\tsc{ipfv-hab.pst.3}	\tsc{1pl-in}	Sanzhi\tsc{-loc}	\tsc{m-}go\tsc{.ipfv-ptcp=1}\\
	\glt	\sqt{He always told us, ``I will go to Sanzhi.'' }

	\ex	\label{ex:Icari people ate it}
	\gll	uc'ran-t-a-l	b-uk-i,	nušːa-l	kːač	a-b-irq'-a-di\\
		Icari\tsc{-pl-obl-erg}	\tsc{n-}eat\tsc{.ipfv-hab.pst.3}	\tsc{1pl-erg}	touch	\tsc{neg-n-}do.\tsc{ipfv-hab.pst-1}\\
	\glt	\sqt{Icari people ate it (the meat of boars), we did not touch it.}

	\ex	\label{ex:If my mother's brother would not have brought me to Chechnya}
	\gll	di-la	aba-la	ucːi-l	du	čaˁčaˁn-t-a-cːe	a-r-uk-utːel		r-ubk'-a-di\\
		\tsc{1sg-gen}	mother\tsc{-gen}	brother\tsc{-erg}	\tsc{1sg}	Chechen\tsc{-pl-obl-in}	\tsc{neg-f-}lead\tsc{.ipfv-cond.pst}	\tsc{f-}die\tsc{.ipfv-hab.pst-1}\\
	\glt	\sqt{If my mother's brother would not have brought me to Chechnya, I would have died.}

\end{exe}


The verb \tit{b-ik'ʷ-ij} \sqt{say}, which was mentioned in the previous section as expressing past time reference by means of the third person \isi{habitual present} suffix for reasons that still await clarification is regularly inflected for the \isi{habitual past}. However, the meaning is not always clearly habitual but seems also to be just a perfective past \refex{ex:Who did you say was the head}.

 
\begin{exe}
	
	\ex	\label{ex:Who did you say was the head}
	\gll	a	presedatel	ča 	ca-w=de	∅-ik'ʷ-a-tːe?\\
		but	head	who	\tsc{cop-m=pst}	\tsc{m-}say\tsc{.ipfv-hab.pst-2sg}\\
	\glt	\sqt{Who (masc.) did you (masc.) say was the head (of the kolkhoz)?}
\end{exe}

The verb \tit{b-ikː-} (\tsc{ipfv}) \sqt{want, like, love}, which lacks a perfective stem, shows exceptional behavior with the habitual forms. The only available forms of the \isi{habitual present} are \tit{dam b-ikː-i} \sqt{I want} and \tit{nišːij b-ikː-i} \sqt{we want} and for \isi{questions} \tit{at b-ikː-i=w?} \sqt{Do you (\tsc{sg}) want?} and \tit{ašːij b-ikː-i=w?} \sqt{Do you (\tsc{pl}) want?}. There are no forms for third person and the second person forms cannot be used in assertions. Furthermore, the \isi{habitual past} expresses irealis modality with the first person, that is, \tit{dam}\slash\tit{nišːij b-ikː-a-di} translates as \sqt{I\slash we would like, I\slash we would want}. It is not used with other persons apart from the first person.


