% 12
\documentclass[output=paper]{LSP/langsci}
\author{Meredith Johnson}
\title{A description of verb-phrase ellipsis in {Hocąk}}

\abstract{In this paper, I argue that \il{Ho-Chunk}Hocąk displays \is{ellipsis, verb-phrase}verb-phrase ellipsis (VPE) and provide the first thorough description of this phenomenon. VPE in Hocąk displays the two defining characteristics of VPE cross-linguistically: it targets all VP-internal material, and it is licensed by a \isi{functional head}. In the case of Hocąk, I propose that the licensing head is active \emph{v}. Furthermore, Hocąk \is{ellipsis, verb-phrase}VPE also shows many other traits of \is{ellipsis, verb-phrase}VPE in other languages: the ellipsis site can be found in coordinated\is{coordination} and adjacent \isi{clauses} in addition to embedded and \isi{adjunct} \isi{clauses}, \is{ellipsis, verb-phrase}VPE is insensitive to the the contents of the \is{verb phrase}VP, and \is{ellipsis, verb-phrase}VPE gives rise to both strict and \isi{sloppy readings}. Lastly, I argue that \is{ellipsis, verb-phrase}VPE in \il{Ho-Chunk}Hocąk is derived by deletion of a full-fledged \is{verb phrase}VP, and that the \isi{ellipsis} site cannot be analyzed as a null \emph{pro} form.
% KEYWORDS: [Ho-Chunk, verb phrase, ellipsis, VPE, light verb]
}
\ChapterDOI{10.17169/langsci.b94.175}

\maketitle

\begin{document}
\let\eachwordtwo=\itshape

\section{Introduction}\label{sec:johnson:1}

 
This purpose of this paper is to both argue that \il{Ho-Chunk}Hocąk displays \is{ellipsis, verb-phrase}verb-phrase ellipsis (\is{ellipsis, verb-phrase}VPE) and to provide the first thorough description of this phenomenon. In \is{ellipsis, verb-phrase}VPE constructions, a \is{verb phrase}VP goes unpronounced when there is an appropriate antecedent \is{verb phrase}VP and a licensing head that identifies the gap. Both of these properties can be seen in the examples of \is{ellipsis, verb-phrase}VPE in \il{Ho-Chunk}Hocąk in \REF{ex:johnson:1} below. In each example, the \is{verb phrase}VP in the second conjunct\is{conjunction} is interpreted as identical to the \is{verb phrase}VP in the first conjunct, even though the former has no phonological realization. Instead, the \isi{light verb} \emph{ųų} takes the place of the \is{verb phrase}VP.
 


\ea\label{ex:johnson:1}
\ea
\glll Cecilga {\ob}{\sVP} wažątirehižą ruwį{\cb} kjane anąga nee šge {\ob}ha'ųų{\cb} kjane.\\
Cecil-ga {} wažątire-hižą {$\emptyset$}-ruwį kjane anąga nee šge {\db}ha-ųų kjane\\
Cecil-\textsc{prop} {} car-\textsc{indef} \textsc{3s/o}-buy \textsc{fut} and I also {\db}\textsc{1s}-do \textsc{fut}\\
\trans `Cecil will buy a car, and I will too.' 
\ex
\glll Cecilga {\ob}{\sVP} xjanąre waši{\cb} anąga Bryanga šge {\ob}ųų{\cb}.\\
Cecil-ga {} xjanąre {$\emptyset$}-waši anąga Bryan-ga šge {\db}{$\emptyset$}-ųų\\
Cecil-\textsc{prop} {} yesterday \textsc{3s}-dance and Bryan-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Cecil danced yesterday, and Bryan did too.'
\z
\z

Throughout this paper, I rely on the set of diagnostics of \is{ellipsis, verb-phrase}VPE established by \citet{Goldberg2005}, and subsequently used for Indonesian by \citet{Fortin2007}. \citet{Goldberg2005} uses characteristics of \ili{English} \is{ellipsis, verb-phrase}VPE to establish a \isi{typology} of \is{ellipsis, verb-phrase}VPE crosslinguistically, noting that ``\ili{English} \is{verb phrase}VP Ellipsis has a characteristic set of behavioral traits, the confluence of which is not found in other types of null anaphora'' \citep[27]{Goldberg2005}. 

Goldberg developed this set of traits in order to diagnose verb-stranding \is{ellipsis, verb-phrase}verb-phrase ellipsis (V\is{ellipsis, verb-phrase}VPE) in a variety of languages, including \ili{Hebrew}, \ili{Irish} and \ili{Swahili}. In V\is{ellipsis, verb-phrase}VPE, the verb has undergone raising before the remainder of the \is{verb phrase}VP is elided. On the surface, V\is{ellipsis, verb-phrase}VPE can be ambiguous between a null \isi{object} analysis or \is{ellipsis, verb-phrase}VPE analysis; thus, some of her diagnostics serve to distinguish these two approaches. An example of V\is{ellipsis, verb-phrase}VPE in \ili{Hebrew} is provided in \REF{ex:johnson:2}:

\let\eachwordtwo=\upshape

\ea\label{ex:johnson:2}
\gll Tazmini {et Dvora} la-misibaʔ Kvar hizmanti.\\
invite.\textsc{2fut} Dvora to.the-party already invite.\textsc{1pst}\\
\trans `Will you invite Dvora to the party? I already invited (Dvora to the party).' \\\citep[14]{Goldberg2005}
\z 

 
\il{Ho-Chunk}Hocąk \is{ellipsis, verb-phrase}VPE does not face this problem: there is an overt \isi{light verb} standing in for the \is{verb phrase}VP, much like \ili{English} \is{ellipsis, verb-phrase}VPE. Nonetheless, the data from \il{Ho-Chunk}Hocąk are consistent with all of the characteristics that Goldberg argues are diagnostic of \is{ellipsis, verb-phrase}VPE crosslinguistically. Furthermore, I show that these traits also distinguish \is{ellipsis, verb-phrase}VPE from other elliptical phenomena found in \il{Ho-Chunk}Hocąk, including \isi{gapping}, \isi{stripping} and null \isi{complement} anaphora. 

This paper is structured as follows. In \sectref{sec:johnson:2}, I establish that putative \is{ellipsis, verb-phrase}VPE in \il{Ho-Chunk}Hocąk displays the two most important characteristics of \is{ellipsis, verb-phrase}VPE: it targets all VP-internal material, and it is licensed by a \isi{functional head}. In the case of \il{Ho-Chunk}Hocąk, I propose that the licensing head for \is{ellipsis, verb-phrase}VPE is active \emph{v}. In \sectref{sec:johnson:3}, I show that \il{Ho-Chunk}Hocąk \is{ellipsis, verb-phrase}VPE displays other traits that have been attributed to \is{ellipsis, verb-phrase}VPE crosslinguistically. \sectref{sec:johnson:4} demonstrates that \is{ellipsis, verb-phrase}VPE in \il{Ho-Chunk}Hocąk must be analyzed as a deletion process, rather than a null \emph{pro} form. \sectref{sec:johnson:5} concludes the paper.
 

\section{Establishing the presence of \is{ellipsis, verb-phrase}VPE in Hocąk}\label{sec:johnson:2}

 
In this section, I show that the construction that I argue instantiates \is{ellipsis, verb-phrase}VPE in \il{Ho-Chunk}Hocąk displays the two defining characteristics of \is{ellipsis, verb-phrase}VPE. In \sectref{sec:johnson:2.1}, I demonstrate that the \isi{ellipsis} site includes all VP-internal material. In \sectref{sec:johnson:2.2}, I show that \is{ellipsis, verb-phrase}VPE is subject to the presence of an appropriate licensing head.
 

\subsection{\emph{ųų} targets the VP}\label{sec:johnson:2.1}

\is{ellipsis, verb-phrase}VPE is possible with both intransitive and transitive verbs, as seen in (\ref{ex:johnson:3}-\ref{ex:johnson:4}) below. \REF{ex:johnson:3a} and \REF{ex:johnson:3b} show that \emph{ųų} can target intransitive VPs. In the examples in \REF{ex:johnson:4} with transitive verbs, the direct \isi{object} is also included in the \isi{ellipsis} site.
 
\let\eachwordtwo=\itshape
 
\ea\label{ex:johnson:3}
\ea\label{ex:johnson:3a} 
\glll Cecilga {\ob}{\sVP} kere{\cb} anąga Matejaga šge {\ob}ųų{\cb}.\\
Cecil-ga {} $\emptyset$-kere, anąga Mateja-ga šge {\db}$\emptyset$-ųų\\
Cecil-\textsc{prop} {} \textsc {3s}-leave and Mateja-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Cecil left, and Mateja did too.' 
\ex\label{ex:johnson:3b} 
\glll Meredithga {\ob}{\sVP} nįįp{\cb} anąga Sarahga šge {\ob}ųų{\cb}.\\
Meredith-ga {} $\emptyset$-nįįp anąga Sarah-ga šge {\db}$\emptyset$-ųų\\
Meredith-\textsc {prop} {} \textsc{3s}-swim and Sarah-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Meredith swam, and Sarah did too.'
\z
\z
 
\ea\label{ex:johnson:4}
\ea\label{ex:johnson:4a} 
\glll Matejaga {\ob}{\sVP} {waisgap sguuhižą} rook'į{\cb} anąga Sarahga šge {\ob}ųų{\cb}.\\
Mateja-ga {} {waisgap sguu-hižą} $\emptyset$-rook'į anąga Sarah-ga šge {\db}$\emptyset$-ųų\\
Mateja-\textsc{prop} {} cake-\textsc{indef} \textsc{3s/o}-bake and Sarah-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Mateja baked a cake, and Sarah did too.' 
\ex\label{ex:johnson:4b} 
\glll Meredithga {\ob}{\sVP} waaruchižą hogiha{\cb} anąga Bryanga šge {\ob}ųų{\cb}.\\
Meredith-ga {} waaruc-hižą $\emptyset$-hogiha anąga Bryan-ga šge {\db}$\emptyset$-ųų\\
Meredith-\textsc{prop} {} table-\textsc{indef} \textsc{3s/o}-paint and Bryan-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Meredith painted a table, and Bryan did too.'
\z
\z

\is{ellipsis, verb-phrase}VPE can also target other internal arguments. Both indirect objects and \isi{resultative} phrases are typically analyzed as VP-internal (see e.g., \citealt{Larson1988} and \citealt{LevinRappaportHovav1995}), and they are also subject to \is{ellipsis, verb-phrase}VPE. The ditransitive example in \REF{ex:johnson:5} shows that both a direct \isi{object} and indirect \isi{object} can be contained in the \isi{ellipsis} site.

 
\ea\label{ex:johnson:5} 
\glll Cecilga {\ob}{\sVP} Meredithga wiiwagaxhižą hok'ų{\cb} anąga Matejaga šge {\ob}ųų{\cb}.\\
Cecil-ga {} Meredith-ga wiiwagax-hižą $\emptyset$-hok'ų anąga Mateja-ga šge {\db}$\emptyset$-ųų\\
Cecil-\textsc{prop} {} Meredith-\textsc{prop} pencil-\textsc{indef} \textsc{3s/o}-give and Mateja-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Cecil gave Meredith a pencil, and Mateja did too.'
\z

In \REF{ex:johnson:6}, we see examples of \is{ellipsis, verb-phrase}VPE with \isi{resultative} constructions in which the direct \isi{object} and result have both been elided. 
 
\ea\label{ex:johnson:6}
\ea\label{ex:johnson:6a} 
\glll Cecilga  {\ob}{\sVP} wažątirehiža šuuc hogiha{\cb} anąga Bryanga šge {\ob}ųų{\cb}.\\
Cecil-ga {} wažątire-hiža šuuc $\emptyset$-hogiha anąga Bryan-ga šge {\db}$\emptyset$-ųų\\
Cecil-\textsc{prop} {} car-\textsc{indef} red \textsc{3s/o}-paint and Bryan-\textsc{prop} also \textsc {s}-do\\
\trans `Cecil painted a car red, and Bryan did too.'
 
\ex\label{ex:johnson:6b} 
\glll Meredithga  {\ob}{\sVP} mąąshiža paras gistak{\cb} anąga Matejaga šge {\ob}ųų{\cb}.\\
Meredith-ga {} mąąs-hiža paras $\emptyset$-gistak anąga Mateja-ga šge {\db}$\emptyset$-ųų\\
Meredith-\textsc{prop} {} metal-\textsc{indef} flat \textsc{3s/o}-hit and Mateja-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Meredith hit metal flat, and Mateja did too.'
\z
\z

\is{ellipsis, verb-phrase}VPE also targets various adjuncts. \REF{ex:johnson:7} shows that \is{ellipsis, verb-phrase}VPE targets VPs containing temporal adjuncts. 
 
\ea\label{ex:johnson:7}
\ea 
\glll Cecilga {\ob}{\sVP} xjanąre waši{\cb} anąga Bryanga šge {\ob}ųų{\cb}.\\
Cecil-ga {} xjanąre $\emptyset$-waši anąga Bryan-ga šge {\db}$\emptyset$-ųų\\
Cecil-\textsc{prop} {} yesterday \textsc{3s}-dance and Bryan-\textsc{prop} also \textsc{3s-}do\\
\trans `Cecil danced yesterday, and Bryan did too.'

\ex 
\glll Meredithga  {\ob}{\sVP} hąąpte'e kšeehižą ruuc{\cb} anąga Matejaga šge {\ob}ųų{\cb}.\\
Meredith-ga {} hąąpte'e kšee-hižą $\emptyset$-ruuc anąga Mateja-ga šge {\db}$\emptyset$-ųų\\
Meredith-\textsc{prop} {} today apple-\textsc{indef  3s/o}-eat and Mateja-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Meredith ate an apple today, and Mateja did too.'
\z
\z

In \REF{ex:johnson:8}, locative adjuncts are included in the \isi{ellipsis} site. 
 
\ea\label{ex:johnson:8}
\ea 
\glll Cecilga {\ob}{\sVP} hosto eja waši{\cb} kjane anąga Bryanga šge {\ob}ųų{\cb} kjane.\\
Cecil-ga {} hosto eja $\emptyset$-waši kjane anąga Bryan-ga šge {\db}$\emptyset$-ųų kjane\\
Cecil-\textsc{prop} {} gathering there \textsc{3s}-dance \textsc{fut} and Bryan-\textsc{prop} also \textsc{3s}-do \textsc{fut}\\
\trans `Cecil will dance at the gathering, and Bryan will too.'
\ex 
\glll Cecilga {\ob}{\sVP} ciinąk eja wažątirehižą ruwį{\cb} anąga Bryanga šge {\ob}ųų{\cb}.\\
Cecil-ga {} ciinąk eja wažątire-hižą $\emptyset$-ruwį anąga Bryan-ga šge {\db}$\emptyset$-ųų\\
Cecil-\textsc{prop} {} city there car-\textsc{indef} \textsc{3s/o}-buy and Bryan-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Cecil bought a car in the city, and Bryan did too.'
\z
\z

\REF{ex:johnson:9} exemplifies \is{ellipsis, verb-phrase}VPE with a \isi{comitative}.
 
\ea\label{ex:johnson:9} 
\glll Cecilga {\ob}{\sVP} hinųkra hakižu waši{\cb} anąga Bryanga šge {\ob}ųų{\cb}.\\
Cecil-ga {} hinųk-ra hakižu $\emptyset$-waši anąga Bryan-ga šge {\db}$\emptyset$-ųų\\
Cecil-\textsc{prop} {} woman-\textsc{def} be.with \textsc{3s}-dance and Bryan-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Cecil danced with the woman, and Bryan did too.'
\z

\REF{ex:johnson:10} demonstrates that various manner adverbs can also be \isi{subject} to \is{ellipsis, verb-phrase}VPE.
 
\ea\label{ex:johnson:10}
\ea 
\glll Bryanga  {\ob}{\sVP} teejąki {nįįtašjak taaxu} racgą{\cb} anąga Sarahga šge {\ob}ųų{\cb}.\\ 
Bryan-ga {} teejąki {nįįtašjak taaxu} $\emptyset$-racgą anąga Sarah-ga šge {\db}$\emptyset$-ųų\\
Bryan-\textsc{prop} {} often coffee \textsc{3s}-drink and Sarah-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Bryan often drinks coffee, and Sarah does too.'
\ex 
\glll Cecilga  {\ob}{\sVP} pįįhį mąąnį{\cb} anąga Bryanga šge {\ob}ųų{\cb}.\\
Cecil-ga {} pįįhį $\emptyset$-mąąnį anąga Bryan-ga šge {\db}$\emptyset$-ųų\\
Cecil-\textsc{prop} {} quietly \textsc{3s}-walk and Bryan-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Cecil walked carefully/quietly, and Bryan did too.'
\ex 
\glll Meredithga {\ob}{\sVP} hikųhe {nįįtašjak taaxu} racgą{\cb} anąga Bryanga šge {\ob}ųų{\cb}.\\
Meredith-ga {} hikųhe {nįįtašjak taaxu} $\emptyset$-racgą anąga Bryan-ga šge {\db}$\emptyset$-ųų\\
Meredith-\textsc{prop} {} quickly coffee \textsc{3s}-drink and Bryan also {\db}\textsc{3s}-do\\
\trans `Meredith drank coffee quickly, and Bryan did too.'
\z
\z

In all of the examples in \REF{ex:johnson:7}--\REF{ex:johnson:10}, the \isi{adjunct} in the antecedent \is{verb phrase}VP is interpreted as being present in the \isi{ellipsis} site, indicating that \emph{ųų} targets the entire \is{verb phrase}VP rather than just the \isi{object}(s).

Lastly, \isi{complement} \isi{clauses} can also be included in \is{ellipsis, verb-phrase}VPE. The example in \REF{ex:johnson:11} has two possible interpretations: either that Meredith also bought a car, or that Meredith also said that Cecil bought a car. Under the second reading, \is{ellipsis, verb-phrase}VPE targets the matrix \is{clauses}clause, eliding the verb and its \isi{complement} \is{clauses}clause.
 
\ea\label{ex:johnson:11} 
\glll Bryanga {\ob}{vp} Cecilga wažątirehižą ruwįže ee{\cb} anąga Meredithga šge {\ob}ųų{\cb}.\\
Bryan-ga {} Cecil-ga wažątire-hižą $\emptyset$-ruwį-že $\emptyset$-ee anąga Meredith-ga šge {\db}$\emptyset$-ųų\\
Bryan-\textsc{prop} {} Cecil-\textsc{prop} car-\textsc{indef} \textsc{3s/o}-buy-\textsc{comp} \textsc{3s}-say and Meredith-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Bryan said that Cecil bought a car, and Meredith did too.'
\z


\subsection{Licensing of VPE}\label{sec:johnson:2.2}

The main characteristic that distinguishes \is{ellipsis, verb-phrase}VPE from other elliptical processes is the presence of an overt licensing head located in the inflectional\is{conjugation} domain above the \is{verb phrase}VP. \is{ellipsis, verb-phrase}VPE in \ili{English} can be licensed by a variety of functional elements, such as \emph{do} in \REF{ex:johnson:12a}, \emph{be} in \REF{ex:johnson:12b}, \emph{have} in \REF{ex:johnson:12c}, \emph{can} in \REF{ex:johnson:12d} and \emph{will} in \REF{ex:johnson:12e}. The obligatory presence of an inflectional\is{conjugation} head has led previous researchers to argue that \is{ellipsis, verb-phrase}VPE is licensed by T/Infl \citep{Bresnan1976,Sag1976,Zagona1988,Lobeck1995}.

 
\ea\label{ex:johnson:12}
\ea\label{ex:johnson:12a}
Lily wore a skirt, and Molly \textbf{did} too.
\ex\label{ex:johnson:12b}
Lily is reading a book, and Molly \textbf{is} too.
\ex\label{ex:johnson:12c}
Lily hasn't finished the book, but Molly \textbf{has}.
\ex\label{ex:johnson:12d}
Lily can ride a bike, and Molly \textbf{can} too.
\ex\label{ex:johnson:12e} 
Lily will leave, and Molly \textbf{will} too.
\z
\z

In contrast, there is no such inflectional\is{conjugation} head found with \isi{stripping} or \isi{gapping}. Stripping is an elliptical phenomenon in which an entire \is{clauses}clause is elided except for a single element that is stranded. This is illustrated in \REF{ex:johnson:13a}. In \isi{gapping} constructions, the verb (and other potential material) is left unpronounced, while there are two elements that are stranded. An example of \isi{gapping} can be seen in \REF{ex:johnson:13b}.
 
\ea\label{ex:johnson:13}
\ea\label{ex:johnson:13a}
Lily came over, and Molly too.
\ex\label{ex:johnson:13b} 
Lily brought bagels, and Molly danishes.
\z
\z

 
In \il{Ho-Chunk}Hocąk, the licensing requirement on \is{ellipsis, verb-phrase}VPE is different: \is{ellipsis, verb-phrase}VPE is conditioned solely by the presence of the \isi{light verb} \emph{ųų}. We have seen this in all of the instances of \is{ellipsis, verb-phrase}VPE given above. The examples in \REF{ex:johnson:14}--\REF{ex:johnson:18} illustrate that \emph{ųų} is indeed a \isi{light verb}: it productively combines with both nouns and verbs to create complex \isi{predicate}s. Based on its distribution, I assume that \emph{ųų} realizes the \isi{functional head} \emph{v} \citep[Examples 14--18]{Hartmann2012}.
 
\renewcommand{\exfont}{\itshape} 
 
\ea\label{ex:johnson:14}
\begin{multicols}{2}
\ea 
mąąnąąpeja\\
\glt `warrior'
\ex 
mąąnąąpeja ųų\\
\glt `be in the military'
\z
\end{multicols}
\z
 
\ea\label{ex:johnson:15}
\begin{multicols}{2}
\ea 
nąąwą\v{g}o\v{g}o\\
\glt `fiddle'
\ex 
nąąwą\v{g}o\v{g}o ųų\\
\glt `play the fiddle'
\z
\end{multicols}
\z
 
\ea\label{ex:johnson:16}
\begin{multicols}{2}
\ea 
waruc\\
\glt `food'
\ex 
waruc ųų\\
\glt `cook, prepare food'
\z
\end{multicols}
\z
 
\ea\label{ex:johnson:17}
\begin{multicols}{2}
\ea 
waagax\\
\glt `paper, letter'
\ex 
waagax ųų\\
\glt `write (a letter)'
\z
\end{multicols}
\z
 
\newpage
\ea\label{ex:johnson:18}
\begin{multicols}{2}
\ea 
hooxiwi\\
\glt `cough' (verb)\\

\ex 
hooxiwi ųų\\
\glt `have a cold'
\z
\end{multicols}
\z

Tense and modals can be present in \is{ellipsis, verb-phrase}VPE constructions; however, they are never obligatory. When present, tense and modals always co-occur with the \isi{light verb} \emph{ųų}. \REF{ex:johnson:19a} shows that the future tense marker \emph{kjane} can follow \emph{ųų}, while \REF{ex:johnson:19b} and \REF{ex:johnson:19c} demonstrate that the modals \emph{ną} and \emph{s'aare} can also appear after \emph{ųų}. When \emph{ųų} is omitted, the result is ungrammatical.

\renewcommand{\exfont}{\upshape} 

\ea\label{ex:johnson:19}
\ea\label{ex:johnson:19a} 
\glll Cecilga wažątirehižą ruwį kjane anąga nee šge *(ha'ųų) \textbf{kjane}.\\
Cecil-ga wažątire-hižą $\emptyset$-ruwį kjane anąga nee šge ha-ųų \textbf{kjane}\\
Cecil-\textsc{prop} car-\textsc{indef} \textsc{3s/o}-buy \textsc{fut} and I also \textsc{1s}-do \textsc{fut}\\
\trans `Cecil will buy a car, and I will too.' 
 
\ex\label{ex:johnson:19b} 
\glll Meredithga hąąke wažątirera {pįį'ų ruxuruknį} nųnįge Matejaga *(ųų) \textbf{ną}.\\
Meredith-ga hąąke wažątire-ra {$\emptyset$-pįį'ų ruxuruk-nį} nųnįge Mateja-ga {\db}$\emptyset$-ųų \textbf{ną}\\
Meredith-\textsc{prop} \textsc{neg} car-\textsc{def} \textsc{3s/o}-fix-\textsc{neg} but Mateja-\textsc{prop} \textsc{3s}-do can\\
\trans `Meredith can't fix the car, but Mateja can.'
 
\ex\label{ex:johnson:19c} 
\glll Meredithga hąąke {nįįtašjak taaxu} ruwįnį nųnįge Matejaga *(ųų) \textbf{s'aare}.\\ 
Meredith-ga hąąke {nįįtašjak taaxu} $\emptyset$-ruwį-nį nųnįge Mateja-ga {\db}$\emptyset$-ųų \textbf{s'aare}\\
Meredith-\textsc{prop} \textsc{neg} coffee \textsc{3s/o}-buy-\textsc{neg} but Mateja-\textsc{prop} \textsc{3s}-do must\\
\trans `Meredith didn't buy coffee but Mateja must have.' 
\z
\z
 
Thus, we see that T/Infl does not play the same role in \is{ellipsis, verb-phrase}VPE licensing in \il{Ho-Chunk}Hocąk as it does in other languages. However, \is{ellipsis, verb-phrase}VPE in \il{Ho-Chunk}Hocąk is constrained by the type of \isi{predicate}. As the examples in \REF{ex:johnson:20} show, \is{ellipsis, verb-phrase}VPE is not licensed with non-agentive verbs:
 

 
\ea\label{ex:johnson:20}
\ea[*] { 
\glll Meredithga kšee gipį anąga Bryanga šge ųų.\\
Meredith-ga kšee $\emptyset$-gipį anąga Bryan-ga šge {\db}$\emptyset$-ųų\\
Meredith-\textsc{prop} apple \textsc{3s}-like and Bryan-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans (Intended: `Meredith likes apples, and Bryan does too.')}\label{ex:johnson:20a}
\ex[*] {
\glll Cecilga wįįxra waaja anąga Meredithga šge ųų.\\
Cecil-ga wįįx-ra wa-$\emptyset$-haja anąga Meredith-ga šge {\db}$\emptyset$-ųų\\
Cecil-\textsc{prop} duck-\textsc{def} \textsc{3o.pl-3s}-see and Meredith-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans (Intended: `Cecil saw the ducks, and Meredith did too.')}\label{ex:johnson:20b}
\ex[*] {
\glll Meredithga hoišą anąga Bryanga šge ųų.\\
Meredith-ga $\emptyset$-hoišą anąga Bryan-ga šge {\db}$\emptyset$-ųų\\
Meredith-\textsc{prop} \textsc{3s}-busy.\textsc{stat} and Bryan-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans (Intended: `Meredith is busy, and Bryan is too.')}\label{ex:johnson:20c}
\ex[*] {
\glll Cecilga hįįcge nųnįge Bryanga hąąke ųųnį.\\
Cecil-ga $\emptyset$-hįįcge nųnįge Bryan-ga hąąke {\db}$\emptyset$-ųų-nį\\
Cecil-\textsc{prop} \textsc{3s}-tired.\textsc{stat} but Bryan-\textsc{prop} \textsc{neg} \textsc{3s}-do-\textsc{neg}\\
\trans (Intended: `Cecil is tired, but Bryan isn't.')}\label{ex:johnson:20d}
\z
\z
 
Like other Siouan languages, \il{Ho-Chunk}Hocąk exhibits an active-stative alignment pattern: the active set of verbal person markers is used to index the \isi{subject} of transitive verbs and active intransitive verbs, while the stative set is used to index the \isi{object} of transitive verbs and the \isi{subject} of stative intransitive verbs. This alignment pattern interacts with \is{ellipsis, verb-phrase}VPE in revealing ways. While \is{ellipsis, verb-phrase}VPE is banned with most stative intransitive verbs, such as those in \REF{ex:johnson:20c} and \REF{ex:johnson:20d}, \is{ellipsis, verb-phrase}VPE is possible with certain stative intransitives when they have an agentive reading. \emph{Hokąre} `to fall in' is normally a stative intransitive verb, but it is possible to use it in \is{ellipsis, verb-phrase}VPE contexts if the \isi{subject} falls \emph{deliberately}, as in \REF{ex:johnson:21}. In this context, \emph{ųų} takes the \emph{active} person marker set. In \REF{ex:johnson:21b}, the verb takes the second person active marker \emph{š-}; the stative marker \emph{nį-} is not permitted. The marker \emph{nį-} is the one that would typically be found on the verb \emph{hokąre}, as shown in \REF{ex:johnson:22}.
 
 
\ea\label{ex:johnson:21}
\ea\label{ex:johnson:21a} 
\glll Meredithga nįį eeja hokąre anąga Bryanga šge ųų.\\
Meredith-ga nįį eeja $\emptyset$-hokąre anąga Bryan-ga šge {\db}$\emptyset$-ųų\\
Meredith-\textsc{prop} water there \textsc{3s}-fall.in and Bryan-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Meredith fell into the water (deliberately), and Bryan did too.'
 
\ex\label{ex:johnson:21b} 
\glll Meredithga nįį eeja hokąre anąga (nee) šge š'ųų/*nį'ųų.\\
Meredith-ga nįį eeja $\emptyset$-hokąre anąga nee šge š-'ųų/nį-'ųų\\
Meredith-\textsc{prop} water there \textsc{3s}-fall.in and you also \textsc{2s}-do\\
\trans `Meredith fell into the water (deliberately), and you did too.'
\z
\z
 
\ea\label{ex:johnson:22} 
\glll Honįkąre.\\
<nį>hokąre\\
<\textsc{2s}>fall.in.\textsc{stat}\\
\trans `You fell in(to something).' \citep{Hartmann2012}
\z
	
This restriction on \is{ellipsis, verb-phrase}VPE is not due to lexical properties of \emph{ųų}: when \emph{ųų} functions as a \isi{light verb}, it can form non-agentive verbs, as in \REF{ex:johnson:23}:

 
\ea\label{ex:johnson:23}
\ea 
\textit{hooxiwi ųų} `have a cold' (stative intransitive)
\vspace{12pt}
\ex 
\textit{roo taakac ųų} `have a fever' (stative intransitive)
\vspace{12pt}
\ex 
\textit{paaxšišik ųų} `have an upset stomach' (stative intransitive)\\
\citep{Hartmann2012}
\z
\z

 
To formalize this restriction on \is{ellipsis, verb-phrase}VPE in \il{Ho-Chunk}Hocąk, I adopt \citegen{Merchant2001} proposal that \isi{ellipsis} takes place when a so-called ``[E]-feature'' is present on the relevant licensing head. In the case of \il{Ho-Chunk}Hocąk, I propose that an [E]-feature is present only on the agentive \emph{v} head.\footnote{This agentivity requirement on a process that affects the \is{verb phrase}VP is not completely unique to \il{Ho-Chunk}Hocąk. For example, \citet{Hallman2004} notes that \ili{English} \emph{do so} replacement is restricted to agentive VPs, even though other uses of \emph{do} are not subject to this constraint (e.g., \emph{Max loves studying \ili{French}, and Mary does (*so) too.}) \citet{Rouveret2012} also shows that \is{ellipsis, verb-phrase}VPE in \ili{Welsh} is licensed uniquely by the \isi{light verb} \emph{gweund}, and furthermore that \is{ellipsis, verb-phrase}VPE is not permitted with stative \isi{predicate}s. The only possibility with stative VPs is V\is{ellipsis, verb-phrase}VPE. However, Rouveret also shows that \emph{gweund} is also incompatible with stative \isi{predicate}s in its non-elliptical uses. This contrasts with the behavior of \emph{ųų} in \il{Ho-Chunk}Hocąk and \emph{do} in \emph{do so} in \ili{English}.}  This accounts for the fact that \is{ellipsis, verb-phrase}VPE is solely conditioned by the presence of the \isi{light verb} (or \emph{v}) \emph{ųų}, and furthermore that only agentive verbs can be elided: if there is a non-agentive \emph{v} present, then \isi{ellipsis} will not be licensed. 

This conclusion is in line with other research that argues that \emph{v} is responsible for licensing \is{ellipsis, verb-phrase}VPE crosslinguistically. Many recent approaches to \isi{ellipsis} have argued for a link between phases and elliptical phenomena (\citealt{Holmberg2001}, \citealt{vanCraenenbroeck2004}, \citealt{Gengel2007},  \citealt{YoshidaGallego2008}, \citealt{Gallego2009}, among others). Specifically, they propose that \isi{ellipsis} results when a phasal head (e.g. \emph{v}, C, D) licenses deletion of its \isi{complement}. These theories are a natural development of Chomsky's (\citeyear{Chomsky2000,Chomsky2001b,Chomsky2004}) theory of phases: if \isi{ellipsis} is PF-deletion, it follows that the units that are sent cyclically to the PF interface are precisely the ones that can be targeted for deletion. More concretely, \citet{Rouveret2012} adopts the phasal analysis of \isi{ellipsis}, and puts forward a theory to predict which languages permit \is{ellipsis, verb-phrase}VPE. He argues that \emph{v} always has an uninterpretable [tense] feature, and that, in languages with \is{ellipsis, verb-phrase}VPE, the [tense] feature is valued on \emph{v} phase-internally. Rouveret proposes that the elements that license \is{ellipsis, verb-phrase}VPE are all merged in \emph{v}, and subsequently move to Infl. All of these approaches are compatible with the \il{Ho-Chunk}Hocąk data, with the caveat that \is{ellipsis, verb-phrase}VPE is more restricted in \il{Ho-Chunk}Hocąk: it is only licensed by active \emph{v}.
 


\section{Crosslinguistic characteristics of VPE}\label{sec:johnson:3}

 
In the previous section, I demonstrated that \il{Ho-Chunk}Hocąk displays the two defining characteristics of \is{ellipsis, verb-phrase}VPE: the elliptical process in question targets the entire \is{verb phrase}VP, and is conditioned by the presence of a licensing head. \citet{Goldberg2005} discusses five other characteristics of \is{ellipsis, verb-phrase}VPE that are not shared by other elliptical phenomena, which are listed in \REF{ex:johnson:24}: 
 
 
\ea\label{ex:johnson:24}
\ea 
Possible in both coordinated\is{coordination} and adjacent CPs
\ex 
Insensitive to contents of elided \is{verb phrase}VP
\ex 
Ellipsis site can be in a syntactic island
\ex 
Ellipsis site can be embedded
\ex 
Presence of strict and \isi{sloppy readings}
\z
\z


 
In the subsections that follow, I show that \il{Ho-Chunk}Hocąk \is{ellipsis, verb-phrase}VPE also generally conforms to this \isi{typology}. In the areas where \il{Ho-Chunk}Hocąk appears to differ from \ili{English}, I demonstrate that this is due to other differences between the two languages that are independent of \isi{ellipsis}.
 

\subsection{Ellipsis licensed in both coordinated\is{coordination} and adjacent CPs}\label{sec:johnson:3.1}

 
\citet{Goldberg2005} notes that \ili{English} \is{ellipsis, verb-phrase}VPE is possible with a variety of sentence types. \is{ellipsis, verb-phrase}VPE is licit when the antecedent \is{verb phrase}VP and elided \is{verb phrase}VP are found in conjoined CPs \REF{ex:johnson:25a}, in adjacent CPs uttered by the same speaker \REF{ex:johnson:25b}, and when the antecedent is in a question and the \isi{ellipsis} site in the answer \REF{ex:johnson:25c}. In this section, I show that the same is true in \il{Ho-Chunk}Hocąk.
 
 
\ea\label{ex:johnson:25}
\ea\label{ex:johnson:25a}
Lily hates beets, but Molly doesn't.
\ex\label{ex:johnson:25b}
Lily hates beets. Molly does too.
\ex\label{ex:johnson:25c} 
Who hates beets? Molly does.
\z
\z

All of the examples of \is{ellipsis, verb-phrase}VPE we saw in \sectref{sec:johnson:2} involved two \isi{clauses} joined by the coordinator\is{coordination} \emph{anąga} `and'. \is{ellipsis, verb-phrase}VPE is also possible with disjunction, as seen in \REF{ex:johnson:26} with \emph{nųnįge} `but'.
 
\ea\label{ex:johnson:26}
\ea 
\glll Cecilga wažątirehižą ruwį nųnįge nee hąąke ha'ųųnį.\\
Cecil-ga wažątire-hižą $\emptyset$-ruwį nųnįge nee hąąke ha-ųų-nį\\
Cecil-\textsc{prop} car-\textsc{indef} \textsc{3s/o}-buy but I \textsc{neg} \textsc{1s}-do-\textsc{neg}\\
\trans `Cecil bought a car, but I didn't.'

\ex 
\glll Sarahga hąąke haas gihinį nųnįge Matejaga ųų.\\
Sarah-ga hąąke haas $\emptyset$-gihi-nį nųnįge Mateja-ga {\db}$\emptyset$-ųų\\
Sarah-\textsc{prop} \textsc{neg} berry \textsc{3s}-pick-\textsc{neg} but Mateja-\textsc{prop} {\db}\textsc{3s}-do\\
\trans `Sarah didn't pick berries, but Mateja did.'
\z
\z


\REF{ex:johnson:27} shows that \is{ellipsis, verb-phrase}VPE is also licit in adjacent CPs. In each example, the antecedent \is{verb phrase}VP is found in the first sentence while the ellipsis\is{ellipsis, verb-phrase} site is in the second sentence.
 
\ea\label{ex:johnson:27}
\ea 
\glll Meredithga waaruchižą hogiha. Bryanga šge ųų.\\
Meredith-ga waaruc-hižą $\emptyset$-hogiha Bryan-ga šge {\db}$\emptyset$-ųų\\
Meredith-\textsc{prop} table-\textsc{indef} \textsc{3s/o}-paint Bryan-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Meredith painted a table. Bryan did too.'
\ex 
\glll Meredithga hąąke {waisgap sguu xuwuxuwuhižą} ruucnį. Bryanga ųų.\\
Meredith-ga hąąke {waisgap sguu xuwuxuwu-hižą} $\emptyset$-ruuc-nį Bryan-ga {\db}$\emptyset$-ųų\\
Meredith-\textsc{prop} \textsc{neg} cookie-\textsc{indef} \textsc{3s/o}-eat-\textsc{neg} Bryan-\textsc{prop} {\db}\textsc{3s}-do\\
\trans `Meredith didn't eat a cookie. Bryan did.'
\z
\z

 
Lastly, \is{ellipsis, verb-phrase}VPE also occurs in question-answer pairs in \il{Ho-Chunk}Hocąk. In \REF{ex:johnson:28a}, a yes-no question contains the antecedent \is{verb phrase}VP and the answer contains the gap. \REF{ex:johnson:28b} demonstrate that the same holds of \emph{wh}-questions.
 
 
\ea\label{ex:johnson:28}
\ea\label{ex:johnson:28a} 
\glll \textup{Question:} {Nįįtašjak taaxu} šuruwį? \textup{Answer:} Ha'ųų.\\
{} {nįįtašjak taaxu} šu-ruwį {} ha-ųų\\
{} coffee \textsc{2s}-buy {} \textsc{1s}-do\\
 \trans Q: `Did you buy coffee?' A: `I did.' 
\ex\label{ex:johnson:28b} 
\glll \textup{Question:} Peežega Cecilga {gišja hii}? \textup{Answer:} Bryanga ųų.\\
{} peežega Cecil-ga $\emptyset$-{gišja hii} {} Bryan-ga {\db}$\emptyset$-ųų.\\
{} who Cecil-\textsc{prop} \textsc{3s/o}-visit {} Bryan-\textsc{prop} {\db}\textsc{3s}-do\\
\trans Q: `Who visited Cecil?' A: `Bryan did.'
\z
\z


\subsection{Ellipsis and the contents of the VP}\label{sec:johnson:3.2}

\citet{Goldberg2005} distinguishes \is{ellipsis, verb-phrase}VPE from null \isi{complement} anaphora (NCA) based on the type of constituent that is elided. In NCA, a matrix verb is stranded and its \isi{complement} is elided. However, NCA is constrained by the contents of the \is{verb phrase}VP: only propositions can be elided. This is illustrated by the contrast between the grammatical NCA examples in \REF{ex:johnson:29a} and \REF{ex:johnson:29c} and the ungrammatical examples in \REF{ex:johnson:29b} and \REF{ex:johnson:29d}:
 
\ea\label{ex:johnson:29}
\ea\label{ex:johnson:29a} 
Pat doesn't know \textbf{that Terry is moving to Japan}, but Robin knows.
\ex[*] {
Pat doesn't know \textbf{how to speak Inuktitut}, but Robin knows.}\label{ex:johnson:29b}
\ex\label{ex:johnson:29c}
Pat forgot \textbf{to close the door}, but Robin remembered.
\ex[*] {
Pat forgot \textbf{the answer}, but Robin remembered. \citep[245]{Fortin2007}}\label{ex:johnson:29d}
\z
\z

In contrast, the grammaticality of \is{ellipsis, verb-phrase}VPE is not dependent on the contents of the \is{verb phrase}VP. The examples in \REF{ex:johnson:30} show that \is{ellipsis, verb-phrase}VPE is possible regardless of whether the \isi{complement} of the \is{verb phrase}VP expresses a proposition or not.
 
\ea\label{ex:johnson:30}
\ea 
Pat doesn't know \textbf{that Terry is moving to Japan}, but Robin does.
\ex 
Pat doesn't know \textbf{how to speak Inuktitut}, but Robin does.
\ex 
Pat forgot \textbf{to close the door}, but Robin didn't.
\ex 
Pat forgot \textbf{the answer}, but Robin didn't.
\z
\z

 
As \citet{Fortin2007} points out, this diagnostic does not serve to distinguish \is{ellipsis, verb-phrase}VPE from NCA in languages with null objects. \il{Ho-Chunk}Hocąk allows both null subjects and objects, as seen in \REF{ex:johnson:31b}:
 

 
\ea
\ea 
\glll Wijųkra šųųkra hoxataprookeeja haja.\\
Wijųk-ra šųųk-ra hoxatap-rook-eeja $\emptyset$-haja\\
cat-\textsc{def} dog-\textsc{def} woods-inside-there \textsc{3s/o}-see\\
\trans `The cat saw the dog in the woods. 
\ex\label{ex:johnson:31b} 
\glll Hoxataprookeeja haja.\\
hoxatap-rook-eeja $\emptyset$-haja\\
woods-inside-there \textsc{3s/o}-see\\
\trans `[The cat] saw [the dog] in the woods.' (\citealt[7]{JohnsonEtAl2013b})
\z
\z

\newpage
Thus, it is not surprising that both propositional and non-propositional verbal complements can be null in \il{Ho-Chunk}Hocąk. In \REF{ex:johnson:32}, the \isi{complement} of the verb \emph{hiperes} `know' can be null both when it is a proposition \REF{ex:johnson:32a} or an embedded question \REF{ex:johnson:32b}. Likewise, both propositional \REF{ex:johnson:33a} and DP \isi{object} \REF{ex:johnson:33b} complements of \emph{wakikųnųnį} `forget' surface as null.
 
\ea\label{ex:johnson:32}
\ea\label{ex:johnson:32a} 
\glll Sarahga Meredithga rookhožura ruucra hiperes, anąga Matejaga šge hireperesšąną.\\
Sarah-ga Meredith-ga rookhožu-ra $\emptyset$-ruuc-ra $\emptyset$-hiperes anąga Mateja-ga šge $\emptyset$-hiperes-šąną\\
Sarah-\textsc{prop} Meredith-\textsc{prop} pie-\textsc{def} \textsc{3s/o}-eat-\textsc{comp} \textsc{3s}-know and Mateja-\textsc{prop} also \textsc{3s}-know-\textsc{decl}\\
\trans `Sarah knows that Meredith ate the pie, and Mateja knows too.'
 
\ex\label{ex:johnson:32b} 
\glll Sarahga jaagu'ų Meredithga kerera hiperes, anąga Matejaga šge hiperesšąną.\\
Sarah-ga jaagu'ų Meredith-ga $\emptyset$-kere-ra $\emptyset$-hiperes anąga Mateja-ga šge $\emptyset$-hiperes-šąną\\
Sarah-\textsc{prop} why Meredith-\textsc{prop} \textsc{3s}-leave-\textsc{comp} \textsc{3s}-know and Mateja-\textsc{prop} also \textsc{3s}-know-\textsc{decl}\\
\trans `Sarah knows why Meredith left, and Mateja knows (why Meredith left) too.'
\z
\z

 
\ea\label{ex:johnson:33}
\ea\label{ex:johnson:33a} 
\glll Bryanga {nįįtašjak taaxu} ruwįra wakikųnųnį, nųnįge Meredithga hąąke wakikųnųnįnį.\\
Bryan-ga {nįįtašjak taaxu} $\emptyset$-ruwį-ra $\emptyset$-wakikųnųnį nųnįge Meredith-ga hąąke $\emptyset$-wakikųnųnį-nį\\
Bryan-\textsc{prop} coffee \textsc{3s}-buy-\textsc{comp} \textsc{3s}-forget but Meredith-\textsc{prop} \textsc{neg} \textsc{3s}-forget-\textsc{neg}\\
\trans `Bryan forgot to buy coffee, but Meredith didn't forget.'
 
\ex\label{ex:johnson:33b} 
\glll Bryanga {waisgap sguura} wakikųnųnį, nųnįge Meredithga hąąke wakikųnųnįnį.\\
Bryan-ga {waisgap sguu-ra} $\emptyset$-wakikųnųnį nųnįge Meredith-ga hąąke $\emptyset$-wakikųnųnį-nį\\
Bryan-\textsc{prop} cake-\textsc{def} \textsc{3s/o}-forget but Meredith-\textsc{prop} \textsc{neg} \textsc{3s/o}-forget-\textsc{neg}\\
\trans `Bryan forgot the cake, but Meredith didn't forget (the cake).'
\z
\z

\newpage
 
Thus, this particular diagnostic does not work for \il{Ho-Chunk}Hocąk due to independent factors. The \isi{complement} of verbs like `know' and `forget' can always be null, presumably due to the availability of \isi{object} \emph{pro} drop.\footnote{A full comparison of NCA and \is{ellipsis, verb-phrase}VPE is not possible in \il{Ho-Chunk}Hocąk. \is{ellipsis, verb-phrase}VPE with verbs like `know' and `forget' is ungrammatical (examples omitted for space purposes) since these verbs are non-agentive.}
 

\subsection{Ellipsis in syntactic islands}\label{sec:johnson:3.3}

\citet{Goldberg2005} notes that the ellipsis site in \is{ellipsis, verb-phrase}VPE constructions can be inside an \isi{adjunct} island, while \isi{gapping} is not permitted in adjuncts. This is shown by the contrast between \REF{ex:johnson:34a} and \REF{ex:johnson:34b} below:
 
\ea
\ea\label{ex:johnson:34a} 
Lily finished her sandwich before Molly did.
\ex[*] {
Lily finished the sandwich before Molly the pizza.}\label{ex:johnson:34b}
\z
\z

 
The same contrast is found in \il{Ho-Chunk}Hocąk. The examples in \REF{ex:johnson:35} show that the gap in \is{ellipsis, verb-phrase}VPE constructions can be found inside \isi{adjunct} \isi{clauses} (which precede the main clause in these examples). In \REF{ex:johnson:35a}, the ellipsis site is in the clause headed by `if', in \REF{ex:johnson:35b} the ellipsis site is in the clause headed by `because', and in \REF{ex:johnson:35c} it is in the clause headed by `before.'
 
 
\ea\label{ex:johnson:35}
\ea\label{ex:johnson:35a} 
\glll Bryanga ųų kjanegi Meredithga Hunterga (nišge) {gišja hii} kjane.\\
Bryan-ga {\db}$\emptyset$-ųų kjane-gi Meredith-ga Hunter-ga (nišge) $\emptyset$-{gišja hii} kjane\\
Bryan-\textsc{prop} \textsc{3s}-do \textsc{fut}-if Meredith-\textsc{prop} Hunter-\textsc{prop} also \textsc{3s/o}-visit \textsc{fut}\\
\trans `Meredith will visit Hunter if Bryan will.'
 
\ex\label{ex:johnson:35b} 
\glll Bryanga hąąke ųųnįge Meredithga (nišge) hąąke Hunterga {gišja hiinį}.\\
Bryan-ga hąąke {\db}$\emptyset$-ųų-nį-ge Meredith-ga (nišge) hąąke Hunter-ga $\emptyset$-{gišja hii-nį}\\
Bryan-\textsc{prop} \textsc{neg} \textsc{3s}-do-\textsc{neg}-because Meredith-\textsc{prop} also \textsc{neg} Hunter-\textsc{prop} \textsc{3s/o}-visit-\textsc{neg}\\
\trans `Meredith didn't visit Hunter because Bryan didn't.'
 
\ex\label{ex:johnson:35c} 
\glll Keenį Sarahga ųųnį Matejaga {waisgap sguu xuwuxuwuhižą} ruucšąną.\\
keenį Sarah-ga {\db}$\emptyset$-ųų-nį Mateja-ga {waisgap sguu xuwuxuwu-hižą} $\emptyset$-ruuc-šąną\\
before Sarah-\textsc{prop} \textsc{3s}-do-\textsc{neg} Mateja-\textsc{prop} cookie-\textsc{indef} \textsc{3s/o}-eat-\textsc{decl}\\
\trans `Mateja ate a cookie before Sarah did.'
\z
\z

In contrast, \isi{gapping} is ungrammatical in adjuncts. \REF{ex:johnson:36} illustrates that the the gap cannot be located in an \isi{adjunct} \is{clauses}clause headed by `if' \REF{ex:johnson:36a}, `because' \REF{ex:johnson:36b} or `before' \REF{ex:johnson:36c}.

 
\ea\label{ex:johnson:36}
\ea[*]{ 
\glll Matejaga rookhožuhižągi Meredithga {waisgap sguuhižą} rook'į kjane.\\
Mateja-ga rookhožu-hižą-gi Meredith-ga {waisgap sguu-hižą} $\emptyset$-rook'į kjane\\
Mateja-\textsc{prop} pie-\textsc{indef}-if Meredith-\textsc{prop} cake-\textsc{indef} \textsc{3s/o}-bake \textsc{fut}\\
\trans (Intended: `Meredith will bake a cake if Mateja will bake a pie.')}\label{ex:johnson:36a}	
\ex[*] {
\glll Sarahga {wažą honąkipįnįhižąge} Matejaga wažątirehižą ruwį.\\
Sarah-ga {wažą honąkipįnį-hižą-ge} Mateja-ga wažątire-hižą $\emptyset$-ruwį\\
Sarah-\textsc{prop} bicycle-\textsc{indef}-because Mateja-\textsc{prop} car-\textsc{indef} \textsc{3s/o}-buy\\
\trans (Intended: `Mateja bought a car because Sarah bought a bicycle.')}\label{ex:johnson:36b}
\ex[*] {
\glll Keenį Bryanga {waisgap sguu xuwuxuwuhižąnį} Meredithga kšeehižą ruucšąną.\\
keenį Bryan-ga {waisgap sguu xuwuxuwu-hižą-nį} Meredith-ga kšee-hižą $\emptyset$-ruuc-šąną\\
before Bryan-\textsc{prop} cookie-\textsc{indef-neg} Meredith-\textsc{prop} apple-\textsc{indef} \textsc{3s/o}-eat-\textsc{decl}\\
\trans (Intended: `Meredith ate an apple before Bryan ate a cookie.')}\label{ex:johnson:36c}
\z
\z

\subsection{Ellipsis in embedded clauses}\label{sec:johnson:3.4}

\citet{Goldberg2005} also shows that the ellipsis site in \is{ellipsis, verb-phrase}VPE constructions can be inside an embedded \is{clauses}clause, while this is not true of other types of \isi{ellipsis}. \REF{ex:johnson:37a} demonstrates that \is{ellipsis, verb-phrase}VPE is licit in an embedded clause, while \REF{ex:johnson:37b}--\REF{ex:johnson:37c} illustrate that neither \isi{gapping} nor \isi{stripping} are possible in an embedded clause.

 
\ea\label{ex:johnson:37}
\ea\label{ex:johnson:37a} 
Lily went to the zoo, and I think (that) Molly did too.
\ex[*] {
Lily went to the zoo, and I think (that) Molly the aquarium.}\label{ex:johnson:37b}
\ex[*] {
Lily went to the zoo, and I think (that) Molly too.}\label{ex:johnson:37c}
\z
\z

 
In \il{Ho-Chunk}Hocąk, \is{ellipsis, verb-phrase}VPE is licit in the \isi{complement} \is{clauses}clause of various matrix verbs, including `know' \REF{ex:johnson:38a}, `want' \REF{ex:johnson:38b}, `think' \REF{ex:johnson:38c} and `say' \REF{ex:johnson:38d}.
 


 
\ea
\ea\label{ex:johnson:38a} 
\glll Bryanga hąąke {nįįtašjak taaxu} ruwįnį, nųnįge Meredithga ųųra yaaperesšąną.\\
Bryan-ga hąąke {nįįtašjak taaxu} $\emptyset$-ruwį-nį nųnįge Meredith-ga {\db}$\emptyset$-ųų-ra <ha>hiperes-šąną\\
Bryan-\textsc{prop} \textsc{neg} coffee \textsc{3s}-buy-\textsc{neg} but Meredith-\textsc{prop} \textsc{3s}-do-\textsc{comp} <\textsc{1s}>know-\textsc{decl}\\
\trans `Bryan didn't buy coffee, but I know Meredith did.'
\ex\label{ex:johnson:38b}
\glll Meredithga hąąke Hunterga {gišja hiinį} nųnįge Bryanga ųų roogų.\\
Meredith-ga hąąke Hunter-ga $\emptyset$-{gišja hii-nį} nųnįge Bryan-ga {\db}$\emptyset$-ųų $\emptyset$-roogų\\
Meredith-\textsc{prop} \textsc{neg} Hunter-\textsc{prop} \textsc{3s/o}-visit-\textsc{neg} but Bryan-\textsc{prop} \textsc{3s}-do \textsc{3s}-want\\
\trans `Meredith didn't visit Hunter, but Bryan wants to.'
\ex\label{ex:johnson:38c}
\glll Matejaga hąąke {wažą honąkipįnįhižą} ruwįnį, nųnįge Cecilga ųųže yaare.\\
Mateja-ga hąąke {wažą honąkipįnį-hižą} $\emptyset$-ruwį-nį nųnįge Cecil-ga {\db}$\emptyset$-ųų-že <ha>hire\\
Mateja-\textsc{prop} \textsc{neg} bicycle-\textsc{indef} \textsc{3s/o}-buy-\textsc{neg} but Cecil-\textsc{prop} \textsc{3s}-do-\textsc{comp} <\textsc{1s}>think\\
\trans `Mateja didn't buy a bicycle, but I think Cecil did.'
\ex\label{ex:johnson:38d}
\glll Sarahga hąąke waarucra hogihanį, nųnįge Meredithga ųųže ee.\\
Sarah-ga hąąke waaruc-ra $\emptyset$-hogiha-nį nųnįge Meredith-ga {\db}$\emptyset$-ųų-že $\emptyset$-ee\\
Sarah-\textsc{prop} \textsc{neg} table-\textsc{def} \textsc{3s/o}-paint-\textsc{neg} but Meredith-\textsc{prop} \textsc{3s}-do-\textsc{comp} \textsc{3s}-say\\
\trans `Sarah didn't paint the table, but Meredith said she did.'
\z
\z

Unlike \ili{English}, \il{Ho-Chunk}Hocąk does not exhibit any constraint on \isi{gapping} in embedded contexts. \REF{ex:johnson:39a} and \REF{ex:johnson:39b} show that the gap can be embedded under the verbs \emph{hire} `think' and \emph{ee} `say', respectively.

 
\ea
\ea\label{ex:johnson:39a} 
\glll Meredithga {wažą honąkipįnįhižą} ruwį anąga Bryanga wažątirehižą yaare.\\
Meredith-ga {wažą honąkipįnį-hižą} $\emptyset$-ruwį anąga Bryan-ga wažątire-hižą <ha>hire\\
Meredith-\textsc{prop} bicycle-\textsc{indef} \textsc{3s/o}-buy and Bryan-\textsc{prop} car-\textsc{indef} \textsc{<1s>}think\\
\trans `Meredith bought a bicycle, and I think that Bryan bought a car.'
\ex\label{ex:johnson:39b}
\glll Meredithga kšeehižą ruuc anąga Matejaga wažązihižą hihe.\\
Meredith-ga kšee-hižą $\emptyset$-ruuc anąga Mateja-ga wažązi-hižą <ha>ee\\
Meredith-\textsc{prop} apple-\textsc{indef} \textsc{3s/o}-eat and Mateja-\textsc{prop} orange-\textsc{indef} <\textsc{1s}>say\\
\trans `Meredith ate an apple and I said that Mateja ate an orange.'
\z
\z

 
The examples in \REF{ex:johnson:40} show that \il{Ho-Chunk}Hocąk exhibits \isi{stripping}. \REF{ex:johnson:40a} illustrates \isi{stripping} with an \isi{object} remnant after the coordinator\is{coordination} `and', while the example in \REF{ex:johnson:40b} has an \isi{object} remnant with disjunction. \REF{ex:johnson:40c} shows that \isi{stripping} is also possible with a \isi{subject} remnant after the coordinator.
 

\ea\label{ex:johnson:40} 
\ea\label{ex:johnson:40a} 
\glll Sarahga šųųkhižą haja, anąga wijukhižą šge.\\
Sarah-ga šųųk-hižą $\emptyset$-haja anąga wijuk-hižą šge\\
Sarah-\textsc{prop} dog-\textsc{indef} \textsc{3s/o}-see and cat-\textsc{indef} also\\
\trans `Sarah saw a dog, and a cat too.'
 \ex\label{ex:johnson:40b}
\glll Meredithga hąąke kšeehižą ruucnį, nųnįge {waisgap sguu xuwuxuwuhižą}.\\
Meredith-ga hąąke kšee-hižą $\emptyset$-ruuc-nį nųnįge {waisgap sguu xuwuxuwu-hižą}\\
Meredith-\textsc{prop} \textsc{neg} apple-\textsc{indef} \textsc{3s/o}-eat \textsc{neg} but cookie-\textsc{indef}\\
\trans `Meredith didn't eat an apple, but a cookie.'
 \ex\label{ex:johnson:40c}
\glll Bryanga {nįįtašjak taaxu} racgą, anąga Matejaga šge.\\
Bryan-ga {nįįtašjak taaxu} $\emptyset$-racgą anąga Mateja-ga šge\\
Bryan-\textsc{prop} coffee \textsc{3s}-drink and Mateja-\textsc{prop} also\\
\trans `Bryan drank coffee, and Mateja too.'
\z
\z


As is the case in \ili{English} and other languages, \isi{stripping} is ungrammatical in embedded \isi{clauses} in \il{Ho-Chunk}Hocąk. This is shown in \REF{ex:johnson:41a} for an \isi{object} remnant with \isi{conjunction}, \REF{ex:johnson:41b} for an \isi{object} remnant with disjunction and \REF{ex:johnson:41c} for a \isi{subject} remnant with \isi{conjunction}.
 
\ea
\ea[*]{\label{ex:johnson:41a}
\glll Matejaga wažahe gipį, anąga kšeexete šge yaare.\\
Mateja-ga wažahe $\emptyset$-gipį anąga kšeexete šge <ha>hire\\
Mateja-\textsc{prop} banana \textsc{3s}-like and pineapple also <\textsc{1s}>think\\
\trans (Intended: `Mateja likes bananas, and I think (she likes) pineapple too.')}
 
\ex[*]{\label{ex:johnson:41b}
\glll Bryanga hąąke wažątirehižą ruwįnį, nųnįge Cecilga {wažą honąkipįnįhižą} ee.\\
Bryan-ga hąąke wažątire-hižą $\emptyset$-ruwį-nį nųnįge Cecil-ga {wažą honąkipįnį-hižą} $\emptyset$-ee\\
Bryan-\textsc{prop} \textsc{neg} car-\textsc{indef} \textsc{3s/o}-buy-\textsc{neg} but Cecil-\textsc{prop} bicycle-\textsc{indef} \textsc{3s}-say\\
\trans (Intended: `Bryan didn't buy a car, but Cecil said (he bought) a bicycle.')
}

\ex[*]{\label{ex:johnson:41c}
\glll Sarahga {waisgap sguuhižą} rook'į, anąga Bryanga Meredithga šge ee.\\
Sarah-ga {waisgap sguu-hižą} $\emptyset$-rook'į anąga Bryan-ga Meredith-ga šge $\emptyset$-ee\\
Sarah-\textsc{prop} cake-\textsc{indef} \textsc{3s/o}-bake and Bryan-\textsc{prop} Meredith-\textsc{prop} also \textsc{3s}-say\\
\trans (Intended: `Sarah baked a cake, and Bryan said Meredith (baked a cake) too.')
}
 
\z
\z

 
To conclude, the possibilities of having an ellipsis site in embedded contexts differ between \ili{English} and \il{Ho-Chunk}Hocąk: \is{ellipsis, verb-phrase}VPE and \isi{gapping} are not differentiated by embedding, but \is{ellipsis, verb-phrase}VPE and \isi{stripping} are. However, \isi{gapping} and \is{ellipsis, verb-phrase}VPE are still distinguished in \isi{adjunct} \isi{clauses}: as we saw in 3.3, \is{ellipsis, verb-phrase}VPE is grammatical in \isi{adjunct} \isi{clauses} \REF{ex:johnson:35} while \isi{gapping} is not \REF{ex:johnson:36}.
 

\subsection{Presence of strict and sloppy readings\is{sloppy readings}}\label{sec:johnson:3.5}

Another characteristic of \is{ellipsis, verb-phrase}VPE is the fact that elided pronouns and anaphors give rise to two different identity readings. The \ili{English} example in \REF{ex:johnson:42} is ambiguous. Under the so-called ``strict'' reading, the referent of the pronoun is identical in both the antecedent and elided \is{verb phrase}VP. Under the ``sloppy'' reading, the pronoun behaves like a variable, and the referent of the anaphor is different for each conjunct\is{conjunction}.
  
\ea\label{ex:johnson:42} 
Lily saw herself in the mirror, and Molly did too.\\
\textit{Strict reading:} Molly saw Lily in the mirror.\\
\textit{Sloppy reading\is{sloppy readings}:} Molly saw herself in the mirror.
\z

\citet{Fortin2007} shows that \isi{stripping} also gives rise to both strict and \isi{sloppy readings}, as in the example in \REF{ex:johnson:43}. However, there is another possible interpretation for the second conjunct\is{conjunction}: the remnant can be interpreted as the \isi{object} of the stripped \is{clauses}clause. Fortin terms this additional reading the ``\isi{object} reading.'' This third reading is unique to \isi{stripping} constructions, as the remnant DP in \is{ellipsis, verb-phrase}VPE is always interpreted as the \isi{subject} of the elided constituent. 
  
\ea\label{ex:johnson:43} 
Lily saw herself in the mirror, and Molly too.\\
\textit{Strict reading:} Molly saw Lily in the mirror.\\
\textit{Sloppy reading\is{sloppy readings}:} Molly saw herself in the mirror.\\
\textit{Object reading:} Lily saw Molly in the mirror.
\z

 
In \il{Ho-Chunk}Hocąk, strict and \isi{sloppy readings} are available with both \is{ellipsis, verb-phrase}VPE and \isi{stripping}, while the additional ``\isi{object} reading'' is possible only with \isi{stripping}. In the examples in \REF{ex:johnson:44}, the antecedent \is{verb phrase}VP contains a \isi{possessed} \isi{object}. \REF{ex:johnson:44a} is an instance of \is{ellipsis, verb-phrase}VPE, and the second conjunct\is{conjunction} has two possible interpretations: either Hunter visited Bryan's mother (strict reading) or Hunter visited his own mother (sloppy reading\is{sloppy readings}). In \REF{ex:johnson:44b}, the second conjunct contains a \isi{stripping} \isi{ellipsis} site. Both the strict and \isi{sloppy readings} are available, but the \isi{object} reading is also possible: the sentence could mean that Bryan visited Hunter.
 
 
\ea\label{ex:johnson:44}
\ea\label{ex:johnson:44a} 
\glll Bryanga hi'ųnį hiira homąkįnį anąga Hunterga šge ųų.\\
Bryan-ga hi'ųnį $\emptyset$-hii-ra $\emptyset$-homąkįnį anąga Hunter-ga šge {\db}$\emptyset$-ųų\\
Bryan-\textsc{prop} mother \textsc{3s-poss-def} \textsc{3s/o}-visit and Hunter-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Bryan visited his mother, and Hunter did too.'
\ex\label{ex:johnson:44b}
\glll Bryanga hi'ųnį hiira homąkįnį anąga Hunterga šge.\\
Bryan-ga hi'ųnį $\emptyset$-hii-ra $\emptyset$-homąkįnį anąga Hunter-ga šge\\
Bryan-\textsc{prop} mother \textsc{3s-poss-def} \textsc{3s/o}-visit and Hunter-\textsc{prop} also\\
\trans `Bryan visited his mother, and Hunter too.'
\z
\z

The examples in \REF{ex:johnson:45} show that the same readings are possible with reflexives. The second conjunct\is{conjunction} of \REF{ex:johnson:45a} contains a \is{ellipsis, verb-phrase}VPE gap, and it has two interpretations: either Meredith hit Mateja (sloppy) or Meredith hit herself (strict). In the \isi{stripping} example in \REF{ex:johnson:45b}, both strict and \isi{sloppy readings} are possible, but so is the ``\isi{object} reading'' under which Mateja hit Meredith.
 
\ea\label{ex:johnson:45}
\ea\label{ex:johnson:45a} 
\glll Matejaga hokijį anąga Meredithga šge ųų.\\
Mateja-ga $\emptyset$<kii>hojį anąga Meredith-ga šge {\db}$\emptyset$-ųų\\
Mateja-\textsc{prop} \textsc{3s}<\textsc{refl}>hit and Meredith-\textsc{prop} also {\db}\textsc{3s}-do\\
\trans `Mateja hit herself, and Meredith did too.'

\ex\label{ex:johnson:45b}
\glll Matejaga hokijį anąga Meredithga šge.\\
Mateja-ga $\emptyset$<kii>hojį anąga Meredith-ga šge\\ 
Mateja-\textsc{prop} \textsc{3s}<\textsc{refl}>hit and Meredith-\textsc{prop} also\\ 
\trans `Mateja hit herself, and Meredith too.'
\z
\z

Thus, while strict and \isi{sloppy readings} are available with both \is{ellipsis, verb-phrase}VPE and \isi{stripping}, \isi{stripping} constructions have the additional reading that \citet{Fortin2007} calls the ``\isi{object} reading''.


\section{Deletion vs. pro-form analysis}\label{sec:johnson:4}

 
In the previous two sections, I presented arguments that \il{Ho-Chunk}Hocąk exhibits \is{ellipsis, verb-phrase}VPE. In this section, I further argue that \is{ellipsis, verb-phrase}VPE in \il{Ho-Chunk}Hocąk is derived by a deletion process. There are two main approaches to any given elliptical phenomena: the \isi{ellipsis} site is either a deleted phrase or a null \emph{pro}-form. Here, I extend two arguments in favor of a deletion approach of \ili{English} \is{ellipsis, verb-phrase}VPE to \il{Ho-Chunk}Hocąk. First, I show that extraction from the \isi{ellipsis} site is possible. Second, I demonstrate that \isi{ellipsis} sites can contain the antecedent to a pronoun outside of the gap.
 

\citet{FiengoMay1994} argue that \ili{English} \is{ellipsis, verb-phrase}VPE is best analyzed as \is{verb phrase}VP deletion. Their argument is based on cases of \isi{object} extraction from the \isi{ellipsis} site. In \REF{ex:johnson:46a}, we see that the \isi{object} of the second \is{clauses}clause has undergone \emph{wh}-movement out of the \isi{ellipsis} site. \REF{ex:johnson:46b} illustrates the phenomenon known as antecedent-contained deletion (ACD). In ACD constructions, the \isi{ellipsis} site is found inside of a \is{clauses, relative}relative clause and is licensed under identity with the matrix \is{verb phrase}VP. The head of the \is{clauses, relative}relative clause (here, \emph{everyone}) is the \isi{object} of the elided \is{verb phrase}VP. In both \REF{ex:johnson:46a} and \REF{ex:johnson:46b}, movement of the \isi{object} in the elided \is{verb phrase}VP has taken place. This is not expected under a \emph{pro}-form analysis of \is{ellipsis, verb-phrase}VPE: a \emph{pro}-form has no internal structure, and thus there should be no \isi{object} position inside the \isi{ellipsis} site that the extracted \isi{object} could have originated in. In contrast, a deletion analysis posits a full-fledged \is{verb phrase}VP in the \isi{ellipsis} site which undergoes deletion at a later stage in the derivation. In the examples in \REF{ex:johnson:46}, the \isi{object} originated inside the elided \is{verb phrase}VP, and underwent movement before deletion took place.
 
\ea\label{ex:johnson:46}
\ea\label{ex:johnson:46a}
I know which book Max read, and \textbf{which book} Oscar didn't.
\ex\label{ex:johnson:46b} 
Dulles suspected everyone who Angleton did. (\citealt[229]{FiengoMay1994}, 257)
\z
\z

Likewise, \il{Ho-Chunk}Hocąk constructions with \emph{ųų} cannot be analyzed as a \emph{pro}-form, as \isi{object} extraction is permitted. \REF{ex:johnson:47a} shows that focused elements can be extracted from the \isi{ellipsis} site, and \REF{ex:johnson:47b} exemplifies the movement of \emph{wh}-words from the \isi{ellipsis} site.\footnote{Like other Siouan languages, \il{Ho-Chunk}Hocąk is a \emph{wh}-in-situ language. However, \emph{wh}-words can undergo focus driven movement.}
 
\ea
\ea\label{ex:johnson:47a} 
\glll Meredithga waagaxra ruw\k{\i}, nųnįge \textbf{wiiwagaxra} hąąke ųųnį.\\
Meredith-ga waagax-ra $\emptyset$-ruwį, nųnįge wiiwagax-ra hąąke {\db}$\emptyset$-ųų-nį\\
Meredith-\textsc{prop} paper-\textsc{def} \textsc{3s/o}-buy but pencil-\textsc{def} \textsc{neg} \textsc{3s}-do-\textsc{neg}\\
\trans `Meredith bought the paper, but the pencil, she didn't.'
 
\ex\label{ex:johnson:47b} 
\glll Jaagu Bryanga ruwįra yaaperesšąną, nųnįge \textbf{jaagu} Hunterga ųųra hąąke yaaperesnį.\\
Jaagu Bryan-ga $\emptyset$-ruwį-ra <ha>hiperes-šąną nųnįge jaagu Hunter-ga {\db}$\emptyset$-ųų-ra hąąke <ha>hiperes-nį\\
what Bryan-\textsc{prop} \textsc{3s/o}-buy-\textsc{comp} <\textsc{1s}>know-\textsc{decl} but what Hunter-\textsc{prop} \textsc{3s}-do-\textsc{comp} \textsc{neg} <\textsc{1s}>know-\textsc{neg}\\
\trans `I know what Bryan bought, but I don't know what Hunter did.'
\z
\z

 
As the example in \REF{ex:johnson:48} shows, ACD is also grammatical in \il{Ho-Chunk}Hocąk. ACD would not be possible if \emph{ųų} were a \emph{pro}-form, since the head of the \is{clauses, relative}relative clause is the \isi{object} of the elided \is{verb phrase}VP.
 
 
\ea\label{ex:johnson:48} 
\glll Bryanga ruwį, jaagu Meredithga ųųra.\\
Bryan-ga $\emptyset$-ruwį jaagu Meredith-ga {\db}$\emptyset$-ųų-ra\\
Bryan-\textsc{prop} \textsc{3s/o}-buy what Meredith-\textsc{prop} \textsc{3s}-do-\textsc{comp}\\
\trans `Bryan bought what(ever) Meredith did.'
\z

 
The second argument in favor of a deletion analysis of \is{ellipsis, verb-phrase}VPE in \il{Ho-Chunk}Hocąk comes from so-called ``missing antecedents.'' \citet{HankamerSag1976} demonstrate that the gap in \ili{English} \is{ellipsis, verb-phrase}VPE constructions can contain the antecedent to a pronoun. In the non-elliptical example in \REF{ex:johnson:49a}, the DP \emph{a camel} in the second conjunct\is{conjunction} serves as the antecedent for the pronoun \emph{it} in the third conjunct. In \REF{ex:johnson:49b}, the \is{verb phrase}VP in the second conjunct is elided, resulting in a missing antecedent for the pronoun \emph{it}. Nonetheless, the sentence is still grammatical. It is important to note that the instance of \emph{a camel} in the first conjunct\is{conjunction} cannot be the antecedent for the pronoun \emph{it}: as \REF{ex:johnson:50} shows, DPs under the scope of \isi{negation} cannot serve as antecedents for pronouns.
 
\ea
\ea\label{ex:johnson:49a}
I've never ridden a camel, but Ivan's ridden a camel$_i$, and he says it$_i$ stank horribly.
\vspace{12pt}
\ex\label{ex:johnson:49b} 
I've never ridden a camel, but Ivan has, and he says it$_i$ stank horribly. (\citealt[403]{HankamerSag1976})
\z
\z
 
\ea\label{ex:johnson:50}
*I've never ridden a camel$_i$, and it$_i$ stank horribly. (\citealt[404]{HankamerSag1976})\\
\z 

\citet{HankamerSag1976} argue that the grammaticality of the example in \REF{ex:johnson:49b} points to a deletion analysis of \is{ellipsis, verb-phrase}VPE. These facts are not readily explained under a \emph{pro}-form analysis: since the \isi{ellipsis} site would not have internal structure at any point in the derivation, the elided \is{verb phrase}VP in \REF{ex:johnson:49b} would never contain the antecedent for the following pronoun. 

 
Examples of \is{ellipsis, verb-phrase}VPE with missing antecedents are also grammatical in \il{Ho-Chunk}Hocąk. In \REF{ex:johnson:51a}, the DP \emph{kšeexetehižą} `a pineapple' in the second conjunct\is{conjunction} is the antecedent for the null pronominal \isi{subject} of the verb \emph{sguu} `sweet'. In \REF{ex:johnson:51b}, the \is{verb phrase}VP containing the antecedent is elided, and the resulting sentence is grammatical. Like \ili{English}, a pronoun cannot find its antecedent in a negated \is{clauses}clause \REF{ex:johnson:52}.
 
 
\ea
\ea\label{ex:johnson:51a} 
\glll Hąkaga kšeexetehižą haacnį, nųnįge Matejaga kšeexetehižą ruuc, anąga sguu ee.\\
hąkaga kšeexete-hižą $\emptyset$<ha>ruuc-nį nųnįge Mateja-ga kšeexete-hižą $\emptyset$-ruuc anąga $\emptyset$-sguu $\emptyset$-ee\\
never pineapple-\textsc{indef} \textsc{3s<1s>}eat-\textsc{neg} but Mateja-\textsc{prop} pineapple-\textsc{indef} \textsc{3s/o}-eat and \textsc{3s}-sweet \textsc{3s}-say\\
\trans `I never ate a pineapple, but Mateja ate a pineapple, and she said it was sweet.'
 
\ex\label{ex:johnson:51b} 
\glll Hąkaga kšeexetehižą haacnį, nųnįge Matejaga ųų, anąga sguu ee.\\
hąkaga kšeexete-hižą $\emptyset$<ha>ruuc-nį nųnįge Mateja-ga {\db}$\emptyset$-ųų anąga $\emptyset$-sguu $\emptyset$-ee\\
never pineapple-\textsc{indef} \textsc{3s<1s>}eat-\textsc{neg} but Mateja-\textsc{prop} \textsc{3s}-do and \textsc{3s}-sweet \textsc{3s}-say\\
\trans `I never ate a pineapple, but Mateja did, and she said it was sweet.'
\z
\z
 
\ea\label{ex:johnson:52} 
\glll Hąkaga kšeexetehižą haacnį anąga sguu.\\
hąkaga kšeexete-hižą $\emptyset$<ha>ruuc-nį anąga $\emptyset$-sguu\\
never pineapple-\textsc{indef} \textsc{3s<1s>}eat-\textsc{neg} and \textsc{3s}-sweet\\
\trans `I never ate a pineapple, and it was sweet.'
\z


Both the extraction facts and pronoun antecedent facts point to an analysis in which the contents of elided VPs in \il{Ho-Chunk}Hocąk are present syntactically, and that the omission of elided VPs is due to a deletion process.

\section{Conclusion}\label{sec:johnson:5}

 
In this paper, I examined an elliptical phenomenon that I argue instantiates \is{ellipsis, verb-phrase}VPE in \il{Ho-Chunk}Hocąk. This process targets all VP-internal material, including direct objects, indirect objects, result phrases, temporal adjuncts, locative adjuncts, comitatives, manner adverbs and \isi{complement} \isi{clauses}. \is{ellipsis, verb-phrase}VPE is conditioned by the presence of a licensing head, which I showed is the \isi{light verb} \emph{ųų} in \il{Ho-Chunk}Hocąk. However, \il{Ho-Chunk}Hocąk \is{ellipsis, verb-phrase}VPE is constrained in that the antecedent verb must be active. I propose that this restriction is due to the fact that active \emph{v} is the licenser. This elliptical process displays many other traits that \citet{Goldberg2005} and \citet{Fortin2007} demonstrate are characteristic of \is{ellipsis, verb-phrase}VPE crosslinguistically. I also briefly discussed that \il{Ho-Chunk}Hocąk \is{ellipsis, verb-phrase}VPE should be analyzed as \is{verb phrase}VP deletion, rather than a \is{verb phrase}VP \emph{pro}-form. This paper constitutes the first in depth description of \is{ellipsis, verb-phrase}VPE in \il{Ho-Chunk}Hocąk, and contributes to the literature on the properties of \is{ellipsis, verb-phrase}VPE crosslinguistically. 
 

\section*{Acknowledgment}

First and foremost, I would like to thank Cecil Garvin\ia{Garvin, Cecil} for sharing his language with me and kindly answering my endless questions. Thanks also to Yafei Li\ia{Li, Yafei}, Bryan Rosen\ia{Rosen, Bryan} and Rand Valentine\ia{Valentine, Rand} for discussion and comments at various stages of this project. A portion of this work was presented at the 87\textsuperscript{th} Annual Meeting of the Linguistic Society of America. I thank the audience there for questions and comments. I would also like to thank an anonymous reviewer and Catherine Rudin\ia{Rudin, Catherine} for comments on an earlier draft of this paper. Lastly, thanks to Iren Hartmann\ia{Hartmann, Iren} for generously providing me access to her Lexique Pro \il{Ho-Chunk}Hocąk \isi{dictionary} to assist with my fieldwork.
\section*{Abbreviations}
\begin{tabularx}{.45\textwidth}{lX}
1, 2, 3 & first, second, third person\\
\textsc{comp} & complementizer \\
 \textsc{decl} & declarative \\
 \textsc{def} & definite \\
 \textsc{fut} & future \\
 \textsc{indef} & indefinite \\
 \textsc{neg} & negative \\
 \textsc{o} & object agreement \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lX}
 \textsc{pl} & plural \\
 \textsc{poss} & possessive \\
 \textsc{prop} & proper name \\
 \textsc{pst} & past tense \\
 \textsc{refl} & reflexive \\
 \textsc{s} & {subject} agreement \\
 \textsc{stat} = stative verb.
\end{tabularx}

{\sloppy
 \printbibliography[heading=subbibliography,notkeyword=this]
}
\end{document}






