\addchap{Acknowledgments}

Firstly, I would like to thank all the native speakers of Tuatschin who offered me their time and patience. Special thanks are due to Tarcisi Hendry, who established the contact with almost all my consultants. The other native speakers who were of great help are, in alphabetical order, Leci Brugger, Nicolina Cathomen-Giossi, Nescha Cavegn, Leci Deflorin, Pascal Gamboni, Lucia Giossi, Vigeli Giossi, Beatrix Graf, Beni Hendry, Cecilia Hendry, Wendelin Hendry, Nina Levy, Laura Loretz, Nicoletta Marino, Ciril Monn, Giusep Monn, and Riccardo Monn.

I would also like to thank Sabine Stoll and Michele Loporcaro for having asked me to participate in their project on the acquisition of Tuatschin (see below).

Martin Haspelmath, Michele Loporcaro, and the five anonymous reviewers of Language Science Press made precious comments about earlier versions of this grammar.

Claudia Cathomas and Tresa Bundi-Pally helped me with some transcriptions, and Sebastian Nordhoff as well as Felix Kopecky helped me to resolve the many difficulties I had with TexStudio and LaTex.

Alys Boote Cooper was kind enough to correct my English.

To all these people I would like to say \textit{in grond grazia fetg!}

Last but not least, many thanks to the Swiss National Science Foundation (project no. 159544 about the acquisition of Tuatschin Sursilvan by children between 2 and 6 years) who covered my hotel and meal 
expenses.


\begin{flushright}
	Wetzikon (Switzerland), November 2020
\end{flushright}
