\chapter*{Appendix A: Texts}\addcontentsline{toc}{chapter}{Appendix A: Texts}\markboth{Appendix A: Texts}{Appendix A: Texts}\label{app-texts}

\section*{The owl and I}\addcontentsline{toc}{section}{The owl and I}\markboth{Appendix A: Texts}{The owl and I}

\noindent This text is a written personal narrative by Shantila Jimi (Kathmandu/Tumok). 

\setcounter{ExNo}{0}

%\ref 42\_leg\_10.001
\exg.phopciba=nuŋ   ka\\
 owl{\sc =com} {\sc 1sg}\\
The owl and I

%\ref 42\_leg\_10.002
\exg.pyak encho         ka  miya   sa-ya-ŋ=niŋa                pyak chocholaplap cok-khuba      saŋ.\\
 very long\_time\_ago {\sc 1sg} little {\sc cop.pst-pst-1sg=ctmp} very mischievous  do{\sc -nmlz} {\sc cop.pst.1sg}\\
Long time ago, when I was little, I was very mischievous.

%\ref 42\_leg\_10.003
\exg.i    len=ga    ceʔya om,  aniŋ=ga  yo taŋŋoca weʔ=na lamdhaŋ=be    ka,  a-ma=nuŋ  a-na leks-i-ŋ=hoŋ, phekme ok-se khe-i-ŋ.\\
 one day=gen matter {\sc cop} {\sc 1pl.excl.poss=gen} across  Tangoca\_field   be{\sc =nmlz.sg} field{\sc =loc} {\sc 1sg} {\sc 1sg.poss-}mother{\sc =com}  {\sc 1sg.poss-}elder\_sister be{\sc -1pl-excl=seq}  vetch  dig\_up{\sc -sup} go{\sc -1pl-excl}\\
This is about one day, my mother, my elder sister and I went to dig our field to plant lentils, there at the Tangoca field.

%\ref 42\_leg\_10.004
\exg.kaniŋ   kondarik=ci=ca                         khet-u-m-ci-m-ŋa=hoŋ                                 lamdhaŋ=be    khe-i-ŋ.\\
 {\sc 1pl[erg]} spade\_with\_long\_handle{\sc =nsg=add} take\_along{\sc -3.P-1pl.A-3nsg.P-1pl.A-excl=seq}  field{\sc  =loc} go{\sc -1pl-excl}\\
We also took some spades with us and went to the field.

%\ref 42\_leg\_10.005
\exg.nna  lamdhaŋ paŋ=bhaŋ    ta-ma      maŋpha maŋdu sa-ya.\\
 that field    house{\sc =abl}  come{\sc -inf} much    far    {\sc cop.pst-pst[3sg]}\\
The field was quite far from our house.

%\ref 42\_leg\_10.006
\exg.kaniŋ lamdhaŋ=be    tas-u-m-ŋa=hoŋ                       phekme ok-saŋ          khe-i=niŋa         nam=ŋa   tuknuŋ    pho-ya.\\
 {\sc 1pl[erg]}   field{\sc =loc} arrive{\sc -3.P[pst]-1pl.A-excl=seq} vetch  dig\_up{\sc -sim} go{\sc  -1pl=ctmp} sun{\sc =erg} thoroughly [sun]burn{\sc -pst[3sg]}\\
When we arrived at the field and began to dig the weed, the sun burnt down on us mercilessly.

%\ref 42\_leg\_10.007
\exg.ŋkhaʔniŋgo nna=be     namciʔmaŋ heʔne=ca        m-ma-ya-n.\\
 but           that{\sc =loc} shady       where{\sc =add} {\sc neg-cop.neg -pst[3sg]-neg}\\
But there was no shady place.

%\ref 42\_leg\_10.008
\exg.ŋkhaʔniŋ kaniŋ whaŋma pirik-pirik    leks-i-ŋ=hoŋ               ibebe    wepma  si-i-ŋ.\\
 that\_time   {\sc 1pl}   sweat   running become{\sc -1pl-excl=seq}  too\_much thirst die{\sc -1pl -excl}\\
At that time we were covered in sweat, and we were very thirsty.

%\ref 42\_leg\_10.009
\exg.nnakha=he    maŋcwa=ca        m-ma-ya-n.\\
 that{\sc =loc} water{\sc =add} {\sc neg-cop.neg-pst[3sg]-neg}\\
There was no water, too.

%\ref 42\_leg\_10.010
\exg.nna  lamdhaŋ=ga    yorok=ŋa        eko mina  hoŋma wa-ya.\\
 that field{\sc =gen} across{\sc =ins} one small river  exist{\sc -pst[3sg]}\\
There was a small stream, a bit farther from the field.

%\ref 42\_leg\_10.011
\exg.ka  a-ma=ŋa                khaʔla   ly-a-ŋ:\\
 {\sc 1sg} {\sc 1sg.poss-}mother{\sc =erg} like\_this say{\sc -pst-1sg.P}\\
Mother told me like this:

%\ref 42\_leg\_10.012
\exg.phuama,         hoŋma=be    khy-a-g=hoŋ             ci=ha                  maŋcwa apt-u           au!\\
 last-born\_girl river{\sc =loc} go{\sc -sbjv-2=seq}  be\_cold{\sc =nmlz.nc} water   bring{\sc -3.P[imp]} {\sc excla}\\
Daughter, go to the stream and bring some cold water, would you?

%\ref 42\_leg\_10.013 - check 3sg.poss.....
\exg.nhaŋa   ka=ca         leks-a=ba                ka-saŋ       con-si-saŋ                    u-yaŋdhaŋ eko a-muk=ŋa             phok-phok      mok-saŋ       khet-u-ŋ=hoŋ                      maŋcwa tap-se         hoŋma=be    khy-a-ŋ.\\
 and\_then {\sc 1sg=add} be{\sc -pst[3sg]=emph} say{\sc -sim} be\_happy{\sc -mddl-sim} {\sc 3sg.poss-}pot
 one {\sc 1sg.poss-}hand{\sc =ins} drumming\_sound beat{\sc -sim} take\_along{\sc -3.P[pst]-1sg.A=seq}  water   take{\sc -sup} river{\sc =loc} go{\sc -pst-1sg}\\
And I responded: \rede{Okay}, and went happily to the stream to fetch water, holding the pot and drumming it with one hand.

%\ref 42\_leg\_10.014
\exg.kheʔ-ma-sy-a=niŋa                  lamdhaŋ=ga    miʔyaŋ yorok=ŋa        tas-uks-a-ŋ=na.\\
 go{\sc -inf-aux.prog-pst=ctmp} field{\sc =gen} little   over\_there{\sc =ins} arrive{\sc -prf-pst[3.P]-1sg.A=nmlz.sg}\\
On the way, I came to a place a little further away from the field.

%\ref 42\_leg\_10.015
\exg.ŋkhaʔniŋ eko khokpu=ga      u-thap=ka              u-sam=be                    usa   keks-a-masa=ha                               nis-u-ŋ.\\
 that\_time   one fig\_tree{\sc =gen} {\sc 3sg.poss-}plant{\sc =gen} {\sc 3sg.poss-}lower\_part{\sc =loc} fruit bear\_fruit{\sc -pst[3sg]-pst.prf =nmlz.nsg} see{\sc -3.P[pst]-1sg.A}\\
Then, I saw some fruits hanging from the lower branches of a fig tree.

%\ref 42\_leg\_10.016
\exg.ka  ŋkha usa   sem-ca-ma            mit-u-ŋ,               ŋkhoŋ sem-saŋ        ekhumdu-ekhumdu co-het-u-ŋ.\\
 {\sc 1sg[erg]} those fruit pluck{\sc -V2.eat-inf} want{\sc -3.P[pst]-1sg.A} thus    pluck{\sc -sim} delicious       eat{\sc -V2.carry.off-3.P[pst]-1sg.A}\\
I wanted to pluck and eat the fruits, and so I did pluck them and ate them, enjoying their taste.

%\ref 42\_leg\_10.017
\exg.usa   sem-saŋ u-sam=be  ca-ma-sy-a-ŋ=niŋ   thapthum=be  u-thap=phaŋ a-muk=pe    makhur=ha             waghui     loʔwa ʈhwaŋ     nam-khuba        a-yuks-a.\\
 fruit pluck{\sc -sim} {\sc 3sg.poss-}under{\sc =loc} eat{\sc -inf-aux.prog-pst-1sg=ctmp} cluelessness{\sc =loc} {\sc 3sg.poss-}plant{\sc =abl}  {\sc 1sg.poss-}arm{\sc =loc} black{\sc =nmlz.nc} chicken\_droppings like   smelling\_awful\_suddenly smell{\sc -nmlz} fall{\sc -V2.come.down-pst[3sg]}\\
While I was eating under the tree, without me noticing, something black, smelling awfully like chicken droppings, dropped on my hand.

%\ref 42\_leg\_10.018
\exg.i=ha                a-yuks-a=ha                                     bhoŋ khaʔla   siŋ=choŋ=be        so-ŋ=niŋa=go,              phopciba=le          weʔ=na!\\
 what{\sc =nmlz.nc} fall{\sc -V2.come.down-pst[3sg]=nmlz.nc} {\sc comp}  like\_this tree{\sc =top=loc} look{\sc -1sg.A=ctmp=top}  owl{\sc =contr} be{\sc [3sg]=nmlz.sg}\\
When I looked up into the tree what had fallen down, there is an owl!

%\ref 42\_leg\_10.019
\exg.nna  phopciba siliklik    a-chik           ekt-u-get-u.\\
 that owl      fuming\_at\_sth {\sc 1sg.poss-}anger be\_angry{\sc -3.P-V2.bring.up-3.P[pst]}\\
That owl made me really angry.

%\ref 42\_leg\_10.020
\exg.lap-ma=ga         bhoŋ khokpu=ga      siŋ=be    thaŋ-ma=ga       cog-a-ŋ.\\
 catch{\sc -inf=gen} {\sc purp} fig\_tree{\sc =gen} tree{\sc =loc} climb{\sc -inf=gen} try{\sc -pst-1sg}\\
I tried to climb the tree to catch it.

%\ref 42\_leg\_10.021
\exg.ŋkhaʔniŋgo nna  siŋ maɖa sa-ya.\\
 but           that tree big   {\sc cop.pst -pst[3sg]}\\
But the tree was huge.

%\ref 42\_leg\_10.022
\exg.ŋkhoŋ=ca        a-chik           ekt-a=na                        belak=ŋa  esap-esap thaŋ-ma    yas-a-ŋ.\\
 thus{\sc =add} {\sc 1sg.poss-}anger be\_angry{\sc -pst[3sg]=nmlz.sg} time{\sc =ins} swiftly   climb{\sc -inf} be\_able{\sc -pst-1sg}\\
Nevertheless, as I was so angry that I managed to climb up.

%\ref 42\_leg\_10.023
\exg.to  thaŋ-a-by-a-ŋ=hoŋ                       khaʔla   so-ŋ=niŋa=go              eko maŋpha=na        hoŋ nis-u-ŋ.\\
 up  climb{\sc -pst-V2.give-pst-1sg=seq}  like\_this look{\sc -1sg.A=ctmp=top}  one much{\sc =nmlz.sg} hole see{\sc -3.P[pst]-1sg.A}\\
When I had climbed up into the tree and looked, I saw a large hole.

%\ref 42\_leg\_10.024
\exg.nhaŋa   phopciba lap-ma=go          men=na,            nna  hoŋ=be    i=ha=le                         weʔ=na        bhoŋ soʔ-ma    mit-het-u-ŋ.\\
 and\_then owl      catch{\sc -inf=top}  {\sc neg.cop=nmlz.sg} that hole{\sc =loc} what{\sc =nmlz.nc=contr} be{\sc [3sg]=nmlz.sg} {\sc comp}  look{\sc -inf} want{\sc -V2.carry.off-3.P[pst]-1sg.A}\\
Then, I wanted to see what was in the hole, even more than I wanted to catch the owl.

%\ref 42\_leg\_10.025
\exg.ŋkhoŋ nna  hoŋ=be    so-ŋ, nna=be=go         phopciba=ga    hop=le           waisa!\\
 thus    that hole{\sc =loc} look{\sc -[3.P;pst]-1sg.A,} that{\sc =loc=top}  owl{\sc =gen} nest{\sc =contr} {\sc cop.pst[3]}\\
Then I looked into that hole, and there was an owl's nest!


%\ref 42\_leg\_10.027
\exg.nna  o-hop=pe=go                  hicci u-in=ci=ca                     ŋ-wa-ya.\\
 that {\sc 3sg.poss-}nest{\sc =loc=top}  two   {\sc 3sg.poss-}egg{\sc =nsg=add} {\sc 3pl-}be{\sc -pst}\\
In that nest there were two eggs, too.

%\ref 42\_leg\_10.028
\exg.ŋkha u-in=ci              taŋ-se            bhoŋ a-muk           end-u-ŋ=niŋa=go  phopciba piciŋgelek-piciŋgelek ok-saŋ pes-a-ra-ya.\\
 those {\sc 3sg.poss-}egg{\sc =nsg} take\_out{\sc -sup} {\sc purp}  {\sc 1sg.poss-}hand insert{\sc -3.P[pst]-1sg.A=ctmp=top}  owl      shrieking\_loudly        peck{\sc -sim} fly{\sc -pst-V2.come-pst[3sg]}\\
When I put my hand into the nest to take out the eggs, the owl flew towards me, screeching; it wanted to peck me.

%\ref 42\_leg\_10.029
\exg.nhaŋa   a-ʈukhuruk=pe            pok=hoŋ             pok           og-a-ŋ.\\
 and\_then {\sc 1sg.poss-}head{\sc =loc} pecking\_sound{\sc =seq}  pecking\_sound peck{\sc -pst-1sg.P}\\
Then it began to peck me on my head.

%\ref 42\_leg\_10.030
\exg.ka  tuknuŋ    kisit-a-ŋ=hoŋ mo   uŋ-ma=ga cog-a-ŋ=niŋa=go a-laŋ=ŋa tu-ŋ=na siŋ=ga    whak   ʈek           eg-a.\\
 {\sc 1sg} thoroughly be\_afraid{\sc -pst-1sg=seq}  down come\_down{\sc -inf=gen} try{\sc -pst-1sg=ctmp=top}
 {\sc 1sg.poss-}foot{\sc =ins} step\_on{\sc -1sg.A[3.P;pst]=nmlz.sg} tree{\sc =gen} branch breaking\_sound break{\sc -pst[3sg]}\\
I was really afraid, and when I tried to get down from the tree, the branch on which I stood cracked.

%\ref 42\_leg\_10.031
\exg.ka  seŋ          kaks-a-kh-eks-a-sa-ŋ=na.\\
 {\sc 1sg} falling\_sound fall\_down{\sc -pst-V2.go-V2.cut-pst-pst.prf-1sg=nmlz.sg}\\
I was about to fall down.

%\ref 42\_leg\_10.032
\exg.heko=na         whak=pe      a-tek              het-u=hoŋ              ka  haŋcaŋcaŋ chu-ya-ŋ.\\
 other{\sc =nmlz.sg} branch{\sc =loc} {\sc 1sg.poss-}clothes hang{\sc -3.P[pst]=seq}  {\sc 1sg} dangling     hang{\sc -pst-1sg}\\
My clothes got caught on another branch, and then I was dangling there.

%\ref 42\_leg\_10.033
\exg.chu-ya-ŋ=niŋa=ca                    phopciba=ŋa   ok-ma      n-lept-i-ya-ŋa-n.\\
 hang{\sc -pst-1sg=ctmp=add} owl{\sc =erg} peck{\sc -inf} {\sc neg-}give\_up{\sc -compl[tr]-pst-1sg.P-neg}\\
Even then, the owl would not stop pecking me.

%\ref 42\_leg\_10.034
\exg.ak=ka          a-ʈukhuruk=pe            tuknuŋ    og-a-ŋ, heli=ca         lond-a, nhaŋa   a-sa             tuknuŋ    tug-a.\\
 {\sc 1sg.poss=gen} {\sc 1sg.poss-}head{\sc =loc} thoroughly peck{\sc -pst-1sg.P} blood{\sc =add} come.out{\sc -pst[3sg],} and\_then {\sc 1sg.poss-}flesh thoroughly hurt{\sc -pst[3sg]}\\
It pecked me on my head, painfully; and I was bleeding, too, and it hurt so much.


%\ref 42\_leg\_10.037
\exg.ka=ca              phopciba iŋ-nhaŋ-ma=ga             cog-a-ŋ, ŋkhaʔniŋgo ka  i=ha=ca                       a-niŋwa=be             ŋ-as-het-u-ŋa-n, a-lawa=ŋa                  naʔ-ya-masa-ŋ.\\
 {\sc 1sg[erg]=add} owl      chase{\sc -V2.send-inf=gen} do{\sc -pst[3.P]-1sg.A} but           {\sc 1sg} what{\sc =nmlz.nc=add} {\sc 1sg.poss-}mind{\sc =loc} {\sc neg-}understand{\sc -V2.carry.off-3.P[pst]-1sg.A-neg,} {\sc 1sg.poss-}soul{\sc =erg} leave{\sc -V2.leave-pst.prf-1sg.P}\\
I also tried to scare off the owl, but then I do not know anything more, because I was frozen in shock.


%\ref 42\_leg\_10.039
\exg.yo         a-ma=nuŋ      a-na=ŋa=chen      heʔne khy-a=na       bhoŋ ibebe     yog-a-ma.\\
 over\_there {\sc 1sg.poss-}mother{\sc =com}  {\sc 1sg.poss-}elder\_sister{\sc =erg=top}  where  go{\sc -pst[3sg]=nmlz.sg} comp  everywhere look\_for{\sc -pst-prf}\\
My mother and my sister came looking where I had gone.


%\ref 42\_leg\_10.040
\exg.hoŋma poʈik yok-saŋ         ta-ya-ma-ci=niŋ=go                  ka  haŋcaŋcaŋ chu-ya-ŋ=na                  nis-a-ma.\\
 river  side   search{\sc -sim} come{\sc -pst-prf-du=ctmp=top}  {\sc 1sg} dangling     hang{\sc -pst-1sg=nmlz.sg} see{\sc -pst-prf}\\
Looking for me at the riverside, they saw me hanging there.

%\ref 42\_leg\_10.041
\exg.nhaŋa   uŋci hani    hani    ta-ya-ma-ci=hoŋ             ka  lab-a-ma.\\
 and\_then {\sc 3nsg}  quickly quickly come{\sc -pst-prf-du=seq} {\sc 1sg} seize{\sc -pst-prf}\\
So they came quickly and got me down.

%\ref 42\_leg\_10.042
\exg.ŋkhiŋ   belak=be   phopciba=ca        ok-saŋ          hop=pe     pes-a-khy-a-ma.\\
  that\_much time{\sc =loc} owl{\sc =add} shriek{\sc -sim} nest{\sc =loc} fly{\sc -pst-V2.go-pst[3sg]-prf}\\
That time, the owl, too, flew away to its nest, screeching.

%\ref 42\_leg\_10.043
\exg.hakhok=ŋa  ka  cend-a-ky-a-ŋ=hoŋ                             so-ŋ=niŋa=go                       ka  luŋkhwak choŋ=be     ips-a-masa.\\
 later{\sc =ins} {\sc 1sg} wake\_up{\sc -pst-V2.come.up-pst-1sg=seq}  look{\sc -1sg.A[3.P;pst]=ctmp=top} {\sc 1sg} stone     on\_top{\sc =loc} sleep{\sc -pst-pst.prf}\\
Later, when I woke up and looked around, I had been sleeping on a rock.

%\ref 42\_leg\_10.044
\exg.nhaŋa   a-ma=nuŋ                a-na=ŋa                      ka  lab-a-masa.\\
 and\_then {\sc 1sg.poss-}mother{\sc =com}  {\sc 1sg.poss-}elder\_sister{\sc =erg} {\sc 1sg} grab{\sc -pst-pst.prf[1.P]}\\
Then, my mother and my sister grabbed me.

%\ref 42\_leg\_10.045
\exg.maŋcwa=le          apt-u           baŋniŋ=go          i=ha=e                         kheps-uks-u=ha?\\
 water{\sc =contr} bring{\sc -3.P[imp]} talking\_about{\sc =top} what{\sc =nmlz.nc=contr} hear{\sc -prf-3.P[pst]=nmlz.nc}\\
When we told you to bring water, what did you hear?

%\ref 42\_leg\_10.046
\exg.imin=hoŋ   ollobak n-sy-a-ga-n=na!\\
 how{\sc =seq}  almost  {\sc neg-}die{\sc -pst-2-neg=nmlz.sg}\\
Somehow you did not die, but almost!

%\ref 42\_leg\_10.047
\exg.lu-saŋ        thind-a          ka  ebbebe            kisit-a-ŋ              khoŋ    ghwa-ghwa     hab-a-ŋ.\\
 say{\sc -sim} scold{\sc -pst[1.P]} {\sc 1sg} shivering\_in\_fear be\_afraid{\sc -pst-1sg} and\_then bawling\_sound cry{\sc -pst-1sg}\\
Saying so, they scolded me, and I was scared and bawled out loudly.

%\ref 42\_leg\_10.048
\exg.ma,     na=ci=ga                 niŋwa=maŋ   om.\\
 mother elder\_sister{\sc =nsg=gen} love{\sc =emph} {\sc cop}\\
Of course my mother and my sister loved me.

%\ref 42\_leg\_10.049
\exg.hakhok=ŋa=go      lem-saŋ           khaʔla   lu-ya:  picchanacha leŋ-ma=hoŋ      pyak chocholaplap men-jok-ma=ha. \\
 later{\sc =ins=top}  persuade{\sc -sim} like\_this say{\sc -pst[1.P]} child       be{\sc -inf=seq}  very mischievous  {\sc neg-}do{\sc -inf[deont]=nmlz.nsg}\\
Later, convincing me, they said the following: \rede{Children should not be naughty.}


%\ref 42\_leg\_10.051
\exg.jeppa nna  len ka       a-ma=nuŋ   a-na=ga  ceʔya y-yen-u-ŋa-n=na=ŋa   ollobak paro=be      tas-u-ŋ=na.\\
 true  that day {\sc 1sg[erg]} {\sc 1sg.poss-}mother{\sc =com} {\sc 1sg.poss-}elder\_sister{\sc =gen} word  {\sc neg-}obey{\sc -3.P[pst]-1sg.A-neg=nmlz.sg=erg.cl} almost     heaven{\sc =loc} reach{\sc -3.P[pst]-1sg.A=nmlz.sg}\\
Really, that day, because I did not listen to my mother and sister, I had almost gone to heaven. 

%\ref 42\_leg\_10.052
\exg.a-ma=nuŋ                a-na=ŋa                      y-yog-a-n=niŋ=bi                                ka  hensen   ŋ-wa-ya-ŋa-n=bi.\\
 {\sc 1sg.poss-}mother{\sc =com}  {\sc 1sg.poss-}elder\_sister{\sc =erg} {\sc neg-}look\_for{\sc -pst[1.P]-neg=ctmp=irr}  {\sc 1sg} nowadays {\sc neg-}be{\sc -pst-1sg-neg=irr}\\
If my mother and sister had not looked for me, I would not be here now.

%\ref 42\_leg\_10.053
\exg.ŋkhaʔniŋgo ka  con-si-saŋ,                    ijaŋ baŋniŋ    nna  len bhaŋto=maŋ      ka  heʔniŋ=ca        chocholaplap n-jog-a-ŋa-n.\\
 but           {\sc 1sg} be\_happy{\sc -mddl-sim} why about    that day  from\_on{\sc =emph} {\sc 1sg} when{\sc =add} mischievous  {\sc neg-}do{\sc -pst-1sg-neg}\\
But I was happy, because from that day on, I never did mischievous things again.

%\ref 42\_leg\_10.054
\exg.ka  haʔniŋ tumha=ci=ga                   ceʔya=ca        yem-ma      tarokt-u-ŋ.\\
 {\sc 1sg} {\sc compar}  be\_ripe{\sc =nmlz.nsg=nsg=gen} word{\sc =add} obey{\sc -inf} start{\sc -3.P[pst]-1sg.A}\\
I began to be obedient to my elders.

%\ref 42\_leg\_10.055
\exg.hensen=ca          phopciba ka-ya=na                    kheps-wa-ŋ          ki   a-niŋwa           imin-imin coŋ-meʔ=na.\\
 nowadays{\sc =add} owl call{\sc -pst[3sg]=nmlz.sg} hear{\sc -npst-1sg.A} or   {\sc 1sg.poss-}mind how-how  do{\sc -npst=nmlz.sg}\\
Even now, when I hear the sound of an owl, I get a strange feeling.

%\ref 42\_leg\_10.056
\exg. a-ʈukhuruk=pe            og-a-ŋ=na                   loʔwa en-si-me-ŋ=na.\\
 {\sc 1sg.poss-}head{\sc =loc} peck{\sc -pst-1sg.P=nmlz.sg} like   perceive{\sc -mddl-npst-1sg=nmlz.sg}\\
It feels like it is still pecking me on the head.





