%\section[]{Namthalungma\\ {\normalsize (oral narrative; speaker: Magman Jimi Linkha)}}\label{namthalung}


\section*{The Namthalungma rocks}\addcontentsline{toc}{section}{The Namthalungma rocks}\markboth{Appendix A: Texts}{The Namthalungma rocks}\label{namthalung}

\noindent This text is an oral narrative of a mythological story. The speaker is Magman Jimi Linkha (Tumok).

\setcounter{ExNo}{0}

\exg.eko liŋkha=ga  teʔma  canuwa=be thaŋ-khy-a-ma=na      rahecha.\\
one a\_clan=gen clan\_sister Canuwa{\sc =loc} go\_away\_in\_marriage{\sc -V2.go-pst[3sg]-prf=nmlz.sg} {\sc mir}\\
One Linkha sister went away to Canuwa village in marriage.

\exg.canuwa  limbukhim=ci=ga taŋme,  liŋkha=ci=ga teʔma  bagdata  nak-se  mamliŋ ta-ya-ma.\\
Canuwa a\_clan{\sc =nsg=gen} daughter-in-law, a\_clan{\sc =nsg=gen} clan\_sister finalization\_of\_marriage ask\_for{\sc -sup} Mamling come{\sc -pst[3sg]-prf}\\
A daughter-in-law of the Limbukhims, a Linkha clan sister, came to Mamling to ask for the bagdata.
%cf. intro, make ref.!

\exg.mamliŋ=be  o-cheba=ci,   u-ppa,   u-pum=ci=ŋa,   ghak, i=ha=we,   ŋkha ghak m-b-hyaks-uks-u.\\
Mamling{\sc =loc} {\sc 3sg.poss-}male\_clan\_relatives{\sc =nsg}, {\sc 3sg.poss-}father, {\sc 3sg.poss-}grandfather{\sc =nsg=erg}, all,  what{\sc =nmlz.nc=loc}, that  all  {\sc 3pl.A-}give{\sc -V2.send-prf-3.P[pst]}\\
In Mamling, her father, the grandfathers, everybody, they gave away everything to her.

\exg.ŋkhaʔniŋgo nna  mamu=ŋa: ka,  na,  eko=chen ka mit-u-ŋ=na,\\
but   that girl{\sc =erg} {\sc 1sg[erg]} this one{\sc =top}  {\sc 1sg[erg]} want{\sc -3.P[pst]-1sg.A=nmlz.sg}\\
But that girl (said): \rede{I want one certain thing.}

\exg.saman  py-haks-a,  na yubak   py-haks-a  lu-ks-u-ci\\
property give{\sc -V2.send-imp[1.P]} this thing give{\sc -V2.send-imp[1.P]} tell{\sc -prf-3.P[pst]-3nsg.P}\\
She told them: \rede{Please send these things with us.}

\exg.nani,  i=na   yubak? n-chimd-uks-u.\\
child, what{\sc =nmlz.sg} thing {\sc 3pl.A-}ask{\sc -prf-3.P[pst]}\\
\rede{Child, what thing?}, they asked her.

\exg.ka na eko luŋkhwak=chen py-haks-a  lu-ks-u-ci,    mamu=ŋa.\\
{\sc 1sg} this one stone{\sc =top}  give{\sc -V2.send-imp[1.P]} tell{\sc -prf-3.P[pst]-3nsg.P} girl{\sc =erg}\\
\rede{Send this one stone with me}, she told them, the girl.

\exg.na=go nda  kheʔ-ma  n-yas-wa-ga-n,   na ŋ-khet-u-n.\\
this{\sc =top}  {\sc 2sg[erg]} carry\_off{\sc -inf} {\sc neg-}be\_able{\sc -npst-2-neg}  this {\sc neg-}carry\_off{\sc -3.P[imp]-neg}\\
\rede{This one, you cannot carry it off, do not carry it off.}

\exg.aniŋ=ga=ca   na encho nhaŋto  wen=ne,   na ŋ-khet-u-n   n-lu-ks-u.\\
{\sc 1pl.excl.poss=gen=add} this long\_time\_ago since exist{\sc [3sg]=nmlz.sg}, this {\sc neg-}carry\_off{\sc -3.P[imp]-neg} {\sc 3pl.A-}tell{\sc -prf-3.P[pst]}\\
\rede{It is ours since times immemorial, do not take it}, they told her.

\exg.mamu=ga  u-niŋwa  ŋ-gaks-a-ma-n.\\
girl{\sc =gen} {\sc 3sg.poss-}mind {\sc neg-}fall\_down{\sc -pst-prf-neg}\\
The girl did not give in.

\exg.na tuknuŋ  u-niŋwa  tug-a-ma,   ka  na meŋ-kheʔ-le, na luŋkhwak meŋ-kheʔ-le ŋ-khe-me-ŋa-n=na   lu-ks-u-ci.\\
this completely {\sc 3sg.poss-}mind have\_pain{\sc -pst[3sg]-prf} {\sc 1sg} this {\sc neg-}carry\_off{\sc -cvb}, this stone {\sc neg-}carry\_off{\sc -cvb} {\sc 3pl-}go{\sc -npst-1sg-neg=nmlz.sg} tell{\sc -prf-3.P[pst]-3nsg.P}\\
She was so sad; \rede{I will not go without taking this stone with me}, she told them.

\exg.nna,  nani,  nna  luŋkhwak=maŋ khet-wa-ga=na   bhoŋ, seʔni=ŋa naʔmasek  lam-ma.\\
that child that stone{\sc =emph} carry\_off{\sc -npst-2=nmlz.sg} {\sc cond},  night\_time{\sc =ins} night\_time walk{\sc -inf[deont]}\\
\rede{This stone, child, if you take away this stone, you have to walk at night.}

\exg.seʔni=ŋa seʔnamphok=pe,  heʔne=ca  me-yuŋ-ma=ga  m-baŋ=be  ta-met-i.\\
night\_time{\sc =ins} whole\_night{\sc =loc}, where{\sc =add} {\sc neg-}put{\sc -inf=gen} {\sc 2sg.poss-}house{\sc =loc} reach{\sc -caus-compl[imp]}\\
\rede{At night, the whole night, without putting it anywhere, deliver it at your home.}

\exg.manhoŋ n-leŋ-me-n=na     n-lu-ks-u.\\
otherwise {\sc neg-}become{\sc -npst[3sg]-neg=nmlz.sg} {\sc 3pl.A-}tell{\sc -prf-3.P[pst]}\\
\rede{Otherwise, it will not be (yours)}, they told her.

\exg.paŋ=be ta-meʔ-ma  n-yas-u-ga-n    bhoŋ, aniŋ=ga=ca   n-leŋ-me-n,   ŋga=ca n-leŋ-me-n    n-lu-ks-u=niŋa, na mamu i  ka-ya-ma=na:    ka ta-met-wa-ŋ=na   ka-ya-ma.\\
house{\sc =loc} arrive{\sc -caus-inf} {\sc neg-}be\_able{\sc -3.P[sbjv]-2-neg} {\sc cond} {\sc 1pl.excl.poss=gen=add} {\sc neg-}become{\sc -npst[3sg]-neg} your{\sc =add} {\sc neg-}become{\sc -npst[3sg]-neg} {\sc 3pl.A-}tell{\sc -prf-3.P[pst]=ctmp} this girl what say{\sc -pst[3sg]-prf=nmlz.sg} {\sc 1sg[erg]} bring{\sc -caus-npst-1sg.A=nmlz.sg} say{\sc -pst[3sg]-prf}\\
When they told her: \rede{If you cannot deliver it home, it will neither be ours nor yours}, what this girl said, she said: \rede{I will bring it home.}

\exg.na mamliŋ baŋna  luŋkhwak m-b-hyaks-uks-u.\\
this Mamling so-called stone {\sc 3pl.A-}give{\sc -V2.send-prf-3.P[pst]}\\
They gave that so-called Mamling stone to her.

\exg.bagdata  nak-se=ha    a-kamnibak=ci=ŋa   nnakha luŋkhwak=ha   i=ha  ghak ŋ-khu-ks-u.\\
finalization\_of\_marriage ask\_for{\sc -sup=nmlz.nsg} {\sc 1sg.poss-}friend{\sc =nsg=erg} that stone{\sc =nmlz.nc} what{\sc =nmlz.nc} all {\sc 3pl.A-}carry{\sc -prf-3.P[pst]}\\
\rede{My friends (who came) in order to ask for the Bagdata, they carry all that belongs to this stone.}

\exg.eko=ŋa eko luŋkhwak, heko=na=ŋa   eko luŋkhwak, heko=na=ŋa   eko luŋkhwak.\\
one{\sc =erg} one stone other{\sc =nmlz.sg=erg} one stone other{\sc =nmlz.sg=erg} one stone\\
One (took) one stone, another one (took) one stone, another one (took) one stone.

\exg.heko=ha=ci=ŋa   i=ha  i=ha  m-bi-ks-a,   ŋkha ghak ŋ-kho-het-u.\\
other{\sc =nmlz.nsg=nsg=erg} what{\sc =nmlz.nc} what{\sc =nmlz.nc} {\sc 3pl.A-}give{\sc -pst.prf} that  all  {\sc 3pl.A-}carry{\sc -V2.carry.off-3.P[pst]}\\
They carried off everything that the others had given her.

\exg.nhaŋa yo, mulgaun=be  n-das-uks-u,   coilikha=be,  coilikha ʈhaun=ko nam.\\
and\_then across Mulgaun{\sc =loc} {\sc 3pl.A-}reach{\sc -prf-3.P[pst]} Coilikha{\sc =loc} Coilikha  place{\sc =top}  name\\
And then they arrived on the other side, in Mulgaun, in Coilikha, Coilikha is the name of the place.

\exg.coilikha=be  n-das-u=niŋa,    nna  mamu=ga  o-phok   tuk-ma tarokt-uks-u.\\
Coilikha{\sc =loc} {\sc 3pl.A-}arrive{\sc -3.P[pst]=ctmp}, that girl{\sc =gen} {\sc 3sg.poss-}stomach hurt{\sc -inf} start{\sc -prf-3.P[pst]}\\
When they arrived in Coilikha, that girl's stomach began to hurt.

\exg.nna  o-phok   tug-a-ma=hoŋ,  pakha yuŋ-a-ma.\\
that {\sc 3sg.poss-}stomach hurt{\sc -pst[3sg]-prf=seq},  outside sit\_down{\sc -pst[3sg]-prf}\\
As her stomach hurt, she went to the toilet.

\exg.pakha=ca  yuŋ-a-ma.\\
outside{\sc =add} sit\_down{\sc -pst[3sg]-prf}\\
She also went to the toilet.

\exg.nna  khatniŋ=ŋa,  pakha=ca  lom-ma cog-a-ma,  u-hi   lom-ma=ca  cog-a-ma,\\
that this\_time{\sc =ins}, outside{\sc =add} come\_out{\sc -inf} try{\sc -pst[3sg]-prf}, {\sc 3sg.poss-}stool come\_out{\sc -inf=add} try{\sc -pst[3sg]-prf}\\
That time, she felt like vomiting and shitting.

\exg.ŋkhatniŋgo  n-lond-a-n.\\
but {\sc neg-}come\_out{\sc -pst[3sg]-neg}\\
But it did not come out.

\exg.kus-uks-u,  kus-uks-u,  kus-u-lo=be   wa=ci ŋ-ga-ya-by-a-ma.\\
wait\_for{\sc -prf-3.P[pst]}, wait\_for-prf-3.P[pst], wait\_for{\sc -3.P[pst]-int.cl=loc} chicken{\sc =nsg} {\sc 3pl-}speak{\sc -pst-V2.give-pst-prf}\\
She waited and waited, and while she waited, the cocks already crowed.

\exg.wa=ci ŋ-ga-ya=hoŋ=go  om leks-a-khy-a\\
chicken{\sc =nsg} {\sc 3pl-}speak{\sc -pst=seq=top}  bright become{\sc -pst-V2.go-pst[3sg]}\\
As the cocks crowed, it  had already dawned.

\exg.ŋkhiŋ-belak wa=ci, wa=ci ŋ-ga-ya=niŋa  eko=ŋa=cen  sui lukt-u-get-uks-u,   eko luŋkhwak uthamalaŋ, limbukhim=ŋa.\\
that\_much-time chicken{\sc =nsg} chicken{\sc =nsg} {\sc 3pl-}speak{\sc -pst=ctmp} one{\sc =erg=top}  quickly run{\sc -3.P-V2.bring.up-prf-3.P} one stone steeply\_uphill a\_clan{\sc =erg}\\
At that time, while the cocks crowed, one ran up with the stone, steeply uphill, one Limbukhim guy.

\exg.eko=na  lukt-u-get-u=hoŋ     to, okhyu, mulgaun=be  ta-uks-u.\\
one{\sc =nmlz.sg} run{\sc -3.P[pst]-V2.bring.up-3.P[pst]=seq}  up Okhyu Mulgaun{\sc =loc} bring{\sc -prf-3.P[pst]}\\
As he ran up with one stone, he brought it there to Mulgaun, to Okhyu.

\exg.na heko=ha=ci=ŋa,   na mo=na  luŋkhwak, heko=ha  luŋkhwak n-yuks-u-sa-ci=ha.\\
this other{\sc =nmlz.nsg=nsg=erg} this down{\sc =nmlz.sg} stone other{\sc =nmlz.nsg} stone {\sc 3pl.A-}keep{\sc -3.P[pst]-pst.prf-3nsg.P=nmlz.nsg}\\
The others, the lower stones, they had put the other stones down there.

\exg.ŋkha them-ma n-jog-uks-u-ci,    them-ma n-yas-uks-u-n-ci-nin.\\
those lift{\sc -inf} {\sc 3pl.A-}try{\sc -prf-3.P[pst]-3nsg.P} lift{\sc -inf} {\sc neg-}be\_able{\sc -prf-3.P[pst]-neg-3nsg.P-3pl.neg}\\
They tried to lift them up, but they could not lift them up.

\exg.them-ma n-yas-u-n-ci-nin,     nna=ga lenlen  ips-a-m=ha,    luŋkhwak.\\
lift{\sc -inf} {\sc neg-}be\_able{\sc -3.P[pst]-neg-3nsg.P-3pl.neg} that{\sc =gen} lying\_horizontally sleep{\sc -pst[3sg]-prf=nmlz.nsg} stone\\
They could not lift them up, the stones were lying (sleeping) there.

\exg.kha pok-ma  n-yas-uks-u-n-ci-nin.\\
these wake\_up{\sc -inf}   {\sc neg-}be\_able{\sc -prf-3.P[pst]-neg-3nsg.P-3pl.neg}\\
They could not wake them up.

\exg.nna=be sum=ci luŋkhwak ŋ-wa-ya-ci.\\
that{\sc =loc} three{\sc =nsg} stone {\sc 3pl-}exist{\sc -pst-3nsg.P}\\
There were three stones.

\exg.eko lenlen,   eko, ŋkha sum=ci sum=ci, carpaʈe=ci.\\
one lying\_horizontally one those three{\sc =nsg} three{\sc =nsg} four-sided{\sc =nsg}\\
One was lying there hugely, those three, the four-sided stones.

\exg.nnakha luŋkhwak=ci  pok-ma  n-yas-uks-u-n-ci-nin,     ŋkhaʔla uŋci yo canuwa=be ŋ-khy-a.\\
those  stone{\sc =nsg} wake\_up{\sc -inf} {\sc neg-}be\_able{\sc -prf-3.P[pst]-neg-3nsg.P-3pl.neg}, like\_that {\sc 3nsg}  across  Canuwa{\sc =loc} {\sc 3pl-}go{\sc -pst}\\
They could not place those stones upright, and like that (i.e., without the stones), they went over there, to Canuwa.

\exg.canuwa  ŋ-khy-a=hoŋ,  heksaŋ i  leks-a  baŋniŋ, na, jaba, na mamu=ŋa luŋkhwak pok-ma  n-yas-u-n, ŋkhiŋ-belak=ŋa na canuwa,  canuwa  men=na, mulgaun, mulgaun=ga  to, okhyu=be  to  ʈhuŋkha=be  eko luŋkhwak waiʔ=na,   dewan  ɖhuŋga baŋna.\\
Canuwa {\sc 3pl-}go{\sc -pst=seq}  later what happen{\sc -pst[3sg]} about this when this girl{\sc =erg} stone wake\_up{\sc -inf} {\sc neg-}be\_able{\sc -3.P[pst]-neg} that\_much-time{\sc =ins} this Canuwa Canuwa {\sc neg.cop=nmlz.sg} Mulgaun Mulgaun{\sc =gen} up Okhyu{\sc =loc} up  steep\_slope{\sc =loc} one stone exist{\sc [3sg]=nmlz.sg} Dewan stone  so-called\\
As they had gone to Canuwa, about what happened later, when the girl could not place the stone upright, at that time, this, Canuwa, not Canuwa, Mulgaun, above Mulgaun, in Okhyu, in that steep place, there is one rock, called Dewan stone.

\exg.nna  dewan  ɖhuŋga baŋna  luŋkhwak sahro cancan sa-ma=na,  pyak cancan, ikhiŋ cancan, nna  cancan=na=bhaŋ  soʔ-ma=niŋa,  mo oʈemma, mʌdes ghak ota=ha=bu.\\
that Dewan stone  so-called stone very  high {\sc cop.pst-prf=nmlz.sg} much high how\_much high that high{\sc =nmlz.sg=abl} look{\sc -inf=ctmp} down plains  plains all  be\_visible{\sc -pst[3sg]=nmlz.nc=hsy}\\
That Dewan stone was really high, it was very high, so high; when one looked down from this high place, the
plains, the Tarai, everything was visible, they say.

\exg.nnakha ghak ot-a-ma,   mo, nna,  saptakosi=ga   u-lap=pe  tori phet-a=ha=ca    ot=ha=bu,    nna=bhaŋ.\\
that all  be\_visible{\sc -pst[3sg]-prf} down that a\_river\_confluence{\sc =gen} {\sc 3sg.poss-}side{\sc =loc} mustard bloom{\sc -pst=nmlz.nc=add} be\_visible{\sc =nmlz.nc=hsy} that{\sc =abl}\\
All was visible, below, even the mustard blooming on the shores of the Saptakosi was visible, from there.

\exg.nna  mamu=ŋa, nna  mamu=ŋa nna  luŋkhwak khet-u=na   din i  leks-a-ma=na    baŋniŋ, haku heko=na taplik tum-si-meʔ=na nna=be.\\
that girl{\sc =erg} that girl{\sc =erg} that stone carry\_off{\sc -3.P[pst]=nmlz.sg} day what happen{\sc -pst[3sg]-prf=nmlz.sg} about now  other{\sc =nmlz.sg} story  find{\sc -mddl-npst[3sg]=nmlz.sg} that{\sc =loc}\\
As for what happened on the day when the girl took away the stone, another story can be found, inside that (story).

\exg.na dewan  ɖhuŋga l-wa=na,   mulgaun=beʔ=na dewan  ɖhuŋga, na cancan=na  luŋkhwak, uŋ sadhai=ca, i=na   ŋ-gam-my-a,    yaksigum,  yaksigum  khy-a=na=bu.\\
this Dewan stone  call{\sc -npst=nmlz.sg} Mulgaun{\sc =loc=nmlz.sg} Dewan stone  this high{\sc =nmlz.sg} stone {\sc 3sg} always{\sc =add} what{\sc =nmlz.sg} {\sc 3pl-}call{\sc -npst=nmlz.nc} place\_to\_meet place\_to\_meet go{\sc -pst[3sg]=nmlz.sg=hsy}\\
They called that high rock the Dewan stone of Mulgaun; that high rock, it was always, what do they call it -- Yaksigum -- it always went to the Yaksigum.

\exg.yaksigum  heʔne khy-a=na   baŋniŋ uŋ cokcokinetham=be khy-a=na=bu\\
place\_to\_meet where  go{\sc -pst[3sg]=nmlz.sg} about  {\sc 3sg} a\_mythical\_place{\sc =loc} go{\sc -pst[3sg]=nmlz.sg=hsy}\\
And about where it went, it went to the place called Bed of Stars, it is said.

\exg.cokcokinetham=be, eko maɖa oʈemma kham wait=na  rahecha!\\
a\_mythical\_place{\sc =loc} one big plains  ground exist{\sc [3sg]=nmlz.sg} {\sc mir}\\
In the Bed of Stars, there is a huge plain area!

\exg.nna  kham=be  naʔmasek=ŋa  uŋ yaksigum  khy-a.\\
that ground=loc evening\_time{\sc =ins} {\sc 3sg} place\_to\_meet go{\sc -pst[3sg]}\\
And at night he went to that place, to the Yaksigum.

\exg.om me-leŋ-le las-a=hoŋ   pheri to  thithi  em-diʔ-ma=bu\\
bright {\sc neg-}become{\sc -cvb} return{\sc -pst[3sg]=seq}  again up  standing\_upright stand{\sc -V2.give-inf[deont]=hsy}\\
Before sunrise, it had to return and stand  upright again.

\exg.nhaŋ  wandik=ŋa=ca  yaksigum  khy-a, om me-leŋ-le ap-ma=hoŋ   thithi  em-diʔ-m=ha.\\
and\_then later{\sc =ins=add} place\_to\_meet go{\sc -pst[3sg]}  bright {\sc neg-}become{\sc -cvb} come{\sc .level-inf[deont]=seq} standing\_upright  stand{\sc -V2.give-inf[deont]=nmlz.nc}\\
The next day it went to the Yaksigum again; it had to come back and stand upright again before sunrise.

\exg.yapmi=ci ni-ma haksaŋ aghi thithi  em-diʔ-ma    sa=na.\\
person{\sc =nsg} see{\sc -inf} {\sc compar} before standing\_upright stand{\sc -V2.give-inf[deont]} {\sc cop.pst=nmlz.sg}\\
It had to stand upright before people would see it.

\exg.ŋkhaʔniŋgo na mamliŋ=bhaŋ khy-a=na   mamu jʌbʌ pakha yuŋ-a, pakha yuŋ-a=hoŋ  uk=ka  o-phok   tug-a, luŋkhwak them-ma n-yas-u-n-ci-n.\\
but   this Mamling{\sc =abl} go{\sc -pst[3sg]=nmlz.sg} girl when outside sit{\sc -pst[3sg]} outside sit\_down{\sc -pst[3sg]=seq}  {\sc 3sg.poss=gen} {\sc 3sg.poss-}stomach hurt{\sc -pst[3sg]} stone lift{\sc -inf} {\sc neg-}be\_able{\sc -3.P[pst]-neg-3nsg.P-neg}\\
But the girl that had gone from Mamling, when she sat outside (to vomit and shit), as she sat outside, her stomach was aching, she could not lift up the stones.

\exg.wa=ci ŋ-ga-ya-by-a=ha.\\
chicken{\sc =nsg} {\sc 3pl-}speak{\sc -pst-V2.give-pst=nmlz.nsg}\\
The cocks had already crowed.

\exg.ŋkhiŋ-belak=ŋa nna  dewan  ɖhuŋga n-l-wa=na   luŋkhwak=ca, eko maʔniŋ yororo    cicaŋgalik   kaks-a-khy-a-ma.\\
that\_much-time{\sc =ins} that Dewan stone  {\sc 3pl.A-}call{\sc -npst=nmlz.sg} stone{\sc =add} one without  falling\_and\_tearing\_along in\_somersaults fall\_down{\sc -pst-V2.go-pst[3sg]-prf}\\
At that time, that stone called Dewan stone, too, it fell down tearing everything along, just like that.

\exg.cicaŋgalik   kaks-a-khy-a-ma, nna  cokcokinetham  n-l-wa=na   kham=ca  uptakham  luptakham leks-a-khy-a-ma.\\
in\_somersaults fall\_down{\sc -pst-V2.go-pst[3sg]-prf} that a\_mythical\_place {\sc 3pl.A-}call{\sc -npst=nmlz.sg} ground{\sc =add} landslide landslide happen{\sc -pst-V2.go-pst[3sg]-prf}\\
It fell down in somersaults, that place called Bed of Stars also slid down, there was a landslide, burying the ground.

\exg.pahiro  khy-a-ma.\\
landslide go{\sc -pst[3sg]-prf}\\
There was a landslide.

\exg.upt-a-khy-a-ma,    nnakha=ci,  mo luŋkhwak=ca  kheʔ-ma  n-yas-u-nin.\\
collapse{\sc -pst-V2.go-pst[3sg]-prf} that{\sc =nsg} down stone{\sc =add} carry\_off{\sc -inf} {\sc 3pl.A-}be\_able{\sc -3.P[pst]-3pl.neg}\\
There was a landslide, and they had not managed to carry off the lower stones, too.

\exg.dewan  ɖhuŋga=ca  kaks-a-khy-a,\\
Dewan stone{\sc =add} fall\_down{\sc -pst-V2.go-pst[3sg]}\\
The Dewan stone fell down.

\exg.cokcokinetham=ca   upt-a-khy-a.\\
a\_mythical\_place{\sc =add} collapse{\sc -pst-V2.go-pst[3sg]}\\
The Bed of Stars collapsed, too.

\exg.nhaŋ  heksaŋ yo paŋ=go  n-das-u.\\
and\_then later across house{\sc =top}  {\sc 3pl.A-}reach{\sc -3.P[pst]}\\
Then, later, they reached the house on the other side.

\exg.paŋ  n-das-u,   nhaŋ  ŋ-wa-ya=hoŋ, heksaŋ i  leks-a  baŋniŋ na luŋkhwak mo puŋda=e ŋkhaʔla wa-ya,  ŋkha namthaluŋma=ci, u-maita=haŋ   ŋ-khet-u=ci=ha     luŋkhwak=ci, ŋkha, ŋkhaʔla ŋ-wa-ya=niŋa   uŋci-niŋwa tug-a-ma  rahecha!\\
house {\sc 3pl.A-}reach{\sc -3.P[pst]} and\_then {\sc 3pl-}live{\sc -pst=seq} later what happen{\sc -pst[3sg]} about this stone down jungle{\sc =loc} like\_that exist{\sc -pst[3sg]} those Namthalungma{\sc =nsg} {\sc 3sg.poss-}natal\_home{\sc =abl} {\sc 3pl.A-}carry\_off{\sc -3.P[pst]-3nsg.P=nmlz.nc} stone{\sc =nsg} those in\_that\_way {\sc 3pl-}live{\sc -pst=ctmp} {\sc 3nsg.poss-}mind hurt{\sc -pst[3sg]-prf} {\sc mir}\\
They reached the house, and as they were there,  as for what happened later, this stone was down there in the jungle, those Namthalungma stones, the stones they had carried away from her maternal
home, as they lived like that, they were sad!

\exg.uŋci-niŋwa ŋkha luŋkhwak=ga  u-niŋwa  tug-a-ma=hoŋ, ten=be  picha=ci  o-phok=ŋa   saʔ-m=ha,  ghau  pok-ma, wha, wha  leŋ-ma=ha,   nhaŋ  koi  koi  sulemwalem  leŋ-khe-khuba, mam=ha yapmi=ci=ga,  u-niŋwa  meʔniŋ, das   men-da-ma=ga   ŋ-wa-ya-ghond-a-ma.\\
{\sc 3nsg.poss-}mind those stone{\sc =gen} {\sc 3sg.poss-}mind hurt{\sc -pst[3sg]-prf=seq} village{\sc =loc} child{\sc =nsg} {\sc 3sg.poss-}stomach{\sc =erg} churn{\sc -inf=nmlz.nc} wound infest{\sc -inf} septic\_wound septic\_wound happen{\sc -inf=nmlz.nc} and\_then some some head\_hanging\_down become{\sc -V2.go-nmlz} big person{\sc =nsg=gen} {\sc 3sg.poss-}mind without mental\_strength {\sc neg-}come{\sc -inf=gen} {\sc 3pl-}live{\sc -pst-V2.roam-pst-prf}\\
As they were sad, as those stones were sad, the children in the village, their stomachs hurt, wounds appeared on them, then, some of them becoming like wilted flowers, the adults, too, they walked around without mental strength.


\exg.na i  leksa,  m-mit-a-ma,  yapmi=ci.\\
this what happen{\sc -pst[3sg]} {\sc 3pl-}think{\sc -pst-prf} person{\sc =nsg}\\
\rede{What happened to this place?} the people thought.

\exg.ŋkhaʔniŋa ikhiŋ leks-a-by-a,   wasik=ca  n-da-me-n=na.\\
that\_time  how\_much happen{\sc -pst-V2.give-pst[3sg]} rain{\sc =add} {\sc neg-}come{\sc -npst[3sg]-neg=nmlz.sg}\\
And then, how many (bad things) happened, it does not rain, too.

\exg.wasik n-da-me-n=niŋa    nam phen-a=na    phen-a=na,    phen-a-nes-a=na.\\
rain  {\sc neg-}come{\sc -npst[3sg]-neg=ctmp} sun shine{\sc -pst[3sg]=nmlz.sg} shine{\sc -pst[3sg]=nmlz.sg}, shine{\sc -pst-V2.lay-pst[3sg]=nmlz.sg}\\
It does not rain, the sun was shining and shining, it kept shining.

\exg.wasik ta-ya-khy-a   bhoŋ ghak i=ha  yaŋ-kheʔ-ma-gari,  ghak yaŋ-het-i=nuŋ=ga    wasik ta-ya-ma.\\
rain  come{\sc -pst-V2.go-pst[3sg]} {\sc cond}  all  what{\sc =nmlz.nc} flush{\sc -V2.take.along-inf-advlz} all  flush{\sc -V2.carry.off-compl=com=gen} rain  come{\sc -pst[3sg]-prf}\\
And when it rained, it rained so much that everything was taken along by the water.

\exg.na wasik=le,  wasik ta-ya=na=ca   la-ma  n-nis-u-n, nam phen-a=na=ca   ucun phen-a=na,    phen-a=na    phen-a=na    leks-a, lan-siʔ-ma    n-nis-u-n.\\
this rain{\sc =contr} rain  come{\sc -pst[3sg]=nmlz.sg=add} return{\sc -inf} {\sc neg-}know{\sc -3.P[pst]-neg} sun shine{\sc -pst[3sg]=nmlz.sg=add} good shine{\sc -pst[3sg]=nmlz.sg} shine{\sc -pst[3sg]=nmlz.sg} shine{\sc -pst[3sg]=nmlz.sg} happen{\sc -pst[3sg]} return{\sc -mddl-inf} {\sc neg-}see{\sc -3.P[pst]-neg}\\
This rain, it rained, and did not know how to stop,  the sun was shining, it happened that it shone nicely,  it did not know how to stop.


\exg.haku i  cogi  ŋ-ga-ya-ma   nnakha mulgaun=be, okhyu=be=ha  yapmi=ci.\\
now  what do{\sc -1pl[sbjv]} {\sc 3pl-}say{\sc -pst-prf} those  Mulgaun{\sc =loc} Okhyu{\sc =loc=nmlz.nsg} person{\sc =nsg}\\
Now what shall we do, said the people from Mulgaun, from Okhyu.

\exg.ŋkhaʔniŋa, eh,  nna=go  haʔlo namthaluŋma=beʔ=na  luŋkhwak leʔlo!\\
that\_time oh that{\sc =top} excla  Namthalungma{\sc =loc=nmlz.sg} stone {\sc contr.emph}\\
At that time, this, gee, it is the stone of Namthalung!

\exg.namthaluŋma=beʔ=na  luŋkhwak nhe  ket-a-ma    eko.\\
Namthalung{\sc =loc=nmlz.sg} stone here bring\_up{\sc -pst[3sg]-prf} one\\
One of the Namthalum stones was brought here!

\exg.na eko=ŋa=go thend-u-get-uks-a=ba,   nna, om leks-a=niŋa,\\
this one{\sc =erg=top}  lift{\sc -3.P[pst]-V2.bring.up-pst.prf=emph} that bright become{\sc -pst[3sg]=ctmp}\\
Someone had carried it up, while it was getting bright.

\exg.wa  ka-ya=na,  ka-ya=na,  thithi end-u-ghet-uks-u=na      luŋkhwak ceŋ  to  ten=be  waiʔ=na   haku=ca.\\
chicken say{\sc -pst[3sg]=nmlz.sg} say{\sc -pst[3sg]=nmlz.sg} standing\_upright insert{\sc -3.P[pst]-V2.take.along-prf-3.P[pst]=nmlz.sg} stone upright up  village{\sc =loc} exist{\sc [3sg]=nmlz.sg} now={\sc add}\\
The cocks crowed, and someone placed the stone there upright; it is there, up high in the village, even now.

\exg.maɖa luŋkhwak om.\\
big stone {\sc cop}\\
It is a big stone.

\exg.nna  ket-uks-u.\\
that bring\_up{\sc -prf-3.P[pst]}\\
He brought it up there.

\exg.haku  na=be  laŋ=ci  muk=ci wa-siŋ-kheʔ-ma=ci.\\
now this{\sc =loc} leg{\sc =nsg} hand{\sc =nsg} wash{\sc -V2.go-inf[deont]=nsg}\\
Now, at this place, one has to go there having washed  one's feet and hands. %V2 go in sequential function!!

\exg.nhaŋa mo kheʔ-ma=hoŋ  thak-m=ha=ci.\\
and\_then down go{\sc -inf=seq}  worship{\sc -inf[deont]=nmlz.nsg=nsg}\\
And then, after going down, one has to worship and pray.

\exg.nna=ca isa=ŋa thak-ma=na   baŋniŋ, liŋkha=ga  teʔma=ŋa  thak-ma=na, ŋkhiŋ=ca  n-leks-a-n   bhoŋ, liŋkha=ga  o-teʔma=ŋa, limbukhim-ga  u-taŋmek  leŋ-khuba=ŋa,\\
that{\sc =add} who{\sc =erg} worship{\sc -inf[deont]=nmlz.sg} about  a\_clan{\sc =gen} clan\_sister{\sc =erg} worship{\sc -inf[deont]=nmlz.sg} that\_much{\sc =add} {\sc neg-}become{\sc -sbjv[3sg]-neg} {\sc cond} a\_clan{\sc =gen} {\sc 3sg.poss-}clan\_sister{\sc =erg} a\_clan{\sc =gen} {\sc 3sg.poss-}daughter-in-law become{\sc -nmlz=erg}\\
Now about who has to do the worship, a Linkha clan sister has to do the worship, and if that much is not possible, a Linkha sister, a prospective Limbukhim daughter-in-law.

\exg.ŋkhiŋ=ca  n-leks-a-n   bhoŋ, liŋkha=ga  teʔma=ŋa=se=hoŋ=ca    leŋ-meʔ=na, thaŋ-meŋ-khe-leʔ=na=hoŋ=ca,      \\
that\_much{\sc =add} {\sc neg-}happen{\sc -sbjv[3sg]-neg} {\sc cond} a\_clan{\sc =gen} clan\_sister{\sc =erg=restr=seq=add} be\_alright{\sc -npst[3sg]=nmlz.sg} go\_away\_in\_marriage-{\sc neg-}go{\sc -cvb=nmlz.sg=seq=add}\\
If that much is not possible, it is also alright if it is only a Linkha sister, even if she is not going to marry.

\exg.thak-ma=na   ŋ-ga-ya-ma=hoŋ   ŋ-khy-a-ma.\\
worship{\sc -inf[deont]=nmlz.sg} {\sc 3pl-}say{\sc -pst-prf=seq}  {\sc 3pl-}go{\sc -pst-prf}\\
She has to do the worship, they said, and went (the people of the girl's maternal home).

\exg.haku imin thak-ma=na   ŋ-ga-ya=niŋa,\\
now  how  worship{\sc -inf[deont]=nmlz.sg} {\sc 3pl-}say{\sc -pst=ctmp}\\
As they said how to do the worship,

\exg.liŋkha=ga  mamu=ŋa,  liŋkha mamu=ŋa thakt-uks-u.\\
a\_clan{\sc =gen} girl{\sc =erg} a\_clan girl{\sc =erg} worship{\sc -prf-3.P[pst]}\\
a Linkha girl did the worship.

\exg.tuknuŋ  nam phen-a=niŋa   kaniŋ=go cinuŋ=le  cah-a=ba  cog-i-ŋ.\\
completely sun shine{\sc -pst[3sg]=ctmp} {\sc 1pl=top} cold{\sc =contr} need{\sc -nativ=emph} do{\sc -1pl[sbjv]-excl}\\
When the sun was shining so much, though we need cold climate.

\exg.miyaŋ taŋkhyaŋ mopmop  cok-t-a-by-a miʔyaŋ cinuŋ cok-t-a-by-a,    ka-saŋ cuwa wahe-saŋ,    samphi wahe-saŋ, nna  n-thakt-uks-u,  nna  thakma=se  n-darokt-u=na, \\
little sky  a\_little\_cloudy do{\sc -ben-imp-V2.give-imp[1.P]} little cold do{\sc -ben-imp-V2.give-imp[1.P]} say{\sc -sim} beer offer[to\_deity]{\sc -sim} ginger offer[to\_deity]{\sc -sim} that {\sc 3pl.A-}worship{\sc -prf-3.P[pst]} that worship{\sc -inf=restr} {\sc 3pl.A-}start{\sc -3.P[pst]=nmlz.sg}\\
Saying: \rede{Please cover the sky a little for us, please make it a bit chilly for us!}, offering beer, offering ginger, they worshipped that stone, and just when they started to worship,

\exg.  n-darokt-uksa=niŋa wasik ta-ma  tarokt-uks-u.\\
{\sc 3pl.A-}start{\sc -pst.prf=ctmp} rain  come{\sc -inf} start{\sc -prf-3.P[pst]}\\
when they started (worshipping), it started to rain.


\exg.nhaŋ  uŋci ŋ-gy-a=hoŋ   wasik ta-ya-by-a,   ucun leks-a.\\
and\_then {\sc 3nsg} {\sc 3pl-}come\_up{\sc -pst=seq}  rain  come{\sc -pst-V2.give-pst[3sg]} good become{\sc -pst[3sg]}\\
And when they came up (from the place of the stones), it was raining for them, it became nice.

\exg.heksaŋ pheri nam, wasik=se  tuknuŋ  ta-ya=hoŋ, tuknuŋ  wasik ta-ya,  haku i  cok-ma,  ŋ-ga-ya-ma.\\
behind  again sun  rain{\sc =restr} thoroughly come{\sc -pst[3sg]=seq} thoroughly rain  come{\sc -pst[3sg]} now  what do{\sc -inf} {\sc 3pl-}say{\sc -pst-prf}\\
Later though, as sun and rain got worse again, it rained quite a lot; \rede{Now what to do?}, they said.


\exg.pheri ŋkhaʔniŋ=ca,  haku, pheri na liŋkha=ga  teʔma, limbukhim=ga  taŋmeʔ=ŋa, nna  thak-ma=na   ŋ-ga-ya-ma.\\
again that\_time{\sc =add} now again this a\_clan{\sc =gen} clan\_sister a\_clan{\sc =gen} daughter-in-law{\sc =erg} that worship{\sc -inf[deont]=nmlz.sg} {\sc 3pl-}say{\sc -pst-prf}\\
At that time, again, that Linkha girl, the daughter-in-law of the Limbukhims, she has to do the worship, they said.

\exg.imin thak-ma,   eh,  na,  haku, na wasik la-ni  bhoŋ thak-ma=na   ŋ-ga-ya-ma   haiko=ha=ci=ŋa.\\
how  worship{\sc -inf[deont]} oh this now this rain  return{\sc -opt} {\sc purp} worship{\sc -inf[deont]=nmlz.sg} {\sc 3pl-}say{\sc -pst-prf} other{\sc =nmlz.nsg=nsg=erg}\\
As for how to worship, now, in order for the rain to go back, one has to do the worship, the others said.

\exg.na wasik la-ni  bhoŋ imin thak-ma=na   baŋniŋ, lu, nam, ikhiŋ miʔwa hond-end-u-g=ha,    haku=go nam=ca  hond-end-u    ka-saŋ n-thakt-uks-u.\\
this rain  return{\sc -opt} purp  how  worship{\sc -inf[deont]=nmlz.sg} about  {\sc init} sun  how\_much tear uncover{\sc -V2.insert-3.P[pst]-2=nmlz.nsg} now{\sc =top}  sun{\sc =add} uncover{\sc -V2.insert-3.P[pst]} say{\sc -sim} {\sc 3pl.A-}worship{\sc -prf-3.P[pst]}\\
For that rain to stop, as for how one has to do the worship: \rede{Oh, Sun [false start], how many tears you (the sky) have dropped, now also make the sun come out!} they said, and did the worship.

\exg.nhaŋ  nam phen-a-ma.\\
and\_then sun shine{\sc -pst[3sg]-prf}\\
And the sun shone.

\exg.ŋkhaʔniŋa nna  luŋkhwak=pu cilleŋ   walleŋ  leŋ-ma   sa=na=bu.\\
that\_time  that stone{\sc =hsy}  lying\_on\_the\_back lying\_on\_the\_front overturn{\sc -inf[deont]} {\sc cop.pst=nmlz.sg=hsy}\\
Thus, that stone had to be turned around to the back and to the front. 

\exg.nna=be sum=ci luŋkhwak ŋ-wa-ya,   eko namthaluŋma, eko lalaluŋma, heko=na=ga  u-niŋ  i=na=ʔlo,   ka=ca  mund-i-ŋ.\\
that{\sc =loc} three{\sc =nsg} stone {\sc 3pl-}exist{\sc -pst} one Namthalungma one Lalalungma other{\sc =nmlz.sg=gen} {\sc 3sg.poss-}name what{\sc =nmlz.sg=excla} {\sc 1sg[erg]=add} forget{\sc -compl-1sg.A}\\
There were three stones, one was Namthalungma, one Lalalungma, and the name of the other stone -- what -- I forgot it, too.

\exg.nnakha luŋkhwak=ci  leŋ-ma=ci=bu, nam phen-ni bhoŋ, nam=bhaŋ leŋ-ma=na,\\
those  stone{\sc =nsg} overturn{\sc -inf[deont]=nsg=hsy} sun shine{\sc -opt} purp sun{\sc =abl} overturn{\sc -inf[deont]=nmlz.sg}\\
These stones have to be turned around, in order for the sun to shine, they have to be turned away from the sun.

\exg.wasik ta-ni  bhoŋ wasik=phaŋ luŋkhwak leŋ-ma=na.\\
rain  come{\sc -opt} comp  rain{\sc =abl} stone overturn{\sc -inf[deont]=nmlz.sg}\\
In order for the rain to come, they had to be turned away from the rain.

\exg.ŋkha mamu=ci=ŋa  ŋkha luŋkhwak n-leks-u-ci=ha=bu, hensen ŋkha munthum=ca  mamu=ci=ŋa  isa=ŋa=ca  n-nis-wa=nin.\\
those girl{\sc =nsg=erg} those stone {\sc 3pl.A-}overturn{\sc -3.P[pst]-3nsg.P=nmlz.nsg=hsy} nowadays those oral\_ritual\_tradition{\sc =add} {\sc girl=nsg=erg} who{\sc =erg=add} {\sc neg-}know{\sc -npst-3pl.neg}\\
Those girls have turned around those stones, it is said, even today, even though none of the girls know our ritual tradition.

\exg.nna  luŋkhwak=ca  ghak yaksaŋ=ŋa kham=ŋa lumd-y-uks-u\\
that stone{\sc =add} all  grass{\sc =erg} soil{\sc =erg} bury{\sc -compl-prf-3.P[pst]}\\
Those stones are covered by grass and earth (nowadays).

\exg.ŋkha luŋkhwak=ci  ŋ-waiʔ=ya,   okhyu=be, mo coilikha=be,\\
those stone{\sc =nsg} {\sc 3pl-}exist{\sc =nmlz.nsg} Okhyu{\sc =loc} down Coilikha{\sc =loc}\\
Those stones are there, in Okhyu, down there, in Coilikha.

\exg.nnakha luŋkhwak=ci=maŋ  n-l-wa=ci=ha    sum=ci namthaluŋma.\\
those  stone{\sc =nsg=emph} {\sc 3pl.A-}call{\sc -npst-3nsg.P=nmlz.nsg} three{\sc =nsg} Namthalungma\\
Those stones are called the three Namthalungma rocks.

\exg.hoŋkha=ci=ga  na tablik om.\\
those\_very{\sc =nsg=gen} this story  {\sc cop}\\
This is their story.


