\addchap{\lsAcknowledgementTitle} 

There are a number of people and institutions who helped with the compilation of this dictionary. 

First, we would like to thank the Volkswagen Foundation for funding the project ‘A comprehensive bilingual talking Ruruuli/Lu\-nya\-la-English dictionary with a descriptive basic grammar for language revitalisation and enhancement of mother-tongue based education’ (2017–2019) which allowed the compilation of this dictionary, as well as `The dynamics of language variation: A study of linguistic repertoires of Ru\-ruu\-li-Lunyala speakers' (2020–2022), which supports further work on the language. Furthermore, we are grateful to the University of Kiel and the Hebrew University of Jerusalem, who supported multiple field trips of the co-authors. 

Many speakers of Ruruuli-Lunyala contributed to the compilation of this dictionary and to the larger corpus from which it is derived. 
The outcomes of their endeavours to preserve and promote a record of their language for future generations as well as for scholars, researchers and other interested people will endure as their legacy. 
It is a legacy for which many generations will be forever grateful. 
Without their untiring dedication to the project, and without their knowledge of the Ru\-ruu\-li\hyp{}Lu\-nya\-la language and traditional culture, the record of Ru\-ruu\-li\hyp{}Lu\-nya\-la could not have been as rich as it is. 
Thank you, the women and men of the Buruuli and Bunyala Kingdoms!

Hundreds of community members contributed to the Ru\-ruu\-li\hyp{}Lu\-nya\-la documentary project. 
However, the ones listed below were crucial to the successful completion of the project in general and the dictionary in particular:

\begin{itemize}
\begin{multicols}{2}
\itemsep -0.3em 
\item Edward Johns Bintubizibu Bugurwa
\item Musa Kato Musoke 
\item Fred Ssekatwe Bukenya 
\item Samson Sez Kyamuwangire 
\item George William Kirya 
\item Sarah Nabukenya 
\item James Wandira Muruuli 
\item Emmanuel Sembatya 
\item Bezaleeri Bukenya
\item Rev. Wilson Galimaka 
\item Kitaakule Mulanda 
\item Debra Kaseera 
\item Martin Ssenkaatuuka 
\item Justine Tamukedde  
\item Patrick Lubega 
\item Victorious Zawede 
\item Stephen Senkonyi 
\item Damali Nakazibwe 
\item Happy Sylvia Naluyulu 
\item Yawe Samuel
\item Jane Nannyoni 
\item Charles Makanga 
\item Fredrick Kabaseke 
\item Irene Naluyiga 
\item Namusisi Robinah 
\item Samuel Butamanya 
\item Yunusu Lwanyaga 
\item Christine Namutebi 
\item Festus Wandera 
\item Inidi Sirikire 
\item Geoffrey Serunkuuma 
\item Kirya Misaki 
\end{multicols}
\end{itemize}

Our gratitude also goes to Dr.\,Matthias Brenzinger formerly at University of Cape Town who served as the project’s mentor and accompanied us during the first field trip. 

We are grateful to Sebastian Bank who assisted with the conversion of the lexical database into the LaTeX document format.

We are indebted to the students from Kiel University for choosing to write their theses on Ru\-ruu\-li\hyp{}Lu\-nya\-la language. Their research findings helped us to understand the different aspects of Ru\-ruu\-li\hyp{}Lu\-nya\-la grammar. 
We are also very grateful to the students at both Kiel University and Hebrew University of Jerusalem who supported us during the many stages of compiling the corpus and the dictionary. Many thanks to Tuarik Cassimo Buanzur (Kiel U.),
Lori Buyanovsky (Hebrew U.\,of Jerusalem), Margareta Ehlers-Kar\-ma\-nova (Kiel U.), Levke Marie Friedrichsen (Kiel U.), Erika Just (Kiel U.), Janina Klingenberg (Kiel U.), Mandy Lorenzen (Kiel U.), Cheryl Akinyi Oluoch (Kiel U.), Eloisa Co\-stanza Ruppert (Kiel U.), Dimitri Schank (Kiel U.), Katharina Sommer (Kiel U.), Anna Veselovsky  (Hebrew U.\,of Jerusalem), Jana Winkler (Kiel U.), Suza Wrze\-szcz (Kiel U.), Isabel Zechner (Kiel U.), and Caroline de Brito Gottlieb (Princeton U.).

Furthermore, we are grateful to the the series editors Adams Bodomo and Firmin Ahoua and two anonymous reviewers for all their helpful suggestions, as well as to Sebastian Nordhoff for his continuous support during the preparation of this volume.

Thank you all indeed!
