\newpage
\begin{letter}{C}
\fmtLemma{caa\-ci}
\fmtCitation{(o)caa\-ci}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{type of woven mat}
\fmtExampleRL{Ocaa\-ci oni naa\-ni ya\-mu\-lu\-ki\-re?}
\fmtExampleEN{Who wove this caaci mat?}

\fmtLemma{caa\-ka}%
\fmtOne{1}
\fmtCitation{(oku)\-caa\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{caa\-ki\-re}
\fmtSense{to start, to begin, to commence}
\fmtExampleRL{Caa\-ka oru\-ba\-zo lwa\-mu.}
\fmtExampleEN{Begin your speech.}

\fmtLemma{caa\-ka}%
\fmtOne{2}
\fmtCitation{(oku)\-caa\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{caa\-ki\-re}
\fmtSense{(of news, emotions) to spread, to become popular}
\fmtExampleRL{Ama\-wuu\-ro ga kwi\-za kwa\-mwe ga\-caa\-ki\-re.}
\fmtExampleEN{The news of his arrival has spread widely.}

\fmtLemma{caa\-nya}
\fmtCitation{(oku)\-caa\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{caa\-nyi\-re}
\fmtSense{to heap, to pile}
\fmtExampleRL{Caa\-nya enkwi ezo ozi\-ree\-te ani.}
\fmtExampleEN{Gather that firewood in heaps and bring it here.}

\fmtLemma{caa\-yi}
\fmtCitation{(o)caa\-yi}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from  Sw.}
\fmtSense{tea}
\fmtExampleRL{Mpe\-rya\-yo eki\-ko\-po kya ca\-yai.}
\fmtExampleEN{Give me a cup of tea.}

\fmtLemma{ca\-bbu\-ka}
\fmtCitation{(oku)\-ca\-bbu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ca\-bbu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to peel off}
\fmtExampleRL{Empuu\-ju ya mu\-saa\-le ya\-ca\-bbu\-ki\-re.}
\fmtExampleEN{The bark of the tree peeled off.}
\fmtSubentry(b)
\fmtSense{(of pottery) to break from}
\fmtExampleRL{Ensu\-wa eca\-bbu\-ki\-re\-ku aka\-tu\-ndu ka\-doo\-li.}
\fmtExampleEN{A small piece has broken off the pot.}

\fmtLemma{ca\-bbu\-la}
\fmtCitation{(oku)\-ca\-bbu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ca\-bbuu\-re}
\fmtSense{to break a piece from}
\fmtExampleRL{Ya\-ba oku ki\-na\-ga onca\-bbuu\-re\-ku aka\-tu\-ndu.}
\fmtExampleEN{Go and break off a piece of the pot for me.}

\fmtLemma{ca\-ci}
\fmtCitation{(o)ca\-ci}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{top prisoner (a prisoner with a high social rank in a prison hierarchy)}
\fmtExampleRL{Oca\-ci yee\-na abba mu\-bbwe.}
\fmtExampleEN{The top prisoner is still a prisoner.}

\fmtLemma{ca\-cu}
\fmtCitation{(o)ca\-cu}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{husk (of cereal crops \eg maize, rice etc.)}
\fmtExampleRL{Ebi\-so\-lo bi\-rya oca\-cu wa bi\-yi\-mba.}
\fmtExampleEN{Animals feed on husks from beans.}

\fmtLemma{ca\-ka\-la}%
\fmtOne{1}
\fmtCitation{(o)ca\-ka\-la}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{retail merchandise}
\fmtExampleRL{OMu\-pii\-na atu\-nda ca\-ka\-la omu ka\-ta\-le.}
\fmtExampleEN{Mupiina sells retail merchandise at the market.}

\fmtLemma{ca\-ka\-la}%
\fmtOne{2}
\fmtCitation{(oku)\-ca\-ka\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ca\-kai\-re}
\fmtSense{to enjoy oneself, to party, to have fun}
\fmtExampleRL{Abaa\-na ba\-ko\-be\-re na\-mbu baa\-bi\-re ku\-ca\-ka\-la.}
\fmtExampleEN{The children have said that they have gone to party.}

\fmtLemma{ca\-mbai\-tu\-ka}
\fmtCitation{(oku)\-ca\-mbai\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ca\-mbai\-tu\-ki\-re}
\fmtSubentry(a)
\fmtSense{(of liquid) to sway, to shake back and forth, to shake from side to side}
\fmtExampleRL{Eki\-do\-mo\-la eki\-ta\-yi\-zwi\-re ki\-ca\-mbai\-tu\-ka mu\-no.}
\fmtExampleEN{A half-fool jerry can shakes from side to side vigorously.}
\fmtSubentry(b)
\fmtSense{to boil}
\fmtExampleRL{Amai\-zi ga\-ku\-ca\-mbai\-tu\-ka.}
\fmtExampleEN{The water is boiling.}
\fmtSubentry(c)
\fmtSense{to be in a civil strife, to have a civil conflict}
\fmtExampleRL{Ki\-ki eki\-zwi\-re\-ku oku\-ca\-mbai\-tu\-ka kwa ba\-ntu?}
\fmtExampleEN{What is the cause of the current civil strife?}

\fmtLemma{ca\-mbai\-tu\-la}
\fmtCitation{(oku)\-ca\-mbai\-tu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ca\-mbai\-twi\-re}
\fmtSubentry(a)
\fmtSense{to stir up}
\fmtExampleRL{Onso\-ke aca\-mbai\-tu\-la mu\-no amai\-zi.}
\fmtExampleEN{A whirlwind makes the water turbulent.}
\fmtSubentry(b)
\fmtSense{to provoke}
\fmtExampleRL{Le\-ka ku\-ca\-mbai\-tu\-la ba\-ntu abo ba\-kwi\-za ku\-ku\-ku\-bba.}
\fmtExampleEN{Don't provoke those people, they may beat you up.}

\fmtLemma{ca\-mu\-ka}
\fmtCrossreference{\textnormal{see also} cambaituka}
\fmtCitation{(oku)\-ca\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ca\-mu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to boil}
\fmtExampleRL{Amai\-zi ga ku\-su\-mba akai\-ta ga\-ca\-mu\-ki\-re.}
\fmtExampleEN{The water for cooking the millet paste has boiled.}
\fmtSubentry(b)
\fmtSense{to be excited}
\fmtExampleRL{Ka mu\-ca\-mu\-ki\-re bu\-ndi eree\-ro?}
\fmtExampleEN{Oh, why are you so excited today?}

\fmtLemma{ca\-nga}
\fmtCitation{(oku)\-ca\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ca\-ngi\-re}
\fmtSource{from  Sw.}
\fmtSubentry(a)
\fmtSense{to shuffle (playing cards)}
\fmtExampleRL{Ca\-nga oca\-ni\-si tu\-zee\-nye.}
\fmtExampleEN{Shuffle the cards and let's play.}
\fmtSubentry(b)
\fmtSense{to confuse}
\fmtExampleRL{Lwa\-ki mu\-ku\-ca\-nga abaa\-na aba\-to?}
\fmtExampleEN{Why do you confuse the young children?}
\fmtSubentry(c)
\fmtSense{to play skilfully (football)}
\fmtExampleRL{OKa\-ba\-ga\-mbe amai\-te oku\-ca\-nga aka\-pii\-ra.}
\fmtExampleEN{Kabagambe knows how to play football skilfully.}
\fmtSubentry(d)
\fmtSense{to clean and sort out the nets (in preparation for fishing)}
\fmtExampleRL{Mu\-wu\-lu\-kye obu\-ti\-mba obwo tu\-bu\-ca\-nge.}
\fmtExampleEN{Get out those nets so that we can have them sorted out and cleaned.}
\fmtSubentry(e)
\fmtSense{fig.\,to have extramarital relationships in a treacherous way}
\fmtExampleRL{Omu\-ka\-li waa\-mwe aca\-nga aba\-sai\-za.}
\fmtExampleEN{His wife has extramarital relations in a treacherous way.}

\fmtLemma{ca\-nga\-ca\-nga}
\fmtCitation{(oku)\-ca\-nga\-ca\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ca\-ngi\-re\-ca\-ngi\-re}
\fmtSense{to mix up things in a confusing way}
\fmtExampleRL{Amai\-te oku\-ca\-nga\-ca\-nga ca\-ni\-si.}
\fmtExampleEN{He knows how to shuffle cards in a confusing way.}

\fmtLemma{ca\-ngi\-sya}
\fmtCitation{(oku)\-ca\-ngi\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ca\-ngi\-si\-rye}
\fmtSense{to shuffle (playing cards)}
\fmtExampleRL{Ca\-ngi\-sya ama\-ta\-tu tu\-zee\-nye.}
\fmtExampleEN{Shuffle the cards and let's play.}

\fmtLemma{ca\-ni\-si}
\fmtCitation{(o)ca\-ni\-si}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{playing card}
\fmtExampleRL{Mmai\-te oku\-zee\-nya oca\-ni\-si.}
\fmtExampleEN{I know how to play cards.}

\fmtLemma{ca\-pa\-ti}
\fmtCitation{(e)ca\-pa\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from  Hindi}
\fmtSense{chapat (UgE), chapatti (a thin pancake cooked on a griddle)}
\fmtExampleRL{Eca\-pa\-ti zi\-gu\-la bi\-taa\-nu.}
\fmtExampleEN{A chapatti costs 500 shillings.}

\fmtLemma{ca\-pu\-la}
\fmtCitation{(oku)\-ca\-pu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ca\-pwi\-re}
\fmtSense{to speak fast}
\fmtExampleRL{Wee\-na oca\-pu\-la mu\-no ebi\-ga\-mbo aba\-ntu ti\-ba\-kwe\-te\-ja.}
\fmtExampleEN{You speak very fast, people cannot follow you.}

\fmtLemma{ce}
\fmtPOS{interj}
\fmtSense{hush}
\fmtExampleRL{Mu\-si\-ri\-ke ce ba\-ta\-ba\-wuu\-ra.}
\fmtExampleEN{Keep quiet, they may overhear you.}

\fmtLemma{cee\-da}
\fmtCitation{(oku)\-cee\-da}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cee\-de\-re}
\fmtSense{to hit, to beat forcefully (with a stick)}
\fmtExampleRL{Nkwi\-za ku\-ku\-cee\-da omwi\-go.}
\fmtExampleEN{I will hit you hard with a stick.}

\fmtLemma{cee\-na}
\fmtCitation{(oku)\-cee\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cee\-ne\-re}
\fmtSense{to curse, to damn}
\fmtExampleRL{Lwa\-ki ocee\-na abaa\-na ba mu\-ga\-ndaa\-wo?}
\fmtExampleEN{Why do you curse your sister's children?}

\fmtLemma{cee\-ra\-na}
\fmtCitation{(oku)\-cee\-ra\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cee\-rai\-ne}
\fmtSense{to insist, to be persistent}
\fmtExampleRL{Cee\-ra\-na mu\-no le\-ke ba\-ku\-sa\-su\-le empii\-ya zaa\-mu.}
\fmtExampleEN{Insist until they pay you your money.}

\fmtLemma{cei\-ce\-cei}
\fmtPOS{ideo}
\fmtSense{total silence, total stillness}
\fmtExampleRL{Nku\-kwe\-ta osi\-ri\-ki\-re\-mbe cei\-cei\-cei!}
\fmtExampleEN{I am calling you and you are totally silent!}

\fmtLemma{ce\-ke\-ce\-ke}
\fmtPOS{adv}
\fmtSense{(of rain) continuously}
\fmtExampleRL{Oi\-ke\-ndi aku\-se\-nse\-ka ce\-ke\-ce\-ke ta\-ku\-kya.}
\fmtExampleEN{It is pouring rain continuously.}

\fmtLemma{ce\-nce\-ge\-ra}
\fmtCitation{(oku)\-ce\-nce\-ge\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ce\-nce\-gei\-re}
\fmtSense{to hop, to jump, to spring, to leap}
\fmtExampleRL{Aya\-bi\-re awo ng’a\-ku\-ce\-nce\-ge\-ra nka eki\-ke\-re.}
\fmtExampleEN{She passed there hopping like a frog.}

\fmtLemma{ce\-pe}
\fmtCitation{(e)ce\-pe}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{pip (a decoration worn on a military uniform to denote rank)}
\fmtExampleRL{Omu\-ku\-lu wa ma\-ye ba\-mwo\-ngei\-re ece\-pe.}
\fmtExampleEN{The army officer has been awarded more pips.}

\fmtLemma{coi\-coi\-coi}
\fmtCitation{(o)coi\-coi\-coi}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{quarrel, argument, disagreement, squabble, fight, dispute}
\fmtExampleRL{Ocoi\-coi\-coi ali\-waa\-wo di?}
\fmtExampleEN{When will the quarrel end?}

\fmtLemma{co\-ndoo\-za}
\fmtCitation{(oku)\-co\-ndoo\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{co\-ndoo\-ze\-re}
\fmtSense{to investigate, to inspect}
\fmtExampleRL{Nku\-co\-ndoo\-za ku bwi\-bbi obu\-li omu ki\-cwe\-ka.}
\fmtExampleEN{I am investigating a theft in this area.}

\fmtLemma{coo\-ba}
\fmtCitation{(oku)\-coo\-ba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{coo\-be\-re}
\fmtSubentry(a)
\fmtSense{to fill up}
\fmtExampleRL{Lee\-ta eki\-do\-mo\-la kya ama\-lwa ekyo tu\-ki\-coo\-be.}
\fmtExampleEN{Bring that jerry can of finger millet brew and let's fill it up.}
\fmtSubentry(b)
\fmtSense{to finish, to complete}
\fmtExampleRL{Ti\-mwai\-re eka nga ti\-mu\-coo\-be\-re lu\-bbi\-mbi lwa\-nywe.}
\fmtExampleEN{You will not go home unless you complete your assignment.}
\fmtSubentry(c)
\fmtSense{to spear}
\fmtExampleRL{Omu\-yii\-gi acoo\-be\-re oki\-so\-lo.}
\fmtExampleEN{The hunter speared an animal while hunting.}

\fmtLemma{coo\-be\-ra}
\fmtCitation{(oku)\-coo\-be\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{coo\-bei\-re}
\fmtSense{to wear a piece of garment on top of the other}
\fmtExampleRL{Oku go\-me\-si eyo coo\-be\-ra\-ku eree\-su.}
\fmtExampleEN{Wear a leesu on top of the gomesi.}

\fmtLemma{coo\-de\-ra}
\fmtCitation{(oku)\-co\-de\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{coo\-dei\-re}
\fmtSense{to limp}
\fmtExampleRL{Omu\-kai\-re aku\-coo\-de\-ra.}
\fmtExampleEN{The old woman is limping.}

\fmtLemma{coo\-ka}
\fmtCitation{(oku)\-coo\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{coo\-ke\-re}
\fmtSense{to collect, to gather}
\fmtExampleRL{Twa\-sa\-ngi\-re ba\-maa\-re oku\-coo\-ka obu\-te\-be omu ka\-sii\-ka.}
\fmtExampleEN{We found them when they had already gathered the chairs in the room.}

\fmtLemma{coo\-lya}
\fmtCitation{(oku)\-coo\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{coi\-rye}
\fmtSense{(of birds) to cackle}
\fmtExampleRL{Obu\-ko\-ko bu\-ku\-coo\-lya.}
\fmtExampleEN{The chickens are cackling.}

\fmtLemma{coo\-ra}
\fmtCitation{(oku)\-coo\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{coi\-re}
\fmtSense{to pile, to heap}
\fmtExampleRL{Ebi\-tee\-re mu\-bi\-coo\-re amwei.}
\fmtExampleEN{Pile the potatoes in one place.}

\fmtLemma{cu\-bbya}
\fmtCitation{(oku)\-cu\-bbya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-bbi\-rye}
\fmtSense{to dip lightly into sauce}
\fmtExampleRL{Cu\-bbya eki\-yaa\-ta omu mu\-ku\-bi oki\-rye.}
\fmtExampleEN{Dip the potato into the sauce and then eat it.}

\fmtLemma{cu\-cu\-ka}
\fmtCitation{(oku)\-cu\-cu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cu\-cu\-ki\-re}
\fmtSense{(of garments) to fade, (of skin) to be pale}
\fmtExampleRL{Olu\-go\-ye lwa\-nge lu\-cu\-cu\-ki\-re.}
\fmtExampleEN{My clothes have faded.}

\fmtLemma{cu\-cu\-ki}
\fmtCitation{\textnormal{(e.g.}\,zi\-cu\-cu\-ki, lu\-cu\-cu\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{faded}
\fmtExampleRL{Olu\-go\-ye olu\-cu\-cu\-ki lwa\-ni?}
\fmtExampleEN{Who is the owner of the faded garment?}

\fmtLemma{cu\-cu\-la}
\fmtCitation{(oku)\-cu\-cu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cu\-cwi\-re}
\fmtSubentry(a)
\fmtSense{(of skin) to be harsh, to be dry}
\fmtSubentry(b)
\fmtSense{to look pale, to lose lustre, to lose colour, to fade}
\fmtExampleRL{Ncu\-cwi\-re olu\-ko\-bba.}
\fmtExampleEN{My skin looks pale.}

\fmtLemma{cu\-cu\-lya}
\fmtCitation{(oku)\-cu\-cu\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-cwi\-rye}
\fmtSense{to make pale, to cause to lose lustre, to tarnish, to make rough}
\fmtExampleRL{Amai\-zi ago ga\-cu\-cu\-lya olu\-ko\-bba.}
\fmtExampleEN{That water makes the skin pale.}

\fmtLemma{cu\-ga}
\fmtCitation{(oku)\-cu\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-gi\-re}
\fmtSense{to have sex with, to have sexual intercourse with}
\fmtExampleRL{Aba\-ta\-mii\-re bai\-ngi be\-ndya mu\-no oku\-cu\-ga.}
\fmtExampleEN{Most drunkards like having sex a lot.}

\fmtLemma{cu\-ku\-nda}
\fmtCitation{(oku)\-cu\-ku\-nda}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-ku\-ndi\-re}
\fmtSense{to shake (with the aim of getting crystallised form to liquid form)}
\fmtExampleRL{Cu\-ku\-nda omu\-ba\-zi gwa mwa\-na le\-ke agu\-nywe.}
\fmtExampleEN{Shake the child's medicine before she takes it.}

\fmtLemma{cu\-lee\-ra}
\fmtCitation{(oku)\-cu\-lee\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cu\-lei\-re}
\fmtSense{to be calm, to be settled, to settle down}
\fmtExampleRL{Cu\-lee\-ra mu\-nywa\-ni wa\-nge ebi\-ntu bi\-kwi\-za ku\-te\-ree\-ra.}
\fmtExampleEN{Be calm, my dear, all will be fine.}

\fmtLemma{cu\-lee\-ri}
\fmtCitation{\textnormal{(e.g.}\,ma\-cu\-lee\-ri, ki\-cu\-lee\-ri, ba\-cu\-lee\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSense{peaceful, quiet, calm}
\fmtExampleRL{Nta\-ka oku\-bba omu kya\-lo eki\-cu\-lee\-ri.}
\fmtExampleEN{I like to live in a peaceful village.}

\fmtLemma{cu\-lee\-rya}
\fmtCitation{(oku)\-cu\-lee\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-lei\-rye}
\fmtSense{to calm down, to make quiet}
\fmtExampleRL{OMu\-se\-ve\-ni akoo\-re mu\-no oku\-cu\-lee\-rya embee\-ra omu Co\-ngo.}
\fmtExampleEN{Museveni has done a lot to calm down the situation in Congo.}

\fmtLemma{cu\-ma}
\fmtCitation{(oku)\-cu\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-mi\-re}
\fmtSubentry(a)
\fmtSense{to throw, to cast}
\fmtExampleRL{Omwa\-na aku\-cu\-ma ama\-bbaa\-le oku nyu\-mba.}
\fmtExampleEN{The child is throwing stones at the house.}
\fmtSubentry(b)
\fmtSense{to throw away, to get rid of, to dispose, to discard}
\fmtExampleRL{Cu\-ma\-yo ebi\-fu\-ka ebyo ndoo\-wo ya\-bi\-rya.}
\fmtExampleEN{Throw away those leftovers, no one will eat them.}

\fmtLemma{cu\-mba}
\fmtCitation{(oku)\-cu\-mba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-mbi\-re}
\fmtSense{to boil, to cook}
\fmtExampleRL{Cu\-mba ama\-tai ago tu\-ga\-nywe.}
\fmtExampleEN{Boil the milk and we'll take it.}

\fmtLemma{cu\-mu\-ka}
\fmtCitation{(oku)\-cu\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-mu\-ki\-re}
\fmtSense{to kindle, to set on fire, to set ablaze}
\fmtExampleRL{Cu\-mu\-ka enyu\-mba eyo.}
\fmtExampleEN{Set that house ablaze.}

\fmtLemma{cuu\-muu\-kii\-rya}
\fmtCitation{(oku)cu\-muu\-kii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-muu\-ki\-rii\-rye}
\fmtSense{to treat with a steam bath (for medical purposes)}
\fmtExampleRL{Omwa\-na oyo alwai\-re, mu\-cuu\-muu\-kii\-rye.}
\fmtExampleEN{That child is sick, treat him with a steam bath.}

\fmtLemma{cu\-nda}
\fmtCitation{(oku)\-cu\-nda}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ci\-ndi\-re}
\fmtSense{to shake vigorously}
\fmtExampleRL{Aku\-cu\-nda ama\-tai le\-ke atoo\-le\-ku ama\-gi\-ta.}
\fmtExampleEN{She is churning milk to get ghee.}

\fmtLemma{cu\-ndu\-la}
\fmtCitation{(oku)\-cu\-ndu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-ndwi\-re}
\fmtSense{to pour away, to have a discharge}
\fmtExampleRL{Ota\-cu\-ndu\-la mai\-zi ago, ga\-wee\-rye ente.}
\fmtExampleEN{Don’t pour away that water, give it to the cows.}

\fmtLemma{cu\-ngu\-la}
\fmtCrossreference{\textnormal{see} cungura}
\fmtDialect{Lunyala}

\fmtLemma{cu\-ngu\-ra}
\fmtDialect{Ruruuli}
\fmtCitation{(oku)\-cu\-ngu\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-ngwi\-re}
\fmtSubentry(a)
\fmtSense{to redeem, to save, to rescue}
\fmtExampleRL{OMu\-sa ya\-cu\-ngwi\-re abaa\-na ba Isi\-ra\-yi\-ri.}
\fmtExampleEN{Moses rescued the children of Israel.}
\fmtSubentry(b)
\fmtSense{to liberate}
\fmtExampleRL{Obu\-zi\-ga bu\-cu\-ngu\-ra ei\-ya\-nga.}
\fmtExampleEN{Bravery liberates the country.}
\fmtSubentry(c)
\fmtSense{to pay a fine to the in-laws to earn recognition for being the father of a child}
\fmtExampleRL{Omu\-si\-ga\-zi ya\-cu\-ngwi\-re owa\-na.}
\fmtExampleEN{The youth paid a fine for the child.}

\fmtLemma{cu\-pa}
\fmtCitation{(e)cu\-pa}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from  Sw.}
\fmtSense{bottle}
\fmtExampleRL{Lee\-ta ecu\-pa tu\-tee\-mu ama\-fu\-ta.}
\fmtExampleEN{Bring the bottle and let's put paraffin in it.}

\fmtLemma{cu\-pu\-la}%
\fmtOne{1}
\fmtCitation{(oku)\-cu\-pu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-pwi\-re}
\fmtSubentry(a)
\fmtSense{to kick}
\fmtExampleRL{Omwo\-jo acu\-pwi\-re omwi\-ra\-waa\-mwe.}
\fmtExampleEN{The boy has kicked his friend.}
\fmtSubentry(b)
\fmtFixedexpression{cu\-pu\-la (o)mu\-ge\-re}
\fmtSense{to kick}
\fmtExampleRL{Ba\-mu\-cu\-pwi\-re omu\-ge\-re oku ka\-bu\-nda.}
\fmtExampleEN{He was kicked in the belly.}

\fmtLemma{cu\-pu\-la}%
\fmtOne{2}
\fmtCitation{(oku)\-cu\-pu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-pwi\-re}
\fmtSense{to make counterfeits, to fake, to forge}
\fmtExampleRL{Aba\-na\-yu\-ga\-nda ba\-cu\-pu\-la ese\-nte.}
\fmtExampleEN{Ugandans make counterfeit money.}

\fmtLemma{cu\-pu\-li}
\fmtCitation{\textnormal{(e.g.}\,ki\-cu\-pu\-li, bi\-cu\-pu\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSense{fake, non-original, counterfeit, inferior}
\fmtExampleRL{Ga\-ni ama\-tai ga\-ti\-ki\-ri ti bi\-cu\-pu\-li ebi\-ri oku ka\-ta\-le.}
\fmtExampleEN{It is real milk not the fake one on the market.}

\fmtLemma{cu\-raa\-na}
\fmtCitation{(oku)\-cu\-raa\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cu\-rai\-ne}
\fmtSubentry(a)
\fmtSense{to live in peace}
\fmtExampleRL{Tu\-ku\-saa\-na ku\-cu\-raa\-na mu Uga\-nda.}
\fmtExampleEN{We deserve to live in peace in Uganda.}
\fmtSubentry(b)
\fmtSense{to be well-off, to be wealthy, to be rich, to be prosperous}
\fmtExampleRL{Wee\-na oi\-za ku\-cu\-raa\-na ni wa\-ko\-la omu\-no.}
\fmtExampleEN{You will be well-off if you work hard.}

\fmtLemma{cu\-raa\-ni\-ki\-sya}
\fmtCitation{(oku)\-cu\-raa\-ni\-ki\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-raa\-ni\-ki\-si\-rye}
\fmtSense{to arbitrate, to adjudicate, to judge, to mediate}
\fmtExampleRL{Cu\-raa\-ni\-ki\-sya aba\-ntu ba\-le\-ke ku\-lwa\-na.}
\fmtExampleEN{Arbitrate between those people so that they stop fighting.}

\fmtLemma{cu\-raa\-nu\-ka}
\fmtCitation{(oku)\-cu\-raa\-nu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-raa\-nu\-ki\-re}
\fmtSense{to fully understand, to comprehend}
\fmtExampleRL{Amai\-te oku\-tu\-so\-me\-sya Olu\-ju\-ngu ni\-tu\-cu\-raa\-nu\-ka ei\-so\-mo.}
\fmtExampleEN{He knows how to teach us English and we fully understand it.}

\fmtLemma{cu\-ra\-ni\-sya}
\fmtCitation{(oku)\-cu\-ra\-ni\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-raa\-ni\-si\-rye}
\fmtSense{to calm down, to pacify, to quiet, to quiet down}
\fmtExampleRL{Fu\-ba oku\-bo\-na ng’o\-ku\-cu\-ra\-ni\-sya aba\-ntu abo.}
\fmtExampleEN{Try to calm down those people.}

\fmtLemma{cu\-ra\-nu\-ra}
\fmtCitation{(oku)\-cu\-ra\-nu\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-ra\-nwi\-re}
\fmtSense{to satisfy, to meet the expectations of}
\fmtExampleRL{Fu\-ba mu\-no oku\-bo\-na ng’o\-ku\-cu\-ra\-nu\-ra aba\-ntu abo.}
\fmtExampleEN{Try to see that you satisfy those people.}

\fmtLemma{cu\-ru\-cu\-ru}%
\fmtOne{1}
\fmtPOS{ideo}
\fmtSense{real, typical, pure}
\fmtExampleRL{Na\-ku\-ta\-ki\-re oku\-swe\-ra Mu\-ruu\-li cu\-ru\-cu\-ru.}
\fmtExampleEN{I would like to marry a typical Muruuli.}

\fmtLemma{cu\-ru\-cu\-ru}%
\fmtOne{2}
\fmtPOS{ideo}
\fmtSense{insufficiently}
\fmtExampleRL{Ompai\-re bwi\-zi cu\-ru\-cu\-ru.}
\fmtExampleEN{You have given me insufficient water.}

\fmtLemma{cu\-ru\-ka}
\fmtCitation{(oku)\-cu\-ru\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cu\-ru\-ki\-re}
\fmtSense{to escape}
\fmtExampleRL{N’o\-ta\-gu\-mya ka\-nyo\-nyi ka\-cu\-ru\-ka.}
\fmtExampleEN{If you don’t secure the bird tightly, it can escape.}

\fmtLemma{cuu\-cu}
\fmtCitation{(o)cuu\-cu}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{dust}
\fmtExampleRL{Emo\-to\-ka edu\-du\-mwi\-re ocuu\-cu.}
\fmtExampleEN{The car has raised dust.}

\fmtLemma{cuu\-ka}%
\fmtOne{1}
\fmtCitation{(e)cuu\-ka}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{ox plough}
\fmtExampleRL{Nje ndu\-mi\-sya cuu\-ka.}
\fmtExampleEN{I use an ox plough to farm.}

\fmtLemma{cuu\-ka}%
\fmtOne{2}
\fmtCitation{(oku)\-cuu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cuu\-ki\-re}
\fmtSense{to emit (smoke), to smoke, to smoulder}
\fmtExampleRL{Eta\-nuu\-lu ya\-nge eta\-ndi\-ki\-re oku\-cuu\-ka.}
\fmtExampleEN{My kiln has begun to smoke.}

\fmtLemma{cuu\-kii\-rya}%
\fmtCitation{(oku)\-cuu\-kii\-rya}
\fmtPOS{v}%
\fmtTrans{ditr}
\fmtPFV{cuu\-ki\-rii\-rye}
\fmtSense{to fumigate, to burn (incense), to treat with steam or smoke from herb (usually by putting them on a clay surface; the practice is meant to drive spirits away, as well as provide a good smell and repel mosquitoes), to do smoke cleansing}
\fmtExampleRL{Cuu\-kii\-rya enyu\-mba ebi\-mbu\-gu\-mbu\-gu awu\-lu\-ke.}
\fmtExampleEN{Fumigate the house so that the vampire bats get out.}

\fmtLemma{cu\-ku\-nda}
\fmtCitation{(oku)\-cuu\-ku\-nda}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cu\-ku\-ndi\-re}
\fmtSense{to shake (with the aim of getting crystallised form to liquid form)}

\fmtLemma{cuu\-ma}
\fmtCitation{(oku)\-cuu\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cuu\-mi\-re}
\fmtSense{to smell, to stink}
\fmtExampleRL{Eki\-na\-ge ki\-ni ki\-ku\-cuu\-ma bu\-ndi ndoo\-wo ayi\-nza oku\-ki\-rya.}
\fmtExampleEN{The tilapia smells badly, no one can eat it.}

\fmtLemma{cuu\-muu\-ka}
\fmtCrossreference{\textnormal{see also} cuuka}
\fmtCitation{(oku)\-cuu\-muu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cuu\-muu\-ki\-re}
\fmtSense{to emit smoke, to smoke}
\fmtExampleRL{Lwa\-ki eta\-nuu\-lu yaa\-mu eku\-cuu\-muu\-ka mu\-no?}
\fmtExampleEN{Why is your kiln emitting so much smoke?}

\fmtLemma{cuu\-nya}
\fmtCitation{(oku)\-cuu\-nya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cuu\-nyi\-re}
\fmtSense{to taste good, to have a pleasant taste}
\fmtExampleRL{Omu\-ku\-bi gwa ncwi gu\-cuu\-nya ku\-sai.}
\fmtExampleEN{The sauce made from fish is delicious.}

\fmtLemma{cuu\-nyi}
\fmtCitation{\textnormal{(e.g.}\,mu\-cuu\-nyi, gu\-cuu\-nyi, ncuu\-nyi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{well spiced}
\fmtExampleRL{Nta\-ka oku\-lya omu\-ku\-bi ogu\-cuu\-nyi.}
\fmtExampleEN{I like eating a well-spiced sauce.}

\fmtLemma{cuu\-pa}
\fmtCitation{(oku)\-cuu\-pa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cuu\-pi\-re}
\fmtSense{to deceive, to defraud, to trick, to swindle}
\fmtExampleRL{We le\-ka ku\-cuu\-pa mwi\-ra\-waa\-mu.}
\fmtExampleEN{Stop deceiving your friend.}

\fmtLemma{cuu\-wa}
\fmtCitation{(oku)\-cuu\-wa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cuu\-wi\-re}
\fmtSense{to be satisfied}
\fmtExampleRL{Ndii\-re ena\-na\-nsi ni ncuu\-wa.}
\fmtExampleEN{I have eaten a pineapple and I am satisfied.}

\fmtLemma{cwa}%
\fmtOne{1}
\fmtCitation{(oku)\-cwa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cwi\-re}
\fmtSense{to break}
\fmtExampleRL{Cwa eki\-saa\-le ekyo ebi\-cwe\-ka bi\-sa\-tu.}
\fmtExampleEN{Break that stick into three pieces.}

\fmtLemma{cwa}%
\fmtOne{2}
\fmtCitation{(oku)\-cwa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cwi\-re}
\fmtSense{to give a verdict, to rule}
\fmtExampleRL{Ema\-mbya ni\-mwo ba\-kwi\-za oku\-cwa omu\-sa\-ngo gwa kwe\-nda enyu\-mba.}
\fmtExampleEN{They will rule on the demolition of the house tomorrow.}

\fmtLemma{cwa}%
\fmtOne{3}
\fmtCrossreference{\textnormal{see also} cwanta}
\fmtCitation{(oku)\-cwa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cwi\-re}
\fmtSense{to spit}
\fmtExampleRL{Naa\-ni acwi\-re awo?}
\fmtExampleEN{Who spat there?}

\fmtLemma{cwa\-cwa\-na}
\fmtCitation{(oku)\-cwa\-cwa\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cwa\-cwai\-ne}
\fmtSense{to chop into pieces, to cut up, to blend}
\fmtExampleRL{Cwa\-cwa\-na enya\-ma eyo ogi\-ga\-bi\-re aba\-ntu.}
\fmtExampleEN{Chop the meat and distribute it to the people.}

\fmtLemma{cwa\-ka}
\fmtCitation{(oku)\-cwa\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cwa\-ki\-re}
\fmtSubentry(a)
\fmtSense{to incite (a dog)}
\fmtExampleRL{Cwa\-ka ombwe\-ne akwa\-te oki\-so\-lo.}
\fmtExampleEN{Incite the dog to catch the animal.}
\fmtSubentry(b)
\fmtSense{to instigate, to prompt, to encourage, to provoke}
\fmtExampleRL{Wa\-li\-wo aku\-cwa\-ka aba\-ntu oku\-jee\-ma?}
\fmtExampleEN{Is there someone who is instigating the people to strike?}

\fmtLemma{cwa\-kii\-rya}
\fmtCitation{(oku)\-cwa\-kii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cwa\-kii\-rii\-rye}
\fmtSense{to instigate, to prompt, to incite, to encourage, to provoke}
\fmtExampleRL{Naa\-ni\-te oyo aku\-cwa\-kii\-rya aba\-ntu oku\-lwa\-na?}
\fmtExampleEN{Who is the one inciting the people to fight?}

\fmtLemma{cwa\-nga\-ni\-za}%
\fmtOne{1}
\fmtCitation{(oku)\-cwa\-nga\-ni\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cwa\-nga\-nii\-ze, cwa\-nga\-ni\-zi\-re}
\fmtSense{to arrange, to organise, to bring order to}
\fmtExampleRL{Cwa\-nga\-ni\-za edii\-ro aba\-ge\-ni ba\-kwi\-za.}
\fmtExampleEN{Get the living room in order, the visitors are coming.}

\fmtLemma{cwa\-nga\-ni\-za}%
\fmtOne{2}
\fmtCitation{(oku)\-cwa\-nga\-ni\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{cwa\-nga\-nii\-ze, cwa\-nga\-ni\-zi\-re}
\fmtSense{to cross, to traverse}
\fmtExampleRL{Onjo\-ka acwa\-nga\-nii\-ze olu\-guu\-do mu\-koo\-to.}
\fmtExampleEN{The snake that has crossed the road is big.}

\fmtLemma{cwa\-nta}
\fmtCitation{(oku)\-cwa\-nta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cwa\-nti\-re}
\fmtSense{to spit}
\fmtExampleRL{Naa\-ni aku\-cwa\-nta awo?}
\fmtExampleEN{Who is spitting there?}

\fmtLemma{cwa\-ntii\-ra}
\fmtCitation{(oku)\-cwa\-ntii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cwa\-ntii\-re}
\fmtSense{to drizzle, to rain lightly}
\fmtExampleRL{Obwi\-ke\-ndi bu\-ta\-ndi\-ki\-re oku\-cwa\-ntii\-ra ti\-tu\-ku\-so\-bo\-la kwa\-ba.}
\fmtExampleEN{It has started to drizzle, we cannot go.}

\fmtLemma{cwe\-ke\-ra}
\fmtCitation{(oku)\-cwe\-ke\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cwe\-kei\-re}
\fmtSense{to vanish, to end, to stop}
\fmtExampleRL{Oku\-lii\-sya ebi\-bu\-nguu\-ra nka ekya\-bu\-suu\-bu\-li omu Uga\-nda kwa\-cwe\-kei\-re irai.}
\fmtExampleEN{Rearing caterpillars as an economic activity in Uganda stopped long time ago.}

\fmtLemma{cwe\-ree\-ra}
\fmtCitation{(oku)\-cwe\-ree\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{cwee\-rei\-re}
\fmtSense{to drizzle, to rain lightly}
\fmtExampleRL{Oi\-ke\-ndi aku\-cwe\-ree\-ra mwa\-nu\-le engo\-ye.}
\fmtExampleEN{It is drizzling, please get the clothes inside the house.}

\fmtLemma{cwi cwi cwi}
\fmtPOS{ideo}
\fmtSense{completely, totally}
\fmtExampleRL{Amai\-zi ga\-wei\-re\-mu cwi cwi cwi.}
\fmtExampleEN{The water is completely gone.}

\fmtLemma{cwi}%
\fmtOne{1}
\fmtPOS{ideo}
\fmtSense{cry of a rat}
\fmtExampleRL{Ome\-se aku\-nga cwi.}
\fmtExampleEN{The rat screams ``cwi".}

\fmtLemma{cwi}%
\fmtOne{2}
\fmtPOS{ideo}
\fmtSense{whole}
\fmtExampleRL{Onko\-ko amu\-lii\-re cwi.}
\fmtExampleEN{He has eaten the whole chicken.}
\end{letter}
