\begin{letter}{E}
\fmtLemma{e}%
\fmtOne{1}
\fmtPOS{part}
\fmtSense{to, in, at, from (locative of noun class 23)}
\fmtExampleRL{Wa\-ya\-ba e Ka\-mpa\-la?}
\fmtExampleEN{Will you go to Kampala?}

\fmtLemma{e-}%
\fmtOne{2}
\fmtPOS{pf}
\fmtSense{augment of noun classes 4, 5, 7, 8, 9, and 10}

\fmtLemma{(-)e}%
\fmtOne{3}
\fmtPOS{sf}
\fmtSense{subjunctive suffix on verbs}

\fmtLemma{e-}%
\fmtOne{4}
\fmtPOS{pf}
\fmtSense{subject agreement prefix of noun class 9 on verbs}

\fmtLemma{e-}%
\fmtOne{5}
\fmtSense{subject and object agreement prefix of noun class 23 on verbs}

\fmtLemma{e}%
\fmtOne{6}
\fmtPOS{interj}
\fmtSense{oh}
\fmtExampleRL{E, oyo naa\-ni?}
\fmtExampleEN{Oh, who is that?}

\fmtLemma{eba\-bii\-ra}
\fmtCitation{(okw)e\-ba\-bii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eba\-bi\-rii\-re}
\fmtSense{to shriek, to scream, to screech}
\fmtExampleRL{Omwa\-na akwe\-ba\-bii\-ra na ki?}
\fmtExampleEN{What is making the baby shriek?}

\fmtLemma{eba\-bu\-lya}
\fmtCitation{(okw)e\-ba\-bu\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eba\-bwi\-rye}
\fmtSense{to peel off}
\fmtExampleRL{Olu\-su\-su lwe\-ba\-bu\-lya ng’a\-mai\-zi ga\-lwo\-ke\-rye.}
\fmtExampleEN{The skin peels off when the water burns it.}

\fmtLemma{eba\-ga\-la}
\fmtCitation{(okw)e\-ba\-ga\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eba\-gai\-re}
\fmtSubentry(a)
\fmtSense{to mount (an animal), to get up on in order to ride it}
\fmtExampleRL{Omu\-sai\-za oyo yee\-ba\-gai\-re emba\-laa\-si.}
\fmtExampleEN{That man mounts the horse.}
\fmtSubentry(b)
\fmtSense{to ride (an animal)}
\fmtExampleRL{Omu\-wa\-ra\-bu yee\-ba\-gai\-re ondo\-go\-yi.}
\fmtExampleEN{The Arab is riding on a donkey.}

\fmtLemma{ebai\-za\-ga\-la}
\fmtCitation{(okw)e\-bai\-za\-ga\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ebai\-za\-gai\-re}
\fmtSubentry(a)
\fmtSense{to belch, to burp}
\fmtExampleRL{Ni\-ya\-mii\-re ento\-nge eya\-se\-mbe\-re\-yo ni yee\-bai\-za\-ga\-la.}
\fmtExampleEN{After she bit into the last bit, she belched.}
\fmtSubentry(b)
\fmtSense{fig. to use vulgar language}
\fmtExampleRL{Omu\-sai\-za ya\-ya\-bi\-re ni yee\-ba\-za\-gaa\-ra omwi\-ra\-waa\-mwe.}
\fmtExampleEN{The man went and used vulgar language before his friend.}

\fmtLemma{eba\-laa\-mii\-rya}
\fmtCitation{(okw)e\-ba\-laa\-mii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eba\-laa\-mi\-rii\-rye}
\fmtSense{to pass via the outskirts of}
\fmtExampleRL{Omu\-sai\-za ngo\-di akwe\-ba\-laa\-mii\-rya oNa\-ka\-so\-ngo\-la.}
\fmtExampleEN{There is a man passing via the outskirts of Nakasongola.}

\fmtLemma{eba\-la\-ngu\-la}
\fmtCitation{(okw)e\-ba\-la\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eba\-la\-ngwi\-re}
\fmtSubentry(a)
\fmtSense{(of the sun) to scorch, to shine brightly with much heat}
\fmtExampleRL{Ei\-sa\-na li\-kwe\-ba\-la\-ngu\-la mu\-no.}
\fmtExampleEN{The sun is shining too much.}
\fmtSubentry(b)
\fmtSense{to cry unceasingly}
\fmtExampleRL{Omwa\-na akwe\-ba\-la\-ngu\-la.}
\fmtExampleEN{The child is crying unceasingly.}

\fmtLemma{eba\-lya}
\fmtCitation{(okw)e\-ba\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ebaa\-rye}
\fmtSense{to thank, to express (one's) gratitude to}
\fmtExampleRL{Nyi\-zi\-re ku\-kwe\-ba\-lya.}
\fmtExampleEN{I've come to thank you.}

\fmtLemma{eba\-mba}
\fmtCitation{(okw)e\-ba\-mba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eba\-mbi\-re}
\fmtSense{to sit with legs spread apart}
\fmtExampleRL{Omu\-ka\-li oyo yee\-ba\-mbi\-re.}
\fmtExampleEN{That woman is sitting with her legs spread apart.}

\fmtLemma{eba\-za\-ga\-la}
\fmtCrossreference{\textnormal{see} ebai\-za\-ga\-la}

\fmtLemma{ebba\-la}
\fmtCitation{(okw)e\-bba\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ebbai\-re}
\fmtSense{to avoid, to keep away from, to stay away from}
\fmtExampleRL{Mwa\-na wa\-nge le\-ke\-ra\-wo oku\-ta\-mbu\-la obwi\-re we\-bba\-le ebi\-zi\-bu.}
\fmtExampleEN{My child, stop walking at night s that you can stay away from problems.}

\fmtLemma{ebbi\-rya}
\fmtCrossreference{\textnormal{see} eibbi\-rya}

\fmtLemma{ebbi\-sa}
\fmtCrossreference{\textnormal{see} ebbisa}

\fmtLemma{ebbu\-bbu\-tya}
\fmtCitation{(okw)e\-bbu\-bbu\-tya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ebbu\-bbu\-ti\-rye}
\fmtSense{to feign foolishness}
\fmtExampleRL{OBi\-sa\-bye\-re yee\-bbu\-bbu\-tya, n’o\-mu\-buu\-lya ky’a\-mai\-te, aku\-ko\-ba nti ta\-ki\-mai\-te.}
\fmtExampleEN{Bisabyere feigns foolishness: when you ask him what he knows, he tells you that he doesn't know it.}

\fmtLemma{ebbuu\-ka}
\fmtCitation{(okw)e\-bbuu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ebbuu\-ki\-re}
\fmtSense{to smear oneself with (e.g.\,vaseline)}
\fmtExampleRL{Omwa\-na yee\-bbuu\-ki\-re ebi\-gi\-ta.}
\fmtExampleEN{The girl has smeared herself with vaseline.}

\fmtLemma{ebbye\-te\-ka}
\fmtCitation{(okw)e\-bbye\-te\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ebbye\-teke\-re}
\fmtSense{to sit (especially in a dirty place), to sit anyhow}
\fmtExampleRL{Le\-ka kwe\-bbye\-te\-ka.}
\fmtExampleEN{Don’t sit in a dirty place.}

\fmtLemma{ebee\-ra}
\fmtCitation{(okw)e\-bee\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ebee\-rei\-re}
\fmtSense{to forget, to fail to think of}
\fmtExampleRL{Otee\-bee\-ra oku\-so\-ma eBbai\-bbu\-li.}
\fmtExampleEN{Don't forget to read the Bible.}

\fmtLemma{ebee\-ri}
\fmtCitation{\textnormal{(e.g.}\,mwe\-bee\-ri. bee\-bee\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSense{forgetful, oblivious}
\fmtExampleRL{Omu\-ka\-li omwe\-bee\-ri ye\-be\-rei\-re omwa\-na waa\-mwe omu bbaa\-si.}
\fmtExampleEN{The forgetful woman forgot her child in the bus.}

\fmtLemma{ebe\-mbe\-ra}
\fmtCitation{(okw)e\-be\-mbe\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ebe\-mbei\-re}
\fmtSense{to lead, to head, to be at the head of, to go first, to precede, to be the best}
\fmtExampleRL{OIsa\-ba\-nya\-la ni\-ye yee\-be\-mbe\-ra oBu\-nya\-la.}
\fmtExampleEN{Isabanyala is the one who leads the Bunyala Kingdom.}

\fmtLemma{ebe\-re\-ka}
\fmtCitation{(okw)e\-be\-re\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ebe\-re\-ke\-re}
\fmtSubentry(a)
\fmtSense{to get stuck onto, to be fixed to}
\fmtExampleRL{Eki\-bu\-nguu\-ra kye\-be\-re\-ke\-re oku nte.}
\fmtExampleEN{The caterpillar is stuck onto the cow.}
\fmtSubentry(b)
\fmtSense{fig. to defile}
\fmtExampleRL{Omu\-sai\-za ye\-be\-re\-ke\-re oku kaa\-na aka\-to.}
\fmtExampleEN{A man defiled a very young girl.}

\fmtLemma{ebe\-re\-ngu\-la}
\fmtCitation{(okw)e\-be\-re\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ebe\-re\-ngwi\-re}
\fmtSense{to shine brightly, to be bright}
\fmtExampleRL{Okwe\-zi ku\-kwe\-be\-re\-ngu\-la.}
\fmtExampleEN{The moon is shining brightly.}

\fmtLemma{ebi\-ro di}
\fmtCrossreference{\textnormal{see} bi\-ro}

\fmtLemma{ebo\-ne\-re\-za}
\fmtCitation{(okw)e\-bo\-ne\-re\-za}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ebo\-ne\-rei\-ze}
\fmtSense{to subject oneself to self-denial (lit.\,to punish oneself)}
\fmtExampleRL{Aba\-ka\-tu\-li\-ki bee\-bo\-ne\-re\-za omu ki\-sii\-bo.}
\fmtExampleEN{The Catholics subject themselves to self-denial during Lent.}

\fmtLemma{ebu\-gii\-rya}
\fmtCitation{(okw)e\-bu\-gii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ebu\-gi\-rii\-rye}
\fmtSense{to encircle}
\fmtExampleRL{Aba\-bi\-ni ba\-bi\-na bee\-bu\-gii\-rya.}
\fmtExampleEN{The dancers dance in circles.}

\fmtLemma{ebu\-kii\-ra}
\fmtCitation{(okw)e\-bu\-kii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ebu\-kii\-re}
\fmtSubentry(a)
\fmtSense{to remember, to recall, to think of}
\fmtExampleRL{Yee\-bu\-kii\-re saa\-ti yaa\-mwe ng’ai\-ki\-re.}
\fmtExampleEN{He remembered about his shirt on arrival.}
\fmtSubentry(b)
\fmtSense{to come to realise, to register, to become aware of, to become conscious of}
\fmtExampleRL{We naa\-bii\-re okwe\-bu\-kii\-ra nti mwi\-bbi, ng’a\-ya\-bi\-re.}
\fmtExampleEN{By the time I came to realise that he was a thief, he had left.}

\fmtLemma{ebu\-kya}
\fmtCitation{(okw)e\-bu\-kya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ebu\-ki\-rye}
\fmtSense{to come to realise suddenly}
\fmtExampleRL{Na\-bbai\-re nkwa\-ba Ki\-ti\-mbwa okwe\-bu\-kya nga ndi Mba\-le.}
\fmtExampleEN{I was going to Kitimba only to realise that I had already reached Mbale.}

\fmtLemma{ebu\-la\-nka\-nia}
\fmtCitation{(okw)e\-bu\-la\-nka\-nia}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ebu\-la\-nkai\-nie}
\fmtSense{to pretend to be busy (in order to avoid doing the assignment)}
\fmtExampleRL{Mwo\-jo le\-ka kwe\-bu\-la\-nka\-nia oli\-na oku\-ko\-la olu\-bbi\-mbi lwa\-mu.}
\fmtExampleEN{Please stop pretending to be busy, you must do your assignment.}

\fmtLemma{ebu\-lu\-nga}
\fmtCitation{(okw)e\-bu\-lu\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ebu\-lu\-ngi\-re}
\fmtSense{to play in dirt}
\fmtExampleRL{Omwa\-na akwe\-bu\-lu\-nga.}
\fmtExampleEN{A child is playing in dirt.}

\fmtLemma{ebu\-lu\-ngi}
\fmtCitation{\textnormal{(e.g.}\,kye\-bu\-lu\-ngi, nee\-bu\-lu\-ngi, lwe\-bu\-lu\-ngi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{oval, cylindrical, spherical}
\fmtExampleRL{Ei\-wu\-li erye\-bu\-lu\-ngi lyo\-nkai ni\-ryo li\-zwa\-mu akaa\-na.}
\fmtExampleEN{It is only the oval egg which hatches.}

\fmtLemma{ebu\-lu\-ngu\-li}
\fmtCitation{\textnormal{(e.g.}\,kye\-bu\-lu\-ngu\-li, nee\-bu\-lu\-ngu\-li, lwe\-bu\-lu\-ngu\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSense{circular, round}
\fmtExampleRL{Omu\-pii\-ra gu\-bba gwe\-bu\-lu\-ngu\-li.}
\fmtExampleEN{A ball is round.}

\fmtLemma{ebu\-lya\-bu\-lya}
\fmtCitation{(okw)e\-bu\-lya\-bu\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ebwi\-rye\-bwi\-rye}
\fmtSense{to be evasive, to not to be straightforward}
\fmtExampleRL{Omu\-ka\-ma oyo ye\-bu\-lya\-bu\-lya.}
\fmtExampleEN{That chief is not straightforward.}

\fmtLemma{ebu\-mba}
\fmtCitation{(okw)e\-bu\-mba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ebu\-mbi\-re}
\fmtSubentry(a)
\fmtSense{to feel isolated, to feel lonely}
\fmtExampleRL{Yee\-bu\-mbi\-re omaa\-mwe ng’a\-fii\-re.}
\fmtExampleEN{She felt lonely when she lost her mum.}
\fmtSubentry(b)
\fmtSense{to be humble, to humble oneself}
\fmtExampleRL{Yee\-bu\-mbii\-re omu mai\-so ga paa\-pa.}
\fmtExampleEN{He humbled himself before the pope.}
\fmtSubentry(c)
\fmtSense{to close (one's legs, wings of a butterfly)}
\fmtExampleRL{Omu\-ka\-li yee\-bu\-mbi\-re, n’ai\-mu\-ka.}
\fmtExampleEN{The woman brought her legs together and got up.}

\fmtLemma{ebu\-lu\-ngu\-la}
\fmtCitation{(okw)e\-bu\-lu\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ebu\-lu\-ngwi\-re}
\fmtSense{to roll, to turn over}
\fmtExampleRL{Emo\-to\-ka ya\-gwi\-re ne yee\-bu\-lu\-ngu\-la.}
\fmtExampleEN{The car fell and rolled over.}

\fmtLemma{ebu\-tu\-ka}
\fmtCitation{(okw)e\-bu\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ebu\-tu\-ki\-re}
\fmtSense{to respond, to answer}
\fmtExampleRL{Omaa\-mwe ya\-mwe\-te\-re ni\-yee\-bu\-tu\-ka.}
\fmtExampleEN{He responded when his mother called him.}

\fmtLemma{ebuu\-lya}
\fmtCitation{(okw)e\-buu\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ebwi\-rye}
\fmtSense{to consult, to inquire, to seek advice}
\fmtExampleRL{Ya\-ba wee\-buu\-lye\-ku oku bai\-ra\-baa\-mu.}
\fmtExampleEN{Go and consult your colleagues.}

\fmtLemma{(-)ebw}
\fmtCrossreference{\textnormal{see} (-)ibw}

\fmtLemma{ebwa}
\fmtCitation{(okw)e\-bwa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ebe\-rwe, ebwe\-re}
\fmtSense{to forget}
\fmtExampleRL{Yee\-bwe\-re okwe\-ta aba\-byai\-re okwi\-za oku mu\-ko\-lo.}
\fmtExampleEN{He forgot to invite the parents to come to the function.}

\fmtLemma{ebyo}
\fmtCrossreference{\textnormal{see} o\textnormal{\fmtOne{3}}}

\fmtLemma{ecaa\-ka}
\fmtCitation{(okw)e\-caa\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ecaa\-ki\-re}
\fmtSubentry(a)
\fmtSense{to do skilfully}
\fmtExampleRL{Aba\-bi\-ni bee\-caa\-ki\-re oku si\-tee\-gi.}
\fmtExampleEN{The dancers danced skilfully on the stage.}
\fmtSubentry(b)
\fmtSense{to riot, to demonstrate}
\fmtExampleRL{Abee\-gi ba Ma\-ke\-re\-re be\-ndya mu\-no okwe\-caa\-ka.}
\fmtExampleEN{Makerere University students like rioting.}

\fmtLemma{eca\-nga}
\fmtCitation{(okw)e\-ca\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eca\-ngi\-re}
\fmtSubentry(a)
\fmtSense{to break down, to stop functioning properly}
\fmtExampleRL{Emo\-to\-ka ye\-ca\-ngi\-re ni tu\-kwa\-ba.}
\fmtExampleEN{The car broke down while going to Kampala.}
\fmtSubentry(b)
\fmtSense{to change one's mind}
\fmtExampleRL{Naa\-bi\-re oku\-mu\-twa\-ra empii\-ya ze twa\-la\-gai\-ne, bai\-tu na\-ya\-ji\-rye yee\-ca\-ngi\-re.}
\fmtExampleEN{I went to take to him the money we had agreed upon, but I found that he had changed his mind.}
\fmtSubentry(c)
\fmtSense{to be upset}
\fmtExampleRL{Omwa\-na yee\-ca\-ngi\-re n’a\-ku\-bba abai\-ra baa\-mwe.}
\fmtExampleEN{The child became upset and beat her friends.}

\fmtLemma{ecu\-ngu\-la}
\fmtCitation{(okw)e\-cu\-ngu\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ecu\-ngwi\-re}
\fmtSense{to liberate oneself, to free oneself}
\fmtExampleRL{Yaa\-li atee\-kei\-bwe okwe\-cu\-ngu\-la.}
\fmtExampleEN{She had to liberate herself from captivity.}

\fmtLemma{ecwa\-cwa\-na}
\fmtCitation{(okw)e\-cwa\-cwa\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ecwa\-cwai\-ne}
\fmtSubentry(a)
\fmtSense{to be furious, to be aggressive}
\fmtExampleRL{Ni\-ba\-mu\-ko\-be\-re\-ku ni yee\-cwa\-cwa\-na.}
\fmtExampleEN{He became furious when they gave him advice.}
\fmtSubentry(b)
\fmtSense{to protest, to demonstrate against}
\fmtExampleRL{Aba\-suu\-bu\-zi be\-cwa\-cwai\-ne olwa mu\-so\-lo ogw’a\-ka\-li\-rii\-rye.}
\fmtExampleEN{The business people have demonstrated against high taxes.}

\fmtLemma{edee\-dya}
\fmtCitation{(okw)e\-dee\-dya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{edee\-de\-rye}
\fmtSubentry(a)
\fmtSense{to swing}
\fmtExampleRL{Nsa\-ngi\-re abaa\-na ba\-kwe\-dee\-dya oku mu\-saa\-le.}
\fmtExampleEN{I found the children on a tree swinging.}
\fmtSubentry(b)
\fmtSense{to behave disrespectfully}
\fmtExampleRL{Mwa\-na we le\-ke\-ra awo okwe\-dee\-dya\-nga oku ba\-ntu.}
\fmtExampleEN{Child, stop behaving disrespectfully to people.}

\fmtLemma{edi}
\fmtCrossreference{\textnormal{see} di}
\fmtSubentry(a)
\fmtSubentry(b)
\fmtSense{there (class 23 distal demonstrative)}
\fmtExampleRL{Aya\-bi\-re edi.}
\fmtExampleEN{He has gone there.}

\fmtLemma{edo\-mya}
\fmtCitation{(okw)e\-do\-mya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{edo\-me\-rye}
\fmtSense{to behave foolishly}
\fmtExampleRL{Le\-ke\-ra\-wo okwe\-do\-mya.}
\fmtExampleEN{Stop behaving foolishly.}

\fmtLemma{edyo\-la}
\fmtCitation{(okw)e\-dyo\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{edyoi\-re}
\fmtSense{(of clouds) to gather}
\fmtExampleRL{Ebii\-gu\-ru bi\-kwe\-dyo\-la tu\-yi\-nza oku\-fu\-na oi\-ke\-ndi.}
\fmtExampleEN{Clouds are gathering, we may get some rain.}

\fmtLemma{ee}
\fmtPOS{interj}
\fmtSense{yes}
\fmtExampleRL{Ee, ya\-ya\-bi\-re.}
\fmtExampleEN{Yes, he went.}

\fmtLemma{efoo\-la}
\fmtCitation{(okw)e\-foo\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{efoi\-re}
\fmtSubentry(a)
\fmtSense{to change}
\fmtExampleRL{Embee\-ra ya bwi\-re e Bu\-laa\-ya yee\-foo\-la bu\-li ki\-see\-ra.}
\fmtExampleEN{The weather in Europe changes frequently.}
\fmtSubentry(b)
\fmtSense{to be unreliable, to be inconsistent}
\fmtExampleRL{Oyo mwe\-ge\-nde\-re\-ze, ye\-foo\-la.}
\fmtExampleEN{Be careful, he is unreliable.}
\fmtSubentry(c)
\fmtSense{to impersonate, to pretend to be}
\fmtExampleRL{Ti mu\-sa\-wo bai\-tu yee\-foo\-la\-mbe.}
\fmtExampleEN{He is not a doctor, he is only pretending to be one.}

\fmtLemma{efu\-ga}
\fmtCitation{(okw)e\-fu\-ga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{efu\-gi\-re}
\fmtSubentry(a)
\fmtSense{to behave oneself}
\fmtExampleRL{Boo\-jo mwe\-fu\-ge mu\-ta\-lwa\-ni\-ra me\-re!}
\fmtExampleEN{Boys, behave yourself and don’t fight for food!}
\fmtSubentry(b)
\fmtSense{to gain independence}
\fmtExampleRL{Uga\-nda yee\-fu\-gi\-re mu 1962.}
\fmtExampleEN{Uganda gained independence in 1962.}

\fmtLemma{efu\-kuu\-ra}
\fmtCitation{(okw)e\-fu\-ku\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{efu\-kwi\-re}
\fmtSense{to change one's mind}
\fmtExampleRL{OMu\-so\-ke ni ya\-boi\-ne opu\-re\-zi\-de\-nti ni yee\-fu\-kuu\-ra.}
\fmtExampleEN{Musoke changed his mind when he saw the president.}

\fmtLemma{efu\-ma\-nga\-ra\-na}
\fmtCitation{(okw)e\-fu\-ma\-nga\-ra\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{efu\-ma\-nga\-rai\-ne}
\fmtSense{to be controversial, to be inconsistent}
\fmtExampleRL{Kyo\-to le\-ka kwe\-fu\-ma\-nga\-ra\-na.}
\fmtExampleEN{Kyoto, stop being controversial.}

\fmtLemma{efu\-ndya}
\fmtCitation{(okw)e\-fu\-ndya}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{efu\-ndi\-rye}
\fmtSubentry(a)
\fmtSense{to move over, to make space}
\fmtExampleRL{Mwe\-fu\-ndye\-mu ka\-doo\-li njee\-na nyi\-ka\-le.}
\fmtExampleEN{Kindly move over so that I can sit.}
\fmtSubentry(b)
\fmtSense{to take possession of (something meant for sharing), to appropriate, to seize}
\fmtExampleRL{Omwa\-la wa\-nge yee\-fu\-ndi\-rye eki\-sii\-ka kyo\-na kyo\-na, ta\-sii\-ma mu\-ntu wo\-ndi kwi\-ngi\-ra\-mu.}
\fmtExampleEN{My daughter took possession of the entire room. She does not allow any other person to enter it.}

\fmtLemma{efu\-nga}
\fmtCitation{(okw)e\-fu\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{efu\-ngi\-re}
\fmtSense{to use pads (during menstruation)}
\fmtExampleRL{Omu\-ka\-li n’a\-bba omu kwe\-zi yee\-fu\-nga.}
\fmtExampleEN{When a woman is menstruating, she uses pads.}

\fmtLemma{ega}
\fmtCitation{(okw)e\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ege\-re}
\fmtSense{to learn}
\fmtExampleRL{Ye\-ga oku\-lii\-sya ente.}
\fmtExampleEN{Learn to rear cattle.}

\fmtLemma{egaa\-la}
\fmtCitation{(okw)e\-gaa\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{gai\-re}
\fmtSense{to sit with legs wide apart}
\fmtExampleRL{Omwa\-la odi yee\-gaa\-la.}
\fmtExampleEN{That girl sits with her legs wide apart.}

\fmtLemma{egaa\-na}
\fmtCitation{(okw)e\-gaa\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{egai\-ne}
\fmtSubentry(a)
\fmtSense{to deny, to reject}
\fmtExampleRL{Obu\-nya\-la baa\-bwe\-gai\-ne omu bi\-ro bya bu\-fu\-gi obwa ma\-twa\-le.}
\fmtExampleEN{They denied being Banyala during the colonial period.}
\fmtSubentry(b)
\fmtSense{to disown, to denounce}
\fmtExampleRL{Omwa\-na ye\-gai\-ne oi\-te waa\-mwe.}
\fmtExampleEN{The child has denounced his father.}

\fmtLemma{ega\-ba\-ngu\-la}
\fmtCitation{(okw)e\-ga\-ba\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ega\-ba\-ngwi\-re}
\fmtSense{to be promiscuous, to be licentious, to be loose}
\fmtExampleRL{Omwa\-la waa\-mu yee\-ga\-ba\-ngu\-la omu ba\-sai\-za.}
\fmtExampleEN{Your daughter is licentious.}

\fmtLemma{ega\-ba\-nia}
\fmtCitation{(okw)e\-ga\-ba\-nia}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ega\-bai\-nie}
\fmtSense{to share, to divide among themselves}
\fmtExampleRL{Mba\-di adi ba\-kwe\-ga\-ba\-nia emi\-ye\-mbe.}
\fmtExampleEN{They are there sharing mangoes amongst themselves.}

\fmtLemma{ega\-bu\-la}
\fmtCitation{(okw)e\-ga\-bu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ega\-bwi\-re}
\fmtSubentry(a)
\fmtSense{to surrender, to give up}
\fmtExampleRL{Ya\-gai\-ne okwe\-ga\-bu\-la egi\-ri ega\-vu\-me\-nti.}
\fmtExampleEN{He refused to surrender to the government.}
\fmtSubentry(b)
\fmtSense{to be promiscuous, to be licentious, to be loose}
\fmtExampleRL{Oyo yee\-ga\-bu\-la bu\-lii\-zo eji ba\-sai\-za.}
\fmtExampleEN{That woman is licentious.}

\fmtLemma{egai\-ta}
\fmtCitation{(okw)e\-gai\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{egai\-ti\-re}
\fmtSubentry(a)
\fmtSense{to join}
\fmtExampleRL{OJu\-sti\-ne yee\-gai\-ti\-re oku D.P.}
\fmtExampleEN{Justine joined the DP party.}
\fmtSubentry(b)
\fmtSense{to unite, to get united, to join together, to cooperate}
\fmtExampleRL{Twe\-gai\-te tu\-wa\-ndii\-ke eki\-dee\-ro.}
\fmtExampleEN{Let's unite and write the dictionary.}
\fmtSubentry(c)
\fmtSense{to have sex}

\fmtLemma{ega\-la\-nzu\-la}
\fmtCitation{(okw)e\-ga\-la\-nzu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ega\-la\-nzwi\-re}
\fmtSubentry(a)
\fmtSense{to overturn, to turn over}
\fmtExampleRL{Emo\-to\-ka yee\-ga\-la\-nzwi\-re.}
\fmtExampleEN{The car overturned.}
\fmtSubentry(b)
\fmtSense{to roll}
\fmtExampleRL{Obwa\-na bu\-kwe\-ga\-la\-nzu\-la omu ita\-ka\-li.}
\fmtExampleEN{The children are rolling in the soil.}

\fmtLemma{ega\-li\-ka}
\fmtCitation{(okw)e\-ga\-li\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ega\-li\-ki\-re}
\fmtSense{to lie on the back}
\fmtExampleRL{Omwa\-na yee\-ga\-li\-ki\-re.}
\fmtExampleEN{The child is lying on its back.}

\fmtLemma{ega\-ma}
\fmtCitation{(okw)e\-ga\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ega\-mi\-re}
\fmtSense{to take shelter from (rain or sun), to find cover from, to shelter oneself from}
\fmtExampleRL{Aya\-bi\-re kwe\-ga\-ma isa\-na omu ki\-wo\-lee\-rya.}
\fmtExampleEN{She has gone to take shelter in the shade.}

\fmtLemma{ega\-mbya}
\fmtCitation{(okw)e\-ga\-mbya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{egambi\-rye}
\fmtSense{to use words irresponsibly}
\fmtExampleRL{Aba\-ta\-mii\-re bee\-ga\-mbya mu\-no.}
\fmtExampleEN{Drunkards use words very irresponsibly.}

\fmtLemma{ega\-raa\-lya}
\fmtDialect{Ruruuli}
\fmtCitation{(okw)e\-ga\-raa\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ega\-rai\-rye}
\fmtSense{to pass time, to waste time}
\fmtExampleRL{Nkwa\-ba ku bii\-ci kwe\-ga\-raa\-lya n’o\-mu\-nywa\-ni wa\-nge.}
\fmtExampleEN{I am going to the beach with my friend to pass time.}

\fmtLemma{ega\-yii\-ra}
\fmtCitation{(okw)e\-ga\-yii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ega\-yi\-rii\-re}
\fmtSource{from archaic}
\fmtSense{to plead, to beg}
\fmtExampleRL{Omwi\-bbi ya\-bee\-ga\-yi\-rii\-re ba\-ta\-mwi\-ta.}
\fmtExampleEN{The thief pleaded with them not to kill him.}

\fmtLemma{ege\-mu\-la}
\fmtCitation{(okw)e\-ge\-mu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ege\-mwi\-re}
\fmtSense{to surrender, to give up}
\fmtExampleRL{Yee\-ge\-mwi\-re ye\-nkai omu ba\-te\-mu.}
\fmtExampleEN{He surrendered to the murderers.}

\fmtLemma{ege\-nde\-re\-za}
\fmtCitation{(okw)e\-ge\-nde\-re\-za}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ege\-nde\-rei\-ze}
\fmtSense{to be careful, to be cautious, to watch out}
\fmtExampleRL{Wee\-ge\-nde\-re\-ze egaa\-li eta\-ku\-ku\-bbi\-ra\-wo.}
\fmtExampleEN{Be careful not to fall off the bicycle.}

\fmtLemma{ege\-ra (a)maa\-ni}
\fmtCitation{(okw)e\-ge\-ra}
\fmtPFV{ege\-re}
\fmtSense{to fight}
\fmtExampleRL{Iza ani twe\-ge\-re amaa\-ni.}
\fmtExampleEN{Come here and let's fight.}

\fmtLemma{ege\-sya}
\fmtCitation{(okw)e\-ge\-sya}
\fmtPOS{v}%
\fmtTrans{ditr}
\fmtPFV{ege\-se\-rye}
\fmtSense{to teach, to educate, to instruct}
\fmtExampleRL{Mwe\-ge\-sye abaa\-na oku\-ra\-mya.}
\fmtExampleEN{Teach the children to worship.}

\fmtLemma{ege\-za}
\fmtCitation{(okw)e\-ge\-za}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ege\-ze\-re}
\fmtSense{to try on, to fit}
\fmtExampleRL{Ya\-ba wee\-ge\-ze omu nkai\-to zi\-di.}
\fmtExampleEN{Go and try on those shoes.}

\fmtLemma{ege\-zaa\-mu}
\fmtCitation{(okw)e\-ge\-zaa\-mu}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ege\-ze\-re\-mu}
\fmtSubentry(a)
\fmtSense{to attempt, to try, to make an effort}
\fmtExampleRL{Ye\-ge\-ze\-re\-mu oku\-ku\-bba omaa\-mwe.}
\fmtExampleEN{He tried to beat his mother.}
\fmtSubentry(b)
\fmtSense{to rehearse, to practice}
\fmtExampleRL{Abe\-mbi ba\-kwe\-ge\-zaa\-mu.}
\fmtExampleEN{The singers are rehearsing.}

\fmtLemma{ege\-zi\-wa\-lya}
\fmtCitation{(okw)e\-ge\-zi\-wa\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ege\-zi\-wai\-rye}
\fmtSense{to make oneself wise, to make a pretense of being wise, to be a wiseacre, to be a know-all}
\fmtExampleRL{Omu\-ka\-li waa\-mu oyo ata\-ka mu\-no okwe\-ge\-zi\-wa\-lya.}
\fmtExampleEN{Your wife likes to be a wiseacre so much.}

\fmtLemma{egi\-mba}
\fmtCitation{(okw)e\-gi\-mba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{egi\-mbi\-re}
\fmtSense{(of clouds) to move, to paddle (in a canoe)}
\fmtExampleRL{Ebi\-cu bi\-kwe\-gi\-mba.}
\fmtExampleEN{The clouds are moving.}

\fmtLemma{egi\-nya}
\fmtCitation{(okw)e\-gi\-nya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{egi\-nyi\-re}
\fmtSense{to be crumpled, to be wrinkled, to be folded, to fold oneself up}
\fmtExampleRL{Olu\-go\-ye lwa\-nge lwe\-gi\-nyi\-re.}
\fmtExampleEN{My cloth is crumpled.}

\fmtLemma{egi\-nyi}
\fmtCitation{\textnormal{(e.g.}\,bwe\-ge\-nyi, lwe\-ge\-nyi, mwe\-ge\-nyi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{wrinkled, with creases}
\fmtExampleRL{Aba\-kai\-re aba\-mwe ba\-bba n’o\-bu\-syo obwe\-gi\-nyi.}
\fmtExampleEN{Some old people have wrinkled faces.}

\fmtLemma{egi\-ri}
\fmtPOS{prep}
\fmtSense{to, towards, in front of, before}
\fmtExampleRL{Mwi\-re egi\-ri oKa\-nca le\-ke mu\-fu\-ne emi\-re\-mbe.}
\fmtExampleEN{Turn to God and you will have peace.}

\fmtLemma{ego}
\fmtCrossreference{\textnormal{see} o\textnormal{\fmtOne{3}}}

\fmtLemma{ego\-da}
\fmtCitation{(okw)e\-go\-da}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ego\-de\-re}
\fmtSense{to bend, to stoop, to lower oneself, to bend oneself}
\fmtExampleRL{Omu\-go\-ngo gwa mu\-kai\-re gwe\-go\-de\-re.}
\fmtExampleEN{The old woman is stooping.}

\fmtLemma{ego\-de\-ra}
\fmtCitation{(okw)e\-go\-de\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ego\-dei\-re}
\fmtSense{to persist, to endure}
\fmtExampleRL{Wee\-go\-de\-re oku nsi\-ka okwi\-kya ni wa\-yi\-kya olu\-bbi\-mbi.}
\fmtExampleEN{Keep going until you finish the part you are digging.}

\fmtLemma{ego\-ga}
\fmtCitation{(okw)e\-go\-ga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ego\-ge\-re}
\fmtSense{to retch, to gag, to make an effort to vomit, to heav}
\fmtExampleRL{Omwa\-na ata\-ndi\-ki\-re okwe\-go\-ga.}
\fmtExampleEN{The child has started to retch.}

\fmtLemma{ego\-ka}
\fmtCrossreference{\textnormal{see} ego\-ga}

\fmtLemma{ego\-lo\-la}
\fmtCitation{(okw)e\-go\-lo\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ego\-loi\-re}
\fmtSubentry(a)
\fmtSense{to stretch out, to stretch oneself}
\fmtExampleRL{Mwi\-kai\-re mu\-no mwa\-be mwe\-go\-lo\-le\-mu.}
\fmtExampleEN{You have been sitting for a long time, go and stretch out.}
\fmtSubentry(b)
\fmtSense{to stand straight on attention}
\fmtExampleRL{Omu\-si\-ri\-ka\-le ye\-go\-loi\-re.}
\fmtExampleEN{The solder is standing straight at attention.}

\fmtLemma{ego\-mba}
\fmtCitation{(okw)e\-go\-mba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ego\-mbe\-re}
\fmtSense{to admire, to desire, to crave}
\fmtExampleRL{Nee\-go\-mba amaa\-ni ga Sau\-da\-h.}
\fmtExampleEN{I admire Saudah's stamina.}

\fmtLemma{egu\-gu\-nga}
\fmtCitation{(okw)e\-gu\-gu\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{egu\-gu\-ngi\-re}
\fmtSense{to riot, to go on strike, to rebel, to demonstrate, to protest}
\fmtExampleRL{Aba Ma\-ke\-re\-re bee\-gu\-gu\-nga mu\-no.}
\fmtExampleEN{Makererians often go on strike.}

\fmtLemma{egu\-la}
\fmtCitation{(okw)e\-gu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eguu\-re}
\fmtSense{to pay in order to get a land title}
\fmtExampleRL{Mwe\-gu\-le egi mu\-ka\-ma wa ita\-ka\-li.}
\fmtExampleEN{Pay and get the land titles from the landlord.}

\fmtLemma{egu\-mbuu\-ra}
\fmtCitation{(okw)e\-gu\-mbuu\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{egu\-mbu\-lii\-re}
\fmtSense{to acquire a bad habit, to start doing (something bad), to fall into the habit of, to become involved in}
\fmtExampleRL{Yee\-gu\-mbuu\-lii\-re oku\-nywa enjaa\-ye.}
\fmtExampleEN{He uses marijuana habitually.}

\fmtLemma{egu\-nga}
\fmtCitation{(okw)e\-gu\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{egu\-ngi\-re}
\fmtSense{to riot, to go on strike, to rebel, to demonstrate, to protest}
\fmtExampleRL{Aba\-ko\-li ba\-kwe\-gu\-nga lwa ku\-ba\-ka\-lya ku\-bbi.}
\fmtExampleEN{Workers are demonstrating against mistreatment.}

\fmtLemma{eguu\-da}
\fmtCrossreference{\textnormal{see} guu\-da}
\fmtCitation{(okw)e\-guu\-da}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eguu\-di\-re}
\fmtSubentry(a)
\fmtSense{to spear oneself}
\fmtExampleRL{Omu\-yii\-gi yee\-guu\-di\-re ei\-su\-mu n’a\-fa.}
\fmtExampleEN{The hunter speared himself to death.}
\fmtSubentry(b)
\fmtSense{to lack appetite}
\fmtExampleRL{Omwo\-yo gwa\-nge gwe\-guu\-di\-re oku\-lya enya\-ma.}
\fmtExampleEN{I have no appetite for meat.}

\fmtLemma{egu\-ya}
\fmtCitation{(okw)e\-gu\-ya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{egu\-yi\-rye}
\fmtSense{to gratify, to please, to cajole, to flatter, to regard as important}
\fmtExampleRL{Oli\-na okwe\-gu\-ya omaa\-wo ni waa\-bba okwe\-ndya aku\-sa\-suu\-re ebi\-sa\-le bya iso\-me\-ro.}
\fmtExampleEN{You have to please your mother if you want her to pay the school fees.}

\fmtLemma{egyo}
\fmtCrossreference{\textnormal{see} o\textnormal{\fmtOne{3}}}

\fmtLemma{ei\-bba}
\fmtCitation{(okw)ei\-bba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eibbi\-re}
\fmtSubentry(a)
\fmtSense{to escape, to run away, to break out, to break free, to slip away}
\fmtExampleRL{Yei\-bbi\-re omu nko\-myo.}
\fmtExampleEN{He escaped from the prison.}
\fmtSubentry(b)
\fmtSense{to move away stealthily, to sneak out}
\fmtExampleRL{Tu\-bbai\-re na\-ye ani bai\-tu atwei\-bbi\-re\-ku n’a\-ya\-ba.}
\fmtExampleEN{We were here together, but he snuck away.}

\fmtLemma{ei\-bba\-ka}
\fmtCitation{(okw)ei\-bba\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eibba\-ki\-re}
\fmtSubentry(a)
\fmtSense{to meddle, to have restless fingers, to have the habit of picking up things}
\fmtSubentry(b)
\fmtSense{to steal}
\fmtExampleRL{Omwa\-na yei\-bba\-ka n’e\-bya bai\-ra\-baa\-mwe.}
\fmtExampleEN{The child steals his friends' property.}

\fmtLemma{ei\-bbii\-ra}
\fmtCitation{(okw)ei\-bbii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eibbi\-rii\-re}
\fmtSense{to do something stealthily, to approach stealthily, to poach}
\fmtExampleRL{Aba\-ntu aba\-mwe bei\-bbii\-ra ni ba\-nywa omwe\-nge.}
\fmtExampleEN{Some people drink alcohol in secret.}

\fmtLemma{ei\-bbi\-rya}
\fmtCrossreference{\textnormal{see} ebbi\-rya}
\fmtCitation{(okw)ei\-bbi\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eibbii\-rye}
\fmtSubentry(a)
\fmtSense{to dive}
\fmtExampleRL{Mmai\-te okwei\-bbi\-rya omu mai\-zi.}
\fmtExampleEN{I know how to dive in water.}
\fmtSubentry(b)
\fmtSense{to become immersed in, to be utterly dedicated to (\eg God, studies)}
\fmtSubentry(c)
\fmtSense{to deny one's identity, to deny a charge or accusation}
\fmtExampleRL{Otei\-bbi\-rya! Ba\-za nti oli Mu\-ruu\-li.}
\fmtExampleEN{Don't deny your identity! Say that you are a Muruuli.}

\fmtLemma{ei\-bbi\-sa}
\fmtCitation{(okw)ei\-bbi\-sa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eibbi\-si\-re}
\fmtSense{to hide, to take cover}
\fmtExampleRL{Oku lu\-ta\-lo twei\-bbi\-sa\-nga mu ka\-pa\-nga.}
\fmtExampleEN{During the war we used to hide in Lantana camara.}

\fmtLemma{ei\-be\-mbe\-ra}
\fmtCrossreference{\textnormal{see} ebe\-mbe\-ra}

\fmtLemma{ei\-ca}
\fmtCitation{(okw)ei\-ca}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eici\-rye}
\fmtSubentry(a)
\fmtSense{to be well off}
\fmtExampleRL{OKa\-ta\-bu\-la yei\-ca ebi\-ro bi\-ni.}
\fmtExampleEN{Katabula is well-off of these days.}
\fmtSubentry(b)
\fmtSense{to exalt oneself, to act in a boastful manner, to be filled with self-conceit}
\fmtExampleRL{OKa\-te\-re\-ga yei\-ca mu\-no.}
\fmtExampleEN{Katerega exalts himself very much.}
\fmtSubentry(c)
\fmtFixedexpression{eica (o)mwo\-yo}
\fmtSense{to be relieved}
\fmtExampleRL{Ni ba\-nko\-bei\-re nti omu\-ka\-ma akwi\-za ni nei\-ca mwo\-yo.}
\fmtExampleEN{I was relieved to learn that the chief is coming.}

\fmtLemma{ei\-caa\-ka}
\fmtCrossreference{\textnormal{see} ecaa\-ka}

\fmtLemma{ei\-cu\-ngu\-la}
\fmtCrossreference{\textnormal{see} ecu\-ngu\-la}

\fmtLemma{ei\-ga\-ma}
\fmtCrossreference{\textnormal{see} ega\-ma}

\fmtLemma{ei\-go\-ga}
\fmtCrossreference{\textnormal{see} ego\-ga}

\fmtLemma{ei\-ka}
\fmtCitation{(okw)ei\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eiki\-re}
\fmtSense{to push (when giving birth)}
\fmtExampleRL{Wei\-ke omwa\-na awu\-lu\-ke.}
\fmtExampleEN{Push the child out.}

\fmtLemma{ei\-ka\-ma}
\fmtCitation{(okw)ei\-ka\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eika\-mi\-re}
\fmtSense{to push (when giving birth), to push hard when defecating (\eg when you are constipated)}
\fmtExampleRL{Wei\-ka\-me omwa\-na awu\-lu\-ke.}
\fmtExampleEN{Push the child out.}

\fmtLemma{ei\-ki\-rii\-rya}
\fmtCitation{(okw)ei\-ki\-rii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eiki\-ri\-rii\-rye}
\fmtSubentry(a)
\fmtSense{to murmur in agreement, to produce indistinct sounds to express affirmation or agreement}
\fmtExampleRL{Ba\-ku\-mu\-ko\-ba akwei\-ki\-rii\-rya\-mbe.}
\fmtExampleEN{They were talking to her and she just kept murmuring in agreement.}
\fmtSubentry(b)
\fmtSense{to moan, to groan}

\fmtLemma{ei\-ki\-rii\-rya\-mu}
\fmtCitation{(okw)ei\-ki\-rii\-rya\-mu}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eiki\-rii\-rye\-mu}
\fmtSense{to be confident, to believe in oneself}
\fmtExampleRL{ORu\-th yei\-ki\-rii\-rya\-mu.}
\fmtExampleEN{Ruth believes in herself.}

\fmtLemma{ei\-ki\-sya\-mu\-lya}
\fmtCitation{(okw)ei\-ki\-sya\-mu\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eiki\-sya\-mwi\-rye}
\fmtSubentry(a)
\fmtSense{to sneeze}
\fmtExampleRL{Omwa\-na yei\-ki\-sya\-mwi\-rye alwai\-re eki\-yi\-nzi.}
\fmtExampleEN{The child is sneezing because of the flu.}
\fmtSubentry(b)
\fmtSense{to be disgusted with, to give up on}

\fmtLemma{ei\-ko\-mai\-ko\-ma}
\fmtCitation{(okw)ei\-ko\-mai\-ko\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eiki\-sya\-mwi\-rye}
\fmtSubentry(a)
\fmtSense{to steal}
\fmtExampleRL{Omwa\-na wa Ga\-nse\-ko ata\-ndi\-ki\-re okwei\-ko\-mai\-ko\-ma.}
\fmtExampleEN{Ganseko’s child has started to steal.}
\fmtSubentry(b)
\fmtSense{to quarrel over trivial matters, to pick on, to nag}
\fmtExampleRL{Wee\-na wei\-ko\-mai\-ko\-ma, eka\-laa\-mu yoo\-na n’e\-ku\-lwa\-ni\-sya?}
\fmtExampleEN{You quarrel over trivial matters, can a mere pencil cause you to quarrel?}

\fmtLemma{ei\-ma\-ga\-la\-lya}
\fmtCrossreference{\textnormal{see} ema\-ga\-la\-lya}

\fmtLemma{ei\-mbya}
\fmtCitation{(okw)ei\-mbya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eimbi\-rye}
\fmtSubentry(a)
\fmtSense{to join hands, to come together, to unite, to come together}
\fmtExampleRL{Amo\-s yee\-yi\-mbi\-rye n’o\-Ka\-to oku\-lu\-ma amai\-do.}
\fmtExampleEN{Amos and Kato joined forces and grew groundnuts.}
\fmtSubentry(b)
\fmtSense{to connive, to conspire, to plot}
\fmtExampleRL{Amo\-s yee\-yi\-mbi\-rye n’o\-Gra\-ce ni bai\-bba egaa\-li ya Nte\-ge.}
\fmtExampleEN{Amos and Grace conspired and stole Ntege's bicycle.}

\fmtLemma{ei\-na\-mi\-ra}
\fmtCrossreference{\textnormal{see} enamira}

\fmtLemma{ei\-na\-ti\-ra}
\fmtCrossreference{\textnormal{see} ena\-ti\-ra}

\fmtLemma{ei\-ni\-ka}
\fmtCitation{(okw)ei\-ni\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eini\-ki\-re}
\fmtSense{to sink, to become submerged, to immerse oneself}
\fmtExampleRL{Obwa\-to bwe twa\-ta\-ka\-nga okwa\-bi\-ra\-mu bwei\-ni\-ki\-re ni bwi\-ta aba\-ntu.}
\fmtExampleEN{The boat which we wanted to travel with was sunk and killed people.}

\fmtLemma{ei\-nu\-la}
\fmtCitation{(okw)ei\-nu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{einwi\-re}
\fmtSense{to do to one's satisfaction}
\fmtExampleRL{Na\-lii\-re ni nei\-nu\-la.}
\fmtExampleEN{I ate to my satisfaction.}

\fmtLemma{ei\-nu\-nu\-la}
\fmtCitation{(okw)ei\-nu\-nu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{einu\-nwi\-re}
\fmtSense{(of a child) to grow teeth, to develop teeth}
\fmtExampleRL{Omwa\-na yei\-nu\-nwi\-re.}
\fmtExampleEN{The child has grown teeth.}

\fmtLemma{ei\-nyai\-nya}
\fmtCitation{(okw)ei\-nyai\-nya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{einyai\-nyi\-re}
\fmtSense{(of a cloth) to wrinkle, to crumple}
\fmtExampleRL{Esaa\-ti ya\-nge yei\-nyai\-nyi\-re.}
\fmtExampleEN{My shirt is wrinkled.}

\fmtLemma{ei\-pa\-mpa\-li\-ka}
\fmtCrossreference{\textnormal{see} epa\-mpa\-li\-ka}

\fmtLemma{ei\-pa\-nka}
\fmtCrossreference{\textnormal{see} epa\-nka}

\fmtLemma{ei\-rai}
\fmtCrossreference{\textnormal{see} irai}

\fmtLemma{ei\-ra\-ku}
\fmtCitation{(okw)ei\-ra\-ku}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eire\-ku}
\fmtSense{to change (for the better), to reform, to improve}
\fmtExampleRL{Mwi\-ra\-wa\-nge wei\-re\-ku.}
\fmtExampleEN{My dear, change for the better.}

\fmtLemma{ei\-rya}
\fmtCitation{(okw)ei\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eiryi\-re}
\fmtSense{(of a ghost or spirit) to haunt, to come back among the living}
\fmtExampleRL{Omu\-zi\-mu gu\-kwei\-rya oJa\-ne.}
\fmtExampleEN{The ghost is haunting Jane.}

\fmtLemma{ei\-sa\-nii\-sii\-rya}
\fmtCitation{(okw)ei\-sa\-nii\-sii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eisa\-ni\-sii\-rye}
\fmtSense{to pretend, compare oneself, imagine oneself, imagine, visualize}
\fmtExampleRL{Ta\-yi\-nza ku\-byei\-sa\-nii\-sii\-rya.}
\fmtExampleEN{He cannot imagine them.}

\fmtLemma{ei\-sa\-ni\-ka}
\fmtCitation{(okw)ei\-sa\-ni\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eisa\-ni\-ki\-re}
\fmtSense{to show off}
\fmtExampleRL{Omu\-ka\-li oni yei\-sa\-ni\-ka mu\-no.}
\fmtExampleEN{This woman shows off a lot.}

\fmtLemma{ei\-si\-si\-ra}
\fmtCrossreference{\textnormal{see} esi\-si\-ra}

\fmtLemma{ei\-suu\-ka}
\fmtCitation{(okw)ei\-suu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eisuu\-ki\-re}
\fmtSense{to return, to come back}
\fmtExampleRL{Wei\-suu\-ke\-yo gy’o\-ku\-zwa.}
\fmtExampleEN{Welcome back from where you had gone.}

\fmtLemma{ei\-suu\-kya\-yo}
\fmtCitation{(okw)ei\-suu\-kya\-yo}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eisuu\-ki\-rye\-yo}
\fmtSense{to welcome}
\fmtExampleRL{Mwei\-suu\-kye\-yo oi\-te waa\-nywe.}
\fmtExampleEN{Welcome your father.}

\fmtLemma{ei\-ta\-ki\-sya}
\fmtCrossreference{\textnormal{see} eta\-ki\-sya}

\fmtLemma{ei\-te\-ja}
\fmtCrossreference{\textnormal{see} ete\-ja}

\fmtLemma{ei\-tii\-ka}
\fmtCrossreference{\textnormal{see} etii\-ka}

\fmtLemma{ei\-ti\-ja}
\fmtCitation{(okw)ei\-ti\-ja}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eiti\-ji\-rye}
\fmtSense{to pretend to be weak}
\fmtExampleRL{We lwa\-ki okwei\-ti\-ja ng’a\-maa\-ni gaa\-mu ota n'a\-go?}
\fmtExampleEN{Why do you pretend to be weak when you are strong?}

\fmtLemma{ei\-to\-nga}
\fmtCrossreference{\textnormal{see} eto\-nga}

\fmtLemma{ei\-tu\-mbya}
\fmtCrossreference{\textnormal{see} etu\-mbya}

\fmtLemma{ei\-ya\-wo}
\fmtCitation{(okw)ei\-ya\-wo}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eiyi\-re\-wo}
\fmtSense{to leave, to depart from, to go away from}
\fmtExampleRL{Iswe twei\-yi\-re\-wo tu\-kwa\-ba Ka\-mpa\-la.}
\fmtExampleEN{We have left for Kampala.}

\fmtLemma{ei\-zi\-ra}
\fmtCitation{(okw)ei\-zi\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eizii\-re}
\fmtSubentry(a)
\fmtSense{(of a member of a clan) to belong to a totem, to adopt as a totem}
\fmtExampleRL{Weei\-zi\-ra mu\-zi\-ro ki?}
\fmtExampleEN{Which totem do you belong to?}
\fmtSubentry(b)
\fmtSense{to regard as an anathema, to hold as an abomination}

\fmtLemma{ei\-zu\-ka\-nya}
\fmtCitation{(okw)ei\-zu\-ka\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eizu\-ka\-nyi\-rye}
\fmtSense{to revise, to edit, to correct}
\fmtExampleRL{Mwei\-zu\-ka\-nye\-nge bye mu\-wa\-ndii\-ki\-re.}
\fmtExampleEN{Revise what you have written.}

\fmtLemma{ei\-zu\-ki\-rya}
\fmtCitation{(okw)ei\-zu\-ki\-rya}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{eizu\-ki\-rye}
\fmtSubentry(a)
\fmtSense{to remember, to recall, to recollect}
\fmtExampleRL{Fu\-ba\-nga mu\-no okwei\-zu\-ki\-rya bye wa\-ye\-ge\-re.}
\fmtExampleEN{Try very hard to remember what you learnt.}
\fmtSubentry(b)
\fmtSense{to realise, to perceive, to understand, to grasp, to comprehend}
\fmtExampleRL{Na\-ya\-bi\-re okwei\-zu\-ki\-rya obu\-kwe\-nda ng’a\-ya\-bi\-re.}
\fmtExampleEN{I understood the message after he had gone.}
\fmtSubentry(c)
\fmtSense{to be self-driven, to be self-motivated}
\fmtExampleRL{Fu\-ba\-nga mu\-no okwei\-zu\-ki\-rya omu mi\-ru\-mo gya\-mu.}
\fmtExampleEN{Try to be self-driven in whatever you do.}

\fmtLemma{ei\-zu\-kya}
\fmtCitation{(okw)ei\-zu\-kya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eizu\-ki\-rye}
\fmtSense{to regret, to be sorry about, to rue}
\fmtExampleRL{Okwi\-za kwei\-zu\-kya nga to\-so\-me\-re.}
\fmtExampleEN{You will regret it if you don’t study.}

\fmtLemma{eja\-ga}
\fmtCitation{(okw)e\-ja\-ga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eja\-gi\-re}
\fmtSubentry(a)
\fmtSense{to be excited, to make merry, to cheer up}
\fmtExampleRL{Olwa\-lee\-ro aba\-ntu be\-ja\-gi\-re ni baa\-boi\-ne oku omu\-ka\-ma.}
\fmtExampleEN{Today the people were excited when they saw the king.}
\fmtSubentry(b)
\fmtSense{to bustle, to be full of activity}
\fmtExampleRL{Na\-mu\-boi\-ne nka akwe\-ja\-ga n'a\-ba\-ge\-ni.}
\fmtExampleEN{I saw her busying herself with visitors.}

\fmtLemma{ejee\-guu\-la}
\fmtCitation{(okw)e\-jee\-guu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ejee\-gwi\-re}
\fmtSense{to behave indecently, to bring oneself into disrepute}
\fmtExampleRL{Le\-ka kwe\-jee\-guu\-la\-nga mu ba\-ntu.}
\fmtExampleEN{Stop behaving indecently in public.}

\fmtLemma{eji}
\fmtPOS{prep}
\fmtSense{to, towards}
\fmtExampleRL{Aya\-bi\-re eji ibaa\-mwe.}
\fmtExampleEN{She has gone to her husband.}

\fmtLemma{eju\-na}
\fmtCitation{(okw)e\-ju\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eju\-ni\-re}
\fmtSense{to give birth, to bear}
\fmtExampleRL{Omwe\-rya\-na wa Na\-mpii\-na yee\-ju\-ni\-re omwa\-na mwo\-jo.}
\fmtExampleEN{Nampiina’s daughter-in-law has given birth to a baby boy.}

\fmtLemma{ejuu\-mu\-ka}
\fmtCitation{(okw)e\-juu\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ejuu\-mu\-ki\-re}
\fmtSense{to lie down, to be turned upside down, to be turned over}
\fmtExampleRL{Ka\-nee\-juu\-mu\-ke\-ku ani.}
\fmtExampleEN{Let me lie down here for a while.}

\fmtLemma{eka}
\fmtCitation{eka}
\fmtPOS{n}%
\fmtNC{9}
\fmtSubentry(a)
\fmtSense{home}
\fmtExampleRL{Ya\-ba ma\-ngu eka.}
\fmtExampleEN{Go home right away.}
\fmtSubentry(b)
\fmtSense{family}
\fmtExampleRL{Eka ya Mu\-so\-ke te\-ta na mpi\-sa.}
\fmtExampleEN{Musoke's family is undisciplined.}

\fmtLemma{ekaa\-ka}
\fmtCitation{(okw)e\-kaa\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ekaa\-ki\-re}
\fmtSubentry(a)
\fmtSense{to crack, to break, to split, to crackle}
\fmtExampleRL{Mpu\-lii\-re omu\-saa\-le nga gu\-kwe\-kaa\-ka.}
\fmtExampleEN{I have heard a tree cracking.}
\fmtSubentry(b)
\fmtSense{to have a pulled muscle, to have sprained ligaments}
\fmtExampleRL{Eki\-nywa ky’o\-mu\-zee\-nyi wa mu\-pii\-ra kye kaa\-ki\-re.}
\fmtExampleEN{The football player has sprained his muscle.}

\fmtLemma{eka\-ka\-ba}
\fmtCitation{(okw)e\-ka\-ka\-ba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eka\-ka\-bi\-re}
\fmtSense{to strive, to try hard, to do one's best}
\fmtExampleRL{Otee\-kwa okwe\-ka\-ka\-ba le\-ke otu\-nge kyo\-kwe\-ndya.}
\fmtExampleEN{You must strive in order to achieve what you want.}

\fmtLemma{eka\-ka\-tya}
\fmtCitation{(okw)e\-ka\-ka\-tya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eka\-ka\-ti\-rye}
\fmtSense{to be sure, to be certain, to assure oneself, to feel confident}
\fmtExampleRL{Nkwe\-ka\-ka\-tya kye nku\-ko\-ba.}
\fmtExampleEN{I am sure of what I am saying.}

\fmtLemma{eka\-laa\-bbii\-rya}
\fmtCitation{(okw)e\-ka\-laa\-bbii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eka\-laa\-bbi\-rii\-rye}
\fmtSense{to make an effort, to strive, to try hard}
\fmtExampleRL{Tu\-kwe\-ka\-la\-bbii\-rya tu\-ma\-le ma\-ngu eki\-dee\-ro ki\-ni.}
\fmtExampleEN{We are striving to finish this dictionary quickly.}

\fmtLemma{eka\-la\-kaa\-sa}
\fmtCrossreference{\textnormal{see also} egu\-gu\-nga}
\fmtCitation{(okw)e\-ka\-la\-kaa\-sa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eka\-la\-kaa\-si\-re}
\fmtSense{to demonstrate, to protest}
\fmtExampleRL{Abee\-gi bee\-ka\-la\-kaa\-sa lwa bu\-ta\-ba\-wu\-lii\-rya.}
\fmtExampleEN{Students demonstrate against not being heard.}

\fmtLemma{eka\-la\-ngu\-la}
\fmtCitation{(okw)e\-ka\-la\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eka\-la\-ngwi\-re}
\fmtSubentry(a)
\fmtSense{to persevere, to persist, to continue, to carry on, to go on, to keep on}
\fmtExampleRL{Boo\-jo twe\-ka\-la\-ngu\-le tu\-ma\-le\-ku olu\-bbi\-mbi lu\-ni.}
\fmtExampleEN{Boys, let us persevere and complete this task.}
\fmtSubentry(b)
\fmtSense{to work hard, to be determined}
\fmtExampleRL{Ka ne\-ka\-la\-ngu\-le ntu\-nge ebya\-nge.}
\fmtExampleEN{Let me work hard and get some wealth for myself.}

\fmtLemma{eka\-li\-rii\-rya}
\fmtCitation{(okw)e\-ka\-li\-rii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eka\-li\-rii\-rye}
\fmtSense{to look keenly, to look intently, to look closely, to stare, to examine}
\fmtExampleRL{Naa\-li ndi\-na okwe\-ka\-li\-rii\-rya nta\-ko\-la nso\-bi.}
\fmtExampleEN{I had to look intently so that I didn’t make a mistake.}

\fmtLemma{eka\-lya}
\fmtCitation{(okw)e\-ka\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ekaa\-rye}
\fmtSense{to pretend to be fine, to turn a blind eye}
\fmtExampleRL{Ba\-mu\-ku\-bbi\-re bai\-tu akwe\-ka\-lya\-mbe.}
\fmtExampleEN{He has been beaten, he is just pretending to be okay.}

\fmtLemma{eka\-mba}
\fmtCitation{(okw)e\-ka\-mba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eka\-mbi\-re}
\fmtSense{to put in effort, to strive}
\fmtExampleRL{Mwa\-na wa\-nge wee\-ka\-mbe oli\-ge omu bwo\-ro.}
\fmtExampleEN{My child, please put in more effort and get out of poverty.}

\fmtLemma{eka\-na\-nsu\-la}
\fmtCitation{(okw)e\-ka\-sa\-nsu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eka\-na\-nswi\-re}
\fmtSense{to treat with revulsion, to show disgust, to show aversion}
\fmtExampleRL{Le\-ke\-ra\-wo okwe\-ka\-na\-nsu\-la oku ba\-lwi\-re.}
\fmtExampleEN{Stop treating the patients with revulsion.}

\fmtLemma{eka\-na\-sa}
\fmtCitation{(okw)e\-ka\-na\-sa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eka\-na\-si\-re}
\fmtSubentry(a)
\fmtSense{to show aversion, to show dislike}
\fmtExampleRL{Aba\-ka\-li ba\-ta\-ka mu\-no okwe\-ka\-na\-sa.}
\fmtExampleEN{Women like showing aversion.}
\fmtSubentry(b)
\fmtSense{to be fussy, to be touchy}
\fmtExampleRL{Lwa\-ki we\-ka\-na\-sa bu\-li lu\-na\-ku?}
\fmtExampleEN{Why are you finicky every day?}

\fmtLemma{eka\-pi\-ka}
\fmtCitation{(okw)e\-ka\-pi\-ka}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{eka\-pi\-ki\-re}
\fmtSubentry(a)
\fmtSense{to wear many clothes, to wear tight clothes}
\fmtExampleRL{Nee\-ka\-pi\-ki\-re engo\-ye isa\-tu.}
\fmtExampleEN{I have worn three garments.}
\fmtSubentry(b)
\fmtSense{to be dressed in formal atti\-re}
\fmtSubentry(c)
\fmtSense{to have a sour face}
\fmtExampleRL{Abbai\-re ye\-ka\-pi\-ki\-re mu\-no ei\-go\-lo.}
\fmtExampleEN{He had a sour face last night.}

\fmtLemma{eka\-yu\-la}
\fmtCitation{(okw)e\-ka\-yu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ekaywi\-re}
\fmtSense{to show resentment, to show dissatisfaction, to show discontentment}
\fmtExampleRL{Lwa\-ki okwe\-ka\-yu\-la ku mwa\-na omu\-to?}
\fmtExampleEN{Why do you show resentment to a child?}

\fmtLemma{eka\-yu\-li}
\fmtCitation{\textnormal{(e.g.}\,mwe\-ka\-yu\-li, bee\-ka\-yu\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSense{irritable, irritable, bad-tempered, short-tempered, touchy, moody}
\fmtExampleRL{Omu\-ko\-li oyo mwe\-ka\-yu\-li.}
\fmtExampleEN{That worker is irritable.}

\fmtLemma{eke\-be\-ja}
\fmtCitation{(okw)e\-ke\-be\-ja}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eke\-be\-je\-re}
\fmtSense{to examine, to inspect, to scrutinise, to investigate, to analyse, to review, to evaluate, to survey}
\fmtExampleRL{Odo\-ki\-ta ali mu ka\-sii\-ka akwe\-ke\-be\-ja omu\-lwai\-re.}
\fmtExampleEN{The doctor is in the examination room examining the patient.}

\fmtLemma{ekee\-ra}
\fmtCitation{(okw)e\-kee\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eke\-rei\-re}
\fmtSubentry(a)
\fmtSense{to be first, to do first}
\fmtExampleRL{Ni mu\-kwi\-ngi\-ra ebbaa\-si, oni ni\-ye yee\-kee\-ra.}
\fmtExampleEN{When boarding a bus, this one will be the first to enter.}
\fmtSubentry(b)
\fmtFixedexpression{mwe\-zi (o)gu\-kwe\-kee\-ra}
\fmtSense{January}

\fmtLemma{eke\-nde\-ke\-ka}
\fmtCitation{(okw)e\-ke\-nde\-ke\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eke\-nde\-ke\-ke\-re}
\fmtSense{to be extra careful}
\fmtExampleRL{Oka\-bo\-na yee\-ke\-nde\-ke\-ka na ku\-wee\-re\-za oKa\-nca.}
\fmtExampleEN{A priest is extra careful while serving God.}

\fmtLemma{eke\-nge\-ra}
\fmtCitation{(okw)e\-ke\-nge\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eke\-ngei\-re}
\fmtSubentry(a)
\fmtSense{to be cautious of, to be suspicious of, to beware of, to be on guard against}
\fmtExampleRL{Omu\-sai\-za odi mwe\-ke\-nge\-re, ami\-nsa ebya\-la.}
\fmtExampleEN{Be cautious of that man, he may affect you with spells.}
\fmtSubentry(b)
\fmtSense{to suspect}
\fmtExampleRL{Nkwe\-ke\-nge\-ra omu\-ntu oyo ya\-ka\-bba omwi\-ti.}
\fmtExampleEN{I suspect that person to be a murderer.}

\fmtLemma{eke\-ree\-ra}
\fmtCrossreference{\textnormal{see} ekee\-ra}

\fmtLemma{ekii\-ka}
\fmtCitation{(okw)e\-kii\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ekii\-ki\-re}
\fmtSubentry(a)
\fmtSense{to lie across, to place oneself sideways}
\fmtExampleRL{Eki\-saa\-le kye\-kii\-ki\-re omu nzi\-ra.}
\fmtExampleEN{A tree is lying across the road.}
\fmtSubentry(b)
\fmtSense{to stand in the way of, to block the way of}
\fmtExampleRL{Omwo\-jo yee\-kii\-ki\-re omu nso\-nga zai\-swe.}
\fmtExampleEN{The young man has stood in the way of our concerns.}
\fmtSubentry(c)
\fmtSense{(of an unborn baby) to lie in a breech position}
\fmtExampleRL{Omwa\-na yee\-kii\-ki\-re omu nda ya maa\-mwei.}
\fmtExampleEN{The baby is presenting itself in a breech position in the mother's womb.}

\fmtLemma{ekii\-ki\-ngi\-ri\-za}
\fmtCitation{(okw)e\-kii\-ki\-ngi\-ri\-za}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ekii\-ki\-ngi\-rii\-ze}
\fmtSense{to exalt oneself, to elevate by praise, to oneself}
\fmtExampleRL{OYe\-su ya\-ko\-be\-re nti bu\-li ye\-kii\-ki\-ngi\-ri\-za ali\-ka\-ka\-nyi\-zi\-bwa.}
\fmtExampleEN{Jesus said that whoever exalts himself will be humbled.}

\fmtLemma{eko\-do\-lya}
\fmtDialect{Ruruuli}
\fmtCitation{(okw)e\-ko\-do\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eko\-doi\-rye}
\fmtSubentry(a)
\fmtSense{to grumble over trivial matters}
\fmtExampleRL{ONa\-mbu\-bi yee\-ko\-do\-lya n’e\-bi\-ntu abi\-taa\-lu\-mu.}
\fmtExampleEN{Nambubi grumbles over trivial matters.}
\fmtSubentry(b)
\fmtSense{to engage in petty theft, to engage in misdemeanour}
\fmtExampleRL{Omwo\-jo odi yee\-ko\-do\-lya.}
\fmtExampleEN{That boy engages in petty theft.}

\fmtLemma{eko\-ko\-ta}
\fmtCitation{(okw)e\-ko\-ko\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eko\-ko\-te\-re}
\fmtSense{to crawl, to creep, to go on all fours}
\fmtExampleRL{Naa\-ni oyo akwe\-ko\-ko\-te\-ra omu maa\-ya?}
\fmtExampleEN{Who is that crawling beneath the bed?}

\fmtLemma{eko\-ko\-tee\-rya}
\fmtCitation{(okw)e\-ko\-ko\-tee\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eko\-ko\-te\-rei\-rye}
\fmtSense{to resist, to oppose, to fight against, to refuse to accept, to object to, to defy}
\fmtExampleRL{Ombu\-li ba\-ku\-mu\-twa\-la oku\-lya ei\-su\-bi, akwe\-ko\-ko\-tee\-rya.}
\fmtExampleEN{They are taking the goat to eat the grass, but it is refusing to go.}

\fmtLemma{eko\-mai\-ko\-ma}
\fmtCitation{(okw)e\-ko\-mai\-ko\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eko\-mai\-ko\-me\-re}
\fmtSubentry(a)
\fmtSense{to quarrel over trivial matters, to quarrel over nothing}
\fmtExampleRL{Wee\-na wee\-ko\-mai\-ko\-ma, eka\-laa\-mu yoo\-na n’e\-ku\-lwa\-ni\-sya?}
\fmtExampleEN{You quarrel over trivial matters, can a mere pencil cause you to quarrel?}
\fmtSubentry(b)
\fmtSense{to steal, to indulge in petty theft}
\fmtExampleRL{Oli ka\-nga di okwe\-ko\-mai\-ko\-ma?}
\fmtExampleEN{When will you stop indulging in petty theft?}

\fmtLemma{eko\-na}
\fmtCitation{(okw)e\-ko\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eko\-ne\-re}
\fmtSense{to stumble, to trip (over)}
\fmtExampleRL{Ye\-ko\-ne\-re n’a\-gwa.}
\fmtExampleEN{He stumbled and fell.}

\fmtLemma{eku\-bbe\-mbe\-ra}
\fmtCitation{(okw)e\-ku\-bbe\-mbe\-ra}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{eko\-mai\-ko\-me\-re}
\fmtSubentry(a)
\fmtSense{to lead, to come first}
\fmtExampleRL{Omu\-taa\-ni waa\-mwe ni\-ye ye\-ku\-bbe\-mbei\-re omu ki\-bii\-na.}
\fmtExampleEN{His son came first in his class.}
\fmtSubentry(b)
\fmtSense{to chair}
\fmtExampleRL{Ya\-ko\-me\-re Ma\-rti\-n okwe\-ku\-bbe\-mbe\-ra olu\-kii\-ko olwo.}
\fmtExampleEN{He appointed Martin to chair the meeting.}

\fmtLemma{eku\-bbi\-ra\-wo}
\fmtCitation{(okw)e\-ku\-bbi\-ra\-wo}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eku\-bbii\-re\-wo}
\fmtSense{to fall heavily}
\fmtExampleRL{Omu\-saa\-le gw’e\-ku\-bbii\-re\-wo.}
\fmtExampleEN{The tree has fallen heavily.}

\fmtLemma{eku\-bii\-ra}
\fmtCitation{(okw)e\-ku\-bii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eku\-bi\-rii\-re}
\fmtSubentry(a)
\fmtSense{to be misaligned, to lean to one side, (of a ship) to list}
\fmtExampleRL{Enzi\-ga ya gaa\-li yee\-ku\-bii\-rii\-re.}
\fmtExampleEN{The bicycle wheel is misaligned.}
\fmtSubentry(b)
\fmtSense{to be partial, to be biased, to be prejudiced, to be partisan}
\fmtExampleRL{Omu\-la\-mu\-li akwe\-ku\-bii\-ra.}
\fmtExampleEN{The judge is partial.}

\fmtLemma{eku\-du\-mba\-lya}
\fmtCitation{(okw)e\-ku\-du\-mba\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eku\-du\-mbai\-rye}
\fmtSense{to exalt oneself, to hold oneself in very high regard, to think highly of oneself, to be proud of oneself}
\fmtExampleRL{Lwa\-ki wee\-ku\-du\-mba\-lya mu\-no nga ndoo\-wo kyo\-li?}
\fmtExampleEN{Why are you exalting yourself when you are a nobody?}

\fmtLemma{eku\-kuu\-tii\-rya}
\fmtCitation{(okw)e\-ku\-kuu\-tii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eku\-kuu\-tii\-rye}
\fmtSense{to follow closely behind, to keep, to close to, to stick to, to press oneself against}
\fmtExampleRL{Ebbaa\-si ni ya\-bbai\-re eri ampi oku\-si\-mbu\-la omu\-ntu owo\-ndi ya\-yi\-zi\-re ni ye\-ku\-kuu\-tii\-rya\-mu.}
\fmtExampleEN{Just as the bus was about to leave, another person squeezed in.}

\fmtLemma{eku\-lii\-sii\-rya}
\fmtCitation{(okw)e\-ku\-lii\-sii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eku\-lii\-sii\-rye}
\fmtSense{to boast, to exalt oneself, to promote oneself, to claim to be older or of higher status than one is}
\fmtExampleRL{Na\-nka we ota\-ka mu\-no okwe\-ku\-lii\-sii\-rya.}
\fmtExampleEN{You really want to exalt yourself.}

\fmtLemma{eku\-lu\-kuu\-nya}
\fmtCitation{(okw)e\-ku\-lu\-kuu\-nya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eku\-lu\-kuu\-nyi\-re}
\fmtSense{to wallow, to grovel, to play in dirt}
\fmtExampleRL{Lwa\-ki okwe\-ku\-lu\-kuu\-nya omu ita\-ka\-li?}
\fmtExampleEN{Why are you playing in dirt?}

\fmtLemma{eku\-lu\-ngi}
\fmtCitation{\textnormal{(e.g.}\,mwe\-ku\-lu\-ngi, gwe\-ku\-lu\-ngi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{round, circular, oval, cylindrical, spherical}
\fmtExampleRL{Omu\-pii\-ra ogu\-sa\-mbi\-bwa gu\-ta gwe\-ku\-lu\-ngi.}
\fmtExampleEN{A ball which is played with is round.}

\fmtLemma{eku\-niai\-mu\-la}
\fmtCitation{(okw)e\-ku\-niai\-mu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eku\-niai\-mwi\-re}
\fmtSense{to fall with a thud, to fall heavily}
\fmtExampleRL{Omu\-ka\-li ni yaa\-li aku\-lu\-ga omu bbaa\-si ni yee\-ku\-niai\-mu\-la eki\-gwo.}
\fmtExampleEN{When the woman was getting off the bus, she fell with a thud.}

\fmtLemma{eku\-nu\-la}
\fmtCitation{(okw)e\-ku\-nu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eku\-nwi\-re}
\fmtSubentry(a)
\fmtSense{to bruise, to be susceptible to bruising}
\fmtExampleRL{Wee\-ku\-nwi\-re mu\-no olu\-su\-su.}
\fmtExampleEN{Your skin is very bruised.}
\fmtSubentry(b)

\fmtLemma{eku\-nuu\-ra}
\fmtCitation{(okw)e\-ku\-nuu\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eku\-nwi\-re}
\fmtSense{to be shy (to do something), to look coy}
\fmtExampleRL{Nee\-ku\-nwi\-re oku\-sa\-nga omu\-sai\-za eya\-ta\-ka\-nga oku\-nswe\-ra.}
\fmtExampleEN{I was shy about meeting the man who wanted to marry me.}

\fmtLemma{eku\-ru\-kuu\-tya}
\fmtCrossreference{\textnormal{see also} ekuu\-tya}
\fmtCitation{(okw)e\-ku\-ruu\-ku\-tya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eku\-ru\-kuu\-ti\-rye}
\fmtSense{to rub oneself}
\fmtExampleRL{Ombe\-re\-ge akwe\-ku\-ru\-kuu\-tya oku mwo\-mo.}
\fmtExampleEN{The pig is rubbing itself against the wall.}

\fmtLemma{eku\-ru\-ndya}
\fmtCitation{(okw)e\-ku\-ru\-ndya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eku\-ru\-ndi\-rye}
\fmtSense{to crawl, to creep, to go on all fours}
\fmtExampleRL{Omu\-yii\-gi eya\-bbai\-re akwe\-ku\-ru\-ndya oku ita\-ka\-li onfu\-lu\-gu\-ndu ya\-mu\-ne\-ne\-re.}
\fmtExampleEN{The hunter who was crawling was bitten by a puff adder.}

\fmtLemma{ekuu\-ma}
\fmtCitation{(okw)e\-kuu\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ekuu\-mi\-re}
\fmtSense{to protect oneself against, to beware of, to watch out for, to guard oneself against}
\fmtExampleRL{Mwe\-kuu\-me endwai\-re ya na\-mu\-nywa.}
\fmtExampleEN{Protect yourself against HIV/AIDS.}

\fmtLemma{ekuu\-tya}
\fmtCitation{(okw)e\-kuu\-tya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ekuu\-ti\-rye}
\fmtSense{to rub oneself}
\fmtExampleRL{Wee\-kuu\-tye oku mwo\-mo.}
\fmtExampleEN{Rub yourself against the wall.}

\fmtLemma{ekwa\-ta}
\fmtCitation{(okw)e\-kwa\-ta}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{ekwai\-te}
\fmtSubentry(a)
\fmtSense{to hold (onto), to adhere (to)}
\fmtExampleRL{Wee\-kwa\-te oku itaa\-gi eryo ota\-gwa.}
\fmtExampleEN{Hold onto that branch so that you don’t fall down.}
\fmtSubentry(b)
\fmtSense{to be compact}
\fmtExampleRL{Ei\-ta\-ka\-li lye\-kwai\-te.}
\fmtExampleEN{The soil is compact.}
\fmtSubentry(c)
\fmtSense{to take for oneself, to seize for oneself, to reserve (\eg a room)}
\fmtExampleRL{Naa\-ni yee\-kwai\-te eki\-sii\-ka ki\-ni?}
\fmtExampleEN{Who reserved this room?}

\fmtLemma{ekwa\-ta\-kwa\-ta}
\fmtCitation{(okw)e\-kwa\-ta\-kwa\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ekwai\-te\-kwai\-te}
\fmtSubentry(a)
\fmtSense{to look for money, to search oneself}
\fmtExampleRL{Wee\-kwa\-te\-kwa\-te, tu\-ju\-ne omwa\-na oni.}
\fmtExampleEN{Look for money and we will help this child.}
\fmtSubentry(b)
\fmtSense{to be shey, to be diffident, to be reluctant}

\fmtLemma{ekwa\-ti}
\fmtCitation{\textnormal{(e.g.}\,mwe\-kwa\-ti, lye\-kwa\-ti\textnormal{)}}
\fmtPOS{adj}
\fmtSense{dark}
\fmtExampleRL{Omu\-si\-ga\-zi azwai\-re saa\-ti ya bbu\-lu mwe\-kwa\-ti.}
\fmtExampleEN{A youth is wearing a dark blue shirt.}

\fmtLemma{ekwa\-ti\-kwa\-ti}
\fmtCitation{\textnormal{(e.g.}\,gwe\-ka\-ti\-kwa\-ti, kye\-kwa\-ti\-kwa\-ti\textnormal{)}}
\fmtPOS{adj}
\fmtSense{light}
\fmtExampleRL{Omwo\-jo azwai\-re saa\-ti ya kye\-nju omwe\-kwa\-ti\-kwa\-ti.}
\fmtExampleEN{The youth is wearing a light-yellow shirt.}

\fmtLemma{ekwe\-ka}
\fmtCrossreference{\textnormal{see also} ebbi\-sa}
\fmtCitation{(okw)e\-kwe\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ekwe\-ke\-re}
\fmtSense{to hide, to conceal oneself}
\fmtExampleRL{Omwa\-na yee\-kwe\-ke\-re mu nta\-bbu.}
\fmtExampleEN{The child has hidden under the bed.}

\fmtLemma{ekya\-li\-gi\-re\-mu}
\fmtPOS{conj}
\fmtSense{thus, so, as a result, consequently}
\fmtExampleRL{Ba\-lwai\-ne ekya\-li\-gi\-re\-mu ni bai\-ta\-nga\-na.}
\fmtExampleEN{They fought and as a result they killed each other.}

\fmtLemma{ekya\-zwi\-re\-mu}
\fmtCrossreference{\textnormal{see also} ekya\-li\-gi\-re\-mu}
\fmtPOS{conj}
\fmtSense{thus, so, as a result, consequently}
\fmtExampleRL{Ba\-lwai\-ne ekya\-zwi\-re\-mu ni bai\-ta\-nga\-na.}
\fmtExampleEN{They fought and as a result they killed each other.}

\fmtLemma{ekyo}
\fmtCrossreference{\textnormal{see} o\textnormal{\fmtOne{3}}}

\fmtLemma{ekyu\-sa}
\fmtCitation{(okw)e\-kyu\-sa (ei\-ndu\-la)}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ekyu\-si\-re}
\fmtSense{to change one's mind, to change sides, to turn oneself}
\fmtExampleRL{Ya\-li asaa\-re\-wo oku\-mu\-su\-mbi\-rwa bai\-tu ni yee\-kyu\-sa.}
\fmtExampleEN{She had decided to marry him, but she changed her mind.}

\fmtLemma{ema}%
\fmtOne{1}
\fmtCitation{(okw)e\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eme\-re}
\fmtSense{to conceive, to become pregnant, to become impregnated}
\fmtExampleRL{Onte wa\-nge aye\-me\-re.}
\fmtExampleEN{My cow has conceived.}

\fmtLemma{ema}%
\fmtOne{2}
\fmtCitation{(okw)e\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eme\-re}
\fmtSense{to step (on, into)}
\fmtExampleRL{Aye\-me\-re oku mwo\-mo n’a\-gwi\-ra\-gu\-rya.}
\fmtExampleEN{He has stepped on the wall and soiled it.}

\fmtLemma{emaa\-ra}
\fmtCitation{(okw)e\-maa\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ema\-lii\-re}
\fmtSense{to be dedicated, to be devoted, to be self-supporting}
\fmtExampleRL{Oyo yee\-maa\-ra mu\-no omu mi\-ru\-mu gya\-mwe.}
\fmtExampleEN{He is dedicated to his work.}

\fmtLemma{ema\-ga\-la\-lya}
\fmtCitation{(okw)e\-ma\-ga\-la\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ema\-ga\-lai\-rye}
\fmtSubentry(a)
\fmtSense{to be unsettled, to be in state of confusion, to be in despair, to be in state of desperation, to look hopeless}
\fmtExampleRL{Omu\-ka\-li akwe\-ma\-ga\-lya oku\-zwa ni ya\-fee\-rei\-rwe oi\-ba waa\-mwe.}
\fmtExampleEN{The woman has been at a loss ever since she lost her husband.}
\fmtSubentry(b)
\fmtSense{to act stupidly, to act silly, to act foolishly}
\fmtExampleRL{Lwa\-ki wee\-ma\-ga\-la\-lya nka ata\-li\-na ma\-ge\-zi?}
\fmtExampleEN{Why do you act like a fool?}

\fmtLemma{ema\-gya}
\fmtCitation{(okw)e\-ma\-gya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ema\-gi\-rye}
\fmtSense{to be unsettled, to be in state of confusion, to be in despair, to be in state of desperation, to look hopeless}
\fmtExampleRL{Ba\-mbi omu\-sai\-za ka\-ti yee\-ma\-gya\-mbe.}
\fmtExampleEN{Oh dear, the man is now in a state of despair.}

\fmtLemma{ema\-la}
\fmtCitation{(okw)e\-ma\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{emaa\-re}
\fmtSense{to be self-sufficient, to be self-supporting}
\fmtExampleRL{Nje nee\-ma\-la ndoo\-wo kye nju\-ra.}
\fmtExampleEN{I am self-sufficient, there is nothing I lack.}

\fmtLemma{ema\-lai\-ma\-la}
\fmtCitation{(okw)e\-ma\-lai\-ma\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{emaa\-rei\-maa\-re}
\fmtSense{to fake innocence}
\fmtExampleRL{Owa\-yi\-ni ya\-mu\-nywi\-re, ati ali awo akwe\-ma\-lai\-ma\-la.}
\fmtExampleEN{He drank the wine, but now he is faking innocence.}

\fmtLemma{ema\-la\-ma\-la}
\fmtCrossreference{\textnormal{see} ema\-lai\-ma\-la}

\fmtLemma{ema\-lii\-ra}
\fmtCitation{(okw)e\-ma\-lii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ema\-li\-rii\-re}
\fmtSense{to be independent, to be self-reliant, to be self-sufficient, to be self-supporting}
\fmtExampleRL{Le\-ke\-ra awo oku\-lo\-woo\-za nti wee\-ma\-lii\-ra omu bu\-li ki\-ntu.}
\fmtExampleEN{Stop thinking that you are independent in everything.}

\fmtLemma{ema\-li\-nga}
\fmtCitation{(okw)e\-ma\-li\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ema\-li\-ngi\-re}
\fmtSense{to waste time}
\fmtExampleRL{Mu\-le\-ke\-ra\-wo okwe\-ma\-li\-nga, mwa\-be mu\-syo\-me amai\-zi.}
\fmtExampleEN{Stop wasting time, go and fetch water.}

\fmtLemma{ema\-li\-nka}
\fmtCitation{(okw)e\-ma\-li\-nka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ema\-li\-nki\-re}
\fmtSense{to elevate oneself, to be conceited, to be arrogant, to overreach oneself, to bite off more than one can chew}
\fmtExampleRL{OMu\-sa ye\-ma\-li\-nka ta\-yi\-ki\-ri\-za nti wa\-li\-wo ayi\-nza oku\-mu\-wa\-ngu\-la.}
\fmtExampleEN{Musa is very conceited, he will never believe that he can be defeated.}

\fmtLemma{ema\-nsu\-li\-sya}
\fmtCitation{(okw)e\-ma\-nsu\-li\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ema\-nsu\-li\-si\-rye}
\fmtSense{to reject pretentiously}
\fmtExampleRL{ONa\-ka\-to omwo\-jo aku\-mu\-ta\-ka, bai\-tu aku\-mwe\-ma\-nsu\-li\-sya\-ku.}
\fmtExampleEN{Nakato loves the boy, but she is rejecting him pretentiously.}

\fmtLemma{emba}
\fmtCitation{(okw)e\-mba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{embe\-re}
\fmtSense{to sing}
\fmtExampleRL{Ya\-ye\-mbe\-re n’o\-mwa\-na waa\-mwe n’a\-wa\-ngu\-la empa\-ka.}
\fmtExampleEN{She sang with her child and won the competition.}

\fmtLemma{emee\-ra}
\fmtCitation{(okw)e\-mee\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eme\-rei\-re}
\fmtSubentry(a)
\fmtSense{to stand}
\fmtExampleRL{Iza oye\-mee\-re ani.}
\fmtExampleEN{Come and stand here.}
\fmtSubentry(b)
\fmtSense{to contest, to compete}
\fmtExampleRL{Aye\-mee\-rei\-re ku ki\-fo kya ise\-nte\-be.}
\fmtExampleEN{He is competing for the post of chairperson.}
\fmtSubentry(c)
\fmtSense{to get an erection}
\fmtExampleRL{Omu\-fee\-rwa ta\-so\-bo\-la kwe\-mee\-ra.}
\fmtExampleEN{An impotent person cannot get an erection.}

\fmtLemma{emee\-ra\-wo}
\fmtCitation{(okw)e\-mee\-ra\-wo}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{eme\-rei\-re\-wo}
\fmtSubentry(a)
\fmtSense{to be established, to be founded}
\fmtExampleRL{Obu\-ka\-ma bwa\-ye\-mee\-rei\-re\-wo ng'e\-bi\-ro bya kwe\-zi 8-10-2004.}
\fmtExampleEN{The Kingdom was established on the 8th of October 2004.}
\fmtSubentry(b)
\fmtSense{to support}
\fmtExampleRL{OKa\-to ya\-ye\-mee\-rei\-re\-wo mu\-no ku lwa\-nge omu lu\-mbe lwa taa\-ta.}
\fmtExampleEN{Kato supported me a lot during my father's funeral.}
\fmtSubentry(c)
\fmtSense{to contest}
\fmtExampleRL{Aye\-mee\-rei\-re\-wo oku ki\-fo kya ise\-nte\-be.}
\fmtExampleEN{He is competing for the post of chairperson.}

\fmtLemma{emee\-rya}
\fmtCitation{(okw)e\-mee\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eme\-rei\-rye}
\fmtSense{to stop}
\fmtExampleRL{Abee\-gi ba\-ge\-ze\-re\-ku okwe\-mee\-rya ebbaa\-si bai\-tu n’e\-tee\-mee\-ra.}
\fmtExampleEN{The students tried to stop the bus, but it didn't stop.}

\fmtLemma{eme\-mu\-la}
\fmtCitation{(okw)e\-me\-mu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eme\-mwi\-re}
\fmtSense{to shine brightly, to burn, to blaze}
\fmtExampleRL{Ei\-sa\-na li\-kwe\-me\-mu\-la mu\-no.}
\fmtExampleEN{The sun is shining brightly.}

\fmtLemma{emi\-gi\-ra\-mu}
\fmtCitation{(okw)e\-mi\-gi\-ra\-mu}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{emi\-gii\-re\-mu}
\fmtSense{to participate, to take part}
\fmtExampleRL{Twe\-mi\-gi\-re\-mu omu ku\-wa\-ndii\-ka eki\-dee\-ro.}
\fmtExampleEN{Let's participate in writing the dictionary.}

\fmtLemma{Emma\-nue\-l}
\fmtCitation{Emma\-nue\-l}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{Biblical name which means God is with us}
\fmtExampleRL{Eri\-mwe oku ma\-ba\-ra ga Ye\-su ni\-ye Emma\-nue\-l.}
\fmtExampleEN{One of Jesus' names is Emmanuel.}

\fmtLemma{emo\-da}
\fmtCitation{(okw)e\-mo\-da}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{emo\-de\-re}
\fmtSense{to make funny faces, to play the clown, to be humorous, to be frivolous, to be comical}
\fmtExampleRL{Abaa\-la abo ba\-ta\-ka mu\-no okwe\-mo\-da.}
\fmtExampleEN{Those girls are fond of making funny faces.}

\fmtLemma{emo\-do\-nko\-la}
\fmtCitation{(okw)e\-mo\-do\-nko\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{emo\-do\-nko\-re}
\fmtSense{to act funny, to play around, to kid around}
\fmtExampleRL{Wee\-na okwe\-mo\-do\-nko\-la bu\-ndi.}
\fmtExampleEN{You are acting funny.}

\fmtLemma{emo\-lo\-nga}
\fmtCitation{(okw)e\-mo\-lo\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{emo\-lo\-nge\-re}
\fmtSense{to isolate oneself}
\fmtExampleRL{Le\-ka kwe\-mo\-lo\-nga okwi\-za ku\-lwa\-la omu\-ti\-ma.}
\fmtExampleEN{Stop isolating yourself, you will suffer from stress.}

\fmtLemma{emu\-la}
\fmtCitation{(okw)e\-mu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{emwi\-re}
\fmtSubentry(a)
\fmtSense{to swear, to curse, to use bad language, to utter profanities, to be vulgur}
\fmtExampleRL{Ata\-ka mu\-no okwe\-mu\-la omu ba\-ntu.}
\fmtExampleEN{He likes using profanities in public a lot.}
\fmtSubentry(b)
\fmtFixedexpression{emu\-la (o)Ka\-nca}
\fmtSense{to blaspheme}
\fmtExampleRL{Oka\-bo\-na ya\-ko\-be\-re oYe\-su nti aku\-we\-mu\-la oKa\-nca.}
\fmtExampleEN{The priest told Jesus that he was profaning God.}

\fmtLemma{emu\-lu\-gu\-nya}
\fmtCitation{(okw)e\-mu\-lu\-gu\-nya (go\-ngoi\-zo\-ka)}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{emu\-lu\-gu\-nyi\-re}
\fmtSense{to complain, to grumble, to protest}
\fmtExampleRL{Mu\-kwe\-mu\-lu\-gu\-nyi\-ra bwe\-ree\-re nywe\-na mwa\-lii\-re ku\-sai.}
\fmtExampleEN{You are complaining for no reason, you all ate enough.}

\fmtLemma{emu\-lu\-la}
\fmtCitation{(okw)e\-mu\-mu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{emu\-lu\-lwi\-re}
\fmtSense{to slip away, to creep away, to walk away stealthily, to sneak out}
\fmtExampleRL{Omu\-ge\-ni omu\-ku\-lu ye\-mu\-lu\-lwi\-re\-mbe.}
\fmtExampleEN{The guest of honour just snuck away.}

\fmtLemma{emu\-nyu\-ngu\-la}
\fmtCitation{(okw)e\-mu\-nyu\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{emu\-nyu\-ngwi\-re}
\fmtSense{(of birds and animals) to play in dust}
\fmtExampleRL{Onko\-wu akwe\-mu\-nyu\-ngu\-la omu cuu\-cu.}
\fmtExampleEN{The guinea fowl is playing in dust.}

\fmtLemma{emya}
\fmtCitation{(okw)e\-mya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eme\-rye}
\fmtSense{(of bull) to mount}
\fmtExampleRL{Onu\-mi oyo aye\-mya ente iku\-mi.}
\fmtExampleEN{That bull mounts ten different cows.}

\fmtLemma{ena}%
\fmtOne{1}
\fmtCitation{(okw)e\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ene\-re}
\fmtSubentry(a)
\fmtSense{to be erect}
\fmtExampleRL{Aka\-sii\-sa ka mu\-sai\-za kaa\-ye\-ne\-re.}
\fmtExampleEN{The man's penis is erect.}
\fmtSubentry(b)
\fmtSense{to have an erection}
\fmtExampleRL{Omu\-sai\-za oyo aku\-bo\-ne\-ka aye\-ne\-re.}
\fmtExampleEN{It seems that man has an erection.}

\fmtLemma{(-)ena}%
\fmtOne{2}
\fmtPOS{sf}
\fmtSense{also, too, as well}
\fmtExampleRL{Wee\-na oli mu\-bbi.}
\fmtExampleEN{You are also bad.}

\fmtLemma{enaa\-ni\-ka}
\fmtCitation{(okw)e\-naa\-ni\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{enaa\-ni\-ki\-re}
\fmtSense{to wear (shoes)}
\fmtExampleRL{Mwe\-naa\-ni\-ki\-re enkai\-to ezi\-sai.}
\fmtExampleEN{You are wearing very nice shoes.}

\fmtLemma{ena\-ga}
\fmtCitation{(okw)e\-na\-ga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ena\-gi\-re}
\fmtSubentry(a)
\fmtSense{to wobble, to stagger}
\fmtExampleRL{Omu\-kai\-re yee\-na\-ga olwa\-ku\-bba amaa\-ni ga\-mu\-wei\-re\-mu.}
\fmtExampleEN{The old woman wobbles because she no longer has the strength.}
\fmtSubentry(b)
\fmtSense{to isolate oneself}
\fmtExampleRL{Le\-ka kwe\-na\-ga\-nga n’o\-zwa oku ba\-ntu.}
\fmtExampleEN{Stop isolating yourself from other people.}
\fmtSubentry(c)
\fmtFixedexpression{ena\-ga lya lu\-ma\-nyo}
\fmtSense{to pretend to be what you are not}
\fmtExampleRL{Oyo akwe\-na\-ga\-mbe lya lu\-ma\-nyo.}
\fmtExampleEN{He is just pretending to be what he is not.}

\fmtLemma{ena\-gi\-za}
\fmtCitation{(okw)e\-na\-gi\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ena\-gii\-ze}
\fmtSense{to drape, wrap}
\fmtExampleRL{Omu\-ka\-li yee\-na\-gii\-ze olu\-go\-ye oku ma\-be\-ga.}
\fmtExampleEN{The woman has draped a cloth on her back.}

\fmtLemma{ena\-mi\-ra}
\fmtCitation{(okw)e\-na\-mi\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ena\-mii\-re}
\fmtSubentry(a)
\fmtSense{to bow down, to bow in respect, to stoop, to bend down, to bend over}
\fmtExampleRL{Yee\-na\-mii\-re omu mai\-so ga Ka\-nca.}
\fmtExampleEN{He bowed before the Lord.}
\fmtSubentry(b)
\fmtSense{to be sad, to feel unhappy, to feel depressed, to be aggrieved}
\fmtExampleRL{Mwa\-na wa\-nge lwa\-ki wee\-na\-mii\-re?}
\fmtExampleEN{Why are you sad, my child?}
\fmtSubentry(c)
\fmtSense{to pass time}
\fmtExampleRL{Aya\-bi\-re ku mu\-caa\-mu kwe\-na\-mi\-ra.}
\fmtExampleEN{He has gone to her neighbours' home to pass time.}

\fmtLemma{ena\-ti\-ra}
\fmtCitation{(okw)e\-na\-ti\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ena\-tii\-re}
\fmtSense{to be sad, to be aggrieved, to be anguished}
\fmtExampleRL{Nee\-na\-tii\-re nga mpu\-lii\-re oku\-fa kwa Pro\-f. Wa\-lu\-si\-mbi.}
\fmtExampleEN{I felt sad when I heard the news of Prof. Walusimbi's death.}

\fmtLemma{ena\-ti\-ri}
\fmtCitation{\textnormal{(e.g.}\,mwe\-ta\-ni\-ri, bee\-ta\-ni\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSense{miserable, unhappy, depressed}
\fmtExampleRL{Lwa\-ki omu\-ka\-li oyo mwe\-na\-ti\-ri?}
\fmtExampleEN{Why is that woman miserable?}

\fmtLemma{enda}%
\fmtOne{1}
\fmtCitation{(okw)e\-nda}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ende\-re}
\fmtSubentry(a)
\fmtSense{to break}
\fmtExampleRL{Ye\-nda\-mu eki\-saa\-le ekyo ebi\-tu\-ndu bi\-bi\-ri.}
\fmtExampleEN{Break that stick into two.}
\fmtSubentry(b)
\fmtSense{to close (a pocket knife, umbrella), to fold}

\fmtLemma{enda}%
\fmtOne{2}
\fmtCitation{\textnormal{(e.g.}\,mwe\-ndi, be\-ndi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{generous}
\fmtExampleRL{Omu\-ntu omwe\-nda aga\-ba mu\-no.}
\fmtExampleEN{A generous person shares a lot.}

\fmtLemma{enda\-wo}
\fmtCrossreference{\textnormal{see} yenda\-wo}
\fmtCitation{(okw)e\-nda\-wo}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ende\-re\-wo}
\fmtSense{to take back (e.g. words), to retract, to refute}
\fmtExampleRL{Ye\-nda\-wo by’o\-ba\-zi\-re.}
\fmtExampleEN{Take back what you have said.}

\fmtLemma{ende}
\fmtDialect{Buyende}
\fmtCitation{\textnormal{(e.g.}\,kye\-nde, bye\-nde\textnormal{)}}
\fmtPOS{adj}
\fmtSense{favourite}
\fmtExampleRL{Omu\-ka\-li omwe\-nde tai\-rya mu\-sa\-ngo.}
\fmtExampleEN{A favourite wife does no wrong.}

\fmtLemma{endee\-rya}
\fmtCitation{(okw)e\-ndee\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ende\-rei\-rye}
\fmtSubentry(a)
\fmtSense{to provoke, to instigate}
\fmtExampleRL{OYo\-kaa\-na aku\-ne\-ndee\-rya tu\-lwa\-ne.}
\fmtExampleEN{John is provoking me to fight.}
\fmtSubentry(b)
\fmtSense{to incite}
\fmtExampleRL{Mu\-ta\-ye\-ndee\-rya mbwe\-ne oyo akwi\-za ku\-ba\-ne\-na.}
\fmtExampleEN{Don’t incite that dog, it will bite them.}

\fmtLemma{ende\-ka}
\fmtCitation{(okw)e\-nde\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ende\-ke\-re}
\fmtSense{to break, to fracture, to bend at the joint}
\fmtExampleRL{Ei\-gu\-mba ni lye\-nde\-ka, omu\-yu\-ngi ali\-yu\-nga.}
\fmtExampleEN{When a bone fractures, an orthopaedic surgeon puts it together.}

\fmtLemma{ende\-ki}
\fmtCitation{\textnormal{(e.g.}\,mwe\-ndi\-ki, lwe\-nde\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{broken}
\fmtExampleRL{Omu\-ko\-nda gwa ki\-ko\-po gwe\-nde\-ki.}
\fmtExampleEN{The handle of the cup is broken.}

\fmtLemma{ende\-ree\-rya}
\fmtCrossreference{\textnormal{see} endee\-rya}

\fmtLemma{endya}
\fmtCitation{(okw)e\-ndya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ende\-rye}
\fmtSense{to want, to desire, to wish}
\fmtExampleRL{Nkwe\-ndya oi\-ze\-ku ewa\-nge onkya\-re\-ku.}
\fmtExampleEN{I want you to come to my place and pay me a visit.}

\fmtLemma{ene\-nya}
\fmtCitation{(okw)e\-ne\-nya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ene\-nye\-re}
\fmtSense{to repent, to blame oneself}
\fmtExampleRL{Mwa\-be egi\-ri oYe\-su mwe\-ne\-nye.}
\fmtExampleEN{Turn to Jesus and repent.}

\fmtLemma{enga}
\fmtCitation{(okw)e\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{enge\-re}
\fmtSense{to ripen, to be ripe}
\fmtExampleRL{Emi\-ye\-mbe gi\-ta\-ndi\-ki\-re okwe\-nga.}
\fmtExampleEN{The mangoes have started to ripen.}

\fmtLemma{enga\-na}
\fmtCrossreference{\textnormal{see also} inga\-na}

\fmtLemma{enge\-ra}
\fmtCitation{(okw)e\-nge\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{engei\-re}
\fmtSense{to ripen, to become ripe}
\fmtExampleRL{Omu\-ye\-mbe gwe\-ngei\-re.}
\fmtExampleEN{The mango is ripe.}

\fmtLemma{enge\-ree\-ra}
\fmtCitation{(okw)e\-nge\-ree\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{enge\-rei\-re}
\fmtSubentry(a)
\fmtSense{to be yellowish, to become yellowish}
\fmtExampleRL{Omu\-cu\-ngwai gwe\-nge\-rei\-re.}
\fmtExampleEN{The orange is yellowish.}
\fmtSubentry(b)
\fmtSense{(of metal) to be red hot, (of a boil or an abscess) to be inflamed}
\fmtExampleRL{Ekyo\-ma ati kye\-nge\-rei\-re.}
\fmtExampleEN{The metal has now become red hot.}

\fmtLemma{enge\-ri}
\fmtCitation{\textnormal{(e.g.}\,mwe\-nge\-ri, lye\-nge\-ri, ze\-nge\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSense{ripe}
\fmtExampleRL{Ofe\-ne gw'o\-lee\-te\-re ti mwe\-nge\-ri.}
\fmtExampleEN{The jackfruit which you have brought is not ripe.}

\fmtLemma{engi}
\fmtCitation{\textnormal{(e.g.}\,mwe\-ngi, lye\-ngi, kye\-ngi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{ripe}
\fmtExampleRL{Ofe\-ne gw'o\-lee\-te\-re mwe\-ngi.}
\fmtExampleEN{The jackfruit which you have brought is ripe.}

\fmtLemma{eni}%
\fmtOne{1}
\fmtPOS{adv}
\fmtSubentry(a)
\fmtSense{this way, here (noun class 23 proximal demonstrative)}
\fmtExampleRL{Ira eni oi\-ka\-le.}
\fmtExampleEN{Come this way and sit down.}
\fmtSubentry(b)
\fmtFixedexpression{eni n'edi}
\fmtSense{here and there}
\fmtExampleRL{Ya\-yo\-mbo\-ke\-re enyu\-mba gi\-koo\-to, eru\-ga eni n’e\-di.}
\fmtExampleEN{He built a big house, it extends from here to there.}

\fmtLemma{eni}%
\fmtOne{2}
\fmtCrossreference{\textnormal{see} ni{\textnormal{\fmtOne{2}}}}

\fmtLemma{enka\-la\-ngu\-la}
\fmtCrossreference{\textnormal{see} ekala\-ngu\-la}

\fmtLemma{enka\-na}
\fmtCrossreference{\textnormal{see} inga\-na}

\fmtLemma{enka\-nia\-nka\-nia}
\fmtCitation{(okw)e\-nka\-nia\-nka\-nia}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{enka\-nia\-ka\-nie}
\fmtSense{to measure equally, to divide equally}
\fmtExampleRL{Mwe\-nka\-nie\-nka\-nie ebi\-ntu ebyo le\-ke mu\-bi\-we aba\-kai\-re.}
\fmtExampleEN{Divide those items equally and give it to the elderly.}

\fmtLemma{eno\-ba}
\fmtCitation{(okw)e\-no\-ba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eno\-be\-re}
\fmtSense{to be disillusioned, to be disappointed}
\fmtExampleRL{Mwi\-ra\-nge omu\-ka\-li yee\-no\-be\-re ni yee\-zi\-nda.}
\fmtExampleEN{My dear, the woman got disillusioned and committed suicide.}

\fmtLemma{enya\-nya\-la}
\fmtCitation{(okw)e\-nya\-nya\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{enya\-nyai\-re}
\fmtSense{to treat with disgust, to treat with aversion, to grimace, to make a facial expression of disgust}
\fmtExampleRL{Le\-ka kwe\-nya\-nya\-la mu\-kai\-re.}
\fmtExampleEN{Don’t treat the old woman with disgust.}

\fmtLemma{enye\-ne\-re}
\fmtCitation{\textnormal{(e.g.}\,bee\-nye\-ne\-re, mwe\-nye\-ne\-re, ke\-nye\-ne\-re\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{actual, real}
\fmtExampleRL{Ni\-ye mu\-ka\-ma wa pee\-ni mwe\-nye\-ne\-re.}
\fmtExampleEN{He is the actual owner of this pen.}
\fmtSubentry(b)
\fmtSense{typical, classic, normal, real, true, actual}
\fmtExampleRL{Oyo Mu\-ruu\-li mwe\-nye\-ne\-re.}
\fmtExampleEN{He is a typical Muruuli.}
\fmtSubentry(c)
\fmtSense{by oneself, alone}
\fmtExampleRL{Yei\-zii\-re mwe\-nye\-ne\-re.}
\fmtExampleEN{He came by himself.}

\fmtLemma{enyi\-ni}
\fmtCitation{\textnormal{(e.g.}\,mwe\-nyi\-ni, bee\-nyi\-ni, lye\-nyi\-ni\textnormal{)}}
\fmtPOS{adj}
\fmtSense{real, true, actual}
\fmtExampleRL{Si\-nga tu\-bbai\-re n’o\-mu\-ntu omu\-wee\-si omwe\-nyi\-ni.}
\fmtExampleEN{I wish we had a real blacksmith.}

\fmtLemma{enzee\-ra}
\fmtCitation{(okw)e\-nzee\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{enzei\-re}
\fmtSense{to wander (about), to roam}
\fmtExampleRL{Omwa\-na oyo akwe\-nzee\-ra omu ma\-nyu\-mba ga ba\-ntu.}
\fmtExampleEN{That child wanders into people’s homes.}

\fmtLemma{enzi}
\fmtCitation{\textnormal{(e.g.}\,mwe\-nzi, be\-nzi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{promiscuous}
\fmtExampleRL{Omu\-sai\-za omwe\-nzi afa ku\-bbi.}
\fmtExampleEN{A promiscuous man faces a bad death.}

\fmtLemma{eŋaa\-la}
\fmtCitation{(okw)e\-ŋaa\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eŋai\-re}
\fmtSense{to sit with legs wide apart}
\fmtExampleRL{Omu\-ka\-li oyo yee\-ŋaa\-la mu\-no.}
\fmtExampleEN{That woman often sits a lot with her legs apart.}

\fmtLemma{eŋa\-ŋa}
\fmtCitation{(okw)e\-ŋa\-ŋa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eŋa\-ŋi\-re}
\fmtSense{to become furious, to growl}
\fmtExampleRL{Oyo n’a\-go\-nge\-ra yee\-ŋa\-ŋa.}
\fmtExampleEN{When he gets annoyed, he becomes furious.}

\fmtLemma{epaa\-kuu\-la}
\fmtCitation{(okw)e\-paa\-kuu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{epaa\-kwi\-re}
\fmtSense{to be arrogant}
\fmtExampleRL{Ba\-bbai\-re ba\-kwe\-paa\-kuu\-la nti ti\-baa\-ya\-be, bai\-re ni baa\-ba.}
\fmtExampleEN{They were arrogantly opposed to leaving, but they eventually left.}

\fmtLemma{epaa\-ruu\-ra}
\fmtCrossreference{\textnormal{see} epaatuu\-ra}

\fmtLemma{epaa\-tuu\-ra}
\fmtCitation{(okw)e\-paa\-tuu\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{epaa\-twi\-re}
\fmtSense{to fall down heavily}
\fmtExampleRL{Omu\-sai\-za azwi\-re oku mo\-to\-ka ni yee\-paa\-tuu\-ra oku bbu\-loo\-ka n’a\-fa.}
\fmtExampleEN{A man got out of the car, fell heavily on a brick, and died.}

\fmtLemma{epa\-ku\-la}
\fmtCitation{(okw)e\-pa\-ku\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{epa\-kwi\-re}
\fmtSense{to have a tendency to pilfer, to steal (things of little value), to be light-fingered}
\fmtExampleRL{Ge\-ndee\-ra omu\-sai\-za yee\-pa\-ku\-la\-pa\-ku\-la.}
\fmtExampleEN{Be careful, that man engages in pilfering.}

\fmtLemma{epa\-mpaa\-ta}
\fmtCitation{(okw)epa\-mpaa\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pa\-mpaa\-ti\-re}
\fmtSense{to search oneself, to check oneself}
\fmtExampleRL{Empii\-ya zi\-mu\-buu\-re\-ku akwe\-pa\-mpaa\-ta.}
\fmtExampleEN{He lost the money and he is searching himself.}

\fmtLemma{epa\-mpa\-li\-ka}
\fmtCrossreference{\textnormal{see also} epaakuu\-la}
\fmtCitation{(okw)e\-pa\-mpa\-li\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{epa\-mpa\-li\-ki\-re}
\fmtSense{to be ostentatious, to pretend to be much better, to act beyond one's capacity, to overreach oneself, to bite off more than one can chew, to be presumptuous, to show off}
\fmtExampleRL{Lwa\-ki wee\-pa\-mpa\-li\-ki\-re?}
\fmtExampleEN{Why are you ostentatious?}

\fmtLemma{epa\-nka}
\fmtCitation{(okw)e\-pa\-nka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{epa\-nki\-re}
\fmtSense{to be ostentatious, to pretend to be much better, to flatter oneself, to boast, to become self-important or egotistical, to put on airs}
\fmtExampleRL{Okwe\-pa\-nka gy’o\-ku\-bba nii\-we pu\-re\-zi\-de\-nti.}
\fmtExampleEN{You are boasting as if you are the president.}

\fmtLemma{epa\-nki}
\fmtCitation{\textnormal{(e.g.}\,mwe\-pa\-nki, bee\-pa\-nki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{arrogant, boastful, pompous, proud}
\fmtExampleRL{Omu\-ntu omwe\-pa\-nki aya\-ka\-lya\-mu aba\-ntu amai\-so.}
\fmtExampleEN{An arrogant person looks down upon other people.}

\fmtLemma{epa\-nkuu\-la}
\fmtCitation{(okw)e\-pa\-nkuu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{epa\-nkwi\-re}
\fmtSense{to boast, to brag, to swagger}
\fmtExampleRL{Yee\-pa\-nkuu\-la nti ti\-yaa\-lye bi\-nyee\-bwa, ti\-ya\-maa\-re n’a\-bi\-rya!}
\fmtExampleEN{He was boasting that he does not eat groundnuts, but he eventually ate them!}

\fmtLemma{epe\-na}
\fmtCitation{(okw)e\-pe\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{epe\-ne\-re}
\fmtSense{to play the ‘dodgy game’}
\fmtExampleRL{Abaa\-la baa\-bi\-re kwe\-pe\-na.}
\fmtExampleEN{The girls have gone to play the dodgy game.}

\fmtLemma{epi\-fa\-ni\-ya}
\fmtCitation{epi\-fa\-ni\-ya}
\fmtPOS{n}%
\fmtNC{9}
\fmtSource{from En.}
\fmtSense{epiphany (the manifestation of Christ to the Gentiles as represented by the Magi)}
\fmtExampleRL{OYe\-su oku\-bo\-ne\-ke\-ra edi aba\-ntu kya\-li epi\-fa\-ni\-ya.}
\fmtExampleEN{It was an epiphany for Jesus to appear to the people.}

\fmtLemma{epii\-ka}
\fmtCitation{(okw)e\-pii\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{epii\-ki\-re}
\fmtSense{to eat too much, to overeat}
\fmtExampleRL{Omwa\-na yee\-pii\-ki\-re eme\-re eki\-da ni ki\-mu\-lu\-ma.}
\fmtExampleEN{The child ate too much food and had an upset stomach.}

\fmtLemma{epi\-ka}
\fmtCitation{(okw)e\-pi\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{epi\-ki\-re}
\fmtSense{to bloat}
\fmtExampleRL{Enda nku\-wuu\-ra eta\-ndi\-ki\-re okwe\-pi\-ka.}
\fmtExampleEN{I feel my stomach has started to bloat.}

\fmtLemma{epu\-ku\-la}
\fmtCitation{(okw)e\-pu\-ku\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{epu\-kwi\-re}
\fmtSense{to rumble, to boom, to thunder}
\fmtExampleRL{Bai\-tu ki\-ki ekyo eki\-kwe\-pu\-ku\-la?}
\fmtExampleEN{What is rumbling like that?}

\fmtLemma{(-)er}
\fmtCrossreference{\textnormal{see} (-)ir}

\fmtLemma{era}%
\fmtOne{1}
\fmtCitation{(okw)e\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eere}
\fmtSubentry(a)
\fmtSense{to be clean}
\fmtExampleRL{Olu\-go\-ye lwa Za\-ri\-na lu\-kwe\-ra.}
\fmtExampleEN{Zarina's cloth is clean.}
\fmtSubentry(b)
\fmtSense{to be holy}
\fmtExampleRL{Omu\-sai\-za akwe\-ra afii\-re.}
\fmtExampleEN{A holy man has died.}

\fmtLemma{era}%
\fmtOne{2}
\fmtCitation{(okw)e\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eere}
\fmtSense{to yield}
\fmtExampleRL{Emi\-ye\-mbe gya\-yee\-re era ati tu\-kwa\-ba ku\-tu\-nda\-ku nzai.}
\fmtExampleEN{The mango tree yielded enough mangoes, we are now planning to sell some abroad.}

\fmtLemma{era}%
\fmtOne{3}
\fmtPOS{conj}
\fmtSense{and, also}
\fmtExampleRL{Omu\-ye\-mbe gwa\-tai\-re\-ku emi\-ye\-mbi gi\-ngi era tu\-ku\-tee\-ka\-tee\-ka ku\-gi\-tu\-nda egi\-mwei enzai.}
\fmtExampleEN{The mango trees yielded enough mangoes, we are now planning to sell some abroad.}

\fmtLemma{eraa\-yii\-rya}
\fmtCitation{(okw)e\-raa\-yii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eraa\-yii\-rye}
\fmtSense{to stand on one's toes}
\fmtExampleRL{Okwe\-raa\-yii\-rya ku\-bo\-na ki?}
\fmtExampleEN{You are standing on your toes, what do you want to see?}

\fmtLemma{era\-bi\-ra}
\fmtCitation{(okw)e\-ra\-bi\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{era\-bii\-re}
\fmtSense{to forget}
\fmtExampleRL{Otee\-ra\-bi\-ra kwi\-za.}
\fmtExampleEN{Don’t forget to come.}

\fmtLemma{era\-li\-ka}
\fmtCitation{(okw)e\-ra\-li\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{era\-li\-kii\-re}
\fmtSubentry(a)
\fmtSense{to lie on the back}
\fmtExampleRL{Oku\-bba\-sya ni wee\-ra\-li\-ki\-re ki\-bbi.}
\fmtExampleEN{It is bad to sleep lying on the back.}
\fmtSubentry(b)
\fmtSense{to toss one's head back (to be able the empty a drink from a cup or bottle)}
\fmtExampleRL{Omwa\-na anywi\-re ama\-tai ne yee\-ra\-li\-ka.}
\fmtExampleEN{The child drank milk and tossed his head back.}

\fmtLemma{era\-li\-kii\-ra}
\fmtCitation{(okw)e\-ra\-li\-kii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{era\-li\-kii\-ri\-re}
\fmtSense{to worry, to be worried, to be concerned, to be anxious}
\fmtExampleRL{Lwa\-ki mwe\-ra\-li\-kii\-ra mu\-no?}
\fmtExampleEN{Why are you so worried?}

\fmtLemma{era\-li\-kii\-ri}
\fmtCitation{\textnormal{(e.g.}\,mwe\-ra\-li\-kii\-ri, bee\-ra\-li\-kii\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSense{worried, troubled, bothered, concerned}
\fmtExampleRL{Aba\-ka\-li bee\-ra\-li\-kii\-ri olwa ma\-ba\-nza.}
\fmtExampleEN{The women are worried because of the debts.}

\fmtLemma{era\-nga}
\fmtCitation{(okw)e\-ra\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{era\-nge\-re}
\fmtSense{to introduce oneself to the clan for marriage}
\fmtExampleRL{Baa\-bi\-re kwe\-ra\-nga bu\-koi.}
\fmtExampleEN{They have gone to the in-laws for a marriage introduction.}

\fmtLemma{era\-te}
\fmtPOS{adv}
\fmtSense{again, still}
\fmtExampleRL{Era\-te oi\-zi\-re?}
\fmtExampleEN{You have come again?}

\fmtLemma{era\-yii\-ra}
\fmtCitation{(okw)e\-ra\-yii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{era\-yi\-rii\-re}
\fmtSense{to swear, to take an oath}
\fmtExampleRL{Nee\-ra\-yi\-rii\-re ti\-nkyai\-ra ani.}
\fmtExampleEN{I swear to never come back here.}

\fmtLemma{(-)e\-re}
\fmtCrossreference{\textnormal{see} -i\-re}
\fmtPOS{sf}
\fmtSense{perfective aspect suffix on verbs}

\fmtLemma{eree\-ra}
\fmtCitation{(okw)e\-ree\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ere\-rei\-re}
\fmtSense{to be next to, to be adjacent to}
\fmtExampleRL{Eki\-ba\-nja kya\-nge kye\-re\-rei\-re iso\-me\-ro.}
\fmtExampleEN{My land is next to the school.}

\fmtLemma{eree\-re}
\fmtCitation{\textnormal{(e.g.}\,gye\-ree\-re, ze\-ree\-re , lwe\-ree\-re\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{empty, bare}
\fmtExampleRL{Enyu\-mba gye\-ree\-re.}
\fmtExampleEN{The house is empty.}
\fmtSubentry(b)
\fmtSense{empty-handed}
\fmtExampleRL{Ata\-mbwi\-re mi\-ko\-no zee\-ree\-re.}
\fmtExampleEN{He went without anything in his hands.}
\fmtSubentry(c)
\fmtSense{only}
\fmtExampleRL{Lee\-ta mi\-gaa\-ti gye\-ree\-re.}
\fmtExampleEN{Bring only bread.}

\fmtLemma{eree\-tee\-rya}
\fmtCitation{(okw)e\-ree\-tee\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eree\-te\-rei\-rye}
\fmtSense{to associate oneself with (especially an unpleasant person)}
\fmtExampleRL{Lwa\-ki okwe\-ree\-tee\-rya omwi\-bbi oyo?}
\fmtExampleEN{Why do you associate yourself with the thief?}

\fmtLemma{ere\-kee\-rya}
\fmtCitation{(okw)e\-re\-kee\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ere\-ke\-rei\-rye}
\fmtSense{to forego, to stop, to give up, to abstain from, to postpone, to put off}
\fmtExampleRL{Mwe\-re\-kee\-rye obu\-sii\-ya\-ni ama\-ka gaa\-nywe ga\-te\-ree\-re.}
\fmtExampleEN{Stop committing adultery so that your families will be stable.}

\fmtLemma{ere\-ma}
\fmtCitation{(okw)e\-re\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ere\-me\-re}
\fmtSense{to become independent, to gain independence}
\fmtExampleRL{Uga\-nda yee\-re\-me\-re mu mwa\-ka gwa 1962.}
\fmtExampleEN{Uganda became independent in 1962.}

\fmtLemma{ere\-raa\-na}
\fmtCitation{(okw)e\-re\-raa\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ere\-rai\-ne}
\fmtSense{to be near, to be close}
\fmtExampleRL{Gu\-la ei\-ta\-ka\-li okwe\-re\-raa\-na oSa\-mwi\-ri.}
\fmtExampleEN{Buy the land near Swamiri's.}

\fmtLemma{eri\-bya}
\fmtCitation{(okw)e\-ri\-bya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eri\-bi\-rye}
\fmtSense{to grip on, to cling on tightly}
\fmtExampleRL{Omwa\-na yee\-ri\-bi\-rye oku maa\-mwe.}
\fmtExampleEN{The child clung tightly to her mother.}

\fmtLemma{eri\-nda}
\fmtCitation{(okw)e\-ri\-nda}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eri\-ndi\-re}
\fmtSense{to be watchful, to be on the alert, to be cautious, to watch out}
\fmtExampleRL{Mwe\-ri\-nde oku kya\-lo ku\-li\-ku abai\-ti.}
\fmtExampleEN{Be on the alert, there are murderers in the village.}

\fmtLemma{eri\-ndu\-ka}
\fmtCitation{(okw)e\-ri\-ndu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eri\-ndu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to be scared, to be frightened}
\fmtExampleRL{Omwo\-jo yee\-ri\-ndu\-ki\-re onjo\-ka.}
\fmtExampleEN{The boy is frightened by the snake.}
\fmtSubentry(b)
\fmtSense{to be surprised by, to be startled by}
\fmtExampleRL{Omu\-ka\-li ya\-ku\-bbi\-re\-wo eki\-ga\-mbo n'e\-twe\-ri\-ndu\-ka.}
\fmtExampleEN{A woman used an abusive word and we were surprised.}

\fmtLemma{eri\-nga\-ti\-ra}
\fmtCitation{(okw)e\-ri\-nga\-ti\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eri\-nga\-ti\-re}
\fmtSense{to be sad, to unhappy, to gloomy, to be overwhelmed with grief}
\fmtExampleRL{Lwa\-ki okwe\-ri\-nga\-ti\-ra?}
\fmtExampleEN{Why are you sad?}

\fmtLemma{ero\-ko\-mpo\-la}
\fmtCitation{(okw)e\-ro\-ko\-mpo\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ero\-ko\-mpoi\-re}
\fmtSense{to talk irresponsibly}
\fmtExampleRL{Aka\-ka\-li ako ka\-ta\-ka mu\-no okwe\-ro\-ko\-mpo\-la n’e\-bi\-ga\-mbo.}
\fmtExampleEN{That small woman really likes to talk irresponsibly.}

\fmtLemma{ero\-nda}
\fmtCitation{(okw)e\-ro\-nda}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ero\-nde\-re}
\fmtSense{to make a fuss, to be picky, to quarrel over small things}
\fmtExampleRL{Omu\-ka\-li wa\-nge yee\-ro\-nda\-lo\-nda na bu\-li ki\-mwei.}
\fmtExampleEN{My wife makes a fuss over anything.}

\fmtLemma{eroo\-ra}
\fmtCitation{(okw)e\-roo\-ra}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{ero\-rei\-re}
\fmtSubentry(a)
\fmtSense{to see for oneself, to look at, to watch}
\fmtExampleRL{Mwi\-ze mwe\-roo\-re oku mu\-syo\-li.}
\fmtExampleEN{Come and look at the night dancer.}
\fmtSubentry(b)
\fmtSense{to despise, to loathe, to look down on, to sneer at}
\fmtExampleRL{Aba\-lu\-ga\-nda ni mu\-lwa\-na\-nga\-na aba\-ntu ba\-bee\-roo\-ra.}
\fmtExampleEN{When relatives fight, people start despising them.}

\fmtLemma{eroo\-tee\-ra}
\fmtCitation{(okw)e\-roo\-tee\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eroo\-te\-rei\-re}
\fmtSense{to have a wet dream}
\fmtExampleRL{Aboo\-jo ba\-te\-ra okwe\-roo\-tee\-ra.}
\fmtExampleEN{Boys usually have wet dreams.}

\fmtLemma{eroo\-tee\-rya}
\fmtCitation{(okw)e\-roo\-tee\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eroo\-tee\-rei\-rye}
\fmtSense{to have a wet dream}
\fmtExampleRL{Aboo\-jo ba\-te\-ra okwe\-roo\-tee\-rya.}
\fmtExampleEN{Boys usually have wet dreams.}

\fmtLemma{eru}
\fmtCitation{\textnormal{(e.g.}\,mweu, gwe\-ru, nje\-ru\textnormal{)}}
\fmtPOS{adj}
\fmtSense{white}
\fmtExampleRL{Enkai\-to za\-nge zee\-ru.}
\fmtExampleEN{My shoes are white.}

\fmtLemma{eru\-guu\-ra}
\fmtCitation{(okw)e\-ru\-guu\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eru\-gwi\-re}
\fmtSense{to encircle, to surround, to go round}
\fmtExampleRL{Baa\-si\-mbi\-re emi\-saa\-le okwe\-ru\-guu\-ra embu\-ga ya Isa\-ba\-nya\-la.}
\fmtExampleEN{They surrounded the Iasabanyala’s palace with trees.}

\fmtLemma{eru\-guu\-ri}
\fmtCitation{\textnormal{(e.g.}\,mwe\-ru\-guu\-ri, gwe\-ru\-guu\-ri, kye\-ru\-guu\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSense{round, circular}
\fmtExampleRL{Omu\-pii\-ra gwa gaa\-li gu\-ta gwe\-ru\-guu\-ri.}
\fmtExampleEN{The bicycle's tyre is round.}

\fmtLemma{eru\-guu\-rya}
\fmtCrossreference{\textnormal{see also} ruguu\-rya}
\fmtCitation{(okw)e\-ru\-guu\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eru\-guu\-rii\-rye}
\fmtSense{to surround, to encircle}
\fmtExampleRL{Bbaa\-si\-mbi\-re emi\-saa\-le okwe\-ru\-guu\-rya enyu\-mba.}
\fmtExampleEN{They planted trees around the house.}

\fmtLemma{eru\-kii\-ri}
\fmtCitation{\textnormal{(e.g.}\,mwe\-ru\-kiii\-ri, beee\-ru\-kii\-ri, kye\-ru\-kii\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSense{whitish}
\fmtExampleRL{Emo\-to\-ka yaa\-mwe gye\-ru\-kii\-ri.}
\fmtExampleEN{His car is whitish.}

\fmtLemma{eru\-ku\-kya}
\fmtCitation{(okw)e\-ru\-ku\-kya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eru\-ku\-ki\-rye}
\fmtSense{to soften, to soak}
\fmtExampleRL{Ye\-ru\-ku\-kya ese\-fu\-lya edi ni wa\-ma\-la ogi\-kuu\-te.}
\fmtExampleEN{Soak the saucepan and clean it afterwards.}

\fmtLemma{eru\-la}%
\fmtOne{1}
\fmtCitation{(okw)e\-ru\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{erwi\-re}
\fmtSense{to clear for cultivation}
\fmtExampleRL{Mwe\-ru\-le eki\-si\-ko tu\-si\-ge ama\-ca\-nde.}
\fmtExampleEN{Clear the bush and we'll grow simsim.}

\fmtLemma{eru\-la}%
\fmtOne{2}
\fmtCitation{(okw)e\-ru\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{erwi\-re}
\fmtSense{to be named after somebody}
\fmtExampleRL{Oni omwa\-na baa\-mwe\-rwi\-re\-mu zei\-za waa\-mwe.}
\fmtExampleEN{This child was named after his grandfather.}

\fmtLemma{eru\-mu\-ka}
\fmtCitation{(okw)e\-ru\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eru\-mu\-ki\-re}
\fmtSense{to have nosebleed, to bleed from the nose}
\fmtExampleRL{Omwa\-na akwe\-ru\-mu\-ka.}
\fmtExampleEN{The child is having a nosebleed.}

\fmtLemma{eru\-ndu\-ka}
\fmtCitation{(okw)e\-ru\-ndu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eru\-ndu\-ki\-re}
\fmtSense{to be frightened, to be scared}
\fmtExampleRL{Wee\-ri\-ndu\-ki\-re ki?}
\fmtExampleEN{What has scared you?}

\fmtLemma{eruu\-ba}
\fmtCitation{(okw)e\-ruu\-ba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eruu\-bi\-re}
\fmtSense{to join, to team up with}
\fmtExampleRL{Wee\-ruu\-be n’a\-bai\-ra\-baa\-mu ba\-di.}
\fmtExampleEN{Join your other colleagues.}

\fmtLemma{eru\-ye\-ru}
\fmtCitation{\textnormal{(e.g.}\,lwe\-ru\-ye\-ru, bye\-ru\-ye\-ru, mwe\-ru\-ye\-ru\textnormal{)}}
\fmtPOS{adj}
\fmtSense{off-white}
\fmtExampleRL{Ya\-guu\-re eki\-tee\-tee\-yi kye\-ru\-ye\-ru.}
\fmtExampleEN{She bought an off-white dress.}

\fmtLemma{erwa\-na\-ku}
\fmtCitation{(okw)e\-rwa\-na\-ku}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{erwai\-ne\-ku}
\fmtSense{to defend oneself}
\fmtExampleRL{Mwe\-ge okwe\-rwa\-na\-ku.}
\fmtExampleEN{Learn to defend yourself.}

\fmtLemma{erya}%
\fmtOne{1}
\fmtCitation{(okw)e\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eryi\-re}
\fmtSense{to be purified, to cleanse oneself}
\fmtExampleRL{Baa\-ya\-ba\-nga ewa Ye\-su oku\-bee\-rya.}
\fmtExampleEN{They used to go to Jesus to be purified.}

\fmtLemma{erya}%
\fmtOne{2}
\fmtCitation{(okw)e\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eryi\-re}
\fmtSense{to have a bumper harvest}
\fmtExampleRL{Omwa\-ka ogu\-wei\-re twa\-yee\-rye eme\-re.}
\fmtExampleEN{We had a bumper harvest last year.}

\fmtLemma{eryo}
\fmtCrossreference{\textnormal{see} o\textnormal{\fmtOne{3}}}

\fmtLemma{esa}
\fmtCitation{(okw)e\-sa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ese\-re}
\fmtSense{to play (a board game)}
\fmtExampleRL{Nje mmai\-te okwe\-sa omwe\-so.}
\fmtExampleEN{I know how to play the mweso board game.}

\fmtLemma{esaa\-si\-ra}
\fmtCitation{(okw)e\-saa\-si\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esaa\-sii\-re}
\fmtSense{to have self-pity, to pity oneself, to feel sorry for oneself}
\fmtExampleRL{Omu\-ntu eyee\-saa\-si\-ra ta\-tu\-nga bi\-ntu.}
\fmtExampleEN{A person who has self-pity doesn't acquire property.}

\fmtLemma{esaa\-yii\-rya}
\fmtCitation{(okw)e\-saa\-yii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esaa\-yi\-rii\-rye}
\fmtSense{to apply makeup}
\fmtExampleRL{Aba\-ka\-li ba\-twa\-la obwi\-re bwi\-ngi nga ba\-kwe\-saa\-yii\-rya.}
\fmtExampleEN{Women spend a lot of time applying makeup.}

\fmtLemma{esa\-bu\-la}
\fmtCrossreference{\textnormal{see} sabu\-la}

\fmtLemma{esa\-lii\-ra}
\fmtCitation{(okw)e\-sa\-lii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esa\-li\-rii\-re}
\fmtSubentry(a)
\fmtSense{to develop a crack}
\fmtExampleRL{Eso\-waa\-ni ya\-nge yee\-sa\-li\-rii\-re.}
\fmtExampleEN{My plate has developed a crack.}
\fmtSubentry(b)
\fmtSense{to turn septic, to become infected}
\fmtExampleRL{Ekyo\-ya kye\-sa\-li\-rii\-re.}
\fmtExampleEN{The wound has turned septic.}

\fmtLemma{esa\-mba}
\fmtCitation{(okw)e\-sa\-mba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{esa\-mbi\-re}
\fmtSense{to avoid, to keep away from, to stay away from, to steer clear of}
\fmtExampleRL{Mwe\-sa\-mbe enta\-lo za ba\-sai\-za abo.}
\fmtExampleEN{Avoid the fights of those men.}

\fmtLemma{esa\-ra\-gu\-la}
\fmtCitation{(okw)e\-sa\-ra\-gu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esa\-ra\-gwi\-re}
\fmtSense{to boil, to reach the boiling point}
\fmtExampleRL{Amai\-zi ga\-kwe\-sa\-ra\-gu\-la.}
\fmtExampleEN{The water is boiling.}

\fmtLemma{esa\-rai\-gu\-la}
\fmtCrossreference{\textnormal{see} esara\-gu\-la}

\fmtLemma{esa\-sa\-ngu\-la}
\fmtCitation{(okw)e\-saa\-nsa\-gu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esaa\-nsa\-gwi\-re}
\fmtSense{to be in the state of despair, to be demoralised}
\fmtExampleRL{Omu\-ka\-li ni ba\-mu\-bi\-kii\-re ne ye\-sa\-sa\-ngu\-la mu\-no.}
\fmtExampleEN{When they announced the death of her husband, the woman fell into a state of despair.}

\fmtLemma{esee\-tya}
\fmtCitation{(okw)e\-see\-tya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esee\-tye\-re}
\fmtSubentry(a)
\fmtSense{to clean one's anus by rubbing it against the ground}
\fmtExampleRL{Abaa\-na aba\-to ba\-ta\-ka mu\-no okwe\-see\-tya.}
\fmtExampleEN{Young children like cleaning their anus by rubbing it against the ground.}
\fmtSubentry(b)
\fmtSense{(of spirits) to move on buttocks}
\fmtExampleRL{Omu\-ntu oni akwe\-see\-tya nk’a\-ba\-ndwa Guu\-da.}
\fmtExampleEN{This person moves on their buttocks as if he were possessed with the spirit called Guuda.}

\fmtLemma{ese\-nda}
\fmtCitation{(okw)e\-se\-nda}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ese\-nde\-re}
\fmtSense{to move away, to give space}
\fmtExampleRL{Wee\-se\-nde\-ku enyu\-ma ka\-doo\-li.}
\fmtExampleEN{Move back slightly.}

\fmtLemma{ese\-ngee\-rya}
\fmtCitation{(okw)e\-se\-ngee\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ese\-nge\-rei\-rye}
\fmtSubentry(a)
\fmtSense{to follow, to go behind, to go after, to walk behind}
\fmtExampleRL{Ya\-se\-nge\-rei\-rye omu\-ka\-li okwi\-ki\-ra kya\-ru\-mwe oku ma\-duu\-ka.}
\fmtExampleEN{She followed the woman all the way to the shops.}
\fmtSubentry(b)
\fmtSense{to beg, to plead, to appeal to, to ask}
\fmtExampleRL{Nkwe\-se\-nge\-rei\-rye bai\-ra\-ba iza ema\-mbya.}
\fmtExampleEN{I am begging you to come tomorrow.}
\fmtSubentry(c)
\fmtSense{to turn to (God)}
\fmtExampleRL{Mwe\-se\-ngee\-rye oKa\-nca aba\-saa\-si\-re ebi\-bbi.}
\fmtExampleEN{Turn to God, he will forgive your sins.}

\fmtLemma{ese\-nge\-sya}
\fmtCitation{(okw)e\-se\-nge\-sya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ese\-nge\-se\-rye}
\fmtSense{to squat, to settle on land unlawfully}
\fmtExampleRL{Omu\-sai\-za oyo ye\-se\-nge\-se\-rye ewa\-nge.}
\fmtExampleEN{That man settled on my land without my permission.}

\fmtLemma{ese\-re\-ka}
\fmtCitation{(okw)e\-se\-re\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ese\-re\-ke\-re}
\fmtSense{(of chicks) to hide under the hen's wings}
\fmtExampleRL{Obu\-ko\-ko bwe\-se\-re\-ka omu bi\-pa\-pa bya bu\-gu\-ma.}
\fmtExampleEN{The chicks hide under their mother's wings.}

\fmtLemma{ese\-re\-ra}
\fmtCitation{(okw)e\-se\-re\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ese\-rei\-re}
\fmtSense{to give (cattle) to drink, to give water to}
\fmtExampleRL{Ente nzee\-se\-re\-ra oku iru\-ba lya\-nge.}
\fmtExampleEN{I give water to the cattle from my well.}

\fmtLemma{esi\-ga}
\fmtCitation{(okw)e\-si\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{esi\-gi\-re}
\fmtSense{to trust, to believe in, to have confidence in}
\fmtExampleRL{Nje nkwe\-si\-ga mu\-no.}
\fmtExampleEN{I trust you a lot.}

\fmtLemma{esi\-gaa\-rya}
\fmtCitation{(okw)e\-si\-gaa\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{esi\-gaa\-li\-rye}
\fmtSense{to keep (to oneself), to save, to put aside, to set aside}
\fmtExampleRL{Wee\-si\-gaa\-rye\-nge\-ku empii\-ya.}
\fmtExampleEN{Save some of your money.}

\fmtLemma{esi\-ga\-ma}
\fmtCitation{(okw)e\-si\-ga\-ma}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{esi\-ga\-mi\-re}
\fmtSubentry(a)
\fmtSense{to cause, to bring about, to lead to, to result in}
\fmtExampleRL{Ebi\-zi\-bu bya\-nge byo\-na bye\-si\-gai\-me ku nii\-we.}
\fmtExampleEN{You are the one causing all my problems.}
\fmtSubentry(b)
\fmtSense{to rely on, to count on}
\fmtExampleRL{Nje nee\-si\-ga\-mi\-re ku bu\-koi\-za oku\-fu\-na ebi\-sa\-le bya iso\-me\-ro.}
\fmtExampleEN{I rely on my maternal uncles’ side for my school fees.}

\fmtLemma{esi\-gi\-ka}
\fmtCitation{(okw)e\-si\-gi\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esi\-gi\-ki\-re}
\fmtSubentry(a)
\fmtSense{to lean (against)}
\fmtExampleRL{Omu\-sai\-za akee\-rwe yee\-si\-gi\-ki\-re oku mu\-saa\-le.}
\fmtExampleEN{The tired man is leaning against a tree.}
\fmtSubentry(b)
\fmtSense{to be trustworthy, to inspire confidence}

\fmtLemma{esi\-gu\-la}
\fmtCitation{(okw)e\-si\-gu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esi\-gwi\-re}
\fmtSense{to rely on, to count on}
\fmtExampleRL{Okwe\-si\-guu\-ra ku Amo\-s ema\-mbya akwa\-ba waa\-bwe.}
\fmtExampleEN{You are relying on Amos, but tomorrow he will be gone.}

\fmtLemma{esi\-gwa}
\fmtCitation{\textnormal{(e.g.}\,mwe\-si\-gwa, bee\-si\-gwa\textnormal{)}}
\fmtPOS{adj}
\fmtSense{faithful, loyal, devoted}
\fmtExampleRL{Ndi mwe\-si\-gwa egi\-ri oKa\-nca wa\-nge.}
\fmtExampleEN{I am faithful to my God.}

\fmtLemma{esii\-ma}
\fmtCitation{(okw)e\-sii\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esii\-mi\-re}
\fmtSense{to be happy, to be glad, to be pleased}
\fmtExampleRL{Twe\-sii\-mi\-re mu\-no olwa nsi yai\-swe ei\-ku\-ra lya Afri\-ca.}
\fmtExampleEN{We are very happy with our country, the pearl of Africa.}

\fmtLemma{esii\-mi\-sya}
\fmtCitation{(okw)e\-sii\-mi\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{esii\-mi\-si\-rye}
\fmtSense{to be proud of}
\fmtExampleRL{Aba\-byai\-re bee\-sii\-mi\-sya omwa\-ra waa\-bwe.}
\fmtExampleEN{The parents are proud of their daughter.}

\fmtLemma{esi\-ma\-gii\-rya}
\fmtCitation{(okw)e\-si\-ma\-gii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esi\-ma\-gii\-rye}
\fmtSense{to doze, to nap, to drowse, to snooze}
\fmtExampleRL{Ni kaa\-maa\-re ni ka\-ta\-ndi\-ka okwe\-si\-ma\-gii\-rya.}
\fmtExampleEN{Eventually she started dozing.}

\fmtLemma{esi\-mba}
\fmtCitation{(okw)e\-si\-mba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esi\-mbi\-re}
\fmtSense{to stand, to plant oneself, to take up a position}
\fmtExampleRL{Otee\-si\-mba ku itee\-wo ng’a\-ku\-lya.}
\fmtExampleEN{Don’t stand next to your father while he is eating.}

\fmtLemma{esi\-mba\-wo}
\fmtCitation{(okw)e\-si\-mba\-wo}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esi\-mbi\-re\-wo}
\fmtSense{to contest, to compete for, to try to win}
\fmtExampleRL{Aku\-ta\-ka kwe\-si\-mba\-wo oku bu\-kwe\-nda bwa paa\-la\-me\-nti.}
\fmtExampleEN{He wants to run for the post of member of parliament.}

\fmtLemma{esi\-mo\-ngo\-ra\-na}
\fmtCitation{(okw)e\-si\-mo\-ngo\-ra\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esi\-mo\-ngo\-rai\-ne}
\fmtSense{to be inconsistent, to changeable, to unpredictable}
\fmtExampleRL{Oyo mwe\-ge\-nde\-re\-ze, ye\-si\-mo\-ngo\-ra\-na.}
\fmtExampleEN{Be careful, he is unpredictable.}

\fmtLemma{esi\-ndyo\-la}
\fmtCitation{(okw)e\-si\-ndyo\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esi\-ndyoi\-re}
\fmtSense{to somersault}
\fmtExampleRL{Omu\-ta\-mii\-re yee\-si\-ndyoi\-re n’a\-gwa.}
\fmtExampleEN{A drunkard somersaulted and fell down.}

\fmtLemma{esi\-rii\-kii\-rya}
\fmtCitation{(okw)e\-si\-rii\-kii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{esi\-rii\-ki\-rii\-rye}
\fmtSubentry(a)
\fmtSense{to not disclose, to conceal, to hide}
\fmtExampleRL{Ndi\-na eki\-zi\-bu lwa kwe\-si\-rii\-kii\-rya.}
\fmtExampleEN{I have a problem, although I have not disclosed it.}
\fmtSubentry(b)
\fmtSense{to pretend to be quiet, to respond with silence, to turn a deaf ear}
\fmtExampleRL{Omu\-ka\-li oyo amai\-te okwe\-si\-rii\-kii\-rya bai\-tu nga mu\-la\-lu ki\-ta\-lo.}
\fmtExampleEN{That woman pretends to be quiet, yet she is very stubborn.}

\fmtLemma{esi\-si\-ra}
\fmtCitation{(okw)e\-si\-si\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esi\-sii\-re}
\fmtSense{to treat with disgust, to have aversion to}
\fmtExampleRL{Le\-ka kwe\-si\-si\-ra mu\-kai\-re.}
\fmtExampleEN{Don’t treat the old woman with disgust.}

\fmtLemma{esi\-si\-wa\-la}
\fmtCitation{(okw)e\-si\-si\-wa\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esi\-si\-wai\-re}
\fmtSense{to tremble all over, to shiver, to have a creepy sensation, to have a tingling sensation, to be shocked, to be traumatise, to be upset, to be disturbed}
\fmtExampleRL{Ni na\-bwoi\-ne omu\-tu\-mbi ni nee\-si\-si\-wa\-la.}
\fmtExampleEN{When I saw a corpse, I was shocked.}

\fmtLemma{eso\-bo\-la}
\fmtCitation{(okw)e\-so\-bo\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eso\-boi\-re}
\fmtSense{to be capable}
\fmtExampleRL{Akwe\-so\-bo\-la, ya\-pa\-ngi\-sya mo\-to\-ka.}
\fmtExampleEN{He is capable, he will hire a car.}

\fmtLemma{eso\-nyi\-wa}
\fmtCitation{(okw)e\-so\-nyi\-wa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eso\-nyi\-wi\-re}
\fmtSense{to ignore, to disregard, to take no notice of, to pay no attention to}
\fmtExampleRL{Wee\-so\-nyi\-we bye ya\-ku\-koo\-re.}
\fmtExampleEN{Ignore what he did to you.}

\fmtLemma{eso\-roo\-lya}
\fmtCitation{(okw)e\-so\-roo\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eso\-roi\-rye}
\fmtSense{to assemble, to gather together}
\fmtExampleRL{Mwe\-so\-roo\-lye ani.}
\fmtExampleEN{Assemble here.}

\fmtLemma{esu\-ku\-la}
\fmtCitation{(okw)e\-su\-ku\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esu\-kwi\-re}
\fmtSense{to move bouncing one's breasts}
\fmtExampleRL{Omu\-ka\-li waa\-mu yee\-su\-ku\-la omu ba\-ntu.}
\fmtExampleEN{Your wife walks with her breasts bouncing in public.}

\fmtLemma{esu\-li\-ka}
\fmtCitation{(okw)e\-su\-li\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esu\-li\-ki\-re}
\fmtSubentry(a)
\fmtSense{to be upside down, to be turned over, to turn oneself upside down}
\fmtSubentry(b)
\fmtSense{to wane}
\fmtExampleRL{Okwe\-zi kwe\-su\-li\-ki\-re.}
\fmtExampleEN{The moon has waned.}

\fmtLemma{esu\-lu\-ka}
\fmtCitation{(okw)e\-su\-lu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esu\-lu\-ki\-re}
\fmtSense{to incline, to tilt, to angle}
\fmtExampleRL{We to\-ku\-bo\-na nga kye\-su\-lu\-ki\-re!}
\fmtExampleEN{Don’t you see that it is tilting!}

\fmtLemma{esu\-nga}
\fmtCitation{(okw)e\-su\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esu\-ngi\-re}
\fmtSense{to look forward to, to be eager, to be keen on}
\fmtExampleRL{Emba\-ga ya Amo\-s tu\-ku\-gye\-su\-nga mu\-no.}
\fmtExampleEN{We are looking forward to Amos’ wedding.}

\fmtLemma{esuu\-bee\-tya}
\fmtCitation{(okw)e\-suu\-bee\-tya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{esuu\-bee\-te\-rye}
\fmtSubentry(a)
\fmtSense{to hang down}
\fmtExampleRL{Omu\-la\-mbo gwa mu\-sai\-za eyee\-zi\-ndi\-re ba\-gu\-sa\-ngi\-re nga gu\-kwe\-suu\-bee\-tya ku mu\-saa\-le.}
\fmtExampleEN{The body of a man who committed suicide was found hanging down from the tree.}
\fmtSubentry(b)
\fmtSense{to swing}
\fmtExampleRL{Mboi\-ne aba\-ka\-li nga ba\-kwe\-suu\-bee\-tya ku kye\-suu\-bo kya baa\-na.}
\fmtExampleEN{I have seen women swinging using the children's swing.}

\fmtLemma{eswa\-nta}
\fmtCitation{(okw)e\-swa\-nta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eswa\-nti\-re}
\fmtSense{to be proud, to be boastful, to be insolent, to become angry, to display one strength}

\fmtLemma{eswe\-kee\-rya}
\fmtCitation{(okw)e\-swe\-kee\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eswe\-ke\-rei\-rye}
\fmtSense{to throw over one's shoulders}
\fmtExampleRL{Aba\-ka\-li bee\-swe\-ke\-rei\-rye engo\-ye obwi\-re bwa mu\-lo\-mbe.}
\fmtExampleEN{The women threw clothes over their shoulders when the weather was cool.}

\fmtLemma{(-)esy}
\fmtCrossreference{\textnormal{see} (-)isy}

\fmtLemma{esye\-tya}
\fmtCrossreference{\textnormal{see} etyetya}

\fmtLemma{eta}
\fmtCitation{(okw)e\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ete\-re}
\fmtSense{to call, to summon}
\fmtExampleRL{OKa\-to aya\-bi\-re kwe\-ta mu\-ga\-nda waa\-mwe.}
\fmtExampleEN{Kato has gone to call his brother.}

\fmtLemma{etaa\-ga}
\fmtCitation{(okw)e\-taa\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{etaa\-gi\-re}
\fmtSense{to need, to want, to like}
\fmtExampleRL{Obbai\-re okwe\-taa\-ga ki?}
\fmtExampleEN{What do you need?}

\fmtLemma{etaa\-ya}
\fmtCitation{(okw)e\-taa\-ya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{etaa\-yi\-re}
\fmtSense{to be free, to feel free}
\fmtExampleRL{Mwe\-taa\-ye, ani waa\-nywe.}
\fmtExampleEN{Feel free, you are at home.}

\fmtLemma{eta\-bu\-la}
\fmtCitation{(okw)e\-ta\-bu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eta\-bwi\-re}
\fmtSense{to mix thoroughly under heat before boiling point}
\fmtExampleRL{Era\-ngi yee\-ta\-bwi\-re ku\-sai.}
\fmtExampleEN{The colour has mixed well.}

\fmtLemma{eta\-bu\-li\-sya}
\fmtCitation{(okw)e\-ta\-bu\-li\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eta\-bu\-li\-si\-rye}
\fmtSense{to pretend to stir}
\fmtExampleRL{Lwa\-ki wee\-ta\-bu\-li\-sye obu\-sye\-ra?}
\fmtExampleEN{Why do you pretend to stir the millet porridge?}

\fmtLemma{eta\-ki\-ra}
\fmtCitation{(okw)e\-ta\-ki\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eta\-kii\-re}
\fmtSense{to attract (an unpleasant reaction, e.g. trouble or beating), to invite for oneself}
\fmtExampleRL{Omwo\-jo omu\-woo\-le yee\-ta\-ki\-ra mii\-go.}
\fmtExampleEN{A stubborn boy invites beatings for himself.}

\fmtLemma{eta\-ki\-sya}
\fmtCitation{(okw)e\-ta\-ki\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eta\-ki\-si\-rye}
\fmtSense{to wish for oneself}
\fmtExampleRL{OWa\-nza\-la yee\-ta\-ki\-sya ye\-nkai bu\-li ki\-ntu eki\-sai.}
\fmtExampleEN{Wanzala wishes all the good things for himself.}

\fmtLemma{eta\-mbuu\-ra}
\fmtCitation{(okw)e\-ta\-mbuu\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eta\-mbu\-lii\-re}
\fmtSense{to go away, to leave, to depart}
\fmtExampleRL{Yee\-ta\-mbu\-lii\-re n’a\-zwa\-wo.}
\fmtExampleEN{She walked away.}

\fmtLemma{eta\-mwa}
\fmtCitation{(okw)e\-ta\-mwa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eta\-mi\-rwe}
\fmtSense{to be fed up, to be disillusioned}
\fmtExampleRL{Ebi\-zi\-bu ni bi\-kii\-ka wee\-ta\-mwa.}
\fmtExampleEN{When there are too many problems you get fed up.}

\fmtLemma{eta\-nta\-la}
\fmtCitation{(okw)e\-ta\-nta\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eta\-ntai\-re}
\fmtSense{to avoid, to keep away from, to stay away from, to steer clear of}
\fmtExampleRL{Wee\-ta\-nta\-le aka\-bbi nga ka\-kya\-li ku\-kwi\-ka\-ku.}
\fmtExampleEN{Avoid the trouble before it happens to you.}

\fmtLemma{eta\-wu\-la}
\fmtCitation{(okw)e\-ta\-wu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eta\-wi\-re}
\fmtSense{to move up and down, to be unsettled, to be restless, to be fidgety}
\fmtExampleRL{Oyo lwa\-ki akwe\-ta\-wu\-la mu\-no?}
\fmtExampleEN{Why is that one so restless?}

\fmtLemma{eta\-yi\-ri\-ra}
\fmtCitation{(okw)e\-ta\-yi\-ri\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eta\-yi\-rii\-re}
\fmtSense{to pester, to plague, to annoy, to bother}
\fmtExampleRL{ONa\-mpii\-na anee\-ta\-yi\-rii\-re\-ku emi\-ru\-ndi gi\-ngi ng’a\-ku\-ta\-ka mu\-se\-nde\-kee\-rye, ni ngaa\-na.}
\fmtExampleEN{Nampiina has approached me several times pestering me to accompany her, but I refused.}

\fmtLemma{ete\-bu\-ka}
\fmtCitation{(okw)e\-te\-bu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ete\-bu\-ki\-re}
\fmtSense{to respond to a call, to react to a call}
\fmtExampleRL{Ni ba\-yee\-ta ei\-ba\-ra lya\-mu, bai\-ra\-ba wee\-te\-bu\-ke.}
\fmtExampleEN{When they call your name please respond.}

\fmtLemma{etee\-ka\-ni\-za}
\fmtCitation{(okw)e\-tee\-ka\-ni\-za}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{etee\-ka\-ni\-zi\-re}
\fmtSense{to prepare oneself, to get prepared, to get ready}
\fmtExampleRL{Twe\-tee\-ka\-ni\-ze mo\-to\-ka ekwi\-za ku\-tu\-twa\-la.}
\fmtExampleEN{Let’s be prepared (lit.\,Let us prepare ourselves), the car is coming to take us}

\fmtLemma{etee\-ka\-tee\-ka}
\fmtCitation{(okw)e\-tee\-ka\-tee\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{etee\-ka\-tee\-ke\-re}
\fmtSense{to prepare oneself, to get prepared, to get ready}
\fmtExampleRL{Tu\-kwe\-tee\-ka\-tee\-ka tu\-kwa\-ba ku mba\-ga.}
\fmtExampleEN{We are getting ready for the wedding.}

\fmtLemma{etee\-rya}
\fmtCitation{(okw)e\-tee\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ete\-rei\-rye}
\fmtSense{to call repeatedly}
\fmtExampleRL{Omu\-ntu akwe\-tee\-rya aka\-yu\-la.}
\fmtExampleEN{A person who calls you repeatedly is annoying.}

\fmtLemma{etee\-sa}
\fmtCitation{(okw)e\-tee\-sa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{etee\-se\-re}
\fmtSense{to agree before a negotiation, to take a common stand, to agree on something}
\fmtExampleRL{Tu\-soo\-ke twe\-tee\-se ale twi\-ti\-ra\-ne Omu\-ka\-ma.}
\fmtExampleEN{Let's agree first, then we meet the King.}

\fmtLemma{etee\-ta}
\fmtCitation{(okw)e\-tee\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{etee\-te\-re}
\fmtSense{to rub off, to clean off, to shake off}
\fmtExampleRL{Wee\-tee\-te ebi\-ta\-ka\-li omu nga\-lo.}
\fmtExampleEN{Clean the dirt off your hands.}

\fmtLemma{ete\-ge\-fu}
\fmtCrossreference{\textnormal{see also} etege\-ki}
\fmtCitation{\textnormal{(e.g.}\,lye\-te\-ge\-fu, bee\-te\-ge\-fu, mwe\-te\-ge\-fu\textnormal{)}}
\fmtPOS{adj}
\fmtSense{ready, prepared, all set}
\fmtExampleRL{Oli mwe\-te\-ge\-fu okwa\-ba?}
\fmtExampleEN{Are you ready to go?}

\fmtLemma{ete\-ge\-ka}
\fmtCitation{(okw)e\-te\-ge\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ete\-ge\-ke\-re}
\fmtSense{to be ready, to be prepared, to be all set}
\fmtExampleRL{Mba\-le\-ke\-re eka nga ba\-kwe\-te\-ge\-ka.}
\fmtExampleEN{I left them at home while they were getting ready.}

\fmtLemma{ete\-ge\-ki}
\fmtCitation{\textnormal{(e.g.}\,mwe\-te\-ge\-ki, bee\-te\-ge\-ki, lye\-te\-ge\-fu\textnormal{)}}
\fmtPOS{adj}
\fmtSense{ready}
\fmtExampleRL{Omwe\-te\-ge\-ki omwei ai\-ze ani.}
\fmtExampleEN{Whoever is ready should come here.}

\fmtLemma{ete\-ge\-re\-za}
\fmtCitation{(okw)e\-te\-ge\-re\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ete\-ge\-rei\-ze}
\fmtSubentry(a)
\fmtSense{to grasp, to comprehend, to understand, to recognize, to identify}
\fmtSubentry(b)
\fmtSense{to assess, to look keenly into, to analyse}
\fmtExampleRL{Ba\-kwe\-te\-ge\-re\-za bi\-zi\-bu bya baa\-na.}
\fmtExampleEN{They are assessing the children's problems.}

\fmtLemma{ete\-gu\-la}
\fmtCitation{(okw)e\-te\-gu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ete\-gwi\-re}
\fmtSense{to leave (in anticipation of a problem), to flee}
\fmtExampleRL{Wee\-te\-gu\-le ng’o\-bwi\-re bu\-kya\-li ba\-ta\-ku\-kwa\-ta.}
\fmtExampleEN{Leave early so that they won't arrest you.}

\fmtLemma{ete\-ja}
\fmtCitation{(okw)i\-te\-ja}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ete\-je\-rye}
\fmtSense{to understand, to realise, to comprehend, to grasp, to be aware of}
\fmtExampleRL{We we\-te\-ja ma\-ngu.}
\fmtExampleEN{You understand quickly.}
\fmtSense{to examine}
\fmtSense{understanding}
\fmtExampleRL{ODi\-ndi mu\-sai\-za mwe\-te\-ja.}
\fmtExampleEN{Dindi is an understanding man.}

\fmtLemma{ete\-ma}
\fmtCitation{(okw)e\-te\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ete\-me\-re}
\fmtSense{to pledge, to promise, to vow, to commit oneself, to promise oneself}
\fmtExampleRL{Twa\-bbai\-re mu lu\-kii\-ko bu\-li omwei ni yee\-te\-ma kye ya\-so\-bo\-la oku\-wee\-rya.}
\fmtExampleEN{We were in a meeting and each of us pledged what one can offer.}

\fmtLemma{ete\-nde\-be\-ka}
\fmtCitation{(okw)e\-te\-nde\-be\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ete\-nde\-beke\-re}
\fmtSense{to humble oneself}
\fmtExampleRL{N’o\-ya\-ba omu mai\-so ga Mu\-ka\-ma, wee\-te\-nde\-be\-ka.}
\fmtExampleEN{When you go to the King, you humble yourself.}

\fmtLemma{ete\-ngee\-ra}
\fmtCitation{(okw)e\-te\-ngee\-ra}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{ete\-nge\-rei\-re}
\fmtSubentry(a)
\fmtSense{to balance a load on the head while walking}
\fmtExampleRL{Yee\-te\-nge\-rei\-re ensu\-wa oku mu\-twe.}
\fmtExampleEN{She is balancing the pot on her head.}
\fmtSubentry(b)
\fmtSense{(of a baby) to attempt to stand}
\fmtExampleRL{Omwa\-na akwe\-te\-ngee\-ra.}
\fmtExampleEN{The child is attempting to stand.}
\fmtSubentry(c)
\fmtSense{to be independent, to be self-reliant}
\fmtExampleRL{Wee\-te\-ngee\-re oko\-le omu\-li\-mu gwa\-mu.}
\fmtExampleEN{Be independent and do your work.}

\fmtLemma{ete\-sya}
\fmtCitation{(okw)e\-te\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ete\-se\-rye}
\fmtSubentry(a)
\fmtSense{to send for, to cause to come, to order (\eg goods)}
\fmtExampleRL{Nyee\-te\-se\-rye oKa\-to ai\-ze ati.}
\fmtExampleEN{I have sent for Kato to come now.}
\fmtSubentry(b)
\fmtSense{to invite}
\fmtExampleRL{Yee\-te\-se\-rye aba\-taa\-ni baa\-mwe bai\-ze.}
\fmtExampleEN{He invited all his sons to come.}

\fmtLemma{etii\-ka}
\fmtCitation{(okw)e\-tii\-ka}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{etii\-ki\-re}
\fmtSubentry(a)
\fmtSense{to carry (on the head), to put on one's head, to carry (of vehicles)}
\fmtExampleRL{Omu\-ka\-li yee\-tii\-ka enkwi oku mu\-twe bu\-li ki\-ro.}
\fmtExampleEN{The woman carries firewood on her head every day.}
\fmtSubentry(b)
\fmtFixedexpression{etii\-ka ki\-koo\-do}
\fmtSense{to be infected with measles}
\fmtExampleRL{Omu irwa\-li\-ro mu\-li\-mu abaa\-na bai\-ngi abee\-tii\-ki\-re.}
\fmtExampleEN{There are many children in the hospital who are suffering from measles.}
\fmtSubentry(c)
\fmtSense{to be infected with measles}
\fmtExampleRL{Yee\-tii\-ki\-re ki\-koo\-do.}
\fmtExampleEN{He got measles.}
\fmtSubentry(d)
\fmtSense{to be at the head of the class}

\fmtLemma{etii\-na}
\fmtCitation{(okw)e\-tii\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{etii\-ni\-re}
\fmtSense{to lack confidence, to be timid, to feel insecure}
\fmtExampleRL{Le\-ka kwe\-tii\-na, eki\-ba\-lo wa\-ki\-ko\-la ku\-sai.}
\fmtExampleEN{Don't be timid, you will perform well in mathematics.}

\fmtLemma{etii\-za}
\fmtCitation{(okw)e\-tii\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{etii\-zi\-re}
\fmtSense{to borrow}
\fmtExampleRL{Ya\-ba wee\-tii\-ze eka\-nzu twa\-be.}
\fmtExampleEN{Go and borrow a kanzu and let's go.}

\fmtLemma{eti\-mba}
\fmtCitation{(okw)e\-ti\-mba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eti\-mbi\-re}
\fmtSense{to hang (a curtain, picture), to set (\eg a mosquito net)}
\fmtExampleRL{Omu\-go\-le ya\-suu\-re ti\-yee\-ti\-mbi\-re ta\-nda\-lu\-wa.}
\fmtExampleEN{The bride slept without setting the mosquito net.}

\fmtLemma{eto}
\fmtCrossreference{\textnormal{see} o\textnormal{\fmtOne{3}}}

\fmtLemma{eto\-mboi\-ta}
\fmtCitation{(okw)e\-to\-mboi\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eto\-mboi\-te\-re}
\fmtSense{to complain, to grumble, to whine, to make a fuss}
\fmtExampleRL{Le\-ke\-ra\-wo okwe\-to\-mboi\-ta\-nga bu\-li kai\-re.}
\fmtExampleEN{Stop complaining all the time.}

\fmtLemma{eto\-na}
\fmtCitation{(okw)e\-to\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eto\-ne\-re}
\fmtSense{to apply makeup, to use decorative cosmetics}
\fmtExampleRL{Abaa\-la aba\-to ba\-ta\-ka mu\-no okwe\-to\-na.}
\fmtExampleEN{Girls love to do makeup.}

\fmtLemma{eto\-nda}
\fmtCrossreference{\textnormal{see also} sengeree\-rya}
\fmtCitation{(okw)e\-to\-nda}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eto\-nde\-re}
\fmtSense{to apologise, to make an apology, to ask forgiveness, to ask for pardon}
\fmtExampleRL{Omu\-ntu owa nge\-so ezi\-sai yee\-to\-nda.}
\fmtExampleEN{A person with good manners apologises.}

\fmtLemma{eto\-nga}
\fmtCitation{(okw)e\-to\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eto\-nge\-re}
\fmtSense{to take upon oneself, to motivate oneself, to find motivation for}
\fmtExampleRL{Ndi\-na ei\-ba\-nza, nkwe\-to\-nga nje\-na ka\-ti oku\-li\-sa\-su\-la.}
\fmtExampleEN{I have a debt, I take it upon myself to pay it.}

\fmtLemma{eto\-ngo\-la}
\fmtCitation{(okw)e\-to\-ngo\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eto\-ngoi\-re}
\fmtSense{to be self-reliant, to be independent}
\fmtExampleRL{Wee\-to\-ngo\-le nk’o\-mu\-sai\-za.}
\fmtExampleEN{Be self-reliant as a man.}

\fmtLemma{eto\-ngo\-lo\-la}
\fmtCitation{(okw)e\-to\-ngo\-lo\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eto\-ngo\-loi\-re}
\fmtSense{to complain, to grumble, to whine, to make a fuss}
\fmtExampleRL{Le\-ke\-ra\-wo okwe\-to\-ngo\-lo\-la\-nga bu\-li kai\-re.}
\fmtExampleEN{Stop complaining all the time.}

\fmtLemma{etoo\-la\-too\-la}
\fmtCitation{(okw)e\-too\-la\-too\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{etoi\-re\-toi\-re}
\fmtSense{to engage in petty theft}
\fmtExampleRL{Lwa\-ki wee\-too\-la\-too\-la?}
\fmtExampleEN{Why do you engage in petty theft?}

\fmtLemma{etoo\-la\-wo}
\fmtCitation{(okw)e\-too\-la\-wo}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{etoi\-re\-wo}
\fmtSense{to go away, to leave, to depart}
\fmtExampleRL{Nje nee\-toi\-re\-wo.}
\fmtExampleEN{I have left.}

\fmtLemma{etoo\-loo\-la}
\fmtCitation{(okw)e\-too\-loo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{etoo\-loi\-re}
\fmtSense{to move round, to do circular movement around, to surround, to encircle}
\fmtExampleRL{Ata\-ndi\-ki\-re okwe\-to\-loo\-la enyu\-mba.}
\fmtExampleEN{He started going round the house.}

\fmtLemma{etoo\-loo\-li}
\fmtCitation{\textnormal{(e.g.}\,gwe\-too\-loo\-li, nee\-too\-loo\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSense{round, circular}
\fmtExampleRL{Omu\-pii\-ra gwa mo\-to\-ka gwe\-too\-loo\-li.}
\fmtExampleEN{The car tyre is round.}

\fmtLemma{etu\-ku\-ta}
\fmtCitation{(okw)e\-tu\-ku\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{etu\-ku\-ti\-re}
\fmtSense{(of blood) to clot, to get bruised, to become swollen}
\fmtExampleRL{We ba\-mu\-ku\-bbi\-re endo\-bo wee\-tu\-ku\-ti\-re.}
\fmtExampleEN{Blood has clotted at the spot where he was kicked.}

\fmtLemma{etu\-ma}
\fmtCitation{(okw)e\-tu\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{etu\-mi\-re}
\fmtSense{to be self-driven, to be motivated}
\fmtExampleRL{Ye\-ga okwe\-tu\-ma.}
\fmtExampleEN{Learn to be self-driven.}

\fmtLemma{etu\-mbya}
\fmtCitation{(okw)e\-tu\-mbya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{etu\-mbi\-rye}
\fmtSense{to hold oneself in very high regard, to think highly of oneself}
\fmtExampleRL{OMu\-bii\-to yee\-tu\-mbya mu\-no.}
\fmtExampleEN{Mubiito thinks highly of himself.}

\fmtLemma{etu\-mi\-nkii\-ri}
\fmtCitation{\textnormal{(e.g.}\,mwe\-tu\-mi\-nkii\-ri, be\-tu\-mi\-ntii\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSense{acting without permission, acting without authority}
\fmtExampleRL{Oli\-re\-ka\-wo di oku\-bba omwe\-tu\-mi\-nkii\-ri?}
\fmtExampleEN{When will you ever stop acting without permission?}

\fmtLemma{etu\-mi\-nkii\-rya}
\fmtCitation{(okw)e\-tu\-mi\-nkii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{etu\-mi\-nki\-rii\-rye}
\fmtSense{to act without permission or without authority}
\fmtExampleRL{Omu\-ntu eyee\-tu\-mi\-nkii\-rya asii\-sa ebi\-ntu.}
\fmtExampleEN{A person who acts without authority spoils things.}

\fmtLemma{etye\-be\-ka}
\fmtCitation{(okw)e\-tye\-be\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{etye\-be\-ke\-re}
\fmtSense{to sit in someone's chair without permission, to sit down (esp. in an inappropriate place)}
\fmtExampleRL{Lwa\-ki okwe\-tye\-be\-ka mu nte\-be ya isa\-ma\-ka?}
\fmtExampleEN{Why do you sit in the chair of the head of the family?}

\fmtLemma{etye\-tya}
\fmtCitation{(okw)e\-tye\-tya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{etye\-te\-rye}
\fmtSense{to clean one's anus by rubbing it against the ground}
\fmtExampleRL{Abaa\-na ba mu kya\-lo bee\-tye\-tya nga ba\-ru\-gi\-re mu byo\-loo\-ni.}
\fmtExampleEN{Children in the village clean their anus by rubbing it against the ground after visiting the toilet.}

\fmtLemma{ewa}
\fmtPOS{prep}
\fmtSense{to, from or at the house or place of (a person)}
\fmtExampleRL{Ya\-ba ewa Se\-z ontoo\-lye\-yo empa\-sa.}
\fmtExampleEN{Go to Sez's place and get me an axe.}

\fmtLemma{ewai\-mya}
\fmtCitation{(okw)e\-wai\-mya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ewai\-mi\-rye}
\fmtSense{to take power by force}
\fmtExampleRL{Ami\-n yee\-wai\-mi\-rye n’a\-foo\-ka opu\-re\-zi\-de\-nti.}
\fmtExampleEN{Amin took power by force and became president.}

\fmtLemma{ewai\-sya}
\fmtCitation{(okw)e\-wai\-sya}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{ewai\-si\-rye}
\fmtSubentry(a)
\fmtSense{to attempt, to try}
\fmtExampleRL{Lwa\-ki wee\-wai\-si\-rya oku by’o\-taa\-so\-bo\-le?}
\fmtExampleEN{Why do you attempt what you cannot do?}
\fmtSubentry(b)
\fmtSense{to be arrogant, to be full of oneself}
\fmtExampleRL{Bai\-ra\-ba mu\-sai\-za we wee\-wai\-sya mu\-no.}
\fmtExampleEN{You are a very arrogant man.}
\fmtSubentry(c)
\fmtSense{to boast, to brag, to swagger}
\fmtExampleRL{Omu\-sai\-za oyo yee\-wai\-sya olwa nte z’a\-li\-na.}
\fmtExampleEN{That man boasts because of the cattle he has.}

\fmtLemma{ewa\-la}
\fmtCitation{(okw)e\-wa\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ewaa\-re}
\fmtSense{to avoid, to keep away from, to stay away from}
\fmtExampleRL{Wee\-wa\-le okwa\-raa\-ta\-na.}
\fmtExampleEN{Avoid exchanging harsh words.}

\fmtLemma{ewa\-nii\-kii\-rya}
\fmtCitation{(okw)e\-wa\-nii\-kii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ewa\-nii\-ki\-rii\-rye}
\fmtSense{to attempt something beyond one's capacities, to be ostentatious, to pretend to be important}
\fmtExampleRL{Lwa\-ki aba\-ntu bee\-wa\-nii\-kii\-rya?}
\fmtExampleEN{Why do people pretend to be important?}

\fmtLemma{ewa\-ni\-ka}
\fmtCitation{(okw)e\-wa\-ni\-ka}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{ewa\-ni\-ki\-re}
\fmtSubentry(a)
\fmtSense{(of the sky) to be clear}
\fmtExampleRL{Ei\-gu\-ru lye\-wa\-ni\-ki\-re.}
\fmtExampleEN{The sky is clear.}
\fmtSubentry(b)
\fmtSense{to attempt something beyond one's capacities, to be ostentatious, to pretend to be important}
\fmtExampleRL{Lwa\-ki wee\-wa\-ni\-ka oku ba\-ntu?}
\fmtExampleEN{Why do you pretend to be important?}

\fmtLemma{ewee\-ka}
\fmtCrossreference{\textnormal{see also} ebaga\-la}
\fmtCitation{(okw)e\-wee\-ka}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{ewee\-ke\-re}
\fmtSubentry(a)
\fmtSense{to ride (an animal)}
\fmtExampleRL{Omu\-gai\-ga yee\-we\-ke\-re oku mba\-laa\-si.}
\fmtExampleEN{The rich man is riding a horse.}
\fmtSubentry(b)
\fmtSense{to get on one's back, to be on one's back}
\fmtExampleRL{Omwa\-na yee\-wee\-ke\-re oku mu\-go\-ngo gwa nyi\-na.}
\fmtExampleEN{The child is on her mother's back.}

\fmtLemma{ewee\-rya}
\fmtCitation{(okw)e\-wee\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ewei\-rye}
\fmtSubentry(a)
\fmtSense{to come stealthily, to sneak in}
\fmtExampleRL{Yai\-ki\-re ani ng’a\-kwe\-wee\-rya nga ta\-kwe\-ndya abai\-ngi ba\-mu\-bo\-ne.}
\fmtExampleEN{He arrived here stealthily because he didn't want many people to see him.}
\fmtSubentry(b)
\fmtSense{to act shyly, to behave in a timid manner}
\fmtExampleRL{Na\-mu\-sa\-ngi\-re omu iduu\-ka lya\-mwe n’a\-ta\-ndi\-ka okwe\-wee\-rya.}
\fmtExampleEN{He acted shy when I met him in his shop.}

\fmtLemma{ewee\-rya\-mu}
\fmtCitation{(okw)e\-wee\-rya\-mu}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ewai\-re\-mu}
\fmtSense{to drink}
\fmtExampleRL{Esaa\-wa ni zai\-ki\-re ei\-ku\-mi n’ei\-bi\-ri ni nee\-wee\-rya\-mu obu\-fi\-ri\-fi\-ri bu\-bi\-ri.}
\fmtExampleEN{When it was 6 p.m. I drank two sachets of local gin.}

\fmtLemma{ewee\-rya\-yo}
\fmtCitation{(okw)e\-wee\-rya\-yo}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ewei\-rya\-yo}
\fmtSense{to dedicate oneself, to sacrifice oneself}
\fmtExampleRL{Twe\-wee\-rye\-yo oku\-wee\-re\-za oKa\-nca.}
\fmtExampleEN{Let's dedicate ourselves to serving God.}

\fmtLemma{ewe\-nye\-ka}
\fmtCitation{(okw)e\-we\-nye\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ewe\-nye\-ke\-re}
\fmtSense{to bend, to be bent}
\fmtExampleRL{Omu\-ta\-yi\-mbwa gwe\-we\-nye\-ke\-re.}
\fmtExampleEN{The iron rod is bent.}

\fmtLemma{ewe\-re\-nkee\-rya}
\fmtCitation{(okw)e\-we\-re\-nkee\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ewe\-re\-nke\-rei\-rye}
\fmtSense{to go far, to settle far away}
\fmtExampleRL{OMa\-ga\-mbo yee\-we\-re\-nke\-rei\-rye omu kya\-lo.}
\fmtExampleEN{Magambo went far away from the village.}

\fmtLemma{ewe\-re\-ra}
\fmtCitation{(okw)e\-we\-re\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ewe\-rei\-re}
\fmtSense{to threaten}
\fmtExampleRL{Lwa\-ki oKa\-ba\-ga\-mbe aku\-nee\-we\-re\-ra oku\-nyi\-ta?}
\fmtExampleEN{Why is Kabagambe threatening to kill me?}

\fmtLemma{ewo\-go\-ma}
\fmtCitation{(okw)e\-wo\-go\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ewo\-go\-me\-re}
\fmtSubentry(a)
\fmtSense{to rest, to take shelter}
\fmtExampleRL{Nje nfu\-ni\-re awa kwe\-wo\-go\-ma.}
\fmtExampleEN{I have gotten a place to rest.}
\fmtSubentry(b)
\fmtSense{to take refuge, to find refuge}
\fmtExampleRL{Aba\-swa\-gi ba bu\-wu\-da\-mo bee\-wo\-go\-me\-re omu Uga\-nda.}
\fmtExampleEN{The refugees have found refuge in Uganda.}

\fmtLemma{ewu\-gu\-la}
\fmtCitation{(okw)e\-wu\-gu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ewu\-gwi\-re}
\fmtSense{to run away, to flee, to leave (in anticipation of a problem)}
\fmtExampleRL{Wee\-wu\-gu\-le\-ku aba\-so\-roo\-lya ba mu\-so\-lo mba\-bo.}
\fmtExampleEN{Run away, the tax collectors are coming.}

\fmtLemma{ewu\-li}
\fmtCitation{\textnormal{(e.g.}\,mwe\-wu\-li, bee\-wu\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSense{proud}
\fmtExampleRL{Ti\-mbo\-na\-nga mu\-tu\-ndi wa duu\-ka mwe\-wu\-li nka Mw. Mu\-so\-ke.}
\fmtExampleEN{I have never seen such a proud shopkeeper as Musoke.}

\fmtLemma{ewu\-lu\-ngi}
\fmtCrossreference{\textnormal{see} ebulungi}

\fmtLemma{ewu\-nda}
\fmtCrossreference{\textnormal{see also} etona}
\fmtCitation{(okw)e\-wu\-nda}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ewu\-ndi\-re}
\fmtSense{to apply makeup, to use decorative cosmetics}
\fmtExampleRL{Abaa\-la aba\-to ba\-ta\-ka mu\-no okwe\-wu\-nda.}
\fmtExampleEN{Girls love to apply makeup.}

\fmtLemma{ewu\-ngu\-la}
\fmtCitation{(okw)e\-wu\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ewu\-ngwi\-re}
\fmtSense{to step out, to exit, to go out}
\fmtExampleRL{Ye\-wu\-ngwi\-re\-mu ka\-doo\-li bai\-tu akwi\-za.}
\fmtExampleEN{He has stepped out briefly, he is coming back.}

\fmtLemma{ewu\-ngu\-lya}
\fmtCitation{(okw)e\-wu\-ngu\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ewu\-ngwi\-rye}
\fmtSense{to move around (searching for something), to move up and down}
\fmtExampleRL{Obwai\-rai akwe\-wu\-ngu\-lya ani.}
\fmtExampleEN{He has been moving around for some time.}

\fmtLemma{ewuu\-nya}
\fmtCitation{(okw)e\-wuu\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ewuu\-nyi\-re}
\fmtSense{to be surprised, to be puzzled, to wonder}
\fmtExampleRL{Nkwe\-wuu\-nyi\-re mu\-no oku\-bba nga to\-mai\-te Kya\-gwi\-re.}
\fmtExampleEN{I am surprised that you don’t know Kyagwire.}

\fmtLemma{ewuu\-ra}
\fmtCitation{(okw)e\-wuu\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ewu\-lii\-re}
\fmtSense{to be proud, to think highly of oneself}
\fmtExampleRL{ONa\-mu\-ka\-sa yee\-wuu\-ra mu\-no.}
\fmtExampleEN{Namukasa is very proud.}

\fmtLemma{eya}%
\fmtOne{1}
\fmtCitation{(okw)e\-ya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eye\-re}
\fmtSense{(of the moon) to be full}
\fmtExampleRL{Okwe\-zi ku\-kwe\-ya.}
\fmtExampleEN{It is now the full moon.}

\fmtLemma{eya}%
\fmtOne{2}
\fmtCitation{(okw)e\-ya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eye\-re}
\fmtSense{to sweep}
\fmtExampleRL{Ye\-ya eki\-sii\-ka kya\-mu.}
\fmtExampleEN{Sweep your room.}

\fmtLemma{eya\-bi\-ra}
\fmtCitation{(okw)e\-ya\-bi\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-bii\-re}
\fmtSense{to leave, to go away}
\fmtExampleRL{Ni mwa\-bba mu\-kya\-li ani, nje nkwe\-ya\-bi\-ra.}
\fmtExampleEN{If you are still here, I am leaving.}

\fmtLemma{eya\-bi\-sya}
\fmtCitation{(okw)e\-ya\-bi\-sya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-bi\-si\-rye}
\fmtSubentry(a)
\fmtSense{to gate-crash, to come to a party uninvited, to crash a party}
\fmtExampleRL{Wee\-ya\-bi\-sya otyai oku bu\-ge\-nyi gye ba\-taa\-kwe\-te\-re?}
\fmtExampleEN{How could you go to the party uninvited?}
\fmtSubentry(b)
\fmtSense{to go somewhere with an intention to spy}
\fmtExampleRL{Ye\-ya\-bi\-si\-rye oku bu\-ge\-nyi.}
\fmtExampleEN{He went to the party with the intention to spy.}

\fmtLemma{eya\-ga\-la}
\fmtCitation{(okw)e\-ya\-ga\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-gai\-re}
\fmtSense{to be happy, to enjoy oneself}
\fmtExampleRL{Aba\-ntu ba mu Ibbaa\-le bee\-ya\-ga\-la.}
\fmtExampleEN{The people of Ibbaale are happy.}

\fmtLemma{eya\-kaa\-ka\-nia}
\fmtCitation{(okw)e\-ya\-kaa\-ka\-nia}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-kaa\-kai\-nie}
\fmtSense{to be pompous, to be egoistic}
\fmtExampleRL{Lwa\-ki okwe\-ya\-kaa\-ka\-nia ebi\-na\-mbu\-la\-mu?}
\fmtExampleEN{Why are you so pompous over such trivial matters?}

\fmtLemma{eya\-ka\-lii\-rya}
\fmtCitation{(okw)e\-ya\-ka\-lii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-ka\-li\-rii\-rye}
\fmtSense{to act without permission}
\fmtExampleRL{Wee\-ya\-ka\-li\-rii\-rye n’o\-ya\-ba\-yo.}
\fmtExampleEN{You acted without my permission and went there.}

\fmtLemma{eya\-ka\-li\-rii\-rya}
\fmtCitation{(okw)e\-ya\-ka\-lii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-ka\-li\-rii\-rye}
\fmtSubentry(a)
\fmtSense{to be complicated, to become unbearable}
\fmtExampleRL{Embee\-ra eni eya\-ka\-lii\-rii\-rye, bai\-tu we oko\-la otyai oku\-bbaa\-wo?.}
\fmtExampleEN{This situation is unbearable, what do you do to survive?}
\fmtSubentry(b)
\fmtSense{to pretend to be more important}

\fmtLemma{eya\-ka\-lya}
\fmtCitation{(okw)e\-ya\-ka\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-kai\-rye}
\fmtSense{to behave, to act}
\fmtExampleRL{Abai\-bbi bee\-ya\-ka\-lya ku\-bbi.}
\fmtExampleEN{Thieves behave badly.}

\fmtLemma{eya\-ma}
\fmtCitation{(okw)e\-ya\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eya\-mi\-re}
\fmtSense{to vow, to swear, to promise, to give one's word}
\fmtExampleRL{Nee\-ya\-mi\-re oku\-gu\-la ente\-be ya Ka\-ni\-sa.}
\fmtExampleEN{I pledged to buy a chair for the church.}

\fmtLemma{eya\-mba}
\fmtCitation{(okw)e\-ya\-mba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-mbi\-re}
\fmtSubentry(a)
\fmtSense{to help oneself}
\fmtExampleRL{Mwe\-ya\-mbe ku\-bba ndoo\-wo ya\-ba\-ya\-mba.}
\fmtExampleEN{Help yourself because no one is willing to help you.}
\fmtSubentry(b)
\fmtSense{to defecate, to poo}
\fmtExampleRL{Nkwa\-bba kwe\-ya\-mba.}
\fmtExampleEN{I am going to defecate.}

\fmtLemma{eya\-mbu\-la}
\fmtCitation{(okw)e\-ya\-mbu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-mbwi\-re}
\fmtSense{to undress, to strip (off)}
\fmtExampleRL{OLu\-ka akwe\-ya\-mbuu\-ra olu\-go\-ye omu ba\-ntu.}
\fmtExampleEN{Luke is undressing in public.}

\fmtLemma{eya\-mpai\-re\-we}
\fmtPOS{interj}
\fmtSense{my goodness (expressing surprise, anger)}
\fmtExampleRL{Eya\-mpai\-re\-we, n’a\-mboi\-ne ai\-ru\-ki\-re\-mbe.}
\fmtExampleEN{My goodness, he simply ran away on seeing me.}

\fmtLemma{eya\-nza}
\fmtCitation{(okw)e\-ya\-nza}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-nzi\-re}
\fmtSense{to be thankful, to be grateful}
\fmtExampleRL{Nee\-ya\-nzi\-re mu\-no olwa bi\-ntu by’o\-mpai\-re.}
\fmtExampleEN{I am very thankful for what you have given me.}

\fmtLemma{eya\-nza\-la}
\fmtCrossreference{\textnormal{see also} ebamba}
\fmtCitation{(okw)e\-ya\-nza\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-nzai\-re}
\fmtSense{to sit with legs spread apart}
\fmtExampleRL{Omu\-ka\-li oyo yee\-ya\-nzai\-re.}
\fmtExampleEN{That woman is sitting with her legs spread apart.}

\fmtLemma{eya\-sa\-mu\-lya}
\fmtCitation{(okw)e\-ya\-sa\-mu\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-sa\-mwi\-rye}
\fmtSubentry(a)
\fmtSense{to sneeze}
\fmtExampleRL{Omwa\-na yee\-ya\-sa\-mwi\-rye alwai\-re eki\-yi\-nzi.}
\fmtExampleEN{The child is sneezing because of the flu.}
\fmtSubentry(b)
\fmtSense{to quarrel, to argue}
\fmtExampleRL{OWa\-nzi\-ge n’o\-mu\-ka\-li waa\-mwe bee\-ya\-sa\-mu\-lya lu\-koo\-ko.}
\fmtExampleEN{Wanzige occasionally quarrels with his wife.}

\fmtLemma{eya\-ta\-yi\-mu}
\fmtCitation{eya\-ta\-yi\-mu}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{airtime, credit which allows one to use a mobile phone network}
\fmtExampleRL{Eya\-ta\-yi\-mu awei\-re\-ku oku si\-mu ya\-nge.}
\fmtExampleEN{My phone is short of airtime.}

\fmtLemma{eya\-tya}
\fmtCitation{(okw)e\-ya\-tya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-ti\-rye}
\fmtSubentry(a)
\fmtSense{to crack, to fracture}
\fmtExampleRL{Omu\-si\-si ate\-nge\-te\-rye n’a\-lee\-te\-ra enyu\-mba okwe\-ya\-tya.}
\fmtExampleEN{The earthquake has caused the house to crack.}
\fmtSubentry(b)
\fmtFixedexpression{kwe\-ya\-tya kwe ita\-ka\-li}
\fmtSense{landslide}
\fmtExampleRL{Okwe\-ya\-tya kwa ita\-ka\-li ku\-ta\-ndi\-ki\-re e Bu\-duu\-da.}
\fmtExampleEN{Landslides have started in Buduuda.}

\fmtLemma{eya\-tya\-mu}
\fmtCitation{(okw)e\-ya\-tya\-mu}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-ti\-rye\-mu}
\fmtSense{to be divided, to disunite}
\fmtExampleRL{Iswe Aba\-pii\-na ti\-tu\-kwi\-za kwe\-ya\-tya\-mu.}
\fmtExampleEN{We the Bapiima will not be divided.}

\fmtLemma{eya\-ya}
\fmtCitation{(okw)e\-ya\-ya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-yi\-re}
\fmtSense{to boast, to brag, to swagger}
\fmtExampleRL{Lwa\-ki okwe\-ya\-ya otyo?}
\fmtExampleEN{Why are you boasting like that?}

\fmtLemma{eya\-ya\-mu\-lya}
\fmtCitation{(okw)e\-ya\-ya\-mu\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eya\-ya\-mwi\-rye}
\fmtSense{to yawn, to yawn noisily and stretch oneself}
\fmtExampleRL{Kwa\-ta oku mu\-nwa nga okwe\-ya\-ya\-mu\-lya.}
\fmtExampleEN{Cover your mouth when yawning.}

\fmtLemma{eya\-yi}
\fmtCitation{\textnormal{(e.g.}\,mwe\-ya\-yi, bee\-ya\-yi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{one who is determined to cause harm}
\fmtExampleRL{Omu\-sai\-za omwe\-ya\-yi ate\-me\-re omwa\-na.}
\fmtExampleEN{A man determined to cause harm has killed the child.}

\fmtLemma{eya\-yi\-ra}
\fmtCitation{(okw)e\-ya\-yi\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eya\-yii\-re}
\fmtSense{to threaten}
\fmtExampleRL{Amwe\-ya\-yii\-re oku\-mwi\-ta.}
\fmtExampleEN{He threatened to kill him.}

\fmtLemma{eye\-bee\-ra}
\fmtCitation{(okw)e\-ye\-bee\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eye\-bee\-rei\-re}
\fmtSense{to forget, to leave behind, to fail to take}
\fmtExampleRL{Wee\-ye\-bee\-rei\-re eka\-laa\-mu yaa\-mu.}
\fmtExampleEN{You have forgotten your pen.}

\fmtLemma{eye\-bee\-ri}
\fmtCitation{\textnormal{(e.g.}\,mwe\-ye\-bee\-ri, bee\-ye\-bee\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSense{forgetful, absentminded}
\fmtExampleRL{Ozei\-za waa\-mu ya\-foo\-ke\-re mwe\-ye\-bee\-ri.}
\fmtExampleEN{Your grandfather has become forgetful.}

\fmtLemma{eye\-bu\-ki\-ra}
\fmtCitation{(okw)e\-ye\-bu\-ki\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eye\-bu\-kii\-re}
\fmtSense{to remember}
\fmtExampleRL{Amaa\-re ni yee\-ye\-bu\-ki\-ra oku\-su\-mba eme\-re.}
\fmtExampleEN{He finally remembered to cook food.}

\fmtLemma{eye\-mee\-ra}
\fmtCitation{(okw)e\-ye\-mee\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eye\-mee\-rei\-re}
\fmtSubentry(a)
\fmtSense{to represent, to stand in for}
\fmtExampleRL{Naa\-ni ye yee\-ye\-mee\-rei\-re oTi\-to?}
\fmtExampleEN{Who represented Titus?}
\fmtSubentry(b)
\fmtSense{to stand surety for, to stand bail for}
\fmtExampleRL{Mwa\-be mwe\-ye\-mee\-re omwi\-ra waa\-nywe azwe omu nko\-myo.}
\fmtExampleEN{Go and stand surety for your friend so that he can be release from the prison.}

\fmtLemma{eye\-nde\-sya}
\fmtCrossreference{\textnormal{see also} etaki\-sya}
\fmtCitation{(okw)e\-ye\-nde\-sya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eye\-nde\-se\-rye}
\fmtSense{to wish for oneself}
\fmtExampleRL{OWa\-nza\-la yee\-ye\-nde\-sya ye\-nkai ebi\-ntu ebi\-sai.}
\fmtExampleEN{Wanzala wishes for himself all the good things.}

\fmtLemma{eye\-rya}
\fmtCitation{(okw)e\-ye\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eye\-rye}
\fmtSubentry(a)
\fmtSense{to tell the truth, to be frank about something unpleasant}
\fmtExampleRL{Ya\-ye\-ye\-rye omu\-ka\-ma waa\-mwe.}
\fmtExampleEN{He told his boss the naked truth.}
\fmtSubentry(b)
\fmtSense{to repent}
\fmtExampleRL{Wee\-ye\-rye omu mai\-so ga Ka\-nca aku\-ga\-nyi\-re.}
\fmtExampleEN{Repent to God to receive his forgiveness.}

\fmtLemma{eye\-ta}
\fmtCitation{(okw)e\-ye\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eye\-te\-re}
\fmtSense{to identify oneself, to call oneself, to consider oneself}
\fmtExampleRL{Aba\-ga\-nda baa\-tu\-gai\-ne okwe\-ye\-ta ama\-ba\-ra aga Ba\-nya\-la.}
\fmtExampleEN{The Baganda forbade us from identifying ourselves by Nyala names.}

\fmtLemma{eyii\-ya}
\fmtCitation{(okw)e\-yii\-ya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eyii\-yi\-re}
\fmtSense{to work for oneself}
\fmtExampleRL{Omwo\-jo oyo eki\-ba\-nja kye ya\-guu\-re, ye\-yii\-yi\-re ye\-nkai.}
\fmtExampleEN{The land which that boy bought was earned by his own hard work.}

\fmtLemma{eyi\-mbya}
\fmtCitation{(okw)e\-yi\-mbya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eyi\-mbi\-rye}
\fmtSense{to be united}
\fmtExampleRL{Aba\-ruu\-li twe\-yi\-mbye twi\-mu\-le oBu\-ruu\-li.}
\fmtExampleEN{Let us, the Baruuli, be united and develop the Buruuli Kingdom.}

\fmtLemma{eyi\-ndu\-la}
\fmtCitation{(okw)e\-yi\-ndu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eyi\-ndwi\-re}
\fmtSubentry(a)
\fmtSense{to change (e.g. colour or appearance), to bleach, to get bleaching}
\fmtExampleRL{Omu\-ka\-li eyaa\-li omwi\-ra\-gu\-zu yee\-yi\-ndwi\-re n’a\-foo\-ka mwe\-ru.}
\fmtExampleEN{The woman who was dark-skinned bleached herself and became brown.}
\fmtSubentry(b)
\fmtSense{to change one's mind, to change one's attitude}
\fmtExampleRL{Naa\-bi\-re oku\-mu\-twa\-ra empii\-ya ze twa\-la\-gai\-ne, bai\-tu n'a\-ya\-ji\-rye yee\-yi\-ndwi\-re.}
\fmtExampleEN{I went to take to him the money we agreed upon, but I found that he had changed his mind.}
\fmtSubentry(c)
\fmtSense{(of weather) to change}
\fmtExampleRL{Obwi\-re bwe\-yi\-ndwi\-re oi\-ke\-ndi ali ampi oku\-too\-nya.}
\fmtExampleEN{The weather has changed, it is about to rain.}
\fmtSubentry(d)
\fmtSense{to turn against}
\fmtExampleRL{Wee\-yi\-ndu\-lii\-re abai\-ra baa\-mu.}
\fmtExampleEN{You turned against your friends.}

\fmtLemma{eyi\-nu\-la}
\fmtCitation{(okw)e\-yi\-nu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eyi\-nwi\-re}
\fmtSense{to feel complacent, to be smug, to be conceited, to be self-satisfied}
\fmtExampleRL{Le\-ka kwe\-yi\-nu\-la na bu\-se\-nte bw’o\-fu\-ni\-re, bu\-kwi\-za ku\-waa\-wo.}
\fmtExampleEN{Do not feel complacent because of that little money you have, it will be gone soon.}

\fmtLemma{eyo}
\fmtCrossreference{\textnormal{see} o\textnormal{\fmtOne{3}}}

\fmtLemma{eyo\-koo\-ka}
\fmtCitation{(okw)e\-yo\-koo\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eyo\-kooke\-re}
\fmtSense{to retch, to gag}
\fmtExampleRL{Ja\-ne akwe\-yo\-koo\-ka lwa mu\-su\-ja.}
\fmtExampleEN{Jane is retching due to malaria.}

\fmtLemma{eyo\-mbo\-ka}
\fmtCitation{(okw)e\-yo\-mbo\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eyo\-mbo\-ke\-re}
\fmtSense{to develop oneself, to progress, to advance}
\fmtExampleRL{Aba\-ntu ba Lu\-wee\-ro ba\-ta\-ndi\-ki\-re okwe\-yo\-mbo\-ka.}
\fmtExampleEN{The people from Luweero have started progressing.}

\fmtLemma{eyo\-nge\-ra\-yo}
\fmtCitation{(okw)e\-yo\-nge\-ra\-yo}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{eyo\-ngei\-re\-yo}
\fmtSubentry(a)
\fmtSense{to increase}
\fmtExampleRL{Omu\-we\-ndo gwa baa\-na ab’o\-ku nzi\-ra gwe\-yo\-ngei\-re\-yo.}
\fmtExampleEN{The number of street children has increased.}
\fmtSubentry(b)
\fmtSense{to go ahead, to continue, to proceed}
\fmtExampleRL{Wee\-yo\-nge\-ra\-yo oya\-ka\-le oku ma\-duu\-ka.}
\fmtExampleEN{Go ahead and pass by the shops.}

\fmtLemma{eyu\-nga}
\fmtCitation{(okw)e\-yu\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eyu\-ngi\-re}
\fmtSense{to meet, to join}
\fmtExampleRL{Amwe\-yu\-ngii\-re\-ku Mu\-ko\-no.}
\fmtExampleEN{He joined him at Mukono.}

\fmtLemma{ezi\-nda}
\fmtCitation{(okw)e\-zi\-nda}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ezi\-ndi\-re}
\fmtSense{to commit suicide}
\fmtExampleRL{Omu\-sai\-za yee\-zi\-ndii\-re ku mu\-saa\-le.}
\fmtExampleEN{The man hanged himself on the tree.}

\fmtLemma{ezi\-ngii\-rya}
\fmtCrossreference{\textnormal{see also} ezi\-ngi\-rii\-rya}
\fmtCitation{(okw)e\-zi\-ngii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ezi\-ngii\-rye}
\fmtSense{to be a burden, to stay close to (a person against his will)}
\fmtExampleRL{Lwa\-ki wee\-zi\-ngii\-rya oku ba\-ntu aba\-ndi?}
\fmtExampleEN{Why are you a burden to other people?}

\fmtLemma{ezi\-ngi\-rii\-rya}
\fmtCrossreference{\textnormal{see also} ezi\-ngii\-rya}
\fmtCitation{(okw)e\-zi\-ngi\-rii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ezi\-ngi\-rii\-rye, ezi\-ngii\-rye}
\fmtSense{to be a burden, to stay close to (a person against his will)}

\fmtLemma{ezi\-ri\-ba\-nga}
\fmtCitation{(okw)e\-zi\-ri\-ba\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ezi\-ri\-ba\-ngi\-re}
\fmtSense{to writhe in pain, to cry in pain}
\fmtExampleRL{Aku\-ngi\-re ni yee\-zi\-ri\-ba\-nga.}
\fmtExampleEN{He wailed and writhed in pain on the ground.}

\fmtLemma{ezi\-ri\-ngii\-rya}
\fmtCitation{(okw)e\-zi\-ri\-ngii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ezi\-ri\-ngi\-rii\-rye}
\fmtSense{to wind around, to wrap oneself around, to coil oneself around}
\fmtExampleRL{Ombu\-li yee\-zi\-ri\-ngi\-rii\-rye oku mu\-saa\-le.}
\fmtExampleEN{A goat has wound itself around a tree with a rope.}

\fmtLemma{ezu\-ka\-nya}
\fmtCitation{(okw)e\-zu\-ka\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ezu\-nkai\-nye}
\fmtSense{to recall, to remember, to revise}
\fmtExampleRL{Mwei\-zu\-ka\-nye\-nge bye\-mu\-wa\-ndii\-ki\-re.}
\fmtExampleEN{Remember what you have written.}

\fmtLemma{ezwa\-mu}
\fmtCitation{(okw)e\-zwa\-mu}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ezwi\-re}
\fmtSense{to decide, to make up one's mind}
\fmtExampleRL{Ye\-zwi\-re\-mu n'a\-mu\-ku\-bba nga kya\-ka\-li\-rii\-rye.}
\fmtExampleEN{It became too much, and he decided to beat him.}

\end{letter}