\begin{letter}{I}

\fmtLemma{iba}
\fmtCitation{(o)i\-ba}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{husband}
\fmtExampleRL{Oi\-ba waa\-mwe mu\-sai\-za mu\-koo\-to.}
\fmtExampleEN{Her husband is a big man.}

\fmtLemma{ibaa\-gi\-ro}
\fmtCitation{(e)i\-baa\-gi\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{abattoir, slaughterhouse}
\fmtExampleRL{Enya\-ma ngi\-guu\-ra mu ibaa\-gi\-ro.}
\fmtExampleEN{I buy meat from the abattoir.}

\fmtLemma{ibaa\-ti}
\fmtCitation{(e)i\-baa\-ti}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSource{from Ar.\,via Sw.}
\fmtSense{iron sheet}
\fmtExampleRL{Eki\-ka kya ibaa\-ti ki\-ni ki\-wa\-ngaa\-li.}
\fmtExampleEN{This type of iron sheet is long lasting.}

\fmtLemma{ibai\-zi\-ro}
\fmtCitation{(e)i\-bai\-zi\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{carpentry workshop}
\fmtExampleRL{Eki\-ko\-nko\-ro nki\-guu\-re mu ibai\-zi\-ro.}
\fmtExampleEN{I bought the stool in the carpentry workshop.}

\fmtLemma{iba\-la}
\fmtCitation{(e)i\-ba\-la}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{stain, spot, smudge}
\fmtExampleRL{Olu\-go\-ye lwa\-mwe lu\-li\-ku ei\-ba\-la eri\-ra\-gu\-zu.}
\fmtExampleEN{His garment has a black stain.}

\fmtLemma{iba\-nza}
\fmtCitation{(e)i\-ba\-nza}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{debt}
\fmtExampleRL{Oki\-mai\-te\-ku nti nta n’ei\-ba\-nza lya\-mu?}
\fmtExampleEN{Do you know that I have your debt?}

\fmtLemma{iba\-ra}
\fmtCitation{(e)i\-ba\-ra}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{name}
\fmtExampleRL{Nko\-be\-ra ei\-ba\-ra lya\-mu erya ki\-ka.}
\fmtExampleEN{Tell me your clan name.}

\fmtLemma{iba\-ti\-ri\-zo}
\fmtCitation{(e)i\-ba\-ti\-ri\-zo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{baptismal font (an article of church furniture used for baptism)}
\fmtExampleRL{Ei\-ba\-ti\-ri\-zo lya ba\-po\-le\-si\-ta\-nti li\-bba mu ka\-ni\-sa.}
\fmtExampleEN{The baptismal font is found in the church.}

\fmtLemma{ibba}
\fmtCitation{(okw)i\-bba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ibbi\-re}
\fmtSense{to steal, to rob}
\fmtExampleRL{Ai\-bba omu\-wo\-go gwa mu\-lu\-mi ampi bu\-li ki\-ro.}
\fmtExampleEN{He steals the farmer's cassava almost every day.}

\fmtLemma{ibbaa\-le}
\fmtCitation{(e)i\-bbaa\-le}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{stone, rock}
\fmtExampleRL{Omwa\-na aku\-cu\-ma ama\-bbaa\-le oku nyu\-mba.}
\fmtExampleEN{The child is throwing stones at the house.}
\fmtSubentry(b)
\fmtSense{(water) ice}
\fmtExampleRL{Amai\-zi ga\-koo\-re ama\-bbaa\-le.}
\fmtExampleEN{The water has turned into ice.}
\fmtSubentry(c)
\fmtSense{Ibbaale is a place of cultural significance where a number of shrines are found, it also has caves believed to be the habitat of the ghost called Ibbaale}

\fmtLemma{ibba\-go}
\fmtCitation{(e)i\-bba\-go}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{draft, preliminary version}
\fmtExampleRL{So\-ma\-mu omu ibba\-go lya\-nge.}
\fmtExampleEN{Read through my draft please.}
\fmtSubentry(b)
\fmtSense{bill, draft law, proposed piece of legislation, proposal}
\fmtExampleRL{Ei\-bba\-go lya itee\-ka li\-twa\-lii\-bwe omu nga\-ngo.}
\fmtExampleEN{The bill has been taken to parliament.}

\fmtLemma{ibba\-nga}
\fmtCitation{(e)i\-bba\-nga}
\fmtPOS{n}%
\fmtNC{5}
\fmtSubentry(a)
\fmtSense{sky}
\fmtExampleRL{Enye\-nyee\-zi zi\-ri mu ibba\-nga.}
\fmtExampleEN{The stars are in the sky.}
\fmtSubentry(b)
\fmtSense{air}
\fmtExampleRL{Ende\-ge eta\-mbuu\-ra mu ibba\-nga.}
\fmtExampleEN{The aeroplane flies in the air.}
\fmtSubentry(c)
\fmtSense{space}
\fmtExampleRL{Wa\-ndii\-ka\-nga nga oku\-le\-ka\-wo ama\-bba\-nga aka\-ti wa bi\-ga\-mbo.}
\fmtExampleEN{Leaving spaces between words when writing.}
\fmtSubentry(d)
\fmtSense{time, period}
\fmtExampleRL{Aba\-ruu\-li ba\-maa\-re ei\-bba\-nga lya mya\-ka iku\-mi na isa\-tu nga bee\-re\-me\-re.}
\fmtExampleEN{The Baruuli have been independent for a period of thirteen years.}

\fmtLemma{ibba\-nge}
\fmtCitation{(e)i\-bba\-nge}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{freedom, liberty}
\fmtExampleRL{OKa\-nca ni\-ye ya\-tu\-wai\-re ei\-bba\-nge lya ku\-si\-nza.}
\fmtExampleEN{It is God who gave us the freedom to worship.}

\fmtLemma{ibba\-ngo}
\fmtCitation{(e)i\-bba\-ngo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{hunchback}
\fmtExampleRL{Ei\-bba\-ngo lya mu\-ka\-li li\-ri\-ku ekyo\-ya.}
\fmtExampleEN{The woman's hunchback is infected with a wound.}
\fmtSubentry(b)
\fmtSense{(of an animal) hump}
\fmtExampleRL{Engu\-ndu zi\-bba n’a\-ma\-bba\-ngo aga\-koo\-to.}
\fmtExampleEN{Bulls usually have big humps.}

\fmtLemma{Ibba\-ngo}%
\fmtOne{1}
\fmtCitation{(o)Ibba\-ngo}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSubentry(c)
\fmtSense{royal name for Bunyala kings}
\fmtExampleRL{Ibba\-ngo iba\-ra lya mu\-ka\-ma wa Bu\-nya\-la.}
\fmtExampleEN{Ibbango is a name given to the king of Bunyala.}

\fmtLemma{ibba\-ni}
\fmtCitation{(e)i\-bba\-ni}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{breast infection, mastitis}
\fmtExampleRL{Omu\-ka\-li alwai\-re ei\-bba\-ni.}
\fmtExampleEN{The woman has mastitis.}

\fmtLemma{ibba\-no}%
\fmtOne{1}
\fmtCitation{(e)i\-bba\-no}
\fmtPOS{n}%
\fmtNC{5}
\fmtSubentry(a)
\fmtSense{misfortune, disaster}
\fmtExampleRL{Ei\-bba\-no li\-gwi\-re oku kya\-lo.}
\fmtExampleEN{A misfortune has befallen the village.}
\fmtSubentry(b)
\fmtSense{tragedy (e.g. when speaking of death)}
\fmtExampleRL{Omu\-ntu okwe\-zi\-nda li\-bba ibba\-no.}
\fmtExampleEN{It is a tragedy for one to commit suicide.}

\fmtLemma{ibba\-no}%
\fmtOne{2}
\fmtPOS{interj}
\fmtSense{used to address mourners, my condolences}
\fmtExampleRL{Ibba\-no!}
\fmtExampleEN{What a pity!}

\fmtLemma{ibbi\-ka}
\fmtCitation{(okw)i\-bbi\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ibbi\-ki\-re}
\fmtSense{to dip, to immerse}
\fmtExampleRL{Ai\-bbi\-ka eki\-ba\-ju omu ma\-lwa.}
\fmtExampleEN{He dips the calabash in the local brew.}

\fmtLemma{ibbi\-ra}
\fmtCitation{(okw)i\-bbi\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ibbii\-re}
\fmtSense{to drown}
\fmtExampleRL{Ya\-yi\-bbii\-re omu nya\-nja.}
\fmtExampleEN{He drowned in the lake.}

\fmtLemma{ibbo}
\fmtCitation{(o)i\-bbo}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{cowpea (\textit{Vigna ungui\-cu\-la\-ta}) powder (made from crushed dried cooked leaves of the cowpea plant)}
\fmtExampleRL{Oibbo wa maa\-ma awu\-mbi\-re.}
\fmtExampleEN{My mother's cowpea powder has been eaten by worms.}

\fmtLemma{ibbo\-mbo}
\fmtCitation{(e)i\-bbo\-mbo}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{\textit{Mo\-mo\-rdi\-ca foe\-ti\-da} (a perennial climbing vine)}
\fmtExampleRL{Ei\-bbo\-mbo ri\-ta\-mba eki\-koo\-ro.}
\fmtExampleEN{\textit{Mo\-mo\-rdi\-ca foetida} cures cough.}

\fmtLemma{ibbo\-ngo}
\fmtCitation{(e)i\-bbo\-ngo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{tooth gap}
\fmtExampleRL{OKya\-ku\-wa ali\-na ama\-bbo\-ngo ya\-gwi\-re amai\-no ni ga\-zwa\-mu.}
\fmtExampleEN{Kyakuwa has tooth gaps: she lost her teeth when she fell.}

\fmtLemma{ibbu\-gu\-mu}
\fmtCitation{(e)i\-bbu\-gu\-mu}
\fmtPOS{n}%
\fmtNC{5}
\fmtSubentry(a)
\fmtSense{warmth, heat, hotness}
\fmtExampleRL{Ebi\-ro bi\-ni e Na\-ka\-so\-ngo\-lo ndoo\-yo bbu\-gu\-mu.}
\fmtExampleEN{These days there is no warmth in Nakasongola.}
\fmtSubentry(b)
\fmtSense{excitement, thrill, pleasure, delight, joy}
\fmtExampleRL{Abee\-gi ba\-fu\-ni\-re ei\-bbu\-gu\-mu.}
\fmtExampleEN{The students have gotten some excitement.}

\fmtLemma{ibbu\-ka}%
\fmtOne{1}
\fmtCitation{(okw)i\-bbu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ibbu\-ki\-re}
\fmtSense{to soak}
\fmtExampleRL{Omwa\-na akwi\-bbu\-ka engo\-ye omu bbe\-nsa\-ni.}
\fmtExampleEN{The child is soaking the clothes in the washing basin.}

\fmtLemma{ibbu\-ka}%
\fmtOne{2}
\fmtPOS{interj}
\fmtSense{expression used in response to a child's sneeze, lit.\,open}
\fmtExampleRL{Omwa\-na ni\-ye ye\-syai\-mu\-la ba\-mwi\-ra\-mu nti, ibbu\-ka!}
\fmtExampleEN{When a child sneezes, people around say ``Ibbuka!".}

\fmtLemma{ibbu\-lu\-ku\-ka}
\fmtCitation{(okw)i\-bbu\-lu\-ku\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ibbu\-lu\-ku\-ki\-re}
\fmtSubentry(a)
\fmtSense{to emerge, to come into view}
\fmtExampleRL{Omu\-saa\-nyi akwi\-bbu\-lu\-ku\-ka omu mai\-zi.}
\fmtExampleEN{The swimmer is emerging from the water.}
\fmtSubentry(b)
\fmtSense{to recover financially}
\fmtExampleRL{Ya\-bbai\-re asee\-ge\-re bai\-tu ati\-bu\-ni ai\-bbu\-lu\-ku\-ki\-re.}
\fmtExampleEN{He incurred a financial loss, but he has now recovered.}

\fmtLemma{ibbu\-lu\-la}
\fmtCitation{(okw)i\-bbu\-lu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ibbu\-lwi\-re}
\fmtSense{to recover something from water}
\fmtExampleRL{Bai\-bbu\-lwi\-re obwa\-to omu mai\-zi.}
\fmtExampleEN{They recovered the boats from the water.}

\fmtLemma{ibbu\-mba}
\fmtCitation{(e)i\-bbu\-mba}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{clay}
\fmtExampleRL{Ei\-bbu\-mba tu\-li\-bu\-mbi\-sya nsu\-wa.}
\fmtExampleEN{We use the clay to mould pots.}

\fmtLemma{ibbu\-tyai}
\fmtCitation{(e)i\-bbu\-tyai}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{fermented millet flour}
\fmtExampleRL{Lee\-ta oi\-bbu\-tyai tu\-mu\-sii\-ke.}
\fmtExampleEN{Bring the fermented millet flour and let's roast it.}

\fmtLemma{ibee\-re}
\fmtCitation{(e)i\-bee\-re}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{breast}
\fmtExampleRL{Omwa\-la amee\-re ama\-bee\-re.}
\fmtExampleEN{The girl has grown breasts.}

\fmtLemma{ibe\-ga}
\fmtCitation{(e)i\-be\-ga}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{back}
\fmtExampleRL{Ta omwa\-na oku ibe\-ga.}
\fmtExampleEN{Put the child on your back.}
\fmtSubentry(b)
\fmtSense{shoulder}
\fmtExampleRL{Omu\-gu\-gu aku\-gu\-twa\-ra ku ibe\-ga.}
\fmtExampleEN{He carries the luggage on his shoulder.}

\fmtLemma{ibii\-ki\-ro}
\fmtCitation{(e)i\-bii\-ki\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{store, storeroom, storehouse}
\fmtExampleRL{Ebi\-yi\-mba bi\-bii\-ke omu ibii\-ki\-ro.}
\fmtExampleEN{Keep the beans in the storeroom.}
\fmtSubentry(b)
\fmtSense{treasury}
\fmtExampleRL{Empii\-ya mu\-zi\-te omu ibii\-ki\-ro.}
\fmtExampleEN{Put the money in the treasury.}
\fmtSubentry(c)
\fmtSense{archive}
\fmtExampleRL{E Nsa\-mbya wa\-li\-yo ei\-bii\-ki\-ro eri\-koo\-to.}
\fmtExampleEN{There is a big archive at Nsambya.}

\fmtLemma{ibi\-ra}
\fmtCitation{(e)i\-bi\-ra}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{hook, fishhook, barb}
\fmtExampleRL{Ei\-bi\-ra li\-kwa\-ta mpuu\-ta.}
\fmtExampleEN{The hook is used to catch the Nile perch.}

\fmtLemma{ibi\-ri}
\fmtCrossreference{used in isolation and with class 9 nouns, see (-)biri}
\fmtPOS{num}
\fmtSense{two}
\fmtExampleRL{Tu\-li ku na\-mba abi\-ri na ibi\-ri.}
\fmtExampleEN{We are number twenty two.}

\fmtLemma{ibu\-ba}
\fmtCitation{(e)i\-bu\-ba}
\fmtPOS{n}%
\fmtNC{5}
\fmtSubentry(a)
\fmtSense{benign tumour, painless tumour}
\fmtExampleRL{Mbai\-re n’ei\-bu\-ba li\-ni oku\-zwa ni na\-bya\-lii\-bwe.}
\fmtExampleEN{I have had this benign tumour since I was born.}
\fmtSubentry(b)
\fmtSense{state of being possessive}
\fmtExampleRL{Aba\-sai\-za ba\-ta n’ei\-bu\-ba oku ba\-ka\-li baa\-bwe.}
\fmtExampleEN{Men are very possessive of their wives.}

\fmtLemma{ibu\-ga}
\fmtCitation{(e)i\-bu\-ga}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{large gourd}
\fmtExampleRL{Ntee\-ra omwe\-nge gwa\-nge omu ibu\-ga.}
\fmtExampleEN{Put my alcohol into the large gourd.}

\fmtLemma{ibu\-la}
\fmtCrossreference{\textnormal{see} ibura}
\fmtDialect{Lunyala}

\fmtLemma{ibu\-mbi\-ro}
\fmtCitation{(e)i\-bu\-mbi\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{potter's workshop, pottery}
\fmtExampleRL{Omu\-bu\-mbi abu\-mbi\-ra mu ibu\-mbi\-ro.}
\fmtExampleEN{The potter works at the pottery workshop.}

\fmtLemma{ibu\-nda}
\fmtCitation{(e)i\-bu\-nda}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{side of the stomach or abdomen}
\fmtExampleRL{Oso\-bo\-la oku\-bo\-na ei\-bu\-nda lya mbu\-li.}
\fmtExampleEN{You can see the side of the goat's abdomen.}

\fmtLemma{ibu\-ra}%
\fmtOne{1}
\fmtDialect{Ruruuli}
\fmtCitation{(e)i\-bu\-ra}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{scarcity, shortage, undersupply, insufficiency, scarceness}
\fmtExampleRL{Ebi\-ro bi\-ni wa\-li\-wo ei\-bu\-ra lya me\-re.}
\fmtExampleEN{There is a shortage of food these days.}

\fmtLemma{ibu\-ra}%
\fmtOne{2}
\fmtDialect{Ruruuli}
\fmtCitation{(okw)i\-bu\-la}
\fmtPOS{v}%
\fmtTrans{ditr}
\fmtPFV{ibwi\-re}
\fmtSense{to serve (food), dish out}
\fmtExampleRL{Ko\-ba omaa\-wo atwi\-buu\-re eme\-re.}
\fmtExampleEN{Tell your mother to serve us food.}

\fmtLemma{ibuu\-ra}
\fmtCitation{(e)i\-buu\-ra}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{deserted house}
\fmtExampleRL{Abaa\-na ba ku nguu\-do bai\-ngi ba\-su\-la mu ibuu\-ra edi.}
\fmtExampleEN{Many street children sleep in that deserted house.}

\fmtLemma{ibu\-ya}
\fmtCitation{(e)i\-bu\-ya}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{occupation, profession, activity}
\fmtExampleRL{Oku\-lii\-sya ni\-kyo ibu\-ya lya Ba\-nya\-la eri\-ku\-lu.}
\fmtExampleEN{Cattle farming is the main activity among the Banyala.}

\fmtLemma{-ibw}
\fmtPOS{sf}
\fmtSense{passive suffix (extension) on verbs}

\fmtLemma{ibwa}
\fmtCitation{(e)i\-bwa}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{wound}
\fmtExampleRL{Ei\-bwa lya mwo\-jo oyo ka li\-rwi\-re oku\-wo\-na!}
\fmtExampleEN{The boy's wound has taken a long time to heal!}

\fmtLemma{ibyai\-re}
\fmtCitation{(e)i\-byai\-re}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{children, offspring, progeny}
\fmtExampleRL{Ata n’ei\-bya\-re ly’a\-baa\-na mu\-kaa\-ga.}
\fmtExampleEN{She has six offspring.}

\fmtLemma{ibya\-ro}
\fmtCitation{(e)i\-bya\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{traditional maternity centre, maternity ward}
\fmtExampleRL{Omu\-ka\-li ba\-mu\-twai\-re mu ibya\-ro ku\-bya\-la.}
\fmtExampleEN{The woman has been taken to the maternity centre to give birth.}

\fmtLemma{ica}%
\fmtOne{1}
\fmtCitation{(okw)i\-ca}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ici\-rye}
\fmtSense{to breathe}
\fmtExampleRL{Omwa\-na wa\-nge akwi\-ca ku\-bbi.}
\fmtExampleEN{My child is not breathing properly.}

\fmtLemma{ica}%
\fmtOne{2}
\fmtCitation{(okw)i\-ca}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ici\-rye}
\fmtSense{to accomplish, to complete, to finish, to finalise}
\fmtExampleRL{Omu\-ko\-li yai\-ci\-rye olu\-bbi\-mbi lwa\-mwe.}
\fmtExampleEN{The worker completed his assignment.}

\fmtLemma{icu\-mbi\-ro}
\fmtCitation{(e)i\-cu\-mbi\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{kitchen}
\fmtExampleRL{Eme\-re eri mu icu\-mbi\-ro.}
\fmtExampleEN{The food is in the kitchen.}

\fmtLemma{icu\-mu}
\fmtCitation{(e)i\-cu\-mu}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{an allergy that causes severe sneezing}
\fmtExampleRL{OGa\-va mu\-lwai\-re wa icu\-mu.}
\fmtExampleEN{Gava has an allergy which makes him sneeze.}

\fmtLemma{icwi}
\fmtPOS{interj}
\fmtSense{oh my goodness (used to express surprise, displeasure, shock, pity)}
\fmtExampleRL{Icwi! Awa\-ngwi\-re atyai aka\-lu\-lu?}
\fmtExampleEN{My goodness, how did he win the election?}

\fmtLemma{idaa\-la}
\fmtCitation{(e)i\-daa\-la}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{step, stage, phase}
\fmtExampleRL{Oku\-ku\-la ku\-bbaa\-mu ama\-daa\-la asa\-tu.}
\fmtExampleEN{Growth goes through three phases.}
\fmtSubentry(b)
\fmtSense{ladder}
\fmtExampleRL{Ndee\-te\-ra ei\-daa\-la nku\-ta\-ka ku\-nii\-na mu\-saa\-le gu\-ni.}
\fmtExampleEN{Bring the ladder, I would like to climb this tree.}
\fmtSubentry(c)
\fmtSense{level, rank}
\fmtExampleRL{Musobo\-la oku\-ku\-laa\-ku\-la\-na oku\-zwa oku idaa\-la li\-mwe okwi\-ra oku li\-ndi.}
\fmtExampleEN{You can progress from one rank to another.}

\fmtLemma{idaa\-li}
\fmtCitation{(e)i\-daa\-li}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{asparagus fern}
\fmtExampleRL{Ama\-daa\-li ga\-bbaa\-ku ama\-wa.}
\fmtExampleEN{The asparagus fern has thorns.}

\fmtLemma{ide\-mbe}
\fmtCitation{(e)i\-de\-mbe}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{freedom, rights, peace}
\fmtExampleRL{Oli\-na ei\-de\-mbe oku\-si\-moo\-la nka n’o\-ku\-ta\-ka.}
\fmtExampleEN{You have the freedom to speak the way you want.}

\fmtLemma{idi\-nda}
\fmtCitation{(e)i\-di\-nda}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{cuff (the end part of a sleeve)}
\fmtExampleRL{Engo\-ye zi\-bbaa\-ku ama\-di\-nda.}
\fmtExampleEN{Clothes have cuffs.}

\fmtLemma{iduu\-ka}
\fmtCitation{(e)i\-duu\-ka}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSource{from Sw.}
\fmtSense{shop, store}
\fmtExampleRL{Ya\-ba oku iduu\-ka ogu\-le su\-kaa\-li.}
\fmtExampleEN{Go to the shop and buy sugar.}

\fmtLemma{ifu\-lo}
\fmtCrossreference{\textnormal{see} ifuro}
\fmtDialect{Lunyala}

\fmtLemma{ifu\-mbe}
\fmtCitation{(o)i\-fu\-mbe}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{African civet (\textit{Civettictis civetta}}
\fmtExampleRL{Ei\-fu\-mbe zi\-rya ama\-paa\-paa\-li.}
\fmtExampleEN{African civets eat pawpaws.}

\fmtLemma{ifu\-mbi\-ro}
\fmtCrossreference{\textnormal{see} isumbiro}

\fmtLemma{ifu\-ro}
\fmtDialect{Ruruuli}
\fmtCitation{(e)i\-fu\-lo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{foam, froth}
\fmtExampleRL{Amai\-zi ga\-koo\-re ei\-fu\-ro.}
\fmtExampleEN{The water foams.}

\fmtLemma{ifu\-ru\-ngu}
\fmtCitation{(o)i\-fu\-ru\-ngu}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{turaco}
\fmtExampleRL{Ei\-fu\-ru\-ngu zi\-bba za bbu\-lu omwe\-kwa\-ti era zi\-rya mu\-no ama\-pee\-ra.}
\fmtExampleEN{The turaco is dark blue and eats a lot of guavas.}

\fmtLemma{iga}
\fmtCitation{(e)i\-ga}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{cooking stone}
\fmtExampleRL{Ekyo\-to ki\-bba n’a\-mai\-ga asa\-tu.}
\fmtExampleEN{A fireplace has three cooking stones.}

\fmtLemma{igaa\-la}%
\fmtCitation{(okw)i\-gaa\-la}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{iga\-lii\-re}
\fmtSubentry(a)
\fmtSense{to block (a passage)}
\fmtExampleRL{Epai\-pu na\-gi\-gaa\-lii\-re amai\-zi ga\-le\-ke ku\-se\-nse\-ka.}
\fmtExampleEN{I blocked the pipe so that the water is not wasted.}
\fmtSubentry(b)
\fmtSense{to patch up, to mend, to repair, to fix}
\fmtExampleRL{Igaa\-la eki\-tu\-li eki\-ri omu yu\-ni\-foo\-mu ya mwa\-na.}
\fmtExampleEN{Patch up the hole in the child's uniform.}
\fmtSubentry(c)
\fmtSense{to dress a wound}
\fmtExampleRL{Omu\-sa\-wu ai\-gaa\-la ebyo\-ya bu\-li ki\-ro.}
\fmtExampleEN{The doctor dresses the wound every day.}
\fmtSubentry(d)
\fmtSense{(of a wound) to heal}
\fmtExampleRL{Ekyo\-ya ki\-gaa\-lii\-re ku\-sai.}
\fmtExampleEN{The wound has healed properly.}

\fmtLemma{iga\-la}
\fmtCrossreference{\textnormal{see} igara}
\fmtDialect{Lunyala}

\fmtLemma{iga\-le}
\fmtCitation{\textnormal{(e.g.}\,kii\-ga\-le, bii\-ga\-le\textnormal{)}}
\fmtPOS{adj}
\fmtSense{covered, closed}
\fmtExampleRL{Ekii\-na kii\-ga\-le.}
\fmtExampleEN{The pit is covered.}

\fmtLemma{iga\-na}
\fmtCitation{(e)i\-ga\-na}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{herd}
\fmtExampleRL{OKi\-rya ata n’ei\-ga\-na lya nte.}
\fmtExampleEN{Kirya has a herd of cattle.}

\fmtLemma{iga\-ni}
\fmtCitation{(e)i\-ga\-ni}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{dedication, effort}
\fmtExampleRL{Omu\-ko\-li wa\-nge ako\-la emi\-ru\-mu n’ei\-ga\-ni.}
\fmtExampleEN{My worker does the job with dedication.}

\fmtLemma{iga\-nzo}
\fmtCitation{(e)i\-ga\-nzo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{fishing ground}
\fmtExampleRL{Baa\-bi\-re ku iga\-nzo ku\-te\-ga ncwi.}
\fmtExampleEN{They have gone to the fishing ground to fish.}
\fmtSubentry(b)
\fmtSense{sudd (an area of floating vegetation in a stretch of the White Nile, thick enough to impede navigation)}
\fmtExampleRL{Aba\-te\-gi aba\-mwei ba\-bba ku iga\-nzo.}
\fmtExampleEN{Some fishermen live on the sudd.}

\fmtLemma{iga\-ra}
\fmtDialect{Ruruuli}
\fmtCitation{(okw)i\-ga\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{igai\-re}
\fmtSense{to close, to shut}
\fmtExampleRL{Iga\-ra olwi\-ge lwa\-nge.}
\fmtExampleEN{Close my door.}

\fmtLemma{iga\-ra\-wo}
\fmtCitation{(okw)i\-ga\-ra\-wo}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{igai\-re\-wo}
\fmtSense{to close, to bring to an end, to end, to conclude, to finish}
\fmtExampleRL{Iga\-ra\-wo olu\-kii\-ko obwi\-re bwa\-bi\-re.}
\fmtExampleEN{Close the meeting, it is late.}

\fmtLemma{igee\-ra}
\fmtCitation{(okw)i\-gee\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ige\-rei\-re}
\fmtSense{to come closer, to approach, to draw closer}
\fmtExampleRL{Ko\-ba aba\-jwa\-ka abo bai\-gee\-re ani.}
\fmtExampleEN{Ask those guards to come closer.}

\fmtLemma{ige\-go}
\fmtCitation{(e)i\-ge\-go}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{molar tooth}
\fmtExampleRL{Ama\-ge\-go ni\-go ge gee\-ra\-yo omu mu\-nwa.}
\fmtExampleEN{The molar teeth are the last ones in the mouth.}

\fmtLemma{ige\-mo}
\fmtCitation{(e)i\-ge\-mo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{cuff (\eg of trousers)}

\fmtLemma{ige\-mu}
\fmtCrossreference{\textnormal{see} igemo}

\fmtLemma{ige\-ree\-ra}
\fmtCitation{(okw)i\-ge\-ree\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ige\-re\-rei\-re}
\fmtSense{to approach, to come toward, to come nearer}
\fmtExampleRL{Omu\-sai\-za oMu\-ruu\-li tai\-ge\-ree\-ra daa\-ni waa\-mwe.}
\fmtExampleEN{A Muruuli man doesn't come close to his mother-in-law.}

\fmtLemma{igi\-gi}
\fmtCitation{(e)i\-gi\-gi}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{curtain}
\fmtExampleRL{Omu\-lya\-ngo gwa\-mwe agu\-tai\-re\-mu ei\-gi\-gi.}
\fmtExampleEN{He has put a curtain in his doorway.}

\fmtLemma{igi\-ka}
\fmtCitation{(okw)i\-gi\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{igi\-ki\-re}
\fmtSubentry(a)
\fmtSense{to support, to prop up}
\fmtExampleRL{Ba\-kwe\-ndya oku\-yi\-gi\-ka enko\-ndo oku nyu\-mba eni.}
\fmtExampleEN{They need to support the house with a pole.}
\fmtSubentry(b)
\fmtFixedexpression{igi\-ka (a)mai\-ga}
\fmtSense{to put in place cooking stones, to set up cooking stones}
\fmtExampleRL{Bai\-gi\-ki\-re amai\-ga omu jo\-koo\-ni.}
\fmtExampleEN{The have set up the cooking stones in the kitchen.}

\fmtLemma{igi\-ki\-rya}
\fmtCitation{(okw)i\-gi\-ki\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{igi\-kii\-rye}
\fmtSense{to support (e.g.\,cooking stones) with a smaller stone}
\fmtExampleRL{Yi\-gi\-ki\-rya eki\-na\-ga omu\-ku\-bi gu\-ta\-su\-ku\-ka.}
\fmtExampleEN{Support the pan with a supporting stone so that the sauce does not spill.}

\fmtLemma{igi\-ki\-ryo}
\fmtCitation{(e)i\-gi\-ki\-ryo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{supporting stone}
\fmtExampleRL{Epi\-pa gi\-yi\-gi\-ke ei\-gi\-ki\-ryo eta\-gwa.}
\fmtExampleEN{Put a supporting stone there so that the barrel does not fall down.}

\fmtLemma{igo\-be}
\fmtCitation{(e)i\-go\-be}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{(of dogs) mating season}
\fmtExampleRL{Embwe\-ne zi\-ri ku igo\-be.}
\fmtExampleEN{The dogs are in their mating season.}

\fmtLemma{igo\-lo}
\fmtCitation{(e)i\-go\-lo}
\fmtPOS{adv}
\fmtSubentry(a)
\fmtSense{in the afternoon}
\fmtExampleRL{Ewaa\-mu nai\-za\-yo igo\-lo.}
\fmtExampleEN{I will come to your place in the afternoon.}
\fmtSubentry(b)
\fmtSense{last evening}
\fmtExampleRL{Ya\-boi\-ne omwe\-be\-mbe\-zi wa iya\-nga ei\-go\-lo.}
\fmtExampleEN{He met the president last evening.}
\fmtSubentry(c)
\fmtSense{yesterday}
\fmtExampleRL{Ei\-go\-lo ya\-bbai\-re\-ku ani.}
\fmtExampleEN{He was here yesterday.}

\fmtLemma{igo\-loi\-go\-lo}
\fmtCitation{(e)i\-go\-loi\-go\-lo}
\fmtPOS{adv}
\fmtSense{early evening, late afternoon}
\fmtExampleRL{Ya\-yi\-ki\-re ani ei\-go\-loi\-go\-lo.}
\fmtExampleEN{He arrived here in the early evening.}

\fmtLemma{igo\-lo\-li\-mwei}
\fmtCitation{(e)i\-go\-lo\-li\-mwei}
\fmtPOS{adv}
\fmtSense{early evening}
\fmtExampleRL{Naa\-bi\-re igo\-lo\-li\-mwei.}
\fmtExampleEN{I left in the early evening.}

\fmtLemma{igo\-mbe}
\fmtCitation{(o)i\-go\-mbe}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSubentry(a)
\fmtSense{western wind}
\fmtExampleRL{Oi\-go\-mbe ala\-nga ike\-ndi.}
\fmtExampleEN{The western wind announces rain.}
\fmtSubentry(b)
\fmtSense{wind that predicts rain}
\fmtExampleRL{Oi\-go\-mbe ala\-nga nti si\-zo\-ni ya\-ku\-si\-ga eri ampi okwi\-ka.}
\fmtExampleEN{The wind that predicts rain is an indicator that the sawing season is about to start.}

\fmtLemma{igo\-mbo\-lo\-la}
\fmtCitation{(e)i\-go\-mbo\-lo\-la}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{sub-county}
\fmtExampleRL{Aya\-bi\-re oku igo\-mbo\-lo\-la.}
\fmtExampleEN{He has gone to the sub-county.}

\fmtLemma{igo\-na}
\fmtCitation{(e)i\-go\-na}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{type of vegetation which grows on a lake shore}

\fmtLemma{igo\-se}
\fmtCitation{(e)i\-go\-se}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{testicle of an animal}
\fmtExampleRL{Ama\-go\-se ga ntaa\-ma ga\-so\-bo\-la oku\-lii\-bwa.}
\fmtExampleEN{The testicles of a sheep can be eaten.}

\fmtLemma{igu\-gu\-nyu}
\fmtCitation{(e)i\-gu\-gu\-nyu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{hip}
\fmtExampleRL{Ei\-gu\-gu\-nyu li\-ku\-ndu\-ma bu\-ndi.}
\fmtExampleEN{My hip hurts.}

\fmtLemma{igu\-ka}
\fmtCitation{(oku)\-i\-gu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{igu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to arrive}
\fmtExampleRL{Omu\-ge\-ni omu\-ku\-lu ai\-gu\-ki\-re.}
\fmtExampleEN{The guest of honour has arrived.}
\fmtSubentry(b)
\fmtSense{to open, to become open}
\fmtExampleRL{Ekya\-nko\-mee\-ro olwi\-ge lwa\-nge lu\-yi\-gu\-ki\-re.}
\fmtExampleEN{Finally, my door is open again.}
\fmtSubentry(c)
\fmtSense{(of a tree) to fall}
\fmtExampleRL{Omu\-saa\-le gu\-yi\-gu\-ki\-re.}
\fmtExampleEN{The tree has fallen down.}

\fmtLemma{igu\-ki\-ra}
\fmtCitation{(okw)i\-gu\-ki\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{igu\-kii\-re}
\fmtSense{to come out}
\fmtExampleRL{N’o\-ya\-kaa\-ra e No\-ngo oi\-gu\-ki\-ra mu Kye\-ri\-ma.}
\fmtExampleEN{When you go via Nongo you come out of Kyerima.}

\fmtLemma{igu\-la}
\fmtCitation{(oku)\-i\-gu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{igwi\-re}
\fmtSense{to open}
\fmtExampleRL{Ya\-yi\-gwi\-re\-wo olwi\-ge.}
\fmtExampleEN{She opened the door.}

\fmtLemma{igu\-mba}
\fmtCitation{(e)i\-gu\-mba}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{bone}
\fmtExampleRL{Omu\-sai\-za aye\-nde\-ke\-re ei\-gu\-mba.}
\fmtExampleEN{The man's bone is broken.}

\fmtLemma{igu\-ru}
\fmtCitation{(e)i\-gu\-ru}
\fmtPOS{n}%
\fmtNC{5}
\fmtSubentry(a)
\fmtSense{sky}
\fmtExampleRL{Oku igu\-ru ndoo\-ku bi\-cu.}
\fmtExampleEN{There are no clouds in the sky.}
\fmtSubentry(b)
\fmtSense{top, place above}
\fmtSubentry(c)
\fmtSense{heaven, paradise}
\fmtExampleRL{OYe\-su atu\-be\-nde\-rei\-re omu igu\-ru.}
\fmtExampleEN{Jesus is waiting for us in heaven.}
\fmtSubentry(d)
\fmtFixedexpression{igu\-ru we lye\-su\-ru\-kii\-re}
\fmtSense{horizon}
\fmtExampleRL{Ei\-gu\-ru we lye\-su\-ru\-kii\-re ti ni\-gyo nka\-ngo ya nsi.}
\fmtExampleEN{The horizon is not the end of the earth.}

\fmtLemma{igwa\-ra}
\fmtCitation{(e)i\-gwa\-ra}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{vuvuzela}
\fmtExampleRL{Aba\-pii\-na ba\-fuu\-wa ama\-gwa\-ra.}
\fmtExampleEN{The Bapiina clan plays the vuvuzela.}

\fmtLemma{ijaa\-bbu}
\fmtCitation{(e)i\-jaa\-bbu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Ar. }
\fmtSense{hijab (a head covering worn in public by some Muslim women)}
\fmtExampleRL{Abbwe\-re ijaa\-bbu.}
\fmtExampleEN{She is wearing a hijab.}

\fmtLemma{ijwi}
\fmtPOS{interj}
\fmtSense{my goodness (used to express shock, surprise)}
\fmtExampleRL{Ijwi\-… Afii\-re ki\-? ata\-bbai\-re mu\-lwai\-re!}
\fmtExampleEN{My goodness, what has he died of? He was not sick.}

\fmtLemma{ika}
\fmtCitation{(okw)i\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iki\-re}
\fmtSubentry(a)
\fmtSense{to arrive at, to come to, to reach}
\fmtExampleRL{Ndi ampi okwi\-ka e wa\-nge.}
\fmtExampleEN{I am about to arrive home.}
\fmtSubentry(b)
\fmtSense{to extend, to stretch}
\fmtExampleRL{Eki\-ba\-nja kya\-nge kii\-ka edi.}
\fmtExampleEN{My land extends up to there.}
\fmtSubentry(c)
\fmtSense{(of taste) to be appropriate, to be of the right measure, to be enough}
\fmtExampleRL{Omuu\-nyu gwi\-ki\-re omu mu\-ku\-bi.}
\fmtExampleEN{There is enough salt in the sauce.}

\fmtLemma{ikaa\-ka}
\fmtCitation{(e)i\-kaa\-ka}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{obstinacy, inflexibility, stubbornness}
\fmtExampleRL{Omu\-sai\-za oyo ya\-gai\-ne oku\-sa\-su\-la aba\-ko\-li baa\-mwe wa ikaa\-ka.}
\fmtExampleEN{That man refused to pay his workers, he is so obstinate.}

\fmtLemma{ikaa\-nya}
\fmtCitation{(okw)i\-kaa\-nya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ikai\-nye}
\fmtSense{to reach an agreement, to reach a consensus}
\fmtExampleRL{Mwi\-kaa\-nye mu\-zwe mu kwa\-raa\-ta\-na.}
\fmtExampleEN{Settle your differences and stop quarrelling.}

\fmtLemma{ikaa\-ra}
\fmtCitation{(okw)i\-kaa\-ra}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{ikai\-re}
\fmtSubentry(a)
\fmtSense{to oppress, to abuse, to maltreat, to ill-treat, to tyrannise}
\fmtExampleRL{Lwa\-ki mu\-ka\-li we oi\-kaa\-ra\-nga abaa\-na ba\-mwi\-ra waa\-mu?}
\fmtExampleEN{Woman, why do you treat your friend's children badly?}
\fmtSubentry(b)
\fmtSense{to sit, to sit down}
\fmtExampleRL{Ikaa\-ra oku ki\-ko\-nko\-ro ekyo ndoo\-wo nte\-be gi\-ndi.}
\fmtExampleEN{Sit on that wooden stool, there are no more seats.}
\fmtSubentry(c)
\fmtSense{to settle, to calm down, to quiet down}
\fmtExampleRL{Ikaa\-ra ansi oso\-me ebi\-ta\-bo bya\-mu.}
\fmtExampleEN{Settle down and read your books.}
\fmtSubentry(d)
\fmtFixedexpression{ikaa\-ra ku lu\-bu\-ga}
\fmtSense{(of females) to live in one's parents' home without being married}
\fmtExampleRL{Omwa\-la akyai\-kai\-re ku lu\-bu\-ga, akya\-li ku\-fu\-na mu\-sai\-za aku\-mu\-twa\-la.}
\fmtExampleEN{The girl is still living at her parents' home, she has not yet got a man to marry her.}

\fmtLemma{ikaa\-ryo}
\fmtCitation{(e)i\-kaa\-ryo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{drying lane}
\fmtExampleRL{Ebi\-tee\-re mu\-bi\-kaa\-rye mu ikaa\-ryo.}
\fmtExampleEN{Dry the potatoes on the drying lane.}

\fmtLemma{ika\-byo}
\fmtCitation{(e)i\-ka\-byo}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{rudeness, lack of manners, impudence, insolence}
\fmtExampleRL{Ei\-ka\-byo ti\-ri\-kwi\-ki\-ri\-zi\-sya kwe\-nde\-bwa ba\-ntu.}
\fmtExampleEN{Rudeness does not make you popular with people.}

\fmtLemma{ikai\-ka\-na}
\fmtCitation{(okw)i\-kai\-ka\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ikai\-kai\-ne}
\fmtSubentry(a)
\fmtSense{to stabilise, to normalise, to calm}
\fmtExampleRL{Enya\-nja ei\-kai\-kai\-ne.}
\fmtExampleEN{The sea has calmed.}
\fmtSubentry(b)
\fmtSense{to calm down, to become tranquil}
\fmtExampleRL{Ba\-nza oi\-kai\-ka\-ne nga okya\-li ku\-ba\-za.}
\fmtExampleEN{Calm down before you speak.}

\fmtLemma{ika\-la}%
\fmtOne{1}
\fmtCitation{(okw)i\-ka\-la}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{ikai\-re}
\fmtSubentry(a)
\fmtSense{to sit (down), to take a seat}
\fmtExampleRL{Ika\-la ansi olye eme\-re.}
\fmtExampleEN{Sit down and eat.}
\fmtSubentry(b)
\fmtSense{to live, to reside, to stay}
\fmtExampleRL{Nyi\-ka\-la Ka\-nju\-ki ku kya\-lo Bu\-yo\-be.}
\fmtExampleEN{I live in Kanjuki-Buyombe village.}
\fmtSubentry(c)
\fmtSense{(of \eg committee) to convene, to meet}
\fmtExampleRL{Olu\-ku\-mbaa\-no lwa Bu\-ruu\-li lwi\-kai\-re.}
\fmtExampleEN{The council of the Baruuli has convened.}
\fmtSubentry(d)
\fmtSense{to write (an exam), to take, to sit}
\fmtExampleRL{Nkwi\-za kwi\-ka\-la ebi\-buu\-lyo bya\-nge ema\-mbya.}
\fmtExampleEN{I will sit my exams tomorrow.}

\fmtLemma{ika\-la}%
\fmtOne{2}
\fmtCitation{(e)i\-ka\-la}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{charcoal}
\fmtExampleRL{Ama\-ka\-la ga\-bu\-sye\-re ebi\-ro bi\-ni mu\-ga\-kee\-ke\-ree\-rye.}
\fmtExampleEN{Charcoal is very expensive these days, use it sparingly.}
\fmtSubentry(b)
\fmtSense{battery}
\fmtExampleRL{Ti\-ndi\-na ma\-ka\-la ga la\-di\-yo ya\-nge.}
\fmtExampleEN{I don’t have batteries for my radio.}

\fmtLemma{ika\-lii\-ra}
\fmtCitation{(okw)i\-ka\-lii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ika\-lii\-re}
\fmtSense{to sit and wait for a long time}
\fmtExampleRL{Nkwi\-za ku\-mwi\-ka\-lii\-ra okwi\-ca ni na\-mu\-bo\-na.}
\fmtExampleEN{I will sit and wait until I see him.}

\fmtLemma{ika\-li\-sya}
\fmtCitation{(okw)i\-ka\-li\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ika\-li\-si\-rye}
\fmtSense{to enthrone, to install}
\fmtExampleRL{Aba\-nya\-la bai\-ka\-li\-si\-rye Omu\-ka\-ma waa\-bwe.}
\fmtExampleEN{The Banyala have enthroned their king.}

\fmtLemma{ika\-nsu}
\fmtCitation{(e)i\-ka\-nsu}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{hot temper, tendency to become angry very quickly}
\fmtExampleRL{Mwa\-na we ei\-ka\-nsu lya\-mu li\-ri\-kwi\-kya a\-lai.}
\fmtExampleEN{My child, your hot temper will destroy you.}

\fmtLemma{ika\-ta}%
\fmtOne{1}
\fmtCitation{(okw)i\-ka\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ika\-ti\-re}
\fmtSubentry(a)
\fmtSense{to press (a boil, a pustule)}
\fmtExampleRL{Ika\-ta ei\-yu\-te otoo\-le\-mu amaa\-ra.}
\fmtExampleEN{Press the boil to extract the pus.}
\fmtSubentry(b)
\fmtSense{to massage}
\fmtExampleRL{Ika\-ta omu\-go\-ngo gwa\-mwe a\-bbee\-ku ka\-le\-ka\-le.}
\fmtExampleEN{Massage his back so that he gets some relief.}
\fmtSubentry(c)
\fmtSense{to massage the belly of a mother of a new born baby to tone the muscles}

\fmtLemma{ika\-ta}%
\fmtOne{2}
\fmtCitation{(okw)i\-ka\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ika\-ti\-re}
\fmtSense{to throw down with a heavy plop}
\fmtExampleRL{Ai\-ka\-ti\-re ensa\-wu ansi.}
\fmtExampleEN{He has thrown down the bag with a heavy plop.}

\fmtLemma{ika\-ti\-ra}
\fmtCitation{(okw)i\-ka\-ti\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ika\-tii\-re}
\fmtSense{to pack, to package}
\fmtExampleRL{Ika\-ti\-ra ensa\-wu ya pa\-mba ku\-sai.}
\fmtExampleEN{Pack the cotton into the sack.}

\fmtLemma{ika\-yu}
\fmtCitation{(e)i\-ka\-yu}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{rudeness, lack of manners, impudence, insolence}
\fmtExampleRL{Ei\-ka\-yu li\-ku\-no\-be\-sya aba\-ntu.}
\fmtExampleEN{Rudeness makes people hate you.}

\fmtLemma{ike\-ka}
\fmtCitation{(e)i\-ke\-ka}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{season of slicing sweet potatoes}
\fmtExampleRL{Twi\-ge\-rei\-re ebi\-ro bya ike\-ka.}
\fmtExampleEN{We are approaching the sweet potatoe slicing season.}

\fmtLemma{ike\-ndi}
\fmtCitation{(o)i\-ke\-ndi}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{rain}
\fmtExampleRL{Oi\-ke\-ndi asi\-nga ku\-gwa mu kwe\-zi Kwai\-se\-nsa\-ncwi.}
\fmtExampleEN{Usually there is heavy rain in May.}

\fmtLemma{ike\-re\-nda}
\fmtCitation{(e)i\-ke\-re\-nda}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{tablet (of medicine), pill}
\fmtExampleRL{Mmii\-re ei\-ke\-re\-nda li\-mwei lyo\-nkai.}
\fmtExampleEN{I have taken only one tablet.}

\fmtLemma{ikii\-ra}%
\fmtOne{1}
\fmtCitation{(okw)i\-kii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iki\-rii\-re}
\fmtSense{to approach, to go near to, to come close to}
\fmtExampleRL{Ya\-ba oi\-kii\-re omu\-ku\-lu omu\-ma\-nyi\-sye ebi\-zi\-bu bya\-mu.}
\fmtExampleEN{Go and approach the head and tell him about your problems.}

\fmtLemma{ikii\-ra}%
\fmtOne{2}
\fmtCitation{(okw)i\-kii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iki\-rii\-re}
\fmtSense{to be completed, to be achieved, to come true}
\fmtExampleRL{Obu\-la\-ngi bwa Yo\-kaa\-na bwai\-ki\-rii\-re.}
\fmtExampleEN{John's prophecy was fulfilled.}

\fmtLemma{ikii\-ra}%
\fmtOne{3}
\fmtCitation{(okw)i\-kii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iki\-rii\-re}
\fmtSense{to be righteous, to be true}

\fmtLemma{ikii\-ri}
\fmtCitation{\textnormal{(e.g.}\,gii\-kii\-ri, bii\-kii\-ri, mwi\-kii\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{(of food) clean, pure, undefiled}
\fmtExampleRL{Eme\-re tu\-maa\-re oku\-gi\-sa\-bi\-ra ati\-bu\-ni gi\-kii\-ri.}
\fmtExampleEN{We have prayed for the food, now it is clean.}
\fmtSubentry(b)
\fmtSense{righteous}
\fmtExampleRL{Ki\-ta\-ka n’o\-bba mwi\-kii\-ri omu mai\-so ga Ka\-nca.}
\fmtExampleEN{It is better to be righteous before God.}

\fmtLemma{ikii\-rya}%
\fmtOne{1}
\fmtCitation{(okw)i\-kii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iki\-rii\-rye}
\fmtSubentry(a)
\fmtSense{to perfect, to polish, to hone, to refine}
\fmtExampleRL{Ya\-ki\-koo\-re emi\-ru\-ndi gi\-ngi okwi\-ca ni ya\-kii\-ki\-rii\-rye.}
\fmtExampleEN{He practiced it so much that he perfected it.}
\fmtSubentry(b)
\fmtSense{to fulfil}
\fmtExampleRL{Lwa\-ki osuu\-bi\-za\-nga n’o\-ta\-yi\-kii\-rya.}
\fmtExampleEN{Why do you make promises, which you never fulfil?}

\fmtLemma{ikii\-rya}%
\fmtOne{2}
\fmtCitation{(okw)i\-kii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iki\-rii\-rye}
\fmtSense{(of crops) to reach maturity, to become ripe}
\fmtExampleRL{Ebi\-duu\-ma bi\-ta\-ndi\-ki\-re okwi\-kii\-rya.}
\fmtExampleEN{Maize is about to ripen.}

\fmtLemma{iki\-ngo}
\fmtCitation{(e)i\-ki\-ngo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{air potato, air yam (\textit{Dio\-sco\-rea bulbi\-fe\-ra})}
\fmtExampleRL{Ama\-ki\-ngo ga\-woo\-ma.}
\fmtExampleEN{Air potatoes are tasty.}

\fmtLemma{iki\-ra}
\fmtCitation{(okw)i\-ki\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ikii\-re}
\fmtSense{to perform rituals for (a new garden)}
\fmtExampleRL{Baa\-bi\-re kwi\-ki\-ra mu\-si\-ri gwa bi\-yaa\-ta.}
\fmtExampleEN{They have gone to perform rituals for the new potato garden.}

\fmtLemma{iki\-ra\-na}
\fmtCitation{(okw)i\-ki\-ra\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iki\-rai\-ne}
\fmtSense{to match perfectly, to be equal, (of a musical instrument) to be in tune}
\fmtExampleRL{Oyo omu\-ka\-li wa\-nge twai\-ki\-rai\-ne.}
\fmtExampleEN{My wife and I are a perfect match.}

\fmtLemma{iki\-rii\-ra}
\fmtCitation{(okw)i\-ki\-rii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iki\-rii\-re}
\fmtSubentry(a)
\fmtSense{to be perfect, to be faultless}
\fmtExampleRL{To\-so\-bo\-la kwi\-ki\-rii\-ra mu bu\-li ki\-mwei.}
\fmtExampleEN{You cannot be perfect in everything.}
\fmtSubentry(b)
\fmtSense{to reach the destination on the same day}
\fmtExampleRL{Aya\-bi\-re lwa\-lee\-ro nai\-ki\-rii\-ra mi\-re\-mbe.}
\fmtExampleEN{He left and arrived on the same day safely.}

\fmtLemma{iki\-rii\-ri}
\fmtCitation{\textnormal{(e.g.}\,gii\-kii\-ri, bii\-kii\-ri, mwi\-kii\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSense{perfect, saint, holy}
\fmtExampleRL{OKa\-nca mwi\-kii\-ri\-ri. Eki\-sa kya\-mwe kya bwi\-ki\-rii\-ri.}
\fmtExampleEN{God is holy. His kindness is saintly.}

\fmtLemma{iki\-ri\-za}
\fmtCitation{(okw)i\-ki\-ri\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iki\-rii\-ze}
\fmtSubentry(a)
\fmtSense{to allow, to let, to permit, to give permission}
\fmtExampleRL{Nyi\-ki\-ri\-za nswe\-re omwa\-ra waa\-mu.}
\fmtExampleEN{Allow me to marry your daughter.}
\fmtSubentry(b)
\fmtSense{to agree, to consent to, to say yes to}
\fmtExampleRL{Na\-mu\-ko\-be\-re ebi\-zi\-bu bya\-nge nai\-ki\-ri\-za oku\-mpa ese\-nte.}
\fmtExampleEN{I told him about my problem, and he agreed to give me money.}
\fmtSubentry(c)
\fmtSense{to accept, to trust}
\fmtExampleRL{We\-mbe iki\-ri\-za by'a\-ku\-ko\-ba le\-ke oya\-ka\-le\-mu.}
\fmtExampleEN{Just accept whatever he is saying, you will succeed.}
\fmtSubentry(d)
\fmtSense{to have faith, to practice (a religion)}
\fmtExampleRL{Nta n’o\-kwi\-ki\-ri\-za kwi\-ngi omu mwa\-na wa\-nge.}
\fmtExampleEN{I have a lot of faith in my child.}

\fmtLemma{iki\-ri\-za\-ga\-nya}
\fmtCitation{(okw)i\-ki\-ri\-za\-ga\-nya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iki\-ri\-za\-gai\-nye}
\fmtSense{to reach a consensus}
\fmtExampleRL{Twi\-ki\-re oku nyi\-ki\-ri\-za\-ga\-nya.}
\fmtExampleEN{We have reached a consensus.}

\fmtLemma{iki\-rya}
\fmtCitation{(okw)i\-ki\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iki\-rii\-rye}
\fmtSense{to keep in a place temporarily}
\fmtExampleRL{Lee\-ta omwa\-na omwi\-ki\-rye ani.}
\fmtExampleEN{Bring the child and keep her here temporarily.}

\fmtLemma{iko\-da}
\fmtCitation{(e)i\-ko\-da}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{rudeness, impoliteness, naughtiness}
\fmtExampleRL{Abaa\-na ba bi\-ro bi\-ni ba\-ta n’ei\-ko\-da.}
\fmtExampleEN{Today's children are naughty.}

\fmtLemma{iko\-ke}
\fmtCitation{(e)i\-ko\-ke}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{ash}
\fmtExampleRL{Oba ekyo\-to ki\-woo\-re, too\-la\-mu ei\-ko\-ke.}
\fmtExampleEN{Remove the ash from the cooking place if it has cooled down.}

\fmtLemma{iko\-me\-ra}
\fmtCitation{(e)i\-ko\-me\-ra}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{prison}
\fmtExampleRL{Omu\-sai\-za ya\-fi\-rii\-re omu iko\-me\-ra.}
\fmtExampleEN{The man died in the prison.}

\fmtLemma{ikoo\-ko}
\fmtCitation{(e)i\-koo\-ko}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{cambium and sapwood parts of a tree}
\fmtExampleRL{Ebi\-ye\-nze bya\-bya\-lii\-re omu ikoo\-ko lya mu\-saa\-le.}
\fmtExampleEN{Cockroaches laid eggs on the bark of the tree.}

\fmtLemma{ikoo\-ro}
\fmtCitation{(e)i\-koo\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{factory, plant}
\fmtExampleRL{Ei\-koo\-ro lya bi\-kai\-zo li\-yii\-re.}
\fmtExampleEN{The sugar factory has burnt down.}

\fmtLemma{iko\-ti}
\fmtCitation{(e)i\-ko\-ti}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{neck}
\fmtExampleRL{Omu\-sai\-za ali\-na ei\-ko\-ti eri\-koo\-to nk’e\-rya mbe\-re\-ge.}
\fmtExampleEN{The man has a big neck like a pig's.}

\fmtLemma{iku\-bbi\-ro}
\fmtCitation{(e)i\-ku\-bbi\-ro}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{thorough beating, good thrashing}
\fmtExampleRL{Baa\-mu\-ku\-bbi\-re n’a\-si\-ga\-la omu iku\-bbi\-ro.}
\fmtExampleEN{He received a good thrashing.}

\fmtLemma{iku\-de}
\fmtCitation{(e)i\-ku\-de}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{hip}
\fmtExampleRL{Omu\-ka\-li ali\-na ama\-ku\-de aga\-koo\-to.}
\fmtExampleEN{The woman has big hips.}

\fmtLemma{iku\-ka}
\fmtCitation{(e)i\-ku\-ka}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{type of water potato, aquatic vegetation}
\fmtExampleRL{Ama\-ku\-ka gai\-ta encwi.}
\fmtExampleEN{Floating water plants kill fish.}

\fmtLemma{iku\-lu\-lu}
\fmtCitation{(o)i\-ku\-lu\-lu}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{ostrich}
\fmtExampleRL{Iku\-lu\-lu ni\-kyo ki\-nyo\-nyi eki\-si\-nga obu\-koo\-to.}
\fmtExampleEN{The ostrich is the biggest bird.}

\fmtLemma{iku\-mbaa\-ni\-ro}
\fmtCitation{(e)i\-ku\-mbaa\-ni\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{social centre}
\fmtExampleRL{Oku bu\-li iru\-gu\-za ku\-bbaa\-ku ei\-ku\-mbaa\-ni\-ro.}
\fmtExampleEN{Every sub-county has a social centre.}
\fmtSubentry(b)
\fmtSense{temple, house of worship, synagogue}
\fmtExampleRL{Omu\-ka\-ma Su\-le\-maa\-ni ya\-yo\-mbo\-kei\-re oKa\-nca ei\-ku\-mbaa\-ni\-ro.}
\fmtExampleEN{King Solomon constructed a temple for God.}

\fmtLemma{iku\-mbe}
\fmtCitation{(e)i\-ku\-mbe}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{public, community}
\fmtExampleRL{Ki\-sai ebi\-fo bya iku\-mbe oku\-bba n’e\-byo\-loo\-ni.}
\fmtExampleEN{It is good for public places to have toilets.}

\fmtLemma{iku\-mi}
\fmtPOS{num}
\fmtSubentry(a)
\fmtSense{ten}
\fmtExampleRL{Omu\-cu\-ngwa gu\-li\-mu emi\-ru\-ndi iku\-mi.}
\fmtExampleEN{An orange comes in ten types.}
\fmtSubentry(b)
\fmtFixedexpression{mwe\-zi gwa iku\-mi}
\fmtSense{October}
\fmtSubentry(c)
\fmtFixedexpression{mwe\-zi gwa iku\-mi na gu\-mwei }
\fmtSense{November}
\fmtSubentry(d)
\fmtFixedexpression{mwe\-zi gwa iku\-mi n’eibi\-ri}
\fmtSense{December}
\fmtSubentry(e)
\fmtFixedexpression{iku\-mi na (-)mwei}
\fmtCitation{\textnormal{(e.g.}\,iku\-mi n'e\-mwei, iku\-mi n'o\-mwei\textnormal{)}}
\fmtSense{eleven}
\fmtSubentry(f)
\fmtFixedexpression{iku\-mi na ibi\-ri}
\fmtSense{twelve}
\fmtSubentry(g)
\fmtFixedexpression{iku\-mi na inai}
\fmtSense{fourteen}
\fmtSubentry(h)
\fmtFixedexpression{iku\-mi na isa\-tu}
\fmtSense{thirteen}
\fmtSubentry(i)
\fmtFixedexpression{iku\-mi na itaa\-nu}
\fmtSense{fifteen}
\fmtSubentry(j)
\fmtSense{sixteen}
\fmtFixedexpression{iku\-mi na mu\-ka\-nga}
\fmtSubentry(k)
\fmtFixedexpression{iku\-mi na mu\-na\-nai}
\fmtSense{eighteen}
\fmtSubentry(l)
\fmtFixedexpression{iku\-mi na mu\-sa\-nju}
\fmtSense{seventeen}
\fmtSubentry(m)
\fmtFixedexpression{iku\-mi na mwe\-ndai}
\fmtSense{nineteen}

\fmtLemma{iku\-ndi}
\fmtCrossreference{\textnormal{see also} mukundi}
\fmtCitation{(e)i\-ku\-ndi}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{navel, umbilicus}
\fmtExampleRL{Ei\-ku\-ndi lya mwa\-na li\-ku\-mu\-lu\-ma.}
\fmtExampleEN{The child's navel hurts.}

\fmtLemma{iku\-ra}
\fmtCitation{(e)i\-ku\-ra}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{pearl}
\fmtExampleRL{OYe\-su ni\-ye ibbaa\-le erya iku\-ra.}
\fmtExampleEN{Jesus is a precious pearl.}
\fmtSubentry(b)
\fmtSense{stone put inside the drum}
\fmtExampleRL{Ei\-bbaa\-le eri\-bba mu\-nda mu ngo\-ma ba\-lye\-ta ei\-ku\-ra.}
\fmtExampleEN{The stone put inside the drum is called 'ikura'.}

\fmtLemma{iku\-ta}
\fmtCitation{(okw)i\-ku\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iku\-ti\-re}
\fmtSense{to be satisfied, to be content}
\fmtExampleRL{Ka nsuu\-bi\-re nti ati oi\-ku\-ti\-re.}
\fmtExampleEN{Let me hope that you are now satisfied.}

\fmtLemma{ikuu\-ku}
\fmtCitation{(o)i\-kuu\-ku}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSubentry(a)
\fmtSense{type of weed}
\fmtExampleRL{Oi\-kuu\-ku mu\-zi\-bu wa ku\-zi\-kya.}
\fmtExampleEN{Ikuuku is difficult to weed.}
\fmtSubentry(b)
\fmtSense{dangerous thing or person}
\fmtExampleRL{OKa\-ju\-ra afuu\-ki\-re ikuu\-ku ta\-kya\-wuu\-ra.}
\fmtExampleEN{Kajura has become very dangerous because he no longer listens.}

\fmtLemma{ikya}
\fmtCitation{(okw)i\-kya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iki\-rye}
\fmtSubentry(a)
\fmtSense{to deliver, to bring, to take, to transport, to courier}
\fmtExampleRL{Ikya obu\-kwe\-nda bu\-ni eji je\-ge\-nge.}
\fmtExampleEN{Deliver this message to the king.}
\fmtSubentry(b)
\fmtSense{to have ready for harvest}
\fmtExampleRL{OMu\-ka\-la\-sa ai\-ki\-rye emi\-ye\-mbe.}
\fmtExampleEN{Mukalasa has his mangoes ready for harvest.}
\fmtSubentry(c)
\fmtSense{to do something for the sake of it}
\fmtExampleRL{Ako\-la kwi\-kya mu\-rwa.}
\fmtExampleEN{He is working for the sake of it.}
\fmtSubentry(d)
\fmtFixedexpression{ikya (o)mwo\-yo}
\fmtSense{to calm down, relax}
\fmtExampleRL{Mwa\-na wa\-nge ikya omwo\-yo oi\-za ku\-te\-ree\-ra.}
\fmtExampleEN{My dear, relax, everything will be okay.}
\fmtSubentry(e)
\fmtFixedexpression{ikya (o)mwo\-yo}
\fmtSense{to sigh, to breathe out, to exhale}
\fmtExampleRL{Ni ya\-mboi\-ne nai\-kya omwo\-yo.}
\fmtExampleEN{He sighed with relief when he saw me.}

\fmtLemma{ima}
\fmtCitation{(okw)i\-ma}
\fmtPOS{v}%
\fmtTrans{ditr}
\fmtPFV{imi\-re}
\fmtSense{to deny, to refuse, to turn down, to reject, to decline}
\fmtExampleRL{Baa\-nyi\-mi\-re omu\-ki\-sa gwa kwa\-ba omu iso\-me\-ro.}
\fmtExampleEN{I was denied the opportunity to go to school.}

\fmtLemma{imaa\-mu}
\fmtCitation{(o)i\-maa\-mu}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from Ar. }
\fmtSense{imam (an Islamic leader)}
\fmtExampleRL{Aba\-si\-raa\-mu ba Ga\-li\-raa\-ya ba\-fu\-ni\-re oi\-maa\-mu owo\-ndi.}
\fmtExampleEN{The Muslims at Galiraa\-ya have got a new imam.}

\fmtLemma{imi}
\fmtCitation{\textnormal{(e.g.}\,mwi\-mi, bee\-mi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{mean, stingy, ungenerous}
\fmtExampleRL{Omu\-ntu omwi\-mi ai\-ma n’o\-waa\-mwe.}
\fmtExampleEN{A stingy person does not want to share even with his relative.}

\fmtLemma{imo\-ko\-le}
\fmtCitation{(e)i\-mo\-ko\-le}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{left side, the left}
\fmtExampleRL{Omu\-ko\-no gwa\-nge ogwa imo\-ko\-le gu\-ndu\-ma.}
\fmtExampleEN{I feel pain in my left arm.}

\fmtLemma{imu\-ka}
\fmtCitation{(okw)i\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{imu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to stand up, to rise, to get on one's feet}
\fmtExampleRL{Imu\-ka twa\-be.}
\fmtExampleEN{Stand up and let's go.}
\fmtSubentry(b)
\fmtSense{to wake up}
\fmtExampleRL{Imu\-ka olye eme\-re.}
\fmtExampleEN{Wake up and eat.}
\fmtSubentry(c)
\fmtSense{(of a war) to erupt}
\fmtExampleRL{Olu\-ta\-lo lwa\-yi\-mu\-kii\-re e Co\-ngo.}
\fmtExampleEN{A war erupted in Congo.}

\fmtLemma{imu\-ki}
\fmtCitation{\textnormal{(e.g.}\,kii\-mu\-ki, bii\-mu\-ki, kye\-mu\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{growing well}
\fmtExampleRL{Ebi\-duu\-ma bya\-nge bii\-mu\-ki.}
\fmtExampleEN{My maize is growing well.}

\fmtLemma{imu\-ki\-ra\-mu}
\fmtCitation{(okw)i\-mu\-ki\-ra\-mu}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{imu\-kii\-re\-mu}
\fmtSense{to act immediately}
\fmtExampleRL{Na\-mwe\-te\-re nai\-mu\-ki\-ra\-mu.}
\fmtExampleEN{I called him and he acted immediately.}

\fmtLemma{imu\-la}
\fmtCitation{(okw)i\-mu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{imwi\-re}
\fmtSubentry(a)
\fmtSense{to lift, to pick up, to raise, to hold up, to elevate}
\fmtExampleRL{Imu\-la ensa\-wu eyo ogye\-tii\-ke oku mu\-twe.}
\fmtExampleEN{Lift that bag and carry it on your head.}
\fmtSubentry(b)
\fmtSense{to wake up, to awaken, to cause to stop sleeping}
\fmtExampleRL{Ya\-ba oi\-mu\-le oi\-tee\-wo alye kyai\-go\-lo.}
\fmtExampleEN{Go and wake up your father for supper.}
\fmtSubentry(c)
\fmtSense{to raise one's hand}
\fmtExampleRL{Eyee\-te\-je\-rye ayi\-mu\-le omu\-ko\-no.}
\fmtExampleEN{Raise your hand if you have understood.}
\fmtSubentry(d)
\fmtSense{to invoke (a spirit), to summon}
\fmtExampleRL{Baa\-bi\-re kwi\-mu\-la mba\-ndwa.}
\fmtExampleEN{They have gone to invoke the spirits.}
\fmtSubentry(e)
\fmtSense{to introduce, to recognise the presence of people by asking them to stand up}
\fmtExampleRL{Yi\-mu\-la aba\-ge\-ni abo le\-ke aba\-ntu ba\-ba\-ma\-nye.}
\fmtExampleEN{Introduce the visitors so that people may know them.}
\fmtSubentry(f)
\fmtSense{to help, to assist, to aid}
\fmtExampleRL{Bu\-li ki\-ro ONya\-nge\-ndo ayi\-mu\-la omu\-lwai\-re.}
\fmtExampleEN{Nyangendo helps a patient every night.}

\fmtLemma{imuu\-ra}
\fmtCitation{(okw)i\-muu\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{imu\-lii\-re}
\fmtSense{to break the hard crust of the potato mound in order to allow water and air to permeate through it}

\fmtLemma{inaa\-mii\-ra}
\fmtCitation{(okw)i\-naa\-mii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ina\-mi\-rii\-re}
\fmtSense{to hunch, to bend (down), to incline one's body}
\fmtExampleRL{Lwa\-ki oyi\-naa\-mii\-ra otyo, omu\-go\-ngo gu\-kwi\-za ku\-lwa\-la.}
\fmtExampleEN{Why are you bending like that? Your back will get injured.}

\fmtLemma{inai}
\fmtCrossreference{used in isolation and with class 9 nouns, see also \textit{nai}}
\fmtPOS{num}
\fmtSense{four}
\fmtExampleRL{Oyi\-nza oku\-kwa\-ta ena\-mba ei\-nai eki\-tu\-ndu.}
\fmtExampleEN{You may get number four and a half.}

\fmtLemma{ina\-la}
\fmtCitation{(okw)i\-na\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{inai\-re}
\fmtSense{to become impoverished, to become poor}
\fmtExampleRL{Omu\-sai\-za oyo yaa\-li mu\-guu\-da bai\-tu ati ata\-ndi\-ki\-re okwi\-na\-la.}
\fmtExampleEN{That man was rich, but he became poor.}

\fmtLemma{ina\-ma}
\fmtCitation{(okw)i\-na\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ina\-mi\-re}
\fmtSubentry(a)
\fmtSense{to bend (down), to incline one's body}
\fmtExampleRL{Akwi\-na\-ma oku\-ko\-ma ebi\-sa\-si\-ro.}
\fmtExampleEN{He is bending down to pick up rubbish.}
\fmtSubentry(b)
\fmtSense{to bow (as a sign of respect or greeting)}
\fmtExampleRL{Aba\-ja\-paa\-ni bai\-na\-ma nga ba\-ku\-la\-mu\-kya.}
\fmtExampleEN{The Japanese bow while greeting.}

\fmtLemma{ina\-mi\-ra}
\fmtCitation{(okw)i\-na\-mi\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ina\-mii\-re}
\fmtSense{to chat, to pass time, to converse, to chatter}
\fmtExampleRL{Aboo\-jo mwi\-ze twi\-na\-mi\-re.}
\fmtExampleEN{Friends, come and let's chat.}

\fmtLemma{ina\-mi\-ri}
\fmtCitation{\textnormal{(e.g.}\,mwi\-na\-mi\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSense{to be sad}
\fmtExampleRL{Ya\-li mwi\-na\-mi\-ri n'a\-tu\-li\-ka ama\-li\-ga.}
\fmtExampleEN{She was so sad, tears flowed down her face.}

\fmtLemma{ina\-mya}
\fmtCitation{(okw)i\-na\-mya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ina\-mi\-rye}
\fmtSense{to bend, to force into a curve or an angle}
\fmtExampleRL{Ina\-mya omu\-ye\-mbe tu\-no\-ko\-le emi\-ye\-mbe.}
\fmtExampleEN{Bend the mango tree so that we can pick the mangoes.}

\fmtLemma{ina\-yi\-na}
\fmtCitation{(okw)i\-na\-yi\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ina\-yi\-ni\-re}
\fmtSense{to loiter, to stroll, to wander, to roam around}
\fmtExampleRL{Ebi\-so\-lo bya\-mu bi\-kwi\-na\-yi\-na kya\-njo.}
\fmtExampleEN{Your animals are roaming around.}

\fmtLemma{ina\-za}
\fmtCitation{(okw)i\-na\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{inai\-ze}
\fmtSubentry(a)
\fmtSense{to destroy, to ruin, to harm, to crash, to demolish}
\fmtExampleRL{Ei\-ba\-ra eri\-bbi lii\-na\-za ebi\-see\-ra bya mwa\-na ebya mu mai\-so.}
\fmtExampleEN{A bad name destroys a child's future.}
\fmtSubentry(b)
\fmtSense{to impoverish, to make poor}
\fmtExampleRL{Ega\-vu\-me\-nti ya Obo\-te ya\-yi\-nai\-ze Obu\-ga\-nda.}
\fmtExampleEN{Obote's government impoverished the Buganda Kingdom.}

\fmtLemma{indu\-ka}
\fmtCitation{(okw)i\-ndu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{indu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to turn, to move around, to rotate}
\fmtExampleRL{Yi\-ndu\-ka olo\-le oku gu\-lyo.}
\fmtExampleEN{Turn and face right.}
\fmtSubentry(b)
\fmtSense{to change direction, to go different way}
\fmtExampleRL{Ya\-yi\-ndu\-ki\-re n’a\-kwa\-ta ogwa bu\-gwa.}
\fmtExampleEN{He changed direction and took the western route.}
\fmtSubentry(c)
\fmtSense{to break away, to separate from}
\fmtExampleRL{OMu\-se\-ve\-ni ya\-yi\-ndu\-ki\-re oku DP.}
\fmtExampleEN{Museveni broke away from the Democratic Party.}

\fmtLemma{indu\-la}
\fmtCitation{(okw)i\-ndu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{indwi\-re}
\fmtSubentry(a)
\fmtSense{to change, to modify}
\fmtExampleRL{Omwo\-jo aku\-yi\-ndu\-la ama\-mi\-ro.}
\fmtExampleEN{The boy's voice is changing due to puberty.}
\fmtSubentry(b)
\fmtSense{to transfer}
\fmtExampleRL{ODa\-ma\-li a\-bbai\-re mu\-ntu mu\-sai, wa\-bu\-la ba\-mu\-yi\-ndwi\-re.}
\fmtExampleEN{Damali has been a good person, but she has been transferred.}
\fmtSubentry(c)
\fmtSense{to avert (disease, bad weather, calamity), to prevent, to ward off}
\fmtExampleRL{Eki\-yi\-ndu\-la ki\-yi\-ndu\-la emi\-sa\-ngo.}
\fmtExampleEN{The Abutilon indicum plant is believed to avert judgement of cases in courts of law.}
\fmtSubentry(d)
\fmtSense{to transfer labour pains to the husband (It is believed that a woman can transfer the labour pains to the father of the child.)}
\fmtExampleRL{Omu\-ka\-li anyi\-ndu\-lii\-re eki\-da.}
\fmtExampleEN{The woman has transferred her labour pains to me.}
\fmtSubentry(e)
\fmtSense{to open (a page of a book)}
\fmtExampleRL{Yi\-ndu\-la oyi\-re oku olu\-pa\-pu\-la lwa 104 olwa ki\-dee\-ro kya Lu\-ga\-nda.}
\fmtExampleEN{Open page 104 of the Luganda dictionary.}
\fmtSubentry(f)
\fmtSense{to interpret, to translate}
\fmtExampleRL{Ayi\-nduu\-ra aba\-ge\-nyi oru\-ba\-zo oku\-zwa omu\-li\-mi olu\-ndi.}
\fmtExampleEN{She interprets speech from a different language for visitors.}

\fmtLemma{indu\-lu\-ge\-nsi\-ya}
\fmtCitation{(e)i\-ndu\-lu\-ge\-nsi\-ya}
\fmtPOS{n}%
\fmtNC{9}
\fmtSource{from En.}
\fmtSense{indulgence}
\fmtExampleRL{N’oju\-na aboo\-ro ng’o\-li mu ki\-sii\-bo ofu\-na ei\-ndu\-lu\-ge\-nsi\-ya oku\-zwa ewa Ka\-nca.}
\fmtExampleEN{When you give alms to the poor, you receive indulgences from God.}

\fmtLemma{ingaa\-ni\-sya}
\fmtCitation{(okw)i\-ngaa\-ni\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{inga\-ni\-si\-rye}
\fmtSense{to exchange, to trade, to swap}
\fmtExampleRL{Aba\-sa\-mbi ba mu\-pii\-ra ba\-te\-ra okwi\-ngaa\-ni\-sya emi\-joo\-zi.}
\fmtExampleEN{Football players usually exchange their jerseys.}

\fmtLemma{ingai}%
\fmtCitation{\textnormal{(e.g.}\,zi\-ngai\textnormal{)}}
\fmtPOS{interrog}
\fmtSubentry(a)
\fmtSense{how much, how many}
\fmtExampleRL{Omu\-wai\-re empii\-ya zi\-ngai?}
\fmtExampleEN{How much money have you given him?}
\fmtSubentry(b)
\fmtFixedexpression{ka ingai}
\fmtSense{how often}
\fmtExampleRL{Ka ingai nga nkwi\-za ani n’o\-ta\-bo\-ne\-ke\-ku?}
\fmtExampleEN{How often have I come here without seeing you?}

\fmtLemma{inga\-na}
\fmtCrossreference{\textnormal{see} enkana}
\fmtCitation{(okw)i\-nga\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ingai\-ne}
\fmtSubentry(a)
\fmtSense{to be equal, to be like, to be the same
}
\fmtExampleRL{Omu\-sai\-za oyo n’o\-mu\-ka\-li waa\-mwe bai\-nga\-na.}
\fmtExampleEN{That man and his wife are equal.}
\fmtSubentry(b)
\fmtSense{to quarrel, to verbally assault, to argue, to fight}
\fmtExampleRL{Bai\-nga\-na mu\-no bai\-tu ba\-si\-ga\-la ba\-kwe\-ndya\-ga\-na.}
\fmtExampleEN{They quarrel a lot, but still love each other.}

\fmtLemma{ingi}
\fmtCitation{\textnormal{(e.g.}\,ki\-ngi, bi\-ngi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{numerous, multiple, many}
\fmtExampleRL{Ebi\-duu\-ma bi\-ngi ebi\-ro bi\-ni.}
\fmtExampleEN{Maize is plentiful these days.}

\fmtLemma{ingi\-ni\-ya}
\fmtCitation{(o)i\-ngi\-ni\-ya}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from En.}
\fmtSense{engineer}
\fmtExampleRL{Omwa\-na wa\-nge aku\-so\-mee\-ra bwa ingi\-ni\-ya.}
\fmtExampleEN{My child is studying to become an engineer.}

\fmtLemma{ingi\-ra}
\fmtCitation{(okw)i\-ngi\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ingii\-re}
\fmtSubentry(a)
\fmtSense{to enter, to go inside}
\fmtExampleRL{Yi\-ngi\-ra oyi\-ka\-le ansi.}
\fmtExampleEN{Enter and sit down!}
\fmtSubentry(b)
\fmtSense{to intervene, to interfe\-re}
\fmtExampleRL{Iza oi\-ngi\-re omu nso\-nga za mu\-to waa\-mu.}
\fmtExampleEN{Come and intervene in the matters concerning your younger brother.}

\fmtLemma{ingi\-rii\-ra}
\fmtCitation{(okw)i\-ngi\-rii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{ingi\-rii\-re}
\fmtSubentry(a)
\fmtSense{to attack, to assault}
\fmtExampleRL{Bbaa\-mwi\-ngi\-rii\-re ne ba\-mwi\-bba ebi\-ntu byo\-na.}
\fmtExampleEN{They attacked her and stole everything from her.}
\fmtSubentry(b)
\fmtSense{to impede, to hinder, to inhibit, to disturb, to interfe\-re}
\fmtExampleRL{Lwa\-ki okwi\-ngi\-rii\-ra entee\-ka\-tee\-ka za mwi\-raa\-wo?}
\fmtExampleEN{Why do you interfere with your colleague’s affairs?}
\fmtSubentry(c)
\fmtSense{(of a man) to live at his wife's home}
\fmtExampleRL{Omu Ba\-ruu\-li omu\-sai\-za okwi\-ngi\-rii\-ra omu\-ka\-li.}
\fmtExampleEN{It is a taboo among the Baruuli for a man to live at his wife's home.}

\fmtLemma{ingi\-rya}%
\fmtOne{1}
\fmtCitation{(okw)i\-ngi\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ingii\-rye}
\fmtSubentry(a)
\fmtSense{to take into, to bring in, to take in}
\fmtExampleRL{OBi\-ntu\-bi\-zi\-bu akwi\-ngi\-rya obu\-sye omu nyu\-mba.}
\fmtExampleEN{Bintubizibu is taking millet flour into the house.}
\fmtSubentry(b)
\fmtSense{to adopt (a child)}
\fmtExampleRL{Omwa\-na oyo na\-mwi\-ngii\-rye omu lu\-bu\-ga lwa\-nge.}
\fmtExampleEN{I adopted that child into my family.}

\fmtLemma{ingi\-rya\-mu}
\fmtCitation{(okw)i\-ngi\-rya\-mu}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ingii\-rye\-mu}
\fmtSense{to include, to incorporate}
\fmtExampleRL{Baa\-bwi\-rye bi\-ga\-mbo ki ebya\-ku\-ta omu nso\-boo\-ro.}
\fmtExampleEN{They asked which words need to be included in the dictionary.}

\fmtLemma{ingi\-sya}
\fmtCitation{(okw)i\-ngi\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ingi\-si\-rye}
\fmtSense{to exchange, to swap, to barter}
\fmtExampleRL{Bai\-ngi\-si\-rye ena\-mba za si\-mu ni bai\-ti\-rai\-ne.}
\fmtExampleEN{They exchanged phone numbers when they met.}

\fmtLemma{ini}
\fmtCitation{(e)i\-ni}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{liver}
\fmtExampleRL{Ei\-ni lya\-mwe li\-rwai\-re.}
\fmtExampleEN{His liver is infected.}

\fmtLemma{ini\-ka}
\fmtCitation{(okw)i\-ni\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ini\-ki\-re}
\fmtSubentry(a)
\fmtSense{to immerse, to dip, to duck, to submerge}
\fmtExampleRL{Ka\-bo\-na ya\-yi\-ni\-ki\-re omwa\-na omu mai\-zi.}
\fmtExampleEN{The priest immersed the child in water.}
\fmtSubentry(b)
\fmtSense{to soak}
\fmtExampleRL{Ini\-ka obu\-sye tu\-ko\-le ama\-lwa.}
\fmtExampleEN{Soak the millet and we will make beer.}

\fmtLemma{ini\-ko}
\fmtCitation{(e)i\-ni\-ko}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{flank hip bone}
\fmtExampleRL{Ya\-ye\-nde\-kei\-re mu ini\-ko.}
\fmtExampleEN{He sustained a fracture of the flank hip bone.}
\fmtSubentry(b)
\fmtSense{pelvic girdle}
\fmtExampleRL{Aye\-nde\-ke\-re ama\-ni\-ko.}
\fmtExampleEN{He has broken his pelvic girdle.}

\fmtLemma{ino\-li}
\fmtCitation{(e)i\-no\-li}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{fat}
\fmtExampleRL{Ontaa\-ma a\-bba n’a\-ma\-no\-li gai\-ngi.}
\fmtExampleEN{The mutton has a lot of fat.}

\fmtLemma{iso\-ngo\-lya nku\-be\-be}
\fmtCitation{(e)i\-so\-ngo\-lya nku\-be\-be}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{sharp-pointed stump or log}
\fmtExampleRL{Ei\-so\-ngo\-lya nku\-be\-be li\-su\-mi\-ta abaa\-na nga ba\-ku\-zee\-nye\-ra omu ki\-si\-ko.}
\fmtExampleEN{Sharp-pointed logs often prick children while they play in the bush.}

\fmtLemma{inu\-ka}
\fmtCitation{(okw)i\-nu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{inu\-ki\-re}
\fmtSense{to end the day's work}
\fmtExampleRL{Obwi\-re bwa kwi\-nu\-ka ni\-bwo bu\-ni.}
\fmtExampleEN{It is time to end the day's work.}

\fmtLemma{inu\-la}
\fmtCitation{(okw)i\-nu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{inwi\-re}
\fmtSense{to ask (someone) to end the day's work}
\fmtExampleRL{Ya\-ba omwi\-nu\-le.}
\fmtExampleEN{Go and ask her to stop working.}

\fmtLemma{inu\-lu\-la}
\fmtCitation{(okw)\-e\-nu\-lu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{inu\-lwi\-re}
\fmtSubentry(a)
\fmtSense{to save oneself from trouble}
\fmtExampleRL{Omu\-li\-raa\-nwa ta\-li\-na se\-nte za kwe\-nu\-lu\-la ku\-zwa mu bu\-zi\-bu bw’a\-li\-mu.}
\fmtExampleEN{My neighbour has no money to save himself from the trouble he is in.}
\fmtSubentry(b)
\fmtSense{to attain independence}
\fmtExampleRL{Uga\-nda ye\-nu\-lwi\-re oku\-zwa omu mi\-ko\-no gya Ba\-ju\-ngu mu 1962.}
\fmtExampleEN{Uganda attained independence from the British in 1962.}

\fmtLemma{inu\-nu\-ka}
\fmtCitation{(okw)i\-nu\-nu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{inu\-nu\-ki\-re}
\fmtSense{to straighten one's back}
\fmtExampleRL{Ni ya\-boi\-ne omu\-ka\-li waa\-mwe n’ai\-nu\-nu\-ka.}
\fmtExampleEN{When he saw his wife, he straightened his back.}

\fmtLemma{ipaa\-paa\-li}
\fmtCitation{(e)i\-paa\-paa\-li}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{papaya}
\fmtExampleRL{Na\-guu\-re ama\-paa\-paa\-li ataa\-nu nga ga baa\-na ba\-nge.}
\fmtExampleEN{I bought five papayas for my children.}

\fmtLemma{ipee\-ra}
\fmtCitation{(e)i\-pee\-ra}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Port.\,via Sw.}
\fmtSense{guava}
\fmtExampleRL{Ba\-mu\-lii\-re\-ku ei\-pee\-ra.}
\fmtExampleEN{He has missed eating guava.}

\fmtLemma{ipee\-sa}
\fmtCitation{(e)i\-pee\-sa}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{button}
\fmtExampleRL{Esaa\-ti ya\-nge eri\-ku ama\-pee\-sa ataa\-nu.}
\fmtExampleEN{My shirt has five buttons.}

\fmtLemma{-ir}
\fmtPOS{sf}
\fmtSense{applicative suffix (extension) on verbs}

\fmtLemma{ira}%
\fmtOne{1}
\fmtCitation{(okw)i\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtSubentry(a)
\fmtSense{to go, to move towards, to move in the direction of}
\fmtExampleRL{Ira edi.}
\fmtExampleEN{Go that way.}
\fmtSubentry(b)
\fmtSense{to come (back), to return}
\fmtExampleRL{Bai\-ra di e Ka\-yu\-nga?}
\fmtExampleEN{When are they returning to Kayunga?}
\fmtSubentry(c)
\fmtFixedexpression{ira (e)nyu\-ma}
\fmtSense{to move in reverse, to move back}
\fmtExampleRL{Emo\-to\-ka ya\-gii\-rye enyu\-ma ku\-bba ye\-ra\-bii\-re ensa\-wu yaa\-mu.}
\fmtExampleEN{He reversed the car because he had forgotten his bag.}
\fmtSubentry(d)
\fmtFixedexpression{ira (e)nyu\-ma}
\fmtSense{to regress, to deteriorate}
\fmtExampleRL{Bbi\-zi\-nee\-nsi yaa\-mwe ei\-re enyu\-ma.}
\fmtExampleEN{His business is not performing well.}

\fmtLemma{ira}%
\fmtOne{2}
\fmtCitation{(e)i\-ra}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{pearl}
\fmtExampleRL{Abaa\-na baa\-boi\-ne ei\-ra oku nya\-nja.}
\fmtExampleEN{The children found a pearl at the lake shore.}

\fmtLemma{ira\-ga\-zu}
\fmtCrossreference{\textnormal{see} iragazu}
\fmtCitation{\textnormal{(e.g.}\,gi\-ra\-gu\-zu, bye\-ra\-ga\-zu, lwi\-ra\-ga\-zu\textnormal{)}}
\fmtPOS{adj}
\fmtSense{black}
\fmtExampleRL{Olu\-go\-ye lwa\-mwe lwi\-ra\-ga\-zu.}
\fmtExampleEN{His cloth is black.}

\fmtLemma{ira\-gu\-lya}
\fmtCitation{(okw)i\-ra\-gu\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ira\-gwi\-rye}
\fmtSubentry(a)
\fmtSense{to soil, to smear, to stain,}
\fmtExampleRL{Omwa\-na ai\-ra\-gwi\-rye saa\-ti ya\-nge.}
\fmtExampleEN{A child has soiled my shirt.}
\fmtSubentry(b)
\fmtSense{to dye black, to blacken, to make black}
\fmtExampleRL{Ku mba\-ga ya mwa\-ra wa\-nge aba\-ka\-li bai\-ra\-gu\-rya ei\-so\-ke.}
\fmtExampleEN{For my daughter’s wedding, the women will dye their hair black.}

\fmtLemma{ira\-gu\-ra}
\fmtCitation{(okw)i\-ra\-gu\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ira\-gwi\-re}
\fmtSense{to be dirty, to be filthy}
\fmtExampleRL{Egi\-raa\-si ekwi\-ra\-gu\-ra.}
\fmtExampleEN{The glass is dirty.}

\fmtLemma{ira\-gu\-ya}
\fmtCitation{(okw)i\-ra\-gu\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ira\-gwi\-rye}
\fmtSense{(of a beard) to grow}
\fmtExampleRL{Omu\-taa\-ni wa\-nge ai\-ra\-gwi\-rye omu\-le\-zu.}
\fmtExampleEN{My son has started growing a beard.}

\fmtLemma{ira\-gu\-zu}
\fmtCrossreference{\textnormal{see} iragazu}

\fmtLemma{irai}
\fmtCitation{(e)i\-rai}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSubentry(a)
\fmtSense{ancient times, old age}
\fmtExampleRL{Mpee\-rya eki\-ta\-bo ekyo ekya irai.}
\fmtExampleEN{Give me that ancient book.}
\fmtSubentry(b)
\fmtSense{once, long ago, in the past, earlier, already}
\fmtExampleRL{Yaa\-bi\-re irai nga okya\-bba\-si\-rye.}
\fmtExampleEN{He went a long time ago while you were still asleep.}
\fmtSubentry(c)
\fmtFixedexpression{irai budi}
\fmtSense{long ago}
\fmtExampleRL{Ei\-rai bu\-di ng’o\-mu\-ka\-li ta\-lya ndu.}
\fmtExampleEN{In the past women used not to eat lungfish.}
\fmtSubentry(d)
\fmtFixedexpression{irai n’(e)i\-rai}
\fmtSense{since time immemorial, very long ago, in the distant past}
\fmtExampleRL{Ei\-rai n’ei\-rai ng’a\-ba\-ka\-li aBa\-ga\-nda ti\-ba\-vu\-ga gaa\-li.}
\fmtExampleEN{Long ago, Baganda women did not ride bicycles.}

\fmtLemma{ira\-ka}%
\fmtOne{1}
\fmtCitation{(e)i\-ra\-ka}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{sound, noise}
\fmtExampleRL{Ki\-ki ekyo eki\-koo\-re ei\-ra\-ka eri\-koo\-to?}
\fmtExampleEN{What made the loud sound?}
\fmtSubentry(b)
\fmtSense{voice}
\fmtExampleRL{Ka\-ngu\-la oku ira\-ka aba\-ntu ba\-ku\-wuu\-re.}
\fmtExampleEN{Speak loudly and clearly so that people hear you.}

\fmtLemma{ira\-ka}%
\fmtOne{2}
\fmtCitation{(e)i\-ra\-ka}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{\textit{Cy\-pe\-rus ro\-tu\-ndus} (coco-grass, Java grass, nut grass)}

\fmtLemma{ira\-lu}
\fmtCitation{(e)i\-ra\-lu}
\fmtPOS{n}%
\fmtNC{5}
\fmtSubentry(a)
\fmtSense{madness, insanity}
\fmtExampleRL{Omu Ka\-mpa\-la mu\-li\-mu aba\-gwi wa ira\-lu bai\-ngi.}
\fmtExampleEN{There are many madmen in Kampala.}
\fmtSubentry(b)
\fmtSense{stubbornness}
\fmtExampleRL{Le\-ke\-ra\-wo ei\-ra\-lu.}
\fmtExampleEN{Stop this stubbornness.}

\fmtLemma{ira\-mi\-ryo}
\fmtCitation{(e)i\-ra\-mi\-ryo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{place of worship (\eg a shrine)}
\fmtExampleRL{Baa\-bi\-re mu ira\-mi\-ryo ku\-ra\-mya.}
\fmtExampleEN{They have gone to the shrine to worship.}

\fmtLemma{ira\-mu}
\fmtCitation{(okw)i\-ra\-mu}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ire\-mu}
\fmtSubentry(a)
\fmtSense{to do again, to repeat}
\fmtExampleRL{Nje okwi\-ra\-mu okwa\-ba ewa mu\-ka\-li oyo, ko\-yi!}
\fmtExampleEN{Me, to go back to that woman's place? No way!}
\fmtSubentry(b)
\fmtSense{to answer, to respond, to reply}
\fmtExampleRL{Odi na\-mwi\-ra\-mu, ``e\-se\-nte nkwi\-za ku\-ku\-wa," nti, ``oo\-h!" nti, ``ee\-h." aa\-h, n’a\-ya\-ba.}
\fmtExampleEN{And the other one responded, ``I will give the money". Then he said ``Ooh!", ``Eeh!", ``Aah", and he left.}
\fmtSubentry(c)
\fmtSense{to repeat (an exam)}
\fmtExampleRL{Ira\-mu omu ki\-bii\-na oso\-bo\-le okwa\-ka\-la oku\-sai.}
\fmtExampleEN{Repeat the class so that you get good grades.}

\fmtLemma{ira\-myo}%
\fmtOne{1}
\fmtCitation{(e)i\-ra\-myo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{church}
\fmtExampleRL{Ei\-ra\-myo lya ba\-ka\-tu\-li\-ki ka li\-sai!}
\fmtExampleEN{The Catholic church is very good.}
\fmtSubentry(b)
\fmtSense{church service}
\fmtExampleRL{Ei\-ra\-myo lya lee\-ro li\-ku\-le\-mbei\-rwe Na\-ka\-zi\-bwe.}
\fmtExampleEN{Today's church service was led by Nakazibwe.}

\fmtLemma{ira\-myo}%
\fmtOne{2}
\fmtPOS{interj}
\fmtSense{the traditional way of thanking the potter}
\fmtExampleRL{Ira\-myo! - Binaga!}
\fmtExampleEN{Let them bake well! – The pots!}

\fmtLemma{ira\-nga}
\fmtCitation{(e)i\-ra\-nga}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{canna lily plant}
\fmtExampleRL{Eki\-me\-ra kya ei\-ra\-nga ki\-zwa\-ku ama\-la\-nga.}
\fmtExampleEN{The canna lily plant produces the canna lily grain.}
\fmtSubentry(b)
\fmtSense{canna lily seed}
\fmtExampleRL{Ama\-ra\-nga ni\-go ba\-tee\-ka omu nsa\-nsi.}
\fmtExampleEN{Canna lily seeds are used to make shakers.}
\fmtSubentry(c)
\fmtSense{steel ball}
\fmtExampleRL{Egaa\-li ebaa\-mu ama\-ra\-nga.}
\fmtExampleEN{The bicycle has steel balls.}

\fmtLemma{ira\-nti}
\fmtCitation{(e)i\-ra\-nti}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{pole supporting a hunting net}
\fmtExampleRL{Ebi\-ti\-mba mu\-bi\-te oku ma\-ra\-nti.}
\fmtExampleEN{Set the hunting nets on the poles.}

\fmtLemma{-ire}
\fmtPOS{sf}
\fmtSense{perfective aspect suffix on verbs}

\fmtLemma{ire\-mbo}%
\fmtCitation{(e)i\-re\-mbo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{(of a house) front side}
\fmtExampleRL{Ya\-ka\-ngii\-re mu ire\-mbo n’a\-yi\-ra\-yo.}
\fmtExampleEN{He stopped in front of the house and went back.}
\fmtSubentry(b)
\fmtSense{home}
\fmtExampleRL{Ki\-ta\-ka bu\-li mu\-ntu n’a\-bba n’ei\-re\-mbo.}
\fmtExampleEN{Every person deserves to have a home.}

\fmtLemma{iri\-bwa}
\fmtCitation{(okw)i\-ri\-bwa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{irii\-bwe}
\fmtSense{to regain consciousness, to come to}
\fmtExampleRL{Abbai\-re aka\-bi\-re ai\-rii\-bwe.}
\fmtExampleEN{A person who had fainted regained consciousness.}

\fmtLemma{iri\-ga}
\fmtCitation{(e)i\-ri\-ga}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{tear drop, tear}
\fmtExampleRL{Taa\-li\-na go\-nze ei\-ri\-ga eri\-mwei omu mai\-so.}
\fmtExampleEN{She doesn’t have even a single teardrop.}

\fmtLemma{irii\-bwa}
\fmtCitation{(e)i\-rii\-bwa}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{cucumber}
\fmtExampleRL{Ama\-lii\-bwa ga\-woo\-ma nk’e\-byo\-ngu.}
\fmtExampleEN{Cucumbers taste like pumpkins.}

\fmtLemma{irii\-ka}
\fmtCitation{(okw)i\-rii\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{irii\-ki\-re}
\fmtSense{to grumble, to complain, to grouse, to whine, to mutter, to protest, to make a fuss}
\fmtExampleRL{Le\-ke\-ra\-wo oku\-kwi\-rii\-ka.}
\fmtExampleEN{Stop grumbling.}

\fmtLemma{irii\-ra}%
\fmtOne{1}
\fmtCitation{(okw)i\-rii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iri\-rii\-re}
\fmtSense{to go and return on the same day}
\fmtExampleRL{Ba\-yi\-ri\-rii\-re e Ka\-mpa\-la.}
\fmtExampleEN{They went to Kampala and returned on the same day.}

\fmtLemma{irii\-ra}%
\fmtOne{2}
\fmtCitation{(okw)i\-rii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iri\-rii\-re}
\fmtSense{to be next}
\fmtExampleRL{Nii\-we okwi\-rii\-ra.}
\fmtExampleEN{You are next.}

\fmtLemma{irii\-ro}
\fmtCitation{(e)i\-rii\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{dining room, sitting room}
\fmtExampleRL{Eme\-re tu\-gi\-rii\-ra mu i\-rii\-ro.}
\fmtExampleEN{We eat food in the dining room.}

\fmtLemma{irii\-rya}
\fmtCitation{(okw)i\-rii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{irii\-rye}
\fmtSubentry(a)
\fmtSense{to return, to replace, to bring back, to give back}
\fmtExampleRL{Nyi\-rii\-rya esaa\-wa ya\-nge.}
\fmtExampleEN{Bring me back my watch.}
\fmtSubentry(b)
\fmtSense{to replace seeds that have failed to germinate}
\fmtExampleRL{Yi\-rii\-rya ebi\-duu\-ma ebyo.}
\fmtExampleEN{Replace the maize seeds that failed to germinate.}
\fmtSubentry(c)
\fmtSense{to give a tip, to tip}
\fmtExampleRL{Omaa\-ma amai\-te okwi\-rii\-rya aba\-ko\-li ba mu woo\-tee\-ri.}
\fmtExampleEN{Mom often gives a tip to hotel workers.}

\fmtLemma{irii\-si\-ryo}
\fmtCitation{(e)i\-rii\-si\-ryo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{animal farm}
\fmtExampleRL{Ata n’ei\-rii\-si\-ryo li\-koo\-to.}
\fmtExampleEN{He has a big animal farm.}

\fmtLemma{iri\-ka}
\fmtCitation{(okw)i\-ri\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iri\-ki\-re}
\fmtSense{to keep together, to gather in one place}
\fmtExampleRL{Enyama, ebi\-yaa\-ta, su\-kaa\-li byo\-na\-byo\-na ba\-bi\-yi\-ri\-ki\-re omu ki\-sii\-ka ki\-mwei.}
\fmtExampleEN{They kept all the meat, potatoes, and sugar together in one room.}

\fmtLemma{iri\-ndu\-ka}
\fmtCitation{(okw)i\-ri\-ndu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iri\-ndu\-ki\-re}
\fmtSense{to move in large numbers}
\fmtExampleRL{Olu\-go\-lo\-go\-mbo lwa mbu\-li lu\-yi\-ri\-ndu\-ki\-re mu bi\-duu\-ma bya\-nge.}
\fmtExampleEN{Many goats moved into my maize garden.}

\fmtLemma{iri\-ngi\-ta}
\fmtCitation{(okw)i\-ri\-ngi\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iri\-ngi\-ti\-re}
\fmtSense{to roll, to move by turning over and over}
\fmtExampleRL{Eki\-bbaa\-le ki\-kwi\-ri\-ngi\-ta.}
\fmtExampleEN{The stone is rolling.}

\fmtLemma{iri\-ngi\-tya}
\fmtCitation{(okw)i\-ri\-ngi\-tya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iri\-ngi\-ti\-rye}
\fmtSense{to roll, to cause to move by turning over and over}
\fmtExampleRL{Yi\-ri\-ngi\-tya eki\-ko\-nge ekyo.}
\fmtExampleEN{Roll that tree stump.}

\fmtLemma{iro}
\fmtCrossreference{\textnormal{see also} irowo}
\fmtCitation{(e)i\-ro}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{craving for smoking, smoking urge}
\fmtExampleRL{Omu\-nywi wa taa\-ba a\-bba n’ei\-ro.}
\fmtExampleEN{A smoker has a craving for cigarettes.}

\fmtLemma{iro\-bo}
\fmtCitation{(e)i\-ro\-bo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{hook, fishhook, barb}
\fmtExampleRL{Ei\-ro\-bo lya\-nge ndi\-te\-ge\-sya ma\-le.}
\fmtExampleEN{I use my hook to catch the mudfish.}

\fmtLemma{iro\-go}
\fmtCitation{(e)i\-ro\-go}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{witchcraft}
\fmtExampleRL{Aba\-ntu aba\-mwe ba\-ko\-la ei\-ro\-go.}
\fmtExampleEN{Some people practice witchcraft.}

\fmtLemma{iro\-mba}
\fmtCitation{(e)i\-ro\-mba}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{big navel}
\fmtExampleRL{Omu\-ka\-li oyo ata n’ei\-ro\-mba.}
\fmtExampleEN{The woman has a big navel.}

\fmtLemma{iro\-ngo}
\fmtCitation{(o)i\-ro\-ngo}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{east wind}
\fmtExampleRL{Oi\-ro\-ngo n’a\-ku\-bba ala\-nga isa\-na.}
\fmtExampleEN{The east wind announces a drought.}

\fmtLemma{iro\-wo}
\fmtCrossreference{\textnormal{see also} iro}
\fmtCitation{(e)i\-ro\-wo}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{craving for smoking, smoking urge}
\fmtExampleRL{Ata n’ei\-ro\-wo lya ku\-nywa taa\-ba.}
\fmtExampleEN{She has a craving to smoke a cigarette.}

\fmtLemma{iru\-ba}
\fmtCitation{(e)i\-ru\-ba}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{well, borehole}
\fmtExampleRL{Baa\-bi\-re ku\-sa\-ka iru\-ba.}
\fmtExampleEN{They have gone to dig the well.}

\fmtLemma{iru\-bu\-za}
\fmtCitation{(okw)i\-ru\-bu\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iru\-bi\-zi\-re, iru\-bwi\-ze}
\fmtSense{to move (animals) from one grazing place to the other}
\fmtExampleRL{Ya\-ba oi\-ru\-bu\-ze embu\-li, obwi\-re bwi\-ki\-re.}
\fmtExampleEN{Go, it is time to transfer the goats.}

\fmtLemma{iru\-gu\-za}
\fmtCitation{(e)i\-ru\-gu\-za}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{sub-county}
\fmtExampleRL{Na\-bya\-lii\-bwe mu iru\-gu\-za lya Ibbaa\-le.}
\fmtExampleEN{I was born in Bbaale sub-county.}

\fmtLemma{iru\-ka}
\fmtCitation{(okw)i\-ru\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iru\-ki\-re}
\fmtSense{to run, to run away, (of a vehicle) to move fast}
\fmtExampleRL{Omwa\-na ai\-ru\-ka mu\-no oku\-si\-nga aba\-ku\-lu.}
\fmtExampleEN{The child runs faster than adults.}

\fmtLemma{iru\-ke}
\fmtCitation{(e)i\-ru\-ke}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{gorilla}
\fmtExampleRL{Amai\-ru\-ke kya\-bu\-la\-mbu\-li omu Uga\-nda.}
\fmtExampleEN{Gorillas are a tourist attraction in Uganda.}

\fmtLemma{iru\-kii\-ra}
\fmtCitation{(okw)i\-ru\-kii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iru\-ki\-rii\-re}
\fmtSense{to help, to rescue, to assist}
\fmtExampleRL{Mwi\-ru\-kii\-re ba\-di aba\-gwi\-re oku bu\-ta\-ndu.}
\fmtExampleEN{Rescue those who were involved in the accident.}

\fmtLemma{iru\-kya}
\fmtCitation{(okw)i\-ru\-kya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iru\-ki\-rye}
\fmtSubentry(a)
\fmtSense{to chase away, to drive away}
\fmtExampleRL{Iru\-kya ebi\-nyo\-nyi omu ndu\-mi\-ro ya bwi\-ta.}
\fmtExampleEN{Chase away those birds from the millet garden.}
\fmtSubentry(b)
\fmtSense{to fire, to dismiss}
\fmtExampleRL{Ba\-mwi\-ru\-kyi\-rye oku mu\-li\-mu mu bu\-ta\-li bwe\-nka\-nya.}
\fmtExampleEN{He was dismissed from his job unfairly.}

\fmtLemma{iru\-ma}%
\fmtOne{1}
\fmtCitation{(o)i\-ru\-ma}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{dove, pigeon}
\fmtExampleRL{Oi\-ru\-ma awoo\-ma mu\-no.}
\fmtExampleEN{Dove is very tasty.}

\fmtLemma{iru\-ma}%
\fmtOne{2}
\fmtCitation{(e)i\-ru\-ma}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{season for farming, sowing season}
\fmtExampleRL{Twi\-ge\-rei\-re ebi\-ro bya iru\-ma.}
\fmtExampleEN{We are approaching the sowing season.}

\fmtLemma{Iru\-ma}%
\fmtOne{3}
\fmtCitation{Iru\-ma}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{Kabaka's landing site}
\fmtExampleRL{Ba\-ta\-nga\-rai\-ne oku mwa\-lo gwa Ka\-ba\-ka ogwe\-te\-bwa Iru\-ma.}
\fmtExampleEN{They met at Kabaka's landing site called Iruma.}

\fmtLemma{iru\-ngu}%
\fmtOne{1}
\fmtCitation{(e)i\-ru\-ngu}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{desert}
\fmtExampleRL{Ki\-zi\-bu oku\-lu\-ma ebi\-too\-ke omu iru\-ngu.}
\fmtExampleEN{It is hard to grow matoke in the desert.}

\fmtLemma{Iru\-ngu}%
\fmtOne{2}
\fmtCitation{(o)i\-ru\-ngu}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{type of spirit responsible for success in hunting and riches}

\fmtLemma{irwa\-li\-ro}
\fmtCrossreference{\textnormal{see} irwaro}

\fmtLemma{irwa\-ro}
\fmtCitation{(e)i\-rwa\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{hospital, ward}
\fmtExampleRL{Ba\-mu\-twai\-re mu irwa\-ro lya Na\-ka\-so\-ngo\-la.}
\fmtExampleEN{He has been taken to Nakasongola hospital.}

\fmtLemma{irya}
\fmtCitation{(okw)i\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{irye}
\fmtSubentry(a)
\fmtSense{to return, to put back, to take back}
\fmtExampleRL{Irya enkai\-to zaa\-mwe ma\-ngu\-ma\-ngu.}
\fmtExampleEN{Return his shoes immediately.}
\fmtSubentry(b)
\fmtFixedexpression{irya (o)bu\-ku\-ku}
\fmtSense{to go back to past grudges}
\fmtExampleRL{OBa\-li\-mwe\-za na\-mu\-ko\-be\-re\-ku, n’a\-nyi\-rii\-rya obu\-ku\-ku.}
\fmtExampleEN{I warned Balimweza, but he kept going back to his past grudges.}
\fmtSubentry(c)
\fmtFixedexpression{irya (o)bwe\-nku\-lu\-mu}
\fmtSense{to regurgitate}
\fmtExampleRL{Embu\-li zi\-mwe oku bi\-so\-lo ebii\-rya obwe\-nku\-lu\-mu.}
\fmtExampleEN{Goats are one of the animals which regurgitate.}
\fmtSubentry(d)
\fmtFixedexpression{irya (a)ma\-gi\-ta}
\fmtSense{to perform the ceremony where the bride goes home to officially visit her relatives after the wedding}
\fmtExampleRL{Omu\-ka\-li aya\-bi\-re kwi\-rya\-yo ma\-gi\-ta.}
\fmtExampleEN{The woman has gone back home to officially visit her relatives after marriage.}
\fmtSubentry(e)
\fmtFixedexpression{irya (o)mu\-sa\-ngo}
\fmtSense{to commit an offence, to commit a crime, to break the law}
\fmtExampleRL{Lwa\-ki okwi\-rya emi\-sa\-ngo bu\-li ka\-see\-ra?}
\fmtExampleEN{Why do you keep committing offences every now and then?}

\fmtLemma{irya\-ku}
\fmtCitation{(okw)i\-rya\-ku}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{irye\-ku}
\fmtSubentry(a)
\fmtSense{to do next}
\fmtExampleRL{Ni na\-ma\-la e wa\-nge, nkwi\-rya\-ku waa\-mu.}
\fmtExampleEN{When I finish working at my home, I will work at yours next.}
\fmtSubentry(b)
\fmtSense{to have (a further child)}
\fmtExampleRL{Omwa\-na tu\-mwi\-rye\-ku.}
\fmtExampleEN{Let us have another child, please.}
\fmtSubentry(c)
\fmtSense{to remarry, to marry again, to marry (another woman)}
\fmtExampleRL{Omu\-ka\-li oku\-mwi\-rya\-ku di?}
\fmtExampleEN{When will you remarry?}

\fmtLemma{irya\-mu}
\fmtCitation{(okw)i\-rya\-mu}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{irye\-mu}
\fmtSubentry(a)
\fmtSense{to refill, to replace}
\fmtExampleRL{Irya\-mu amai\-zi omu ki\-do\-mo\-la.}
\fmtExampleEN{Refill the jerry can with water.}
\fmtSubentry(b)
\fmtSense{to replant}
\fmtExampleRL{Ebi\-duu\-ma bya\-nge bya\-gai\-ne oku\-me\-ra, nkwi\-rya\-mu bi\-ndi.}
\fmtExampleEN{My maize did not germinate well, so I am replanting.}
\fmtSubentry(c)
\fmtFixedexpression{irya\-mu (a)maa\-ni}
\fmtSense{to encourage, to support}
\fmtExampleRL{Aba\-wa\-gi\-zi bai\-rya\-mu amaa\-ni etii\-mu yaa\-bwe.}
\fmtExampleEN{The fans encourage their team.}

\fmtLemma{irya\-wo}
\fmtCitation{(okw)i\-rya\-wo}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{irye\-wo}
\fmtSubentry(a)
\fmtSense{to shut, to close}
\fmtExampleRL{Ai\-za eka nai\-rya\-wo olwi\-ge.}
\fmtExampleEN{He comes home and shuts the door.}
\fmtSubentry(b)
\fmtSense{to restore, to revive, to bring back}
\fmtExampleRL{Twi\-rye\-wo emi\-rwa, enge\-so n’e\-mi\-kii\-ro\-kii\-ro gyai\-swe.}
\fmtExampleEN{Let us revive our norms, values and traditions.}

\fmtLemma{irya\-yo}
\fmtCitation{(okw)i\-rya\-yo}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{irye\-yo}
\fmtSense{to take back, to return, to put back, to pay back}
\fmtExampleRL{Irya\-yo omu\-kee\-ka ogwo omu nyu\-mba.}
\fmtExampleEN{Take that mat back to the house.}

\fmtLemma{isa}
\fmtCitation{(e)i\-sa}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{(of a person) category, kind, type}
\fmtExampleRL{Nku\-te mu isa ki?}
\fmtExampleEN{What kind of a person are you?}

\fmtLemma{isaa\-ba\-tuu\-bi}
\fmtCitation{(o)i\-saa\-ba\-tuu\-bi}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{glutton}
\fmtExampleRL{Oyo omu\-sai\-za isaa\-ba\-tuu\-bi.}
\fmtExampleEN{That man is a glutton.}

\fmtLemma{isaa\-mi\-ro}
\fmtCitation{(e)i\-saa\-mi\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{bark cloth making workshop}
\fmtExampleRL{Ka\-twa\-be omu isaa\-mi\-ro nku\-saa\-mi\-re\-yo olu\-bu\-go.}
\fmtExampleEN{Let us go to the bark cloth making workshop and I will make bark cloth.}

\fmtLemma{isaa\-nii\-ko}
\fmtCitation{(e)i\-saa\-nii\-ko}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{used banana leaves}
\fmtExampleRL{Ei\-saa\-nii\-ko li\-ni li\-na\-ge omu lu\-su\-ku.}
\fmtExampleEN{Throw away the used banana leaves at the banana plantation.}

\fmtLemma{isaa\-si}
\fmtCitation{(e)i\-saa\-si}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{pity, mercy}
\fmtExampleRL{OYe\-su agi\-ra ei\-saa\-si oku ba\-ntu aba\-ku\-bo\-naa\-bo\-na.}
\fmtExampleEN{Jesus has mercy upon people who are suffering.}

\fmtLemma{isaa\-yi\-rii\-ryo}
\fmtCitation{(e)i\-sai\-ri\-rii\-ryo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{processing plant}
\fmtExampleRL{Opa\-mba tu\-mu\-twa\-le omu isaa\-yi\-rii\-ryo ba\-mu\-ko\-le\-mu enko\-si.}
\fmtExampleEN{Let us bring the cotton to the processing plant and they will make it into fibre.}

\fmtLemma{isaa\-zi}
\fmtCitation{(e)i\-saa\-zi}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{milking parlour}
\fmtExampleRL{Ya\-ba omu isaa\-zi ondee\-te\-re\-yo ama\-tai.}
\fmtExampleEN{Go to the milking parlour and bring me some milk.}

\fmtLemma{isa\-ba\-guu\-da}
\fmtCitation{(o)i\-sa\-ba\-guu\-da}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{richest person}
\fmtExampleRL{Oi\-sa\-ba\-guu\-da wa ku\-ni ni\-ye naa\-ni?}
\fmtExampleEN{Who is the richest person here?}

\fmtLemma{isa\-ba\-ka\-bo\-na}
\fmtCitation{(o)i\-sa\-ba\-ka\-bo\-na}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{high priest, archpriest}
\fmtExampleRL{Oisa\-ba\-ka\-bo\-na akwi\-za olwa\-lee\-ro oku\-taa\-ku abaa\-na emi\-ko\-no.}
\fmtExampleEN{The high priest is coming today to carry out the children's confirmation.}

\fmtLemma{isa\-ba\-la\-bi\-ri\-zi}
\fmtCitation{(o)i\-sa\-ba\-la\-bi\-ri\-zi}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{archbishop (of the Church of Uganda)}
\fmtExampleRL{Oi\-sa\-ba\-la\-bi\-ri\-zi ni\-ye akuu\-ra aba\-la\-bi\-ri\-zi.}
\fmtExampleEN{The Archbishop is the head of all the bishops.}

\fmtLemma{isa\-ba\-loo\-lee\-ri}
\fmtCitation{(o)i\-sa\-ba\-loo\-lee\-ri}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{archbishop (of the Church of Uganda)}

\fmtLemma{Isa\-ba\-nya\-la}
\fmtCitation{(o)Isa\-Ba\-nya\-la}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{Isabanyala (the traditional leader of Banyala)}
\fmtExampleRL{OIsa\-ba\-nya\-la li\-mwei oku ma\-ba\-ra ga mu\-ka\-ma wa Bu\-nya\-la.}
\fmtExampleEN{Isabanyala is one of the names given to the cultural leader of the Banyala.}

\fmtLemma{Isa\-ba\-ruu\-li}
\fmtCitation{(o)Isa\-ba\-ruu\-li}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{Isabaruuli (the traditional leader of Baluuri)}
\fmtExampleRL{OIsa\-ba\-ruu\-li li\-mwei oku ma\-ba\-ra ga mu\-ka\-ma wa Bu\-ruu\-li.}
\fmtExampleEN{Isabaruuli is one of the names given to the cultural leader of the Baruuli.}

\fmtLemma{isa\-ba\-wa\-ndii\-ki}
\fmtCitation{(o)i\-sa\-ba\-wa\-ndii\-ki}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{secretary general}
\fmtExampleRL{OBa\-la\-ba ni\-ye isa\-ba\-wa\-ndii\-ki wa Bu\-ka\-ma bwa Bu\-nya\-la.}
\fmtExampleEN{Balaba is the secretary general of the Bunyala kingdom.}

\fmtLemma{isa\-ga}
\fmtCitation{(e)i\-sa\-ga}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{twig}
\fmtExampleRL{Ku\-mbaa\-nia ama\-sa\-ga tu\-su\-mbe ocaa\-yi.}
\fmtExampleEN{Collect the twigs and we will make tea.}

\fmtLemma{isa\-ga\-ma}
\fmtCitation{(e)i\-sa\-ga\-ma}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{blood}
\fmtExampleRL{Aba\-ka\-ro\-mo\-ja ba\-lya ei\-sa\-ga\-ma lya nte.}
\fmtExampleEN{The Karamojong eat cow blood.}

\fmtLemma{isa\-ju}
\fmtCitation{(e)i\-sa\-ju}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{fat}
\fmtExampleRL{Abaa\-na ba\-ta\-ka mu\-no oku\-lya ama\-sa\-ju.}
\fmtExampleEN{Children like to eat fats very much.}

\fmtLemma{isa\-le}
\fmtCitation{(e)i\-sa\-le}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{sea shell}
\fmtExampleRL{Ama\-sa\-le gai\-ngi e Na\-ma\-sa\-le.}
\fmtExampleEN{There are many seashells at the Namasale landing site.}

\fmtLemma{isa\-ma\-ka}
\fmtCrossreference{\textnormal{see} isemaka}

\fmtLemma{isa\-mba}%
\fmtOne{1}
\fmtCitation{(o)i\-sa\-mba}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{eagle}
\fmtExampleRL{Oi\-sa\-mba alii\-re akaa\-na ka nko\-ko.}
\fmtExampleEN{The eagle ate the chick.}

\fmtLemma{isa\-mba}%
\fmtOne{2}
\fmtCitation{(e)i\-sa\-mba}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSource{from Sw.}
\fmtSense{sha\-mba, plantation}
\fmtExampleRL{Nta n’ei\-sa\-mba lya njaa\-ye.}
\fmtExampleEN{I have a marijuana plantation.}

\fmtLemma{isa\-mbi\-ro}
\fmtCitation{(e)i\-sa\-mbi\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{playground, football pitch}
\fmtExampleRL{Ei\-sa\-mbi\-ro lya mu\-pii\-ra li\-ni li\-fu\-ndi.}
\fmtExampleEN{This football pitch is small.}

\fmtLemma{isa\-mbi\-ryo}
\fmtCitation{(e)i\-sa\-mbi\-ryo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{spur (of a cock)}
\fmtExampleRL{Empa\-nga zi\-bba n’a\-ma\-sa\-mbi\-ryo oku ma\-gu\-lu.}
\fmtExampleEN{Cocks have spurs on their legs.}

\fmtLemma{Isa\-mbu\-la}
\fmtCitation{(e)Isa\-mbu\-la}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{July}
\fmtExampleRL{Ndii\-za omu kwe\-zi kwa Isa\-mbu\-la.}
\fmtExampleEN{I will come in July.}

\fmtLemma{isa\-mbya}
\fmtCitation{(e)i\-sa\-mbya}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{second rain season, second season for planting (which usually begins from around August)}
\fmtExampleRL{Nku\-bboo\-ka bi\-yaa\-ta bya isa\-mbya.}
\fmtExampleEN{I am planting potatoes for the second rainy season.}

\fmtLemma{isa\-mi}
\fmtCitation{(e)i\-sa\-mi}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{lake fly}
\fmtExampleRL{Ei\-sa\-mi li\-bba li\-ngi mu bi\-ro bya kya\-nda.}
\fmtExampleEN{There are too many lake flies during the dry season.}

\fmtLemma{isa\-na}%
\fmtOne{1}
\fmtCitation{(e)i\-sa\-na}
\fmtPOS{n}%
\fmtNC{5}
\fmtSubentry(a)
\fmtSense{sun, sunshine}
\fmtExampleRL{N’o\-lo\-la omu isa\-na ensaa\-le z’ai\-sa\-na zi\-kwi\-ta amai\-so.}
\fmtExampleEN{If you look directly into the sun, the rays will injure your eyes.}
\fmtSubentry(b)
\fmtSense{drought, dry spell}
\fmtExampleRL{Ei\-sa\-na ka lya\-ka\-li\-rii\-rye!}
\fmtExampleEN{The drought is excessive.}
\fmtSubentry(c)
\fmtFixedexpression{isa\-na li\-gwi\-re}
\fmtSense{at dusk}
\fmtExampleRL{Nai\-za nga ei\-sa\-na li\-gwi\-re.}
\fmtExampleEN{I will come at dusk.}
\fmtSubentry(d)
\fmtFixedexpression{isa\-na li\-mii\-re}
\fmtSense{at dusk}
\fmtExampleRL{Ya\-fii\-re nga ei\-sa\-na li\-mii\-re.}
\fmtExampleEN{He died at dusk.}
\fmtSubentry(e)
\fmtFixedexpression{isa\-na lya kya\-nda}
\fmtSense{dry season}
\fmtExampleRL{Omu ike\-ka tu\-bba n’ei\-sa\-na lya kya\-ndya.}
\fmtExampleEN{We are experiencing a dry season during the sweet potato slicing season.}
\fmtSubentry(f)
\fmtSense{hepatitis}
\fmtExampleRL{Oyo mu\-lwai\-re wa isa\-na.}
\fmtExampleEN{That person is suffering from hepatitis.}

\fmtLemma{isa\-na}%
\fmtOne{2}
\fmtCitation{(okw)i\-sa\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{isai\-ne}
\fmtSense{to look like, to be like, to resemble}
\fmtExampleRL{We oi\-sa\-na itee\-wo.}
\fmtExampleEN{You look like your dad.}

\fmtLemma{isa\-nga}
\fmtCitation{(e)i\-sa\-nga}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{tusk}
\fmtExampleRL{Onzo\-gi a\-bba n’a\-ma\-sa\-nga.}
\fmtExampleEN{The elephant has tusks.}

\fmtLemma{isa\-ni}
\fmtCitation{(e)i\-sa\-ni}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{statue, idol}
\fmtExampleRL{Ei\-sa\-ni lya Bi\-ki\-ra Ma\-ria li\-sai.}
\fmtExampleEN{The statue of Holy Mary is nice.}

\fmtLemma{isa\-nii\-sii\-rya}
\fmtCitation{(okw)i\-sa\-nii\-sii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{isa\-nii\-si\-rii\-rye}
\fmtSense{to look familiar, to look recognisable}
\fmtExampleRL{Eki\-fo ki\-ni nku\-ki\-sa\-nii\-sii\-rya.}
\fmtExampleEN{This place looks familiar to me.}

\fmtLemma{isa\-ni\-ka}
\fmtCitation{(okw)i\-sa\-ni\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{isa\-ni\-ki\-re}
\fmtSense{to look like, to resemble}

\fmtLemma{isa\-nsa}
\fmtCitation{(e)i\-sa\-nsa}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{type of wild fruit which grows in a cluster and is red when ripe}
\fmtSense{type of wild cooking matoke}

\fmtLemma{isa\-nso}
\fmtCitation{(e)i\-sa\-nso}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{twig}
\fmtExampleRL{Ama\-sa\-nso ga mu\-saa\-le ge\-nde\-ke\-re.}
\fmtExampleEN{The tree's twigs are broken.}

\fmtLemma{isa\-nyu}
\fmtCitation{(e)i\-sa\-nyu}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{happiness, joy}
\fmtExampleRL{Ota\-sii\-sa isa\-nyu lyai\-swe.}
\fmtExampleEN{Don’t ruin our happiness.}

\fmtLemma{isa\-ra}
\fmtCitation{(e)i\-sa\-ra}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{stale local brew}
\fmtExampleRL{Ti\-nta\-ka ku\-nywa isa\-ra.}
\fmtExampleEN{I don’t like drinking stale local brew.}

\fmtLemma{isa\-sa}
\fmtCitation{(e)i\-sa\-sa}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{blacksmith's workshop}
\fmtExampleRL{Omu\-ka\-li ali omu kwe\-zi tai\-ki\-ri\-zi\-bwa kwi\-ka\-la omu isa\-sa.}
\fmtExampleEN{A woman who is menstruating is not allowed to sit in a blacksmith's workshop.}

\fmtLemma{isa\-si}
\fmtCitation{(e)i\-sa\-si}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{bullet}
\fmtExampleRL{Ba\-ku\-bbi\-re ongi\-ri ei\-sa\-si.}
\fmtExampleEN{They shot the warthog with a bullet.}

\fmtLemma{isa\-tu}
\fmtCrossreference{used in isolation and with class 9 nouns, see (-)satu}
\fmtCitation{(e)i\-sa\-tu}
\fmtPOS{num}
\fmtSense{three}
\fmtExampleRL{Na\-mba isa\-tu ku gu\-mo\-so.}
\fmtExampleEN{Number three is on the left hand.}

\fmtLemma{isa\-za}
\fmtCitation{(e)i\-sa\-za}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{county}
\fmtExampleRL{Na\-bya\-lii\-bwe omu isa\-za lya Bu\-ko\-ma\-nsi\-mbi.}
\fmtExampleEN{I was born in Bukomansimbi county.}

\fmtLemma{ise\-ba\-lo\-ngo}
\fmtCitation{(o)i\-se\-ba\-lo\-ngo}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{father of twins}
\fmtExampleRL{Oi\-se\-ba\-lo\-ngo n’a\-fa, azi\-ki\-bwa igo\-loi\-go\-lo.}
\fmtExampleEN{When a father of twins dies, he is buried in the evening.}

\fmtLemma{ise\-bi\-ntu}
\fmtCitation{(o)i\-se\-bi\-ntu}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{property owner, rich person}
\fmtExampleRL{OBu\-gu\-rwa ni\-ye ise\-bi\-ntu wa ku\-ni.}
\fmtExampleEN{Bugurwa is the property owner in this area.}

\fmtLemma{ise\-ma\-ka}
\fmtCitation{(o)i\-se\-ma\-ka}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{family head}
\fmtExampleRL{Ise\-ma\-ka ye\-nkai ni\-ye ai\-ki\-ri\-zi\-bwa okwi\-kaa\-ra omu nte\-be.}
\fmtExampleEN{It is only the head of the family who is allowed to sit in the chair.}

\fmtLemma{ise\-ma\-kaa\-bi\-ri}
\fmtCitation{(o)i\-se\-ma\-kaa\-bi\-ri}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{polygamous man}
\fmtExampleRL{Oi\-se\-ma\-kaa\-bi\-ri ta\-bba na mi\-re\-mbe.}
\fmtExampleEN{A polygamous man is never at peace.}

\fmtLemma{ise\-ma\-tee\-ka}
\fmtCitation{(o)i\-se\-ma\-tee\-ka}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{constitution}
\fmtExampleRL{Oi\-se\-ma\-tee\-ka wai\-swe asai\-ne akuu\-mi\-bwe ku\-sai.}
\fmtExampleEN{Our constitution should be well protected.}

\fmtLemma{ise\-nde\-mwa\-ki}
\fmtCitation{(o)i\-se\-nde\-mwa\-ki}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{go-getter, achiever, high achiever, man/woman of action}
\fmtExampleRL{Abai\-se\-nde\-mwa\-ki ba\-bba ba\-wa\-ngu\-zi.}
\fmtExampleEN{Go-getters are always successful.}

\fmtLemma{ise\-nga}
\fmtCitation{(o)i\-se\-nga}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{paternal aunt}
\fmtExampleRL{Aya\-bi\-re n’oi\-se\-nga waa\-mwe.}
\fmtExampleEN{She has gone with her paternal aunt.}

\fmtLemma{ise\-nku\-lu}
\fmtCitation{(o)i\-se\-nku\-lu}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{leader, head}
\fmtExampleRL{Ni\-ye ise\-nku\-lu wa ki\-to\-ngo\-le kya bya\-bwo\-mi.}
\fmtExampleEN{He is the head of the health department.}

\fmtLemma{ise\-nte\-be}
\fmtCitation{(o)i\-se\-nte\-be}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{chairperson, chairman, chairwoman}
\fmtExampleRL{Ni\-ye ise\-nte\-be wa di\-si\-tu\-li\-ki\-ti ya Isi\-ngi\-ro.}
\fmtExampleEN{He is the chairman of Isingiro district.}

\fmtLemma{ishi}
\fmtPOS{interj}
\fmtSense{yuck, ew, ick (expression of strong distaste or disgust)}
\fmtExampleRL{Ishi, to\-so\-bo\-la ku\-cu\-lee\-ra!}
\fmtExampleEN{Yuck, can't you settle down!}

\fmtLemma{isi\-bya\-la}
\fmtCitation{(o)i\-si\-bya\-la}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{father-in-law}
\fmtExampleRL{Oi\-si\-bya\-la wa\-nge mu\-lwai\-re mu\-no.}
\fmtExampleEN{My father-in-law is very sick.}

\fmtLemma{isi\-ga}
\fmtCitation{(e)i\-si\-mba}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{planting season}

\fmtLemma{isi\-mba}%
\fmtOne{1}
\fmtCitation{(o)i\-si\-mba}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{genet}
\fmtExampleRL{Oi\-si\-mba alii\-re onko\-ko wa\-nge.}
\fmtExampleEN{The genet has eaten my hen.}

\fmtLemma{isi\-mba}%
\fmtOne{2}
\fmtCitation{(e)i\-si\-mba}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{planting season}
\fmtExampleRL{OMu\-se\-ve\-n ya\-bya\-lii\-bwe omu bi\-ro bya isi\-mba.}
\fmtExampleEN{Museven was born during the planting season.}

\fmtLemma{isi\-mbi}
\fmtCitation{(e)i\-si\-mbi}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{bedroom}
\fmtExampleRL{Ba\-kwi\-ra\-mu oku\-saa\-yii\-rya ei\-si\-mbi.}
\fmtExampleEN{They are redecorating the bedroom.}

\fmtLemma{Isi\-ra\-yi\-ri}
\fmtCitation{Isi\-ra\-yi\-ri}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{Israel}
\fmtExampleRL{OMu\-sa yaa\-kai\-rye abaa\-na ba Isi\-ra\-yi\-ri omu nya\-nja\-gi\-koo\-to eye\-te\-bwa Re\-d sea.}
\fmtExampleEN{Moses led the Israelites through the Red Sea.}

\fmtLemma{isi\-re}
\fmtPOS{adv}
\fmtSense{nonsensically}
\fmtExampleRL{Omu\-sai\-za oyo aba\-za isi\-re bu\-li ka\-see\-ra.}
\fmtExampleEN{That man always talks nonsensically.}

\fmtLemma{isi\-sye}
\fmtCitation{(o)i\-si\-sye}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{type of water weed which grows underneath water and is breeding ground for small fish}

\fmtLemma{iso\-goo\-ro}
\fmtCitation{(e)i\-so\-goo\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{brewing container}
\fmtExampleRL{Ei\-so\-goo\-ro lya\-nge lii\-zwi\-re ensa\-nde.}
\fmtExampleEN{My brewing container is full of juice.}

\fmtLemma{isoi\-gi}
\fmtCitation{(e)i\-soi\-gi}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{dry banana leaf}
\fmtExampleRL{Aba\-sa\-gi\-ki ba NRM ba\-zwa\-la ei\-soi\-gi nka aba\-syo\-li.}
\fmtExampleEN{NRM supporters wear dry banana leaves like night dancers.}

\fmtLemma{iso\-ke}
\fmtCitation{(e)i\-so\-ke}
\fmtPOS{n}%
\fmtNC{5}
\fmtSubentry(a)
\fmtSense{hair}
\fmtExampleRL{Ge\-mba ei\-so\-ke oku mu\-twe ogwa mwa\-na.}
\fmtExampleEN{Shave off the child's hair.}
\fmtSubentry(b)
\fmtFixedexpression{iso\-ke lya bi\-ra\-ndi}
\fmtSense{artificially curled hair}
\fmtExampleRL{Omu\-ka\-li wa\-nge ata n’ei\-so\-ke lya bi\-ra\-ndi.}
\fmtExampleEN{My wife has curled hair.}

\fmtLemma{iso\-mba}
\fmtCitation{(o)i\-so\-mba}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSubentry(a)
\fmtSense{giant pouched rat}
\fmtExampleRL{Ei\-so\-mba zi\-ma\-lii\-re ebi\-ko\-po.}
\fmtExampleEN{Giant pouched rats have carried away all my cups.}
\fmtSubentry(b)

\fmtLemma{iso\-me\-ro}
\fmtCitation{(e)i\-so\-me\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{school}
\fmtExampleRL{Ayo\-mbo\-ke\-re ei\-so\-me\-ro lya lu\-ka\-le.}
\fmtExampleEN{He built a community school.}

\fmtLemma{iso\-mo}
\fmtCitation{(e)i\-so\-mo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{subject, lesson}
\fmtExampleRL{Mu\-kwi\-ra mu iso\-mo ki?}
\fmtExampleEN{Which lesson are you going to?}

\fmtLemma{iso\-nge\-ryo}
\fmtCitation{(e)i\-so\-nge\-ryo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{canine tooth}
\fmtExampleRL{Ongi\-ri oni mu\-ku\-lu, ali\-na ei\-so\-nge\-ryo li\-na\-mu\-lai.}
\fmtExampleEN{This warthog is so old, it has a long canine tooth.}

\fmtLemma{iso\-ngi\-no}
\fmtCitation{(e)i\-so\-ngi\-no}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{hip}
\fmtExampleRL{ONa\-ncwe\-zi ata n’a\-ga\-so\-ngi\-no aga\-koo\-to.}
\fmtExampleEN{Nancwezi has big hips.}

\fmtLemma{iso\-ngo\-lyo}
\fmtCrossreference{\textnormal{see} isongeryo}

\fmtLemma{iso\-nko}
\fmtCitation{(e)i\-so\-nko}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{snail shell}
\fmtExampleRL{Eki\-ri\-ŋa\-li\-ŋa ki\-bba mu iso\-nko.}
\fmtExampleEN{A snail lives in a shell.}

\fmtLemma{iso\-roo\-ryo}
\fmtCitation{(e)i\-so\-roo\-ryo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{collection centre (\eg coffee, pineapple)}
\fmtExampleRL{Ei\-so\-roo\-ryo lya mwa\-ni li\-ri Kye\-ri\-ma.}
\fmtExampleEN{The coffee collection centre is in Kyerima.}

\fmtLemma{isu\-bi}
\fmtCitation{(e)i\-su\-bi}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{grass}
\fmtExampleRL{Ente te\-zi\-kya\-li\-na isu\-bi olwa kya\-nda.}
\fmtExampleEN{The cows no longer have grass because of the drought.}

\fmtLemma{isu\-mbi\-ro}
\fmtCitation{(e)i\-su\-mbi\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{kitchen}
\fmtExampleRL{Ba\-li mu isu\-mbi\-ro ba\-ku\-lya me\-re.}
\fmtExampleEN{They are in the kitchen eating.}

\fmtLemma{isu\-mo}
\fmtCrossreference{\textnormal{see} isumu}

\fmtLemma{isu\-mu}
\fmtCitation{(e)i\-su\-mu}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{spear}
\fmtExampleRL{Mpee\-rya ei\-su\-mu eri\-ndi.}
\fmtExampleEN{Give me another spear.}

\fmtLemma{isu\-nsuu\-ro}
\fmtCitation{(e)i\-su\-nsuu\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{ginnery}
\fmtExampleRL{Opa\-mpa mu\-twa\-le oku isu\-nsuu\-ro nga obwi\-re bu\-kya\-li kwa\-ka\-la.}
\fmtExampleEN{Take the cotton to the ginnery before it is too late.}

\fmtLemma{isuu\-ka}%
\fmtOne{1}
\fmtCitation{(okw)i\-suu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{isuu\-ki\-re}
\fmtSense{to congratulate}
\fmtExampleRL{Omwa\-na oyo mwi\-suu\-kye emi\-so\-mo.}
\fmtExampleEN{Congratulate that child for completing her studies.}

\fmtLemma{isuu\-ka}%
\fmtOne{2}
\fmtCitation{(okw)i\-suu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{isuu\-ki\-re}
\fmtSense{to recover (from an illness), to get better}
\fmtExampleRL{Yai\-suu\-ki\-re obu\-lwai\-re.}
\fmtExampleEN{He recovered from the illness.}

\fmtLemma{isuu\-ka}%
\fmtOne{3}
\fmtCitation{(e)i\-suu\-ka}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{bed sheet}
\fmtExampleRL{Nguu\-re ama\-suu\-ka ga\-yaa\-ka.}
\fmtExampleEN{I bought new bed sheets.}

\fmtLemma{isuu\-ka\-yo}
\fmtPOS{interj}
\fmtSense{Welcome back!}
\fmtExampleRL{Isuu\-ka\-yo oku iso\-me\-ro.}
\fmtExampleEN{Welcome back from school!}

\fmtLemma{isuu\-kya}
\fmtCitation{(okw)\-ei\-suu\-kya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{eisuu\-ki\-rye}
\fmtSense{to congratulate}
\fmtExampleRL{Twa\-be twi\-suu\-kye omwa\-na oku\-fu\-na edi\-gu\-li.}
\fmtExampleEN{Let’s go and congratulate the graduate.}

\fmtLemma{isuu\-ya}
\fmtCitation{(e)i\-suu\-ya}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{prognathism (abnormally extended jaw)}
\fmtExampleRL{Omu\-sai\-za ali\-na ama\-suu\-ya aga\-koo\-to.}
\fmtExampleEN{The man has a bad case of prognathism.}

\fmtLemma{iswe}
\fmtPOS{pro}
\fmtSense{we, us (first person plural independent pronoun)}
\fmtExampleRL{Iswe tu\-ku\-ta\-ka kwe\-ga omu ki\-fo kya kwe\-cwa\-cwa\-na.}
\fmtExampleEN{As for us, we want to learn instead of going on strike.}

\fmtLemma{iswei\-gi}
\fmtCitation{(e)i\-swei\-gi}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{dry banana leaf}
\fmtExampleRL{Omu\-lwai\-re ya\-bba\-sya\-nga ansi oku iswei\-gi.}
\fmtExampleEN{The sick person could sleep on dried banana leaves.}

\fmtLemma{iswe\-na}
\fmtPOS{pro}
\fmtSubentry(a)
\fmtSense{all of us}
\fmtExampleRL{Iswe\-na tu\-kwi\-za.}
\fmtExampleEN{All of us are coming.}
\fmtSubentry(b)
\fmtSense{we too, we as well}
\fmtExampleRL{Iswe\-na tu\-lya ama\-too\-ke bu\-li ki\-ro.}
\fmtExampleEN{We too eat matoke every day.}

\fmtLemma{-isy}
\fmtPOS{sf}
\fmtSense{causative suffix (extension) on verbs}

\fmtLemma{isya}
\fmtCitation{(okw)i\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{isi\-rye}
\fmtSense{to brew}
\fmtExampleRL{Ya\-yi\-si\-rye ama\-lwa aga\-sai aga ba\-ge\-ni.}
\fmtExampleEN{He brewed very nice finger millet brew for the visitors.}

\fmtLemma{isya\-du}
\fmtPOS{adv}
\fmtSense{scornfully, contemptuously, mockingly, in a sneering manner, in a superior manner}
\fmtExampleRL{Omwo\-jo oyo aba\-za isya\-du.}
\fmtExampleEN{That boy speaks scornfully.}

\fmtLemma{isye\-na}
\fmtPOS{pro}
\fmtSense{all of us}
\fmtExampleRL{Twa\-bbaa\-yo isye\-na.}
\fmtExampleEN{All of us will be there.}

\fmtLemma{isyo\-ma}
\fmtCitation{(e)i\-syo\-ma}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{creeping aerial yam}
\fmtExampleRL{Nei\-zi\-ra ma\-syo\-ma.}
\fmtExampleEN{I belong to the creeping aerial yam clan.}

\fmtLemma{ita}%
\fmtOne{1}
\fmtCitation{(okw)i\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iti\-re}
\fmtSubentry(a)
\fmtSense{to kill, to murder, to put to death}
\fmtExampleRL{Twi\-ta njo\-ka za ka\-suu\-me zo\-nkai.}
\fmtExampleEN{We only kill dangerous snakes.}
\fmtSubentry(b)
\fmtSense{to slaughter, to butcher, to kill, to slay}
\fmtExampleRL{Omu\-ge\-ni omu\-la\-ge oso\-bo\-la oku\-mwi\-ti\-ra ombu\-li.}
\fmtExampleEN{You can slaughter a goat for an invited guest.}
\fmtSubentry(c)
\fmtSense{to destroy}

\fmtLemma{ita}%
\fmtOne{2}
\fmtCitation{(okw)i\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iti\-re}
\fmtSense{to plant (sweet potato vines)}
\fmtExampleRL{Abaa\-la nywe\-na mwa\-be mu\-yi\-ti\-re ebi\-yaa\-ta.}
\fmtExampleEN{All of you girls, go and plant the potato vines.}

\fmtLemma{itaa\-gi}
\fmtCitation{(e)i\-taa\-gi}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{branch (\eg of a tree or organisation)}
\fmtExampleRL{Ei\-taa\-gi lya mu\-saa\-le lye\-nde\-ke\-re.}
\fmtExampleEN{The tree branch broke.}

\fmtLemma{itaa\-li}
\fmtCitation{(e)i\-taa\-li}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{maatali drums used primarily by Muslims at social and religious functions}
\fmtExampleRL{Aba\-si\-raa\-mu bu\-ku\-bba ama\-taa\-li oku mi\-ko\-lo.}
\fmtExampleEN{Moslems play mataali drums during functions.}

\fmtLemma{itaa\-nu}
\fmtCrossreference{used in isolation and with class 9 nouns, see (-)taanu}
\fmtCitation{(e)i\-taa\-nu}
\fmtPOS{num}
\fmtSense{five}

\fmtLemma{ita\-ba\-lo\-ngo}
\fmtCitation{(o)i\-ta\-ba\-lo\-ngo}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{father of twins}
\fmtExampleRL{Oi\-ta\-ba\-lo\-ngo ya\-ka\-zwa ani aya\-bi\-re Ma\-sa\-ka.}
\fmtExampleEN{The father of the twins has just left, he has gone to Masaka.}

\fmtLemma{ita\-bi}
\fmtCitation{(e)i\-ta\-bi}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{branch, bough, subdivision or section (of a company or association)}
\fmtExampleRL{Omu\-saa\-le gu\-ni gu\-li\-ku ama\-ta\-bi gai\-ngi.}
\fmtExampleEN{This tree has many branches.}

\fmtLemma{ita\-faa\-li}
\fmtCitation{(e)i\-ta\-faa\-li}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{brick}
\fmtExampleRL{Lee\-ta ani ei\-ta\-faa\-li li\-mwei.}
\fmtExampleEN{Bring one brick here.}

\fmtLemma{ita\-ka\-li}
\fmtCitation{(e)i\-ta\-ka\-li}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{land, soil, ground}
\fmtExampleRL{Nguu\-re ei\-ta\-ka\-li obu\-kai\-re iku\-mi.}
\fmtExampleEN{I bought the land for ten million.}

\fmtLemma{ita\-ko}
\fmtCitation{(e)i\-ta\-ko}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSource{from Sw.}
\fmtSense{buttock, cheek}
\fmtExampleRL{Bo\-na omu\-sai\-za ali\-na ama\-ta\-ko aga\-koo\-to.}
\fmtExampleEN{Look at the man with big buttocks.}

\fmtLemma{ita\-ma}
\fmtCitation{(e)i\-ta\-ma}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{cheek}
\fmtExampleRL{Omwa\-la ata n’a\-ma\-ta\-ma ga\-koo\-to.}
\fmtExampleEN{The girl has big cheeks.}
\fmtSubentry(b)
\fmtSense{blade (of a spear)}
\fmtExampleRL{Ei\-su\-mu lya\-mwe li\-ri\-ku ama\-ta\-ma ga\-ga\-zi.}
\fmtExampleEN{His spear has big blades.}

\fmtLemma{ita\-mbi\-ro}
\fmtCitation{(e)i\-ta\-mbi\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{hospital, clinic}
\fmtExampleRL{Omu\-lwai\-re mu\-mu\-twa\-le omu ita\-mbi\-ro.}
\fmtExampleEN{Take the patient to hospital.}
\fmtSubentry(b)
\fmtSense{altar}
\fmtExampleRL{Omu\-buu\-li\-zi ya\-fee\-rei\-re ku ita\-mbi\-ro.}
\fmtExampleEN{The priest died at the altar.}

\fmtLemma{ita\-mii\-ro}
\fmtCitation{(e)i\-ta\-mii\-ro}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{drunkenness}
\fmtExampleRL{Ei\-ta\-mii\-ro lya\-mu\-su\-li\-si\-rye mu bbaa\-la.}
\fmtExampleEN{Drunkenness made him sleep in the bar.}

\fmtLemma{ita\-mu\-to}
\fmtCitation{(o)i\-ta\-mu\-to}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{paternal uncle}
\fmtExampleRL{Na\-ku\-rii\-re wa ita mu\-to.}
\fmtExampleEN{I grew up in my paternal uncle's home.}

\fmtLemma{ita\-nga}
\fmtCitation{(e)i\-ta\-nga}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{tell-tale (a piece of fabric on a sailboat that shows the direction of the wind)}
\fmtExampleRL{Bbwa ei\-ta\-nga oku lya\-to tu\-so\-bo\-le oku\-ma\-nya ompu\-ngo gy’a\-ku\-li\-ga.}
\fmtExampleEN{Tie a tell-tale on the boat so that we know where the wind is blowing from.}

\fmtLemma{Ita\-nge}
\fmtCitation{(o)Ita\-nge}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{Father, God}
\fmtExampleRL{Ka\-to\-nda otaa\-ta, Ka\-to\-nda mwa\-na ne Ka\-to\-nda omwo\-yo akwe\-ra, amii\-na.}
\fmtExampleEN{God the Father, the Son, and the Holy Spirit, amen.}

\fmtLemma{ita\-ngi}
\fmtCrossreference{\textnormal{see} itaagi}

\fmtLemma{ite}
\fmtCitation{(o)i\-te}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{father}
\fmtExampleRL{Oi\-te wa\-nge ya\-li\-na ente bi\-nai.}
\fmtExampleEN{My father had four hundred cows.}

\fmtLemma{itee\-ka}
\fmtCitation{(e)i\-tee\-ka}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{law}
\fmtExampleRL{Ama\-tee\-ka ga Uga\-nda ba\-ga\-kyu\-sa\-kyu\-sa bu\-li bwi\-re.}
\fmtExampleEN{Uganda's laws are changed every now and then.}

\fmtLemma{itee\-wo}
\fmtCitation{(o)i\-tee\-wo}
\fmtPOS{n}%
\fmtNC{1}
\fmtSense{your father}
\fmtExampleRL{Ya\-ba oi\-mu\-le oi\-tee\-wo alye ekyai\-go\-lo.}
\fmtExampleEN{Go and wake up your father for supper.}

\fmtLemma{ite\-gu\-la}
\fmtCitation{(e)i\-te\-gu\-la}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{roofing tile}
\fmtExampleRL{Wa\-si\-gai\-re\-wo ei\-te\-gu\-la li\-mwei.}
\fmtExampleEN{There is one roofing tile left.}

\fmtLemma{ite\-mbo}
\fmtCitation{(e)i\-te\-mbo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{young banana plantation}
\fmtExampleRL{Ngo\-di aku\-zu\-ba mu ite\-mbo lya\-mwe.}
\fmtExampleEN{He is there weeding his young banana plantation.}

\fmtLemma{ite\-nde\-ke\-ro}
\fmtCitation{(e)i\-te\-nde\-ke\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{training institute, college, university}
\fmtExampleRL{Omwa\-na aya\-bi\-re mu ite\-nde\-ke\-ro lya bu\-lu\-mi.}
\fmtExampleEN{The child has joined the Institute of Agriculture.}

\fmtLemma{ite\-nte}
\fmtCitation{(e)i\-te\-nte}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{stubbornness, obstinacy}
\fmtExampleRL{Omwa\-na wa ite\-nte ba\-mu\-ku\-bbi\-re.}
\fmtExampleEN{The stubborn child was beaten.}

\fmtLemma{ite\-ri}
\fmtCitation{(o)i\-te\-ri}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{dayflower (\textit{Commelina})}
\fmtExampleRL{Oi\-te\-ri ya\-mee\-re ampi omu ki\-tu\-nga.}
\fmtExampleEN{A dayflower grew near the swamp.}

\fmtLemma{itii\-ra}
\fmtCitation{(okw)i\-tii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iti\-rii\-re}
\fmtSense{to irrigate, to water}
\fmtExampleRL{Itii\-ra ebi\-ru\-me bya\-mu le\-ke obi\-fu\-ne\-mu.}
\fmtExampleEN{Irrigate your crops if you want to benefit from them.}

\fmtLemma{iti\-ki\-ra}
\fmtCitation{(okw)i\-ti\-ki\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iti\-kii\-re}
\fmtSense{to be thorough, to be efficient}
\fmtExampleRL{Ti\-ya\-li mwi\-ti\-ki\-re omu mu\-li\-mu gwa\-mwe.}
\fmtExampleEN{He was not thorough in his work.}

\fmtLemma{iti\-ki\-ri}%
\fmtOne{1}
\fmtCitation{\textnormal{(e.g.}\,kye\-ti\-ki\-ri, byee\-ti\-ki\-ri, lwe\-ti\-ki\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSense{real, true, original}
\fmtExampleRL{Ga\-ni ama\-tai ga\-ti\-ki\-ri ti bi\-cu\-pu\-li ebi\-ri oku ka\-ta\-le.}
\fmtExampleEN{It is real milk, not the fake one on the market.}

\fmtLemma{iti\-ki\-ri}%
\fmtOne{2}
\fmtCitation{\textnormal{(e.g.}\,mwi\-ti\-ki\-ri, bee\-ti\-ki\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSense{dark-skinned, swarthy}
\fmtExampleRL{Oi\-te wa\-nge mu\-sai\-za mwi\-ti\-ki\-ri.}
\fmtExampleEN{My father is a dark-skinned man.}

\fmtLemma{iti\-mba}
\fmtCitation{(o)i\-ti\-mba}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{python}
\fmtExampleRL{Ya\-boi\-ne oi\-ti\-mba omu ki\-si\-ko ei\-go\-lo.}
\fmtExampleEN{He saw a python in the bush yesterday.}

\fmtLemma{iti\-nde}
\fmtCitation{(e)i\-ti\-nde}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{clod, lump of soil}
\fmtExampleRL{Omu\-lu\-mi aku\-te\-ma ma\-ti\-nde.}
\fmtExampleEN{The farmer is breaking clods of soil.}

\fmtLemma{iti\-ra}%
\fmtOne{1}
\fmtCitation{(okw)i\-ti\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{itii\-re}
\fmtSense{to have a running stomach, to have an upset stomach}
\fmtExampleRL{Omwa\-na wa\-nge akwi\-ti\-ra.}
\fmtExampleEN{My child has a running stomach.}

\fmtLemma{iti\-ra}%
\fmtOne{2}
\fmtCitation{(okw)i\-ti\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{itii\-re}
\fmtSense{to fumigate a container with grass}
\fmtExampleRL{Ekya\-nzi ba\-ki\-ti\-ra na ki\-cu\-nga.}
\fmtExampleEN{A milk gourd is cleaned with smoke from a small kiln.}

\fmtLemma{iti\-ra}%
\fmtOne{3}
\fmtCitation{(okw)i\-ti\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{itii\-re}
\fmtSense{to cull, to thin out}
\fmtExampleRL{Obwi\-ta ni bu\-bba bwi\-ngi ba\-bwi\-ti\-ra.}
\fmtExampleEN{When there are too many millet plants in one place, they are thinned out.}

\fmtLemma{iti\-ra\-na}
\fmtCitation{(okw)i\-ti\-ra\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iti\-rai\-ne}
\fmtSense{to meet face to face, to encounter}
\fmtExampleRL{Nyi\-ti\-rai\-ne n’o\-nyi\-bya\-la wa\-nge.}
\fmtExampleEN{I met my mother-in-law face-to-face.}

\fmtLemma{iti\-ri\-ga\-na}
\fmtCitation{(okw)i\-ti\-ri\-ga\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iti\-ri\-gai\-ne}
\fmtSense{to lose control}
\fmtExampleRL{Owa boo\-da\-boo\-da yai\-ti\-ri\-gai\-ne epi\-ki\-pi\-ki.}
\fmtExampleEN{The bodaboda driver lost control.}

\fmtLemma{ito\-faa\-li}
\fmtCrossreference{\textnormal{see} itafaali}

\fmtLemma{itoi\-go}
\fmtCitation{(e)i\-toi\-go}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{rainy season which lasts from March to June, first planting season}
\fmtExampleRL{Ebi\-yaa\-ta tu\-bi\-bboo\-ka kwa itoi\-go.}
\fmtExampleEN{We grow sweet potatoes during the rainy season.}

\fmtLemma{ito\-ndo}
\fmtCitation{(e)i\-to\-ndo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{drop, spot, blotch}
\fmtExampleRL{Ndoo\-wo wa\-de ei\-to\-ndo lya mai\-zi eree\-ro.}
\fmtExampleEN{There is not a single drop of water today.}

\fmtLemma{itoo\-ke}
\fmtCitation{(e)i\-too\-ke}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{banana}
\fmtExampleRL{Ba\-na\-yu\-ga\-nda ba\-ta\-ka mu\-no oku\-lya ama\-too\-ke.}
\fmtExampleEN{Ugandans really love eating matoke.}

\fmtLemma{itu\-bba}
\fmtCitation{(e)i\-tu\-bba}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{wooden plate}
\fmtExampleRL{Ama\-tu\-bba ga\-buu\-re ebi\-ro bi\-ni.}
\fmtExampleEN{Wooden plates are rare these days.}
\fmtSubentry(b)
\fmtSense{boat-shaped vessel used for making local brew}
\fmtExampleRL{Naa\-bya ei\-tu\-bba.}
\fmtExampleEN{Wash the vessel used for making the local brew.}

\fmtLemma{itu\-gu\-nda}
\fmtCitation{(e)i\-tu\-gu\-nda}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{\textit{Vangueria apiculata} (shrub or small tree with edible fruit)}
\fmtExampleRL{Ei\-tu\-gu\-nda ki\-ka kya ki\-ba\-la kya mu ki\-si\-ko eki\-ta ensi\-go egi\-gu\-mu.}
\fmtExampleEN{\textit{Vangueria apiculate} is a type of wild fruit with a very hard seed.}

\fmtLemma{itu\-ka}
\fmtCitation{(okw)i\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ituki\-re}
\fmtSense{to recur, to reappear, to happen repeatedly}
\fmtExampleRL{Obu\-lwai\-re bwa mu\-sai\-za oyo bwi\-tu\-ka bu\-li lu\-va\-nyu\-ma lwa kwe\-zi.}
\fmtExampleEN{That man’s illness recurs every month.}

\fmtLemma{itu\-ki\-rya}
\fmtCitation{(okw)i\-tu\-ki\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{itu\-kii\-rye}
\fmtSense{to reopen (an investigation, inquiry, case), to revisit, to repeat, to revive}
\fmtExampleRL{Epo\-lii\-si etu\-ki\-rye omu\-sa\-ngo.}
\fmtExampleEN{The police have reopened the investigation.}

\fmtLemma{itu\-la}%
\fmtOne{1}
\fmtCitation{(okw)i\-tu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{itwi\-re}
\fmtSubentry(a)
\fmtSense{to reopen (an investigation, inquiry, case), to revisit, to repeat, to revive}
\fmtExampleRL{Na\-bbai\-re mmai\-te nti omu\-sa\-ngo gwa\-wei\-re, ya\-gwi\-twi\-re bu\-yaa\-ka.}
\fmtExampleEN{I thought the case was closed, but he reopened it.}
\fmtSubentry(b)
\fmtSense{to exacerbate, to aggravate, to make worse}
\fmtExampleRL{Omwo\-jo yai\-twi\-re obu\-lwai\-re bwa maa\-mwe.}
\fmtExampleEN{The young man aggravated his mother's sickness.}

\fmtLemma{itu\-la}%
\fmtOne{2}
\fmtCitation{(okw)i\-tu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{itwi\-re}
\fmtSense{to beat patterns (into bark cloth with a finely grooved mallet as the last step in making bark cloth)}
\fmtExampleRL{Itu\-la olu\-bu\-go olwo lu\-go\-nde.}
\fmtExampleEN{Beat patterns into bark cloth with a finely grooved mallet.}

\fmtLemma{itu\-la}%
\fmtOne{3}
\fmtCitation{(e)i\-tu\-la}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{type of mallet with small curves}
\fmtExampleRL{Nko\-le\-sya itu\-la okwi\-tu\-la.}
\fmtExampleEN{I use a mallet with small curves to soften the bark cloth.}

\fmtLemma{itu\-lu\-ku\-ka}
\fmtCitation{(okw)i\-tu\-lu\-ku\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{itu\-lu\-ku\-ki\-re}
\fmtSubentry(a)
\fmtSense{(of water) to flow from (a container), to run out, to trickle down, to drip}
\fmtExampleRL{Amai\-zi ga\-kwi\-tu\-lu\-ku\-ka oku\-zwa omu ki\-do\-mo\-la.}
\fmtExampleEN{Water is running out of the jerry can.}
\fmtSubentry(b)
\fmtSense{(of tears) to flow}
\fmtSubentry(c)
\fmtSense{(of hair) to fall down}
\fmtSubentry(d)
\fmtSense{(of ghee, butter) to become clear}
\fmtExampleRL{Ama\-gi\-ta gai\-tu\-lu\-ku\-ki\-re ku\-sai mu\-no.}
\fmtExampleEN{The ghee has completely separated from the milk solids.}

\fmtLemma{itu\-lu\-la}
\fmtCitation{(okw)i\-tu\-lu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{itu\-lwi\-re}
\fmtSense{to decant, to transfer, to tip, to pour into, to pour out, to drain out, to strain}
\fmtExampleRL{Akwi\-tu\-lu\-la amai\-zi omu ko\-nte\-yi\-na egi\-ndi.}
\fmtExampleEN{She is decanting the water into another container.}

\fmtLemma{itu\-ma}
\fmtCitation{(e)i\-tu\-ma}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{meanness, nastiness, unkindness}
\fmtExampleRL{Aba\-ntu bee\-yo ba\-ta n’ei\-tu\-ma.}
\fmtExampleEN{People from that area are bad-hearted.}

\fmtLemma{itu\-mbi}
\fmtCitation{(e)i\-tu\-mbi}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{midnight}
\fmtExampleRL{Ya\-ta\-ndi\-ki\-re oku\-ta\-na\-ka mu itu\-mbi.}
\fmtExampleEN{He started vomiting at midnight.}

\fmtLemma{itu\-ndi}
\fmtCitation{(e)i\-tu\-ndi}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{high demand}
\fmtExampleRL{Ati\-bu\-ni amai\-do ga\-li ku itu\-ndi.}
\fmtExampleEN{These days groundnuts are in high demand.}

\fmtLemma{itu\-ndi\-ro}
\fmtCitation{(e)i\-tu\-ndi\-ro rya ma\-tu\-fa\-t}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{petrol station, fuel station, gas station, filling station}
\fmtExampleRL{Ei\-tu\-ndi\-ro ly’a\-ma\-fu\-ta li\-kwai\-te omuu\-ro.}
\fmtExampleEN{The petrol station has caught fire.}

\fmtLemma{itu\-ndu\-baa\-li}
\fmtCitation{(e)i\-tu\-ndu\-baa\-li}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{tarpaulin, tarp}
\fmtExampleRL{Atu\-gu\-lii\-re ei\-tu\-ndu\-baa\-li erya ku\-le\-ja amai\-zi.}
\fmtExampleEN{He bought us a tarpaulin for tapping water.}

\fmtLemma{itu\-ndu\-ro}
\fmtCrossreference{\textnormal{see} isumu}

\fmtLemma{itu\-ngo}
\fmtCitation{(e)i\-tu\-ngo}
\fmtPOS{n}%
\fmtNC{5}
\fmtSubentry(a)
\fmtSense{riches, wealth}
\fmtExampleRL{Omwo\-jo aka\-nyi\-rye ei\-tu\-ngo.}
\fmtExampleEN{That boy has accumulated a lot of wealth.}
\fmtSubentry(b)
\fmtSense{domestic animal (cattle, sheep or goats)}
\fmtExampleRL{Omu\-sai\-za ali\-na ei\-tu\-ngo lya bi\-so\-lo.}
\fmtExampleEN{That man has many domestic animals.}

\fmtLemma{itwa\-le}
\fmtCitation{(e)i\-twa\-le}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{administrative unit in the Islamic faith at parish level}
\fmtExampleRL{Shei\-kh Mu\-sa ni\-ye akuu\-ra ei\-twa\-le lya Bu\-ko\-ma\-nsi\-mbi.}
\fmtExampleEN{Sheikh Musa is the one who leads the parish administrative unit for Muslims in Bukomansibi.}
\fmtSubentry(b)
\fmtSense{constituency}
\fmtExampleRL{Omu Ka\-yu\-nga mu\-li\-mu ama\-twa\-le asa\-tu.}
\fmtExampleEN{There are three constituencies in Kayunga.}
\fmtSubentry(c)
\fmtSense{colony, dependency, protectorate}
\fmtExampleRL{Ke\-nya lya\-li itwa\-le lya Bu\-nge\-re\-za.}
\fmtExampleEN{Kenya was a British colony.}

\fmtLemma{iwa}
\fmtCitation{(e)i\-wa}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{thorn, prickle, spike}
\fmtExampleRL{Ama\-wa ga\-nsu\-mi\-ti\-re nga nku\-zwa omu ndu\-mi\-ro.}
\fmtExampleEN{I was pricked by thorns on my way back from the garden.}

\fmtLemma{iwa\-ko}
\fmtCitation{(e)i\-wa\-ko}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{pregnancy (of an animal), conception}
\fmtExampleRL{Ombu\-li oyo ali\-na ei\-wa\-ko ti\-mu\-ku\-so\-bo\-la ku\-mu\-baa\-ga.}
\fmtExampleEN{That goat is pregnant, you cannot have it slaughtered.}

\fmtLemma{iwa\-ndii\-ki\-ro}
\fmtCitation{(e)i\-wa\-ndii\-ki\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{secretariat}
\fmtExampleRL{Ei\-wa\-ndii\-ki\-ro eri\-ku\-lu erya NRM li\-ri ku nzi\-ra ya Kya\-do\-ndo.}
\fmtExampleEN{The NRM secretariat is along the Kyadondo road.}

\fmtLemma{iwa\-ni\-ka}
\fmtCitation{(e)i\-wa\-ni\-ka}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{treasury}
\fmtExampleRL{Omu iwa\-ni\-ka lya Isa\-ba\-ruu\-li mu\-li\-mu empii\-ya zi\-ngi.}
\fmtExampleEN{There is a lot of money in the Isabaruuli's treasury.}
\fmtSubentry(b)
\fmtSense{mortuary, morgue}
\fmtExampleRL{Emi\-ra\-mbo egi\-ri omu iwa\-ni\-ka gi\-ku\-ju\-nda.}
\fmtExampleEN{The bodies in the mortuary are rotting.}

\fmtLemma{iwe\-se\-ro}
\fmtCitation{(e)i\-we\-se\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{metal fabrication workshop, blacksmith's workshop or plant}
\fmtExampleRL{Omu iwe\-se\-ro lya Mu\-ka\-la\-zi mu\-li\-mu enyi\-ge eza byo\-ma zi\-ngi.}
\fmtExampleEN{There is a number of metal doors in Mukalazi's metal fabrication workshop.}

\fmtLemma{iwo\-lu}
\fmtCitation{(e)i\-wo\-lu}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{leftovers}
\fmtExampleRL{Mpee\-rya oku iwo\-lu ndye\-ku.}
\fmtExampleEN{Give me the leftovers and I'll eat some.}

\fmtLemma{iwo\-ndo}
\fmtCitation{(e)i\-wo\-ndo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{coconut-like wild fruit}
\fmtExampleRL{Enke\-mbo zi\-ta\-ka mu\-no oku\-lya ama\-wo\-ndo.}
\fmtExampleEN{Monkeys are fond of eating coconut-like wild fruit.}

\fmtLemma{iwoo\-bee\-si}
\fmtCitation{(e)i\-woo\-bee\-si}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{hook}
\fmtExampleRL{Ama\-wo\-bee\-si ga\-te\-ga ncwi.}
\fmtExampleEN{Hooks are used to catch fish.}

\fmtLemma{iwoo\-le}
\fmtCitation{(e)i\-woo\-le}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{stubbornness, obstinacy}
\fmtExampleRL{Ei\-woo\-le lya\-mwe lya\-mu\-lu\-gii\-re\-ku oku\-fee\-rwa omu\-lu\-mu.}
\fmtExampleEN{His stubbornness cost him his job.}

\fmtLemma{iwoo\-ryo}
\fmtCitation{(e)i\-woo\-ryo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{court of justice}
\fmtExampleRL{Ya\-le\-ge\-re omu\-sa\-ngo omu iwoo\-ryo.}
\fmtExampleEN{He reported the case at the court of law.}

\fmtLemma{iwo\-woo\-lo}
\fmtCitation{(e)i\-wo\-woo\-lo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{lung (of an animal)}
\fmtExampleRL{Nta\-ka mu\-no oku\-lya ama\-woo\-woo\-lo.}
\fmtExampleEN{I like eating animal lungs.}

\fmtLemma{iwu\-la}%
\fmtOne{1}
\fmtCitation{(okw)i\-wu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iwii\-re}
\fmtSense{to dig out, to scoop out, to excavate}
\fmtExampleRL{Yi\-wu\-la ei\-ta\-ka\-li oku\-zwa omu kii\-na tu\-bu\-mbe ebbu\-loo\-ka.}
\fmtExampleEN{Dig out the soil so that we can lay bricks.}

\fmtLemma{iwu\-la}%
\fmtOne{2}
\fmtCrossreference{\textnormal{see} ibura}

\fmtLemma{iwu\-li}
\fmtCitation{(e)i\-wu\-li}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{egg}
\fmtExampleRL{Onko\-ko ana\-gi\-re ama\-wu\-li gai\-ngi.}
\fmtExampleEN{The hen laid many eggs.}
\fmtSubentry(b)
\fmtFixedexpression{iwu\-li li\-zi\-rii\-rwa}
\fmtSense{unhatched egg}
\fmtExampleRL{Ei\-wu\-li eri\-zi\-rii\-rwa lya\-ti\-ki\-re.}
\fmtExampleEN{The unhatched egg cracked.}

\fmtLemma{iwu\-we}
\fmtCitation{(e)i\-wu\-we}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{lung (of a human being)}
\fmtExampleRL{Omu\-ntu a\-bba n’a\-ma\-wu\-we abi\-ri.}
\fmtExampleEN{A human being has two lungs.}

\fmtLemma{iya}%
\fmtOne{1}
\fmtCitation{(okw)i\-ya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iyi\-re}
\fmtSubentry(a)
\fmtSense{to get, to obtain, to acqui\-re}
\fmtExampleRL{OKa\-wee\-si empii\-ya azii\-ya mu ku\-te\-ga bi\-na\-ge.}
\fmtExampleEN{Kaweesi gets money from fishing tilapia.}
\fmtSubentry(b)
\fmtSense{to take away, to remove}
\fmtExampleRL{Iya\-wo eki\-do\-mo\-la ki\-ni oki\-twa\-le omu nyu\-mba.}
\fmtExampleEN{Take away this jerry can and take it into the house.}
\fmtSubentry(c)
\fmtSense{to uproot, to root out}
\fmtExampleRL{Naa\-ni eya\-yii\-re endo\-kwai za\-nge oku\-zwa mu bbee\-di?}
\fmtExampleEN{Who uprooted my trees?}
\fmtSubentry(d)
\fmtSense{to harvest (by uprooting, \eg groundnut, mushroom)}
\fmtExampleRL{Baa\-bi\-re kwi\-ya mai\-do.}
\fmtExampleEN{They have gone to harvest groundnuts.}
\fmtSubentry(e)
\fmtSense{(in agriculture) to thin out, to trim}
\fmtExampleRL{Aya\-bi\-re kwi\-ya pa\-mba.}
\fmtExampleEN{He has gone to thin out the cotton plants.}
\fmtSubentry(f)
\fmtSense{to set free, to release}
\fmtExampleRL{Omu\-bbwe ba\-mwi\-yi\-re omu nko\-myo.}
\fmtExampleEN{The prisoner has been released from prison.}

\fmtLemma{iya}%
\fmtOne{2}
\fmtCitation{(okw)i\-ya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iyi\-re}
\fmtSense{to toddle, to take the first step to walk}
\fmtExampleRL{Omwa\-na ata\-ndi\-ki\-re okwi\-ya ama\-gu\-lu.}
\fmtExampleEN{The baby has taken its first steps.}

\fmtLemma{iya}%
\fmtOne{3}
\fmtCitation{(e)i\-ya}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{shelf}
\fmtExampleRL{Ama\-duu\-ka ga\-bbaa\-mu ama\-ya.}
\fmtExampleEN{Shops have shelves.}

\fmtLemma{iya}%
\fmtOne{4}
\fmtCitation{(e)i\-ya}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{axe}
\fmtExampleRL{Ei\-ya lya mu\-bai\-zi lya\-buu\-re.}
\fmtExampleEN{The carpenter’s axe is lost.}

\fmtLemma{iyai}
\fmtPOS{interj}
\fmtSense{goodness (expression of surprise)}
\fmtExampleRL{Iyai! Omwa\-na anyi\-ru\-ki\-re\-ku.}
\fmtExampleEN{Goodness! The child has run away.}

\fmtLemma{iya\-ku}
\fmtCitation{(okw)i\-ya\-ku}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iyi\-re\-ku}
\fmtSubentry(a)
\fmtSense{to take away, to take part of, to take off a piece}
\fmtExampleRL{Yayi\-re\-ku ebi\-fi bi\-sa\-tu ebya nya\-ma n’a\-bi\-ga\-bi\-ra abaa\-li omu bwe\-taa\-vu.}
\fmtExampleEN{She took away three pieces of meat and shared them with the needy.}
\fmtSubentry(b)
\fmtSense{to remove (\eg a leader), to overthrow}
\fmtExampleRL{Baa\-mwi\-re\-ku ewoo\-fii\-si yaa\-mwe lwa bu\-nya\-ku\-li.}
\fmtExampleEN{He was removed from his office because of corruption.}

\fmtLemma{iya\-li}
\fmtCitation{(e)i\-ya\-li}
\fmtPOS{n}%
\fmtNC{5}
\fmtSubentry(a)
\fmtSense{envy}
\fmtExampleRL{Aba\-ntu aba\-ta n’ei\-ya\-li ti\-ba\-lya\-ba mu igu\-ru.}
\fmtExampleEN{The envious will never go to heaven.}
\fmtSubentry(b)
\fmtSense{jealousy}
\fmtExampleRL{Oi\-ba waa\-mwe ata n’ei\-ya\-li.}
\fmtExampleEN{Her husband is jealous. (lit.\,Her husband has jealousy.)}

\fmtLemma{iya\-mu}
\fmtCitation{(okw)i\-ya\-mu}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iyi\-re\-mu}
\fmtSense{to remove, to take off}
\fmtExampleRL{Iya\-mu enkai\-to zaa\-mu le\-ke oi\-ngi\-re omu\-zi\-gi\-ti.}
\fmtExampleEN{Remove your shoes and then enter the mosque.}

\fmtLemma{iya\-na}
\fmtCitation{(okw)i\-ya\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{iya\-ni\-re}
\fmtSense{to make oneself look important}
\fmtExampleRL{Lwa\-ki wee\-yi\-ya\-na?}
\fmtExampleEN{Why do you want to make yourself look important?}

\fmtLemma{iya\-nga}%
\fmtOne{1}
\fmtCitation{(e)i\-ya\-nga}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{indigenous group, ethnic group, race, tribe}
\fmtExampleRL{Uga\-nda eri\-na ama\-nga gai\-ngi.}
\fmtExampleEN{Uganda has many different indigenous groups.}
\fmtSubentry(b)
\fmtSense{nation, country, state}
\fmtExampleRL{Ei\-ya\-nga lya Ke\-nya lii\-zi\-re\-mu aka\-ca\-mbai\-tu\-ko.}
\fmtExampleEN{The country of Kenya is unstable.}

\fmtLemma{iya\-ngwe}
\fmtCitation{(e)i\-ya\-ngwe}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{noon, midday}
\fmtExampleRL{Lwa\-ki ota\-mbwi\-re omu iya\-ngwe?}
\fmtExampleEN{Why did you travel at noon?}

\fmtLemma{iya\-no}
\fmtCitation{(e)i\-ya\-no}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{taboo, culturally or socially unacceptable act}
\fmtExampleRL{Ei\-rai omu\-ka\-li oku\-lya ondu ya\-bba\-nga akoo\-re iya\-no.}
\fmtExampleEN{Long ago it was culturally inappropriate for a woman to eat a lungfish.}
\fmtSubentry(b)
\fmtSense{terrible event, catastrophe, tragedy}
\fmtExampleRL{Ga\-bba maa\-no oi\-te wa mwa\-na oku\-mu\-twe\-ka enda.}
\fmtExampleEN{It is a tragedy when a father impregnates his own daughter.}

\fmtLemma{iya\-wo}
\fmtCitation{(okw)i\-ya\-wo}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{iyi\-re\-wo}
\fmtSense{to remove, to take away, to put away}
\fmtExampleRL{Iya\-wo ebi\-ta\-bo bya\-mu oku mee\-za.}
\fmtExampleEN{Remove your books from the table.}

\fmtLemma{iye}%
\fmtOne{1}
\fmtCitation{(e)i\-ye}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{army, regiment, armed force}
\fmtExampleRL{Ei\-ye lya Uga\-nda li\-te\-nde\-ke ku\-sai.}
\fmtExampleEN{Uganda's army is professionally trained.}

\fmtLemma{iye}%
\fmtOne{2}
\fmtPOS{interj}
\fmtSense{yes (expresses agreement)}
\fmtExampleRL{Iye, njee\-na nkwi\-za ku\-bbaa\-yo.}
\fmtExampleEN{Yes, I will also be there.}

\fmtLemma{iye\-mbe}%
\fmtOne{1}
\fmtCitation{(e)i\-ye\-mbe}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{fetish, spirit associated with a fetish, spell, hex}
\fmtExampleRL{Aba\-ntu aba\-mwe e Ka\-yu\-nga ba\-ta n’a\-me\-mbe.}
\fmtExampleEN{Some people in Kayunga have fetishes.}

\fmtLemma{iye\-mbe}%
\fmtOne{2}
\fmtCitation{(e)i\-ye\-mbe}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{horn}
\fmtExampleRL{Ame\-mbe ga nte wa\-nge ga\-so\-ngo\-li.}
\fmtExampleEN{My cow's horns are straight.}

\fmtLemma{iye\-ngo}
\fmtCitation{(e)i\-ye\-ngo}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{wave}
\fmtExampleRL{Ei\-ye\-ngo li\-ku\-bbi\-re akaa\-to ni kaa\-ti\-ka.}
\fmtExampleEN{The wave hit the canoe and it broke.}
\fmtSubentry(b)
\fmtSense{air wave}
\fmtExampleRL{Ama\-ra\-ka ga la\-di\-yo ga\-ka\-la mu ma\-ye\-ngo.}
\fmtExampleEN{Radio sounds are transmitted through air waves.}

\fmtLemma{iye\-rya}
\fmtCitation{(e)i\-ye\-rya}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{harvest, harvesting season}
\fmtExampleRL{Tu\-li mu iye\-rya lya me\-re.}
\fmtExampleEN{We are in the harvesting season.}

\fmtLemma{iyi\-mbe}
\fmtCrossreference{\textnormal{see} yimbe}

\fmtLemma{iyoo\-ro}
\fmtCitation{(e)i\-yoo\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{hardcore stone}
\fmtExampleRL{Ba\-lo\-nde\-re ama\-yoo\-ro oku\-zwa omu ki\-ro\-mbe.}
\fmtExampleEN{They picked hardcore stones from the stone quarry.}

\fmtLemma{iyo\-roo\-ro}
\fmtCitation{(e)i\-yo\-roo\-ro}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{nursery bed}
\fmtExampleRL{Ewai\-swe wa\-li\-yo ei\-yo\-roo\-ro lya mi\-cu\-ngwai.}
\fmtExampleEN{There is a nursery bed of orange trees in our area.}

\fmtLemma{iyu\-te}
\fmtCitation{(e)i\-yu\-te}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{boil, pustule}
\fmtExampleRL{Alwai\-re ei\-yu\-te oku ita\-ma.}
\fmtExampleEN{He has a boil on his cheek.}

\fmtLemma{iyuu\-ni}
\fmtCitation{(e)i\-yuu\-ni}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{yam}
\fmtExampleRL{Ama\-yuu\-ni ge ba\-su\-mbi\-re eree\-ro ti\-ga\-yii\-re.}
\fmtExampleEN{The yams they prepared today were not well cooked.}

\fmtLemma{iza}
\fmtCitation{(okw)i\-za}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{izi\-re}
\fmtSubentry(a)
\fmtSense{to come, to appear, to show up}
\fmtExampleRL{Ai\-za bu\-li lwa Sa\-nde n’a\-nkya\-ra.}
\fmtExampleEN{Every Sunday, she comes here and visits me.}
\fmtSubentry(b)
\fmtSense{will, shall (used to build remote habitual future)}
\fmtExampleRL{Ebi\-ko\-lwa bya\-mwe ebya ki\-sa bi\-kwi\-za kwi\-zu\-ki\-rwa\-nga.}
\fmtExampleEN{His acts of kindness will always be remembered.}

\fmtLemma{iza\-za}
\fmtCitation{(e)i\-za\-za}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{shrine}
\fmtExampleRL{Ei\-za\-za lya Ba\-yi\-nda li\-ri ku Iru\-ba lya Na\-ki\-to\-ko\-lo.}
\fmtExampleEN{The Bayinda's shrine is found at Nakitokolo.}

\fmtLemma{izi\-moo\-ra}
\fmtCitation{(e)i\-zi\-moo\-ra}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{stretch mark}
\fmtExampleRL{Omu\-ka\-li oyo ali\-na ama\-zi\-moo\-ra oku nda.}
\fmtExampleEN{That woman has stretch marks on her stomach.}

\fmtLemma{izo}
\fmtCitation{(e)i\-zo}
\fmtPOS{adv}
\fmtSubentry(a)
\fmtSense{tomorrow, yesterday, day from now}
\fmtExampleRL{OJo\-y alii\-za izo.}
\fmtExampleEN{Joy will come tomorrow.}
\fmtSubentry(b)
\fmtFixedexpression{izo ma\-mbya}
\fmtSense{day after tomorrow, day before yesterday, two days from today}
\fmtExampleRL{Yai\-za izo ma\-mbya.}
\fmtExampleEN{He will come the day after tomorrow.}

\fmtLemma{izo\-bo\-ne}
\fmtCitation{(e)i\-zo\-bo\-ne}
\fmtPOS{n}
\fmtSense{colourfulness}
\fmtExampleRL{Emba\-ga izo\-bo\-ne ti lya\-ngu ku\-tee\-ka\-tee\-ka nga to\-li\-na se\-nte.}
\fmtExampleEN{A colourful wedding is not easy to organise when you do not have money.}

\fmtLemma{izu\-ba}%
\fmtOne{1}
\fmtCitation{(e)i\-zu\-ba}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{weed}
\fmtExampleRL{Ei\-zu\-ba lii\-ti\-re ebi\-duu\-ma.}
\fmtExampleEN{The weed was destroyed.}

\fmtLemma{izu\-ba}%
\fmtOne{2}
\fmtCitation{(e)i\-zu\-ba}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{well}
\fmtExampleRL{Ei\-zu\-ba lyai\-swe lya\-kaa\-re olwa kya\-nda.}
\fmtExampleEN{Our well has dried up because of the drought.}

\fmtLemma{izu\-ki\-ra}
\fmtCitation{(okw)i\-zu\-ki\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{izu\-kii\-re}
\fmtSense{to remember, to recall}
\fmtExampleRL{Okwi\-zu\-ki\-ra kye na\-ku\-ko\-be\-re?}
\fmtExampleEN{Do you remember what I told you?}

\fmtLemma{izu\-ki\-rya}
\fmtCitation{(okw)i\-zu\-ki\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{izu\-kii\-rye}
\fmtSense{to remind, to make one think of, to cause one to remember}
\fmtExampleRL{Taa\-yo ei\-ga\-ni okwi\-zu\-ki\-rya abaa\-na abo oku\-so\-ma.}
\fmtExampleEN{Make sure that you remind those children to study.}

\fmtLemma{izu\-ku\-rya}
\fmtCitation{(okw)i\-zu\-ku\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{izu\-ku\-rye}
\fmtSense{to become a grandparent, to have grandchildren}
\fmtExampleRL{Ata\-ndi\-ki\-re okwi\-zu\-ku\-rya.}
\fmtExampleEN{He became a grand\-parent.}

\fmtLemma{izu\-la}
\fmtCitation{(okw)i\-zu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{izwi\-re}
\fmtSense{to be full with, to be filled with}
\fmtExampleRL{Jaa\-ga ei\-zwi\-re amai\-zi.}
\fmtExampleEN{The jug is full of water.}

\fmtLemma{izu\-li}
\fmtCitation{\textnormal{(e.g.}\,kii\-zu\-li, gii\-zu\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSense{full}
\fmtExampleRL{Mpa eki\-ko\-po kii\-zu\-li.}
\fmtExampleEN{Give me a full cup.}

\fmtLemma{izu\-lya}
\fmtCitation{(okw)i\-zu\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{izwi\-rye}
\fmtSubentry(a)
\fmtSense{to fill, to make full, to charge}
\fmtExampleRL{Yai\-zwi\-rye ejaa\-ga ama\-tai.}
\fmtExampleEN{She filled the jug with milk.}
\fmtSubentry(b)
\fmtSense{to flood with, to overflow with, to become submerged with water}
\fmtExampleRL{OKa\-mpa\-la ai\-zu\-lya amai\-zi omu bi\-ro bya ike\-ndi.}
\fmtExampleEN{Kampala floods during the rainy season.}

\fmtLemma{izwe\-ra}
\fmtCitation{(okw)i\-zwe\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{izwei\-re}
\fmtSense{(to lock with a security bar across the door}
\fmtExampleRL{Naa\-ni yai\-zwe\-ra olwi\-ge lwa\-nge?}
\fmtExampleEN{Who locked my door with a security bar?}

\fmtLemma{izwe\-ri}
\fmtCitation{(e)izwe\-ri}
\fmtPOS{adv}
\fmtSubentry(a)
\fmtSense{day after tomorrow, day before yesterday, two days from today}
\fmtExampleRL{Abaa\-na ba\-lya\-ba oku iso\-me\-ro izwe\-ri.}
\fmtExampleEN{The children will go to school the day after tomorrow.}
\fmtSubentry(b)
\fmtSense{recently, not long ago, a short time ago}
\fmtExampleRL{Yaa\-bi\-re izwe\-ri.}
\fmtExampleEN{He has left recently.}
\end{letter}