\begin{letter}{J}

\fmtLemma{jaa\-ga}
\fmtCitation{(e)jaa\-ga}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{jug, pitcher}
\fmtExampleRL{Ejaa\-ga gye na\-boi\-ne ya\-bbai\-re gye\-ree\-re.}
\fmtExampleEN{The jug that I saw was empty.}

\fmtLemma{jaa\-mu}
\fmtCitation{(o)jaa\-mu}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{traffic jam}
\fmtExampleRL{E Ka\-mpa\-la wa\-bbaa\-yo ojaa\-mu mwi\-ngi.}
\fmtExampleEN{There are always a lot of traffic jams in Kampala.}

\fmtLemma{jaa\-nya}
\fmtCitation{(oku)\-jaa\-nya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{jaa\-nyi\-re}
\fmtSubentry(a)
\fmtSense{to worsen, to get worse}
\fmtExampleRL{Omu\-lwai\-re ba\-mu\-twi\-re bu\-yi\-ndi nga embee\-ra ejaa\-nyi\-re.}
\fmtExampleEN{The patient's condition worsened, and he was flown to India.}
\fmtSubentry(b)
\fmtSense{to go septic, to become infected, to fester}
\fmtExampleRL{Ekyo\-ya ki\-jaa\-nyi\-re.}
\fmtExampleEN{The wound became infected.}

\fmtLemma{jaa\-ruu\-ka}
\fmtCitation{(oku)\-jaa\-ru\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{jaa\-ru\-ki\-re}
\fmtSense{to be diluted, to be thinned down}
\fmtExampleRL{Yo\-nge\-ra\-mu amai\-zi omu bi\-yi\-mba bi\-jaa\-ruu\-ke.}
\fmtExampleEN{Add water to the beans to dilute them.)}

\fmtLemma{jaa\-ruu\-la}
\fmtCitation{(oku)\-jaa\-ruu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{jaa\-ru\-lwi\-re}
\fmtSense{to dilute, to thin down}
\fmtExampleRL{Ajaa\-ruu\-lwi\-re omu\-ku\-bi.}
\fmtExampleEN{She diluted the sauce.}

\fmtLemma{jaa\-ta}
\fmtCitation{(oku)\-jaa\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{jaa\-ti\-re}
\fmtSubentry(a)
\fmtSense{to be slippery, to be smooth}
\fmtExampleRL{Oi\-ke\-ndi atoo\-nye\-re mwi\-ngi olu\-bu\-ga ni\-ru\-jaa\-ta.}
\fmtExampleEN{It rained heavily and the compound became slippery.}
\fmtSubentry(b)
\fmtSense{to slip, to slide, to glide}

\fmtLemma{ja\-bbu\-la}
\fmtCitation{(oku)\-ja\-bbu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ja\-bbwi\-re}
\fmtSubentry(a)
\fmtSense{to dig hurriedly by leaving some weeds in the ground}
\fmtExampleRL{Aja\-bbwi\-re\-mbe olu\-bbi\-mbi.}
\fmtExampleEN{He just dug hurriedly.}
\fmtSubentry(b)
\fmtSense{to do in a hasty manner, to do in a slapdash manner}
\fmtExampleRL{Omu ru\-ba\-zo lwa\-nge nkwa\-ba ku\-ja\-bbu\-la\-ja\-bbu\-la\-mbe.}
\fmtExampleEN{I will only make a quick summary of the key points in my speech.}

\fmtLemma{ja\-ga\-la\-la}
\fmtCitation{(oku)\-ja\-ga\-la\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ja\-ga\-lai\-re}
\fmtSource{from Lug.} 
\fmtSubentry(a)
\fmtSense{(of hair) to be unkempt, to be messy, to stand up straight and stiff}
\fmtExampleRL{Ei\-so\-ke li\-mu\-ja\-ga\-lai\-re oku mu\-twe.}
\fmtExampleEN{The hair looks unkempt.}
\fmtSubentry(b)
\fmtSense{to be in uproar, to protest, to be in turmoil}
\fmtExampleRL{Aba\-ntu ba\-ja\-ga\-lai\-re olu\-zwa\-nyu\-ma lwa ki\-ra\-ngo.}
\fmtExampleEN{The crowd was in an uproar after the announcement.}

\fmtLemma{jai\-ga}
\fmtCitation{(oku)\-jai\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{jai\-gi\-re}
\fmtSubentry(a)
\fmtSense{to chew}
\fmtExampleRL{Ki\-sai oku\-jai\-ga eme\-re n’e\-go\-nda ng’o\-kya\-li ku\-gi\-mi\-ra.}
\fmtExampleEN{It's good to chew food so that it softens before you swallow it.}
\fmtSubentry(b)
\fmtSense{to talk over matters, to carry on a discussion with a view to reaching a decision}

\fmtLemma{ja\-ke\-ti}
\fmtCitation{(e)ja\-ke\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{jacket, coat}
\fmtExampleRL{Okwa ike\-ndi nzwa\-la ja\-ke\-ti.}
\fmtExampleEN{I wear a jacket during the rainy season.}

\fmtLemma{ja\-la\-ja}
\fmtCitation{(e)ja\-la\-ja}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{ferry station}
\fmtExampleRL{Eki\-dye\-ri kii\-ki\-re oku ja\-la\-ja.}
\fmtExampleEN{The ferry has reached the ferry station.}

\fmtLemma{ja\-lu\-ki}
\fmtCitation{\textnormal{(e.g.}\,mu\-ja\-lu\-ki, ba\-ja\-lu\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{joyous, cheerful, jolly}
\fmtExampleRL{Abbai\-re mwa\-na mu\-ja\-lu\-ki bu\-li nkya.}
\fmtExampleEN{She has always been a joyous child.}

\fmtLemma{ja\-mbi\-ya}
\fmtCitation{(e)ja\-mbi\-ya}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Sw.}
\fmtSense{panga, machete, long-bladed dagger}
\fmtExampleRL{Eja\-mbi\-ya za\-nge zi\-buu\-re.}
\fmtExampleEN{My machetes are missing.}

\fmtLemma{ja\-mbu\-la}
\fmtCitation{(o)ja\-mbu\-la}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{java plum}
\fmtExampleRL{Abaa\-na ba\-ta\-ka oku\-lya eja\-mbu\-la.}
\fmtExampleEN{Children like eating java plums.}

\fmtLemma{ja\-na}
\fmtCitation{(e)ja\-na}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{holy sanctuary (for Moslems ), heaven}
\fmtExampleRL{Aba\-ko\-li ba bi\-sai ba\-lya\-ba omu ja\-na.}
\fmtExampleEN{Those who do good deeds will go to heaven.}

\fmtLemma{ja\-na\-za}
\fmtCitation{(e)ja\-na\-za}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{stretcher used by Moslems to carry the dead}
\fmtExampleRL{Oru\-fa ba\-mu\-ta mu ja\-na\-za nga ba\-kwa\-ba oku\-mu\-lii\-ka.}
\fmtExampleEN{They put the dead man on a stretcher to take him for burial.}

\fmtLemma{ja\-nga}
\fmtCitation{(oku)\-ja\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ja\-ngi\-re}
\fmtSense{to make merry, to have fun}
\fmtExampleRL{ODau\-di n’o\-Na\-bwe\-ze ba\-ku\-ja\-nga.}
\fmtExampleEN{David and Nabweze are having fun.}

\fmtLemma{ja\-nja\-ba}
\fmtCitation{(oku)\-ja\-nja\-ba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ja\-nja\-bi\-re}
\fmtSense{to treat, to care (for the sick), to nurse, to attend}
\fmtExampleRL{Amai\-te oku\-ja\-nja\-ba aba\-lwai\-re.}
\fmtExampleEN{He knows how to care for the sick.}

\fmtLemma{ja\-ra\-jai\-ta}
\fmtCitation{(oku)\-ja\-ra\-jai\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ja\-ra\-jai\-ti\-re}
\fmtSense{(of a wound) to ooze (blood or pus)}
\fmtExampleRL{Ba\-mu\-sa\-mbi\-re omu ki\-rwai\-re ni ki\-ja\-ra\-jai\-ta omu\-saa\-yi.}
\fmtExampleEN{They kicked him in his wound and it oozed out blood.}

\fmtLemma{ja\-ru\-ka}
\fmtCitation{(oku)\-ja\-ru\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ja\-ru\-ki\-re}
\fmtSense{to be lively, to be cheerful}
\fmtExampleRL{Omwa\-na akya\-li ku\-ja\-ru\-ka, ya\-bba mu\-lwai\-re?}
\fmtExampleEN{The child is not lively, is she sick?}

\fmtLemma{ja\-ru\-ki}
\fmtCitation{\textnormal{(e.g.}\,mu\-ja\-ru\-ki, ba\-ja\-ru\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{confident, feeling confidence in oneself, self-assured}
\fmtExampleRL{Omwa\-na oni mu\-ja\-ru\-ki nku\-ka\-ka\-sa ya\-ka\-la yi\-nta\-vi\-yu.}
\fmtExampleEN{This child is confident, I am sure she will pass the interview.}

\fmtLemma{ja\-ruu\-ka}
\fmtCitation{(oku)\-ja\-ruu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ja\-ruu\-ki\-re}
\fmtSense{to be well-fermented}
\fmtExampleRL{Ama\-lwa ga\-kya\-li ku\-ja\-ruu\-ka.}
\fmtExampleEN{The millet beer has not yet fermented well.}

\fmtLemma{ja\-wu\-ki}
\fmtCitation{\textnormal{(e.g.}\,mu\-ja\-wu\-ki, ba\-ja\-wu\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{uncultured, ill-mannered, savage, uncivilised}
\fmtExampleRL{Oni omu\-sai\-za mu\-ja\-wu\-ki aku\-bba omu\-ka\-li waa\-mwe bu\-lii\-zo.}
\fmtExampleEN{That man is uncivilized: he beats his wife every day.}

\fmtLemma{je\-ba}
\fmtCitation{(oku)\-je\-ba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{je\-be\-re}
\fmtSubentry(a)
\fmtSense{to get drenched, to get soaked, to become soggy, to become mushy}
\fmtExampleRL{Oi\-ke\-ndi amu\-ku\-bbi\-re n’a\-je\-ba.}
\fmtExampleEN{He has been drenched by rain.}
\fmtSubentry(b)
\fmtSense{to weaken due to illness}
\fmtExampleRL{Omwa\-na aje\-be\-re lwa ki\-bi\-ri\-bi\-ri.}
\fmtExampleEN{The child is weak due to malaria.}
\fmtSubentry(c)
\fmtSense{to lack consistency}
\fmtExampleRL{Omu\-li\-mu gwa akoo\-re gu\-je\-be\-rei\-re.}
\fmtExampleEN{The job he has done lacks consistency.}

\fmtLemma{jee\-ja}
\fmtCitation{(oku)\-jee\-ja}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{jee\-je\-rye}
\fmtSubentry(a)
\fmtSense{to imitate, to mirror, to act like, to impersonate}
\fmtExampleRL{Toi\-ra\-nga\-mu ku\-jee\-ja ba\-ntu ba\-ku\-lu, n’o\-li\-kii\-ra\-mu oli\-ku\-bbi\-bwa.}
\fmtExampleEN{Do not imitate your elders ever again, or else you will be beaten.}
\fmtSubentry(b)
\fmtSense{to tease, to make fun of, to lough at, to mock, to jeer at}
\fmtExampleRL{Ba\-mu\-jee\-je\-re mu\-no obwi\-bbi na\-bu\-zwa\-ku.}
\fmtExampleEN{They teased him so much that he gave up stealing.}

\fmtLemma{jee\-jee\-ra}
\fmtCitation{(oku)\-jee\-jee\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{jee\-je\-rei\-re}
\fmtSense{to mock, to mimic scornfully, to ape}
\fmtExampleRL{Omwo\-jo ya\-bbai\-re ago\-ngei\-re n’a\-jee\-jee\-ra oi\-te waa\-mwe.}
\fmtExampleEN{The boy was upset and mocked his father scornfully.}

\fmtLemma{jee\-ke}
\fmtCitation{(e)jee\-ke}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{car jerk}
\fmtExampleRL{Ejee\-ke ya mo\-to\-ka eye\-nde\-ke\-re.}
\fmtExampleEN{The car jack has broken down.}

\fmtLemma{jee\-ma}
\fmtCitation{(oku)\-jee\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{jee\-me\-re}
\fmtSense{to disobey, to defy, to go against, to resist, to refuse}
\fmtExampleRL{Omwa\-na oku\-jee\-me\-ra aba\-kai\-re baa\-mwe ki\-bba ki\-bbi.}
\fmtExampleEN{It is very bad for a child to disobey his parents.}

\fmtLemma{jee\-mu}
\fmtCitation{\textnormal{(e.g.}\,mu\-jee\-mu, ba\-jee\-mu\textnormal{)}}
\fmtPOS{adj}
\fmtSense{disobedient, insubordinate}
\fmtExampleRL{Ti\-nta\-ka baa\-na ba\-jee\-mu.}
\fmtExampleEN{I hate disobedient children.}

\fmtLemma{jee\-mu\-ka}
\fmtCitation{(oku)\-jee\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{jee\-mu\-ki\-re}
\fmtSense{to surrender, to give up, to yield}
\fmtExampleRL{Omu\-ka\-li a\-bbai\-re ajee\-me\-re ajee\-mu\-ki\-re.}
\fmtExampleEN{The woman who had resisted has finally surrendered.}

\fmtLemma{jee\-mu\-lu\-ku\-ka}
\fmtCitation{(oku)\-jee\-mu\-lu\-ku\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{jee\-mu\-lu\-ku\-ki\-re}
\fmtSense{to surrender, to end a rebellion, to capitulate}
\fmtExampleRL{Aba\-yee\-ke\-ra ba\-jee\-mu\-lu\-ku\-ki\-re.}
\fmtExampleEN{The rebels have ended the rebellion.}

\fmtLemma{jee\-ru}
\fmtCitation{(e)jee\-ru}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{white colour}
\fmtExampleRL{Ejee\-ru ni\-ye la\-ngi gye nsi\-nga oku\-ta\-ka.}
\fmtExampleEN{White is my favourite colour.}

\fmtLemma{je\-ge\-ja}
\fmtCitation{(o)je\-ge\-ja}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{flat-headed borer}
\fmtExampleRL{Oje\-ge\-ja afuu\-ki\-re ki\-sii\-sa kya mu\-ta\-waa\-na edi emi\-saa\-le.}
\fmtExampleEN{Flat-headed borers have become a devastating pest for trees.}

\fmtLemma{je\-ge\-je\-ge}
\fmtCitation{(o)je\-ge\-je\-ge}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{coin}
\fmtExampleRL{Baa\-mwi\-rii\-rye ba\-la\-nsi wa je\-ge\-je\-ge.}
\fmtExampleEN{The balance (UgE)/change he was given was in coins.}

\fmtLemma{je\-ge\-nge}
\fmtCitation{(o)je\-ge\-nge}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSubentry(a)
\fmtSense{leader, highest in rank}
\fmtExampleRL{Oje\-ge\-nge wa iya\-nga li\-ni ni\-ye Mu\-se\-ve\-ni.}
\fmtExampleEN{The leader of this nation is Museveni.}
\fmtSubentry(b)
\fmtSense{title of Isabanyala}
\fmtExampleRL{Oje\-ge\-nge wa Bu\-nya\-la ni\-ye Rt. Ma\-j. Ba\-cke\-r Ki\-me\-ze Mpa\-gi.}
\fmtExampleEN{The king of the Banyala is called Rt. Maj. Backer Kimeze Mpagi.}

\fmtLemma{je\-je\-ka}
\fmtCitation{(oku)\-je\-je\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{je\-je\-ke\-re}
\fmtSense{to speak ill of, to put down, to badmouth}
\fmtExampleRL{OMu\-si\-si ba\-mu\-je\-je\-ke\-re\-ku ebi\-ga\-mbo ebi\-bbi.}
\fmtExampleEN{They spoke ill of Musisi.}

\fmtLemma{je\-jja}
\fmtCrossreference{\textnormal{see} jajja}

\fmtLemma{je\-ne\-ro}%
\fmtOne{1}
\fmtCitation{(e)je\-ne\-ro}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from English, archaic}
\fmtSubentry(a)
\fmtSense{ginnery}
\fmtExampleRL{Aya\-bi\-re oku je\-ne\-ro oku\-su\-nsu\-la pa\-mpa.}
\fmtExampleEN{He has gone to the ginnery to gin cotton.}
\fmtSubentry(b)
\fmtSense{petrol station, fuel station, gas station, filling station}
\fmtExampleRL{Ya\-ba oku je\-ne\-ro ofu\-ne ama\-fu\-ta.}
\fmtExampleEN{Go to the petrol station and get paraffin.}

\fmtLemma{je\-ne\-ro}%
\fmtOne{2}
\fmtCitation{(o)je\-ne\-ro}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from En.}
\fmtSense{five star general officer}
\fmtExampleRL{Oje\-ne\-ro ya\-kwa\-ta eki\-fo kya ka\-bi\-ri mu bu\-lu\-lu mu iye lya Ame\-ri\-ka.}
\fmtExampleEN{A five-star general is the second highest rank in the US Army.}

\fmtLemma{je\-nje\-be\-ka}
\fmtCitation{(oku)\-je\-nje\-be\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{je\-nje\-be\-ke\-re}
\fmtSense{to sit where one is not supposed to sit}
\fmtExampleRL{Omu\-sai\-za yee\-je\-nje\-be\-ke\-re omu ki\-fo kya mu\-ge\-ni omu\-ku\-lu.}
\fmtExampleEN{The man positioned himself in the guest of honour's seat.}

\fmtLemma{je\-nje\-ka}%
\fmtOne{1}
\fmtCitation{(oku)\-je\-nje\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{je\-nje\-ke\-re}
\fmtSubentry(a)
\fmtSense{to tilt into a suitable position}
\fmtExampleRL{Omwa\-na aku\-je\-nje\-ka eki\-do\-mo\-la ki\-ta\-gwa.}
\fmtExampleEN{The child is tilting the jerry can into a suitable position so that it does not fall.}
\fmtSubentry(b)
\fmtSense{to tie loosely, to place loosely}

\fmtLemma{je\-nje\-ka}%
\fmtOne{2}
\fmtCitation{(oku)\-je\-nje\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{je\-nje\-ke\-re}
\fmtSense{to trudge along (walk slowly and with heavy steps, typically because of exhaustion or harsh conditions), to trot}
\fmtExampleRL{Ba\-je\-nje\-ke\-re okwi\-ra eka nga ti\-ba\-ba\-zi\-re\-wo ki\-ga\-mbo kyo\-na.}
\fmtExampleEN{They trudged home without a word being spoken.}

\fmtLemma{je\-re\-ba}
\fmtCitation{(oku)\-je\-re\-ba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{je\-re\-be\-re}
\fmtSense{to be slippery, to be soft}
\fmtExampleRL{Olu\-bu\-ga lu\-je\-re\-be\-re olwa oi\-ke\-ndi atoo\-nye\-re.}
\fmtExampleEN{The courtyard has become slippery due to the rain.}

\fmtLemma{je\-re\-be\-ra}
\fmtCitation{(oku)\-je\-re\-be\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{je\-re\-bei\-re}
\fmtSense{to soften}
\fmtExampleRL{Epaa\-paa\-li li\-ni li\-je\-re\-bei\-re.}
\fmtExampleEN{Why did the papaya become soft?}

\fmtLemma{je\-re\-ga}
\fmtCitation{(oku)\-je\-re\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{je\-re\-ge\-re}
\fmtSense{to make fun of, to mock, to tease, to deride, to belittle}
\fmtExampleRL{Abaa\-na ba\-ku\-je\-re\-ga omu\-ga\-nda waa\-bwe.}
\fmtExampleEN{The children are making fun of their sibling.}

\fmtLemma{ji\-gai\-ja}
\fmtCitation{(oku)\-ji\-gai\-ja}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ji\-gai\-ji\-rye}
\fmtSense{to give birth to a first born}
\fmtExampleRL{Nji\-gai\-ji\-rye mwo\-jo.}
\fmtExampleEN{I have given birth to my first-born baby boy.}

\fmtLemma{ji\-gi\-ja}
\fmtCitation{(o)ji\-gi\-ja}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{cow}
\fmtExampleRL{Aguu\-re oji\-gi\-ja wa nte.}
\fmtExampleEN{He bought a cow.}

\fmtLemma{jii\-ni}%
\fmtOne{1}
\fmtCitation{(e)jii\-ni}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSource{from En.}
\fmtSense{evil spirit, demon}
\fmtExampleRL{Ama\-jii\-ni ga\-ga\-dya aba\-ntu.}
\fmtExampleEN{Evil spirits disturb people.}

\fmtLemma{jii\-ni}%
\fmtOne{2}
\fmtCitation{(e)jii\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{denim (type of fabric)}
\fmtExampleRL{Empa\-le yaa\-mwe ya jii\-ni.}
\fmtExampleEN{His trousers are made of denim.}

\fmtLemma{ji\-mba}
\fmtCitation{(oku)\-ji\-mba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ji\-mbi\-re}
\fmtSense{to arrest, to detain}
\fmtExampleRL{Omu\-sai\-za ba\-mu\-ji\-mbi\-re ni ba\-mu\-twa\-la oku po\-lii\-si.}
\fmtExampleEN{The man was arrested and taken to the police station.}

\fmtLemma{ji\-nga\-ji\-nga}
\fmtCitation{(oku)\-ji\-nga\-ji\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ji\-ngi\-re\-ji\-ngi\-re}
\fmtSense{to forge, to fake, to tamper}
\fmtExampleRL{Aba\-si\-ri\-ka\-le ba\-ji\-ngi\-re\-ji\-ngi\-re omu\-ko\-no gwa mi\-ni\-si\-ta.}
\fmtExampleEN{The soldiers forged the minister's signature.}

\fmtLemma{ji\-ngii\-ra}
\fmtCitation{(oku)\-ji\-ngii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ji\-ngi\-rii\-re}
\fmtSense{to forge, to fake, to tamper}
\fmtExampleRL{Ebi\-ro bi\-ni aba\-fe\-re ba\-ji\-ngii\-ra emi\-ko\-no gya ba\-ntu.}
\fmtExampleEN{These days, con men forge people's signatures.}

\fmtLemma{ji\-ri\-koo\-ti}
\fmtCitation{(e)ji\-ri\-koo\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{rain coat}
\fmtExampleRL{Okwa nju\-ra nzwa\-la ji\-ri\-koo\-ti.}
\fmtExampleEN{I wear a raincoat during the rainy season.}

\fmtLemma{jja\-jja}
\fmtCitation{(o)jja\-jja}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from Lug.} 
\fmtSense{grandparent}

\fmtLemma{jo\-bo\-ja}
\fmtCitation{(oku)\-jo\-bo\-ja}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{jo\-bo\-je\-rye}
\fmtSense{to mumble, to speak unclearly, to speak unintelligibly}
\fmtExampleRL{Omwa\-na oyo ajo\-bo\-ja\-mbe ta\-ru\-nguu\-ra bi\-ga\-mbo.}
\fmtExampleEN{That child just mumbles instead of speaking clearly.}

\fmtLemma{jo\-ko}
\fmtCitation{(e)jo\-ko}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{yoke}
\fmtExampleRL{Onte ba\-mu\-bbwe\-re\-mu ejo\-ko.}
\fmtExampleEN{They put a yoke around the cow's neck.}

\fmtLemma{jo\-koo\-ni}
\fmtCitation{(e)jo\-koo\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{kitchen, room for cooking}
\fmtExampleRL{Nje nsu\-mbi\-ra mu jo\-koo\-ni.}
\fmtExampleEN{I cook in the kitchen.}

\fmtLemma{jo\-lo\-nga}
\fmtCrossreference{\textnormal{see} joronga}

\fmtLemma{jo\-njo\-ba}%
\fmtOne{1}
\fmtCitation{(oku)\-jo\-njo\-ba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{jo\-njo\-bai\-re}
\fmtSense{(of weather) to turn grey, to get worse}
\fmtExampleRL{Obwi\-re ka bu\-jo\-njo\-bai\-re mu\-no!}
\fmtExampleEN{The weather has turned very grey!}

\fmtLemma{jo\-njo\-ba}%
\fmtOne{2}
\fmtCitation{(o)jo\-njo\-ba}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{frail person, weak person, fragile person, feeble person}
\fmtExampleRL{Omu\-sai\-za oyo jo\-njo\-ba.}
\fmtExampleEN{That man is frail.}

\fmtLemma{jo\-njo\-ba\-la}
\fmtCitation{(oku)\-jo\-njo\-ba\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{jo\-njo\-bai\-re}
\fmtSubentry(a)
\fmtSense{to be soaked in water, to be wet}
\fmtExampleRL{Onko\-ko agwi\-re omu mai\-zi n’a\-jo\-njo\-ba\-la.}
\fmtExampleEN{The chicken fell in the water and got wet.}
\fmtSubentry(b)
\fmtSense{(of weather) to be humid}
\fmtExampleRL{Obwi\-re bu\-jo\-njo\-bai\-re.}
\fmtExampleEN{It is humid. (lit.\,The weather is humid.).}

\fmtLemma{joo\-ga}
\fmtCitation{(oku)\-joo\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{joo\-ge\-re}
\fmtSense{to insult, to abuse, to offend, to slander, to show contempt for, to defy}
\fmtExampleRL{Omu\-ka\-li odi amaa\-re eki\-see\-ra ki\-na\-mu\-lai nga aku\-joo\-ga oi\-ba waa\-mwe.}
\fmtExampleEN{That woman has spent a long time insulting her husband.}

\fmtLemma{joo\-ge\-re\-za}
\fmtCitation{(oku)\-joo\-ge\-re\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{joo\-ge\-rei\-ze}
\fmtSense{to treat with contempt, to treat without respect}
\fmtExampleRL{Ojoo\-ge\-re\-za mu\-no abai\-ra baa\-mu.}
\fmtExampleEN{You treat your colleagues with a lot of contempt.}

\fmtLemma{joo\-ja}
\fmtCitation{(oku)\-joo\-ja}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{joo\-je\-rye}
\fmtSense{to disturb, to annoy, to bother, to upset}
\fmtExampleRL{Naa\-ni oyo aku\-joo\-ja abai\-raa\-mwe?}
\fmtExampleEN{Who is that disturbing his friends?}

\fmtLemma{jo\-ro\-nga}
\fmtCitation{(oku)\-jo\-ro\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{jo\-ro\-nge\-re}
\fmtSource{from Lug.} 
\fmtSense{to disrespect, to show a lack of respect, to be contemptuous of, to treat contemptuously, to despise}
\fmtExampleRL{Omu\-ka\-li waa\-mu omu\-jo\-ro\-nge\-re mu\-no.}
\fmtExampleEN{You treated your wife with a lot of contempt.}

\fmtLemma{ju\-bi\-ri\-sya}
\fmtCitation{(oku)\-ju\-bi\-ri\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-bi\-ri\-si\-rye}
\fmtSense{to exaggerate, to overstate}
\fmtExampleRL{Ali\-mu oka\-te\-mba, ata\-ka mu\-no oku\-ju\-bi\-ri\-sya emboo\-zi yaa\-mwe.}
\fmtExampleEN{She is dramatic, she likes to exaggerate her stories.}

\fmtLemma{ju\-ga}
\fmtCitation{(oku)\-ju\-ga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ju\-gi\-re}
\fmtSense{(of bull) to moo when sexually aroused}
\fmtExampleRL{Onte aku\-ju\-ga.}
\fmtExampleEN{The bull is mooing.}

\fmtLemma{ju\-gu\-mi\-ra}
\fmtCitation{(oku)\-ju\-gu\-mi\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ju\-gu\-mii\-re}
\fmtSense{to shiver, to shudder, to tremble}
\fmtExampleRL{Enju\-ra emu\-ku\-bbi\-re era ati\-bu\-ni oku\-ju\-gu\-mi\-ra.}
\fmtExampleEN{He was drenched by the rain and is now shivering.}

\fmtLemma{ju\-gu\-ta}
\fmtCitation{(oku)\-ju\-gu\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-gu\-ti\-re}
\fmtSubentry(a)
\fmtSense{to beat thoroughly, to give a good thrashing, to beat violently}
\fmtExampleRL{Nkwi\-za oku\-ku\-ju\-gu\-ta emii\-go.}
\fmtExampleEN{I will beat you thoroughly.}
\fmtSubentry(b)
\fmtSense{to fan (the flames), to blow (on a fire to keep it going)}
\fmtExampleRL{Ju\-gu\-ta mu\-no omuu\-ro omu kyo\-to gu\-ka\-nye.}
\fmtExampleEN{Fan the flames so that we get the fire burning.}

\fmtLemma{ju\-gu\-tii\-ra}
\fmtCitation{(oku)\-ju\-gu\-tii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-gu\-ti\-rii\-rye}
\fmtSense{to blackmail}
\fmtExampleRL{Ba\-mu\-ju\-gu\-ti\-rii\-rye\-mbe.}
\fmtExampleEN{He was just blackmailed.}

\fmtLemma{ju\-ju\-nja}
\fmtCitation{(oku)\-ju\-ju\-nja}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-ju\-nji\-rye}
\fmtSense{to mistreat, to abuse, to ill treat, to molest}
\fmtExampleRL{Abai\-bbi ba\-ju\-ju\-nji\-re mu\-no aba\-ntu.}
\fmtExampleEN{The thieves have mistreated people a lot.}

\fmtLemma{ju\-lai\-bba}
\fmtCitation{(o)ju\-lai\-bba}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{tall and big person}
\fmtExampleRL{Oju\-lai\-bba wa mu\-ka\-li aku\-bbi\-re oi\-baa\-mwe.}
\fmtExampleEN{A big and tall woman has beaten her husband.}

\fmtLemma{ju\-lii\-ra}
\fmtCitation{(oku)\-ju\-lii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ju\-li\-rii\-re}
\fmtSense{to appeal, to petition, to make request, to plead}
\fmtExampleRL{OMu\-so\-ke ta\-ma\-tii\-re n’a\-nsa\-la ya mu\-sa\-ngo aju\-lii\-re.}
\fmtExampleEN{Musoke has appealed to a higher court because he is not satisfied with the way the judgement was passed.}

\fmtLemma{ju\-lu\-ngu\-ta\-na}
\fmtCitation{(oku)\-ju\-lu\-ngu\-ta\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ju\-lu\-ngu\-tai\-ne}
\fmtSense{to be mixed up, to be in a mess, to be upside down}
\fmtExampleRL{Ebi\-ntu byo\-na bi\-ju\-lu\-ngu\-tai\-ne.}
\fmtExampleEN{Everything is upside down.}

\fmtLemma{ju\-lu\-ngu\-ta\-ni}
\fmtCitation{\textnormal{(e.g.}\,gi\-ju\-lu\-ngu\-ta\-ni, gu\-ju\-lu\-nga\-ta\-ni\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{dilute, thinned, watered down}
\fmtExampleRL{Ensa\-nde ampai\-re gi\-ju\-lu\-ngu\-ta\-ni.}
\fmtExampleEN{He gave me diluted juice.}
\fmtSubentry(b)
\fmtSense{lacking in substance}

\fmtLemma{ju\-lu\-ngu\-ta\-nya}
\fmtCitation{(oku)\-ju\-lu\-ngu\-ta\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-lu\-ngu\-tai\-nye}
\fmtSense{to dilute, to make thinner}
\fmtExampleRL{Ni wa\-ju\-lu\-ngu\-ta\-nya ensa\-nde ndoo\-wo ya\-gi\-nywa.}
\fmtExampleEN{No one will drink that juice if you dilute it.}

\fmtLemma{Ju\-ma}
\fmtCitation{Ju\-ma}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSubentry(a)
\fmtSense{Friday}
\fmtExampleRL{Aba\-si\-raa\-mu ti\-ba\-ko\-la ku Ju\-ma.}
\fmtExampleEN{Moslems do not work on Friday.}
\fmtSubentry(b)
\fmtSense{Friday prayers}
\fmtExampleRL{Aba\-si\-raa\-mu ba\-ya\-bi\-re ku mu\-zi\-ki\-ti ku\-saa\-la ju\-ma.}
\fmtExampleEN{The Muslims have gone to the mosque for Friday prayers.}

\fmtLemma{ju\-mba}
\fmtCitation{(oku)\-ju\-mba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ju\-mbi\-re}
\fmtSense{to bloom, to blossom}
\fmtExampleRL{Emi\-saa\-le gi\-ju\-mbi\-re.}
\fmtExampleEN{The trees have blossomed.}

\fmtLemma{ju\-mbi\-ra}
\fmtCitation{(oku)\-ju\-mbi\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-mbii\-re}
\fmtSense{to be enthusiastic about, to be eager about, to be zealous in, to take a great interest in, to be diligent in}
\fmtExampleRL{Oyo aju\-mbi\-ra mu\-no emi\-so\-mo gya\-mwe.}
\fmtExampleEN{That one is enthusiastic about his studies.}

\fmtLemma{ju\-mbu\-ka}
\fmtCitation{(oku)\-ju\-mbu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-mbu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to put under cover, to hide}
\fmtSubentry(b)
\fmtSense{to put under warm ashes (\eg food for later use)}
\fmtExampleRL{Aju\-mbu\-ki\-re omu\-wo\-go omu bu\-nyo\-ta.}
\fmtExampleEN{He has put cassava under warm ashes.}

\fmtLemma{ju\-mbu\-la}
\fmtCitation{(oku)\-ju\-mbu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-mbwi\-re}
\fmtSubentry(a)
\fmtSense{to unhide, to uncover}
\fmtSubentry(b)
\fmtSense{to remove out of hot ash}
\fmtExampleRL{Nkwa\-ba ku\-ju\-mbu\-la mu\-wo\-go wa\-nge mu ki\-koo\-mi.}
\fmtExampleEN{I am going to remove my cassava from the bonfire.}

\fmtLemma{ju\-mbu\-tu\-ka}
\fmtCitation{(oku)\-ju\-mbu\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ju\-mbu\-tu\-ki\-re}
\fmtSense{to rush out, to dash out, to burst out}
\fmtExampleRL{Ompi\-ri ya\-ju\-mbu\-tu\-ki\-re omu nyu\-mba.}
\fmtExampleEN{The snake rushed out of the house.}

\fmtLemma{ju\-mbu\-tu\-la}
\fmtCitation{(oku)\-ju\-mbu\-tu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-mbu\-twi\-re}
\fmtSense{to cause to rush out, to drive out, to chase away}
\fmtExampleRL{Epo\-lii\-si eju\-mbu\-twi\-re abai\-bbi omu loo\-gi.}
\fmtExampleEN{The police chased thieves from the lodge.}

\fmtLemma{ju\-mu}
\fmtCitation{\textnormal{(e.g.}\,li\-ju\-mu, bi\-ju\-mu\textnormal{)}}
\fmtPOS{adj}
\fmtSense{fertile}
\fmtExampleRL{Ei\-ta\-ka\-li li\-ni li\-ju\-mu li\-ba\-lya ebi\-coo\-li ebi\-sai.}
\fmtExampleEN{This land is fertile, it yields better maize.}

\fmtLemma{ju\-mu\-ka}
\fmtCitation{(oku)\-ju\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ju\-mu\-ki\-re}
\fmtSense{(of soil) to become fertile}
\fmtExampleRL{Ei\-ta\-ka\-li li\-ju\-mu\-ki\-re olwa bi\-ju\-mu\-sa bye tu\-bbai\-re tu\-su\-ka\-wo.}
\fmtExampleEN{The soil has become fertile due to the fertilisers we have been adding.}

\fmtLemma{ju\-mu\-kya}
\fmtCitation{(oku)\-ju\-mu\-kya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-mu\-ki\-rye}
\fmtSense{to fertilise}
\fmtExampleRL{Ju\-mu\-kya ei\-ta\-ka\-li lya\-mu n’e\-bi\-syo\-do bya nko\-ko.}
\fmtExampleEN{Fertilise your soil using chicken droppings.}

\fmtLemma{ju\-na}
\fmtCitation{(oku)\-ju\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-ni\-re}
\fmtSense{to help, to assist, to aid, to support}
\fmtExampleRL{Ju\-na omwa\-na oyo aye\-mee\-re mu\-lwai\-re.}
\fmtExampleEN{Help that child to stand, she is sick.}

\fmtLemma{ju\-naa\-na}
\fmtCitation{(oku)\-ju\-naa\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-nai\-ne}
\fmtSense{to hold responsible, to accuse of, to blame, to charge with}
\fmtExampleRL{Mu\-ku\-mu\-ju\-naa\-na ki?}
\fmtExampleEN{What are you accusing her of?}

\fmtLemma{ju\-nda}
\fmtCitation{(oku)\-ju\-nda}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ju\-ndi\-re}
\fmtSense{to rot, to decay, to spoil, to go bad, to decompose}
\fmtExampleRL{Ova\-ke\-do wa\-nge aju\-ndi\-re.}
\fmtExampleEN{My avocado has gone bad.}

\fmtLemma{ju\-ndi}
\fmtCitation{\textnormal{(e.g.}\,mu\-ju\-ndi, bi\-ju\-ndi, mu\-ju\-ndi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{rotten, decayed, spoiled, rancid}
\fmtExampleRL{Epaa\-paa\-li li\-ni li\-ju\-ndi.}
\fmtExampleEN{This pawpaw is rotten.}

\fmtLemma{ju\-ngu}
\fmtCitation{\textnormal{(e.g.}\,ki\-ju\-ngu, bi\-ju\-nju\textnormal{)}}
\fmtPOS{adj}
\fmtSense{artificial, hybrid}
\fmtExampleRL{Ekyo eki\-duu\-ma eki\-ju\-ngu.}
\fmtExampleEN{That maize is a hybrid.}

\fmtLemma{ju\-ngu\-lu\-ku\-ka}
\fmtCitation{(oku)\-ju\-ngu\-lu\-ku\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ju\-ngu\-lu\-ku\-ki\-re}
\fmtSense{to be diluted, to be thinned}
\fmtExampleRL{Omwe\-nge gu\-ju\-ngu\-lu\-ku\-ki\-re ndoo\-wo ya\-gu\-gu\-la.}
\fmtExampleEN{The millet beer has become diluted, no one will buy it.}

\fmtLemma{ju\-ngu\-lu\-la}
\fmtCitation{(oku)\-ju\-ngu\-lu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-ngu\-lwi\-re}
\fmtSense{to dilute, to thin}
\fmtExampleRL{Wee\-na oju\-ngu\-lwi\-re mu\-no ensa\-nde.}
\fmtExampleEN{You have diluted the banana juice too much.}

\fmtLemma{ju\-nguu\-ra}
\fmtCitation{(oku)\-ju\-nguu\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-ngui\-re}
\fmtSense{to dilute excessively}
\fmtExampleRL{Ota\-ju\-nguu\-ra ma\-tai.}
\fmtExampleEN{Do not dilute the milk excessively.}

\fmtLemma{ju\-ru\-juu\-ta}
\fmtCitation{(oku)\-ju\-ru\-juu\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-ru\-juu\-ti\-re}
\fmtSense{to put out, to turn off, to put out (\eg a light), to extinguish something burning}
\fmtExampleRL{Ju\-ru\-juu\-ta omuu\-ro.}
\fmtExampleEN{Turn off the fire.}

\fmtLemma{ju\-ru\-mu\-ka}
\fmtCitation{(oku)\-ju\-ru\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ju\-ru\-mu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to tingle (of a body part), to prickle}
\fmtExampleRL{Na\-lii\-re omu\-ye\-mbe ogu\-bbi\-si ama\-ta\-ma ga\-ku\-nju\-ru\-mu\-ka.}
\fmtExampleEN{I ate an unripe mango, and my cheeks are tingling.}
\fmtSubentry(b)
\fmtSense{to tremble all over, to shiver, to have a creepy feeling (in a body part), to be frightened, to be startled, to be shocked, to experience revulsion}
\fmtExampleRL{Omu\-bi\-ri gu\-nju\-ru\-mu\-ki\-re nga mboi\-ne onjo\-ka mu nyu\-mba.}
\fmtExampleEN{I trembled all over when I saw a snake in the house.}

\fmtLemma{ju\-ru\-nga}
\fmtCitation{(oku)\-ju\-ru\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ju\-ru\-ngi\-re}
\fmtSense{to humiliate, to cause to lose face}
\fmtExampleRL{Le\-ke\-ra\-wo oku\-ju\-ru\-nga abai\-raa\-mu.}
\fmtExampleEN{Stop humiliating your friends.}

\fmtLemma{ju\-ta}
\fmtCitation{(oku)\-ju\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ju\-ti\-re}
\fmtSense{to grow fat, to get obese}
\fmtExampleRL{Onke\-re\-mba aju\-ti\-re.}
\fmtExampleEN{The baby has grown fat.}

\fmtLemma{juu\-le}
\fmtCitation{(o)juu\-le}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from Lug.} 
\fmtSense{diluted brew}
\fmtExampleRL{Omwe\-nge gwa juu\-le ti\-gu\-bba gu\-sai.}
\fmtExampleEN{Diluted beer is not good.}

\fmtLemma{juu\-mu\-ka}
\fmtCitation{(oku)\-juu\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{juu\-mu\-ki\-re}
\fmtSense{to overturn, to turn upside down, to turn over, to turn face down}
\fmtExampleRL{Juu\-mu\-ka ensu\-wa eyo eka\-le.}
\fmtExampleEN{Turn the pot upside down so that it dries.}

\fmtLemma{juu\-mu\-la}
\fmtCitation{(oku)\-juu\-mu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{juu\-mwi\-re}
\fmtSense{to turn upright, to turn right side up, to orient}
\fmtExampleRL{Juu\-mu\-la engo\-ma.}
\fmtExampleEN{Turn the drum upright.}

\fmtLemma{juu\-ra}
\fmtCitation{(oku)\-juu\-ra}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{ju\-lii\-re}
\fmtSubentry(a)
\fmtSense{to testify, to give evidence, to bear witness}
\fmtExampleRL{Nyi\-zi\-re ku\-juu\-ra mu\-sai\-za oyo ku\-bba ta\-li\-na mu\-sa\-ngo.}
\fmtExampleEN{I have come to testify for this man because he is innocent.}
\fmtSubentry(b)
\fmtSense{to appeal}
\fmtExampleRL{Lwa\-ki to\-juu\-ra omu koo\-ti egi\-ku\-lu?}
\fmtExampleEN{Why don’t you appeal to the high court?}

\fmtLemma{juu\-ru}%
\fmtOne{1}
\fmtCitation{(o)juu\-ru}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{concentrated honey}
\fmtExampleRL{Nta\-ka mu\-no oku\-nuu\-na ojuu\-ru wa bwo\-ki.}
\fmtExampleEN{I like licking concentrated honey.}

\fmtLemma{juu\-ta}%
\fmtOne{2}
\fmtCitation{(oku)\-juu\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{juu\-ti\-re}
\fmtSense{to throb, to pulsate rapidly or painfully, to cause pain in a series of regular beats}
\fmtExampleRL{Eki\-zi\-mba ki\-ku\-njuu\-ta bu\-ndi ti\-nku\-so\-bo\-la n'a\-ku\-bba\-sya.}
\fmtExampleEN{The boil is throbbing with so much pain that I cannot sleep.}

\fmtLemma{jwa}
\fmtCitation{(oku)\-jwa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{jwi\-re}
\fmtSubentry(a)
\fmtSense{to come out, to ooze out}
\fmtExampleRL{Amai\-zi ga\-ku\-jwa omu ita\-ka\-li.}
\fmtExampleEN{Water is coming out of the ground.}
\fmtSubentry(b)
\fmtFixedexpression{jwa (o)mu\-saa\-yi}
\fmtSense{to bleed}
\fmtExampleRL{Ota\-ka\-ku\-la ki\-rwai\-re ki\-kwi\-za oku\-jwa omu\-saa\-yi.}
\fmtExampleEN{You shouldn't scratch your wound, it will start bleeding.}
\fmtSubentry(c)
\fmtFixedexpression{jwa (o)mu\-ju\-le}
\fmtSense{to have a nosebleed}
\fmtExampleRL{Omu\-kai\-re aku\-jwa omu\-ju\-le.}
\fmtExampleEN{The woman has a nosebleed.}

\fmtLemma{jwa\-wu\-ka}
\fmtCitation{(oku)\-jwa\-wu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{jwa\-wu\-ki\-re}
\fmtSense{to become a devotee}
\fmtExampleRL{Aba\-ga\-nda bai\-ngi ba\-jwa\-wu\-ki\-re omu 1970s.}
\fmtExampleEN{Many Baganda became devotees in 1970s.}

\fmtLemma{jwa\-wu\-ki}
\fmtCitation{\textnormal{(e.g.}\,mu\-ja\-wu\-ki, ba\-ja\-wu\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{devout, pious, religious}
\fmtExampleRL{OMo\-se\-s mu\-jwa\-wu\-ki.}
\fmtExampleEN{Moses is devout.}

\fmtLemma{jwa\-ya}
\fmtCitation{(o)jwa\-ya}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{gazelle}
\fmtExampleRL{Ojwa\-ya awoo\-ma mu\-no omu\-ku\-bi.}
\fmtExampleEN{Gazelle makes very good soup.}

\fmtLemma{jwe\-ka}
\fmtCitation{(oku)\-jwe\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{jwe\-ke\-re}
\fmtSubentry(a)
\fmtSense{to adorn (a heir or a traditional leader in traditional garments)}
\fmtExampleRL{Omu\-gwe\-twa ba\-mu\-jwe\-ke\-re eki\-fu\-ndu\-kwa eky’o\-bu\-gwe\-te.}
\fmtExampleEN{The heir has been adorned in his official garment of inheritance.}
\fmtSubentry(b)
\fmtSense{to award (a degree), to crown}
\fmtExampleRL{Omwa\-na wa\-nge ba\-ku\-mu\-jwe\-ke\-re lee\-ro edi\-gu\-li.}
\fmtExampleEN{My daughter was awarded a degree today.}

\fmtLemma{jwe\-nge}
\fmtCitation{(o)jwe\-nge}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{canoe}
\fmtExampleRL{Abai\-bbi ba\-ta\-mbu\-lii\-re mu jwe\-nge ne ba\-sa\-la okwi\-ra oku ki\-sa\-mba eki\-ndi.}
\fmtExampleEN{The thieves used a canoe and crossed to another island.}

\fmtLemma{jwe\-te\-ka}
\fmtCitation{(oku)\-jwe\-te\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{jwe\-te\-ke\-re}
\fmtSense{to make up (stories or lies, often with a purpose to slander or to blackmail)}
\fmtExampleRL{Aku\-jwe\-te\-ka\-mbe ebi\-ga\-mbo oku nyi\-na waa\-mwe.}
\fmtExampleEN{He is just blackmailing his mother.}

\fmtLemma{jwi}
\fmtCitation{\textnormal{(e.g.}\,mu\-jwi, ki\-jwi\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{giant}
\fmtExampleRL{Aba\-sai\-za aba\-jwi ba\-bba n’a\-maa\-ni.}
\fmtExampleEN{Giant men are strong.}
\fmtSubentry(b)
\fmtFixedexpression{ku\-bba (a)ma\-jwi}
\fmtSense{to beg (lit.\,to beat the knees against the ground)}
\fmtExampleRL{Nkwa\-ba ku\-ku\-bbi\-ra koi\-za wa\-nge ma\-jwi ansaa\-si\-re.}
\fmtExampleEN{I am going to beg my uncle for forgiveness.}

\fmtLemma{jwi\-ga}
\fmtCitation{(oku)\-jwi\-ga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{jwi\-gi\-re}
\fmtSubentry(a)
\fmtSense{to expand, to develop, to widen, to grow, to advance}
\fmtExampleRL{Bbi\-zi\-nee\-si yaa\-mu ejwi\-gi\-re omu bi\-ro eby'a\-kai\-re.}
\fmtExampleEN{Your business has steadily expanded over the years.}
\fmtSubentry(b)
\fmtSense{to prostrate, to stoop, to bend down, to bend over, to bow in respect}
\fmtExampleRL{Jwi\-ga Omu\-ka\-ma wai\-swe.}
\fmtExampleEN{Prostrate before our king please.}
\fmtSubentry(c)
\fmtSense{to be of good quality, to look smart, to be neat}
\fmtExampleRL{Tu\-kwe\-ndya ebya\-nye\-ge\-sya bi\-jwi\-ge omu Bu\-duu\-li.}
\fmtExampleEN{We would like the education system to be of good quality in the Buruuli region.}

\fmtLemma{jwi\-gi\-sya}%
\fmtCitation{(oku)\-jwi\-gi\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{jwi\-gi\-si\-rye}
\fmtSubentry(a)
\fmtSense{to extend, to expand, to spread out, to open}
\fmtExampleRL{Tu\-ku\-ta\-ka ku\-jwi\-gi\-sya bya\-nye\-ge\-sya edi aboo\-ro boo\-na\-boo\-na.}
\fmtExampleEN{We would like to extend free education to all the poor.}
\fmtSubentry(b)
\fmtSense{to develop, to improve, to reinforce, to widen}
\fmtExampleRL{Tu\-jwi\-gi\-sye Obu\-ka\-ma bwa Bu\-nya\-la.}
\fmtExampleEN{Let us develop the Bunyala Kingdom.}

\fmtLemma{jwi\-ma}
\fmtCitation{(oku)\-jwi\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{jwi\-mi\-re}
\fmtSense{to shriek, to scream, to moan (in pain), to groan}
\fmtExampleRL{Ombe\-re\-ge ajwi\-ma nga ba\-kwa\-ba ku\-mu\-sa\-la.}
\fmtExampleEN{The pig screams when it is about to be slaughtered.}
\end{letter}

