\begin{letter}{K}
\fmtLemma{ka}%
\fmtOne{1}
\fmtPOS{part}
\fmtSubentry(a)
\fmtSense{let, allow (precedes verbs to express hortative and optative meanings)}
\fmtExampleRL{Ka tu\-bo\-ne eki\-kwi\-ra\-ku.}
\fmtExampleEN{Let's see what happens.}
\fmtSubentry(b)
\fmtFixedexpression{Ka nya\-gwe Ba\-ga\-nda.}
\fmtExampleEN{May you be taken by the Baganda.}
\fmtSense{(a curse, as traditionally the Baganda used to take the Banyala into forced labour, thus, whenever someone annoys you, you would curse this person to be taken by the Baganda into slavery.)}

\fmtLemma{ka}%
\fmtOne{2}
\fmtCitation{(e)ka}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{home, family}
\fmtExampleRL{Akwi\-ra eka eki\-ro kya lee\-ro.}
\fmtExampleEN{He is coming home tonight.}

\fmtLemma{ka}%
\fmtOne{3}
\fmtCitation{\textnormal{(e.g.}\,mu\-ka, gu\-ka, mi\-ka\textnormal{)}}
\fmtPOS{adj}
\fmtSense{concentrated}
\fmtExampleRL{Omwe\-nge gu\-ni gu\-ka.}
\fmtExampleEN{This beer is concentrated.}

\fmtLemma{ka}%
\fmtOne{4}
\fmtCrossreference{\textnormal{see} nka}

\fmtLemma{ka}%
\fmtOne{5}
\fmtPOS{interj}
\fmtSense{oh (used to express amusement)}
\fmtExampleRL{Ka mu\-ca\-mu\-ki\-re bu\-ndi eree\-ro?}
\fmtExampleEN{Oh, why are you so excited today?}

\fmtLemma{ka-}%
\fmtOne{6}
\fmtPOS{pf}
\fmtSense{subject and object agreement prefix of class 12 on verbs}

\fmtLemma{kaa\-bba}
\fmtCitation{(e)kaa\-bba}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{Kaabba (a building at the centre of Islam's most important mosque Al-Masjid Al-Ḥarām in Mecca)}
\fmtExampleRL{Aba\-si\-raa\-mu bee\-ru\-guu\-rya ekaa\-bba.}
\fmtExampleEN{Moslems go around the Kaabba.}

\fmtLemma{kaa\-bu\-ga}
\fmtCitation{(oku)\-kaa\-bu\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kaa\-bu\-gi\-re}
\fmtSubentry(a)
\fmtSense{to slash unwanted natural vegetation slightly (as a way of clearing land for cultivation), to cut natural vegetation slightly}
\fmtExampleRL{Aya\-bi\-re ku\-kaa\-bu\-ga wa ku\-lu\-ma pa\-mba.}
\fmtExampleEN{He has gone to slash the garden where he will plant cotton.}
\fmtSubentry(b)
\fmtSense{to close with green reeds (the initial phases in the building process), to lay the foundations for, to begin}
\fmtSubentry(c)
\fmtSense{to make a draft, to draft}

\fmtLemma{kaa\-bu\-yi}
\fmtCitation{(o)kaa\-bu\-yi}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{meat from the breast of a goat or ox}
\fmtExampleRL{Oka\-tei\-ca alya ka\-buu\-yi.}
\fmtExampleEN{A poor person eats meat from the breast of a goat or ox.}

\fmtLemma{kaa\-bwe}
\fmtCrossreference{\textnormal{see} (-)abwe}

\fmtLemma{kaa\-da}
\fmtCitation{(e)kaa\-da}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSubentry(a)
\fmtSense{card, ID, credit card, bank card, certificate}
\fmtExampleRL{Ba\-mu\-wai\-re ekaa\-da egi\-myu\-ki lwa ku\-ku\-bba ndo\-bo.}
\fmtExampleEN{He was given a red card for foul play.}
\fmtSubentry(b)
\fmtFixedexpression{kaa\-da ya bu\-bya\-le}
\fmtSense{birth certificate}
\fmtExampleRL{Ki\-ku\-ka\-ta\-ku oku\-la\-nga ekaa\-da yaa\-mu eya bu\-bya\-le oku\-so\-bo\-la oku\-fu\-na enda\-nga\-mu\-ntu.}
\fmtExampleEN{It is a requirement to show your birth certificate before you get a national ID.}
\fmtSubentry(c)
\fmtFixedexpression{kaa\-da ya ku\-ko\-mya}
\fmtSense{voter's card}
\fmtExampleRL{Ekaa\-da ya\-nge eya ku\-ko\-mya nji\-re\-ke\-re eka.}
\fmtExampleEN{I've left my voter's card at home.}

\fmtLemma{kaa\-di}
\fmtCrossreference{\textnormal{see} kaada}

\fmtLemma{kaa\-fi\-yu}
\fmtCitation{(e)kaa\-fi\-yu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{curfew}
\fmtExampleRL{Aba\-po\-lii\-si baa\-koo\-re ekaa\-fi\-yu obwi\-re.}
\fmtExampleEN{The police enforced a curfew at night.}

\fmtLemma{kaa\-ka}%
\fmtOne{1}
\fmtCitation{(oku)\-kaa\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kaa\-ki\-re}
\fmtSense{to slap hard, to strike, to smack}
\fmtExampleRL{Ya\-mu\-kaa\-ki\-re olu\-yi.}
\fmtExampleEN{He slapped him hard.}

\fmtLemma{kaa\-ka}%
\fmtOne{2}
\fmtCitation{\textnormal{(e.g.}\,mu\-kaa\-ka, ba\-kaa\-ka, ki\-kaa\-ka\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{(of a person) tall and fat}
\fmtExampleRL{OGa\-li\-ma\-ka yaa\-li mu\-sai\-za mu\-kaa\-ka.}
\fmtExampleEN{Galimaka was a tall and fat man.}
\fmtSubentry(b)
\fmtSense{assertive}
\fmtExampleRL{Oi\-taa\-mwe yaa\-li mu\-sai\-za mu\-kaa\-ka.}
\fmtExampleEN{His father was a very assertive man.}

\fmtLemma{kaa\-ka\-lya\-nga\-no}
\fmtCitation{(a)kaa\-ka\-lya\-nga\-no}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{commotion, disturbance, uproar, disorder, confusion, chaos, mayhem, unrest}
\fmtExampleRL{Akaa\-ka\-lya\-nga\-no ka\-caa\-ki\-re omu Ka\-mpa\-la.}
\fmtExampleEN{There is a lot of commotion in Kampala.}

\fmtLemma{kaa\-ka\-ti}
\fmtPOS{adv}
\fmtSense{now}
\fmtExampleRL{Kaa\-ka\-ti tu\-ta\-ndi\-ki\-re ani.}
\fmtExampleEN{Now, let us begin here.}

\fmtLemma{kaa\-ki}
\fmtCitation{(e)kaa\-ki}
\fmtPOS{n}%
\fmtNC{9}
\fmtSource{from En.}
\fmtSense{khaki fabric}
\fmtExampleRL{Engo\-ye za kaa\-ki zi\-bba zi\-zi\-bu okwo\-za n’o\-bba nga to\-li\-na kyo\-ma kyo\-za.}
\fmtExampleEN{Khaki clothes are difficult to wash if you don't have a washing machine.}

\fmtLemma{kaa\-ki\-mwe}
\fmtCrossreference{\textnormal{see} kaakimwei}

\fmtLemma{kaa\-ki\-mwei}
\fmtPOS{adv}
\fmtSense{thoroughly, at all, absolutely, totally, once and for all}
\fmtExampleRL{Ya\-ya\-bii\-re kaa\-ki\-mwei.}
\fmtExampleEN{He left once and for all.}

\fmtLemma{kaa\-ku\-la}
\fmtCitation{(oku)\-kaa\-ku\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kaa\-kwi\-re}
\fmtSubentry(a)
\fmtSense{to grab, to seize, to grasp, to snatch away, to take away}
\fmtExampleRL{Omwi\-bbi anka\-kwi\-re\-ku ensa\-wu ya\-nge n’ai\-ru\-ka.}
\fmtExampleEN{The thief grabbed my bag and ran away.}
\fmtSubentry(b)
\fmtSense{to clear (e.g.\,a path)}
\fmtExampleRL{Kaa\-ku\-la omu\-wa\-nda omu ki\-si\-ko aba\-ntu baa\-ka\-le\-wo.}
\fmtExampleEN{Clear a path in the bush for people to pass.}

\fmtLemma{kaa\-la ake\-ree\-ra}
\fmtCitation{(a)kaa\-la ake\-ree\-ra}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{baby finger, pinkie}
\fmtExampleRL{Akaa\-la akee\-re\-ra ni\-ko ka\-si\-nga\-mu obu\-doo\-li oku\-bya\-la byo\-na\-byo\-na.}
\fmtExampleEN{The baby finger is the smallest of all the fingers.}

\fmtLemma{kaa\-laa\-la}
\fmtCitation{(o)kaa\-laa\-la}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{\textit{Bi\-dens pi\-lo\-sa}}
\fmtExampleRL{Okaa\-laa\-la awo\-nya ebi\-rwai\-re.}
\fmtExampleEN{\textit{Bidens pilosa} is used to treat burns.}

\fmtLemma{kaa\-li}
\fmtCitation{(a)kaa\-li}
\fmtPOS{n}%
\fmtNC{12}
\fmtSubentry(a)
\fmtSense{crib, basket-like structure from papyrus which serves as a crib for babies}
\fmtSubentry(b)
\fmtSense{bed of a newborn and newborn's mother}
\fmtExampleRL{Akaa\-li ti\-ka\-li\-na ku\-bba\-si\-bwa\-ku mu\-sai\-za.}
\fmtExampleEN{It is taboo for a man to sleep in the bed of a woman who has recently given birth.}

\fmtLemma{kaa\-ma}
\fmtCitation{(o)kaa\-ma}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{type of wild yam eaten during famine}
\fmtExampleRL{Okaa\-ma ba\-te\-ra ku\-mu\-lya kwa nza\-la.}
\fmtExampleEN{The wild yam is usually eaten during famine.}

\fmtLemma{kaa\-mu}
\fmtCrossreference{\textnormal{see} (-)amu}

\fmtLemma{kaa\-mu\-je}
\fmtCitation{(o)kaa\-mu\-je}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{squirrel}
\fmtExampleRL{Okaa\-mu\-je alii\-re amai\-do gai\-swe goo\-na\-goo\-na.}
\fmtExampleEN{The squirrel has eaten all our groundnuts.}

\fmtLemma{kaa\-mu\-kuu\-ku\-lu}
\fmtCitation{(o)kaa\-mu\-kuu\-ku\-lu}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{ringdove (\textit{Streptopelia ri\-so\-ria})}
\fmtExampleRL{Okaa\-mu\-kuu\-ku\-lu abo\-ne\-ka ku\-sai.}
\fmtExampleEN{The ringdove looks nice.}

\fmtLemma{kaa\-mu\-sye\-ke\-re}
\fmtCitation{(o)kaa\-mu\-sye\-ke\-re}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{type of insect which lives in water and burns one's skin if one comes into contact with it}

\fmtLemma{kaa\-mwe\-(i)}
\fmtCrossreference{\textnormal{see} amwe\textnormal{{\fmtOne{1}}}}

\fmtLemma{kaa\-na}
\fmtCitation{(a)kaa\-na}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{small child}
\fmtSubentry(b)
\fmtSense{chick}
\fmtExampleRL{Obwa\-na bwa nko\-ko ni bu\-ta\-fu\-na bbu\-gu\-mu eri\-ku\-ma\-la bu\-fa.}
\fmtExampleEN{If chicks do not get enough warmth they die.}
\fmtSubentry(c)
\fmtSense{girl identified for marriage (lit.\,chick)}
\fmtExampleRL{Omwo\-jo wa\-nge akuu\-re bai\-ra\-ba mpee\-rya akaa\-na ka nko\-ko.}
\fmtExampleEN{My son is now a man, would you please give him your daughter in marriage?}

\fmtLemma{kaa\-nywe}
\fmtCrossreference{\textnormal{see} nywe}

\fmtLemma{kaa\-sa}
\fmtCitation{(o)kaa\-sa}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{white-strip headed calf}
\fmtExampleRL{Kaa\-sa wa\-nge alwai\-re.}
\fmtExampleEN{My white-striped headed calf is sick.}

\fmtLemma{kaa\-to}
\fmtCitation{(a)kaa\-to}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{canoe, boat}
\fmtExampleRL{Aba\-ma\-gye ba\-kwa\-ta obwa\-to bwe ba\-sa\-nga oku nya\-nja.}
\fmtExampleEN{The army men confiscate canoes they find on the lake.}
\fmtSubentry(b)
\fmtSense{feeder (for animals)}
\fmtExampleRL{Ente zi\-ku\-lii\-ra mu kaa\-to.}
\fmtExampleEN{The cows eat from a feeder.}
\fmtSubentry(c)
\fmtFixedexpression{kaa\-to ka ingi\-ni}
\fmtSense{engine boat, motorboat}
\fmtExampleRL{Akaa\-to ka ingi\-ni ka\-kwa\-ndaa\-za enya\-nja.}
\fmtExampleEN{An engine boat is patrolling the lake.}
\fmtSubentry(d)
\fmtFixedexpression{kaa\-to ka mu\-uro}
\fmtSense{engine boat, motorboat}

\fmtLemma{kaa\-tuu\-ka}%
\fmtOne{1}
\fmtCitation{(oku)\-kaa\-tuu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kaa\-tuu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to go at random, to wander blindly, to wander aimlessly}
\fmtSubentry(b)
\fmtSense{to walk hurriedly}
\fmtExampleRL{Lwa\-ki oku\-kaa\-tuu\-ka mu\-no?}
\fmtExampleEN{Why are you walking so hurriedly?}
\fmtSubentry(c)
\fmtSense{to become unfriendly, to become hostile}
\fmtExampleRL{Omu\-sai\-za akaa\-tuu\-ki\-re, mu\-zwe\-re.}
\fmtExampleEN{The man has become unfriendly, get away from him.}

\fmtLemma{kaa\-tuu\-ka}%
\fmtOne{2}
\fmtCitation{(oku)\-kaa\-tuu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kaa\-tuu\-ki\-re}
\fmtSubentry(a)
\fmtSense{(of beer) to ferment excessively, to become sour, (of herbs, medicine) to go bad}
\fmtExampleRL{Omwe\-nge gu\-kaa\-tuu\-ki\-re.}
\fmtExampleEN{The beer was overfermented.}
\fmtSubentry(b)
\fmtSense{to become an old maid, to pass the marriageable age}

\fmtLemma{kaa\-tuu\-ki}
\fmtCitation{\textnormal{(e.g.}\,mu\-kaa\-tuu\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{(of beer) sour, old, (of local herbs) expired}
\fmtExampleRL{Omu\-ba\-zi gu\-ni gu\-kaa\-tuu\-ki to\-gu\-wa mwa\-na.}
\fmtExampleEN{This medicine has expired. Do not give it to the child.}

\fmtLemma{kaa\-wa}%
\fmtOne{1}
\fmtCitation{(o)kaa\-wa}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from Sw.}
\fmtSense{coffee}
\fmtExampleRL{Taa\-ku okaa\-wa oku ma\-tai.}
\fmtExampleEN{Add coffee to the milk.}

\fmtLemma{Kaa\-wa}%
\fmtOne{2}
\fmtCitation{(o)Kaa\-wa}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from Hebrew}
\fmtSense{Eve}
\fmtExampleRL{OKaa\-wa yaa\-li mu\-ka\-li wa Ada\-mu bai\-ka\-la\-nga mu lu\-su\-ku Ede\-n nga ba\-kya\-li ku\-sii\-sa wa Ka\-nca.}
\fmtExampleEN{Eve was Adam's wife, and both lived in Eden before they sinned against God.}

\fmtLemma{kaa\-ya\-na}
\fmtCitation{(oku)\-kaa\-ya\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-yai\-ne}
\fmtSense{to argue, to fight, to quarrel, to protest, to dispute}
\fmtExampleRL{Aba\-ka\-li ba\-bbai\-re ba\-ku\-kaa\-ya\-na oku\-zwa ma\-mbya.}
\fmtExampleEN{The women have been arguing since morning.}

\fmtLemma{kaa\-yi}
\fmtCitation{(a)kaa\-yi}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{banana leaf tied around a pot (Traditionally, the delegation visiting the in-laws for a marriage ceremony carries a pot of local brew with a banana leaf tied around it. It forms the basis of all the negotiations that are involved in this function.)}
\fmtExampleRL{Bbwa akaa\-yi omu ki\-taa\-bbu kya ma\-lwa twa\-be e bu\-koi.}
\fmtExampleEN{Tie a banana leaf around a gourd of alcohol and let's go to the in-laws.}

\fmtLemma{ka\-ba}
\fmtCitation{(oku)\-ka\-ba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-bi\-re}
\fmtSense{to faint, to become unconscious, to pass out, to lose consciousness}
\fmtExampleRL{Omwa\-na ya\-lwai\-re eki\-bi\-ri\-bi\-ri n’a\-ka\-ba.}
\fmtExampleEN{The child had malaria and fell unconscious.}

\fmtLemma{ka\-ba\-da}
\fmtCitation{(e)ka\-ba\-da}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{cupboard}
\fmtExampleRL{Ebi\-ko\-po bi\-ri mu ka\-ba\-da ti mu se\-fu\-li\-ya.}
\fmtExampleEN{The cups are kept in the cupboard, not in the saucepan.}

\fmtLemma{ka\-ba\-ju}
\fmtCitation{(a)ka\-ba\-ju}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{ladle (usually a piece of a broken calabash), scooping spoon}
\fmtExampleRL{Lee\-ta aka\-ba\-ju tu\-se\-ne omu\-ku\-bi.}
\fmtExampleEN{Bring the ladle and we will serve the sauce.}

\fmtLemma{ka\-ba\-ka}
\fmtCitation{(o)ka\-ba\-ka}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSubentry(a)
\fmtSense{king}
\fmtExampleRL{Oka\-ba\-ka wa Bu\-ga\-nda akwi\-za e Ka\-yu\-nga.}
\fmtExampleEN{The king of Buganda is coming to Kayunga.}
\fmtSubentry(b)
\fmtSense{title given to the king of Buganda Kingdom}
\fmtExampleRL{Omu\-ka\-ma wa Bu\-ga\-nda ba\-mwe\-ta Ka\-ba\-ka.}
\fmtExampleEN{Kabaka is the title given to the king of Buganda.}

\fmtLemma{ka\-ba\-la}
\fmtCitation{(oku)\-ka\-ba\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-bai\-re}
\fmtSense{to spade over, to dig out}
\fmtExampleRL{Ka\-ba\-la otoo\-le\-yo lu\-cwa\-mba.}
\fmtExampleEN{Dig out that couch grass.}

\fmtLemma{ka\-ba\-la\-ga\-la}
\fmtCitation{(o)ka\-ba\-la\-ga\-la}
\fmtPOS{n}%
\fmtNC{1a/14}
\fmtSource{from Lug.} 
\fmtSense{pancake (from ripe sweet bananas and cassava flour with some pepper)}
\fmtExampleRL{Obu\-ka\-ba\-la\-ga\-la bwa bi\-ri bi\-ni ti\-bu\-woo\-ma.}
\fmtExampleEN{The pancakes baked these days are not tasty.}

\fmtLemma{Ka\-ba\-le\-ga}
\fmtCitation{(o)Ka\-ba\-le\-ga}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{Kabalega (a king of Bu\-nyoro)}
\fmtExampleRL{OKa\-ba\-le\-ga yaa\-li mu\-zi\-ra ya Bu\-nyo\-ro.}
\fmtExampleEN{Kabalega was a hero of the Bu\-nyoro Kingdom.}

\fmtLemma{Ka\-ba\-mbai\-gu\-ru}
\fmtCitation{(o)Ka\-ba\-mbai\-gu\-ru}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSubentry(a)
\fmtSense{creator of heavens}
\fmtExampleRL{OKa\-nca ni\-ye Ka\-ba\-mbai\-gu\-ru.}
\fmtExampleEN{God is the creator of the heavens.}
\fmtSubentry(b)
\fmtSense{one of the titles of Bu\-nyoro and Tooro kings}
\fmtExampleRL{Angi\-ri\-za oKa\-ba\-mbai\-gu\-ru.}
\fmtExampleEN{Long live the king of Bu\-nyoro-Tooro.}

\fmtLemma{ka\-ba\-ndo\-le}
\fmtCitation{(e)ka\-ba\-ndo\-le}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{double-decker bus}
\fmtExampleRL{Aba\-nya\-la ba\-ta\-ka mu\-no oku\-ta\-mbuu\-ra omu ka\-ba\-ndo\-le.}
\fmtExampleEN{The Banyala love to travel in double-decker buses.}

\fmtLemma{ka\-ba\-ngo}
\fmtCitation{(a)ka\-ba\-ngo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{disobedience, misbehaviour}
\fmtExampleRL{Aka\-ba\-ngo ka\-yi\-nza oku\-lee\-ta obu\-zu\-bu.}
\fmtExampleEN{Disobedience may lead to problems.}

\fmtLemma{ka\-bba}
\fmtCitation{(e)ka\-bba}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{yoke (of a shirt), the upper-back part (of a shirt)}
\fmtExampleRL{Esaa\-ti n'e\-ta\-bbaa\-ku ka\-bba ebba bbu\-la\-wu\-zi.}
\fmtExampleEN{If a shirt does not have a design on the yoke, then it is a blouse.}

\fmtLemma{ka\-bbaa\-ni}
\fmtCitation{(a)ka\-bbaa\-ni}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{incense}
\fmtExampleRL{Cuu\-kii\-rya obu\-bbaa\-ni omu nyu\-mba.}
\fmtExampleEN{Burn the incense in the house.}

\fmtLemma{ka\-bbaa\-ra}
\fmtCitation{(a)ka\-bbaa\-ra}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{parched land}
\fmtExampleRL{Aka\-bbaa\-ra ti\-ka\-me\-ra\-ku ki\-ntu.}
\fmtExampleEN{Parched land is not good for growing crops.}

\fmtLemma{ka\-bbaa\-sa}
\fmtCrossreference{\textnormal{see also} bba\-sa}
\fmtCitation{(a)ka\-bbaa\-sa}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSource{from Sw.}
\fmtSense{burial shroud (wrapped in an envelope-like style), burial sheets}
\fmtExampleRL{Oru\-fa ba\-mu\-bbwa ka\-bbaa\-sa.}
\fmtExampleEN{They wrapped the corpse in an envelope-like style.}

\fmtLemma{ka\-bba\-ngo}
\fmtCitation{(a)ka\-bba\-ngo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{hunch back}
\fmtExampleRL{Osaa\-lii\-rwe to\-boi\-ne mu\-ntu wa ka\-bba\-ngo.}
\fmtExampleEN{You have missed seeing a person with a hunchback.}
\fmtSubentry(b)
\fmtSense{nickname given to someone with a hunchback}
\fmtExampleRL{OKa\-bba\-ngo ai\-zi\-re.}
\fmtExampleEN{Kabbango has come.}

\fmtLemma{ka\-bba\-ni}%
\fmtOne{1}
\fmtCitation{(a)ka\-bba\-ni}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{squeeze dance (dancing style where two people dance embracing each other)}
\fmtExampleRL{Ba\-ku\-bi\-na ka\-bba\-ni.}
\fmtExampleEN{They are dancing the squeeze dance.}

\fmtLemma{ka\-bba\-ni}%
\fmtOne{2}
\fmtCitation{(a)ka\-bba\-ni}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{protuberance, protruding stomach, bulging belly (\eg of a pregnant woman)}
\fmtExampleRL{Omu\-ka\-li ali\-na enda y’a\-ka\-bba\-ni.}
\fmtExampleEN{The pregnant woman has a belly which is growing upward.}

\fmtLemma{ka\-bba\-ra\-ra}
\fmtCrossreference{\textnormal{see} kabbaara}

\fmtLemma{ka\-bbe\-bbe\-nee\-rwa}
\fmtCitation{(a)ka\-bbe\-bbe\-nee\-rwa}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{cartilage (of an animal), gristle}
\fmtExampleRL{Nta\-ka mu\-no oku\-lya aka\-bbe\-bbe\-nee\-rwa.}
\fmtExampleEN{I like eating animal cartilage.}

\fmtLemma{ka\-bbe\-bbe\-nu}
\fmtCitation{(a)ka\-bbe\-bbe\-nu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{edge, boundary, fringe, margin}
\fmtExampleRL{Aye\-me\-rei\-re ku ka\-bbe\-bbe\-nu ka lya\-to.}
\fmtExampleEN{He is standing on the edge of the boat.}

\fmtLemma{ka\-bbe\-bbe\-nwa}
\fmtCrossreference{\textnormal{see} kabuuro}

\fmtLemma{ka\-bbi}
\fmtCitation{(a)ka\-bbi}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{danger, emergency, risk, crisis}
\fmtExampleRL{Nku\-suu\-bi\-ra\-yo aka\-bbi omu nyu\-mba.}
\fmtExampleEN{I sense danger in the house.}

\fmtLemma{ka\-bbi\-bbii\-ga\-no}
\fmtCitation{(a)ka\-bbi\-bbii\-ga\-no}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{chaos, tumult, disorder, turmoil}
\fmtExampleRL{Aka\-bbi\-bbii\-ga\-no ka kai\-ngi omu nyu\-mba!}
\fmtExampleEN{There is a lot of chaos in the house!}

\fmtLemma{ka\-bbii\-ni\-ki\-ryo}
\fmtCitation{(a)ka\-bbii\-ni\-ki\-ryo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{cover, lid}
\fmtExampleRL{Atei\-re\-ku aka\-bbii\-ni\-ki\-ryo oku kii\-na.}
\fmtExampleEN{He has put a cover on the hole.}
\fmtSubentry(b)
\fmtSense{knee cap}
\fmtExampleRL{Oku\-jwi ku\-bbaa\-ku aka\-bbii\-ni\-ki\-ryo.}
\fmtExampleEN{The knee has a part called a knee cap.}

\fmtLemma{ka\-bbo\-nge}
\fmtCitation{(o)ka\-bbo\-nge}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{small pinkish variety of groundnuts}
\fmtExampleRL{Ndwi\-re oku\-bo\-na ebi\-nyee\-bwa bya ka\-bbo\-nge.}
\fmtExampleEN{It's been a long time since I saw the small pinkish variety of groundnut.}

\fmtLemma{ka\-bbu\-li}
\fmtCitation{(e)ka\-bbu\-li}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{grave of a Muslim}
\fmtExampleRL{Eka\-bbu\-li za ba\-si\-raa\-mu abaa\-fii\-re mu nsi\-ngwa ni\-zo zi\-ni.}
\fmtExampleEN{These are the graves of the Muslims who died during the war.}

\fmtLemma{ka\-bbuu\-ru}
\fmtCitation{(a)ka\-bbuu\-ru}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{gang, group, band}
\fmtExampleRL{Oku kya\-lo ki\-ni ku\-li\-ku aka\-bbuu\-ru ka bai\-bbi.}
\fmtExampleEN{There is a gang of thieves in this village.}

\fmtLemma{ka\-bbwe\-ne}
\fmtCitation{(a)ka\-bbwe\-ne}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{puppy}
\fmtExampleRL{Ampai\-re aka\-bbwe\-ne nka\-ku\-li\-sye.}
\fmtExampleEN{He has given me a puppy to rear.}

\fmtLemma{ka\-bbwi}
\fmtCitation{(a)ka\-bbwi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{bolt}
\fmtExampleRL{Aka\-bbwi ka lwi\-ge lwa\-nge ka\-zwi\-re\-ku.}
\fmtExampleEN{The bolt of my door has fallen off.}

\fmtLemma{ka\-bbwi\-bbwi}
\fmtCitation{(a)ka\-bbwi\-bbwi}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{dawn, very early in the morning}
\fmtExampleRL{Ya\-zi\-ndu\-ki\-re mu ka\-bbwi\-bbwi.}
\fmtExampleEN{He left very early in the morning.}

\fmtLemma{ka\-bbwi\-ku\-bbwi\-ku}
\fmtCitation{(a)ka\-bbwi\-ku\-bbwi\-ku}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{dawn, very early morning}
\fmtExampleRL{Abo aba\-ku\-ta\-ka oku\-te\-ga ebi\-na\-ge ba\-kee\-ra\-mbe mu ka\-bbwi\-ku\-bbwi\-ku.}
\fmtExampleEN{Those who want to fish tilapia have to get up at dawn.}

\fmtLemma{ka\-be\-nje}
\fmtCitation{(a)ka\-be\-nje}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSource{from Lug.} 
\fmtSense{accident, misfortune}
\fmtExampleRL{Omu\-sai\-za ya\-fu\-ni\-re aka\-be\-nje.}
\fmtExampleEN{The man was involved in an accident.}

\fmtLemma{ka\-be\-zo}
\fmtCitation{(a)ka\-be\-zo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{scooping plate, banana leaf used as a plate for serving food}
\fmtExampleRL{Kwa\-ta aka\-be\-zo ome\-ge eme\-re.}
\fmtExampleEN{Get a scooping plate and serve the food.}
\fmtSubentry(b)
\fmtSense{saucer, plate}

\fmtLemma{ka\-bi\-ne\-ti}
\fmtCitation{(e)ka\-bi\-ne\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{cabinet, office}
\fmtExampleRL{Opu\-re\-zi\-de\-nti aye\-ye\-re eka\-bi\-ne\-ti yaa\-mwe.}
\fmtExampleEN{The president has swept his cabinet.}

\fmtLemma{ka\-bi\-ri}
\fmtPOS{num}
\fmtSubentry(a)
\fmtSense{second}
\fmtExampleRL{Ya\-bbai\-re wa ka\-bi\-ri oku bee\-gi ki\-ku\-mi.}
\fmtExampleEN{He came second out of a hundred students.}
\fmtSubentry(b)
\fmtSense{twice, two times}
\fmtExampleRL{Abbai\-re ani ka\-bi\-ri.}
\fmtExampleEN{He has been here twice.}
\fmtSubentry(c)
\fmtFixedexpression{mwe\-zi gwa ka\-bi\-ri}
\fmtSense{February}

\fmtLemma{ka\-bi\-rii\-ti}
\fmtCitation{(a)ka\-bi\-rii\-ti}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSource{from Sw.}
\fmtSense{match (for starting a fire), matchstick}
\fmtExampleRL{Mu\-we\-rye\-yo aka\-bi\-rii\-ti ka\-mwei aku\-me omuu\-ro.}
\fmtExampleEN{Give him a matchstick, and he will light the fire.}

\fmtLemma{ka\-bo\-go}
\fmtCitation{(o)ka\-bo\-go}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{dark-black goat}
\fmtExampleRL{Oka\-bo\-go wa\-nge awa\-ki\-re.}
\fmtExampleEN{My dark-black goat has conceived.}

\fmtLemma{ka\-bo\-go\-lo\-le}
\fmtCitation{(a)ka\-bo\-go\-lo\-le}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{big house, large building}
\fmtExampleRL{Enyu\-mba ka\-bo\-go\-lo\-le ti gya\-ngu kwo\-mbo\-ka.}
\fmtExampleEN{It is not easy to construct a big house.}

\fmtLemma{ka\-bo\-na}
\fmtCitation{(o)ka\-bo\-na}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{priest}
\fmtExampleRL{Ba\-ka\-bo\-na aba\-ka\-tu\-li\-ki ba\-so\-ma mi\-sa bu\-li Sa\-nde.}
\fmtExampleEN{Catholic priests lead masses in church every Sunday.}

\fmtLemma{ka\-bo\-ne\-ro}
\fmtCitation{(a)ka\-bo\-ne\-ro}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{sign, signal}
\fmtExampleRL{OYe\-su ali\-taa\-ku aka\-bo\-ne\-ro oku baa\-na baa\-mwe.}
\fmtExampleEN{Jesus will send a sign to his people.}
\fmtSubentry(b)
\fmtSense{symbol}
\fmtExampleRL{Ako aka\-bo\-ne\-ro ka\-ma\-nyi\-sya ku\-bbaa\-wo kwa irwa\-ro.}
\fmtExampleEN{That symbol signifies the presence of a hospital.}
\fmtSubentry(c)
\fmtSense{score, mark}
\fmtExampleRL{Afu\-ni\-re obu\-bo\-ne\-ro bwi\-ngi.}
\fmtExampleEN{He got high scores.}
\fmtSubentry(d)
\fmtSense{symptom}
\fmtExampleRL{Oku\-ta\-na\-ka oluu\-si ka\-bba ka\-bo\-ne\-ro ka na\-mu\-nywa.}
\fmtExampleEN{Vomiting is sometimes a symptom of HIV/AIDS.}

\fmtLemma{ka\-bu\-go ka mwa\-na}
\fmtCitation{(a)ka\-bu\-go ka mwa\-na}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{baby sling}
\fmtExampleRL{Mu\-mpee\-rye aka\-bu\-go ka mwa\-na.}
\fmtExampleEN{Give her the baby sling for me please.}

\fmtLemma{ka\-bu\-gu}
\fmtCitation{(a)ka\-bu\-gu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{type of shrub used to baskets for fish}
\fmtExampleRL{Aka\-bu\-gu tu\-ka\-bbwe\-sya ki\-ba\-ga\-lo.}
\fmtExampleEN{We use the kabugu shrub to tie the basket for fish.}

\fmtLemma{ka\-bu\-ki}
\fmtCitation{\textnormal{(e.g.}\,mu\-ka\-bu\-ki, ba\-ka\-bu\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{(of a person) creative}
\fmtExampleRL{Omu\-ka\-li wa\-nge mu\-ka\-li mu\-ka\-bu\-ki.}
\fmtExampleEN{My wife is a creative woman.}

\fmtLemma{ka\-bu\-ki\-ka\-bu\-ki}
\fmtCitation{\textnormal{(e.g.}\,mu\-ka\-bu\-li\-ka\-bu\-li, ba\-ka\-bu\-li\-ka\-bu\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSense{sharp, keen, acute, quick-witted, clever, intelligent}
\fmtExampleRL{Omu\-ka\-li wa\-nge mu\-ka\-bu\-ki\-ka\-bu\-ki.}
\fmtExampleEN{My wife is clever.}

\fmtLemma{Ka\-bu\-la}
\fmtCitation{(o)Ka\-bu\-la}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{Kabula (one of the counties that make Buganda kingdom)}
\fmtExampleRL{Ka\-bu\-la li\-mwei oku ma\-sa\-za ga Bu\-ga\-nda.}
\fmtExampleEN{Kabula is one of Buganda's counties.}

\fmtLemma{ka\-bu\-lii\-ro}
\fmtCrossreference{\textnormal{see} kabuuro}

\fmtLemma{ka\-bu\-nda}
\fmtCitation{(a)ka\-bu\-nda}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{belly}
\fmtExampleRL{Ba\-mu\-cu\-pwi\-re omu\-ge\-re oku ka\-bu\-nda.}
\fmtExampleEN{He was kicked in the belly.}

\fmtLemma{ka\-bu\-ra\-bu\-zo}
\fmtPOS{adv}
\fmtSense{wrongly, incorrectly (e.g. of fastening with buttons)}
\fmtExampleRL{Ama\-pee\-sa ga saa\-ti aga\-bbwe\-re ka\-bu\-ra\-bu\-zo.}
\fmtExampleEN{He buttoned his shirt wrongly.}

\fmtLemma{ka\-buu\-lya}
\fmtCitation{(a)ka\-buu\-lya}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{question mark}
\fmtExampleRL{Taa\-ku aka\-buu\-lya ku ki\-buu\-lyo.}
\fmtExampleEN{Put a question mark at the end of the question.}

\fmtLemma{ka\-buu\-ro}
\fmtCitation{(a)ka\-buu\-ro}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{long period of absence}
\fmtExampleRL{Aka\-buu\-ro ka\-bbai\-re ka\-maa\-ni oku\-zwa omaa\-wo ni ya\-fii\-re.}
\fmtExampleEN{It has been a long period of absence since your mother passed away.}

\fmtLemma{ka\-buu\-ti}
\fmtCitation{(e)ka\-buu\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{coat, winter coat}
\fmtExampleRL{Eka\-buu\-ti te\-ya\-ka\-la\-mu mai\-zi.}
\fmtExampleEN{This coat is water resistant.}

\fmtLemma{ka\-bwi\-bwi}
\fmtCitation{(o)ka\-bwi\-bwi}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{fruit fly}
\fmtExampleRL{Obu\-bwi\-bwi bu\-ta\-ka mu\-no ame\-nju.}
\fmtExampleEN{Fruit flies like yellow bananas a lot.}

\fmtLemma{ka\-bwi\-re}
\fmtPOS{adv}
\fmtSense{at dawn}
\fmtExampleRL{Nai\-za ka\-bwi\-re.}
\fmtExampleEN{I will come at dawn.}

\fmtLemma{ka\-cai}
\fmtCitation{(a)ka\-cai}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{bad manners, insolence, rudeness, bad manners}
\fmtExampleRL{Aka\-cai mu\-ze gu\-bbi.}
\fmtExampleEN{Insolence is a bad habit.}

\fmtLemma{ka\-ce\-nce}
\fmtCitation{(a)ka\-ce\-nce}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{loin cloth (worn by unmarried girls)}
\fmtExampleRL{Omwa\-la yee\-bbwe\-re aka\-ce\-nce.}
\fmtExampleEN{The girl is wearing a loincloth.}
\fmtSubentry(b)
\fmtSense{raffia skirt, grass skirt}
\fmtExampleRL{Mu\-lee\-te obu\-ce\-nce twe\-bbu\-we tu\-bi\-ne.}
\fmtExampleEN{Bring the raffia skirts, we will put them on and dance.}

\fmtLemma{ka\-cu\-lyai}
\fmtCitation{(o)ka\-cu\-lyai}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSubentry(c)
\fmtSense{black bulbul (\textit{Hypsipetes leucoce\-pha\-lus})}
\fmtExampleRL{Oka\-cu\-lyai amaa\-re\-wo oka\-la\-li wa\-nge.}
\fmtExampleEN{The black bulbul has eaten all my red pepper.}

\fmtLemma{ka\-cu\-ngwa}
\fmtCrossreference{\textnormal{see} kacungwai}

\fmtLemma{ka\-cu\-ngwai}
\fmtCitation{(a)ka\-cu\-ngwai}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{orange colour}
\fmtExampleRL{Nye\-ndya mu\-no era\-ngi ya ka\-cu\-ngwai.}
\fmtExampleEN{I like the orange colour a lot.}

\fmtLemma{ka\-cuu\-cu}
\fmtPOS{adv}
\fmtSense{very speedily, quickly}
\fmtExampleRL{Avu\-gi\-re egaa\-li ka\-cuu\-cu.}
\fmtExampleEN{He rode the bicycle very fast.}

\fmtLemma{ka\-cuu\-cuu\-ke}
\fmtCitation{(o)ka\-cuu\-cuu\-ke}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{skin rash}
\fmtExampleRL{Alwai\-re oka\-cuu\-cuu\-ke.}
\fmtExampleEN{He is suffering from a skin rash.}

\fmtLemma{ka\-cuu\-di}%
\fmtOne{1}
\fmtCitation{(a)ka\-cuu\-di}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{drill}
\fmtExampleRL{Nya\-zi\-ka aka\-juu\-di kaa\-mu.}
\fmtExampleEN{Lend me your drill.}

\fmtLemma{ka\-cuu\-di}%
\fmtOne{2}
\fmtCitation{(a)ka\-cuu\-di}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{coccyx}
\fmtExampleRL{Ngwi\-re aka\-cuu\-di ni ka\-ndu\-ma.}
\fmtExampleEN{I have fallen and my coccyx hurts.}

\fmtLemma{ka\-cuu\-li}
\fmtCitation{(a)ka\-cuu\-li}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{sandpaper}
\fmtExampleRL{Ko\-le\-sya aka\-cuu\-li onsa\-yi\-rii\-rye erya\-to lya\-nge.}
\fmtExampleEN{Use the sandpaper to smooth my boat for me.}

\fmtLemma{ka\-cuu\-muu\-ko}
\fmtCitation{(a)ka\-cuu\-muu\-ko}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{good smell, aroma}
\fmtExampleRL{Bai\-tu aka\-cuu\-muu\-ko ako ka\-li\-gi\-re ayi?}
\fmtExampleEN{Where is the good smell coming from?}

\fmtLemma{ka\-cwa\-no}
\fmtCitation{(a)ka\-cwa\-no}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{trouble, problem, difficulty}
\fmtExampleRL{Ota\-tu\-lee\-te\-ra ka\-cwa\-no.}
\fmtExampleEN{Do not cause us trouble.}

\fmtLemma{ka\-cwa\-nte}
\fmtCitation{(o)ka\-cwa\-nte}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{beloved person}
\fmtExampleRL{Oka\-cwa\-nte waa\-mwe afii\-re.}
\fmtExampleEN{His beloved one is dead.}

\fmtLemma{ka\-da ka ru\-ba\-ngo}
\fmtCitation{(a)ka\-da ka ru\-ba\-ngo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{blood relative of the same father}
\fmtExampleRL{Tu\-li ka\-da ka ru\-ba\-ngo.}
\fmtExampleEN{We are relatives of the same father.}
\fmtSubentry(a)
\fmtSense{blood relatives of the same mother}
\fmtExampleRL{Tu\-li ka\-da ka nai\-ru\-nda.}
\fmtExampleEN{We are blood relatives of the same mother.}
\fmtSubentry(b)
\fmtSense{solidarity, unity, harmony}
\fmtExampleRL{Ka tu\-kuu\-me aka\-da ka nai\-ru\-nda okwi\-ka oku nka\-ngo.}
\fmtExampleEN{Let us keep this solidarity till the end.}

\fmtLemma{ka\-daa\-li}%
\fmtOne{1}
\fmtCitation{(a)ka\-daa\-li}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{bridal crown, bridal head cover}
\fmtExampleRL{Omu\-go\-le azwai\-re ka\-daa\-li.}
\fmtExampleEN{The bride is wearing a bridal crown.}

\fmtLemma{ka\-daa\-li}%
\fmtOne{2}
\fmtCitation{(o)ka\-daa\-li}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{\textit{Asparagus africanus} (climbing asparagus fern, a climbing plant with thorny main stems)}
\fmtExampleRL{Oka\-daa\-li ka\-ndo\-be\-re.}
\fmtExampleEN{The thorn of \textit{Asparagus africanus} pricked me.}

\fmtLemma{ka\-dai\-dai}
\fmtPOS{adv}
\fmtSense{thoroughly, properly}
\fmtExampleRL{Ya\-nza\-la ebi\-yi\-mba ebyo ka\-dai\-dai.}
\fmtExampleEN{Spread those beans thoroughly.}

\fmtLemma{ka\-dai\-ka\-dai}
\fmtPOS{adv}
\fmtSense{without rest}
\fmtExampleRL{Ei\-zo twa\-koo\-re ka\-dai\-ka\-dai.}
\fmtExampleEN{Yesterday we worked without rest.}

\fmtLemma{ka\-da\-ka\-da\-ni}
\fmtCitation{\textnormal{(e.g.}\,mu\-ka\-da\-ka\-da\-ni, ba\-ka\-da\-ka\-da\-ni\textnormal{)}}
\fmtPOS{adj}
\fmtSense{innovative, creative (of a person)}
\fmtExampleRL{OSa\-ma\-li mu\-ka\-li mu\-ka\-da\-ka\-da\-ni.}
\fmtExampleEN{Samali is an innovative lady.}

\fmtLemma{ka\-da\-li\-da\-li}
\fmtCitation{(a)ka\-da\-li\-da\-li}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{tinea pedis, athlete's foot}
\fmtExampleRL{Aka\-da\-li\-da\-li ka\-ku\-ndu\-ma.}
\fmtExampleEN{I am suffering from athlete's foot.}

\fmtLemma{ka\-da\-mba\-la}
\fmtCitation{(oku)\-ka\-da\-mba\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-da\-mbai\-re}
\fmtSense{(of skin) to stiffen, to harden, to grow hard, to become stiff}
\fmtExampleRL{Ekya\-nzo ki\-ka\-da\-mbu\-li\-re bu\-ndi.}
\fmtExampleEN{The skin has stiffened so much.}

\fmtLemma{ka\-da\-mba\-li}
\fmtCitation{\textnormal{(e.g.}\,ki\-ka\-da\-mba\-li, lu\-ka\-da\-mba\-li, gi\-ka\-da\-mba\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSense{stiff, hard, inflexible}
\fmtExampleRL{Twa\-la ekya\-nzo eki\-ka\-da\-mba\-li ekyo ewa omu\-ko\-li wa nkai\-to aki\-go\-ndye.}
\fmtExampleEN{Take that stiff skin to the shoemaker for softening.}

\fmtLemma{ka\-da\-ngi\-do}
\fmtCrossreference{\textnormal{see also} kadonkiiro}
\fmtCitation{(a)ka\-da\-ngi\-do}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{Adam's apple}
\fmtExampleRL{Ya\-lwai\-re koo\-ko\-lo wa ka\-da\-ngi\-do.}
\fmtExampleEN{He is suffering from cancer of the Adam's apple.}

\fmtLemma{ka\-da\-ni\-ka\-da\-ni}
\fmtCitation{\textnormal{(e.g.}\,mu\-ka\-da\-ni\-ka\-da\-ni, ba\-ka\-da\-ni\-ka\-ka\-da\-ni\textnormal{)}}
\fmtPOS{adj}
\fmtSense{very active}
\fmtExampleRL{Oyo ni mu\-sai\-za mu\-ka\-da\-ni\-ka\-da\-ni.}
\fmtExampleEN{This man is very active.}

\fmtLemma{ka\-de}
\fmtCitation{(e)ka\-de}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{gomesi worn for manual labour}
\fmtExampleRL{Omu\-ka\-li wa\-nge mmu\-gu\-lii\-re eka\-de.}
\fmtExampleEN{I have bought my wife a gomesi to use for manual labour.}

\fmtLemma{ka\-de\-be\-de}%
\fmtOne{1}
\fmtCitation{(a)ka\-de\-be\-de}
\fmtPOS{n}%
\fmtNC{12}
\fmtSubentry(a)
\fmtSense{marshland}
\fmtExampleRL{Ente nku\-zi\-twa\-la mu ka\-de\-be\-de.}
\fmtExampleEN{I am taking the cows to the marshland.}
\fmtSubentry(b)
\fmtSense{river bank}
\fmtExampleRL{Aka\-de\-be\-de ba\-ka\-lu\-mi\-re\-ku ebi\-yaa\-ta.}
\fmtExampleEN{They have grown sweet potatoes on the riverbank.}

\fmtLemma{ka\-de\-be\-de}%
\fmtOne{2}
\fmtPOS{adv}
\fmtSense{full to the brim, stuffed}
\fmtExampleRL{Ensa\-wo agii\-zwi\-rye ka\-de\-be\-de.}
\fmtExampleEN{He stuffed the bag full.}

\fmtLemma{ka\-dee\-bee\-do}
\fmtCitation{(a)ka\-dee\-bee\-do}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{hessian square}
\fmtExampleRL{Amai\-do n’a\-ga\-twa\-la mu ka\-dee\-bee\-do.}
\fmtExampleEN{I will take the groundnuts using a hessian square.}

\fmtLemma{ka\-dee\-ya}
\fmtCitation{(a)ka\-dee\-ya}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{hessian square}
\fmtExampleRL{Obu\-dee\-ya ti\-bu\-kya\-bo\-ne\-ka mu\-ni.}
\fmtExampleEN{Hessian squares are scarce here.}

\fmtLemma{ka\-de\-nge}
\fmtCitation{(e)ka\-de\-nge}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{patrolling vehicle}
\fmtExampleRL{Ka\-de\-nge za po\-lii\-si ni\-zo zi\-kwi\-za.}
\fmtExampleEN{The police patrolling vehicles are coming.}

\fmtLemma{ka\-di}
\fmtCrossreference{\textnormal{see} (-)di}

\fmtLemma{ka\-dii\-si}
\fmtCitation{(a)ka\-dii\-si}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{local pot for drinking beer}
\fmtExampleRL{Tu\-nywe\-ra ama\-lwa omu ka\-dii\-si.}
\fmtExampleEN{We drink local beer from a kadiisi pot.}

\fmtLemma{ka\-do\-di}
\fmtCitation{(a)ka\-do\-di}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{type of drum beat usually played by the Ba\-gi\-su}
\fmtExampleRL{Aba\-gi\-su ba\-ku\-bba aka\-do\-di oku mu\-ko\-lo ogwa ku\-sa\-la emba\-lu.}
\fmtExampleEN{The Bagisu play the ``akadodi" during the circumcision ceremony.}

\fmtLemma{ka\-do\-ma}
\fmtCitation{(o)ka\-do\-ma}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{stingless wild bee}
\fmtExampleRL{Abaa\-na ba\-ta\-ka mu\-no okwa\-ku\-la ojuu\-lu wa ka\-do\-ma.}
\fmtExampleEN{The children love harvesting honey of stingless wild bees.}

\fmtLemma{ka\-do\-nkii\-ro}
\fmtCrossreference{\textnormal{see also} kadangido}
\fmtCitation{(a)ka\-do\-nkii\-ro}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{Adam's apple}
\fmtExampleRL{Amai\-zi ga\-mwa\-bi\-re omu ka\-do\-nkii\-ro.}
\fmtExampleEN{The water has choked him in the Adam's apple.}

\fmtLemma{ka\-doo\-li}
\fmtPOS{adv}
\fmtSense{nearly, almost}
\fmtExampleRL{Omwo\-jo oyo ya\-lwai\-re ka\-doo\-li afe.}
\fmtExampleEN{The boy fell sick and nearly died.}

\fmtLemma{ka\-do\-yi}
\fmtCitation{(o)ka\-do\-yi}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{expert, professional}
\fmtExampleRL{Tu\-li\-na aba\-ka\-do\-yi ba ndu\-mi ba\-doo\-li omu Uga\-nda.}
\fmtExampleEN{We have few language experts in Uganda.}

\fmtLemma{ka\-du\-li}
\fmtCitation{(a)ka\-du\-li}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{larva (of a flea, jiggers)}
\fmtExampleRL{Aba\-nju\-nzai ba\-tai\-za ani ba\-yi\-nza oku\-tu\-na\-gi\-ra obu\-du\-li bwa nju\-nzai.}
\fmtExampleEN{Those with jiggers should not come here because they may leave jigger larva here.}
\fmtSubentry(b)
\fmtSense{marijuana seed}
\fmtExampleRL{Ya\-tei\-re obu\-du\-li omu bu\-se\-ra bu\-li omu n’a\-ta\-mii\-ra.}
\fmtExampleEN{He put marijuana seeds in the millet porridge, and everyone got intoxicated.}

\fmtLemma{ka\-fa\-nso\-ni}
\fmtCitation{(a)ka\-fa\-nso\-ni}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{self-respect, dignity, pride, self-esteem}
\fmtExampleRL{Wee\-na bbaa\-mu aka\-fa\-nso\-ni.}
\fmtExampleEN{Have some self-respect.}

\fmtLemma{ka\-fee\-ko}
\fmtCitation{(a)ka\-fee\-ko}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{baby bump (a protrusion of a pregnant woman's abdomen especially when it first becomes noticeable to other people)}
\fmtExampleRL{Bo\-na omu\-ka\-li ali\-na aka\-fee\-ko.}
\fmtExampleEN{I saw a woman with a baby bump.}

\fmtLemma{ka\-fii\-fi}
\fmtCitation{(o)ka\-fii\-fi}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{waste material from cotton, sub-standard}
\fmtExampleRL{Oka\-fii\-fi mu\-mu\-na\-ge\-yo.}
\fmtExampleEN{Throw away the sub-standard cotton.}

\fmtLemma{ka\-fo}
\fmtCitation{(a)ka\-fo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{place, plot}

\fmtLemma{ka\-foo\-ke\-ro}
\fmtCitation{(a)ka\-foo\-ke\-ro}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{respiratory tract infections in babies caused by the respiratory syncytial virus (RSV)}

\fmtLemma{ka\-fu\-bba}
\fmtCrossreference{\textnormal{see} kasubba}

\fmtLemma{ka\-ga}%
\fmtOne{1}
\fmtCitation{(oku)\-ka\-ga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-gi\-re}
\fmtSubentry(a)
\fmtSense{to go for a walk, to stroll, to promenade}
\fmtExampleRL{Iza tu\-ka\-ge\-ku.}
\fmtExampleEN{Come, let's go for a walk.}
\fmtSubentry(b)
\fmtSense{to walk elegantly}
\fmtExampleRL{Ai\-zi\-re aku\-ka\-ga.}
\fmtExampleEN{He came walking elegantly.}

\fmtLemma{ka\-ga}%
\fmtOne{2}
\fmtCitation{(oku)\-ka\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-gi\-re}
\fmtSense{to detect, to discover, to identify, to notice}
\fmtExampleRL{Mbwe\-ne aka\-ga olu\-su lwa mwi\-bbi.}
\fmtExampleEN{The dog detects the thief's smell.}

\fmtLemma{ka\-gaa\-ga}
\fmtCitation{(a)ka\-gaa\-ga}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{bracelet}
\fmtExampleRL{Ka nzwa\-le\-ku aka\-gaa\-ga kaa\-mu.}
\fmtExampleEN{Let me wear your bracelet for a while.}

\fmtLemma{ka\-ga\-bo}
\fmtCitation{(o)ka\-ga\-bo}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{calf with red or black and white patches}
\fmtExampleRL{Oka\-ga\-bo wa\-nge alwai\-rwe.}
\fmtExampleEN{My calf with red and white patches is sick.}

\fmtLemma{ka\-ga\-da}
\fmtCitation{(a)ka\-ga\-da}
\fmtPOS{n}%
\fmtNC{12}
\fmtSubentry(a)
\fmtSense{(of woman) sexual desire disorder}
\fmtExampleRL{Omu\-ka\-li wa ka\-ga\-da ta\-ke\-rwa.}
\fmtExampleEN{A woman with a sexual desire disorder does not get sexually satisfied.}
\fmtSubentry(b)
\fmtSense{person who takes long to ejaculate}
\fmtExampleRL{Ti\-ki\-sai su\-mbi\-rwa mu\-sai\-za ka\-ga\-da.}
\fmtExampleEN{It is not good to marry a man who takes a long time to ejaculate.}

\fmtLemma{ka\-ga\-ndo}
\fmtCitation{(a)ka\-ga\-ndo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{white thorn acacia (\textit{Acacia hockii}, thorny bush with yellow flowers, its leaves are used for making a red dye)}

\fmtLemma{ka\-ge\-mbe\-ge\-mbe}
\fmtCitation{(a)ka\-ge\-mbe\-ge\-mbe}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{scapula bone, shoulder blade}
\fmtExampleRL{Ba\-mu\-cu\-pwi\-re n'a\-ye\-nde\-ka aka\-ge\-mbe\-ge\-mbe.}
\fmtExampleEN{He was beaten and broke his scapula bone.}

\fmtLemma{ka\-ge\-mbi}
\fmtCitation{(a)ka\-ge\-mbi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{razor blade}
\fmtExampleRL{Ya\-ba ondee\-te\-re aka\-ge\-mbi.}
\fmtExampleEN{Go and bring me a razor blade.}

\fmtLemma{ka\-ge\-ndai\-ga\-na}
\fmtCitation{(a)ka\-ge\-ndai\-ga\-na}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{mongoose}
\fmtExampleRL{Obu\-ge\-ndai\-ga\-na bu\-ga\-bai\-ne ompi\-ri.}
\fmtExampleEN{The mongooses have shared a snake.}

\fmtLemma{ka\-gi\-na}
\fmtCitation{(a)ka\-gi\-na}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{larva of a jigger, nit, egg of a louse}
\fmtExampleRL{Onju\-nzai ana\-gi\-re obu\-gi\-na.}
\fmtExampleEN{The jigger produced larvae.}

\fmtLemma{ka\-gi\-ngo}
\fmtPOS{adj}
\fmtSense{giant, big, huge}
\fmtExampleRL{Oka\-gi\-ngo wa mu\-sai\-za ngo\-yo akwi\-za.}
\fmtExampleEN{A giant man is coming.}
\fmtSubentry(a)
\fmtSense{type of god, spirit}
\fmtExampleRL{Aba\-nya\-la ba\-la\-mya oKa\-gi\-ngo.}
\fmtExampleEN{The Banyala worship a spirit called Kagingo.}
\fmtSubentry(b)
\fmtSense{God}
\fmtExampleRL{OKa\-gi\-ngo atu\-wai\-re oi\-ke\-ndi.}
\fmtExampleEN{God has blessed us with rain.}

\fmtLemma{ka\-gi\-ra\-nso\-ni}
\fmtCitation{(a)ka\-gi\-ra\-nso\-ni}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{\textit{Biophytum petersianum}}

\fmtLemma{ka\-go\-ba}
\fmtCitation{(a)ka\-go\-ba}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{win (\eg in sport), victory, triumph}
\fmtExampleRL{Yaa\-ka\-wa\-ngu\-la obu\-go\-ba bu\-taa\-nu.}
\fmtExampleEN{He has five wins.}

\fmtLemma{ka\-go\-bei\-swa}
\fmtCitation{(o)ka\-go\-bei\-swa}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{cowpea (\textit{Vigna unguiculata})}
\fmtExampleRL{Oka\-go\-bei\-swa aka\-lo\-se\-bwa oku\-ta\-mba ebi\-rwai\-re mu mi\-nwa gya baa\-na.}
\fmtExampleEN{\textit{Vigna un\-gui\-culata} is used to cure oral sores in infants.}

\fmtLemma{Ka\-go\-le}
\fmtCitation{(o)Ka\-go\-le}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{type of god, spirit (the messenger god for Nabuzaana)}
\fmtExampleRL{OKa\-go\-le akoo\-ra nk’o\-mu\-kwe\-nda wa Na\-bu\-zaa\-na.}
\fmtExampleEN{Kagole works as a messenger to Nabuzaana.}

\fmtLemma{ka\-go\-ma}
\fmtCitation{(o)ka\-go\-ma}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{eagle}
\fmtExampleRL{Oka\-go\-ma ai\-ta enku\-du.}
\fmtExampleEN{The eagle kills tortoises.}

\fmtLemma{ka\-go\-ndo}
\fmtCitation{(o)ka\-go\-ndo}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{calf with spotted skin}
\fmtExampleRL{Ya\-guu\-re onte oka\-go\-ndo.}
\fmtExampleEN{He bought a calf with spotted skin.}

\fmtLemma{ka\-go\-ngo}
\fmtCitation{(a)ka\-go\-ngo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{widthwise direction, direction parallel with a thing's width}
\fmtExampleRL{Eki\-do\-mo\-la kye\-ti\-ki\-re ku ka\-go\-ngo.}
\fmtExampleEN{Carry the jerry can widthwise.}

\fmtLemma{ka\-goo\-be}
\fmtCitation{(o)ka\-goo\-be}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{brown and black-striped calf}
\fmtExampleRL{Nku\-too\-rya ka\-goo\-be.}
\fmtExampleEN{I am looking for a brown and black-striped calf.}

\fmtLemma{ka\-go\-ro}
\fmtCitation{(o)ka\-go\-ro}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSubentry(a)
\fmtSense{lightening, thunder}
\fmtExampleRL{Oka\-go\-ro ya\-ku\-bbi\-re ente zai\-swe zoo\-na\-zoo\-na n’e\-zi\-fa.}
\fmtExampleEN{The lightning struck all our cows dead.}
\fmtSubentry(b)
\fmtSense{name given to a boy born after a lightening struck his family}
\fmtExampleRL{OKa\-go\-ro ngo\-yo akwi\-za.}
\fmtExampleEN{Kagoro is coming.}
\fmtSubentry(c)
\fmtSense{one of the gods among the Banyala/Baruuli}
\fmtExampleRL{OKa\-go\-ro ata mba\-ndwa wa Ba\-duu\-li.}
\fmtExampleEN{Kagoro is a god of the Baruuli.}

\fmtLemma{ka\-gu\-li}%
\fmtOne{1}
\fmtCitation{(a)ka\-gu\-li}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{roof apex}
\fmtExampleRL{Aka\-gu\-li ka nyu\-mba ya\-nge ti\-ka\-gu\-mi\-re.}
\fmtExampleEN{The apex of my house is not strong.}

\fmtLemma{ka\-gu\-li}%
\fmtOne{2}
\fmtCitation{(a)ka\-gu\-li}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{dock (enclosure in a criminal court where a defendant is placed), small cage, ring (for boxing)}
\fmtExampleRL{Ya\-ku\-ngii\-re omu ka\-gu\-li.}
\fmtExampleEN{He cried in the dock.}

\fmtLemma{ka\-gu\-lu}
\fmtCitation{(o)ka\-gu\-lu}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{loose woman}
\fmtExampleRL{Ti ki\-sai ku\-bba\-ndu\-la ka\-gu\-lu.}
\fmtExampleEN{It is not good to marry a loose woman.}

\fmtLemma{ka\-gu\-mba}
\fmtCitation{(a)ka\-gu\-mba}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{grudge, bitterness}
\fmtExampleRL{Omwo\-jo ata n’a\-ka\-gu\-mba oku mwo\-yo.}
\fmtExampleEN{That boy has a grudge.}

\fmtLemma{ka\-gu\-ngu}
\fmtCitation{(a)ka\-gu\-ngu}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{smell of a he-goat}
\fmtExampleRL{Omu\-sai\-za oyo awu\-nya aka\-gu\-ngu.}
\fmtExampleEN{That man smells like a he-goat.}

\fmtLemma{ka\-gu\-nguu\-li}
\fmtCitation{(a)ka\-gu\-nguu\-li}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{steep slope}
\fmtExampleRL{Aka\-so\-zi ka\-di ka\-li\-ku aka\-gu\-nguu\-li.}
\fmtExampleEN{That hill has a steep slope.}

\fmtLemma{ka\-guu\-li}
\fmtCitation{(a)ka\-guu\-li}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{hill}
\fmtExampleRL{Aba\-ntu ba\-fee\-ra oku ka\-guu\-li ka\-ni.}
\fmtExampleEN{People die on this hill.}

\fmtLemma{ka\-gwei\-gwei}
\fmtCitation{(a)ka\-gwei\-gwei}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{sisal thread}
\fmtExampleRL{Aka\-bbe\-re\-ge nka\-bwe\-re n’a\-ka\-gwei\-gwei.}
\fmtExampleEN{I tied the piglet with a sisal thread.}

\fmtLemma{kai\-ga}
\fmtDialect{Ruruuli}
\fmtCitation{(o)kai\-ga}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{bonfire, fire made by children to roast potatoes}

\fmtLemma{kai\-gai\-re}
\fmtCitation{(a)kai\-gai\-re}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{vaginismus (painful contraction of the vagina in response to sexual intercourse)}
\fmtExampleRL{Aba\-ka\-li aba\-kai\-gai\-re ti\-ba\-so\-la ku\-bba na ba\-sai\-za.}
\fmtExampleEN{Women who suffer from vaginismus cannot have sex.}

\fmtLemma{kai\-go}
\fmtCitation{(a)kai\-go}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{stick, cane}

\fmtLemma{kai\-goi\-go}
\fmtCitation{(a)kai\-goi\-go}
\fmtPOS{n}%
\fmtNC{12}
\fmtSubentry(a)
\fmtSense{blind man's buff (a children's game where one person is blindfolded and chases other participants), children's game, in which if a player is found without a stick, he is beaten until he obtains one}
\fmtExampleRL{Abaa\-na ba\-nyu\-mi\-rwa oku\-zee\-nya akai\-goi\-go.}
\fmtExampleEN{The children like to play blind man's buff.}
\fmtSubentry(b)
\fmtSense{sadist, person who enjoys beating others}
\fmtExampleRL{Omu\-so\-me\-sa kai\-goi\-go aku\-bba abaa\-na.}
\fmtExampleEN{The sadistic teacher beats his pupil.}

\fmtLemma{kai\-gu\-ru}
\fmtCitation{(a)kai\-gu\-ru}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{thunder}
\fmtExampleRL{Owu\-lii\-re akai\-gu\-ru?}
\fmtExampleEN{Did you hear the thunder?}

\fmtLemma{kai\-gu\-rwi\-gu\-ru}%
\fmtOne{1}
\fmtCitation{(a)kai\-gu\-rwi\-gu\-ru}
\fmtPOS{n}%
\fmtNC{12}
\fmtSubentry(a)
\fmtSense{small dark cloud which signals a storm with wind or hail}
\fmtExampleRL{Akai\-gu\-rwi\-gu\-ru ni ka\-kwai\-te ni ma\-nya nti oi\-ke\-ndi omu\-bbi akwa\-ba ku\-too\-nya.}
\fmtExampleEN{When a I saw a small dark cloud, I knew that the rain would not be good.}
\fmtSubentry(b)
\fmtSense{form of swearing (lit.\,'by the powers in heaven')}
\fmtExampleRL{Nda\-yi\-ra okai\-gu\-rwi\-gu\-ru.}
\fmtExampleEN{I swear by the powers in heaven.}

\fmtLemma{Kai\-ja\-na\-zo}
\fmtCitation{(o)Kai\-ja\-na\-zo}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{name given to a child born during a period of hardship}
\fmtExampleRL{OKai\-ja\-na\-zo anye\-te\-re.}
\fmtExampleEN{Kaijanazo has become a young adult.}

\fmtLemma{kai\-re}%
\fmtOne{1}
\fmtCitation{(a)kai\-re}
\fmtPOS{n}%
\fmtNC{12}
\fmtSubentry(a)
\fmtSense{time}
\fmtExampleRL{Akai\-re ka\-li ampi okwi\-ka aka\-kwa\-ba.}
\fmtExampleEN{It is almost time to go.}
\fmtSubentry(b)
\fmtSense{opportunity, chance}
\fmtExampleRL{Mpee\-rya akai\-re tu\-si\-moo\-le\-mu.}
\fmtExampleEN{Give me an opportunity and let's talk a bit please.}

\fmtLemma{kai\-re}%
\fmtOne{2}
\fmtCitation{\textnormal{(e.g.}\,mu\-kai\-re, ki\-kai\-re, ka\-kai\-re\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{old}
\fmtExampleRL{Emo\-to\-ka yaa\-mwe gi\-kai\-re.}
\fmtExampleEN{His car is old.}
\fmtSubentry(b)
\fmtSense{elderly, aged}
\fmtExampleRL{Omu\-ka\-li omu\-kai\-re ta\-saa\-na ku\-ko\-la mu\-no.}
\fmtExampleEN{It is not good for an elderly woman to work a lot.}

\fmtLemma{kai\-ri\-ka}
\fmtCitation{(oku)\-kai\-ri\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kai\-ri\-ki\-re}
\fmtSense{to grow old, to depreciate, to break down}
\fmtExampleRL{Olu\-go\-ye lwa\-nge lu\-kai\-ri\-ki\-re.}
\fmtExampleEN{My clothing has got old.}

\fmtLemma{kai\-ri\-ki}
\fmtCitation{\textnormal{(e.g.}\,mu\-kai\-ri\-ki, ba\-kai\-ri\-ki, lu\-kai\-ri\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{old, aged}
\fmtExampleRL{Ombwe\-ne waa\-mwe mu\-kai\-ri\-ki.}
\fmtExampleEN{His dog is old.}

\fmtLemma{kai\-ri\-kya}
\fmtCitation{(oku)\-kai\-ri\-kya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kai\-ri\-ki\-rye}
\fmtSense{(of harsh conditions) to make one look old}
\fmtExampleRL{Obu\-see\-ge bwo\-na bu\-kai\-ri\-kya omu\-ntu.}
\fmtExampleEN{Poverty also makes one look old.}

\fmtLemma{kai\-ri\-za}
\fmtCitation{(a)kai\-ri\-za}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{pubic hair}
\fmtExampleRL{Akoo\-jo ka\-nge ka\-mee\-re obwi\-ri\-za.}
\fmtExampleEN{My son has grown pubic hair.}

\fmtLemma{kai\-sa\-ba\-sa\-bai}
\fmtCitation{(a)kai\-sa\-ba\-sa\-bai}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{\textit{Mu\-sa × paradisiaca} banana (the hybrid between \textit{Musa acuminata} and \textit{Mu\-sa balbisiana})}

\fmtLemma{kai\-sa\-bi\-sa\-bi}
\fmtCitation{(o)kai\-sa\-ba\-sa\-bai}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{\textit{Aspa\-ra\-gus africanus}}

\fmtLemma{kai\-si}
\fmtPOS{interj}
\fmtSense{by the way}
\fmtExampleRL{Kai\-si nii\-we naa\-ni?}
\fmtExampleEN{By the way, what is your name?}

\fmtLemma{kai\-si\-te}
\fmtPOS{conj}
\fmtSense{yet, despite that, in spite of that, for all that, all the same}
\fmtExampleRL{Na\-ya\-ba oku\-mu\-swa\-ga\-na kai\-si\-te nga mu\-le\-ke\-re eka.}
\fmtExampleEN{I went to look for him, but I had left him at home.}

\fmtLemma{kai\-swe}
\fmtCrossreference{\textnormal{see} (-)aiswe}

\fmtLemma{kai\-ta}
\fmtPOS{num}
\fmtSense{billion}
\fmtExampleRL{Atu\-ngi\-re akai\-ta ka mpii\-ya.}
\fmtExampleEN{He has a billion shillings.}
\fmtSubentry(a)
\fmtSense{millet (grain)}
\fmtExampleRL{Akai\-ta ka\-me\-ra kwa ike\-ndi.}
\fmtExampleEN{Millet grows during the rainy season.}
\fmtSubentry(b)
\fmtSense{millet meal, millet paste}
\fmtExampleRL{Omu\-ge\-ni mu\-mye\-ne\-re kai\-ta.}
\fmtExampleEN{Make millet meal for the visitor.}

\fmtLemma{kai\-ta\-ba\-go\-le}
\fmtCitation{(a)kai\-ta\-ba\-go\-le}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{elbow}
\fmtExampleRL{Ya\-mu\-ko\-ne\-se\-rye kai\-ta\-ba\-go\-le.}
\fmtExampleEN{He knocked him with an elbow.}

\fmtLemma{kai\-tai\-ga\-na}
\fmtCitation{(a)kai\-tai\-ga\-na}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{bird trap}
\fmtExampleRL{Akai\-tai\-ga\-na kai\-ta enko\-wu bu\-lii\-zo.}
\fmtExampleEN{The bird trap kills a guinea fowl every day.}

\fmtLemma{kai\-zo\-lya\-nzo\-gi}
\fmtCitation{(e)kai\-zo\-lya\-nzo\-gi}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{\textit{Dracaena steudneri} (northern large-leaved dragon-tree, traditionally used to treat rheumatic pain)}

\fmtLemma{kai\-zu\-li}
\fmtCitation{(e)kai\-zu\-li}
\fmtPOS{n}%
\fmtNC{12}
\fmtSubentry(a)
\fmtSense{rear side of a house, place behind a house}
\fmtExampleRL{Oma wa\-nge ya\-si\-mbi\-re omu\-cu\-ngwa ekai\-zu\-li.}
\fmtExampleEN{My mother grew an orange tree behind the house.}
\fmtSubentry(b)
\fmtSense{toilet, washroom, bathroom, restroom, lavatory, WC}
\fmtExampleRL{Aya\-bi\-re omu kai\-zu\-li.}
\fmtExampleEN{He has gone to the toilet.}

\fmtLemma{ka\-ja\-ga\-la\-lo}
\fmtCitation{(a)ka\-ja\-ga\-la\-lo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{commotion, instability, disturbance, disorder, turmoil}
\fmtExampleRL{Abe\-mbi ba\-lee\-te\-re aka\-ja\-ga\-la\-lo omu ki\-bu\-ga.}
\fmtExampleEN{The musicians have caused a commotion in town.}

\fmtLemma{ka\-ja\-nja}
\fmtCitation{(a)ka\-ja\-nja}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{overfamiliarity, intrusiveness}
\fmtExampleRL{Ntoo\-la\-ku aka\-nja\-nja.}
\fmtExampleEN{Stop your intrusiveness. (lit.\,Take away your intrusiveness.)}

\fmtLemma{ka\-ja\-njaa\-lo}
\fmtCitation{(a)ka\-ja\-njaa\-lo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{clitoris}
\fmtExampleRL{Aba\-kya\-la ni\-bo ba\-bba n’a\-ka\-ja\-nja\-lo.}
\fmtExampleEN{Women have a clitoris.}

\fmtLemma{ka\-je\-ge\-re}
\fmtCitation{(a)ka\-je\-ge\-re}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{plait (of hair), braid, a length of hair made up of three or more interlaced strands}
\fmtExampleRL{Omu\-ka\-li abbwe\-re ka\-je\-ge\-re.}
\fmtExampleEN{The woman has made a plait.}

\fmtLemma{ka\-jo\-njo}
\fmtCitation{(a)ka\-jo\-njo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{rheum from eyes (mucus discharged from the eyes which gathers as a crust in the corners of the eyes), sleepy sand, eye goop, sleep dust}
\fmtExampleRL{Naa\-ba omu mai\-so oga\-too\-le\-mu obu\-jo\-njo.}
\fmtExampleEN{Wash your eyes and get rid of the sleepy sand.}

\fmtLemma{ka\-ju\-mbu\-ra}
\fmtCitation{(o)ka\-ju\-mbu\-ra}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{straggling or climbing plant covered with stinging hairs (\textit{Tragia benthamii})}
\fmtExampleRL{Oka\-ju\-mbu\-ra asii\-ya n’a\-bba nga aku\-ko\-ne\-re oku lu\-su\-su.}
\fmtExampleEN{\textit{Tragia benthamii} stings when it comes in contact with the skin.}

\fmtLemma{Ka\-ju\-ra}
\fmtCitation{(o)Ka\-ju\-ra}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{name given to a boy born when it is raining}
\fmtExampleRL{OKa\-ju\-ra ya\-bya\-lii\-bwe omu nju\-ra.}
\fmtExampleEN{Kajura was born when it is raining.}

\fmtLemma{ka\-juu\-la}
\fmtCitation{(a)ka\-juu\-la}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{sand fly}
\fmtExampleRL{Obu\-juu\-la bu\-ne\-na omu mu\-twe ng’o\-ku\-lu\-ma.}
\fmtExampleEN{Sand flies bite your head when you are digging.}

\fmtLemma{ka\-jwe\-jwe\-te}%
\fmtOne{1}
\fmtCitation{(o)ka\-jwe\-jwe\-te}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{concentrated banana juice (\eg juice made from beer banana)}
\fmtExampleRL{Ka\-jwe\-jwe\-te ba\-mu\-wa ba\-lwai\-re.}
\fmtExampleEN{Concentrated juice is given to sick people.}

\fmtLemma{Ka\-jwe\-jwe\-te}%
\fmtOne{2}
\fmtCitation{(o)Ka\-jwe\-jwe\-te}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{name given to the Bu\-nyala queen}

\fmtLemma{ka\-jwi\-syo}
\fmtCitation{(a)ka\-jwi\-so}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{abnormal uterine bleeding, bleeding between monthly periods, extremely heavy period}
\fmtExampleRL{Omu Bbai\-bbu\-li mu\-li\-mu omu\-ka\-li eyaa\-li omu\-lwai\-re wa ka\-jwi\-syo.}
\fmtExampleEN{In the Bible there was a woman who was suffering from haemorrhage.}

\fmtLemma{ka\-ka}
\fmtCitation{(oku)\-ka\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-ki\-re}
\fmtSense{to force, to make, to pressure}
\fmtExampleRL{Na\-mu\-ka\-ki\-re\-mbe okwi\-za.}
\fmtExampleEN{I just forced her to come.}

\fmtLemma{ka\-kaa\-ga}
\fmtCitation{(o)ka\-kaa\-ga}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSubentry(a)
\fmtSense{constellation of the Pleiades}
\fmtSubentry(b)
\fmtSense{six thousand}
\fmtExampleRL{Mu\-sa\-swi\-re empii\-ya ezi\-ku\-loo\-ma omu ka\-kaa\-ga.}
\fmtExampleEN{I have paid him almost six thousand shillings.}

\fmtLemma{ka\-kaa\-ki}
\fmtCitation{(a)ka\-kaa\-ki}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{microphone}
\fmtExampleRL{Ti\-tu\-li\-na ka\-kaa\-ki ka\-ku\-ba\-zi\-ra\-mu.}
\fmtExampleEN{We do not have a microphone to speak into.}

\fmtLemma{ka\-kaa\-lu\-ka\-na}
\fmtCitation{(oku)\-ka\-kaa\-lu\-ka\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-kaa\-lu\-kai\-ne}
\fmtSense{to work hard}
\fmtExampleRL{Omu\-ka\-li oyo aka\-kaa\-lu\-kai\-ne oku\-ku\-lya abaa\-na baa\-mu.}
\fmtExampleEN{That woman has worked hard to raise her children.}

\fmtLemma{ka\-kaa\-lu\-ka\-ni}
\fmtCitation{\textnormal{(e.g.}\,mu\-ka\-kaa\-lu\-ka\-ni\textnormal{)}}
\fmtPOS{adj}
\fmtSense{hard-working}
\fmtExampleRL{Mu\-kaa\-gwa wa\-nge mu\-ka\-li mu\-ka\-kaa\-lu\-ka\-ni.}
\fmtExampleEN{My sister-in-law is a hard-working woman.}

\fmtLemma{ka\-kaa\-nia}
\fmtCitation{(oku)\-ka\-kaa\-nia}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-kai\-nye}
\fmtSense{to kindle, to ignite, to lighten}
\fmtExampleRL{Ya\-ka\-kai\-ni\-rye omuu\-ro.}
\fmtExampleEN{She kindled the fire.}

\fmtLemma{ka\-kaa\-tii\-rya}
\fmtCitation{(oku)\-ka\-kaa\-tii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-kaa\-tii\-rye, ka\-kaa\-ti\-rii\-rye}
\fmtSense{to emphasise, to highlight, to stress, to underline}
\fmtExampleRL{Ka\-kaa\-tii\-rya enso\-nga egyo le\-ke ba\-gye\-te\-je.}
\fmtExampleEN{Emphasise that point so that people understand it properly.}

\fmtLemma{ka\-kaa\-ti\-ka}
\fmtCitation{(oku)\-ka\-kaa\-ti\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-kaa\-ti\-ki\-re}
\fmtSense{to impose, to enforce}
\fmtExampleRL{Lwa\-ki mu\-ku\-tu\-ka\-kaa\-ti\-ka\-ku bye tu\-ta\-kwe\-ndya?}
\fmtExampleEN{Why do you impose on us things that we do not want?}

\fmtLemma{ka\-ka\-bbu\-ka}
\fmtCitation{(oku)\-ka\-ka\-bbu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-ka\-bbu\-ki\-re}
\fmtSense{to flake off, to peel off}
\fmtExampleRL{Era\-ngi eku\-ka\-ka\-bbu\-ka.}
\fmtExampleEN{The paint is peeling off.}

\fmtLemma{ka\-ka\-bbu\-la}
\fmtCitation{(oku)\-ka\-ka\-bbu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-ka\-bbwi\-re}
\fmtSense{to tear off, to peel off}
\fmtExampleRL{Aka\-ka\-bbwi\-re\-ku eki\-su\-su oku mu\-saa\-le.}
\fmtExampleEN{She peeled off the bark of the tree.}

\fmtLemma{ka\-kai\-re}
\fmtPOS{num}
\fmtSense{one million}
\fmtExampleRL{Mmu\-wai\-re aka\-kai\-re ka\-mwei.}
\fmtExampleEN{I have given him one million.}

\fmtLemma{ka\-ka\-ka\-mpa}
\fmtCitation{(a)ka\-ka\-ka\-mpa}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{dried nasal mucus, booger}
\fmtExampleRL{Ya\-ba otoo\-le\-mu obu\-ka\-ka\-ka\-mpa omu nyi\-ndo.}
\fmtExampleEN{Go and remove the booger.}

\fmtLemma{ka\-ka\-ku\-la}
\fmtCitation{(oku)\-ka\-ka\-ku\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-kakwi\-re}
\fmtSense{to remove by force, to topple, to bring down}
\fmtExampleRL{Ega\-vu\-me\-nti ya Co\-ngo ba\-gi\-ka\-ka\-kwi\-re\-ku.}
\fmtExampleEN{The government of Congo was toppled.}

\fmtLemma{ka\-ka\-mba\-maa\-gi}
\fmtCrossreference{\textnormal{see also} kadonkiiro}
\fmtCitation{(a)ka\-ka\-mba\-maa\-gi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{type of fishing net used for trapping tilapia}

\fmtLemma{ka\-ka\-na}%
\fmtOne{1}
\fmtCitation{(o)ka\-ka\-na}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{pied kingfisher (\textit{Ceryle rudis})}
\fmtExampleRL{Oka\-ka\-na alya encwi.}
\fmtExampleEN{The kingfisher eats fish.}

\fmtLemma{ka\-ka\-na}%
\fmtOne{2}
\fmtCitation{(oku)\-ka\-ka\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-kai\-ne}
\fmtSense{to calm down, to stabilise, to normalise}
\fmtExampleRL{Ka\-ka\-na, ota\-ju\-mbu\-tu\-ki\-ra bi\-ga\-mbo.}
\fmtExampleEN{Calm down, don’t speak anyhow.}

\fmtLemma{ka\-ka\-nga}
\fmtCitation{(a)ka\-ka\-nga}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{(of a house) roof apex}
\fmtExampleRL{Aka\-ka\-nga ka nyu\-mba baa\-ka\-bwe\-re ku\-bbi.}
\fmtExampleEN{The roof apex was poorly constructed.}

\fmtLemma{ka\-ka\-nja}
\fmtCitation{(a)ka\-ka\-nja}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{full stop}
\fmtExampleRL{Mpa\-ndii\-ki\-re ebi\-ga\-mbo bi\-sa\-tu ni ntaa\-ku aka\-ka\-nja.}
\fmtExampleEN{I wrote three words and then put a full stop.}

\fmtLemma{ka\-ka\-nya\-la}
\fmtCitation{(oku)\-ka\-ka\-nya\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-ka\-nyai\-re}
\fmtSubentry(a)
\fmtSense{to be hard, to be stiff (\eg due to drying out)}
\fmtExampleRL{Ekya\-nzo ki\-ka\-ka\-nyai\-re.}
\fmtExampleEN{The hide has stiffened.}
\fmtSubentry(b)
\fmtSense{(of food) to harden, to become cold}
\fmtExampleRL{Eme\-re ya ma\-too\-ke eka\-ka\-nya\-la nga to\-gi\-swe\-ke\-re ku\-sai.}
\fmtExampleEN{If you don’t cover cooked matoke properly, it hardens.}
\fmtSubentry(c)
\fmtSense{to undergo rigor mortis}
\fmtSubentry(d)
\fmtSense{to be difficult, to become obstinate, to be persistent, to be rebellious}
\fmtExampleRL{Ki\-bbi omwa\-na oku\-ka\-ka\-nya\-la nga aba\-kai\-re baa\-mwe ba\-mu\-ko\-be\-re\-ku.}
\fmtExampleEN{It is bad for a child to be obstinate when his parents try to counsel him.}

\fmtLemma{ka\-ka\-nya\-li}
\fmtCitation{\textnormal{(e.g.}\,lu\-ka\-ka\-nya\-li, ka\-ka\-ka\-nya\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{(of clothes) stiff, dry}
\fmtExampleRL{Amai\-zi ga\-ni ga\-ko\-le\-se\-bwa oku\-lee\-ta obu\-ka\-ka\-nya\-li omu ngo\-ye.}
\fmtExampleEN{This water is used to make fabrics stiff.}
\fmtSubentry(b)
\fmtSense{(of food) cool, cold}
\fmtSubentry(c)
\fmtSense{obstinate, stubborn}
\fmtExampleRL{Omu\-sai\-za oyo mu\-ka\-ka\-nya\-li.}
\fmtExampleEN{That man is obstinate.}

\fmtLemma{ka\-ka\-ra}
\fmtCitation{(o)ka\-ka\-ra}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSubentry(a)
\fmtSense{black cow with brown stripes from head to tail}
\fmtSubentry(b)
\fmtSense{compulsory labour, forced labour}
\fmtExampleRL{Ti\-nko\-la mu\-lu\-mo gwa ka\-ka\-ra.}
\fmtExampleEN{I don’t do forced labour.}

\fmtLemma{ka\-ka\-ra\-ka\-ka\-ra}
\fmtPOS{adv}
\fmtSense{forcibly, using force, involuntary}
\fmtExampleRL{Oyo atu\-ko\-le\-sya ebya ka\-ka\-ra\-ka\-ka\-ra.}
\fmtExampleEN{That one forces us to do things.}

\fmtLemma{ka\-ka\-sa}
\fmtCitation{(oku)\-ka\-ka\-sa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-ka\-si\-re}
\fmtSense{to confirm, to ascertain, to assure}
\fmtExampleRL{Nyi\-zi\-re ani ku\-ka\-ka\-sa oba nga wai\-za oku mba\-ga.}
\fmtExampleEN{I have come here to confirm whether you will attend the wedding.}

\fmtLemma{ka\-ka\-sya}
\fmtCitation{(oku)\-ka\-ka\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-ka\-si\-rye}
\fmtSense{to brief, to inform, to update, to notify}
\fmtExampleRL{Omu\-ku\-lu wa iso\-me\-ro yai\-zi\-re oku\-ka\-sya abaa\-na.}
\fmtExampleEN{The head teacher came to inform the children.}

\fmtLemma{ka\-ka\-ta}
\fmtCitation{(oku)\-ka\-ka\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-ka\-ti\-re}
\fmtSense{to be certain, to be sure, to be confirmed, to become assured, to be self-reliant}
\fmtExampleRL{Omwa\-na aka\-ka\-ti\-re omu ku\-te\-ga encwi.}
\fmtExampleEN{The child proved to be good at fishing.}

\fmtLemma{ka\-ka\-ti}
\fmtCitation{\textnormal{(e.g.}\,mu\-ka\-ka\-ti, lu\-ka\-ka\-ti, ba\-ka\-ka\-ti\textnormal{)}}
\fmtPOS{adj}
\fmtSense{firm, strong}
\fmtExampleRL{Olu\-ju\-ngu lu\-ka\-ka\-ti omu Uga\-nda.}
\fmtExampleEN{English is strong in Uganda.}

\fmtLemma{ka\-ka\-tya}
\fmtCitation{(oku)\-ka\-ka\-tya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-ka\-ti\-rye}
\fmtSubentry(a)
\fmtSense{to fulfil, to accomplish}
\fmtExampleRL{Ka\-ka\-tya bye wa\-suu\-bii\-ze.}
\fmtExampleEN{Fulfil your promise.}
\fmtSubentry(b)
\fmtSense{to reinforce, to emphasise, to uphold}
\fmtExampleRL{Ka\-ka\-tya ebbi\-zi\-nee\-si yaa\-mu ng’o\-ku\-fu\-na eroo\-ni omu bba\-nka.}
\fmtExampleEN{Reinforce your business with a loan from the bank.}
\fmtSubentry(c)
\fmtSense{to prove, to demonstrate}
\fmtExampleRL{Ka\-ka\-tya nti oli mu\-sai\-za.}
\fmtExampleEN{Prove that you are a man.}
\fmtSubentry(d)
\fmtSense{to make certain, to make sure}
\fmtExampleRL{Ka\-ka\-tya emi\-so\-mo gya\-mu.}
\fmtExampleEN{Be sure that you are studying well.}

\fmtLemma{ka\-ka\-wee\-si}
\fmtCitation{(a)ka\-ka\-wee\-si}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{cricket (an insect)}
\fmtExampleRL{Aka\-ka\-wee\-si ka\-ku\-ko\-koo\-ma.}
\fmtExampleEN{The cricket is making a lot of noise.}

\fmtLemma{ka\-kee\-do}
\fmtCitation{(a)ka\-kee\-do}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{fibre from banana petiole (stalk) for making baskets}
\fmtExampleRL{Obu\-kee\-do tu\-bu\-lu\-ki\-sya kii\-bbo.}
\fmtExampleEN{We weave the basket using fibre from banana petioles.}

\fmtLemma{ka\-kee\-kee\-yu}
\fmtCitation{(a)ka\-kee\-kee\-yu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{drill}
\fmtExampleRL{Ko\-le\-sya oke\-kee\-yu ofu\-mu\-le eki\-wu\-lu omu lu\-baa\-wo.}
\fmtExampleEN{Use the drill and make a hole in the timber.}

\fmtLemma{ka\-kee\-si}
\fmtCitation{(a)ka\-kee\-si}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{small black cricket}
\fmtExampleRL{Aka\-kee\-si ka\-ku\-ku\-ngi\-ra omu nyu\-mba.}
\fmtExampleEN{The small black cricket is chirping in the house.}

\fmtLemma{ka\-ke\-kee\-yo}
\fmtCitation{(a)ka\-ke\-kee\-yo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{sieve, strainer}
\fmtExampleRL{Kwa\-ta aka\-ke\-kee\-yo oku\-nguu\-te obwi\-ta.}
\fmtExampleEN{Use the sieve and sift the millet.}

\fmtLemma{ka\-ke\-re\-nge}%
\fmtOne{1}
\fmtCitation{(o)ka\-ke\-re\-nge}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{greyish calf}
\fmtExampleRL{Onya\-na wa\-nge oka\-ke\-re\-nge asaa\-re.}
\fmtExampleEN{My greyish calf is in heat.}

\fmtLemma{ka\-ke\-re\-nge}%
\fmtOne{2}
\fmtCitation{(a)ka\-ke\-re\-nge}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{goat dropping}
\fmtExampleRL{Obu\-ke\-re\-nge bwi\-zwi\-re omu lu\-bu\-ga.}
\fmtExampleEN{The goat droppings are all over the compound.}

\fmtLemma{ka\-kii\-ko}
\fmtCitation{(a)ka\-kii\-ko}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{commission of inquiry, committee}
\fmtExampleRL{Opu\-re\-zi\-de\-nti atai\-re\-wo aka\-kii\-ko ka\-too\-lee\-rye oku byai\-ta\-ka\-li.}
\fmtExampleEN{The president has instituted a commission of land inquiry.}

\fmtLemma{ka\-kii\-nii\-ko}%
\fmtOne{1}
\fmtCitation{(a)ka\-kii\-nii\-ko}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{food crumb}
\fmtExampleRL{Obu\-kii\-nii\-ko bu\-lii\-bwa nko\-ko.}
\fmtExampleEN{Food crumbs are eaten by chicken.}
\fmtSubentry(b)
\fmtSense{food dregs}
\fmtExampleRL{Obu\-kii\-nii\-ko onja\-ngo abu\-lii\-re.}
\fmtExampleEN{The cat has eaten the food dregs.}

\fmtLemma{ka\-kii\-nii\-ko}%
\fmtOne{2}
\fmtCitation{(a)ka\-kii\-nii\-ko}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{habit of mocking others}
\fmtExampleRL{Omu\-ka\-li oyo ata n’a\-ka\-kii\-nii\-ko.}
\fmtExampleEN{That woman has a habit of mocking others.}

\fmtLemma{ka\-ki\-ndo}
\fmtCitation{(a)ka\-ki\-ndo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{crease (a line on cloth produced \eg by ironing)}
\fmtExampleRL{Ago\-loi\-re esaa\-ti n’ei\-za\-mu obu\-ki\-ndo.}
\fmtExampleEN{He has produced a crease on the shirt by ironing.}

\fmtLemma{ka\-ki\-ri\-ngi}
\fmtCitation{(a)ka\-ki\-ri\-ngi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{goat droppings}
\fmtExampleRL{Bu\-ni bu\-kwi\-sa\-na bu\-ki\-ri\-ngi bwa mbu\-li.}
\fmtExampleEN{These resemble goat droppings.}

\fmtLemma{ka\-ki\-sa}
\fmtCitation{(a)ka\-ki\-sa}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{chance, opportunity}
\fmtExampleRL{Ko\-le\-sya ku\-sai aka\-ki\-sa ko\-fu\-ni\-re.}
\fmtExampleEN{Make the best use of the opportunity you have gotten.}

\fmtLemma{ka\-ki\-zi}
\fmtCitation{(a)ka\-ki\-zi}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{infatuation, short-lived passion}
\fmtExampleRL{Aba\-sai\-za ba\-ta n’a\-ka\-ki\-zi.}
\fmtExampleEN{Men have short-lived passion.}

\fmtLemma{ka\-ko\-be}
\fmtCitation{(o)ka\-ko\-be}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{purple}
\fmtExampleRL{Esaa\-ti yaa\-mwe ya ka\-ko\-be.}
\fmtExampleEN{His shirt is purple.}

\fmtLemma{ka\-ko\-bo\-li}
\fmtCitation{(a)ka\-ko\-bo\-li}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{tannia (\textit{Xanthosoma sagittifolium})}
\fmtExampleRL{Obu\-ko\-bo\-li bu\-woo\-ma mu\-no.}
\fmtExampleEN{Tannia tastes good.}

\fmtLemma{ka\-ko\-dyo}
\fmtCitation{(a)ka\-ko\-dyo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{trick, skill, technique}
\fmtExampleRL{Ako ka\-ko\-dyo ka\-yaa\-ka.}
\fmtExampleEN{That is a new trick.}

\fmtLemma{ka\-ko\-ko}
\fmtCitation{(a)ka\-ko\-ko}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{chick}
\fmtExampleRL{Aka\-ko\-ko ka\-gwi\-re omu kii\-na.}
\fmtExampleEN{The chick has fallen into a pit.}

\fmtLemma{ka\-ko\-ko\-la}
\fmtCitation{(a)ka\-ko\-ko\-la}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{elbow}
\fmtExampleRL{Aka\-ko\-ko\-la ka\-ndu\-ma.}
\fmtExampleEN{My elbow hurts.}

\fmtLemma{ka\-ko\-lwa}
\fmtCrossreference{\textnormal{see also} wai\-tuu\-ru}
\fmtCitation{(a)ka\-ko\-lwa}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{banded mongoose}
\fmtExampleRL{Obu\-ko\-lwa bu\-lya mu\-no enjo\-ka.}
\fmtExampleEN{The mongoose eat snakes a lot.}

\fmtLemma{ka\-ko\-ma}
\fmtCitation{(a)ka\-ko\-ma}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{way of wrapping a dead body tightly}
\fmtExampleRL{Oru\-fa ba\-mu\-bbwe\-re ka\-ko\-ma.}
\fmtExampleEN{The body was wrapped tightly.}

\fmtLemma{ka\-ko\-mo}
\fmtCitation{(a)ka\-ko\-mo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{bracelet}
\fmtExampleRL{Aba\-ka\-li be\-ndya mu\-no oku\-zwa\-la obu\-ko\-mo.}
\fmtExampleEN{Women like wearing bracelets a lot.}
\fmtSubentry(b)
\fmtSense{way of wrapping a dead body tightly}
\fmtExampleRL{Oru\-fa ba\-mu\-bbwe\-re ka\-ko\-mo.}
\fmtExampleEN{The body was tied tightly.}

\fmtLemma{ka\-ko\-ndo}%
\fmtOne{1}
\fmtCitation{(a)ka\-ko\-ndo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{latch}
\fmtExampleRL{Ai\-gai\-re\-wo olwi\-ge n’a\-taa\-ku n’a\-ka\-ko\-ndo.}
\fmtExampleEN{He locked the door with a latch.}
\fmtSubentry(b)
\fmtSense{belt loop}
\fmtExampleRL{Ako\-ko\-ndo ka mpa\-le yaa\-mwe ka\-zwi\-re\-ku.}
\fmtExampleEN{His belt loop is torn off.}
\fmtSubentry(c)
\fmtSense{high-heeled shoe, high-heel}
\fmtExampleRL{Azwa\-la nkai\-to za ka\-ko\-ndo.}
\fmtExampleEN{She wears high-heeled shoes.}

\fmtLemma{ka\-ko\-ndo}%
\fmtOne{2}
\fmtCitation{(o)ka\-ko\-ndo}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{calf with horns pointing or bending forwards}
\fmtExampleRL{Onya\-na wa\-nge oka\-ko\-ndo abuu\-re.}
\fmtExampleEN{My calf with horns that point forward is lost.}

\fmtLemma{ka\-ko\-ngoi\-zo\-le}
\fmtCitation{(a)ka\-ko\-ngoi\-zo\-le}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{ankle}
\fmtExampleRL{Amu\-ku\-bbi\-re ei\-bbaa\-le n’a\-mwa\-tya aka\-ko\-ngoi\-zo\-le.}
\fmtExampleEN{He hit him with a stone and broke his ankle.}

\fmtLemma{ka\-ko\-nko\-lo}
\fmtCitation{(a)ka\-ko\-nko\-lo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{wooden chair (a log traditionally used by the head of the family as a seat)}
\fmtExampleRL{Oi\-se\-ma\-ka ai\-kai\-re oku ka\-ko\-nko\-lo.}
\fmtExampleEN{The head of the family is sitting in a wooden chair.}

\fmtLemma{ka\-ko\-nwa}
\fmtCitation{(a)ka\-ko\-nwa}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{half-cooked matoke}
\fmtExampleRL{Ama\-too\-ke tu\-lii\-re ka\-ko\-nwa.}
\fmtExampleEN{We ate half-cooked matoke.}

\fmtLemma{ka\-koo\-lo}
\fmtCitation{(a)ka\-koo\-lo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{cough}
\fmtExampleRL{Abaa\-na baa\-mwe boo\-na\-boo\-na ba\-lwai\-re ba ka\-koo\-lo.}
\fmtExampleEN{All her children are suffering from a cough.}

\fmtLemma{ka\-koo\-we\-koo\-we}
\fmtCitation{(a)ka\-koo\-we\-koo\-we}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{eye lash}
\fmtExampleRL{Omu\-kai\-re odi ata n’o\-bu\-koo\-we\-koo\-we bwi\-ngi.}
\fmtExampleEN{That old man has thick eyelashes.}

\fmtLemma{ka\-ko\-pa}
\fmtCitation{(a)ka\-ko\-pa}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{small type of yam}
\fmtExampleRL{Ti\-nku\-ta\-ka ku\-lya bu\-ko\-pa obwo ku\-bba bu\-ku\-gu\-ma.}
\fmtExampleEN{I don’t want to eat those yams because they are very hard.}

\fmtLemma{ka\-ko\-si}
\fmtCitation{(a)ka\-ko\-si}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{sisal thread}
\fmtExampleRL{Obu\-ko\-si mbu\-ka\-mba\-mu omu\-gu\-wa.}
\fmtExampleEN{I am making a rope from sisal threads.}

\fmtLemma{ka\-ku\-ba\-mpa\-nga}
\fmtCitation{(o)ka\-ku\-ba\-mpa\-nga}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{hawk}
\fmtExampleRL{Oka\-ku\-ba\-mpa\-nga atee\-ra alya obu\-nyo\-nyi bwi\-raa\-mwe.}
\fmtExampleEN{The hawk eats its fellow birds.}

\fmtLemma{ka\-ku\-bba\-mai\-zi}
\fmtCitation{(a)ka\-ku\-bba\-mai\-zi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{the extreme part of a chicken wing composed of the metacarpals and the phalanges}
\fmtExampleRL{Nta\-ka mu\-no oku\-lya aka\-ku\-bba\-mai\-zi.}
\fmtExampleEN{I love eating chicken wings.}

\fmtLemma{ka\-ku\-bba\-nsi\-ri}
\fmtCitation{(o)ka\-ku\-bba\-nsi\-ri}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{basil (\textit{Ocimum basilicum})}
\fmtExampleRL{Oka\-ku\-bba\-nsi\-ri ata\-mba endwai\-re zi\-ngi mu\-no.}
\fmtExampleEN{Basil cures many diseases.}

\fmtLemma{ka\-ku\-du}
\fmtCitation{(o)ka\-ku\-du}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSubentry(a)
\fmtSense{beetle}
\fmtExampleRL{Oka\-ka\-ku\-du abu\-mba ebi\-sai n’a\-bi\-yi\-ri\-ngi\-tya.}
\fmtExampleEN{A beetle makes small balls from dung, then it rolls them around.}
\fmtSubentry(b)
\fmtSense{Volkswagen cars}
\fmtExampleRL{Obu\-ka\-ku\-du bu\-bba bu\-mo\-to\-ka bu\-gu\-mu.}
\fmtExampleEN{Volkswagen cars are strong.}

\fmtLemma{ka\-ku\-ju}
\fmtCitation{(a)ka\-ku\-ju}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{sliced dried young pumpkin}
\fmtExampleRL{Obu\-ku\-ju ba\-bu\-ko\-la\-mu omu\-ku\-bi.}
\fmtExampleEN{They make sauce from sliced dried young pumpkin.}

\fmtLemma{ka\-ku\-ku}
\fmtCitation{(a)ka\-ku\-ku}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{grudge, resentment, bitterness}
\fmtExampleRL{OPa\-tri\-ck ata n’a\-ka\-ku\-ku oku Ja\-ne.}
\fmtExampleEN{Patrick holds a grudge against Jane.}

\fmtLemma{ka\-ku\-la}
\fmtCitation{(oku)\-ka\-ku\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-kwi\-re}
\fmtSubentry(a)
\fmtSense{to scratch, to scrape}
\fmtExampleRL{Ota\-ka\-ku\-la ki\-rwai\-re ki\-kwi\-za oku\-jwa omu\-saa\-yi.}
\fmtExampleEN{You shouldn't scratch your wound, it will start bleeding.}
\fmtSubentry(b)
\fmtSense{(of animal) to dig (a hole), to burrow, to tunnel}
\fmtExampleRL{Ome\-se aku\-ka\-ku\-la ekii\-na omu nyu\-mba.}
\fmtExampleEN{The rat is digging a hole in a house.}

\fmtLemma{ka\-ku\-le}
\fmtCitation{(a)ka\-ku\-le}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{loud high-pitched laughter, a burst of laughter (often scornful in character)}
\fmtExampleRL{Aba\-ka\-li ba\-ku\-bbi\-re aka\-le\-ka omwa\-na n’a\-yi\-mu\-ka.}
\fmtExampleEN{The women laughed with a loud high-pitched laughter which woke the child up.}

\fmtLemma{ka\-ku\-lo}
\fmtCrossreference{\textnormal{see also} kakulwe}
\fmtCitation{(a)ka\-ku\-lo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{tadpole}
\fmtExampleRL{Obu\-ku\-lo bu\-bba mu mai\-zi.}
\fmtExampleEN{Tadpoles live in water.}

\fmtLemma{ka\-ku\-lwe}
\fmtCrossreference{\textnormal{see also} kakulo}
\fmtCitation{(a)ka\-ku\-lwe}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{tadpole}
\fmtExampleRL{Obu\-ku\-lwe bu\-bba mu mai\-zi.}
\fmtExampleEN{Tadpoles live in water.}

\fmtLemma{ka\-ku\-nguu\-ta}
\fmtCitation{(a)ka\-ku\-nguu\-ta}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{sieve, strainer}

\fmtLemma{ka\-ku\-nku\-mu\-ka}
\fmtCitation{(a)ka\-ku\-nku\-mu\-ka}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{food crumb, food dregs}
\fmtExampleRL{Oka\-pa alii\-re obu\-ku\-nku\-mu\-ka bwo\-na.}
\fmtExampleEN{The cat has eaten all the food crumbs.}

\fmtLemma{ka\-ku\-nku\-mu\-kwai}
\fmtCrossreference{\textnormal{see also} kakunkumuka}
\fmtCitation{(a)ka\-ku\-nku\-mu\-kwai}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{food crumb, food dregs}

\fmtLemma{ka\-kwa\-nso\-kwa\-nso}
\fmtCitation{(a)ka\-kwa\-nso\-kwa\-nso}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{\textit{Rhus natalensis}}
\fmtExampleRL{Abaa\-na ba\-ta\-ka mu\-no oku\-lya obu\-kwa\-so\-kwa\-so.}
\fmtExampleEN{Children like to eat the fruit of \textit{Rhus natalensis}.}

\fmtLemma{ka\-kwe\-so}
\fmtCitation{(a)ka\-kwe\-so}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{embroidery on gomesi}
\fmtExampleRL{We\-bbu\-we ego\-me\-si ote\-ree\-ze aka\-kwe\-so.}
\fmtExampleEN{Wear a gomesi and arrange for embroidery on it.}

\fmtLemma{ka\-kya\-ru\-mwe}
\fmtPOS{adv}
\fmtSense{totally, completely, seriously, at all}
\fmtExampleRL{Ye\-be\-rei\-re ka\-kya\-ru\-mwe okwi\-ra oku\-syo\-ma ebi\-wa\-ndii\-ko bya\-mwe.}
\fmtExampleEN{He completely forgot to come back.}

\fmtLemma{ka\-la}
\fmtCitation{(oku)\-ka\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kaa\-re}
\fmtSense{to dry}
\fmtExampleRL{Obwi\-ta bu\-ka\-la na isa\-na.}
\fmtExampleEN{Millet is dried by the sunshine.}

\fmtLemma{ka\-laa\-mu}
\fmtCitation{(e)ka\-laa\-mu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Ar. via Sw.}
\fmtSubentry(a)
\fmtSense{pencil, pen}
\fmtExampleRL{So\-ngo\-lya eka\-laa\-mu ya\-nge!}
\fmtExampleEN{Sharpen my pencil!}
\fmtSubentry(b)
\fmtSense{pencil}
\fmtExampleRL{Abaa\-na aba\-to ba\-wa\-ndii\-ki\-sya ka\-laa\-mu gi\-ka\-lu.}
\fmtExampleEN{Pupils use pencils to write.}
\fmtSubentry(c)
\fmtSense{eye pencil}
\fmtExampleRL{Eka\-laa\-mu y’a\-mai\-so ba\-gi\-sii\-ga ku bi\-ti\-gi.}
\fmtExampleEN{The eye pencil is used to define the eyebrow.}
\fmtSubentry(d)
\fmtSense{pen}
\fmtExampleRL{Abaa\-na ba\-bi\-bii\-na byai\-gu\-ru ba\-wa\-ndii\-ki\-sya ka\-laa\-mu ya bwi\-no.}
\fmtExampleEN{Pupils in the upper classes use a pen to write.}

\fmtLemma{ka\-laa\-yi}
\fmtCitation{(e)ka\-laa\-yi}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{metal basin}
\fmtExampleRL{Ei\-rai twa\-naa\-bi\-ra\-nga mu ka\-laa\-yi.}
\fmtExampleEN{Long ago we used to bathe in a metal basin.}

\fmtLemma{ka\-la\-ba\-ndu\-ka}
\fmtCitation{(oku)\-ka\-la\-ba\-ndu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-la\-ba\-ndu\-ki\-re}
\fmtSense{(of babies) to crawl on one's hands and legs but without using the knees}
\fmtExampleRL{Omwa\-na aku\-ka\-la\-ba\-ndu\-nka.}
\fmtExampleEN{A baby is crawling on its hands and feet.}

\fmtLemma{ka\-la\-bu\-la}
\fmtCitation{(oku)\-ka\-la\-bu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-la\-bwi\-re}
\fmtSense{to scratch, to scrape, to scar}
\fmtExampleRL{Aka\-la\-bwi\-re eko\-mpyu\-ta ya\-nge ati te\-ku\-bo\-ne\-ka ku\-sai.}
\fmtExampleEN{He scratched my laptop, it looks ugly now.}

\fmtLemma{ka\-lai\-ka\-lai}
\fmtCitation{(o)ka\-lai\-ka\-lai}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{tall and slender person}
\fmtExampleRL{Akaa\-ra ka Koo\-too\-tyo ka\-lai\-ka\-lai.}
\fmtExampleEN{Kootootyo’s daughter is tall and slender.}

\fmtLemma{ka\-la\-ka\-ta}
\fmtCitation{(oku)\-ka\-la\-ka\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-la\-ka\-ti\-re}
\fmtSense{to scrape, to scratch}
\fmtExampleRL{ONgo\-yo aku\-ka\-la\-ka\-ta eki\-na\-ga.}
\fmtExampleEN{Ngoyo is scraping out the cooking pan.}

\fmtLemma{ka\-la\-la}%
\fmtOne{1}
\fmtCitation{(a)ka\-la\-la}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{piece, slice, chunk, wedge}
\fmtExampleRL{Mpe\-rya\-ku aka\-la\-la ka mwo\-go.}
\fmtExampleEN{Give me a piece of cassava.}

\fmtLemma{ka\-la\-la}%
\fmtOne{2}
\fmtCitation{(oku)\-ka\-la\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-lai\-re}
\fmtSense{to argue constantly}
\fmtExampleRL{Lwa\-ki oka\-la\-lya otyo?}
\fmtExampleEN{Why are you constantly arguing?}

\fmtLemma{ka\-la\-li}
\fmtCitation{(o)ka\-la\-li}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{hot chilli, pepper}
\fmtExampleRL{Enyo\-nyi zi\-rya mu\-no oka\-la\-li.}
\fmtExampleEN{Birds eat a lot of hot chillies.}

\fmtLemma{ka\-la\-li\-ke}
\fmtCitation{(o)ka\-la\-li\-ke}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{\textit{Gre\-wia da\-mine} (species of flowering plant)}

\fmtLemma{ka\-la\-mba\-ndu\-ka}
\fmtCitation{(oku)\-ka\-la\-mba\-ndu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-la\-mba\-ndu\-ki\-re}
\fmtSense{to do secretly, to do surreptitiously, to do stealthily}
\fmtExampleRL{Ba\-mwa\-jii\-rye nga aku\-ka\-la\-mba\-ndu\-ka n’o\-ru\-fa.}
\fmtExampleEN{He was found stealthily carrying a dead body.}

\fmtLemma{ka\-la\-mbi\-rya}
\fmtCitation{(oku)\-ka\-la\-mbi\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-la\-mbi\-rii\-rye}
\fmtSense{to be obstinate, to stubborn, to insistent}
\fmtExampleRL{ONya\-nge\-ndo aka\-la\-mbi\-rya mu\-no empa\-ka.}
\fmtExampleEN{Nyangendo is very obstinate.}

\fmtLemma{ka\-la\-ngu\-ka}
\fmtCitation{(oku)\-ka\-la\-ngu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-la\-ngu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to stay long, to last}
\fmtExampleRL{ODe\-bu\-la aka\-la\-ngu\-kii\-re omu bu\-fu\-mbo.}
\fmtExampleEN{Deborah has stayed long in her marriage.}
\fmtSubentry(b)
\fmtSense{to become overdried, to overdry}
\fmtExampleRL{Ebi\-go\-go bi\-ka\-la\-ngu\-ki\-re.}
\fmtExampleEN{The banana fibres have overdried.}

\fmtLemma{ka\-la\-so}
\fmtCitation{(a)ka\-la\-so}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{arrow}
\fmtExampleRL{Omwi\-bbi ba\-mu\-ku\-bbi\-re aka\-la\-so oku ku\-gu\-lu.}
\fmtExampleEN{They shot the thief in the leg with an arrow.}

\fmtLemma{ka\-la\-ta}
\fmtCitation{(o)ka\-la\-ta}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{card game}
\fmtExampleRL{Abaa\-na aba\-to ba\-bai\-bbi\-ra mu\-no omu ka\-la\-ta.}
\fmtExampleEN{Children are often cheated in the kalata card game.}

\fmtLemma{ka\-le}
\fmtPOS{interj}
\fmtSense{alright, ok}
\fmtExampleRL{Ka\-le, ka nyi\-ze.}
\fmtExampleEN{Alright, let me come.}

\fmtLemma{ka\-le\-bwe\-le\-bwe}
\fmtCitation{(a)ka\-le\-bwe\-le\-bwe}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{edge, margin, side}
\fmtExampleRL{Toi\-kaa\-ra ku ka\-le\-bwe\-le\-bwe ka bwa\-to.}
\fmtExampleEN{Don’t sit on the edge of the boat.}

\fmtLemma{ka\-lee\-ga}
\fmtCitation{(a)ka\-lee\-ga}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{fastener}
\fmtExampleRL{Aka\-lee\-ga ka gaa\-li ya\-nge ka\-gwi\-re.}
\fmtExampleEN{The fastener for my bicycle is lost.}
\fmtSubentry(b)
\fmtSense{bra, brassie\-re}
\fmtExampleRL{Nfu\-ndi\-rya aka\-lee\-ga ako.}
\fmtExampleEN{Make that bra smaller for me.}

\fmtLemma{ka\-le\-ge\-te}
\fmtCitation{(a)ka\-le\-ge\-te}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{sachet, pack (usually wrapped with banana fibres)}
\fmtExampleRL{Mbwe\-re obu\-le\-ge\-te bu\-sa\-tu obwa mwa\-ni.}
\fmtExampleEN{I have wrapped three sachets of coffee beans.}

\fmtLemma{ka\-le\-ge\-ya}
\fmtCitation{(a)ka\-le\-ge\-ya}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{(of maize) anther}
\fmtExampleRL{Oduu\-ma alee\-te\-re obu\-le\-ge\-ya.}
\fmtExampleEN{The maize got anther.}
\fmtSubentry(b)
\fmtSense{wattle (a fleshy protuberance hanging on either side of the throat of goat)}
\fmtExampleRL{Ombu\-li wa\-nge ali\-na obu\-le\-ge\-ya.}
\fmtExampleEN{My goat has wattles.}

\fmtLemma{ka\-lei\-tu\-ko}
\fmtCitation{(a)ka\-lei\-tu\-ko}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{the habit of wandering, roaming}
\fmtExampleRL{Omu\-ka\-li oka\-lei\-tu\-ko ta\-yo\-roo\-ra baa\-na ku\-sai.}
\fmtExampleEN{A woman with the habit of roaming cannot raise children properly.}

\fmtLemma{ka\-le\-ju}
\fmtCitation{(a)ka\-le\-ju}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{chin}
\fmtExampleRL{Endoo\-rwa\-mu ya mo\-to\-ka ya\-mu\-saa\-re\-ku aka\-le\-ju.}
\fmtExampleEN{The mirror of the car cut off his chin.}

\fmtLemma{ka\-le\-ka}
\fmtCitation{(a)ka\-le\-ka}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{loud high-pitched laughter}
\fmtExampleRL{Aba\-ka\-li ba\-ku\-bbi\-re aka\-le\-ka omwa\-na n’a\-yi\-mu\-ka.}
\fmtExampleEN{The women laughed with a loud high-pitched laughter which woke the child up.}
\fmtSubentry(b)
\fmtSense{sharp whistle (produced with a palm)}
\fmtExampleRL{Ku\-bba aka\-le\-ka aba\-lii\-sya ba\-ma\-nye gye tu\-li.}
\fmtExampleEN{Blow a high-pitched whistle to let the cattle keepers know where we are.}

\fmtLemma{ka\-le\-ka\-le}
\fmtPOS{adv}
\fmtSense{a little better, slightly better, somewhat better}
\fmtExampleRL{Omu\-lwai\-re ali ka\-le\-ka\-le.}
\fmtExampleEN{The patient is a little better.}

\fmtLemma{ka\-le\-ka\-le\-ku}
\fmtPOS{adv}
\fmtSense{a little better, slightly better, somewhat better}
\fmtExampleRL{Omu\-lwai\-re na\-mu\-le\-ke\-re ali ka\-le\-ka\-le.}
\fmtExampleEN{I left when the patient was feeling a little better.}

\fmtLemma{Ka\-le\-kwa}
\fmtCitation{(o)Ka\-le\-kwa}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{name given to a girl born when her parent is dead}
\fmtExampleRL{Oka\-le\-kwa akuu\-re.}
\fmtExampleEN{Kalekwa has grown up.}

\fmtLemma{ka\-le\-ma\-ngu\-ndu}
\fmtCitation{(a)ka\-le\-ma\-ngu\-ndu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{\textit{Garde\-nia thun\-bergia} (white gardenia, lit.\,one-which-defeats-the-elephant)}
\fmtExampleRL{Aka\-le\-ma\-ngu\-ndu ka\-ta\-mba endwai\-re zi\-ngi.}
\fmtExampleEN{Gardenia thunbergia treats many diseases.}

\fmtLemma{ka\-le\-mba}
\fmtCitation{(o)ka\-le\-mba}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{name of a calf with black and white-dotted patches on the back}
\fmtExampleRL{Na\-mu\-gu\-lii\-re onte ka\-le\-mba.}
\fmtExampleEN{I bought him a calf with black and white-dotted patches on its back.}

\fmtLemma{ka\-le\-nda}
\fmtCitation{(e)ka\-le\-nda}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{calendar}
\fmtExampleRL{Mba\-ga\-bii\-re eka\-le\-nda ya mwa\-ka ogu\-kwi\-za.}
\fmtExampleEN{I gave them the calendar for the coming year.}

\fmtLemma{ka\-le\-nge}
\fmtCitation{(o)ka\-le\-nge}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{mist}
\fmtExampleRL{Oka\-le\-nge ala\-nga isa\-na.}
\fmtExampleEN{The mist signals sunshine.}

\fmtLemma{ka\-le\-ni}
\fmtPOS{conj}
\fmtSense{hence, therefore, for that matter}
\fmtExampleRL{Ka\-le\-ni, ka tu\-be\-ndee\-re ebyai\-rii\-ra.}
\fmtExampleEN{Therefore let us wait for what will happen next.}

\fmtLemma{ka\-le\-ra}
\fmtCitation{(a)ka\-le\-ra}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{u\-mbilical cord}
\fmtExampleRL{Mpee\-rya aka\-le\-ra ka mwa\-na wa\-nge tu\-ko\-le emi\-ko\-lo gya kwa\-ru\-la.}
\fmtExampleEN{Give me my child's u\-mbilical cord and we will carry out the initiation ceremony for her.}

\fmtLemma{ka\-le\-zu}
\fmtCrossreference{\textnormal{see} kaleju}

\fmtLemma{ka\-li}
\fmtCitation{\textnormal{(e.g.}\,ga\-ka\-li, bi\-ka\-li, zi\-ka\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{all (in a phrase with a numeral, \eg all the four)}
\fmtExampleRL{Emo\-to\-ka ezo zi\-ka\-li inai za\-nge.}
\fmtExampleEN{All four of those cars are mine.}
\fmtSubentry(b)
\fmtSense{both}
\fmtExampleRL{Ya\-kwai\-te omu\-gu\-wa n’e\-mi\-ko\-no gi\-ka\-li ibi\-ri.}
\fmtExampleEN{She grabbed the rope with both hands.}
\fmtSubentry(c)
\fmtSense{as one is, in the original state}
\fmtExampleRL{To\-ba\-li\-nda ku\-zwa\-la ba\-lee\-te nka ni ba\-ka\-li.}
\fmtExampleEN{Do not wait for them to get dressed, bring them as they are.}

\fmtLemma{ka\-li\-cu\-lee\-ta}
\fmtCitation{(e)ka\-li\-cu\-lee\-ta}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{calculator}
\fmtExampleRL{Eka\-li\-cu\-lee\-ta emba\-li\-sya ku\-bbi.}
\fmtExampleEN{The calculator gives me wrong results.}

\fmtLemma{ka\-li\-daa\-li}
\fmtCitation{(o)ka\-li\-daa\-li}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSource{from Ar. via Sw.}
\fmtSense{mustard seed}
\fmtExampleRL{Oka\-li\-daa\-li ai\-ra ku\-sai ewaa\-bwe.}
\fmtExampleEN{Mustard grows well in their area.}

\fmtLemma{ka\-li\-foo\-mu}
\fmtCitation{(o)ka\-li\-foo\-mu}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{chloroform}
\fmtExampleRL{Ba\-mu\-ku\-bbi\-re oka\-li\-foo\-mu mwi\-ngi n’a\-fa.}
\fmtExampleEN{They administered him a lot of chloroform and he died.}

\fmtLemma{ka\-li\-fuu\-wa}
\fmtCitation{(o)ka\-li\-fuu\-wa}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{perfume, scent}
\fmtExampleRL{Aba\-ka\-li ba\-ta\-ka mu\-no okwe\-ku\-bba oka\-li\-fuu\-wa.}
\fmtExampleEN{Women really love wearing perfumes.}

\fmtLemma{ka\-li\-ga}
\fmtCitation{(o)ka\-li\-ga}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{pure concentrated brew}
\fmtExampleRL{Anywi\-re oka\-li\-ga olwa\-lee\-ro.}
\fmtExampleEN{Today he has drunk pure concentrated brew.}

\fmtLemma{ka\-lii\-bwa\-ba\-lii\-sya}
\fmtCitation{(a)ka\-lii\-bwa\-ba\-lii\-sya}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{type of wild small grapes eaten by people who are grazing animals}

\fmtLemma{Ka\-lii\-sya}
\fmtCitation{(o)Ka\-lii\-sya}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSubentry(a)
\fmtSense{type of god, spirit responsible for cattle keeping}
\fmtExampleRL{Aba\-lii\-sya ba\-wo\-nge\-ra ka\-lii\-sya.}
\fmtExampleEN{Cattle-keepers offer sacrifices to the spirit called Kaliisa.}
\fmtSubentry(b)
\fmtSense{God}
\fmtExampleRL{Oka\-lii\-sya wa mwi\-gu\-ru atu\-ga\-bi\-rii\-re ku\-sai.}
\fmtExampleEN{The God of the heavens has blessed us mightily.}

\fmtLemma{ka\-li\-kii\-rya}
\fmtCitation{(a)ka\-li\-kii\-rya}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{darkness}
\fmtExampleRL{Wa\-li\-yo aka\-li\-kii\-rya, ti\-wa\-so\-bo\-le ku\-mu\-bo\-na ku\-sai.}
\fmtExampleEN{It is a bit dark (lit.\,there is some darkness), you won’t be able to see him properly.}

\fmtLemma{ka\-li\-ma\-ki\-za}
\fmtCitation{(o)ka\-li\-ma\-ki\-za}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{controversial person}
\fmtExampleRL{Omwa\-la oyo ka\-li\-ma\-ki\-za.}
\fmtExampleEN{That girl is a controversial person.}

\fmtLemma{ka\-li\-mi}%
\fmtOne{1}
\fmtCitation{(a)ka\-li\-mi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{bud, sprout, shoot}
\fmtExampleRL{Ni wa\-bo\-na omu\-cu\-ngwai nga gu\-lee\-te\-re obu\-li\-mi bwi\-ngi, obu\-mwei bu\-no\-ko\-le\-ku.}
\fmtExampleEN{If you see the orange tree producing many buds, prune some off.}
\fmtSubentry(b)
\fmtSense{hand (a pointer on a clock)}
\fmtExampleRL{Aka\-li\-mi ka saa\-wa ka\-tii\-ki\-re.}
\fmtExampleEN{The clock hand has broken.}

\fmtLemma{ka\-li\-mi}%
\fmtOne{2}
\fmtCitation{(a)ka\-li\-mi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{convincing tongue, silver tongue}
\fmtExampleRL{Owa\-ndi\-ra ata na ka\-li\-mi.}
\fmtExampleEN{Wandera has a silver tongue.}

\fmtLemma{ka\-li\-mi\-ka\-ga\-mba}
\fmtPOS{pro}
\fmtSense{myself}
\fmtExampleRL{Ni\-nje ka\-li\-mi\-ka\-ga\-mba.}
\fmtExampleEN{It is me myself.}

\fmtLemma{ka\-li\-mpa\-ka}
\fmtCitation{(o)ka\-li\-mpa\-ka}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{argumentative person}
\fmtExampleRL{Omu\-ka\-li waa\-mwe ka\-li\-mpa\-ka.}
\fmtExampleEN{His wife is an argumentative person.}

\fmtLemma{ka\-li\-na}
\fmtCrossreference{\textnormal{see} kalinabiri}

\fmtLemma{ka\-li\-na\-bi\-ri}
\fmtCitation{(e)ka\-li\-na\-bi\-ri}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{skyscraper}

\fmtLemma{Ka\-li\-ndai\-ru\-ba}
\fmtCitation{(o)ka\-li\-ndai\-ru\-ba}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{officer in charge of communal wells}
\fmtExampleRL{Oka\-li\-ndai\-ru\-ba yee\-na asa\-su\-lwe\-nge.}
\fmtExampleEN{The officer in charge of communal wells should also be paid.}

\fmtLemma{ka\-li\-vaa\-li}
\fmtCitation{(e)ka\-li\-vaa\-li}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{cavalry}
\fmtExampleRL{Oye\-su ba\-mu\-ba\-mbii\-re Ka\-li\-vaa\-li.}
\fmtExampleEN{Jesus was crucified at Calvary.}

\fmtLemma{ka\-li\-ya}
\fmtCitation{(e)ka\-li\-ya}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{carrier (of a bicycle)}
\fmtExampleRL{Ta eka\-li\-ya oku gaa\-li ya\-nge.}
\fmtExampleEN{Put a carrier on my bicycle.}

\fmtLemma{ka\-loi\-so}
\fmtCitation{(a)ka\-loi\-so}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{stare, long fixed look}
\fmtExampleRL{Aka\-loi\-so ko\-nku\-bbi\-re kaa\-maa\-re\-ko emi\-re\-mbe.}
\fmtExampleEN{Your stare is making me uncomfortable.}

\fmtLemma{ka\-loo\-li}
\fmtCitation{(o)ka\-loo\-li}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{marabou stork}
\fmtExampleRL{Oka\-loo\-li alya encwi.}
\fmtExampleEN{The marabou stork eats fish.}

\fmtLemma{ka\-loo\-lo}
\fmtCitation{(a)ka\-loo\-lo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{chicken flea, chicken lice (\textit{Ceratophyllus gallinea})}
\fmtExampleRL{Obu\-loo\-lo bu\-bba ku nko\-ko.}
\fmtExampleEN{Chicken fleas live on chicken.}

\fmtLemma{ka\-loo\-sa}
\fmtCitation{(a)ka\-loo\-sa}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{aroma, scent}
\fmtExampleRL{Omu\-ku\-bi gu\-li\-mu aka\-loo\-sa.}
\fmtExampleEN{The sauce has a good aroma.}
\fmtSubentry(b)
\fmtSense{perfume}
\fmtExampleRL{Omu\-ka\-li ye\-ku\-bbi\-re aka\-loo\-sa aka\-sai.}
\fmtExampleEN{The woman is wearing a nice perfume.}

\fmtLemma{ka\-lu}
\fmtCitation{(e.g.ri\-ka\-li, bi\-ka\-lu, mu\-ka\-lu)}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{dry}
\fmtExampleRL{Ei\-su\-bi li\-ka\-lu.}
\fmtExampleEN{The grass is dry.}
\fmtSubentry(b)
\fmtSense{without milk}
\fmtExampleRL{Nku\-nywa caa\-yi mu\-ka\-lu.}
\fmtExampleEN{I am drinking tea without milk.}
\fmtSubentry(c)
\fmtSense{tangible, concrete}
\fmtExampleRL{Oi\-se\-ma\-ka atee\-kwa oku\-bbaa\-wo n’e\-bi\-ntu ebi\-ka\-lu.}
\fmtExampleEN{The family head must possess some tangible property.}
\fmtSubentry(d)
\fmtSense{remorseless, heartless, pitiless, merciless, cruel}
\fmtExampleRL{OKa\-la\-ngwa mu\-ka\-lu wa mwo\-yo.}
\fmtExampleEN{Kalangwa is a remorseless person.}

\fmtLemma{ka\-lu\-bii\-rya}
\fmtCitation{(oku)\-ka\-lu\-bii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-lu\-bi\-rii\-rye}
\fmtSense{to complicate matters for}
\fmtExampleRL{Lwa\-ki oku\-nka\-lu\-bii\-rya?}
\fmtExampleEN{Why are you complicating matters for me?}

\fmtLemma{ka\-lu\-ka\-lu}
\fmtCitation{\textnormal{(e.g.}\,mu\-ka\-lu\-ka\-lu, ba\-ka\-lu\-ka\-lu\textnormal{)}}
\fmtPOS{adj}
\fmtSense{confident, self-assured}
\fmtExampleRL{Ai\-za ku\-ku\-laa\-ku\-la\-na ku\-bba abbai\-re mwa\-na mu\-ka\-lu\-ka\-lu.}
\fmtExampleEN{He will succeed because he has been a confident boy.}

\fmtLemma{ka\-lu\-lu}
\fmtCitation{(a)ka\-lu\-lu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{lot, betting, gambling}
\fmtExampleRL{Aboo\-jo ba\-ta\-ka mu\-no oku\-zee\-nya aka\-lu\-lu.}
\fmtExampleEN{The boys like playing the lot game.}
\fmtSubentry(b)
\fmtSense{vote, election}
\fmtExampleRL{Bu\-li ka\-lu\-lu ka mu\-ntu ka\-bba ka mu\-ga\-so.}
\fmtExampleEN{Everyone’s vote is important.}
\fmtSubentry(c)
\fmtFixedexpression{ku\-bba (a)ka\-lu\-lu}
\fmtSense{to cast a vote}
\fmtExampleRL{Tu\-maa\-re oku\-ku\-bba obu\-lu\-lu.}
\fmtExampleEN{We have finished casting our votes.}

\fmtLemma{ka\-lu\-mba}
\fmtCitation{(a)ka\-lu\-mba}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{(of a boat) bow}
\fmtExampleRL{Omu\-gi\-mbi tai\-ka\-la mu ka\-lu\-mba ai\-ka\-la mu nsaa\-sa.}
\fmtExampleEN{The oarsman does not sit at the stern of the boat, he sits at the bow.}

\fmtLemma{ka\-lu\-nga}
\fmtDialect{Ruruuli}
\fmtCitation{(a)ka\-lu\-nga}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{black and red seed from a kalunga plant (type of herbal plant with red seeds)}

\fmtLemma{ka\-lu\-su}
\fmtCitation{(o)ka\-lu\-su}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{foot-and-mouth disease}
\fmtExampleRL{Ente zi\-rwai\-re oka\-lu\-su.}
\fmtExampleEN{The cattle are suffering from foot-and-mouth disease.}

\fmtLemma{ka\-luu\-lu}
\fmtCitation{(a)ka\-luu\-lu}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{joyous ululation}
\fmtExampleRL{Ba\-sa\-ga\-ru\-ki\-re ni ba\-ku\-bba n’a\-ka\-luu\-lu.}
\fmtExampleEN{They were very happy and made a joyous ululation.}

\fmtLemma{ka\-lya}
\fmtCitation{(oku)\-ka\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kaa\-rye}
\fmtSense{to dry, to make dry}
\fmtExampleRL{Ka\-lya ebya\-nzo ebyo bi\-ta\-ju\-nda.}
\fmtExampleEN{Dry those hides or else they will rot.}

\fmtLemma{ka\-lya\-bu\-ga\-lu\-la}
\fmtCitation{(e)ka\-lya\-bu\-ga\-lu\-la}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSubentry(a)
\fmtSense{type of hammer used in metal forging}
\fmtSubentry(b)
\fmtSense{god, spirit worshipped by blacksmiths commonly among the Bagala clan}
\fmtExampleRL{Aba\-ga\-la ba\-wo\-nge\-ra oKa\-la\-bu\-ga\-lu\-la.}
\fmtExampleEN{The Bagala offer sacrifices to Kalyabugalula.}

\fmtLemma{ka\-lyai\-ko\-ti}
\fmtCitation{(o)ka\-lyai\-ko\-ti}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{type of fetish and type of spirit responsible for giving wealth and child birth}

\fmtLemma{ka\-lyai\-so\-ke}
\fmtCitation{(o)ka\-lyai\-so\-ke}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{Mediterranean house gecko}
\fmtExampleRL{Aka\-lyai\-so\-ke ka\-su\-la mu nyu\-mba.}
\fmtExampleEN{The Mediterranean house gecko lives in houses.}

\fmtLemma{ka\-ma}
\fmtCitation{(oku)\-ka\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-mi\-re}
\fmtSense{to milk}
\fmtExampleRL{Aya\-bi\-re ku\-ka\-ma nte.}
\fmtExampleEN{He has gone to milk the cow.}

\fmtLemma{ka\-maa\-nya}
\fmtCitation{(o)ka\-maa\-nya}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{(of a person) very evil}
\fmtExampleRL{Ka\-maa\-nya we fa n’a\-ma\-go\-mbe ga\-bu\-le.}
\fmtExampleEN{You evil one, may you die and never be buried.}

\fmtLemma{ka\-mai\-tu\-ka}
\fmtCitation{(oku)\-ka\-mai\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-mai\-tu\-ki\-re}
\fmtSense{to do without thinking it through}
\fmtExampleRL{Ebi\-ro ebi\-ngi aka\-mai\-tu\-ka\-mbe.}
\fmtExampleEN{Most times he does things without thinking them through.}

\fmtLemma{ka\-ma\-la}
\fmtCrossreference{\textnormal{see} kamara}

\fmtLemma{ka\-ma\-nsuu\-li}
\fmtCitation{(a)ka\-ma\-nsuu\-li}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{trap}
\fmtExampleRL{Aka\-ma\-nsuu\-li kai\-ti\-re onko\-wu.}
\fmtExampleEN{The trap has killed a guinea fowl.}

\fmtLemma{ka\-ma\-nyii\-ro}
\fmtCitation{(a)ka\-ma\-nyii\-ro}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{tactlessness, lack of sensitivity}
\fmtExampleRL{Aka\-ma\-nyii\-ro ka\-bya\-la obu\-ta\-tii\-na.}
\fmtExampleEN{Tactlessness leads to contempt.}

\fmtLemma{ka\-ma\-nyi\-syo}
\fmtCitation{(a)ka\-ma\-nyi\-syo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{symptom, indicator, sign}
\fmtExampleRL{Ako ka\-ma\-nyi\-syo ka bu\-lwai\-re bwa bi\-soo\-ne.}
\fmtExampleEN{That is a symptom of syphilis.}
\fmtSubentry(b)
\fmtSense{signpost}
\fmtExampleRL{Omu Ka\-yu\-nga mu\-li\-mu obu\-ma\-nyi\-syo bwi\-ngi.}
\fmtExampleEN{There are many signposts in Kayunga.}
\fmtSubentry(c)
\fmtSense{notification}
\fmtExampleRL{Aka\-ma\-nyi\-syo kai\-zi\-re oku si\-mu ya\-nge.}
\fmtExampleEN{A notification appeared on my phone.}
\fmtSubentry(d)
\fmtSense{punctuation mark}
\fmtExampleRL{N’o\-bba oku\-wa\-ndii\-ka otaa\-mu obu\-ma\-nyi\-syo.}
\fmtExampleEN{When you are writing, use punctuation marks.}

\fmtLemma{ka\-ma\-ra}
\fmtCitation{(o)ka\-ma\-ra}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{county chief}
\fmtExampleRL{Ni\-nje ka\-ma\-ra wa\-bwa\-ka\-ma\-ra bu\-ni.}
\fmtExampleEN{I am the county chief of this county.}

\fmtLemma{ka\-ma\-ra\-byo\-na}
\fmtCitation{(o)ka\-ma\-ra\-byo\-na}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSubentry(a)
\fmtSense{prime minister}
\fmtExampleRL{Ni\-ye Ka\-ma\-ra\-byo\-na wa Bu\-ka\-ma bwa Bu\-ruu\-li.}
\fmtExampleEN{He is the prime minister of Buruuli kingdom.}
\fmtSubentry(b)
\fmtSense{Lord}
\fmtExampleRL{Nku\-be\-ndee\-ra ka\-ma\-ra\-byo\-na anju\-ne.}
\fmtExampleEN{I am waiting upon the Lord for his help.}

\fmtLemma{ka\-ma\-su}
\fmtCitation{(a)ka\-ma\-su}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{rat trap}
\fmtExampleRL{Aka\-ma\-su ka\-kwai\-te ome\-se.}
\fmtExampleEN{The rat trap has caught a rat.}

\fmtLemma{ka\-ma\-su\-ka}
\fmtCitation{(a)ka\-ma\-su\-ka}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{new moon}
\fmtExampleRL{Aka\-ma\-su\-ka ka\-ku\-bo\-ne\-ka ku\-sai.}
\fmtExampleEN{The new moon looks nice.}

\fmtLemma{ka\-ma\-zi\-ma}
\fmtPOS{interj}
\fmtSense{imagine, just suppose (an exclamation used to express surprise)}
\fmtExampleRL{Ka\-ma\-zi\-ma, na\-mu\-wuu\-li\-re nga aku\-ba\-za!}
\fmtExampleEN{Imagine, I heard him say so!}

\fmtLemma{ka\-mba}
\fmtCitation{(oku)\-ka\-mba}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{ka\-mbi\-re}
\fmtSubentry(a)
\fmtSense{to weave}
\fmtExampleRL{Mmai\-te oku\-ka\-mba omu\-kee\-ka.}
\fmtExampleEN{I know how to weave a mat.}
\fmtSubentry(b)
\fmtSense{to knit}
\fmtExampleRL{Ta\-ndi\-ka oku\-ka\-mba eswe\-ta.}
\fmtExampleEN{Start knitting the sweater.}
\fmtSubentry(c)
\fmtSense{to plait (hair), to braid}
\fmtExampleRL{Omwa\-ra wa\-nge aye\-ge\-re oku\-ka\-mba ei\-so\-ke.}
\fmtExampleEN{My daughter has learnt to plait hair.}
\fmtSubentry(d)
\fmtSense{to have sticky skin}
\fmtExampleRL{Naa\-ba enga\-lo zaa\-mu ng’o\-kya\-li ku\-nkwa\-ta\-ko ku\-bba zi\-ku\-ka\-mba.}
\fmtExampleEN{Wash your hands before you touch me, they are sticky.}

\fmtLemma{ka\-mbai\-ga}
\fmtCitation{(oku)\-ka\-mbai\-ga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-mbai\-gi\-re}
\fmtSense{to have a tart taste (\eg of unripe mangoes)}
\fmtExampleRL{Omu\-koo\-ge gu\-ku\-ka\-mbai\-ga.}
\fmtExampleEN{The tamarind has a tart taste.}

\fmtLemma{ka\-mba\-maa\-gi}
\fmtCrossreference{\textnormal{see also} kaka\-mbamaagi}
\fmtCitation{(o)ka\-mba\-maa\-gi}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{type of fishing net used for trapping tilapia}
\fmtExampleRL{Ka\-mba\-maa\-gi ka\-ti\-mba ka bi\-na\-ge.}
\fmtExampleEN{Ka\-mbamaagi is a type of fishnet for fishing tilapia.}

\fmtLemma{ka\-mbii\-ra}
\fmtCitation{(oku)\-ka\-mbii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-mbii\-re}
\fmtSource{from archaic}
\fmtSense{to show direction, to direct, to show the way}
\fmtExampleRL{Amo\-s mu\-ka\-mbii\-re enzi\-ra ekwa\-ba ewa Edwa\-rd.}
\fmtExampleEN{Show Amos the way to Edward’s home.}

\fmtLemma{ka\-mbu\-gu\-mbu\-gu}
\fmtCitation{(a)ka\-mbu\-gu\-mbu\-gu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{kind of insect-eating bat, small bat}
\fmtExampleRL{Obu\-mbu\-gu\-mbu\-gu bu\-lee\-ta endwai\-re zi\-ngi omu ba\-ntu.}
\fmtExampleEN{Bats cause many diseases in humans.}

\fmtLemma{ka\-mbu\-la}
\fmtCitation{(a)ka\-mbu\-la}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{type of herb used to cleanse someone with bad luck or misfortune}
\fmtExampleRL{Azwi\-re omu bu\-bbwe ba\-mu\-naa\-bya ka\-mbu\-la.}
\fmtExampleEN{A person from the prison is cleansed with the kambura herb.}

\fmtLemma{ka\-mbuu\-la}
\fmtCitation{(oku)\-ka\-mbuu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-mbwi\-re}
\fmtSense{to unroof, to remove the roof, to uncover}
\fmtExampleRL{Ka\-mbuu\-la enyu\-mba ako otee\-ku ama\-baa\-ti aga\-yaa\-ka ge twa\-guu\-re.}
\fmtExampleEN{Unroof that house and put on the new iron sheets we bought.}

\fmtLemma{ka\-mbu\-wa\-la}
\fmtCitation{(oku)\-ka\-mbu\-wa\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-mbu\-wai\-re}
\fmtSense{to be harsh, to impolite, to unfriendly, to unkind}
\fmtExampleRL{Lwa\-ki oka\-mbu\-wa\-la mu\-no?}
\fmtExampleEN{Why are you always harsh?}

\fmtLemma{ka\-mbwe}
\fmtCitation{\textnormal{(e.g.}\,mu\-ka\-mbwe, zi\-ka\-mbwe, mu\-ka\-mbwe\textnormal{)}}
\fmtPOS{adj}
\fmtSense{fierce}
\fmtExampleRL{Onta\-le abba mu\-ka\-mbwe.}
\fmtExampleEN{A lion is fierce.}

\fmtLemma{ka\-me\-ga}
\fmtCitation{(a)ka\-me\-ga}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{scooping plate}
\fmtExampleRL{Lee\-ta aka\-me\-ga mba\-ga\-bi\-re eme\-re.}
\fmtExampleEN{Bring the scooping plate and I'll serve you food.}

\fmtLemma{ka\-me\-kai}
\fmtPOS{interrog}
\fmtSense{how often, how many times}
\fmtExampleRL{Ka\-me\-kai nga nkwi\-za ani n’o\-ta\-bo\-ne\-ke\-ku?}
\fmtExampleEN{How many times have I come here without seeing you?}

\fmtLemma{ka\-me\-ngo}
\fmtCitation{(a)ka\-me\-ngo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{enlarged spleen usually due to untreated malaria or a tumour, splenomegaly}
\fmtExampleRL{Omwa\-na waa\-mwe mu\-lwai\-re wa ka\-me\-ngo.}
\fmtExampleEN{Her son is suffering from splenomegaly.}

\fmtLemma{ka\-me\-ra}
\fmtCitation{(e)ka\-me\-ra}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{camera}
\fmtExampleRL{Eka\-me\-ra yaa\-mwe ewu\-lu\-ki\-rya\-wo ebii\-sa\-ni.}
\fmtExampleEN{His camera produces photos instantly.}

\fmtLemma{ka\-mi\-gii\-ryo}
\fmtCitation{(a)ka\-mi\-gii\-ryo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSubentry(a)
\fmtSense{crowded place, narrow place}
\fmtExampleRL{Nje ti\-nta\-ka ku\-ta\-mbuu\-ra mu mbee\-ra ya ka\-mi\-gii\-ryo.}
\fmtExampleEN{I don’t want to walk in crowded places.}
\fmtSubentry(b)
\fmtSense{selfishness, egoism, self-centeredness}
\fmtExampleRL{Ako kaa\-li ka\-mi\-gii\-ryo ke\-nyi\-ni.}
\fmtExampleEN{That was an act of pure selfishness.}

\fmtLemma{ka\-mi\-gi\-syo}
\fmtCitation{(a)ka\-mi\-gi\-syo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{clamp for castration, burdizzo}
\fmtExampleRL{Aka\-mi\-gi\-syo ba\-ka\-ko\-le\-sya nga ba\-ku\-saa\-ma ebi\-so\-lo.}
\fmtExampleEN{A burdizzo is used during the castration of animals.}

\fmtLemma{ka\-mi\-go}
\fmtCitation{(a)ka\-mi\-go}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{congestion, overcrowding}
\fmtExampleRL{Ota\-yi\-ngi\-ra mu nyu\-mba ku\-bba aka\-mi\-go kai\-ngi.}
\fmtExampleEN{Don’t enter the house because it is too crowded.}

\fmtLemma{ka\-mii\-mo}
\fmtPOS{adj}
\fmtSense{tight, small, tight fitting}
\fmtExampleRL{Omwo\-jo azwai\-re empa\-le ka\-mii\-mo.}
\fmtExampleEN{The boy is wearing tight trousers.}

\fmtLemma{ka\-mi\-ro}
\fmtCitation{(a)ka\-mi\-ro}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{windpipe, trachea}
\fmtExampleRL{Amai\-zi ga\-mwa\-bi\-re omu ka\-mi\-ro.}
\fmtExampleEN{The water got into his windpipe.}
\fmtSubentry(b)
\fmtSense{excessive greed}
\fmtExampleRL{Omu\-ka\-li oyo ata n’a\-ka\-mi\-ro.}
\fmtExampleEN{That woman is very greedy.}

\fmtLemma{ka\-moo\-li}
\fmtCitation{(a)ka\-moo\-li}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{ventilation shaft, ventilation system}
\fmtExampleRL{Ome\-se ayi\-ngii\-re mu ka\-moo\-li.}
\fmtExampleEN{The rat entered through the ventilation shift.}

\fmtLemma{ka\-mpa\-gu\-ki}
\fmtCitation{(a)ka\-mpa\-gu\-ki}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{hiccup}
\fmtExampleRL{Mu\-lwai\-re wa\-ka\-mpa\-gu\-ki.}
\fmtExampleEN{She’s suffering from hiccups.}

\fmtLemma{ka\-mpa\-la\-la}
\fmtCitation{(o)ka\-mpa\-la\-la}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{large grasshopper}
\fmtExampleRL{Onko\-ko aku\-lya oka\-mpa\-la\-la.}
\fmtExampleEN{The hen is eating a large grasshopper.}

\fmtLemma{ka\-mpa\-nga}
\fmtCrossreference{\textnormal{see also} mpanga}
\fmtCitation{(a)ka\-mpa\-nga}
\fmtPOS{n}%
\fmtNC{12 (or 1a)/14}
\fmtSense{cockerel, young rooster, young cock}
\fmtExampleRL{Nguu\-re\-yo aka\-mpa\-nga.}
\fmtExampleEN{I have bought a young cockerel.}

\fmtLemma{ka\-mpe\-wo}
\fmtCitation{(a)ka\-mpe\-wo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{fresh air}
\fmtExampleRL{Awu\-lu\-ki\-re nzai ku\-wuu\-ra ku\-ka\-mpe\-wo.}
\fmtExampleEN{He has gone outside to feel the fresh air.}

\fmtLemma{ka\-mpu\-lu}
\fmtCitation{(a)ka\-mpu\-lu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{hide-and-seek (children's game)}
\fmtExampleRL{Abaa\-na aba\-to ba\-ta\-ka oku\-zee\-nya aka\-mpu\-lu.}
\fmtExampleEN{Children like to play hide-and-seek.}

\fmtLemma{ka\-mpu\-ngo}
\fmtCitation{(a)ka\-mpu\-ngo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{breeze, gentle fresh wind}
\fmtExampleRL{Aka\-mpu\-ngo enzai ka\-ku\-fuu\-wa mpo\-la\-mpo\-la.}
\fmtExampleEN{A breeze is blowing outside.}

\fmtLemma{ka\-mpu\-ni}
\fmtCitation{(e)ka\-mpu\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{company, firm, enterprise}
\fmtExampleRL{Aguu\-re eka\-mpu\-ni ya mu\-ba\-zi.}
\fmtExampleEN{He has bought a pharmaceutical company.}

\fmtLemma{ka\-mpwa\-ci\-mpwa\-ci}
\fmtCitation{(a)ka\-mpwa\-ci\-mpwa\-ci}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{snack}
\fmtExampleRL{Abaa\-na ba\-ta\-ka mu\-no oku\-lya obu\-mpwa\-nci\-mpwa\-nci oku iso\-me\-ro,}
\fmtExampleEN{Children love to eat snacks when they are at school.}

\fmtLemma{ka\-mu\-ce\-nce\-ne}
\fmtCitation{(o)ka\-mu\-ce\-nce\-ne}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{Mediterranean house gecko (\textit{Hemidactylus turcicus})}
\fmtExampleRL{Oka\-mu\-ce\-nce\-ne mwa\-ndi mu\-no.}
\fmtExampleEN{The Mediterranean house gecko is a very fierce lizard.}

\fmtLemma{ka\-mu\-ka}
\fmtCitation{(oku)\-ka\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-mu\-ki\-re}
\fmtSense{to be half-dry}
\fmtExampleRL{Olu\-go\-ye lu\-ka\-mu\-ki\-re.}
\fmtExampleEN{The cloth is half-dry.}

\fmtLemma{ka\-mu\-ko\-dyai}
\fmtCitation{(o)ka\-mu\-ko\-dyai}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{type of leafy plant with pinkish flowers used as vegetables}
\fmtExampleRL{Oka\-mu\-ko\-dyai mu\-ku\-bi gwa nza\-la.}
\fmtExampleEN{The pinkish leafy plant is usually used to make sauce during a period of famine.}

\fmtLemma{ka\-mu\-ku\-nu}
\fmtCitation{(o)ka\-mu\-ku\-nu}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{cassava brown streak}
\fmtExampleRL{Oka\-mu\-ku\-nu afoo\-la ebi\-yaa\-ta ni bi\-sa\-na ebi\-su\-mbe\-ku.}
\fmtExampleEN{Cassava brown streak makes cassava tubers look as though they were cooked.}

\fmtLemma{ka\-mu\-la}
\fmtCitation{(oku)\-ka\-mu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-mwi\-re}
\fmtSense{to squeeze out (liquid), to extract}
\fmtExampleRL{Ya\-ka\-mwi\-re omu\-cu\-gwai.}
\fmtExampleEN{She squeezed juice out from the orange.}

\fmtLemma{ka\-mu\-la\-li}
\fmtCitation{(o)ka\-mu\-la\-li}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{hot chilli pepper}
\fmtExampleRL{Ntee\-ra oka\-mu\-la\-li omu mu\-ku\-bi gwa\-nge.}
\fmtExampleEN{Put hot chilli in my sauce.}

\fmtLemma{ka\-mu\-nye}
\fmtCitation{(e)ka\-mu\-nye}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{fourteen seater minibus used as means of public transport}
\fmtExampleRL{Eka\-mu\-nye ni\-zo zi\-si\-nga oku\-ta\-mbu\-lya aba\-ntu omu Uga\-nda.}
\fmtExampleEN{A fourteen seater minibus is the most common means to transport people in Uganda.}

\fmtLemma{ka\-mu\-sa\-lya}
\fmtCitation{(a)ka\-mu\-sa\-lya}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{temporally mental disorder}
\fmtExampleRL{Omu\-ka\-li oyo ali\-ku aka\-mu\-sa\-lya.}
\fmtExampleEN{That woman has a temporary mental disorder.}

\fmtLemma{ka\-mu\-sye\-ku}
\fmtCitation{(a)ka\-mu\-sye\-ku}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{powder made from cow peas leaves and used as a spice for sauces}
\fmtExampleRL{Omu\-ku\-bi gu\-li\-mu ka\-mu\-sye\-ku gu woo\-ma.}
\fmtExampleEN{Sauce with leaf powder is delicious.}

\fmtLemma{ka\-mwa\-no}
\fmtCitation{(a)ka\-mwa\-no}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{traditional razor blade (piece of reed traditionally used to cut the u\-mbilical cord)}
\fmtExampleRL{Ebi\-ro aba\-ka\-li ti\-ba\-kya\-le\-sya ka\-mwa\-no oku\-sa\-la omu\-ku\-ndi gwa nke\-re\-mba zaa\-bwe.}
\fmtExampleEN{These days women no longer use traditional razor blades to cut the u\-mbilical cord of their babies.}

\fmtLemma{ka\-mwe\-nyu\-mwe\-nyu}
\fmtCitation{(a)ka\-mwe\-nyu\-mwe\-nyu}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{smile}
\fmtExampleRL{N’a\-mbwoi\-ne n’ai\-za\-ku aka\-mwe\-nyu\-mwe\-nyu.}
\fmtExampleEN{When she saw me, she smiled.}

\fmtLemma{ka\-mwe\-syo}
\fmtCitation{(a)ka\-mwe\-syo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{shaver, razor blade}
\fmtExampleRL{Lee\-ta aka\-mwe\-syo nku\-ge\-mbe omu\-le\-zu.}
\fmtExampleEN{Bring the shaver and I'll shave off your beard.}

\fmtLemma{Ka\-mwi\-zi\-nga}
\fmtCitation{(o)Ka\-mwi\-zi\-nga}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{name of a female born on an island}
\fmtExampleRL{OKwa\-mwi\-zi\-nga abba mwa\-na eya\-bya\-lii\-bwe oku ki\-zi\-nga.}
\fmtExampleEN{Okwamwizinga is a name given to a female born from an island.}

\fmtLemma{ka\-naa\-nai}
\fmtPOS{num}
\fmtSense{eight thousand}
\fmtExampleRL{Empa\-le ngi\-guu\-re ka\-naa\-nai.}
\fmtExampleEN{I bought the trousers for eight thousand shillings.}

\fmtLemma{ka\-na\-ba\-lii\-sya}
\fmtCitation{(a)ka\-na\-ba\-lii\-sya}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{caecum of a goat}
\fmtExampleRL{Aka\-na\-ba\-lii\-sya ka mbu\-li ka\-woo\-ma mu\-no.}
\fmtExampleEN{The caecum of a goat is delicious.}

\fmtLemma{ka\-na\-ga}
\fmtCitation{(a)ka\-na\-ga}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{cooking pot}
\fmtExampleRL{Omu\-ku\-bi ba\-gu\-su\-mbi\-ra mu ka\-na\-ga.}
\fmtExampleEN{The sauce is cooked in a cooking pot.}

\fmtLemma{ka\-na\-ga\-ba\-ra\-si}
\fmtCitation{(a)ka\-na\-ga\-ba\-ra\-si}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{siesta, afternoon sleep, nap}
\fmtExampleRL{Afu\-ni\-re\-mu aka\-na\-ga\-ba\-ra\-si.}
\fmtExampleEN{He has taken a siesta.}

\fmtLemma{ka\-nai}
\fmtPOS{num}
\fmtSubentry(a)
\fmtSense{fourth}
\fmtSubentry(b)
\fmtFixedexpression{mwe\-zi ga ka\-nai}
\fmtSense{April}
\fmtSubentry(c)
\fmtFixedexpression{buu\-lya ka\-nai n'aka\-taa\-no}
\fmtSense{to interrogated, to ask to provide details}

\fmtLemma{ka\-na\-mai\-zi}
\fmtCitation{(a)na\-mai\-zi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{outer mushroom}
\fmtExampleRL{Obu\-na\-mai\-zi bu\-me\-ra kwa ike\-ndi.}
\fmtExampleEN{Outer mushrooms grow during the rainy season.}

\fmtLemma{ka\-na\-nai}
\fmtPOS{num}
\fmtSense{eight thousand}

\fmtLemma{ka\-na\-nsoo\-ro}
\fmtCitation{(a)ka\-na\-nsoo\-ro}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{small whitish edible mushroom (\textit{Termitomyces microcarpus})}
\fmtExampleRL{Obu\-na\-nsoo\-ro bu\-te\-ra ku\-me\-ra ku ki\-swa.}
\fmtExampleEN{\textit{Termitomyces microcarpus} usually grows on termite mounds.}

\fmtLemma{ka\-nca}%
\fmtOne{1}
\fmtCitation{(oku)\-ka\-nca}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-nci\-re}
\fmtSense{to create, to generate, to produce}
\fmtExampleRL{To\-so\-bo\-la ku\-ka\-nca\-wo lu\-li\-mi lu\-ta\-bbai\-re\-wo.}
\fmtExampleEN{You can’t create a language out of nothing.}

\fmtLemma{Ka\-nca}%
\fmtOne{2}
\fmtCitation{(o)Ka\-nca}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{God, creator}
\fmtExampleRL{OKa\-nca mu\-sai.}
\fmtExampleEN{God is good.}

\fmtLemma{ka\-nda}%
\fmtOne{1}
\fmtPOS{adv}
\fmtSense{maybe, perhaps, possibly}
\fmtExampleRL{Okwi\-za ka\-nda bbe\-? Ka\-nda wai\-za je bu\-kwi\-za.}
\fmtExampleEN{Are you coming or not? Perhaps you will come later.}

\fmtLemma{ka\-nda}%
\fmtOne{2}
\fmtPOS{conj}
\fmtSense{unless, or}
\fmtExampleRL{Oli mu\-nywa\-ni wa\-nge ka\-nda nga oli omwe oku ba\-ku\-ndo\-ga.}
\fmtExampleEN{You are my friend, unless you are among those who are bewitching me.}

\fmtLemma{ka\-nda}%
\fmtOne{3}
\fmtCitation{(oku)\-ka\-nda}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-ndi\-re}
\fmtSense{to knead}
\fmtExampleRL{Aku\-ka\-nda ŋaa\-no.}
\fmtExampleEN{He is kneading flour to make dough.}

\fmtLemma{ka\-nda\-bu\-la}
\fmtCitation{(oku)\-ka\-nda\-bu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-nda\-bwi\-re}
\fmtSense{to unsew, to undo the sewing, to take apart the sewing}
\fmtExampleRL{Ka\-nda\-bu\-la eki\-tee\-tee\-yi ekyo oki\-ga\-zi\-ye.}
\fmtExampleEN{Take apart that dress and make it bigger.}

\fmtLemma{ka\-nda\-ki}
\fmtPOS{interj}
\fmtSense{expression of surprise}
\fmtExampleRL{Odi mwa\-la ka\-nda\-ki?}
\fmtExampleEN{What kind of girl is that?}

\fmtLemma{ka\-nde\-re\-nde\-re}
\fmtCitation{(o)ka\-nde\-re\-nde\-re}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{newly born baby}
\fmtExampleRL{Enye\-nyee\-zi edi eku\-la\-nga nti wa\-li\-wo oka\-nde\-re\-nde\-re aka\-bya\-lii\-bwe.}
\fmtExampleEN{That star announces that there is a newborn baby.}

\fmtLemma{ka\-ndi\-da}
\fmtCitation{(o)ka\-ndi\-da}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{candida}
\fmtExampleRL{Ai\-da\-h alwai\-re oka\-ndi\-da.}
\fmtExampleEN{Aidah is suffering from candida.}

\fmtLemma{ka\-ndi\-ga\-ma\-ra\-ka}
\fmtCitation{(a)ka\-ndi\-ga\-ma\-ra\-ka}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{microphone}
\fmtExampleRL{Aka\-ndi\-ga\-ma\-ra\-ka ke ba\-ko\-le\-se\-rye oku mba\-ga ka\-bbai\-re ka\-fu.}
\fmtExampleEN{The microphone which they used at the party was faulty.}

\fmtLemma{ka\-ndu\-la}
\fmtCitation{(oku)\-ka\-ndu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-ndwi\-re}
\fmtSense{to rake}
\fmtExampleRL{Ka\-ndu\-la ebi\-su\-bi ebyo.}
\fmtExampleEN{Rake that grass.}

\fmtLemma{ka\-nga}
\fmtCitation{(oku)\-ka\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-ngi\-re}
\fmtSense{to stop, to end}
\fmtExampleRL{Ka\-nga okwi\-za\-nga e wa\-nge.}
\fmtExampleEN{Stop coming to my home.}

\fmtLemma{ka\-nga\-ba\-la}
\fmtCitation{(oku)\-ka\-nga\-ba\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-nga\-bai\-re}
\fmtSense{to stand on tiptoes}
\fmtExampleRL{Ka\-nga\-ba\-la otoo\-le\-yo ekii\-sa\-ni ki\-di.}
\fmtExampleEN{Stand on tiptoes and get that photograph.}

\fmtLemma{ka\-ngai\-ta\-na}
\fmtCitation{(oku)\-ka\-ngai\-ta\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-ngai\-tai\-ne}
\fmtSense{to stagger, to walk unsteadily}
\fmtExampleRL{Omu\-ka\-li aku\-ka\-ngai\-ta\-na nka ata\-mii\-re.}
\fmtExampleEN{The woman is staggering as though she is drunk.}

\fmtLemma{ka\-ngai\-tu\-ka}
\fmtCitation{(oku)\-ka\-ngai\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-ngai\-tu\-ki\-re}
\fmtSense{to stumble, to trip}
\fmtExampleRL{Omwa\-na aka\-ngai\-tu\-ki\-re n’a\-gwa.}
\fmtExampleEN{The child has stumbled and fallen down.}

\fmtLemma{ka\-nga\-la}%
\fmtOne{1}
\fmtCitation{(a)ka\-nga\-la}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{skull}
\fmtExampleRL{Mboi\-ne aka\-nga\-la ka nke\-mbo.}
\fmtExampleEN{I have seen a monkey skull.}

\fmtLemma{ka\-nga\-la}%
\fmtOne{2}
\fmtCitation{(o)ka\-nga\-la}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from Ar. via Sw.}
\fmtSource{from archaic}
\fmtSense{gin}
\fmtExampleRL{Oka\-nga\-la aca\-mu\-la obwo\-ngo.}
\fmtExampleEN{Gin stimulates the brain.}

\fmtLemma{ka\-nge}
\fmtCrossreference{\textnormal{see} ange}

\fmtLemma{ka\-ngi\-rya}
\fmtCitation{(oku)\-ka\-ngi\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-ngii\-rye}
\fmtSense{to stop, to pause}
\fmtExampleRL{So\-ma oka\-ngi\-rye ani.}
\fmtExampleEN{Read and stop here.}

\fmtLemma{ka\-ngu\-la}
\fmtCitation{(oku)\-ka\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-ngwi\-re}
\fmtSense{to speak loudly and clearly}
\fmtExampleRL{Ka\-ngu\-la oku ira\-ka tu\-wuu\-re.}
\fmtExampleEN{Speak loudly and clearly for us to hear.}

\fmtLemma{ka\-ngu\-ra}
\fmtPOS{adv}
\fmtSense{frankly, honestly, straightforwardly}
\fmtExampleRL{Mu\-si\-moo\-le\-nge ka\-ngu\-ra.}
\fmtExampleEN{You should always speak frankly.}

\fmtLemma{ka\-ni}
\fmtCrossreference{\textnormal{see} ni{\textnormal{\fmtOne{2}}}}

\fmtLemma{ka\-nia\-la}
\fmtCitation{(a)ka\-nia\-la}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{premature banana}
\fmtExampleRL{Obu\-nia\-la ti\-bu\-nu\-ra me\-re.}
\fmtExampleEN{Premature bananas are not good for food.}

\fmtLemma{ka\-nii\-no}
\fmtCitation{(a)ka\-nii\-no}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{uphill}
\fmtExampleRL{Ta\-mbu\-la nga olee\-ge\-re oku ka\-nii\-no ako okwi\-za ku\-bo\-na ewoo\-tee\-ri.}
\fmtExampleEN{Walk straight uphill, you will see the hotel.}

\fmtLemma{ka\-ni\-ka}
\fmtCitation{(oku)\-ka\-ni\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-ni\-ki\-re}
\fmtSense{to repair, to fix, to mend}
\fmtExampleRL{Emo\-to\-ka yai\-swe ba\-gi\-ka\-ni\-ki\-re ku\-sai.}
\fmtExampleEN{Our car was well repaired.}

\fmtLemma{ka\-ni\-ndi}
\fmtCitation{(a)ka\-ni\-ndi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{spring, source, wellspring}
\fmtExampleRL{Amai\-zi ga\-zwa omu bu\-ni\-ndi.}
\fmtExampleEN{Water flows from springs.}

\fmtLemma{ka\-ni\-sa}
\fmtCitation{(e)ka\-ni\-sa}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Ar. via Sw.}
\fmtSense{church}
\fmtExampleRL{Nko\-be\-ra ei\-ba\-ra lya ka\-ni\-sa yaa\-nywe.}
\fmtExampleEN{Tell me the name of your church.}

\fmtLemma{ka\-nja}
\fmtCitation{(oku)\-ka\-nja}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-nji\-rye}
\fmtSense{to stop, to quit, to cease}
\fmtExampleRL{Ka\-nja oku\-yi\-rii\-ka ata\-li nso\-nga oli mu\-ntu mu\-ku\-lu.}
\fmtExampleEN{Stop quarrelling for no reason, you are an adult.}

\fmtLemma{ka\-nka\-na}
\fmtCitation{(oku)\-ka\-nka\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-nkai\-ne}
\fmtSense{to tremble, to shake, to shiver}
\fmtExampleRL{Lwa\-ki oku\-ka\-nka\-na?}
\fmtExampleEN{Why are you trembling?}

\fmtLemma{ka\-nkei\-re}
\fmtPOS{adj}
\fmtSense{huge, enormous}
\fmtExampleRL{Omu\-sai\-za ata n’o\-ka\-nkei\-re wa nyu\-mba.}
\fmtExampleEN{A man has a huge house.}

\fmtLemma{ka\-nki\-ka\-su\-ke\-yo}
\fmtCitation{(o)ka\-nki\-ka\-su\-ke\-yo}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{grinding stone}
\fmtExampleRL{Oka\-nki\-ka\-su\-ke\-yo oyo aku\-sya ku\-sai.}
\fmtExampleEN{That grinding stone grinds well.}

\fmtLemma{ka\-nsa}
\fmtCitation{(o)ka\-nsa}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{cancer}
\fmtExampleRL{Oka\-nsa akwi\-ta aba\-na\-yu\-ga\-nda bai\-ngi ebi\-ro bi\-ni.}
\fmtExampleEN{Cancer is killing many Ugandans these days.}

\fmtLemma{ka\-nsu\-ka}
\fmtCitation{(oku)\-ka\-nsu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-nsu\-ki\-re}
\fmtSense{to be grumpy, to be bad-temperate, to be irritable}
\fmtExampleRL{Ye ati\-bu\-ni lwa\-ki oku\-ka\-nsu\-ka na\-nje?}
\fmtExampleEN{Why are you grumpy towards me?}

\fmtLemma{ka\-nsu\-su\-le}
\fmtCitation{(a)ka\-nsu\-su\-le}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{pinky finger, little finger}
\fmtExampleRL{Aka\-nsu\-su\-le ka\-ku\-ndu\-ma.}
\fmtExampleEN{My little finger hurts.}

\fmtLemma{ka\-nsu\-swe}
\fmtCitation{(a)ka\-nsu\-swe}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{bud}
\fmtExampleRL{Eki\-me\-ra kya\-nge ki\-ree\-te\-re aka\-nsu\-nswe.}
\fmtExampleEN{My plant has developed a bud.}

\fmtLemma{ka\-nta}
\fmtCitation{(oku)\-ka\-nta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-nti\-re}
\fmtSense{to make a clicking sound in order to communicate with an animal (\eg a cow when it is being milked)}
\fmtExampleRL{Naa\-ni oyo aku\-ka\-nta?}
\fmtExampleEN{Who is making a clicking sound?}

\fmtLemma{ka\-ntii\-ni}
\fmtCitation{(e)ka\-ntii\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{canteen, cafeteria}
\fmtExampleRL{Ya\-ba oku ka\-ntii\-ni ogu\-le oswi\-ti.}
\fmtExampleEN{Go to the canteen and buy sweets.}

\fmtLemma{ka\-ntu}
\fmtCitation{(a)ka\-ntu}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{mental problem}
\fmtExampleRL{Mu\-ka\-li\-we oli\-ku aka\-ntu?}
\fmtExampleEN{Woman, do you have a mental problem?}

\fmtLemma{ka\-nuu\-nu}
\fmtCrossreference{\textnormal{see} kanuunwai}

\fmtLemma{ka\-nuu\-nwai}
\fmtCitation{(a)ka\-nuu\-nwai}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{nectar sack}
\fmtExampleRL{Eki\-mu\-li ki\-bbaa\-ku aka\-nuu\-nwai.}
\fmtExampleEN{A flower has a nectar sack.}

\fmtLemma{ka\-nya}
\fmtCitation{(oku)\-ka\-nya}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{ka\-nyi\-re}
\fmtSubentry(a)
\fmtSense{to be many, to accumulate, to increase}
\fmtExampleRL{Tu\-ku\-tu\-ku zi\-ka\-nyi\-re omu Uga\-nda ebi\-ro bi\-ni.}
\fmtExampleEN{These days three-wheeled motorcycles have become popular in Uganda.}
\fmtSubentry(b)
\fmtSense{to persist, to continue, to carry on, to go on}
\fmtExampleRL{Ka\-nya ku\-ko\-la okwi\-za ku\-bba ku\-sai.}
\fmtExampleEN{Continue working and you will be successful.}
\fmtSubentry(c)
\fmtSense{to accumulate, to gather, to collect}
\fmtExampleRL{Omwa\-mi aka\-nyi\-rye ei\-tu\-ngo li\-ngi.}
\fmtExampleEN{The chief has accumulated a lot of wealth.}

\fmtLemma{ka\-nya\-la\-la}
\fmtCrossreference{\textnormal{see} kanyala}

\fmtLemma{ka\-nya\-mu\-ntu}
\fmtCitation{(a)ka\-nya\-mu\-ntu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{mask (worn as a form of disguise)}
\fmtExampleRL{Ye\-bbwe\-re aka\-nya\-mu\-ntu omu bu\-syo.}
\fmtExampleEN{He is wearing a mask.}

\fmtLemma{ka\-nya\-ngo}
\fmtCitation{(a)ka\-nya\-ngo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{dry spell}
\fmtExampleRL{Ewai\-swe aka\-nya\-ngo ka\-tu\-li ku\-bbi e Lwa\-mpa\-nga.}
\fmtExampleEN{The dry spell is treating us badly in Lwampanga village.}
\fmtSubentry(b)
\fmtSense{late morning bright sunlight}
\fmtExampleRL{Ya\-ta\-mbwi\-re mu ka\-nya\-ngo.}
\fmtExampleEN{He left during the late morning.}

\fmtLemma{ka\-nya\-ngo\-nya\-ngo}
\fmtCitation{(a)ka\-nya\-ngo\-nya\-ngo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{intermittent sunshine}
\fmtExampleRL{Akwo\-ta ka\-nya\-ngo\-nya\-ngo.}
\fmtExampleEN{She is busking in the intermittent sunshine.}

\fmtLemma{ka\-nya\-ta}
\fmtCitation{(a)ka\-nya\-ta}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{frenulum of tongue, lingual frenulum}
\fmtExampleRL{Omwa\-na baa\-mu\-saa\-re aka\-nya\-ta kaa\-mwe.}
\fmtExampleEN{They cut the child's lingual frenulum.}

\fmtLemma{ka\-nye\-nyee\-zi}
\fmtCitation{(a)ka\-nye\-nyee\-zi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{fluorescent beetle, firefly}
\fmtExampleRL{Aka\-nye\-nyee\-zi kaa\-ka obwi\-re.}
\fmtExampleEN{The firefly emits light at night.}

\fmtLemma{ka\-nyi\-sya}
\fmtCitation{(oku)\-ka\-nyi\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-nyi\-syi\-re}
\fmtSense{to have a considerable amount, to have a considerable quantity}
\fmtExampleRL{Omu\-yu\-nja\-ga\-to ta\-ka\-nyi\-sya ma\-too\-ke mu lu\-su\-ku.}
\fmtExampleEN{A person who harvests young matoke cannot have food in plenty.}

\fmtLemma{ka\-nyu\-ga}
\fmtCitation{(oku)\-ka\-nyu\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-nyu\-gi\-re}
\fmtSense{to throw (\eg a ball)}
\fmtExampleRL{Nka\-nyu\-gi\-ra omu\-pii\-ra ogwo.}
\fmtExampleEN{Throw that ball to me.}

\fmtLemma{ka\-nzu}
\fmtCitation{(e)ka\-nzu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Ar. via Sw.}
\fmtSense{kanzu (a white or cream coloured robe worn by men in the African Great Lakes region, a traditional attire of the Baganda introduced to the Buganda Kingdom by Arab traders)}
\fmtExampleRL{Omu\-si\-raa\-mu azwai\-re eka\-nzu.}
\fmtExampleEN{The Muslim is dressed in a kanzu.}

\fmtLemma{ka\-pa}%
\fmtOne{1}
\fmtCitation{(o)ka\-pa}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSource{from Sw.}
\fmtSense{cat}
\fmtExampleRL{Ya\-boi\-ne oka\-pa ei\-go\-lo nga aku\-lya ome\-se.}
\fmtExampleEN{He saw a cat eating a rat yesterday.}

\fmtLemma{ka\-pa}%
\fmtOne{2}
\fmtCitation{\textnormal{(e.g.}\,gi\-ka\-pa, bi\-ka\-pa, mu\-ka\-pa\textnormal{)}}
\fmtPOS{adj}
\fmtSense{lean}
\fmtExampleRL{Nsa\-ngi\-re\-yo nya\-ma gi\-ka\-pa yo\-nkai.}
\fmtExampleEN{I found only lean meat.}

\fmtLemma{ka\-pa\-nga}
\fmtCitation{(o)ka\-pa\-nga}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{\textit{Lantana camara} (species of flowering plant within the verbena family)}
\fmtExampleRL{Oka\-pa\-nga aku\-la ma\-ngu.}
\fmtExampleEN{\textit{Lantana camara} grows very fast.}

\fmtLemma{ka\-pa\-nka}
\fmtCrossreference{\textnormal{see} kapanga}
\fmtCitation{(o)ka\-pa\-nka}
\fmtPOS{n}%
\fmtNC{1a/10}

\fmtLemma{ka\-pe\-nda}
\fmtCitation{(o)ka\-pe\-nda}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSubentry(a)
\fmtSense{pigeon pea (\textit{Cajanus cajan})}
\fmtExampleRL{Ka\-pe\-nda alu\-mi\-bwa mu\-no omu Uga\-nda.}
\fmtExampleEN{Pigeon peas are widely grown in Uganda.}
\fmtSubentry(b)
\fmtFixedexpression{ka\-pe\-nda wa na\-wai\-ra}
\fmtSense{variety of ratoon pigeon pea (\textit{Cajanus cajan}) with large seeds}
\fmtExampleRL{Aba\-ntu ba\-ta\-ka mu\-no oku\-lu\-ma ka\-pe\-nda wa na\-wai\-ra ku\-bba aba\-la mu\-no.}
\fmtExampleEN{People prefer to grow ratoon pigeon peas because their yield is high.}

\fmtLemma{ka\-pi\-ci}
\fmtCitation{(o)ka\-pi\-ci}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{cabbage}
\fmtExampleRL{Omu\-ku\-bi twa\-lya gwa ka\-pi\-ci.}
\fmtExampleEN{We will eat cabbage as sauce.}

\fmtLemma{ka\-pii\-ra}
\fmtCitation{(a)ka\-pii\-ra}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{condom}
\fmtExampleRL{Ko\-le\-sya aka\-pii\-ra ku\-sai okwe\-ki\-nga endwai\-re ya na\-mu\-nywa.}
\fmtExampleEN{Use the condom effectively in order to prevent HIV/AIDS.}
\fmtSubentry(b)
\fmtSense{football (ball used in football)}
\fmtExampleRL{OKa\-ba\-ga\-mbe amai\-te oku\-ca\-nga aka\-pii\-ra.}
\fmtExampleEN{Kabagambe knows how to play football skilfully.}

\fmtLemma{ka\-pi\-ka}
\fmtCitation{(oku)\-ka\-pi\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-pi\-ki\-re}
\fmtSense{to dress tightly, to put on (many items of clothing)}
\fmtExampleRL{Omwa\-na aku\-ku\-nga ku\-bba omu\-ka\-pi\-ki\-re mu\-no.}
\fmtExampleEN{The child is crying because you have dressed him tightly.}

\fmtLemma{ka\-pi\-mpi\-du}
\fmtCitation{(o)ka\-pi\-mpi\-du}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{short person}
\fmtExampleRL{Bo\-na ka\-pi\-mpi\-du ka mu\-sai\-za ka\-ku\-bi\-na.}
\fmtExampleEN{Look at the very short man dancing.}

\fmtLemma{Ka\-pi\-mpi\-ni}
\fmtCitation{(o)Ka\-pi\-mpi\-ni}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{type of god, spirit}
\fmtExampleRL{OKa\-pi\-mpi\-ni omwe oku mba\-dwa z’Aba\-nya\-la.}
\fmtExampleEN{Kapimpini is one of the gods among the Banyala.}

\fmtLemma{ka\-pi\-pi}
\fmtCitation{(a)ka\-pi\-pi}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{vengeance, revenge, retaliation}
\fmtExampleRL{Nku\-tee\-ka\-tee\-ka okwe\-ma\-la\-ku aka\-pi\-pi ke mu\-li\-na\-ku.}
\fmtExampleEN{I am planning to wreak vengeance on him.}

\fmtLemma{ka\-pi\-ri\-pii\-yo}%
\fmtOne{1}
\fmtCitation{(a)ka\-pi\-ri\-pii\-yo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{drill}
\fmtExampleRL{Ko\-le\-sya aka\-pi\-ri\-pii\-yo ote eki\-tu\-li omu lu\-baa\-wo.}
\fmtExampleEN{Use the drill and make a hole in the timber.}
\fmtSubentry(b)
\fmtSense{traditional blender}
\fmtExampleRL{Aka\-pi\-ri\-pii\-yo ka\-ko\-le\-se\-bwa oku\-cwa\-cwa\-na ebi\-yi\-mba.}
\fmtExampleEN{A traditional blender is used to blend beans.}
\fmtSubentry(c)
\fmtSense{needle for sewing machine}
\fmtExampleRL{Taa\-ku aka\-pi\-ri\-pii\-yo aka\-ndi oku kya\-laa\-ni.}
\fmtExampleEN{Insert another needle in the sewing machine.}

\fmtLemma{ka\-pi\-ri\-pii\-yo}%
\fmtOne{2}
\fmtCitation{(a)ka\-pi\-ri\-pii\-yo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{lighter}
\fmtExampleRL{Aba\-ntu ba mu Ka\-mpa\-la ba\-ko\-le\-sya bu\-pi\-ri\-pii\-yo oku\-ko\-lee\-rya osi\-ga\-la.}
\fmtExampleEN{People in Kampala use lighters to light cigarettes.}

\fmtLemma{ka\-pu\-ka}
\fmtCitation{(oku)\-ka\-pu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-pu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to die suddenly or unexpectedly}
\fmtExampleRL{Omu\-sai\-za ya\-ka\-pu\-kii\-re omu ka\-ta\-le.}
\fmtExampleEN{The man died suddenly at the market.}
\fmtSubentry(b)
\fmtSense{to break suddenly (\eg of a tree)}
\fmtExampleRL{Omu\-saa\-le gu\-ka\-pu\-ki\-re.}
\fmtExampleEN{The tree suddenly broke.}
\fmtSubentry(c)
\fmtSense{to speak anyhow}
\fmtExampleRL{Mwo\-jo le\-ke\-ra\-wo oka\-pu\-ka mu\-no.}
\fmtExampleEN{My friend, stop speaking anyhow.}

\fmtLemma{ka\-pu\-la}
\fmtCitation{(oku)\-ka\-pu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-pwi\-re}
\fmtSubentry(a)
\fmtSense{to spread}
\fmtExampleRL{Le\-ke\-ra\-wo oku\-ka\-pu\-la ebi\-ga\-mbo oku ba\-kwe\-nda ba mu\-ka\-ma.}
\fmtExampleEN{Stop spreading lies about clergymen.}
\fmtSubentry(b)
\fmtFixedexpression{ka\-pu\-la (o)lu\-yi}
\fmtSense{to slap hard, to smack hard, to strike hard}
\fmtExampleRL{Mmu\-ka\-pwi\-re olu\-yi.}
\fmtExampleEN{I slapped her hard.}

\fmtLemma{ka\-pu\-ndo}
\fmtCitation{(a)ka\-pu\-ndo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{mineral lick, salt lick}
\fmtExampleRL{Twa\-la ente ezo omu ka\-pu\-ndo.}
\fmtExampleEN{Take the cows to the salt lick.}

\fmtLemma{ka\-pu\-tee\-ni}
\fmtCitation{(o)ka\-pu\-tee\-ni}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from En.}
\fmtSense{captain}
\fmtExampleRL{Naa\-ni oka\-pu\-tee\-ni wa tii\-mu yaa\-nywe?}
\fmtExampleEN{Who is your team captain?}

\fmtLemma{ka\-ra\-bwa\-ku}
\fmtCitation{(o)ka\-ra\-bwa\-ku}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{mentally disturbed person}
\fmtExampleRL{Oka\-ra\-bwa\-ku ai\-ti\-re omwa\-na.}
\fmtExampleEN{A mentally disturbed person killed a child.}

\fmtLemma{ka\-rai}
\fmtPOS{adv}
\fmtSense{early, early in the day, ahead of time}
\fmtExampleRL{Ya\-ba e Ka\-mpa\-la ka\-rai.}
\fmtExampleEN{Go to Kampala early in the day.}

\fmtLemma{ka\-ra\-nga}
\fmtCitation{(oku)\-ka\-ra\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-ra\-ngi\-re}
\fmtSense{to roast (grains) in husks}
\fmtExampleRL{Ka\-ra\-nga ebi\-nyee\-bwa tu\-lye.}
\fmtExampleEN{Roast the groundnuts and let's eat.}

\fmtLemma{ka\-ra\-ngu\-ka}
\fmtCitation{(oku)\-ka\-ra\-ngu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-ra\-ngu\-ki\-re}
\fmtSense{to fight hard, to work extremely hard}
\fmtExampleRL{We\-ka\-ra\-ngu\-le nka aku\-ni\-ga ombwe\-ne.}
\fmtExampleEN{Fight as hard as someone strangling a dog. (a proverb)}

\fmtLemma{ka\-ra\-pyo}
\fmtCitation{(a)ka\-ra\-pyo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{pestle for mashing soft food}
\fmtExampleRL{Mpee\-rya aka\-ra\-pyo ngoi\-te eme\-re.}
\fmtExampleEN{Give me the pestle and I'll mash the food.}

\fmtLemma{ka\-ro\-loi\-so}
\fmtCitation{(a)ka\-ro\-loi\-so}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{astonishment, shock, wonder}
\fmtExampleRL{Ya\-mu\-loo\-rei\-re n’a\-ka\-ro\-loi\-so.}
\fmtExampleEN{He looked at him in astonishment.}

\fmtLemma{ka\-ro\-ti}
\fmtCitation{(e)ka\-ro\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{carrot}
\fmtExampleRL{Eka\-ro\-ti zi\-woo\-me\-sya omu\-ku\-bi.}
\fmtExampleEN{Carrots make the sauce tasty.}

\fmtLemma{ka\-ru}
\fmtCitation{\textnormal{(e.g.}\,mu\-ka\-ru, ba\-ka\-ru\textnormal{)}}
\fmtPOS{adj}
\fmtSense{mean, miserly, stingy}
\fmtExampleRL{Ti\-nta\-ka ku\-swe\-rwa omu\-sai\-za omu\-ka\-ru.}
\fmtExampleEN{I don't want to marry a mean man.}

\fmtLemma{ka\-ru\-ma}
\fmtCitation{(a)ka\-ru\-ma}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{menstrual pain, cramps}
\fmtExampleRL{Aka\-ru\-ma ka\-te\-ra ku\-mwi\-zi\-ra nga ali mu kwe\-zi.}
\fmtExampleEN{She usually gets cramps during menstruation.}

\fmtLemma{ka\-ru\-mba}
\fmtCitation{(o)ka\-ru\-mba}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{type of wasp (black and small)}
\fmtExampleRL{Oka\-ru\-mba ane\-ne\-re omu\-sai\-za.}
\fmtExampleEN{A wasp stung the man.}

\fmtLemma{ka\-ru\-ndi\-ru\-ndi}
\fmtCitation{(a)ka\-ru\-ndi\-ru\-ndi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{rod of a bicycle peddle}
\fmtExampleRL{OSe\-mba\-tya ya\-guu\-re aka\-ru\-ndi\-ru\-ndi aka\-ya.}
\fmtExampleEN{Sembatya bought a new bicycle peddle-rod.}

\fmtLemma{ka\-ruu\-kuu\-le}
\fmtCitation{(a)ka\-ruu\-kuu\-le}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{danger, problem}
\fmtExampleRL{Atu\-lee\-tei\-re obu\-ruu\-kuu\-le.}
\fmtExampleEN{He has caused us problems.}

\fmtLemma{ka\-saa\-le}
\fmtCitation{(a)ka\-saa\-le}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{branch, stick}
\fmtExampleRL{OSu\-sa\-n ya\-ku\-mbai\-nye obu\-saa\-le n’a\-ya\-kya omuu\-ro.}
\fmtExampleEN{Susan gathered sticks and lit a fire.}
\fmtSubentry(b)
\fmtSense{arrow}
\fmtExampleRL{Omwi\-bbi ba\-mu\-la\-si\-re aka\-saa\-le.}
\fmtExampleEN{The thief was shot with an arrow.}

\fmtLemma{ka\-saa\-na}
\fmtCitation{(a)ka\-saa\-na}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{\textit{Acacia hockii} (a widespread acacia shrub or small tree}
\fmtExampleRL{Aka\-saa\-na ba\-ko\-ko\-le\-sya omu mi\-ko\-lo gya bya\-bu\-wa\-ngwa.}
\fmtExampleEN{\textit{Acacia hockii} is used in some cultural ceremonies.}

\fmtLemma{ka\-saa\-saa\-yi}
\fmtCitation{(o)ka\-saa\-saa\-yi}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{maroon colour}
\fmtExampleRL{Yu\-ni\-foo\-mu yaa\-bwe ya ka\-saa\-yi\-saa\-yi.}
\fmtExampleEN{Their uniform is maroon.}

\fmtLemma{Ka\-saa\-to}
\fmtCitation{(o)Ka\-saa\-to}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{name of a mountain in Kamuli district believed to the home for the spirit called Kasaato}

\fmtLemma{ka\-saa\-yi}
\fmtCitation{(o)ka\-saa\-yi}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{red colour}
\fmtExampleRL{Esaa\-ti yaa\-mwe ya ka\-saa\-yi.}
\fmtExampleEN{His shirt is red.}

\fmtLemma{ka\-sa\-mbai\-tu\-ko}
\fmtCitation{(a)ka\-sa\-mbai\-tu\-ko}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{commotion, turmoil, chaos}
\fmtExampleRL{Wa\-gwi\-re\-wo aka\-sa\-mbai\-tu\-ko omu ki\-bu\-ga.}
\fmtExampleEN{A commotion erupted in the city.}

\fmtLemma{ka\-sa\-na}
\fmtCitation{(a)ka\-sa\-na}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{sunshine}
\fmtExampleRL{Aka\-sa\-na ka\-ta\-ndi\-ki\-re okwe\-me\-mu\-la.}
\fmtExampleEN{The sunshine became bright.}

\fmtLemma{ka\-sa\-ndo}
\fmtCitation{(a)ka\-sa\-ndo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{sap}
\fmtExampleRL{Omu\-tu\-ba gu\-bbaa\-ko gu\-zwa\-ku obu\-sa\-ndo bwi\-ngi.}
\fmtExampleEN{The mutuba tree produces a lot of sap.}

\fmtLemma{Ka\-sa\-nga\-ki}
\fmtCitation{(o)Ka\-sa\-nga\-ki}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{name given to a child born after the death of a grandparent}

\fmtLemma{ka\-sa\-nju}%
\fmtOne{1}
\fmtPOS{num}
\fmtSense{seven thousand}
\fmtExampleRL{Olu\-go\-ye olwo na\-lu\-guu\-re si\-ri\-ngi ka\-sa\-nju.}
\fmtExampleEN{I bought that cloth for seven thousand shillings.}

\fmtLemma{ka\-sa\-nju}%
\fmtOne{2}
\fmtCitation{(a)ka\-sa\-nju}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{forced labour}
\fmtExampleRL{Aba\-si\-be ba\-ba\-ko\-le\-sya ka\-sa\-nju.}
\fmtExampleEN{The prisoners are made to do forced labour.}

\fmtLemma{ka\-sa\-nvu}
\fmtCrossreference{\textnormal{see} kasanju}

\fmtLemma{ka\-sa\-nyi\-re}
\fmtCitation{(o)ka\-sa\-nyi\-re}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{famine}

\fmtLemma{ka\-sa\-ra}
\fmtCitation{(o)ka\-sa\-ra}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{cow with horns hanging downwards}
\fmtExampleRL{Oka\-sa\-ra abo\-ne\-ka ku\-sai.}
\fmtExampleEN{The cow with horns hanging downwards looks nice.}

\fmtLemma{ka\-sa\-si\-ro}
\fmtCitation{(o)ka\-sa\-si\-ro}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{trash, rubbish}
\fmtExampleRL{Omu Ka\-mpa\-la mu\-li\-mu oka\-sa\-si\-ro mwi\-ngi.}
\fmtExampleEN{There is a lot of trash in Kampala.}

\fmtLemma{ka\-sa\-tu}
\fmtPOS{num}
\fmtSubentry(a)
\fmtSense{third}
\fmtExampleRL{Gu\-ni mwa\-ka gwa ka\-sa\-tu nga aku\-so\-ma oku yu\-ni\-va\-si\-te.}
\fmtExampleEN{This is her third year at the university.}
\fmtSubentry(b)
\fmtFixedexpression{mwe\-zi gwa ka\-sa\-tu}
\fmtSense{March}

\fmtLemma{ka\-sa\-wu}
\fmtCitation{(a)ka\-sa\-wu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{money purse, wallet}
\fmtExampleRL{Ya\-nya\-gi\-re aka\-sa\-wu ka\-nge n’ai\-ru\-ka.}
\fmtExampleEN{He grabbed my money purse and ran away.}

\fmtLemma{ka\-se\-de}
\fmtCitation{(a)ka\-se\-de}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{sweet potato slice}
\fmtExampleRL{Nku\-ta\-ka oku\-lya oku ka\-se\-de.}
\fmtExampleEN{I feel like eating dried potato.}

\fmtLemma{ka\-see\-ra}
\fmtCitation{(a)ka\-see\-ra}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{short period of time}
\fmtSubentry(b)
\fmtFixedexpression{bu\-li ka\-see\-ra}
\fmtSense{every time, always}
\fmtExampleRL{Owa\-bi\-sa\-nzi\-re ase\-ke\-sya aba\-ntu bu\-li ka\-see\-ra.}
\fmtExampleEN{A joker always makes people laugh.}

\fmtLemma{ka\-see\-se}
\fmtCitation{(o)ka\-see\-se}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{type of crude gin, it used to come from Kaseese}
\fmtExampleRL{Ka\-see\-se mwe\-nge ogwa\-li\-gi\-re omu di\-si\-tu\-li\-ki\-ti ya Ka\-see\-se.}
\fmtExampleEN{Kaseese gin originated in the Kaseese district.}

\fmtLemma{ka\-se\-ke\-ndi}
\fmtCitation{(a)ka\-se\-ke\-ndi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{second (a sixtieth of a minute of time)}
\fmtExampleRL{Waa\-kaa\-ka\-la\-wo obu\-se\-ke\-ndi mu\-sa\-nju.}
\fmtExampleEN{Seven seconds have passed.}

\fmtLemma{ka\-se\-ke\-ra\-nyo\-nyi}
\fmtCitation{(a)ka\-se\-ke\-ra\-nyo\-nyi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{\textit{Lantana trifolia} (herbal plant with small sweet purple berries used to treat hiccups)}

\fmtLemma{ka\-se\-ke\-re}
\fmtCitation{(a)ka\-se\-ke\-re}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{sty (an inflamed swelling on the edge of an eyelid caused by bacterial infection)}
\fmtExampleRL{Aka\-se\-ke\-re ka\-ku\-ndu\-ma mu\-no.}
\fmtExampleEN{The sty hurts me a lot.}

\fmtLemma{ka\-se\-mbo}
\fmtCrossreference{\textnormal{see also} mutaragi}
\fmtCitation{(a)ka\-se\-mbo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{token of appreciation, incentive, gratuity}
\fmtExampleRL{Ya\-koo\-re ku\-sai ni mmu\-wee\-rya\-yo aka\-se\-mbo.}
\fmtExampleEN{He performed well and I gave him a token of my appreciation.}

\fmtLemma{ka\-se\-ngei\-za}
\fmtCitation{(a)ka\-se\-ngei\-za}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{strainer}
\fmtExampleRL{Se\-ngei\-ze\-sya ka\-se\-ngei\-za.}
\fmtExampleEN{Filter it with the strainer.}

\fmtLemma{ka\-se\-nse\-ke\-ra}
\fmtCitation{(o)ka\-se\-nse\-ke\-ra}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{kind of large flying ant capable of inflicting a serious sting, type of pest which attacks coffee plants (It is black, small and it stings.)}

\fmtLemma{ka\-se\-re\-ge\-te}
\fmtCitation{(a)ka\-se\-re\-ge\-te}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{small grey termite}
\fmtExampleRL{Obu\-se\-re\-ge\-te bu\-lya mu\-no emi\-saa\-le.}
\fmtExampleEN{Grey termites eat a lot of trees.}

\fmtLemma{ka\-se\-re\-nge\-to}
\fmtCitation{(a)ka\-se\-re\-nge\-to}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{slope, decline}
\fmtExampleRL{Aba\-ntu ba\-fu\-na mu\-no obu\-ta\-ndwa oku ka\-se\-re\-nge\-to ka\-di.}
\fmtExampleEN{Many people have an accident on that slope.}

\fmtLemma{ka\-sii\-bu\-lii\-ryo}
\fmtCitation{(a)ka\-sii\-bu\-lii\-ryo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{farewell, goodbye}
\fmtExampleRL{Aku\-sa\-mba mu\-pii\-ra gwa ka\-sii\-bu\-lii\-ryo.}
\fmtExampleEN{He is playing a farewell football game.}

\fmtLemma{ka\-sii\-di}
\fmtCitation{(a)ka\-sii\-bu\-lii\-ryo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{stubborness, obstinacy}
\fmtExampleRL{Omwa\-na oni ata n’a\-ka\-sii\-di.}
\fmtExampleEN{This child is pig-headed.}

\fmtLemma{ka\-sii\-mo}
\fmtCrossreference{\textnormal{see also} kasembo}
\fmtCitation{(a)ka\-sii\-mo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{token of appreciation, incentive, gratuity}
\fmtExampleRL{Aba\-zi\-rwai\-ne\-ku ba\-ba\-wai\-re aka\-sii\-mo kaa\-bwe.}
\fmtExampleEN{The war veterans were given a token of appreciation.}

\fmtLemma{ka\-sii\-sa}%
\fmtOne{1}
\fmtCitation{(a)ka\-sii\-sa}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{insect, bug}
\fmtExampleRL{Mo\-do\-la erii\-so nku\-too\-le\-mu aka\-sii\-sa.}
\fmtExampleEN{Open your eye so that I can remove the insect.}
\fmtSubentry(b)
\fmtSense{bacterium, germ, microbe}
\fmtExampleRL{Obu\-sii\-sa bu\-kwi\-za ku\-kwi\-ngi\-ra omu kyo\-ya ni waa\-bba nga to\-ki\-swe\-ke\-re\-ku ku\-sai.}
\fmtExampleEN{Bacteria will enter your wound if you don't cover it properly.}
\fmtSubentry(c)
\fmtSense{virus}
\fmtExampleRL{Omu Uga\-nda mu\-li\-mu obu\-sii\-sa obwa mu\-taa\-wa\-na bwi\-ngi.}
\fmtExampleEN{There are many deadly viruses in Uganda.}
\fmtSubentry(d)
\fmtSense{HIV virus}
\fmtExampleRL{Aka\-sii\-sa ka na\-mu\-nywa ka\-lee\-ta osi\-rii\-mu.}
\fmtExampleEN{HIV causes AIDS.}

\fmtLemma{ka\-sii\-sa}%
\fmtOne{2}
\fmtCitation{(a)ka\-sii\-sa}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{penis}
\fmtExampleRL{Aka\-sii\-sa ka mwa\-na ka\-li\-ku eki\-rwai\-re.}
\fmtExampleEN{The child’s penis is infected.}

\fmtLemma{ka\-sii\-si\-ra}
\fmtCitation{(a)ka\-sii\-si\-ra}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{hut}
\fmtExampleRL{Asu\-la mu ka\-sii\-si\-ra.}
\fmtExampleEN{He sleeps in a hut.}

\fmtLemma{ka\-sii\-te}
\fmtCitation{(a)ka\-sii\-te}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{(prison) cell}
\fmtExampleRL{Ba\-mu\-twai\-re mu ka\-sii\-te.}
\fmtExampleEN{He was taken to the cell.}

\fmtLemma{ka\-sii\-ya\-ni}
\fmtCitation{(a)ka\-sii\-ya\-ni}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{nymphomania (excessive sexual desire in a woman)}
\fmtExampleRL{Omu\-ka\-li wa ka\-sii\-ya\-ni ta\-ke\-rwa.}
\fmtExampleEN{A woman with nymphomania does not get sexually satisfied.}

\fmtLemma{ka\-si\-mba}
\fmtCitation{(a)ka\-si\-mba}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{genet}
\fmtExampleRL{Aka\-si\-mba ka\-lya enko\-ko.}
\fmtExampleEN{The genet eats hens.}

\fmtLemma{ka\-si\-mu}
\fmtCitation{(a)ka\-si\-mu}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{small hoe}
\fmtExampleRL{Mpee\-rya aka\-si\-mu nsa\-ke ekii\-na.}
\fmtExampleEN{Give me a small sized hoe and I'll dig a hole.}

\fmtLemma{ka\-si\-ra\-bo}
\fmtCitation{(a)ka\-si\-ra\-bo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{oribi (a small antelope)}
\fmtExampleRL{Aka\-si\-ra\-bo ti\-nka\-bo\-na\-nga\-ko ka\-wuu\-ra\-mbe.}
\fmtExampleEN{I have never seen an oribi, I have only heard about it.}

\fmtLemma{ka\-si\-rii\-kii\-ro}%
\fmtOne{1}
\fmtCitation{(a)ka\-si\-rii\-kii\-ro}
\fmtPOS{n}%
\fmtNC{12}
\fmtSubentry(a)
\fmtSense{span of time, period, interval}
\fmtExampleRL{Aka\-si\-rii\-kii\-ro ka\-bbai\-re ka maa\-ni oku\-zwa ni twa\-si\-si\-nkai\-ne.}
\fmtExampleEN{It has been a long time since we met.}
\fmtSubentry(b)
\fmtSense{period of silence, quietness, truce, lull}
\fmtExampleRL{Nkuu\-ra\-yo aka\-si\-rii\-kii\-ro.}
\fmtExampleEN{I can hear a moment of silence.}

\fmtLemma{ka\-si\-ru\-mu\-ko}
\fmtCitation{(a)ka\-si\-ri\-mu\-ko}
\fmtPOS{n}%
\fmtNC{6}
\fmtSense{slope}
\fmtExampleRL{Mu\-le\-ke\-re nga ai\-ki\-re adi oku ka\-si\-ri\-mu\-ko.}
\fmtExampleEN{I left when he had reached the slope.}

\fmtLemma{ka\-si\-si}
\fmtCitation{(a)ka\-si\-si}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{small gourd}
\fmtExampleRL{Omwe\-nge gu\-nywe\-re mu ka\-si\-si.}
\fmtExampleEN{Take the alcohol from the small gourd.}

\fmtLemma{ka\-si\-ta}
\fmtPOS{conj}
\fmtSense{as long as, provided that}
\fmtExampleRL{Ndoo\-wo nso\-nga ka\-si\-ta waa\-bba nga wai\-ra mu bwi\-re.}
\fmtExampleEN{There is no problem as long as you come back in time.}

\fmtLemma{ka\-so\-be\-za}
\fmtCitation{(a)ka\-so\-be\-za}
\fmtPOS{n}
\fmtSense{ill behaved person}
\fmtExampleRL{Omwa\-na ka\-so\-be\-za ai\-bbi\-re se\-nte z’a\-maa\-mwe.}
\fmtExampleEN{The ill-behaved child stole his mother’s money.}
\fmtSense{wrongly, incorrectly (\eg to button wrongly)}

\fmtLemma{ka\-so\-lo}
\fmtCitation{(a)ka\-so\-lo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{three legged ladder}
\fmtExampleRL{Lee\-ta aka\-so\-lo nii\-ni\-re\-ku.}
\fmtExampleEN{Bring the three-legged ladder and I will climb.}

\fmtLemma{ka\-so\-lya}
\fmtCitation{(a)ka\-so\-lya}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{roof}
\fmtExampleRL{Aka\-nyo\-nyi kee\-me\-rei\-re oku ka\-so\-lya ka nyu\-mba.}
\fmtExampleEN{The bird is standing on the roof of the house.}

\fmtLemma{ka\-so\-me\-re\-zi}
\fmtCitation{(a)ka\-so\-me\-re\-zi}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{early morning sunshine}
\fmtExampleRL{Bu\-lii\-zo aka\-sa\-na aka\-so\-me\-re\-zi ka\-bba ka\-sai.}
\fmtExampleEN{The early morning sunshine is always good.}

\fmtLemma{ka\-so\-nga}
\fmtCitation{(a)ka\-so\-nga}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{tip}
\fmtExampleRL{Ka\-ni ni\-ko ka\-so\-nga ka isu\-mu.}
\fmtExampleEN{This is the tip of the spear.}
\fmtSubentry(b)
\fmtSense{apex (tip of a leaf)}
\fmtExampleRL{Ka\-ni ni\-ko ka\-so\-nga ka ki\-koo\-la.}
\fmtExampleEN{This is the apex of the leaf.}

\fmtLemma{ka\-so\-noo\-ko}
\fmtPOS{adv}
\fmtSense{slowly}
\fmtExampleRL{Ata\-mbu\-la ka\-soo\-no\-ko nka eki\-taa\-nii\-na.}
\fmtExampleEN{He walks slowly like a snail.}

\fmtLemma{ka\-soo\-bo}
\fmtPOS{adv}
\fmtSense{slowly}
\fmtExampleRL{Lwa\-ki oku\-ta\-mbu\-la ka\-soo\-bo?}
\fmtExampleEN{Why are you walking slowly?}

\fmtLemma{ka\-soo\-nya}%
\fmtOne{1}
\fmtCitation{(o)ka\-soo\-nya}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{the famine of 1918, which claimed many lives}
\fmtExampleRL{Ya\-fee\-rei\-re mu ka\-soo\-nya wa 1918.}
\fmtExampleEN{He died during the famine of 1918.}

\fmtLemma{ka\-soo\-nya}%
\fmtOne{2}
\fmtCitation{(a)ka\-soo\-nya}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{old-fashioned way}
\fmtExampleRL{Oku\-suu\-bu\-la aba\-ntu nko\-la ya ka\-soo\-nya.}
\fmtExampleEN{Selling people into slavery is an old-fashioned business.}

\fmtLemma{ka\-so\-roo\-ro}
\fmtCitation{(a)ka\-so\-roo\-ro}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{discrimination, intolerance}
\fmtExampleRL{Aba\-ntu aba\-li\-na aka\-so\-roo\-ro ba ka\-bbi omu ba\-ntu.}
\fmtExampleEN{Those who practice discrimination are dangerous in a society.}

\fmtLemma{ka\-so\-zi}
\fmtCitation{(a)ka\-so\-zi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{hill}
\fmtExampleRL{Aka\-so\-zi ka\-di ka\-li\-ku aka\-gu\-nguu\-li.}
\fmtExampleEN{That hill has a steep slope.}

\fmtLemma{ka\-su\-bba}
\fmtCitation{(a)ka\-su\-bba}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{tuberculosis}
\fmtExampleRL{Obu\-lwai\-re bwa ka\-su\-bba bwi\-ngi omu Uga\-nda.}
\fmtExampleEN{Tuberculosis is rampant in Uganda.}

\fmtLemma{ka\-su\-ka}
\fmtCitation{(oku)\-ka\-su\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-su\-ki\-re}
\fmtSense{to throw, to toss}
\fmtExampleRL{Nka\-su\-ki\-ra enso\-ne eyo.}
\fmtExampleEN{Throw that knife to me.}

\fmtLemma{ka\-su\-le}
\fmtCrossreference{\textnormal{see also} kasunsu}
\fmtCitation{(a)ka\-su\-le}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{crest, tuft of feathers on the head of a bird}
\fmtExampleRL{Ose\-ko\-ko abbaa\-ku aka\-su\-le.}
\fmtExampleEN{A turkey has a crest.}

\fmtLemma{ka\-su\-lu}
\fmtCitation{(o)ka\-su\-lu}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{elephant snout fish}
\fmtExampleRL{Ka\-su\-lu ki\-na\-ge ki\-koo\-to\-mu.}
\fmtExampleEN{The elephant snout fish is relatively big.}

\fmtLemma{ka\-su\-lu\-bba\-na}
\fmtCitation{(o)ka\-su\-lu\-bba\-na}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{elephant snout fish (\textit{Mormyrus ka\-nu\-me})}
\fmtExampleRL{Olwa\-lee\-ro twa\-lya ka\-su\-lu\-bba\-na.}
\fmtExampleEN{Today we will eat the elephant snout fish.}

\fmtLemma{ka\-su\-nsu}
\fmtCrossreference{\textnormal{see also} kasule}
\fmtCitation{(a)ka\-su\-nsu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{crest, tuft of feathers on the head of a bird}
\fmtExampleRL{Wai\-su\-ngee\-rya abba n’a\-ka\-su\-nsu ku mu\-twe.}
\fmtExampleEN{A hornbill has a small crest on the head.}

\fmtLemma{ka\-suu\-me}
\fmtCitation{(a)ka\-suu\-me}
\fmtPOS{n}%
\fmtNC{12}
\fmtSubentry(a)
\fmtSense{trouble, danger, problems}
\fmtExampleRL{E Mi\-gye\-ra wa\-li\-yo aka\-suu\-me.}
\fmtExampleEN{There is trouble in Migyera.}
\fmtSubentry(b)
\fmtSense{accident}
\fmtExampleRL{Aba\-ntu abaa\-gwi\-re oku ka\-suu\-me ba\-fu\-ni\-re ebi\-wu\-ta.}
\fmtExampleEN{The people who were involved in the accident sustained injuries.}

\fmtLemma{ka\-suu\-mi}
\fmtCitation{(a)ka\-suu\-mi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{small hoe used to plant especially groundnuts}
\fmtExampleRL{Ko\-le\-sya ka\-suu\-mi oku\-si\-mba amai\-do.}
\fmtExampleEN{Use a small hoe to plant groundnuts.}

\fmtLemma{ka\-suu\-ni}
\fmtCitation{(a)ka\-suu\-ni}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{small hoe, warren hoe}
\fmtExampleRL{Mpee\-rya aka\-suu\-ni nsa\-ke ekii\-na.}
\fmtExampleEN{Give me a warren hoe and I will dig a hole.}

\fmtLemma{ka\-suu\-nyo}
\fmtCitation{(a)ka\-suu\-nyo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{whip, lash}
\fmtExampleRL{Omwa\-na na\-mu\-suu\-nyi\-re obu\-suu\-nyo bu\-sa\-tu bwo\-nkai.}
\fmtExampleEN{I only whipped the child three times.}

\fmtLemma{ka\-suu\-nyu}
\fmtCrossreference{\textnormal{see} kasuu\-nyo}

\fmtLemma{ka\-suu\-ya}
\fmtCitation{(a)ka\-suu\-ya}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{moustache}
\fmtExampleRL{Ba\-mu\-ge\-mbe\-re\-ku aka\-suu\-ya.}
\fmtExampleEN{They shaved off his moustache.}

\fmtLemma{ka\-suu\-yai}
\fmtCrossreference{\textnormal{see} kasuuyi}

\fmtLemma{ka\-suu\-yi}
\fmtCitation{(a)ka\-suu\-yi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{whisker}
\fmtExampleRL{Onja\-ngwa ali\-na obu\-suu\-yi.}
\fmtExampleEN{The cat has whiskers.}

\fmtLemma{ka\-swe\-ko}
\fmtCitation{ka\-swe\-ko}
\fmtPOS{adv}
\fmtSense{evening, dusk}
\fmtExampleRL{Yai\-zi\-re ka\-swe\-ko.}
\fmtExampleEN{He came at dusk.}

\fmtLemma{ka\-swi\-ri\-ri}
\fmtCitation{(a)ka\-swi\-ri\-ri}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{whisker}
\fmtExampleRL{Onja\-ngo abba n’o\-bu\-swi\-ri\-ri.}
\fmtExampleEN{The cat has whiskers.}

\fmtLemma{ka\-sya\-tuu\-ko}
\fmtCitation{(a)ka\-sya\-tuu\-ko}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{commotion}
\fmtExampleRL{Ei\-so\-me\-ro lii\-zi\-re\-mu aka\-sya\-tuu\-ko.}
\fmtExampleEN{A commotion erupted in the school.}

\fmtLemma{ka\-sye\-ne\-ne}
\fmtPOS{adv}
\fmtSubentry(a)
\fmtSense{evening, dusk}
\fmtExampleRL{Ya\-ya\-bii\-re mu ka\-sye\-ne\-ne.}
\fmtExampleEN{He left at dusk.}
\fmtSubentry(b)
\fmtSense{young brown she-goat}
\fmtExampleRL{Ombu\-li abyai\-re ka\-sye\-ne\-ne.}
\fmtExampleEN{The goat has produced a young brown she-goat.}

\fmtLemma{ka\-syo\-lo\-ko\-ta}
\fmtCitation{(oku)\-ka\-syo\-lo\-ko\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-syo\-lo\-ko\-te\-re}
\fmtSense{(of a chameleon) to hiss}
\fmtExampleRL{Owu\-lii\-re owa\-lu\-bu n’a\-ku\-syo\-lo\-ko\-ta ?}
\fmtExampleEN{Have you heard the chameleon hissing?}

\fmtLemma{ka\-syo\-lo\-ko\-to}%
\fmtOne{1}
\fmtCitation{(a)ka\-syo\-lo\-ko\-to}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{hiss of a chameleon}
\fmtExampleRL{Owu\-lii\-re aka\-syo\-lo\-ko\-to ka wa\-lu\-bu?}
\fmtExampleEN{Have you heard the hiss of a chameleon?}

\fmtLemma{ka\-syo\-lo\-ko\-to}%
\fmtOne{2}
\fmtPOS{adv}
\fmtSense{stealthily, in a cautious manner}
\fmtExampleRL{Omu\-sai\-za ya\-ya\-bi\-re ka\-syo\-lo\-ko\-to.}
\fmtExampleEN{The man left stealthily.}

\fmtLemma{ka\-syo\-ma}
\fmtCitation{(o)ka\-syo\-ma}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{purple colour}
\fmtExampleRL{Azwai\-re ebbu\-la\-wu\-zi ya ka\-syo\-ma.}
\fmtExampleEN{She’s wearing a purple blouse.}

\fmtLemma{ka\-syo\-no}
\fmtCitation{(a)ka\-syo\-no}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{\textit{Acacia hockii} or \textit{Acacia brevispica} (widespread acacia shrub or small tree )}
\fmtExampleRL{Aka\-syo\-no ka\-ta\-mba eki\-koo\-lo.}
\fmtExampleEN{\textit{Aca\-cia hockii} cures a cough.}

\fmtLemma{ka\-taa\-go}
\fmtCitation{(a)ka\-taa\-go}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{spleen}
\fmtExampleRL{Aka\-taa\-go ka mbu\-li ka\-woo\-ma.}
\fmtExampleEN{Goat spleen is tasty.}

\fmtLemma{ka\-taa\-nu}
\fmtCrossreference{\textnormal{see also} (-)taanu}
\fmtPOS{num}
\fmtSubentry(a)
\fmtSense{fifth}
\fmtSubentry(b)
\fmtFixedexpression{mwezi gwa kataa\-nu}
\fmtSense{May}

\fmtLemma{ka\-taa\-sa}
\fmtCitation{(a)ka\-taa\-sa}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{hand wash basin}
\fmtExampleRL{Enga\-lo zi\-naa\-bi\-re mu ka\-taa\-sa.}
\fmtExampleEN{Wash your hands in the hand wash basin.}

\fmtLemma{ka\-tai\-de}
\fmtPOS{conj}
\fmtSubentry(a)
\fmtSense{since, because, as}
\fmtExampleRL{Ka\-tai\-de nka ni\-waa\-li omu lu\-kii\-ko nko\-be\-ra ebya\-lu\-mu.}
\fmtExampleEN{Since you attended the meeting, tell me what transpired.}
\fmtSubentry(b)
\fmtSense{firstly, first}
\fmtExampleRL{Ka\-tai\-de mpee\-rya\-mu akai\-re mbo\-ne eki\-kwi\-ra\-ku.}
\fmtExampleEN{First give me some time and we'll see what happens.}
\fmtSubentry(c)
\fmtSense{alright, okay}
\fmtExampleRL{Ka\-tai\-de iza tu\-bo\-ne.}
\fmtExampleEN{Okay, come and we will see.}
\fmtSubentry(d)
\fmtSense{by the way}
\fmtExampleRL{Ka\-tai\-de naa\-ni ya\-ku\-tu\-mi\-re ani?}
\fmtExampleEN{By the way, who sent you here?}

\fmtLemma{ka\-tai\-ka\-tai}
\fmtPOS{adv}
\fmtSubentry(a)
\fmtSense{repeatedly, frequently}
\fmtExampleRL{Abbai\-re ai\-za ani ka\-tai\-ka\-tai.}
\fmtExampleEN{He has been coming here frequently.}
\fmtSubentry(b)
\fmtSense{in gulps}
\fmtExampleRL{Amai\-zi aga\-nywi\-re ka\-tai\-ka\-tai.}
\fmtExampleEN{He drank the water in gulps.}

\fmtLemma{ka\-ta\-la}%
\fmtOne{1}
\fmtCitation{(a)ka\-ta\-la}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{border between two villages}
\fmtExampleRL{Mmwa\-kai\-re\-ku omu ka\-ta\-la aka\-li aka\-ti wa Rwa\-ngaa\-ga n’ oKa\-ba\-ku.}
\fmtExampleEN{I passed by him at the border between Rwangaaga and Kabaku.}

\fmtLemma{ka\-ta\-la}%
\fmtOne{2}
\fmtCitation{(e)ka\-ta\-la}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{compulsory assignment, assigned duty}
\fmtExampleRL{Ma\-li\-rii\-rye eka\-ta\-la ya\-nge.}
\fmtExampleEN{I have completed my compulsory assignment.}

\fmtLemma{ka\-ta\-la\-ta\-la}
\fmtCrossreference{\textnormal{see} katala}

\fmtLemma{ka\-ta\-le}
\fmtCitation{(a)ka\-ta\-le}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{market}
\fmtExampleRL{Ntu\-nda ngo\-ye omu ka\-ta\-le.}
\fmtExampleEN{I sell clothes at the market.}

\fmtLemma{ka\-ta\-mbaa\-la}
\fmtCitation{(a)ka\-ta\-mbaa\-la}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{headscarf}
\fmtExampleRL{Yee\-bbwe\-re aka\-ta\-mbaa\-la oku mu\-twe.}
\fmtExampleEN{She is wearing a headscarf.}
\fmtSubentry(b)
\fmtSense{handkerchief}
\fmtExampleRL{Fu\-lya omu ka\-ta\-mbaa\-la.}
\fmtExampleEN{Blow your nose using the handkerchief.}

\fmtLemma{ka\-ta\-mii\-ra\-yo\-byo}
\fmtCitation{(o)ka\-ta\-mii\-ra\-yo\-byo}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{drunkard, alcoholic}
\fmtExampleRL{Oyo ka\-ta\-mii\-ra\-yo\-byo alwa\-na mu\-no.}
\fmtExampleEN{That drunkard fights a lot.}

\fmtLemma{ka\-ta\-ndaa\-lo}
\fmtCitation{(a)ka\-ta\-ndaa\-lo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{rack for utensils}
\fmtExampleRL{Naa\-bya esaa\-ni ozi\-te oku ka\-ta\-ndaa\-lo.}
\fmtExampleEN{Wash the plates and put them on the rack.}

\fmtLemma{ka\-ta\-ra\-gi}
\fmtCrossreference{\textnormal{see also} mutaragi}
\fmtCitation{(a)ka\-ta\-ra\-gi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{\textit{Gardenia thunbergia} (white gardenia)}

\fmtLemma{ka\-ta\-rai\-ga}
\fmtCitation{(a)ka\-ta\-rai\-ga}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{type of a bird trap long enough to trap a number of birds at the same time}
\fmtExampleRL{Na\-te\-ge\-re aka\-ta\-rai\-ga ni nkwa\-ti\-sya enko\-wu.}
\fmtExampleEN{I set a bird trap and it caught a guinea fowl.}

\fmtLemma{ka\-ta\-ru\-nka\-na}
\fmtCitation{(o)ka\-ta\-ru\-nka\-na}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{wanderer, one who roams about}
\fmtExampleRL{Oka\-ta\-ru\-nka\-na ta\-ku\-laa\-ku\-la\-na.}
\fmtExampleEN{A wanderer does not prosper.}

\fmtLemma{ka\-tee\-ra}
\fmtCitation{(o)ka\-tee\-ra}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSubentry(a)
\fmtSense{hawk (common term for all birds of prey)}
\fmtExampleRL{Obu\-tee\-ra bu\-ku\-lya obu\-ko\-ko bwai\-swe ebi\-ro bi\-ni.}
\fmtExampleEN{These days hawks eat our chicks.}
\fmtSubentry(b)
\fmtSense{kite (a type of bird)}
\fmtExampleRL{Oka\-tee\-ra alii\-re aka\-ko\-ko ka\-nge.}
\fmtExampleEN{A kite has preyed on my chick.}

\fmtLemma{ka\-tee\-si\-gwa}
\fmtCitation{(o)ka\-tee\-si\-gwa}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{untrustworthy person}
\fmtExampleRL{Omu\-ka\-li oyo ka\-tee\-si\-gwa ku\-bba ta\-kuu\-ma bya\-ma.}
\fmtExampleEN{That woman is untrustworthy because she does not keep secrets.)}

\fmtLemma{ka\-tee\-ya\-mba}
\fmtCitation{(o)ka\-tee\-ya\-mba}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{person with disability, disabled person}
\fmtExampleRL{Ju\-na abaa\-na aba\-ka\-tee\-ya\-mba.}
\fmtExampleEN{Help children with disabilities.}

\fmtLemma{ka\-tei\-ca}
\fmtCitation{(o)ka\-tei\-ca}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{very poor person}
\fmtExampleRL{Oka\-tei\-ca asa\-bi\-re omu\-ki\-njaa\-ge eki\-fi ky’e\-nya\-ma.}
\fmtExampleEN{A very poor person begged the butcher for a piece of meat.}

\fmtLemma{ka\-te\-mba}
\fmtCitation{(o)ka\-te\-mba}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{theatre play, drama, performance}
\fmtExampleRL{E Na\-ka\-so\-ngo\-la ndoo\-yo ki\-fo gye ba\-zee\-nyee\-ra oka\-te\-mba.}
\fmtExampleEN{There is no place in Nakasongola where they perform theatre plays.}

\fmtLemma{ka\-te\-mee\-ro}
\fmtCitation{(a)ka\-te\-mee\-ro}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{long distance}
\fmtExampleRL{Wa\-li\-wo aka\-te\-mee\-ro okwi\-ka ewaa\-mwe.}
\fmtExampleEN{It is a long distance to reach his home.}

\fmtLemma{ka\-te\-ngee\-to}
\fmtCitation{(a)ka\-te\-ngee\-to}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{swampy ground}
\fmtExampleRL{Abaa\-na ba\-gwa omu ka\-te\-ngee\-to.}
\fmtExampleEN{Children get stuck in the swampy ground.}

\fmtLemma{ka\-te\-ngee\-tyo}
\fmtCitation{(a)ka\-te\-ngee\-tyo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{tremor, trembling, shaking}
\fmtExampleRL{Omu\-te\-nga alee\-te\-re aka\-te\-ngee\-tyo.}
\fmtExampleEN{The earthquake caused a tremor.}

\fmtLemma{ka\-te\-ngu}
\fmtCitation{(o)bu\-te\-ngu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{tattered cloth}
\fmtExampleRL{Olu\-go\-ye lwa\-mu lwa\-foo\-ke\-re bu\-te\-ngu.}
\fmtExampleEN{Your cloth is tattered.}

\fmtLemma{ka\-te\-re\-ga}
\fmtCitation{(a)ka\-te\-re\-ga}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{small dark tilapia}
\fmtExampleRL{Ti\-ndya ka\-te\-re\-ga.}
\fmtExampleEN{I don’t like eating small dark tilapia.}

\fmtLemma{ka\-ti\-ba}
\fmtCitation{(a)ka\-ti\-ba}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{frame for moulding bricks}
\fmtExampleRL{Aki\-ta\-ba aka\-ga\-zi ka\-ko\-la ebbu\-loo\-ka ezi\-koo\-to.}
\fmtExampleEN{A big moulding frame produces big bricks.}

\fmtLemma{Ka\-ti\-go}
\fmtCitation{(o)Ka\-ti\-go}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{one of the gods among Banyala/Baruuli}
\fmtExampleRL{Aba\-ndi ba\-ra\-mya Ka\-ti\-go.}
\fmtExampleEN{Other people worship Katigo.}

\fmtLemma{ka\-tii\-ki}
\fmtCitation{(a)ka\-tii\-ki}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{broken piece, fragment}
\fmtExampleRL{Aka\-tii\-ki ka gi\-raa\-si ka\-so\-bo\-la okwi\-ta omu\-ntu.}
\fmtExampleEN{A broken piece of glass can kill a person.}

\fmtLemma{ka\-tii\-ri\-bba\-lu\-ka}
\fmtPOS{adv}
\fmtSense{too much (used to describe someone who eats too much)}
\fmtExampleRL{Omu\-sai\-za oyo alya ka\-tii\-ri\-bba\-lu\-ka.}
\fmtExampleEN{The man eats too much.}

\fmtLemma{ka\-ti\-ki\-si\-mu}
\fmtCitation{(e)ka\-ti\-ki\-si\-mu}
\fmtPOS{n}%
\fmtNC{9}
\fmtSource{from En.}
\fmtSense{catechism}
\fmtExampleRL{So\-ma eka\-ti\-ki\-si\-mu yaa\-mu.}
\fmtExampleEN{Read your catechism.}

\fmtLemma{ka\-ti\-ki\-ti\-ki}
\fmtCitation{(a)ka\-ti\-ki\-ti\-ki}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{second (a sixtieth of a minute of time)}
\fmtExampleRL{Ai\-ru\-kii\-re\-wo obu\-ti\-ki\-ti\-ki iku\-mu.}
\fmtExampleEN{He covered the distance in ten seconds.}

\fmtLemma{ka\-ti\-ko}
\fmtCitation{(a)ka\-ti\-ko}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{mushroom}
\fmtExampleRL{Obu\-ti\-ko bu\-mee\-re.}
\fmtExampleEN{The mushrooms have germinated.}

\fmtLemma{ka\-ti\-ma ka gaa\-li}
\fmtCitation{(a)ka\-ti\-ma ka gaa\-li}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{valve}
\fmtExampleRL{Lee\-ta aka\-ti\-ma ka gaa\-li tu\-bo\-ne\-ku.}
\fmtExampleEN{Please bring the bicycle valve and we will have a look.}

\fmtLemma{ka\-ti\-mba}
\fmtCitation{(a)ka\-ti\-mba}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{fishing net}
\fmtExampleRL{Ako aka\-ti\-mba ka\-te\-ga ncwi.}
\fmtExampleEN{That net traps fish.}
\fmtSubentry(b)
\fmtSense{mosquito net}
\fmtExampleRL{Nsu\-la mu ka\-ti\-mba.}
\fmtExampleEN{I sleep under a mosquito net.}

\fmtLemma{ka\-ti\-ndi\-ro}
\fmtCitation{(a)ka\-ti\-ndi\-ro}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{rack (for utensils), shelf}
\fmtExampleRL{Tee\-ka esu\-waa\-ni oku ka\-ti\-ndi\-ro.}
\fmtExampleEN{Put the plates on the rack.}

\fmtLemma{ka\-ti\-nko}
\fmtCitation{(a)ka\-ti\-nko}
\fmtPOS{n}%
\fmtNC{12}
\fmtSubentry(a)
\fmtSense{fad, infatuation, short lived passion}
\fmtExampleRL{Omu\-ka\-li oyo ata n’a\-ka\-ti\-nko.}
\fmtExampleEN{That woman is infatuated.}
\fmtSubentry(b)
\fmtSense{intrusiveness, importunity}
\fmtExampleRL{Mu\-too\-le\-wo aka\-ti\-nko kaa\-nywe.}
\fmtExampleEN{Stop your intrusiveness.}

\fmtLemma{ka\-ti\-nvu\-ma}
\fmtCrossreference{\textnormal{see also} mukai}
\fmtCitation{(a)ka\-ti\-nvu\-ma}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{blue-seeded snoutbean, hairy snoutbean (\textit{Rhynchosia hirta})}
\fmtExampleRL{Aba\-ga\-ya bei\-zi\-ra ka\-ti\-nvu\-ma.}
\fmtExampleEN{Abagaya belong to the \textit{Rhynchosia hirta} clan.}

\fmtLemma{Ka\-to}
\fmtCitation{(o)Ka\-to}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{name given to the younger male twin}
\fmtExampleRL{Omu\-lo\-ngo owa\-ka\-bi\-ri ba\-mwe\-ta Ka\-to.}
\fmtExampleEN{The second twin is called Kato.}

\fmtLemma{ka\-to\-go}
\fmtCitation{(a)ka\-to\-go}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{mixture of foods prepared together (\eg beans with matoke)}
\fmtExampleRL{Omu\-ge\-nyi yaa\-lya ka\-to\-go.}
\fmtExampleEN{The visitor will eat a mixture of foods.}

\fmtLemma{Ka\-to\-nda}
\fmtCitation{(o)Ka\-to\-nda}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{God}
\fmtExampleRL{OKa\-to\-nda ali\-wo obwi\-re bwo\-na.}
\fmtExampleEN{God is present all the time.}

\fmtLemma{ka\-to\-nde}
\fmtCitation{(a)ka\-to\-nde}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{central part of a room between two poles}
\fmtExampleRL{Aka\-to\-nde ka nyu\-mba ba\-ko\-mbo\-ke\-re ku\-sai.}
\fmtExampleEN{The central part of the room was well constructed.}

\fmtLemma{ka\-too\-go}
\fmtCitation{(a)ka\-too\-go}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{cloth with stripes running vertically}
\fmtExampleRL{OMo\-se\-s azwai\-re esaa\-ti eya ka\-too\-go.}
\fmtExampleEN{Moses is wearing a cloth with vertical stripes.}

\fmtLemma{ka\-too\-ke}
\fmtCitation{(o)ka\-too\-ke}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{crude local gin}
\fmtExampleRL{Oka\-too\-ke ata\-mii\-rya mu\-no.}
\fmtExampleEN{Crude gin makes one very drunk.}

\fmtLemma{ka\-tu\-ba}
\fmtCitation{(a)ka\-tu\-ba}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{frame for moulding bricks}
\fmtExampleRL{Nkoo\-ra\-yo aka\-tu\-ba ka bbu\-loo\-ka.}
\fmtExampleEN{Make for me a frame for moulding bricks.}

\fmtLemma{ka\-tu\-ku}
\fmtCitation{(o)ka\-tu\-ku}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{light-skinned woman}
\fmtExampleRL{Omu\-ka\-li na\-swei\-re ka\-tu\-ku.}
\fmtExampleEN{I married a light-skinned woman.}

\fmtLemma{ka\-tu\-mbu\-gu\-lu}
\fmtCitation{(a)ka\-tu\-mbu\-gu\-lu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{drumstick (the lower joint of the leg of a cooked chicken)}
\fmtExampleRL{Aka\-tu\-mbu\-gu\-lu ka nko\-ko ka\-woo\-ma.}
\fmtExampleEN{The chicken drumstick is tasty.}

\fmtLemma{ka\-tu\-nda}
\fmtCitation{(a)ka\-tu\-nda}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{passion fruit, maracuya, granadilla (\textit{Passiflora edulis})}
\fmtExampleRL{OMa\-ndy aye\-ndya mu\-no oku\-lya obu\-tu\-nda.}
\fmtExampleEN{Mandy really loves eating passion fruit.}

\fmtLemma{ka\-tu\-ndai}
\fmtCrossreference{\textnormal{see} katunda}

\fmtLemma{ka\-tu\-ngu\-lu}
\fmtCitation{(a)ka\-tu\-ngu\-lu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{onion}
\fmtExampleRL{Aka\-tu\-ngu\-lu ba\-ka\-nsyei\-re.}
\fmtExampleEN{They have sold me expensive onions.}

\fmtLemma{ka\-tu\-ngu\-lu\-cu\-mu}
\fmtCitation{(a)ka\-tu\-ngu\-lu\-cu\-mu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{garlic}
\fmtExampleRL{Oka\-tu\-ngu\-lu\-cu\-mu awoo\-me\-sya omu\-ku\-bi.}
\fmtExampleEN{Garlic makes the sauce delicious.}

\fmtLemma{ka\-tu\-nku\-ma}
\fmtCitation{(a)ka\-tu\-nku\-ma}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{bitter berry, pea eggplant (\textit{Solanum torvum})}
\fmtExampleRL{Nta\-ka oku\-lya oka\-tu\-nku\-ma.}
\fmtExampleEN{I like eating bitter berries.}

\fmtLemma{ka\-tuu\-ka}
\fmtCitation{(a)ka\-tuu\-ka}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{orchitis (inflammation of testicles)}
\fmtExampleRL{Omu\-sai\-za oyo mu\-lwai\-re wa ka\-tuu\-ka.}
\fmtExampleEN{That man is suffering from orchitis.}

\fmtLemma{ka\-two\-ge\-re}
\fmtCitation{(o)ka\-two\-ge\-re}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{multi-coloured maize}
\fmtExampleRL{Omwa\-ka gu\-ni na\-lu\-ma ka\-two\-ge\-re ye\-nkai.}
\fmtExampleEN{This season I only grew multi-coloured maize.}

\fmtLemma{ka\-vu\-yo}
\fmtCitation{(a)ka\-vu\-yo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{confusion, mess}
\fmtExampleRL{Naa\-ni ya\-li mu\-wa\-ku\-li wa ka\-vu\-yo?}
\fmtExampleEN{Who was the instigator of the strike?}

\fmtLemma{ka\-waa\-li}
\fmtCitation{(o)ka\-waa\-li}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{smallpox}
\fmtExampleRL{Aba\-ntu baa\-fii\-re ka\-waa\-li.}
\fmtExampleEN{People died of smallpox.}

\fmtLemma{ka\-wa\-go}
\fmtCitation{(a)ka\-wa\-go}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{gall bladder}
\fmtExampleRL{Aka\-wa\-go ka ndo\-go\-yi ka\-bba ka\-koo\-to.}
\fmtExampleEN{A horse's gall bladder is big.}

\fmtLemma{ka\-wai\-si\-mba}
\fmtCitation{(a)ka\-wai\-si\-mba}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{genet}
\fmtExampleRL{Aka\-wai\-si\-mba ka\-wu\-nya mu\-no.}
\fmtExampleEN{A genet really stinks.}

\fmtLemma{ka\-wai\-ye\-nje}
\fmtCitation{(a)ka\-wai\-ye\-nje}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{dizziness}
\fmtExampleRL{Nkuu\-ra\-mu aka\-wai\-ye\-nje.}
\fmtExampleEN{I am feeling somewhat dizzy.}

\fmtLemma{ka\-wa\-ku\-ra}
\fmtCitation{(o)ka\-wa\-ku\-ra}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{troublemaker, agitator}
\fmtExampleRL{Oyo ka\-wa\-ku\-ra nta\-lo.}
\fmtExampleEN{That one is a troublemaker.}

\fmtLemma{ka\-wa\-lu\-twe}
\fmtCitation{(o)ka\-wa\-lu\-twe}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{stunted, small, undersized}
\fmtExampleRL{Akaa\-na kaa\-mwe ka\-wa\-lu\-twe.}
\fmtExampleEN{His child is undersized.}

\fmtLemma{ka\-wa\-lu\-zo\-ka}
\fmtCitation{(o)ka\-wa\-lu\-zo\-ka}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{(of a child) malnourished}
\fmtExampleRL{Akaa\-la ka\-wa\-lu\-zo\-ka.}
\fmtExampleEN{The girl is malnourished.}

\fmtLemma{ka\-wa\-ndi\-ki\-syo}
\fmtCitation{(a)ka\-wa\-ndi\-ki\-syo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{pen (a general term for any writing device)}
\fmtExampleRL{Mpee\-rya aka\-wa\-ndii\-ki\-syo.}
\fmtExampleEN{Give me a pen.}

\fmtLemma{ka\-wa\-ngii\-ryo}%
\fmtOne{1}
\fmtCitation{(a)ka\-wa\-ngii\-ryo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{handle}
\fmtExampleRL{Ensi\-ka ezwi\-re\-mu aka\-wa\-ngii\-ryo.}
\fmtExampleEN{The hoe's handle has come off.}

\fmtLemma{ka\-wa\-ngii\-ryo}%
\fmtOne{2}
\fmtCitation{(a)ka\-wa\-ngii\-ryo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{habit of accusing others falsely, habit of slander}
\fmtExampleRL{Aba\-ntu aba\-mwe ba\-ta n’a\-ka\-wa\-ngii\-ryo.}
\fmtExampleEN{Some people have a habit of accusing others falsely.}

\fmtLemma{ka\-wa\-ngo}
\fmtCitation{(a)ka\-wa\-ngo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{mental disorder}
\fmtExampleRL{Abee\-ka egyo aba\-ku\-si\-nga ba\-ta n’a\-ka\-wa\-ngo.}
\fmtExampleEN{Most of the members of that family have a mental disorder.}

\fmtLemma{ka\-wa\-wa}
\fmtCitation{(o)ka\-wa\-wa}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{horsefly, gadfly}
\fmtExampleRL{Oka\-wa\-wa aga\-dya ebi\-so\-lo.}
\fmtExampleEN{Horse flies disturb animals.}

\fmtLemma{ka\-wee\-fu\-be}
\fmtCitation{(o)ka\-wee\-fu\-be}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{effort, attempt}
\fmtExampleRL{Tu\-li ku ka\-wee\-fu\-be wa ku\-ma\-lii\-rya nyu\-mba.}
\fmtExampleEN{We are making an effort to complete the house.}

\fmtLemma{ka\-wee\-si}
\fmtCitation{(o)ka\-wee\-si}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{small black cricket (It is believed to be an omen of death if heard in a house at night.)}
\fmtExampleRL{Oka\-wee\-si ala\-nga oku\-fa.}
\fmtExampleEN{The small black cricket is an omen of death.}

\fmtLemma{ka\-wee\-we}
\fmtCitation{(a)ka\-wee\-we}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{whisper}
\fmtExampleRL{Na\-wuu\-li\-re nga ba\-ku\-nyu\-mya mu ka\-wee\-we.}
\fmtExampleEN{I heard them speaking in whispers.}

\fmtLemma{ka\-wii\-ru\-wii\-ru}
\fmtCitation{(a)ka\-wii\-ru\-wii\-ru}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{wheelbarrow}
\fmtExampleRL{Amai\-za ba\-go\-so\-mee\-ra mu ka\-wii\-ru\-wii\-ru.}
\fmtExampleEN{They carry the water with the wheelbarrow.}

\fmtLemma{ka\-wii\-wi}
\fmtCitation{(a)ka\-wii\-wi}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{lake fly}
\fmtExampleRL{Omu bi\-ro bi\-ni mu\-bbaa\-mu oka\-wii\-wi mwi\-ngi.}
\fmtExampleEN{There are too many lake flies during this season.}

\fmtLemma{ka\-wo}
\fmtCitation{(a)ka\-wo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{pea (\textit{Pisum sativum})}

\fmtLemma{ka\-wo\-lee\-ro}
\fmtCitation{(a)ka\-wo\-lee\-ro}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{silence}
\fmtExampleRL{Aka\-wo\-lee\-ro kaa\-ka\-bbaa\-wo\-mbe.}
\fmtExampleEN{The silence has just started.}

\fmtLemma{ka\-wo\-ndoo\-ro}
\fmtCitation{(a)ka\-wo\-ndoo\-ro}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{vagina opening}
\fmtExampleRL{Ata n’e\-ki\-rwai\-re oku ka\-wo\-ndoo\-ro.}
\fmtExampleEN{She has an infection of the vagina opening.}

\fmtLemma{ka\-woo\-ro}
\fmtCitation{(a)ka\-woo\-ro}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{condition in babies which requires them to be fed frequently, greed in babies}

\fmtLemma{ka\-woo\-wo}
\fmtCitation{(a)ka\-woo\-wo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{perfume, scent}
\fmtExampleRL{Ye\-ku\-bbi\-re aka\-woo\-wo aka\-sai.}
\fmtExampleEN{She is wearing a nice perfume.}

\fmtLemma{ka\-wu\-ka}
\fmtCitation{(a)ka\-wu\-ka}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{insect}
\fmtExampleRL{Aka\-wu\-ka ka\-mu\-ne\-ne\-re.}
\fmtExampleEN{An insect has bitten him.}
\fmtSubentry(b)
\fmtSense{virus}
\fmtExampleRL{Aka\-wu\-ka ka Co\-ro\-na ka\-nyoo\-ke\-re mu nsi yoo\-na.}
\fmtExampleEN{The corona virus has become rampant in the entire world.}

\fmtLemma{ka\-wu\-la\-wu\-la}
\fmtCitation{(a)ka\-wu\-la\-wu\-la}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{lima bean (\textit{Phaseolus lunatus})}
\fmtExampleRL{Ebi\-koo\-la bya ka\-wu\-la\-wu\-la bi\-ta\-mba endwai\-re ya ma\-twi omu baa\-na.}
\fmtExampleEN{Leaves of lima bean are used to cure ear infections in children.}

\fmtLemma{ka\-wu\-li}
\fmtCrossreference{\textnormal{see} iwuli}
\fmtCitation{(a)ka\-wu\-li}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{fart egg, fairy egg, small yolkless egg laid by a hen beginning to lay}

\fmtLemma{ka\-wu\-lii\-rya}
\fmtCitation{(a)ka\-wu\-lii\-rya}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{blood pressure monitor, BP machine}
\fmtExampleRL{Omu irwa\-lii\-ro ndoo\-mu ka\-wu\-lii\-rya.}
\fmtExampleEN{There is no BP machine in the hospital.}

\fmtLemma{ka\-wu\-lu}
\fmtCitation{(a)ka\-wu\-lu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{hole, opening}
\fmtExampleRL{Enki\-nzo ebbaa\-ku aka\-wu\-lu.}
\fmtExampleEN{The sewing needle has a hole.}
\fmtSubentry(b)
\fmtSense{ventilator, fan}
\fmtExampleRL{enyu\-mba ebbaa\-ku obu\-wu\-lu.}
\fmtExampleEN{The house should have fans.}
\fmtSubentry(c)
\fmtSense{hide-and-seek (children's game)}
\fmtExampleRL{Abaa\-na ba\-te\-ra oku\-zee\-nya aka\-wu\-lu ni ba\-li oku iso\-me\-ro.}
\fmtExampleEN{Children often play hide-and-seek at school.}

\fmtLemma{ka\-wu\-lu\-ku\-tu}
\fmtCitation{(a)ka\-wu\-lu\-ku\-tu}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{otitis (inflammation of the ear)}
\fmtExampleRL{Oyo mu\-lwai\-re wa ka\-wu\-lu\-ku\-tu.}
\fmtExampleEN{He is suffering from otitis.}

\fmtLemma{ka\-wu\-lu\-la}
\fmtCitation{(a)ka\-wu\-lu\-la}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{Kivu tree frog (\textit{Leptopelis kivuensis}, a small green frog believed to be very poisonous)}
\fmtExampleRL{Ombu\-li wa\-nge alii\-re aka\-wu\-lu\-la.}
\fmtExampleEN{My goat has eaten a Kivu tree frog.}

\fmtLemma{ka\-wu\-lu\-wu\-lu}
\fmtCitation{(o)ka\-wu\-lu\-wu\-lu}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{bitter berries}
\fmtExampleRL{Oku\-ta\-ku\-nya ka\-wu\-lu\-wu\-lu.}
\fmtExampleEN{You are chewing bitter berries.}

\fmtLemma{ka\-wu\-mpu\-li}
\fmtCitation{(o)ka\-wu\-mpu\-li}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSubentry(a)
\fmtSense{plague (a contagious bacterial disease)}
\fmtExampleRL{Endwai\-re ya ka\-wu\-mpu\-li ei\-ta aba\-ntu bai\-ngi mu ka\-see\-ra ka\-doo\-li mu\-no.}
\fmtExampleEN{Plague kills many people in a short span of time.}
\fmtSubentry(b)
\fmtSense{type of god or spirit which causes a plague}

\fmtLemma{ka\-wu\-na}
\fmtCitation{(a)ka\-wu\-na}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{taro (\textit{Co\-lo\-casia escu\-len\-ta})}
\fmtExampleRL{Ti\-nku\-ta\-ka ku\-lya bu\-wu\-na.}
\fmtExampleEN{I don’t want to eat taro.}

\fmtLemma{ka\-wu\-nga}
\fmtCitation{(o)ka\-wu\-nga}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{maize meal (mingled into paste)}
\fmtExampleRL{Ema\-mbya twa\-lya ka\-wu\-nga.}
\fmtExampleEN{Tomorrow we will eat a maize meal.}

\fmtLemma{ka\-wu\-nge}
\fmtCitation{(o)ka\-wu\-nge}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{type of banana used for brewing local beer}
\fmtExampleRL{Embii\-re eni tu\-gye\-ta ka\-wu\-nge.}
\fmtExampleEN{This type of banana is called ``kawunge".}

\fmtLemma{ka\-wu\-ngo}
\fmtCitation{(a)ka\-wu\-ngo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{mental disorder}
\fmtExampleRL{ONa\-mu\-si\-si ataa\-ku aka\-wu\-ngo.}
\fmtExampleEN{Namusisi has a mental disorder.}

\fmtLemma{ka\-wu\-ngu\-ko}
\fmtPOS{adv}
\fmtSense{fart egg, fairy egg, small yolkless egg laid by a hen beginning to lay}
\fmtExampleRL{Abbai\-re\-ku ani ka\-wu\-ngu\-ko n’a\-ya\-ba.}
\fmtExampleEN{He was here briefly and left.}

\fmtLemma{ka\-wu\-nya}
\fmtCitation{(o)ka\-wu\-nya}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{basil (\textit{Ocimum basilicum})}
\fmtExampleRL{Oka\-wu\-nya ata\-mba eki\-nzi.}
\fmtExampleEN{Basil cures flu.}

\fmtLemma{ka\-wuu\-jo}
\fmtCitation{(a)ka\-wuu\-jo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{spoon formed from a banana leaf for eating soup, banana leaf spoon}
\fmtSubentry(b)
\fmtSense{ventilator, fan}
\fmtExampleRL{Bu\-li ma\-sii\-ni etee\-kwa oku\-bba n’a\-ka\-wuu\-jo.}
\fmtExampleEN{Every machine must have a fan.}

\fmtLemma{ka\-wuu\-kuu\-mi}
\fmtCitation{(o)ka\-wuu\-kuu\-mi}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{grain weevil (\textit{Sitophilus granarius})}
\fmtExampleRL{Oka\-wu\-kuu\-mi alii\-re ebi\-yi\-mba bya\-nge.}
\fmtExampleEN{The grain weevil has destroyed my beans.}

\fmtLemma{ka\-wuu\-na\-mii\-rya}
\fmtCitation{(a)ka\-wuu\-na\-mii\-rya}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{total silence}
\fmtExampleRL{Wa\-bbai\-re\-wo aka\-wuu\-na\-mii\-rya oIsa\-ba\-nya\-la ni ya\-ki\-re.}
\fmtExampleEN{There was total silence when Isabanyala arrived.}

\fmtLemma{ka\-yaa\-ta (a)ka\-ju\-ngu}
\fmtCitation{(a)ka\-yaa\-ta}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{Irish potato (UgE), potato (\textit{Solonum tubersorum})}
\fmtExampleRL{Obu\-yaa\-ta obu\-ju\-ngu bu\-zwa Ire\-la\-nd.}
\fmtExampleEN{The Irish potato originated from Ireland.}

\fmtLemma{ka\-yaa\-yaa\-na}
\fmtCitation{(a)ka\-yaa\-yaa\-na}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{\textit{Vernonia cinerea} (medicinal herb believed to bring blessings if used, it is usually used for bathing to bring good fortune)}
\fmtExampleRL{Oka\-yaa\-ya\-na ba\-naa\-bya enke\-re\-mba zi\-bbe n’e\-mi\-ki\-sa.}
\fmtExampleEN{\textit{Vernonia cinerea} is used to bathe babies to give them luck.}

\fmtLemma{ka\-ya\-ya}
\fmtCitation{(a)ka\-ya\-ya}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{body odour}
\fmtExampleRL{Omwa\-la oyo ata n’a\-ka\-ya\-ya.}
\fmtExampleEN{That girl has body odour.}

\fmtLemma{ka\-yi\-nja}
\fmtCitation{(o)ka\-yi\-nja}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{type of banana used to make local brew}

\fmtLemma{Ka\-yi\-zi}
\fmtCitation{(o)Ka\-yi\-zi}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{spirit or god responsible for hunting}
\fmtExampleRL{OKa\-yi\-zi ni ya\-wa aba\-yi\-gi emi\-ki\-sa.}
\fmtExampleEN{It is Kayizi who gives luck to hunters.}

\fmtLemma{ka\-yi\-zu\-li}
\fmtCrossreference{\textnormal{see} kaizuli}

\fmtLemma{ka\-yo\-vu}
\fmtCitation{(o)ka\-yo\-vu}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{banana weevil (\textit{Cosmopolites sordidus})}
\fmtExampleRL{Oka\-yo\-vu azi\-kya ensu\-ku.}
\fmtExampleEN{The banana weevil destroys banana plantations.}

\fmtLemma{ka\-yu\-ka}
\fmtCitation{(oku)\-ka\-yu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ka\-yu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to be extremely angry, to become furious}
\fmtExampleRL{Oi\-se\-ma\-ka aka\-yu\-ki\-re bu\-ndi llwa bu\-ta\-lya kya\-mi\-sa\-na.}
\fmtExampleEN{The head of the family is extremely angry because he missed lunch.}
\fmtSubentry(b)
\fmtSense{to be exhausted, to be very tired}
\fmtExampleRL{Nka\-yu\-ki\-re olwa mi\-ri\-mu egi\-ngi gye nkoo\-re.}
\fmtExampleEN{I am very exhausted from work.}
\fmtSubentry(c)
\fmtSense{(of sunshine) to shine very brightly}
\fmtExampleRL{Ei\-sa\-na li\-ka\-yu\-ki\-re.}
\fmtExampleEN{The sun shines very brightly.}

\fmtLemma{ka\-yu\-la}
\fmtCitation{(oku)\-ka\-yu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ka\-ywi\-re}
\fmtSense{to annoy, to disturb, to anger}
\fmtExampleRL{Ya\-ba omu\-ka\-yu\-le tu\-bo\-ne kye ya\-ko\-la.}
\fmtExampleEN{Go and annoy her and see what she will do.}

\fmtLemma{ka\-yuu\-ki\-yuu\-ki}
\fmtCitation{(o)ka\-yuu\-ki\-yuu\-ki}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{\textit{Lantana trifolia} (plant with small sweet purple berries used to treat hiccups)}

\fmtLemma{ka\-yuu\-ya}
\fmtCitation{(a)ka\-yuu\-ya}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{sickle}
\fmtExampleRL{Naa\-ni eyai\-bbi\-re aka\-yuu\-ya ka\-nge aka\-sa\-la ei\-su\-bi?}
\fmtExampleEN{Who stole my sickle which I use to cut grass?}

\fmtLemma{ka\-ze\-ru}
\fmtCitation{(o)ka\-ze\-ru}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{white calf}
\fmtExampleRL{Enya\-na ezi\-ka\-ze\-ru ba\-zi\-twai\-re.}
\fmtExampleEN{The white calves have been taken away.}

\fmtLemma{ka\-zi}
\fmtCitation{(a)ka\-zi}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{blood relationship}
\fmtExampleRL{Ndoo\-wo ka\-zi aka\-ti waa\-mwe n’o\-Sse\-mba\-tya.}
\fmtExampleEN{There is no relationship between her and Ssembatya.}

\fmtLemma{ka\-zi\-myei\-zi\-mye}
\fmtCitation{(a)ka\-zi\-myei\-zi\-mye}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{dusk, twilight}
\fmtExampleRL{Ya\-ya\-bi\-re mu ka\-zi\-myei\-zi\-mye.}
\fmtExampleEN{He left at dusk.}

\fmtLemma{ka\-zi\-ndaa\-lo}
\fmtCitation{(a)ka\-zi\-ndaa\-lo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{funnel}
\fmtExampleRL{Aka\-zi\-ndaa\-lo tu\-ka\-ko\-le\-sya oku\-su\-ka ama\-fu\-ta omu taa\-la.}
\fmtExampleEN{We use a funnel to pour paraffin in the lantern.}
\fmtSubentry(b)
\fmtSense{microphone}
\fmtExampleRL{Lee\-ta aka\-zi\-ndaa\-lo nsi\-moo\-re\-ku.}
\fmtExampleEN{Bring the microphone and I will use it to speak.}

\fmtLemma{ka\-zi\-nda\-ba\-ka\-ma}
\fmtCitation{(a)ka\-zi\-nda\-ba\-ka\-ma}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{tonsillitis}
\fmtExampleRL{Mu\-lwai\-re wa ka\-zi\-nda\-ba\-ka\-ma.}
\fmtExampleEN{He is suffering from tonsillitis.}

\fmtLemma{ka\-zi\-yo}
\fmtCitation{(a)ka\-zi\-yo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{sticks used to play a small drum or xylophone}
\fmtExampleRL{Ya\-ba omu ka\-ta\-le ogu\-le obu\-zi\-yo.}
\fmtExampleEN{Go to the market and buy sticks used to play a small sized drum.}

\fmtLemma{ka\-zu\-mu}
\fmtCitation{(a)ka\-zu\-mu}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{whirlwind, tornado, devil wind}
\fmtExampleRL{Aka\-zu\-mu ka\-ka\-mbwi\-re enyu\-mba.}
\fmtExampleEN{The tornado took the roof off of the house.}

\fmtLemma{ka\-zu\-mu\-zu\-mu}
\fmtCrossreference{\textnormal{see} kazumu}

\fmtLemma{ka\-zyo}
\fmtCrossreference{\textnormal{see} kaziyo}

\fmtLemma{ke\-bbo}
\fmtCitation{(e)ke\-bbo}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{cable}
\fmtExampleRL{Eni ke\-bbo ya ma\-sa\-nya\-la\-ze ome\-se agi\-gu\-gu\-ni\-re.}
\fmtExampleEN{This power cable has been eaten by a rat.}

\fmtLemma{ke\-be\-ra}
\fmtCitation{(oku)\-ke\-be\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ke\-bei\-re}
\fmtSubentry(a)
\fmtSense{to check, to look at}
\fmtExampleRL{Ke\-be\-ra omu ka\-ba\-da obo\-ne ni mwa\-bba mu\-li\-mu osu\-kaa\-li.}
\fmtExampleEN{Check in the cupboard and see whether there is any sugar.}
\fmtSubentry(b)
\fmtSense{to examine, to study, to inspect, to test}
\fmtExampleRL{Ke\-be\-ra omwa\-na oyo ku\-sai le\-ke wee\-te\-gye eki\-ku\-mu\-lu\-ma.}
\fmtExampleEN{Examine that child properly and find out what she is suffering from.}

\fmtLemma{kee\-ka}
\fmtCitation{(oku)\-kee\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kee\-ke\-re}
\fmtSense{to bother, to disturb, to trouble}
\fmtExampleRL{Bai\-ra\-nge ti ki\-sai ku\-kee\-ka mu\-ntu bu\-li bwi\-re.}
\fmtExampleEN{It is not good to bother someone every now and then.}

\fmtLemma{kee\-ke}
\fmtCitation{(o)kee\-ke}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{sticky sauce}
\fmtExampleRL{Asu\-mbi\-re okee\-ke ni tu\-lya.}
\fmtExampleEN{He prepared a sticky sauce, and we ate.}

\fmtLemma{kee\-kee\-ya}
\fmtCitation{(oku)\-kee\-kee\-ya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kee\-kee\-ye\-re}
\fmtSense{to sieve, to filter}
\fmtExampleRL{Nta\-ka mu\-no oku\-kee\-kee\-ya obu\-sye.}
\fmtExampleEN{I like sieving powdered millet.}

\fmtLemma{kee\-ke\-ree\-rya}
\fmtCitation{(oku)\-kee\-ke\-ree\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kee\-ke\-rei\-rye}
\fmtSense{to use sparingly}
\fmtExampleRL{Ki\-ntu ki\-sai oku\-kee\-ke\-ree\-rya eme\-re.}
\fmtExampleEN{It is good to use food sparingly.}

\fmtLemma{kee\-ki}
\fmtCitation{(e)kee\-ki}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtExampleRL{Obu\-wu\-nga obu\-nie\-nu\-ki ba\-bu\-su\-mba\-mu kee\-ki.}
\fmtExampleEN{Fine flour is used for baking cakes.}

\fmtLemma{kee\-ra}
\fmtCitation{(oku)\-kee\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kei\-re}
\fmtSense{to be early}
\fmtExampleRL{Kee\-ra mu\-no ema\-mbya.}
\fmtExampleEN{Come very early tomorrow.}

\fmtLemma{kee\-ree\-rya}
\fmtCitation{(oku)\-kee\-ree\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kee\-rei\-rye}
\fmtSense{to make empty promises to}
\fmtExampleRL{Aba bya\-bu\-fu\-zi ba\-kee\-ree\-rya aba\-ntu.}
\fmtExampleEN{Politicians make empty promises.}

\fmtLemma{kee\-re\-rwa}
\fmtCitation{(oku)\-ku\-kee\-re\-rwa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtSource{from Sw.}
\fmtPFV{kee\-rei\-rwe}
\fmtSense{to be too late to catch (\eg a public vehicle), to miss, to have a delay}
\fmtExampleRL{Ya\-kee\-rei\-rwe ebbaa\-si n’e\-mwa\-ka\-la\-ku.}
\fmtExampleEN{He came late and missed the bus.}

\fmtLemma{kee\-si}
\fmtCitation{(e)kee\-si}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSubentry(a)
\fmtSense{suitcase}
\fmtSubentry(b)
\fmtSense{coffin}
\fmtExampleRL{Omu\-fu ba\-mu\-lii\-ka mu kee\-si.}
\fmtExampleEN{A dead person is buried in a coffin.}

\fmtLemma{kee\-wu\-lyo\-wu\-lyo}
\fmtCitation{(a)kee\-wu\-lyo\-wu\-lyo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{fontanelle}
\fmtExampleRL{Ti\-nkuu\-ra akee\-wu\-lyo\-wu\-lyo kaa\-mwe nga ka\-ku\-bba.}
\fmtExampleEN{I cannot feel the pulse of her fontanelle.}

\fmtLemma{kee\-ya}
\fmtCitation{(oku)\-kee\-ya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kee\-ye\-re}
\fmtSense{to reduce, to decrease, to minimise}
\fmtExampleRL{Obu\-lwai\-re bu\-kee\-ya obwi\-ngi bw’a\-ba\-ntu.}
\fmtExampleEN{Diseases reduce the size of the population.}

\fmtLemma{kee\-yee\-yo}
\fmtCitation{(a)kee\-yee\-yo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{\textit{Sida tenuicarpa} (small bush with yellow flowers, parts of the bush are used for making brooms)}

\fmtLemma{kee\-yi}
\fmtCitation{\textnormal{(e.g.}\,ga\-kee\-yi, bu\-kee\-yi, tu\-kee\-yi\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{little, small amount}
\fmtExampleRL{Mu\-si\-gai\-re\-mu obwi\-ta bu\-kee\-yi.}
\fmtExampleEN{There is little millet flour left.}
\fmtSubentry(b)
\fmtSense{scarce, a few}
\fmtExampleRL{Obu\-tu\-ngu\-lu eni ewai\-swe bu\-kee\-yi.}
\fmtExampleEN{Onions are very scarce here.}

\fmtLemma{kee\-yo}
\fmtCrossreference{\textnormal{see} keeyeeyo}

\fmtLemma{kee\-zi ka\-fii\-re}
\fmtCitation{(a)kee\-zi ka\-fii\-re}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{moonset}
\fmtExampleRL{Yai\-kii\-re mu kee\-zi ka\-fii\-re.}
\fmtExampleEN{She came at moonset.}

\fmtLemma{kei\-kei\-kei}
\fmtPOS{adv}
\fmtSense{until morning, whole night}
\fmtExampleRL{Oi\-ke\-ndi ya\-too\-nye\-re obwi\-re kei\-kei\-kei.}
\fmtExampleEN{It rained throughout the night.}

\fmtLemma{kei\-ta}
\fmtCitation{(oku)\-kei\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kei\-te\-re}
\fmtSense{to spy, to observe, to gather intelligence}
\fmtExampleRL{Aya\-bi\-re ku\-ba\-kei\-ta aso\-bo\-le oku\-ma\-nya oba nga bai\-bbi.}
\fmtExampleEN{She has gone to spy on them to find out if they are robbers.}

\fmtLemma{ke\-ka}
\fmtCitation{(oku)\-ke\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ke\-ke\-re}
\fmtSense{to slice, to cut, to chop}
\fmtExampleRL{Nku\-ke\-ka mu\-wo\-go wa ku\-ko\-la\-mu bu\-sye obwa na\-mi\-ra\-ntyo.}
\fmtExampleEN{I am slicing cassava for making flour.}

\fmtLemma{ke\-kee\-rya}
\fmtCitation{(oku)\-ke\-kee\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ke\-ke\-rei\-rye}
\fmtSense{to slice, to cut, to chop}
\fmtExampleRL{Baa\-bi\-re ku\-ke\-kee\-rya mu\-wo\-go.}
\fmtExampleEN{They went to slice cassava.}

\fmtLemma{ke\-kee\-ya}
\fmtCitation{(oku)\-ke\-kee\-ya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ke\-kee\-ye\-re}
\fmtSense{to sieve}
\fmtExampleRL{Naa\-ni eya\-ke\-kee\-ya obwi\-ta olwa\-lee\-ro?}
\fmtExampleEN{Who will sieve the millet today?}

\fmtLemma{ke\-kee\-yu}
\fmtCitation{(a)ke\-kee\-yu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{sieve, filter}
\fmtExampleRL{Omu\-so\-me\-sya ya\-guu\-re aka\-kee\-kee\-yu akee\-kee\-ye obu\-sye bwa\-mwe.}
\fmtExampleEN{The teacher bought a sieve to use it to sift the flour.}

\fmtLemma{ke\-ke\-ra}
\fmtCitation{(oku)\-ke\-ke\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ke\-kei\-re}
\fmtSubentry(a)
\fmtSense{to cackle}
\fmtExampleRL{Onko\-ko oyo ake\-ke\-ra mu\-no.}
\fmtExampleEN{That chicken cackles a lot.}
\fmtSubentry(b)
\fmtSense{(of guns) to fi\-re}
\fmtExampleRL{Embu\-ndu zi\-ku\-ke\-ke\-ra e Lu\-we\-ro.}
\fmtExampleEN{The guns are firing in Luweero.}

\fmtLemma{ke\-ma}
\fmtCitation{(oku)\-ke\-ma}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{ke\-me\-re}
\fmtSubentry(a)
\fmtSense{to groan, to moan, to whimper}
\fmtExampleRL{Omu\-lwai\-re ka aku\-ke\-ma bu\-ndi!}
\fmtExampleEN{The patient is groaning a lot!}
\fmtSubentry(b)
\fmtSense{to coax, to flatter, to persuade, to sweet talk}
\fmtExampleRL{Mu\-ke\-me obo\-ne ekya\-zwa\-yo.}
\fmtExampleEN{Coax him and see his reaction.}

\fmtLemma{ke\-me\-nkee\-rya}
\fmtCitation{(oku)\-ke\-me\-nkee\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ke\-me\-nkei\-rye}
\fmtSense{to coax, to flatter, to persuade, to sweet talk}
\fmtExampleRL{Lwa\-ki oku\-nke\-me\-nkee\-rya otyo?}
\fmtExampleEN{Why are you flattering me to that extent?}

\fmtLemma{ke\-mi}
\fmtCitation{(o)mu\-ke\-mi}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{a person who seduces others}
\fmtExampleRL{OSi\-taa\-ni ni\-ye yaa\-li omu\-ke\-mi wa Kaa\-wa oku\-lya eki\-ba\-la ekya\-gaa\-nii\-bwe.}
\fmtExampleEN{Satan is the one who seduced Eve to eat the forbidden fruit.}

\fmtLemma{ke\-mi\-si\-tu\-le}
\fmtCitation{(e)ke\-mi\-si\-tu\-le}
\fmtPOS{n}%
\fmtNC{9}
\fmtSource{from En.}
\fmtSense{chemistry}
\fmtExampleRL{Ku\-bbai\-gu\-lya\-ku ocuu\-cu ali oku ki\-ta\-bo kya\-mu ekya ke\-mi\-si\-tu\-le ota\-ndi\-ke oku\-so\-ma.}
\fmtExampleEN{Dust off the old chemistry book and start studying.}

\fmtLemma{ke\-ndai}
\fmtPOS{num}
\fmtSense{nine thousand}
\fmtExampleRL{Ali\-na embu\-li ke\-ndai.}
\fmtExampleEN{She has nine thousand goats.}

\fmtLemma{ke\-nde\-bu\-ka}
\fmtCitation{(oku)\-ke\-nde\-bu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ke\-nde\-bu\-ki\-re}
\fmtSense{to grow slim, to lose weight, to slim}
\fmtExampleRL{Omwa\-la waa\-mwe ake\-nde\-bu\-ki\-re lwa na\-mu\-nywa.}
\fmtExampleEN{His girl has lost weight due to HIV.}

\fmtLemma{ke\-ndee\-ra}
\fmtCitation{(oku)\-ke\-ndee\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ke\-ndei\-re}
\fmtSense{to reduce, to make smaller, to decrease, to turn down}
\fmtExampleRL{Ke\-ndee\-rya\-ku oku nda\-sa\-ma\-ra\-ka.}
\fmtExampleEN{Turn the radio down a little.}

\fmtLemma{ke\-ne\-ke\-ra}
\fmtCitation{(oku)\-ke\-ne\-ke\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ke\-ne\-kei\-re}
\fmtSense{to filter, to sieve, to filtrate}
\fmtExampleRL{Ke\-ne\-ke\-ra amai\-zi ago ga\-kya\-fu.}
\fmtExampleEN{Filter that water, it's dirty.}

\fmtLemma{ke\-ne\-nu\-ka}
\fmtCitation{(oku)\-ke\-ne\-nu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ke\-ne\-nu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to be emaciated}
\fmtExampleRL{Omu\-lwai\-re ake\-ne\-nu\-ki\-re ali\-na oku\-lya eme\-re ezi\-mbi\-sya omu\-bi\-ri.}
\fmtExampleEN{The patient is emaciated, he needs to eat food rich in proteins.}
\fmtSubentry(b)
\fmtSense{to be satisfied, to be content, to be persuaded}
\fmtExampleRL{Omu\-sai\-za waa\-mu wa\-mu\-ko\-bei\-re n’a\-ke\-ne\-nu\-ka.}
\fmtExampleEN{You talked to your husband and he was satisfied.}

\fmtLemma{ke\-ne\-nu\-la}
\fmtCitation{(oku)\-ke\-ne\-nu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ke\-ne\-nwi\-re}
\fmtSubentry(a)
\fmtSense{to filter, to strain, to sieve}
\fmtExampleRL{Ke\-ne\-nu\-la\-mu omu\-je\-ne\-ne.}
\fmtExampleEN{Filter the honey.}
\fmtSubentry(b)
\fmtSense{to sort out, to select}
\fmtExampleRL{Ke\-ne\-nu\-la\-mu ebi\-ga\-mbo by’a\-ma\-ku\-ru byo\-nkai.}
\fmtExampleEN{Sort out only the most important information.}

\fmtLemma{ke\-nga}
\fmtCitation{(oku)\-ke\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ke\-nge\-re}
\fmtSense{to foresee, to anticipate}
\fmtExampleRL{Na\-bi\-ke\-nge\-re bu\-kya\-li ni nyi\-ru\-ka.}
\fmtExampleEN{I foresaw what was coming and ran away.}

\fmtLemma{ke\-ngei\-ta}
\fmtCitation{(oku)\-ke\-ngei\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ke\-ngei\-te\-re}
\fmtSense{to evade the main path or entrance}
\fmtExampleRL{Ake\-ngei\-te\-re ei\-re\-mbo n’a\-ya\-ka\-la oku nta\-ndo.}
\fmtExampleEN{He avoided the main path and used the rear door.}

\fmtLemma{ke\-nge\-re}
\fmtCitation{(e)ke\-nge\-re}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{bicycle bell}
\fmtExampleRL{Egaa\-li ya\-nge gi\-na\-mbu\-la ke\-nge\-re.}
\fmtExampleEN{My bicycle has no bell.}

\fmtLemma{ke\-pe\-rya}
\fmtCitation{(oku)\-ke\-pe\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ke\-pei\-rye}
\fmtSense{to limp}
\fmtExampleRL{Lwa\-ki omwa\-na oyo aku\-ke\-pe\-rya?}
\fmtExampleEN{Why is that child limping?}
\fmtSense{to wander about, to roam around, to rove}
\fmtExampleRL{Omwa\-na oyo ake\-pe\-rya oku kya\-lo.}
\fmtExampleEN{That child roams around the village.}

\fmtLemma{ke\-re\-sya}
\fmtCitation{(oku)\-ke\-re\-sya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ke\-re\-se\-rye}
\fmtSense{to be disgusted, to be fed up}
\fmtExampleRL{Omu\-ka\-li oni anke\-re\-se\-rye.}
\fmtExampleEN{I am fed up with this woman.}

\fmtLemma{ke\-re\-zi\-ya}
\fmtCitation{(e)ke\-re\-zi\-ya}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{Catholic church}
\fmtExampleRL{Bo\-mbo\-ke\-re eke\-re\-zi\-ya egi\-yaa\-ka.}
\fmtExampleEN{They have built a new Catholic church.}

\fmtLemma{ke\-rwa}
\fmtCitation{(oku)\-ke\-rwa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kee\-rwe}
\fmtSubentry(a)
\fmtSense{to be tired}
\fmtExampleRL{Abba\-si\-rye n’a\-wee\-ra\-yo ku\-ba\-nga ya\-kee\-rwe mu\-no.}
\fmtExampleEN{She slept deeply because she was too tired.}
\fmtSubentry(b)
\fmtSense{to be fed up with, to be sick of, to be annoyed with}
\fmtExampleRL{Nkee\-rwe ebi\-ntu by’o\-ku nko\-la.}
\fmtExampleEN{I am fed up with the way you are treating me.}

\fmtLemma{ke\-rwi}
\fmtPOS{adj}
\fmtSense{very tired, exhausted}
\fmtExampleRL{Aba\-ntu ba\-ke\-rwi olwa bwi\-bbi obwa\-ka\-li\-rii\-rye omu kya\-lo.}
\fmtExampleEN{The people are a very tired of the robberies going on in the village.}

\fmtLemma{ke\-sa}
\fmtCitation{(oku)\-ke\-sa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ke\-se\-re}
\fmtSense{to harvest (millet or sorghum)}
\fmtExampleRL{Nkwa\-ba ku\-ke\-sa bwi\-ta.}
\fmtExampleEN{I am going to harvest millet.}

\fmtLemma{ke\-tu\-la}
\fmtCitation{(oku)\-ke\-tu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ke\-twi\-re}
\fmtSense{to cut off a piece}
\fmtExampleRL{Nke\-tuu\-ra\-ku ka\-doo\-li.}
\fmtExampleEN{Cut me a piece please.}

\fmtLemma{ke\-wu\-lye\-wu\-lye}
\fmtCitation{(a)ka\-wu\-lye\-wu\-lye}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{fontanelle}
\fmtExampleRL{Ake\-wu\-lye\-wu\-lye ka mwa\-na ti\-ka\-ku\-bba ku\-sai.}
\fmtExampleEN{The child’s fontanelle doesn’t pulse normally.}

\fmtLemma{ke\-yi\-mbyo}
\fmtCitation{(a)ke\-yi\-mbyo}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{central point of heir growth}

\fmtLemma{ki-}%
\fmtOne{1}
\fmtPOS{pf}
\fmtSense{subject and object agreement prefix of class 7 on verbs}

\fmtLemma{ki}%
\fmtOne{2}
\fmtPOS{interrog}
\fmtSense{what, which}
\fmtExampleRL{Oku\-ko\-ba ki?}
\fmtExampleEN{What are you saying?}

\fmtLemma{ki\-baa\-gi}
\fmtCitation{(e)ki\-baa\-gi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{slaughtering machete}
\fmtExampleRL{Ti\-ndi\-na ki\-baa\-gi kya ku\-baa\-ga onte.}
\fmtExampleEN{I don't have a machete to slaughter the cow.}

\fmtLemma{ki\-baa\-ra}
\fmtCrossreference{\textnormal{see also} mbaara}
\fmtCitation{(e)ki\-baa\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{tsetse fly}
\fmtExampleRL{Ebi\-baa\-ra bi\-ree\-ta endwai\-re ya na\-mpo\-nge\-ra.}
\fmtExampleEN{Tsetse flies cause sleeping sickness.}

\fmtLemma{ki\-ba\-ga\-lo}
\fmtCitation{(e)ki\-ba\-ga\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{basket for fish}
\fmtExampleRL{Ebi\-ba\-ga\-lo ba\-bi\-ta\-mbu\-rya\-mu ncwi.}
\fmtExampleEN{Baskets are used to carry fish.}

\fmtLemma{ki\-ba\-ga\-nio}
\fmtCitation{(e)ki\-ba\-ga\-nio}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{share, portion, part, ration, quota}
\fmtExampleRL{Bu\-li mu\-ntu atwa\-le eki\-ba\-ga\-nio kya\-mwe.}
\fmtExampleEN{Every person should take her or his share.}

\fmtLemma{ki\-bai\-ze}
\fmtCitation{(e)ki\-bai\-ze}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{piece of furniture}
\fmtExampleRL{Nkoo\-ra eki\-bai\-ze eki\-ndi ki\-ni ki\-ka\-yi\-ri\-ki\-re.}
\fmtExampleEN{Make me another piece of furniture, this one has grown old.}

\fmtLemma{ki\-bai\-zo}
\fmtCitation{(e)ki\-bai\-zo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{part, piece}
\fmtExampleRL{Mpee\-rya\-ku eki\-bai\-zo kya mu\-wo\-go oni.}
\fmtExampleEN{Give me a piece of this cassava.}

\fmtLemma{ki\-ba\-ju}
\fmtCitation{(e)ki\-ba\-ju}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{calabash}
\fmtExampleRL{Ya\-boi\-ne eki\-ba\-ju eki\-koo\-to omu ka\-ta\-le.}
\fmtExampleEN{He saw a large calabash in the market.}

\fmtLemma{ki\-ba\-la}
\fmtCitation{(e)ki\-ba\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fruit}
\fmtExampleRL{Uga\-nda eta n’e\-bi\-ba\-la ebya nja\-wu\-lo bi\-ngi.}
\fmtExampleEN{Uganda has a variety of fruits.}

\fmtLemma{ki\-ba\-lo}
\fmtCitation{(e)ki\-ba\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{calculation, computation, adding up, mathematics}
\fmtExampleRL{Eki\-ba\-lo omwo\-jo kye ya\-koo\-re ki\-fu.}
\fmtExampleEN{The calculation done by the boy is wrong.}

\fmtLemma{ki\-ba\-mba}
\fmtCitation{(e)ki\-ba\-mba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{extraordinary happening, misfortune, disaster, tragedy}
\fmtExampleRL{Kya\-li ki\-ba\-mba omaa\-ma oku\-bba\-sya n’o\-mu\-taa\-ni.}
\fmtExampleEN{It was a tragedy for a mother to sleep with her own son.}

\fmtLemma{ki\-ba\-ndwa}
\fmtCitation{(o)ki\-ba\-ndwa}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSubentry(a)
\fmtSense{traditional priest}
\fmtExampleRL{Oki\-ba\-ndwa ai\-ki\-re n’a\-ba\-ntu aba\-ndi ab’e\-bi\-ti\-ni\-sa.}
\fmtExampleEN{The traditional priest has arrived with many other honoraries.}
\fmtSubentry(b)
\fmtSense{possessed (by a demon, a spirit)}
\fmtExampleRL{Abo ba\-ki\-ba\-ndwa Ba\-bu\-ruu\-li.}
\fmtExampleEN{Those are the people from Buruuli possessed by spirits.}

\fmtLemma{ki\-ba\-ngo}
\fmtCitation{(e)ki\-ba\-ngo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{trap used to fish mudfish}
\fmtExampleRL{Eki\-ba\-ngo kya\-nge ki\-buu\-re.}
\fmtExampleEN{My trap for fishing mudfish is missing.}

\fmtLemma{ki\-ba\-nja}
\fmtCitation{(e)ki\-ba\-nja}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{plot of land}
\fmtExampleRL{Nguu\-re eki\-ba\-nja.}
\fmtExampleEN{I have bought a plot of land.}

\fmtLemma{ki\-ba\-raa\-zi}
\fmtCitation{(e)ki\-ba\-raa\-zi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{tragedy, pity, misfortune, disaster}
\fmtExampleRL{Ki\-ba\-raa\-zi kya maa\-ni ekya ku\-fii\-rwa owa\-ki\-tii\-ni\-sya Pe\-te\-r Nyo\-mbi.}
\fmtExampleEN{It is a pity to lose Hon. Peter Nyombi.}
\fmtSubentry(b)
\fmtSense{broken heart}
\fmtExampleRL{ONa\-ndi\-ra ata n’e\-ki\-ba\-raa\-zi.}
\fmtExampleEN{Nandira is heartbroken.}

\fmtLemma{ki\-ba\-ya}
\fmtCitation{(e)ki\-ba\-ya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{saddle}
\fmtExampleRL{Omwa\-na amu\-gi\-ngii\-re mu ki\-ba\-ya.}
\fmtExampleEN{She’s carrying the baby in a saddle.}

\fmtLemma{ki\-bba}
\fmtCitation{(e)ki\-bba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bundle of firewood, pile of firewood}
\fmtExampleRL{Ki\-ni ki\-bba kya nkwi oNa\-mu\-si\-si ni\-ye ya\-ki\-ree\-te\-re.}
\fmtExampleEN{This bundle of firewood was brought by Namusisi.}

\fmtLemma{ki\-bbaa\-ki}
\fmtCitation{(e)ki\-bbaa\-ki}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{shoulder bone}
\fmtExampleRL{Omu\-sai\-za odi ata n’e\-ki\-bbaa\-ki eki\-koo\-to.}
\fmtExampleEN{That man has a broad shoulder bone.}

\fmtLemma{ki\-bbaa\-ki\-ro}
\fmtCitation{(e)ki\-bbaa\-ki\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fermentation place}
\fmtExampleRL{Ki\-ni ni\-kyo ki\-bbaa\-ki\-ro kya bu\-sye.}
\fmtExampleEN{This is the fermentation place for millet.}

\fmtLemma{ki\-bbaa\-la}
\fmtCitation{(e)ki\-bbaa\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{tombstone}
\fmtExampleRL{Ti\-ndi ba\-ni abe\-rya\-na aba\-ku\-ko\-ba\-nga oku\-sa\-nga onyi\-bya\-la akii\-ri nsa\-nga eki\-bbaa\-la.}
\fmtExampleEN{I am not like these daughters-in-law who would say ``Instead of getting a mother-in-law, I'd rather get a tombstone".}

\fmtLemma{ki\-bbaa\-le}
\fmtCitation{(e)ki\-bbaa\-le}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{stone}
\fmtExampleRL{Eki\-bbaa\-le ki\-kwi\-ri\-ngi\-ta.}
\fmtExampleEN{The stone is rolling.}
\fmtSubentry(b)
\fmtSense{biceps}
\fmtExampleRL{Eki\-bbaa\-le kya mu\-ko\-no ogu\-lyo ki\-ku\-ndu\-ma.}
\fmtExampleEN{My right biceps ache.}
\fmtSubentry(c)
\fmtSense{hernia}
\fmtExampleRL{Ya\-fii\-re bu\-lwai\-re bwa ki\-bbaa\-le.}
\fmtExampleEN{He died of a hernia.}

\fmtLemma{ki\-bbaa\-ta}
\fmtCitation{(e)ki\-bbaa\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{testicle}
\fmtExampleRL{Onu\-me wa\-nge ba\-mwi\-yi\-re\-mu ama\-bbaa\-ta.}
\fmtExampleEN{My bull's testicles were removed.}

\fmtLemma{ki\-bba\-lu}
\fmtCitation{(e)ki\-bba\-lu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sore on the skin believed to be caused by syphilis}
\fmtExampleRL{Ebi\-bba\-lu obu\-ndi bi\-ree\-te\-bwa bi\-soo\-ni.}
\fmtExampleEN{Skin sores are sometimes caused by syphilis.}

\fmtLemma{ki\-bba\-lu\-ka}
\fmtCitation{(e)ki\-bba\-lu\-ka}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{sudden death}
\fmtExampleRL{Afii\-re ki\-bba\-lu\-ka.}
\fmtExampleEN{He died a sudden death.}

\fmtLemma{ki\-bba\-mbo}
\fmtCitation{(e)ki\-bba\-mbo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{parcel, package, bundle}
\fmtExampleRL{Si\-nda\-bu\-la eki\-bba\-mbo tu\-bo\-ne ebi\-ri mu\-nda.}
\fmtExampleEN{Open the parcel and let's see what is inside.}

\fmtLemma{ki\-bba\-mbya}%
\fmtOne{1}
\fmtCitation{(e)ki\-bba\-mbya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{daybreak, dawn, sunrise}
\fmtExampleRL{Wai\-za mu ki\-bba\-mbya.}
\fmtExampleEN{You will come at daybreak.}
\fmtSubentry(b)
\fmtFixedexpression{ki\-bba\-mbya (o)ku\-sa\-la}
\fmtSense{sunrise}
\fmtExampleRL{Oki\-bba\-mbya aku\-sa\-la.}
\fmtExampleEN{The sun is rising.}

\fmtLemma{ki\-bba\-mbya}%
\fmtOne{2}
\fmtCitation{(e)ki\-bba\-mbya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{patch, cover}
\fmtExampleRL{Ata n’e\-ki\-bba\-mbya oku ita\-ma ekya\-li\-gi\-re oku ka\-be\-nje ke ya\-fu\-ni\-re.}
\fmtExampleEN{He has a patch on the cheek which was caused by the accident he sustained.}

\fmtLemma{ki\-bba\-ni}
\fmtCitation{(e)ki\-bba\-ni}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{kitchen rack, kitchen shelf}
\fmtExampleRL{Mu\-te ama\-too\-ke oku ki\-bba\-ni.}
\fmtExampleEN{Put the banana on the kitchen rack.}
\fmtSubentry(b)
\fmtSense{cluster or mantle of bees, bees bearding together}
\fmtExampleRL{Enzo\-ki zi\-coo\-ke\-re adi ni zi\-ko\-la eki\-bba\-ni.}
\fmtExampleEN{The bees have clustered together and made a mantle.}
\fmtSubentry(c)
\fmtSense{continent}
\fmtExampleRL{Ensi eri\-ku ebi\-bba\-ni bi\-me\-kai?}
\fmtExampleEN{How many continents are there on this planet?}

\fmtLemma{ki\-bba\-si\-rwa}
\fmtCitation{(e)ki\-bba\-si\-rwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bedding}
\fmtExampleRL{Ata\-mbu\-la n’e\-bi\-bba\-si\-rwa bya\-mwe.}
\fmtExampleEN{He walks around with his bedding.}

\fmtLemma{ki\-bba\-si\-ryo}
\fmtCrossreference{\textnormal{see} ki\-bba\-syo}

\fmtLemma{ki\-bba\-syo}
\fmtCitation{(e)ki\-bba\-syo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{hostel, dormitory, guest house, accommodation}
\fmtExampleRL{Ani wa\-li\-wo eki\-bba\-syo kya ba\-ge\-ni.}
\fmtExampleEN{There is a visitors' hostel here.}
\fmtSubentry(b)
\fmtSense{(of animal) habitat}
\fmtExampleRL{Onga\-bi bu\-mu\-ju\-mbu\-twi\-re omu ki\-ba\-syo kya\-mwe.}
\fmtExampleEN{The bushbuck has been chased out of its habitat.}

\fmtLemma{ki\-bba\-ta\-ta}
\fmtCitation{(e)ki\-bba\-ta\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{heap of faeces, pile of shit}
\fmtExampleRL{Naa\-ni aniei\-re ani eki\-bba\-ta\-ta kya ma\-bbi?}
\fmtExampleEN{Who has excreted here a heap of faeces?}

\fmtLemma{ki\-bbe ki\-tyo}
\fmtPOS{interj}
\fmtSense{amen (expression used show confirmation or agreement)}
\fmtExampleRL{Ki\-bbe ki\-tyo.}
\fmtExampleEN{Amen. (lit.\,Be it that way.)}

\fmtLemma{ki\-bbee\-ro}
\fmtCitation{(e)ki\-bbee\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{mixture of foods prepared together}
\fmtExampleRL{Tu\-lii\-re ki\-bbee\-ro kya ma\-too\-ke, enya\-ma, ebi\-yi\-mba n’e\-bi\-nyee\-bwa.}
\fmtExampleEN{We ate a mixture of matoke, meat, beans and groundnuts.}

\fmtLemma{Ki\-bbee\-wo}
\fmtCitation{Ki\-bbee\-wo}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{name given to a child whose paternity is doubted}

\fmtLemma{ki\-bbe\-te\-bbe\-te}
\fmtCitation{(e)ki\-bbe\-te\-bbe\-te}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{thick soup}
\fmtExampleRL{Omu\-ku\-bi asu\-mbi\-re ki\-bbe\-te\-bbe\-te.}
\fmtExampleEN{She has prepared a thick soup.}

\fmtLemma{ki\-bbi}
\fmtCitation{(e)ki\-bbi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sin, evil, corruption, wrongdoing}
\fmtExampleRL{Ka\-nca atu\-ga\-nyi\-re ebi\-bbi byai\-swe.}
\fmtExampleEN{God should forgive our sins.}

\fmtLemma{ki\-bbii\-ni\-ki\-ra}
\fmtCitation{(e)ki\-bbii\-ni\-ki\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{cover, lid}
\fmtExampleRL{Ekyo\-loo\-ni ki\-tee\-ku eki\-bbii\-ni\-ki\-ra.}
\fmtExampleEN{Put a cover on the latrine.}
\fmtSubentry(b)
\fmtSense{bird trap which looks like a cover}
\fmtExampleRL{Eki\-nyo\-nyi nki\-te\-ge\-se\-rye ki\-bbii\-ni\-ki\-ra.}
\fmtExampleEN{I trapped the bird with a trap.}

\fmtLemma{ki\-bbo\-mboi\-zi}
\fmtCitation{(e)ki\-bbo\-mboi\-zi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{ash pit}
\fmtExampleRL{Ei\-ko\-ke ndi\-su\-ki\-re ku ki\-bbo\-mboi\-zi.}
\fmtExampleEN{I poured the ash into the ash pit.}

\fmtLemma{ki\-bbo\-nde}
\fmtCitation{(e)ki\-bbo\-nde}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{one-day-fermented brew, fresh local beer}
\fmtExampleRL{Abaa\-na aba\-to ba\-nywa ki\-bbo\-nde ku\-bba ti\-ki\-taa\-mi\-rya.}
\fmtExampleEN{Young children drink fresh fermented beer because it does not make them drunk.}

\fmtLemma{ki\-bbo\-nge}
\fmtCitation{(e)ki\-bbo\-nge}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{shank (lower part of an animal's foreleg)}
\fmtExampleRL{Onte abba n’e\-ki\-bbo\-nge ki\-koo\-to.}
\fmtExampleEN{The cow has a big shank.}

\fmtLemma{ki\-bboo\-ke}
\fmtCitation{(e)ki\-bboo\-ke}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sweet potato mound}
\fmtExampleRL{Ya\-ka\-bboo\-ka ebi\-bboo\-ke bi\-taa\-nu oku\-li\-ga ni bwa\-kei\-re.}
\fmtExampleEN{He has dug only five sweet potato mounds since morning.}

\fmtLemma{ki\-bbu\-bbu}
\fmtCitation{(o)ki\-bbu\-bbu}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{mute person}
\fmtExampleRL{Oki\-bbu\-bbu ta\-ba\-za era ta\-wuu\-ra.}
\fmtExampleEN{A mute person can neither talk nor hear.}

\fmtLemma{ki\-bbu\-tu\-tu}
\fmtCitation{(e)ki\-bbu\-tu\-tu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{cluster of sprouting mushrooms}
\fmtExampleRL{Ani wa\-li\-wo eki\-bbu\-tu\-tu.}
\fmtExampleEN{There is a cluster of sprouting mushrooms.}

\fmtLemma{ki\-bbuu\-ra}
\fmtCitation{(e)ki\-bbuu\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{heap, pile}
\fmtExampleRL{Ya\-guu\-re eki\-bbuu\-ra kya mu\-wo\-go ku bbee\-yi la\-yi\-si.}
\fmtExampleEN{She bought a heap of cassava cheaply.}
\fmtSubentry(b)
\fmtSense{area with dense grass, compact grass}
\fmtExampleRL{Adi wa\-li\-wo eki\-bbuu\-ra ky’ei\-su\-bi.}
\fmtExampleEN{There is an area with dense grass.}
\fmtSubentry(c)
\fmtSense{settlement of closely related people}
\fmtExampleRL{Aba\-gi\-su ba\-bba omu ki\-bbuu\-ra.}
\fmtExampleEN{The Gisu people stay in settlements of closely related people.}

\fmtLemma{ki\-bbuu\-ru}
\fmtCitation{(e)ki\-bbuu\-ru}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(d)
\fmtSense{clique, gang}
\fmtExampleRL{Awo wa\-taa\-wo eki\-bbuu\-ru ky’a\-bai\-bbi.}
\fmtExampleEN{There is a gang of thieves there.}

\fmtLemma{ki\-bbuu\-ta}
\fmtCitation{(o)ki\-bbuu\-ta}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{good looking young piglet}
\fmtExampleRL{Eki\-bbuu\-ta kya mbe\-re\-ge ki\-ni ki\-gu\-la se\-nte zi\-nkai?}
\fmtExampleEN{How much does this good-looking piglet cost?}

\fmtLemma{ki\-bee\-re}%
\fmtOne{1}
\fmtCitation{(e)ki\-bee\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{black monket thorn, Senegalia burkei from the acacia burkie Benth family}

\fmtLemma{ki\-bee\-re}%
\fmtOne{2}
\fmtCitation{(e)ki\-bee\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{udder}
\fmtExampleRL{Onte wa\-nge ali\-na eki\-bee\-re ki\-koo\-to bai\-tu ta\-zwa\-mu ma\-tai ga\-ku\-ma\-la.}
\fmtExampleEN{My cow has a big udder, but it does not produce enough milk.}

\fmtLemma{ki\-be\-ga\-be\-ga}
\fmtCitation{(e)ki\-be\-ga\-be\-ga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{shoulder}
\fmtExampleRL{Eki\-bba\-mbo aki\-gi\-ngii\-re ku ki\-be\-gaa\-be\-ga.}
\fmtExampleEN{He is carrying the parcel on his shoulder.}

\fmtLemma{ki\-be\-nde}%
\fmtOne{1}
\fmtCitation{(e)ki\-be\-nde}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fishing basket}
\fmtExampleRL{Ebi\-na\-ge mbii\-ti\-sya bi\-be\-nde.}
\fmtExampleEN{I fish tilapia using a fishing basket.}

\fmtLemma{ki\-be\-nde}%
\fmtOne{2}
\fmtCitation{(e)ki\-be\-nde}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{iron ore}
\fmtExampleRL{Ekyo\-ma ki\-zwa mu ki\-be\-nde.}
\fmtExampleEN{Iron is smelted from iron ore.}

\fmtLemma{ki\-be\-nde}%
\fmtOne{3}
\fmtCitation{(e)ki\-be\-nde}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{raffia skirt (a dancing costume for girls or women)}
\fmtExampleRL{Omu\-bi\-ni mu\-bbwe eki\-be\-nde atu\-bi\-ni\-re\-mu.}
\fmtExampleEN{Tie a raffia skirt around the dancer and she will dance for us.}

\fmtLemma{ki\-be\-ro}
\fmtCitation{(e)ki\-be\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{(of a person) thigh, lap}
\fmtExampleRL{Ata n’e\-ki\-be\-ro ki\-koo\-to.}
\fmtExampleEN{She has big thighs.}

\fmtLemma{ki\-bii\-ki\-ro}
\fmtCitation{(e)ki\-bii\-ki\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{store}
\fmtExampleRL{Ki\-ni ni\-kyo ki\-bii\-ki\-ro kya mwa\-ni.}
\fmtExampleEN{This is the coffee store.}

\fmtLemma{ki\-bii\-na}
\fmtCitation{(e)ki\-bii\-na}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{crowd}
\fmtExampleRL{Ye\-te ki\-di eki\-bii\-na kya ba\-ntu?}
\fmtExampleEN{What about the other crowd of people?}
\fmtSubentry(b)
\fmtSense{association, organisation, cooperative, party}
\fmtExampleRL{Tu\-ta n’e\-ki\-bii\-na kyai\-swe ekya ba\-ka\-li aba\-suu\-bu\-li.}
\fmtExampleEN{We have an association for female traders.}
\fmtSubentry(c)
\fmtSense{classroom, class}
\fmtExampleRL{Ai\-ngii\-re mu ki\-bii\-na ku\-so\-me\-sya baa\-na.}
\fmtExampleEN{She entered the classroom to teach the pupils.}
\fmtSubentry(d)
\fmtFixedexpression{ki\-bii\-na ky’a\-bai\-ki\-ri\-za}
\fmtSense{flock (a Christian congregation or body of believers)}
\fmtExampleRL{Eka\-ni\-sa ki\-bba ki\-bii\-na ky’a\-bai\-ki\-ri\-za.}
\fmtExampleEN{The church is a flock of believers.}

\fmtLemma{ki\-bi\-mbe}
\fmtCitation{(e)ki\-bi\-mbe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fishing net for lung fish}
\fmtExampleRL{Ngu\-li\-sya\-ku eki\-bi\-mbe n’a\-ku\-sa\-su\-la ni nfu\-ni\-re ese\-nte.}
\fmtExampleEN{Please buy me a fishing net for lung fish, I will pay you when I get the money.}

\fmtLemma{ki\-bi\-ngo}
\fmtCitation{(e)ki\-bi\-ngo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{reed torch (sticks tied together and lit to provide light at night)}
\fmtExampleRL{Ebi\-bi\-ngo tu\-bii\-ti\-sya enswa.}
\fmtExampleEN{We use the traditional reed torch when harvesting flying termites.}

\fmtLemma{ki\-bi\-ra}
\fmtCitation{(e)ki\-bi\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{forest, wood(s)}
\fmtExampleRL{Omwi\-bbi ali mu ki\-bi\-ra.}
\fmtExampleEN{The thief is in the forest.}

\fmtLemma{ki\-bi\-ri\-bi\-ri}
\fmtCitation{(e)ki\-bi\-ri\-bi\-ri}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{fever, malaria}
\fmtExampleRL{Omwa\-na alwai\-re ki\-bi\-ri\-bi\-ri ni\-kyo ki\-ku\-mu\-ku\-nja.}
\fmtExampleEN{The child has a fever, that is what is making her cry.}

\fmtLemma{ki\-bi\-rii\-ti}
\fmtCitation{(e)ki\-bi\-rii\-ti}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{match box}
\fmtExampleRL{Lee\-ta eki\-bi\-rii\-ti le\-ke twa\-kye omuu\-ro tu\-su\-mbe.}
\fmtExampleEN{Bring the match box, and we'll make fire for cooking.}

\fmtLemma{ki\-bi\-rii\-zi}
\fmtCitation{(e)ki\-bi\-rii\-zi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{\textit{Vernonia amygdalina} (plant whose leaves and roots are used to treat malaria)}
\fmtExampleRL{Eki\-bi\-rii\-zi ki\-ta\-mba eki\-bi\-ri\-bi\-ri.}
\fmtExampleEN{Vernonia amygdalina cures malaria.}

\fmtLemma{ki\-bo\-be}
\fmtCitation{(e)ki\-bo\-be}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{thick-yellowish mucus usually with a bad smell}
\fmtExampleRL{Omwa\-na ba\-mu\-twai\-re mu irwa\-ro nga aku\-zwa\-mu eki\-bo\-be.}
\fmtExampleEN{They took the child to the hospital because his mucus was thick and yellow.}

\fmtLemma{ki\-bo\-ndo}
\fmtCitation{(e)ki\-bo\-ndo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{cocoon}
\fmtExampleRL{Omwa\-na baa\-mwo\-koi\-re, na\-ye ti\-baa\-mu\-toi\-re\-mu bi\-bo\-ndo kya nju\-nzai.}
\fmtExampleEN{Jiggers were removed from the child's feet, but the cocoons were not.}

\fmtLemma{ki\-bo\-ne\-re\-zo}
\fmtCitation{(e)ki\-bo\-ne\-re\-zo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{punishment, penalty, disciplining}
\fmtExampleRL{Eki\-bo\-ne\-re\-zo ya\-fu\-ni\-re kya maa\-ni.}
\fmtExampleEN{He got a severe punishment.}

\fmtLemma{ki\-bo\-nwo\-mwei}
\fmtCitation{(o)ki\-bo\-nwo\-mwei}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{comet, meteorite, shooting star}
\fmtExampleEN{A shooting star fulfils a wish.}

\fmtLemma{ki\-boo\-be\-ryo}
\fmtCitation{(e)ki\-boo\-be\-ryo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{traditional pot used as a food flask}
\fmtExampleRL{Eki\-boo\-be\-ryo ki\-ko\-le\-se\-bwa oku\-bu\-gu\-mya eme\-re.}
\fmtExampleEN{The traditional food flask is used to keep food warm.}

\fmtLemma{ki\-boo\-bo}
\fmtCitation{(e)ki\-boo\-bo}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{loud cry}
\fmtExampleRL{Ya\-ku\-bbi\-re eki\-boo\-bo ni ya\-boi\-ne omu\-la\-mbo gwa ite waa\-mwe.}
\fmtExampleEN{She cried out loudly on seeing the body of her father.}

\fmtLemma{ki\-bu}
\fmtCitation{(e)ki\-bu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{meal}
\fmtExampleRL{Aboo\-ro ba\-lya eki\-bu ki\-mwei.}
\fmtExampleEN{The poor eat one meal a day.}

\fmtLemma{ki\-bu\-ga}%
\fmtOne{1}
\fmtCitation{(e)ki\-bu\-ga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{city, town}
\fmtExampleRL{Ka\-mpa\-la ni\-kyo ki\-bu\-ga kya Uga\-nda eki\-ku\-lu.}
\fmtExampleEN{Kampala is Uganda's capital city.}

\fmtLemma{ki\-bu\-gu}%
\fmtOne{2}
\fmtCrossreference{\textnormal{see also} kimbugumbugu}
\fmtCitation{(o)ki\-bu\-gu}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{vampire bat}
\fmtExampleRL{Oki\-bu\-gu alya emi\-ye\-mbe.}
\fmtExampleEN{Vampire bats eat mangoes.}

\fmtLemma{ki\-bu\-gu\-tu}
\fmtCitation{(e)ki\-bu\-gu\-tu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{coarse bark cloth, unrefined bark cloth}
\fmtExampleRL{Ye\-bbwe\-re eki\-bu\-gu\-tu.}
\fmtExampleEN{He is wearing coarse bark cloth.}

\fmtLemma{ki\-bu\-la}
\fmtCitation{(e)ki\-bu\-la}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{direction where Muslims face during their prayer, direction of Khabba}
\fmtExampleRL{Aba\-si\-raa\-mu ba\-lo\-lee\-ra omu ki\-bu\-la nga ba\-ku\-saa\-la.}
\fmtExampleEN{Moslems face in the direction of Khabba during prayers.}

\fmtLemma{ki\-bu\-lu\-ga}
\fmtCitation{(e)ki\-bu\-lu\-ga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sauce made from roasted groundnuts}
\fmtExampleRL{Eki\-bu\-lu\-ga ki\-woo\-ma.}
\fmtExampleEN{Sauce made from roasted groundnuts is tasty.}

\fmtLemma{ki\-bu\-lwa}
\fmtCitation{(o)ki\-bu\-lwa}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from taboo}
\fmtSense{without pubic hair, a person who naturally does not grow pubic hair}
\fmtExampleRL{Omu\-sai\-za oyo ata ki\-bu\-lwa.}
\fmtExampleEN{That man has no pubic hair.}

\fmtLemma{ki\-bu\-mbe}
\fmtCitation{(e)ki\-bu\-mbe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{piece of pottery}
\fmtExampleRL{Awo waa\-bba\-nga\-wo eki\-bu\-mbe kya mu\-ka\-ma.}
\fmtExampleEN{There used to be a piece of pottery for the chief.}

\fmtLemma{ki\-bu\-nguu\-ra}
\fmtCitation{(e)ki\-bu\-nguu\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{caterpillar}
\fmtExampleRL{Eki\-bu\-nguu\-ra ki\-tii\-ni\-sya abaa\-na.}
\fmtExampleEN{The caterpillar scares children.}

\fmtLemma{ki\-bu\-nwo\-mwei}
\fmtCitation{(e)ki\-bu\-nwo\-mwei}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{flying saucer}
\fmtExampleRL{Eki\-bu\-nwo\-mwei kya\-kai\-re.}
\fmtExampleEN{The flying saucer has passed.}

\fmtLemma{ki\-buu\-lyo}
\fmtCitation{(e)ki\-buu\-lyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{question, examination}
\fmtExampleRL{Ambwi\-rye ebi\-buu\-lyo bi\-sa\-tu.}
\fmtExampleEN{She asked me three questions.}

\fmtLemma{ki\-buu\-ra}
\fmtCitation{(o)ki\-buu\-ra}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{(of a woman) with tiny breasts}
\fmtExampleRL{Aba\-ka\-li ba\-ki\-buu\-ra ba\-bba n’a\-ma\-tai gai\-ngi.}
\fmtExampleEN{Women with small breasts usually have a lot of breast milk.}

\fmtLemma{ki\-buu\-zo}
\fmtCrossreference{\textnormal{see} kibuulyo}

\fmtLemma{ki\-bya}
\fmtCitation{(e)ki\-bya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{clay dish}
\fmtExampleRL{Nje ndii\-ra mu ki\-bya.}
\fmtExampleEN{I eat from a clay dish.}

\fmtLemma{ki\-ca\-nde\-ca\-nde}
\fmtCitation{(e)ki\-ca\-nde\-ca\-nde}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sunflower}
\fmtExampleRL{Obu\-tto tu\-mwi\-ya omu bi\-ca\-nde\-ca\-nde.}
\fmtExampleEN{We make oil from sunflowers.}

\fmtLemma{ki\-ca\-pa\-ti ibbu\-mba}
\fmtCrossreference{\textnormal{see also} capati}
\fmtCitation{(e)ki\-ca\-pa\-ti}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{clay sheet}
\fmtExampleRL{Nkoo\-ra\-yo eki\-ca\-pa\-ti kya ibbu\-mba nta\-ka ku\-bu\-mba ki\-na\-ga.}
\fmtExampleEN{Make me a clay sheet, I would like to make a pot.}

\fmtLemma{ki\-co\-la}
\fmtCitation{(e)ki\-co\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{curse, spell}
\fmtExampleRL{Omu\-sai\-za oyo ata n’e\-ki\-co\-la.}
\fmtExampleEN{That man has a spell on him.}

\fmtLemma{ki\-co\-nco}
\fmtCitation{(e)ki\-co\-nco}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{talent}
\fmtExampleRL{Ata n’e\-ki\-co\-nco kya kwe\-mba.}
\fmtExampleEN{She has a talent for music.}

\fmtLemma{ki\-coo\-bi}
\fmtCitation{(e)ki\-coo\-bi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{big pot used to keeping local brew}
\fmtExampleRL{Okwe\-te mu\-su\-ke omu ki\-coo\-bi.}
\fmtExampleEN{Pour the local brew into the big pot.}

\fmtLemma{ki\-coo\-li}
\fmtCitation{(e)ki\-coo\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{maize}
\fmtExampleRL{Ya\-ni\-ka ebi\-coo\-li bi\-ka\-le.}
\fmtExampleEN{Put the maize out to dry.}

\fmtLemma{ki\-cu}
\fmtCrossreference{\textnormal{see also} kiiguru}
\fmtCitation{(e)ki\-cu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{cloud}
\fmtExampleRL{Oku igu\-ru ndoo\-ku bi\-cu.}
\fmtExampleEN{There are no clouds in the sky.}
\fmtSubentry(b)
\fmtFixedexpression{ki\-cu kya ike\-ndi}
\fmtSense{nimbus cloud}

\fmtLemma{ki\-cu\-lo}
\fmtCitation{(e)ki\-cu\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{twitch (of any part of the body), muscle spasm}
\fmtExampleRL{Nku\-wuu\-ra eki\-cu\-lo oku rii\-so.}
\fmtExampleEN{I feel a twitch in my eye.}

\fmtLemma{ki\-cu\-mu\-ko}%
\fmtOne{1}
\fmtCitation{(e)ki\-cu\-mu\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bellows}
\fmtExampleRL{Eki\-cu\-mu\-ko ki\-rai?}
\fmtExampleEN{Where are the bellows for lighting the fire?}

\fmtLemma{ki\-cu\-mu\-ko}%
\fmtOne{2}
\fmtCitation{(e)ki\-cu\-mu\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{labelling tool}
\fmtExampleRL{Lee\-ta eki\-cu\-mu\-ko tu\-la\-mbe ebi\-do\-mo\-la byai\-swe.}
\fmtExampleEN{Bring the labelling tool and let's label our jerry cans.}

\fmtLemma{ki\-cuu\-cu}
\fmtPOS{adv}
\fmtSense{hastily, hurriedly, speedily}
\fmtExampleRL{Yai\-ru\-ki\-re ki\-cuu\-cu.}
\fmtExampleEN{He ran hastily.}

\fmtLemma{ki\-cwa\-ntai}
\fmtCitation{(e)ki\-cwa\-ntai}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sputum}
\fmtExampleRL{Eki\-cwa\-ntai kya\-la\-ngi\-re bu\-lwai\-re bwa ka\-su\-bba.}
\fmtExampleEN{The sputum revealed tuberculosis.}

\fmtLemma{ki\-cwe\-ka}
\fmtCitation{(e)ki\-cwe\-ka}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{part, portion, segment}
\fmtExampleRL{Eki\-cwe\-ka ki\-ni ki\-kya\-li ki\-ga\-zi.}
\fmtExampleEN{This part is still wide.}
\fmtSubentry(b)
\fmtSense{region, area}
\fmtExampleRL{Eki\-cwe\-ka kya bu\-gwai\-sa\-na ki\-koo\-to.}
\fmtExampleEN{The western region is big.}
\fmtSubentry(c)
\fmtSense{half}
\fmtExampleRL{Enca ba\-gya\-wii\-re\-mu ebi\-cwe\-ka bi\-bi\-ri.}
\fmtExampleEN{The country was divided into two halves.}
\fmtSubentry(d)
\fmtSense{(chapter of the Bible}
\fmtExampleRL{So\-ma omu Yo\-kaa\-na eki\-cwe\-ka ekya ka\-sa\-tu, olu\-nyi\-ri\-ri olwa iku\-mi na mu\-kaa\-ga.}
\fmtExampleEN{Read John, chapter three, verse sixteen.}

\fmtLemma{ki\-da}
\fmtCitation{(e)ki\-da}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{belly, stomach, abdomen}
\fmtExampleRL{Ya\-wu\-tai\-zi\-bwe eki\-da ei\-go\-lo.}
\fmtExampleEN{She hurt her belly yesterday.}

\fmtLemma{ki\-daa\-da}
\fmtCitation{(e)ki\-daa\-da}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{cirrhosis}
\fmtExampleRL{Mu\-lwai\-re wa ki\-daa\-da.}
\fmtExampleEN{He is suffering from cirrhosis.}

\fmtLemma{ki\-da\-ga\-do}
\fmtCitation{(e)ki\-da\-ga\-do}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{inner flap of a gomesi}
\fmtExampleRL{Aba\-ka\-li ba\-bii\-ka ese\-nte omu ki\-da\-ga\-do.}
\fmtExampleEN{Women keep money in the inner flap of a gomesi.}

\fmtLemma{ki\-da\-mbu\-la}
\fmtCitation{(e)ki\-da\-mbu\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from archaic}
\fmtSense{coin}
\fmtExampleRL{Mpee\-rya\-yo ebi\-da\-mbu\-la bi\-taa\-no.}
\fmtExampleEN{Give me five coins.}

\fmtLemma{ki\-da\-nda\-li}
\fmtCitation{(e)ki\-da\-nda\-li}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{social centre, community centre}
\fmtExampleRL{Baa\-bi\-re mu ki\-da\-nda\-li.}
\fmtExampleEN{They have gone to the social centre.}
\fmtSubentry(b)
\fmtSense{bar}
\fmtExampleRL{Aya\-bi\-re mu ki\-da\-nda\-li ku\-nywa mwe\-nge.}
\fmtExampleEN{She has gone to the bar to drink alcohol.}

\fmtLemma{ki\-dee\-ro}
\fmtCitation{(e)ki\-dee\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{granary}
\fmtExampleRL{Obwi\-ta bu\-te omu ki\-dee\-ro.}
\fmtExampleEN{Put the millet in the granary.}
\fmtSubentry(b)
\fmtSense{dictionary, lexicon}
\fmtExampleRL{Tu\-ku\-ko\-la ki\-dee\-ro kya Lu\-nya\-la n’o Ru\-ruu\-li.}
\fmtExampleEN{We are making a Ruruuli-Lu\-nyala dictionary.}
\fmtSubentry(c)
\fmtFixedexpression{ki\-dee\-ro kya ma\-ku\-ru}
\fmtSense{encyclopaedia}
\fmtExampleRL{Too\-lya eki\-ga\-mbo ekyo omu ki\-dee\-ro kya ma\-ku\-ru.}
\fmtExampleEN{Look up that word in the encyclopaedia.}

\fmtLemma{ki\-de\-wu}
\fmtCitation{(e)ki\-de\-wu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{weevil which eats dry meat}
\fmtExampleRL{Eki\-de\-wu ki\-bba ki\-ra\-gu\-zu.}
\fmtExampleEN{The weevil which eats dry meat is black.}

\fmtLemma{ki\-di}
\fmtCrossreference{\textnormal{see} (-)di}

\fmtLemma{ki\-di\-ba}
\fmtCitation{(e)ki\-di\-ba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{water pool}

\fmtLemma{ki\-di\-ma}
\fmtCitation{(e)ki\-di\-ma}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{hoe}
\fmtExampleRL{Akwai\-te ki\-di\-ma.}
\fmtExampleEN{He is digging. (lit.\,He is handling the hoe.)}

\fmtLemma{ki\-do\-mo}
\fmtCitation{(e)ki\-do\-mo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{bay, inlet, gulf}
\fmtExampleRL{Enya\-nja ya kyo\-ga eri\-na eki\-do\-mo.}
\fmtExampleEN{Lake Kyoga has a bay.}
\fmtSubentry(b)
\fmtSense{milk container}
\fmtExampleRL{Ama\-tai nga\-su\-ki\-re mu ki\-do\-mo.}
\fmtExampleEN{I poured the milk into the milk container.}

\fmtLemma{ki\-do\-mo\-la}
\fmtCitation{(e)ki\-do\-mo\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{jerry can}
\fmtExampleRL{Eki\-do\-mo\-la ki\-ni kya mai\-zi ota\-taa\-mu ma\-tai.}
\fmtExampleEN{This jerry can is for water, do not put milk in it.}

\fmtLemma{ki\-do\-ndo\-mpaa\-la}
\fmtCrossreference{\textnormal{see} kidongompaara}

\fmtLemma{ki\-do\-ndoo\-ro}
\fmtCitation{(e)ki\-do\-ndoo\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{type of pot (used to store maize, millet flour, as well as money for the occasion of unexpected needs)}
\fmtExampleRL{Omu nyu\-mba mu\-li\-mu eki\-do\-ndoo\-ro kya bu\-sye.}
\fmtExampleEN{There is a pot of millet flour in the house.}
\fmtSubentry(b)
\fmtSense{traditional purse}
\fmtExampleRL{Ese\-nte za me\-re nzi\-tai\-re mu ki\-do\-ndoo\-ro.}
\fmtExampleEN{I kept the money for the food in the purse.}

\fmtLemma{ki\-do\-ngo}
\fmtCitation{(e)ki\-do\-ngo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{music}
\fmtExampleRL{Eki\-do\-ngo ekyo ki\-ku\-ko\-koo\-mee\-ra ayi?}
\fmtExampleEN{Where is that music coming from?}
\fmtSubentry(b)
\fmtSense{musical instrument}
\fmtExampleRL{Eki\-do\-ngo ekyo mmai\-te oku\-ki\-ku\-bba.}
\fmtExampleEN{I know how to play that musical instrument.}
\fmtSubentry(c)
\fmtSense{disco}
\fmtExampleRL{Omwo\-jo ya\-lai\-re mu ki\-do\-ngo.}
\fmtExampleEN{The boy spent the night in the disco.}

\fmtLemma{ki\-do\-ngo\-mpaa\-ra}
\fmtCitation{(e)ki\-do\-ngo\-mpaa\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{type of poisonous caterpillar with spikes}
\fmtExampleRL{Obu\-lu\-ngu bwa ki\-do\-ngo\-mpaa\-ra bu\-bba bwa\-ndi.}
\fmtExampleEN{Caterpillar spikes are very dangerous.}

\fmtLemma{ki\-doo\-me}%
\fmtOne{1}
\fmtCitation{(e)ki\-doo\-me}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{calabash}
\fmtExampleRL{Amai\-zi nga\-su\-ki\-re mu ki\-doo\-me.}
\fmtExampleEN{I poured the water into a calabash.}

\fmtLemma{ki\-doo\-me}%
\fmtOne{2}
\fmtCitation{(e)ki\-doo\-me}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{path which goes through a papyrus swamp}
\fmtExampleRL{Eki\-doo\-me kya\-wo ki\-nya\-mu\-lai.}
\fmtExampleEN{The path through the papyrus swamp is long.}

\fmtLemma{ki\-du\-ba}
\fmtCitation{(e)ki\-du\-ba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{water pool}
\fmtExampleRL{Omwa\-na aku\-saa\-nyi\-ra mu ki\-du\-ba.}
\fmtExampleEN{The child is swimming in the water pool.}

\fmtLemma{ki\-du\-ka\-no}
\fmtCitation{(e)ki\-du\-ka\-no}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{diarrhoea}
\fmtExampleRL{Ya\-wei\-re\-mu amai\-zi omu mu\-bi\-ri lwa ki\-du\-ka\-no.}
\fmtExampleEN{He lost a lot of water due to diarrhoea.}

\fmtLemma{ki\-du\-ma}
\fmtCitation{(e)ki\-du\-ma}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{lung fish trap (trap used for trapping fish}

\fmtLemma{ki\-du\-mu}
\fmtCitation{(e)ki\-du\-mu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(b)
\fmtSense{culvert, drain, gutter}
\fmtExampleRL{Ebi\-du\-mu bi\-ju\-na oku\-ga\-du\-la amai\-zi oku\-zwa mu nzi\-ra.}
\fmtExampleEN{Culverts help to channel water away from the roads.}

\fmtLemma{ki\-du\-nduu\-lu}
\fmtCitation{(e)ki\-du\-nduu\-lu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{basket for fish}
\fmtExampleRL{Encwi zi\-te mu ki\-du\-nduu\-lu.}
\fmtExampleEN{Put the fish into the fish basket.}

\fmtLemma{ki\-duu\-du}
\fmtCitation{(e)ki\-duu\-du}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{coucal}
\fmtExampleRL{Eki\-duu\-du ki\-mai\-te oku\-pi\-ma obwi\-re.}
\fmtExampleEN{Coucals know how to keep time.}

\fmtLemma{ki\-duu\-ku\-lu}
\fmtCitation{(e)ki\-duu\-ku\-lu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{granary}
\fmtExampleRL{Eki\-duu\-ku\-lu kya\-nge ki\-kai\-ri\-ki\-re.}
\fmtExampleEN{My granary has gotten old.}

\fmtLemma{ki\-duu\-la}
\fmtCitation{(e)ki\-duu\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{fun group}
\fmtExampleRL{Omu\-sai\-za oyo ata\-ka mu\-no okwa\-ba omu bi\-duu\-la.}
\fmtExampleEN{That man really likes joining fun groups.}
\fmtSubentry(b)
\fmtSense{concert, concert hall}
\fmtExampleRL{Aya\-bi\-re mu ki\-duu\-la ku\-bi\-na.}
\fmtExampleEN{She has gone to a concert to dance.}

\fmtLemma{ki\-duu\-li}
\fmtCitation{(e)ki\-duu\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{the Kiruuli way, ways of the Baruuli people, the Kiruli style}
\fmtExampleRL{Abi\-ni\-re ki\-duu\-li.}
\fmtExampleEN{She danced in the Kiruuli style.}

\fmtLemma{ki\-duu\-ma}
\fmtCitation{(e)ki\-duu\-ma}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{maize}
\fmtExampleRL{Ontaa\-ma aku\-lya eki\-duu\-ma.}
\fmtExampleEN{The sheep is eating maize.}

\fmtLemma{ki\-duu\-ma\-duu\-ma}
\fmtCitation{(e)ki\-duu\-ma\-duu\-ma}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{plantation where maize has been harvested}
\fmtExampleRL{Ebi\-yaa\-ta mbi\-bboo\-ke\-re mu ki\-duu\-ma\-duu\-ma.}
\fmtExampleEN{I grew potatoes on the plantation that originally had maize.}

\fmtLemma{ki\-dye\-ri}
\fmtCitation{(e)ki\-dye\-ri}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{ferry}
\fmtExampleRL{Ata\-mbu\-lii\-re ku ki\-dye\-ri.}
\fmtExampleEN{She went with a ferry.}

\fmtLemma{ki\-faa\-fa}
\fmtCitation{(o)ki\-faa\-fa}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSubentry(a)
\fmtSense{animal which habitually loses its young ones}
\fmtExampleRL{Oki\-faa\-fa mu\-tu\-ndi\-re.}
\fmtExampleEN{I sold the animal which habitually loses its young ones.}
\fmtSubentry(b)
\fmtSense{carnivorous hen}
\fmtExampleRL{Onko\-ko oki\-faa\-fa alya ama\-wu\-li gaa\-bi\-ka.}
\fmtExampleEN{A carnivorous hen eats eggs which she lays.}
\fmtSubentry(c)
\fmtSense{useless person}
\fmtExampleRL{Odi omu\-ka\-li ki\-faa\-fa.}
\fmtExampleEN{That woman is useless.}

\fmtLemma{ki\-faa\-na\-nyi}%
\fmtOne{1}
\fmtCitation{(e)ki\-faa\-na\-nyi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{diagram, drawing, picture}
\fmtExampleRL{Ku\-bba ani eki\-faa\-na\-nyi kya Paa\-pa.}
\fmtExampleEN{Draw a picture of the Pope here.}
\fmtSubentry(b)
\fmtSense{photo}
\fmtExampleRL{Ki\-ni ki\-faa\-na\-nyi kya ngu\-li maa\-mwe.}
\fmtExampleEN{This is a photo of her late mother.}

\fmtLemma{ki\-fe\-ne}
\fmtCitation{(e)ki\-fe\-ne}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{jackfruit}
\fmtExampleRL{Eki\-fe\-ne ki\-bba n’a\-ma\-sa\-nda gai\-ngi.}
\fmtExampleEN{The jackfruit has a lot of sap.}

\fmtLemma{ki\-fi}
\fmtCitation{(e)ki\-fi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{piece of meat, slice of meat}
\fmtExampleRL{Nfu\-ni\-re ebi\-fi bi\-bi\-ri.}
\fmtExampleEN{I got two pieces of meat.}

\fmtLemma{ki\-fii\-ni}
\fmtCitation{(e)ki\-fii\-ni}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bad luck, misfortune}
\fmtExampleRL{Omwo\-jo oyo ata n’e\-ki\-fii\-ni.}
\fmtExampleEN{That boy has bad luck.}

\fmtLemma{ki\-fo}
\fmtCitation{(e)ki\-fo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{place, area}
\fmtExampleRL{Mmu\-fu\-nii\-re eki\-fo eki\-ndi ekya\-ku\-bba\-sya\-mu.}
\fmtExampleEN{I have secured another place for him to stay in.}
\fmtSubentry(b)
\fmtSense{fort, castle}
\fmtExampleRL{Bu\-li mu\-ka\-ma abba n’e\-ki\-fo kya\-mwe.}
\fmtExampleEN{Every king must have his own castle.}
\fmtSubentry(c)
\fmtSense{post, position}
\fmtExampleRL{Aye\-me\-rei\-re ku ki\-fo kya ise\-nte\-be.}
\fmtExampleEN{He is running for the post of chairperson.}
\fmtSubentry(d)
\fmtFixedexpression{(o)mu ki\-fo kya}
\fmtSense{instead of}
\fmtExampleRL{Abaa\-na ba\-ku\-gu\-lu\-ka\-mbe omu ki\-fo kya ku\-so\-ma.}
\fmtExampleEN{The children are simply jumping instead of studying.}

\fmtLemma{ki\-fu\-bba}
\fmtCrossreference{\textnormal{see} kisubba}

\fmtLemma{ki\-fu\-gi}
\fmtCitation{(e)ki\-fu\-gi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{veranda}
\fmtExampleRL{Ai\-kai\-re ku ki\-fu\-gi kya nyu\-mba yaa\-mwe.}
\fmtExampleEN{She is sitting on the veranda of her house.}

\fmtLemma{ki\-fu\-ka}
\fmtCitation{(e)ki\-fu\-ka}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{overnight leftovers}
\fmtExampleRL{Omu\-ge\-nyi ataa\-li mu\-la\-ge alya bi\-fu\-ka.}
\fmtExampleEN{An expected guest eats leftovers. (a Ruruuli-Lu\-nyala proverb)}

\fmtLemma{ki\-fu\-ndi\-kwa}
\fmtCitation{(e)ki\-fu\-ndi\-kwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{knot, tie}
\fmtExampleRL{Ya\-boi\-ne eki\-fu\-ndi\-kwa kya lu\-bu\-go.}
\fmtExampleEN{He saw a knot made of bark cloth.}

\fmtLemma{ki\-fu\-ndu\-kwa}
\fmtCitation{(e)ki\-fu\-ndu\-kwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{knot}
\fmtExampleRL{Omu\-kai\-re abbwe\-re eki\-fu\-ndu\-kwa ku go\-me\-si.}
\fmtExampleEN{The old woman had tied a knot on the gomesi.}
\fmtSubentry(b)
\fmtSense{official garment}
\fmtExampleRL{Omu\-gwe\-twa ba\-mu\-jwe\-ke\-re eki\-fu\-ndu\-kwa eky’o\-bu\-gwe\-te.}
\fmtExampleEN{The heir has been adorned in his official garment of inheritance.}

\fmtLemma{ki\-fu\-njo}
\fmtCitation{(e)ki\-fu\-njo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{rectum}
\fmtExampleRL{Ama\-bbi ga\-bba mu ki\-fu\-njo.}
\fmtExampleEN{Faeces are found in the rectum.}

\fmtLemma{ki\-fu\-ru\-kwa}
\fmtCitation{(e)ki\-fu\-ru\-kwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{abandoned house}
\fmtExampleRL{Ndi\-na abaa\-na be na\-boi\-ne nga ba\-ba\-si\-rye omu ki\-fu\-ru\-kwa.}
\fmtExampleEN{I saw some children sleeping in an abandoned house.}

\fmtLemma{ki\-fuu\-tuu\-lyo}
\fmtCitation{(e)ki\-fuu\-tuu\-lyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{location of a scuffle \eg for a fight or struggle}
\fmtExampleRL{Eki\-fuu\-tuu\-lyo ki\-ri mu mu\-si\-ri gwa bwi\-ta.}
\fmtExampleEN{The location of the scuffle was in the millet field.}

\fmtLemma{ki\-gaa\-li}%
\fmtOne{1}
\fmtCitation{(e)ki\-gaa\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wheelbarrow, wooden cart}
\fmtExampleRL{Ombu\-li omu\-lwai\-re ba\-mu\-twa\-lii\-re omu ki\-gaa\-li.}
\fmtExampleEN{The sick goat was carried in a wooden cart.}

\fmtLemma{ki\-gaa\-li}%
\fmtOne{2}
\fmtCitation{(e)ki\-gaa\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{foot fissure, foot crack, cracked heel}
\fmtExampleRL{Oyo mu\-lwai\-re wa bi\-gaa\-li.}
\fmtExampleEN{He is suffering from foot fissures.}

\fmtLemma{ki\-ga\-ba\-nyo}
\fmtCitation{(e)ki\-ga\-ba\-nyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{piece of gourd}
\fmtExampleRL{Eki\-ga\-ba\-nyo ki\-se\-na mwe\-nge.}
\fmtExampleEN{A piece of gourd is used to scoop the local brew.}

\fmtLemma{ki\-ga\-bo}
\fmtCitation{(o)ki\-ga\-bo}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{black cow with white patches}
\fmtExampleRL{Oki\-ga\-bo mu\-lwai\-re.}
\fmtExampleEN{The black cow with white patches is sick.}

\fmtLemma{ki\-ga\-di\-ga\-di}
\fmtCitation{(e)ki\-ga\-di\-ga\-di}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wheelbarrow, cart}
\fmtExampleRL{Ofe\-ne ba\-mu\-twa\-lli\-re omu ki\-ga\-di\-ga\-di.}
\fmtExampleEN{The jackfruit was transported in a wheelbarrow.}

\fmtLemma{ki\-ga\-gi}
\fmtCitation{(e)ki\-ga\-gi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{aloe vera}
\fmtExampleRL{Eki\-ga\-gi ki\-ta\-mba endwai\-re zi\-ngi.}
\fmtExampleEN{Aloe vera treats many diseases.}

\fmtLemma{ki\-ga\-la}
\fmtCitation{(e)ki\-ga\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sheath}
\fmtExampleRL{Ei\-su\-mu ali\-toi\-re omu ki\-ga\-la.}
\fmtExampleEN{He removed the spear from its sheath.}

\fmtLemma{ki\-ga\-la\-nga}
\fmtCitation{(e)ki\-ga\-la\-nga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{hem (for smoking opium)}
\fmtExampleRL{Atei\-re enjaa\-ye omu ki\-ga\-la\-nga.}
\fmtExampleEN{He put the opium in the hem.}

\fmtLemma{ki\-ga\-mbo}
\fmtCitation{(e)ki\-ga\-mbo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{word, issue}
\fmtExampleRL{Si\-moo\-la\-yo ebi\-ga\-mbo bi\-sa\-tu.}
\fmtExampleEN{Say three words.}
\fmtSubentry(b)
\fmtFixedexpression{ki\-ga\-mbo kya Ka\-nca}
\fmtSense{gospel}
\fmtExampleRL{Aya\-bi\-re ku\-lii\-sya ki\-ga\-mbo kya Ka\-nca.}
\fmtExampleEN{He has gone to preach the gospel.}

\fmtLemma{ki\-ga\-nda}%
\fmtOne{1}
\fmtCitation{(e)ki\-ga\-nda}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{pile, heap, stack}
\fmtExampleRL{Ki\-ni ki\-ga\-nda kya yo\-byo.}
\fmtExampleEN{This is a pile of spider weed.}

\fmtLemma{ki\-ga\-nda}%
\fmtOne{2}
\fmtCitation{(e)ki\-ga\-nda}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{Ganda way, style, code}
\fmtExampleRL{Enzwa\-la eni ya ki\-ga\-nda.}
\fmtExampleEN{This is the Ganda style of dressing.}

\fmtLemma{ki\-ga\-ngo}
\fmtCitation{(e)ki\-ga\-ngo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{temporary structure to serve as a tent or hall}

\fmtLemma{ki\-ga\-ni\-co}
\fmtCitation{(e)ki\-ga\-ni\-co}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{poem, verse}
\fmtExampleRL{Amai\-te oku\-to\-nto\-ma ebi\-ga\-ni\-co.}
\fmtExampleEN{He knows how to recite poems.}

\fmtLemma{ki\-ga\-nja}
\fmtCitation{(e)ki\-ga\-nja}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{palm (of a hand)}
\fmtExampleRL{Aba\-la\-gu\-li ba\-ke\-be\-ra ebi\-see\-ra ebi\-ri mu ki\-ga\-nja ne ba\-ko\-be\-ra omu\-ntu emi\-ki\-sa gya\-mwe.}
\fmtExampleEN{Fortune tellers use the lines on one's palm to foretell luck.}
\fmtSubentry(b)
\fmtSense{sole (of the foot)}
\fmtExampleRL{Ama\-wa ga\-mu\-su\-mi\-ti\-re oku ki\-ga\-nja.}
\fmtExampleEN{Thorns pricked the sole of his foot.}
\fmtSubentry(c)
\fmtSense{palm (as a measure)}
\fmtExampleRL{Olu\-baa\-wo lu\-ni ndu\-pi\-mi\-re\-ku ebi\-ga\-nja bi\-nai.}
\fmtExampleEN{This timber measures four palms wide.}

\fmtLemma{ki\-ga\-nyi\-ro}
\fmtCitation{(e)ki\-ga\-nyi\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{forgiveness, amnesty}
\fmtExampleRL{OMu\-se\-ve\-ni ya\-wai\-re eki\-ga\-nyi\-ro edi abaa\-li ba\-ye\-kee\-ra ba ADF.}
\fmtExampleEN{Museveni gave amnesty to the former ADF rebels.}

\fmtLemma{ki\-ga\-ra\-ga\-mba}
\fmtCitation{(e)ki\-ga\-ra\-ga\-mba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{scale (of fish and reptiles)}
\fmtExampleRL{Eki\-na\-ge ki\-bba n’e\-bi\-ga\-ra\-ga\-mba ebi\-koo\-to.}
\fmtExampleEN{Tilapia has big scales.}

\fmtLemma{ki\-ge\-ga}
\fmtCitation{(e)ki\-ge\-ga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{basket}
\fmtExampleRL{Su\-ka ebi\-yi\-mba omu ki\-ge\-ga.}
\fmtExampleEN{Pour the beans into the basket.}
\fmtSubentry(b)
\fmtSense{suitcase}
\fmtExampleRL{Omu\-ka\-li ata\-mbwi\-re n’a\-ka\-ge\-ga kaa\-mwe.}
\fmtExampleEN{The woman has left with her small suitcase.}

\fmtLemma{ki\-ge\-ndee\-rwa}
\fmtCitation{(e)ki\-ge\-ndee\-rwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{intention, intent, goal, objective}
\fmtExampleRL{Ebi\-ge\-ndee\-rwa bya\-mwe edi ei\-ya\-nga bi\-sai.}
\fmtExampleEN{He has good intentions for this country.}

\fmtLemma{ki\-ge\-nyi}
\fmtCitation{(e)ki\-ge\-nyi}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{meal for the spirits, feast for the spirits}
\fmtExampleRL{Ba\-ki\-ba\-ndwa ba\-koo\-ra emba\-ndwa eki\-ge\-nyi bu\-li lu\-zwa\-nyu\-ma lwa kai\-re.}
\fmtExampleEN{Traditional priests periodically prepare a feast for the spirits.}

\fmtLemma{ki\-ge\-re}
\fmtCitation{(e)ki\-ge\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from Lug.} 
\fmtSubentry(a)
\fmtSense{foot}
\fmtSubentry(b)
\fmtSense{peddle (of a bicycle)}
\fmtExampleRL{Eki\-ge\-re kya gaa\-li yaa\-mu ki\-tii\-ki\-re.}
\fmtExampleEN{The peddle of your bicycle has broken.}

\fmtLemma{ki\-ge\-ro}
\fmtCitation{(e)ki\-ge\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{size, weight}
\fmtExampleRL{Ti\-mmai\-te ki\-ge\-ro kya nkai\-to gy’a\-zwa\-la.}
\fmtExampleEN{I don’t know his shoe size.}
\fmtSubentry(b)
\fmtSense{appropriateness, suitability}
\fmtExampleRL{Aba\-ntu aba\-mwe ba\-kwe\-buu\-lya oku ki\-ge\-ro kya\-mwe oku bya\-kwe\-si\-mba\-wo.}
\fmtExampleEN{Some people are questioning the appropriateness of his candidature.}

\fmtLemma{ki\-gii\-si\-ri}
\fmtCitation{(e)ki\-gii\-si\-ri}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{rumen}
\fmtExampleRL{Aba\-ga\-nda ba\-ta\-ka mu\-no oku\-lya eki\-gii\-si\-ri.}
\fmtExampleEN{The Baganda love eating rumen.}

\fmtLemma{ki\-gi\-mu\-sa}
\fmtCitation{(e)ki\-gi\-mu\-sa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fertiliser}
\fmtExampleRL{Ei\-ta\-ka\-li li\-tee\-mu eki\-gi\-mu\-sa.}
\fmtExampleEN{Add fertilisers to the soil.}

\fmtLemma{ki\-gi\-ngi}
\fmtCitation{(e)ki\-gi\-ngi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wheel stopper}
\fmtExampleRL{Taa\-yo eki\-gi\-ngi emo\-to\-ka ere\-ke ku\-see\-see\-tu\-ka.}
\fmtExampleEN{Use a wheel stopper to stop the vehicle from sliding.}

\fmtLemma{ki\-gi\-ngi\-ryo}
\fmtCitation{(e)ki\-gi\-ngi\-ryo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{stretcher, gurney}
\fmtExampleRL{Ba\-mu\-twa\-lii\-re ku ki\-gi\-ngi\-ryo ku\-bba mu\-lwai\-re.}
\fmtExampleEN{He has been taken away on a stretcher because he is sick.}

\fmtLemma{ki\-gi\-no}
\fmtCitation{(e)ki\-gi\-no}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{larva, maggot}
\fmtExampleRL{Omu\-ye\-mbe gwe\-nguu\-re gu\-bbai\-re\-mu eki\-gi\-no.}
\fmtExampleEN{The mango I bought had a maggot inside.}
\fmtSubentry(b)
\fmtSense{maggot which causes myiasis}
\fmtExampleRL{Si\-nga eki\-gi\-no ki\-ku\-yi\-ngi\-ra omu mu\-bi\-ri ki\-yi\-nza oku\-lee\-te\-ra eki\-rwai\-re.}
\fmtExampleEN{Maggots can cause myiasis if they get into human skin.}

\fmtLemma{ki\-go}
\fmtCitation{(e)ki\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{fence}
\fmtExampleRL{Eka yaa\-mwe eri\-ku eki\-go.}
\fmtExampleEN{His house has a fence.}
\fmtSubentry(b)
\fmtSense{fort, fortress, citadel, stronghold}
\fmtExampleRL{Eki\-go kya mu\-ka\-ma oKoo\-jo kya\-li Ki\-ra\-sa.}
\fmtExampleEN{King Koojo’s fort was in Kirasa.}

\fmtLemma{ki\-go\-do}
\fmtCitation{(e)ki\-go\-do}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{skin (edible part of the skin of an animal, \eg the fatty part of pork)}
\fmtExampleRL{Ombe\-re\-ge abba n’e\-ki\-go\-do eki\-no\-li.}
\fmtExampleEN{The pig has fatty skin.}

\fmtLemma{ki\-go\-ga}
\fmtCitation{(e)ki\-go\-ga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{storehouse for threshed grain,}
\fmtExampleRL{Ama\-too\-ke mu\-ga\-tee\-ke oku ki\-go\-ga.}
\fmtExampleEN{Put the bananas into the granary.}

\fmtLemma{ki\-go\-go}
\fmtCitation{(e)ki\-go\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{banana plant, matoke plant}
\fmtExampleRL{Eki\-go\-go ti\-ki\-bba na mu\-zi gu\-lai.}
\fmtExampleEN{The banana plant has no tap root.}
\fmtSubentry(b)
\fmtSense{banana stem}
\fmtExampleRL{Mu\-ko\-le\-sye ebi\-go\-go oku\-ko\-la enko\-ndo za ki\-ga\-ngo.}
\fmtExampleEN{Use banana stems as poles to build a shade.}
\fmtSubentry(c)
\fmtSense{banana fibre}
\fmtExampleRL{Nko\-le\-sya bi\-go\-go oku\-lu\-ka ebii\-bbo.}
\fmtExampleEN{I use banana fibre to make baskets.}

\fmtLemma{ki\-goi\-goi}
\fmtCitation{(e)ki\-goi\-goi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sisal (\textit{Agave sisalana})}
\fmtExampleRL{Ega\-vu\-me\-nti etu\-ku\-bi\-rii\-ze oku\-lu\-ma ebi\-goi\-goi.}
\fmtExampleEN{The government is encouraging us to grow sisal.}

\fmtLemma{ki\-goi\-gwe}
\fmtCitation{(e)ki\-goi\-gwe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sisal bundle}
\fmtExampleRL{Eki\-goi\-gwe ki\-zi\-ri\-ngu\-ki\-re.}
\fmtExampleEN{The sisal bundle has uncoiled.}

\fmtLemma{ki\-goi\-ti}
\fmtCitation{(e)ki\-goi\-ti}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{pestle}
\fmtExampleRL{Eki\-goi\-ti kya\-nge ki\-kai\-ri\-ki\-re.}
\fmtExampleEN{My mashing pestle has grown old.}

\fmtLemma{ki\-go\-mba}
\fmtCitation{(e)ki\-go\-mba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{dried matoke}
\fmtExampleRL{Ebi\-go\-mba ba\-bi\-rya kwa nza\-la.}
\fmtExampleEN{Dried matoke is eaten during famine.}

\fmtLemma{ki\-go\-mbe}
\fmtCitation{(e)ki\-go\-mbe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{cemetery, graveyard, burial ground}
\fmtExampleRL{Ki\-ni ni\-kyo ki\-go\-mbe kya ba\-ka\-ma.}
\fmtExampleEN{This is the cemetery for royalty.}
\fmtSubentry(b)
\fmtSense{calf hutch}
\fmtExampleRL{Enya\-na zi\-ngii\-re omu ki\-go\-mbe.}
\fmtExampleEN{The calves entered the calf hutch.}

\fmtLemma{ki\-go\-ndi}
\fmtCitation{(e)ki\-go\-ndi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sagging back of the head}
\fmtExampleRL{Omu\-ka\-li ata n’e\-ki\-go\-ndi.}
\fmtExampleEN{The back of the woman's head is saggy.}

\fmtLemma{ki\-go\-ndya}
\fmtCitation{(e)ki\-go\-ndya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bare stem of a banana bunch}
\fmtExampleRL{Ebi\-go\-ndya ba\-bi\-ta ansi omu ki\-na\-ga nga ba\-kwa\-ba oku\-su\-mba.}
\fmtExampleEN{They put the bare banana stems into a saucepan when they are going to cook (them by steaming).}

\fmtLemma{ki\-go\-nge\-ryo}
\fmtCitation{(e)ki\-go\-nge\-ryo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{nuisance, annoyance, problem, annoying situation}
\fmtExampleRL{Olwa\-lee\-ro nfu\-ni\-re ebi\-go\-nge\-ryo bi\-bi\-ri.}
\fmtExampleEN{Today I was faced with two annoying situations.}

\fmtLemma{ki\-goo\-di}
\fmtCitation{(e)ki\-goo\-di}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{circle, ring}
\fmtExampleRL{Tu\-ko\-le eki\-goo\-di le\-ke tu\-bi\-ne.}
\fmtExampleEN{Let us make a circle and dance.}

\fmtLemma{ki\-goo\-do}
\fmtCrossreference{\textnormal{see} kigoodi}

\fmtLemma{ki\-goo\-li}%
\fmtOne{1}
\fmtCitation{(e)ki\-goo\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{hair style with hair on top and no hair at the bottom of the head}
\fmtExampleRL{Aba\-si\-ga\-zi ba bi\-ro bi\-ni be\-ndya mu\-no oku\-sa\-la eki\-goo\-li.}
\fmtExampleEN{Young people these days love to shave their hair in a style which leaves hair on top but no hair on the lower part of the head.}

\fmtLemma{ki\-goo\-li}%
\fmtOne{2}
\fmtCitation{(e)ki\-goo\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{lake plant with potato-like tubers eaten during famine}

\fmtLemma{ki\-gu}
\fmtCitation{(e)ki\-gu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{U-turn}
\fmtExampleRL{Ya\-koo\-re eki\-gu omu nzi\-ra aka\-ti.}
\fmtExampleEN{He made a U-turn in the middle of the road.}

\fmtLemma{ki\-gu\-bbi}
\fmtPOS{adj}
\fmtSense{dirty}
\fmtExampleRL{Omu\-sai\-za oyo bu\-lii\-zo abba ki\-gu\-bbi.}
\fmtExampleEN{That man is always dirty.}

\fmtLemma{ki\-gu\-gu\-nu\-bu\-lyo}
\fmtCitation{(e)ki\-gu\-gu\-nu\-bu\-lyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{peeled off bark of a tree}
\fmtExampleRL{Ebi\-gu\-gu\-nu\-bu\-lyo bya mu\-saa\-le ba\-bi\-su\-mbi\-sya.}
\fmtExampleEN{Peeled off tree bark is used for cooking.}

\fmtLemma{ki\-gu\-gu\-nya}
\fmtCitation{(e)ki\-gu\-gu\-nya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bottom of a pot}
\fmtExampleRL{Eki\-gu\-gu\-nya ki\-ku\-too\-nya.}
\fmtExampleEN{The bottom of the pot is leaking.}

\fmtLemma{ki\-gu\-ju}
\fmtCitation{(e)ki\-gu\-ju}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{young pumpkin}
\fmtExampleRL{Tu\-lii\-re bi\-gu\-ju olwa\-lee\-ro.}
\fmtExampleEN{We ate young pumpkins today.}

\fmtLemma{ki\-gu\-lu}
\fmtCitation{(e)ki\-gu\-lu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{foot}
\fmtExampleRL{Eyi\-wa li\-mu\-su\-mi\-ti\-re omu ki\-gu\-lu.}
\fmtExampleEN{His foot was pricked by a thorn.}

\fmtLemma{ki\-gu\-lu\-gu\-nya}
\fmtCitation{(e)ki\-gu\-lu\-gu\-nya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bare stem of a banana bunch}
\fmtExampleRL{Eki\-gu\-lu\-gu\-nya kya ma\-too\-ke nki\-ni.}
\fmtExampleEN{Here is the bare stalk from the banana bunch.}

\fmtLemma{ki\-gu\-lu\-ke}
\fmtCitation{(e)ki\-gu\-lu\-k(e)}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{patches of uncleared grass}
\fmtExampleRL{Endu\-mi\-ro eri\-mu ebi\-gu\-lu\-ke bi\-ngi mu\-no.}
\fmtExampleEN{The garden has many patches of uncleared grass.}

\fmtLemma{ki\-gu\-ngi\-ro}
\fmtCitation{(e)ki\-gu\-ngi\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{manger, feed trough}
\fmtExampleRL{OYe\-su ba\-mu\-bba\-si\-si\-rye mu ki\-gu\-ngi\-ro.}
\fmtExampleEN{Jesus was laid in the manger to sleep.}

\fmtLemma{ki\-gu\-nguu\-li}
\fmtCitation{(e)ki\-gu\-nguu\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{hump, bump}
\fmtExampleRL{Enzi\-ra eri\-mu ebi\-gu\-nguu\-li.}
\fmtExampleEN{There are bumps in the road.}

\fmtLemma{ki\-gu\-ru\-gu\-nya}
\fmtCitation{(e)ki\-gu\-ru\-gu\-nya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{hump, bump}
\fmtExampleRL{Enzi\-ra eya\-ba e Ibbaa\-le eri\-mu ebi\-gu\-ru\-gu\-nya.}
\fmtExampleEN{The road to Bbaale has many bumps.}

\fmtLemma{ki\-gu\-ru\-sa\-ni}
\fmtCitation{(e)ki\-gu\-ru\-sa\-ni}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{plant used to make beer straws (\textit{Clerodendrum capitatum} and \textit{Peucedanum grantii})}
\fmtExampleRL{Ebi\-gu\-ru\-sa\-ni ba\-bi\-ko\-la\-mu ense\-ke.}
\fmtExampleEN{They make straws from \textit{Clerodendrum capitatum}.}

\fmtLemma{ki\-gu\-ru\-suu\-ru}
\fmtCitation{(e)ki\-gu\-ru\-suu\-ru}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{\textit{Solanum incanum} (sodom apple)}
\fmtExampleRL{Ebi\-gu\-ru\-suu\-ru bi\-ko\-le\-se\-bwa nka omu\-ba\-zi.}
\fmtExampleEN{Sodom apples serve as medicine.}

\fmtLemma{ki\-gu\-ti}
\fmtCitation{(o)ki\-gu\-ti}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{Kiguti banana cultivar (usually eaten when ripe)}
\fmtExampleRL{Nku\-ta\-ka ku\-lya ku ki\-gu\-ti.}
\fmtExampleEN{I like eating banana of the Kiguti cultivar.}

\fmtLemma{ki\-guu\-nwa}
\fmtCitation{(e)ki\-guu\-nwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{ringworm, dermatophytosis, tinea (a contagious itching skin disease)}
\fmtExampleRL{Abaa\-na ba\-lwa\-la mu\-no ebi\-guu\-nwa.}
\fmtExampleEN{Children are prone to suffering from ringworm.}

\fmtLemma{ki\-gwa\-byo}
\fmtCitation{(e)ki\-gwa\-byo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{long knife used to prune banana leaves}
\fmtExampleRL{Enda\-ga\-la nzi\-sa\-li\-sya ki\-gwa\-byo.}
\fmtExampleEN{I use the pruning knife to cut banana leaves.}

\fmtLemma{ki\-gwai\-go\-lo}
\fmtCitation{(o)ki\-gwai\-go\-lo}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{person who goes to sleep early}
\fmtExampleRL{Omu\-sai\-za waa\-wo ki\-gwai\-go\-lo.}
\fmtExampleEN{That man goes to bed early.}

\fmtLemma{ki\-gwa\-ta}
\fmtCitation{(e)ki\-gwa\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{half gourd}
\fmtExampleRL{Omwe\-nge ngu\-se\-ne\-sya ki\-gwa\-ta.}
\fmtExampleEN{I use the half gourd to serve beer.}

\fmtLemma{ki\-gwei\-gwei}
\fmtCitation{(e)ki\-gwei\-gwei}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sisal, sisal leaf}
\fmtExampleRL{Obu\-ko\-si tu\-bu\-woo\-ga mu bi\-gwei\-gwei.}
\fmtExampleEN{Sisal threads are made from sisal leaves.}

\fmtLemma{ki\-gwe\-ki}
\fmtCitation{(e)ki\-gwe\-ki}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{knife-like tool used to clean a mingling stick}
\fmtExampleRL{Mpee\-rya eki\-gwe\-ki nsaa\-yii\-rye omu\-ga\-yi.}
\fmtExampleEN{Give me the knife-like tool and I'll clean the mingling stick.}

\fmtLemma{ki\-gwi}
\fmtCitation{(e)ki\-gwi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{woven trap}
\fmtExampleRL{Eki\-gwi kya\-kwai\-te enka\-yi\-ba ibi\-ri.}
\fmtExampleEN{The woven trap caught two doves.}

\fmtLemma{ki\-gwo}
\fmtCitation{(e)ki\-gwo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{fall (of a person)}
\fmtExampleRL{Agwi\-re eki\-gwo eki\-koo\-to nga akwi\-ngi\-ra mu nyu\-mba.}
\fmtExampleEN{She has had a heavy fall while entering the house.}
\fmtSubentry(b)
\fmtSense{wrestling}
\fmtExampleRL{Ti\-nta\-ka mu\-zee\-nyo gwa ki\-gwo.}
\fmtExampleEN{I hate the sport of wrestling.}

\fmtLemma{kii\-ba\-kii\-ba}
\fmtCitation{(oku)\-kii\-ba\-kii\-ba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kii\-bi\-re\-kii\-bi\-re}
\fmtSense{to evade the main path or entrance}
\fmtExampleRL{OKa\-to aku\-kii\-ba\-kii\-ba oku lu\-bu\-ga.}
\fmtExampleEN{Kato is avoiding the courtyard.}

\fmtLemma{kii\-bbai\-ta\-ka\-li}
\fmtCitation{(e)kii\-bbai\-ta\-ka\-li}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{land grabbing}
\fmtExampleRL{Ekii\-bbai\-ta\-ka\-li kya\-ka\-li\-rii\-rye omu Uga\-nda.}
\fmtExampleEN{Land grabbing is widespread in Uganda.}

\fmtLemma{kii\-bbo}
\fmtCitation{(e)kii\-bbo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{basket}
\fmtExampleRL{Ntee\-ra eme\-re omu kii\-bbo.}
\fmtExampleEN{Put the food into a basket for me.}

\fmtLemma{kii\-bu\-lo}
\fmtCitation{(e)kii\-bu\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{meal, food}
\fmtExampleRL{Ntee\-ke\-ra\-tee\-ke\-ra ekii\-bu\-lo, enza\-la eku\-ndu\-ma.}
\fmtExampleEN{Prepare me a meal, I am hungry.}

\fmtLemma{kii\-ca}%
\fmtOne{1}
\fmtCitation{(e)kii\-ca}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fruit of sausage tree (\textit{Kigelia africana})}
\fmtExampleRL{Ekii\-ca ki\-ta\-mba epu\-le\-sa.}
\fmtExampleEN{The fruit of the sausage tree cures high blood pressure.}

\fmtLemma{kii\-ca}%
\fmtOne{2}
\fmtCitation{(e)kii\-ca}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{parish (as a tier of administration)}
\fmtExampleRL{Nye\-te\-ra omwi\-ca wa kii\-ca ki\-ni.}
\fmtExampleEN{Call the chief of this parish for me.}

\fmtLemma{kii\-ga\-ra}
\fmtCitation{(o)kii\-ga\-ra}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSubentry(a)
\fmtSense{deaf person}
\fmtExampleRL{Okii\-ga\-ra asi\-moo\-la bai\-tu ta\-wuu\-ra.}
\fmtExampleEN{A deaf person talks but does not hear.}
\fmtSubentry(b)
\fmtSense{type of god, spirit responsible for making people deaf}

\fmtLemma{kii\-go}
\fmtCitation{(a)kii\-go}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{stick, cane}
\fmtExampleRL{Ya\-swa\-bwi\-re omwa\-na obwi\-go ku\-bba ya\-so\-me\-re ku\-bbi.}
\fmtExampleEN{He caned the child because he did not perform well in his studies.}

\fmtLemma{kii\-gu\-la\-lwi\-ge}
\fmtCitation{(e)kii\-gu\-la\-lwi\-ge}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{ceremonial local brew given to the bride's parents by the groom-to-be during the introduction ceremony}
\fmtExampleRL{Aba\-koi ba\-ree\-te\-re eki\-na\-ga eki\-koo\-to ekya kii\-gu\-la\-lwi\-ge.}
\fmtExampleEN{Their in-laws brought a big pot of ceremonial local brew.}

\fmtLemma{kii\-gu\-ru}
\fmtCrossreference{\textnormal{see also} kicu}
\fmtCitation{(e)kii\-gu\-ru}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{cloud}
\fmtExampleRL{Ebii\-gu\-ru bi\-zwa\-mu oi\-ke\-ndi.}
\fmtExampleEN{Clouds produce rain.}
\fmtSubentry(b)
\fmtFixedexpression{kii\-gu\-ru kya (e)ike\-ndi}
\fmtSense{nimbus cloud}
\fmtExampleRL{Ekii\-gu\-ru kya ike\-ndi ki\-kwai\-te twa\-ka\-fu\-na oi\-ke\-ndi eree\-ro.}
\fmtExampleEN{A nimbus cloud has formed, we may get rain today.}

\fmtLemma{kii\-ja\-nya}
\fmtCitation{(oku)\-kii\-ja\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kii\-jai\-nye}
\fmtSense{to mistreat, to ill-treat, to maltreat}
\fmtExampleRL{Omu\-ka\-li oyo akii\-ja\-nya abaa\-na baa\-mwe.}
\fmtExampleEN{That woman mistreats her children.}

\fmtLemma{kii\-ju\-ki\-ryo}
\fmtCitation{(e)kii\-ju\-ki\-ryo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{monument, memorial}
\fmtExampleRL{Aba\-ruu\-li baa\-ya\-mbo\-ke\-re ekii\-ju\-ki\-ryo kya Paa\-pa.}
\fmtExampleEN{The Baruuli constructed a monument in honour of the Pope.}

\fmtLemma{kii\-ka}
\fmtCrossreference{\textnormal{see also} mwika}
\fmtCitation{(e)kii\-ka}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{smoke, fume}

\fmtLemma{kii\-ka}%
\fmtOne{1}
\fmtCitation{(e)kii\-ka}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{craving, longing, yearning, desire, urge}
\fmtExampleRL{Ndi\-na ekii\-ka ekya ku\-lya oku ayi\-si\-ku\-rii\-mu.}
\fmtExampleEN{I crave ice cream.}

\fmtLemma{kii\-ka}%
\fmtOne{2}
\fmtCitation{(oku)\- kii\-ka}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{kii\-ki\-re}
\fmtSubentry(a)
\fmtSense{to position sideways, to put sideways}
\fmtExampleRL{Omwa\-ngo gu\-kii\-ke\-mbe.}
\fmtExampleEN{Just put the frame sideways.}
\fmtSubentry(b)
\fmtSense{to block, to obstruct, to congest}
\fmtExampleRL{Omu\-saa\-le ogwa\-gwi\-re ei\-go\-lo gu\-kii\-ki\-re mu\-lu\-guu\-do olu\-kwa\-ba e Ka\-mpa\-la.}
\fmtExampleEN{The road to Kampala is blocked by a big tree which fell during the rain.}
\fmtSubentry(c)
\fmtSense{to feel restless especially due to eating too much}
\fmtExampleRL{Omwa\-na ya\-lii\-re n’a\-kii\-ka.}
\fmtExampleEN{The child became restless due to eating too much.}

\fmtLemma{kii\-ka}%
\fmtOne{3}
\fmtCitation{(oku)\-kii\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kii\-ki\-re}
\fmtSense{to attend (e.g.\,a meeting)}
\fmtExampleRL{Aya\-bi\-re ku\-kii\-ka mu lu\-kii\-ko.}
\fmtExampleEN{He has gone to attend a meeting.}

\fmtLemma{kii\-kaa\-ro}
\fmtCitation{(e)kii\-kaa\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{seat}
\fmtExampleRL{Lee\-te\-ra omu\-ge\-nyi ekii\-kaa\-ro.}
\fmtExampleEN{Bring a seat for the visitor.}
\fmtSubentry(b)
\fmtSense{residence}
\fmtExampleRL{Ekii\-kaa\-ro kya\-mu ki\-ta ayi?}
\fmtExampleEN{Where is your place of residence?}

\fmtLemma{kii\-ka\-li}
\fmtCitation{(e)kii\-ka\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{residence, home, quarters}
\fmtExampleRL{Ki\-ni ni\-kyo kii\-ka\-li kya Ba\-ga\-ya.}
\fmtExampleEN{This is the residence of the Bagaya clan.}

\fmtLemma{kii\-ki}
\fmtPOS{quant}
\fmtSense{a lot of, too many, too much}
\fmtExampleRL{Empii\-ya ata na zi\-kii\-ki.}
\fmtExampleEN{She has a lot of money.}

\fmtLemma{kii\-kii\-ra}
\fmtCitation{(oku)\-kii\-kii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ki\-kii\-re, kii\-kii\-re}
\fmtSense{to represent, to speak on behalf of}
\fmtExampleRL{Ya\-ba otu\-kii\-kii\-re.}
\fmtExampleEN{Go and represent us.}

\fmtLemma{kii\-kya}
\fmtCitation{(oku)\-kii\-kya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kii\-ki\-rye}
\fmtSense{to overdo}
\fmtExampleRL{Eme\-re ya\-gi\-su\-mbi\-re n’a\-kii\-kya.}
\fmtExampleEN{She overcooked the food.}

\fmtLemma{kii\-na}%
\fmtOne{1}
\fmtCitation{(e)kii\-na}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{pit, hole}
\fmtExampleRL{Ombu\-li agwi\-re omu kii\-na.}
\fmtExampleEN{A goat has fallen into a pit.}
\fmtSubentry(b)
\fmtSense{grave, tomb}
\fmtExampleRL{Ba\-ba\-lii\-ki\-re mu kii\-na ki\-mwei.}
\fmtExampleEN{They were buried in one grave.}

\fmtLemma{kii\-na}%
\fmtOne{2}
\fmtCitation{(oku)\-kii\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kii\-ni\-re}
\fmtSense{to deride, to sneer at, to mock, to speak sarcastically to}
\fmtExampleRL{Lwa\-ki oku\-mu\-kii\-na na\-mbu mu\-guu\-da?}
\fmtExampleEN{Why do you mock her for being rich?}

\fmtLemma{kii\-na\-mo}
\fmtCitation{(e)kii\-na\-mo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{valley}
\fmtExampleRL{Omu\-gwai\-ra\-lu ayo\-mbo\-ke\-re enyu\-mba mu kii\-na\-mo.}
\fmtExampleEN{The madman has built a house in a valley.}

\fmtLemma{kii\-nii\-ka}
\fmtCitation{(oku)\-kii\-nii\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kii\-nii\-ki\-re}
\fmtSense{to laugh mockingly}
\fmtExampleRL{Oyo aku\-ku\-kii\-nii\-ka\-mbe.}
\fmtExampleEN{That one is just laughing at you mockingly.}

\fmtLemma{kii\-ni\-ki\-ra}
\fmtCitation{(oku)\-kii\-ni\-ki\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kii\-ni\-kii\-re}
\fmtSense{to cover}
\fmtExampleRL{Kii\-ni\-ki\-ra omu\-ba\-zi ogwo.}
\fmtExampleEN{Cover that medicine.}

\fmtLemma{kii\-nu\-ku\-ka}
\fmtCitation{(oku)\-kii\-nu\-ku\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kii\-nu\-ku\-ki\-re}
\fmtSense{to be uncovered}
\fmtExampleRL{Eki\-na\-ga ki\-kii\-nu\-ku\-ki\-re.}
\fmtExampleEN{The pot is uncovered.}

\fmtLemma{kii\-nu\-ku\-la}
\fmtCitation{(oku)\-kii\-nu\-ku\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kii\-nu\-kwi\-re}
\fmtSense{to uncover, to unveil, to remove a lid}
\fmtExampleRL{Kii\-nu\-ku\-la\-ku tu\-bo\-ne eki\-ri mu\-nda.}
\fmtExampleEN{Uncover it and we'll see what is inside.}

\fmtLemma{kii\-nyi}
\fmtCitation{(e)kii\-nyi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{private parts, loin area}
\fmtExampleRL{Obwi\-ri\-za bu\-me\-ra ku kii\-nyi.}
\fmtExampleEN{Pubic hair develops around the private parts.}

\fmtLemma{Kii\-ra}
\fmtCitation{(o)Kii\-ra}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{River Nile}
\fmtExampleRL{Omwi\-ga gwa Kii\-ra gu\-ta\-ndi\-ki\-ra mu Uga\-nda.}
\fmtExampleEN{The River Nile starts in Uganda.}

\fmtLemma{kii\-roi\-ro}
\fmtCitation{(e)kii\-roi\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fireworks}
\fmtExampleRL{Oku lwa\-ku\-ba\-nza aba\-ntu ba\-na\-ga ebii\-roi\-ro.}
\fmtExampleEN{People shoot fireworks on New Year's Eve.}

\fmtLemma{kii\-ru\-ka}
\fmtCitation{(e)kii\-ru\-ka}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{vehicle, car}
\fmtExampleRL{Aki\-ri gu\-la\-yo ekii\-ru\-ka ekya\-twa\-la\-nga baa\-na oku iso\-me\-ro.}
\fmtExampleEN{At least buy a vehicle for taking the children to school.}

\fmtLemma{Kii\-rya}
\fmtCitation{(o)Kii\-rya}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{name given to a boy who follows twins}
\fmtExampleRL{OKii\-rya ai\-ra ku ba\-lo\-ngo.}
\fmtExampleEN{Kiirya was born after the twins.}

\fmtLemma{kii\-sa\-ni}
\fmtCitation{(e)kii\-sa\-ni}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{image}
\fmtExampleRL{Ekii\-sa\-ni kya Bi\-ki\-ra Ma\-ria Ki\-sai.}
\fmtExampleEN{The image of the Holy Mary is nice.}
\fmtSubentry(b)
\fmtSense{idol, statue}
\fmtExampleRL{Aba\-ntu aba\-mwe ba\-ra\-mya ebii\-sa\-ni.}
\fmtExampleEN{Some people worship idols.}
\fmtSubentry(c)
\fmtSense{photograph, photo}
\fmtExampleRL{Ekii\-sa\-ni ki\-ni ki\-bo\-ne\-ke\-re ku\-sai.}
\fmtExampleEN{This photograph looks nice.}

\fmtLemma{kii\-ti}
\fmtCitation{(e)kii\-ti}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{weapon}
\fmtExampleRL{Lwa\-ki ota\-mbu\-la n’e\-kii\-ti omu ba\-ntu?}
\fmtExampleEN{Why do you go around with a weapon in public?}

\fmtLemma{kii\-ti\-ro}
\fmtCitation{(e)kii\-ti\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{place of massacre}
\fmtExampleRL{OLu\-wee\-ro ya\-foo\-ke\-re ekii\-ti\-ro omu 1985.}
\fmtExampleEN{Luweero was the site of a massacre in 1985.}

\fmtLemma{kii\-tuu\-ru}
\fmtCitation{(e)kii\-tuu\-ru}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{shadow, shade}
\fmtExampleRL{Ikaa\-ra\-ku awo omu kii\-tuu\-ru.}
\fmtExampleEN{Sit in the shade for a while please.}
\fmtSubentry(b)
\fmtSense{umbrella organisation (an organisation that coordinates the activities of a number of member organisations)}
\fmtExampleRL{Ba\-li mu kii\-tuu\-ru ki\-mwe.}
\fmtExampleEN{They are part of the same umbrella organisation.}

\fmtLemma{kii\-yo}
\fmtCitation{(e)kii\-yo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{miracle, wonder, marvel, supernatural phenomenon}
\fmtExampleRL{Ebi\-ro bi\-ni aba\-su\-mba aba\-mwe ba\-ko\-la ebii\-yo ebi\-pa\-ngii\-re.}
\fmtExampleEN{These days some pastors perform staged fake miracles.}

\fmtLemma{Kii\-za}
\fmtCitation{(o)Kii\-za}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{name given to a child who follows twins}

\fmtLemma{kii\-zi\-ka}
\fmtCitation{(oku)\-kii\-zi\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kii\-zi\-ki\-re}
\fmtSense{to lay (down), to put (down), to set (down)}
\fmtExampleRL{Kii\-zi\-ka omwa\-na omu ki\-ta\-bu.}
\fmtExampleEN{Lay the child down in bed.}

\fmtLemma{kii\-zu\-ki\-ryo}
\fmtCitation{(e)kii\-zu\-ki\-ryo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{commemoration}
\fmtExampleRL{Abu\-mbi\-re ekii\-sa\-ni kya kii\-zu\-ki\-ryo.}
\fmtExampleEN{He sculpted a commemoration obelisk.}

\fmtLemma{ki\-jaa\-ye\-jaa\-ye}
\fmtCitation{(e)ki\-jaa\-ye\-jaa\-ye}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{plant used to treat malaria (its leaves looks like leaves of a marijuana plant)}

\fmtLemma{ki\-ja\-lee\-ro}
\fmtCitation{(e)ki\-ja\-lee\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{lima bean (\textit{Phaseolus lunatus})}
\fmtExampleRL{Ebi\-ja\-lee\-ro ki\-ra\-ndi\-ra ku mu\-saa\-le.}
\fmtExampleEN{Lima bean wines creep on trees.}

\fmtLemma{Ki\-ja\-mbi\-ya}
\fmtCitation{(o)Ki\-ja\-mbi\-ya}
\fmtCrossreference{\textnormal{see also} ja\-mbi\-ya}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{nickname of the former president of Uganda Idi Amin Dada believed to have killed many innocent people, it derives from \textit{jambiya} `machete'}

\fmtLemma{ki\-ja\-njaa\-lo}
\fmtCitation{(e)ki\-ja\-njaa\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bean}
\fmtExampleRL{Ebi\-ja\-njaa\-lo bya\-ki\-rye.}
\fmtExampleEN{The beans have flowered.}

\fmtLemma{ki\-je\-bu}
\fmtCitation{(e)ki\-je\-bu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{intrusiveness, pushiness}
\fmtExampleRL{Eki\-je\-bu ti mpi\-sa gi\-sai omu baa\-na.}
\fmtExampleEN{Intrusiveness is not a good practice among children.}

\fmtLemma{ki\-je\-ge\-ju}
\fmtCitation{(e)ki\-je\-ge\-ju}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{beetle}
\fmtExampleRL{Oje\-ge\-ju ane\-na emi\-saa\-le.}
\fmtExampleEN{Beetles eat trees.}

\fmtLemma{ki\-je\-gu}
\fmtCitation{(e)ki\-je\-gu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{intrusive memory}
\fmtExampleRL{Nno\-ba eki\-je\-gu ekya\-kwi\-zu\-ki\-ra omu\-ku\-nzi wa\-nge eya\-fii\-re.}
\fmtExampleEN{I hate the intrusive memory of my beloved one who died.}

\fmtLemma{ki\-jii\-ko}
\fmtCitation{(e)ki\-jii\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from Sw.}
\fmtSense{spoon}
\fmtExampleRL{Eki\-jii\-ko ki\-ni ki\-se\-na su\-kaa\-li.}
\fmtExampleEN{This spoon is used to scoop sugar.}

\fmtLemma{ki\-jo}
\fmtCitation{(e)ki\-jo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from Buyende dialect}
\fmtSense{sweet potato}
\fmtExampleRL{Tu\-kwe\-ndya ku\-lya bi\-jo.}
\fmtExampleEN{We would like to eat sweet potatoes.}

\fmtLemma{ki\-jo\-loo\-to}
\fmtCitation{(e)ki\-jo\-loo\-to}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{monster, beast}
\fmtExampleRL{Omu\-sai\-za oyo ai\-sa\-na ki\-jo\-loo\-to.}
\fmtExampleEN{That man looks like a monster.}

\fmtLemma{ki\-jo\-ro\-ko\-ko}%
\fmtOne{1}
\fmtCitation{(e)ki\-jo\-ro\-ko\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{straw container}
\fmtExampleRL{Eki\-jo\-ro\-ko\-ko kye mbii\-ka\-mu ense\-ke ki\-ri ayi?}
\fmtExampleEN{Where is the container where I keep my straws?}

\fmtLemma{ki\-jo\-ro\-ko\-to}%
\fmtOne{2}
\fmtCitation{(e)ki\-jo\-ro\-ko\-to}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{tree where cleansing takes place}
\fmtExampleRL{Omu\-lwai\-re mu\-mu\-twa\-le oku ki\-jo\-ro\-ko\-to.}
\fmtExampleEN{Take the patient to the cleansing tree.}

\fmtLemma{ki\-ju}
\fmtCitation{(e)ki\-ju}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{shade, goat house}
\fmtExampleRL{Eki\-ju kya mbu\-li ki\-ge\-gei\-re.}
\fmtExampleEN{The goat's house is very dirty.}

\fmtLemma{ki\-ju\-gu\-mi\-ro}
\fmtCitation{(e)ki\-ju\-gu\-mi\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{shivering, trembling}
\fmtExampleRL{Eki\-ju\-gu\-mi\-ro ti\-kya\-wee\-wo okwi\-ca nga onywi\-re eki\-ntu eki\-kwo\-ca.}
\fmtExampleEN{The shivering will not stop until you drink something warm.}

\fmtLemma{ki\-ju\-gu\-ta}
\fmtCitation{(e)ki\-ju\-gu\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bellows (a device with an air bag that emits a stream of air used for blowing air into a fire)}
\fmtExampleRL{Nkwe\-ndya\-yo eki\-ju\-gu\-ta eki\-yaa\-ka kya kyo\-to.}
\fmtExampleEN{I need a new bellows for my furnace.}

\fmtLemma{ki\-ju\-li\-za}
\fmtCitation{(e)ki\-ju\-li\-za}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{after party meal, meal served the day after the main function}
\fmtExampleRL{Mu\-su\-mbi\-re aba\-ge\-nyi eki\-ju\-li\-za baa\-be.}
\fmtExampleEN{Cook an after-party meal so that the visitors can leave.}

\fmtLemma{ki\-ju\-mbi}
\fmtCitation{(e)ki\-ju\-mbi}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{rainy season}
\fmtExampleRL{OMu\-ju\-mbi ya\-bya\-lii\-bwe mu ki\-ju\-mbi.}
\fmtExampleEN{Muju\-mbi was born during the rainy season.}

\fmtLemma{ki\-ju\-mbu\-ra}
\fmtCitation{(e)ki\-ju\-mbu\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{cow parsnip, hogweed}
\fmtExampleRL{Eki\-ju\-mbu\-ra kyo\-ca si\-nga ki\-ku\-koo\-na oku lu\-su\-su.}
\fmtExampleEN{Cow parsnip can burn if it comes into contact with skin.}

\fmtLemma{ki\-ju\-mbu\-tu\-ko}
\fmtPOS{adv}
\fmtSense{hurriedly, hastily, speedily, quickly}
\fmtExampleRL{Po\-lii\-si ya\-mu\-twai\-re ki\-ju\-mbu\-tu\-ko.}
\fmtExampleEN{He was taken hurriedly by the police.}

\fmtLemma{ki\-ju\-mu\-la}
\fmtCitation{(e)ki\-ju\-mu\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{jatropha}
\fmtExampleRL{Ebi\-ju\-mu\-la bi\-ko\-la nk’o\-bu\-bo\-ne\-ro bwa nsa\-lo.}
\fmtExampleEN{Jatropha plants are used as boundary markers.}

\fmtLemma{ki\-ju\-ndi}
\fmtCitation{(e)ki\-ju\-ndi}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{bad smell}
\fmtExampleRL{Eki\-ju\-ndi ka kya maa\-ni!}
\fmtExampleEN{The smell is too much!}

\fmtLemma{ki\-ju\-ngu}
\fmtCitation{(e)ki\-ju\-ngu}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{European style}
\fmtExampleRL{Eyo nko\-la ya ki\-ju\-ngu.}
\fmtExampleEN{That is the European style of doing things.}

\fmtLemma{ki\-juu\-li}
\fmtCitation{(e)ki\-juu\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{dewlap (a loose flap of the skin hanging from the neck of some animals)}
\fmtExampleRL{Ongu\-ndu wa\-nge ali\-na eki\-juu\-li ki\-koo\-to.}
\fmtExampleEN{The bull has a big dewlap.}

\fmtLemma{ki\-ka}
\fmtCitation{(e)ki\-ka}
\fmtPOS{n}%
\fmtNC{7}
\fmtSubentry(a)
\fmtSense{kind, type, sort}
\fmtExampleRL{Oli mu\-ntu ki\-ka ki?}
\fmtExampleEN{What kind of person are you?}
\fmtSubentry(b)
\fmtSense{clan}
\fmtExampleRL{Eki\-ka kya\-nge nta Mu\-ko\-mbe.}
\fmtExampleEN{I belong to the Kikombe clan.}

\fmtLemma{ki\-kaa\-bu\-go}
\fmtCitation{(e)ki\-kaa\-bu\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{piece of land from which cotton has been harvested}
\fmtSubentry(b)
\fmtSense{house structure woven with reeds before filling it with mud}
\fmtSubentry(c)
\fmtSense{draft}
\fmtExampleRL{Ko\-la eki\-kaa\-bu\-go kya bba\-lu\-wa.}
\fmtExampleEN{Write a draft of the letter.}

\fmtLemma{ki\-kaa\-gi\-ri}
\fmtCitation{(e)ki\-kaa\-gi\-ri}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{style of the Bakaagiri, denoting something done smartly and nicely}

\fmtLemma{ki\-kaa\-kaa\-te}
\fmtCitation{(e)ki\-kaa\-kaa\-te}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{piece of cloth or bark cloth tied around the waist as a sign of mourning}

\fmtLemma{ki\-kaa\-li}
\fmtCitation{(e)ki\-kaa\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{enclosure, palace}
\fmtExampleRL{Eki\-kaa\-li kya Mu\-ka\-ma wa Bu\-nyo\-ro ki\-ri Ka\-ru\-zii\-ka.}
\fmtExampleEN{The King of Bu\-nyoro's palace is in Karuziika.}
\fmtSubentry(b)
\fmtSense{casino}
\fmtExampleRL{Baa\-bi\-re mu ki\-kaa\-li ku\-zee\-nya ca\-ni\-si.}
\fmtExampleEN{They have gone to the casino to play cards.}

\fmtLemma{ki\-kaa\-ra}
\fmtCitation{(e)ki\-kaa\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{residence for people of the Bushbuck clan}

\fmtLemma{ki\-kaa\-ta}
\fmtCitation{(e)ki\-kaa\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{dry flaky skin patch, fungal skin infection}
\fmtExampleRL{Aba\-ntu aba\-bba oku nya\-nja ba\-bba ne bi\-kaa\-ta.}
\fmtExampleEN{People who live near the lake often develop dry flaky patches on their skin.}

\fmtLemma{ki\-kai\-jo}
\fmtCrossreference{\textnormal{see} kikaizo}

\fmtLemma{ki\-kai\-zo}
\fmtCitation{(e)ki\-kai\-zo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sugarcane}
\fmtExampleRL{Eki\-kai\-zo ki\-nu\-lii\-rya nga ki\-kuu\-re.}
\fmtExampleEN{Sugar cane is sweet when it has matured.}

\fmtLemma{ki\-ka\-kaa\-ra}
\fmtCitation{(e)ki\-ka\-kaa\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{giant, extremely hard working person}
\fmtExampleRL{Bo\-na eki\-ka\-kaa\-ra kya mu\-sai\-za aku\-bi\-na omu\-no!}
\fmtExampleEN{Look at the giant man dancing with big strides!}

\fmtLemma{ki\-ka\-ka\-mu}
\fmtCitation{(e)ki\-ka\-ka\-mu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{dried nasal mucus, booger}
\fmtExampleRL{Ya\-ba otoo\-le\-mu ebi\-ka\-ka\-mu omu nyi\-ndo yaa\-mu.}
\fmtExampleEN{Go and clean the dried mucus from your nose.}
\fmtSubentry(b)
\fmtSense{hardened layer of mingled food which gets stuck to the pan}
\fmtExampleRL{Too\-la\-mu ebi\-ka\-ka\-mu omu se\-fu\-li\-ya eya\-mye\-ne\-re obwi\-ta.}
\fmtExampleEN{Scoop out the hardened layer from the saucepan in which the millet was mingled.}

\fmtLemma{ki\-ka\-ka\-ra}
\fmtCitation{(e)ki\-ka\-ka\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{Pseudarthria hookeri}

\fmtLemma{ki\-ka\-la}
\fmtCitation{(e)ki\-ka\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{smell of a he-goat}
\fmtExampleRL{Omu\-ka\-li oyo awu\-nya eki\-ka\-la.}
\fmtExampleEN{That woman smells like a he-goat.}

\fmtLemma{ki\-ka\-la\-ka\-la}
\fmtCitation{(e)ki\-ka\-la\-ka\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{overgrown bush, fallow}
\fmtExampleRL{Nta\-ndi\-ki\-re oku\-saa\-wa omu ki\-ka\-la\-ka\-la.}
\fmtExampleEN{I started slashing the overgrown bush.}

\fmtLemma{ki\-ka\-la\-ka\-mpa}
\fmtCitation{(e)ki\-ka\-la\-ka\-mpa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{scale (of fish and reptiles)}
\fmtExampleRL{Ompyo abba n’e\-bi\-ka\-la\-ka\-mpa.}
\fmtExampleEN{The crocodile has scales.}

\fmtLemma{ki\-ka\-li\-fuu\-wa}
\fmtCitation{(e)ki\-ka\-li\-fuu\-wa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{lemon grass, cymbocogon}
\fmtExampleRL{Eki\-ka\-li\-fuu\-wa ki\-cuu\-nyi\-sya ocaa\-yi.}
\fmtExampleEN{The lemon grass spices up tea.}

\fmtLemma{ki\-ka\-lu}
\fmtCitation{(e)ki\-ka\-lu}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{tucked-in manner}
\fmtExampleRL{OSe\-bwa\-to mu\-sai\-za mu\-ku\-ni bu\-lii\-zo aku\-bba esaa\-ti eki\-ka\-lu.}
\fmtExampleEN{Sebwato is a smart man, he always tucks in his shirt.}

\fmtLemma{ki\-ka\-ma}
\fmtCitation{(e)ki\-ka\-ma}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{milking container}
\fmtExampleRL{Lee\-ta ani eki\-ka\-ma nka\-me ente.}
\fmtExampleEN{Bring the milking container here and I'll milk.}

\fmtLemma{ki\-ka\-mi\-ro}
\fmtCitation{(e)ki\-ka\-mi\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fungal skin infection}
\fmtExampleRL{Ebi\-ka\-mi\-ro bi\-kwa\-ta baa\-na ba\-to.}
\fmtExampleEN{Children often suffer from fungal skin infections.}

\fmtLemma{ki\-ka\-mu\-lo}
\fmtCitation{(e)ki\-ka\-mu\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{residue, remainder (especially of squashed bananas and grass after beer making)}
\fmtExampleRL{Na\-ga\-yo ebi\-ka\-mu\-lo bya nsa\-nde ebyo.}
\fmtExampleEN{Throw away the banana leaf residues.}
\fmtSubentry(b)
\fmtSense{unironed item of clothing, item of clothing with creases, wrinkled eitem of clothes}
\fmtExampleRL{Twe\-wuu\-nyi\-re oku\-bo\-na omu\-go\-le ng’a\-zwai\-re bi\-ka\-mu\-lo.}
\fmtExampleEN{We were all surprised to see the bride dressed in wrinkled clothes.}

\fmtLemma{ki\-ka\-nda}
\fmtCitation{(e)ki\-ka\-nda}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{basket}
\fmtExampleRL{Obwi\-ta bu\-te mu ki\-ka\-nda.}
\fmtExampleEN{Put the millet into the basket.}
\fmtSubentry(b)
\fmtSense{fish tied into bundles (\eg a bundle of sprat)}
\fmtExampleRL{Nku\-ta\-ka ku\-gu\-la\-yo ki\-ka\-nda kya mpe\-re.}
\fmtExampleEN{I would like to buy a bundle of sprats.}

\fmtLemma{ki\-ka\-nde}
\fmtCitation{(e)ki\-ka\-nde}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{overgrown bush, fallow}
\fmtExampleRL{Aku\-lu\-ma mu ki\-ka\-nde.}
\fmtExampleEN{He is clearing the overgrown bush.}

\fmtLemma{ki\-ka\-ndu\-lyo}
\fmtCitation{(e)ki\-ka\-ndu\-lyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{garden fork}
\fmtExampleRL{Ebi\-su\-bi aku\-bi\-so\-roo\-le\-sya ki\-ka\-ndu\-lyo.}
\fmtExampleEN{He is gathering grass with a garden fork.}

\fmtLemma{ki\-ka\-nja}
\fmtCitation{(e)ki\-ka\-nja}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{residue of local brew, impurities in a beverage}
\fmtExampleRL{Su\-ka\-yo ebi\-ka\-nja bya ma\-lwa.}
\fmtExampleEN{Dispose of the residue of the local brew.}

\fmtLemma{ki\-ka\-nya\-ba\-ntu}
\fmtCitation{(e)ki\-ka\-nya\-ba\-ntu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{city, town}
\fmtExampleRL{Aya\-bi\-re mu ki\-ka\-nya\-ba\-ntu.}
\fmtExampleEN{He has gone to the city.}

\fmtLemma{ki\-ka\-pu}
\fmtCitation{(e)ki\-ka\-pu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{woven handbag}
\fmtExampleRL{Aba\-kai\-re ba\-ta\-ka mu\-no oku\-ta\-mbu\-la n’e\-bi\-ka\-pu.}
\fmtExampleEN{Older people like carrying woven handbags.}

\fmtLemma{ki\-ka\-ra}%
\fmtOne{1}
\fmtCitation{(e)ki\-ka\-ra}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{bad smell, body odour}
\fmtExampleRL{Omu\-sai\-za oyo aku\-wu\-nya eki\-ka\-ra.}
\fmtExampleEN{That man has bad body odour.}

\fmtLemma{ki\-ka\-ra}%
\fmtOne{2}
\fmtCitation{(o)ki\-ka\-ra}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{cow with black and brown stripes from tail to head with patches around the face}

\fmtLemma{ki\-ka\-ra\-kaa\-ta}
\fmtCitation{(e)ki\-ka\-ra\-kaa\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{scaly skin}
\fmtExampleRL{Ki\-bbi omu\-ntu oku\-bba n’e\-bi\-ka\-ra\-kaa\-ta.}
\fmtExampleEN{It is bad for a person to have scaly skin.}
\fmtSubentry(b)
\fmtSense{foot fissure, foot crack, cracked heel}
\fmtExampleRL{Aba\-lu\-mi bai\-ngi ba\-ta n’e\-bi\-ka\-ra\-kaa\-ta.}
\fmtExampleEN{Many farmers have heel fissures.}

\fmtLemma{ki\-ka\-ra\-ngo}
\fmtCitation{(e)ki\-ka\-ra\-ngo}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{musky smell}
\fmtExampleRL{Eki\-ka\-ra\-ngo ki\-bba kya njo\-ka.}
\fmtExampleEN{A musky smell comes from snakes.}

\fmtLemma{ki\-ka\-ra\-nzi\-ra}
\fmtCitation{(e)ki\-ka\-ra\-nzi\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{roadside, side of a road}
\fmtExampleRL{Ai\-kai\-re oku ki\-ka\-ra\-nzi\-ra ki\-di.}
\fmtExampleEN{He is sitting on the other side of the road.}
\fmtSubentry(b)
\fmtSense{the other side of the road, the opposite side of the road}
\fmtExampleRL{Ya\-ba oku ki\-ka\-ra\-nzi\-ra ki\-di.}
\fmtExampleEN{Go to the other side of the road.}

\fmtLemma{ki\-ka\-ra\-pwa}
\fmtCitation{(e)ki\-ka\-ra\-pwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{corn, hard skin, callus, a thickened and hardened part of the skin or soft tissue, especially in an area that has been subjected to friction}

\fmtLemma{ki\-ka\-ra\-tu}
\fmtCitation{(e)ki\-ka\-ra\-tu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{dry itchy skin, scaly skin}
\fmtExampleRL{Omu\-lwai\-re ali\-na eki\-ka\-ra\-tu.}
\fmtExampleEN{The patient has scaly skin.}

\fmtLemma{ki\-kei\-sya}%
\fmtOne{1}
\fmtCitation{(e)ki\-kei\-sya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{Black kite clan}
\fmtExampleRL{Eki\-ka kya\-mwe ki\-kei\-sya.}
\fmtExampleEN{He is from the Black kite clan.}

\fmtLemma{ki\-kei\-sya}%
\fmtOne{2}
\fmtCitation{(e)ki\-kei\-sya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{all-night dance}
\fmtExampleRL{Ba\-suu\-re mu ki\-kei\-sya.}
\fmtExampleEN{They had an all-night dance.}

\fmtLemma{ki\-ke\-ke}
\fmtCitation{(e)ki\-ke\-ke}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{slice, piece, chunk, share, portion}
\fmtExampleRL{Mwa\-ji\-rye ya\-ka\-ke\-ka\-wo ebi\-ke\-ke bi\-sa\-tu byo\-nkai.}
\fmtExampleEN{I found him when he had made only three slices.}

\fmtLemma{ki\-ke\-ne\-ke\-ro}
\fmtCitation{(e)ki\-ke\-ne\-ke\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{filtering container}
\fmtExampleRL{Lee\-ta ani eki\-ke\-ne\-kee\-ro nse\-ngei\-ze muu\-nyu.}
\fmtExampleEN{Bring the filtering container here and I will filter the salt.}

\fmtLemma{ki\-ke\-re}
\fmtCitation{(e)ki\-ke\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{frog, toad}
\fmtExampleRL{Aba\-go\-nya bei\-zi\-ra ki\-ke\-re.}
\fmtExampleEN{The Bagonya belong to the frog totem.}

\fmtLemma{ki\-ke\-se\-ke\-se}
\fmtCitation{(e)ki\-ke\-se\-ke\-se}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{field from which millet has been harvested}
\fmtExampleRL{Embu\-li zi\-twa\-le ozi\-bbwe omu ki\-ke\-se\-ke\-se.}
\fmtExampleEN{Take the goats to the field where millet was harvested.}

\fmtLemma{ki\-ke\-so}
\fmtCitation{(e)ki\-ke\-so}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{pruning shears, a pair of shears used to harvest millet (a cutting instrument like larger scissors)}
\fmtExampleRL{Obwi\-ta ba\-bu\-ke\-se\-sya ki\-ke\-so.}
\fmtExampleEN{They use pruning shears to cut the millet plants.}

\fmtLemma{ki\-ke\-wu}
\fmtCitation{(e)ki\-ke\-wu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{straw container}
\fmtExampleRL{Nde\-te\-ra eki\-ke\-wu ntee\-mu olu\-se\-ke.}
\fmtExampleEN{Bring me the straw contained and I'll put in my straw.}

\fmtLemma{ki\-ki}
\fmtPOS{interrog}
\fmtSense{what, which}
\fmtExampleRL{Ki\-ki eki\-kwa\-ba omu mai\-so?}
\fmtExampleEN{What is going on?}

\fmtLemma{ki\-kii\-ni\-ki\-ra}
\fmtCitation{(e)ki\-kii\-ni\-ki\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bottle top, lid, cover}
\fmtExampleRL{Lee\-ta eki\-kii\-ni\-ki\-ra kya ki\-do\-mo\-la.}
\fmtExampleEN{Bring the lid of the jerry can.}

\fmtLemma{ki\-ki\-na\-la}
\fmtCitation{(oku)\-ki\-ki\-na\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ki\-ki\-nai\-re}
\fmtSubentry(a)
\fmtSense{(of body part) to stick out (\eg buttocks)}
\fmtExampleRL{Omwa\-la aku\-ta\-mbu\-la ng’a\-ka\-bi\-na ka\-ki\-ki\-nai\-re.}
\fmtExampleEN{The girl is walking with her buttocks sticking out.}
\fmtSubentry(b)
\fmtSense{to disobey, to disregard rules}
\fmtExampleRL{Omwa\-na aki\-ki\-nai\-re oku ma\-waa\-mwe.}
\fmtExampleEN{The child disobeys her mother.}

\fmtLemma{ki\-ko\-bba}
\fmtCitation{(e)ki\-ko\-bba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{skin}
\fmtExampleRL{Ya\-yii\-re n’a\-bba\-mbu\-ka eki\-ko\-bba.}
\fmtExampleEN{He got burnt and his skin peeled off.}
\fmtSubentry(b)
\fmtSense{safety belt}
\fmtExampleRL{Omu\-vu\-gi wa mo\-to\-ka ye\-bbwe\-re eki\-ko\-bba.}
\fmtExampleEN{The driver fastened his safety belt.}

\fmtLemma{ki\-ko\-boi}
\fmtCitation{(o)ki\-ko\-boi}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{chatterbox, chatterer}
\fmtExampleRL{Oki\-ko\-boi abbai\-re\-ku ani n’a\-ku\-ku\-buu\-lya.}
\fmtExampleEN{The chatterbox was here briefly asking for you.}

\fmtLemma{ki\-koi\-ge}
\fmtCitation{(e)ki\-koi\-ge}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{cleared ground after a sweet potato harvest}
\fmtExampleRL{Ebi\-nyee\-bwa na\-bi\-ru\-ma mu ki\-koi\-ge.}
\fmtExampleEN{I will plant groundnuts in the cleared garden where sweet potatoes were harvested.}

\fmtLemma{ki\-koi\-ro}
\fmtCitation{(e)ki\-koi\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{type of illegal fish trap used to trap small fish}

\fmtLemma{ki\-ko\-ke}%
\fmtOne{1}
\fmtCitation{(e)ki\-ko\-ke}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{grey}
\fmtExampleRL{Enyu\-mba yaa\-mwe ya\-gi\-sii\-gi\-re la\-ngi ya ki\-ko\-ke.}
\fmtExampleEN{He painted his house grey.}

\fmtLemma{ki\-ko\-ke}%
\fmtOne{2}
\fmtCitation{(o)ki\-ko\-ke}
\fmtPOS{n}%
\fmtNC{1}
\fmtSense{grey-feathered chicken}

\fmtLemma{ki\-ko\-ke\-ko\-ke}
\fmtCitation{(e)ki\-ko\-ke\-ko\-ke}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{greyish colour}
\fmtExampleRL{Onte waa\-mwe wa ki\-ko\-ke\-ko\-ke.}
\fmtExampleEN{His cow is greyish.}

\fmtLemma{ki\-ko\-ko\-ro}
\fmtCitation{(e)ki\-ko\-ko\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{hoof cap (\eg of a cow)}
\fmtExampleRL{Eki\-ko\-ko\-ro kya nte ki\-rwai\-re.}
\fmtExampleEN{The hoof cap is infected.}

\fmtLemma{ki\-ko\-ko\-to}
\fmtCitation{(e)ki\-ko\-ko\-to}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{force, power, strength, effort}
\fmtExampleRL{Eki\-ko\-ko\-to ki\-so\-bo\-la owa mpa\-ka.}
\fmtExampleEN{The use of force can manage a pig-headed person.}

\fmtLemma{ki\-ko\-kwei\-zi}
\fmtCitation{(e)ki\-ko\-kwei\-zi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from archaic}
\fmtSubentry(a)
\fmtSense{sisal plant}
\fmtExampleRL{Eki\-ko\-kwei\-zi ki\-kuu\-re ku\-sai.}
\fmtExampleEN{The sisal plant has grown well.}
\fmtSubentry(b)
\fmtSense{sisal bundle}
\fmtExampleRL{Aguu\-re eki\-ko\-kwei\-zi ki\-mwei.}
\fmtExampleEN{He bought one sisal bundle.}

\fmtLemma{ki\-ko\-kwo\-to\-ka}
\fmtCitation{(e)ki\-ko\-kwo\-to\-ka}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{reptile, collective name for snakes, lizards, and crocodile}
\fmtExampleRL{Omu nya\-nja mu\-bbaa\-mu ebi\-ko\-kwo\-to\-ka.}
\fmtExampleEN{There are reptiles in the lake.}

\fmtLemma{ki\-ko\-li\-mo}
\fmtCitation{(e)ki\-ko\-li\-mo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{curse, jinx, spell, evil eye}
\fmtExampleRL{Eki\-ko\-li\-mo kya Ada\-m ki\-kya\-ta\-mbuu\-ra oku mwa\-na wa mu\-ntu.}
\fmtExampleEN{Adam's curse still haunts human beings.}

\fmtLemma{ki\-ko\-lo}
\fmtCitation{(e)ki\-ko\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{stem, trunk, stalk of a tree}
\fmtExampleRL{Eki\-ko\-lo kya mu\-saa\-le gu\-ni ki\-koo\-to.}
\fmtExampleEN{The trunk of this tree is big.}

\fmtLemma{ki\-ko\-lo\-be\-ro}
\fmtCitation{(e)ki\-ko\-lo\-be\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{high degree crime, law breaking}
\fmtExampleRL{E Na\-lu\-fee\-nya baa\-ko\-lei\-re\-yo ebi\-ko\-lo\-be\-ro bi\-ngi.}
\fmtExampleEN{They committed a lot of crimes in Nalufeenya.}

\fmtLemma{ki\-ko\-lo\-ko\-ndo}
\fmtCitation{(e)ki\-ko\-lo\-ko\-ndo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bare banana stalk of the bunch, stem of a bunch of plantains}

\fmtLemma{ki\-ko\-lwa}
\fmtCitation{(e)ki\-ko\-lwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{act, performance, action, deed}
\fmtExampleRL{Ebi\-ko\-lwa bya\-mwe ebya ki\-sa bi\-kwi\-za kwi\-zu\-ki\-rwa\-nga.}
\fmtExampleEN{His acts of kindness will always be remembered.}
\fmtSubentry(b)
\fmtSense{example, model}
\fmtExampleRL{Ko\-le\-sya bi\-ko\-lwa.}
\fmtExampleEN{Live by examples.}

\fmtLemma{ki\-ko\-mo}
\fmtCitation{(e)ki\-ko\-mo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{copper}
\fmtExampleRL{Omu\-ka\-li azwai\-re ne\-ki\-lee\-si ya ki\-ko\-ma.}
\fmtExampleEN{The woman is wearing a necklace made of copper.}
\fmtSubentry(b)
\fmtSense{bangle, bracelet}
\fmtExampleRL{Aba\-ka\-li ba\-ta\-ka mu\-no oku\-zwa\-la ebi\-ko\-mo.}
\fmtExampleEN{Women love wearing bangles.}
\fmtSubentry(c)
\fmtSense{round object, circle, ring-shaped object}
\fmtExampleRL{Alii\-za ng’o\-kwe\-zi kwa ki\-ko\-mo.}
\fmtExampleEN{She will come when it is the full moon.}
\fmtSubentry(d)
\fmtSense{limit, boundary, range}
\fmtExampleRL{Nko\-be\-ra eki\-ko\-mo kya bbee\-yi nte oni.}
\fmtExampleEN{Tell me the price range of this cow.}
\fmtSubentry(e)
\fmtSense{22-degree halo (an optical phenomenon in the form of a ring with a radius of approximately 22° around the Moon)}

\fmtLemma{ki\-ko\-mpe}%
\fmtOne{1}
\fmtCitation{(e)ki\-ko\-mpe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from Sw.}
\fmtSubentry(a)
\fmtSense{cup, mug}
\fmtExampleRL{Ta omu\-ba\-zi mu ki\-ko\-mpe.}
\fmtExampleEN{Put the medicine in the cup.}
\fmtSubentry(b)
\fmtSense{trophy, award, price}
\fmtExampleRL{Ba\-wa\-ngwi\-re eki\-ko\-mpe.}
\fmtExampleEN{They won a trophy.}

\fmtLemma{ki\-ko\-mpe}%
\fmtOne{2}
\fmtCitation{(e)ki\-ko\-mpe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{chalice, ciborium}
\fmtExampleRL{Omu ka\-ni\-sa mu\-bbaa\-mu ebi\-ko\-mpe.}
\fmtExampleEN{There are chalices in the church.}

\fmtLemma{ki\-ko\-ndi}
\fmtCitation{(e)ki\-ko\-ndi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{guinea grass (\textit{Megathyrsus maximus})}
\fmtExampleRL{Nya\-jii\-rye omu\-sai\-za mu\-kai\-re nga aku\-bbwa ombu\-li oku ki\-ko\-ndi.}
\fmtExampleEN{I found an old man tying a goat to guinea grass.}
\fmtSubentry(b)
\fmtSense{flower of guinea grass (\textit{Megathyrsus maximus})}
\fmtExampleRL{Obu\-ko\-ndi bwe\-ya ku\-sai omu nyu\-mba.}
\fmtExampleEN{Guinea grass flowers sweep the house very well.}

\fmtLemma{ki\-ko\-ndo}
\fmtCitation{(e)ki\-ko\-ndo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{pole (\eg electricity pole)}
\fmtExampleRL{Ei\-zo ni\-mwo twa\-si\-mbi\-re ebi\-ko\-ndo bya ku\-bbwa ose\-nge\-nge.}
\fmtExampleEN{It was yesterday that we erected the poles to hold barbed wire.}
\fmtSubentry(b)
\fmtSense{milepost}
\fmtExampleRL{Ku nzi\-ra za Ka\-mpa\-la ku\-li\-ku ebi\-ko\-ndo.}
\fmtExampleEN{There are mileposts on Kampala roads.}
\fmtSubentry(c)
\fmtFixedexpression{ki\-kondo kya goolo}
\fmtSense{goalpost}
\fmtExampleRL{Eki\-saa\-we ki\-bbaa\-mu ebi\-ko\-ndo bya goo\-lo.}
\fmtExampleEN{The playing field has goalposts.}

\fmtLemma{ki\-ko\-ndoo\-ro}
\fmtCitation{(e)ki\-ko\-ndoo\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{inner flap of a gomesi}
\fmtExampleRL{Aba\-ka\-li ba\-bii\-ka ese\-nte omu ki\-ko\-ndoo\-ro.}
\fmtExampleEN{Women keep money in the inner flap of a gomesi.}

\fmtLemma{ki\-ko\-nge}
\fmtCitation{(e)ki\-ko\-nge}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{tree stump}
\fmtExampleRL{Wee\-ge\-nde\-re\-ze, awo wa\-li\-wo eki\-ko\-nge.}
\fmtExampleEN{Be careful, there is a tree stump there.}
\fmtSubentry(b)
\fmtSense{wealthy person, famous person, big shot, dignitary}
\fmtExampleRL{Abo ni\-bo bi\-ko\-nge bya Na\-ka\-so\-ngo\-la.}
\fmtExampleEN{Those are the wealthy people from Nakasongola.}

\fmtLemma{ki\-ko\-ngoo\-ro}
\fmtCitation{(e)ki\-ko\-ngoo\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{maize cob from which the kernels have been removed}
\fmtExampleRL{Eki\-ko\-ngoo\-ro ki\-kuu\-ta ebi\-ka\-ra\-kaa\-ta ku\-sai mu\-no.}
\fmtExampleEN{A maize cob cleans heel fissures quite well.}

\fmtLemma{ki\-ko\-ni}
\fmtCitation{(e)ki\-ko\-ni}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{group of houses constructed facing each other}
\fmtExampleRL{Omu\-ka\-ma wa ki\-no\-ni ni\-ye naa\-ni?}
\fmtExampleEN{Who is the owner of the houses constructed facing each other?}

\fmtLemma{ki\-ko\-nko\-li}
\fmtCitation{(e)ki\-ko\-nko\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{log, trunk}
\fmtExampleRL{Ki\-ni ki\-ko\-nko\-li kya taa\-ta kya ko\-le\-sya nk’a\-ka\-te\-be.}
\fmtExampleEN{This is the father's log, which he uses as a stool.}

\fmtLemma{ki\-ko\-nko\-ro}
\fmtCitation{(e)ki\-ko\-nko\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wooden stool}
\fmtExampleRL{Ikaa\-ra oku ki\-ko\-nko\-ro ekyo ndoo\-wo nte\-be gi\-ndi.}
\fmtExampleEN{Sit on that wooden stool, there are no more seats.}

\fmtLemma{ki\-ko\-nwa}
\fmtCitation{(e)ki\-ko\-nwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{half-baked item (\eg pot, bricks)}
\fmtExampleRL{Angu\-li\-si\-rye eki\-ko\-nwa bya bbu\-loo\-ka.}
\fmtExampleEN{He sold me half-baked bricks.}

\fmtLemma{ki\-koo\-de\-re}
\fmtCitation{(e)ki\-koo\-de\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{drying lane for simsim (UgE)/sesame}
\fmtExampleRL{Ama\-ca\-nde ba\-ga\-kaa\-rya ku ki\-koo\-de\-re.}
\fmtExampleEN{Simsim is dried on a simsim drying lane.}

\fmtLemma{ki\-koo\-do}
\fmtCitation{(e)ki\-koo\-do}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{measles}
\fmtExampleRL{Omwa\-na yee\-tii\-ki\-re eki\-koo\-do.}
\fmtExampleEN{The child is suffering from measles.}

\fmtLemma{ki\-koo\-ko}
\fmtCitation{(e)ki\-koo\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{outer part of the tree, bark}
\fmtExampleRL{Mu\-too\-le\-ku eki\-koo\-ko oku mu\-saa\-le tu\-sa\-le embaa\-wo.}
\fmtExampleEN{Remove the bark of the tree so that we get timber.}

\fmtLemma{ki\-koo\-koo\-ma}
\fmtCitation{(e)ki\-koo\-koo\-ma}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{\textit{Ver\-no\-nia auriculifera} (shrub with large soft leaves used as toilet paper}
\fmtExampleRL{Ebi\-koo\-la bya ki\-koo\-koo\-ma ni\-kyo tu\-ko\-le\-sya okwe\-su\-suu\-la omu kyo\-loo\-ni.}
\fmtExampleEN{We use \textit{Ver\-no\-nia auriculifera} as toilet paper.}

\fmtLemma{ki\-koo\-la}
\fmtCitation{(e)ki\-koo\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{leaf}
\fmtExampleRL{No\-ko\-la ebi\-koo\-la bya mu\-ye\-mbe tu\-ko\-le omu\-ba\-zi gwa ki\-koo\-ro.}
\fmtExampleEN{Get mango leaves and let's prepare cough medicine.}
\fmtSubentry(b)
\fmtSense{sweet potato vine for planting}
\fmtExampleRL{Aya\-bi\-re Ka\-wa\-nda ku\-gu\-la bi\-koo\-la ebi\-ro\-ngo\-see\-mu.}
\fmtExampleEN{He has gone to Kawanda (Agricultural Centre) to buy modified sweet potato vines.}

\fmtLemma{ki\-koo\-lo}
\fmtCrossreference{\textnormal{see} kikooro}
\fmtDialect{Lu\-nyala}

\fmtLemma{ki\-koo\-mi}
\fmtCitation{(e)ki\-koo\-mi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bonfire, smoky slow-burning fire, burning of trash or grass}
\fmtExampleRL{Eki\-ku\-wu\-nya bi\-koo\-la ebyo\-kei\-bwe oku\-zwa omu ki\-koo\-mi.}
\fmtExampleEN{That is a smell of burning leaves from a garden bonfire.}

\fmtLemma{ki\-koo\-ro}
\fmtDialect{Ruruuli}
\fmtCitation{(e)ki\-koo\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{cough}
\fmtExampleRL{Eki\-koo\-ro ki\-kwa\-ba ku\-nyi\-ta.}
\fmtExampleEN{The cough is going to kill me.}

\fmtLemma{ki\-koo\-ryo}
\fmtCitation{(e)ki\-koo\-ryo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{common hereditary features, birth marks}
\fmtExampleRL{Abaa\-na ba Na\-jje\-mba ba\-li\-na eki\-koo\-ryo.}
\fmtExampleEN{Najjemba's children have clear common hereditary features.}
\fmtSubentry(b)
\fmtSense{resemblance, correspondence, similarity}
\fmtExampleRL{Ki\-ta\-na\-mu eki\-koo\-ryo n’o\-ku bi\-ntu bye nku\-lo\-woo\-za.}
\fmtExampleEN{It bears some resemblance to the things I have in mind.}
\fmtSubentry(c)
\fmtSense{person vulnerable to infections}
\fmtExampleRL{Oyo omwa\-na ki\-koo\-ryo.}
\fmtExampleEN{That child is vulnerable to infections.}

\fmtLemma{ki\-koo\-yi}
\fmtCitation{(e)ki\-koo\-yi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{loin cloth for women, long piece of cloth worn by Baganda women under a gomesi}
\fmtExampleRL{Ego\-me\-si enyu\-mi\-ra\-mu eki\-koo\-yi.}
\fmtExampleEN{The gomesi looks nice when worn with a loin cloth.}

\fmtLemma{ki\-ko\-po}
\fmtCitation{(e)ki\-ko\-po}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{cup, mug}
\fmtExampleRL{Ya\-su\-ki\-re amai\-zi omu ki\-ko\-po n'a\-wee\-rya omwa\-na waa\-mwe.}
\fmtExampleEN{She poured water into the cup and gave it to her child.}
\fmtSubentry(b)
\fmtSense{trophy, award, price}
\fmtExampleRL{Etii\-mu yai\-swe ni\-yo ya\-twai\-re eki\-ko\-po.}
\fmtExampleEN{It was our team that took the trophy.}

\fmtLemma{ki\-ko\-si}
\fmtCitation{(e)ki\-ko\-si}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{sisal fiber}
\fmtSubentry(b)
\fmtSense{sisal string, rope}
\fmtExampleRL{Mpee\-rya eki\-ko\-si mbwe ensa\-wu z’a\-ma\-nda.}
\fmtExampleEN{Give me a rope so that I can tie the sacks of charcoal.}

\fmtLemma{ki\-ko\-wu}
\fmtCitation{(e)ki\-ko\-wu}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{guineafowl colour}
\fmtExampleRL{Azwai\-re saa\-ti ya ki\-ko\-wu.}
\fmtExampleEN{He is wearing a shirt in a guineafowl colour.}

\fmtLemma{ki\-ku}%
\fmtOne{1}
\fmtCrossreference{\textnormal{see also} kikwi}
\fmtCitation{(e)ki\-ku}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bed bug (\textit{Cimex lectularius})}
\fmtExampleRL{Ebi\-ku bya\-ka\-li\-rii\-rye omu ki\-ta\-nda.}
\fmtExampleEN{There are too many bed bugs in the bed.}

\fmtLemma{ki\-ku}%
\fmtOne{2}
\fmtCitation{(e)ki\-ku}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{heavy psychological burden one cannot disclose easily}
\fmtExampleRL{Mwi\-ra wa\-nge ndi\-na eki\-ku kye nta\-ku\-so\-bo\-la ku\-nyu\-mya.}
\fmtExampleEN{My friend, I have a heavy burden which I cannot disclose.}

\fmtLemma{Ki\-ku\-be}
\fmtCitation{(o)Ki\-ku\-be}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{child born outside of wedlock}
\fmtExampleRL{Ki\-ku\-be mwa\-na abya\-li\-bwa enzei wa bu\-fu\-mbo.}
\fmtExampleEN{Kikube is a name given to a child born out of wedlock.}

\fmtLemma{ki\-ku\-bi\-ro}
\fmtCitation{(e)ki\-ku\-bi\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{overgarment, apron}
\fmtExampleRL{Aba\-ka\-li aba irai ba\-zwa\-la\-nga ebi\-ku\-bi\-ro.}
\fmtExampleEN{Long ago women used to wear a wrapper cloth on top of a gomesi.}

\fmtLemma{ki\-ku\-ju}
\fmtCitation{(e)ki\-ku\-ju}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sauce made from young pumpkin}
\fmtExampleRL{Ebi\-ku\-ju me\-re eju\-na abaa\-na oku\-ku\-la ku\-sai.}
\fmtExampleEN{The sauce made from young pumpkin helps children grow healthy.}

\fmtLemma{ki\-ku\-lu\-bba}
\fmtCitation{(e)ki\-ku\-lu\-bba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{pumice stone (used to remove dry and excess skin from the bottom of the foot as well as calluses)}
\fmtExampleRL{Naa\-ni atwai\-re eki\-ku\-lu\-bba kya\-mu?}
\fmtExampleEN{Who has taken my pumice stone?}

\fmtLemma{ki\-ku\-lu\-ku\-to}
\fmtCitation{(e)ki\-ku\-lu\-ku\-to}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{haemorrhage}
\fmtExampleRL{Omu\-ka\-li alwai\-re eki\-ku\-lu\-ku\-to.}
\fmtExampleEN{The woman is suffering from haemorrhaging.}

\fmtLemma{ki\-ku\-lu\-kya\-mu\-ti}
\fmtCitation{(e)ki\-ku\-lu\-kya\-mu\-ti}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{type of herb that cures stomach ulcer}
\fmtExampleRL{Omu\-ba\-zi ogu\-zwa omu bi\-koo\-la bya ki\-ku\-lu\-kya\-mu\-ti gu\-ta\-mba alu\-sa.}
\fmtExampleEN{The medicine from the leaves of kikulukyamuti treats ulcers.}

\fmtLemma{ki\-ku\-mi}
\fmtPOS{num}
\fmtSense{one hundred}
\fmtExampleRL{Eki\-ku\-mi ampi te\-ki\-kya\-ta na mu\-ga\-so omu Uga\-nda.}
\fmtExampleEN{A hundred shillings is almost useless in Uganda.}

\fmtLemma{ki\-ku\-nduu\-ru}
\fmtCitation{(e)ki\-ku\-nduu\-ru}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fetish, talisman, charm}
\fmtExampleRL{Mu\-na\-ge\-yo ebi\-ku\-nduu\-ru omu nya\-nja.}
\fmtExampleEN{Throw the fetishes into the lake.}

\fmtLemma{ki\-ku\-nku\-du}
\fmtCitation{(o)ki\-ku\-nku\-du}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{animal or person without a limb or tail}
\fmtExampleRL{Ngo\-ngee\-ra omaa\-ma nta go\-ngei\-re ki\-ku\-nku\-du mu\-ki\-ra.}
\fmtExampleEN{I am sad that my mother is dead but you also do not have a tail. (from a folktale)}

\fmtLemma{ki\-ku\-nu\-bu\-lyo}
\fmtCitation{(e)ki\-ku\-nu\-bu\-lyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bruise}
\fmtExampleRL{Egaa\-li ya\-mu\-ku\-bbi\-re n’a\-fu\-na ebi\-ku\-nu\-bu\-lyo.}
\fmtExampleEN{He had a bicycle accident and got some bruises.}

\fmtLemma{ki\-ku\-nuu\-re}
\fmtCitation{(e)ki\-ku\-nuu\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bruise}
\fmtExampleRL{Omu\-ta\-mii\-re ayi\-zwi\-re bi\-ku\-nuu\-re.}
\fmtExampleEN{The drunkard is covered in bruises.}

\fmtLemma{Ki\-ku\-rwe}
\fmtCitation{(o)Ki\-ku\-rwe}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{member of the Otter clan}

\fmtLemma{ki\-ku\-tu}
\fmtCitation{(e)ki\-ku\-tu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{trap for hunting small animals (e.g.\,cane rats) (\eg cane rats)}
\fmtExampleRL{Eki\-ku\-tu ki\-kwai\-te omu\-swi.}
\fmtExampleEN{The trap has caught a cane rat.}

\fmtLemma{ki\-kuu\-bo}
\fmtCitation{(e)ki\-kuu\-bo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{stripe, line}
\fmtExampleRL{Esaa\-ti ya\-nge eri\-mu ebi\-kuu\-bo.}
\fmtExampleEN{My shirt is striped. (lit.\,My shirt has stripes.)}
\fmtSubentry(b)
\fmtSense{corridor, aisle, passage}
\fmtExampleRL{Abai\-bbi ba\-nya\-gi\-ra mu ki\-kuu\-bo kya ma\-duu\-ka.}
\fmtExampleEN{Thieves usually grab people's things in the shops' corridors.}

\fmtLemma{ki\-kuu\-te}
\fmtCitation{(e)ki\-kuu\-te}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{cattle trail}
\fmtExampleRL{Ente zi\-ta\-mbuu\-ra mu ki\-kuu\-te.}
\fmtExampleEN{Cows walk via a cattle trail.}

\fmtLemma{ki\-kwa\-nga\-la}
\fmtCitation{(e)ki\-kwa\-nga\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fake, forgery, sham}
\fmtExampleRL{Ba\-mu\-gu\-li\-si\-rye se\-mi\-nti wa ki\-kwa\-nga\-la.}
\fmtExampleEN{They sold him fake cement.}

\fmtLemma{ki\-kwa\-so}
\fmtCitation{(e)ki\-kwa\-so}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{safety pin}
\fmtExampleRL{Eki\-kwa\-so kyo\-ko\-la nju\-nzai.}
\fmtExampleEN{Safety pins are used to remove jiggers from the body.}

\fmtLemma{ki\-kwa\-ta\-ntaa\-ma}
\fmtCitation{(e)ki\-kwa\-ta\-ntaa\-ma}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{love grass (\textit{Setaria verticillata})}

\fmtLemma{ki\-kwa\-te}
\fmtCitation{(e)ki\-kwa\-te}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{poem, lyrics, verse}
\fmtExampleRL{Abaa\-na ba\-ba\-zi\-re eki\-kwa\-te kya\-bwe.}
\fmtExampleEN{The children recited their poem.}

\fmtLemma{ki\-kwa\-to}
\fmtCitation{(e)ki\-kwa\-to}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{handle}
\fmtExampleRL{Eki\-ko\-po ki\-bbaa\-ku eki\-kwa\-to.}
\fmtExampleEN{A cup has a handle.}
\fmtSubentry(b)
\fmtSense{spout of a kettle or teapot}

\fmtLemma{ki\-kwe}%
\fmtOne{1}
\fmtCitation{(e)ki\-kwe}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{Parkinson’s disease}
\fmtExampleRL{Obu\-lwai\-re bwa ki\-kwe bu\-tii\-ni\-bwa mu\-no omu Ba\-ruu\-li.}
\fmtExampleEN{Parkinson’s disease is feared a lot among the Baruuli.}

\fmtLemma{ki\-kwe}%
\fmtOne{2}
\fmtCitation{(e)ki\-kwe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{taboo word}
\fmtExampleRL{Aba\-ruu\-li ba\-tii\-na oku\-si\-moo\-la ebi\-ga\-mbo ebya ki\-kwe.}
\fmtExampleEN{The Baruuli are afraid of saying taboo words.}

\fmtLemma{ki\-kwe\-kwe\-to}
\fmtCitation{(e)ki\-kwe\-kwe\-to}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{raid, foray, operation, manoeuvre, scouting expedition, reconnaissance operation}
\fmtExampleRL{Aba po\-lii\-si baa\-koo\-re eki\-kwe\-kwe\-to.}
\fmtExampleEN{The policemen staged an operation.}

\fmtLemma{ki\-kwe\-mpe}
\fmtCitation{(e)ki\-kwe\-mpe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{old bark cloth}
\fmtExampleRL{Omu\-sai\-za yee\-bbwe\-re eki\-kwe\-mpe.}
\fmtExampleEN{The man is wearing an old bark cloth.}
\fmtSubentry(b)
\fmtSense{embroidery, binding on the edge of cloth, decoration on the edge of \eg a gomesi}

\fmtLemma{ki\-kwi}
\fmtCrossreference{\textnormal{see also} kiku}
\fmtCitation{(e)ki\-kwi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bed bug (\textit{Cimex lectularius})}
\fmtExampleRL{Ebi\-kwi bi\-ga\-dya aba\-ntu ba Ka\-mpa\-la.}
\fmtExampleEN{Bed bugs disturb people in Kampala.}

\fmtLemma{ki\-kwi\-za}
\fmtCitation{(e)ki\-kwi\-za}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bellows used in metal forging}
\fmtExampleRL{Ko\-le\-sya eki\-kwi\-za okwe\-mba omuu\-ro two\-ce ebyo\-ma.}
\fmtExampleEN{Use the bellows to blow the fire so that we can melt metals.}

\fmtLemma{ki\-kwoi\-ko\-lyo}
\fmtCitation{(e)ki\-kwoi\-ko\-lyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{riddle, puzzle}
\fmtExampleRL{Mpee\-rya\-yo eki\-kwoi\-ko\-lyo ki\-mwe mu Lu\-nya\-la.}
\fmtExampleEN{Give me a riddle in Lu\-nyala.}

\fmtLemma{ki\-kwoi\-kwe}
\fmtCitation{(e)ki\-kwoi\-kwe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{riddle, puzzle}
\fmtExampleRL{Aba\-nya\-la ba\-li\-na ebi\-kwoi\-kwe bi\-ngi.}
\fmtExampleEN{The Banyala have many short riddles.}

\fmtLemma{ki\-kwoi\-kwoi}
\fmtCrossreference{\textnormal{see} kikwoikwe}

\fmtLemma{ki\-kye\-taa\-wo}%
\fmtOne{1}
\fmtCitation{(e)ki\-kye\-taa\-wo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{unexpected event, surprise}
\fmtExampleRL{Eki\-kye\-taa\-wo ki\-sii\-sa entee\-ka\-tee\-ka z’a\-ba\-ntu.}
\fmtExampleEN{An unexpected event disrupts peoples' plans.}

\fmtLemma{ki\-kye\-taa\-wo}%
\fmtOne{2}
\fmtCitation{(e)ki\-kye\-taa\-wo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{non-government organisation, NGO}
\fmtExampleRL{Ba\-koo\-re ebi\-kye\-taa\-wo bi\-ngi bai\-tu ti\-bi\-ko\-la.}
\fmtExampleEN{They formed many non-governmental organisations, but they are not functioning.}

\fmtLemma{ki\-maa\-mo}
\fmtCitation{(e)ki\-maa\-mo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{handle (\eg of a knife, panga)}
\fmtExampleRL{Eki\-maa\-mo kya nso\-ne ki\-zwi\-re\-ku.}
\fmtExampleEN{The handle of the knife has broken off.}

\fmtLemma{ki\-maa\-mu}
\fmtCrossreference{\textnormal{see} kimaamo}

\fmtLemma{ki\-ma\-je\-ge\-re}
\fmtCitation{(e)ki\-ma\-je\-ge\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{grader}
\fmtExampleRL{Ebi\-ma\-je\-ge\-re bi\-ko\-le\-se\-bwa oku\-ko\-la enzi\-ra.}
\fmtExampleEN{Graders are used to construct roads.}

\fmtLemma{ki\-ma\-ma}
\fmtCitation{(e)ki\-ma\-ma}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{intuitive behavior, incautiousness}
\fmtExampleRL{Omu\-vu\-gi ba\-mu\-si\-ngi\-si\-rye ogwa ku\-vu\-gi\-sya eki\-ma\-ma.}
\fmtExampleEN{The driver was convicted of reckless driving.}

\fmtLemma{ki\-ma\-nsuu\-li}
\fmtCitation{(e)ki\-ma\-nsuu\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bow-like trap}
\fmtExampleRL{Eki\-ma\-nsuu\-li kii\-ta n’e\-bi\-nyo\-nyi.}
\fmtExampleEN{The bow-like trap also kills birds.}

\fmtLemma{ki\-ma\-nte}
\fmtCitation{(e)ki\-ma\-nte}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wall}
\fmtExampleRL{Eki\-ma\-nte kya\-mwe kii\-zi\-re\-mu olu\-ka\-la\-la.}
\fmtExampleEN{His wall has developed a crack.}

\fmtLemma{ki\-ma\-nyi\-syo}
\fmtCitation{(e)ki\-ma\-nyi\-syo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{announcement}
\fmtExampleRL{Ya\-ka\-lya eki\-ma\-nyi\-syo oku la\-di\-yo.}
\fmtExampleEN{Place an announcement on the radio.}
\fmtSubentry(b)
\fmtSense{notice board, billboard}
\fmtExampleRL{Oku iso\-me\-ro lyai\-swe tu\-ta n’e\-ki\-ma\-nyi\-syo.}
\fmtExampleEN{We have a notice board at our school.}
\fmtSubentry(c)
\fmtSense{poster, banner}
\fmtExampleRL{Mba\-di ba\-ku\-ti\-mba ebi\-ma\-nyi\-syo.}
\fmtExampleEN{They are there pinning up posters.}
\fmtSubentry(d)
\fmtSense{signpost}
\fmtExampleRL{Ako ka\-ma\-nyi\-syo ka iso\-me\-ro.}
\fmtExampleEN{That is a school signpost.}

\fmtLemma{Ki\-ma\-nywe\-nda}
\fmtCitation{(o)Ki\-ma\-nywe\-nda}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{name of a boy whose father is a relative to a mother}

\fmtLemma{ki\-mbaa\-ra}%
\fmtOne{1}
\fmtCitation{(e)ki\-mbaa\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{tsetse fly}
\fmtExampleRL{Onte ane\-nei\-bwe eki\-mbaa\-ra.}
\fmtExampleEN{The cow was bitten by a tsetse fly.}

\fmtLemma{ki\-mbaa\-ra}%
\fmtOne{2}
\fmtCitation{(o)ki\-mbaa\-ra}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{blind person}
\fmtExampleRL{Oki\-mbaa\-ra oyo abba\-ndi\-re enko\-nge.}
\fmtExampleEN{The blind person hit a tree stump.}

\fmtLemma{ki\-mbi}
\fmtCitation{(e)ki\-mbi}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{garbage can, dustbin}
\fmtExampleRL{Ei\-ko\-ke li\-su\-ke oku ki\-mbi.}
\fmtExampleEN{Throw the ash into the garbage can.}

\fmtLemma{ki\-mbu\-gu\-mbu\-gu}
\fmtCrossreference{\textnormal{see also} kibugu}
\fmtCitation{(e)ki\-mbu\-gu\-mbu\-gu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{vampire bat (probably yellow-winged bat)}
\fmtExampleRL{Eki\-mbu\-gu\-mbu\-gu ki\-nywa omu\-saa\-yi.}
\fmtExampleEN{The vampire bat sucks blood.}

\fmtLemma{ki\-me}%
\fmtOne{1}
\fmtCitation{(e)ki\-me}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{private parts, loin area, genitals}
\fmtExampleRL{Obwi\-ri\-za bu\-me\-ra oku ki\-me.}
\fmtExampleEN{Pubic hair develops around the private parts.}

\fmtLemma{ki\-me}%
\fmtOne{2}
\fmtCitation{(e)ki\-me}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{dew}
\fmtExampleRL{Eki\-me ki\-gwi\-re.}
\fmtExampleEN{The dew has formed.}

\fmtLemma{ki\-me\-go}
\fmtCitation{(e)ki\-me\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{helping of food, serving of food}
\fmtExampleRL{Ntwai\-re eki\-me\-go ki\-mwei.}
\fmtExampleEN{I took one helping.}

\fmtLemma{ki\-me\-ki\-me}
\fmtCitation{(e)ki\-m(e)ki\-me}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{fading dew}
\fmtExampleRL{Obwi\-re bwa ki\-me\-ki\-me.}
\fmtExampleEN{The dew is beginning to clear.}

\fmtLemma{ki\-me\-me\-ya}%
\fmtOne{1}
\fmtCitation{(o)ki\-me\-me\-ya}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSubentry(a)
\fmtSense{naked neck chicken}
\fmtExampleRL{Nguu\-re ki\-me\-me\-ya.}
\fmtExampleEN{I have bought a naked neck chicken.}
\fmtSubentry(b)
\fmtSense{undercut hair style (with hair on top while the back and sides are closely cut)}
\fmtExampleRL{Aba\-si\-ga\-zi ba\-ta\-ka mu\-no oku\-sa\-la ebi\-me\-me\-ya.}
\fmtExampleEN{Young people love undercuts.}

\fmtLemma{ki\-me\-ra}
\fmtCitation{(e)ki\-me\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{plant, crops}
\fmtExampleRL{Nsi\-mbi\-re ebi\-ka bya bi\-me\-ra ebya nja\-wu\-lo.}
\fmtExampleEN{I have planted different varieties of crops.}

\fmtLemma{ki\-me\-ree\-ri}
\fmtCitation{(e)ki\-me\-ree\-ri}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{post-harvest sprouts, volunteer crops}
\fmtExampleRL{Bi\-ni bi\-me\-ree\-ri bya duu\-ma.}
\fmtExampleEN{These are post-harvest sprouts of maize.}

\fmtLemma{ki\-me\-ree\-rwa}
\fmtCitation{(e)ki\-me\-ree\-rwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{post-harvest sprouts, volunteer plant or crops}

\fmtLemma{ki\-mo\-go}
\fmtCitation{(e)ki\-mo\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{piece, slice, portion}
\fmtExampleRL{Mpee\-rya\-ku eki\-mo\-go kya ki\-duu\-ma.}
\fmtExampleEN{Give me a piece of maize.}

\fmtLemma{ki\-mo\-le}
\fmtCitation{(e)ki\-mo\-le}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{torch made of reeds tied together}
\fmtExampleRL{Ebi\-mo\-le baa\-bi\-ko\-le\-sya\-nga nk’e\-taa\-la obwi\-re.}
\fmtExampleEN{Torches made of reeds were used to provide light at night.}

\fmtLemma{ki\-mo\-mo}%
\fmtOne{1}
\fmtCitation{(e)ki\-mo\-mo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{type of white ant}
\fmtExampleRL{Abaa\-na ba\-ta\-ka oku\-lya ebi\-mo\-mo.}
\fmtExampleEN{Children like eating ants.}

\fmtLemma{ki\-mo\-mo}%
\fmtOne{2}
\fmtPOS{adv}
\fmtSense{(of speaking) inaudibly, unclear}
\fmtExampleRL{Oyo omu\-ka\-li aku\-ba\-za ki\-mo\-mo.}
\fmtExampleEN{That woman speaks inaudibly.}

\fmtLemma{ki\-mo\-te}
\fmtCitation{(o)ki\-mo\-te}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{refined bark cloth (dark brown in color and very soft)}
\fmtExampleRL{Oki\-mo\-te agu\-lwa ba\-guu\-da.}
\fmtExampleEN{Refined bark cloth is expensive.}

\fmtLemma{ki\-mpa\-ngwe}
\fmtCitation{(o)ki\-mpa\-ngwe}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{storm, strong wind, tempest (over a body of water)}
\fmtExampleRL{Oki\-mpa\-ngwe aku\-bbi\-re ebi\-go\-go.}
\fmtExampleEN{The strong wind has destroyed the banana plantation.}

\fmtLemma{ki\-mpa\-nu\-ka}
\fmtPOS{adv}
\fmtSense{hurriedly, hastily, quickly}
\fmtExampleRL{Ba\-mu\-twai\-re ki\-mpa\-nu\-ka.}
\fmtExampleEN{They took him hurriedly.}
\fmtSense{uninvited visitor, uninvited guest}
\fmtExampleRL{Oki\-mpa\-nu\-ka alya bu\-sye\-ree\-re.}
\fmtExampleEN{An uninvited visitor eats whatever he finds.}

\fmtLemma{ki\-mpi\-ri\-gi\-nyi}
\fmtCitation{(o)ki\-mpi\-ri\-gi\-nyi}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{very short person, dwarf}
\fmtExampleRL{Oki\-mpi\-ri\-gi\-nyi ngo\-yo akwi\-za oku\-syo\-ma ese\-nte zaa\-mwe.}
\fmtExampleEN{The dwarf is coming to pick up his money.}

\fmtLemma{ki\-mu\-li}
\fmtCitation{(e)ki\-mu\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{flower}
\fmtExampleRL{Ki\-ni ki\-mu\-li kya mu\-cu\-ngwai.}
\fmtExampleEN{This is an orange flower.}

\fmtLemma{ki\-mu\-nku}
\fmtCitation{(e)ki\-mu\-nku}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{piece of hard dry soil}
\fmtExampleRL{Anka\-su\-kii\-re eki\-mu\-nku kya ita\-ka\-li.}
\fmtExampleEN{She threw a piece of hard dry soil at me.}

\fmtLemma{ki\-mwe}
\fmtCrossreference{\textnormal{see} kimwei and (-)mwei}

\fmtLemma{ki\-mwei}
\fmtPOS{adv}
\fmtSubentry(a)
\fmtSense{completely, totally, entirely}
\fmtExampleRL{Wee\-ye\-be\-rei\-re eka\-laa\-mu yaa\-mu ki\-mwei.}
\fmtExampleEN{You completely forgot your pen.}
\fmtSubentry(b)
\fmtSense{forever, permanently, for good}
\fmtExampleRL{Ki\-twa\-re ki\-mwei.}
\fmtExampleEN{Take it forever.}

\fmtLemma{ki\-mwei\-ki\-mwei}
\fmtCrossreference{\textnormal{see} (-)mwei (-)mwei}

\fmtLemma{ki\-mwe\-mwe\-nge}
\fmtCitation{(e)ki\-mwe\-mwe\-nge}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{smile}
\fmtExampleRL{N’a\-mboi\-ne n’ai\-za\-ku ki\-mwe\-mwe\-nge.}
\fmtExampleEN{When she saw me, she smiled.}

\fmtLemma{ki\-naa\-bi\-ro}
\fmtCitation{(e)ki\-naa\-bi\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bathroom, washroom}
\fmtExampleRL{Aku\-naa\-bi\-ra mu ki\-naa\-bi\-ro.}
\fmtExampleEN{She is having a bath in the bathroom.}

\fmtLemma{ki\-naa\-nai}
\fmtCrossreference{\textnormal{see} kinanai}

\fmtLemma{ki\-naa\-nyu\-ma\-nyu\-ma}
\fmtCrossreference{\textnormal{see also} kyanyumaanyuma}
\fmtPOS{adv}
\fmtSense{in reverse, backwards}
\fmtExampleRL{Aku\-ta\-mbu\-la ki\-naa\-nyu\-maa\-nyu\-ma.}
\fmtExampleEN{He walked backwards.}

\fmtLemma{ki\-na\-bu\-ke\-ze}
\fmtPOS{adv}
\fmtSense{thoroughly, to exhaustion}
\fmtExampleRL{Na\-koo\-re ki\-na\-bu\-ke\-ze.}
\fmtExampleEN{I worked to exhaustion.}

\fmtLemma{ki\-na\-ga}%
\fmtOne{1}
\fmtCitation{(e)ki\-na\-ga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{pot, soucepan}
\fmtExampleRL{Ya\-bu\-mbi\-re eki\-na\-ga eki\-koo\-to ei\-go\-lo.}
\fmtExampleEN{He moulded a big pot yesterday.}

\fmtLemma{ki\-na\-ge}%
\fmtOne{2}
\fmtCitation{(e)ki\-na\-ge}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{tilapia (also used generically to refer to any fish)}
\fmtExampleRL{Ebi\-na\-ge ni\-byo bi\-si\-nga oku\-lii\-bwa omu Uga\-nda.}
\fmtExampleEN{Tilapia is the most commonly eaten fish in Uganda.}

\fmtLemma{ki\-nai\-bee\-re}
\fmtCitation{(o)ki\-nai\-bee\-re}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSubentry(a)
\fmtSense{relative from one's mother's side}
\fmtSubentry(b)
\fmtSense{one of siblings who have different fathers}

\fmtLemma{ki\-nai\-gu\-du}
\fmtCitation{(e)ki\-nai\-gu\-du}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{merry making, fun, party}
\fmtExampleRL{Tu\-kwa\-ba mu ki\-nai\-gu\-du.}
\fmtExampleEN{We are going to have fun.}

\fmtLemma{ki\-na\-li\-ga}
\fmtCitation{(e)ki\-na\-li\-ga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{trial-and-error manner of hunting covering many villages}
\fmtExampleRL{Omu\-yi\-go ba\-koo\-re gwa ki\-na\-li\-ga.}
\fmtExampleEN{They are hunting in a trial-and-error manner covering many villages.}

\fmtLemma{ki\-na\-ma\-di}
\fmtCitation{(e)ki\-na\-ma\-di}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bag for meat}
\fmtExampleRL{Mpee\-rya eki\-na\-ma\-di kya\-mu nya\-be ngu\-le enya\-ma.}
\fmtExampleEN{Give me your meat bag and I'll go and buy meat.}

\fmtLemma{ki\-na\-mbu\-la\-mu}
\fmtCitation{(e)ki\-na\-mbu\-la\-mu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{useless item, object without a purpose}

\fmtLemma{ki\-na\-mbu\-la\-mu\-ga\-so}
\fmtCitation{(e)ki\-na\-mbu\-la\-mu\-ga\-so}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{useless item, object without a purpose}
\fmtExampleRL{Ekyo\-ma ekyo ki\-na\-mbu\-la\-mu\-ga\-so ki\-na\-ge\-yo.}
\fmtExampleEN{That metal is useless, throw it away.}

\fmtLemma{ki\-na\-mo}
\fmtCitation{(e)ki\-na\-mo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{valley}
\fmtExampleRL{Ei\-rwa\-ro ba\-lyo\-mbo\-ke\-re mu ki\-na\-mo.}
\fmtExampleEN{They built the hospital in a valley.}

\fmtLemma{ki\-na\-nai}
\fmtPOS{num}
\fmtSense{eighty}
\fmtExampleRL{Ompyo abba n’a\-mai\-no ki\-na\-nai.}
\fmtExampleEN{A crocodile has eighty teeth.}

\fmtLemma{ki\-na\-na\-nsi}
\fmtCitation{(e)ki\-na\-na\-nsi}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{genital warts}
\fmtExampleRL{Omu\-ka\-li oyo mu\-lwai\-re wa ki\-na\-na\-nsi.}
\fmtExampleEN{That woman suffers from genital warts.}

\fmtLemma{ki\-ndaa\-zi}
\fmtCitation{(e)ki\-ndaa\-zi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{doughnut}
\fmtExampleRL{Abaa\-na ba\-ta\-ka mu\-no oku\-lya ebi\-ndaa\-zi.}
\fmtExampleEN{Children like eating donuts very much.}

\fmtLemma{ki\-ndyo\-la}
\fmtCitation{(oku)\-ki\-ndyo\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ki\-ndyoi\-re}
\fmtSense{to roll one's eyes in an insulting manner}
\fmtExampleRL{Oboi\-ne na\-ku\-ki\-ndyo\-la?}
\fmtExampleEN{Have you seen how she is rolling her eyes insultingly?}

\fmtLemma{ki\-nga}%
\fmtOne{1}
\fmtCitation{(oku)\-ki\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ki\-ngi\-re}
\fmtSubentry(a)
\fmtSense{to close, to shut}
\fmtExampleRL{Ki\-nga olwi\-ge lwa\-mu.}
\fmtExampleEN{Close your door.}
\fmtSubentry(b)
\fmtSense{to veil, to cover}
\fmtExampleRL{Omu\-ka\-li ye\-ki\-ngi\-rii\-rye olu\-go\-ye oku mu\-twe.}
\fmtExampleEN{The woman covered her head.}
\fmtSubentry(c)
\fmtSense{to prevent, to avoid}
\fmtExampleRL{Naa\-ba enga\-lo wee\-ki\-nge endwai\-re.}
\fmtExampleEN{Wash your hands in order to prevent diseases.}
\fmtSubentry(d)
\fmtSense{to repulse spells}
\fmtExampleRL{Oi\-se\-ma\-ka waa\-wo ya\-ki\-ngi\-re ama\-ka gaa\-mwe.}
\fmtExampleEN{The head of the family repulsed his home against spells.}
\fmtSubentry(e)
\fmtSense{to treat (a corpse, \eg by embalming)}
\fmtExampleRL{Oru\-fa mu\-mu\-ki\-nge ata\-wu\-nya.}
\fmtExampleEN{Treat the corpse so that it does not smell.}

\fmtLemma{ki\-nga}%
\fmtOne{2}
\fmtCitation{(e)ki\-nga}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{immunisation, prevention, vaccination}
\fmtExampleRL{Enki\-nga y’o\-bu\-lwai\-re bwa lu\-se\-ru eta\-ndi\-ki\-re.}
\fmtExampleEN{The immunisation against measles has started.}

\fmtLemma{ki\-ngii\-rya}
\fmtCitation{(oku)\-ki\-ngii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ki\-ngi\-rii\-rye}
\fmtSense{to protect, to defend, to cover}
\fmtExampleRL{Yai\-mwi\-re enki\-nga ai\-gu\-ru oku\-ki\-ngii\-rya omu\-twe gwa\-mwe.}
\fmtExampleEN{He held up his shield to protect his head from the attack.}

\fmtLemma{ki\-ngu\-la}
\fmtCitation{(oku)\-ki\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ki\-ngwi\-re}
\fmtSense{to open, to uncover}
\fmtExampleRL{Ki\-ngu\-la olwi\-ge.}
\fmtExampleEN{Open the door.}

\fmtLemma{ki\-ni}
\fmtCrossreference{\textnormal{see} ni{\textnormal{\fmtOne{2}}}}

\fmtLemma{ki\-ni\-ga}
\fmtCitation{(e)ki\-ni\-ga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{anger, fury, jealousy, jealous rage}
\fmtExampleRL{Eki\-ni\-ga eki\-ngi ki\-see\-nya ama\-ka.}
\fmtExampleEN{A lot of anger destroys families.}

\fmtLemma{ki\-ni\-ga\-mbwe\-ne}
\fmtPOS{adv}
\fmtSense{determinedly, resolutely}
\fmtExampleRL{Wee\-ka\-la\-ngu\-le ki\-ni\-ga\-mbwe\-ne.}
\fmtExampleEN{Work resolutely.}

\fmtLemma{ki\-nii\-ka}
\fmtCitation{(oku)\-ki\-nii\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ki\-nii\-ki\-re}
\fmtSense{to mock, to make fun of, to laugh at}
\fmtExampleRL{Ti\-nye\-ndya omu\-ntu aki\-nii\-ka abai\-ra baa\-mwe.}
\fmtExampleEN{I hate a person who mocks others.}

\fmtLemma{ki\-nka\-ta}
\fmtCitation{(e)ki\-nka\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{hot pads, pot holders, trivet}
\fmtExampleRL{Bii\-ki\-rya eki\-na\-ga oku ki\-nka\-ta.}
\fmtExampleEN{Put the saucepan on the trivet.}

\fmtLemma{ki\-nkoo\-na}
\fmtCrossreference{\textnormal{see also} wakikona}
\fmtDialect{Ruruuli}
\fmtCitation{(o)ki\-nkoo\-na}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{pied crow (\textit{Corvus albus})}
\fmtExampleRL{Oki\-nkoo\-na ali\-na ebi\-paa\-pi ebye\-ru n’e\-bi\-ra\-gu\-zu.}
\fmtExampleEN{The plumage of the pied crow is black and white.}

\fmtLemma{ki\-no\-be}
\fmtCitation{(o)ki\-no\-be}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{abandoned person}
\fmtExampleRL{Omwa\-na oyo ki\-no\-be ba\-mbi.}
\fmtExampleEN{Oh dear, that is an abandoned child.}

\fmtLemma{ki\-no\-la}
\fmtCitation{(e)ki\-no\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{type of mushroom (brown inside and cream on the outside)}

\fmtLemma{ki\-no\-nko}
\fmtCrossreference{\textnormal{see also} kinoso\-ka}
\fmtCitation{(e)ki\-no\-nko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sound of cracking one's knuckles}
\fmtExampleRL{Ami\-gi\-re ekya\-la ni mu\-zwa\-mu eki\-no\-nko.}
\fmtExampleEN{He pressed his finger hard and cracked his knuckles.}

\fmtLemma{ki\-no\-so}
\fmtCitation{(e)ki\-no\-so}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fresh groundnuts boiled in their shells}
\fmtExampleRL{Ebi\-no\-so bi\-woo\-ma mu\-no.}
\fmtExampleEN{Fresh groundnuts boiled in their shells are very tasty.}

\fmtLemma{ki\-no\-so\-ka}
\fmtCrossreference{\textnormal{see also} kinonko}
\fmtCitation{(e)ki\-no\-so\-ka}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sound of cracking one's knuckles}
\fmtExampleRL{Abaa\-na ba\-ta\-ka mu\-no oku\-no\-so\-la ebi\-no\-so\-ka.}
\fmtExampleEN{Children love cracking their knuckles.}

\fmtLemma{ki\-nti\-ki\-ti\-ki}
\fmtCitation{(e)ki\-nti\-ki\-ti\-ki}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{owl}
\fmtExampleRL{Eki\-nti\-ki\-ti\-ki ki\-ku\-ku\-nga.}
\fmtExampleEN{The owl is hooting.}

\fmtLemma{ki\-ntu}%
\fmtOne{1}
\fmtCitation{(e)ki\-ntu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{thing, something}
\fmtExampleRL{Ta ani eki\-ntu ekyo.}
\fmtExampleEN{Put that thing here.}
\fmtSubentry(b)
\fmtSense{foetus, unborn baby, pregnancy}
\fmtExampleRL{Omu\-ka\-li ata n’e\-ki\-ntu.}
\fmtExampleEN{The woman is pregnant. (lit.\,The woman is with a thing.)}

\fmtLemma{Ki\-ntu}%
\fmtOne{2}
\fmtCitation{(o)Ki\-ntu}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{the first Muganda}
\fmtExampleRL{Ki\-ntu mbu ni\-ye Mu\-ga\-nda eye\-ke\-rei\-re.}
\fmtExampleEN{It is believed that Kintu was the first Muganda.}

\fmtLemma{ki\-nu\-li\-nu\-li}
\fmtCitation{(e)ki\-nu\-li\-nu\-li}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{sweetish flavour, slight sweetness}
\fmtExampleRL{Omu caa\-yi mu\-bbai\-re\-mu ki\-nu\-li\-nu\-li.}
\fmtExampleEN{The tea was somewhat sweet.}

\fmtLemma{ki\-nu\-mpu\-nu\-mpu}
\fmtCitation{(e)ki\-nu\-mpu\-nu\-mpu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{anther (of banana fruit), flower sheath of the plantain}
\fmtExampleRL{Ebi\-nu\-mpu\-nu\-mpu bi\-ngi oku ma\-too\-ke.}
\fmtExampleEN{There are many anthers on the banana.}

\fmtLemma{ki\-nu\-nko}
\fmtCitation{(e)ki\-nu\-nko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{horny wall of a hoof}
\fmtExampleRL{Eki\-nu\-nko ni\-jo nkai\-to ya ki\-so\-lo.}
\fmtExampleEN{The horny wall of a hoof serves as a shoe for an animal.}

\fmtLemma{ki\-nu\-nu\-li\-ze}
\fmtCitation{(e)ki\-nu\-nu\-li\-ze}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sweetish, slightly sweet}
\fmtExampleRL{Nta\-ka oku\-lya ebi\-ntu bya ki\-nu\-nu\-li\-ze.}
\fmtExampleEN{I like eating slightly sweet things.}

\fmtLemma{ki\-nuu\-lo}
\fmtCitation{(e)ki\-nuu\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{inner part of the hoof}
\fmtExampleRL{Aba\-ntu bai\-ngi ba\-ta\-ka mu\-no oku\-lya ebi\-nuu\-lo.}
\fmtExampleEN{Many people love eating the inner part of the hoof.}

\fmtLemma{ki\-nyaa\-nya}
\fmtCitation{(e)ki\-nyaa\-nya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{impetigo (common skin infection in infants and children)}
\fmtExampleRL{Aba\-nke\-re\-mba ba\-te\-ra oku\-lwa\-la ebi\-nyaa\-nya.}
\fmtExampleEN{New-borns are often affected by impetigo.}

\fmtLemma{ki\-nyaa\-ro}
\fmtCitation{(e)ki\-nyaa\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{urinal}
\fmtExampleRL{Ya\-ba onyaa\-le omu ki\-nyaa\-ro.}
\fmtExampleEN{Go and urinate in the urinal.}

\fmtLemma{ki\-nyaa\-ta}
\fmtCitation{(e)ki\-nyaa\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{food without sauce or relish}
\fmtExampleRL{Ebi\-nyaa\-ta bi\-rii\-bwa boo\-ro.}
\fmtExampleEN{Food without sauce is eaten by the poor.}

\fmtLemma{ki\-nya\-ba\-ntu}
\fmtCitation{(e)ki\-nya\-ba\-ntu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{ghost in a shape of an object}
\fmtExampleRL{Eki\-nya\-ba\-ntu ki\-tii\-ni\-sya abaa\-na.}
\fmtExampleEN{A ghost in a shape of an object scares children.}
\fmtSubentry(b)
\fmtSense{cartoon, animation}
\fmtExampleRL{Eki\-nya\-ba\-ntu mu ki\-ta\-bo kya\-se\-ke\-se\-rye abaa\-na.}
\fmtExampleEN{The cartoon in the book made the children laugh.}

\fmtLemma{ki\-nya\-ba\-yi\-nda}
\fmtCitation{(e)ki\-nya\-ba\-yi\-nda}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{herb used to treat stomach pain}

\fmtLemma{ki\-nyai}
\fmtCitation{(e)ki\-nyai}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{tomentose burbark (\textit{Triumfetta tomentosa}, tropical shrub, the fibre from the bark of the lower part of the stem is widely used for making ropes and strings, these ropes are in turn frequently used for tying reeds in roof constructions)}

\fmtLemma{Ki\-nya\-la}
\fmtCitation{(e)Ki\-nya\-la}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{Banyala way, Banyala style}
\fmtExampleRL{Abi\-ni\-re Ki\-nya\-la.}
\fmtExampleEN{She danced in the style of the Banyala people.}

\fmtLemma{ki\-nya\-la\-lo}
\fmtCitation{(e)ki\-nya\-la\-lo}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{diarrhoea}
\fmtExampleRL{Agwi\-re\-mu eki\-nya\-la\-lo.}
\fmtExampleEN{He has diarrhoea.}

\fmtLemma{ki\-nya\-mu\-ntu}
\fmtCitation{(e)ki\-nya\-mu\-ntu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{doll}
\fmtExampleRL{Omwa\-na wa\-nge mu\-gu\-lii\-re eki\-nya\-mu\-ntu.}
\fmtExampleEN{I bought a doll for my child.}

\fmtLemma{ki\-nya\-mu\-nya\-mu}
\fmtCitation{(e)ki\-nya\-mu\-nya\-mu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{monster, beast, demon}
\fmtExampleRL{Eki\-nya\-mu\-nya\-mu ki\-rya abaa\-na aba mpa\-ka.}
\fmtExampleEN{The monster eats pig-headed children.}

\fmtLemma{ki\-nya\-ngo\-nya\-ngo}
\fmtCitation{(e)ki\-nya\-ngo\-nya\-ngo}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{morning sunshine}

\fmtLemma{ki\-nya\-nya\-nzi}
\fmtCitation{(e)ki\-nya\-nya\-nzi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{cicada}
\fmtExampleRL{Eki\-nya\-nya\-nzi ni\-kyo ki\-sii\-sa eki\-si\-nga oku\-ko\-koo\-ma.}
\fmtExampleEN{The cicada makes the loudest sound among insects.}

\fmtLemma{ki\-nya\-nyi\-ndo}
\fmtCitation{(e)ki\-nya\-nyi\-ndo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{nasal voice, speaking through the nose (\eg in case of a stopped nose)}

\fmtLemma{ki\-nyee\-bwa}
\fmtCitation{(e)ki\-nyee\-bwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{groundnut, groundnut paste, peanut}
\fmtExampleRL{Ebi\-nyee\-bwa byo\-na bi\-mee\-re ku\-sai.}
\fmtExampleEN{All the groundnuts have germinated well.}

\fmtLemma{ki\-nyee\-ri}
\fmtCitation{(e)ki\-nyee\-ri}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{runoff water, streams of water during or after the rain}
\fmtExampleRL{Oi\-ke\-ndi ya\-too\-nye\-re ni mbo\-na n’e\-bi\-nyee\-ri bi\-ngi.}
\fmtExampleEN{It rained and I saw a lot of runoff water.}

\fmtLemma{ki\-nye\-ge}
\fmtCitation{(e)ki\-nye\-ge}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wave}
\fmtExampleRL{Ebi\-nye\-ge bi\-ku\-bbi\-re erya\-to.}
\fmtExampleEN{The waves moved the boat.}

\fmtLemma{ki\-nye\-ge\-nye\-ge}
\fmtCitation{(e)ki\-nye\-ge\-nye\-ge}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{excitement, enthusiasm, impulse, urge, drive}
\fmtExampleRL{Ya\-kwa\-tii\-bwe eki\-nye\-ge\-nye\-ge nga aboi\-ne omwa\-na aku\-bi\-na.}
\fmtExampleEN{He was filled with excitement when he saw the child dancing.}

\fmtLemma{ki\-nye\-nze}
\fmtCitation{(e)ki\-nye\-nze}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{mole cricket}
\fmtExampleRL{Ebi\-nye\-nze bi\-sii\-sa mu\-no ebi\-ru\-me.}
\fmtExampleEN{Mole crickets damage crops badly.}

\fmtLemma{ki\-nyi\-ra}
\fmtCitation{(e)ki\-nyi\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{large fruit bat, large bat}
\fmtExampleRL{Ebi\-nyi\-ra bi\-rya mu\-no ebi\-ba\-la.}
\fmtExampleEN{The large fruit bat eats fruits a lot.}

\fmtLemma{ki\-nyo\-lo}
\fmtCitation{(e)ki\-nyo\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{door bolt, door bar, door handle}
\fmtExampleRL{Iga\-la\-wo obbwe n’e\-ki\-nyo\-lo.}
\fmtExampleEN{Shut the door and fasten it with the door bolt.}

\fmtLemma{ki\-nyo\-lo\-lo}
\fmtCitation{(e)ki\-nyo\-lo\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{tape measure, measuring tape}
\fmtExampleRL{Ntii\-zaa\-ku eki\-nyo\-lo\-lo kya\-mu.}
\fmtExampleEN{Lend me your tape measure, please.}

\fmtLemma{ki\-nyo\-mo}
\fmtCitation{(e)ki\-nyo\-mo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{breeding ground for lung fish and mud fish}
\fmtExampleRL{Endu zee\-bbi\-sa mu bi\-nyo\-mo okwa kya\-nda.}
\fmtExampleEN{Lung fish hide in the breeding grounds during the dry spell.}

\fmtLemma{ki\-nyo\-ndo}
\fmtCrossreference{\textnormal{see} kiyondo}

\fmtLemma{ki\-nyo\-nyi}
\fmtCitation{(e)ki\-nyo\-nyi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bird}
\fmtExampleRL{Uga\-nda eri\-mu ebi\-nyo\-nyi bya mi\-ri\-ngo gi\-ngi.}
\fmtExampleEN{There are many varieties of birds in Uganda.}

\fmtLemma{ki\-nyoo\-zi}%
\fmtOne{1}
\fmtCitation{(o)ki\-nyoo\-zi}
\fmtPOS{n}%
\fmtSource{from Ar. via Sw.}
\fmtNC{1a/2}
\fmtSense{barber}

\fmtLemma{ki\-nyoo\-zi}%
\fmtOne{2}
\fmtCitation{(e)ki\-nyoo\-zi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{type of banana eaten only when it is ripe}

\fmtLemma{Ki\-nyo\-ro}
\fmtCitation{(e)Ki\-nyo\-ro}
\fmtPOS{adv}%
\fmtNC{7}
\fmtSense{the Kinyoro way, the Kinyoro style}
\fmtExampleRL{Enke\-nde eyo ya Ki\-nyo\-ro.}
\fmtExampleEN{That is a Kinyoro way of dancing.}

\fmtLemma{ki\-nyu\-mba kya mbu\-li}
\fmtCrossreference{\textnormal{see also} nyumu}
\fmtSense{goat shed}
\fmtExampleRL{Eki\-nyu\-mba kya mbu\-li ki\-ge\-gei\-re.}
\fmtExampleEN{The goat shed is very dirty.}

\fmtLemma{ki\-nyu\-mo}
\fmtCrossreference{\textnormal{see} kinyumu}

\fmtLemma{ki\-nyu\-mu}
\fmtCitation{(e)ki\-nyu\-mu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{amusement, entertainment, merry making, fun}
\fmtExampleRL{Nta\-ka mu\-no ebi\-nyu\-mu.}
\fmtExampleEN{I like to have fun.}

\fmtLemma{ki\-nyu\-si}%
\fmtOne{1}
\fmtCitation{(e)ki\-nyu\-si}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{eschar, necrotic tissue (a dry, dark scab or falling away of dead skin, \eg caused by a burn)}

\fmtLemma{ki\-nyu\-si}%
\fmtOne{2}
\fmtCitation{(e)ki\-nyu\-si}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{gist, central point, essence, central theme, heart of the matter}
\fmtExampleRL{Ika oku ki\-nyu\-si.}
\fmtExampleEN{Get to the point.}

\fmtLemma{ki\-nyu\-zi}
\fmtCitation{(e)ki\-nyu\-zi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{brake pad (of bicycle)}
\fmtExampleRL{Eki\-nyu\-zi kya gaa\-li ya\-nge ki\-gwi\-re.}
\fmtExampleEN{The brake pad of my bicycle is lost.}

\fmtLemma{ki\-nywa}
\fmtCitation{(e)ki\-nywa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{bundle of fire wood}
\fmtExampleRL{Eki\-nywa kya nkwi kye yee\-ti\-ki\-re ei\-go\-lo kya\-bbai\-re ki\-ku\-le\-mee\-ra.}
\fmtExampleEN{The bundle of wood she carried yesterday was very heavy.}
\fmtSubentry(b)
\fmtSense{muscle}
\fmtExampleRL{Aba\-zee\-nyi ba mu\-pii\-ra ba\-li\-na ebi\-nywa ebi\-gu\-mu.}
\fmtExampleEN{Football players have well-developed muscles.}

\fmtLemma{ki\-nywa\-bwi\-no}
\fmtCitation{(e)ki\-nywa\-bwi\-no}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{blotting paper}
\fmtExampleRL{Eki\-nywa\-bwi\-no kyo\-no\-ke\-re\-mu aka\-ti.}
\fmtExampleEN{The blotting paper is torn in two.}
\fmtSubentry(b)
\fmtSense{ink pad}
\fmtExampleRL{Ini\-ka esi\-ta\-mpu omu ki\-nywa\-bwi\-no.}
\fmtExampleEN{Place the stamp in the inkpad.}

\fmtLemma{ki\-nywa\-nsa\-nde}
\fmtCitation{(e)ki\-nywa\-nsa\-de}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sugar ant}
\fmtExampleRL{Ebi\-nywa\-nsa\-nde bi\-ta\-ka mu\-no oku\-nuu\-na osu\-kaa\-li.}
\fmtExampleEN{Sugar ants really love licking sugar.}

\fmtLemma{ki\-nywe\-ro}
\fmtCitation{(e)ki\-nywe\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{drinking bowl for birds}
\fmtExampleRL{Eki\-nywe\-ro kya nko\-ko ki\-sai\-ne oku\-kuu\-mi\-bwa nga ki\-yo\-njo.}
\fmtExampleEN{The chickens' drinking bowl should be kept clean.}
\fmtSubentry(b)
\fmtSense{bar, drinking place}
\fmtExampleRL{Ama\-ka gaa\-mwe ya\-ga\-foi\-re ki\-nywe\-ro.}
\fmtExampleEN{He turned his home into a bar.}

\fmtLemma{ki\-nzaa\-li}
\fmtCitation{(e)ki\-nzaa\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{turmeric plant, curry powder}
\fmtExampleRL{Eki\-nzaa\-li ki\-woo\-me\-sya omu\-ku\-bi.}
\fmtExampleEN{Curry powder makes sauce delicious.}
\fmtSubentry(b)
\fmtSense{dark yellow colour}
\fmtExampleRL{Ovu\-lo yaa\-mwe ya ki\-nzaa\-li.}
\fmtExampleEN{His overall suit is dark yellow.}

\fmtLemma{ki\-nzi}
\fmtCitation{(e)ki\-nzi}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{flu, influenza}
\fmtExampleRL{Mu\-lwai\-re wa ki\-nzi.}
\fmtExampleEN{He is suffering from the flu.}

\fmtLemma{ki\-nzi\-ka}
\fmtCitation{(oku)\-ki\-nzi\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ki\-nzi\-ki\-re}
\fmtSense{to tilt, to angle, to lean, to slope}
\fmtExampleRL{Ki\-nzi\-ka ebbi\-ni\-ka edi.}
\fmtExampleEN{Tilt that kettle.}

\fmtLemma{ki\-paa\-li}
\fmtCitation{(e)ki\-paa\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{papyrus mat}
\fmtExampleRL{Ai\-kai\-re ku ki\-paa\-li.}
\fmtExampleEN{He is sitting on a papyrus mat.}
\fmtSubentry(b)
\fmtSense{carpet, rug}
\fmtExampleRL{Ei\-rii\-ro lya\-mwe ali\-tai\-re\-mu eki\-paa\-li.}
\fmtExampleEN{He put a carpet in his dining room.}

\fmtLemma{ki\-paa\-paa\-li}
\fmtCitation{(e)ki\-paa\-paa\-li}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{orange colour, colout of pawpaw}
\fmtExampleRL{Enkai\-to za\-nge za ki\-paa\-paa\-li.}
\fmtExampleEN{My shoes are orange.}

\fmtLemma{ki\-paa\-pi}
\fmtCitation{(e)ki\-paa\-pi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{patch}
\fmtExampleRL{Esaa\-ti yaa\-mwe eri\-ku eki\-paa\-pi.}
\fmtExampleEN{His shirt has a patch.}

\fmtLemma{ki\-paa\-ti\-paa\-ti}
\fmtPOS{adv}
\fmtSense{mercilessly}
\fmtExampleRL{Ai\-ta ki\-paa\-ti\-paa\-ti.}
\fmtExampleEN{He kills mercilessly.}

\fmtLemma{ki\-pa\-ku}
\fmtCitation{(o)ki\-pa\-ku}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{one who acts hurriedly}
\fmtExampleRL{Ki\-pa\-ku abbai\-re\-ku ani.}
\fmtExampleEN{The person who acts hurriedly has been here.}

\fmtLemma{ki\-pa\-mpa}
\fmtCitation{(e)ki\-pa\-mpa}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{smell of rotten potato}
\fmtExampleRL{Omu nyu\-mba mu\-li\-mu eki\-pa\-mpa.}
\fmtExampleEN{It smells of rotten potatos in the house.}

\fmtLemma{ki\-pa\-mpa\-ga\-lo}
\fmtCitation{(e)ki\-pa\-mpa\-ga\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{mudguard (of a bicycle)}
\fmtExampleRL{Eki\-pa\-mpa\-ga\-lo kya gaa\-li ya\-nge ki\-zwi\-re\-ku.}
\fmtExampleEN{The mudguard has fallen off my bicycle.}

\fmtLemma{ki\-pa\-mpa\-gu\-lo}
\fmtCitation{(e)ki\-pa\-mpa\-gu\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{shell of a snail}
\fmtExampleRL{Eki\-taa\-nii\-na ki\-bba mu ki\-pa\-mpa\-gu\-lo.}
\fmtExampleEN{A snail lives in a shell.}

\fmtLemma{ki\-pa\-nde}
\fmtCitation{(e)ki\-pa\-nde}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{signpost, sign}
\fmtExampleRL{Ki\-ni eki\-pa\-nde ki\-ra\-ngii\-ra wa mu\-wee\-si.}
\fmtExampleEN{This signpost points to the blacksmith.}
\fmtSubentry(b)
\fmtSense{notice board}
\fmtExampleRL{Ya\-ba ote ekii\-sa\-ni ki\-ni oku ki\-pa\-nde.}
\fmtExampleEN{Go and put this photo on the notice board.}
\fmtSubentry(c)
\fmtSense{billboard, poster}
\fmtExampleRL{Adi wa\-li\-wo eki\-pa\-nde eki\-koo\-to.}
\fmtExampleEN{There is a big billboard there.}
\fmtSubentry(d)
\fmtSense{placard, banner}
\fmtExampleRL{Mboi\-ne aba\-ntu ni ba\-kwai\-te ebi\-pa\-nde.}
\fmtExampleEN{I saw people holding placards.}
\fmtSubentry(e)
\fmtSense{poster}
\fmtExampleRL{Mu\-ta\-ti\-mba bi\-pa\-nde bya\-nywe ku ki\-sii\-ka ki\-ni.}
\fmtExampleEN{Do not place your posters on this wall.}
\fmtSubentry(f)
\fmtSense{medical card}
\fmtExampleRL{Aba\-ka\-li ba nda ba\-ba\-wa ebi\-pa\-nde.}
\fmtExampleEN{Women who are pregnant are given medical cards.}

\fmtLemma{ki\-pa\-nga}
\fmtCitation{(e)ki\-pa\-nga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{machete, panga}
\fmtExampleRL{Mpee\-rya\-ku eki\-pa\-nga kya\-mu nte\-me emi\-saa\-le.}
\fmtExampleEN{Help me with your panga so that I can cut the trees.}

\fmtLemma{ki\-pa\-pa}
\fmtCitation{(e)ki\-pa\-pa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wing}
\fmtExampleRL{Onko\-ko abba n’e\-bi\-pa\-pa.}
\fmtExampleEN{The hen has wings.}

\fmtLemma{ki\-pa\-ra\-ti}
\fmtCitation{(e)ki\-pa\-ra\-ti}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{permit, license}
\fmtExampleRL{Okwi\-za ku\-bbwe\-bwa ni\-wa vu\-ga nga to\-li\-na ki\-pa\-ra\-ti.}
\fmtExampleEN{You will be arrested if you drive without a permit.}
\fmtSubentry(b)
\fmtSense{immunisation card, vaccination card, vaccination certificate, yellow card}
\fmtExampleRL{Eki\-pa\-ra\-ti kya Na\-mu\-bi\-ru ki\-buu\-re.}
\fmtExampleEN{Namubiru's immunisation card is lost.}

\fmtLemma{ki\-pi\-mo}
\fmtCitation{(e)ki\-pi\-mo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{measurement, quantity, amount}
\fmtExampleRL{Wa\-ko\-le\-se\-rye ki\-pi\-mo ki ekya su\-kaa\-li?}
\fmtExampleEN{What quantity of sugar did you order?}

\fmtLemma{ki\-poo\-li}
\fmtCitation{(e)ki\-poo\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{paste (\eg of groundnut, simsim (UgE) or white ants)}
\fmtExampleRL{Nta\-ka mu\-no oku\-lya eki\-poo\-li kya ma\-ca\-nde.}
\fmtExampleEN{I love eating simsim paste.}

\fmtLemma{ki\-poo\-poo\-yo}
\fmtCitation{(e)ki\-poo\-poo\-yo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{beanless coffee, beanless groundnuts, beanless beans}
\fmtExampleRL{Ebi\-poo\-poo\-yo bya mwa\-ni ti\-bi\-tu\-ndi\-bwa.}
\fmtExampleEN{Coffee without beans cannot be sold.}

\fmtLemma{ki\-pu\-mpu\-li\-ra}
\fmtCitation{(e)ki\-pu\-mpu\-li\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{apron-like skirt}
\fmtExampleRL{Omu\-ka\-li azwai\-re eki\-pu\-mpu\-li\-ra.}
\fmtExampleEN{The woman is wearing an apron-like skirt.}
\fmtSubentry(b)
\fmtSense{skirt}
\fmtExampleRL{Eki\-pu\-mpu\-li\-ra ky'a\-guu\-re ki\-ku\-mwi\-ka ku\-sai.}
\fmtExampleEN{The skirt she bought fits her well.}

\fmtLemma{ki\-ra}
\fmtCitation{(oku)\-ki\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kii\-re}
\fmtSense{to be better than}
\fmtExampleRL{Oyo omwa\-na aki\-ra abai\-ra baa\-mwe.}
\fmtExampleEN{That child is better than her friends.}

\fmtLemma{ki\-raa\-lo}
\fmtCitation{(e)ki\-raa\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from En.}
\fmtSubentry(a)
\fmtSense{kraal}
\fmtExampleRL{Ente zi\-bba\-sya mu ki\-raa\-lo.}
\fmtExampleEN{The cattle stays in a kraal.}
\fmtSubentry(b)
\fmtSense{herd}
\fmtExampleRL{Odi omu\-sai\-za mu\-guu\-da ata n’e\-ki\-raa\-lo kya nte.}
\fmtExampleEN{That man is rich, he has a herd of cattle.}

\fmtLemma{ki\-raa\-mo}
\fmtCitation{(e)ki\-raa\-mo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{will, testament}
\fmtExampleRL{Ki\-sai omu\-ntu oku\-ko\-la eki\-raa\-mo ng’a\-kya\-li ku\-fa.}
\fmtExampleEN{It is good for a person to make a will before he dies.}

\fmtLemma{ki\-raa\-nyu\-ma}
\fmtCitation{(e)ki\-raa\-nyu\-ma}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{going backwards, retrogression}
\fmtExampleRL{Eki\-raa\-nyu\-ma ki\-bbi nka omu bi\-ro ebya si\-rii\-mu.}
\fmtExampleEN{Retrogression is bad especially in this era of HIV pandemic.}

\fmtLemma{ki\-raa\-ro}
\fmtCitation{(e)ki\-raa\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{inn}
\fmtExampleRL{OYe\-su ya\-bya\-lii\-bwe mu ki\-raa\-ro kya ba\-ge\-nyi.}
\fmtExampleEN{Jesus was born in an inn.}

\fmtLemma{ki\-raa\-to}
\fmtCitation{(e)ki\-raa\-to}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{sheath}
\fmtExampleRL{Empi\-ri\-ma agi\-tai\-re mu ki\-raa\-to.}
\fmtExampleEN{He has put the sword into the sheath.}
\fmtSubentry(b)
\fmtSense{tyre sandal, tyre shoe}
\fmtExampleRL{Aba\-ka\-ra\-mo\-ja ba\-ta\-ka mu\-no ebi\-raa\-to.}
\fmtExampleEN{Karamojongs like tyre sandals.}

\fmtLemma{ki\-ra\-bo}
\fmtCitation{(e)ki\-ra\-bo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{gift, present}
\fmtExampleRL{Ka mu\-wee\-rye\-yo eki\-ra\-bo.}
\fmtExampleEN{Let me give her a gift.}
\fmtSubentry(b)
\fmtSense{offertory, offering}
\fmtExampleRL{Ki\-sai oku\-waa\-yo eki\-ra\-bo omu ira\-myo.}
\fmtExampleEN{It is good to make an offering in church.}

\fmtLemma{ki\-ra\-gaa\-no}
\fmtCitation{(e)ki\-ra\-gaa\-no}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{marriage vow}
\fmtExampleRL{Aba\-go\-le ba\-ku\-bbi\-re eki\-ra\-gaa\-no.}
\fmtExampleEN{The couple made a marriage vow.}

\fmtLemma{ki\-ra\-ga\-la}
\fmtCitation{(o)ki\-ra\-ga\-la}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSubentry(a)
\fmtSense{green colour}
\fmtExampleRL{Ebi\-koo\-la bi\-te\-ra ku\-bba bya ki\-ra\-ga\-la.}
\fmtExampleEN{Leaves are usually green.}
\fmtSubentry(b)
\fmtFixedexpression{ki\-ra\-ga\-la (o)mu\-kwe\-nyi}
\fmtSense{light green}
\fmtExampleRL{Ba\-zwa\-la yu\-ni\-foo\-mu ya ki\-ra\-ga\-la omu\-kwe\-nyi oku iso\-me\-ro lwa\-mwe.}
\fmtExampleEN{They wear a light green uniform at her school.}

\fmtLemma{ki\-ra\-ga\-ni\-syo}
\fmtCitation{(e)ki\-ra\-ga\-ni\-syo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{agreement, arrangement}
\fmtExampleRL{Ba\-koo\-re eki\-ra\-gaa\-ni\-syo obu\-tai\-ra\-mu ku\-lwa\-na\-nga\-na.}
\fmtExampleEN{They have made an agreement to never fight each other.}

\fmtLemma{ki\-ra\-gi\-ro}
\fmtCitation{(e)ki\-ra\-gi\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{instruction, directive, directions}
\fmtExampleRL{Ale\-ke\-re ebi\-ra\-gi\-ro ebya baa\-na oku\-bba\-sya nga bu\-kya\-li.}
\fmtExampleEN{He left instructions for the kids to go to bed early.}
\fmtSubentry(b)
\fmtSense{law, order}
\fmtExampleRL{Ebi\-ra\-gi\-ro bye Uga\-nda bi\-ta bi\-zi\-bu oku\-ta mu nko\-la.}
\fmtExampleEN{Uganda's laws are difficult to implement.}

\fmtLemma{ki\-rai}
\fmtCitation{(e)ki\-rai}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{yam}
\fmtExampleRL{Eki\-rai ki\-rwa\-wo oku\-ya.}
\fmtExampleEN{A yam takes long to cook.}

\fmtLemma{ki\-ra\-ka}
\fmtCitation{(e)ki\-ra\-ka}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{patch}
\fmtExampleRL{Ata n’e\-ki\-ra\-ka oku mpa\-le yaa\-mwe.}
\fmtExampleEN{He has a patch on his trousers.}

\fmtLemma{ki\-ra\-mbi}
\fmtCitation{(e)ki\-ra\-mbi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{corn rows hair style}
\fmtSubentry(b)
\fmtFixedexpression{ku\-\-bba (e)bi\-ra\-mbi}
\fmtSense{to curl one's hair}
\fmtExampleRL{Omu\-ka\-li aku\-bbi\-re ebi\-ra\-mbi.}
\fmtExampleEN{The woman has curled her hair.}

\fmtLemma{ki\-ra\-mu}
\fmtCitation{(oku)\-ki\-ra\-mu}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ki\-re\-mu}
\fmtSense{to exceed, to surpass}
\fmtExampleRL{Ki\-ni ki\-ku\-ki\-ra\-mu obwi\-ngi bwa saa\-ti ze twa\-la\-gi\-rii\-ze.}
\fmtExampleEN{This exceeds the number of shirts we ordered.}

\fmtLemma{ki\-ra\-mu\-kyo}
\fmtCrossreference{\textnormal{see also} lamukya}
\fmtCitation{(e)ki\-ra\-mu\-kyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{greeting}
\fmtExampleRL{Ti\-mmai\-te ki\-ra\-mu\-kyo kya ba\-si\-raa\-mu.}
\fmtExampleEN{I do not know the Islamic way of greeting.}

\fmtLemma{ki\-ra\-myo}
\fmtCitation{(e)ki\-ra\-myo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{incantation, magic spell}
\fmtExampleRL{Ba\-mu\-ku\-bbi\-re eki\-ra\-myo nga mu\-lwai\-re.}
\fmtExampleEN{They performed an incantation when she was sick.}
\fmtSubentry(b)
\fmtSense{prayer}
\fmtExampleRL{Bu\-li omwei amai\-te eki\-ra\-myo kya mu\-ka\-ma wai\-swe.}
\fmtExampleEN{Everyone knows the Lord's prayer.}

\fmtLemma{ki\-ra\-ngi}
\fmtCitation{(e)ki\-ra\-ngi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{stage of growth}
\fmtExampleRL{OPe\-te\-r ali mu ki\-ra\-ngi ky’a\-ba\-sai\-za.}
\fmtExampleEN{Peter is now at the stage of becoming a man.}
\fmtSubentry(b)
\fmtSense{medium height}
\fmtExampleRL{Omu\-ka\-li wa\-nge wa ki\-ra\-ngi.}
\fmtExampleEN{My wife is of medium height.}

\fmtLemma{ki\-ra\-ngii\-ro}
\fmtCrossreference{\textnormal{see also} kirango}
\fmtCitation{(e)ki\-ra\-ngii\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{official announcement}
\fmtExampleRL{Oje\-ge\-nge akoo\-re eki\-ra\-ngii\-ro.}
\fmtExampleEN{The supreme leader has made an official announcement.}

\fmtLemma{ki\-ra\-ngo}
\fmtCrossreference{\textnormal{see also} kirangiiro}
\fmtCitation{(e)ki\-ra\-ngo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{official announcement}
\fmtExampleRL{Aba\-ntu ba\-ja\-ga\-lai\-re olu\-zwa\-nyu\-ma lwa ki\-ra\-ngo.}
\fmtExampleEN{The crowd was in uproar after the announcement.}

\fmtLemma{ki\-ra\-so}
\fmtCrossreference{\textnormal{see also} kalaso}
\fmtCitation{(e)ki\-ra\-so}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSubentry(a)
\fmtSense{arrow}
\fmtSubentry(b)
\fmtSense{bleeding of a animal when hit with an arrow or other pointed weapon}

\fmtLemma{ki\-ra\-wu\-li}
\fmtCitation{(e)ki\-ra\-wu\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from Ar. via Sw.}
\fmtSense{lantern glass, lantern shade, lamp shade}
\fmtExampleRL{Eki\-ra\-wu\-li kya taa\-la kya\-ti\-ki\-re.}
\fmtExampleEN{The lantern glass has broken.}

\fmtLemma{ki\-ra\-yi\-ro}
\fmtCitation{(e)ki\-ra\-yi\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{oath, promise}
\fmtExampleRL{Opu\-re\-zi\-de\-nti ali\-na oku\-ku\-bba eki\-ra\-yi\-ro ku mu\-ko\-lo gwa ku mu\-la\-yi\-za.}
\fmtExampleEN{The president has to take an oath during the swearing-in ceremony.}

\fmtLemma{ki\-re\-be}
\fmtPOS{adv}
\fmtSense{other, another}
\fmtExampleRL{Ya\-ba oswe\-re mu ki\-ka ki\-re\-be.}
\fmtExampleEN{Go and marry someone from the other clan.}

\fmtLemma{ki\-re\-bee\-ki}
\fmtCitation{(e)ki\-re\-bee\-ki}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{residence for people of the Bushbuck clan}
\fmtExampleRL{Aba\-re\-bee\-ki bai\-ka\-la mu Ki\-re\-bee\-ki.}
\fmtExampleEN{The Bushbuck clan members stay in Kirebeeki.}

\fmtLemma{ki\-ree\-go}%
\fmtOne{1}
\fmtCitation{(e)ki\-ree\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{failure to thrive (insufficient weight gain or inappropriate weight loss in children)}

\fmtLemma{ki\-ree\-go}%
\fmtOne{2}
\fmtCitation{(e)ki\-ree\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{expensive item, high-priced item}
\fmtExampleRL{Ama\-fu\-ta ga bi\-ree\-go ebi\-ro bi\-ni.}
\fmtExampleEN{Fuel is expensive these days.}

\fmtLemma{ki\-re\-ga}
\fmtCitation{(e)ki\-re\-ga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{ferrule}
\fmtExampleRL{Ta eki\-re\-ga oku isu\-mu.}
\fmtExampleEN{Put a ferrule on the bottom of a spear.}

\fmtLemma{ki\-re\-ju}
\fmtCitation{(e)ki\-re\-ju}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{beard}
\fmtExampleRL{Aba\-ta\-bbu\-li\-ki ba\-bba n’e\-bi\-re\-ju ebi\-koo\-to.}
\fmtExampleEN{The Tabliques have long thick beards.}

\fmtLemma{ki\-re\-ka}
\fmtCitation{(e)ki\-re\-ka}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{yellowish banana leaf, banana leaf turning yellow}
\fmtExampleRL{Eki\-go\-go ki\-na\-gi\-re ebi\-re\-ka.}
\fmtExampleEN{The banana's leaves have turned yellow.}

\fmtLemma{ki\-re\-mba}
\fmtCitation{(o)ki\-re\-mba}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{black or brown cow with patches of white dots}
\fmtExampleRL{Oki\-re\-mba wa\-nge ba\-mwi\-bbi\-re.}
\fmtExampleEN{My black cow with patches of white dots got stolen.}

\fmtLemma{ki\-re\-nge}
\fmtCitation{(e)ki\-re\-nge}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{lower part of a leg of herbivorous animal, hoof (UgE)}
\fmtExampleRL{Eki\-re\-nge kya nte ki\-woo\-ma.}
\fmtExampleEN{The lower part of the cow's leg is tasty.}
\fmtSubentry(b)
\fmtSense{bottom hem of garment}
\fmtExampleRL{Ego\-me\-si ebbaa\-ku eki\-re\-nge.}
\fmtExampleEN{A gomesi has a hem.}

\fmtLemma{ki\-re\-nge\-re\-nge}
\fmtCitation{(e)ki\-re\-nge\-re\-nge}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{reed-like plant growing near the lake shore}
\fmtExampleRL{Ebi\-re\-nge\-re\-nge bi\-bba ampi n’e\-nya\-nja.}
\fmtExampleEN{Reed-like plants grow near the lake shore.}
\fmtSubentry(b)
\fmtSense{reed, straw}
\fmtSubentry(c)
\fmtSense{straw container}
\fmtExampleRL{Obu\-se\-ke mu\-bu\-bi\-ki\-re mu ki\-re\-nge\-re\-nge.}
\fmtExampleEN{Keep the straws in the straw container.}

\fmtLemma{ki\-re\-rya}
\fmtCitation{(e)ki\-re\-rya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{coffee without beans}
\fmtExampleRL{Ebi\-re\-rya bi\-bbi.}
\fmtExampleEN{Coffee without beans is bad.}

\fmtLemma{ki\-re\-zu}
\fmtCrossreference{\textnormal{see} kireju}

\fmtLemma{ki\-rii\-sa}
\fmtCitation{(e)ki\-rii\-sa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{nutrient, fertiliser}
\fmtExampleRL{Aya\-bi\-re Ka\-mpa\-la ku\-gu\-la bi\-rii\-sa.}
\fmtExampleEN{He has gone to Kampala to buy fertiliser.}

\fmtLemma{ki\-ri\-ki\-ta\-na}
\fmtCitation{(oku)\-ki\-ri\-ki\-ta\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ki\-ri\-ki\-tai\-ne}
\fmtSense{to act by trial and error}
\fmtExampleRL{OBa\-kwa\-na aki\-ri\-ki\-ta\-na mu\-no.}
\fmtExampleEN{Bakwana does things by trial and error.}

\fmtLemma{ki\-ri\-maa\-wa}
\fmtCitation{(e)ki\-ri\-maa\-wa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{large type of lemons}
\fmtExampleRL{Eki\-ri\-maa\-wa ki\-ta\-mba eki\-koo\-lo.}
\fmtExampleEN{The large lemon cures a cough.}

\fmtLemma{ki\-ri\-mba}
\fmtCitation{(e)ki\-ri\-mba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{cluster of fruits}
\fmtExampleRL{Ano\-koi\-re eki\-ri\-mba kya ma\-pee\-ra.}
\fmtExampleEN{She picked a cluster of guavas.}
\fmtSubentry(b)
\fmtSense{way of sitting with legs crossed}
\fmtExampleRL{Mmwi\-ti\-rai\-ne ai\-kai\-re ki\-ri\-mba.}
\fmtExampleEN{I found him sitting with his legs crossed.}

\fmtLemma{ki\-ri\-mbi}
\fmtCitation{(e)ki\-ri\-mbi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{river bank, lake shore, beach}
\fmtExampleRL{OMo\-se\-s abba ampi n’e\-ki\-ri\-mbi.}
\fmtExampleEN{Moses lives close to the riverbank.}

\fmtLemma{ki\-ri\-me}
\fmtCrossreference{\textnormal{see} kirume}

\fmtLemma{ki\-ri\-myo}
\fmtCrossreference{\textnormal{see also} kirumyo}
\fmtDialect{Lu\-nyala}
\fmtCitation{(e)ki\-ri\-myo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{hole digger, excavator}
\fmtExampleRL{Mpee\-rya eki\-ri\-myo nsa\-ke ebii\-na bya nyu\-mba.}
\fmtExampleEN{Give me a digger and I'll dig the foundation for the house.}

\fmtLemma{ki\-ri\-nda}
\fmtCitation{(e)ki\-ri\-nda}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{loin cloth for women}
\fmtExampleRL{Eki\-ri\-nda kya mu\-ka\-li ki\-bba mu\-nda.}
\fmtExampleEN{A loincloth for women is worn under the clothes.}

\fmtLemma{ki\-ri\-ndi}
\fmtCitation{(e)ki\-ri\-ndi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{headgear used in traditional worship}
\fmtExampleRL{Ba\-ki\-ba\-ndwa ba\-zwa\-la ebi\-ri\-ndi.}
\fmtExampleEN{Traditional priests wear headgear.}

\fmtLemma{ki\-ri\-ndi\-ro}
\fmtCitation{(e)ki\-ri\-ndi\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{waiting place (for a hunter waiting to spear an animal)}
\fmtExampleRL{Omwi\-gi ali ku ki\-ri\-ndi\-ro ku\-su\-mi\-ta ki\-so\-lo.}
\fmtExampleEN{The hunter is stationed in the waiting place to spear the animal.}
\fmtSubentry(b)
\fmtSense{bus station, waiting room}
\fmtExampleRL{Na\-mwa\-ji\-rye ku ki\-ri\-ndi\-ro ei\-zo.}
\fmtExampleEN{I met him at the bus station yesterday.}

\fmtLemma{ki\-ri\-ndwe}
\fmtCitation{(e)ki\-ri\-ndwe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{net for sliver fish}
\fmtExampleRL{Obu\-fu\-lu mbu\-te\-ge\-sya ki\-ri\-ndwe.}
\fmtExampleEN{I trap the sliver fish using a silver fish net.}

\fmtLemma{ki\-ri\-ta}
\fmtCrossreference{\textnormal{see also} girita}
\fmtCitation{(e)ki\-ri\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{razor blade}
\fmtExampleRL{Ndee\-te\-ra eki\-ri\-ta eyo nge\-mbe omu\-le\-zu.}
\fmtExampleEN{Bring me that razor blade and I'll shave off my beard.}

\fmtLemma{ki\-ri\-ŋa\-li\-ŋa}
\fmtCitation{(e)ki\-ri\-ŋa\-li\-ŋa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{snail}
\fmtExampleRL{Eki\-ri\-ŋa\-li\-ŋa ki\-re\-ka omu\-ku\-ru\-lo bu\-li we kya\-kai\-re.}
\fmtExampleEN{A snail leaves a mark wherever it passes.}

\fmtLemma{ki\-ro}%
\fmtOne{1}
\fmtCitation{(e)ki\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{day}
\fmtExampleRL{Waa\-kai\-re\-wo ebi\-ro bi\-bi\-ri bu\-kya omwa\-na awa\-mbi\-bwa.}
\fmtExampleEN{Two days have gone by since the child was kidnapped.}

\fmtLemma{ki\-ro}%
\fmtOne{2}
\fmtCitation{(e)ki\-ro}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{kilogram}
\fmtExampleRL{Nguu\-re eki\-ro isa\-tu eza bu\-wu\-nga.}
\fmtExampleEN{I have bought three kilograms of maize floor.}

\fmtLemma{ki\-ro\-byo}
\fmtCitation{(e)ki\-ro\-byo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{long pruning stick}
\fmtExampleRL{Lee\-ta eki\-ro\-byo tu\-sa\-le enda\-ga\-la.}
\fmtExampleEN{Bring the long pruning stick and we'll cut banana leaves.}

\fmtLemma{ki\-ro\-ga\-mu\-tuu\-bi}
\fmtCitation{(e)ki\-ro\-ga\-mu\-tuu\-bi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{big mug}
\fmtExampleRL{Anywee\-ra mu ki\-ro\-ga\-mu\-tuu\-bi.}
\fmtExampleEN{She drinks from a big mug.}

\fmtLemma{ki\-ro\-ko\-dyo}
\fmtCitation{(e)ki\-ro\-ko\-dyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{long pruning stick, long shaft with a knife at the end used for pruning plantains}
\fmtExampleRL{Lee\-ta eki\-ro\-ko\-dyo tu\-wa\-nu\-le emi\-ye\-mbe.}
\fmtExampleEN{Bring the long pruning stick and we'll harvest mangoes.}

\fmtLemma{ki\-ro\-koo\-ro}
\fmtCitation{(e)ki\-ro\-koo\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{pair of vocal cords}
\fmtExampleRL{Omu\-sai\-za oyo ali\-na eki\-ro\-koo\-ro eki\-koo\-to.}
\fmtExampleEN{That man has big vocal cords.}

\fmtLemma{ki\-ro\-mbe}
\fmtCitation{(e)ki\-ro\-mbe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{mine, stone quarry}
\fmtExampleRL{Omu\-se\-nyu ngu\-sa\-ka mu ki\-ro\-mbe kya mu\-se\-nyu.}
\fmtExampleEN{I dig sand from the sand mine.}

\fmtLemma{Ki\-roo\-ma}
\fmtCitation{(e)Ki\-roo\-ma}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{Roman Catholic church}
\fmtExampleRL{Opaa\-pa ni\-ye akuu\-ra aba\-ka\-tu\-li\-ki ab’e\-Ki\-roo\-ma.}
\fmtExampleEN{The pope is the head of the Roman Catholic church.}

\fmtLemma{ki\-roo\-ro}
\fmtCitation{(e)ki\-roo\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{shrine, temple}
\fmtExampleRL{Ba\-kwa\-ba mu ki\-roo\-ro ku\-ba\-ndwa.}
\fmtExampleEN{They are going to the shrine to worship.}
\fmtSubentry(b)
\fmtSense{tent}
\fmtExampleRL{Abbwe\-re adi eki\-roo\-ro kya\-mwe ku\-sai.}
\fmtExampleEN{He has set up his tent properly.}

\fmtLemma{ki\-roo\-to}%
\fmtOne{1}
\fmtCitation{(e)ki\-roo\-to}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{vulture}
\fmtExampleRL{Ebi\-roo\-to bi\-rya nya\-ma.}
\fmtExampleEN{Vultures eat meat.}

\fmtLemma{ki\-roo\-to}%
\fmtOne{2}
\fmtCitation{(e)ki\-roo\-to}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{dream}
\fmtExampleRL{Na\-fu\-ni\-re eki\-roo\-to eki\-bbi ei\-zo.}
\fmtExampleEN{I had a bad dream yesterday.}

\fmtLemma{ki\-roo\-wa}
\fmtCitation{(e)ki\-roo\-wa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{jatropha (a flowering plant in the spurge family)}
\fmtExampleRL{Si\-mba eki\-roo\-wa oku lu\-sa\-lo\-sa\-lo lwa ki\-ba\-nja kya\-nge.}
\fmtExampleEN{Plant a jatropha plant at the boundary of my land.}

\fmtLemma{ki\-ro\-ri\-ro\-li}
\fmtPOS{adv}
\fmtSense{shamefully, shyly, dishonourably, ashamedly}
\fmtExampleRL{Aku\-lo\-la ki\-ro\-ri\-ro\-li nka omu\-ka\-li gwe ba\-kwai\-te n’o\-mu\-sai\-za.}
\fmtExampleEN{She looks ashamed, like a woman caught committing adultery.}

\fmtLemma{ki\-ro\-woo\-za}
\fmtCitation{(e)ki\-ro\-woo\-za}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{thought, idea, opinion}
\fmtExampleRL{Aba\-sa\-wu ba\-ko\-be\-re nti oku\-fa kwa\-mwe kwa\-li\-gi\-re ku bi\-ro\-woo\-za ebya\-lee\-tei\-bwe omu\-woo\-jo.}
\fmtExampleEN{The doctors said that the main cause of his death was stress due loneliness.}

\fmtLemma{ki\-ro\-woo\-zo}
\fmtCitation{(e)ki\-ro\-woo\-zo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{thought, idea}
\fmtExampleRL{Eki\-ro\-woo\-zo kya\-nge kya\-wu\-kai\-ne oku kya\-mwe.}
\fmtExampleEN{My idea differs from his.}

\fmtLemma{ki\-ru\-ke}
\fmtCitation{(e)ki\-ru\-ke}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{woven product}
\fmtExampleRL{Ekii\-bbo ki\-bba ki\-ru\-ke.}
\fmtExampleEN{A basket is a woven product.}

\fmtLemma{ki\-ru\-mba}
\fmtCitation{(e)ki\-ru\-mba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wasp}
\fmtExampleRL{Eki\-ru\-mba ki\-nne\-ne\-re.}
\fmtExampleEN{I have been stung by a wasp.}

\fmtLemma{ki\-ru\-mbu}
\fmtCitation{(e)ki\-ru\-mbu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{(of an animal)}
\fmtExampleRL{Onte ba\-mu\-ku\-bbi\-re enki\-nzo oku ki\-ru\-mbu.}
\fmtExampleEN{The cow was given an injection in the thigh.}

\fmtLemma{ki\-ru\-me}
\fmtCitation{(e)ki\-ru\-me}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{crop}
\fmtExampleRL{Ebi\-ru\-me byo\-me\-re olwa ei\-sa\-na eri\-ngi.}
\fmtExampleEN{The crops have dried out because of too much sun.}

\fmtLemma{ki\-ru\-mu\-ko}
\fmtCitation{(e)ki\-ru\-mu\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{cupping horn (used for treatment), instrument or vessel used for cupping or bleeding}
\fmtExampleRL{Aba\-su\-mu ba\-ko\-le\-sya ebi\-ru\-mu\-ko oku\-ja\-nja\-ba aba\-ntu.}
\fmtExampleEN{Traditional healers use cupping horns to treat people.}

\fmtLemma{ki\-ru\-myo}
\fmtCrossreference{\textnormal{see also} kirimyo}
\fmtDialect{Ruruuli}
\fmtCitation{(e)ki\-ru\-myo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{hole digger, excavator}
\fmtExampleRL{Eki\-ru\-myo ki\-sa\-ka bii\-na.}
\fmtExampleEN{We use a hole digger to dig holes.}

\fmtLemma{ki\-ru\-ndu}%
\fmtOne{1}
\fmtCitation{(e)ki\-ru\-ndu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{\textit{Antiaris toxicaria} (a tree in the mulberry and fig family)}
\fmtExampleRL{Eki\-ru\-ndu ki\-zwa\-mu embaa\-wo zi\-na\-ki.}
\fmtExampleEN{\textit{Antiaris toxicaria} produces soft timber.}

\fmtLemma{ki\-ru\-ndu}%
\fmtOne{2}
\fmtCitation{(e)ki\-ru\-ndu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{troop (of monkeys), group (of monkeys)}
\fmtExampleRL{Ani wa\-su\-la\-wo eki\-ru\-ndu kya nke\-mbo.}
\fmtExampleEN{A troop of monkeys sleeps here.}

\fmtLemma{ki\-ru\-ngo}
\fmtCitation{(e)ki\-ru\-ngo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{spice, flavour, flavouring, seasoning}
\fmtExampleRL{Taa\-mu ebi\-ru\-ngo omu mu\-ku\-bi.}
\fmtExampleEN{Add spices to the sauce please.}

\fmtLemma{ki\-ruu\-bii\-rwa}
\fmtCitation{(e)ki\-ruu\-bii\-rwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{aim, goal, objective}
\fmtExampleRL{Nko\-be\-ra eki\-ruu\-bii\-rwa kya\-mu nso\-bo\-le oku\-ku\-wa\-bu\-la oku\-sai.}
\fmtExampleEN{Tell me your goal so that I can guide you properly.}

\fmtLemma{ki\-ruu\-li}
\fmtCitation{(e)ki\-ruu\-li}
\fmtPOS{adv}%
\fmtNC{7}
\fmtSense{the Baruuli way, the Baruuli style}
\fmtExampleRL{Aba\-bi\-ni ba\-ni ba\-ku\-bi\-na ki\-ruu\-li.}
\fmtExampleEN{These dancers are dancing the Baruuli rhythm.}

\fmtLemma{ki\-ruu\-suu\-ra}
\fmtPOS{adv}
\fmtSense{militarily, in a military manner}
\fmtExampleRL{OKa\-gwa ata\-mbu\-la ki\-ruu\-suu\-ra.}
\fmtExampleEN{Kagwa walks in a military manner.}

\fmtLemma{ki\-ru\-yi}
\fmtCitation{(e)ki\-ru\-yi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wrath, fury, anger, rancor, grudge}
\fmtExampleRL{Omu\-sai\-za odi ata n'e\-ki\-ru\-yi ki\-ngi ku mwa\-na omu\-to.}
\fmtExampleEN{That man is very angry with the child.}

\fmtLemma{ki\-ru\-yi\-ru\-yi}
\fmtCrossreference{\textnormal{see also} kirwirwi}
\fmtCitation{(e)ki\-ru\-yi\-ru\-yi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{urge, desi\-re}
\fmtExampleRL{Amo\-s ata n’e\-ki\-ru\-yi\-ru\-yi kya ku\-lwa\-na.}
\fmtExampleEN{Amos has an urge to fight.}

\fmtLemma{ki\-rwa}
\fmtCitation{(e)ki\-rwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{common water hyacinth (\textit{Eichhornia crassipes})}
\fmtExampleRL{Eki\-rwa ki\-bba ku mai\-zi.}
\fmtExampleEN{Water hyacinth is found on water.}

\fmtLemma{ki\-rwai\-re}
\fmtCitation{(e)ki\-rwai\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wound, injury}
\fmtExampleRL{Mu\-twa\-le ewa mu\-sa\-wu ba\-mu\-ta\-mbe eki\-rwai\-re oku ku\-gu\-lu.}
\fmtExampleEN{Take him to the doctor so that the wound on his leg can be treated.}

\fmtLemma{ki\-rwa\-ni}
\fmtCitation{(e)ki\-rwa\-ni}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{weapon}
\fmtExampleRL{Ata\-mbu\-la n’e\-ki\-rwa\-ni bu\-li bwi\-re.}
\fmtExampleEN{He walks with a weapon every day.}

\fmtLemma{ki\-rwa\-yi\-re}
\fmtCrossreference{\textnormal{see} kirwai\-re}

\fmtLemma{ki\-rwi\-rwi}
\fmtCrossreference{\textnormal{see also} kiruyiruyi}
\fmtCitation{(e)ki\-rwi\-rwi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{urge, desi\-re}
\fmtExampleRL{Ata n’e\-ki\-rwi\-rwi kya ku\-lwa\-na.}
\fmtExampleEN{He has an urge to fight.}

\fmtLemma{Ki\-rya\-baa\-na}
\fmtCitation{(o)Ki\-rya\-baa\-na}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{name given to a child born when all his or her elder siblings are dead}
\fmtExampleRL{Omwa\-la oyo aye\-te\-bwa Ki\-rya\-baa\-na ku\-bba aba\-ga\-nda boo\-na baa\-fi\-re.}
\fmtExampleEN{That girl is called Kiryabaana because all her siblings are dead.}

\fmtLemma{Ki\-rya\-bwi\-re}
\fmtCitation{(o)Ki\-rya\-bwi\-re}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{proverbial name given to someone born at night (lit.\,What eats at night trusts its eyes.)}

\fmtLemma{ki\-rya\-ma}
\fmtCitation{(e)ki\-rya\-ma}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{Congo signal or Congo grass, a forage crop (\textit{Brachiaria ruziziensis})}
\fmtExampleRL{Ebi\-ra\-mya bi\-zu\-bu oku\-lu\-ma.}
\fmtExampleEN{Congo signal is difficult to clear.}

\fmtLemma{ki\-rya\-nyo\-nyi}
\fmtCitation{(e)ki\-rya\-nyo\-nyi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fruit of a tree of the genus \textit{Ficus}}

\fmtLemma{ki\-sa}%
\fmtOne{1}
\fmtCitation{(e)ki\-sa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{kindness, compassion, sympathy, grace, favour, mercy}
\fmtExampleRL{Eki\-sa kya Ka\-nca ki\-ngi.}
\fmtExampleEN{God's kindness is paramount.}

\fmtLemma{ki\-sa}%
\fmtOne{2}
\fmtCitation{(e)ki\-sa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{labour pain}
\fmtExampleRL{Omu\-ka\-li aku\-lu\-mwa ebi\-sa bya ku\-bya\-la.}
\fmtExampleEN{The woman is feeling labour pains.}

\fmtLemma{ki\-sa}%
\fmtOne{3}
\fmtCitation{(o)ki\-sa}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{white cow}
\fmtExampleRL{Onte wa\-nge oki\-sa afii\-re.}
\fmtExampleEN{My white cow has died.}

\fmtLemma{ki\-saa\-bo}%
\fmtOne{1}
\fmtCitation{(e)ki\-saa\-bo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{calabash}
\fmtExampleRL{Ama\-tai ga\-cu\-ndi\-re mu ki\-saa\-bo.}
\fmtExampleEN{Process the milk in the calabash by shaking it.}

\fmtLemma{ki\-saa\-bo}%
\fmtOne{2}
\fmtCitation{(e)ki\-saa\-bo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{mud}
\fmtExampleRL{Bai\-ra\-ba mu\-ta\-ye\-ma mu bi\-saa\-bo.}
\fmtExampleEN{Don’t step into the mud, please.}

\fmtLemma{ki\-saa\-gaa\-te}
\fmtCitation{(e)ki\-saa\-gaa\-te}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fence}
\fmtExampleRL{Abbwe\-re eki\-saa\-gaa\-te oku ka\-ni\-sa.}
\fmtExampleEN{He built a fence around the church.}

\fmtLemma{ki\-saa\-le}
\fmtCitation{(e)ki\-saa\-le}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{log, stick, branch}
\fmtExampleRL{Lee\-ta eki\-saa\-le tu\-ku\-bbe onjo\-ka.}
\fmtExampleEN{Bring a stick and we will beat the snake.}

\fmtLemma{ki\-saa\-nii\-ko}
\fmtCitation{(e)ki\-saa\-nii\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{garbage, trash, rubbish, waste}
\fmtExampleRL{Ebi\-saa\-nii\-ko bi\-ngi oku lu\-bu\-ga.}
\fmtExampleEN{There is a lot of garbage around the courtyard.}

\fmtLemma{ki\-saa\-rwe\-wo}
\fmtCitation{(e)ki\-saa\-rwe\-wo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{decision, recommendation, verdict}
\fmtExampleRL{Ki\-ni ni\-kyo kya\-saa\-rwe\-wo ei\-go\-lo.}
\fmtExampleEN{This is the decision we arrived at yesterday.}

\fmtLemma{ki\-saa\-to}
\fmtCitation{(e)ki\-saa\-to}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{skin}
\fmtExampleRL{Onte ba\-mu\-baa\-gi\-re\-ku eki\-saa\-to.}
\fmtExampleEN{The cow's skin has been removed.}

\fmtLemma{ki\-saa\-we}
\fmtCitation{(e)ki\-saa\-we}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{field}
\fmtExampleRL{Tu\-ku\-si\-mba pa\-si\-pa\-la\-mu mu ki\-saa\-we.}
\fmtExampleEN{We are planting paspalum on the field.}
\fmtSubentry(b)
\fmtSense{playground, court, pitch, sports field, playing field}
\fmtExampleRL{Aba\-sa\-mbi ba mu\-pii\-ra bai\-ngii\-re omu ki\-saa\-we.}
\fmtExampleEN{The football players have entered the sports field.}
\fmtSubentry(c)
\fmtSense{airport, airfield}
\fmtExampleRL{Eki\-saa\-we kya nde\-ge ki\-ri Nte\-be.}
\fmtExampleEN{The airport is at Ntebe.}

\fmtLemma{ki\-sa\-ga\-zi}
\fmtCitation{(e)ki\-sa\-ga\-zi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{elephant grass (\textit{Pennisetum purpureum})}
\fmtExampleRL{Oku isa\-na ente zi\-rya bi\-sa\-ga\-zi.}
\fmtExampleEN{During a drought cows eat elephant grass.}

\fmtLemma{ki\-sa\-go}
\fmtCitation{(e)ki\-sa\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{injury, wound}
\fmtExampleRL{Ya\-fu\-ni\-re eki\-sa\-go oku mu\-bi\-ri.}
\fmtExampleEN{He sustained an injury on the body.}

\fmtLemma{ki\-sa\-ka}
\fmtCitation{(e)ki\-sa\-ka}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{thicket, bush}
\fmtExampleRL{Onga\-bi ai\-ngii\-re mu ki\-sa\-ka.}
\fmtExampleEN{The bushbuck entered the thicket.}

\fmtLemma{ki\-sa\-ki}
\fmtCitation{(e)ki\-sa\-ki}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{cave}
\fmtExampleRL{Abai\-rai ba\-su\-la\-nga mu bi\-sa\-ki.}
\fmtExampleEN{Ancient people used to sleep in caves.}
\fmtSubentry(b)
\fmtSense{water shade}
\fmtExampleRL{Oku ki\-sa\-ki tu\-le\-mbe\-ke\-ra\-wo amai\-zi.}
\fmtExampleEN{We tap water from the water shade.}

\fmtLemma{ki\-sa\-ki\-ro}
\fmtCitation{(e)ki\-sa\-ki\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{crop (a pouch in a bird's gullet where food is prepared for digestion)}
\fmtExampleRL{Ki\-ni ni\-kyo ki\-sa\-ki\-ro kya nko\-ko.}
\fmtExampleEN{This is the crop of the hen.}

\fmtLemma{ki\-sa\-mba}
\fmtCitation{(e)ki\-sa\-mba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{island}
\fmtExampleRL{Abai\-bbi ba\-ta\-mbu\-lii\-re mu jwe\-nge ne ba\-sa\-la okwi\-ra oku ki\-sa\-mba eki\-ndi.}
\fmtExampleEN{The thieves used a canoe and crossed to another island.}
\fmtSubentry(b)
\fmtSense{sudd (floating vegetation)}
\fmtExampleRL{Eki\-sa\-mba ki\-bba ku mai\-zi.}
\fmtExampleEN{Sudd is found on water.}

\fmtLemma{ki\-sa\-mba\-ga\-nyi}
\fmtCitation{(e)ki\-sa\-mba\-ga\-nyi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{mat made from banana fibre}
\fmtExampleRL{Asu\-la ku ki\-sa\-mba\-ga\-nyi.}
\fmtExampleEN{He sleeps on a mat made from banana fibre.}

\fmtLemma{ki\-sa\-mbai\-si\-ko}
\fmtCitation{(e)ki\-sa\-mbai\-si\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{down payment for a traditional treatment}
\fmtExampleRL{Mpee\-rya eki\-sa\-mbai\-si\-ko nku\-swa\-ga\-ni\-re omu\-ba\-zi.}
\fmtExampleEN{Make a down payment and I'll get herbal medicine for you.}

\fmtLemma{ki\-sa\-mbai\-zi}
\fmtCitation{(e)ki\-sa\-mbai\-zi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{wetland, wetlands}
\fmtExampleRL{Aba\-ntu ba\-sii\-si\-re ebi\-sa\-mbai\-zi.}
\fmtExampleEN{People have destroyed the wetlands.}
\fmtSubentry(b)
\fmtSense{brewing vat}
\fmtExampleRL{Omwe\-nge ba\-gu\-yi\-si\-rya mu ki\-sa\-mbai\-zi.}
\fmtExampleEN{The local beer is brewed in a brewing vat.}

\fmtLemma{ki\-sa\-mba\-zi}
\fmtCitation{(e)ki\-sa\-mba\-zi}
\fmtPOS{n}%
\fmtNC{7}
\fmtSubentry(c)
\fmtSense{diluted alcohol, watered down alcohol}

\fmtLemma{ki\-sa\-mbi}
\fmtCitation{(e)ki\-sa\-mbi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{thigh (\eg of a person, goat, cow)}
\fmtExampleRL{Aguu\-re ki\-sa\-mbi kya nte.}
\fmtExampleEN{He has bought a thigh part of a cow.}

\fmtLemma{ki\-sa\-mbi\-ri\-zi}
\fmtCitation{(e)ki\-sa\-mbi\-ri\-zi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bicycle peddle}
\fmtExampleRL{Eki\-sa\-mbi\-ri\-zi ki\-zwi\-re\-ku.}
\fmtExampleEN{The bicycle peddle has broken off.}

\fmtLemma{ki\-sa\-mbo}
\fmtCitation{(e)ki\-sa\-mbo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{overgrown patch of weeds, area under stubble}
\fmtExampleRL{Bu\-ni bwi\-re bwa\-ku\-lo\-ngoo\-sa ki\-sa\-mbo.}
\fmtExampleEN{It is time to clear the overgrown patch of weeds.}

\fmtLemma{Ki\-sa\-mbu}%
\fmtOne{1}
\fmtCitation{(e)Ki\-sa\-mbu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{home or area of people who belong to the Heron clan}
\fmtExampleRL{Aba\-sa\-mbu bai\-ka\-la Ki\-sa\-mbu.}
\fmtExampleEN{The members of the Heron clan reside in the place called Ekisambo.}

\fmtLemma{ki\-sa\-mbu}%
\fmtOne{2}
\fmtCrossreference{\textnormal{see} kisambo}

\fmtLemma{ki\-sa\-mpa}
\fmtCitation{(e)ki\-sa\-mpa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{mat woven from banana fibres}
\fmtExampleRL{Nkoo\-ra\-yo eki\-sa\-mpa.}
\fmtExampleEN{Make me a mat from banana fibres.}

\fmtLemma{ki\-sa\-na}
\fmtCitation{(e)ki\-sa\-na}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{brilliant yellow sun at dawn or dusk}
\fmtExampleRL{Eki\-sa\-na ki\-zwi\-re\-yo.}
\fmtExampleEN{The brilliant yellow sun has come out.}
\fmtSubentry(b)
\fmtSense{neonatal jaundice}
\fmtExampleRL{Oyo omwa\-na mu\-lwai\-re wa ki\-sa\-na.}
\fmtExampleEN{That child is suffering from neonatal jaundice.}

\fmtLemma{ki\-sa\-na\-ni}
\fmtCitation{(e)ki\-sa\-na\-ni}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{photo, picture}
\fmtExampleRL{Omwa\-la ase\-mei\-re omu ki\-sa\-na\-ni.}
\fmtExampleEN{The girl looks beautiful in the photo.}

\fmtLemma{ki\-sa\-nga}
\fmtCitation{(e)ki\-sa\-nga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{device used for moulding}
\fmtExampleRL{Ngu\-li\-sya\-yo eki\-sa\-nga.}
\fmtExampleEN{Sell me a moulding device.}

\fmtLemma{ki\-sa\-nsa}%
\fmtOne{1}
\fmtCitation{(e)ki\-sa\-nsa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wild onion}
\fmtExampleRL{Ebi\-sa\-nsa bi\-woo\-ma mu\-no.}
\fmtExampleEN{Wild onions are delicious.}

\fmtLemma{ki\-sa\-nsa}%
\fmtOne{2}
\fmtCitation{(e)ki\-sa\-nsa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{ileum (the third portion of the small intestine)}
\fmtExampleRL{Ebi\-sa\-nsa bya onte bi\-woo\-ma oku\-lya.}
\fmtExampleEN{Cows' ileums are good for eating.}

\fmtLemma{ki\-sa\-nsa}%
\fmtOne{3}
\fmtCitation{(e)ki\-sa\-nsa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{vaginal yeast infection}
\fmtExampleRL{Ebyo\-loo\-ni bya iku\-mbe ni\-byo bi\-sii\-ga endwai\-re y’a\-ki\-sa\-nsa.}
\fmtExampleEN{Public toilets are the main cause of vaginal yeast infections.}

\fmtLemma{ki\-sa\-nsa\-ga\-la}
\fmtCitation{(e)ki\-sa\-nsa\-ga\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{husk (of simsim (UgE)/sesame)}
\fmtExampleRL{Ebi\-sa\-nsa\-ga\-la bya ma\-ca\-nde bi\-gi\-mu\-sa ei\-ta\-ka\-li.}
\fmtExampleEN{Simsim husks fertilise the soil.}

\fmtLemma{ki\-sa\-nzi\-re}
\fmtCitation{(e)ki\-sa\-nzi\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{joke, trick}
\fmtExampleRL{Bye nku\-ku\-ko\-be\-ra ma\-zi\-ma ti bi\-sa\-nzi\-re.}
\fmtExampleEN{I am telling you the truth, it is not a joke.}

\fmtLemma{ki\-sa\-ra\-li}
\fmtCitation{(e)ki\-sa\-ra\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{overgrown tasteless food crop (mainly tubers)}
\fmtExampleRL{Eki\-yaa\-ta ky'o\-mpai\-re ki\-sa\-ra\-li.}
\fmtExampleEN{The cassava you have given me is overgrown.}

\fmtLemma{ki\-sa\-ra\-sa\-mbwa}
\fmtCitation{(e)ki\-sa\-ra\-sa\-mbwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{maize chaff}
\fmtExampleRL{Embu\-li zi\-ku\-lya bi\-sa\-ra\-sa\-mbwa.}
\fmtExampleEN{The goats are eating maize chaff.}

\fmtLemma{ki\-sa\-sa\-la}
\fmtCitation{(e)ki\-sa\-sa\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{honeycomb}
\fmtExampleRL{Ebi\-sa\-sa\-la bi\-bbaa\-mu ojuu\-ru.}
\fmtExampleEN{Honeycombs have concentrated honey.}

\fmtLemma{ki\-sa\-si}
\fmtCitation{(e)ki\-sa\-si}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{rain shed, roof overhang}
\fmtExampleRL{Ikaa\-ra\-ku omu ki\-sa\-si.}
\fmtExampleEN{Sit a while under the rain shed.}

\fmtLemma{ki\-sa\-sii\-ro}
\fmtCitation{(e)ki\-sa\-sii\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{laying nest}
\fmtExampleRL{Koo\-ra onko\-ko eki\-sa\-sii\-ro.}
\fmtExampleEN{Make a laying nest for the chicken.}

\fmtLemma{ki\-sa\-si\-ro}
\fmtCitation{(e)ki\-sa\-si\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{litter, trash, rubbish}
\fmtExampleRL{Ebi\-sa\-si\-ro bi\-ngi mu\-no omu Ka\-mpa\-la.}
\fmtExampleEN{There is a lot of litter in Kampala.}

\fmtLemma{ki\-sa\-ya}
\fmtCitation{(e)ki\-sa\-ya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{outcry, yell, cry, groan, wail, alarm}
\fmtExampleRL{Mbai\-re nkwa\-ba ku\-ta\-ndi\-ka ku\-wa\-ndii\-ka, omwe\-ge\-sya n’a\-nku\-bba eki\-sa\-ya.}
\fmtExampleEN{I was about to start writing when the teacher frightened me with a yell.}

\fmtLemma{ki\-see\-ra}%
\fmtOne{1}
\fmtCitation{(e)ki\-see\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{time, period}
\fmtExampleRL{Ebi\-see\-ra bya kwa\-ba e Ge\-rma\-ny bii\-ki\-re.}
\fmtExampleEN{The time to go to Germany has come.}

\fmtLemma{ki\-see\-ra}%
\fmtOne{2}
\fmtCitation{(e)ki\-see\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{wrinkle, line}
\fmtExampleRL{Omu\-ka\-li ata n’e\-bi\-see\-ra omu iko\-ti.}
\fmtExampleEN{The woman has lines on her neck.}
\fmtSubentry(b)
\fmtSense{palm lines, palmar creases}
\fmtExampleRL{Aba\-la\-gu\-li ba\-ko\-le\-sya ebi\-see\-ra bya mu ki\-ga\-nja oku\-ma\-nya omu\-ki\-sa gwa mu\-ntu.}
\fmtExampleEN{Fortune tellers use one's palm lines to foretell one's luck.}

\fmtLemma{ki\-see\-re}
\fmtCitation{(e)ki\-see\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{knife without a handle, blade of a knife}
\fmtExampleRL{Nde\-te\-ra ani eki\-see\-re.}
\fmtExampleEN{Bring me the knife without a handle.}

\fmtLemma{ki\-sei\-ze\-re}
\fmtCitation{(e)ki\-sei\-ze\-re}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{mound populated by big flying termites}
\fmtExampleRL{Eki\-swa eki\-se\-ze\-re ba\-ki\-sa\-ki\-re\-wo.}
\fmtExampleEN{The termite mound has been dug up.}

\fmtLemma{ki\-se\-ke\-sa}
\fmtCitation{(e)ki\-se\-ke\-sa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{type of furry caterpillar (It is poisonous and sometimes has red bristles at ends.)}
\fmtExampleRL{Eki\-se\-ke\-sa ki\-nku\-bbi\-re obu\-lu\-ngu.}
\fmtExampleEN{A hairy poisonous caterpillar has irritated me with its short stinging hair.}

\fmtLemma{ki\-se\-ke\-se\-ke}
\fmtCitation{(e)ki\-se\-ke\-se\-ke}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wrist}
\fmtExampleRL{Omu\-ko\-no gwe\-we\-te\-ra mu ki\-se\-ke\-se\-ke.}
\fmtExampleEN{The arm bends at the wrist.}

\fmtLemma{ki\-se\-mbo}
\fmtCitation{(e)ki\-se\-mbo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{gift, present}
\fmtExampleRL{Ya\-mu\-to\-nei\-re onte nga eki\-se\-mbo ky’a\-ma\-bya\-li\-bwa.}
\fmtExampleEN{He gave her a cow as a birthday gift.}
\fmtSubentry(b)
\fmtSense{reward, award, remuneration, reimbursement}
\fmtExampleRL{Ba\-mu\-wai\-re eki\-se\-mbo kya ku\-kee\-ra\-nga.}
\fmtExampleEN{He was given an award for coming early.}

\fmtLemma{ki\-se\-mee\-ri}
\fmtCitation{(e)ki\-se\-mee\-ri}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fresh local beer}
\fmtExampleRL{Abaa\-na n’a\-ba\-ka\-li be\-ndya mu\-no ebi\-se\-mee\-ri.}
\fmtExampleEN{Children and women love drinking fresh local beer.}

\fmtLemma{ki\-se\-nge}
\fmtCitation{(e)ki\-se\-nge}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{wall}
\fmtExampleRL{Eki\-se\-nge kya nyu\-mba yaa\-mwe ki\-gu\-mu.}
\fmtExampleEN{The wall of his house is strong.}
\fmtSubentry(b)
\fmtSense{bedroom}
\fmtExampleRL{Ya\-ba omu ki\-se\-nge ondee\-te\-re eBbai\-bbu\-li ya\-nge.}
\fmtExampleEN{Go to the bedroom and bring my Bible.}

\fmtLemma{ki\-se\-ngee\-rya}
\fmtCitation{(e)ki\-se\-ngee\-rya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wagon of a trailer}
\fmtExampleRL{Eki\-se\-ngee\-rya kya mo\-to\-ka kye\-nde\-ke\-re\-ku.}
\fmtExampleEN{The wagon of the trailer was disconnected.}

\fmtLemma{ki\-se\-nge\-re}
\fmtCitation{(e)ki\-se\-nge\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{slag, waste matter separated from metals during the smelting, scoria}

\fmtLemma{ki\-si\-bo}
\fmtCitation{(e)ki\-si\-bo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{flock (of goats)}
\fmtExampleRL{Enzo\-ki za\-ne\-ne\-re eki\-si\-bo kya mbu\-li.}
\fmtExampleEN{Bees stung a whole flock of goats.}
\fmtSubentry(b)
\fmtSense{goat shed}
\fmtExampleRL{Omu\-ta\-mii\-re ya\-suu\-re mu ki\-si\-bo.}
\fmtExampleEN{The drunkard slept in the goat shed.}

\fmtLemma{ki\-si\-gaa\-ro}
\fmtCrossreference{\textnormal{see} kisigaarwa}

\fmtLemma{ki\-si\-gaa\-rwa}
\fmtCitation{(e)ki\-si\-gaa\-rwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{residue, remains, leftover part}
\fmtExampleRL{Ebi\-si\-gaa\-rwo bya nyu\-mba eyii\-re bi\-bba bi\-na\-mbu\-la mu\-ga\-so.}
\fmtExampleEN{The remains of a burnt house are useless.}

\fmtLemma{ki\-si\-ge}
\fmtCitation{(e)ki\-si\-ge}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{eyebrow}
\fmtExampleRL{Aba\-ka\-li ba\-sii\-ga eka\-laa\-mu oku bi\-si\-ge.}
\fmtExampleEN{Women color their eyebrows.}

\fmtLemma{ki\-sii\-bo}
\fmtCitation{(e)ki\-sii\-bo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{Lent}
\fmtExampleRL{Aba\-ka\-tu\-li\-ki bee\-bo\-ne\-re\-za omu ki\-sii\-bo.}
\fmtExampleEN{The Catholics subject themselves to self-denial during Lent.}
\fmtSubentry(b)
\fmtSense{Ramadan (month of fasting in the Muslim year)}
\fmtExampleRL{Aba\-si\-raa\-mu ba\-kwa\-ba ku\-ta\-ndi\-ka ki\-sii\-bo.}
\fmtExampleEN{Moslems are about to start the month of Ramadan.}

\fmtLemma{ki\-sii\-ka}
\fmtCitation{(e)ki\-sii\-ka}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{wall}
\fmtExampleRL{Ya\-ba ogu\-buu\-re ebi\-sii\-ka bya nyu\-mba.}
\fmtExampleEN{Go and make the walls of the house smooth.}
\fmtSubentry(b)
\fmtSense{room}
\fmtExampleRL{Enyu\-mba yaa\-mu eri\-mu ebi\-sii\-ka bi\-ngai?}
\fmtExampleEN{How many rooms does your house have?}
\fmtSubentry(c)
\fmtSense{bedroom}
\fmtExampleRL{Ya\-ba obba\-sye omu ki\-sii\-ka.}
\fmtExampleEN{Go and sleep in the bedroom.}
\fmtSubentry(d)
\fmtSense{honeymoon}
\fmtExampleRL{Aba\-go\-le ba\-kya\-li mu ki\-sii\-ka.}
\fmtExampleEN{The bride and bridegroom are still on their honeymoon.}

\fmtLemma{ki\-sii\-mi\-syo}
\fmtCitation{(e)ki\-sii\-mi\-syo}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{incentive, motivation, motive, reason, stimulus, impetus, encouragement, impulse}
\fmtExampleRL{Aba\-so\-me\-sya ba Ma\-ke\-re\-re ba\-ba\-yi\-mi\-re aka\-si\-mi\-syo kaa\-bwe.}
\fmtExampleEN{Makerere lecturers were denied their incentives.}

\fmtLemma{ki\-sii\-mo}
\fmtCitation{(e)ki\-sii\-mo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{gift}
\fmtExampleRL{Aba\-ntu ba\-ta\-ndi\-ki\-re oku\-gu\-la ebi\-sii\-mo bya Se\-ku\-ku\-lu.}
\fmtExampleEN{People have started buying Christmas gifts.}

\fmtLemma{ki\-sii\-rii\-rwa}
\fmtCitation{(e)ki\-sii\-rii\-rwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{burnt food, burnt residue}
\fmtExampleRL{Ki\-ni ki\-sii\-rii\-rwa kya ki\-duu\-ma.}
\fmtExampleEN{This is the residue of burnt maize.}

\fmtLemma{ki\-sii\-sa}
\fmtCitation{(e)ki\-sii\-sa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{insect, bug}
\fmtExampleRL{Eki\-sii\-sa ki\-mu\-ne\-ne\-re erii\-so.}
\fmtExampleEN{An insect has bitten her on the eye.}

\fmtLemma{ki\-sii\-so}
\fmtCitation{(e)ki\-sii\-so}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sin}
\fmtExampleRL{Oyo ata n’e\-bi\-sii\-so bi\-ngi.}
\fmtExampleEN{He commits many sins.}

\fmtLemma{Ki\-sii\-ta}
\fmtCitation{(e)Ki\-sii\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{residence of the people of the Sheep clan}
\fmtExampleRL{Aba\-sii\-ta bai\-ka\-la Ki\-sii\-ta.}
\fmtExampleEN{The members of the Sheep clan stay at Kisiita.}

\fmtLemma{ki\-sii\-yi}
\fmtCitation{(e)ki\-sii\-yi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{eczema}
\fmtExampleRL{Eki\-sii\-yi bu\-lwai\-re bwa lu\-su\-su obu\-bbi.}
\fmtExampleEN{Eczema is a terrible skin ailment.}

\fmtLemma{ki\-si\-ki}%
\fmtOne{1}
\fmtCitation{(e)ki\-si\-ki}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{log}
\fmtExampleRL{Ba\-na\-gi\-re ebi\-si\-ki omu lu\-bu\-ga lwa\-nge.}
\fmtExampleEN{They threw logs into my compound.}

\fmtLemma{ki\-si\-ki}%
\fmtOne{2}
\fmtCitation{(o)ki\-si\-ki}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{frigid person, person lacking the desire for sexual activity}
\fmtExampleRL{Omu\-ka\-li oyo ki\-si\-ki.}
\fmtExampleEN{That woman is frigid.}

\fmtLemma{ki\-si\-ki\-ri}
\fmtCitation{(e)ki\-si\-ki\-ri}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{respectable thing, reputable thing}
\fmtExampleRL{Oyo ki\-si\-ki\-ri kya mu\-ka\-li.}
\fmtExampleEN{She is a respectable woman.}

\fmtLemma{ki\-si\-ko}
\fmtCitation{(e)ki\-si\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bush}
\fmtExampleRL{Tu\-kwa\-ba ku\-lu\-ma ki\-si\-ko.}
\fmtExampleEN{We are going to clear the bush.}

\fmtLemma{ki\-si\-nda\-bu\-li\-syo}
\fmtCrossreference{\textnormal{see} kisindabulyo}

\fmtLemma{ki\-si\-nda\-bu\-lyo}%
\fmtOne{1}
\fmtCitation{(e)ki\-si\-nda\-bu\-lyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{key}
\fmtExampleRL{Ti\-tu\-ku\-so\-bo\-la ku\-ki\-ngu\-la lwi\-ge lu\-ni awa\-ta\-li ki\-si\-ndya\-bu\-lyo.}
\fmtExampleEN{We cannot open this door without the key.}
\fmtSubentry(b)
\fmtSense{opener (\eg a bottle opener)}

\fmtLemma{ki\-si\-nda\-bu\-zo}
\fmtCrossreference{\textnormal{see} kisindabulyo}

\fmtLemma{ki\-si\-nde}
\fmtCitation{(e)ki\-si\-nde}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{path, track (of an animal)}
\fmtExampleRL{Ekyo eki\-si\-nde kya nte.}
\fmtExampleEN{That is a cow path.}

\fmtLemma{ki\-si\-nga}
\fmtCitation{(e)ki\-si\-nga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{copper ore, copper}
\fmtExampleRL{Omu Uga\-nda ti\-mu\-kya\-lu\-mu bi\-si\-ngo.}
\fmtExampleEN{Uganda no longer has copper.}
\fmtSubentry(b)
\fmtSense{necklace}
\fmtExampleRL{Omwa\-la azwai\-re eki\-si\-nga.}
\fmtExampleEN{The girl is wearing a necklace.}

\fmtLemma{ki\-si\-ngo}
\fmtCitation{(e)ki\-si\-ngo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{evil spirit}
\fmtExampleRL{Omwe\-gi ali\-ku eki\-si\-ngo.}
\fmtExampleEN{The student is possessed by an evil spirit.}
\fmtSubentry(b)
\fmtSense{decorated short stick used by traditional priests}
\fmtExampleRL{Aba\-ki\-ba\-ndwa ba\-kwa\-ta ebi\-si\-ngo nga ba\-ku\-ra\-mya.}
\fmtExampleEN{Traditional priests hold a decorated short stick while worshipping.}

\fmtLemma{ki\-si\-ni\-ni}%
\fmtOne{1}
\fmtCitation{(e)ki\-si\-ni\-ni}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{ember, glowing piece of burning wood}
\fmtExampleRL{Ombwe\-ne ba\-mu\-ku\-bbi\-re eki\-si\-ni\-ni.}
\fmtExampleEN{The dog was hit with a glowing piece of burning wood.}

\fmtLemma{ki\-si\-ni\-ni}%
\fmtOne{2}
\fmtCitation{(e)ki\-si\-ni\-ni}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{grieving period, mourning period}
\fmtExampleRL{Eki\-si\-ni\-ni kya\-li\-wa di?}
\fmtExampleEN{When will that mourning period end?}

\fmtLemma{ki\-si\-nja}
\fmtCitation{(e)ki\-si\-nja}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{barbel (\textit{Labeobarbus bynni})}
\fmtExampleRL{Eki\-si\-nja ki\-no\-la mu\-no.}
\fmtExampleEN{The barbel is very fatty.}

\fmtLemma{ki\-si\-pi}
\fmtCitation{(e)ki\-si\-pi}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{herpes zoster, shingles}
\fmtExampleRL{Eki\-si\-pi kya\-mu\-ku\-bbi\-re omu bu\-syo.}
\fmtExampleEN{He is suffering from facial shingles.}

\fmtLemma{ki\-si\-ra}
\fmtCitation{(e)ki\-si\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{basket}
\fmtExampleRL{Eki\-si\-ra kii\-zwi\-re ena\-na\-nsi.}
\fmtExampleEN{The basket is full of pineapples.}

\fmtLemma{ki\-si\-raa\-ni}
\fmtCitation{(e)ki\-si\-raa\-ni}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bad luck, misfortune, disaster}
\fmtExampleRL{Eki\-si\-raa\-ni ki\-zwa ku ki\-ki?}
\fmtExampleEN{What causes a misfortune?}

\fmtLemma{ki\-si\-rii\-kii\-ri}
\fmtPOS{adv}
\fmtSense{quietly, silently}
\fmtExampleRL{Ya\-ta\-mbwi\-re ki\-si\-rii\-kii\-ri ndoo\-wo ya\-mu\-boi\-ne.}
\fmtExampleEN{He walked away quietly, no one saw him.}

\fmtLemma{ki\-si\-ryo\-ngo}
\fmtCitation{(e)ki\-si\-ryo\-ngo}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{smell of burnt food}
\fmtExampleRL{Eme\-re eku\-wu\-nya eki\-si\-ryo\-ngo.}
\fmtExampleEN{The food smells burnt.}

\fmtLemma{ki\-si\-si}
\fmtCitation{(e)ki\-si\-si}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{gourd (used to hold local brew)}
\fmtExampleRL{Eki\-si\-si kya\-nge kya\-ti\-ki\-re.}
\fmtExampleEN{My gourd is broken.}

\fmtLemma{ki\-si\-si\-mu\-ko}
\fmtCitation{(e)ki\-si\-si\-mu\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wakefulness phase between sleep cycles}
\fmtExampleRL{Abai\-bbi ba\-lu\-mbii\-re mu ki\-si\-si\-mu\-ko kya ka\-sa\-tu.}
\fmtExampleEN{The thieves attacked during the third waking phase of the sleep cycle.}

\fmtLemma{ki\-so}
\fmtCitation{(e)ki\-so}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{machete, panga}
\fmtExampleRL{Omu\-si\-ri\-ka\-le ata\-mbu\-la n’e\-ki\-so.}
\fmtExampleEN{The soldier carries (lit.\,walks with) a machete.}

\fmtLemma{ki\-so\-byo}
\fmtCitation{(e)ki\-so\-byo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{wrong, crime, mistake, wrongdoing}
\fmtExampleRL{Akoo\-re ebi\-so\-byo bya maa\-ni oku\-ma\-la emya\-ka gi\-ngi.}
\fmtExampleEN{He has committed grave mistakes over the years.}
\fmtSubentry(b)
\fmtSense{foul (in sports)}

\fmtLemma{ki\-so\-do\-ka}
\fmtCitation{(e)ki\-so\-do\-ka}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{coccidiosis}
\fmtExampleRL{Enko\-ko zi\-rwai\-re eki\-so\-do\-ka.}
\fmtExampleEN{The chicken are suffering from coccidiosis.}

\fmtLemma{ki\-so\-ga\-so\-ga}
\fmtCrossreference{\textnormal{see also} nsogasoga}
\fmtCitation{(e)ki\-so\-ga\-so\-ga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{castor oil plant, castor bean (\textit{Ricinus communis})}

\fmtLemma{ki\-so\-ke}
\fmtCitation{(e)ki\-so\-ke}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{gill}
\fmtExampleRL{Ebi\-so\-ke bi\-bba mu ncwi.}
\fmtExampleEN{Gills are found in fish.}

\fmtLemma{ki\-so\-ko\-mi}
\fmtCitation{(e)ki\-so\-ko\-mi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{type of white larva of a beetle which usually lives in decomposing vegetation}
\fmtExampleRL{Ebi\-so\-ko\-mi bi\-bba mu mi\-saa\-le egi\-ju\-ndi.}
\fmtExampleEN{The larvae live in the decomposing wood of a tree.}

\fmtLemma{ki\-so\-lo}
\fmtCitation{(e)ki\-so\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{animal}
\fmtExampleRL{Eki\-so\-lo ki\-ngii\-re omu nyu\-mba.}
\fmtExampleEN{The animal entered the house.}

\fmtLemma{ki\-so\-mee\-ri}%
\fmtOne{1}
\fmtCitation{(e)ki\-so\-mee\-ri}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{evening sunshine}
\fmtExampleRL{Aba\-kai\-re ba\-ta\-ka mu\-no okwo\-ta eki\-so\-mee\-ri.}
\fmtExampleEN{Old people love to bask in the evening sunshine.}

\fmtLemma{ki\-so\-mee\-ri}%
\fmtOne{2}
\fmtCitation{(e)ki\-so\-mee\-ri}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{sweet fresh local beer}
\fmtExampleRL{Ebi\-so\-mee\-ri bi\-si\-nga ku\-nywe\-bwa ba\-ka\-li.}
\fmtExampleEN{Sweet fresh local beer is usually consumed by women.}

\fmtLemma{ki\-so\-nga}%
\fmtOne{1}
\fmtCitation{(o)ki\-so\-nga}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{cockerel}
\fmtExampleRL{Oki\-so\-nga wa\-nge ala\-mpi oku\-koo\-ka.}
\fmtExampleEN{My cockerel is about to crow.}

\fmtLemma{Ki\-so\-nga}%
\fmtOne{2}
\fmtCitation{(e)Ki\-so\-nga}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{residence of people of the Grasshopper clan}
\fmtExampleRL{Aba\-so\-nga bai\-ka\-la Ki\-so\-nga.}
\fmtExampleEN{Members of the Grasshopper clan stay in Kisonga.}

\fmtLemma{ki\-so\-nko}
\fmtCitation{(e)ki\-so\-nko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{snail shell}
\fmtExampleRL{Ebi\-so\-nko bi\-ko\-la eme\-re ya nko\-ko.}
\fmtExampleEN{Snail shells are used to make chicken feed.}

\fmtLemma{ki\-soo\-ne}
\fmtCitation{(e)ki\-soo\-ne}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{syphilis}
\fmtExampleRL{Aba\-nya\-la ba\-lwa\-la mu\-no ebi\-soo\-ne.}
\fmtExampleEN{The Banyala often suffer from syphilis.}

\fmtLemma{ki\-soo\-ro}
\fmtCitation{(e)ki\-soo\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{termite mound on which small white mushrooms grow}
\fmtExampleRL{Ebi\-soo\-ro ba\-bi\-ro\-le\-ra ku\-sai omu ki\-cwe\-ka kyai\-swe.}
\fmtExampleEN{Termite mounds on which small white mushrooms grow are well protected in the community.}

\fmtLemma{ki\-soo\-roo\-ro}
\fmtCitation{(e)ki\-soo\-roo\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{\textit{Senna hirsuta}, \textit{Senna occidentalis} or \textit{Senna ramosissimum} (type of shrub)}
\fmtExampleRL{Enzo\-ka zi\-ku\-ndu\-ma, mu\-ndee\-te\-re ebi\-koo\-la bya bi\-soo\-roo\-ro nta\-ku\-nye\-ku.}
\fmtExampleEN{I am suffering from a stomach ache, please get leaves of \textit{Senna hirsuta} so that I can chew them.}

\fmtLemma{ki\-so\-so\-nko\-ro}
\fmtCitation{(e)ki\-so\-so\-nko\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{egg shell}
\fmtExampleRL{Eki\-so\-so\-nko\-ro kya iwu\-li kya\-ti\-ka ma\-ngu.}
\fmtExampleEN{An egg shell breaks easily.}
\fmtSubentry(b)
\fmtSense{empty box, sachet, tube}
\fmtExampleRL{Mpee\-rya eki\-so\-so\-nko\-ro kya ki\-bi\-rii\-ti ekyo.}
\fmtExampleEN{Get me an empty match box.}

\fmtLemma{ki\-so\-to\-ko}
\fmtCitation{(e)ki\-so\-to\-ko}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{fowl typhoid}
\fmtExampleRL{Obu\-lwai\-re bwa ki\-so\-to\-ko bu\-kwai\-te enko\-ko zoo\-na.}
\fmtExampleEN{All the chickens are infected with fowl typhoid.}

\fmtLemma{ki\-su}
\fmtCitation{(e)ki\-su}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{nest}
\fmtExampleRL{Ekyo ki\-su kya ki\-syo.}
\fmtExampleEN{That is a weaver bird's nest.}

\fmtLemma{ki\-su\-bba}
\fmtCitation{(e)ki\-su\-bba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{chest}
\fmtExampleRL{Eki\-su\-bba ki\-ku\-ndu\-ma mu\-no.}
\fmtExampleEN{I have a serious pain in the chest.}
\fmtSubentry(b)
\fmtSense{chest pain}
\fmtExampleRL{Ba\-ku\-mu\-ja\-nja\-ba ki\-su\-bba omu irwa\-lo.}
\fmtExampleEN{He is being treated for chest pain in the hospital.}
\fmtSubentry(c)
\fmtSense{cough}
\fmtExampleRL{Ebi\-su\-bba bi\-ru\-mi\-re aba\-ntu ebi\-ro bi\-ni.}
\fmtExampleEN{People are suffering from a cough these days.}

\fmtLemma{ki\-su\-bi}%
\fmtOne{1}
\fmtCitation{(e)ki\-su\-bi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{type of banana used to make local brew}

\fmtLemma{ki\-su\-bi}%
\fmtOne{2}
\fmtCitation{(e)ki\-su\-bi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{grass}
\fmtExampleRL{Ka\-ndu\-la ebi\-su\-bi ebyo.}
\fmtExampleEN{Rake that grass.}
\fmtSubentry(b)
\fmtFixedexpression{ki\-su\-bi ku rii\-so}
\fmtSense{hurriedly (lit.\,grass in the eye), on a very short notice, unprepared}
\fmtExampleRL{Omu\-ka\-li ba\-mu\-twai\-re n’e\-ki\-su\-bi oku rii\-so.}
\fmtExampleEN{The woman was taken hurriedly.}
\fmtSubentry(c)
\fmtSense{lemon grass (\textit{Cymbopogon citratus})}
\fmtExampleRL{Ntee\-ra\-ku eki\-su\-bi oku caa\-yi.}
\fmtExampleEN{Put some lemon grass in my tea.}

\fmtLemma{ki\-su\-bii\-ro}
\fmtCitation{(e)ki\-su\-bii\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{story, tale, narrative}
\fmtExampleRL{Iza nku\-su\-bii\-re eki\-su\-bii\-ro.}
\fmtExampleEN{Come and I will tell you a story.}

\fmtLemma{Ki\-su\-kwe}
\fmtCitation{(e)Ki\-su\-kwe}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{residence of the Mushroom clan}
\fmtExampleRL{Aba\-su\-kwe bai\-ka\-la Ki\-su\-kwe.}
\fmtExampleEN{Members of the Mushroom clan live in Kisukwe.}

\fmtLemma{ki\-su\-le}
\fmtCitation{(e)ki\-su\-le}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{crest of a hen}
\fmtExampleRL{Onko\-ko oni wa ki\-su\-le.}
\fmtExampleEN{This hen has a crest.}

\fmtLemma{ki\-su\-lo}
\fmtCitation{(e)ki\-su\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{students' hall, dormitory}
\fmtExampleRL{Abee\-gi ba\-su\-la mu ki\-su\-lo.}
\fmtExampleEN{The students stay in the dormitory.}

\fmtLemma{ki\-su\-lu\-la}
\fmtCitation{(e)ki\-su\-lu\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{general name for creeping plant (\eg beans, sweet potatoes, passion fruit)}
\fmtExampleRL{Omu ndu\-mi\-ro mu\-li\-mu ebi\-su\-lu\-la.}
\fmtExampleEN{There are many creeping plants in the garden.}

\fmtLemma{ki\-su\-mu\-lu\-zo}
\fmtCitation{(e)ki\-su\-mu\-lu\-zo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{key, opener (\eg a bottle opener)}
\fmtExampleRL{Naa\-ni ya\-bii\-ki\-re eki\-su\-mu\-lu\-zo kya ki\-yu\-ngu?}
\fmtExampleEN{Who kept the key for the kitchen?}

\fmtLemma{ki\-su\-nku}
\fmtCitation{(e)ki\-su\-nku}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{granary cover}
\fmtExampleRL{Eki\-su\-nku ompu\-ngo aki\-na\-gi\-re.}
\fmtExampleEN{The cover of the granary was blown away by the wind.}

\fmtLemma{ki\-su\-nsu\-li}
\fmtCitation{(e)ki\-su\-nsu\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{(of a male bird) comb, cock crown}
\fmtExampleRL{Ompa\-ngu\-zi ata n’e\-ki\-su\-nsu\-li.}
\fmtExampleEN{The cock has a comb.}

\fmtLemma{ki\-su\-nsuu\-ro}
\fmtCitation{(e)ki\-su\-nsuu\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{ginnery}
\fmtExampleRL{Eki\-su\-nsuu\-ro kya Na\-ka\-so\-ngo\-la ti\-ki\-kya\-ko\-la.}
\fmtExampleEN{The ginnery in Nakosongola is no longer operating.}

\fmtLemma{ki\-su\-nsu\-zo}
\fmtCitation{(e)ki\-su\-nsu\-zo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{hair comb}
\fmtExampleRL{Ntii\-za\-ku eki\-su\-nsu\-zo nsu\-nsu\-le ei\-so\-ke lya\-nge.}
\fmtExampleEN{Lend me a comb and I'll do my hair.}

\fmtLemma{ki\-su\-ru\-la}
\fmtCitation{(e)ki\-su\-ru\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{creeping plant, sweet potato vine}
\fmtExampleRL{Embe\-re\-ge zi\-rya mu\-no ebi\-su\-ru\-la.}
\fmtExampleEN{Pigs like to eat sweet potato vines.}

\fmtLemma{ki\-su\-su}
\fmtCitation{(e)ki\-su\-su}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{bark (of a tree)}
\fmtExampleRL{Omu\-to\-ma ba\-gu\-toi\-re\-ku eki\-su\-su.}
\fmtExampleEN{They peeled the bark off of the bark cloth tree.}
\fmtSubentry(b)
\fmtSense{peel, skin (\eg of banana)}
\fmtExampleRL{Embu\-li zi\-rya ebi\-su\-su bya ma\-too\-ke.}
\fmtExampleEN{The goats eat banana peels.}
\fmtSubentry(c)
\fmtSense{nutshell, chaff, husk}
\fmtExampleRL{Ebi\-su\-su bya mai\-do bi\-sai.}
\fmtExampleEN{The shells of groundnuts are nice.}

\fmtLemma{ki\-su\-su\-nku}
\fmtCrossreference{\textnormal{see also} kisusu}
\fmtCitation{(e)ki\-su\-su\-nku}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{nutshell, chaff, husk}
\fmtExampleRL{Na\-ga\-yo ebi\-su\-su bya bi\-yi\-mba.}
\fmtExampleEN{Throw away the bean husks.}

\fmtLemma{ki\-su\-suu\-ryo}
\fmtCitation{(e)ki\-su\-suu\-ryo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{door mat}
\fmtExampleRL{Su\-suu\-rya ei\-ta\-ka\-li oku ki\-su\-suu\-ryo.}
\fmtExampleEN{Wipe off the soil on the door mat.}

\fmtLemma{ki\-suu\-bi\-ra\-ba\-lii\-sya}
\fmtCitation{(e)ki\-suu\-bi\-ra\-ba\-lii\-sya}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{late afternoon sunshine}
\fmtExampleRL{Ebi\-so\-lo bi\-ta\-ka mu\-no oku\-lya omu ki\-suu\-bi\-ra\-ba\-lii\-sya.}
\fmtExampleEN{Animals love eating during the late afternoon sunshine.}

\fmtLemma{ki\-suu\-mwa}
\fmtCrossreference{\textnormal{see} kisuumwasuumwa}

\fmtLemma{ki\-suu\-mwa\-suu\-mwa}
\fmtCitation{(e)ki\-suu\-mwa\-suu\-mwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{scrapings or pieces of banana pith}
\fmtExampleRL{Ebi\-suu\-mwa\-suu\-mwa ba\-bii\-ka\-ti\-sya abaa\-na aba\-to aba\-li\-na eki\-nzi.}
\fmtExampleEN{Scrapings of banana pith are used to massage kids with flu.}

\fmtLemma{ki\-suu\-ra}
\fmtCitation{(o)ki\-suu\-ra}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{goat with fur on the forehead, tuft}
\fmtExampleRL{Oki\-suu\-ra abo\-ne\-ka ku\-sai.}
\fmtExampleEN{The goat with fur on its forehead looks nice.}

\fmtLemma{ki\-su\-yi}
\fmtCitation{(e)ki\-su\-yi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{trap for cane edible}
\fmtExampleRL{Eki\-su\-yi ti\-ki\-te\-ga mbi\-ri.}
\fmtExampleEN{The cane rat trap cannot trap a hippopotamus.}

\fmtLemma{ki\-swa}
\fmtCitation{(e)ki\-swa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{flying termites' mound}
\fmtExampleRL{Enswa zi\-bba mu ki\-swa.}
\fmtExampleEN{Flying termites live in a termite hill.}

\fmtLemma{ki\-swai\-zi}
\fmtCitation{(e)ki\-swai\-zi}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{diluted local brew}
\fmtExampleRL{Eki\-swai\-zi te\-ki\-ta\-mii\-rya.}
\fmtExampleEN{Diluted local brew does not make anyone drunk.}

\fmtLemma{ki\-swei\-ta}%
\fmtOne{1}
\fmtCitation{(o)ki\-swei\-ta}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{woman in menopause}
\fmtExampleRL{Oki\-swei\-ta atee\-kwa ku\-bba mu\-ka\-li mu\-ku\-lu.}
\fmtExampleEN{A woman in menopause must be elderly.}

\fmtLemma{ki\-swei\-ta}%
\fmtOne{2}
\fmtCitation{(o)ki\-swei\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{oversize (of breasts)}
\fmtExampleRL{Ama\-bee\-re gaa\-mwe ga ki\-swei\-ta.}
\fmtExampleEN{Her breasts are oversized.}

\fmtLemma{Ki\-sya\-ga}%
\fmtOne{1}
\fmtCitation{(o)Ki\-sya\-ga}
\fmtPOS{n}%
\fmtSense{title given to kings of Baruuri and Banyala}

\fmtLemma{ki\-sya\-ga}%
\fmtOne{2}
\fmtCitation{(o)ki\-sya\-ga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{signal grass (\textit{Brachiaria decumbens})}
\fmtExampleRL{Eki\-sya\-ga ki\-me\-ra ampi n’e\-nya\-nja.}
\fmtExampleEN{\textit{Bra\-chia\-ria decumbens} grows near the lake.}

\fmtLemma{ki\-syo}
\fmtCitation{(e)ki\-syo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{weaver bird}
\fmtExampleRL{Oki\-syo alya oduu\-ma.}
\fmtExampleEN{The weaver bird eats maize.}

\fmtLemma{ki\-syo\-do}
\fmtCitation{(e)ki\-syo\-do}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{chicken droppings}
\fmtExampleRL{Eki\-syo\-do kya nko\-ko ki\-ku\-wu\-nya.}
\fmtExampleEN{The chicken droppings smell.}

\fmtLemma{ki\-taa\-bbu}
\fmtCitation{(e)ki\-taa\-bbu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{gourd or calabash filled with local brew}
\fmtExampleRL{Mu\-lee\-te eki\-taa\-bbu kya ma\-lwa.}
\fmtExampleEN{Bring the calabash full of local brew.}

\fmtLemma{ki\-taa\-bi\-ri}
\fmtCitation{(e)ki\-taa\-bi\-ri}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{drying rack for simsim (UgE)/sesame}
\fmtExampleRL{Ba\-yo\-mbo\-ke\-re eki\-taa\-bi\-ri kya ma\-ca\-nde eki\-koo\-to.}
\fmtExampleEN{They have constructed a big drying rack for simsim.}

\fmtLemma{ki\-taa\-di\-ri}
\fmtCitation{(e)ki\-taa\-di\-ri}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(b)
\fmtSense{raised platform to store hay, platform, pulpit}

\fmtLemma{Ki\-taa\-ku\-le}
\fmtCitation{(o)Ki\-taa\-ku\-le}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{name of a boy whose siblings are dead}
\fmtExampleRL{Omwa\-na oyo aye\-te\-bwa Ki\-taa\-ku\-le ku\-bba aba\-ga\-nda boo\-na baa\-fi\-re.}
\fmtExampleEN{That child is called Kitaakule because all his siblings are dead.}

\fmtLemma{ki\-taa\-nii\-na}
\fmtCitation{(e)ki\-taa\-nii\-na}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{snail}
\fmtExampleRL{Wa\-ya\-ja ebi\-taa\-nii\-na bi\-ngi oi\-ke\-ndi ng’a\-too\-nye\-re.}
\fmtExampleEN{You will find many snails after it has rained.}

\fmtLemma{ki\-taa\-ra}
\fmtCitation{(e)ki\-taa\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{type of hair style}

\fmtLemma{ki\-taa\-ru}
\fmtCitation{(e)ki\-taa\-ru}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{drying lane for cassava or sweet potato}
\fmtExampleRL{Ebi\-tee\-re bi\-te oku ki\-taa\-ru.}
\fmtExampleEN{Put the sliced sweet potatoes on the drying lane.}

\fmtLemma{ki\-ta\-ba}
\fmtCitation{(e)ki\-ta\-ba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{puddle, waterlogged pit, temporary gathering of water, stagnant water after rain}
\fmtExampleRL{Ota\-ye\-ma mu ki\-ta\-ba.}
\fmtExampleEN{Do not step into a puddle.}

\fmtLemma{ki\-ta\-bo}
\fmtCitation{(e)ki\-ta\-bo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from Ar.\,via Sw.}
\fmtSense{book}
\fmtExampleRL{Ndee\-te\-ra eki\-ta\-bo ekya bya\-faa\-yo bya Uga\-nda.}
\fmtExampleEN{Bring me the book on Uganda's history.}

\fmtLemma{ki\-ta\-bu}
\fmtCitation{(e)ki\-ta\-bu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bed}
\fmtExampleRL{Abba\-si\-rye mu ki\-ta\-bu kya\-mwe.}
\fmtExampleEN{He slept in his bed.}

\fmtLemma{ki\-ta\-fee\-ri}
\fmtCitation{(e)ki\-ta\-fee\-ri}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{soursop, graviola}
\fmtExampleRL{Eki\-ta\-fee\-ri ki\-ba\-la eki\-woo\-ma.}
\fmtExampleEN{The soursop is a sweet fruit.}

\fmtLemma{ki\-ta\-ga\-tya}
\fmtCitation{(e)ki\-ta\-ga\-tya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{traditional oven}
\fmtExampleRL{Omu\-ku\-bi agu\-tai\-re mu ki\-ta\-ga\-tya.}
\fmtExampleEN{He put the sauce into the traditional oven.}

\fmtLemma{ki\-ta\-ka\-ba\-la}
\fmtCitation{(e)ki\-ta\-ka\-ba\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{large clay pot, large pot for water, large vessel}
\fmtExampleRL{Mwi\-zu\-lye eki\-ta\-ka\-ba\-la kya\-nge amai\-zi.}
\fmtExampleEN{Fill my large pot with water.}

\fmtLemma{ki\-ta\-ka\-li}
\fmtCitation{(e)ki\-ta\-ka\-li}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{brown colour}
\fmtExampleRL{Olu\-go\-ye lwa\-mwe lwa ki\-ta\-ka\-li.}
\fmtExampleEN{Her dress is brown.}

\fmtLemma{ki\-ta\-ka\-li\-ta\-ka\-li}
\fmtCitation{(e)ki\-ta\-ka\-li\-ta\-ka\-li}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{reddish-brown colour}
\fmtExampleRL{Aguu\-re olu\-go\-ye lwa ki\-ta\-ka\-li\-ta\-ka\-li.}
\fmtExampleEN{He bought a reddish-brown garment.}

\fmtLemma{ki\-ta\-ki\-bwa}
\fmtCitation{(e)ki\-ta\-ki\-bwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{need, want, requirement}
\fmtExampleRL{Tu\-li\-na ebi\-ta\-ki\-bwa omu bwo\-mi bwai\-swe.}
\fmtExampleEN{We have needs in our lives.}

\fmtLemma{ki\-ta\-le}
\fmtCitation{(o)ki\-ta\-le}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{white cow}
\fmtExampleRL{Aba\-lii\-sya ba\-ta\-ka oku\-lii\-sya zi\-ki\-ta\-le.}
\fmtExampleEN{Cattle keepers like keeping white cows.}

\fmtLemma{ki\-ta\-lo}%
\fmtOne{1}
\fmtPOS{adv}
\fmtSubentry(a)
\fmtSense{too much, very much, very fast}
\fmtExampleRL{Ki\-ni ki\-ta\-lo!}
\fmtExampleEN{It is too much!}
\fmtSubentry(b)
\fmtSense{too bad, so sad}
\fmtExampleRL{Ka ki\-ta\-lo mu\-no oku\-fee\-rwa omaa\-wo.}
\fmtExampleEN{It is a pity that you have lost your mom.}

\fmtLemma{ki\-ta\-lo}%
\fmtOne{2}
\fmtCitation{(e)ki\-ta\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wonder, marvel}

\fmtLemma{ki\-ta\-maa\-li}%
\fmtOne{1}
\fmtCitation{(e)ki\-ta\-maa\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{type of tree that grows on the shores}
\fmtSubentry(b)
\fmtSense{floating log}
\fmtExampleRL{Abaa\-na ba\-ku\-gi\-mbi\-ra ku bi\-ta\-maa\-li.}
\fmtExampleEN{Children are floating on logs.}

\fmtLemma{ki\-ta\-mbaa\-la}
\fmtCitation{(e)ki\-ta\-mbaa\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{sash}
\fmtExampleRL{Mpee\-rya eki\-ta\-mbaa\-la mbi\-ne\-mu.}
\fmtExampleEN{Give me a sash and I'll dance a bit.}
\fmtSubentry(b)
\fmtSense{turban}
\fmtExampleRL{Aba\-wa\-ra\-bu ba\-bbwe\-re ebi\-ta\-mbaa\-la.}
\fmtExampleEN{Arabs wear turbans.}
\fmtSubentry(c)
\fmtSense{head scarf}
\fmtExampleRL{Omu\-ka\-li yee\-bbwe\-re eki\-ta\-mbaa\-la.}
\fmtExampleEN{The woman is wearing a head scarf.}
\fmtSubentry(d)
\fmtSense{waistband}
\fmtExampleRL{Go\-me\-si aba\-ka\-li ba\-zi\-bwe\-sya bi\-ta\-mba\-la.}
\fmtExampleEN{Women tie the gomesi using a waistband.}
\fmtSubentry(e)
\fmtSense{tablecloth}
\fmtExampleRL{Emee\-za yaa\-mwe eri\-ku eki\-ta\-mbaa\-la.}
\fmtExampleEN{His dining table is covered with a table cloth.}

\fmtLemma{ki\-ta\-mbi\-ro}
\fmtCitation{(e)ki\-ta\-mbi\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{church mass, church service}
\fmtExampleRL{Eki\-ta\-mbi\-ro kya mi\-sa ki\-wei\-re.}
\fmtExampleEN{The church service has concluded.}

\fmtLemma{ki\-ta\-mbo}
\fmtCitation{(e)ki\-ta\-mbo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{evil spirit (which turns people into cannibals)}
\fmtExampleRL{Omu\-sai\-za odi ali\-ku eki\-ta\-mbo.}
\fmtExampleEN{That man is possessed by an evil spirit which turns people into cannibals.}

\fmtLemma{ki\-ta\-mbu\-li\-syo}
\fmtCitation{(e)ki\-ta\-mbu\-li\-syo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{identity card}
\fmtExampleRL{Ki\-sai oku\-bba n’e\-ki\-ta\-mbu\-li\-syo.}
\fmtExampleEN{It is good to have an identity card.}

\fmtLemma{ki\-ta\-mii\-rya}
\fmtCitation{(e)ki\-ta\-mii\-rya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{alcohol, alcoholic beverag}
\fmtExampleRL{Mu\-le\-ke\-ra\-wo oku\-nywa ebi\-ta\-mii\-rya.}
\fmtExampleEN{Stop drinking alcohol!}

\fmtLemma{ki\-ta\-nda}
\fmtCitation{(e)ki\-ta\-nda}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bed}
\fmtExampleRL{Eki\-ta\-nda kya\-mwe ki\-ri\-mu ebi\-ku.}
\fmtExampleEN{His bed has bed bugs.}

\fmtLemma{ki\-ta\-ni}
\fmtCitation{(e)ki\-ta\-ni}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{placenta}
\fmtExampleRL{Lii\-ka eki\-ta\-ni ekyo embwe\-ne zi\-ta\-ki\-rya.}
\fmtExampleEN{Bury that placenta so that the dogs do not eat it.}

\fmtLemma{ki\-ta\-nta\-la\-nke\-mbo}
\fmtCitation{(e)ki\-ta\-nta\-la\-nke\-mbo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{type of yam}

\fmtLemma{ki\-ta\-nvuu\-wo}
\fmtCitation{(e)ki\-ta\-nvuu\-wo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sash}
\fmtExampleRL{Aba\-ka\-li be\-bbwa ebi\-ta\-nvuu\-wo oku go\-me\-si.}
\fmtExampleEN{Women wear a sash on a gomesi. (lit.\,Women tie themselves with a sash on a gomesi.)}

\fmtLemma{ki\-ta\-ra\-sa}
\fmtCitation{(e)ki\-ta\-ra\-sa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{whitewash (win in a game in which the loser scores no points)}
\fmtExampleRL{OMu\-sii\-ta ba\-mu\-ku\-bbi\-re ki\-ta\-ra\-sa.}
\fmtExampleEN{Musiita whitewashed.}

\fmtLemma{ki\-te\-be}
\fmtCitation{(e)ki\-te\-be}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{headquarters}
\fmtExampleRL{Eki\-te\-be kya po\-lii\-si ki\-ta Na\-gu\-ru.}
\fmtExampleEN{The police headquarters are located at Naguru.}

\fmtLemma{ki\-tee\-ke\-ro}
\fmtCitation{(e)ki\-tee\-ke\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{place where a pot is kept}
\fmtExampleRL{Omu ki\-tee\-ke\-ro kya nsu\-wa wa\-bbaa\-wo ebi\-ke\-re.}
\fmtExampleEN{There are frogs where the pot is kept.}

\fmtLemma{ki\-tee\-ra}
\fmtCitation{(e)ki\-tee\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{banana plant under which a placenta is buried}
\fmtExampleRL{Eki\-tee\-ra ki\-bbaa\-ku emi\-zi\-zo gi\-ngi.}
\fmtExampleEN{The banana plant where the placenta is buried has many taboos.}

\fmtLemma{ki\-tee\-re}
\fmtCitation{(e)ki\-tee\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sweet potato slice}
\fmtExampleRL{Ebi\-tee\-re bi\-kaa\-re ku\-sai.}
\fmtExampleEN{The sweet potato slices have dried well.}

\fmtLemma{ki\-tee\-so}
\fmtCitation{(e)ki\-tee\-so}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{suggestion, proposal}
\fmtExampleRL{Lee\-ta eki\-tee\-so kya\-mu.}
\fmtExampleEN{Bring forward your suggestion.}
\fmtSubentry(b)
\fmtSense{minutes, summarised record of the proceedings at a meeting, notes, transcript}
\fmtExampleRL{Wa\-ndii\-ka ebi\-tee\-so bya lu\-kii\-ko.}
\fmtExampleEN{Write down the minutes of the meeting.}

\fmtLemma{ki\-tee\-tee\-yi}%
\fmtOne{1}
\fmtCitation{(e)ki\-tee\-tee\-yi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{dress}
\fmtExampleRL{OZa\-ri\-na ali\-na ebi\-tee\-tee\-yi ebi\-sai.}
\fmtExampleEN{Zarina has beautiful dresses.}

\fmtLemma{ki\-tee\-tee\-yi}%
\fmtOne{2}
\fmtCitation{(e)ki\-tee\-tee\-yi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{round object, circle}
\fmtExampleRL{Ayo\-mbe\-ke\-re enyu\-mba ya ki\-tee\-tee\-yi.}
\fmtExampleEN{He has constructed a round house.}

\fmtLemma{ki\-te\-ma}
\fmtCitation{(o)ki\-te\-ma}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{slaughterer, butcher}
\fmtExampleRL{Aba\-si\-raa\-mu ba\-ko\-la nka ba\-ki\-te\-ma.}
\fmtExampleEN{Moslems work as slaughterers.}

\fmtLemma{ki\-te\-mbe}
\fmtCitation{(e)ki\-te\-mbe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{enset, Ethiopian banana, false banana (\textit{Ensete ventricosum})}
\fmtExampleRL{Ebi\-te\-mbe ti\-bi\-zwa\-ku ma\-too\-ke ga\-ku\-lya.}
\fmtExampleEN{\textit{Ensete ventricosum} does not yield edible fruit.}
\fmtSubentry(b)
\fmtSense{headgear of traditional healers}
\fmtExampleRL{Aba\-su\-mu bee\-bbwa ebi\-te\-mbe.}
\fmtExampleEN{Traditional healers wear headgears.}

\fmtLemma{ki\-te\-nde}
\fmtCitation{(e)ki\-te\-nde}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{soft stone, soft rock}
\fmtExampleRL{Tu\-gu\-le\-yo ama\-bbaa\-le ga ki\-te\-nde aga nyu\-mba.}
\fmtExampleEN{Let us buy soft stones for the house.}

\fmtLemma{ki\-te\-ndu\-ka}
\fmtCitation{(e)ki\-te\-ndu\-ka}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{epilepsy}
\fmtExampleRL{Ba\-mbi ya\-lwai\-re ki\-te\-ndu\-ka.}
\fmtExampleEN{Oh dear, he is suffering from epilepsy.}

\fmtLemma{ki\-te\-ngee\-to}
\fmtCitation{(e)ki\-te\-ngee\-to}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bog, marshland}
\fmtExampleRL{Emo\-to\-ka ya\-gwi\-re omu ki\-te\-ngee\-to.}
\fmtExampleEN{The car fell in a bog.}

\fmtLemma{ki\-te\-ngee\-tya mu\-twe}
\fmtCitation{(e)ki\-te\-ngee\-tya mu\-twe}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{nodding disease}
\fmtExampleRL{Obu\-lwai\-re bwa ki\-te\-ngee\-tya mu\-twe bu\-maa\-re\-wo aba\-ntu.}
\fmtExampleEN{The nodding disease is killing people.}

\fmtLemma{ki\-te\-nge\-ja}
\fmtCrossreference{\textnormal{see also} kitengeta}
\fmtCitation{(e)ki\-te\-nge\-ja}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{water lettuce (\textit{Pistia stratiotes})}
\fmtExampleRL{OKyo\-ga ali\-ku ebi\-te\-nge\-ja bi\-ngi.}
\fmtExampleEN{There are lots and lots of \textit{Pistia stratiotes} on Lake Kyoga.}

\fmtLemma{ki\-te\-nge\-ke\-ro}
\fmtCitation{(e)ki\-te\-nge\-ke\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{base ring where a pot is positioned}
\fmtExampleRL{Sa\-la ei\-su\-bi oko\-le\-mu eki\-te\-nge\-ke\-ro kya nsu\-wa.}
\fmtExampleEN{Cut the grass to make a base ring from it to set the pot on.}

\fmtLemma{ki\-te\-nge\-ta}
\fmtCrossreference{\textnormal{see also} kitengeja}
\fmtCitation{(e)ki\-te\-ngee\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{water lettuce (\textit{Pistia stratiotes})}
\fmtExampleRL{Eki\-te\-nge\-ka ki\-ba\-la oku mai\-zi.}
\fmtExampleEN{\textit{Pistia stratiotes} grows on water.}

\fmtLemma{ki\-te\-ngi}
\fmtCitation{(e)ki\-te\-ngi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{kitengi (type of clothing)}
\fmtExampleRL{Aba\-na\-ge\-ria ba\-ta\-ka mu\-no oku\-zwa\-la ebi\-te\-ngi.}
\fmtExampleEN{Nigerians really like wearing bitengi.}

\fmtLemma{ki\-te\-ngu}
\fmtCrossreference{\textnormal{see also} kitentege\-re}
\fmtCitation{(e)ki\-te\-ngu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{rag, tattled cloth}
\fmtExampleRL{Omu\-gwai\-ra\-lu azwai\-re ebi\-te\-ngu.}
\fmtExampleEN{The madman is wearing rags.}

\fmtLemma{ki\-te\-nte\-ge\-re}
\fmtCitation{(e)ki\-te\-nte\-ge\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{rough bark cloth (the first layer of the barkcloth tree, used for burials and other purposes which do not require a finer type of bark cloth)}
\fmtExampleRL{Ei\-rai abaa\-na bee\-swe\-ka\-nga bi\-te\-nte\-ge\-re.}
\fmtExampleEN{In the past, children used to cover themselves with rough bark cloth.}
\fmtSubentry(b)
\fmtSense{rag, tattled clothing}
\fmtExampleRL{Aboo\-ro ba\-zwa\-la bi\-te\-nte\-ge\-re.}
\fmtExampleEN{The poor wear tattered items of clothing.}

\fmtLemma{ki\-te\-te\-nge}
\fmtCitation{(e)ki\-te\-te\-nge}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{millet chaff}
\fmtExampleRL{Yo\-ca ebi\-te\-te\-nge ebyo.}
\fmtExampleEN{Burn that millet chaff.}

\fmtLemma{ki\-ti\-gi}%
\fmtOne{1}
\fmtCitation{(e)ki\-ti\-gi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{eyelid}
\fmtExampleRL{Aba\-ka\-li ba bi\-ro bi\-ni ba\-sii\-ga ebi\-ti\-gi.}
\fmtExampleEN{Women of this generation colour their eyelids (i.e. wear eyeshadow).}

\fmtLemma{ki\-ti\-gi}%
\fmtOne{2}
\fmtCitation{(e)ki\-ti\-gi}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{insolence, bad manners, impoliteness}
\fmtExampleRL{Eki\-ti\-gi omu baa\-na ki\-bbi mu\-no.}
\fmtExampleEN{Insolence in children is bad.}

\fmtLemma{ki\-tii\-bwa}
\fmtCitation{(e)ki\-tii\-bwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{respect, honour}
\fmtExampleRL{Ntaa\-mu eki\-tii\-bwa.}
\fmtExampleEN{Show me some respect.}

\fmtLemma{ki\-tii\-ni}
\fmtCitation{(e)ki\-tii\-ni}
\fmtNC{7}
\fmtSense{cowardness}
\fmtExampleRL{Endwa\-na yaa\-mwe ya kii\-ti\-ni.}
\fmtExampleEN{His way of fighting is cowardly.}

\fmtLemma{ki\-tii\-ni\-sya}
\fmtCitation{(e)ki\-tii\-ni\-sya}
\fmtPOS{n}%
\fmtNC{7}
\fmtSubentry(a)
\fmtSense{honour, respect}
\fmtExampleRL{Aba\-ka\-ma ba\-ta ne ki\-tii\-ni\-sya ki\-ngi.}
\fmtExampleEN{Kings demand a lot of respect.}
\fmtSubentry(b)
\fmtSense{glory, fame, praise}
\fmtExampleRL{OKa\-nca aku\-saa\-ni\-ra eki\-tii\-ni\-sya kyo\-na kyo\-na.}
\fmtExampleEN{God deserves all the glory.}

\fmtLemma{ki\-tii\-ri}
\fmtCitation{(e)ki\-tii\-ri}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from taboo}
\fmtSense{anus}
\fmtExampleRL{Eki\-tii\-ri ki\-ku\-mu\-si\-ya.}
\fmtExampleEN{His anus is itching.}

\fmtLemma{ki\-tii\-yo}
\fmtCitation{(e)ki\-tii\-yo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{spade}
\fmtExampleRL{Lee\-ta ani eki\-tii\-yo tu\-ta\-bu\-le omu\-se\-nyu.}
\fmtExampleEN{Bring the spade and we will mix the sand.}

\fmtLemma{ki\-ti\-ki\-ti}
\fmtCitation{(o)ki\-ti\-ki\-ti}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{excitement, enthusiasm, craze, fad}
\fmtExampleRL{Oki\-ti\-ki\-ti owa ku\-lii\-sya obu\-nguu\-ra ya\-cwe\-kei\-re.}
\fmtExampleEN{The caterpillar farming craze vanished.}

\fmtLemma{ki\-ti\-mba}
\fmtCitation{(e)ki\-ti\-mba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{net (for hunting or fishing)}
\fmtExampleRL{Onso\-lo agwi\-re omu ki\-ti\-mba kya\-nge.}
\fmtExampleEN{The animal was trapped in my net.}

\fmtLemma{ki\-ti\-na}%
\fmtOne{1}
\fmtCitation{(e)ki\-ti\-na}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{stem, trunk}
\fmtExampleRL{Eki\-ti\-na kya mu\-saa\-le ki\-fii\-re.}
\fmtExampleEN{The trunk of the tree is rotten.}

\fmtLemma{ki\-ti\-na}%
\fmtOne{2}
\fmtCitation{(e)ki\-ti\-na}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{ancestry, lineage, descent}
\fmtExampleRL{Eki\-ti\-na kya\-mu ki\-ta ayi?}
\fmtExampleEN{Where is your ancestry?}

\fmtLemma{ki\-ti\-nde}
\fmtDialect{Ruruuli}
\fmtCitation{(e)ki\-ti\-nde}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{temporary hut}
\fmtExampleRL{Aba\-ntu oku lu\-mbe ba\-bba\-sya mu bi\-ti\-nde.}
\fmtExampleEN{People sleep in temporary huts during the celebration of funeral rites.}

\fmtLemma{ki\-ti\-ndi}
\fmtSubentry(a)
\fmtCrossreference{\textnormal{see} kitinde}
\fmtSubentry(b)
\fmtSense{log placed on the grave to stop cannibals from exhuming the dead body (It is believed that cannibals can perform ceremonies and have a body exhumed without digging up so that it comes out by itself.)}

\fmtLemma{ki\-ti\-ndi\-ryo}
\fmtCitation{(e)ki\-ti\-ndi\-ryo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{heel}
\fmtExampleRL{Ye\-ko\-ne\-re eki\-ti\-ndi\-ryo oku ibbaa\-le.}
\fmtExampleEN{He knocked his heel on the stone.}

\fmtLemma{ki\-ti\-ndyo\-le}
\fmtCitation{(e)ki\-ti\-ndyo\-le}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{mound (\eg for sweet potatoes)}
\fmtExampleRL{Ebi\-yaa\-ta ba\-bi\-si\-mba omu bi\-ti\-ndyo\-le.}
\fmtExampleEN{Sweet potatoes are grown in potato mounds.}

\fmtLemma{ki\-to\-gi}
\fmtCitation{(e)ki\-to\-gi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{collar}
\fmtExampleRL{Esaa\-ti yaa\-mwe eyo\-no\-ke\-re eki\-to\-gi.}
\fmtExampleEN{The collar of his shirt is torn.}

\fmtLemma{ki\-to\-le}
\fmtCitation{(e)ki\-to\-le}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{portion, piece (especially of cooked matoke)}
\fmtExampleRL{Afu\-ni\-re ebi\-to\-le bya me\-re ya ma\-too\-ke bi\-bi\-ri.}
\fmtExampleEN{He got two pieces of cooked matoke.}

\fmtLemma{ki\-to\-nde}
\fmtCitation{(e)ki\-to\-nde}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{creature}
\fmtExampleRL{Tu\-li bi\-to\-nde bya Ka\-nca.}
\fmtExampleEN{We are God’s creatures.}

\fmtLemma{ki\-to\-ndee\-rye}
\fmtCitation{(e)ki\-to\-ndee\-rye}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{post-harvest sprout}
\fmtExampleRL{Omu ndu\-mi\-ro mu\-li\-mu ebi\-to\-ndee\-rye.}
\fmtExampleEN{There are post-harvest sprouts in the garden.}

\fmtLemma{ki\-to\-ne}
\fmtCitation{(e)ki\-to\-ne}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{talent}
\fmtExampleRL{Tu\-ta n’e\-bi\-to\-ne bi\-ngi.}
\fmtExampleEN{We are gifted with many talents.}

\fmtLemma{ki\-to\-ngo\-le}
\fmtCitation{(e)ki\-to\-ngo\-le}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{organisation, company}
\fmtExampleRL{Nko\-la mu ki\-to\-ngo\-le kya bya\-bwo\-mi.}
\fmtExampleEN{I work in the health department.}
\fmtSubentry(b)
\fmtSense{ministry, department}
\fmtExampleRL{Omu\-ku\-lu oyo azwi\-re mu ki\-to\-ngo\-le kya bya\-bwo\-mi.}
\fmtExampleEN{That official is from the ministry of health.}

\fmtLemma{ki\-to\-nto}
\fmtCitation{(e)ki\-to\-nto}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{\textit{Crassocephalum vitellinum}}
\fmtExampleRL{Aba\-ka\-li aba\-wa\-nda ba\-naa\-ba ebi\-to\-nto.}
\fmtExampleEN{Pregnant women bathe with a potion of \textit{Crassocephalum vitellinum}.}

\fmtLemma{ki\-to\-nto\-go\-ro}
\fmtCitation{(e)ki\-to\-nto\-go\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fruit of a tree of the genus \textit{Ficus}}

\fmtLemma{ki\-to\-nto\-me}
\fmtCitation{(e)ki\-to\-nto\-me}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{poem, verse}
\fmtExampleRL{Amai\-te oku\-to\-nto\-ma ebi\-to\-nto\-me.}
\fmtExampleEN{He knows how to recite poems.}

\fmtLemma{ki\-too\-be\-ro}
\fmtCitation{(e)ki\-too\-be\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{variety of foods mixed together, thick sauce}
\fmtExampleRL{Alii\-re ki\-too\-be\-ro.}
\fmtExampleEN{He ate a variety of foods mixed together.}

\fmtLemma{ki\-too\-go}
\fmtCitation{(e)ki\-too\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{papyrus reed (\textit{Cyperus papyrus})}
\fmtExampleRL{Ebi\-too\-go bi\-kuu\-re mu\-no.}
\fmtExampleEN{The papyrus reeds have overgrown.}

\fmtLemma{ki\-too\-ke}
\fmtCitation{(e)ki\-too\-ke}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{banana plant}
\fmtExampleRL{Aku\-si\-mba bi\-too\-ke.}
\fmtExampleEN{He is planting banana plants.}
\fmtSubentry(b)
\fmtSense{yellow millet}
\fmtExampleRL{Obwi\-ta bwa ki\-too\-ke bu\-woo\-ma mu\-no.}
\fmtExampleEN{The yellow millet variety is very delicious.}

\fmtLemma{ki\-too\-le\-too\-le}
\fmtCitation{(e)ki\-too\-le\-too\-le}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{ill-mannered person, uncivilised person, stupid person}
\fmtExampleRL{Ti\-nso\-bo\-la ku\-swe\-ra mu\-sai\-za ki\-too\-le\-too\-le.}
\fmtExampleEN{I cannot marry an ill-mannered man.}

\fmtLemma{ki\-tu\-gwa}
\fmtCitation{(e)ki\-tu\-gwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{domesticated animal}
\fmtExampleRL{Ebi\-tu\-gwa bya bbee\-yi oku\-li\-sya.}
\fmtExampleEN{Domesticated animals are expensive to keep.}

\fmtLemma{ki\-tu\-ku}
\fmtCitation{(e)ki\-tu\-ku}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{light-skinned}
\fmtExampleRL{Omu\-ka\-li ki\-tu\-ku abo\-ne\-ka ku\-sai.}
\fmtExampleEN{A light-skinned woman looks pretty.}

\fmtLemma{ki\-tu\-li}
\fmtCitation{(e)ki\-tu\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{hole}
\fmtExampleRL{Igaa\-la eki\-tu\-li eki\-ri omu yu\-ni\-foo\-mu ya mwa\-na.}
\fmtExampleEN{Patch up the hole in the child's uniform.}

\fmtLemma{ki\-tu\-mbi\-si\-sya}
\fmtCrossreference{\textnormal{see} kitu\-mbisya}

\fmtLemma{ki\-tu\-mbi\-sya}
\fmtCitation{(e)ki\-tu\-mbi\-sya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{baking powder}
\fmtExampleRL{Emi\-gaa\-ti ba\-gi\-taa\-mu eki\-tu\-mbi\-sya.}
\fmtExampleEN{They add baking powder to bread.}

\fmtLemma{ki\-tu\-mbu\-gu\-lu}
\fmtCitation{(e)ki\-tu\-mbu\-gu\-lu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{chicken thigh}
\fmtExampleRL{Nta\-ka mu\-no oku\-lya eki\-tu\-mbu\-gu\-lu.}
\fmtExampleEN{I like eating chicken thighs.}

\fmtLemma{ki\-tu\-ndu}
\fmtCitation{(e)ki\-tu\-ndu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{part, piece, portion}
\fmtExampleRL{Nsaa\-ra\-ku eki\-tu\-ndu oku ki\-ba\-nja kya\-mu.}
\fmtExampleEN{Give me a part of your land.}
\fmtSubentry(b)
\fmtSense{area, region, locality}
\fmtExampleRL{Nko\-be\-ra ebi\-ku\-fa omu ki\-tu\-ndu kya waa\-nywe.}
\fmtExampleEN{Tell me what is happening in your area.}
\fmtSubentry(c)
\fmtSense{province}
\fmtExampleRL{Aba\-ntu bai\-ngi omu ki\-tu\-ndu ekyo.}
\fmtExampleEN{There are many people in that province.}
\fmtSubentry(d)
\fmtSense{half, 50 per cent}
\fmtExampleRL{Ensi ba\-gi\-ku\-twi\-re\-mu ebi\-tu\-ndu bi\-bi\-ri.}
\fmtExampleEN{They divided the country into two halves.}
\fmtSubentry(f)
\fmtSense{bible passage, scripture}
\fmtExampleRL{Nso\-me\-ra\-yo eki\-tu\-ndu ki\-mwei oku\-zwa omu Bbai\-bbu\-li.}
\fmtSubentry(g)
\fmtExampleEN{Read me a scripture from the Bible.}
\fmtSense{paragraph}
\fmtExampleRL{Aso\-me\-re ebi\-tu\-ndu bi\-sa\-tu byo\-nkai.}
\fmtExampleEN{He has only read three paragraphs.}

\fmtLemma{ki\-tu\-nga}
\fmtCitation{(e)ki\-tu\-nga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{puddle, temporary gathering of water, stagnant water after rain}
\fmtExampleRL{Emo\-to\-ka emu\-mi\-sii\-re amai\-zi ga ki\-tu\-nga.}
\fmtExampleEN{A vehicle splashed him with muddy water.}
\fmtSubentry(b)
\fmtSense{water pool, swimming pool}
\fmtExampleRL{Ewa\-nge wa\-li\-yo eki\-tu\-nga ky’a\-mai\-zi.}
\fmtExampleEN{I have a swimming pool at my place.}

\fmtLemma{ki\-tu\-nga\-tu\-nga}
\fmtCitation{(e)ki\-tu\-nga\-tu\-nga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{seasonal pond}
\fmtExampleRL{Ente zi\-ta\-ka mu\-no oku\-nywa amai\-zi ga\-mu bi\-tu\-nga\-tu\-nga.}
\fmtExampleEN{Cows like drinking water from seasonal water ponds.}

\fmtLemma{ki\-tu\-ngo}
\fmtCitation{(e)ki\-tu\-ngo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{product from papyrus reed}
\fmtExampleRL{Nkoo\-ra\-yo eki\-tu\-ngo ki\-mwe.}
\fmtExampleEN{Make me a product from papyrus reeds.}

\fmtLemma{ki\-tu\-ngwa}
\fmtCitation{(e)ki\-tu\-ngwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{domestic animal}
\fmtExampleRL{Ebyo ni bi\-tu\-ngwa bya Na\-mpii\-na.}
\fmtExampleEN{Those are Nampiina’s domestic animals.}

\fmtLemma{ki\-tuu\-go}
\fmtCitation{(e)ki\-tuu\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{type of palm tree}
\fmtExampleRL{Eki\-tuu\-go kya\-mee\-re oku lu\-bu\-ga lwa\-nge.}
\fmtExampleEN{A palm tree grew in my courtyard.}

\fmtLemma{ki\-tuu\-ke}
\fmtCitation{(e)ki\-tuu\-ke}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{knot}
\fmtExampleRL{Eki\-tuu\-ke kya\-mu\-le\-me\-re oku\-sa\-nda\-bu\-la.}
\fmtExampleEN{He failed to untie the knot.}

\fmtLemma{ki\-tuu\-lo}
\fmtCrossreference{\textnormal{see} kituuro}

\fmtLemma{ki\-tuu\-ro}
\fmtCitation{(e)ki\-tuu\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{abandoned place, deserted place}
\fmtExampleRL{Ya\-suu\-re mu ki\-tuu\-ro.}
\fmtExampleEN{He slept in an abandoned place.}
\fmtSubentry(b)
\fmtSense{royal cemetery}
\fmtExampleRL{Ebi\-tuu\-ro bya ba\-ka\-ma bi\-ri Mi\-sa\-nga.}
\fmtExampleEN{The royal cemetery is found at Misanga.}

\fmtLemma{ki\-tuu\-ti}
\fmtCitation{(e)ki\-tuu\-ti}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{pulpit, pavilion, high table}
\fmtExampleRL{Ai\-kai\-re nge\-di ku ki\-tuu\-ti.}
\fmtExampleEN{He sits that way at the pulpit.}

\fmtLemma{ki\-tuu\-to}
\fmtCitation{(e)ki\-tuu\-to}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{wind sound}
\fmtExampleRL{Wuu\-ra eki\-tuu\-to kya ki\-mpa\-ngwe.}
\fmtExampleEN{Listen to the sound of the wind from the storm.}

\fmtLemma{ki\-tuu\-twa}
\fmtCitation{(e)ki\-tuu\-twa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{knot}
\fmtExampleRL{Abbwe\-re eki\-tuu\-twa oku lu\-go\-ye lwa\-mwe.}
\fmtExampleEN{He tied a knot in his cloth.}
\fmtSubentry(b)
\fmtSense{hairdo in the shape of a knot, knot hairstyle}
\fmtExampleRL{Ei\-so\-ke abbwe\-re lya ki\-tuu\-twa.}
\fmtExampleEN{She has plaited a knot hairdo.}

\fmtLemma{ki\-tu\-wo}
\fmtCrossreference{\textnormal{see} kitwo}

\fmtLemma{ki\-two}
\fmtCitation{(e)ki\-two}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{gift, present}
\fmtExampleRL{OKa\-nca aga\-ba eki\-two kya bwo\-mi.}
\fmtExampleEN{God gives the gift of life.}

\fmtLemma{ki\-two\-ni}
\fmtCitation{(e)ki\-two\-ni}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{position, responsibility, title}
\fmtExampleRL{Nko\-be\-ra eki\-two\-ni kya\-mu omu bu\-ka\-ma.}
\fmtExampleEN{Tell me your position in the kingdom.}

\fmtLemma{ki\-vu}
\fmtCitation{(e)ki\-vu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{tsetse fly}
\fmtExampleRL{Onte ane\-nei\-bwe eki\-vu.}
\fmtExampleEN{The cow was bitten by a tsetse fly.}

\fmtLemma{ki\-vu\-ga}
\fmtCitation{(e)ki\-vu\-ga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{musical instrument}
\fmtExampleRL{Ebi\-vu\-ga bye mu\-lee\-te\-re ka bi\-ngi!}
\fmtExampleEN{You have brought too many musical instruments!}

\fmtLemma{ki\-waa\-ki}
\fmtCitation{(e)ki\-waa\-ki}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{afterbirth pain due to involution, menstrual pain}
\fmtExampleRL{Ona\-ka\-zai\-re aku\-lu\-mwa eki\-waa\-ki.}
\fmtExampleEN{The mother is suffering from afterbirth pain.}

\fmtLemma{ki\-waa\-li\-mu}
\fmtCitation{(e)ki\-waa\-li\-mu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{first administrative level in the Islamic faith}

\fmtLemma{ki\-waa\-we}
\fmtDialect{Ruruuli}
\fmtCitation{(e)ki\-waa\-we}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{shell (\eg of a snail or tortoise)}
\fmtExampleRL{Eki\-taa\-nii\-na ki\-bba mu ki\-waa\-we.}
\fmtExampleEN{The snail lives in a shell.}
\fmtSubentry(b)
\fmtSense{lung}
\fmtExampleRL{Ebi\-waa\-we bya\-mwe omu\-sa\-wo ya\-ko\-be\-re bi\-rai\-rwe.}
\fmtExampleEN{The doctor said that his lungs are infected.}
\fmtSubentry(c)
\fmtSense{outer race of ball bearings (of a bicycle)}

\fmtLemma{ki\-waa\-wo}
\fmtCrossreference{\textnormal{see} kiwaawe}
\fmtDialect{Lu\-nyala}

\fmtLemma{ki\-wa\-byo}
\fmtCitation{(e)ki\-wa\-byo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{long knife used to prune banana leaves}
\fmtExampleRL{Lee\-ta eki\-wa\-byo tu\-sa\-le enda\-ga\-la.}
\fmtExampleEN{Bring the long knife and we'll cut the banana leaves.}

\fmtLemma{ki\-wai\-wo\-lo}
\fmtCitation{(e)ki\-wai\-wo\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{butterfly}
\fmtExampleRL{Eki\-wai\-wo\-lo ki\-gwi\-re omu mu\-ku\-bi.}
\fmtExampleEN{A butterfly has fallen into the sauce.}

\fmtLemma{ki\-wai\-ye\-nze}
\fmtCrossreference{\textnormal{see} kinyenze}

\fmtLemma{ki\-wa\-ka\-si}
\fmtCitation{(e)ki\-wa\-ka\-si}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{piece of carved log used as a stool}
\fmtExampleRL{Ebi\-wa\-ka\-si mu\-bi\-ta\-yi\-re ayi?}
\fmtExampleEN{Where did you put the carved logs we use as stools?}

\fmtLemma{ki\-wa\-ndii\-ko}
\fmtCrossreference{\textnormal{see} wandiika}
\fmtCitation{(e)ki\-wa\-ndii\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{document}
\fmtExampleRL{Ai\-zi\-re n’e\-ki\-wa\-ndii\-ko kya opu\-re\-zi\-de\-nti.}
\fmtExampleEN{He has come with a document from the president.}

\fmtLemma{ki\-wa\-ngii\-rye}
\fmtCitation{(e)ki\-wa\-ngii\-rye}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{accusation, allegation}
\fmtExampleRL{Ekyo ki\-wa\-ngii\-rye\-mbe, omu\-sai\-za ta\-li\-na mu\-sa\-ngo.}
\fmtExampleEN{That is just an allegation, the man is innocent.}

\fmtLemma{ki\-wa\-rai\-zi\-ma}
\fmtPOS{adv}
\fmtSense{aimlessly, in confusion, with great effort}
\fmtExampleRL{Yai\-ru\-ki\-re ki\-wa\-rai\-zi\-ma awo\-ne oku\-mwi\-ta.}
\fmtExampleEN{He ran away in confusion to avoid being killed.}

\fmtLemma{ki\-wa\-yi}
\fmtCitation{(e)ki\-wa\-yi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{portion, piece, slice (\eg of cassava)}
\fmtExampleRL{Ampai\-re ebi\-wa\-yi bya mwo\-go bi\-sa\-tu.}
\fmtExampleEN{He gave me three pieces of cassava.}
\fmtSubentry(b)
\fmtSense{group, faction, section}
\fmtExampleRL{Aba\-si\-raa\-mu ba\-li\-mu ebi\-wa\-yi bi\-bi\-ri.}
\fmtExampleEN{There are two factions among Muslims.}

\fmtLemma{ki\-wee\-we}
\fmtCitation{(e)ki\-wee\-we}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{panic, anxiety, fear}
\fmtExampleRL{Ali\-mu eki\-wee\-we ku\-bba epo\-lii\-si eku\-mu\-swa\-ga\-nii\-rya\-ku.}
\fmtExampleEN{He is in a panic because the police are investigating him.}

\fmtLemma{ki\-we\-ro}
\fmtCitation{(e)ki\-we\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{patch, rag, piece of old cloth}
\fmtExampleRL{Empa\-le yaa\-mwe ba\-gi\-tai\-re\-ku ebi\-we\-ro.}
\fmtExampleEN{They put patches onto his trousers.}

\fmtLemma{ki\-we\-wei\-ta}
\fmtCitation{(e)ki\-we\-wei\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{hypocrite}
\fmtExampleRL{Omu\-ka\-li oyo ki\-we\-wei\-ta to\-so\-bo\-la kwe\-si\-ga by’a\-ba\-za.}
\fmtExampleEN{That woman is a hypocrite, you can never trust what she says.}

\fmtLemma{ki\-wi\-wii\-tu\-ko}
\fmtPOS{adv}
\fmtSense{hastily, hurriedly, quickly, fast}
\fmtExampleRL{Ya\-ya\-bi\-re ki\-wi\-wii\-tu\-ko.}
\fmtExampleEN{He went hastily.}

\fmtLemma{ki\-wo\-go}
\fmtCitation{(o)ki\-wo\-go}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{female dark brown animal}

\fmtLemma{ki\-wo\-lee\-rya}
\fmtCitation{(e)ki\-wo\-lee\-rya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{shadow, shade}
\fmtExampleRL{Ngo\-di adi omu ki\-wo\-lee\-rya.}
\fmtExampleEN{He is there in the shade.}

\fmtLemma{ki\-wo\-lo\-go\-ma}
\fmtCitation{(e)ki\-wo\-lo\-go\-ma}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{large hole (\eg as ones found in a termite mound)}
\fmtExampleRL{Mwa\-na wa\-nge to\-zee\-nyi\-ra oku ki\-wo\-lo\-go\-ma mu\-bbaa\-mu enjo\-ka.}
\fmtExampleEN{My child, do not play around those large holes, they harbour snakes.}
\fmtSubentry(b)
\fmtSense{waterfall}
\fmtExampleRL{E Ka\-la\-ga\-la wa\-taa\-yo eki\-wo\-lo\-go\-ma.}
\fmtExampleEN{There is a waterfall in Kalagala.}

\fmtLemma{ki\-wo\-ngo\-le}
\fmtCitation{(e)ki\-wo\-ngo\-le}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{hutch, kraal}
\fmtExampleRL{Enya\-na zi\-su\-la mu ki\-wo\-ngo\-le.}
\fmtExampleEN{Calves stay in a hutch.}
\fmtSubentry(b)
\fmtSense{house without a door}
\fmtExampleRL{Ti\-mu\-ku\-bba\-sya omu nyu\-mba ya ki\-wo\-ngo\-le.}
\fmtExampleEN{Don’t sleep in a house without a door.}

\fmtLemma{ki\-wo\-ngwa}
\fmtCitation{(e)ki\-wo\-ngwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sacrifice, offering, ritual slaughter}
\fmtExampleRL{Aba\-ba\-ndwa ba\-ta\-mbi\-ra onko\-ko nk’e\-ki\-wo\-ngwa.}
\fmtExampleEN{The traditional priests offer a chicken as a sacrifice.}

\fmtLemma{ki\-woo\-pi}
\fmtCitation{(e)ki\-woo\-pi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{parish}
\fmtExampleRL{Nko\-be\-ra omu\-woo\-pi wa ki\-woo\-pi ki\-ni.}
\fmtExampleEN{Tell me who the parish chief of this parish is.}

\fmtLemma{ki\-woo\-woo\-lo}
\fmtCitation{(e)ki\-woo\-woo\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{lung (of an animal)}
\fmtExampleRL{Ebi\-woo\-woo\-lo bi\-woo\-ma omu\-ku\-bi.}
\fmtExampleEN{Animal lungs make delicious sauce.}

\fmtLemma{ki\-wo\-to\-ka}
\fmtCitation{(e)ki\-wo\-to\-ka}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{banana wilt}
\fmtExampleRL{Ebi\-go\-go bi\-ri\-mu obu\-lwai\-re bwa ki\-wo\-to\-ka.}
\fmtExampleEN{The banana plantation is infected with banana wilt.}

\fmtLemma{ki\-wo\-woi\-ta}
\fmtCitation{(o)ki\-wo\-woi\-ta}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{rumourmonger}
\fmtExampleRL{Aba\-ki\-wo\-woi\-ta ba\-ku\-ta\-bu\-la\-ta\-bu\-la aba\-ntu.}
\fmtExampleEN{Rumourmongers cause misunderstandings among people.}

\fmtLemma{ki\-wo\-wo\-ngo\-le}
\fmtPOS{adv}
\fmtSense{empty}
\fmtExampleRL{Eraa\-di\-yo ba\-gii\-bbi\-re\-mu ewa\-ya n’e\-si\-ga\-la ki\-wo\-wo\-ngo\-le.}
\fmtExampleEN{They stole the wires from the radio, and it remained empty.}

\fmtLemma{ki\-wu\-du\-wu\-du}
\fmtCitation{(e)ki\-wu\-du\-wu\-du}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{beheaded body, headless corpse}
\fmtExampleRL{Nya\-ji\-rye eki\-wu\-du\-wu\-du oku nzi\-ra.}
\fmtExampleEN{I found a beheaded body along the road.}

\fmtLemma{ki\-wu\-jo}
\fmtCitation{(e)ki\-wu\-jo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fan, ventilator}
\fmtExampleRL{Emo\-to\-ka ebbaa\-mu eki\-wu\-jo.}
\fmtExampleEN{The car has a fan.}

\fmtLemma{ki\-wu\-lu}
\fmtCitation{(e)ki\-wu\-lu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{hole, crack}
\fmtExampleRL{Enyu\-mba ba\-gi\-fu\-mwi\-re\-mu eki\-wu\-lu.}
\fmtExampleEN{They dug a hole in the wall of the house.}
\fmtSubentry(b)
\fmtSense{nostril}
\fmtExampleRL{Encwi zi\-bbaa\-ku obu\-wu\-lu.}
\fmtExampleEN{Fish have nostrils.}
\fmtSubentry(c)
\fmtSense{mortar}
\fmtExampleRL{Amai\-do nga\-wo\-nde\-ra mu ki\-wu\-lo.}
\fmtExampleEN{I pound groundnuts in the mortar.}

\fmtLemma{ki\-wu\-naa\-mii\-rya}
\fmtCitation{(e)ki\-wu\-naa\-mii\-rya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wonder, miracle, surprise}
\fmtExampleRL{Ki\-wu\-naa\-mii\-rya ombwe\-ne oku\-ka\-ma onte.}
\fmtExampleEN{It is a wonder for a dog to milk a cow. (Ruruuli-Lu\-nyala proverb)}

\fmtLemma{ki\-wu\-ta}
\fmtCitation{(e)ki\-wu\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fresh wound, injury}
\fmtExampleRL{Aba\-ntu abaa\-gwi\-re oku ka\-suu\-me ba\-fu\-ni\-re ebi\-wu\-ta.}
\fmtExampleEN{The people who were involved in an accident sustained injuries.}

\fmtLemma{ki\-wuu\-bii\-ro}
\fmtCitation{(e)ki\-wuu\-bii\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{accelerator, gas pedal}
\fmtExampleRL{Epi\-ki\-pi\-ki ebbaa\-ku eki\-wuu\-bii\-ro.}
\fmtExampleEN{The motorcycle has an accelerator.}

\fmtLemma{ki\-yaa\-ta}
\fmtCitation{(e)ki\-yaa\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sweet potato}
\fmtExampleRL{Aba\-ruu\-li n’Aba\-nya\-la ba\-li\-ma mu\-no ebi\-yaa\-ta.}
\fmtExampleEN{The Baruuli and Banyala grow a lot of sweet potatoes.}

\fmtLemma{ki\-ya\-gi}
\fmtCitation{(e)ki\-ya\-gi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{necklace}
\fmtExampleRL{Abaa\-la ba\-zwa\-la ebi\-ya\-gi.}
\fmtExampleEN{Girls wear necklaces.}

\fmtLemma{ki\-yee\-ro}
\fmtCitation{(e)ki\-yee\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{place where something was burnt}
\fmtExampleRL{Ki\-ni ni\-kyo ki\-yee\-ro kya mo\-to\-ka.}
\fmtExampleEN{This is where the car got burnt.}

\fmtLemma{ki\-ye\-nje}
\fmtCitation{(e)ki\-ye\-nje}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{cockroach}
\fmtExampleRL{Enko\-ko zi\-rya ebi\-ye\-nje.}
\fmtExampleEN{Chickens eat cockroaches.}

\fmtLemma{ki\-ye\-nke}
\fmtCitation{(e)ki\-ye\-nke}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{cataract}
\fmtExampleRL{Ata n’e\-ki\-ye\-nke oku rii\-so.}
\fmtExampleEN{He is suffering from a cataract. (lit.\,He has a cataract on the eye.)}

\fmtLemma{ki\-ye\-nze}
\fmtCitation{(e)ki\-ye\-nze}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{cockroach}
\fmtExampleRL{Enko\-ko zi\-rya ebi\-ye\-nze.}
\fmtExampleEN{Chickens eat cockroaches.}

\fmtLemma{ki\-yi\-go}
\fmtCitation{(e)ki\-yi\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{bathroom}
\fmtExampleRL{Anaa\-bi\-ra mu ki\-yi\-go.}
\fmtExampleEN{He bathes in the bathroom.}
\fmtSubentry(b)
\fmtSense{urinal}
\fmtExampleRL{Eki\-yi\-go kya ba\-sai\-za ki\-ku\-wu\-nya mu\-no.}
\fmtExampleEN{The men's urinal smells a lot.}

\fmtLemma{ki\-yi\-mba}
\fmtCitation{(e)ki\-yi\-mba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bean}
\fmtExampleRL{Eki\-yi\-mba ki\-mee\-re.}
\fmtExampleEN{The beans have germinated.}

\fmtLemma{ki\-yi\-mbye}
\fmtCitation{(e)ki\-yi\-mbye}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{heap, pile}
\fmtExampleRL{Engo\-ye zi\-bbai\-re mu ki\-yi\-mbye.}
\fmtExampleEN{The clothes were in a heap.}

\fmtLemma{ki\-yi\-nda}
\fmtCitation{(e)ki\-yi\-nda}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{residence or place for people of the Monkey clan}

\fmtLemma{ki\-yi\-ndu\-la}
\fmtCitation{(e)ki\-yi\-ndu\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{Indian abutilon (\textit{Abutilon indicum})}
\fmtExampleRL{Eki\-yi\-ndu\-la ki\-yi\-ndu\-la emi\-sa\-ngo.}
\fmtExampleEN{\textit{Abutilon indicum} is believed to avert judgement of cases in courts of law.}

\fmtLemma{ki\-yi\-ngu\-la}
\fmtCitation{(e)ki\-yi\-ngu\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{big piece, chunk (of meat)}
\fmtExampleRL{Mba\-wai\-re ebi\-yi\-ngu\-la bya nya\-ma.}
\fmtExampleEN{I gave them big pieces of meat.}

\fmtLemma{ki\-yi\-nzi}
\fmtCitation{(e)ki\-yi\-nzi}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{flu}
\fmtExampleRL{Mu\-lwai\-re wa ki\-yi\-nzi.}
\fmtExampleEN{He is suffering from the flu.}

\fmtLemma{Ki\-yi\-ra}
\fmtCitation{(o)Ki\-yi\-ra}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{River Nile}
\fmtExampleRL{Omwi\-ga Ki\-yi\-ra gu\-si\-bu\-ka omu Uga\-nda.}
\fmtExampleEN{The River Nile starts in Uganda.}

\fmtLemma{ki\-yo\-bo\-ryo}
\fmtDialect{Ruruuli}
\fmtCitation{(e)ki\-yo\-boo\-ryo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{old skin of reptiles after shedding, cast off outer cuticle of insects}
\fmtExampleRL{Eki\-yo\-boo\-ryo ki\-ra\-nga nti awo wa\-li\-wo onjo\-ka.}
\fmtExampleEN{A snakeskin shows that there is a snake in the area.}

\fmtLemma{ki\-yo\-ndo}
\fmtCitation{(e)ki\-yo\-ndo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{air plant, cathedral bells (\textit{Kalanchoe pinnata} or \textit{Bryophyllum pinnatum}, a succulent plant type used to treat skin infections)}
\fmtExampleRL{Eki\-yo\-ndo ki\-ta\-mba ona\-bu\-ge\-re.}
\fmtExampleEN{\textit{Kalanchoe pinnata} cures foot fungus.}

\fmtLemma{ki\-yo\-nga}
\fmtCitation{(e)ki\-yo\-nga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sparks of bushfire, charred remains, burned-out remains}
\fmtExampleRL{Ebi\-yo\-nga kya\-ko\-lei\-rye enyu\-mba ya\-nge.}
\fmtExampleEN{The sparks from the bushfire ignited my house.}

\fmtLemma{ki\-yo\-ngo\-be\-ro}
\fmtCitation{(e)ki\-yo\-ngo\-be\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{misery, unhappiness, distress}
\fmtExampleRL{Mu\-zwe\-re, akya\-li mu ki\-yo\-ngo\-be\-ro.}
\fmtExampleEN{Leave him, he is still in a miserable state.}

\fmtLemma{ki\-yu\-daa\-ya}
\fmtCitation{(e)ki\-yu\-daa\-ya}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{Judaism, Jewish way}
\fmtExampleRL{Ara\-mya ki\-yu\-daa\-ya.}
\fmtExampleEN{He prays in the Jewish way.}

\fmtLemma{ki\-yu\-lu}
\fmtCitation{(e)ki\-yu\-lu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{papyrus grass (\textit{Cyperus alternifolius})}
\fmtExampleRL{Eki\-yu\-lu ki\-ni nkwa\-ba ku\-ki\-ko\-lu\-kla\-mu ekii\-bbo.}
\fmtExampleEN{I am going to make a basket from this papyrus.}

\fmtLemma{ki\-yuu\-ge}
\fmtCitation{(e)ki\-yuu\-ge}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{kind of small water lily}
\fmtExampleRL{Ebi\-yuu\-ge bi\-sii\-si\-re enchwi ezi\-ri omu enya\-nja.}
\fmtExampleEN{The water lily has destroyed the fish in the lake.}

\fmtLemma{Ki\-zaa\-zi}
\fmtCitation{(o)Ki\-zaa\-zi}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{type of spirit responsible for giving children}
\fmtExampleRL{Aba\-nya\-la ba\-ko\-ba nti oKi\-zaa\-zi aga\-ba embya\-lo.}
\fmtExampleEN{The Banyala believe that Kizaazi is responsible for blessing people with children.}

\fmtLemma{ki\-zee\-nye\-ro}
\fmtCitation{(e)ki\-zee\-nye\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{playground}
\fmtExampleRL{Embi\-ro tu\-zii\-ru\-ki\-ra mu ki\-zee\-nye\-ro.}
\fmtExampleEN{We do athletics at the playground.}

\fmtLemma{ki\-zi\-bi\-ti}
\fmtCitation{(e)ki\-zi\-bi\-ti}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{evidence, proof}
\fmtExampleRL{Omu\-sa\-ngo ti\-gu\-li\-ku eki\-zi\-bi\-ti.}
\fmtExampleEN{There is no proof for this case.}

\fmtLemma{ki\-zi\-bu}
\fmtCrossreference{\textnormal{see} kizubu}

\fmtLemma{ki\-zi\-go}%
\fmtOne{1}
\fmtCitation{(e)ki\-zi\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{vaseline, petrolleum jelly}
\fmtExampleRL{Akwe\-sii\-ga bi\-zi\-go.}
\fmtExampleEN{She is smearing herself with vaseline.}

\fmtLemma{ki\-zi\-go}%
\fmtOne{2}
\fmtCitation{(e)ki\-zi\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{place where animal is cornered in hunting, animal's hideout}
\fmtExampleRL{Oki\-so\-lo baa\-mwi\-tii\-re oku ki\-zi\-go.}
\fmtExampleEN{The animal was killed in its hideout.}

\fmtLemma{ki\-zii\-ro}
\fmtCitation{(e)ki\-zii\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{suffocation, asphyxia, interruption of breathing}
\fmtExampleRL{Ya\-fii\-re eki\-zii\-ro.}
\fmtExampleEN{He died of suffocation.}

\fmtLemma{ki\-zi\-mba}
\fmtCitation{(e)ki\-zi\-mba}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{boil, abscess, pustule}
\fmtExampleRL{Ata n’e\-ki\-zi\-mba oku ita\-ma.}
\fmtExampleEN{He has a boil on his cheek.}
\fmtSubentry(b)
\fmtSense{sore, wound, injury}
\fmtExampleRL{Omwa\-na ata n’o\-bu\-zi\-mba oku mu\-bi\-ri.}
\fmtExampleEN{The child has sores on its body.}
\fmtSubentry(c)
\fmtSense{fibroid}
\fmtExampleRL{Ya\-wo\-ne\-re ebi\-zi\-mba bya mu\-nda.}
\fmtExampleEN{He was cured of fibroids.}

\fmtLemma{ki\-zi\-mbe}
\fmtCitation{(e)ki\-zi\-mbe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{building}
\fmtExampleRL{Ya\-ba oku ki\-zi\-mbe ki\-di eki\-yaa\-ka.}
\fmtExampleEN{Go to that new building.}

\fmtLemma{ki\-zi\-mbu\-lu\-ku\-sya}
\fmtCitation{(e)ki\-zi\-mbu\-lu\-ku\-sya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{yeast}
\fmtExampleRL{Eki\-zi\-mbu\-lu\-ku\-sya oki\-tei\-re ayi?}
\fmtExampleEN{Where did you put the yeast?}
\fmtSubentry(b)
\fmtSense{baking powder}
\fmtExampleRL{Emi\-gaa\-ti ba\-gi\-taa\-mu eki\-zi\-mbu\-lu\-ku\-sya.}
\fmtExampleEN{They add baking powder to bread.}

\fmtLemma{ki\-zi\-nga}
\fmtCitation{(e)ki\-zi\-nga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{island}
\fmtExampleRL{Ebi\-ro bi\-ni abba ku ki\-zi\-nga.}
\fmtExampleEN{These days he lives on the island.}

\fmtLemma{ki\-zi\-ngi\-ri}
\fmtCitation{(e)ki\-zi\-ngi\-ri}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{grauer's broadbill, African green broadbill (\textit{Pseudocalyptomena graueri})}
\fmtExampleRL{Eki\-zi\-ngi\-ri ki\-rya ama\-ca\-nde.}
\fmtExampleEN{The grauer's broadbill eats simsim.}

\fmtLemma{ki\-zi\-rii\-rwa}
\fmtCitation{(e)ki\-zi\-rii\-rwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{unhatched egg}
\fmtExampleRL{Ebi\-zi\-rii\-rwa bi\-wu\-nya mu\-no.}
\fmtExampleEN{Unhatched eggs smell very bad.}

\fmtLemma{ki\-zi\-ro}
\fmtCitation{(e)ki\-zi\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{swarm of mosquito}
\fmtExampleRL{E Ka\-yu\-nga wa\-li\-yo eki\-zi\-ro.}
\fmtExampleEN{There is a swarm of mosquitoes in Kayunga.}

\fmtLemma{ki\-zi\-zi}
\fmtCitation{(e)ki\-zi\-zi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{gap filler, stuffing to prevent something from leaking}
\fmtExampleRL{Eki\-do\-mo\-la ki\-tee\-mu eki\-zi\-zi ki\-re\-me ku\-too\-nya.}
\fmtExampleEN{Put in gap fillers so that the jerry can stops leaking.}
\fmtSubentry(b)
\fmtSense{roof beam}
\fmtExampleRL{Enyu\-mba tu\-gi\-taa\-ku enzi\-zi.}
\fmtExampleEN{We put roof beams onto the house.}

\fmtLemma{ki\-zi\-zi\-ri}
\fmtCitation{(e)ki\-zi\-zi\-ri}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{unhatched egg}
\fmtExampleRL{Ebi\-zi\-zi\-ri bi\-ta\-mba olu\-se\-ru.}
\fmtExampleEN{Unhatched eggs cure measles.}

\fmtLemma{ki\-zo\-ka}
\fmtCitation{(e)ki\-zo\-ka}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{colic}
\fmtExampleRL{Omu\-kai\-re wa\-nge aku\-lu\-mwa eki\-zo\-ka ki\-mu\-le\-me\-se\-rye oku\-lya.}
\fmtExampleEN{My parent is suffering from colic and cannot eat.}

\fmtLemma{ki\-zu\-bu}
\fmtCitation{(e)ki\-zu\-bu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{problem, issue}
\fmtExampleRL{Twa\-li tu\-ku\-vu\-ga emo\-to\-ka nee\-fu\-na eki\-zu\-bu omu yi\-ngi\-ni.}
\fmtExampleEN{While driving, the car had a mechanical problem in its engine.}

\fmtLemma{ki\-zu\-mo}
\fmtCitation{(e)ki\-zu\-mo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{rudeness, abuse, insult}
\fmtExampleRL{Ei\-ba\-ra eryo lya\-foo\-ke\-re ki\-zu\-mo.}
\fmtExampleEN{That name has become an insult.}

\fmtLemma{ki\-zu\-ngi\-ri\-zi}
\fmtCitation{(e)ki\-zu\-ngi\-ri\-zi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{rocket, spaceship, satelite}
\fmtExampleRL{Eki\-zu\-ngi\-ri\-zi kii\-ru\-ka mu\-no.}
\fmtExampleEN{The rocket flies very fast.}

\fmtLemma{ki\-zu\-nzuu\-li}
\fmtCitation{(e)ki\-zu\-nzuu\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{type of reddish beetle}
\fmtExampleRL{Eki\-zu\-nzuu\-li ki\-su\-la mu ki\-wu\-li kya mu\-saa\-le.}
\fmtExampleEN{The kizunzuuli beetle sleeps in a hole in a tree.}

\fmtLemma{ki\-zwa\-li\-bwa}
\fmtCrossreference{\textnormal{see also} kizwalo}
\fmtCitation{(e)ki\-zwa\-li\-bwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{robe (often of a king)}
\fmtExampleRL{Ebi\-zwa\-li\-bwa bya mu\-ka\-ma bi\-bba bya nja\-wu\-lo.}
\fmtExampleEN{The king's robes are special.}

\fmtLemma{ki\-zwa\-lo}
\fmtCitation{(e)ki\-zwa\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{attire, garment, article of clothing, cloth}
\fmtExampleRL{Eki\-zwa\-lo kya ba\-sai\-za Aba\-ruu\-li ka\-nzu.}
\fmtExampleEN{The attire for Baruuli men is a kanzu.}
\fmtSubentry(b)
\fmtFixedexpression{ki\-zwa\-lo kya mu\-ka\-ma}
\fmtSense{royal robe}
\fmtExampleRL{Eki\-zwa\-lo kya mu\-ka\-ma ki\-bba kya nja\-wu\-lo.}
\fmtExampleEN{The king's robe is special.}

\fmtLemma{ki\-zwa\-mbi\-ryo}
\fmtCitation{(e)ki\-zwa\-mbi\-ryo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{opening in a fishing net, hole in a fishing net}
\fmtExampleRL{Eki\-na\-ge ki\-ngii\-ra mu ki\-zwa\-mbi\-ryo.}
\fmtExampleEN{The tilapia entered through the opening in the fishing net.}

\fmtLemma{ki\-zwa\-ro}
\fmtCitation{(e)ki\-zwa\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{item of clothes, garments, dress, robe}
\fmtExampleRL{Nta n’e\-bi\-zwa\-ro bi\-sa\-tu.}
\fmtExampleEN{I have three articles of clothing.}

\fmtLemma{ki\-ŋaa\-ŋaa\-li}
\fmtCitation{(e)ki\-ŋaa\-ŋaa\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{jerry can or pot without a top}
\fmtExampleRL{Ama\-lwa ga\-su\-ke mu ki\-ŋaa\-ŋaa\-li.}
\fmtExampleEN{Pour the brew into the jerry can without a top.}

\fmtLemma{ko}
\fmtPOS{pro}
\fmtSense{it (noun class 12 independent pronoun)}

\fmtLemma{ko\-ba}
\fmtCitation{(oku)\-ko\-ba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-be\-re}
\fmtSense{to say, to tell, to speak}
\fmtExampleRL{Ya\-ba oko\-be oBu\-ku\-nu\-rwa ai\-ze ani.}
\fmtExampleEN{Go and tell Bukunurwa to come here.}

\fmtLemma{ko\-ba\-ku}
\fmtCitation{(oku)\-ko\-ba\-ku}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-be\-re\-ku}
\fmtSense{to counsel, to give advice}
\fmtExampleRL{Omu\-ntu n’a\-ko\-la eki\-bbi omu\-ko\-ba\-ku.}
\fmtExampleEN{When someone does wrong, counsel him.}

\fmtLemma{ko\-coi}%
\fmtOne{1}
\fmtCitation{(o)ko\-coi}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{method of planting groundnuts whereby a small hole is dug and seeds are planted forthwith}

\fmtLemma{ko\-coi}%
\fmtOne{2}
\fmtPOS{adj}
\fmtSense{related to the method of planting groundnuts whereby a small hole is dug and seeds are planted forthwith}
\fmtExampleRL{Omu\-ka\-li n’a\-si\-mba amai\-do ako\-le\-sya aka\-si\-ka aka\-ko\-coi.}
\fmtExampleEN{When a woman is planting groundnuts she uses a small dedicated hoe.}

\fmtLemma{ko\-de}
\fmtCitation{(o)ko\-de}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{disabled person}
\fmtExampleRL{Oko\-de ni\-ye aye\-be\-mbe\-ra ei\-so\-me\-ro li\-ni.}
\fmtExampleEN{It is the disabled person who heads this school.}

\fmtLemma{ko\-do\-do\-ka}
\fmtCitation{(oku)\-ko\-do\-do\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-do\-do\-ke\-re}
\fmtSense{to be sticky}
\fmtExampleRL{Omu\-ku\-bi gw’a\-Ba\-su\-da\-ni gu\-ko\-do\-do\-ka.}
\fmtExampleEN{The Sudanese sauce is sticky.}

\fmtLemma{ko\-do\-la}
\fmtCitation{(oku)\-ko\-do\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtSense{to remove, to take out}
\fmtExampleRL{Ontu\-ku\-lu ako\-do\-la amai\-so ga bi\-na\-ge.}
\fmtExampleEN{The otter removes the eyes from tilapias.}

\fmtLemma{ko\-do\-lya}
\fmtCitation{(oku)\-ko\-do\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-doi\-rye}
\fmtSubentry(a)
\fmtSense{to lower, to bring down, to take down (something hanging)}
\fmtSubentry(b)
\fmtSense{to pluck out, to pull out}
\fmtExampleRL{Ongo\-nge ako\-doi\-rei\-mu amai\-so ga bi\-na\-ge.}
\fmtExampleEN{The otter plucked out the eyes of the tilapia.}
\fmtSubentry(c)
\fmtSense{to harvest selectively}
\fmtExampleRL{Omu\-pa\-ka\-si ya\-ko\-do\-lya\-nga\-mbe ebi\-yaa\-ta.}
\fmtExampleEN{The labourer used to harvest sweet potatoes selectively.}

\fmtLemma{ko\-do\-ma\-la}
\fmtCitation{(oku)\-ko\-do\-ma\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-do\-mai\-re}
\fmtSense{to be idle and disorderly}
\fmtExampleRL{Oku\-ko\-do\-ma\-la mu\-sa\-ngo gw’a\-maa\-ni omu Uga\-nda.}
\fmtExampleEN{To be idle and disorderly is a serious offense in Uganda.}

\fmtLemma{ko\-do\-mee\-rya}
\fmtCitation{(oku)\-ko\-do\-mee\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-do\-mei\-rye}
\fmtSubentry(a)
\fmtSense{to beg, to ask for, to request}
\fmtExampleRL{Wee\-na oko\-do\-mee\-rya mu\-no.}
\fmtExampleEN{You beg unnecessarily, my friend.}
\fmtSubentry(b)
\fmtSense{to wait eagerly, wait impatiently}
\fmtExampleRL{Bu\-li lu\-na\-ku aya\-ba ku woo\-tee\-ri ako\-do\-me\-rye eme\-re.}
\fmtExampleEN{Every day he goes to the hotel and waits eagerly for food.}

\fmtLemma{ko\-doo\-la}
\fmtCitation{(oku)\-ko\-doo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-doi\-re}
\fmtSense{to scoop, to dig out}
\fmtExampleRL{Ko\-doo\-la ebi\-gi\-ta wee\-sii\-ge.}
\fmtExampleEN{Scoop out the vaseline and apply it on your body.}

\fmtLemma{koi\-ro}
\fmtCitation{(e)koi\-ro}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{lantern, torch, lamp, flashlight}
\fmtExampleRL{Nku\-ta\-ka koi\-ro ya\-ku\-ko\-le\-sya obwi\-re.}
\fmtExampleEN{I need a lantern to use at night.}

\fmtLemma{koi\-za}
\fmtCitation{(o)koi\-za}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{maternal uncle}
\fmtExampleRL{Okoi\-za waa\-mwe mu\-lai.}
\fmtExampleEN{His maternal uncle is tall.}

\fmtLemma{ko\-ja}
\fmtCrossreference{\textnormal{see} koiza}
\fmtCitation{(o)ko\-ja}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{maternal uncle}

\fmtLemma{ko\-koo\-la}
\fmtCitation{(oku)\-ko\-koo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-koi\-re}
\fmtSense{to clean (a filthy place), to wash, to cleanse}
\fmtExampleRL{Ko\-koo\-la\-wo ama\-bbi ago.}
\fmtExampleEN{Clean away those faeces.}
\fmtSense{to contract a disease}
\fmtExampleRL{Omu\-sai\-za ya\-ko\-koi\-re enzu\-ku n’a\-gi\-sii\-ga omu\-ka\-li waa\-mwe.}
\fmtExampleEN{The man contracted gonorrhoea and infected his wife with it.}

\fmtLemma{ko\-koo\-li\-ka}
\fmtCitation{(oku)\-ko\-koo\-li\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-koo\-li\-ki\-re}
\fmtSense{to cackle for a long time in fear}
\fmtExampleRL{Lwa\-ki enko\-ko zi\-ko\-koo\-li\-ka?}
\fmtExampleEN{Why have the hens been cackling in fear for such a long time?}

\fmtLemma{ko\-koo\-li\-ka\-na}
\fmtCitation{(oku)\-ko\-koo\-li\-ka\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-koo\-li\-kai\-ne}
\fmtSense{(of a hen) to cluck}
\fmtExampleRL{Enko\-ko zi\-ko\-koo\-li\-ka\-na mu\-no nga zi\-maa\-re oku\-bii\-ka.}
\fmtExampleEN{Hens cluck loudly after laying eggs.}

\fmtLemma{ko\-koo\-ma}
\fmtCitation{(oku)\-ko\-koo\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-koo\-me\-re}
\fmtSense{to shout, to scream}
\fmtExampleRL{Ko\-koo\-ma mu\-no ba\-ku\-wuu\-re.}
\fmtExampleEN{Shout loudly so that they can hear you.}

\fmtLemma{ko\-koo\-tee\-rya}
\fmtCitation{(oku)\-ko\-koo\-tee\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-koo\-te\-rei\-rye}
\fmtSense{to draw water carefully}
\fmtExampleRL{Ko\-koo\-tee\-rya otwi\-zi tu\-nywe\-ku.}
\fmtExampleEN{Draw the remaining water carefully so that we may drink some.}

\fmtLemma{ko\-koo\-za}
\fmtCitation{(oku)\-ko\-koo\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-koo\-ze\-re}
\fmtSense{to cause trouble, to be a trouble maker}
\fmtExampleRL{Omwa\-na oyo ako\-koo\-za mu\-no enta\-lo.}
\fmtExampleEN{That child causes a lot of trouble.}

\fmtLemma{ko\-ko\-ta}
\fmtCitation{(oku)\-ko\-ko\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-ko\-te\-re}
\fmtSubentry(a)
\fmtSense{to pull, to drag, to put out, to draw}
\fmtExampleRL{Ko\-ko\-ta mpo\-la\-mpo\-la okwi\-ca ne wa\-ki\-too\-la\-yo.}
\fmtExampleEN{Pull slowly until you get it out.}
\fmtSubentry(b)
\fmtSense{to fish using seine nets, to fish indiscriminately, to fish wildly}
\fmtExampleRL{Le\-ke\-ra\-wo oku\-ko\-ko\-ta encwi.}
\fmtExampleEN{Stop fishing indiscriminately.}
\fmtSubentry(c)
\fmtSense{nets for seine-haul fishing, sein nets (nets which hang vertically in the water with the bottom edge held down by weights and the top edge buoyed by floats)}
\fmtExampleRL{Ega\-vu\-me\-nti ya\-gai\-ne eko\-ko\-ta omu nya\-nja.}
\fmtExampleEN{The government prohibited the use of fishing nets which trap everything indiscriminately.}

\fmtLemma{ko\-kwo\-to\-ka}
\fmtCitation{(oku)\-ko\-kwo\-to\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-kwo\-to\-ke\-re}
\fmtSubentry(a)
\fmtSense{(of a snake) to crawl, to creep}
\fmtExampleRL{Oi\-ti\-mba ako\-kwo\-to\-ka mpo\-la.}
\fmtExampleEN{The python crawls slowly.}
\fmtSubentry(b)
\fmtSense{to move in slow motion}
\fmtExampleRL{Aba\-ntu omu fi\-ri\-mu ba\-ku\-ko\-kwo\-ta\-ka\-mbe.}
\fmtExampleEN{The people in the film are just moving in slow motion.}

\fmtLemma{ko\-la}
\fmtCitation{(oku)\-ko\-la}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{koo\-re}
\fmtSubentry(a)
\fmtSense{to do, to make}
\fmtExampleRL{Tu\-ku\-ko\-la ki\-dee\-ro kya Lu\-nya\-la n’o Ru\-ruu\-li.}
\fmtExampleEN{We are making a Ruruuli-Lu\-nyala dictionary.}
\fmtSubentry(b)
\fmtSense{to work}
\fmtExampleRL{Lwa\-ki ota\-ku\-ko\-la?}
\fmtExampleEN{Why aren't you working?}
\fmtSubentry(c)
\fmtSense{to handle, to take care of, to manage, to be in charge of}
\fmtExampleRL{Ko\-la oku nso\-nga zi\-ni.}
\fmtExampleEN{Handle these cases.}

\fmtLemma{ko\-la\-ga\-na}
\fmtCitation{(oku)\-ko\-la\-ga\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-la\-gai\-ne}
\fmtSense{to associate with, to work together, to cooperate}
\fmtExampleRL{Ti\-nko\-la\-ga\-na na mu\-ntu mu\-go\-bya.}
\fmtExampleEN{I don’t associate with a treacherous person.}

\fmtLemma{ko\-la\-ku}
\fmtCitation{(oku)\-ko\-la\-ku}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{koo\-re\-ku}
\fmtSense{to treat, to attend (to a patient)}
\fmtExampleRL{Ko\-la\-ku ba\-ntu abo.}
\fmtExampleEN{Attend to those people.}

\fmtLemma{ko\-la\-nsi}
\fmtCitation{(e)ko\-la\-nsi}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{tarmac}
\fmtExampleRL{Olu\-guu\-do lwa\-bwe lwa ko\-la\-nsi.}
\fmtExampleEN{Their road is a tarmac one.}

\fmtLemma{ko\-lee\-ra}
\fmtCitation{(oku)\-ko\-lee\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-lei\-re}
\fmtSense{to catch fire, to be lit, to lighted, to be ignited}
\fmtExampleRL{Ei\-su\-bi li\-ko\-lei\-re omuu\-ro.}
\fmtExampleEN{The grass has caught fire.}

\fmtLemma{ko\-lee\-rya}
\fmtCitation{(oku)\-ko\-lee\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-lei\-rye}
\fmtSubentry(a)
\fmtSense{to light, to kindle, to ignite}
\fmtExampleRL{Ko\-lee\-rya omuu\-ro.}
\fmtExampleEN{Light the fire.}
\fmtSubentry(b)
\fmtSense{to ignite (an engine), to start (\eg car or boat)}
\fmtExampleRL{Ko\-lee\-rya mo\-to\-ka twa\-be.}
\fmtExampleEN{Start the car and let's go.}
\fmtSubentry(c)
\fmtFixedexpression{ko\-leerya (o)lu\-ga\-mbo}
\fmtSense{to start a rumour}
\fmtExampleRL{Naa\-ni ya\-ko\-lei\-rye olu\-ga\-mbo olwo?}
\fmtExampleEN{Who started this rumour?}

\fmtLemma{ko\-lee\-rye}
\fmtCitation{\textnormal{(e.g.}\,ki\-ko\-lee\-rye\textnormal{)}}
\fmtPOS{adj}
\fmtSense{artificial}
\fmtExampleRL{Ebi\-ro bi\-ni aba\-ntu ba\-zwa\-la obu\-ko\-we obu\-ko\-lee\-rye.}
\fmtExampleEN{These days people wear artificial eyelashes.}

\fmtLemma{ko\-le\-gi}
\fmtCitation{(e)ko\-le\-gi}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{college, institution of higher education}
\fmtExampleRL{Eni ko\-le\-gi ya bu\-so\-me\-sya.}
\fmtExampleEN{This is a teacher’s college.}

\fmtLemma{ko\-le\-sya}
\fmtCitation{(oku)\-ko\-le\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-le\-sye\-re}
\fmtSense{to use, to utilise}
\fmtExampleRL{Ti\-ndi\-na bbu\-la\-ngi\-ti ya ku\-ko\-le\-sya.}
\fmtExampleEN{I have no blanket to use.}

\fmtLemma{ko\-li}
\fmtCitation{\textnormal{(e.g.}\,mu\-ko\-lo, ba\-ko\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSense{hard-working}
\fmtExampleRL{Omu\-ka\-li oyo mu\-ko\-li mu\-no.}
\fmtExampleEN{That woman is very hard-working.}

\fmtLemma{ko\-li\-ma}
\fmtCitation{(oku)\-ko\-li\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-li\-mi\-re}
\fmtSense{to curse, to cast a spell, to utter curses}
\fmtExampleRL{Omu\-ka\-li oyo ata\-ka mu\-no oku\-ko\-li\-ma.}
\fmtExampleEN{That woman likes cursing a lot.}

\fmtLemma{ko\-lo\-ba}
\fmtCitation{(oku)\-ko\-lo\-ba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-lo\-be\-re}
\fmtSense{to work for a wage, to be employed, to work for a living}
\fmtExampleRL{Aba\-ruu\-li ti\-ba\-ta\-ka ku\-ko\-lo\-ba.}
\fmtExampleEN{The Baruuli hate working for a wage.}

\fmtLemma{ko\-lo\-ko\-lo}
\fmtCitation{(e)ko\-lo\-ko\-lo}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{cup used to measure local brew}
\fmtExampleRL{Nse\-ne\-ra\-yo eko\-lo\-ko\-lo ya kwe\-te.}
\fmtExampleEN{Give me a cup of local brew.}

\fmtLemma{ko\-lo\-ko\-ta}
\fmtCitation{(oku)\-ko\-lo\-ko\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-lo\-ko\-te\-re}
\fmtSense{to scrape (off), to scratch, to grate}
\fmtExampleRL{Ko\-lo\-ko\-ta omu\-wo\-go.}
\fmtExampleEN{Scrape off the inner layer of the cassava.}
\fmtSense{to remove (fur, bristle), to skin off}
\fmtExampleRL{Ko\-lo\-ko\-ta ombe\-re\-ge tu\-mu\-su\-mbe.}
\fmtExampleEN{Remove the pig's bristle and let's cook it.}

\fmtLemma{ko\-lo\-ndo\-lya}
\fmtCitation{(oku)\-ko\-lo\-ndo\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-lo\-ndoi\-rye}
\fmtSense{to spit, to bring up mucus}
\fmtExampleRL{Naa\-ni oyo ako\-lo\-ndoi\-re awo?}
\fmtExampleEN{Who spit there?}

\fmtLemma{ko\-ma}
\fmtCitation{(oku)\-ko\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-me\-re}
\fmtSubentry(a)
\fmtSense{to choose, to select}
\fmtExampleRL{Oku\-so\-bo\-la okwe\-ko\-me\-rya enta\-mba gy’o\-kwe\-ndya oku\-fu\-na.}
\fmtExampleEN{You can choose for yourself the best treatment you wish to have.}
\fmtSubentry(b)
\fmtSense{to pick up, to gather, to take up}
\fmtExampleRL{Ya\-ko\-me\-re olu\-pa\-pu\-la n’a\-ya\-ba.}
\fmtExampleEN{She picked up the paper and left.}
\fmtSubentry(c)
\fmtSense{to elect}
\fmtExampleRL{Baa\-bi\-re ku\-ko\-ma ise\-nte\-be.}
\fmtExampleEN{They have gone to elect the chairperson.}

\fmtLemma{ko\-ma\-ga}
\fmtCitation{(oku)\-ko\-ma\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-ma\-gi\-re}
\fmtSense{to make bark cloth, to beat barkcloth}
\fmtExampleRL{Aba\-ruu\-li te\-ba\-mai\-te ku\-ko\-ma\-ga.}
\fmtExampleEN{The Baruuli do not know how to make barkcloth.}

\fmtLemma{ko\-mba}
\fmtCitation{(oku)\-ko\-mba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-mbe\-re}
\fmtSubentry(a)
\fmtSense{to lick}
\fmtExampleRL{Le\-ke\-ra\-wo oku\-ko\-mba eki\-na\-ga.}
\fmtExampleEN{Stop licking the saucepan.}
\fmtSubentry(b)
\fmtSense{to taste, to try}
\fmtExampleRL{Ko\-mba\-ku owuu\-re obu\-nu\-li.}
\fmtExampleEN{Taste it and notice the sweetness.}

\fmtLemma{ko\-mbi}
\fmtCitation{\textnormal{(e.g.}\,mu\-ko\-mbi, ba\-ko\-mbi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{with a habit of licking}
\fmtExampleRL{Ombwe\-ne mu\-ko\-mbi wa ma\-tai.}
\fmtExampleEN{The dog has a habit of licking milk.}

\fmtLemma{ko\-mo\-la}
\fmtCitation{(oku)\-ko\-mo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-moi\-re}
\fmtSubentry(a)
\fmtSense{to trim, to cut}
\fmtExampleRL{Ko\-mo\-la olu\-ko\-me\-ra lwa lu\-bu\-ga.}
\fmtExampleEN{Trim the hedge of the compound.}
\fmtSubentry(b)
\fmtSense{to cut out, to cut around the edge}
\fmtExampleRL{Ko\-mo\-la olu\-go\-ye olwo omu mu\-so\-no gw’o\-kwe\-ndya.}
\fmtExampleEN{Cut that cloth into your desired shape.}
\fmtSubentry(c)
\fmtSense{to circumcise}
\fmtExampleRL{Mu\-ko\-mo\-le aba\-sai\-za boo\-na\-boo\-na.}
\fmtExampleEN{Circumcise all the men.}

\fmtLemma{ko\-mpyu\-ta}
\fmtCitation{(e)ko\-mpyu\-ta}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{computer}
\fmtExampleRL{Yo\-nja eko\-mpyu\-ta yaa\-mu.}
\fmtExampleEN{Clean your computer.}

\fmtLemma{ko\-mya}
\fmtCitation{(oku)\-ko\-mya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-me\-rye}
\fmtSense{to appoint, to choose, to elect, to vote}
\fmtExampleRL{Mwe\-ko\-mye\-mu oje\-ge\-nge.}
\fmtExampleEN{Choose among yourselves a supreme leader.}

\fmtLemma{ko\-na}%
\fmtOne{1}
\fmtCitation{(oku)\-ko\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-ne\-re}
\fmtSubentry(a)
\fmtSense{(of eggs) to be unhatched}
\fmtExampleRL{Ama\-wu\-li ga\-ko\-ne\-re.}
\fmtExampleEN{The eggs did not hatch.}
\fmtSubentry(b)
\fmtSense{(of food) to be half-cooked}
\fmtExampleRL{Eme\-re gye\-twa\-lii\-re ya\-bbai\-re eko\-ne\-re.}
\fmtExampleEN{The food we ate was half-cooked.}
\fmtSubentry(c)
\fmtSense{to be slow in cooking, to refuse to ripen}
\fmtSubentry(d)
\fmtSense{to be stunted}
\fmtExampleRL{Omwa\-na wa Pa\-tri\-ck ya\-ko\-ne\-re.}
\fmtExampleEN{Patrick's child is stunted.}

\fmtLemma{ko\-na}%
\fmtOne{2}
\fmtCitation{(oku)\-ko\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-ne\-re}
\fmtSense{to hit, to strike, to knock}
\fmtExampleRL{Emo\-to\-ka ka\-doo\-li emu\-ko\-ne.}
\fmtExampleEN{She was about to be hit by a car.}

\fmtLemma{ko\-nda\-ki\-ta}
\fmtCitation{(o)ko\-nda\-ki\-ta}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from En.}
\fmtSense{conductor}
\fmtExampleRL{Ebbaa\-si ebbaa\-mu oko\-nda\-ki\-ta.}
\fmtExampleEN{A bus must have a conductor.}

\fmtLemma{ko\-ndo}%
\fmtOne{1}
\fmtCitation{(o)ko\-ndo}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{corridor, hall}
\fmtExampleRL{Ayei\-me\-rei\-re nge\-di mu ko\-ndo.}
\fmtExampleEN{He is standing there in the corridor.}

\fmtLemma{ko\-ndo}%
\fmtOne{2}
\fmtCitation{(o)ko\-ndo}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{robber, thief, burglar, mugger}
\fmtExampleRL{Aba\-ko\-ndo ba\-ta\-ndi\-ki\-re okwi\-ta aba\-ntu.}
\fmtExampleEN{The robbers have started killing people.}

\fmtLemma{ko\-ndo\-mu}
\fmtCitation{(e)ko\-ndo\-mu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{condom}
\fmtExampleRL{Eko\-ndo\-mu zi\-ki\-nga endwai\-re ya na\-mu\-nywa.}
\fmtExampleEN{Condoms prevent the HIV infection.}

\fmtLemma{ko\-nee\-ra}
\fmtCitation{(oku)\-ko\-nee\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-nei\-re, ko\-ne\-rei\-re}
\fmtSense{to make compact (soil)}
\fmtExampleRL{Ko\-nee\-ra ei\-ta\-ka\-li li\-gu\-me.}
\fmtExampleEN{Compact that soil so that it becomes hard.}

\fmtLemma{ko\-nga}
\fmtCitation{(oku)\-ko\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-nge\-re}
\fmtSense{to sniff, to smell}
\fmtExampleRL{Ombwe\-ne ako\-nge\-re olu\-su lwa bai\-bbi.}
\fmtExampleEN{The dog has smelled the thieves.}

\fmtLemma{ko\-nge\-ra}
\fmtCitation{(oku)\-ko\-nge\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-ngei\-re}
\fmtSense{to be very dirty, to be filthy}
\fmtExampleRL{Bo\-na omu\-sai\-za ako\-ngei\-re omu\-no.}
\fmtExampleEN{Look at the man who is very dirty.}

\fmtLemma{ko\-ngo}
\fmtCitation{(e)ko\-ngo}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{kongo (Ateso musical instrument), thumb piano}
\fmtExampleRL{Mmai\-te oku\-ku\-bba eko\-ngo ya Ba\-te\-so.}
\fmtExampleEN{I know how to play the kongo.}

\fmtLemma{ko\-ngo\-ja}
\fmtCitation{(oku)\-ko\-ngo\-ja}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-ngo\-je\-re}
\fmtSubentry(a)
\fmtSense{to carry (another person) on one's shoulders, to drive for or serve as chauffeur for (esp. the Kabaka)}
\fmtExampleRL{Bo\-na omu\-ntu aku\-ko\-ngoi\-ze.}
\fmtExampleEN{Get someone to carry you on their shoulders.}
\fmtSubentry(b)
\fmtSense{to be infected (with measles)}
\fmtExampleRL{Omwa\-na ya\-ku\-ngu\-ji\-re olu\-nku\-se\-nse.}
\fmtExampleEN{The child is infected measles.}

\fmtLemma{ko\-ngo\-la}
\fmtCitation{(oku)\-ko\-ngo\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-ngoi\-re}
\fmtSubentry(a)
\fmtSense{to remove from a plant, to pluck off (\eg beans, groundnuts), to pick}
\fmtExampleRL{Abaa\-na ba\-ya\-bi\-re ku\-ko\-ngo\-la mai\-do.}
\fmtExampleEN{The children have gone to pluck groundnuts.}
\fmtSubentry(b)
\fmtSense{to evade one's duty, to kill time instead of working}
\fmtExampleRL{OMo\-se\-s ako\-ngo\-la mu\-no bai\-ra\-baa\-mwe ta\-ye\-ndya ku\-ko\-la.}
\fmtExampleEN{Moses is not doing his part of work, he does not want to work.}

\fmtLemma{ko\-ngu}
\fmtCitation{(a)ko\-ngu}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{young pumpkin}
\fmtExampleRL{Okwa nza\-la tu\-lya obwo\-ngu.}
\fmtExampleEN{We eat young pumpkin during a famine.}

\fmtLemma{ko\-ni}
\fmtCitation{\textnormal{(e.g.}\,mu\-ko\-ni, ba\-ko\-ni\textnormal{)}}
\fmtPOS{adj}
\fmtSense{with a habit of beating}
\fmtExampleRL{Omwa\-na oni mu\-ko\-ni wa bai\-raa\-mwe.}
\fmtExampleEN{This child has a habit of beating his pals.}

\fmtLemma{ko\-ni\-mba}
\fmtCitation{(oku)\-ko\-ni\-mba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-ni\-mbi\-re}
\fmtSense{to be malnourished, to be skinny, to be slender}
\fmtExampleRL{Omwi\-zu\-ku\-lu wa\-nge ako\-ni\-mbi\-re.}
\fmtExampleEN{My grandchild is malnourished.}

\fmtLemma{ko\-njei}
\fmtCrossreference{\textnormal{see also} lukonge}
\fmtCitation{(e)ko\-njei}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{algae, seaweed}
\fmtExampleRL{Oko\-njei ame\-ra mu ki\-fo eki\-fu\-ki oku mba\-la\-za.}
\fmtExampleEN{Algae germinates in a cold place on verandas.}

\fmtLemma{ko\-nko\-ma\-la}
\fmtCitation{(oku)\-ko\-nko\-ma\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-nko\-mai\-re}
\fmtSense{to wait in vain, to be kept waiting}
\fmtExampleRL{Mwo\-jo to\-li\-na kya ku\-ko\-la oko\-nko\-mai\-re\-mbe.}
\fmtExampleEN{Son, don't you have anything to do, you are just waiting in vain.}

\fmtLemma{ko\-nko\-na}
\fmtCitation{(oku)\-ko\-nko\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-nko\-ne\-re}
\fmtSense{to knock (on a door)}
\fmtExampleRL{Ki\-sai oku\-ko\-nko\-na nga okya\-li kwi\-ngi\-ra.}
\fmtExampleEN{It is good to knock on the door before you enter.}

\fmtLemma{ko\-no\-ko\-no}
\fmtCitation{(o)ko\-no\-ko\-no}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{larva found in decomposing grass or trees, maggot}
\fmtExampleRL{Ebi\-ko\-no\-ko\-no bi\-bba mu mi\-saa\-le.}
\fmtExampleEN{Maggots are found in trees.}

\fmtLemma{ko\-noo\-lya}
\fmtCitation{(oku)\-ko\-noo\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-noi\-rye}
\fmtSense{to tear off a piece, to break off, to bend}
\fmtExampleRL{To\-ko\-noo\-lya nda\-ga\-la za\-nge!}
\fmtExampleEN{Don’t tear pieces off my banana leaves.}

\fmtLemma{ko\-nte\-yi\-na}
\fmtCitation{(e)ko\-nte\-yi\-na}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{container, box}
\fmtExampleRL{Mu\-tee\-re omu\-gaa\-ti omu ko\-nte\-yi\-na atwa\-le oku iso\-me\-ro.}
\fmtExampleEN{Put the bread into the container so that she takes it to school.}

\fmtLemma{ko\-nto\-lya}
\fmtCitation{(oku)\-ko\-nto\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-ntoi\-rye}
\fmtSense{to click with one's tongue, to click, to make a clicking sound in the throat as a sign of sorrow or disapproval}
\fmtExampleRL{We\-we, oko\-nto\-lya nka mu\-syo\-li?}
\fmtExampleEN{My dear, are you clicking your tongue like a night dancer?}

\fmtLemma{ko\-nyo\-bo\-yi}
\fmtCitation{(e)ko\-nyo\-bo\-yi}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{type of fishing net}

\fmtLemma{koo\-byo}
\fmtCitation{(o)koo\-byo}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{water grass with long narrow leaves which cause bruises if it contacts skin}
\fmtExampleRL{Okoo\-byo ame\-ra ampi n’a\-mai\-zi.}
\fmtExampleEN{Water grass with long narrow leaves grows near the stream of water.}

\fmtLemma{koo\-di}%
\fmtOne{1}
\fmtCitation{(e)koo\-di}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{call for attention (when entering instead of knocking), request for permission to enter}
\fmtExampleRL{Naa\-ni aku\-ku\-bba ekoo\-di?}
\fmtExampleEN{Who is calling for attention?}

\fmtLemma{koo\-di}%
\fmtOne{2}
\fmtCitation{(o)koo\-di}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{first crude alcohol during filtering}
\fmtExampleRL{Okoo\-di anywe\-bwa ba\-su\-mbi ba wa\-ra\-gi.}
\fmtExampleEN{The first crude alcohol during filtering is drunk by brewers.}

\fmtLemma{koo\-ka}
\fmtCitation{(oku)\-koo\-ka}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{koo\-ke\-re}
\fmtSubentry(a)
\fmtSense{to sing}
\fmtExampleRL{Abe\-mbi ba\-ku\-koo\-ka.}
\fmtExampleEN{The musicians are singing.}
\fmtSubentry(b)
\fmtSense{(of a cock) to crow}
\fmtExampleRL{Ompa\-ngu\-zi wa\-nge akoo\-ke\-re emi\-ru\-ndi gi\-ngi olwa\-lee\-ro.}
\fmtExampleEN{My cock has crowed several times today.}
\fmtSubentry(c)
\fmtSense{to break voice (during puberty)}
\fmtExampleRL{Omwo\-jo wa\-nge ata\-ndi\-ki\-re oku\-koo\-ka.}
\fmtExampleEN{My son's voice has started breaking.}

\fmtLemma{koo\-ki}
\fmtCitation{(a)koo\-ki}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{type of small wild bee}
\fmtExampleRL{Obwo\-ki bu\-bba bwa\-ndi.}
\fmtExampleEN{Small wild bees are very aggressive.}

\fmtLemma{koo\-ko}
\fmtCitation{(o)koo\-ko}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{cocoa}
\fmtExampleRL{E Mbaa\-le ba\-lu\-ma okoo\-ko.}
\fmtExampleEN{Cocoa is grown in Mbale.}

\fmtLemma{koo\-ko\-la}
\fmtCitation{(oku)\-koo\-ko\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{koo\-koi\-re}
\fmtSubentry(a)
\fmtSense{to pull down, to bend down, to lower}
\fmtExampleRL{Koo\-ko\-la ei\-taa\-gi lya mu\-saa\-le.}
\fmtExampleEN{Pull down the tree branch.}
\fmtSubentry(b)
\fmtSense{to tear off a piece, to break off}
\fmtExampleRL{Abaa\-na abaa\-ya\-kai\-re awo baa\-koo\-koi\-re enda\-ga\-la za bi\-go\-go bya\-nge.}
\fmtExampleEN{The children who passed there partially broke the leaves of my bananas.}

\fmtLemma{koo\-ko\-li\-yoo\-ko}
\fmtPOS{ideo}
\fmtSense{cock-a-doodle-doo (sound of a cock crowing)}
\fmtExampleRL{Ompa\-ngu\-zi akoo\-ka ‘koo\-ko\-li\-yoo\-ko\-’.}
\fmtExampleEN{The cock crows cock-a-doodle-doo.}

\fmtLemma{koo\-ko\-lo}
\fmtCitation{(o)koo\-ko\-lo}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{cancer}
\fmtExampleRL{Okoo\-ko\-lo bu\-lwai\-re bu\-lu\-ma mu\-no.}
\fmtExampleEN{Cancer is a very painful disease.}

\fmtLemma{koo\-koo\-ma}
\fmtCitation{(oku)\-koo\-koo\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{koo\-koo\-me\-re}
\fmtSense{to shout, to scream, to call out}
\fmtExampleRL{Naa\-ni oyo aku\-ko\-koo\-ma?}
\fmtExampleEN{Who is that shouting?}

\fmtLemma{koo\-koo\-nya}
\fmtCitation{(oku)\-koo\-koo\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{koo\-koo\-nye\-re}
\fmtSense{to tantalise, to make a gesture of offering something without intending to do so, to tease by pretending to give}
\fmtExampleRL{Omu\-ka\-li oyo aku\-nkoo\-koo\-nya.}
\fmtExampleEN{That woman is teasing me.}

\fmtLemma{koo\-ko\-to\-zi}
\fmtCitation{(o)koo\-ko\-to\-zi}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{pied crow (\textit{Corvus albus})}
\fmtExampleRL{Okoo\-ko\-to\-zi ayi\-sa\-na ku\-sai.}
\fmtExampleEN{The pied crow is a nice-looking bird.}

\fmtLemma{koo\-la}
\fmtCitation{(oku)\-koo\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{koi\-re}
\fmtSense{to cough, to whoop}
\fmtExampleRL{Lwa\-ki oku\-koo\-la otyo?}
\fmtExampleEN{Why do you cough like that?}

\fmtLemma{koo\-me\-ra}
\fmtCitation{(oku)\-koo\-me\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{koo\-mei\-re}
\fmtSubentry(a)
\fmtSense{to keep (food) in a traditional oven}
\fmtExampleRL{Koo\-me\-ra eme\-re yai\-swe eta\-sii\-si\-ka\-la.}
\fmtExampleEN{Keep the food in the traditional oven so that it does not go bad.}
\fmtSubentry(b)
\fmtSense{to make a fire in the shed of domestic animals to repel flies and other insects, to make bonfi\-re}

\fmtLemma{koo\-mi}
\fmtCitation{(e)koo\-mi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{bonfire, smouldering fi\-re}
\fmtExampleRL{Aba\-laa\-lo ba\-ki\-rya ente zaa\-bwe ekoo\-mi omu ki\-raa\-ro.}
\fmtExampleEN{Pull down the tree branch.}

\fmtLemma{koo\-na}
\fmtCitation{(e)koo\-na}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{corner, sharp bend}
\fmtExampleRL{Ya\-ba mpo\-la omu mai\-so ali\-yo ekoo\-na.}
\fmtExampleEN{Go slowly, there is a corner.}

\fmtLemma{koo\-ni\-mba}
\fmtCitation{(oku)\-koo\-ni\-mba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{koo\-ni\-mbi\-re}
\fmtSense{to be malnourished}
\fmtExampleRL{Omwa\-na akoo\-ni\-mbi\-re mu\-no.}
\fmtExampleEN{The child is very malnourished.}

\fmtLemma{koo\-no\-ka}
\fmtCitation{(oku)\-koo\-no\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{koo\-no\-ke\-re}
\fmtSense{(of a banana plant) to break partially, to be dislocated, to droop, to bend over, to snap}
\fmtExampleRL{Ei\-too\-ke lya\-nge li\-koo\-no\-ke\-re.}
\fmtExampleEN{My banana plant is bent.}

\fmtLemma{koo\-no\-la}
\fmtCitation{(oku)\-koo\-no\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{koo\-noi\-re}
\fmtSense{to remove, to pick}
\fmtExampleRL{Ya\-ba okoo\-no\-le olu\-la\-ga\-la olee\-te.}
\fmtExampleEN{Go pick a banana leaf and bring it.}

\fmtLemma{koo\-ro}
\fmtCitation{(a)koo\-ro}
\fmtPOS{n}%
\fmtNC{12}
\fmtSense{greed}
\fmtExampleRL{Omwa\-na ali\-na akoo\-ro.}
\fmtExampleEN{The child is greedy. (lit.\,The child has greed).}

\fmtLemma{koo\-si}
\fmtCitation{(e)koo\-si}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{course, class}
\fmtExampleRL{Aku\-ko\-la koo\-si ya bu\-so\-me\-sya.}
\fmtExampleEN{He is taking a teaching course.}

\fmtLemma{koo\-ti}%
\fmtOne{1}
\fmtCitation{(e)koo\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{coat}
\fmtExampleRL{Omu\-sai\-za azwai\-re ekoo\-ti.}
\fmtExampleEN{The man is wearing a coat.}

\fmtLemma{koo\-ti}%
\fmtOne{2}
\fmtCitation{(e)koo\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{court}
\fmtExampleRL{Omu\-la\-mu\-li ayi\-kai\-re omu koo\-ti.}
\fmtExampleEN{The judge is sitting in the court.}

\fmtLemma{koo\-ti}%
\fmtOne{3}
\fmtPOS{interj}
\fmtSense{without me, leave me out of this (an expression used to show that one does not want to be associated with someone or something)}
\fmtExampleRL{Koo\-ti ka mba\-zwe\-re.}
\fmtExampleEN{Without me! Let me leave them.}

\fmtLemma{koo\-to}
\fmtCitation{\textnormal{(e.g.}\,mu\-koo\-to, gi\-koo\-to, bi\-koo\-to\textnormal{)}}
\fmtPOS{adj}
\fmtSense{big, large, long}
\fmtExampleRL{Eki\-ba\-nja kya\-nge ki\-koo\-to.}
\fmtExampleEN{My piece of land is big.}

\fmtLemma{koo\-wa}
\fmtCitation{(a)koo\-wa}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{bow-like device used to remove water from a boat, scoop}
\fmtExampleRL{Lee\-ta akoo\-wa two\-we amai\-zi omu bwa\-to.}
\fmtExampleEN{Bring the scoop so that we can remove water from the boat.}

\fmtLemma{koo\-ya}
\fmtCitation{(a)koo\-ya}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSense{nasal hair, nose hair}
\fmtExampleRL{Akoo\-ya ka mu nyi\-ndo oke\-yi\-ya we\-nkai.}
\fmtExampleEN{Nasal hair can only be removed by the person himself. (a proverb)}

\fmtLemma{koo\-yo}
\fmtCitation{(a)koo\-yo}
\fmtPOS{n}%
\fmtNC{12/14}
\fmtSubentry(a)
\fmtSense{longing, craving, yearning}
\fmtExampleRL{Nkuu\-ra akoo\-yo ka ku\-bo\-na oku ba\-ntu ba\-nge.}
\fmtExampleEN{I feel a longing to see my people.}
\fmtSubentry(b)
\fmtSense{banana leaf sprout}
\fmtExampleRL{Akoo\-yo ka ki\-go\-go ka\-wu\-lu\-ki\-re\-yo.}
\fmtExampleEN{A banana leaf sprout has come out.}

\fmtLemma{ko\-pa}
\fmtCitation{(oku)\-ko\-pa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-pe\-re}
\fmtSource{from En.}
\fmtSense{to copy, to imitate}
\fmtExampleRL{Oyo ata\-ka mu\-no oku\-ko\-pa abai\-raa\-mwe.}
\fmtExampleEN{That one is fond of copying other people's work.}

\fmtLemma{ko\-po\-la}
\fmtCitation{(oku)\-ko\-po\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-poi\-re}
\fmtSubentry(a)
\fmtSense{to cut, to slash}
\fmtExampleRL{Ako\-poi\-re omwi\-raa\-mwe enso\-ne.}
\fmtExampleEN{He cut his friend with a knife.}
\fmtSubentry(b)
\fmtSense{to slander, to tell lies about, to speak ill of, to sully one's reputation}
\fmtExampleRL{Wa\-mu\-boi\-ne ng’a\-ku\-ko\-la atyo\-? Lwa\-ki oku\-mu\-ko\-po\-la\-ku ebi\-ga\-mbo?}
\fmtExampleEN{Did you see her do it? Why are you slandering her?}
\fmtSubentry(c)
\fmtSense{to use obscene words, to swear, to curse}
\fmtExampleRL{Lwa\-ki oku\-ko\-po\-la ebi\-ga\-mbo omu mai\-so g’a\-ba\-ntu aba\-ku\-lu?}
\fmtExampleEN{Why are you using obscene words in the presence of your elders?}

\fmtLemma{ko\-po\-lya (a)mai\-so}
\fmtCitation{(oku)\-ko\-po\-lya amai\-so}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-poi\-rye}
\fmtSense{to blink (one's eyes) repeatedly in a way that shows that one is guilty}

\fmtLemma{ko\-pwe}
\fmtCitation{(o)ko\-pwe}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSubentry(a)
\fmtSense{person who blinks abnormally}
\fmtExampleRL{Omu\-sai\-za oko\-pwe ai\-zi\-re.}
\fmtExampleEN{The man who blinks abnormally has come.}
\fmtSubentry(b)
\fmtSense{person who talks ill of others}
\fmtExampleRL{Oko\-pwe abba mu\-ntu mu\-bbi.}
\fmtExampleEN{A person who talks ill of others is a bad person.}

\fmtLemma{ko\-pwei}
\fmtPOS{ideo}
\fmtSense{like this}
\fmtExampleRL{Ya\-mu\-te\-me\-re eki\-so ko\-pwei.}
\fmtExampleEN{He cut him with a panga like this.}

\fmtLemma{ko\-ro}
\fmtCitation{(o)ko\-ro}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{goat}
\fmtExampleRL{Ya\-mpai\-re engai\-si ya ko\-ro omwei.}
\fmtExampleEN{As a fine, he paid me one male goat.}
\fmtSense{male goat, billy goat}
\fmtExampleRL{Ya\-mpai\-re engai\-si ya ko\-ro omwei.}
\fmtExampleEN{As a fine, he paid me one male goat.}

\fmtLemma{ko\-ro\-ga}
\fmtCitation{(oku)\-ko\-ro\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-ro\-ge\-re}
\fmtSense{to stir, to mix, to stir up}
\fmtExampleRL{Omwa\-na waa\-mu ali adi aku\-ko\-ro\-ga buu\-gi.}
\fmtExampleEN{Your child is stirring the maize porridge there.}

\fmtLemma{ko\-roo\-ta}
\fmtCitation{(oku)\-ko\-roo\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ko\-roo\-te\-re}
\fmtSubentry(a)
\fmtSense{to wheeze, to breathe with a whistling or rattling sound in the chest, to as a result of obstruction in the air passages}
\fmtExampleRL{Omu\-sai\-za aku\-ko\-roo\-ta nk’o\-mbe\-re\-ge.}
\fmtExampleEN{The man is wheezing like a pig.}
\fmtSubentry(b)
\fmtSense{(of a hen) to crow because of difficulty in swallowing}
\fmtExampleRL{Onko\-ko waa\-mu aku\-ko\-roo\-ta.}
\fmtExampleEN{Your hen is crowing as a result of having difficulty in swallowing.}
\fmtSubentry(c)
\fmtSense{to moan, to groan (in intense pain)}

\fmtLemma{ko\-ru}
\fmtCitation{(o)ko\-ru}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{slug}
\fmtExampleRL{Oko\-ru ata\-mbu\-la mpo\-la mu\-no.}
\fmtExampleEN{The slug moves very slowly.}

\fmtLemma{ko\-to\-lya}
\fmtCitation{(oku)\-ko\-to\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-toi\-rye}
\fmtSense{(of a hen) to cackle signalling the presence of food}
\fmtExampleRL{Onko\-ko aku\-ko\-to\-lya bwa\-na ku me\-re.}
\fmtExampleEN{The hen is cackling to signal the presence of food to her chicks.}

\fmtLemma{ko\-yi}
\fmtPOS{interj}
\fmtSubentry(a)
\fmtSense{expression of surprise}
\fmtExampleRL{Ale ko\-yi oku\-ku\-bo\-na e wa\-nge ani.}
\fmtExampleEN{What a surprise to see you at my place!}
\fmtSubentry(b)
\fmtSense{no way, not in the world}
\fmtExampleRL{Nje okwi\-ra\-mu okwa\-ba ewa mu\-ka\-li oyo, ko\-yi!}
\fmtExampleEN{Me, to go back to that woman's place? No way!}
\fmtSubentry(c)
\fmtSense{expression of fear, disgust}
\fmtExampleRL{COVID-19 bulwaire koyi.}
\fmtExampleEN{The COVID-19 disease is terrible.}

\fmtLemma{ko\-zo\-la}
\fmtCitation{(oku)\-ko\-zo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-zoi\-re}
\fmtSubentry(a)
\fmtSense{to strip, to strip off bark (from a tree), to peel (skin), to skin, to flay}
\fmtExampleRL{Ko\-zo\-la omu\-saa\-le gu\-di.}
\fmtExampleEN{Strip off the bark from that tree.}
\fmtSubentry(b)
\fmtSense{to have (the rope around the neck) loosened by itself}
\fmtExampleRL{Onte ako\-zoi\-re omu\-gwa.}
\fmtExampleEN{The cow loosened the rope from around its neck itself.}
\fmtSubentry(c)
\fmtSense{to pull off (\eg some of the grass from a grass-thatched house in order to use it to light fire)}
\fmtExampleRL{Ko\-zo\-la\-ku ei\-su\-bi oku nyu\-mba eyo.}
\fmtExampleEN{Pull off some of the grass from a grass-thatched house.}

\fmtLemma{ko\-zya}
\fmtCitation{(oku)\-ko\-zya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ko\-ze\-rye}
\fmtSense{to dip (food into sauce)}
\fmtExampleRL{Ko\-zya eme\-re omu mu\-ku\-bi.}
\fmtExampleEN{Dip the food into the sauce.}

\fmtLemma{Kri\-sto}
\fmtCitation{(o)Kri\-sto}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{Christ}
\fmtExampleRL{OKri\-sto mwa\-na wa Ka\-nca.}
\fmtExampleEN{Christ is the son of God.}

\fmtLemma{ku}%
\fmtOne{1}
\fmtCitation{oku}
\fmtPOS{prep}
\fmtSubentry(a)
\fmtSense{at}
\fmtExampleRL{Mba ku woo\-tee\-ri ya Lu\-na\-r.}
\fmtExampleEN{I live at Lunar hotel.}
\fmtSubentry(b)
\fmtSense{on, onto}
\fmtExampleRL{Esaa\-ni ngi\-bi\-kii\-rye oku nsu\-wa.}
\fmtExampleEN{I kept the plate on top of the pot.}
\fmtSubentry(c)
\fmtSense{about, with reference to, referring to, with regard to}
\fmtExampleRL{Oku\-ba\-za ku naa\-ni?}
\fmtExampleEN{Whom are you talking about?}
\fmtSubentry(d)
\fmtSense{over, above}
\fmtExampleRL{Ama\-tai ga\-bi\-mbi\-re ni ga\-se\-nse\-ka oku se\-fu\-li\-ya.}
\fmtExampleEN{The milk has boiled over the saucepan.}

\fmtLemma{ku-}%
\fmtOne{2}
\fmtPOS{pf}
\fmtSense{object agreement prefix of the second person singular on verbs}

\fmtLemma{ku-}%
\fmtOne{3}
\fmtPOS{pf}
\fmtSense{subject and object agreement prefix of class 15 on verbs}

\fmtLemma{ku-}%
\fmtOne{4}
\fmtPOS{pf}
\fmtSense{subject agreement prefix of class 17 on verbs}

\fmtLemma{ku-}%
\fmtOne{5}
\fmtPOS{pf}
\fmtSense{progressive aspect prefix on verbs}

\fmtLemma{ku\-ba}
\fmtCitation{ku\-ba}
\fmtPOS{conj}
\fmtSense{because}
\fmtExampleRL{Ya\-ki\-re ma\-ngu ku\-ba amai\-te okwi\-ru\-ka.}
\fmtExampleEN{He arrived early because he knows how to run.}

\fmtLemma{ku\-ba\-gi\-za}
\fmtCitation{(oku)\-ku\-ba\-gi\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ba\-gii\-ze}
\fmtSense{to console, to comfort, to sympathise with}
\fmtExampleRL{Ka nya\-be nku\-ba\-gi\-ze oNa\-mpi\-ja oku\-fee\-rwa omwa\-la.}
\fmtExampleEN{Let me go and console Nampija on the occasion of the loss of her daughter.}

\fmtLemma{ku\-ba\-la}
\fmtCitation{(o)ku\-ba\-la}
\fmtPOS{n}%
\fmtNC{15}
\fmtSubentry(a)
\fmtSense{mathematics, maths}
\fmtExampleRL{Abee\-gi aba\-si\-nga ti\-be\-ndya iso\-mo lya ku\-ba\-la.}
\fmtExampleEN{The majority of students don’t like mathematics.}
\fmtSubentry(b)
\fmtFixedexpression{ku\-ba\-la ba\-ntu}
\fmtSense{census}
\fmtExampleRL{Nkwa\-ba ku\-ba\-la ba\-ntu.}
\fmtExampleEN{I am going to conduct a census.}

\fmtLemma{ku\-ba\-ndwa}
\fmtCitation{(o)ku\-ba\-ndwa}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{traditional worshiping}
\fmtExampleRL{Oku\-ba\-ndwa ku\-gai\-ta Aba\-nya\-la n’Aba\-ruu\-li.}
\fmtExampleEN{Traditional worshipping brings together both the Banyala and the Baruuli.}

\fmtLemma{ku\-ba\-nga}
\fmtCrossreference{\textnormal{see also} kuba and kubba}
\fmtPOS{conj}
\fmtSense{because}
\fmtExampleRL{Abba\-si\-rye n’a\-wee\-ra\-yo ku\-ba\-nga ya\-kee\-rwe mu\-no.}
\fmtExampleEN{She slept deeply because she was too tired.}

\fmtLemma{ku\-ba\-ti\-za}
\fmtCitation{(o\-k)u\-ba\-ti\-za}
\fmtPOS{n}%
\fmtNC{15}
\fmtSource{from En.}
\fmtSense{baptism}

\fmtLemma{ku\-bba}
\fmtCitation{(oku)\-ku\-bba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-bbi\-re}
\fmtSubentry(a)
\fmtSense{to beat, to hit, to punch}
\fmtExampleRL{Nkwi\-za ku\-ku\-bba ni wai\-ra\-mu okwe\-ya\-ka\-lya otyo omu\-ru\-ndi ogu\-ndi.}
\fmtExampleEN{I will beat you if you happen to behave like that next time.}
\fmtSubentry(b)
\fmtSense{to attract, to draw, to lure}
\fmtExampleRL{Omwa\-la oyo aku\-bba mu\-no aba\-sai\-za.}
\fmtExampleEN{That girl is so attractive to men.}
\fmtSubentry(c)
\fmtSense{to drench, to soak, to wet through}
\fmtExampleRL{Oi\-ke\-ndi aku\-bbi\-re abaa\-na.}
\fmtExampleEN{The rain has drenched the children.}
\fmtSubentry(d)
\fmtSense{to draw, to make a drawing, to sketch}
\fmtExampleRL{Ku\-bba ani ekii\-sa\-ni kya mbwe\-ne.}
\fmtExampleEN{Draw here a picture of a dog.}
\fmtSubentry(e)
\fmtSense{to polish (\eg shoes)}
\fmtExampleRL{Ku\-bba enkai\-to yaa\-mu!}
\fmtExampleEN{Polish your shoes!}
\fmtSubentry(f)
\fmtSense{to inject}
\fmtExampleRL{Onte mmaa\-re oku\-mu\-ku\-bba enki\-nzo.}
\fmtExampleEN{I have finished injecting the cow.}
\fmtSubentry(g)
\fmtSense{to photograph, to take a photo, to take a picture}
\fmtExampleRL{Omu\-sai\-za oyo aku\-ku\-bba bii\-sa\-ni.}
\fmtExampleEN{That man is taking photos.}
\fmtSubentry(h)
\fmtSense{(of a saucepan) to heat up}
\fmtExampleRL{Eki\-na\-ga ki\-ku\-bbi\-re.}
\fmtExampleEN{The saucepan is heated up.}
\fmtSubentry(i)
\fmtFixedexpression{ku\-bba (a)ka\-le\-ka}
\fmtSense{(of leaves) to become yellowish (a sign that they are ready to be harvested)}
\fmtExampleRL{Amai\-do ga\-ku\-bbi\-re aka\-le\-ka.}
\fmtExampleEN{The groundnuts' leaves are yellowish.}
\fmtSubentry(j)
\fmtFixedexpression{ku\-bba (e)kya\-la}
\fmtSense{to initiate marriage talks, to start marriage talks}
\fmtExampleRL{Ya\-ku\-bbi\-re ekya\-la aka\-ti wa Bre\-nda n’o\-To\-m.}
\fmtExampleEN{He initiated the marriage talks between Brenda and Tom.}
\fmtSubentry(k)
\fmtFixedexpression{ku\-bba (a)kai\-so}
\fmtSense{to stare at romantically}
\fmtExampleRL{OYu\-su\-f aku\-bbi\-re oDe\-bu\-la akai\-so.}
\fmtExampleEN{Yusuf is romantically staring at Deborah.}
\fmtSubentry(l)
\fmtLemma{ku\-bba (e)mpo\-do}%
\fmtFixedexpression{ku\-bba (e)mpo\-do}
\fmtSense{to expose oneself indecently}
\fmtExampleRL{Omu\-sai\-za ya\-tu\-ku\-bbi\-re mpo\-do nga n’a\-baa\-na ba\-ku\-bo\-na.}
\fmtExampleEN{The man exposed himself indecently to us in the presence of children.}

\fmtLemma{ku\-bba}%
\fmtOne{2}
\fmtCrossreference{\textnormal{see} kuba}

\fmtLemma{ku\-bbai\-gu\-lya}
\fmtCitation{(oku)\-ku\-bbai\-gu\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-bbai\-gwi\-rye}
\fmtSense{to dust off}
\fmtExampleRL{Ku\-bbai\-gu\-lya\-ku ocuu\-cu ali oku ki\-ta\-bo kya\-mu ekya ke\-mi\-si\-tu\-le ota\-ndi\-ke oku\-so\-ma.}
\fmtExampleEN{Dust off the old chemistry book and start studying.}

\fmtLemma{ku\-bba\-nga\-nya}
\fmtCitation{(oku)\-ku\-bba\-nga\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-bba\-ngai\-nye}
\fmtSense{to discuss, to talk about}
\fmtExampleRL{Mwi\-ze tu\-ku\-bba\-nga\-nye ebi\-ro\-woo\-zo.}
\fmtExampleEN{Come and let's discuss ideas.}

\fmtLemma{ku\-bbi}
\fmtPOS{adv}
\fmtSubentry(a)
\fmtSense{badly, naughtily, disobediently}
\fmtExampleRL{Omwo\-jo oni yee\-ya\-ka\-lya ku\-bbi.}
\fmtExampleEN{This young man behaves badly.}
\fmtSubentry(b)
\fmtSense{with a habit of battering, with a habit of beating}
\fmtExampleRL{Omu\-sai\-za odi mu\-ku\-bbi wa ba\-ka\-li.}
\fmtExampleEN{That man has a habit of beating women.}

\fmtLemma{ku\-bbi\-ra\-wo}
\fmtCitation{(oku)\-ku\-bbi\-ra\-wo}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-bbi\-re\-wo}
\fmtSubentry(a)
\fmtSense{to throw down heavily}
\fmtExampleRL{Omu\-wo\-go ya\-nde\-me\-rei\-re ne mmu\-ku\-bbi\-ra\-wo.}
\fmtExampleEN{The cassava was heavy for me and I threw it down heavily.}
\fmtSubentry(b)
\fmtFixedexpression{ku\-bbi\-ra\-wo (e)ki\-ga\-mbo}
\fmtSense{to utter a profanity, to swear, to curse}
\fmtExampleRL{Omu\-ka\-li ya\-ku\-bbi\-re\-wo eki\-ga\-mbo n'e\-twe\-ri\-ndu\-ka.}
\fmtExampleEN{A woman used a swear word, and we were surprised.}

\fmtLemma{ku\-bbi\-rya}
\fmtCitation{(oku)\-ku\-bbi\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-bbii\-rye}
\fmtSense{to close (a hole in the fence)}
\fmtExampleRL{Ku\-bbi\-rya olu\-ko\-ro\-go olwo.}
\fmtExampleEN{Fix the hole in the fence.}

\fmtLemma{ku\-bbi\-rya\-wo}
\fmtCitation{(oku)\-ku\-bbi\-rya\-wo}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-bbii\-rye\-wo}
\fmtSense{to clear weeds especially on water}
\fmtExampleRL{Nku\-bbi\-rya\-wo eki\-te\-nge\-ja ekyo oku mai\-zi.}
\fmtExampleEN{Kindly clear that water lettuce from the water for me.}

\fmtLemma{ku\-bbu ku\-bbu}
\fmtPOS{interj}
\fmtSense{really, exactly}
\fmtExampleRL{Oni ni\-ye maa\-ma eya\-mbya\-yi\-re ku\-bbu ku\-bbu.}
\fmtExampleEN{This is my real mother.}

\fmtLemma{ku\-bbu\-la}%
\fmtOne{1}
\fmtCitation{(oku)\-ku\-bbu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-bbwi\-re}
\fmtSubentry(a)
\fmtSense{to press hard on a belly of a pregnant woman}
\fmtExampleRL{Ozei\-za ya\-ku\-bbwi\-re omu\-ka\-li.}
\fmtExampleEN{Grandmother pressed the woman's belly.}
\fmtSubentry(b)
\fmtSense{to give birth}
\fmtExampleRL{Nku\-ku\-la\-yii\-ri\-ra omaa\-ma eya\-nku\-bbwi\-re.}
\fmtExampleEN{I swear by my mother who gave birth to me.}
\fmtSubentry(c)
\fmtSense{to remove (mingled food) from a saucepan or pot to a serving basket}
\fmtExampleRL{Ku\-bbu\-la aka\-wu\-nga ka baa\-na oku saa\-ni.}
\fmtExampleEN{Put the children's maize paste onto the plate.}

\fmtLemma{ku\-bbu\-la}%
\fmtOne{2}
\fmtCitation{(o\-k)u\-bbu\-la}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{haemorrhage}
\fmtExampleRL{Omu\-ka\-li alwai\-re eku\-bbu\-la.}
\fmtExampleEN{The woman is suffering from haemorrhaging.}

\fmtLemma{ku\-bi\-gi\-za}
\fmtCrossreference{\textnormal{see also} kubiriza}
\fmtCitation{(oku)\-ku\-bi\-gi\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-bi\-gi\-zi\-re}
\fmtSense{to chair (a meeting)}
\fmtExampleRL{Wee\-we, nku\-bi\-gi\-zaa\-ku oru\-ku\-ra\-to.}
\fmtExampleEN{Chair the meeting for me please.}

\fmtLemma{ku\-bi\-ri\-za}
\fmtCrossreference{\textnormal{see also} kubigiza}
\fmtCitation{(oku)\-ku\-bi\-ri\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-bi\-rii\-ze}
\fmtSubentry(a)
\fmtSense{to chair (a meeting)}
\fmtExampleRL{Wee\-we, nku\-bi\-ri\-zaa\-ku oru\-ku\-ra\-to.}
\fmtExampleEN{Chair the meeting for me please.}
\fmtSubentry(b)
\fmtSense{to encourage, to persuade}
\fmtExampleRL{Ku\-bi\-ri\-za abaa\-na abo ba\-ta\-ya\-ba.}
\fmtExampleEN{Encourage those children not to go.}

\fmtLemma{ku\-co\-ndoo\-za}
\fmtCitation{(o)ku\-co\-ndoo\-za}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{investigation}
\fmtExampleRL{We nii\-we wa\-koo\-re ogwa ku\-co\-ndoo\-za omu\-sa\-ngo gwa aba\-vu\-gi ba boo\-da\-boo\-da aba\-kwi\-ta aba\-ntu?}
\fmtExampleEN{Is it you who did the investigation of the killings by the bodaboda driver?}

\fmtLemma{ku\-cu\-ngu\-rwa}
\fmtCitation{(o)ku\-cu\-ngu\-rwa}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{salvation}
\fmtExampleRL{Oku\-cu\-ngu\-rwa kwa mu\-ntu kwa\-zwi\-re oku Ye\-su Kri\-sto.}
\fmtExampleEN{The salvation of man came through Jesus Christ.}

\fmtLemma{ku\-da}
\fmtCitation{(oku)\-ku\-da}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-di\-re}
\fmtSense{to float}
\fmtExampleRL{Eki\-ke\-re eki\-fu ki\-ku\-ku\-da ai\-gu\-ru oku mai\-zi.}
\fmtExampleEN{A dead frog is floating on the water.}

\fmtLemma{ku\-daa\-la}
\fmtCitation{(oku)\-ku\-daa\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-dai\-re}
\fmtSense{to mock, to ridicule, to scorn}
\fmtExampleRL{Lwa\-ki oku\-ku\-daa\-li\-ra omwa\-na wa\-nge bwo\-tyo?}
\fmtExampleEN{Why are you making fun of my child like that?}

\fmtLemma{ku\-daa\-li}
\fmtCitation{\textnormal{(e.g.}\,mu\-ku\-daa\-li, ba\-ku\-daa\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSense{boastful, overbearing}
\fmtExampleRL{Na\-bbai\-re n’o\-mu\-sai\-za omu\-ku\-daa\-li oku\-ma\-la emya\-ka isa\-tu.}
\fmtExampleEN{I lived with a boastful person for almost three years.}

\fmtLemma{ku\-di}
\fmtCrossreference{\textnormal{see} (-)di}

\fmtLemma{ku\-dii\-ri\-sya}
\fmtCitation{(oku)\-ku\-dii\-ri\-sya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-dii\-ri\-si\-rye}
\fmtSense{to be stingy, to be a miser, to be a penny-pincher}
\fmtExampleRL{Omu\-sai\-za oyo adi\-rii\-sya mu\-no ta\-guu\-ra ba\-ntu ba\-ma\-ka gaa\-mu me\-re nga ate ali\-na empii\-ya zi\-ngi.}
\fmtExampleEN{That man is a miser, he does not even buy food for his family, yet he has money.}

\fmtLemma{ku\-du\-ka}
\fmtCitation{(oku)\-ku\-du\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-du\-ki\-re}
\fmtSense{(of eyes) to protrude}
\fmtExampleRL{Omu\-ntu n’a\-gwa omu nya\-nja aku\-du\-ka amai\-so.}
\fmtExampleEN{When a person drowns in water, his eyes protrude.}

\fmtLemma{ku\-du\-ka\-na}
\fmtCitation{(o)ku\-du\-ka\-na}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{diarrhoea}
\fmtExampleRL{Omwa\-na alwai\-re ku\-du\-ka\-na.}
\fmtExampleEN{The child is suffering from diarrhoea.}

\fmtLemma{ku\-du\-la}
\fmtCitation{(oku)\-ku\-du\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-dwi\-re}
\fmtSense{to pluck out, to pull out}
\fmtExampleRL{Eya\-gwi\-re omu nya\-nja ebi\-na\-ge bi\-mu\-ku\-dwi\-re\-mu amai\-so.}
\fmtExampleEN{The tilapia plucked out the eyes of the person who drowned.}

\fmtLemma{ku\-du\-ma\-la}
\fmtCitation{(oku)\-ku\-du\-ma\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-du\-mai\-re}
\fmtSense{to protrude, to stick out}
\fmtExampleRL{Ebi\-yaa\-ta bi\-ku\-du\-mai\-re omu ita\-ka\-li.}
\fmtExampleEN{The sweet potatoes are protruding from the soil.}

\fmtLemma{ku\-du\-ma\-lya}
\fmtCitation{(oku)\-ku\-du\-ma\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-du\-mai\-rye}
\fmtSense{to bend, to bow, to arch}
\fmtExampleRL{Obwoi\-ne n’a\-ku\-ku\-du\-ma\-lya omu\-go\-ngo!}
\fmtExampleEN{Do you see how he is bending his back!}

\fmtLemma{ku\-du\-mba\-la}
\fmtCitation{(oku)\-ku\-du\-mba\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-du\-mbai\-re}
\fmtSense{to be disobedient}
\fmtExampleRL{Mwa\-na\-we ota\-ndi\-ki\-re oku\-ku\-du\-mba\-la, ti\-na\-ku\-so\-bo\-le.}
\fmtExampleEN{I cannot bear to stay with you since you have started being disobedient.}

\fmtLemma{ku\-du\-mu\-ka}
\fmtCitation{(oku)\-ku\-du\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-du\-mu\-ki\-re}
\fmtSense{to swell, to become large}
\fmtExampleRL{ONsa\-mba ya\-ku\-ku\-du\-mu\-ki\-re omu\-twe ni ya\-gwi\-re oku ka\-be\-nje.}
\fmtExampleEN{Nsa\-mba's head was swollen after he was involved in the accident.}

\fmtLemma{ku\-du\-ngu\-la}
\fmtCitation{(oku)\-ku\-du\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-du\-ngwi\-re}
\fmtSense{to mingle}
\fmtExampleRL{Ku\-du\-nguu\-ra aba\-ge\-ni eme\-re ba\-lye.}
\fmtExampleEN{Mingle millet for the visitors to eat.}

\fmtLemma{ku\-fu\-ka\-na}
\fmtCitation{(o)ku\-fu\-ka\-na}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{wrestling}
\fmtExampleRL{Ali\-yo empa\-ka za ku\-fu\-ka\-na oku iru\-gu\-za.}
\fmtExampleEN{There is a wrestling competition in the sub-county.}

\fmtLemma{ku\-fu\-lu}
\fmtCitation{(e)ku\-fu\-lu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Ar. via Sw.}
\fmtSense{lock, padlock}
\fmtExampleRL{Bbwa olwi\-ge n’e\-ku\-fu\-lu!}
\fmtExampleEN{Lock the door with a padlock!}

\fmtLemma{ku\-gai\-ta}
\fmtCitation{(o)ku\-gai\-ta}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{holy matrimony, wedding}
\fmtExampleRL{Aloo\-wo oku\-gai\-ta aba\-go\-le oku\-kwa\-ba omu mai\-so ku ka\-ni\-sa.}
\fmtExampleEN{A wedding is taking place at the church.}

\fmtLemma{ku\-gai\-ti\-bwa}
\fmtCitation{(o)ku\-gai\-ti\-bwa}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{holy matrimony, wedding}
\fmtExampleRL{Oku\-gai\-ti\-bwa kwa\-bwe baa\-ku\-ko\-lei\-re Na\-ka\-so\-ngo\-la.}
\fmtExampleEN{Their wedding was held in Nakasongola.}

\fmtLemma{ku\-ge\-za}
\fmtCrossreference{\textnormal{see also} geza}
\fmtCitation{(o)ku\-ge\-za}
\fmtPOS{adv}
\fmtSense{for example}
\fmtExampleRL{Oku\-ge\-za, aBa\-nya\-la bai\-ngi ba\-si\-moo\-la Lu\-ga\-nda.}
\fmtExampleEN{For example, many Banyala speak Luganda.}

\fmtLemma{ku\-gi\-di\-ma}
\fmtCitation{(oku)\-ku\-gi\-di\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-gi\-di\-mi\-re}
\fmtSense{(of children) to play (especially at night when the moon is bright), to romp, to cavort about}
\fmtExampleRL{Abaa\-na ba\-ta\-ka mu\-no oku\-gi\-di\-ma okwa kwe\-zi.}
\fmtExampleEN{Children especially like playing when the moon is up.}

\fmtLemma{ku\-gi\-ra}
\fmtCitation{(oku)\-ku\-gi\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-gii\-re}
\fmtSubentry(a)
\fmtSense{to restrict, to limit, to hinder, to prohibit}
\fmtExampleRL{OMu\-se\-ve\-ni ya\-ku\-gi\-re aba\-ntu okwa\-ba mu ta\-wu\-ni lwa Co\-ro\-na.}
\fmtExampleEN{Museveni prohibited people from going to town because of the Coronavirus.}
\fmtSubentry(b)
\fmtSense{to fence, to build a fence around}
\fmtExampleRL{Ku\-gi\-ra omu\-si\-ri.}
\fmtExampleEN{Build a fence around the garden.}
\fmtSubentry(c)
\fmtSense{to put a hem or a binding on an item of clothing, to hem}
\fmtExampleRL{Twa\-la esaa\-ti ya\-nge ewa mu\-tu\-ngi ba\-gi\-ku\-gi\-re.}
\fmtExampleEN{Take my shirt to the tailor so that he can put a new hem on it.}

\fmtLemma{ku\-gu}
\fmtCitation{\textnormal{(e.g.}\,mu\-ku\-gu, ba\-ku\-gu\textnormal{)}}
\fmtPOS{adj}
\fmtSense{experienced, knowledgeable}
\fmtExampleRL{Omu\-sai\-za oyo mu\-ku\-gu omu bya bu\-lu\-mi.}
\fmtExampleEN{That man is an experienced agriculturalist.}

\fmtLemma{ku\-gu\-gu\-la}
\fmtCitation{(o)ku\-gu\-gu\-la}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{disagreement, heated argument}
\fmtExampleRL{Oku\-gu\-gu\-la oku\-li omu eka egyo ku\-ku\-zwa ku ki\-ki?}
\fmtExampleEN{What is the cause of the heated argument in that home?}

\fmtLemma{ku\-gu\-gu\-la\-na}
\fmtCitation{(o)ku\-gu\-gu\-la\-na}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{misunderstanding, misreading, false impression}
\fmtExampleRL{Ba\-ku\-bo\-ne\-ka ba\-li\-mu oku\-gu\-gu\-la\-na aka\-ti waa\-bwe.}
\fmtExampleEN{There seems to be some misunderstanding between them.}

\fmtLemma{ku\-gu\-ka}
\fmtCitation{(oku)\-ku\-gu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-gu\-ki\-re}
\fmtSense{to become an expert, to become a professional}
\fmtExampleRL{Aba\-ntu bai\-ngi ba\-ku\-gu\-ki\-re omu ku\-wa\-ndii\-ka enku\-lu\-ze.}
\fmtExampleEN{Many people have become experts in writing dictionaries.}

\fmtLemma{ku\-gu\-ki}
\fmtCitation{\textnormal{(e.g.}\,mu\-ku\-gu\-ki, ba\-ku\-gu\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{expert, professional}
\fmtExampleRL{Tu\-nga omwo\-roo\-li wa baa\-na nga mu\-ku\-gu\-ki.}
\fmtExampleEN{Get a professional caretaker for the children.}

\fmtLemma{ku\-gu\-lu}
\fmtCitation{(o)ku\-gu\-lu}
\fmtPOS{n}%
\fmtNC{15/6}
\fmtSubentry(a)
\fmtSense{foot (of a human), leg}
\fmtExampleRL{Ama\-gu\-lu ga\-nge ga\-zi\-mbi\-re.}
\fmtExampleEN{My feet are swollen.}
\fmtSubentry(b)
\fmtSense{foot (of a bed)}
\fmtExampleRL{Oku\-gu\-lu kwa ki\-ta\-nda kwe\-nde\-ke\-re.}
\fmtExampleEN{The foot of the bed has broken.}
\fmtSubentry(c)
\fmtSense{footstep, track, footprint}
\fmtExampleRL{Mboi\-ne ama\-gu\-lu g’a\-nte omu ndu\-mi\-ro.}
\fmtExampleEN{I have seen animals' tracks in the garden.}

\fmtLemma{ku\-jwi}
\fmtCitation{(o)ku\-jwi}
\fmtPOS{n}%
\fmtNC{15/6}
\fmtSense{knee}
\fmtExampleRL{Ti\-yai\-ru\-ki\-re ku\-bba oku\-jwi kwa\-mwe ku\-lwai\-re.}
\fmtExampleEN{He did not run because his knee is infected.}

\fmtLemma{ku\-ka}
\fmtCitation{(oku)\-ku\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ki\-re}
\fmtSense{to remove (grass from a cleared piece of land)}
\fmtExampleRL{Ku\-ku\-ka ei\-su\-bi oli\-too\-le omu ita\-ka\-li.}
\fmtExampleEN{Remove the grass from the cleared piece of land.}

\fmtLemma{ku\-ku\-la}
\fmtCitation{(oku)\-ku\-ku\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-kwi\-re}
\fmtSense{to develop mould, to become mouldy}
\fmtExampleRL{Omu\-gaa\-ti gu\-ku\-kwi\-re.}
\fmtExampleEN{The bread became mouldy.}

\fmtLemma{ku\-ku\-mba}
\fmtCitation{(oku)\-ku\-ku\-mba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ku\-mbi\-re}
\fmtSense{to empty (\eg a house by robbers)}
\fmtExampleRL{Ba\-ku\-ku\-mbi\-re ebi\-ntu bya\-mwe byo\-na\-byo\-na.}
\fmtExampleEN{They emptied the house of all his property.}

\fmtLemma{ku\-ku\-mu\-ka}
\fmtCitation{(oku)\-ku\-ku\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-ku\-mu\-ki\-re}
\fmtSense{to develop, to grow, to expand}
\fmtExampleRL{OKa\-yu\-nga aku\-ku\-ku\-mu\-ka ma\-ngu.}
\fmtExampleEN{Kayunga is developing fast.}

\fmtLemma{ku\-ku\-mu\-la}
\fmtCitation{(oku)\-ku\-ku\-mu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-mwi\-re}
\fmtSense{to construct, to build}
\fmtExampleRL{Omwa\-na aku\-ku\-mwi\-re enyu\-mba ku mbi\-ro.}
\fmtExampleEN{The child has constructed a house speedily.}

\fmtLemma{ku\-ku\-nu\-ka}
\fmtCitation{(oku)\-ku\-ku\-nu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-ku\-nu\-ki\-re}
\fmtSense{to emerge from a hiding place}
\fmtExampleRL{Aku\-ku\-nu\-ki\-re\-yo n'ai\-ru\-ka.}
\fmtExampleEN{He emerged and ran away.}

\fmtLemma{ku\-ku\-nu\-la}
\fmtCitation{(oku)\-ku\-ku\-nu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ku\-nwi\-re}
\fmtSense{to take out (from a hiding place), to extract (from a hiding place)}
\fmtExampleRL{Amu\-ku\-ku\-nwi\-re\-yo gy'a\-bbai\-re ye\-bbi\-si\-re.}
\fmtExampleEN{He took him out of his hiding place.}

\fmtLemma{ku\-ku\-ta}
\fmtCitation{(oku)\-ku\-ku\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-ku\-ti\-re}
\fmtSense{to act stealthily, to act secretly, to sneak}
\fmtExampleRL{Lwa\-ki oku\-ku\-ta otyo nga okwi\-ngi\-ra omu nyu\-mba?}
\fmtExampleEN{Why are you entering the house stealthily?}

\fmtLemma{ku\-ku\-ti\-ra}
\fmtCitation{(oku)\-ku\-ku\-ti\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ku\-tii\-re}
\fmtSense{(of a cock) to prepare a place where a hen can lay eggs}
\fmtExampleRL{Ompa\-ngu\-zi aku\-ku\-ti\-ra obu\-gu\-ma we ya\-bii\-ka.}
\fmtExampleEN{The cock looks for a place where the hen will lay eggs.}

\fmtLemma{ku\-ku\-tya}
\fmtCitation{(oku)\-ku\-ku\-tya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ku\-ti\-rye}
\fmtSense{to smuggle, to import illegally}
\fmtExampleRL{Mbaa\-ji\-rye nga ba\-ku\-ku\-ku\-tya ebii\-bbe.}
\fmtExampleEN{I met them smuggling stolen property.}

\fmtLemma{ku\-kuu\-la}
\fmtCitation{(oku)\-ku\-kuu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-kui\-re}
\fmtSense{to clean (especially a filthy place), to cleanse, to remove dirt}
\fmtExampleRL{Ya\-ba oku\-kuu\-le eki\-nyu\-mba kya nko\-ko.}
\fmtExampleEN{Go and clean the poultry house.}

\fmtLemma{ku\-la}
\fmtCitation{(oku)\-ku\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kuu\-re}
\fmtSubentry(a)
\fmtSense{to grow, to get bigger}
\fmtExampleRL{Omwa\-na oyo akuu\-re ma\-ngu.}
\fmtExampleEN{That child has grown very fast.}
\fmtSubentry(b)
\fmtSense{to advance, to improve, to make progress}
\fmtExampleRL{Aba\-ntu ba\-kwi\-za ku\-ma\-maa\-la emya\-ka gi\-ngi si\-nga ote\-ki\-no\-lo\-gi\-ya akwa\-ba aku\-la.}
\fmtExampleEN{People will live long as technology advances.}
\fmtSubentry(c)
\fmtSense{(of cereal crops) to form bunches}
\fmtExampleRL{Omu\-we\-mba gu\-kuu\-re ma\-ngu.}
\fmtExampleEN{The sorghum has formed bunches so quickly. (lit.\,The sorghum has grown.)}
\fmtSubentry(d)
\fmtSense{(of butter, ghee, brew) to be ready}
\fmtExampleRL{Ama\-gi\-ta ga\-kuu\-re.}
\fmtExampleEN{The ghee is ready for consumption.}

\fmtLemma{ku\-laa\-ku\-la\-na}
\fmtCitation{(oku)\-ku\-laa\-ku\-la\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-la\-ku\-lai\-ne}
\fmtSense{to develop, to grow, to expand}
\fmtExampleRL{Ekya\-lo kyai\-swe ki\-ku\-laa\-ku\-lai\-ne.}
\fmtExampleEN{Our village has developed.}

\fmtLemma{ku\-lee\-ti}
\fmtCitation{(e)ku\-lee\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{crate}
\fmtExampleRL{Mu\-wee\-rye eku\-lee\-ti za soo\-da itaa\-nu.}
\fmtExampleEN{Give him five crates of soda.}

\fmtLemma{ku\-le\-ku\-la}
\fmtCitation{(o)ku\-le\-ku\-la}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{oyster nut}
\fmtExampleRL{Oku\-le\-ku\-la ala\-ndi\-ra ku mu\-saa\-le.}
\fmtExampleEN{The oyster nut creeps up the tree.}

\fmtLemma{ku\-le\-mbe\-ra}
\fmtCitation{(oku)\-ku\-le\-mbe\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-mbei\-re}
\fmtSense{to lead, to guide, to head}
\fmtExampleRL{Ku\-le\-mbe\-ra\-mu abai\-raa\-wo mwa\-be oku ka\-ni\-sa.}
\fmtExampleEN{Lead your friends and go to the church.}

\fmtLemma{ku\-li\-ga}
\fmtCitation{(o)ku\-lu\-ga}
\fmtPOS{conj}
\fmtSense{since, ever since}
\fmtExampleRL{Oku\-li\-ga ni yaa\-bi\-re e Ka\-mpa\-la tai\-ra\-nga.}
\fmtExampleEN{He has never returned since he went to Kampala.}

\fmtLemma{Ku\-li\-si\-to}
\fmtCitation{(o)Ku\-li\-si\-to}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{Christ, messiah, Jesus}
\fmtExampleRL{Aba\-ku\-li\-si\-taa\-yo ba\-ra\-mya Ku\-li\-si\-to.}
\fmtExampleEN{Christians worship Christ.}

\fmtLemma{ku\-li\-sya}
\fmtCitation{(oku)\-ku\-li\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-li\-si\-rye}
\fmtSense{to raise, to bring up (a child)}
\fmtExampleRL{ONa\-ka\-le\-be gwe\-ye\-si\-gi\-re nti ali\-ku\-li\-sya abaa\-na ba\-mwe mu ma\-zi\-ra aga\-so\-bo\-ro\-ko\-ki ano\-be\-re.}
\fmtExampleEN{The woman whom he trusted to raise his children in a proper way has divorced him.}

\fmtLemma{ku\-lu}
\fmtCitation{(e.g.u\-ku\-lu, ba\-ku\-lu,gi\-ku\-lu)}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{old, elderly, aged, mature}
\fmtExampleRL{Oyo mu\-sai\-za mu\-ku\-lu.}
\fmtExampleEN{That is an old man.}
\fmtSubentry(b)
\fmtSense{elder}
\fmtExampleRL{Oyo ni\-ye mwo\-nyo\-ko wa\-nge omu\-ku\-lu.}
\fmtExampleEN{That is my elder brother.}
\fmtSubentry(c)
\fmtSense{main, major, principal, chief, primary, most important}
\fmtExampleRL{Mpee\-rya enso\-nga egi\-ku\-lu lwa\-ki ti\-wai\-zi\-re.}
\fmtExampleEN{Tell me the main reason why you did not come.}
\fmtSubentry(d)
\fmtSense{great, important}
\fmtExampleRL{Oka\-nca mu\-ku\-lu.}
\fmtExampleEN{God is great.}

\fmtLemma{ku\-lu\-ba\-la}
\fmtCitation{(oku)\-ku\-lu\-ba\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-lu\-bai\-re}
\fmtSense{(of food) to be half-cooked}
\fmtExampleRL{Ebi\-yaa\-ta bi\-ku\-lu\-bai\-re.}
\fmtExampleEN{The sweet potatoes are half-cooked.}

\fmtLemma{ku\-lu\-bu\-ka}
\fmtCitation{(oku)\-ku\-lu\-bu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-lu\-bu\-ki\-re}
\fmtSense{to be scratched}
\fmtExampleRL{Emo\-to\-ka ya\-nge eku\-lu\-bu\-ki\-re.}
\fmtExampleEN{My car is scratched.}

\fmtLemma{ku\-lu\-bu\-la}
\fmtCitation{(oku)\-ku\-lu\-bu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-lu\-bwi\-re}
\fmtSense{to scratch, to scar}
\fmtExampleRL{Owa\-bo\-da\-bo\-da aku\-lu\-bwi\-re emo\-to\-ka ya\-nge.}
\fmtExampleEN{The bodaboda rider has scratched my car.}

\fmtLemma{ku\-lu\-ga}
\fmtCrossreference{\textnormal{see} kuliga}

\fmtLemma{ku\-lu\-ku\-ta}
\fmtCitation{(oku)\-ku\-lu\-ku\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-lu\-ku\-ti\-re}
\fmtSense{(of water) to move, to flow, to run}
\fmtExampleRL{Amai\-zi ga\-ku\-ku\-lu\-ku\-ta mpo\-la.}
\fmtExampleEN{The water is flowing calmly.}

\fmtLemma{ku\-lu\-lu}%
\fmtOne{1}
\fmtCitation{(o)ku\-lu\-lu}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{ostrich}
\fmtExampleRL{Oku\-lu\-lu ni\-kyo ki\-nyo\-ni eki\-si\-nga obu\-koo\-to.}
\fmtExampleEN{The ostrich is the biggest bird.}

\fmtLemma{ku\-lu\-lu}%
\fmtOne{2}
\fmtCitation{(e)ku\-lu\-lu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{small drum}
\fmtExampleRL{Eku\-lu\-lu ba\-gi\-ku\-bi\-sya bu\-saa\-le.}
\fmtExampleEN{The small drum is played using small sticks.}

\fmtLemma{ku\-lu\-mba\-la}
\fmtCitation{(oku)\-ku\-lu\-mba\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-lu\-mbai\-re}
\fmtSense{to jut out, to stick out, to bulge out}
\fmtExampleRL{Onkai\-na abba n’o\-mu\-go\-ngo ogwa\-ku\-lu\-mbai\-re.}
\fmtExampleEN{A camel has a back which juts out.}

\fmtLemma{ku\-lu\-ndu\-lya}
\fmtCrossreference{\textnormal{see} kurundya}
\fmtCitation{(oku)\-ku\-lu\-ndu\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-lu\-ndwi\-rye}
\fmtSense{to drag, to pull, to tug}
\fmtExampleRL{Omwa\-na aku\-ku\-lu\-ndya odo\-le ema\-be\-ga waa\-mu.}
\fmtExampleEN{The child is dragging a toy behind him.}

\fmtLemma{ku\-lu\-pya}
\fmtCitation{\textnormal{(e.g.}\,mu\-ku\-lu\-pya, ba\-ku\-lu\-pya\textnormal{)}}
\fmtPOS{adj}
\fmtSense{(of a person) dishonest, untruthful}
\fmtExampleRL{Omwo\-jo omu\-ku\-lu\-pya ta\-so\-bo\-la ku\-ba\-za ma\-zi\-ma.}
\fmtExampleEN{A dishonest person, he will never tell the truth.}

\fmtLemma{ku\-lya}
\fmtCitation{(oku)\-ku\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kuu\-rye}
\fmtSense{to promote, to give promotion, to advance}
\fmtExampleRL{OZi\-yaa\-di baa\-mu\-kuu\-rye.}
\fmtExampleEN{Ziyaadi was promoted.}

\fmtLemma{ku\-ma}
\fmtCitation{(oku)\-ku\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-mi\-re}
\fmtSense{to make fire, to ignite}
\fmtExampleRL{Ku\-ma omuu\-ro two\-ce ebi\-coo\-li.}
\fmtExampleEN{Make a fire and let's roast the maize.}

\fmtLemma{ku\-ma\-nya\-nga\-na}
\fmtCitation{(o)ku\-ma\-nya\-nga\-na}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{reunion}
\fmtExampleRL{Abaa\-na aba\-so\-mei\-re\-ku e Ga\-yaa\-za ba\-kwa\-ba ku\-ma\-nya\-nga\-na ma\-mbya.}
\fmtExampleEN{The Gayaza old girls are to hold their reunion tomorrow.}

\fmtLemma{ku\-mba}
\fmtCitation{(oku)\-ku\-mba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-mbi\-re}
\fmtSense{to march, to stride}
\fmtExampleRL{Aba\-si\-ri\-ka\-le ba\-ku\-mba bu\-li ma\-mbya.}
\fmtExampleEN{Soldiers march every morning.}

\fmtLemma{ku\-mbaa\-na}
\fmtCitation{(oku)\-ku\-mbaa\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-mbai\-ne}
\fmtSense{to assemble, to gather, to come together, to get together}
\fmtExampleRL{Tu\-kwa\-ba ku\-ku\-mbaa\-ni\-ra ku Lu\-na\-r.}
\fmtExampleEN{We are going to meet at Lunar Guest House.}

\fmtLemma{ku\-mbaa\-nia}
\fmtCitation{(oku)\-ku\-mbaa\-nia}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-mbai\-nie}
\fmtSense{to bring together, to collect, to gather}
\fmtExampleRL{Omu Bu\-nya\-la mwa\-lu\-mu aba\-ka\-ma aba\-koo\-re omu\-li\-mu ogw’a\-maa\-ni oku\-ku\-mbaa\-nia Aba\-nya\-la.}
\fmtExampleEN{In Bu\-nyala there were chiefs who worked hard to bring together all the Banyala people.}

\fmtLemma{ku\-mwei\-ku\-mwei}
\fmtPOS{adv}
\fmtSubentry(a)
\fmtSense{one after the other, close together, close to each other}
\fmtExampleRL{Ama\-nyu\-mba ga\-li ku\-mwei\-ku\-mwei.}
\fmtExampleEN{The houses are close to each other.}
\fmtSubentry(b)
\fmtSense{in a squeezed way, crampedly, overcrowdedly}
\fmtExampleRL{Aba\-ntu bai\-ka\-la ku\-mwei\-ku\-mwei omu ta\-ki\-si ezaa\-ba e Na\-ka\-so\-ngo\-la.}
\fmtExampleEN{People are always squeezed in the taxis which go to Nakasongola.}

\fmtLemma{ku\-naa\-mii\-rya}
\fmtCitation{(oku)\-ku\-naa\-mii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-naa\-mi\-rii\-rye}
\fmtSense{to lean over, to bend over (so that the head is lower than the knees)}
\fmtExampleRL{Mwa\-na we abai\-ra baa\-mu ba\-ku\-lu\-ma, oku\-naa\-mi\-rii\-rye\-mbe?}
\fmtExampleEN{Child, your friends are digging and you are just bending over?}

\fmtLemma{ku\-na\-la}
\fmtCitation{(oku)\-ku\-na\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-nai\-re}
\fmtSubentry(a)
\fmtSense{to protrude, to stick out, to bulge out, to coming out}
\fmtExampleRL{Ebi\-yaa\-ta bi\-ku\-ku\-nai\-re\-mbe omu bi\-ti\-ndyo\-le.}
\fmtExampleEN{The sweet potatoes are just protruding from the mounds.}
\fmtSubentry(b)
\fmtSense{to sleep without covering oneself}
\fmtExampleRL{Omwa\-na aku\-ku\-nai\-re omu ki\-ta\-bu.}
\fmtExampleEN{The child sleeps without covering himself in the bed.}
\fmtSubentry(c)
\fmtSense{to be restless (esp. during sleep), to have a sleepless night}
\fmtExampleRL{Ebi\-ro bi\-ni nsu\-la nku\-ku\-nai\-re.}
\fmtExampleEN{These days I sleep restlessly.}

\fmtLemma{ku\-na\-ma}
\fmtCitation{(oku)\-ku\-na\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-na\-mi\-re, ku\-nai\-me}
\fmtSubentry(a)
\fmtSense{to bend (down), to bow, to arch}
\fmtExampleRL{Aku\-nai\-me n’a\-ko\-ma\-wo eka\-laa\-mu.}
\fmtExampleEN{He bent down and picked a pencil.}
\fmtSubentry(b)
\fmtSense{to expose one’s nakedness}
\fmtExampleRL{Ki\-bbi mu\-no omu\-ntu oku\-ku\-na\-ma.}
\fmtExampleEN{It is bad to expose one's nakedness.}

\fmtLemma{ku\-ndaa\-za}
\fmtCitation{(oku)\-ku\-ndaa\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ndai\-ze, ku\-ndaa\-zi\-re}
\fmtSense{to cover up indecent exposure (\eg of breasts)}

\fmtLemma{ku\-ndu\-ka}
\fmtCitation{(oku)\-ku\-ndu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-ndu\-ki\-re}
\fmtSense{(of eyes) to pop out}
\fmtExampleRL{Amai\-so g’a\-mu\-fu ga\-ku\-ndu\-ki\-re.}
\fmtExampleEN{The eyes of the dead person have popped out.}

\fmtLemma{ku\-ndu\-la}
\fmtCitation{(oku)\-ku\-ndu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ndwi\-re}
\fmtSubentry(a)
\fmtSense{to open wide (one's eyes)}
\fmtExampleRL{Lwa\-ki oku\-ku\-ndu\-la amai\-so nk’o\-ka\-pa?}
\fmtExampleEN{Why are you opening your eyes wide like a cat?}
\fmtSubentry(b)
\fmtFixedexpression{ku\-ndu\-la (a)mai\-so}
\fmtSense{to stare}
\fmtExampleRL{Omwa\-la ba\-ku\-mu\-ko\-ba aku\-ku\-ndu\-la\-mbe mai\-so.}
\fmtExampleEN{They are speaking to the girl, and she is just staring.}

\fmtLemma{ku\-nga}
\fmtCitation{(oku)\-ku\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-ngi\-re}
\fmtSense{to cry, to weep}
\fmtExampleRL{Oku\-ku\-ngi\-ra bwe\-ree\-re.}
\fmtExampleEN{You are crying for no reason.}

\fmtLemma{ku\-ngu\-ba\-ga}
\fmtCitation{(oku)\-ku\-ngu\-ba\-ga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-ngu\-ba\-gi\-re}
\fmtSense{to mourn, to grieve}
\fmtExampleRL{Tu\-kwa\-ba ku\-ku\-ngu\-ba\-gi\-ra mu\-ntu wai\-swe.}
\fmtExampleEN{We are going to mourn our comrade.}

\fmtLemma{ku\-ngu\-ba\-la}
\fmtCitation{(oku)\-ku\-ngu\-ba\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-ngu\-bai\-re}
\fmtSense{to protrude, to be uneven}
\fmtExampleRL{Ye\-te ekyo eki\-yaa\-ta eki\-ku\-ngu\-bai\-re lwa\-ki to\-ki\-swe\-ka?}
\fmtExampleEN{Why don’t you cover that sweet potato which is protruding?}

\fmtLemma{ku\-ngu\-la}
\fmtCitation{(oku)\-ku\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ngwi\-re}
\fmtSense{to harvest}
\fmtExampleRL{Ebi\-ro bya ku\-ku\-ngu\-la bye twa\-lu\-mi\-re bii\-ki\-re.}
\fmtExampleEN{The time has come to harvest what we planted.}

\fmtLemma{ku\-nguu\-ta}
\fmtCitation{(oku)\-ku\-nguu\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ngu\-ti\-re}
\fmtSense{to sieve}
\fmtExampleRL{Abbai\-re aku\-ku\-nguu\-ta bwi\-ta bwa\-mwe oku\-li\-ga ema\-mbya.}
\fmtExampleEN{She has been sieving her millet since morning.}

\fmtLemma{ku\-ni}%
\fmtOne{1}
\fmtCrossreference{\textnormal{see} ni{\textnormal{\fmtOne{2}}}}

\fmtLemma{ku\-ni}%
\fmtOne{2}
\fmtCitation{\textnormal{(e.g.}\,mu\-ku\-ni, ba\-ku\-ni, gi\-ku\-ni\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{clever, smart, intelligent}
\fmtExampleRL{Omwa\-la oni mu\-ku\-ni mu\-no.}
\fmtExampleEN{This girl is very smart.}
\fmtSubentry(b)
\fmtSense{neat, well dressed, elegant}
\fmtExampleRL{OSe\-bwa\-to mu\-sai\-za mu\-ku\-ni bu\-lii\-zo aku\-bba esaa\-ti eki\-ka\-lu.}
\fmtExampleEN{Sebwato is a well-dressed man, he always tucks in his shirt.}

\fmtLemma{ku\-nia\-yi\-mu\-ka}
\fmtCitation{(oku)\-ku\-nia\-yi\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-nia\-yi\-mu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to rush, to hurry, to dash}
\fmtExampleRL{Omu\-sai\-za n’a\-maa\-re n’a\-ku\-nia\-yi\-mu\-ka n’a\-ya\-ba.}
\fmtExampleEN{When the man finished eating food, he rushed away.}
\fmtSubentry(b)
\fmtSense{to walk in an awkward manner}
\fmtExampleRL{Aku\-nia\-yi\-mu\-ki\-re\-mbe na\-ya\-ba.}
\fmtExampleEN{He just went away.}
\fmtSubentry(c)
\fmtSense{to fall flat, to fall down}
\fmtExampleRL{Omu\-ta\-mii\-re aku\-nia\-yi\-mu\-ki\-re.}
\fmtExampleEN{The drunkard has fallen down.}

\fmtLemma{ku\-ni\-ka}
\fmtCitation{(oku)\-ku\-ni\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ni\-ki\-re}
\fmtSense{to plant (potato vines)}
\fmtExampleRL{Ya\-ba oku\-ni\-ke ebi\-koo\-la bya bi\-yaa\-ta.}
\fmtExampleEN{Go and plant the sweet potato vines.}

\fmtLemma{ku\-nku\-bya}
\fmtCitation{(oku)\-ku\-nku\-bya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-nku\-bi\-rye}
\fmtSubentry(a)
\fmtSense{to cause to do hurriedly, to rush, to hurry}
\fmtExampleRL{OMu\-ka\-sa aku\-nku\-bya aba\-ntu.}
\fmtExampleEN{Mukasa causes people to act hurriedly.}
\fmtSubentry(b)
\fmtSense{to arouse expectations in}
\fmtExampleRL{Lwa\-ki oku\-ku\-nku\-bya abai\-raa\-wo?}
\fmtExampleEN{Why do you arouse expectations in your pals?}

\fmtLemma{ku\-nku\-mu\-la}
\fmtCitation{(oku)\-ku\-nku\-mu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ku\-mwi\-re}
\fmtSubentry(a)
\fmtSense{to shake off, to throw off}
\fmtExampleRL{Ku\-nku\-mu\-la\-ku ei\-ta\-ka\-li oku olu\-go\-ye olwo.}
\fmtExampleEN{Shake the dust off that cloth.}
\fmtSubentry(b)
\fmtSense{to urinate}
\fmtExampleRL{Twa\-la omwa\-na oyo enzai aku\-nku\-mu\-le.}
\fmtExampleEN{Take that child outside to urinate.}

\fmtLemma{ku\-nku\-na}%
\fmtOne{1}
\fmtCitation{(oku)\-ku\-nku\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-nku\-nai\-re, ku\-nku\-ni\-re}
\fmtSense{to shiver, to tremble}
\fmtExampleRL{Twa\-mwa\-ji\-rye nga aku\-ku\-nku\-na olwa mu\-lo\-mbe.}
\fmtExampleEN{We found him shivering from the cold.}

\fmtLemma{ku\-nku\-na}%
\fmtOne{2}
\fmtCitation{(oku)\-ku\-nku\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-nku\-ni\-re}
\fmtSense{to walk alone}
\fmtExampleRL{Bo\-na omu\-sai\-za omu\-kai\-re n’a\-ku\-ku\-nku\-na.}
\fmtExampleEN{Have a look at the old man walking alone.}

\fmtLemma{ku\-nku\-ni\-za}
\fmtCitation{(oku)\-ku\-nku\-ni\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-nku\-nii\-ze}
\fmtSubentry(a)
\fmtSense{to press hard (with an intention of stopping the bleeding)}
\fmtExampleRL{Ku\-nki\-ni\-za olu\-bba\-le olwo lu\-le\-ke\-ra\-wo oku\-jwa omu\-saa\-yi.}
\fmtExampleEN{Press hard on that cut so that it doesn't bleed a lot.}
\fmtSubentry(b)
\fmtSense{to emphasise, to focus on}
\fmtExampleRL{Ku\-nku\-ni\-za enso\-nga ya kwe\-ga oku\-ko\-la.}
\fmtExampleEN{Focus on learning how to work.}
\fmtSubentry(c)
\fmtSense{to keep quiet (with an intention of disobeying)}
\fmtExampleRL{Omwa\-na oyo n’o\-mu\-ko\-ba\-ku aku\-nku\-ni\-za\-mbe.}
\fmtExampleEN{When you talk to that child, he just keeps quiet, but he won't obey.}

\fmtLemma{ku\-nku\-nu\-ka}
\fmtCitation{(oku)\-ku\-nku\-nu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-nku\-nu\-ki\-re}
\fmtSense{to scuttle, to run hurriedly}
\fmtExampleRL{Omwi\-bbi aku\-ku\-nku\-nu\-ka\-mbe.}
\fmtExampleEN{The thief is just scuttling away.}

\fmtLemma{ku\-nu\-ka}
\fmtCitation{(oku)\-ku\-nu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-nu\-ki\-re}
\fmtSense{to get a bruise}
\fmtExampleRL{Ya\-gwi\-re eki\-gwo n’a\-ku\-nu\-ka oku mu\-ku\-no.}
\fmtExampleEN{He fell and got a bruise on his hand.}

\fmtLemma{ku\-nu\-la}
\fmtCitation{(oku)\-ku\-nu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-nwi\-re}
\fmtSense{to scrub, to wash properly}
\fmtExampleRL{Ku\-ku\-nu\-la ese\-fu\-li\-ya zee\-re.}
\fmtExampleEN{Scrub the saucepan properly.}

\fmtLemma{ku\-nu\-mba}
\fmtCitation{(oku)\-ku\-nu\-mba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-nu\-mbi\-re}
\fmtSense{to walk freely, to behave innocently}
\fmtExampleRL{Nku\-kwe\-ta we\-te oku\-ku\-nu\-mba\-mbe!}
\fmtExampleEN{I am calling and you are simply walking away.}

\fmtLemma{ku\-nuu\-kii\-rya}
\fmtCitation{(oku)\-ku\-nuu\-kii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-nuu\-ki\-rii\-rye}
\fmtSubentry(a)
\fmtSense{to stand on tiptoes, to tip toe}
\fmtExampleRL{Na\-bbai\-re ntee\-kei\-bwe oku\-ku\-nu\-kii\-rya le\-ke mbo\-ne enzai wa lu\-ko\-lo\-go.}
\fmtExampleEN{I had to stand on tiptoe in order to look over the fence.}
\fmtSubentry(b)
\fmtSense{be near (to), to approach}
\fmtExampleRL{Tu\-ku\-nuu\-ki\-rya Na\-ka\-so\-ngo\-la.}
\fmtExampleEN{We are approaching Nakasongola.}

\fmtLemma{ku\-nya}
\fmtCitation{(oku)\-ku\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-nyi\-re}
\fmtSubentry(a)
\fmtSense{to make (juice from banana), to crush, to tread on (when making beer), to squeeze}
\fmtExampleRL{Mmai\-te oku\-ku\-nya ensa\-nde.}
\fmtExampleEN{I know how to make juice from bananas.}
\fmtSubentry(b)
\fmtSense{to rub, to rub together}
\fmtSubentry(c)
\fmtFixedexpression{ku\-nya (e)mi\-si\-nde}
\fmtSense{to race, to run}
\fmtExampleRL{Ni ya\-boi\-ne omu\-si\-ri\-ka\-le n'a\-ku\-nya emi\-si\-nde.}
\fmtExampleEN{He raced away when he saw a soldier.}

\fmtLemma{ku\-nyu\-la}
\fmtCitation{(oku)\-ku\-nyu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-nywi\-re}
\fmtSense{to take off (a piece), to tear off a piece}
\fmtExampleRL{Nku\-nyuu\-ra\-ku aka\-tu\-ndu oku ca\-pa\-ti ompe\-ku.}
\fmtExampleEN{Tear off a piece of chapati for me and give it to me please.}

\fmtLemma{ku\-pa}
\fmtCitation{\textnormal{(e.g.}\,gi\-ku\-pa, ka\-ku\-pa, bi\-ku\-pa\textnormal{)}}
\fmtPOS{adj}
\fmtSource{from Buyende}
\fmtSense{blunt, not sharp}
\fmtExampleRL{Enso\-ne eni gi\-ku\-pa.}
\fmtExampleEN{This knife is blunt.}

\fmtLemma{ku\-puu\-li\-nka}
\fmtCitation{(o)ku\-puu\-li\-nka}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{explosion}
\fmtExampleRL{Emo\-to\-ka eya\-sii\-si\-kai\-re ya\-ree\-te\-re oku\-puu\-li\-nka.}
\fmtExampleEN{The damaged car led to an explosion.}

\fmtLemma{ku\-pya\-ku\-pya}
\fmtCitation{(oku)\-ku\-pya\-ku\-pya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-pi\-rye\-ku\-pi\-rye}
\fmtSense{to be unscrupulous, to be unprincipled, to be dishonest}
\fmtExampleRL{Le\-ke\-ra\-wo oku\-ku\-pya\-ku\-pya.}
\fmtExampleEN{Stop being unscrupulous.}

\fmtLemma{ku\-raa\-ti\-ra}
\fmtCitation{(oku)\-ku\-raa\-ti\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-raa\-tii\-re}
\fmtSense{to convince, to talk into, to appeal to, to urge}
\fmtExampleRL{Ya\-ba oku\-raa\-ti\-re Aba\-yi\-nda ba\-tu\-we\-rye omwa\-la waa\-bwe.}
\fmtExampleEN{Go and convince the Bayinda to give us their daughter.}

\fmtLemma{ku\-ra\-ta}
\fmtCitation{(oku)\-ku\-ra\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ra\-ti\-re}
\fmtSense{to follow, to come after}
\fmtExampleRL{Edwa\-rd ni\-ye ye\-ku\-bbe\-mbe\-ra oSu\-sa\-n n’a\-mu\-ku\-ra\-ta.}
\fmtExampleEN{Edward leads and Susan follows him.}

\fmtLemma{ku\-ri\-gi\-ra}
\fmtCitation{(oku)\-ku\-ri\-gi\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ri\-gii\-re}
\fmtSense{to court (a person), to pursue, to run after, to woo}
\fmtExampleRL{OPau\-l aku\-ku\-ri\-gi\-ra abaa\-la ba\-bi\-ri omu\-lu\-ndi gu\-mwei.}
\fmtExampleEN{Paul is courting two girls at the same time.}

\fmtLemma{ku\-ru\-haa\-ni}
\fmtCitation{(e)ku\-ru\-haa\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{Koran}
\fmtExampleRL{Omu\-si\-raa\-mu akwai\-te eku\-ru\-haa\-ni yoo\-na\-yoo\-na omu mu\-twe.}
\fmtExampleEN{The Muslim has memorised the whole Koran.}

\fmtLemma{ku\-ru\-kuu\-nya}
\fmtCitation{(oku)\-ku\-ru\-kuu\-nya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-ru\-kuu\-nyi\-re}
\fmtSubentry(a)
\fmtSense{to roll around in the dirt}
\fmtExampleRL{Ombe\-re\-ge akwe\-ku\-ru\-kuu\-nya mu ita\-ka\-li.}
\fmtExampleEN{The pig is rolling in the dirt.}
\fmtSubentry(b)
\fmtSense{to write untidily}
\fmtExampleRL{Omwa\-na awa\-ndii\-ka ng’a\-ku\-ru\-kuu\-nya.}
\fmtExampleEN{The child writes untidily.}

\fmtLemma{ku\-ru\-kuu\-tya}%
\fmtOne{1}
\fmtCitation{(oku)\-ku\-ru\-kuu\-tya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-ru\-kuu\-ti\-rye}
\fmtSense{to unblock, to unclog, to unplug, to remove blockage}
\fmtExampleRL{Mpee\-rya aka\-saa\-le nku\-ru\-kuu\-tye omu nyu\-ngu.}
\fmtExampleEN{Give me a stick so that I can unblock the pipe.}

\fmtLemma{ku\-ru\-kuu\-tya}%
\fmtOne{2}
\fmtCitation{(oku)\-ku\-ru\-kuu\-tya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-ru\-kuu\-ti\-rye}
\fmtSense{to write untidily}
\fmtExampleRL{Omwe\-ge\-sya wai\-swe ya\-tu\-ku\-bbi\-ra\-nga oku\-ku\-ru\-kuu\-tya omu bi\-ta\-bu.}
\fmtExampleEN{Our teacher used to beat us whenever we wrote untidily in our notebooks.}

\fmtLemma{ku\-ru\-ndu\-ka}
\fmtCitation{(oku)\-ku\-ru\-ndu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-ru\-ndu\-ki\-re}
\fmtSense{to crawl, to creep, to go on hands and knee}
\fmtExampleRL{Omwa\-na ata\-ndi\-ki\-re oku\-ku\-ru\-ndu\-ka.}
\fmtExampleEN{The child has started crawling.}

\fmtLemma{ku\-ru\-ndya}
\fmtCitation{(oku)\-ku\-ru\-ndya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-ru\-ndi\-rye}
\fmtSense{to pull, to drag}
\fmtExampleRL{Omwa\-na aku\-ku\-ru\-ndya eki\-saa\-le eki\-re\-mee\-ri.}
\fmtExampleEN{The child is pulling a heavy log.}

\fmtLemma{ku\-ru\-see\-di}
\fmtCitation{(e)ku\-ru\-see\-di}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{crusade, campaign, movement}
\fmtExampleRL{Aba\-lo\-ko\-le baa\-bi\-re ku ku\-ru\-see\-di.}
\fmtExampleEN{The born-again Christians have gone on a campaign.}

\fmtLemma{ku\-rwa}
\fmtCitation{(o)ku\-rwa}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{red soil}
\fmtExampleRL{Ya\-sa\-ki\-re ekii\-na n’ai\-ka oku ku\-rwa.}
\fmtExampleEN{He dug a hole until he reached red soil.}

\fmtLemma{ku\-saa\-ga}
\fmtCitation{(o)ku\-saa\-ga}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{joke, prank, stunt, jest}
\fmtExampleRL{Okwo ti ku\-saa\-ga.}
\fmtExampleEN{That is not a joke.}

\fmtLemma{ku\-sai}
\fmtPOS{adv}
\fmtSense{well, nicely, correctly, properly}
\fmtExampleRL{Okoo\-re ku\-sai.}
\fmtExampleEN{You have done well.}

\fmtLemma{ku\-se\-mbe\-ra}
\fmtCrossreference{\textnormal{see also} sembera}
\fmtCitation{(o)ku\-se\-mbe\-ra}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{Holy Communion, Eucharist}
\fmtExampleRL{Ki\-ni ni\-kyo ki\-see\-ra kya ku\-se\-mbe\-ra.}
\fmtExampleEN{This is the time for Holy Communion.}

\fmtLemma{ku\-se\-nsa}
\fmtCitation{(oku)\-ku\-se\-nsa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ku\-se\-nse\-re}
\fmtSense{to be infected with measles}
\fmtExampleRL{Abaa\-na bai\-ngi ba\-ku\-se\-nse\-re.}
\fmtExampleEN{Many children got measles.}

\fmtLemma{ku\-si\-gi\-ki\-ra oku}
\fmtPOS{conj}
\fmtSense{according to, as stated by, as claimed by, based on, due to, in terms of}
\fmtExampleRL{Oku\-si\-gi\-ki\-ra oku ru\-ba\-zo lwa\-mwe mu\-ntu mu\-sai.}
\fmtExampleEN{According to his speech, he is a good person.}

\fmtLemma{ku\-sii\-ma}
\fmtCitation{(o)ku\-sii\-ma}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{appreciation, acknowledgement, admiration, affection}
\fmtExampleRL{Oku\-sii\-ma kwa\-mwe kwa\-mwe\-nde\-se\-rye opu\-re\-zi\-de\-nti.}
\fmtExampleEN{His appreciation endeared him to the president.}

\fmtLemma{ku\-si\-nga}
\fmtCrossreference{\textnormal{see also} singa}
\fmtCitation{(o)ku\-si\-nga}
\fmtPOS{conj}
\fmtSubentry(a)
\fmtSense{instead of, as an alternative}
\fmtExampleRL{Oku\-si\-nga oku\-zee\-nya ya\-ba olu\-me.}
\fmtExampleEN{Go and dig instead of playing.}
\fmtSubentry(b)
\fmtSense{mostly, mainly, for the most part, largely, chiefly, predominantly, principally, primarily}
\fmtExampleRL{Oku\-si\-nga mu\-no Aba\-ga\-nda ni\-bo ba\-si\-moo\-la ba\-tyo.}
\fmtExampleEN{It is mostly the Baganda who speak like that.}

\fmtLemma{ku\-si\-nzii\-ra}
\fmtCitation{(o)ku\-si\-nzii\-ra}
\fmtPOS{conj}
\fmtSense{according to, as stated by, as claimed by, depending on, based on}
\fmtExampleRL{Oku\-si\-nzii\-ra oku bi\-ntu bye tu\-boi\-ne, Su\-za\-n ni\-ye mu\-bbi.}
\fmtExampleEN{Based on what we have seen, it is Suzan who is wrong.}

\fmtLemma{ku\-soo\-ka}
\fmtSubentry(a)
\fmtCrossreference{\textnormal{see} sooka}
\fmtSubentry(b)
\fmtFixedexpression{ku\-soo\-ke\-ra kaa\-ki\-mwei}
\fmtSense{in the first place, first of all}
\fmtExampleRL{Oku\-soo\-ke\-ra kaa\-ki\-mwei, oyo mu\-lwai\-re.}
\fmtExampleEN{In the first place, he is sick.}

\fmtLemma{ku\-te\-ki\-si}
\fmtCitation{(o)ku\-te\-ki\-si}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{nail polish, cutex (UgE)}
\fmtExampleRL{Oku\-te\-ki\-si asii\-gi\-bwa ku no\-ne.}
\fmtExampleEN{Nail polish is applied on nails.}

\fmtLemma{ku\-ti\-ya}
\fmtCitation{(e)ku\-ti\-ya}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{sac, sack, bag}
\fmtExampleRL{Ama\-yuu\-ni aga\-twa\-lii\-re mu ku\-ti\-ya.}
\fmtExampleEN{He has taken the yams in a sack.}

\fmtLemma{ku\-tu}
\fmtPOS{ideo}
\fmtSense{describing how something is broken}
\fmtExampleRL{Atii\-ki\-rye omu\-gu\-wa ku\-tu.}
\fmtExampleEN{He has torn up the rope completely.}

\fmtLemma{ku\-tu\-bba}
\fmtCitation{(e)ku\-tu\-bba}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{Islamic sermon that precedes the Friday prayer}
\fmtExampleRL{Eku\-tu\-bba y’a\-ba\-si\-raa\-mu eta\-ndi\-ki\-re.}
\fmtExampleEN{The Islamic sermon has started.}

\fmtLemma{ku\-tu\-ki}
\fmtCitation{\textnormal{(e.g.}\,gi\-tu\-ki, bi\-tu\-ki, lu\-tu\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{broken}
\fmtExampleRL{Ota\-vu\-ga gaa\-li eyo ku\-bba gi\-ku\-tu\-ki.}
\fmtExampleEN{Don’t ride that bicycle because it is broken.}

\fmtLemma{ku\-tu\-la}
\fmtCitation{(oku)\-ku\-tu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-twi\-re}
\fmtSubentry(a)
\fmtSense{to break, to break of, to split, to cut off, to separate, to divide}
\fmtExampleRL{Ensi ba\-gi\-ku\-twi\-re\-mu ebi\-tu\-ndu bi\-bi\-ri.}
\fmtExampleEN{They divided the country into two halves.}
\fmtSubentry(b)
\fmtSense{to tear (a rope)}
\fmtExampleRL{Onte aku\-twi\-re omu\-gu\-wa.}
\fmtExampleEN{The cow has torn the rope.}
\fmtSubentry(c)
\fmtSense{to decide, to decide on}
\fmtExampleRL{Ku\-tu\-la ma\-ngu twa\-be mu mai\-so.}
\fmtExampleEN{Decide quickly and let's go ahead.}

\fmtLemma{ku\-tu\-lya}
\fmtCitation{(oku)\-ku\-tu\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-twi\-rye}
\fmtSense{to gulp, to swallow quickly}
\fmtExampleRL{Aku\-ku\-tu\-lya\-mbe nka aku\-nywa amai\-zi.}
\fmtExampleEN{He just gulps the water.}

\fmtLemma{ku\-twa}
\fmtCitation{(oku)\-ku\-twa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ku\-twi\-re}
\fmtSense{to quicken the process of making beer, to accelerate the process of making beer, to rush the process of making beer}
\fmtExampleRL{Ti ki\-sai ku\-ku\-twa mwe\-nge.}
\fmtExampleEN{It is not good to rush the process of making beer.}

\fmtLemma{ku\-twi}
\fmtCitation{(o)ku\-twi}
\fmtPOS{n}%
\fmtNC{15/6}
\fmtSense{ear}
\fmtExampleRL{Oku\-twi kwa\-nge ku\-ku\-ndu\-ma.}
\fmtExampleEN{My ear hurts.}

\fmtLemma{ku\-twi\-ku\-twi}
\fmtPOS{adv}
\fmtSense{right now, immediately}
\fmtExampleRL{Iza ku\-twi\-ku\-twi.}
\fmtExampleEN{Come right now.}

\fmtLemma{kuu\-ba}
\fmtCitation{(oku)\-kuu\-ba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kuu\-bi\-re}
\fmtSubentry(a)
\fmtSense{to rub, to massage}
\fmtExampleRL{Iza onkuu\-be ng’o\-ku\-ko\-le\-sya omu\-ba\-zi gu\-ni.}
\fmtExampleEN{Come and rub me with these herbs.}
\fmtSubentry(b)
\fmtSense{to smear}
\fmtExampleRL{Kuu\-ba obu\-sai obwo oku ki\-sii\-ka.}
\fmtExampleEN{Smear the cow dung on the wall.}
\fmtSubentry(c)
\fmtSense{to smooth, to smoothen, to make even, to make level}
\fmtExampleRL{Kuu\-ba olu\-baa\-wo olwo ku\-sai nga oku\-ko\-le\-sya la\-nda.}
\fmtExampleEN{Smooth that timber properly using a hand plane.}
\fmtSubentry(d)
\fmtSense{to climb (a tree without branches)}
\fmtExampleRL{Ki\-zi\-bu oku\-kuu\-ba omu\-saa\-le ogu\-ta\-li\-na ma\-taa\-gi.}
\fmtExampleEN{It is difficult to climb a tall tree without branches.}

\fmtLemma{kuu\-buu\-lya}
\fmtCitation{(oku)\-kuu\-buu\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kuu\-bui\-rye}
\fmtSense{to remove the midrib of a banana leaf, to peel the mid-rib off of a plantain leaf, to remove the stem from the leaf}
\fmtExampleRL{Ku\-kuu\-buu\-lya enda\-ga\-la ezo.}
\fmtExampleEN{Remove the midribs of the banana leaves.}

\fmtLemma{kuu\-ka}
\fmtCitation{(oku)\-kuu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kuu\-ki\-re}
\fmtSense{to be satisfied (esp. with food and drink), to be pleased, to be sated, to be full}
\fmtExampleRL{Onte anywi\-re amai\-zi n’a\-kuu\-ka.}
\fmtExampleEN{The cow drank water and is satisfied.}

\fmtLemma{kuu\-ku\-la}
\fmtCitation{(oku)\-kuu\-ku\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kuu\-kwi\-re}
\fmtSense{to dig out, to uproot, to pull out, to pull up}
\fmtExampleRL{Baa\-bi\-re ku\-kuu\-ku\-la mu\-wo\-go.}
\fmtExampleEN{They have gone to dig out the cassava.}

\fmtLemma{kuu\-la}
\fmtCitation{(oku)\-kuu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kwi\-re}
\fmtSense{to dig out, to uproot, to pull out, to pull up}
\fmtExampleRL{Kuu\-la amai\-do ago.}
\fmtExampleEN{Uproot those groundnuts.}

\fmtLemma{kuu\-ma}
\fmtCitation{(oku)\-kuu\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kuu\-mi\-re}
\fmtSubentry(a)
\fmtSense{to keep, to protect, to save, to guard, to watch, to watch over, to protect}
\fmtExampleRL{Aku\-kuu\-ma mai\-do gaa\-mwe bu\-ta\-gai\-bba.}
\fmtExampleEN{He is protecting his groundnuts against thieves.}
\fmtSubentry(b)
\fmtSense{to wait}
\fmtExampleRL{Kuu\-ma okwi\-ca ni yai\-za.}
\fmtExampleEN{Wait until he comes.}

\fmtLemma{kuu\-mu\-la}
\fmtCitation{(oku)\-kuu\-mu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kuu\-mwi\-re}
\fmtSubentry(a)
\fmtSense{to hum, to sing with closed lips}
\fmtExampleRL{Mu\-le\-ke\-ra\-wo oku\-kuu\-mu\-la le\-ke mwe\-mbe ku\-sai.}
\fmtExampleEN{Stop humming and sing properly.}
\fmtSubentry(b)
\fmtSense{to grunt, to mumble, to murmur}
\fmtExampleRL{Omu\-sai\-za oyo ni mu\-la\-mu\-kya akuu\-mu\-la\-mbe.}
\fmtExampleEN{When I greet this man, he just grunts.}

\fmtLemma{kuu\-mu\-lya}
\fmtCitation{(oku)\-kuu\-mu\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kuu\-mwi\-rye}
\fmtSense{to take off a piece, to tear off a piece}
\fmtExampleRL{Nkuu\-mu\-lya\-ku oku ca\-pa\-ti ndye\-ku.}
\fmtExampleEN{Tear off a piece of chapati for me so that I can eat a bit of it.}

\fmtLemma{kuu\-nyuu\-ka}%
\fmtOne{1}
\fmtCitation{(oku)\-kuu\-nyuu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kuu\-nyuu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to lose hair}
\fmtExampleRL{Ombwe\-ne akuu\-nyuu\-ki\-re obwo\-ya.}
\fmtExampleEN{The dog has lost some of its hair.}
\fmtSubentry(b)
\fmtSense{(of clothing) to get rough texture, to become rough}
\fmtExampleRL{Olu\-go\-ye lwa\-nge lu\-kuu\-nyuu\-ki\-re.}
\fmtExampleEN{My clothing got a rough texture.}
\fmtSubentry(c)
\fmtSense{to run off, to slip away}
\fmtExampleRL{Akuu\-nyuu\-ki\-re n'e\-ki\-tee\-tee\-yi kya Na\-lu\-woo\-za.}
\fmtExampleEN{He has run off with Naluwooza's dress.}

\fmtLemma{kuu\-ra\-ta}
\fmtCitation{(oku)\-kuu\-ra\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kuu\-ra\-ti\-re}
\fmtSense{to praise, to compliment, to admi\-re}
\fmtExampleRL{Mu\-ta\-ka oku\-kuu\-ra\-ta aba\-ka\-li ni w’e\-ki\-ta\-kwe\-taa\-gi\-sya.}
\fmtExampleEN{You like complimenting women even when it is not necessary.}

\fmtLemma{kuu\-ra\-tya}
\fmtCitation{(oku)\-ku\-raa\-tya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kuu\-ra\-ti\-rye}
\fmtSense{to canvass (for a party), to campaign, to solicit (votes)}
\fmtExampleRL{Mpee\-rya se\-nte nku\-kuu\-ra\-ti\-rye aba\-ntu.}
\fmtExampleEN{Give me money and I will solicit votes for you.}

\fmtLemma{kuu\-rya}
\fmtCitation{(oku)\-kuu\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kuu\-rye}
\fmtSubentry(a)
\fmtSense{to raise}
\fmtExampleRL{Omaa\-ma eya\-nkuu\-rye afii\-re.}
\fmtExampleEN{The auntie who raised me has passed on.}
\fmtSubentry(b)
\fmtSense{to train, to coach, to tutor, to mentor}
\fmtExampleRL{Naa\-ni eya\-ku\-kuu\-rye ni wee\-yu\-ngi\-re oku ka\-mpu\-ni eni?}
\fmtExampleEN{Who mentored you when you joined this company?}

\fmtLemma{kuu\-sii\-ra}
\fmtCitation{(oku)\-kuu\-sii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kuu\-sii\-re}
\fmtSense{to be very dirty, to be stained}
\fmtExampleRL{Eki\-na\-ga ki\-ni ki\-kuu\-sii\-re mu\-no.}
\fmtExampleEN{This cooking pan is excessively dirty.}

\fmtLemma{kuu\-ta}
\fmtCitation{(oku)\-kuu\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kuu\-ti\-re}
\fmtSense{to scrub, to clean by scrubing, to brush}
\fmtExampleRL{Kuu\-ta eki\-naa\-bi\-ro.}
\fmtExampleEN{Scrub the bathroom.}

\fmtLemma{kuu\-ti\-ra}
\fmtCitation{(oku)\-kuu\-ti\-ra}
\fmtPOS{v}%
\fmtTrans{ditr}
\fmtPFV{kuu\-tii\-re}
\fmtSubentry(a)
\fmtSense{to advise, to recommend, to warn}
\fmtExampleRL{Ba\-kuu\-ti\-re ba\-ta\-ya\-ba bwi\-re.}
\fmtExampleEN{Warn them not to walk at night.}
\fmtSubentry(b)
\fmtSense{to emphasise, to lay stress}
\fmtExampleRL{Kuu\-ti\-ra aba\-ntu enso\-nga ya bu\-yo\-njo.}
\fmtExampleEN{Emphasise the issue of cleanliness to the people.}

\fmtLemma{kuu\-ti\-rya}
\fmtCitation{(oku)\-kuu\-ti\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kuu\-tii\-rye}
\fmtSense{to write sloppily, to write untidily}
\fmtExampleRL{Omwa\-na oni akuu\-ti\-rya mu\-no omu ki\-ta\-bo kya\-mwe.}
\fmtExampleEN{This child writes very sloppily in his book.}

\fmtLemma{kuu\-tuu\-ka}
\fmtCitation{(oku)\-kuu\-tuu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kuu\-tuu\-ki\-re}
\fmtSense{to be at loggerheads, to have disagreement with, to be in opposition to}
\fmtExampleRL{Aba\-sai\-za abo ba\-bbai\-re ba\-ku\-kuu\-tuu\-ka oku\-ma\-la ei\-bba\-nga li\-koo\-to.}
\fmtExampleEN{Those men have been at loggerheads for a long time.}

\fmtLemma{ku\-yi\-ku\-yi}
\fmtCitation{\textnormal{(e.g.}\,mu\-ku\-yi\-ku\-yi, ba\-ku\-yi\-ku\-yi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{untrustworthy, dodgy}
\fmtExampleRL{Omu\-fu\-mu odi mu\-ku\-yi\-ku\-yi.}
\fmtExampleEN{That herbalist is untrustworthy.}

\fmtLemma{ku\-zu\-mu}
\fmtCitation{(e)ku\-zu\-mu}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{world of the dead, underworld}
\fmtExampleRL{Eku\-zu\-mu ki\-fo ki\-bbi mu\-no.}
\fmtExampleEN{The world of the dead is a very terrible place.}

\fmtLemma{ku\-zwa}
\fmtCrossreference{\textnormal{see also} zwa}
\fmtPOS{prep}
\fmtSubentry(a)
\fmtSense{from, out of}
\fmtExampleRL{Lee\-ta ekyai\-jo kya ki\-saa\-le oku\-zwa mu si\-too\-wa.}
\fmtExampleEN{Bring a piece of wood from the store.}
\fmtSubentry(b)
\fmtSense{since}
\fmtExampleRL{Abbai\-re na\-mbu\-la-buculeeri okuzwa lwe baitire omwara.}
\fmtExampleEN{He has been restless since they killed his daughter.}
\fmtSubentry(c)
\fmtFixedexpression{ku\-zwa (e)irai}
\fmtSense{since long ago, for ages}
\fmtExampleRL{Oku\-zwa ei\-rai n’ei\-rai aba\-ka\-li ti\-ba\-lya\-nga nko\-ko.}
\fmtExampleEN{Since time immemorial women have not eaten chicken.}

\fmtLemma{kwa}%
\fmtOne{1}
\fmtPOS{ideo}
\fmtSense{sound of blades rubbing against each other}
\fmtExampleRL{Mpuu\-lii\-re eki\-ntu nga ki\-kwa\-kwa\-ya, kwa.}
\fmtExampleEN{I heard the sound of blades rubbing against each other.}

\fmtLemma{kwa}%
\fmtOne{2}
\fmtCitation{(o)kwa}
\fmtPOS{prep}
\fmtSense{during, in the course of, throughout the time of, for the time of, in the time of}
\fmtExampleRL{Endu zee\-bbi\-sa mu bi\-nyo\-mo okwa kya\-nda.}
\fmtExampleEN{Lung fish hide in the breeding grounds during the dry spell.}

\fmtLemma{Kwa\-bi\-duu\-ma}
\fmtCitation{(o)Kwa\-bi\-duu\-ma}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{June}
\fmtExampleRL{Okwe\-zi Kwa\-bi\-duu\-ma ku\-ta\-ndi\-ki\-re.}
\fmtExampleEN{June has started.}

\fmtLemma{kwa\-bwe}
\fmtCrossreference{\textnormal{see} (-)abwe}

\fmtLemma{kwa\-du}
\fmtCitation{(o)kwa\-du}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{scorpion}
\fmtExampleRL{Okwa\-du amu\-ne\-ne\-re.}
\fmtExampleEN{A scorpion bit him.}

\fmtLemma{kwa\-du\-la}
\fmtCitation{(oku)\-kwa\-du\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kwa\-dwi\-re}
\fmtSense{to scratch, to scrape}
\fmtExampleRL{Amu\-kwa\-dwi\-re omu mai\-so.}
\fmtExampleEN{She scratched him in the face.}

\fmtLemma{Kwai\-ke\-ka\-bi\-tee\-re}
\fmtCitation{(o)Kwai\-ke\-ka\-bi\-tee\-re}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{February}
\fmtExampleRL{Tu\-ke\-ka Kwai\-ke\-ka\-bi\-tee\-re.}
\fmtExampleEN{We slice sweet potatoes in February.}
\fmtSense{February}
\fmtExampleRL{Nai\-za mu Kwai\-ke\-ka\-bi\-tee\-re.}
\fmtExampleEN{I will come in February.}

\fmtLemma{Kwai\-ke\-sa}
\fmtCitation{(o)Kwai\-ke\-sa}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{June}
\fmtExampleRL{Ya\-ya\-bi\-re mu kwe\-zi Kwai\-ke\-sa.}
\fmtExampleEN{She left in June.}

\fmtLemma{Kwai\-se\-nsa\-ncwi}
\fmtCitation{(o)Kwai\-se\-nsa\-ncwi}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{May}
\fmtExampleRL{Encwi zi\-se\-nsa Kwai\-se\-nsa\-ncwi.}
\fmtExampleEN{Fish increases in May.}

\fmtLemma{Kwai\-si\-ga}
\fmtCitation{(o)Kwai\-si\-ga}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{March}
\fmtExampleRL{Tu\-li mu\-kwe\-zi Kwai\-si\-ga.}
\fmtExampleEN{It is March. (lit.\,We are in March.)}

\fmtLemma{kwai\-swe}
\fmtCrossreference{\textnormal{see} (-)aiswe}

\fmtLemma{kwa\-ku\-la}
\fmtCitation{(oku)\-kwa\-ku\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kwa\-kwi\-re}
\fmtSense{to remove by force, to topple, to overthrow}
\fmtExampleRL{Obo\-te baa\-mu\-kwa\-kwi\-re\-ku nga aya\-bi\-re e Si\-nga\-po\-re.}
\fmtExampleEN{Obote was toppled when he had gone to Singapore.}

\fmtLemma{kwa\-kwa\-ya}
\fmtCitation{(oku)\-kwa\-kwa\-ya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwa\-kwa\-yi\-re}
\fmtSense{to rustle}
\fmtExampleRL{Ombwe\-ne aku\-kwa\-kwa\-ya omu bi\-swoi\-gi.}
\fmtExampleEN{The dog is rustling in the dry banana leaves.}

\fmtLemma{Kwa\-mai\-do}
\fmtCitation{(o)Kwa\-mai\-do}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{July}
\fmtExampleRL{Amai\-do tu\-gai\-ya mu kwe\-zi Kwa\-mai\-do.}
\fmtExampleEN{We harvest groundnuts in July.}

\fmtLemma{Kwa\-mai\-zi}
\fmtCitation{(o)Kwa\-mai\-zi}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{May}
\fmtExampleRL{Mu\-li\-ta\-ndi\-ka omu kwe\-zi Kwa\-mai\-zi.}
\fmtExampleEN{You will start in May.}

\fmtLemma{Kwa\-mba\-ga}
\fmtCitation{(o)Kwa\-mba\-ga}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{December}
\fmtExampleRL{Twi\-ge\-rei\-re okwe\-zi Kwa\-mba\-ga.}
\fmtExampleEN{We are approaching the month of December.}

\fmtLemma{Kwa\-mpii\-ra}
\fmtCitation{(o)Kwa\-mpii\-ra}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{January}
\fmtExampleRL{Ei\-sa\-na li\-bba li\-ngi Okwa\-mpii\-ra.}
\fmtExampleEN{There is always a lot of sunshine in January.}

\fmtLemma{kwa\-mu}
\fmtCrossreference{\textnormal{see} (-)amu}

\fmtLemma{kwa\-mu\-li\-kii\-rya}
\fmtCrossreference{\textnormal{see also} mulikiirya}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{darkness before the new moon}
\fmtExampleRL{Encwi zi\-bo\-ne\-ka kwa\-mu\-li\-kii\-rya.}
\fmtExampleEN{Fish is plentiful during the darkness before the new moon.}

\fmtLemma{kwa\-mwe\-(i)}
\fmtCrossreference{\textnormal{see} amwe\textnormal{{\fmtOne{1}}}}

\fmtLemma{kwa\-na}
\fmtCitation{(oku)\-kwa\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kwai\-ne}
\fmtSense{to woo, to romantically pursue, to befriend, to make friends with}
\fmtExampleRL{Ata\-ka mu\-no oku\-kwa\-na emi\-ka\-go.}
\fmtExampleEN{He likes befriending people.}

\fmtLemma{kwa\-nge}
\fmtCrossreference{\textnormal{see} ange}

\fmtLemma{kwa\-nju\-la}
\fmtCrossreference{\textnormal{see} aslo anjula}
\fmtCitation{(o)kwa\-nju\-la}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{traditional engagement, introduction ceremony (a cultural function where the girl introduces her fiancé to the clan members)}
\fmtExampleRL{Baa\-bi\-re ku mu\-ko\-lo gwa kwa\-nju\-la.}
\fmtExampleEN{They went to the introduction ceremony.}

\fmtLemma{Kwa\-nsei\-ze\-re}
\fmtCrossreference{\textnormal{see also} nseize\-re}
\fmtCitation{(o)Kwa\-nsei\-ze\-re}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{April}
\fmtExampleRL{Ensei\-ze\-re zi\-gya Kwa\-nsei\-ze\-re.}
\fmtExampleEN{Big flying termites fly in April.}

\fmtLemma{Kwa\-nse\-nde\-kee\-rya\-mpee\-fu}
\fmtCrossreference{\textnormal{see also} mpeefu}
\fmtCitation{(o)Kwa\-nse\-nde\-kee\-rya\-mpee\-fu}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{September}
\fmtExampleRL{Onte wa\-nge ya\-byai\-re mu Kwa\-nse\-nde\-kee\-rya\-mpee\-fu.}
\fmtExampleEN{My cow delivered in September.}

\fmtLemma{Kwa\-nsu\-la\-mii\-ra}
\fmtCrossreference{\textnormal{see also} nsulamiira}
\fmtCitation{(o)Kwa\-nsu\-la\-mii\-ra}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{October}
\fmtExampleRL{Okwe\-zi Kwa\-nsu\-la\-mii\-ra ku\-ta\-ndi\-ki\-re.}
\fmtExampleEN{October has started.}

\fmtLemma{Kwa\-nsye\-ne\-ne}
\fmtCrossreference{\textnormal{see also} nsyenene}
\fmtCitation{(o)Kwa\-nsye\-ne\-ne}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{November}
\fmtExampleRL{Ya\-bya\-lii\-bwe mu kwe\-zi Kwa\-nsye\-ne\-ne.}
\fmtExampleEN{He was born in November.}

\fmtLemma{kwa\-nya}
\fmtCitation{(oku)\-kwa\-nya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwa\-nyi\-re}
\fmtSense{to sprout, to germinate, to shoot (up)}
\fmtExampleRL{Obwi\-ta bu\-ta\-ndi\-ki\-re oku\-kwa\-nya.}
\fmtExampleEN{Millet has started sprouting.}

\fmtLemma{kwa\-nywe}
\fmtCrossreference{\textnormal{see} nywe}

\fmtLemma{Kwa\-nzo\-gi}
\fmtCrossreference{\textnormal{see also} nzogi}
\fmtCitation{(o)Kwa\-nzo\-gi}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{August}
\fmtExampleRL{Okwe\-zi Kwa\-nzo\-gi ku\-bba ku\-lai.}
\fmtExampleEN{August is always long.}

\fmtLemma{kwa\-raa\-ta\-na}
\fmtCrossreference{\textnormal{see also} araatana}
\fmtCitation{(o)kwa\-raa\-ta\-na}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{misunderstanding, quarrel}
\fmtExampleRL{Ki\-ki eki\-re\-te\-re\-wo okwa\-raa\-ta\-na?}
\fmtExampleEN{What caused the misunderstanding?}

\fmtLemma{kwa\-ta}
\fmtCitation{(oku)\-kwa\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kwai\-te}
\fmtSubentry(a)
\fmtSense{to hit (a target)}
\fmtExampleRL{Aka\-saa\-le ka\-kwai\-te ku\-sai.}
\fmtExampleEN{The dart has hit the target.}
\fmtSubentry(b)
\fmtSense{to hold}
\fmtExampleRL{Kwa\-ta ombwe\-ne oyo ata\-ne\-na ba\-ntu.}
\fmtExampleEN{Hold that dog so that it does not bite anyone.}
\fmtSubentry(c)
\fmtSense{to obey, to observe}
\fmtExampleRL{Kwa\-ta ama\-tee\-ka ga Ka\-nca.}
\fmtExampleEN{Observe the Lord's commandments.}
\fmtSubentry(d)
\fmtSense{to catch, to arrest, to take into custody, to detain}
\fmtExampleRL{Kwa\-ta omwi\-bbi oyo.}
\fmtExampleEN{Arrest that thief.}
\fmtSubentry(e)
\fmtSense{to rape, to sexual assault, to sexual abuse}
\fmtExampleRL{Omu\-sai\-za eya\-kwai\-te omwa\-la ba\-mu\-bbwe\-re.}
\fmtExampleEN{The man who raped the girl has been sent to prison.}
\fmtSubentry(f)
\fmtSense{(of a spirit) to possess}
\fmtExampleRL{Akwa\-tii\-bwe omu\-zu\-mu.}
\fmtExampleEN{He is possessed by an evil spirit.}
\fmtSubentry(g)
\fmtSense{(of a desease) to affect (often in passive)}
\fmtExampleRL{Ya\-kwa\-tii\-bwe embi\-ro.}
\fmtExampleEN{He has diarrhoea.}
\fmtSubentry(i)
\fmtSense{to solidify, to form}
\fmtExampleRL{Amai\-zi ga\-kwai\-te ama\-bbaa\-le.}
\fmtExampleEN{Water has turned into ice.}
\fmtSubentry(h)
\fmtSense{to be firm, to grow well}
\fmtExampleRL{Omu\-saa\-le gwe twa\-si\-mbi\-re gu\-kwai\-te ku\-sai.}
\fmtExampleEN{The tree which we planted is growing well.}
\fmtSubentry(j)
\fmtSense{to fit tightly}
\fmtExampleRL{Empa\-le emu\-kwai\-te.}
\fmtExampleEN{The trousers fit him tightly.}
\fmtSubentry(k)
\fmtSense{to understand, to apprehend, to grasp, to see, to get}
\fmtExampleRL{Kwa\-ta bye nku\-ku\-ko\-ba.}
\fmtExampleEN{Understand what I am telling you.}
\fmtSubentry(l)
\fmtSense{to memorise, to learn}
\fmtExampleRL{Bai\-ra\-ba\-nge tu\-kwa\-te olwe\-mbo lu\-ni.}
\fmtExampleEN{My friends let us learn this song.}
\fmtSubentry(m)
\fmtFixedexpression{kwa\-ta (o)muu\-ro}
\fmtSense{to catch fire, to begin to burn}
\fmtExampleRL{Ei\-su\-bi li\-kwai\-te omuu\-ro.}
\fmtExampleEN{The grass has caught fire.}


\fmtLemma{kwa\-ta\-ga\-na}
\fmtCrossreference{\textnormal{see} kwatangana}

\fmtLemma{kwa\-ta\-mpo\-la}
\fmtCitation{\textnormal{(e.g.}\,mu\-kwa\-ta\-mpo\-la\textnormal{)}}
\fmtPOS{adj}
\fmtSense{very cautious, careful}
\fmtExampleRL{Omu\-sai\-za omu\-kwa\-ta\-mpo\-la asoo\-ka ku\-lo\-woo\-za kya kwa\-ba oku\-ba\-za.}
\fmtExampleEN{A cautious man, he thinks through whatever he says.}

\fmtLemma{kwa\-ta\-nga\-na}
\fmtCitation{(oku)\-kwa\-ta\-nga\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwa\-ta\-gai\-ne}
\fmtSubentry(a)
\fmtSense{to hold each other, to seize each other, to fight each other}
\fmtExampleRL{Aba\-ka\-li ba\-kwa\-ta\-gai\-ne omwei n'a\-fa.}
\fmtExampleEN{The women fought each other and one of them died}
\fmtSubentry(b)
\fmtSense{to have one's hands full}
\fmtExampleRL{Mboi\-ne OSa\-mue\-l mu ka\-ta\-le nga akwa\-ta\-gai\-ne.}
\fmtExampleEN{I have seen Samuel in the market with his hands full.}
\fmtSubentry(c)
\fmtSense{to cooperate, to collaborate, to work together, to work side by side}
\fmtExampleRL{Ki\-sai aba\-lu\-ga\-nda oku\-kwa\-ta\-nga\-na.}
\fmtExampleEN{It is good for relatives to cooperate.}
\fmtSubentry(d)
\fmtSense{to come together, to get together}
\fmtExampleRL{Ki\-sai aba\-lu\-ga\-nda oku\-kwa\-ta\-na.}
\fmtExampleEN{It is good for relatives to come together.}
\fmtSubentry(f)
\fmtSense{to play (against or with)}
\fmtExampleRL{OKa\-yu\-nga akwa\-ta\-nga\-na n’o\-Iga\-nga omu mu\-pii\-ra.}
\fmtExampleEN{Kayunga will play against Iganga in football.}
\fmtSubentry(f)
\fmtSense{to have a connection (to or with), to have a relation (to), to come to an agreement (with)}
\fmtExampleRL{Tu\-kwa\-ta\-ga\-na omu mwo\-yo.}
\fmtExampleEN{We are connected in spirit.}

\fmtLemma{kwa\-ta\-ni\-za}
\fmtCitation{(oku)\-kwa\-ta\-ni\-za}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwa\-ta\-nii\-ze}
\fmtSubentry(a)
\fmtSense{to join hands, to cooperate, to collaborate, to work together, to work side by side}
\fmtExampleRL{Mwi\-ze tu\-kwa\-ta\-ni\-ze two\-mbo\-ke olu\-ko\-ro\-go.}
\fmtExampleEN{Come, we'll work together and build a fence.}
\fmtSubentry(b)
\fmtSense{to do many things at the same time, to multitask}
\fmtExampleRL{Wee\-na\-te olwa\-lee\-ro okwa\-ta\-nii\-ze bu\-ndi.}
\fmtExampleEN{You are doing many things today.}

\fmtLemma{kwa\-ti}
\fmtCitation{\textnormal{(e.g.}\,mwe\-kwa\-ti, ge\-kwa\-ti\textnormal{)}}
\fmtPOS{adj}
\fmtSense{thick}
\fmtExampleRL{Omwa\-la asu\-mbbi\-re omu\-ku\-bi gwe\-kwa\-ti.}
\fmtExampleEN{The girl cooked thick sauce.}

\fmtLemma{kwa\-ti\-bwa}
\fmtCrossreference{\textnormal{see} kwata}

\fmtLemma{kwa\-ti\-ra}
\fmtCitation{(oku)\-kwa\-ti\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwa\-tii\-re}
\fmtSense{to be sticky}
\fmtExampleRL{Ama\-sa\-nda ga fe\-ne ga\-kwa\-ti\-ra.}
\fmtExampleEN{Jackfruit sap is sticky.}
\fmtSense{to get stuck}

\fmtLemma{kwa\-ti\-ra\-ku}
\fmtCitation{(oku)\-kwa\-ti\-ra\-ku}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kwa\-tii\-re\-ku}
\fmtSense{to help, to assist}
\fmtExampleRL{Ka nyi\-ze nku\-kwa\-ti\-re\-ku ekii\-bbo.}
\fmtExampleEN{Let me come and help you with the basket.}

\fmtLemma{kwa\-ti\-sya}
\fmtCitation{(oku)\-kwa\-ti\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kwa\-ti\-si\-rye}
\fmtSubentry(a)
\fmtSense{to stick together, to glue together}
\fmtExampleRL{Ko\-le\-sya ma\-sa\-nda oku\-kwa\-ti\-sya empa\-pu\-la.}
\fmtExampleEN{Use glue to stick sheets of paper together.}
\fmtSubentry(b)
\fmtSense{to catch, to capture}
\fmtExampleRL{Ya\-te\-ge\-re ebi\-na\-ge n’a\-kwa\-ti\-sya ondu omu\-koo\-to.}
\fmtExampleEN{He set a trap and caught a big lung fish.}
\fmtSubentry(c)
\fmtSense{to hand over, to pass, to yield, to give away}
\fmtExampleRL{Kwa\-ti\-sya omwe\-be\-mbe\-zi omu\-yaa\-ka obu\-ka\-ma.}
\fmtExampleEN{Hand over power to the new leader.}

\fmtLemma{kwa\-ya}%
\fmtOne{1}
\fmtCitation{(e)kwa\-ya}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{choir}
\fmtExampleRL{Eyo ekwa\-ya eye\-mba ku\-sai.}
\fmtExampleEN{That choir sings well.}

\fmtLemma{kwa\-ya}%
\fmtOne{2}
\fmtCitation{(oku)\-kwa\-ya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwa\-yi\-re}
\fmtSense{to feel a pricking sensation, to feel a stabbing pain}
\fmtExampleRL{Nku\-wuu\-ra eki\-ntu eki\-ku\-nkwa\-ya omu ma\-twi.}
\fmtExampleEN{I have a stabbing pain in my ears.}

\fmtLemma{kwe\-ka}
\fmtCitation{(oku)\-kwe\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kwe\-ke\-re}
\fmtSense{to hide, to conceal, to put out of sight, to tuck away}
\fmtExampleRL{Kwe\-ka empii\-ya za\-nge ba\-ta\-zii\-bba.}
\fmtExampleEN{Hide my money or else they will steal it.}

\fmtLemma{kwe\-kee\-ra}
\fmtCrossreference{\textnormal{see also} ekeera}
\fmtCitation{(o)kwe\-kee\-ra}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{beginning, start}
\fmtExampleRL{Omu kwe\-kee\-ra baa\-li ba\-doo\-li ati\-bu\-ni bai\-ngi.}
\fmtExampleEN{In the beginning, they were few, but now they are many.}

\fmtLemma{kwe\-ku\-ka}
\fmtCitation{(oku)\-kwe\-ku\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwe\-ku\-ki\-re}
\fmtSense{to come out of hiding}
\fmtExampleRL{Omwi\-bbi akwe\-ku\-ki\-re.}
\fmtExampleEN{The thief has come out of hiding.}

\fmtLemma{kwe\-ku\-la}
\fmtCitation{(oku)\-kwe\-ku\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kwe\-kwi\-re}
\fmtSense{to find, to locate, to spot}
\fmtExampleRL{Omwi\-bbi ba\-mu\-kwe\-kwi\-re omu ka\-sa\-ka.}
\fmtExampleEN{They found the thief in the thicket.}

\fmtLemma{kwe\-mba}
\fmtCrossreference{\textnormal{see also} emba}
\fmtCitation{(o)kwe\-mba}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{singing, music}
\fmtExampleRL{Ata n’e\-ki\-co\-nco kya kwe\-mba.}
\fmtExampleEN{She has a talent for music.}

\fmtLemma{kwe\-nya}
\fmtCitation{(oku)\-kwe\-nya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwe\-nye\-re}
\fmtSense{to have dry skin, to be pale, to lose lustre, to lose colour, to fade}
\fmtExampleRL{K’o\-kwe\-nye\-re bu\-ndi?}
\fmtExampleEN{Why are you so pale?}

\fmtLemma{kwe\-nya\-kwe\-nya}
\fmtCitation{(oku)\-kwe\-nya\-kwe\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kwe\-nye\-re\-kwe\-nye\-re}
\fmtSense{to deceive, to coax, to cheat, to lure}
\fmtExampleRL{Le\-ke\-ra awo oku\-tu\-kwe\-nya\-kwe\-nya!}
\fmtExampleEN{Stop deceiving us!}

\fmtLemma{kwe\-nyi}
\fmtCitation{\textnormal{(e.g.}\,mu\-kwe\-nyi, ba\-kwe\-nyi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{dry-skinned, pale}
\fmtExampleRL{Omu\-sai\-za oyo abba mu\-kwe\-nyi, ta\-nyi\-ri\-ra.}
\fmtExampleEN{That man is always pale, he is never attractive.}

\fmtLemma{kwe\-nyu\-la}
\fmtCitation{(oku)\-kwe\-nyu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kwe\-nywi\-re}
\fmtSense{to grab, to seize, to grasp, to snatch}
\fmtExampleRL{Ba\-mu\-kwe\-nywi\-re\-ku aka\-sa\-wu ka se\-nte ni bai\-ru\-ka.}
\fmtExampleEN{They grabbed the money purse from her and ran away.}

\fmtLemma{kwe\-ra}
\fmtCrossreference{\textnormal{see} mwoyo (a)kwera and era}

\fmtLemma{kwe\-re\-ma}
\fmtCitation{(o)kwe\-re\-ma}
\fmtPOS{n}%
\fmtNC{15}
\fmtSense{independence}
\fmtExampleRL{Uga\-nda ya\-fu\-ni\-re okwe\-re\-ma mu mwa\-ka gwa 1962.}
\fmtExampleEN{Uganda became independent in 1962.}

\fmtLemma{kwe\-ri}
\fmtPOS{interj}
\fmtSource{from Sw.}
\fmtSense{really}
\fmtExampleRL{Kwe\-ri aba\-sai\-za abo bai\-za?}
\fmtExampleEN{Really, will those men come?}

\fmtLemma{kwe\-se\-ra}
\fmtCitation{(oku)\-kwe\-se\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwe\-sei\-re}
\fmtSubentry(a)
\fmtSense{to dance in an appropriate tempo, to run at an appropriate speed}
\fmtExampleRL{Omwi\-ru\-ki wa mbi\-ro aku\-kwe\-se\-ra ama\-le\-yo emi\-ru\-ndi gya\-mwe.}
\fmtExampleEN{An athlete is running at the appropriate speed so as to complete the race.}
\fmtSubentry(b)
\fmtSense{to rush (carrying a heavy load)}
\fmtExampleRL{Mmu\-boi\-ne aku\-kwe\-se\-ra nga yee\-tii\-ki\-re ensa\-wu ya muu\-nyu.}
\fmtExampleEN{I saw him rushing while carrying a sack of salt.}

\fmtLemma{kwe\-te}
\fmtCitation{(e)kwe\-te}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{type of local beer (made from millet, cassava and maize floor)}
\fmtExampleRL{Ti\-nywa kwe\-te.}
\fmtExampleEN{I don’t drink local beer.}

\fmtLemma{kwe\-tu\-ka}
\fmtCitation{(oku)\-kwe\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwe\-tu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to move slowly, to push along, to slide along}
\fmtExampleRL{Omu\-sa\-wu yaa\-kai\-ka ati\-mbe era ati\-bu\-ni tu\-ta\-ndi\-ki\-re oku\-kwe\-tu\-ka.}
\fmtExampleEN{The doctor has just arrived, and we are now moving slowly.}
\fmtSubentry(b)
\fmtSense{to move on the buttocks}
\fmtExampleRL{Omwa\-na wa na\-nka akwe\-tu\-ka mai\-ka\-la.}
\fmtExampleEN{So-and-so’s child crawls on his buttocks.}

\fmtLemma{kwe\-ya}
\fmtCitation{(oku)\-kwe\-ya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwe\-ye\-re}
\fmtSense{(of a dress) to sweep the floor, to brush the floor, to drag along the ground, to touch the ground}
\fmtExampleRL{Olu\-go\-ye lwa\-mwe lu\-ku\-kwe\-ya ansi.}
\fmtExampleEN{Her dress is sweeping the floor.}

\fmtLemma{kwe\-ya\-tya}
\fmtCrossreference{\textnormal{see} eyatya}

\fmtLemma{kwe\-zi}%
\fmtOne{1}
\fmtCrossreference{\textnormal{see also} mwezi}
\fmtCitation{(o)kwe\-zi}
\fmtPOS{n}%
\fmtNC{15}
\fmtSubentry(a)
\fmtSense{moon}
\fmtExampleRL{Okwe\-zi ku\-ta\-ndi\-ki\-re okwa\-ba omu nkoo\-ra.}
\fmtExampleEN{The moon has started waning.}
\fmtSubentry(b)
\fmtSense{month}
\fmtSubentry(c)
\fmtFixedexpression{kwe\-zi ku\-li (o)mu iya\-ngwe}
\fmtSense{(of the moon) to be full}
\fmtExampleRL{Okwe\-zi ku\-li mu iya\-ngwe.}
\fmtExampleEN{The moon is full.}
\fmtSubentry(d)
\fmtFixedexpression{kwezi kumwei}
\fmtSense{fast-growing variety of sweet potatoes (lit.\,sweet potatoes of one month)}
\fmtExampleRL{Ebi\-yaa\-ta bya kwe\-zi ku\-mwei bi\-nu\-ra mu\-no.}
\fmtExampleEN{The fast-growing variety of sweet potatoes is very delicious.}
\fmtSubentry(e)
\fmtFixedexpression{li (o)mu kwezi}
\fmtSense{to have mensturation, to menstruate}
\fmtExampleRL{Aka\-ru\-ma ka\-te\-ra ku\-mwi\-zi\-ra nga ali mu kwe\-zi.}
\fmtExampleEN{She usually gets cramps during menstruation.}

\fmtLemma{kwe\-zi}%
\fmtOne{2}
\fmtCitation{(o)kwe\-zi}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{type of banana}
\fmtExampleRL{Ama\-too\-ke ga kwe\-zi ti\-ga\-woo\-ma.}
\fmtExampleEN{The bananas called 'kwezi' are not tasty.}

\fmtLemma{kwi\-kya}
\fmtCitation{(o)kwi\-ca}
\fmtPOS{conj}
\fmtSense{until, till}
\fmtExampleRL{Wee\-go\-de\-re oku nsi\-ka okwi\-kya ni wa\-yi\-kya olu\-bbi\-mbi.}
\fmtExampleEN{Endure until you finish the part you are digging.}

\fmtLemma{kwi\-ni}
\fmtCitation{(o)kwi\-ni}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from En.}
\fmtSense{queen}
\fmtExampleRL{Okwi\-ni wa Bu\-nge\-re\-za akai\-ri\-ki\-re.}
\fmtExampleEN{The queen of England is very old.}

\fmtLemma{kwi\-ya\-ku}
\fmtCrossreference{\textnormal{see also} iyaku}
\fmtPOS{conj}
\fmtSense{unless, without, except that, apart from}

\fmtLemma{kwo}
\fmtSense{it (class 15 independent pronoun)}

\fmtLemma{kwoi\-ga}
\fmtCitation{(oku)\-kwoi\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kwoi\-ge\-re}
\fmtSense{to dig up}
\fmtExampleRL{Ka nya\-be nkwoi\-ge ebi\-yaa\-ta.}
\fmtExampleEN{Let me go and dig up the potatoes.}

\fmtLemma{kwoi\-gwa}
\fmtCitation{(oku)\-kwoi\-gwa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kwoi\-gwe\-re}
\fmtSense{to harrow, to draw a harrow over}
\fmtExampleRL{Etu\-ra\-ki\-ta eku\-kwoi\-ga endu\-mi\-ro ya Na\-lu\-yu\-lu.}
\fmtExampleEN{The tractor is harrowing Naluyulu's garden.}

\fmtLemma{kwoi\-ko\-lya}
\fmtCitation{(oku)\-kwoi\-ko\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwo\-koi\-rye}
\fmtSense{to riddle, to pose riddles}
\fmtExampleRL{Nje mmai\-te oku\-kwoi\-ko\-lya.}
\fmtExampleEN{I know how to pose a riddle.}

\fmtLemma{kwo\-kwo\-to\-ka}
\fmtCitation{(oku)\-kwo\-kwo\-to\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwo\-kwo\-to\-ke\-re}
\fmtSense{to move on one's buttocks}
\fmtExampleRL{Ya\-te\-mei\-bwe\-ku ama\-gu\-lu era ati\-bu\-ni akwo\-kwo\-to\-ka\-mbe.}
\fmtExampleEN{He had his legs amputated, he now moves on his buttocks.}

\fmtLemma{kwo\-kya}
\fmtCitation{(oku)\-kwo\-ky}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwo\-ke\-rye}
\fmtSense{to be hot}
\fmtExampleRL{Mwa\-be mu\-laa\-tye eme\-re eni omu ikoo\-ke eri\-kwo\-ca.}
\fmtExampleEN{Go and keep this food in the ash which is hot.}

\fmtLemma{kwo\-to\-ka}
\fmtCitation{(oku)\-kwo\-to\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kwo\-to\-ke\-re}
\fmtSense{(of a baby) to crawl}
\fmtExampleRL{Omwa\-na aku\-kwo\-to\-ka kwa\-ba wa nyi\-na.}
\fmtExampleEN{The baby is crawling towards her mother.}

\fmtLemma{kya}%
\fmtOne{1}
\fmtCitation{(oku)\-kya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kei\-re}
\fmtSubentry(a)
\fmtSense{to calm down, to reduce, to become smaller}
\fmtExampleRL{Ebi\-zu\-bu bi\-kei\-re\-mu.}
\fmtExampleEN{The problems have gotten smaller.}
\fmtSubentry(b)
\fmtSense{(of rain) to stop, to end}
\fmtExampleRL{Oi\-ke\-ndi agai\-ne oku\-kya.}
\fmtExampleEN{The rain has failed to stop falling.}
\fmtSubentry(c)
\fmtSense{to dawn, (of a day) to begin}
\fmtExampleRL{Obwi\-re bu\-ku\-kya ma\-ngu.}
\fmtExampleEN{Dawn is early.}

\fmtLemma{kya-}%
\fmtOne{2}
\fmtPOS{pf}
\fmtSense{still, no longer, not anymore, not yet [persistive aspect prefix on verbs]}
\fmtExampleRL{Empu\-li\-zi\-ga\-nya aka\-ti wai\-swe te\-kya\-li\-wo.}
\fmtExampleEN{There is still no communication between us.}

\fmtLemma{Kya\-ba\-zi\-nga}
\fmtCitation{(o)Kya\-ba\-zi\-nga}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{the title of Busoga king}
\fmtExampleRL{OKya\-ba\-zi\-nga mu\-ka\-ma wa Bu\-so\-ga.}
\fmtExampleEN{Kyabazinga is the king of Busoga.}

\fmtLemma{kya\-bi\-ree\-di}
\fmtPOS{interj}
\fmtSense{God forbid}
\fmtExampleRL{Kya\-bi\-ree\-di nje oku\-nywa enjaa\-ye.}
\fmtExampleEN{God forbid that I find myself smoking marijuana.}

\fmtLemma{kya\-bu\-la\-mbu\-li}
\fmtCitation{(e)kya\-bu\-la\-mbu\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{tourist attraction}
\fmtExampleRL{Uga\-nda eta n’e\-bya\-bu\-la\-mbu\-li ebi\-ku\-ka\-nya\-ku.}
\fmtExampleEN{Uganda has quite a number of tourist attractions.}

\fmtLemma{kya\-bu\-su\-bu\-zi}
\fmtCrossreference{\textnormal{see} kyabusuubuli}

\fmtLemma{kya\-bu\-suu\-bu\-li}
\fmtCitation{(e)kya\-bu\-suu\-bu\-li}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{merchandise, goods, produce}
\fmtExampleRL{Ewaa\-nywe mu\-ta\-mbuu\-ra mu kya\-bu\-suu\-bu\-li ki?}
\fmtExampleEN{What merchandise do you deal in where you live?}

\fmtLemma{kya\-bwe}
\fmtCrossreference{\textnormal{see} (-)abwe}

\fmtLemma{kya\-bwi\-re}
\fmtCitation{(e)kya\-bwi\-re}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{sex done at night}
\fmtExampleRL{Baa\-bi\-re ku kya\-bwi\-re.}
\fmtExampleEN{They have gone to have sex.}

\fmtLemma{kya\-co}
\fmtCitation{(e)kya\-co}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{flower}
\fmtExampleRL{Ekya\-co kya mu\-cu\-ngwai ki\-bba kye\-ru.}
\fmtExampleEN{The flower of an orange is always white.}

\fmtLemma{kya\-faa\-yo}
\fmtCitation{(e)kya\-faa\-yo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{fact}
\fmtExampleRL{Mmai\-te ekya\-faa\-yo ki\-mwe oku Uga\-nda.}
\fmtExampleEN{I know one fact about Uganda.}
\fmtSubentry(b)
\fmtSense{history}
\fmtExampleRL{Nko\-be\-ra ebya\-faa\-yo bya\-mu nga nkya\-li ku\-ku\-wa mu\-li\-mu.}
\fmtExampleEN{Tell me your history before I employ you.}

\fmtLemma{kya\-fu}
\fmtCitation{\textnormal{(e.g.}\,mu\-kya\-fu, ga\-kya\-fu\textnormal{)}}
\fmtPOS{adj}
\fmtSource{from Lu\-ga\-nda}
\fmtSense{dirty, filthy}
\fmtExampleRL{Ke\-ne\-ke\-ra amai\-zi ago ga\-kya\-fu.}
\fmtExampleEN{Filter that water, it's dirty.}

\fmtLemma{kya\-gu}
\fmtCitation{(e)kya\-gu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{cluster, hand of bananas, bunch of bananas}
\fmtExampleRL{Ei\-too\-ke li\-ni li\-ri\-na ebya\-gu ama\-ku\-mi abi\-ri.}
\fmtExampleEN{This banana has twenty clusters.}

\fmtLemma{Kya\-gwi\-re}
\fmtCitation{(o)Kya\-gwi\-re}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{Bu\-nyala/Buruuli name given to the god of lake Kyoga}
\fmtExampleRL{OKya\-gwi\-re ni\-ye mu\-ka\-ma wa Kyo\-ga.}
\fmtExampleEN{Kyagwire is the spirit responsible for Lake Kyoga.}

\fmtLemma{kyai\-go\-lo}
\fmtPOS{adv}
\fmtSubentry(a)
\fmtSense{in the evening}
\fmtExampleRL{Nai\-za kyai\-go\-lo.}
\fmtExampleEN{I will come in the evening.}
\fmtSubentry(b)
\fmtSense{supper}
\fmtExampleRL{Nku\-lya kyai\-go\-lo n’a\-baa\-na ba\-nge.}
\fmtExampleEN{I am having supper with my children.}

\fmtLemma{kyai\-jo}
\fmtCitation{(e)kyai\-jo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{part or piece of something}
\fmtExampleRL{Lee\-ta ekyai\-jo kya ki\-saa\-le oku\-zwa mu si\-too\-wa.}
\fmtExampleEN{Bring a piece of wood from the store.}

\fmtLemma{kyai\-rai}%
\fmtOne{1}
\fmtCitation{(e)kyai\-rai}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{story of long ago, tale about the past, narrative}
\fmtExampleRL{Nko\-be\-ra oku byai\-rai bya\-mu ebya bu\-to.}
\fmtExampleEN{Tell me any of your childhood stories.}

\fmtLemma{kyai\-rai}%
\fmtOne{2}
\fmtPOS{adv}
\fmtSense{afterward, finally, later}
\fmtExampleRL{Na\-mu\-li\-ndi\-re esaa\-wa itaa\-nu ekyai\-rai n’ai\-za.}
\fmtExampleEN{I waited for him for five hours and finally he came.}

\fmtLemma{kyai\-swe}
\fmtCrossreference{\textnormal{see} (-)aiswe}

\fmtLemma{kya\-ka}
\fmtCrossreference{\textnormal{see also} kyata}
\fmtCitation{(e)kya\-ka}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{heartburn}
\fmtExampleRL{Ndii\-re ebi\-tee\-re ati ekya\-ka ki\-ku\-nyi\-ta.}
\fmtExampleEN{I ate sliced potatoes and now I have heartburn.}

\fmtLemma{kya\-kaa\-ro}
\fmtCitation{(e)kya\-kaa\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{trail, track, spoor, trace}
\fmtExampleRL{Ki\-ni kya\-kaa\-ro kya njo\-ka.}
\fmtExampleEN{This is a snake trail.}
\fmtSubentry(b)
\fmtSense{official entrance}
\fmtExampleRL{Ki\-ni ni\-kyo kya\-kaa\-ro kya mu\-ka\-ma.}
\fmtExampleEN{This is the official entrance for the king.}

\fmtLemma{kya\-ka\-bbi}
\fmtCitation{(e)kya\-ka\-bbi}
\fmtPOS{adv}
\fmtSense{unfortunately, unluckily, regrettably, unhappily}
\fmtExampleRL{Ekya\-ka\-bbi ya\-fu\-ni\-re enda nga mu\-lwai\-re.}
\fmtExampleEN{Unfortunately, she got pregnant while she was sick.}

\fmtLemma{kya\-ka\-bi\-ri}
\fmtCitation{(e)kya\-ka\-bi\-ri}
\fmtPOS{adv}
\fmtSubentry(a)
\fmtSense{secondly, second}
\fmtExampleRL{Ekya\-ka\-bi\-ri, nkwe\-ndya obbee\-mu ama\-ge\-zi.}
\fmtExampleEN{Secondly, I want you to be wise.}
\fmtSubentry(b)
\fmtSense{primary two}
\fmtExampleRL{Aso\-ma ki\-bii\-na kya\-ka\-bi\-ri.}
\fmtExampleEN{He is in primary two.}

\fmtLemma{kya\-ka\-nai}
\fmtCrossreference{\textnormal{see also} (-)nai}
\fmtCitation{(e)kya\-ka\-nai}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{primary four}
\fmtExampleRL{Ali mu ki\-bii\-na kya\-ka\-nai.}
\fmtExampleEN{She is in primary four.}

\fmtLemma{kya\-ka\-sa\-tu}
\fmtCrossreference{\textnormal{see also} (-)satu}
\fmtCitation{(e)kya\-ka\-sa\-tu}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{primary three}
\fmtExampleRL{Azwi\-re mu ki\-bii\-na kya\-ka\-sa\-tu.}
\fmtExampleEN{He has completed primary three.}

\fmtLemma{kya\-ka\-taa\-nu}
\fmtCitation{(e)kya\-ka\-taa\-nu}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{primary five}
\fmtExampleRL{Alwi\-re mu kya\-ka\-taa\-nu.}
\fmtExampleEN{She stayed a long time in primary five.}

\fmtLemma{kya\-ki}
\fmtCitation{(e)kya\-ki}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{flower}
\fmtExampleRL{Nta\-ka ebya\-ki mu\-no.}
\fmtExampleEN{I really love flowers.}

\fmtLemma{kya\-ku\-ba\-nza}
\fmtCrossreference{\textnormal{see also} banza}
\fmtCitation{(e)kya\-ku\-ba\-nza}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{primary one}
\fmtExampleRL{Aso\-ma kya\-ku\-ba\-nza.}
\fmtExampleEN{He is in primary one.}

\fmtLemma{kya\-ku\-bo\-ne\-ra\-ku}
\fmtCitation{(e)kya\-ku\-bo\-ne\-ra\-ku}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{example}
\fmtExampleRL{Mpee\-rya\-yo ebya\-ku\-bo\-ne\-ra\-ku bi\-sa\-tu.}
\fmtExampleEN{Give me three examples.}

\fmtLemma{kya\-ku\-lwa\-ni\-sya}
\fmtCrossreference{\textnormal{see also} lwanisya}
\fmtCitation{(e)kya\-ku\-lwa\-ni\-sya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{weapon}
\fmtExampleRL{Ru\-ssia eta n’e\-bya\-ku\-lwa\-ni\-sa eby’a\-maa\-ni.}
\fmtExampleEN{Russia has very strong weapons.}

\fmtLemma{kya\-ku\-lya}
\fmtCitation{(e)kya\-ku\-lya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{food, something to eat}
\fmtExampleRL{Mpee\-rya oku kya\-ku\-lya.}
\fmtExampleEN{Give me something to eat.}

\fmtLemma{kya\-ku\-nywa}
\fmtCrossreference{\textnormal{see also} nywa}
\fmtCitation{(e)kya\-ku\-nywa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{drink, beverage}
\fmtExampleRL{Mpee\-rya oku kya\-ku\-nywa.}
\fmtExampleEN{Give me something to drink.}

\fmtLemma{kya\-kwe\-ku\-mi\-sya}
\fmtCitation{(e)kya\-kwe\-ku\-mi\-sya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{shield, protector}
\fmtExampleRL{Eki\-ri\-ŋa\-li\-ŋa ki\-ko\-le\-sya obu\-li\-mbo nka ekya\-kwe\-ku\-mi\-sya.}
\fmtExampleEN{A snail uses slime for protection.}
\fmtSense{reserve, fall back, standby}

\fmtLemma{kya\-kwe\-kwa\-ti\-sya}
\fmtCitation{(e)kya\-kwe\-kwa\-ti\-sya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{excuse, pretext}
\fmtExampleRL{Ekyo kya\-kwe\-kwa\-ti\-sya, ti\-wa\-koo\-re mu\-li\-mu gwa\-mu.}
\fmtExampleEN{That is an excuse, you did not do your work.}

\fmtLemma{kya\-kwe\-wuu\-nya}
\fmtCitation{(e)kya\-kwe\-wuu\-nya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wonder, miracle}
\fmtExampleRL{To\-ku\-bo\-na nga ekyo kya\-kwe\-wuu\-nya?}
\fmtExampleEN{Don’t you see, that is a miracle.}

\fmtLemma{kya\-kyo}
\fmtCitation{(e)kya\-kyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{flower of spider weed (\textit{Cleome gynandra})}
\fmtExampleRL{Ebya\-kyo bya yo\-byo bi\-bba bye\-ru.}
\fmtExampleEN{Spider weed flowers are white.}

\fmtLemma{kya\-la}%
\fmtOne{1}
\fmtCitation{(e)kya\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{finger}
\fmtExampleRL{Aye\-nde\-ke\-re ekya\-la kya\-mwe.}
\fmtExampleEN{He broke his finger.}
\fmtSubentry(b)
\fmtSense{fingernail}
\fmtExampleRL{Ata n’e\-kya\-la eki\-koo\-to.}
\fmtExampleEN{He has a big fingernail.}
\fmtSubentry(c)
\fmtSense{claw}
\fmtExampleRL{Ombwe\-ne abba n’e\-bya\-la.}
\fmtExampleEN{A dog has claws.}
\fmtSubentry(d)
\fmtSense{fingerprint, thumbprint}
\fmtExampleRL{Atai\-re ekya\-la oku ki\-wa\-ndii\-ko.}
\fmtExampleEN{He put a fingerprint on the document.}
\fmtSubentry(e)
\fmtSense{thumb}
\fmtExampleRL{Ekya\-la eki\-sai\-za ekya mwa\-na ki\-rwai\-re.}
\fmtExampleEN{The baby's thumb is infected.}

\fmtLemma{kya\-la}%
\fmtOne{2}
\fmtCitation{(e)kya\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{ghost, type of spirit which is believed to assume physical form}
\fmtExampleRL{Ekya\-la ekyai\-zi\-re oku kya\-lo kyai\-swe ki\-kwa\-ta aba\-ka\-li.}
\fmtExampleEN{The ghost that came to our village rapes women.}

\fmtLemma{kya\-la}%
\fmtOne{3}
\fmtCitation{(oku)\-kya\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kyai\-re}
\fmtSense{to visit}
\fmtExampleRL{Aya\-bi\-re ku\-kya\-la mu ba\-ga\-nda.}
\fmtExampleEN{He has gone to visit his relatives.}

\fmtLemma{kya\-laa\-ni}
\fmtCitation{(e)kya\-laa\-ni}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from Persian via Sw.}
\fmtSense{sewing machine}
\fmtExampleRL{Mmu\-gu\-lii\-re ekya\-laa\-ni.}
\fmtExampleEN{I bought a sewing machine for him.}

\fmtLemma{kya\-li}
\fmtCrossreference{\textnormal{see also} kya- \textnormal{and} li}
\fmtCitation{\textnormal{(e.g.}\,nkya\-li, akya\-li, tu\-kya\-li\textnormal{)}}
\fmtSense{before (doing something)}
\fmtExampleRL{Ya\-mu\-sa\-ngi\-re akya\-li ku\-fa.}
\fmtExampleEN{She came before he had passed on.}

\fmtLemma{kya\-li\-sya}
\fmtCitation{(oku)\-kya\-li\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kya\-li\-si\-rye}
\fmtSense{to host, to receive (guests)}
\fmtExampleRL{Nte\-ra mu\-no oku\-kya\-li\-sya aba\-ge\-nyi.}
\fmtExampleEN{I host guests very often.}

\fmtLemma{kya\-lo}%
\fmtOne{1}
\fmtCitation{(e)kya\-lo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{village}
\fmtExampleRL{Aya\-bi\-re omu kya\-lo ku\-bo\-na ba\-byai\-re baa\-mwe.}
\fmtExampleEN{He has gone to the village to see his parents.}
\fmtSubentry(b)
\fmtSense{Local Council one (LC1, a tier of administration)}
\fmtExampleRL{Ekya\-lo ki\-ni ki\-ko\-le\-bwa aba\-ntu lu\-ku\-mi.}
\fmtExampleEN{This LC1 is made up of one thousand people.}

\fmtLemma{kya\-lo}%
\fmtOne{2}
\fmtCitation{(o)kya\-lo}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{vocalist, singer}
\fmtExampleRL{Aba\-kya\-lo ba\-ya\-bi\-re e Ka\-mpa\-la kwe\-mba.}
\fmtExampleEN{The vocalists went to Kampala to sing.}

\fmtLemma{kya\-lu\-be\-re\-be\-rye}
\fmtCitation{(e)kya\-lu\-be\-re\-be\-rye}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{primary one}
\fmtExampleRL{Naa\-ni aku\-so\-ma ekya\-lu\-be\-rye\-be\-rye?}
\fmtExampleEN{Who is studying in primary one?}

\fmtLemma{kya\-lu\-ka}
\fmtCitation{(oku)\-kya\-lu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kya\-lu\-ki\-re}
\fmtSense{(of a visitor) to end a visit, to depart, to leave}
\fmtExampleRL{Omu\-ge\-ni akya\-lu\-ki\-re.}
\fmtExampleEN{The visitor has ended her visit.}

\fmtLemma{kya\-lu\-mwei}
\fmtPOS{adv}
\fmtSense{once, one time}
\fmtExampleRL{Ai\-za kya\-lu\-mwei omu okwe\-zi.}
\fmtExampleEN{He comes once a month.}

\fmtLemma{kya\-ma}%
\fmtOne{1}
\fmtCitation{(oku)\-kya\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kya\-mi\-re}
\fmtSubentry(a)
\fmtSense{to bend}
\fmtExampleRL{Ekyo\-ma ki\-ni kya\-kya\-mi\-re.}
\fmtExampleEN{This metal is bent.}
\fmtSubentry(b)
\fmtSense{to turn, to change direction, to make a turn}
\fmtExampleRL{Ni wai\-ka omu mai\-so kya\-ma oku mu\-ko\-no ogu\-mo\-so.}
\fmtExampleEN{Go straight and then turn to the left.}
\fmtSubentry(c)
\fmtSense{to go astray, to err}
\fmtExampleRL{Ki\-bbi oku\-kya\-ma n’o\-zwa oku mu\-la\-mwa.}
\fmtExampleEN{It is bad to go astray and leave the right path.}

\fmtLemma{kya\-ma}%
\fmtOne{2}
\fmtCitation{(e)kya\-ma}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{secret}
\fmtExampleRL{Ti\-ndi\-na kya\-ma kyo\-na\-kyo\-na.}
\fmtExampleEN{I don’t have any secrets.}

\fmtLemma{kya\-ma\-ge\-ro}
\fmtCitation{(e)kya\-ma\-ge\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{miracle, wonder}
\fmtExampleRL{OYe\-su ya\-koo\-re ebya\-me\-ge\-ro bi\-ngi mu\-no.}
\fmtExampleEN{Jesus performed many miracles.}

\fmtLemma{kya\-ma\-gu\-zi}
\fmtCitation{(e)kya\-ma\-gu\-zi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{merchandise, goods, commodity}
\fmtExampleRL{Alee\-te\-re ebya\-ma\-gu\-zi bya\-mwe omu ka\-ta\-le.}
\fmtExampleEN{He brought his goods to the market.}

\fmtLemma{kya\-ma\-mbya}
\fmtCitation{(e)kya\-ma\-mbya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{breakfast}
\fmtExampleRL{Tu\-ku\-lya kya ma\-mbya.}
\fmtExampleEN{We are having breakfast.}

\fmtLemma{kya\-mbai\-tu\-ka}
\fmtCitation{(oku)\-kya\-mbai\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kya\-mbai\-tu\-ki\-re}
\fmtSense{to splash}
\fmtExampleRL{Amai\-zi ga\-ku\-kya\-mbai\-tu\-ka oku\-zwa mu nsu\-wa.}
\fmtExampleEN{Water is splashing out of the pot.}

\fmtLemma{kya\-mbu}
\fmtCitation{(e)kya\-mbu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{landing site}
\fmtExampleRL{Emee\-ri ei\-ki\-re oku kya\-mbu.}
\fmtExampleEN{The ship has reached the landing site.}

\fmtLemma{kya\-mi}
\fmtCitation{\textnormal{(e.g.}\,ki\-kya\-mi, lu\-kya\-mi)}
\fmtPOS{adj}
\fmtSense{bent}
\fmtExampleRL{Olu\-kii\-kii\-di lwa mee\-za lu\-kya\-mi.}
\fmtExampleEN{The edge of the table is not straight.}

\fmtLemma{kya\-mii\-ro}
\fmtCitation{(e)kya\-mii\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{announcement, notification, broadcasting}
\fmtExampleRL{Mu\-wuu\-re ekya\-mii\-ro ki\-ni.}
\fmtExampleEN{Listen to this announcement.}

\fmtLemma{kya\-mi\-sa\-na}
\fmtCitation{(e)kya\-mi\-sa\-na}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{lunch}
\fmtExampleRL{Iza tu\-lye ekya\-mi\-sa\-na.}
\fmtExampleEN{Come and we will have lunch.}

\fmtLemma{kya\-mu}%
\fmtOne{1}
\fmtCrossreference{\textnormal{see} (-)amu}

\fmtLemma{kya\-mu}%
\fmtOne{2}
\fmtCitation{\textnormal{(e.g.}\,mu\-kya\-mu, ki\-kya\-mu, bi\-kya\-mu\textnormal{)}}
\fmtPOS{adj}
\fmtSense{wrong, incorrect, false}
\fmtExampleRL{Oku nso\-nga edi nii\-we mu\-kya\-mu.}
\fmtExampleEN{You are the one who is wrong about that matter.}

\fmtLemma{kya\-mu\-ga\-yai\-re}%
\fmtOne{1}
\fmtCitation{(e)kya\-mu\-ga\-yai\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{family planning, birth control}
\fmtExampleRL{Ali ku mbya\-la ya kya\-mu\-ga\-yai\-re.}
\fmtExampleEN{She is on birth control.}

\fmtLemma{kya\-mu\-ga\-yai\-re}%
\fmtOne{2}
\fmtCitation{(e)kya\-mu\-ga\-yai\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{lazy person}
\fmtExampleRL{Omwa\-la kya\-mu\-ga\-yai\-re amaa\-re n’a\-ya\-ka\-la ebi\-buu\-lyo.}
\fmtExampleEN{The lazy girl has finally passed the examination.}

\fmtLemma{kya\-mu\-gi\-sa}
\fmtCrossreference{\textnormal{see} kyamukisa}

\fmtLemma{kya\-mu\-ka}
\fmtCitation{(oku)\-kya\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kya\-mu\-ki\-re}
\fmtSense{to take a turn, to turn}
\fmtExampleRL{Kya\-mu\-ki\-ra\-ku awo le\-ke osi\-mbe.}
\fmtExampleEN{Take a turn there and then go straight.}

\fmtLemma{kya\-mu\-kaa\-ga}
\fmtCitation{(e)kya\-mu\-kaa\-ga}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{primary six}
\fmtExampleRL{Nku\-so\-ma kya\-mu\-kaa\-ga.}
\fmtExampleEN{I am in primary six.}

\fmtLemma{kya\-mu\-ki\-sa}
\fmtCitation{(e)kya\-mu\-ki\-sa}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{fortune, good luck}
\fmtExampleRL{Kya\-mu\-ki\-sa oRu\-th ya\-wu\-mwi\-re nka pu\-lo\-fee\-sa.}
\fmtExampleEN{It is fortunate that Ruth retired as a professor.}

\fmtLemma{kya\-mu\-kwe\-nu}
\fmtCitation{(o)kya\-mu\-kwe\-nu}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{flirt, coquette}
\fmtExampleRL{Okya\-mu\-kwe\-nu aku\-bu\-lu\-ngu\-ta\-nya abaa\-na.}
\fmtExampleEN{A flirt is making the children disorganised.}

\fmtLemma{kya\-mu\-la}
\fmtCitation{(oku)\-kya\-mu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kya\-mwi\-re}
\fmtSubentry(a)
\fmtSense{to straighten, to make straight}
\fmtExampleRL{Ofu\-ndi akya\-mwi\-re ekyo\-ma eki\-bbai\-re kye\-go\-de\-re.}
\fmtExampleEN{The builder has straightened the bent metal.}
\fmtSubentry(b)
\fmtSense{to turn}
\fmtExampleRL{Etu\-ra\-ki\-ta ebbai\-re eku\-lu\-ma ani gi\-kya\-mu\-le eru\-me nge\-di.}
\fmtExampleEN{Turn the tractor which was ploughing this side, so that it can plough the other side.}
\fmtSubentry(c)
\fmtSense{to stop, to keep from, to prevent from}
\fmtExampleRL{Kya\-mu\-la omwa\-na ale\-ke kwa\-ba eyo.}
\fmtExampleEN{Stop the child from going that way.}

\fmtLemma{kya\-mu\-sa\-nju}
\fmtCitation{(e)kya\-mu\-sa\-nju}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{primary seven}
\fmtExampleRL{Ama\-li\-rii\-rye ekya\-mu\-sa\-nju.}
\fmtExampleEN{She has completed primary seven.}

\fmtLemma{kya\-mwe\-(i)}
\fmtCrossreference{\textnormal{see} amwe\textnormal{{\fmtOne{1}}}}

\fmtLemma{kya\-na}%
\fmtOne{1}
\fmtCitation{(e)kya\-na}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{course particles (of cereals)}
\fmtExampleRL{Eki\-poo\-li ki\-si\-gai\-re\-mu ebya\-na.}
\fmtExampleEN{Some particles remained in the groundnut paste.}
\fmtSubentry(b)
\fmtSense{pupa, chrysalis}
\fmtExampleRL{Ti\-ndya kya\-na kya nzo\-ki.}
\fmtExampleEN{I don’t eat bees’ pupa.}
\fmtSubentry(c)
\fmtFixedexpression{kya\-na kya ike\-ndi}
\fmtSense{raindrop}
\fmtExampleRL{Nku\-wuu\-ra ebya\-na bya ike\-ndi.}
\fmtExampleEN{I hear raindrops.}
\fmtSubentry(d)
\fmtSense{naughty behaviour, bad behaviour, disobedience}
\fmtExampleRL{Omwa\-la owa kya\-na ta\-so\-bo\-la ku\-su\-mbi\-rwa!}
\fmtExampleEN{A disobedient girl cannot be married!}

\fmtLemma{kya\-na}%
\fmtOne{2}
\fmtCitation{(e)kya\-na}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wagon, car trailer}
\fmtExampleRL{Ekya\-na kya mo\-to\-ka kye\-nde\-ke\-re\-ku.}
\fmtExampleEN{The car trailer is disconnected.}

\fmtLemma{kya\-nda}
\fmtCitation{(e)kya\-nda}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{dry season, drought, dry spell}
\fmtExampleRL{Ekya\-nda ki\-ta\-ndi\-ki\-re.}
\fmtExampleEN{The drought has started.}

\fmtLemma{kya\-nda\-nda}
\fmtCitation{(e)kya\-nda\-nda}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{tepid temperature, warmish temperature, mild temperature}
\fmtExampleRL{Olwa\-le\-ro lwa kya\-nda\-nda.}
\fmtExampleEN{Today is tepid.}

\fmtLemma{kya\-ndii\-ko}
\fmtCitation{(e)kya\-ndii\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{document, form, paper}
\fmtExampleRL{Ai\-zi\-re n’e\-kya\-ndii\-ko oku\-zwa owa faa\-za.}
\fmtExampleEN{He came with a document from a Catholic priest.}

\fmtLemma{kya\-ndya}
\fmtCitation{(e)kya\-ndya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{food, edible things}
\fmtExampleRL{Ebya\-ndya bi\-ku\-tu\-ga\-dya ewai\-swe.}
\fmtExampleEN{We have challenges with food at home.}

\fmtLemma{kya\-ngaa\-ra}
\fmtCitation{(e)kya\-ngaa\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{empty space, empty room, open area, opening, emptiness}
\fmtExampleRL{Abaa\-na bai\-kai\-re mu kya\-ngaa\-ra.}
\fmtExampleEN{The children sit in an empty room.}
\fmtSubentry(b)
\fmtSense{house without shutters}
\fmtExampleRL{Aba\-mu\-le\-kwa ba\-raa\-ra mu kya\-ngaa\-ra.}
\fmtExampleEN{The orphans sleep in a house without shutters.}

\fmtLemma{kya\-nge}
\fmtCrossreference{\textnormal{see} ange}

\fmtLemma{kya\-ngii\-re}
\fmtCitation{(e)kya\-ngii\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{accusation, allegation}
\fmtExampleRL{Ekyo kya\-ngii\-re.}
\fmtExampleEN{That is just an allegation.}

\fmtLemma{kya\-ngwe}
\fmtCitation{(e)kya\-ngwe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sponge (for bathing)}
\fmtExampleRL{Na\-bi\-sya ekya\-ngwe.}
\fmtExampleEN{Use a sponge to bathe yourself.}

\fmtLemma{kya\-ni\-ki\-ro}
\fmtCitation{(e)kya\-ni\-ki\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{drying place}
\fmtExampleRL{Ya\-ni\-ki\-ra obwi\-ta oku kya\-ni\-ki\-ro.}
\fmtExampleEN{Dry the millet at the drying place.}

\fmtLemma{kya\-njo}
\fmtCitation{(e)kya\-njo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{nonsense, foolishness}
\fmtExampleRL{Oyo aku\-ba\-za bi\-ga\-mbo bya kya\-njo.}
\fmtExampleEN{That one is speaking nonsense.}
\fmtSubentry(b)
\fmtSense{recklessness, carelessness}
\fmtExampleRL{Le\-ke\-ra\-wo oku\-ko\-la ebya kya\-njo.}
\fmtExampleEN{Stop doing things recklessly.}
\fmtSubentry(c)
\fmtSense{incompetence}
\fmtExampleRL{Ti\-nta\-ka ku\-bba na mu\-ka\-li wa kya\-njo.}
\fmtExampleEN{I do not want to live with an incompetent woman.}
\fmtSubentry(d)
\fmtSense{poor quality}
\fmtExampleRL{OBu\-gu\-mi\-rwa ya\-koo\-re ebya kya\-njo.}
\fmtExampleEN{Bugumirwa did the work shoddily.}
\fmtSubentry(e)
\fmtSense{triviality, insignificance}
\fmtExampleRL{Ebyo by'a\-ku\-ba\-za bya kya\-njo.}
\fmtExampleEN{What he is saying is trivial.}
\fmtSubentry(f)
\fmtSense{just, simply, anyhow}
\fmtExampleRL{Ya\-ta\-mbwi\-re kya\-njo.}
\fmtExampleEN{He simply walked away.}

\fmtLemma{kya\-nka\-la\-na}
\fmtCitation{(oku)\-kya\-nka\-la\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kya\-nka\-lai\-ne}
\fmtSubentry(a)
\fmtSense{to be annoyed}
\fmtExampleRL{Aba\-ntu ba\-kya\-nka\-lai\-ne ng’o\-Bo\-bi wi\-ne ba\-mu\-kwai\-te.}
\fmtExampleEN{People were annoyed when Bobi Wine was arrested.}
\fmtSubentry(b)
\fmtSense{to be disorderly, to be disorganised}
\fmtExampleRL{Eki\-sii\-ka kya mwa\-la ki\-kya\-nka\-lai\-ne.}
\fmtExampleEN{The girl's room is disorganised.}

\fmtLemma{kya\-nko\-mee\-ro}
\fmtCitation{(e)kya\-nko\-mee\-ro}
\fmtPOS{adv}
\fmtSense{finally, eventually, ultimately, in the end}
\fmtExampleRL{Ekya\-nko\-mee\-ro olwi\-ge lu\-yi\-gu\-ki\-re.}
\fmtExampleEN{Finally, the door opened.}

\fmtLemma{kya\-nkoo\-le}
\fmtCitation{(o)kya\-nkoo\-le}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{common water hyacinth (\textit{Eichhornia crassipes})}
\fmtExampleRL{Okya\-nkoo\-le ye\-yo\-ngei\-re oku nya\-nja.}
\fmtExampleEN{Water hyacinth has spread in the lake.}

\fmtLemma{kya\-nya}%
\fmtOne{1}
\fmtCitation{(e)kya\-nya}
\fmtPOS{n}%
\fmtNC{7}
\fmtSubentry(a)
\fmtSense{chance}
\fmtExampleRL{Mpee\-rya ekya\-nya nsi\-moo\-le\-ku.}
\fmtExampleEN{Give me a chance to talk.}
\fmtSubentry(b)
\fmtSense{opportunity, occasion, opening}
\fmtExampleRL{Ebya\-nya bya ku\-ko\-la oku la\-di\-yo bi\-wei\-re\-wo.}
\fmtExampleEN{There are no more opportunities to work on the radio.}
\fmtSubentry(c)
\fmtSense{room, unoccupied area, space}

\fmtLemma{kya\-nya}%
\fmtOne{2}
\fmtCitation{(oku)\-kya\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kya\-nyi\-re}
\fmtSense{to pile up, to heap}
\fmtExampleRL{Kya\-nya ei\-soi\-gi eryo amwei.}
\fmtExampleEN{Pile up those dry banana leaves together.}

\fmtLemma{kya\-nyu\-maa\-nyu\-ma}
\fmtCrossreference{\textnormal{see also} kinaanyuma\-nyu\-ma}
\fmtPOS{adv}
\fmtSense{in reverse, backwards}
\fmtExampleRL{Emo\-to\-ka aku\-gi\-vu\-ga kya\-nyu\-maa\-nyu\-ma.}
\fmtExampleEN{He is driving the car in reverse.}

\fmtLemma{kya\-nywe}
\fmtCrossreference{\textnormal{see} nywe}

\fmtLemma{kya\-nzi}
\fmtCitation{(e)kya\-nzi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wooden milk jug}
\fmtExampleRL{Ama\-tai nga\-su\-ki\-re mu kya\-nzi.}
\fmtExampleEN{I put the milk into the wooden milk jug.}

\fmtLemma{kya\-nzo}
\fmtCitation{(e)kya\-nzo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{skin, hide}
\fmtExampleRL{Ekya\-nzo ki\-zwa ku bi\-so\-lo.}
\fmtExampleEN{Skin is removed from animals.}

\fmtLemma{kya\-pa}%
\fmtOne{1}
\fmtCitation{(e)kya\-pa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{mark, spot}
\fmtExampleRL{Onte wa\-nge ali\-ku ekya\-pa oku mu\-twe.}
\fmtExampleEN{My cow has a mark on its head.}

\fmtLemma{kya\-pa}%
\fmtOne{2}
\fmtCitation{(e)kya\-pa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from Persian via Sw.}
\fmtSense{land title}
\fmtExampleRL{Nfu\-ni\-ra ekya\-pa kya ita\-ka\-li lya\-nge.}
\fmtExampleEN{Secure a land title on my land for me.}

\fmtLemma{kya\-rai\-zu\-ma}
\fmtPOS{adv}
\fmtSense{aimlessly, without purpose, without direction}
\fmtExampleRL{Aya\-bi\-re kya\-rai\-zu\-ma.}
\fmtExampleEN{He has left without having a real destination.}

\fmtLemma{kya\-ru\-mwe}
\fmtPOS{adv}
\fmtSense{for good, forever, definitively}
\fmtExampleRL{Ya\-ya\-bii\-re kya\-ru\-mwe.}
\fmtExampleEN{He left for good.}

\fmtLemma{kya\-sa}%
\fmtOne{1}
\fmtCitation{(o)kya\-sa}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{cow with white stripes on the head}
\fmtExampleRL{Onte wa\-nge kya\-sa.}
\fmtExampleEN{My cow has white stripes on its head.}

\fmtLemma{kya\-sa}%
\fmtOne{2}
\fmtCitation{(e)kya\-sa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{century}
\fmtExampleRL{Ekya\-sa kyai\-ku\-mi n’o\-mwe\-ndai kya\-wei\-re\-ku ku\-sai.}
\fmtExampleEN{The 19th century ended very well.}

\fmtLemma{kya\-si}
\fmtCitation{(e)kya\-si}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from Sw.}
\fmtSubentry(a)
\fmtSense{shot, gunfire, blast, bang}
\fmtExampleRL{Mpu\-lii\-re ekya\-si kya mbu\-ndu.}
\fmtExampleEN{I heard the sound of a gunshot.}
\fmtSubentry(b)
\fmtSense{sneeze}
\fmtExampleRL{Omu\-ka\-li aya\-si\-mwi\-re ekya\-si.}
\fmtExampleEN{The woman sneezed loudly.}

\fmtLemma{kya\-ta}
\fmtCrossreference{\textnormal{see also} kyaka}
\fmtCitation{(e)kya\-ta}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{heartburn}
\fmtExampleRL{Ebi\-yaa\-ta n’e\-bi\-yi\-mba bi\-ree\-ta ekya\-ta.}
\fmtExampleEN{Sweet potatoes and beans cause heartburn.}

\fmtLemma{kya\-wa}
\fmtCitation{(oku)\-kya\-wa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kya\-wi\-re}
\fmtSense{to hate, to despise}
\fmtExampleRL{Omu\-ka\-li oyo akya\-wa aba\-la\-mu baa\-mwe mu\-no.}
\fmtExampleEN{That woman really hates her in-laws.}

\fmtLemma{kya\-wa\-ndii\-kii\-rwe}
\fmtCitation{(e)kya\-wa\-ndii\-kii\-rwe}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{scripture}
\fmtExampleRL{Aso\-me\-re ekya\-wa\-ndii\-kii\-rwe n’o\-bu\-gu\-mu.}
\fmtExampleEN{He read the scripture with confidence.}

\fmtLemma{kya\-wo\-ngo}
\fmtCitation{(e)kya\-wo\-ngo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{witchcraft, sorcery, magic}
\fmtExampleRL{Ebya\-wo\-ngo twa\-bi\-kee\-rwe.}
\fmtExampleEN{We are fed up with witchcraft.}

\fmtLemma{kya\-yi\-ta}
\fmtCitation{(e)kya\-yi\-ta}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{banana peel}
\fmtExampleRL{Embu\-li zi\-ta\-ka oku\-lya ebya\-yi\-ta.}
\fmtExampleEN{Goats like eating banana peels.}

\fmtLemma{kya\-zo}
\fmtCitation{(e)kya\-zo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{papyrus mat}
\fmtExampleRL{Ekya\-zo ba\-ki\-si\-muu\-ra\-ku ma\-gu\-lu.}
\fmtExampleEN{The papyrus mat is used for wiping the feet.}

\fmtLemma{kye\-bbi\-sii\-ri}
\fmtPOS{adv}
\fmtSense{stealthily}
\fmtExampleRL{Oyo ata\-mbu\-la kye\-bbi\-sii\-ri.}
\fmtExampleEN{That one walks stealthily.}

\fmtLemma{kye\-be\-re\-ka}
\fmtCitation{(e)kye\-be\-re\-ka}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{gate crasher}
\fmtExampleRL{Ba\-kye\-be\-ra\-ka ba\-sii\-si\-re omu\-ko\-lo.}
\fmtExampleEN{The gatecrashers ruined the ceremony.}

\fmtLemma{kye\-bi\-ki\-ro}
\fmtCitation{(e)kye\-bi\-ki\-ro}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{premonition, feeling in one's bones, bad omen, foreboding}
\fmtExampleRL{Ei\-zo ya\-fu\-ni\-re ekye\-bi\-ki\-ro.}
\fmtExampleEN{Yesterday he had a premonition.}

\fmtLemma{kye\-bo\-ne\-ro}
\fmtCitation{(e)kye\-bo\-ne\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wonder, miracle, marvel}
\fmtExampleRL{Kye\-bo\-ne\-ro nk’e\-nke\-nde ya wa\-lu\-bu.}
\fmtExampleEN{It's a wonder to see a chameleon dancing. (lit.\,It is a wonder to see the waist of a chameleon.) (Ruruuli-Lu\-nyala proverb)}

\fmtLemma{kye\-dee\-dyo}
\fmtCitation{(e)kye\-dee\-dyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{swing (a seat suspended by ropes or chains)}
\fmtExampleRL{Ai\-kai\-re ku kye\-dee\-dyo.}
\fmtExampleEN{She sits on a swing.}

\fmtLemma{kye\-fee\-ra}
\fmtCitation{(e)kye\-fee\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{animal dead of natural causes}
\fmtExampleRL{Aba\-si\-raa\-mu ti\-ba\-lya kye\-fee\-ra.}
\fmtExampleEN{Moslems do not eat animals which died of natural causes.}

\fmtLemma{kye\-ge\-se\-ryo}
\fmtCitation{(e)kye\-ge\-se\-ryo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{college}
\fmtExampleRL{Aya\-bi\-re mu kye\-ge\-se\-ryo kwe\-ga bya\-ma\-sa\-nya\-la\-ze.}
\fmtExampleEN{She has gone to college to train as an electrician.}

\fmtLemma{kyei\-bbi\-sei\-bbi\-se}
\fmtPOS{adv}
\fmtSense{secretly, stealthily}
\fmtExampleRL{Lwa\-ki ebi\-ntu bya\-mu obi\-ko\-la kyei\-bbi\-sei\-bbi\-se?}
\fmtExampleEN{Why do you do things secretly?}

\fmtLemma{kyei\-ru\-ndu\-ko}
\fmtCitation{(e)kyei\-ru\-ndu\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{shock, fright, scare, fear, alarm}
\fmtExampleRL{Afu\-ni\-re ekyei\-ru\-ndu\-ko n’a\-mu\-boi\-ne.}
\fmtExampleEN{He got a shock when he saw him.}

\fmtLemma{kye\-jo}
\fmtCitation{(e)kye\-jo}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{naughty behaviour, bad manners, disobedience}
\fmtExampleRL{Mwa\-na\-we ota wa kye\-jo.}
\fmtExampleEN{You child, you are naughty.}

\fmtLemma{kye\-ko}
\fmtCitation{(e)kye\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sash, belt, waistband}
\fmtExampleRL{Ekye\-ko ki\-gu\-mu.}
\fmtExampleEN{The sash is durable.}

\fmtLemma{kye\-ko\-la\-mu\-jwi\-ga}
\fmtCitation{(e)kye\-ko\-la\-mu\-jwi\-ga}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{tiger}
\fmtExampleRL{Okye\-ko\-la\-mu\-jwi\-ga tu\-mwa\-ja mu zoo.}
\fmtExampleEN{Tigers are found in the zoo.}

\fmtLemma{kye\-ku\-ru\-ndya}
\fmtCitation{(e)kye\-ku\-ru\-ndya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{reptile}
\fmtExampleRL{Ebye\-ku\-ru\-ndya bi\-bba mu mai\-zi n’o\-ku\-lu\-ka\-lu.}
\fmtExampleEN{Reptiles are found in water and on land.}

\fmtLemma{kye\-ku\-ru\-ngi}
\fmtCitation{(e)kye\-ku\-ru\-ngi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{circle}
\fmtExampleRL{Ku\-bba ekye\-ku\-ru\-ngi omu ki\-ta\-bo kya\-mu.}
\fmtExampleEN{Draw a circle in your book.}

\fmtLemma{kye\-mee\-ro}
\fmtCitation{(e)kye\-mee\-ro}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{sudden shoulder pain}
\fmtExampleRL{Obu\-lwai\-re bwa kye\-mee\-ro bu\-lu\-ma mu\-no.}
\fmtExampleEN{A sudden shoulder pain hurts a lot.}

\fmtLemma{kye\-me\-ro}
\fmtCitation{(e)kye\-me\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{foot}
\fmtExampleRL{Omu\-ntu aye\-mee\-ra ku bwe\-me\-ro.}
\fmtExampleEN{A person stands on his feet.}
\fmtSubentry(b)
\fmtSense{sole (of feet)}
\fmtExampleRL{Omu\-ntu ata n’e\-kye\-me\-ro eki\-bba\-mbaa\-ta\-li abba wa ki\-fii\-ni.}
\fmtExampleEN{A person with flat feet is believed to have bad luck.}

\fmtLemma{kye\-mo}
\fmtCitation{(e)kye\-mo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{footstep, track, footprint}
\fmtExampleRL{Ki\-ni ekye\-mo kya ntaa\-ma.}
\fmtExampleEN{This is the footprint of a sheep.}

\fmtLemma{kye\-na}
\fmtCitation{(oku)\-kye\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kye\-ne\-re}
\fmtSense{to curse}
\fmtExampleRL{Lwa\-ki okye\-na abaa\-na baa\-mu?}
\fmtExampleEN{Why do you curse your children?}

\fmtLemma{kye\-nda}
\fmtCitation{(e)kye\-nda}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{intestine}
\fmtExampleRL{Ebye\-nda bya nte bi\-bba bi\-rai.}
\fmtExampleEN{Intestines of a cow are long.}
\fmtSubentry(b)
\fmtSense{offal, entrails and internal organs of an animal, innards}
\fmtExampleRL{Ata\-ka mu\-no oku\-lya ebye\-nda.}
\fmtExampleEN{He likes eating offal.}

\fmtLemma{kye\-ndai}
\fmtCitation{(e)kye\-ndai}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{ninety}
\fmtExampleRL{Omwa\-na ya\-fu\-ni\-re kye\-ndai ku ki\-ku\-mi mu ku\-ba\-la.}
\fmtExampleEN{The child got ninety out of one hundred in mathematics.}

\fmtLemma{kye\-nde\-bwa}
\fmtCitation{(e)kye\-nde\-bwa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{need, necessity, requirement}
\fmtExampleRL{Eme\-re kye\-nde\-bwa kya bwo\-mi.}
\fmtExampleEN{Food is a necessity in life.}

\fmtLemma{kye\-ndo}
\fmtCitation{(e)kye\-ndo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{gourd (used to carry local brew)}
\fmtExampleRL{Mpee\-rya\-yo ekye\-ndo kya ma\-lwa.}
\fmtExampleEN{Give me a gourd of local brew.}

\fmtLemma{kye\-ndu\-do}
\fmtCitation{(e)kye\-ndu\-do}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wooden hand-washing basin}
\fmtExampleRL{Omu\-ge\-ni mu\-wee\-rye ekye\-ndu\-do anaa\-be omu nga\-lo.}
\fmtExampleEN{Bring a wooden hand-washing basin so that the visitor may wash his hands.}

\fmtLemma{kye\-nduu\-do}
\fmtCitation{(e)kye\-nduu\-do}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{one-litre wooden container used as a unit of measure}
\fmtExampleRL{Nko\-le\-sya kye\-nduu\-do oku\-pi\-ma obwi\-ngi bwa mwe\-nge.}
\fmtExampleEN{I use a one-litre wooden container to determine the amount of beer.}

\fmtLemma{kye\-ngee\-zi}
\fmtCitation{(e)kye\-ngee\-zi}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{dim light}
\fmtExampleRL{Omu nyu\-mba mu\-li\-mu kye\-ngee\-zi.}
\fmtExampleEN{There is dim light in the house.}

\fmtLemma{kye\-nge\-zi}
\fmtCitation{(e)kye\-nge\-zi}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{brightness, illumination}
\fmtExampleRL{Ekye\-nge\-zi kya laa\-pu\-too\-pu ki\-sii\-sa amai\-so.}
\fmtExampleEN{The brightness of the laptop affects the eyes.}

\fmtLemma{kye\-nju}%
\fmtOne{1}
\fmtCrossreference{\textnormal{see also} lyenju}
\fmtCitation{(e)kye\-nju}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{yellow banana}
\fmtExampleRL{Omu Uga\-nda mu\-li\-mu ebi\-ka bya bye\-nju bi\-ngi.}
\fmtExampleEN{There are different varieties of yellow bananas in Uganda.}

\fmtLemma{kye\-nju}%
\fmtOne{2}
\fmtCitation{(e)kye\-nju}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{yellow colour}
\fmtExampleRL{Olu\-go\-ye lwa\-mwe lwa kye\-nju.}
\fmtExampleEN{His clothing is yellow in colour.}

\fmtLemma{kye\-nka\-ni\-nka\-ni}
\fmtPOS{adv}
\fmtSense{equally, evenly}
\fmtExampleRL{Mu\-ga\-ba\-ne empii\-ya ezo kye\-nka\-ni\-nka\-ni.}
\fmtExampleEN{Share that money equally.}

\fmtLemma{kye\-nyi}
\fmtCitation{(e)kye\-nyi}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{forehead}
\fmtExampleRL{Ekye\-nyi kya\-mwe ki\-ga\-zi.}
\fmtExampleEN{His forehead is wide.}

\fmtLemma{kye\-nyu\-ma\-nyu\-ma}
\fmtPOS{adv}
\fmtSense{backwards}
\fmtExampleRL{Lwa\-ki ota\-mbu\-la kye\-nyu\-ma\-nyu\-ma?}
\fmtExampleEN{Why are you walking backwards?}

\fmtLemma{kye\-re\-nga}
\fmtCitation{(e)kye\-re\-nga}
\fmtPOS{n}
\fmtSense{disobedience, insubordination}
\fmtExampleRL{Omwa\-na kye\-re\-nga mu\-zi\-bu oku\-ku\-lya.}
\fmtExampleEN{A disobedient child is difficult to raise.}

\fmtLemma{kye\-re\-za}
\fmtCitation{(e)kye\-re\-za}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{half gourd (especially for drawing water from a pot or boat)}
\fmtExampleRL{Mpee\-re\-za ekye\-re\-za nse\-ne amai\-zi.}
\fmtExampleEN{Pass me the half gourd so that I can draw some water.}

\fmtLemma{kye\-ro}
\fmtCitation{(e)kye\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{piece of clothing, garment}
\fmtExampleRL{Ekye\-ro ekyo ki\-too\-le\-wo awo.}
\fmtExampleEN{Remove that piece of clothing.}

\fmtLemma{kye\-roo\-rwa}
\fmtCitation{(o)kye\-roo\-rwa}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{ill-mannered person, uncivilised person}
\fmtExampleRL{We\-mbe wa\-foo\-ke\-re kye\-roo\-rwa.}
\fmtExampleEN{No wonder you became an ill-mannered person.}

\fmtLemma{Kye\-ru\-pe}
\fmtCitation{(o)Kye\-ru\-pe}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{white person}
\fmtExampleRL{Iswe aba\-ju\-ngu tu\-bee\-ta ba kye\-ru\-pe.}
\fmtExampleEN{We refer to white people as Kyerupe.}

\fmtLemma{kye\-sa\-gu\-ro}
\fmtCitation{(e)kye\-sa\-gu\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{pillow}
\fmtExampleRL{Mpee\-rya ekye\-sa\-gu\-ro kya\-nge mba\-sye.}
\fmtExampleEN{Give me my pillow and I will sleep.}

\fmtLemma{kye\-se\-ro}
\fmtCitation{(e)kye\-se\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{watering trough, drinking trough}
\fmtExampleRL{Ente zi\-kwa\-ba ku kye\-se\-ro ku\-nywa mai\-zi.}
\fmtExampleEN{The cows are going to the watering trough to drink water.}

\fmtLemma{kye\-sii\-mi\-re}
\fmtPOS{adv}
\fmtSense{willingly, voluntarily, readily}
\fmtExampleRL{Omu\-lu\-mu tu\-ku\-gu\-ko\-la kye\-sii\-mi\-re.}
\fmtExampleEN{We are doing the work willingly.}

\fmtLemma{kye\-suu\-bi\-re}
\fmtCitation{(e)kye\-suu\-bi\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{secret box, safe}
\fmtExampleRL{Omu kye\-suu\-bi\-re mu\-bbaa\-mu bi\-ku\-lu.}
\fmtExampleEN{It is only important things that are kept in the secret box.}

\fmtLemma{kye\-suu\-bo}
\fmtCitation{(e)kye\-suu\-bo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{swing (a seat suspended by ropes or chains)}
\fmtExampleRL{Abaa\-na ba\-li oku kye\-suu\-bo ba\-ku\-zee\-nya.}
\fmtExampleEN{The children are playing on the swing.}

\fmtLemma{kye\-swe\-ko}
\fmtCitation{(e)kye\-swe\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{bedding}
\fmtExampleRL{Ebye\-swe\-ko bya\-mwe bi\-kai\-ri\-ki\-re.}
\fmtExampleEN{His bedding has grown very old.}
\fmtSubentry(b)
\fmtSense{supper}
\fmtExampleRL{Tu\-maa\-re oku\-lya ekye\-swe\-ko.}
\fmtExampleEN{We have finished eating supper.}

\fmtLemma{kye\-taa\-go}
\fmtCitation{(e)kye\-taa\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{necessity, need, requirement}
\fmtExampleRL{Ekye\-taa\-go kyai\-swe mai\-zi.}
\fmtExampleEN{Our need is water.}

\fmtLemma{kye\-taa\-wo}%
\fmtOne{1}
\fmtCitation{(e)kye\-taa\-wo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{non-government organisation, NGO}
\fmtExampleRL{Omu Uga\-nda mu\-li\-mu ebye\-taa\-wo bi\-ngi.}
\fmtExampleEN{There are many NGOs in Uganda.}

\fmtLemma{Kye\-taa\-wo}%
\fmtOne{2}
\fmtCitation{(o)Kye\-taa\-wo}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{name given to a child who is conceived unexpectedly}
\fmtExampleRL{OKye\-taa\-wo mwo\-jo mu\-sai ate mu\-go\-ndi.}
\fmtExampleEN{Kyetaawo is a handsome boy and humble at the same time.}

\fmtLemma{kye\-te\-ngo}
\fmtCrossreference{\textnormal{see also} kyetaago}
\fmtCitation{(e)kye\-te\-ngo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from archaic}
\fmtSense{necessity, need, requirement}
\fmtExampleRL{Ekye\-te\-ngo kyai\-swe kya bya\-nye\-ge\-sya ebi\-sai.}
\fmtExampleEN{Our need is a good education system.}

\fmtLemma{kye\-tii\-ki\-re}
\fmtCitation{(e)kye\-tii\-ki\-re}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from flora}
\fmtSense{\textit{Leonotis nepetifolia}}
\fmtExampleRL{Mpee\-rya ekye\-tii\-ki\-re nta\-mbe oka\-ru\-ma.}
\fmtExampleEN{Get me \textit{Leonotis nepetifolia} and I'll treat cramps.}

\fmtLemma{kye\-wa\-ma\-la}
\fmtCitation{(e)kya\-wa\-ma\-la}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{\textit{Te\-tra\-denia riparia} (a species of flowering plant, the essential oil of its leaves have antimalarial contents)}
\fmtExampleRL{Ekya\-wa\-ma\-la ki\-ta\-mba omu\-su\-ja.}
\fmtExampleEN{\textit{Tetradenia riparia} is used to treat malaria.}

\fmtLemma{kye\-wu\-mbye\-wu\-mbye}
\fmtCitation{(e)kye\-wu\-mbye\-wu\-mbye}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{toy made from clay (especially made by children)}
\fmtExampleRL{Abaa\-na ba\-koo\-re ebye\-wu\-mbye\-wu\-mbye bi\-ngi mu\-no.}
\fmtExampleEN{The children have made many toys out of clay.}

\fmtLemma{kye\-wu\-ngo}
\fmtCitation{(e)kye\-wu\-ngo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{type of spirit, charm}
\fmtExampleRL{Omu\-sai\-za ali\-ku ekye\-wu\-ngo.}
\fmtExampleEN{The man is possessed by a spirit.}

\fmtLemma{kye\-wuu\-nyo}
\fmtCitation{(e)kye\-wuu\-nyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wonder, miracle}
\fmtExampleRL{Ki\-ni kye\-wu\-nyo omu\-si\-raa\-mu okwa\-ba omu ke\-re\-zi\-ya.}
\fmtExampleEN{It is a miracle for a Moslem to go to church.}

\fmtLemma{kye\-ye\-nde\-re}
\fmtPOS{adv}
\fmtSense{by choice, voluntary, freely}
\fmtExampleRL{Oku\-lu\-mwa enza\-la ng’e\-me\-re eri\-wo ki\-bba kye\-ye\-nde\-re.}
\fmtExampleEN{One can eat freely when food is available.}

\fmtLemma{kye\-ye\-ra}
\fmtCitation{(e)kye\-ye\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{high broad forehead}
\fmtExampleRL{Omu\-sai\-za ata n’e\-kye\-ye\-ra eki\-ga\-zi.}
\fmtExampleEN{The man has a high broad forehead.}

\fmtLemma{kye\-ye\-te\-bu\-ca}
\fmtCitation{(e)kye\-ye\-te\-bu\-ca}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{echo}
\fmtExampleRL{Mpu\-lii\-re ekye\-ye\-te\-bu\-ca.}
\fmtExampleEN{I heard an echo.}

\fmtLemma{kye\-ye\-ya\-mu\-lyo}
\fmtCitation{(e)kye\-ye\-ya\-mu\-lyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{yawn}
\fmtExampleRL{Aku\-bbi\-re ebye\-ye\-ya\-mu\-lyo bi\-sa\-tu.}
\fmtExampleEN{He yawned three times.}

\fmtLemma{kye\-yo}%
\fmtOne{1}
\fmtCitation{(e)kye\-yo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{broom, besom}
\fmtExampleRL{Ekye\-yo kya\-nge ki\-ka\-yi\-ri\-ki\-re.}
\fmtExampleEN{My broom has worn out.}

\fmtLemma{kye\-yo}%
\fmtOne{2}
\fmtCitation{(e)kye\-yo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wage employment}
\fmtExampleRL{Aya\-bi\-re mu Ame\-ri\-ka ku\-ku\-bba kye\-yo.}
\fmtExampleEN{He went to America for wage employment.}

\fmtLemma{kye\-yo\-bo\-lyo}
\fmtCitation{(e)kye\-yo\-boo\-lyo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{discarded skin of a snake after shedding}
\fmtExampleRL{Ekye\-ye\-yo\-bo\-lyo ki\-ra\-nga nti onjo\-ka ali ampi.}
\fmtExampleEN{The discarded skin of a snake shows that the snake is near.}

\fmtLemma{kyo}
\fmtPOS{pro}
\fmtSense{it (noun class 7 independent pronoun)}

\fmtLemma{kyo\-ba}%
\fmtOne{1}
\fmtCitation{(oku)\-kyo\-ba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kyo\-be\-re}
\fmtSense{to mix}
\fmtExampleRL{Kyo\-ba omu\-wo\-go omu bwi\-ta.}
\fmtExampleEN{Mix the cassava flour with the millet flour.}

\fmtLemma{kyo\-ba}%
\fmtOne{2}
\fmtCitation{(oku)\-kyo\-ba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kyo\-be\-re}
\fmtSense{to reach (one's goal)}
\fmtExampleRL{Ko\-la mu\-no okyo\-be w’o\-ku\-ta\-ka.}
\fmtExampleEN{Work hard and you will reach your goal.}

\fmtLemma{Kyo\-ga}
\fmtCitation{(o)Kyo\-ga}
\fmtPOS{n}
\fmtSense{Lake Kyoga}

\fmtLemma{kyo\-ge\-ro}%
\fmtOne{1}
\fmtCitation{(e)kyo\-ge\-ro}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{traditional bath solution for a baby}
\fmtExampleRL{Abaa\-na aba\-to tu\-ba\-naa\-bya ekyo\-ge\-ro.}
\fmtExampleEN{We bathe children in a traditional bath solution.}
\fmtSubentry(b)
\fmtSense{basin in which a baby is washed (originally made of clay, now motly of metal)}

\fmtLemma{kyo\-go\-yo\-go}
\fmtCitation{(e)kyo\-go\-yo\-go}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wild cassava}
\fmtExampleRL{Ekyo\-go\-yo\-go ki\-kuu\-ra mu ki\-bi\-ra era ti\-ki\-rii\-bwa.}
\fmtExampleEN{Wild cassava grows in the forest and is not edible.}

\fmtLemma{kyo\-ke\-ryo}
\fmtCitation{(e)kyo\-ke\-ryo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{kiln, oven, furnace}
\fmtExampleRL{Ama\-nda aga\-twoi\-re mu kyo\-ke\-ryo.}
\fmtExampleEN{He removed the charcoal from the kiln.}

\fmtLemma{kyo\-ko}
\fmtCitation{(e)kyo\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{pond}
\fmtExampleRL{Abaa\-na ba\-ku\-saa\-nya omu kyo\-ko.}
\fmtExampleEN{The children are swimming in the pond.}

\fmtLemma{kyo\-le}
\fmtCitation{(e)kyo\-le}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{cave}
\fmtExampleRL{Ongu\-li Ga\-da\-fi ya\-yo\-mbo\-ke\-re ekyo\-le omu ma\-ka gaa\-mwe.}
\fmtExampleEN{The late Gadafi dug a cave in his house.}
\fmtSubentry(b)
\fmtSense{burrow, underground hole (of porcupines or moles)}
\fmtExampleRL{Ose\-ke\-su asu\-la mu kyo\-le.}
\fmtExampleEN{The porcupine lives in a hole.}

\fmtLemma{kyo\-lo\-go\-ma}
\fmtCitation{(e)kyo\-lo\-go\-ma}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{pit, large hole in the ground}
\fmtExampleRL{Ombu\-li agwi\-re omu kyo\-lo\-go\-ma.}
\fmtExampleEN{A goat has fallen into a pit.}

\fmtLemma{kyo\-loo\-ni}
\fmtCitation{(e)kyo\-loo\-ni}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from Sw.}
\fmtSense{pit latrine, toilet, restroom, washroom, bathroom}
\fmtExampleRL{Aya\-bi\-re mu kyo\-loo\-ni.}
\fmtExampleEN{He has gone to the toilet.}

\fmtLemma{kyo\-ma}
\fmtCitation{(e)kyo\-ma}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSource{from Sw.}
\fmtSubentry(a)
\fmtSense{iron, metal}
\fmtExampleRL{Ensi\-ka ba\-gi\-ko\-la mu kyo\-ma.}
\fmtExampleEN{A hoe is made of iron.}
\fmtSubentry(b)
\fmtSense{machine, apparatus, appliance, device}
\fmtExampleRL{Ekyo\-ma eki\-sa\-la embaa\-wo kya\-sii\-si\-kai\-re.}
\fmtExampleEN{The machine which cuts timber broke down.}
\fmtSubentry(c)
\fmtSense{milling machine}
\fmtExampleRL{Aya\-bi\-re ku kyo\-ma ku\-sya bwi\-ta.}
\fmtExampleEN{He has gone to the milling machine to grind his millet.}
\fmtSubentry(d)
\fmtFixedexpression{kyo\-ma ka\-li\-ma\-ge\-zi}
\fmtSense{computer}
\fmtExampleRL{Aba\-ntu ba mwi\-jo ogu\-kai\-re ti\-ba\-mai\-te ku\-ko\-le\-sya byo\-ma bi\-ka\-li\-ma\-ge\-zi.}
\fmtExampleEN{People of the older generation do not know how to use a computer properly.}
\fmtSubentry(e)
\fmtFixedexpression{kyo\-ma kyo\-za}
\fmtSense{washing machine}
\fmtExampleRL{Engo\-ye za kaa\-ki zi\-bba zi\-zi\-bu okwo\-za n’o\-bba nga to\-li\-na kyo\-ma kyo\-za.}
\fmtExampleEN{Khaki clothes are difficult to wash if you don't have a washing machine.}

\fmtLemma{kyo\-mbe\-ko}
\fmtCitation{(e)kyo\-mbe\-ko}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{building, construction}
\fmtExampleRL{Ya\-ba oku kyo\-mbe\-ko ki\-di eki\-yaa\-ka onguu\-re olu\-go\-ye olu\-ya.}
\fmtExampleEN{Go to the other new building and buy me a new dress.}

\fmtLemma{kyo\-mbo\-ke}
\fmtCrossreference{\textnormal{see} kyombeko}

\fmtLemma{kyo\-mya}
\fmtCitation{(o)kyo\-mya}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSubentry(a)
\fmtSense{type of spirit among the Banyoro}
\fmtExampleRL{Kyo\-mya abba mba\-ndwa ya Ba\-nyo\-ro.}
\fmtExampleEN{Kyomya is a spirit belonging to the Banyoro.}
\fmtSubentry(b)
\fmtSense{very rich person}
\fmtExampleRL{Omu\-sai\-za kyo\-mya afii\-re.}
\fmtExampleEN{The rich man is dead.}

\fmtLemma{kyo\-na}
\fmtCitation{see (-)ona}

\fmtLemma{kyo\-ndi}
\fmtPOS{interj}
\fmtSubentry(a)
\fmtSense{response of a hunter who did not have a successful hunting}
\fmtExampleRL{Wei\-gu\-le! – Kyo\-ndi.}
\fmtExampleEN{Welcome back! – I was unsuccessful.}
\fmtSubentry(b)
\fmtSense{lack of success, failure, nonsuccess, inefficiency, fruitlessness}
\fmtExampleRL{Olwa\-lee\-ro lu\-bbai\-re lwa kyo\-ndi.}
\fmtExampleEN{Today I was unsuccessful in all my endeavours.}

\fmtLemma{kyo\-ndo}%
\fmtOne{1}
\fmtCitation{(e)kyo\-ndo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{gulf}
\fmtExampleRL{Enya\-nja\-gi\-koo\-to eya Re\-d Sea etaa\-ku ekyo\-ndo eki\-koo\-to.}
\fmtExampleEN{The Red Sea has a big gulf.}

\fmtLemma{kyo\-ndo}%
\fmtOne{2}
\fmtCitation{(e)kyo\-ndo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{\textit{Bryo\-phy\-llum pinnatum} (air plant, cathedral bells, a type of plant used to treat skin infections)}

\fmtLemma{kyo\-ngu}
\fmtCitation{(e)kyo\-ngu}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{pumpkin}
\fmtExampleRL{Ekyo\-ngu ki\-kuu\-re.}
\fmtExampleEN{The pumpkin is ripe.}

\fmtLemma{kyo\-nkai}
\fmtPOS{conj}
\fmtSense{yet, however, although, though, even though}
\fmtExampleRL{Ya\-ya\-bi\-re kyo\-nkai nga mu\-lwai\-re.}
\fmtExampleEN{He left even though he was sick.}

\fmtLemma{kyo\-noo\-no}
\fmtCitation{(e)kyo\-noo\-no}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sin, wrongdoing}
\fmtExampleRL{Ata n’e\-byo\-noo\-no bi\-ngi.}
\fmtExampleEN{He has committed many sins.}

\fmtLemma{kyo\-nzee\-ra}
\fmtCitation{(e)kyo\-nzee\-ra}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{witchcraft, charm, talisman, fetish, juju}
\fmtExampleRL{Ebyo\-nzee\-ra ba\-bi\-na\-ga mu ma\-sa\-nga\-nzi\-ra.}
\fmtExampleEN{They throw fetishes in the road junction.}
\fmtSubentry(b)
\fmtSense{family member responsible for solving the family's predicaments}
\fmtExampleRL{Bu\-li eka ebba n’o\-kyo\-nzee\-ra waa\-bwe.}
\fmtExampleEN{Every family has a family member responsible for solving the family's predicaments.}

\fmtLemma{kyo\-si}%
\fmtOne{1}
\fmtCitation{(e)kyo\-si}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{bad luck, misfortune}
\fmtExampleRL{Ekyo\-si ki\-nde\-me\-se\-rye oku\-ta\-mbu\-la olwa\-lee\-ro.}
\fmtExampleEN{A misfortune prevented me from travelling today.}
\fmtSubentry(b)
\fmtSense{kwashiorkor (a form of malnutrition caused by protein deficiency in the diet), deficiency in a child caused by premature weaning (believed to occur when the mother conceives another child)}
\fmtExampleRL{Omwa\-na wa mu\-ga\-nda wa\-nge ya\-lwai\-re ekyo\-se.}
\fmtExampleEN{My sister's child is experiencing nutritional deficiency due to premature weaning.}

\fmtLemma{kyo\-si}%
\fmtOne{2}
\fmtCitation{(e)kyo\-si}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{off duty, day off, respite}
\fmtExampleRL{Ekyo\-si kya\-nge ki\-bbai\-re kya lee\-ro.}
\fmtExampleEN{Today was my day off.}

\fmtLemma{kyo\-ta}
\fmtCitation{(oku)\-kyo\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kyo\-te\-re}
\fmtSense{to dig mounds (for sweet potato)}
\fmtExampleRL{Aya\-bi\-re ku\-kyo\-ta bi\-yaa\-ta.}
\fmtExampleEN{He has gone to dig sweet potato mounds.}

\fmtLemma{kyo\-te\-ryo}
\fmtCitation{(e)kyo\-te\-ryo}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{sauna}
\fmtExampleRL{Aya\-bi\-re mu kyo\-te\-ryo.}
\fmtExampleEN{He has gone to the sauna.}

\fmtLemma{kyo\-to}%
\fmtOne{1}
\fmtCitation{(e)kyo\-to}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{fireplace}
\fmtExampleRL{Ka\-kaa\-nia omuu\-ro omu kyo\-to.}
\fmtExampleEN{Stir the fire in the fireplace.}

\fmtLemma{kyo\-to}%
\fmtOne{2}
\fmtCitation{(e)kyo\-to}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSubentry(a)
\fmtSense{floating vegetation}
\fmtExampleRL{Emi\-sa\-nga wa\-li\-yo ekyo\-to.}
\fmtExampleEN{There is floating vegetation at Misanga.}
\fmtSubentry(b)
\fmtSense{name of a male born on an island}
\fmtExampleRL{OKyo\-to mu\-bya\-le wa ku iga\-nzo.}
\fmtExampleEN{Kyoto was born on an island.}

\fmtLemma{kyo\-to}%
\fmtOne{3}
\fmtCitation{(e)kyo\-to}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{front derailleur (part of a bicycle)}
\fmtExampleRL{Egaa\-li ya\-nge esii\-si\-kai\-re omu kyo\-to.}
\fmtExampleEN{My bicycle has mechanical problems with the front derailleur.}

\fmtLemma{kyo\-wa}
\fmtCitation{(e)kyo\-wa}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{device used to scoop water from the boat}
\fmtExampleRL{Ko\-le\-sya ekyo\-wa okwo\-wa amai\-zi.}
\fmtExampleEN{Use that scoop to scoop water from the boat.}

\fmtLemma{kyo\-wo}
\fmtCitation{(e)kyo\-wo}
\fmtPOS{n}%
\fmtNC{7}
\fmtSubentry(a)
\fmtSense{mist, fog}
\fmtExampleRL{Ekyo\-wo nka kya\-ka\-li\-rii\-rye olwa\-lee\-ro!}
\fmtExampleEN{There is heavy mist today.}
\fmtSubentry(b)
\fmtSense{humidity}
\fmtExampleRL{Obwi\-re bwa kyo\-wo.}
\fmtExampleEN{The weather is humid.}
\fmtSubentry(c)
\fmtFixedexpression{kwe\-zi ku\-li\-ko (e)kyo\-wo}
\fmtSense{(of the moon) to be pale}
\fmtExampleRL{Okwe\-zi ku\-li\-ku ekyo\-wo.}
\fmtExampleEN{The moon is pale.}
\fmtSubentry(d)
\fmtSense{athlete's foot, tinea pedis (type of fungal infection of the feet)}
\fmtExampleRL{Ekyo\-wo ki\-kya\-li ku\-wo\-na.}
\fmtExampleEN{The tinea pedis is yet to heal.}

\fmtLemma{kyo\-ya}%
\fmtOne{1}
\fmtCitation{(e)kyo\-ya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{wound}
\fmtExampleRL{Ali\-na ekyo\-ya ku ku\-gu\-lu.}
\fmtExampleEN{He has a wound on the leg.}

\fmtLemma{kyo\-ya}%
\fmtOne{2}
\fmtCitation{(e)kyo\-ya}
\fmtPOS{n}%
\fmtNC{7/8}
\fmtSense{feather}
\fmtExampleRL{Onko\-ko azwi\-re\-ku ekyo\-ya.}
\fmtExampleEN{The hen has shed a feather.}

\fmtLemma{kyo\-yo}
\fmtCitation{(e)kyo\-yo}
\fmtPOS{n}%
\fmtNC{7}
\fmtSense{fault, deficiency, imperfection}
\fmtExampleRL{Omu\-lu\-mu gwa\-nge mu\-gu\-koo\-re na kyo\-yo.}
\fmtExampleEN{You did my work shoddily.}

\fmtLemma{kyo\-zi}
\fmtCitation{(o)kyo\-zi}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{black cow}
\fmtExampleRL{Okyo\-zi ala\-yi?}
\fmtExampleEN{Where is the black cow?}

\fmtLemma{kyu\-ka}
\fmtCitation{(oku)\-kyu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kyu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to turn, to turn around}
\fmtExampleRL{Kyu\-ka oi\-re oku gu\-mo\-so.}
\fmtExampleEN{Turn to the left.}
\fmtSubentry(b)
\fmtSense{to change}
\fmtExampleRL{Ebi\-ntu bi\-kyu\-ki\-re mu\-no ebi\-ro bi\-ni.}
\fmtExampleEN{Things have changed a lot lately.}
\fmtSubentry(c)
\fmtSense{to convert, to change one's religious faith}
\fmtExampleRL{Ya\-kyu\-ki\-re n’a\-foo\-ka omu\-lo\-ko\-le.}
\fmtExampleEN{He converted to the born-again faith.}

\fmtLemma{kyu\-ku\-la}
\fmtCitation{(oku)\-kyu\-ku\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kyu\-kwi\-re}
\fmtSense{to wean off}
\fmtExampleRL{Kyu\-ku\-la omwa\-na oyo oku ma\-bee\-re.}
\fmtExampleEN{Wean that child.}

\fmtLemma{kyu\-ngu\-ra}
\fmtCitation{(oku)\-kyu\-ngu\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kyu\-ngwi\-re}
\fmtSense{to redeem, to reclaim, to free}
\fmtExampleRL{Naa\-ni ya\-kyu\-ngu\-ra ei\-ya\-nga lyai\-swe?}
\fmtExampleEN{Who will ever redeem our nation?}

\fmtLemma{kyu\-nya}
\fmtCitation{(oku)\-kyu\-nya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{kyu\-nyi\-re}
\fmtSense{to be delicious, to be tasty}
\fmtExampleRL{Omu\-ku\-bi gu\-ni gu\-kyu\-ni\-re.}
\fmtExampleEN{This sauce is delicious.}

\fmtLemma{kyu\-nyi}
\fmtCitation{\textnormal{(e.g.}\,gu\-kyu\-nyi, mu\-kyu\-nyi\textnormal{)}}
\fmtPOS{adj}
\fmtSense{tasty, delicious, yummy}
\fmtExampleRL{Omu\-ku\-bi gu\-ni gu\-kyu\-nyi.}
\fmtExampleEN{This sauce is delicious.}

\fmtLemma{kyu\-sa}
\fmtCitation{(oku)\-kyu\-sa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{kyu\-si\-re}
\fmtSubentry(a)
\fmtSense{to change, to replace}
\fmtExampleRL{Oba\-na\-saa\-ya\-nsi ati\-bu\-ni ba\-so\-la oku\-kyu\-sa emi\-so\-koo\-lo.}
\fmtExampleEN{Scientists can now transplant bone marrow.}
\fmtSubentry(b)
\fmtSense{to transfer (\eg from one job to another)}
\fmtSubentry(c)
\fmtSense{to translate, to interprete}
\fmtExampleRL{Omu\-kyu\-sa aye\-me\-rei\-re ampi n’o\-mu\-ntu gwa ku\-kyu\-si\-ra.}
\fmtExampleEN{The interpreter is standing near the person he is interpreting for.}

\end{letter}