\begin{letter}{P}

\fmtLemma{paa\-ka}%
\fmtOne{1}
\fmtCitation{(e)paa\-ka}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSubentry(a)
\fmtSense{national park, game reserve}
\fmtExampleRL{E Na\-ki\-to\-ma ali\-yo epaa\-ka ya pyo\-ko.}
\fmtExampleEN{There is a game reserve of rhinos at Nakitoma.}
\fmtSubentry(b)
\fmtSense{taxi park, parking lot, parking area}
\fmtExampleRL{Emo\-to\-ka mu\-zi\-te omu paa\-ka.}
\fmtExampleEN{Put the motorcars in the park.}

\fmtLemma{paa\-ka}%
\fmtOne{2}
\fmtCitation{(oku)\-paa\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{paa\-ki\-re}
\fmtSubentry(a)
\fmtSense{to praise, to express admiration for, to speak highly of}
\fmtExampleRL{Aba\-nya\-la ba\-paa\-ka mu\-no Oje\-ge\-nge waa\-bwe.}
\fmtExampleEN{The Banyala praise their king a lot.}
\fmtSubentry(b)
\fmtSense{to flatter}
\fmtExampleRL{Aba\-ku\-mu\-paa\-ka ni\-bo baa\-mu\-lee\-te\-ra obu\-suu\-me.}
\fmtExampleEN{It is those who are flattering him that have caused him trouble.}

\fmtLemma{paa\-la}
\fmtCitation{(oku)\-paa\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pai\-re}
\fmtSubentry(a)
\fmtSense{to disperse hurriedly, to run in all directions}
\fmtExampleRL{Aba\-ntu baa\-pai\-re ebbo\-mu ni ya\-vu\-gi\-re.}
\fmtExampleEN{People dispersed hurriedly when the bomb exploded.}
\fmtSubentry(b)
\fmtSense{to run away from home}
\fmtExampleRL{Omwa\-la ni ya\-toi\-re enda, ya\-pai\-re n’o\-mwo\-jo waa\-mwe.}
\fmtExampleEN{The girl who got pregnant ran away from home with her boy.}

\fmtLemma{paa\-la\-me\-nti}
\fmtCitation{(e)paa\-la\-me\-nti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{parliament}
\fmtExampleRL{Aya\-kai\-re\-mu oku bu\-kwe\-nda bwa paa\-la\-me\-nti.}
\fmtExampleEN{He has gone through as the member of parliament.}

\fmtLemma{paa\-lu\-nka}
\fmtCitation{(oku)\-paa\-lu\-nka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{paa\-lu\-nki\-re}
\fmtSense{to pop}
\fmtExampleRL{Oso\-ya\-bi\-ni n'a\-ka\-la apaa\-lu\-nka.}
\fmtExampleEN{When soya bean seeds get dry, they pop.}

\fmtLemma{paa\-pa}
\fmtCitation{(o)paa\-pa}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from En.}
\fmtSense{pope}
\fmtExampleRL{Opaa\-pa ni\-ye akuu\-ra Aba\-ka\-tu\-li\-ki ab’Eki\-roo\-ma.}
\fmtExampleEN{The pope is the head of the Roman Catholics.}

\fmtLemma{paa\-paa\-li}
\fmtCrossreference{\textnormal{see} ipaapaali}

\fmtLemma{paa\-ru\-ka}
\fmtCitation{(oku)\-paa\-ru\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{paa\-ru\-ki\-re}
\fmtSubentry(a)
\fmtSense{to be in excess, to increase}
\fmtExampleRL{Emi\-we\-ndo gya ma\-fu\-ta gi\-paa\-ru\-ki\-re.}
\fmtExampleEN{The cost of fuel has excessively increased.}
\fmtSubentry(b)
\fmtSense{(of sun) to shine brightly with a lot of heat}
\fmtExampleRL{Ei\-sa\-na li\-paa\-ru\-ki\-re.}
\fmtExampleEN{It is very hot.}

\fmtLemma{paa\-si}
\fmtCitation{(e)paa\-si}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Sw.}
\fmtSense{iron (a device for smoothing clothes), flat iron}
\fmtExampleRL{Igee\-rya ani epaa\-si ngo\-lo\-le olu\-go\-ye.}
\fmtExampleEN{Bring the flat iron, and I will iron my cloth.}

\fmtLemma{Paa\-si\-ka}
\fmtCitation{(e)paa\-si\-ka}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Sw.}
\fmtSense{Easter Sunday}
\fmtExampleRL{Nai\-za\-yo oku Paa\-si\-ka.}
\fmtExampleEN{I will come there on Easter Sunday.}

\fmtLemma{paa\-si\-ka\-la\-mu}
\fmtCrossreference{\textnormal{see} pasipalamu}

\fmtLemma{paa\-si\-poo\-ta}
\fmtCitation{(e)paa\-si\-poo\-ta}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{passport}
\fmtExampleRL{Aya\-bi\-re ku\-fu\-na paa\-si\-poo\-ta.}
\fmtExampleEN{He has gone to get a passport.}

\fmtLemma{paa\-tii\-ka}
\fmtCitation{(oku)\-paa\-tii\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{paa\-tii\-ki\-re}
\fmtSubentry(a)
\fmtSense{to patch, to put a patch on}
\fmtExampleRL{Eki\-tee\-tee\-yi ya\-ki\-paa\-tii\-ke\-re\-ku eki\-we\-ro.}
\fmtExampleEN{She patched her dress with a piece of cloth.}
\fmtSubentry(b)
\fmtSense{to slander, to denigrate, to bash}
\fmtExampleRL{Ba\-ku\-mu\-paa\-tii\-ka\-ku ebi\-ga\-mbo.}
\fmtExampleEN{They are slandering him.}
\fmtSubentry(c)
\fmtSense{to nickname, to give a nickname to, to call by a nickname}
\fmtExampleRL{Omu\-so\-me\-sa ba\-mu\-paa\-tii\-ki\-re ei\-ba\-ra lya Ka\-do\-mo\-la.}
\fmtExampleEN{The teacher was nicknamed Kadomola.}

\fmtLemma{paa\-tii\-ka\-ku}
\fmtCitation{(oku)\-paa\-tii\-ka\-ku}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{paa\-tii\-ki\-re\-ku}
\fmtSense{to falsely accuse}
\fmtExampleRL{Omu\-su\-mba ba\-mu\-paa\-tii\-ke\-re\-ku emi\-sa\-ngo.}
\fmtExampleEN{The pastor was falsely accused of the offences.}

\fmtLemma{paa\-wu\-nda}
\fmtCitation{(e)paa\-wu\-nda}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{pound, pound sterling}
\fmtExampleRL{Empii\-ya ya\-zi\-mu\-we\-rei\-re mu paa\-wu\-nda.}
\fmtExampleEN{He gave him money in pounds.}

\fmtLemma{paa\-ya}
\fmtCitation{(oku)\-paa\-ya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{paa\-yi\-re}
\fmtSense{to slap, to smack, to strike}
\fmtExampleRL{Paa\-ya omu\-sai\-za oyo olu\-yi.}
\fmtExampleEN{Slap that man.}

\fmtLemma{paa\-za}
\fmtCitation{(oku)\-paa\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{paa\-zi\-re}
\fmtSense{to increase (a price), to hike}
\fmtExampleRL{Aba\-vu\-gi ba ta\-ki\-si ba\-paa\-za ebi\-sa\-le bya nta\-mbu\-la mu bi\-see\-la bya Se\-ku\-ku\-lu.}
\fmtExampleEN{The taxi drivers hike up transport fares during the Christmas season.}

\fmtLemma{pa\-cu\-la}
\fmtCitation{(oku)\-pa\-cu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pa\-cwi\-re}
\fmtSubentry(a)
\fmtSense{to slap hard, to strike, to punch}
\fmtExampleRL{OSe\-ngo\-nzi baa\-mu\-pa\-cwi\-re olu\-wi.}
\fmtExampleEN{Sengonzi was slapped hard.}
\fmtSubentry(b)
\fmtSense{to cut a piece of}
\fmtExampleRL{Mpa\-cuu\-ra\-ku aka\-su\-su ka mu\-saa\-le ogwo.}
\fmtExampleEN{Cut for me a piece of the bark from that tree.}

\fmtLemma{Pa\-dre}
\fmtCitation{(o)Pa\-dre}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from Latin}
\fmtSense{priest (of the Catholic church)}
\fmtExampleRL{Aba Nko\-ko\-nje\-ru ba\-ba\-lee\-tei\-re oPa\-dre.}
\fmtExampleEN{They have brought a Catholic priest for Nkokojeru church.}

\fmtLemma{pai pai pai}
\fmtPOS{ideo}
\fmtSense{used to express intensity}
\fmtExampleRL{Baa\-mu\-suu\-nyi\-re pai pai pai.}
\fmtExampleEN{He was caned thoroughly.}

\fmtLemma{pai\-ku\-la}
\fmtCitation{(oku)\-pai\-ku\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pai\-kwi\-re}
\fmtSense{to beat hard, to strike, to batter}
\fmtExampleRL{Ya\-mu\-pai\-kwi\-re olu\-yi n’a\-zwa\-mu amai\-no.}
\fmtExampleEN{He beat him thoroughly and he lost his teeth.}

\fmtLemma{pai\-pu}
\fmtCitation{(e)pai\-pu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{pipe}
\fmtExampleRL{Epai\-pu ya mai\-zi edo\-mo\-ke\-re.}
\fmtExampleEN{The water pipe got a hole.}

\fmtLemma{pa\-ja\-ma}
\fmtCitation{(e)pa\-ja\-ma}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{underwear, underpants}
\fmtExampleRL{Aba\-sai\-za ba\-zwa\-la epa\-ja\-ma.}
\fmtExampleEN{Men wear underpants.}

\fmtLemma{pa\-ka}
\fmtPOS{prep}
\fmtSense{till, until, up to}
\fmtExampleRL{Ti\-nai\-ze pa\-ka nga oi\-ki\-re.}
\fmtExampleEN{I won’t come until you arrive.}

\fmtLemma{pa\-ka\-la}
\fmtCitation{(oku)\-pa\-ka\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pa\-kai\-re}
\fmtSense{to rattle, to clatter, to clink}
\fmtExampleRL{Eki\-do\-mo\-la kya\-li ki\-ku\-pa\-ka\-la\-mbe oku gaa\-li.}
\fmtExampleEN{The jerry can was rattling against the bicycle.}

\fmtLemma{pa\-ka\-sa}
\fmtCitation{(oku)\-pa\-ka\-sa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pa\-ka\-si\-re}
\fmtSense{to work for wages}
\fmtExampleRL{Ya\-ya\-bi\-re mu Ame\-ri\-ka ku\-pa\-ka\-sa.}
\fmtExampleEN{He went to America to work as a casual labourer.}

\fmtLemma{pa\-ki\-ra}
\fmtCitation{(oku)\-pa\-ki\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pa\-kii\-re}
\fmtSource{from En.}
\fmtSubentry(a)
\fmtSense{to load, to pack, to fill}
\fmtExampleRL{Osu\-kaa\-li mu\-pa\-ki\-re ku loo\-le.}
\fmtExampleEN{You load sugar onto the lorry.}
\fmtSubentry(b)
\fmtSense{to pack, to stow, to put away}
\fmtExampleRL{Ensa\-wo ngi\-pa\-kii\-re\-mu empii\-ya.}
\fmtExampleEN{I packed money into a sack.}
\fmtSubentry(c)
\fmtSense{to usher (passengers in a vehicle)}
\fmtExampleRL{Pa\-ki\-ra aba\-ta\-mbu\-li mu ta\-ki\-si.}
\fmtExampleEN{Usher the passengers into the taxi.}

\fmtLemma{pa\-ki\-ti}
\fmtCitation{(e)pa\-ki\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{packet, pack}
\fmtExampleRL{Nguu\-ra\-yo epa\-ki\-ti ya si\-ga\-la.}
\fmtExampleEN{Buy for me a packet of cigarette.}

\fmtLemma{pa\-ku\-la}
\fmtCitation{(oku)\-pa\-ku\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pa\-kwi\-re}
\fmtSense{to snatch, to take away without permission}
\fmtExampleRL{Ya\-mpa\-kwi\-re\-ku omu\-ye\-mbe gwa\-nge.}
\fmtExampleEN{He snatched my mango from me.}

\fmtLemma{pa\-ku\-li}
\fmtCitation{(e)pa\-ku\-li}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{bowl, dish}
\fmtExampleRL{Tee\-ka eme\-re omu pa\-ku\-li tu\-ga\-bu\-le aba\-ge\-ni.}
\fmtExampleEN{Put the food in the dish and we will serve the visitors.}

\fmtLemma{pa\-ku\-lu\-la}
\fmtCitation{(oku)\-pa\-ku\-lu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pa\-ku\-lwi\-re}
\fmtSense{to offload, to unload, to remove}
\fmtExampleRL{Pa\-ku\-lu\-la ensa\-wo ezo ku mo\-to\-ka.}
\fmtExampleEN{Offload those sacks from the vehicle.}

\fmtLemma{pa\-la}
\fmtCrossreference{\textnormal{see also} mukapa}
\fmtCitation{(e)pa\-la}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{rubber tree (\textit{Hevea brasiliensis})}
\fmtExampleRL{Omu\-saa\-le gwa pa\-la gu\-zwa\-mu obu\-le\-mbo.}
\fmtExampleEN{\textit{Hevea brasiliensis} produces rubber.}

\fmtLemma{pa\-la\-la}
\fmtCitation{(oku)\-pa\-pa\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pa\-pai\-re}
\fmtSubentry(a)
\fmtSense{to stridulate}
\fmtExampleRL{Eki\-nye\-nze ki\-pa\-lai\-re ni kyaa\-ba.}
\fmtExampleEN{A cricket stridulated and flew away.}
\fmtSubentry(b)
\fmtSense{(of a motorcycle) to roar noisily}
\fmtExampleRL{Epi\-ki\-pi\-ki eku\-pa\-la\-la.}
\fmtExampleEN{A motorcycle is roaring noisily.}

\fmtLemma{pa\-la\-mpa}
\fmtCitation{(oku)\-pa\-la\-mpa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pa\-la\-mpi\-re}
\fmtSense{to climb, to go up}
\fmtExampleRL{Omwa\-na aku\-pa\-la\-mpa omu\-saa\-le akwi\-za ku\-gwa.}
\fmtExampleEN{The child who is climbing a tree will fall down.}

\fmtLemma{pa\-lee\-di}
\fmtCitation{(e)pa\-lee\-di}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{parade, procession, march}
\fmtExampleRL{Abaa\-na ba\-li oku pa\-lee\-di.}
\fmtExampleEN{The children are on parade.}

\fmtLemma{pa\-li\-ya\-me\-nti}
\fmtCitation{(e)paa\-li\-ya\-me\-nti}
\fmtPOS{n}%
\fmtNC{9}
\fmtSource{from En.}
\fmtSense{parliament}
\fmtExampleRL{Pa\-li\-ya\-me\-nti ni\-rwo lu\-kii\-ko lwa ba\-kwe\-nda.}
\fmtExampleEN{The parliament is the meeting of the people's representatives.}

\fmtLemma{pa\-mba}
\fmtCitation{(e)pa\-mba}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Persian via Sw.}
\fmtSense{cotton}
\fmtExampleRL{Aba\-ntu ba Na\-ka\-so\-ngo\-la ba\-lu\-ma mu\-no opa\-mba.}
\fmtExampleEN{People from Nakasongola grow a lot of cotton.}

\fmtLemma{pa\-mi\-ti}
\fmtCitation{(e)pa\-mi\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{driving permit, permit, authorisation, license}
\fmtExampleRL{Ka nfu\-ne epa\-mi\-ti eku\-nyi\-ki\-ri\-zi\-sya oku\-ta\-mbu\-lya ente.}
\fmtExampleEN{Let me get a permit that allows me to move with cattle.}

\fmtLemma{pa\-mpa}
\fmtCrossreference{\textnormal{see} pamba}

\fmtExampleEN{That man grew a lot of cotton.}

\fmtLemma{pa\-mpaa\-ta}
\fmtCitation{(oku)\-pa\-mpaa\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pa\-mpaa\-ti\-re}
\fmtSense{to grope, to feel blindly, to fumble around}
\fmtExampleRL{Omu mu\-li\-kii\-rya omu\-ntu apa\-mpaa\-ta\-mbe.}
\fmtExampleEN{People just grope about in darkness.}

\fmtLemma{pa\-mpu}
\fmtCitation{(e)pa\-mpu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{to pump}
\fmtExampleRL{Ngu\-li\-sya epa\-mpu ya mai\-zi.}
\fmtExampleEN{Sell to me a water pump.}

\fmtLemma{pa\-nga}
\fmtCitation{(oku)\-pa\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pa\-ngi\-re}
\fmtSource{from Ar. via Sw.}
\fmtSubentry(a)
\fmtSense{to arrange}
\fmtExampleRL{Pa\-nga emi\-saa\-le egyo.}
\fmtExampleEN{Arrange those trees.}
\fmtSubentry(b)
\fmtSense{to come up with, to think up}
\fmtExampleRL{Mwi\-ra\-nge to\-faa\-yo tu\-kwi\-za ku\-pa\-nga\-pa\-nga.}
\fmtExampleEN{My darling take heart, we will come up with something.}

\fmtLemma{pa\-ngi\-sya}
\fmtCitation{(oku)\-pa\-ngi\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pa\-ngi\-si\-rye}
\fmtSubentry(a)
\fmtSense{to hire, to lease, to eploy, to engage}
\fmtExampleRL{Amaa\-re oku\-pa\-ngi\-sya emo\-to\-ka.}
\fmtExampleEN{She has already hired a car.}
\fmtSubentry(b)
\fmtSense{to rent}
\fmtExampleRL{Ati aku\-pa\-ngi\-sya enyu\-mba omu Ka\-yu\-nga.}
\fmtExampleEN{He is now renting a house in Kayunga.}

\fmtLemma{pa\-pa}
\fmtCitation{(oku)\-pa\-pa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pa\-pi\-re}
\fmtSubentry(a)
\fmtSense{to act hurriedly, to be in a rush}
\fmtExampleRL{Mu\-sai\-za we opa\-pi\-ra mu\-no!}
\fmtExampleEN{You man, you act so hurriedly!}
\fmtSubentry(b)
\fmtSense{to panic, to be alarmed, to lose one's nerve}
\fmtExampleRL{Ni ya\-mboi\-ne n’a\-tii\-na n’a\-ta\-ndi\-ka oku\-pa\-pa.}
\fmtExampleEN{When he saw me, he got scared and started to panic.}

\fmtLemma{pa\-paa\-lu\-ka}
\fmtCrossreference{\textnormal{see} papaaruka}
\fmtDialect{Lunyala}

\fmtLemma{pa\-paa\-ru\-ka}
\fmtDialect{Ruruuli}
\fmtCitation{(oku)\-pa\-paa\-ru\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pa\-paa\-ru\-ki\-re}
\fmtSense{to fly (away) hurriedly, to run away, to run off}
\fmtExampleRL{Onko\-wu apa\-paa\-lu\-ki\-re.}
\fmtExampleEN{A guinea fowl flew away hurriedly.}

\fmtLemma{pa\-pa\-la}
\fmtCitation{(oku)\-pa\-pa\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pa\-pai\-re}
\fmtSense{(of a bird) to flap wings}
\fmtExampleRL{Eki\-nyo\-nyi ki\-ku\-pa\-pa\-la.}
\fmtExampleEN{The bird is flapping its wings.}

\fmtLemma{pa\-pa\-nku\-la}
\fmtCitation{(oku)\-pa\-pa\-nku\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pa\-pa\-nkwi\-re}
\fmtSense{to take without permission, to steal}
\fmtExampleRL{Lwa\-ki opa\-pa\-nkwi\-re egaa\-li ya ba\-ndi?}
\fmtExampleEN{Why did you take somebody’s bicycle without permission?}

\fmtLemma{pa\-ra\-cuu\-ti}
\fmtCitation{(e)pa\-ra\-cuu\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{canoe}
\fmtExampleRL{Epa\-ra\-cuu\-ti ya Ka\-bo\-nge\-ra eku\-too\-nya.}
\fmtExampleEN{Kabongera's canoe leaks.}

\fmtLemma{pa\-ra\-joo\-ki}
\fmtCitation{\textnormal{(e.g.}\,lu\-pa\-ra\-joo\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{one who borrows a lot from others}
\fmtExampleRL{Olu\-nya\-la lu\-li\-mi lu\-pa\-ra\-joo\-ki.}
\fmtExampleEN{Lunyala borrows a lot from other languages.}

\fmtLemma{pa\-ra\-rya}
\fmtCitation{(oku)\-pa\-ra\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pa\-rai\-rye}
\fmtSense{(of termites) to come in small quantity during the harvest season}
\fmtExampleRL{Enswa zaa\-pa\-rai\-rye\-mbe.}
\fmtExampleEN{Flying termites just came in small quantity.}

\fmtLemma{pa\-si\-pa\-si}
\fmtPOS{ideo}
\fmtSense{completely, entirely, totally}
\fmtExampleRL{Ensa\-wo agi\-pa\-kii\-re pa\-si\-pa\-si.}
\fmtExampleEN{He has filled the sack completely.}

\fmtLemma{pa\-si\-pa\-la\-mu}
\fmtCitation{(e)pa\-si\-pa\-la\-mu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{paspalum grass, crowngrass}
\fmtExampleRL{Tu\-ku\-si\-mba pa\-si\-pa\-la\-mu mu ki\-saa\-we.}
\fmtExampleEN{We are planting paspalum on the field.}

\fmtLemma{pa\-su\-ka}
\fmtCitation{(oku)\-pa\-su\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pa\-su\-ki\-re}
\fmtSense{(of a sesame) to open, to split}
\fmtExampleRL{Ama\-ca\-nde ga\-pa\-su\-ki\-re.}
\fmtExampleEN{The sesame has opened.}

\fmtLemma{pa\-ta}
\fmtCitation{(e)pa\-ta}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Sw.}
\fmtSense{door hinge}
\fmtExampleRL{Olwi\-ge lu\-tee\-kwa oku\-bbaa\-ku epa\-ta.}
\fmtExampleEN{A door must have a door hinge.}

\fmtLemma{pa\-ta\-na}
\fmtCitation{(oku)\-pa\-ta\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pa\-tai\-ne}
\fmtSource{from Sw.}
\fmtSense{to come to terms, to come to an agreement, to reach an understanding (esp. with regard to work)}
\fmtExampleRL{Nje mmaa\-re oku\-pa\-ta\-na olu\-bbi\-mbi.}
\fmtExampleEN{As for me, I already reached an agreement for work.}

\fmtLemma{pa\-tu\-ro}
\fmtCitation{(e)pa\-tu\-ro}
\fmtPOS{n}%
\fmtNC{9}
\fmtSource{from En.}
\fmtSense{patrol, guard, watch, monitoring}
\fmtExampleRL{Epi\-ki\-pi\-ki ya pa\-tu\-ro eya\-kai\-re\-wo ani.}
\fmtExampleEN{The patrol motorcycle has passed here.}

\fmtLemma{pa\-wu\-da}
\fmtCitation{(o)pa\-wu\-da}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{powder (a cosmetic), face powder}
\fmtExampleRL{Pa\-wu\-da ya\-du\-bi\-re.}
\fmtExampleEN{Face powder is out of fashioned.}

\fmtLemma{pa\-wu\-nda}
\fmtCitation{(e)pa\-wu\-nda}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{pound sterling}
\fmtExampleRL{Aba\-ju\-ngu ba\-ko\-le\-sya pa\-wu\-nda.}
\fmtExampleEN{The British use pound sterling.}

\fmtLemma{pa\-yi\-ni}
\fmtCitation{(e)pa\-yi\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{pine tree}
\fmtExampleRL{Pa\-yi\-ni mu\-saa\-le gu\-sai.}
\fmtExampleEN{The pine is a good tree.}

\fmtLemma{pee\-ka}
\fmtCitation{(oku)\-pee\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pee\-ke\-re}
\fmtSense{to pester, to demand incessantly}
\fmtExampleRL{Omu\-ka\-li aku\-mpee\-ka tu\-kyle ewaa\-bwe.}
\fmtExampleEN{The woman is pestering me to visit their home.}

\fmtLemma{pee\-ni}
\fmtCitation{(e)pee\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{pen}
\fmtExampleRL{Abeegesya ba\-ko\-le\-sya pee\-ni egi\-yoo\-ki oku\-go\-lo\-la ebi\-ta\-bo by’a\-bee\-gi.}
\fmtExampleEN{Teachers always use red pen to correct the pupils' work.}

\fmtLemma{pee\-ruu\-ka}
\fmtCitation{(oku)\-pee\-ruu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pee\-ruu\-ki\-re}
\fmtSense{to become pale}
\fmtExampleRL{Olu\-su\-su lwa\-mwe lu\-pee\-ruu\-ki\-re.}
\fmtExampleEN{Her skin became pale.}

\fmtLemma{pee\-sa}
\fmtCitation{(e)pee\-sa}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Hindi via Sw.}
\fmtSense{button}
\fmtExampleRL{Ntee\-ra epee\-sa oku bbu\-la\-wu\-zi.}
\fmtExampleEN{Fix for me the button on my blouse?}

\fmtLemma{pee\-vu\-me\-nti}
\fmtCitation{(e)pee\-vu\-me\-nti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{pavement, sidewalk}
\fmtExampleRL{Ya\-mu\-bba\-ndii\-re ku pee\-vu\-me\-nti.}
\fmtExampleEN{He hit him on the pavement.}

\fmtLemma{pe\-ka}
\fmtCitation{(oku)\-pe\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pe\-ke\-re}
\fmtSubentry(a)
\fmtSense{to chase, to pursue, to run after}
\fmtExampleRL{Pe\-ka onko\-ko oyo okwi\-kya ni wa\-mu\-kwa\-ta.}
\fmtExampleEN{Chase that hen until you catch it.}
\fmtSubentry(b)
\fmtSense{to send away}
\fmtExampleRL{Omwo\-jo oyo oi\-te\-waa\-mwe ya\-mu\-pe\-ke\-re eka.}
\fmtExampleEN{That boy was sent away from home by his father.}

\fmtLemma{pe\-ku\-ka}
\fmtCitation{(oku)\-pe\-ku\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pe\-ku\-ki\-re}
\fmtSense{to talk nonsense, to say irrelevant things}
\fmtExampleRL{Aku\-pe\-ku\-ka na bi\-ga\-mbo.}
\fmtExampleEN{She is talking nonsense.}
\fmtSubentry(a)
\fmtSense{to hurry, to rush}
\fmtExampleRL{Aku\-pe\-ku\-ka ku\-sa\-nga mo\-to\-ka.}
\fmtExampleEN{He is hurrying to catch a vehicle.}
\fmtSubentry(b)
\fmtSense{(of a heart) to beat fast}
\fmtExampleRL{Omwo\-yo gu\-ku\-mpe\-ku\-ka.}
\fmtExampleEN{My heart is beating fast.}

\fmtLemma{pe\-na\-ti}
\fmtCitation{(e)pe\-na\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{penalty kick}
\fmtExampleRL{Ba\-ku\-si\-mu\-la\-nga\-na pe\-na\-ti.}
\fmtExampleEN{They are shooting a penalty kick.}

\fmtLemma{Pe\-ne\-te\-nsi\-ya}
\fmtCitation{(e)Pe\-ne\-te\-nsi\-ya}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{Sacrament of Penance and Reconciliation (one of the seven sacraments of the Catholic Church)}
\fmtExampleRL{Akwe\-ga sa\-ka\-la\-me\-nto ya Pe\-ne\-te\-nsi\-ya.}
\fmtExampleEN{He is studying the Sacrament of Penance and Reconciliation.}

\fmtLemma{pe\-ne\-ti}
\fmtCrossreference{\textnormal{see} penati}

\fmtLemma{pe\-nso\-ni}
\fmtCitation{(e)pe\-nso\-ni}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{pension, retirement (benefits)}
\fmtExampleRL{Akya\-li ku\-tu\-nga pe\-nso\-ni yaa\-mwe.}
\fmtExampleEN{He has not yet got his pension.}

\fmtLemma{pe\-nte\-koo\-ti}
\fmtCitation{(e)pe\-nte\-koo\-ti}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{Pentecost}
\fmtExampleRL{Omwo\-yo akwe\-ra ya\-nii\-nu\-ki\-re ku lu\-na\-ku lwa pe\-nta\-koo\-ti.}
\fmtExampleEN{The holy spirit descended on the Pentecost day.}

\fmtLemma{pe\-re\-mu\-lo}
\fmtDialect{Ruruuli}
\fmtCitation{(o)pe\-re\-mu\-lo}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{evil spirit}
\fmtExampleRL{Epe\-re\-mu\-lo zii\-ta aba\-ntu.}
\fmtExampleEN{Evil spirits kill people.}

\fmtLemma{pe\-te\-ro}
\fmtCrossreference{\textnormal{see also}  lucwamba}
\fmtCitation{(o)pe\-te\-ro}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{\textit{Cynodon dactylon} (Bermuda grass, couch grass)}

\fmtLemma{pe\-tu\-la}
\fmtCitation{(oku)\-pe\-tu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pe\-twi\-re}
\fmtSense{to separate stones from the grains}
\fmtExampleRL{Aba\-ka\-li ba\-ya\-bi\-re ku\-pe\-tu\-la bi\-yi\-mba.}
\fmtExampleEN{The women have gone to separate stones from the beans.}

\fmtLemma{pe\-tu\-roo\-li}
\fmtCitation{(e)pe\-tu\-roo\-li}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{petrol, gasoline, fuel, gas}
\fmtExampleRL{Emo\-to\-ka ezi\-mwei zaa\-ba\-mu pe\-tu\-roo\-li.}
\fmtExampleEN{Some vehicles use petrol.}

\fmtLemma{pii\-ka}
\fmtCitation{(oku)\-pii\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pii\-ki\-re}
\fmtSubentry(a)
\fmtSense{to put into, to pour into}
\fmtExampleRL{Pii\-ka amai\-zi mu cu\-pa.}
\fmtExampleEN{Pour the water into the bottle.}
\fmtSubentry(b)
\fmtSense{to give an excessive amount (\eg food)}
\fmtExampleRL{Omwa\-na owa ka\-do\-ko baa\-mu\-pii\-ka eme\-re.}
\fmtExampleEN{A greedy child is given an excessive amount of food.}

\fmtLemma{pii\-si\-pii\-si}
\fmtCitation{(e)pii\-si\-pii\-si}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{screwdriver}
\fmtExampleRL{Epii\-si\-pii\-si ya\-nge ekai\-ri\-ki\-re.}
\fmtExampleEN{My screwdriver is worn out.}

\fmtLemma{pii\-tii\-ka}
\fmtCitation{(oku)\-pii\-tii\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pii\-tii\-ki\-re}
\fmtSubentry(a)
\fmtSense{to stuff, to cram in carelessly}
\fmtExampleRL{Lwa\-ki oku\-pii\-tii\-ka ebi\-ntu ebi\-ngi omu ka\-sii\-ka aka\-doo\-li?}
\fmtExampleEN{Why do you cram many things in a small room?}
\fmtSubentry(b)
\fmtSense{to write without spacing letters}
\fmtExampleRL{Le\-ka ku\-pii\-tii\-ka bi\-ga\-mbo.}
\fmtExampleEN{Stop writing words without spacing them.}

\fmtLemma{pi\-ka}
\fmtCitation{(oku)\-pi\-ka}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{pi\-ki\-re}
\fmtSubentry(a)
\fmtSense{to reach, to arrive at}
\fmtExampleRL{Mpi\-ki\-re omu lwe\-ru.}
\fmtExampleEN{I have reached clear waters.}
\fmtSubentry(b)
\fmtSense{to pump up, to add air pressure}
\fmtExampleRL{Pi\-ka omwi\-ka omu gaa\-li.}
\fmtExampleEN{Pump up the bicycle tyre. (lit.\,Add air pressure to the bicycle tyre.)}

\fmtLemma{pi\-kii\-rya}
\fmtCitation{(oku)\-pi\-kii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pi\-ki\-rii\-rye}
\fmtSubentry(a)
\fmtSense{to pester, to annoy with persistent requests}
\fmtExampleRL{Mu\-mu\-pi\-kii\-rye aya\-be.}
\fmtExampleEN{You pester him to go away.}
\fmtSubentry(b)
\fmtSense{to incite, to encourage, to stimulate, to instigate, to provoke}
\fmtExampleRL{Lwa\-ki oku\-pi\-kii\-rya abaa\-na ba\-lwa\-ne?}
\fmtExampleEN{Why are you inciting the children into fighting?}

\fmtLemma{pi\-ki\-pi\-ki}
\fmtCitation{(e)pi\-ki\-pi\-ki}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Sw.}
\fmtSense{motorcycle}
\fmtExampleRL{Omu\-ka\-li aku\-vu\-ga epi\-ki\-pi\-ki.}
\fmtExampleEN{A woman is riding a motorcycle.}

\fmtLemma{pi\-ma}
\fmtCitation{(oku)\-pi\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pi\-mi\-re}
\fmtSubentry(a)
\fmtSense{to measure}
\fmtExampleRL{Pi\-ma enyu\-mba yaa\-mu.}
\fmtExampleEN{Measure the size of your house.}
\fmtSubentry(b)
\fmtSense{to survey (land)}
\fmtExampleRL{Aba\-sa\-ve\-ya ba\-ku\-pi\-ma ei\-ta\-ka\-li lya\-nge.}
\fmtExampleEN{The surveyors are surveying my land.}

\fmtLemma{pi\-ni}
\fmtCitation{(e)pi\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{pin (earpin, nose pin)}
\fmtExampleRL{Atai\-re aka\-pi\-ni oku ku\-twi.}
\fmtExampleEN{He has put an ear pin in the ear.}

\fmtLemma{pi\-nka}
\fmtCitation{(o)pi\-nka}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{pink colour}
\fmtExampleRL{Esa\-pa\-tu za\-nge za pi\-nka.}
\fmtExampleEN{My sandals are pink.}

\fmtLemma{pi\-pa}
\fmtCitation{(e)pi\-pa}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Port.\,via Sw.}
\fmtSense{drum, canister, barrel, tank}
\fmtExampleRL{Amai\-zi ga\-su\-ke omu pi\-pa.}
\fmtExampleEN{Pour the water into the drum.}

\fmtLemma{pi\-pi\-ra}
\fmtCitation{(oku)\-pi\-pi\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pi\-pii\-re}
\fmtSubentry(a)
\fmtSense{to hobble}
\fmtSubentry(b)
\fmtSense{(of a faulty engine) to sputter, to pop}
\fmtExampleRL{Epi\-ki\-pi\-ki eku\-pi\-pi\-ra.}
\fmtExampleEN{The motorcycle engine is sputtering.}

\fmtLemma{pi\-pi\-rya}
\fmtCitation{(oku)\-pi\-pi\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pi\-pi\-rii\-rye}
\fmtSense{to escape, to break out, to break free, to free oneself}
\fmtExampleRL{Omu\-sai\-za baa\-mu\-bbwe\-re bai\-tu ya\-pi\-pi\-rya\-nga\-mbe.}
\fmtExampleEN{A man was tied with a rope, but he freed himself.}

\fmtLemma{pi\-ri\-ki\-ta}
\fmtCitation{(oku)\-pi\-ri\-ki\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pi\-ri\-ki\-ti\-re}
\fmtSense{to glitter, to shine, to sparkle, to twinkle, to glimmer}
\fmtExampleRL{Emo\-to\-ka egi\-yaa\-ka epi\-ri\-ki\-ta.}
\fmtExampleEN{A new car shines.}

\fmtLemma{pi\-ri\-mpa}
\fmtCitation{(oku)\-pi\-ri\-mpa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pi\-ri\-mpi\-re}
\fmtSense{to run very fast, to bolt, to sprint}
\fmtExampleRL{Oso\-bo\-la oku\-pi\-ri\-mpa eki\-ku\-mi?}
\fmtExampleEN{Can you run a hundred meters?}

\fmtLemma{pi\-ri\-mpya}%
\fmtOne{1}
\fmtCitation{(oku)\-pi\-ri\-mpya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pi\-ri\-mpi\-rye}
\fmtSense{to chase, to pursue, to run after, to follow}
\fmtExampleRL{Ya\-pi\-ri\-mpi\-rye omwi\-bbi okwi\-ca ni ya\-mu\-kwai\-te.}
\fmtExampleEN{He chased the thief swiftly till he caught him.}

\fmtLemma{pi\-ri\-mpya}%
\fmtOne{2}
\fmtCitation{(oku)\-pi\-ri\-mpya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pi\-ri\-mpi\-rye}
\fmtSense{to run at a terrific speed, to bolt, to sprint}
\fmtExampleRL{Omwa\-na waa\-mu amai\-te oku\-pi\-ri\-mpya eki\-ku\-mi.}
\fmtExampleEN{Your child knows how to sprint one hundred meters.}

\fmtLemma{pi\-ri\-pii\-ya}
\fmtCitation{(oku)\-pi\-ri\-pii\-ya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pi\-ri\-pii\-yi\-re}
\fmtSubentry(a)
\fmtSense{to drill, to bore, to bore a hole in}
\fmtExampleRL{Pi\-ri\-pii\-ya olu\-baa\-wo olu\-tee\-mu eki\-wu\-lu.}
\fmtExampleEN{Drill the timber.}
\fmtSubentry(b)
\fmtSense{to rub sticks together to make fi\-re}
\fmtExampleRL{Aku\-pi\-ri\-pii\-ya le\-ke aya\-kye omuu\-ro.}
\fmtExampleEN{He is rubbing the sticks together in order to make fire.}
\fmtSubentry(c)
\fmtSense{to administer an enema, to give an enema}
\fmtExampleRL{Omwa\-na ba\-mu\-pi\-ri\-pii\-yi\-re n’a\-ta\-ndi\-ka oku\-wu\-lu\-ka.}
\fmtExampleEN{They have administered an enema to the child, and he started passing stool.}

\fmtLemma{pi\-ri\-ri\-ya}
\fmtCitation{(o)pi\-ri\-ri\-ya}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from archaic}
\fmtSense{gin}

\fmtLemma{pi\-rya\-pi\-rya}
\fmtCitation{(oku)\-pi\-rya\-pi\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pii\-rye\-pii\-rye}
\fmtSense{to glitter, to shine, to sparkle, to twinkle, to glint, to gleam, to shimmer, to glimmer}
\fmtExampleRL{Bo\-na ei\-sa\-na li\-ku\-pi\-rya\-pi\-rya.}
\fmtExampleEN{Look at the sun, it is shining.}

\fmtLemma{pi\-si\-to}
\fmtCitation{(e)pi\-si\-to}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{pistol}
\fmtExampleRL{Epi\-si\-to mbu\-ndu gi\-doo\-li.}
\fmtExampleEN{A pistol is a small gun.}

\fmtLemma{pi\-ti\-rya}
\fmtCitation{(oku)\-pi\-ti\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pi\-tii\-rye}
\fmtSense{to walk in an exhausted way}
\fmtExampleRL{Mwi\-ra\-nge lwa\-ki oku\-pi\-ti\-rya?}
\fmtExampleEN{My dear, why are you walking as if you were exhausted?}

\fmtLemma{pi\-ya\-no}
\fmtCitation{(e)pi\-ya\-no}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{piano}
\fmtExampleRL{Naa\-ni amai\-te oku\-zee\-nya epi\-ya\-no.}
\fmtExampleEN{Who knows how to play the piano?}

\fmtLemma{po\-ka}
\fmtCitation{(oku)\-po\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{po\-ke\-re}
\fmtSubentry(a)
\fmtSense{to pump}
\fmtExampleRL{Mwa\-be mu\-po\-ke amai\-zi gai\-ze.}
\fmtExampleEN{Go and pump the water.}
\fmtSubentry(b)
\fmtSense{to fill (with air using a pump), to pump (with air)}
\fmtExampleRL{Aku\-po\-ka omwi\-ka omu gaa\-li.}
\fmtExampleEN{He is pumping air into the bicycle tire.}
\fmtSubentry(c)
\fmtSense{to pack, to put things in, to package}
\fmtExampleRL{Mu\-po\-ke engo\-ye za\-nge omu ku\-ti\-ya.}
\fmtExampleEN{Pack my clothes in the bag.}

\fmtLemma{po\-lii\-si}
\fmtCitation{(e)po\-lii\-si}
\fmtPOS{n}%
\fmtNC{9}
\fmtSource{from En.}
\fmtSubentry(a)
\fmtSense{police force}
\fmtExampleRL{Epo\-lii\-si ya Uga\-nda eko\-la emi\-ru\-mu gi\-ngi.}
\fmtExampleEN{The Ugandan police does many things.}
\fmtSubentry(b)
\fmtSense{police station}
\fmtExampleRL{Eki\-te\-be eki\-ku\-lu kya po\-lii\-si ya Uga\-nda ki\-ri Na\-gu\-lu.}
\fmtExampleEN{The main police station in Uganda is in Nagulu.}

\fmtLemma{poo\-ca}
\fmtCitation{(oku)\-poo\-ca}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{poo\-ce\-rye}
\fmtSense{to suffer, to be in pain}
\fmtExampleRL{Baa\-gwi\-re oku bu\-ta\-ndu era ati\-bu\-ni ba\-li mu irwa\-ro ba\-ku\-poo\-ca.}
\fmtExampleEN{They had an accident and now they are in the hospital suffering.}

\fmtLemma{poo\-li\-yo}
\fmtCitation{(o)poo\-li\-yo}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{poliomyelitis, polio}
\fmtExampleRL{Oyo mu\-lwai\-re wa poo\-li\-yo.}
\fmtExampleEN{That one is a poliomyelitis patient.}

\fmtLemma{ppu}
\fmtPOS{ideo}
\fmtSubentry(a)
\fmtSense{sound of a falling object, sound of a thud}
\fmtExampleRL{Omu\-saa\-le gwa\-gwi\-re ppuu!}
\fmtExampleEN{The tree fell heavily.}
\fmtSubentry(b)
\fmtFixedexpression{ppu na ppu}
\fmtSense{from time immemorial}
\fmtExampleRL{Ppu na ppu aba\-ka\-li ti\-baa\-lya\-nga nko\-ko.}
\fmtExampleEN{From time immemorial women have not been eating chicken.}

\fmtLemma{pu\-laa\-ni}
\fmtCitation{(e)pu\-laa\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{plan, blueprint, diagram}
\fmtExampleRL{Eni ni\-gyo pu\-laa\-ni ya nyu\-mba ya\-nge.}
\fmtExampleEN{This is the plan of my house.}

\fmtLemma{pu\-la\-si\-ta}
\fmtCitation{(e)pu\-la\-si\-ta}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{plaster (for cuts and wounds)}
\fmtExampleRL{Ta epu\-la\-si\-ta oku kyo\-ya.}
\fmtExampleEN{Dress your wound with a plaster.}

\fmtLemma{pu\-le\-sa}
\fmtCitation{(e)pu\-le\-sa}
\fmtPOS{n}%
\fmtNC{9}
\fmtSource{from En.}
\fmtSense{high blood pressure}
\fmtExampleRL{Ekii\-ca ki\-ta\-mba epu\-le\-sa.}
\fmtExampleEN{The fruit of the sausage tree cures high blood pressure.}

\fmtLemma{pu\-li\-gaa\-too\-li}
\fmtCitation{(e)pu\-li\-gaa\-too\-li}
\fmtPOS{n}%
\fmtNC{9}
\fmtSource{from En.}
\fmtSense{(Catholic) purgatory}
\fmtExampleRL{Omwo\-yo gwa ru\-fa gu\-kya\-li mu pu\-li\-gaa\-too\-li.}
\fmtExampleEN{The spirit of the late is still in the purgatory.}

\fmtLemma{pu\-lii\-da}
\fmtCitation{(o)pu\-lii\-da}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from En.}
\fmtSense{pleader, defending lawyer}
\fmtExampleRL{Yasabi\-re omu\-la\-mu\-li amwi\-ki\-ri\-ze ye\-kee\-re afu\-ne opu\-lii\-da.}
\fmtExampleEN{He requested the judge to first allow him to get a pleader.}

\fmtLemma{pu\-lo\-fee\-sa}
\fmtCitation{(o)pu\-lo\-fee\-sa}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from En.}
\fmtSense{professor}
\fmtExampleRL{Omu\-ko\-li wa pu\-lo\-fee\-sa mwa\-saa\-mii\-ri.}
\fmtExampleEN{The professor's worker is irresponsible.}

\fmtLemma{pu\-lo\-gu\-ra\-mu}
\fmtCitation{(e)pu\-lo\-gu\-ra\-mu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{program}
\fmtExampleRL{Ni\-nje mu\-ju\-gi\-ri\-za wa pu\-lo\-gu\-ra\-mu eni.}
\fmtExampleEN{I am the coordinator of this program.}

\fmtLemma{pu\-lo\-ti}
\fmtCitation{(e)pu\-lo\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{plot, piece of ground, allotment}
\fmtExampleRL{Epu\-lo\-ti ya\-nge gi\-ga\-zi.}
\fmtExampleEN{My plot is big.}

\fmtLemma{pu\-lu\-la}
\fmtCitation{(oku)\-pu\-lu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pu\-lwi\-re}
\fmtSense{to have severe diarrhoea}
\fmtExampleRL{Omwa\-na mu\-lwai\-re, aku\-pu\-lu\-la.}
\fmtExampleEN{The child is sick, he is having severe diarrhoea.}

\fmtLemma{pu\-pu\-la}
\fmtCitation{(oku)\-pu\-pu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pu\-pwi\-re}
\fmtSense{(of a stomach) to grumble, to growl, to rumble}
\fmtExampleRL{Enda ya mwa\-na eku\-pu\-pu\-la oku\-zwa ma\-mbya.}
\fmtExampleEN{The child's stomach has been grumbling since morning.}

\fmtLemma{pu\-pu\-lya}
\fmtCitation{(oku)\-pu\-pu\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pu\-pwi\-rye}
\fmtSense{to bubble}
\fmtExampleRL{Amai\-zi ga\-ta\-ndi\-ki\-re oku\-pu\-pu\-lya.}
\fmtExampleEN{The water has started bubbling.}

\fmtLemma{pu\-re\-zi\-de\-nti}
\fmtCitation{(o)pu\-re\-zi\-de\-nti}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from En.}
\fmtSense{president}
\fmtExampleRL{Uga\-nda ebbai\-re na ba\-pu\-re\-zi\-de\-nti bai\-ngi.}
\fmtExampleEN{Uganda has had many presidents.}

\fmtLemma{pu\-ru\-ka}
\fmtCitation{(oku)\-pu\-ru\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pu\-ru\-ki\-re}
\fmtSubentry(a)
\fmtSense{(of a bird) to escape, to flee}
\fmtExampleRL{Aka\-nyo\-nyi ka\-pu\-ru\-ki\-re.}
\fmtExampleEN{The bird has escaped.}
\fmtSubentry(b)
\fmtSense{to have diarrhoea}
\fmtExampleRL{Omwa\-na aku\-pu\-ru\-ka abbai\-re ki?}
\fmtExampleEN{What has happened to a child who is having diarrhoea?}

\fmtLemma{pu\-ru\-nju\-ka}
\fmtCitation{(oku)\-pu\-ru\-nju\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pu\-ru\-nju\-ki\-re}
\fmtSense{(of a bird) to evade capture}
\fmtExampleRL{Aka\-nyo\-nyi ka\-pu\-ru\-nju\-ki\-re ni kaa\-ba.}
\fmtExampleEN{A bird evaded capture and flew away.}

\fmtLemma{pu\-si}
\fmtCitation{(o)pu\-si}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSource{from En.}
\fmtSense{cat}

\fmtLemma{puu\-cu}
\fmtCitation{(o)puu\-cu}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{pangolin}
\fmtExampleRL{Opuu\-cu abba wa nso\-ni mu\-no.}
\fmtExampleEN{The pangolin is a shy animal.}

\fmtLemma{puu\-la}
\fmtCitation{(oku)\-puu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{pui\-re}
\fmtSense{to throb with pain, to ache}
\fmtExampleRL{Erii\-no li\-ku\-mpuu\-la mu\-no.}
\fmtExampleEN{The tooth is aching a lot.}

\fmtLemma{puu\-li\-nka}
\fmtCitation{(oku)\-puu\-li\-nka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{puu\-li\-nki\-re}
\fmtSubentry(a)
\fmtSense{to burst, to split open, to explode}
\fmtExampleRL{Omu\-pii\-ra gu\-puu\-li\-nki\-re.}
\fmtExampleEN{The ball has burst.}
\fmtSubentry(b)
\fmtSense{to explode, to burst}
\fmtExampleRL{Ebbo\-mu epuu\-li\-nki\-re.}
\fmtExampleEN{The bomb has exploded.}
\fmtSubentry(c)
\fmtSense{to escape, to run away, to flee, to break away}
\fmtExampleRL{Omu\-bbwe apuu\-li\-nki\-re.}
\fmtExampleEN{The prisoner has escaped.}
\fmtSubentry(d)
\fmtSense{to get out hurriedly}
\fmtExampleRL{Lwa\-ki omu\-sai\-za oyo aku\-puu\-li\-nka oku\-zwa mu nyu\-mba?}
\fmtExampleEN{Why is that man hurriedly getting out of the house?}

\fmtLemma{puu\-luu\-la}
\fmtCitation{(oku)\-puu\-luu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{puu\-lwi\-re}
\fmtSubentry(a)
\fmtSense{to thrash, to hit grains or seeds out of husks}
\fmtExampleRL{Omaa\-ma aku\-puu\-luu\-la obwi\-ta omu lu\-bu\-ga.}
\fmtExampleEN{My mother is threshing corn in the courtyard.}
\fmtSubentry(b)
\fmtSense{to beat severely, to thrash, to batter}
\fmtExampleRL{Ba\-mu\-puu\-lwi\-re emii\-go.}
\fmtExampleEN{They severely beat him.}
\fmtSubentry(c)
\fmtSense{to beat the dust out of}
\fmtExampleRL{Aku\-puu\-luu\-la ebbu\-la\-ngi\-ti.}
\fmtExampleEN{He is beating the dust out of the blanket.}

\fmtLemma{puu\-mu\-ka}
\fmtCitation{(oku)\-puu\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{puu\-mu\-ki\-re}
\fmtSense{release, to discharge}
\fmtExampleRL{Emo\-to\-ka ezi\-kai\-re zi\-puu\-mu\-ka omwi\-ka ogu\-sii\-sa obu\-to\-nde.}
\fmtExampleEN{Old cars release thick fumes, which destroy the environment.}

\fmtLemma{puu\-muu\-la}
\fmtCitation{(oku)\-puu\-muu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{puu\-mwi\-re}
\fmtSense{to strike heavily, to hit, to beat}
\fmtExampleRL{Ekii\-ti\-mba mu\-ki\-puu\-muu\-le emii\-go.}
\fmtExampleEN{Strike the python heavily.}

\fmtLemma{puu\-pa}
\fmtCitation{(oku)\-puu\-pa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{puu\-pi\-re}
\fmtSubentry(a)
\fmtSense{(of fish) to go bad}
\fmtExampleRL{Encwi zi\-puu\-pa nga zi\-rwi\-re\-wo.}
\fmtExampleEN{The fish go bad when they are not used for too long.}
\fmtSubentry(b)
\fmtSense{fish that has gone bad}
\fmtExampleRL{Omu\-lo\-ndo ba\-mu\-kwai\-te nga aku\-tu\-nda puu\-pa.}
\fmtExampleEN{The fisherman was caught selling fish which has gone bad.}

\fmtLemma{puu\-ya}
\fmtCitation{(oku)\-puu\-ya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{puu\-yi\-re}
\fmtSense{to smear (with cow dung)}
\fmtExampleRL{Ema\-mbya nkwa\-ba ku\-puu\-ya nyu\-mba ya\-nge.}
\fmtExampleEN{Tomorrow I am going to smear my house with cow dung.}

\fmtLemma{pwa}
\fmtPOS{ideo}
\fmtSense{in a manner of a shattered object}
\fmtExampleRL{Egi\-raa\-si egwi\-re ansi n’e\-ya\-ti\-ka pwa.}
\fmtExampleEN{The glass fell down and broke completely.}
\fmtSense{very talkative}
\fmtExampleRL{ONa\-ka\-kaa\-wa ata pwa pwa pwa.}
\fmtExampleEN{Nakakaawa is very talkative.}

\fmtLemma{pwa\-nya}
\fmtCitation{(oku)\-pwa\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pwa\-nyi\-re}
\fmtSense{to skewer, to put (on a skewer)}
\fmtExampleRL{Pwa\-nya enya\-ma eni oku nsaa\-le.}
\fmtExampleEN{Put this meat on the skewer.}

\fmtLemma{pye\-ke\-ta}
\fmtCitation{(oku)\-pye\-ke\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{pye\-ke\-te\-re}
\fmtSense{to winnow, to remove chaff from grain using a winnower}
\fmtExampleRL{Ndi ani nku\-pye\-ke\-ta mu\-we\-mba gwa\-nge.}
\fmtExampleEN{I am here winnowing my sorghum.}

\fmtLemma{pyo\-ko}
\fmtCitation{(o)pyo\-ko}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{rhinoceros, rhino}
\fmtExampleRL{Opyo\-ko abba mu\-koo\-to era wa iye\-mbe li\-mwei.}
\fmtExampleEN{The rhinoceros is a big animal and has one horn.}
\end{letter}
