\begin{letter}{R}

\fmtLemma{raa\-ka}
\fmtCitation{(oku)\-raa\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{raa\-ki\-re}
\fmtSense{(of fish) to spoil, to go bad}
\fmtExampleRL{Encwi zi\-raa\-ki\-re.}
\fmtExampleEN{The fish has spoilt.}

\fmtLemma{raa\-ra}
\fmtCrossreference{\textnormal{see} laala}

\fmtLemma{raa\-tya}
\fmtCitation{(oku)\-raa\-tya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{raa\-ti\-rye}
\fmtSense{to put under warm ashes (\eg food, to be kept for later use), to put under cover, to hide}
\fmtExampleRL{Tu\-maa\-re oku\-lya mwa\-be mu\-raa\-tye eme\-re eni.}
\fmtExampleEN{We have finished eating, so go and put the food under warm ashes.}

\fmtLemma{ra\-di\-yo}
\fmtCrossreference{\textnormal{see} ladiyo}

\fmtLemma{ra\-gaa\-na}
\fmtCrossreference{\textnormal{see} lagaana}
\fmtDialect{Ruruuli}

\fmtLemma{ra\-gaa\-ni\-sya}
\fmtCrossreference{\textnormal{see} lagaanisya}
\fmtDialect{Ruruuli}

\fmtLemma{Ra\-ma\-nzaa\-ni}
\fmtCitation{Ra\-ma\-nzaa\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{Ramadan}
\fmtExampleRL{Omu\-si\-raa\-mu odi ali mu\-kwe\-zi kwa Ra\-ma\-nzaa\-ni.}
\fmtExampleEN{That Moslem is observing the month of Ramadan.}

\fmtLemma{ra\-mya}
\fmtCitation{(oku)\-ra\-mya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ra\-mi\-rye}
\fmtSense{to pray, to worship}
\fmtExampleRL{Mwi\-je tu\-ra\-mye oKa\-nca.}
\fmtExampleEN{Come and let's pray to God.}

\fmtLemma{ra\-ngi}
\fmtCrossreference{\textnormal{see} langi}

\fmtLemma{ra\-pya}
\fmtCitation{(oku)\-ra\-pya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ra\-pi\-rye}
\fmtSubentry(a)
\fmtSense{to roam, to wander, to ramble}
\fmtExampleRL{Abaa\-na ba ku nguu\-do ba\-ra\-pya omu ki\-bu\-ga bu\-li ki\-ro.}
\fmtExampleEN{The street children roam the city every day.}
\fmtSubentry(b)
\fmtSense{to fornicate, to have sexual intercourse with someone one is not married to}
\fmtExampleRL{Omwa\-la oyo ata\-ndi\-ki\-re oku\-ra\-pya.}
\fmtExampleEN{That girl has started to fornicate.}

\fmtLemma{ra\-ra\-nga}
\fmtCitation{(oku)\-ra\-ra\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ra\-ra\-ngi\-re}
\fmtSense{to go into exile}
\fmtExampleRL{Abai\-ti\-re oMa\-ga\-ra ba\-ra\-ra\-ngai\-re.}
\fmtExampleEN{Those who killed Magara went into exile.}
\fmtSense{to spread, to propagate}
\fmtExampleRL{Eki\-ga\-mbo kya mu\-ka\-ma ni kya\-ba kye\-yo\-nge\-ra oku\-ra\-ra\-nga.}
\fmtExampleEN{The word of God continued to spread.}

\fmtLemma{ra\-ra\-ngya}
\fmtCitation{(oku)\-ra\-ra\-ngya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ra\-ra\-ngi\-rye}
\fmtSense{to spread (information), to disseminate, to make public, to make known}
\fmtExampleRL{Na\-ni oyo aku\-ra\-ra\-ngya ama\-wuu\-ro ga ku\-fa?}
\fmtExampleEN{Who is spreading the news of death?}

\fmtLemma{re\-vu\-ra\-ndi}
\fmtCitation{(o)re\-vu\-ra\-ndi}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from En.}
\fmtSense{Reverend (form of address to members of the clergy)}

\fmtLemma{rii\-no}
\fmtDialect{Ruruuli}
\fmtCitation{(e)rii\-no}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{tooth}
\fmtExampleRL{Erii\-no li\-ku\-mu\-lu\-ma.}
\fmtExampleEN{He has a toothache.}

\fmtLemma{rii\-so}
\fmtCitation{(e)rii\-so}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSubentry(a)
\fmtSense{eye}
\fmtExampleRL{Amai\-so gaa\-mwe ga\-koo\-to.}
\fmtExampleEN{His eyes are big.}
\fmtSubentry(b)
\fmtSense{bud (of a tree)}
\fmtExampleRL{Omu\-saa\-le gwi\-rye amai\-so.}
\fmtExampleEN{The buds of the tree have sprouted.}
\fmtSubentry(c)
\fmtSense{concave container used to keep grease for bicycles in case of emergency}
\fmtSubentry(d)
\fmtSense{vanilla bean, vanilla pod}
\fmtExampleRL{Ova\-ni\-la atai\-re\-ku amai\-so.}
\fmtExampleEN{The vanilla is bearing beans.}

\fmtLemma{ri\-ndu\-kya}
\fmtCrossreference{\textnormal{see} lindukya}
\fmtDialect{Ruruuli}

\fmtLemma{ro\-fu}
\fmtCrossreference{\textnormal{see} lofu}

\fmtLemma{ro\-ka}
\fmtCitation{(oku)\-ro\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ro\-ke\-re}
\fmtSense{to sprout, to germinate}
\fmtExampleRL{Emwa\-ni zi\-ro\-ke\-re ma\-ngu.}
\fmtExampleEN{The coffee seedlings have sprouted quite fast.}

\fmtLemma{ro\-kya}
\fmtCitation{(oku)\-ro\-kya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ro\-ke\-rye}
\fmtSubentry(a)
\fmtSense{to beat thoroughly, to give a good thrashing}
\fmtExampleRL{Oyo ba\-mu\-ro\-ke\-rye\-mbe.}
\fmtExampleEN{They have beaten him thoroughly.}
\fmtSubentry(b)
\fmtSense{to prune trees to allow good sprouts to grow}
\fmtExampleRL{Aba\-li\-mi\-sya ba\-tu\-so\-me\-se\-rye oku\-ro\-kya ebi\-ko\-nge.}
\fmtExampleEN{The agricultural officers taught us how to prune trees in order to allow new sprouts to grow.}

\fmtLemma{roo\-ba\-roo\-ba\-na}
\fmtCitation{(oku)\-roo\-ba\-roo\-ba\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{roo\-ba\-roo\-bai\-ne}
\fmtSense{to be covered with, to be immersed}
\fmtExampleRL{Omwa\-na aroo\-ba\-roo\-bai\-ne ebi\-nya\-nga\-ta.}
\fmtExampleEN{The child is covered in mud.}

\fmtLemma{roo\-ka\-roo\-ka}
\fmtCrossreference{\textnormal{see} lookalooka}
\fmtDialect{Ruruuli}

\fmtLemma{Ru\-ba\-ngo}
\fmtCitation{(o)Ru\-ba\-ngo}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{the person believed to be the ancestral father of the Banyala and Baruuli}
\fmtExampleRL{Aba\-ruu\-li n’Aba\-nya\-la bai\-zu\-ku\-lu ba Ru\-ba\-ngo.}
\fmtExampleEN{The Baruuli and the Banyala are Rubango's grandchildren.}

\fmtLemma{ru\-ba\-zo}
\fmtCrossreference{\textnormal{see} lubazo}

\fmtLemma{ru\-bba\-bbya}
\fmtCrossreference{\textnormal{see} lubbabbya}
\fmtDialect{Ruruuli}

\fmtLemma{ru\-bba\-nga}
\fmtCitation{(o)ru\-bba\-nga}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{type of spirit, god}
\fmtExampleRL{Oru\-bba\-nga ta\-si\-moo\-la, bai\-tu aga\-ba emi\-ki\-sa.}
\fmtExampleEN{Rubbanga does not speak, but he gives blessings.}

\fmtLemma{ru\-bi\-ngo}
\fmtCrossreference{\textnormal{see} lubingo}

\fmtLemma{ru\-bo\-go}
\fmtCitation{(o)ru\-bo\-go}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{brown male animal (a bull or he-goat)}
\fmtExampleRL{Oru\-bo\-go wa maa\-ma ba\-mu\-twai\-re.}
\fmtExampleEN{Mother's brownish bull is taken.}

\fmtLemma{ru\-bo\-na}
\fmtPOS{adv}
\fmtSubentry(a)
\fmtSense{red-handed}
\fmtExampleRL{Na\-mu\-kwai\-te ru\-bo\-na n’o\-mu\-ka\-mu\-sai\-za.}
\fmtExampleEN{I caught him red-handed with a married woman.}
\fmtSubentry(b)
\fmtSense{eyewitness, observer, witness}
\fmtExampleRL{Nii\-we oli ru\-bo\-na?}
\fmtExampleEN{Are you an eyewitness?}

\fmtLemma{ru\-bu\-ga}
\fmtCrossreference{\textnormal{see} lubuga}
\fmtDialect{Ruruuli}

\fmtLemma{ru\-cwa\-mba}
\fmtCrossreference{\textnormal{see} lucwamba}
\fmtDialect{Ruruuli}

\fmtLemma{ru\-di\-gi\-do}
\fmtCrossreference{\textnormal{see} ludigido}
\fmtDialect{Ruruuli}

\fmtLemma{ru\-fa}
\fmtCitation{(o)ru\-fa}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{dead person, corpse, dead body, the deceased}
\fmtExampleRL{Oru\-fa ba\-ku\-mu\-lee\-ta lee\-ro.}
\fmtExampleEN{They are bringing the corpse today.}

\fmtLemma{ru\-fu}
\fmtCrossreference{\textnormal{see} lufu}

\fmtLemma{ru\-ga}
\fmtCitation{(oku)\-ru\-ga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ru\-gi\-re}
\fmtSense{to give up, to surrender}
\fmtExampleRL{Bai\-ra\-ba nto\-be\-re n’e\-nsi, bai\-tu ti\-na\-gi\-ru\-ge\-ku.}
\fmtExampleEN{My dear, I have toiled in this world, but I am not giving up.}

\fmtLemma{ru\-gaa\-zu}
\fmtCitation{(o)ru\-gaa\-zu}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{male animal (a bull or he-goat) with a brownish skin}
\fmtExampleRL{Ni na\-ya\-kai\-re ebi\-buu\-lyo ni ba\-mpa oru\-gaa\-zu wa nte.}
\fmtExampleEN{When I passed the exams, they gave me a brown bull.}

\fmtLemma{Ru\-ga\-ba}
\fmtDialect{Ruruuli}
\fmtCitation{(o)Ru\-ga\-ba}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{God}
\fmtExampleRL{ORu\-ga\-ba otu\-wai\-re enju\-ra omwa\-ka gu\-ni.}
\fmtExampleEN{God has blessed us with rain this year.}

\fmtLemma{ru\-ga\-bo}
\fmtCitation{(o)ru\-ga\-bo}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{bull with black and white or brown and black patches}
\fmtExampleRL{Oru\-ga\-bo wa nte abo\-ne\-ka ku\-sai.}
\fmtExampleEN{Rugabo looks good.}

\fmtLemma{ru\-ga\-nga\-mai\-ru}
\fmtCitation{(o)ru\-ga\-nga\-mai\-ru}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{one of the gods among the Banyal and the Baruuli}

\fmtLemma{ru\-ge\-nyi}
\fmtCrossreference{\textnormal{see} lugenyi}
\fmtDialect{Ruruuli}

\fmtLemma{ru\-go\-ndo}
\fmtCitation{(o)ru\-go\-ndo}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{striped white and brown bull}
\fmtExampleRL{Oru\-go\-ndo odi mu\-na\-mu\-koo\-to.}
\fmtExampleEN{That spotted white and brown bull is very big.}

\fmtLemma{ru\-goo\-be}
\fmtCitation{(o)ru\-goo\-be}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{striped black and brown cow}
\fmtExampleRL{Oru\-goo\-be ataa\-mu emya\-laa\-lo.}
\fmtExampleEN{Rugoobe has black and brown stripes.}

\fmtLemma{ru\-gu\-ru}
\fmtPOS{adv}
\fmtSubentry(a)
\fmtSense{north, toward the north, to the north}
\fmtExampleRL{Ya\-ya\-bi\-re ru\-gu\-ru.}
\fmtExampleEN{He went to the north.}
\fmtSubentry(b)
\fmtSense{upward, upwards, up, to the top, skywards}
\fmtExampleRL{Aya\-bi\-re ru\-gu\-ru.}
\fmtExampleEN{He has gone upwards.}

\fmtLemma{ru\-guu\-ra}
\fmtCitation{(oku)\-ru\-guu\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ru\-gui\-re}
\fmtSense{to surround, to besiege}
\fmtExampleRL{Mwe\-ru\-guu\-re enyu\-mba mu\-mu\-bbwe.}
\fmtExampleEN{Surround the house and arrest him.}

\fmtLemma{ru\-guu\-rya}
\fmtCitation{(oku)\-ru\-guu\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ru\-guu\-rii\-rye}
\fmtSubentry(a)
\fmtSense{to wrap around, to coil around, to encircle, to surround}
\fmtExampleRL{Ru\-guu\-rya aka\-ko\-si ka\-ni oku mu\-twe.}
\fmtExampleEN{Wrap around this string on the head.}
\fmtSubentry(b)
\fmtSense{to rotate}
\fmtExampleRL{Omwa\-na aku\-ru\-guu\-rya nzi\-ga ya gaa\-li.}
\fmtExampleEN{The child is rotating the wheel of the bicycle.}

\fmtLemma{Ru\-ha\-nga}
\fmtCrossreference{\textnormal{see} Ruwanga}

\fmtLemma{ru\-ju\-mba}
\fmtCitation{(o)ru\-ju\-mba}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{brown and black chicken}
\fmtExampleRL{Oru\-ju\-mba tu\-mu\-wai\-re omu\-lwai\-re.}
\fmtExampleEN{We have given a brown and black chicken to a sick person.}

\fmtLemma{ru\-ke\-re\-nge}
\fmtCitation{(o)ru\-ke\-re\-nge}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{greyish bull}
\fmtExampleRL{Ya\-guu\-re onte oru\-ke\-re\-nge.}
\fmtExampleEN{He bought a greyish bull.}

\fmtLemma{ru\-ko\-me}
\fmtCrossreference{\textnormal{see} lukome}

\fmtLemma{ru\-ko\-ndo}
\fmtCitation{(o)ru\-ko\-ndo}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{bull with horns growing forward}
\fmtExampleRL{Tu\-boi\-ne onte ru\-ko\-ndo.}
\fmtExampleEN{We have seen a bull with horns growing forward.}

\fmtLemma{ru\-ko\-ro\-go}
\fmtCrossreference{\textnormal{see} lukorogo}

\fmtLemma{ru\-ku\-ba}
\fmtCitation{(o)ru\-ku\-ba}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{thunder}
\fmtExampleRL{Oru\-ku\-ba omu\-wu\-lii\-re n’a\-ba\-lu\-ki\-re?}
\fmtExampleEN{Did you hear the thunder?}

\fmtLemma{ru\-ku\-ngo}
\fmtCrossreference{\textnormal{see} lukungo}


\fmtLemma{ru\-ku\-ra\-to}
\fmtCitation{(o)ru\-ku\-ra\-to}
\fmtPOS{n}%
\fmtNC{11}
\fmtSense{council, meeting, assembly, parliamentary session}
\fmtExampleRL{Oru\-ku\-ra\-to lu\-bbai\-re\-mu aba\-ntu ba\-doo\-li.}
\fmtExampleEN{There were few people in the meeting.}

\fmtLemma{ru\-kuu\-ma\-nfuu\-zi}
\fmtCrossreference{\textnormal{see} rukuumansuuzi}
\fmtDialect{Lunyala}

\fmtLemma{ru\-kuu\-ma\-nsuu\-zi}
\fmtDialect{Ruruuli}
\fmtCitation{(o)ru\-kuu\-ma\-nsuu\-zi}
\fmtNC{1a/2}
\fmtSense{person who cares for the orphans}
\fmtExampleRL{Eri\-ya ni\-ye ru\-kuu\-ma\-nsuu\-zi.}
\fmtExampleEN{Eriya is the person who cares for the orphans.}

\fmtLemma{ru\-la}
\fmtCitation{(o)ru\-la}
\fmtPOS{n}%
\fmtNC{11}
\fmtSense{lineage}
\fmtExampleRL{Oru\-la lwa\-mu lwa\-fii\-re.}
\fmtExampleEN{That child is from my lineage.}

\fmtLemma{ru\-le\-mba}
\fmtCitation{(o)ru\-le\-mba}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{bull with black or brown skin and white dotted patches}
\fmtExampleRL{Oru\-le\-mba ya\-mu\-tu\-ndi\-re aka\-kai\-re ka\-mwei n’o\-mso\-ro\-ngo.}
\fmtExampleEN{He sold a ``rulemba" for over one million.}

\fmtLemma{ru\-maa\-da}
\fmtCrossreference{\textnormal{see} lumaada}

\fmtLemma{ru\-ma\-nyo}
\fmtCrossreference{\textnormal{see} lumanyo}

\fmtLemma{ru\-ngi\-ra}
\fmtCrossreference{\textnormal{see} lungira}
\fmtDialect{Kiryando\-ngo}

\fmtLemma{ru\-nguu\-ra}
\fmtCrossreference{\textnormal{see} lunguura}
\fmtDialect{Ruruuli}

\fmtLemma{ru\-ngu\-wa}
\fmtCitation{(oku)\-ru\-ngu\-wa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ru\-ngu\-wi\-re}
\fmtSubentry(a)
\fmtSense{to sort (objects), to sort out}
\fmtExampleRL{Omu bi\-ntu ebyo ru\-ngu\-wa\-mu ebya\-mu.}
\fmtExampleEN{Sort out your things from those ones.}
\fmtSubentry(b)
\fmtSense{to discriminate against}
\fmtExampleRL{Aba\-guu\-da ba\-ru\-ngu\-wa aboo\-ro.}
\fmtExampleEN{The rich discriminate against the poor.}

\fmtLemma{ru\-ngwa\-ru\-ngwa}
\fmtCitation{(oku)\-ru\-ngwa\-ru\-ngwa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{ru\-ngu\-wi\-re\-ru\-ngu\-wi\-re}
\fmtSense{to walk absent-mindedly, to walk preoccupied}
\fmtExampleRL{Omu\-kai\-re oyo lwa\-ki aku\-ru\-ngwa\-ru\-ngwa.}
\fmtExampleEN{Why is that old woman moving in an absent-minded way?}

\fmtLemma{ru\-nwa}
\fmtCitation{(o)ru\-nwa}
\fmtPOS{n}%
\fmtNC{11}
\fmtSense{quarrelsomeness, argumentativity}
\fmtExampleRL{Omu\-ka\-li oyo wa ru\-nwa.}
\fmtExampleEN{That woman is quarrelsome.}

\fmtLemma{Ru\-ruu\-li}
\fmtCitation{(o)Ru\-ruu\-li}
\fmtPOS{n}%
\fmtNC{11}
\fmtSense{the Ruruuli language}
\fmtExampleRL{Oru\-ruu\-li n’Olu\-nya\-la zii\-saa\-na\-nga\-na.}
\fmtExampleEN{Ruruuli and Lunyala are similar.}

\fmtLemma{ru\-sa}
\fmtCitation{(o)ru\-sa}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{greyish bull or he-goat}
\fmtExampleRL{Ombu\-li oru\-sa aya\-kai\-re ani.}
\fmtExampleEN{A greyish he-goat has passed here.}

\fmtLemma{ru\-sa\-ra}
\fmtCitation{(o)ru\-sa\-ra}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{bull or he-goat with saggy horns hanging downwards}
\fmtExampleRL{ODa\-ma\-li ya\-tii\-na\-nga onte oru\-sa\-ra.}
\fmtExampleEN{Damali used to be afraid of a bull with saggy horns.}

\fmtLemma{ru\-su\-zi}
\fmtCitation{(o)ru\-su\-zi}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSubentry(a)
\fmtSense{genet}
\fmtExampleRL{Oru\-su\-zi alya enko\-ko.}
\fmtExampleEN{The genet eats hens.}
\fmtSubentry(b)
\fmtSense{undomesticated cat, wild cat, feral cat}
\fmtExampleRL{Oru\-su\-zi ama\-lii\-re enko\-ko za\-nge.}
\fmtExampleEN{The undomesticated cat has eaten all my chicken.}

\fmtLemma{ru\-taa\-ra}
\fmtCitation{(o)ru\-taa\-ra}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{bull with horns pointing sideways}
\fmtExampleRL{Oru\-taa\-ra wa nte ba\-mwe\-ndya mu\-no.}
\fmtExampleEN{Rutaara is liked so much.}

\fmtLemma{ru\-ta\-nga}
\fmtCitation{(o)ru\-ta\-nga}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{bull or he-goat with black-and-white spotted skin}

\fmtLemma{ru\-ti\-mba}
\fmtCitation{(o)ru\-ti\-mba}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{greyish bull with stripes}

\fmtLemma{ruu\-ba}
\fmtDialect{Ruruuli}
\fmtCitation{(oku)\-ruu\-ba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ruu\-bi\-re}
\fmtSubentry(a)
\fmtSense{to mix, to blend, to mingle, to combine}
\fmtExampleRL{Omwo\-mbo\-ki ya\-ruu\-bi\-re ose\-mi\-nti n’o\-mu\-se\-nyu ale n’a\-ya\-mbo\-ka omwo\-mo.}
\fmtExampleEN{The builder mixed cement with sand before building the wall.}
\fmtSubentry(b)
\fmtSense{powdered millet with yeast}
\fmtExampleRL{Oruu\-ba ni\-ye ayi\-sya ma\-lwa.}
\fmtExampleEN{The powered millet with yeast is the one used to make finger millet beer.}

\fmtLemma{ruu\-bya}
\fmtDialect{Ruruuli}
\fmtCitation{(oku)\-ruu\-bya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ruu\-bi\-rye}
\fmtSense{to mix (with), to stir (with or into), to mix up}
\fmtExampleRL{Ruu\-bya osu\-kaa\-li omu ma\-tai.}
\fmtExampleEN{Mix milk with sugar.}

\fmtLemma{ruu\-tya}
\fmtCitation{(oku)\-ruu\-tya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ruu\-ti\-rye}
\fmtSense{to throw, to cast away, to toss away}
\fmtExampleRL{Ruu\-tya\-yo ei\-bbaa\-le eryo.}
\fmtExampleEN{Throw away that stone.}

\fmtLemma{ruu\-ya}
\fmtCitation{(oku)\-ruu\-ya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ruu\-yi\-rye}
\fmtSubentry(a)
\fmtSense{to wrap}
\fmtExampleRL{Ruu\-ya aka\-vee\-ra ako oku ki\-mu\-li.}
\fmtExampleEN{Wrap that flower with a polythene bag.}
\fmtSubentry(b)
\fmtSense{to wind a rope around (\eg a pole)}
\fmtExampleRL{Ruu\-ya omu\-gu\-wa oku ki\-ko\-ndo.}
\fmtExampleEN{Wind the rope on the pole.}

\fmtLemma{ru\-waa\-si}
\fmtCrossreference{\textnormal{see} luwaasi}
\fmtDialect{Ruruuli}

\fmtLemma{Ru\-wa\-nga}
\fmtCitation{(o)Ru\-wa\-nga}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{God}
\fmtExampleRL{Oru\-wa\-nga atu\-wai\-re enju\-ra.}
\fmtExampleEN{God has blessed us with rain.}

\fmtLemma{ru\-ze\-ru}
\fmtCitation{(o)ru\-ze\-ru}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{white male cow, goat}
\fmtExampleRL{Aguu\-re onte oru\-ze\-ru.}
\fmtExampleEN{He bought a white male cow.}

\fmtLemma{rwa\-nda\-ga\-da}
\fmtCitation{(o)rwa\-nda\-ga\-da}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{very tall person}
\fmtExampleRL{OKa\-ye ata rwa\-nda\-ga\-da.}
\fmtExampleEN{Kaye is very tall.}

\fmtLemma{rwa\-rwa\-ngu\-la}
\fmtCitation{(oku)\-rwa\-rwa\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{rwa\-rwa\-ngwi\-re}
\fmtSense{to sound an alarm, to make a very loud alarm}
\fmtExampleRL{Naa\-ni aku\-rwa\-rwa\-ngu\-la enduu\-lu?}
\fmtExampleEN{Who is making such a loud alarm?}

\fmtLemma{rwa\-ta}
\fmtCitation{(o)rwa\-ta}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{cold-hearted person}
\fmtExampleRL{Ba\-rwa\-ta bai\-ngi mu nsi mu\-ni.}
\fmtExampleEN{There are many cold-hearted people in this world.}

\fmtLemma{rwa\-ta\-maa\-to}
\fmtCitation{(o)rwa\-ta\-maa\-to}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{strong wind on lakes}
\fmtExampleRL{Orwa\-maa\-to aya\-ti\-rye obwa\-to.}
\fmtExampleEN{The strong wind has broken the boats.}

\fmtLemma{rwei\-ndu\-la}
\fmtCitation{(o)rwei\-ndu\-la}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{untrustworthy person, trustless person}
\fmtExampleRL{Mu\-tee\-si\-ga mu\-ka\-li oyo ku\-bba ata rwe\-yi\-ndu\-la.}
\fmtExampleEN{Don’t trust the woman because she is untrustworthy.}

\fmtLemma{Rwe\-nzo\-ri}
\fmtCitation{(o)Rwe\-nzo\-ri}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{the Rwenzori Mountains range}
\fmtExampleRL{Anii\-ni\-re olu\-so\-zi lwa Rwe\-nzo\-ri.}
\fmtExampleEN{He has climbed the Rwenzori Mountains.}

\fmtLemma{rwi\-to}
\fmtCrossreference{\textnormal{see} lwito}

\fmtLemma{rwo\-da\-ma}
\fmtCitation{(oku)\-rwo\-da\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{rwo\-dai\-me, rwo\-da\-mi\-re}
\fmtSense{to squat, to crouch (down), to sit on one's heels}
\fmtExampleRL{Omu\-ka\-li arwo\-dai\-me oku nzi\-ra.}
\fmtExampleEN{The woman is squatting alongside the road.}

\fmtLemma{rwo\-zi}
\fmtCitation{(o)rwo\-zi}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{black bull or he-goat}

\fmtLemma{rya\-buu\-kii\-rya}
\fmtCitation{(oku)\-rya\-buu\-kii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{rya\-buu\-ki\-rii\-rye}
\fmtSense{to act recklessly}
\fmtExampleRL{Lwa\-ki orya\-buu\-kii\-rya bu\-lii\-zo?}
\fmtExampleEN{Why do you do things recklessly every now and then?}
\end{letter}