\begin{letter}{S}

\fmtLemma{saa\-baa\-na}
\fmtCitation{(oku)\-saa\-baa\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-bai\-ne}
\fmtSubentry(a)
\fmtSense{to be generously smeared with, to be stained all over with}
\fmtExampleRL{Omwa\-na asaa\-bai\-ne ama\-gi\-ta.}
\fmtExampleEN{The child is smeared with a lot of jelly.}
\fmtSubentry(b)
\fmtSense{to be drunk}

\fmtLemma{Saa\-ba\-di\-ko\-ni}
\fmtCitation{(o)Saa\-ba\-di\-ko\-ni}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{archdeacon (of the Church of Uganda)}
\fmtExampleRL{OSaa\-ba\-di\-ko\-ni ni\-ye akuu\-ra obu\-saa\-ba\-di\-ko\-ni.}
\fmtExampleEN{The archdeacon is the head of the archdeaconry.}

\fmtLemma{saa\-ba\-la}
\fmtCitation{(oku)\-saa\-ba\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-bai\-re}
\fmtSense{to embark, to get onto (a car or bus)}
\fmtExampleRL{Erya\-to eri\-sa\-ba\-lya aba\-ntu ti\-ri\-ri\-na ku\-ta\-mbu\-lya mbu\-li.}
\fmtExampleEN{A boat which carries people should not be used to ferry goats.}

\fmtLemma{saa\-bo}
\fmtCitation{(o)saa\-bo}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{mud}
\fmtExampleRL{Epai\-pu ezwi\-re osaa\-bo.}
\fmtExampleEN{The pipe clogged with mud.}

\fmtLemma{saa\-ga}%
\fmtOne{1}
\fmtCitation{(oku)\-saa\-ga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{saa\-gi\-re}
\fmtSubentry(a)
\fmtSense{to remain, to stay (behind)}
\fmtExampleRL{Na\-saa\-gi\-re lwa\-ku\-bba na\-bbai\-re mu\-lwai\-re.}
\fmtExampleEN{I stayed behind because I was sick.}
\fmtSubentry(b)
\fmtSense{to be in surplus, to be left over, to be omitted, to be left out}
\fmtExampleRL{Ba\-tei\-re ebi\-ntu omu nyu\-mba ebi\-ndi bi\-saa\-gi\-re.}
\fmtExampleEN{They have put things into the house, and some have remained.}

\fmtLemma{saa\-ga}%
\fmtOne{2}
\fmtCitation{(oku)\-saa\-ga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{saa\-gi\-re}
\fmtSense{to tell jokes, to crack jokes}
\fmtExampleRL{Abaa\-na ba\-ku\-saa\-ga.}
\fmtExampleEN{The children are telling jokes.}

\fmtLemma{saa\-kaa\-la}
\fmtCitation{(oku)\-saa\-kaa\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{saa\-kai\-re}
\fmtSense{(of voice) to be hoarse}
\fmtExampleRL{Lwa\-ki ei\-ra\-ka lya\-mu li\-saa\-kai\-re?}
\fmtExampleEN{Why is your voice hoarse?}

\fmtLemma{saa\-ko}
\fmtCrossreference{\textnormal{see also} kigoodo}
\fmtCitation{(e)saa\-ko}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{circle, ring}

\fmtLemma{saa\-la}
\fmtCitation{(e)saa\-la}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{prayer}
\fmtExampleRL{OMu\-sii\-ta ni\-ye ata\-ndi\-ki\-re esaa\-la.}
\fmtExampleEN{It is Musiita who started the prayer.}

\fmtLemma{saa\-li}
\fmtCitation{(e)saa\-li}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{saari (a traditional rectangle of woven cloth)}
\fmtExampleRL{Aguu\-re eki\-koo\-yi ekya ki\-ka kya saa\-li.}
\fmtExampleEN{She bought a saari undergarment.}

\fmtLemma{saa\-li\-rwa}
\fmtCitation{(oku)\-saa\-li\-rwa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-lii\-rwe}
\fmtSense{to feel bad about, to be disappointed, to be sorry/regretful}
\fmtExampleRL{Nsaa\-lii\-rwe nya\-ji\-rye mo\-to\-ka ende\-ke\-re.}
\fmtExampleEN{I was disappointed that by the time I arrived the bus had left.}

\fmtLemma{saa\-li\-za}
\fmtCitation{(oku)\-saa\-li\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-lii\-ze}
\fmtSense{to make someone feel a desire for something, to admi\-re}
\fmtExampleRL{Obu\-sai bwa\-mwe bu\-mu\-saa\-li\-za.}
\fmtExampleEN{He admires her beauty.}

\fmtLemma{saa\-ma}%
\fmtOne{1}
\fmtCitation{(oku)\-saa\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-mi\-re}
\fmtSubentry(a)
\fmtSense{to process (bark cloth), to make (bark cloth)}
\fmtExampleRL{Aba\-nya\-la ba\-saa\-ma\-nga embu\-go.}
\fmtExampleEN{The Banyala used to make barkcloth.}
\fmtSubentry(b)
\fmtSense{(of a caterpillar with a thorn-like spike) to spike}
\fmtExampleRL{Mbai\-re nkwa\-ka\-la\-wo ensaa\-mo n’e\-nsaa\-ma.}
\fmtExampleEN{The caterpillar spiked me when I was passing by.}

\fmtLemma{saa\-ma}%
\fmtOne{2}
\fmtCitation{(oku)\-saa\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-mi\-re}
\fmtSense{to castrate}
\fmtExampleRL{Iza onsaa\-mi\-re onte wa\-nge.}
\fmtExampleEN{Come and castrate my bull.}

\fmtLemma{saa\-mu\-la}
\fmtCitation{(oku)\-saa\-mu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-mwi\-re}
\fmtSense{to push (away)}
\fmtExampleRL{Eki\-ke\-re ba\-ki\-sa\-mwi\-re enzai wa nyu\-mba.}
\fmtExampleEN{They pushed a frog outside of the house.}

\fmtLemma{saa\-mu\-lya}
\fmtCitation{(oku)\-saa\-mu\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-mwi\-rye}
\fmtSense{to splash, to splatter}
\fmtExampleRL{Saa\-mu\-lya ebi\-nya\-nga\-ta ebyo.}
\fmtExampleEN{Splash that mud.}

\fmtLemma{saa\-muu\-ka}
\fmtCitation{(oku)\-saa\-muu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{saa\-muu\-ki\-re}
\fmtSense{to splash, to splatter, to spray}
\fmtExampleRL{Amai\-zi ga\-saa\-muu\-ki\-re ku lu\-go\-ye.}
\fmtExampleEN{The water has splashed onto the cloth.}

\fmtLemma{saa\-na}
\fmtCitation{(oku)\-saa\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sai\-ne}
\fmtSubentry(a)
\fmtSense{to be obliged, to be required, to should}
\fmtExampleRL{Osai\-ne okwi\-za omu lu\-kii\-ko ei\-go\-lo ya\-lee\-ro.}
\fmtExampleEN{You are obliged to come to the meeting this afternoon.}
\fmtSubentry(b)
\fmtSense{to be suitable, to deserve, to be worthy of, to be entitled to, to have a right to}
\fmtExampleRL{Osaa\-na ku\-bba mu\-ka\-ma.}
\fmtExampleEN{You are suitable to be a leader.}

\fmtLemma{saa\-ni}
\fmtCitation{(e)saa\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{plate}
\fmtExampleRL{Lee\-ta esaa\-ni tu\-lii\-re\-ku eme\-re.}
\fmtExampleEN{Bring the plates and let's eat food.}

\fmtLemma{saa\-nii\-ka}
\fmtCitation{(oku)\-saa\-nii\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-nii\-ki\-re}
\fmtSense{to cover (food with leaves)}
\fmtExampleRL{Saa\-nii\-ka eme\-re n’e\-nda\-ga\-la.}
\fmtExampleEN{Cover the food with banana leaves.}

\fmtLemma{saa\-ni\-ki\-ra}
\fmtCitation{(oku)\-saa\-ni\-ki\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{saa\-ni\-kii\-re}
\fmtSense{to cover, to put cover on}
\fmtExampleRL{Saa\-ni\-ki\-ra oku nsu\-wa.}
\fmtExampleEN{Cover the pot, please.}

\fmtLemma{saa\-ni\-ra}
\fmtCitation{(oku)\-saa\-ni\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-nii\-re}
\fmtSense{to be worthy of, to deserve, to be suitable for, to be fit for, to suit, to merit}
\fmtExampleRL{Edwa\-rd asaa\-nii\-re oku\-bba pu\-re\-si\-de\-nti.}
\fmtExampleEN{Edward is suitable to be a president.}

\fmtLemma{saa\-ni\-rya}
\fmtCitation{(oku)\-saa\-ni\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-nii\-rye}
\fmtSubentry(a)
\fmtSense{to make level, to make smooth, to put straight}
\fmtExampleRL{Olu\-bu\-ga olwo lu\-saa\-ni\-rye.}
\fmtExampleEN{Level that courtyard.}
\fmtSubentry(b)
\fmtSense{to straighten out (a situation), to pacify, to settle}

\fmtLemma{saa\-nuu\-ka}
\fmtCitation{(oku)\-saa\-nuu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{saa\-nuu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to melt, to become liquefied by heat}
\fmtExampleRL{Ebi\-ba\-ngo n’o\-byo\-ca n’o\-muu\-ro ogwi\-ngi bi\-saa\-nuu\-ka.}
\fmtExampleEN{When you heat the ore with very high temperatures it melts.}
\fmtSubentry(b)
\fmtFixedexpression{(e)nzo\-ka saa\-nuu\-ka}
\fmtSense{to have sharp abrupt abdominal pain}
\fmtExampleRL{Enzo\-ka zi\-nsaa\-nuu\-ki\-re omu ki\-da.}
\fmtExampleEN{I've got sharp abdominal pain.}

\fmtLemma{saa\-nuu\-la}
\fmtCitation{(oku)\-saa\-nuu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-nwi\-re}
\fmtSense{to melt, to liquefy by heat}
\fmtExampleRL{Nku\-saa\-nuu\-la ki\-do\-mo\-la.}
\fmtExampleEN{I am melting a jerry can.}

\fmtLemma{saa\-nya}
\fmtCitation{(oku)\-saa\-nya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{saa\-nyi\-re}
\fmtSubentry(a)
\fmtSense{to swim}
\fmtExampleRL{Abaa\-na ba\-ku\-saa\-nya.}
\fmtExampleEN{The children are swimming.}
\fmtSubentry(b)
\fmtSense{(of a pot) to be damaged during baking}
\fmtExampleRL{Ebi\-na\-ga bya mu\-ru\-ndi gu\-ni bya\-saa\-nyi\-re.}
\fmtExampleEN{The pots were damaged during baking.}

\fmtLemma{saa\-ra}%
\fmtOne{1}
\fmtCitation{(oku)\-saa\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-re}
\fmtSubentry(a)
\fmtSense{to cut}
\fmtExampleRL{Abai\-bbi baa\-saa\-re ewa\-ya za si\-mu.}
\fmtExampleEN{Thieves cut the telephone cables.}
\fmtSubentry(b)
\fmtSense{to prune}
\fmtExampleRL{Nkwa\-ba ku\-saa\-ra bi\-go\-go bya\-nge.}
\fmtExampleEN{I am going to prune my banana plants.}

\fmtLemma{saa\-ra}%
\fmtOne{2}
\fmtDialect{Ruruuli}
\fmtCitation{(oku)\-saa\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{saa\-re}
\fmtSense{to taste bitter}
\fmtExampleRL{Eki\-ga\-gi ki\-saa\-ra mu\-no.}
\fmtExampleEN{Aloe vera is a very bitter plant.}

\fmtLemma{saa\-rwa}
\fmtCitation{(oku)\-saa\-rwa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-rwe}
\fmtSense{to regret, to be sorry about}
\fmtExampleRL{Ndi\-na oku\-saa\-rwa ku\-bba tai\-zi\-re.}
\fmtExampleEN{I regret that he didn’t come.}

\fmtLemma{saa\-sa}
\fmtCitation{(oku)\-saa\-sa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-si\-re}
\fmtSense{(of a body part) to hurt, to ache}
\fmtExampleRL{Eki\-da ki\-ku\-nsaa\-sa mu\-no.}
\fmtExampleEN{I have abdominal pain. (lit.\,The stomach hurts me.)}

\fmtLemma{saa\-saa\-na}%
\fmtOne{1}
\fmtCitation{(oku)\-saa\-saa\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{saa\-sai\-ne}
\fmtSense{to dawn, (of a day) to begin}
\fmtExampleRL{Twai\-ki\-re obwi\-re bu\-ku\-saa\-saa\-na.}
\fmtExampleEN{We arrived at dawn. (lit.\,We arrived when it was dawning.)}

\fmtLemma{saa\-saa\-na}%
\fmtOne{2}
\fmtCitation{(oku)\-saa\-saa\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{saa\-sai\-ne}
\fmtSense{to disperse, to scatter}
\fmtExampleRL{Aba\-ntu ba\-saa\-sai\-ne omu\-ko\-lo gu\-wei\-re.}
\fmtExampleEN{The function has come to an end and the people have dispersed.}

\fmtLemma{saa\-si\-ra}
\fmtCitation{(oku)\-saa\-si\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-sii\-re}
\fmtSubentry(a)
\fmtSense{to forgive, to pardon, to feel no resentment toward, to feel no malice toward, to harbour no grudge against}
\fmtExampleRL{Saa\-si\-ra\-nga aku\-koo\-re oku\-bbi.}
\fmtExampleEN{Forgive whoever has wronged you.}
\fmtSubentry(b)
\fmtSense{to have pity to, to show compassion for}
\fmtExampleRL{Ki\-sai oku\-saa\-si\-ra aba\-ntu aba\-lwai\-re.}
\fmtExampleEN{It is good to show compassion for the sick.}

\fmtLemma{saa\-ti}
\fmtCitation{(e)saa\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSubentry(a)
\fmtSense{shirt}
\fmtExampleRL{OKa\-sa\-ndo azwai\-re esaa\-ti egi\-tu\-ku\-li\-ki.}
\fmtExampleEN{Kasando is wearing a red shirt.}
\fmtSubentry(b)
\fmtFixedexpression{saa\-ti ya mi\-ko\-no gi\-mpi}
\fmtSense{short-sleeved shirt}
\fmtExampleRL{Aboo\-jo ba\-zwa\-la saa\-ti ya mi\-ko\-no gi\-mpi.}
\fmtExampleEN{Boys wear short-sleeved shirts.}
\fmtSubentry(c)
\fmtFixedexpression{saa\-ti ya miko\-no gi\-rai}
\fmtSense{long-sleeved shirt}
\fmtExampleRL{Aba\-sai\-za ba\-zwa\-la saa\-ti za mi\-ko\-no gi\-rai.}
\fmtExampleEN{Men wear long-sleeved shirts.}

\fmtLemma{saa\-wa}%
\fmtOne{1}
\fmtCitation{(e)ssaa\-wa}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSubentry(a)
\fmtSense{time, moment, period}
\fmtExampleRL{Okwi\-ra saa\-wa ime\-ka?}
\fmtExampleEN{At what time are you coming?}
\fmtSubentry(b)
\fmtSense{clock, watch}
\fmtExampleRL{Esaa\-wa ya ku mwo\-mo gi\-koo\-to.}
\fmtExampleEN{The wall clock is big.}

\fmtLemma{saa\-wa}%
\fmtOne{2}
\fmtCitation{(oku)\-saa\-wa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{saa\-wi\-re}
\fmtSense{to cut grass using a slasher, to slash natural vegetation as a way of clearing land}
\fmtExampleRL{Saa\-wa oku lu\-bu\-ga lwa\-nge.}
\fmtExampleEN{Slash my compound.}

\fmtLemma{saa\-wa\-saa\-wa}
\fmtPOS{adv}
\fmtSense{as expected, expectedly}
\fmtExampleRL{Ai\-ki\-re saa\-wa\-saa\-wa.}
\fmtExampleEN{She has arrived as expected.}

\fmtLemma{saa\-wu\-la}
\fmtCitation{(oku)\-saa\-wu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-wii\-re}
\fmtSense{to remove (bush, to vegetation), to slash (bush), to clear bush}
\fmtExampleRL{Saa\-wu\-la eki\-si\-ko ekyo obu\-ta\-ka\-la Sa\-nde.}
\fmtExampleEN{Clear that bush by Sunday.}

\fmtLemma{saa\-ya\-nsi}
\fmtCitation{(o)saa\-ya\-nsi}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{science}
\fmtExampleRL{Nku\-so\-ma saa\-ya\-nsi oku kye\-ge\-se\-ryo.}
\fmtExampleEN{I am studying science at the university.}

\fmtLemma{saa\-yii\-rya}
\fmtCitation{(oku)\-saa\-yii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-yii\-rye}
\fmtSubentry(a)
\fmtSense{to repair, to mend, to fix}
\fmtExampleRL{Saa\-yii\-rya enzi\-ga ya gaa\-li ya\-nge.}
\fmtExampleEN{Repair my bicycle wheel.}
\fmtSubentry(b)
\fmtSense{to improve, to modernise, to redecorate, to renovate}
\fmtExampleRL{Okwo oku\-ju\-gu\-ta kwo\-ba\-za kwa\-li ku\-saa\-yii\-rye\-mu.}
\fmtExampleEN{The kind of metal forging you are talking about has been improved.}
\fmtSubentry(c)
\fmtSense{to clear (land, plot)}
\fmtExampleRL{Aku\-saa\-yii\-rya ndu\-mi\-ro yaa\-mwe.}
\fmtExampleEN{He is clearing his garden.}
\fmtSubentry(d)
\fmtSense{to edit, to clean up}
\fmtExampleRL{Saa\-yii\-rya eki\-dee\-ro kyai\-swe.}
\fmtExampleEN{Edit our dictionary please.}

\fmtLemma{saa\-yi\-rii\-rya}
\fmtCitation{(oku)\-saa\-yi\-rii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-yi\-rii\-rye}
\fmtSense{to trim, to cut}
\fmtExampleRL{Ya\-ba osaa\-yi\-rii\-rye ei\-so\-ke lya\-mu.}
\fmtExampleEN{Go and trim your hair.}

\fmtLemma{sa\-ba}
\fmtCitation{(oku)\-sa\-ba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-bi\-re}
\fmtSubentry(a)
\fmtSense{to ask for, to request}
\fmtExampleRL{Sa\-ba ky’o\-ku\-ta\-ka nki\-ku\-we\-rye.}
\fmtExampleEN{Ask for what you want, and I'll give it to you.}
\fmtSubentry(b)
\fmtSense{to pray to}
\fmtExampleRL{Ema\-mbya nkwa\-ba ku\-sa\-ba oKa\-nca.}
\fmtExampleEN{Tomorrow I am going to pray to God.}
\fmtSubentry(c)
\fmtSense{to ask to marry, woo away (by following the traditional official courtship of the Baruuli and Banyala)}
\fmtExampleRL{Aya\-bi\-re ku\-sa\-ba kaa\-na ka nko\-ko.}
\fmtExampleEN{He has gone for the courtship ceremony. (lit.\,He has gone to ask for a chic.)}

\fmtLemma{sa\-bbaa\-wa}
\fmtCitation{(e)sa\-bbaa\-wa}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSubentry(a)
\fmtSense{machine gun}
\fmtSubentry(b)
\fmtSense{target shooting}
\fmtExampleRL{Omu\-taa\-ni wa\-nge amai\-te oku\-ku\-bba esa\-bbaa\-wa.}
\fmtExampleEN{My son knows how to shoot at a target.}

\fmtLemma{sa\-bbii\-ti}
\fmtCitation{(e)sa\-bbii\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSubentry(a)
\fmtSense{week}
\fmtExampleRL{Abbai\-re ani sa\-bbii\-ti gi\-ra\-mba.}
\fmtExampleEN{He has been here for a full week.}
\fmtSubentry(b)
\fmtSense{the Sabbath}
\fmtExampleRL{Esa\-bbii\-ti ebbaa\-wo Lwa\-mu\-kaa\-ga.}
\fmtExampleEN{The Sabbath is on Saturday.}

\fmtLemma{sa\-bbuu\-ni}
\fmtCitation{(o)sa\-bbuu\-ni}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from Ar. via Sw.}
\fmtSense{soap}
\fmtExampleRL{Osa\-bbuu\-ni oni ta\-yo\-za ku\-sai.}
\fmtExampleEN{This soap doesn’t clean well.}

\fmtLemma{sa\-bii\-rya}
\fmtCitation{(oku)\-sa\-bii\-rya}
\fmtPOS{v}%
\fmtTrans{ditr}
\fmtPFV{sa\-bi\-rii\-rye}
\fmtSense{to beg, to ask for, to plead for}
\fmtExampleRL{We\-te to\-ke\-rwa ku\-sa\-bii\-rya se\-nte omaa\-ma waa\-mu ?}
\fmtExampleEN{Don’t you get tired of begging your mother for money?}

\fmtLemma{sa\-bi\-ka}
\fmtCitation{(oku)\-sa\-bi\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-bi\-ki\-re}
\fmtSense{to wrap, to cover e.g.\,in paper}
\fmtExampleRL{Sa\-bi\-ka eki\-ta\-bo kya\-mu.}
\fmtExampleEN{Wrap your book.}

\fmtLemma{sa\-bu\-la}
\fmtCitation{(oku)\-sa\-bu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-bwi\-re}
\fmtSubentry(a)
\fmtSense{to dress smartly, to put on smart clothes}
\fmtExampleRL{Asa\-bu\-la engo\-ye ezi\-sai.}
\fmtExampleEN{She dresses smartly.}
\fmtSubentry(b)
\fmtSense{to whip, to lash}
\fmtExampleRL{Na\-mu\-sa\-bwi\-re emii\-go.}
\fmtExampleEN{I whipped him.}

\fmtLemma{sa\-bu\-lee\-nya}
\fmtCitation{(o)sa\-bu\-lee\-nya}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{Nile perch}

\fmtLemma{sa\-daa\-ka}
\fmtCitation{(oku)\-sa\-daa\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-daa\-ki\-re}
\fmtSubentry(a)
\fmtSense{to offer a sacrifice, to offer or kill as a religious sacrifice}
\fmtExampleRL{Aba\-si\-raa\-mu ba\-ta\-ka mu\-no oku\-sa\-daa\-ka.}
\fmtExampleEN{Muslims like offering sacrifices.}
\fmtSubentry(b)
\fmtSense{sacrifice, ritual slaughter, offering}
\fmtExampleRL{Ya\-zwa\-li\-ki\-re esaa\-da\-ka y’e\-nta\-ma edi oKa\-nca.}
\fmtExampleEN{He sacrificed a sheep to God.}

\fmtLemma{sa\-faa\-li}
\fmtCitation{(e)sa\-faa\-li}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Ar. via Sw.}
\fmtSense{journey, trip, tour}
\fmtExampleRL{Aya\-bi\-re sa\-faa\-li.}
\fmtExampleEN{He has gone on a journey.}

\fmtLemma{sa\-ga\-la}
\fmtCitation{(oku)\-sa\-ga\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-gai\-re}
\fmtSense{to be loose, to be slack, to wiggle}
\fmtExampleRL{Omu\-ki\-ra gwa ntaa\-ma gu\-sa\-ga\-la.}
\fmtExampleEN{The sheep's tail wiggles.}

\fmtLemma{sa\-ga\-li}
\fmtCitation{\textnormal{(e.g.}\,gi\-sa\-ga\-li, mu\-sa\-ga\-li, bi\-sa\-ga\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSense{loose, detached, unfastened}
\fmtExampleRL{Emi\-su\-maa\-li gya ente\-be eyo gi\-sa\-ga\-li.}
\fmtExampleEN{The nails of that chair have become loose.}

\fmtLemma{sa\-ga\-lya}
\fmtCitation{(oku)\-sa\-ga\-lya}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{sa\-gai\-rye}
\fmtSubentry(a)
\fmtSense{to shake}
\fmtExampleRL{ONa\-ka\-le\-be aku\-sa\-ga\-lya nsa\-nsi.}
\fmtExampleEN{What's-her-name is playing the shakers. (lit.\,What's-her-name is shaking the shakers.)}
\fmtSubentry(b)
\fmtSense{to wiggle, to let loose}
\fmtExampleRL{Aku\-sa\-ga\-lya ka\-bi\-na.}
\fmtExampleEN{She is wiggling her bum.}
\fmtSubentry(c)
\fmtSense{to rustle speedily}
\fmtExampleRL{Omu\-swi gu\-sa\-ga\-lya nga gu\-kwa\-ka\-la mu bi\-su\-bi.}
\fmtExampleEN{The cane rat rustles speedily when it runs through grass.}

\fmtLemma{sa\-ga\-ru\-ka}
\fmtCitation{(oku)\-sa\-ga\-ru\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-ga\-ru\-ki\-re}
\fmtSense{to be joyful, to jubilate, to rejoice}
\fmtExampleRL{Oku mba\-ga aba\-ntu ba\-sa\-ga\-ru\-ka.}
\fmtExampleEN{People rejoiced during the wedding.}

\fmtLemma{sa\-ga\-ru\-ki}
\fmtCitation{\textnormal{(e.g.}\,sa\-ga\-ru\-ki, ba\-sa\-ga\-ru\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{joyous, cheerful, jolly}
\fmtExampleRL{Ya\-swei\-re omu\-ka\-li omu\-sa\-ga\-ru\-ki.}
\fmtExampleEN{He married a joyous wife.}

\fmtLemma{sa\-gi\-ka}
\fmtCitation{(oku)\-sa\-gi\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-gi\-ki\-re}
\fmtSense{to support, to help, to aid, to assist, to contribute to}
\fmtExampleRL{Aba\-nya\-ra mu\-sa\-gi\-ke entee\-ka\-ni\-za ya ku\-jwi\-gi\-sya OLu\-nya\-la.}
\fmtExampleEN{The Banyara should support the program for the development of the Lunyala language.}

\fmtLemma{sai}
\fmtCitation{\textnormal{(e.g.}\,ki\-sai, mu\-sai, lu\-sai\textnormal{)}}
\fmtPOS{adj}
\fmtSense{good, nice, kind, beautiful, handsome}
\fmtExampleRL{Oli mu\-ka\-li mu\-sai.}
\fmtExampleEN{You are a nice woman.}

\fmtLemma{sai\-su\-la}
\fmtCitation{(oku)\-sai\-su\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sai\-swi\-re}
\fmtSense{to destroy (a bird's nest)}
\fmtExampleRL{Owa\-mboo\-li asai\-su\-la ebi\-nyu\-mba bya ki\-syo.}
\fmtExampleEN{A black kite destroys the weaverbird's nest.}

\fmtLemma{sai\-su\-lya}
\fmtCitation{(oku)\-sai\-su\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sai\-swi\-rye}
\fmtSubentry(a)
\fmtSense{to release, to untie, to unwind}
\fmtExampleRL{Ya\-ba osai\-su\-lye ombu\-li odi.}
\fmtExampleEN{Go and unwind that goat.}
\fmtSubentry(b)
\fmtSense{to separate, to take apart, to break up}
\fmtExampleRL{Ya\-ba osai\-su\-lye ba\-di aba\-ku\-lwa\-na.}
\fmtExampleEN{Go and separate those who are fighting.}

\fmtLemma{sai\-sya}
\fmtCitation{(oku)\-sai\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sai\-si\-rye}
\fmtSense{to release (a prison)}
\fmtExampleRL{Omu\-bbwe ba\-mu\-sai\-si\-rye.}
\fmtExampleEN{The prisoner has been released.}

\fmtLemma{sai\-za}
\fmtCitation{\textnormal{(e.g.}\,ki\-sai\-za, bi\-sai\-za, gu\-sai\-za\textnormal{)}}
\fmtPOS{adj}
\fmtSense{male}
\fmtExampleRL{Omu\-paa\-paa\-li ogu\-sai\-za gu\-taa\-ku oba\-paa\-paa\-li bu\-dyo\-li.}
\fmtExampleEN{A male pawpaw plant yields tiny pawpaws.}

\fmtLemma{sai\-zaa\-la}
\fmtCrossreference{\textnormal{see} sai\-zaa\-la}
\fmtDialect{Buyende}

\fmtLemma{sai\-za\-ku\-la}
\fmtCitation{(oku)\-sai\-za\-ku\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sai\-za\-kwi\-re}
\fmtSense{(of a boy) to become an adult}
\fmtExampleRL{Abbai\-re mu\-si\-ga\-zi bai\-tu ati asai\-za\-kwi\-re.}
\fmtExampleEN{He was a boy, now he has become an adult.}

\fmtLemma{sai\-zi}
\fmtCitation{(e)sai\-zi}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{size}
\fmtExampleRL{Esai\-zi yaa\-mwe ti\-ngi\-mai\-te.}
\fmtExampleEN{I don’t know her size.}

\fmtLemma{sa\-ju\-wa\-la}
\fmtCitation{(oku)\-sa\-ju\-wa\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-ju\-wai\-re}
\fmtSense{to fatten, to get heavier, to grow fatter}
\fmtExampleRL{Onko\-ko asa\-ju\-wai\-re.}
\fmtExampleEN{The chicken has fattened.}

\fmtLemma{sa\-ka}
\fmtCitation{(oku)\-sa\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-ki\-re}
\fmtSubentry(a)
\fmtSense{to dig (a hole or a pit), to excavate, to mine, to drill (e.g.\,for oil)}
\fmtExampleRL{Baa\-sa\-ki\-re ekii\-na mu nzi\-ra aba\-ntu ne\-ba\-fee\-ra\-mu.}
\fmtExampleEN{They dug a hole in the road, and it killed people.}
\fmtSubentry(b)
\fmtSense{to dig up (e.g.\,potatoes)}
\fmtExampleRL{Omu\-ka\-li aya\-bi\-re mu ndu\-mi\-ro ku\-sa\-ka bi\-yaa\-ta.}
\fmtExampleEN{The woman has gone to the garden to dig up potatoes.}

\fmtLemma{sa\-kaa\-ra}
\fmtCitation{(oku)\-sa\-kaa\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-kai\-re}
\fmtSubentry(a)
\fmtSense{to thatch, to cover (a roof or a building) with straw or a similar material}
\fmtExampleRL{Boo\-jo mu\-sa\-kaa\-re enyu\-mba.}
\fmtExampleEN{Please, thatch the house.}
\fmtSubentry(b)
\fmtSense{to roof, to cover with roof}
\fmtExampleRL{Ba\-ku\-sa\-kaa\-ra di enyu\-mba?}
\fmtExampleEN{When are they roofing the house?}

\fmtLemma{sa\-ka\-la\-me\-nto}
\fmtCitation{(e)sa\-ka\-la\-me\-nto}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{sacrament}
\fmtExampleRL{Na\-maa\-re esa\-ka\-la\-me\-nto ya ku\-ba\-ti\-zi\-bwa.}
\fmtExampleEN{I received the sacrament of baptism.}

\fmtLemma{sa\-la}
\fmtCitation{(oku)\-sa\-la}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{saa\-re}
\fmtSubentry(a)
\fmtSense{to cut, to slice}
\fmtExampleRL{Sa\-la ofe\-ne oni ebi\-tu\-ndu bi\-sa\-tu.}
\fmtExampleEN{Cut the jackfruit into three parts.}
\fmtSubentry(b)
\fmtSense{(of a tool) to be sharp}
\fmtExampleRL{Enso\-ne eni eku\-sa\-la mu\-no.}
\fmtExampleEN{This knife is very sharp.}
\fmtSubentry(c)
\fmtSense{to slaughter}
\fmtExampleRL{Nkwa\-ba ku\-sa\-la nko\-ko.}
\fmtExampleEN{I am going to slaughter a hen.}
\fmtSubentry(d)
\fmtSense{to put marks, to draw a boundary}
\fmtExampleRL{Sa\-la era\-yi\-ni omu ki\-saa\-we.}
\fmtExampleEN{Put marks in the playground.}
\fmtSubentry(e)
\fmtSense{to make a loss, to make losses, to incur losses}
\fmtExampleRL{Ebyai\-tu\-ndi bye\-ya\-su\-bwi\-re bya\-mu\-saa\-re.}
\fmtExampleEN{He made losses from his merchandise.}
\fmtSubentry(f)
\fmtSense{to ask for a price, to agree on a bride price}
\fmtExampleRL{Ba\-mu\-sa\-lii\-re omu\-twa\-lo gwa nya\-na iku\-mu.}
\fmtExampleEN{He was asked to pay a bride price worth ten heifers.}
\fmtSubentry(g)
\fmtSense{to determine a fine, to be fined}
\fmtExampleRL{Ba\-mu\-sa\-lii\-re emi\-twa\-lo asa\-tu.}
\fmtExampleEN{She was fined for three hundred thousand.}
\fmtSubentry(h)
\fmtSense{to officiate (a game), to referee}
\fmtExampleRL{Odii\-fi\-ri amai\-te oku\-sa\-la omu\-pii\-ra.}
\fmtSubentry(i)
\fmtExampleEN{The referee knows how to officiate a football match.}
\fmtSense{(of an animal) to be on heat, to be in heat}
\fmtExampleRL{Onte wa\-nge asaa\-re.}
\fmtExampleEN{My cow is on heat.}

\fmtLemma{sa\-la\-ka\-ta}
\fmtCitation{(oku)\-sa\-la\-ka\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-la\-ka\-ti\-re}
\fmtSense{(of water) to boil, to simmer}
\fmtExampleRL{Amai\-zi ga\-ta\-ndi\-ki\-re oku\-sa\-la\-ka\-ta.}
\fmtExampleEN{The water has started simmering.}

\fmtLemma{sa\-la\-la}
\fmtCitation{(oku)\-sa\-la\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-lai\-re}
\fmtSense{(of root tubers) to become watery (as a result of being left for too long in the ground)}
\fmtExampleRL{Ebi\-yaa\-ta ni bi\-sa\-la\-la ti\-bi\-woo\-ma.}
\fmtExampleEN{When potatoes are spoilt, they lose the taste.}

\fmtLemma{sa\-la\-li}
\fmtCitation{\textnormal{(e.g.}\,gu\-sa\-la\-li, bi\-sa\-la\-li, ka\-sa\-la\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSense{(of root tubers) spoilt}
\fmtExampleRL{Omwo\-go ogu\-sa\-la\-li ni bw’o\-gu\-su\-mba ti\-gu\-go\-nda.}
\fmtExampleEN{No matter how long you cook the spoilt cassava, it does not soften.}

\fmtLemma{sa\-la\-lu\-mu\-ka}
\fmtPOS{adv}
\fmtSense{speedily, quickly}
\fmtExampleRL{Emi\-ru\-mo ogi\-ko\-la sa\-la\-lu\-mu\-ka.}
\fmtExampleEN{You do your work speedily.}

\fmtLemma{sa\-la\-mu}
\fmtCitation{(oku)\-sa\-la\-mu}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-re\-mu}
\fmtSense{to partition, to divide, to split}
\fmtExampleRL{Sa\-la\-mu eki\-sii\-ka ekyo ebi\-tu\-ndu bi\-bi\-ri.}
\fmtExampleEN{Partition the room into two parts.}

\fmtLemma{sa\-la\-sai\-ta}
\fmtCitation{(oku)\-sa\-la\-sai\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-lai\-ti\-re}
\fmtSense{to rustle}
\fmtExampleRL{Ombwe\-ne aku\-sa\-la\-sai\-ta omu bi\-go\-go.}
\fmtExampleEN{The dog is rustling in the banana plantation.}

\fmtLemma{sa\-la\-sai\-ti\-ra}
\fmtCitation{(oku)\-sa\-la\-sai\-ti\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-lai\-sai\-tii\-re}
\fmtSense{to trespass}
\fmtExampleRL{Obwoi\-ne oku\-sa\-rai\-ti\-ra nka ne\-ku\-li oku\-bbi?}
\fmtExampleEN{Have you learnt how bad trespassing is?}

\fmtLemma{sa\-la\-sa\-la}
\fmtCitation{(oku)\-sa\-la\-sa\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-re\-saa\-re}
\fmtSense{to chop, to cut up}
\fmtExampleRL{Sa\-la\-sa\-la obbi\-ri\-ŋa\-nya oyo.}
\fmtExampleEN{Chop those eggplants.}

\fmtLemma{sa\-la\-wo}
\fmtCitation{(oku)\-sa\-la\-wo}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{saa\-re\-wo}
\fmtSense{to decide, to determine}
\fmtExampleRL{Sa\-la\-wo ekya\-ku\-ko\-la.}
\fmtExampleEN{Decide on what to do.}

\fmtLemma{sa\-la\-za}
\fmtCitation{(oku)\-sa\-la\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-lai\-ze}
\fmtSense{to hurt (a person), to offend, to hurt someone's feelings}
\fmtExampleRL{Eki\-ga\-mbo kye ba\-mu\-ko\-be\-re ki\-mu\-sa\-lai\-ze.}
\fmtExampleEN{He has been hurt by what he has been told.}

\fmtLemma{sa\-lii\-ra}
\fmtCitation{(oku)\-sa\-lii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-li\-rii\-re}
\fmtSubentry(a)
\fmtSense{to carve, to cut}
\fmtExampleRL{Ya\-sa\-li\-rii\-re olu\-baa\-wo.}
\fmtExampleEN{He carved the wood.}
\fmtSubentry(b)
\fmtFixedexpression{sa\-lii\-ra (a)ka\-le\-ra}
\fmtSense{to cut (the umbilical cord)}
\fmtExampleRL{Omu\-bya\-li\-sya asa\-lii\-ra aka\-le\-ra.}
\fmtExampleEN{The midwife cuts the umbilical cord.}

\fmtLemma{sa\-lii\-rya}
\fmtCitation{(oku)\-sa\-lii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-li\-rii\-rye}
\fmtSense{to be biased, to judge unfairly, to pass an unfair judgment on, to judge unjustly}
\fmtExampleRL{Dii\-fi\-ri asa\-li\-rii\-rye omu\-pii\-ra.}
\fmtExampleEN{The referee unjustly judged the football.}

\fmtLemma{sa\-li\-nkii\-rya}
\fmtCitation{(oku)\-sa\-li\-nkii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-li\-nki\-rii\-rye}
\fmtSense{to take a shortcut}
\fmtExampleRL{Ngo\-di aku\-sa\-li\-nkii\-rya ta\-ku\-ta\-ka tu\-mu\-bo\-ne.}
\fmtExampleEN{There he is taking a shortcut to avoid being seen.}

\fmtLemma{sa\-lu\-ti}
\fmtCitation{(e)sa\-lu\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{salute}
\fmtExampleRL{Ome\-ja ba\-mu\-ku\-bbii\-re esa\-lu\-ti.}
\fmtExampleEN{The major was saluted. (lit.\,They have made a salute for the major).}

\fmtLemma{sa\-ma}
\fmtCitation{(oku)\-sa\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-mi\-re}
\fmtSense{to make an incision (for treatment purposes or to draw blood)}
\fmtExampleRL{Na\-ya\-bi\-re wa mu\-sa\-mi n’a\-nsa\-ma emi\-sa\-le n’a\-ntaa\-mu omu\-ba\-zi.}
\fmtExampleEN{I went to the healer who made an incision in the skin and put herbs in it.}

\fmtLemma{sa\-mba}
\fmtCitation{(oku)\-sa\-mba}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{sa\-mbi\-re}
\fmtSubentry(a)
\fmtSense{to stamp, to clomp}
\fmtExampleRL{Ya\-sa\-mbi\-re oku ita\-ka\-li ne tu\-loo\-wo\-za nti mbu\-ndu.}
\fmtExampleEN{He stamped hard and we thought it was a gunshot.}
\fmtSubentry(b)
\fmtSense{to kick}
\fmtExampleRL{Ya\-sa\-mbi\-re olwi\-ge ni\-rwe\-gu\-la.}
\fmtExampleEN{He kicked the door open.}
\fmtSubentry(c)
\fmtSense{to extract banana juice by treading banana with one's feet}
\fmtSubentry(d)
\fmtSense{to trample (soil in order to make it soft before making bricks), to tread on}
\fmtExampleRL{Aba\-ku\-bbi ba bbu\-loo\-ka ba\-sa\-mba ei\-ta\-ka\-li.}
\fmtExampleEN{Brick makers tread on the soil.}
\fmtSubentry(e)
\fmtSense{to wean off (by kicking the calf)}
\fmtExampleRL{Onte aku\-sa\-mba onya\-na.}
\fmtExampleEN{The cow is weaning off the calf.}
\fmtSubentry(f)
\fmtSense{(of birds) to fight}
\fmtExampleRL{Enko\-ko zi\-ku\-sa\-mba\-nga\-na.}
\fmtExampleEN{The chickens are fighting.}
\fmtSubentry(g)
\fmtFixedexpression{sa\-mba omu\-pii\-ra}
\fmtSense{to play football}
\fmtExampleRL{Mba\-di ba\-ku\-sa\-mba mu\-pii\-ra.}
\fmtExampleEN{They are there playing football.}

\fmtLemma{sa\-mba\-ga\-la}
\fmtCitation{(oku)\-sa\-mba\-ga\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-mba\-gai\-re}
\fmtSense{to thrash about, to thrash around, to jerk}
\fmtExampleRL{Onsya ni ba\-mu\-su\-mi\-ti\-re ya\-sa\-mba\-gai\-re.}
\fmtExampleEN{After the duiker was speared, it thrashed around.}

\fmtLemma{sa\-mba\-nga\-na}
\fmtCitation{(oku)\-sa\-mba\-nga\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-mba\-nga\-nai\-ne}
\fmtSense{to do kickboxing}
\fmtExampleRL{Omu\-zee\-nyo gwa ku\-sa\-mba\-nga\-na gu\-bbi.}
\fmtExampleEN{Kickboxing is bad.}

\fmtLemma{sa\-mba\-tii\-ra}
\fmtCitation{(oku)\-sa\-mba\-tii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-mba\-ti\-rii\-re}
\fmtSense{to frequent, to visit regularly, to be a regular visitor to}
\fmtExampleRL{Ye okwa\-ba ku\-sa\-mba\-tii\-ra ma\-ka ga naa\-ni?}
\fmtExampleEN{Whose home are you frequenting like that?}

\fmtLemma{sa\-mbii\-ra}
\fmtCitation{(oku)\-sa\-mbii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-mbi\-rii\-re}
\fmtSense{to trod on}
\fmtExampleRL{Mu\-le\-ke\-ra\-wo oku\-sa\-mbii\-ra eki\-sa\-mba\-ga\-nyi kya\-nge?}
\fmtExampleEN{What are you treading on my mat?}

\fmtLemma{sa\-mbu\-la}
\fmtCitation{(oku)\-sa\-mbu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-mbwi\-re}
\fmtSense{to clear fields}
\fmtExampleRL{Tu\-li ampi oku\-sa\-mbu\-la.}
\fmtExampleEN{We are about to clear the fields.}

\fmtLemma{sa\-mu\-ka}
\fmtCitation{(oku)\-sa\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-mu\-ki\-re}
\fmtSense{(of rain) to ease up, to abade, to subside}
\fmtExampleRL{Oi\-ke\-ndi asa\-mu\-ki\-re.}
\fmtExampleEN{The rain has eased up.}

\fmtLemma{sa\-nda}
\fmtCitation{(e)sa\-nda}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Sw.}
\fmtSense{cloths used to bury Muslims}
\fmtExampleRL{Aba\-si\-raa\-mu oru\-fa ba\-mu\-bbwa mu sa\-nda.}
\fmtExampleEN{Dead Muslims are wrapped in sanda.}

\fmtLemma{sa\-ndaa\-la}
\fmtCitation{(oku)\-sa\-ndaa\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-ndai\-re}
\fmtSubentry(a)
\fmtSense{to be full to the brim}
\fmtExampleRL{Eki\-ko\-po ki\-sa\-ndai\-re.}
\fmtExampleEN{The cup is full to the brim.}
\fmtSubentry(b)
\fmtSense{to be very satiated (with food)}
\fmtExampleRL{Akaa\-na ka\-lii\-re ni ka\-sa\-ndaa\-la, ti\-ka\-kya\-ta\-ka kwe\-yo\-nge\-ra nya\-ma.}
\fmtExampleEN{The child is so satisfied that he does not want any more meat.}

\fmtLemma{sa\-nda\-bu\-ki}
\fmtCrossreference{\textnormal{see also} sanduuke}
\fmtCitation{(e)sa\-nda\-bu\-ki}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Ar. via Sw.}
\fmtSense{suitcase}
\fmtExampleRL{Nta n’e\-sa\-nda\-bu\-ki ibi\-ri omu nyu\-mba ya\-nge.}
\fmtExampleEN{I have two suitcases in my house.}

\fmtLemma{sa\-nda\-bu\-la}
\fmtCitation{(oku)\-sa\-nda\-bu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-nda\-bwi\-re}
\fmtSense{to untie (e.g.\,shoe laces), to unlock}
\fmtExampleRL{Sa\-nda\-bu\-la enkai\-to ozi\-too\-le\-mu.}
\fmtExampleEN{Untie the shoes and remove them.}

\fmtLemma{sa\-ndai\-ga}
\fmtCitation{(oku)\-sa\-ndai\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-ndai\-gi\-re}
\fmtSubentry(a)
\fmtSense{to make light incisions into the skin}
\fmtExampleRL{Omu\-fu\-mu amu\-sa\-ndai\-gi\-re omu\-ba\-zi.}
\fmtExampleEN{He has made a light incision to administer medicine.}
\fmtSubentry(b)
\fmtSense{to put a border on (a mat)}
\fmtExampleRL{Omu\-ka\-li aku\-sa\-ndai\-ga mu\-kee\-ka gwa\-mwe.}
\fmtExampleEN{A woman is putting a border on her mat.}

\fmtLemma{Sa\-nde}
\fmtCitation{(e)Sa\-nde}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSubentry(a)
\fmtSense{Sunday}
\fmtExampleRL{Aba\-mu\-po\-lo\-te\-si\-ta\-nti ba\-ra\-mya ku Sa\-nde.}
\fmtExampleEN{The Protestants pray on Sunday.}
\fmtSubentry(b)
\fmtSense{week}
\fmtExampleRL{Nai\-za mu sa\-nde ekwi\-za.}
\fmtExampleEN{I will come next week.}

\fmtLemma{sa\-nduu\-ke}
\fmtCitation{(e)sa\-nduu\-ke}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Ar. via Sw.}
\fmtSubentry(a)
\fmtSense{suitcase, case}
\fmtSubentry(b)
\fmtSense{coffin, casket, sarcophagus}
\fmtExampleRL{Aba\-fu ba\-ba\-lii\-ka omu sa\-nduu\-ke.}
\fmtExampleEN{The dead are buried in coffins.}

\fmtLemma{sa\-nduu\-ki}
\fmtCrossreference{\textnormal{see} sanduuke}

\fmtLemma{sa\-nga}
\fmtCitation{(oku)\-sa\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-ngi\-re}
\fmtSubentry(a)
\fmtSense{to meet, to encounter, to run into}
\fmtExampleRL{Na\-ku\-sa\-nga Iri\-ma.}
\fmtExampleEN{I will meet you at Irima.}
\fmtSubentry(b)
\fmtSense{to find}
\fmtExampleRL{Aba\-ma\-gye ba\-kwa\-ta obwa\-to bwe ba\-sa\-nga oku nya\-nja.}
\fmtExampleEN{The soldiers confiscate boats they find on the lake.}

\fmtLemma{sa\-ngu\-la}
\fmtCitation{(oku)\-sa\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-ngwi\-re}
\fmtSense{to erase, to remove, to delete}
\fmtExampleRL{Sa\-ngu\-la ebi\-ga\-mbo ebyo.}
\fmtExampleEN{Erase those words.}

\fmtLemma{sa\-ni\-ya}
\fmtCitation{(e)sa\-ni\-ya}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Sw.}
\fmtSense{tray}
\fmtExampleRL{Enya\-ma gi\-su\-ke oku sa\-ni\-ya.}
\fmtExampleEN{Put the meat on the tray.}

\fmtLemma{sa\-nja}
\fmtCitation{(e)sa\-nja}
\fmtPOS{n}%
\fmtNC{5}
\fmtSense{dry banana leaf}

\fmtLemma{sa\-nju}
\fmtPOS{num}
\fmtSense{seven}
\fmtExampleRL{Mme\-te\-re emi\-ru\-ndi mu\-sa\-nju nga ta\-ku\-nnye\-ga.}
\fmtExampleEN{I called him seven times, but he did not answer me.}

\fmtLemma{sa\-nsa\-li\-ga}
\fmtCitation{(oku)\-sa\-nsa\-li\-ga}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{sa\-nsa\-li\-gi\-re}
\fmtSubentry(a)
\fmtSense{to be scattered}
\fmtExampleRL{Ebi\-ta\-bo bya\-nge bya\-sa\-nsa\-li\-gi\-re.}
\fmtExampleEN{My books are scattered.}
\fmtSubentry(b)
\fmtSense{to spread out}
\fmtExampleRL{Sa\-nsa\-li\-ga obwi\-ta oku kya\-ni\-ki\-ro.}
\fmtExampleEN{Spread out the millet on the drying place.}

\fmtLemma{sa\-nsuu\-rya}
\fmtCitation{(oku)\-sa\-nsuu\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-nswi\-rye}
\fmtSense{to scatter, to spread, to distribute}
\fmtExampleRL{Ya\-sa\-nsuu\-rye engo\-ye zaa\-mwe zoo\-na zoo\-na omu ki\-sii\-ka bai\-tu ti\-ya\-boi\-ne\-mu paa\-si\-poo\-ta yaa\-mwe.}
\fmtExampleEN{He scattered his clothes all over the room but failed locate his passport.}

\fmtLemma{sa\-nya\-la\-la}
\fmtCitation{(oku)\-sa\-nya\-la\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-nya\-lai\-re}
\fmtSense{to be numb, to grow stiff, to feel paralysed}
\fmtExampleRL{Oku\-gu\-lu kwa\-nge ku\-sa\-nya\-lai\-re.}
\fmtExampleEN{My leg has become numb.}

\fmtLemma{sa\-nyu\-ka}
\fmtCitation{(oku)\-sa\-nyu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-nyi\-ki\-re}
\fmtSense{to be glad, to be happy}
\fmtExampleRL{Nkwi\-za ku\-sa\-nyu\-ka nga eki\-dee\-ro kya Ru\-ruu\-li-Lunyala kiweire.}
\fmtExampleEN{I will be happy when the Ruruuli-Lunyala dictionary is completed.}

\fmtLemma{sa\-nyu\-ki}
\fmtCitation{\textnormal{(e.g.}\,mu\-sa\-nyu\-ki, ba\-sa\-nyu\-ki, tu\-sa\-nyu\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{happy, joyous, cheerful}
\fmtExampleRL{Omu\-la\-bi\-ri\-zi mu\-sa\-nyu\-ki.}
\fmtExampleEN{The bishop is happy.}

\fmtLemma{sa\-nyu\-sa}
\fmtCitation{(oku)\-sa\-nyu\-sa}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{sa\-nyu\-si\-re}
\fmtSubentry(a)
\fmtSense{to be interesting}
\fmtExampleRL{Oru\-ruu\-li n’o\-Lu\-nya\-la zi\-sa\-nyu\-sa.}
\fmtExampleEN{Ruruuli and Lunyala are interesting languages.}
\fmtSubentry(b)
\fmtSense{to entertain, to please}
\fmtExampleRL{Aya\-bi\-re ku\-sa\-nyu\-sa ba\-ntu.}
\fmtExampleEN{She has gone to entertain people.}

\fmtLemma{sa\-nzi\-ra}
\fmtCitation{(oku)\-sa\-nzi\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-nzii\-re}
\fmtSense{to make a joke, to jest, to speak in a joking way}
\fmtExampleRL{Oi\-ba\-wa\-nge asa\-nzi\-ra mu\-no n’a\-baa\-na.}
\fmtExampleEN{My husband loves to make jokes with children.}

\fmtLemma{sa\-pa\-tu}
\fmtCitation{(e)sa\-pa\-tu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Port. via Sw.}
\fmtSense{sandal, slipper}
\fmtExampleRL{Esa\-pa\-tu ezo nzi\-gu\-lii\-re ise\-nga wa\-nge.}
\fmtExampleEN{I have bought a pair of sandals for my aunt.}

\fmtLemma{sa\-pu\-li}
\fmtCitation{(e)sa\-pu\-li}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from French via En.}
\fmtSense{chaplet (a string of beads), rosary}
\fmtExampleRL{Omu\-ka\-tu\-li\-ki azwai\-re esa\-pu\-li.}
\fmtExampleEN{A catholic is wearing a rosary.}

\fmtLemma{sa\-raa\-ra}
\fmtCitation{(oku)\-sa\-raa\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-rai\-re}
\fmtSense{to be numb}
\fmtExampleRL{Eki\-sa\-mbi kya\-nge ki\-sa\-rai\-re.}
\fmtExampleEN{My thigh is numb.}

\fmtLemma{sa\-ra\-la}
\fmtCrossreference{\textnormal{see} salala}

\fmtLemma{sa\-ra\-li}
\fmtCitation{\textnormal{(e.g.}\,mu\-sa\-ra\-li, ki\-sa\-ra\-li, gi\-sa\-ra\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSense{(of root tubers) spoilt}
\fmtExampleRL{Omwo\-go ogu\-sa\-ra\-li ni bw’o\-gu\-su\-mba ti\-gu\-go\-nda.}
\fmtExampleEN{No matter how long you cook the spoilt cassava, it does not soften.}

\fmtLemma{sa\-sa\-ma\-lya}
\fmtCitation{(oku)\-sa\-sa\-ma\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-sa\-mai\-rye}
\fmtSense{to incite, to encourage, to stimulate, to instigate, to provoke}
\fmtExampleRL{Lwa\-ki osa\-sa\-ma\-lya aba\-ntu?}
\fmtExampleEN{Why are you inciting people?}

\fmtLemma{sa\-sa\-mbu\-la}
\fmtCitation{(oku)\-sa\-sa\-mbu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-sa\-mbwi\-re}
\fmtSense{to peel off}
\fmtExampleRL{Lee\-ta enso\-ne tu\-sa\-sa\-mbu\-le emo\-le.}
\fmtExampleEN{Bring a knife and let's peel off reeds.}

\fmtLemma{sa\-si\-ka}
\fmtCitation{(oku)\-sa\-si\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-si\-ki\-re}
\fmtSense{to flower, to blossom, to bloom}
\fmtExampleRL{Omu\-saa\-le gu\-sa\-si\-ki\-re mu\-no.}
\fmtExampleEN{The tree has flowered a lot.}

\fmtLemma{sa\-si\-ra}
\fmtCitation{(oku)\-sa\-si\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sa\-sii\-re}
\fmtSense{to sprinkle}
\fmtExampleRL{Omu\-ba\-zi gu\-ni gu\-sa\-si\-re omu ire\-mbo lya\-mu.}
\fmtExampleEN{Sprinkle these herbs in the way right in front of your house.}

\fmtLemma{sa\-su\-la}
\fmtCitation{(oku)\-sa\-su\-la}
\fmtPOS{v}%
\fmtTrans{ditr}
\fmtPFV{sa\-swi\-re}
\fmtSense{to pay back, to pay, to repay}
\fmtExampleRL{Ka nku\-sa\-su\-le empii\-ya z’o\-ku\-nto\-nga.}
\fmtExampleEN{Let me pay you the money I owe you.}

\fmtLemma{satu}
\fmtCitation{\textnormal{(e.g.}\,ba\-sa\-tu\textnormal{)}}
\fmtPOS{num}
\fmtSense{three}
\fmtExampleRL{We\-nya\-yo aba\-te\-gi nka ba\-sa\-tu b’o\-mai\-te.}
\fmtExampleEN{Mention about three fishermen you know.}

\fmtLemma{sa\-tu\-ka}
\fmtCitation{(oku)\-sa\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sa\-tu\-ki\-re}
\fmtSense{to crack, to split, to fracture}
\fmtExampleRL{Eki\-ko\-po kya ibbu\-mba ki\-sa\-tu\-ki\-re.}
\fmtExampleEN{The cup has cracked.}

\fmtLemma{saa\-ya\-nsi}
\fmtCitation{(e)sa\-ya\-nsi}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{science}
\fmtExampleRL{Omwa\-na n’a\-so\-ma osaa\-ya\-nsi aso\-bo\-la oku\-fuu\-ka do\-ki\-ta.}
\fmtExampleEN{When a child studies sciences he can become a doctor.}

\fmtLemma{sa\-yi\-ri}
\fmtCitation{(o)sa\-yi\-ri}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{type of peas (smaller than cow peas)}

\fmtLemma{sa\-yi\-ri\-ra}
\fmtCitation{(oku)\-sa\-yi\-ri\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{saa\-yi\-rii\-rye}
\fmtSense{to smoothen, to make smooth}
\fmtExampleRL{Ko\-le\-sya aka\-cuu\-li onsa\-yi\-rii\-rye erya\-to lya\-nge.}
\fmtExampleEN{Use the sandpaper to smoothen for me my boat.}

\fmtLemma{se\-bo}
\fmtCitation{(o)se\-bo}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{sir}
\fmtExampleRL{Se\-bo nku\-ko\-be\-re ama\-zi\-ma?}
\fmtExampleEN{Sir, should I tell you the truth?}

\fmtLemma{se\-cu\-ngwa}
\fmtCitation{(o)se\-cu\-ngwa}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{grapefruit}
\fmtExampleRL{Ose\-cu\-ngwa abba mu\-koo\-to bai\-tu ta\-nu\-lii\-rya.}
\fmtExampleEN{The grapefruit is big but not sweet.}

\fmtLemma{se\-da}
\fmtCitation{(e)se\-da}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{Passover Seder (a Jewish ritual feast that marks the beginning of the Jewish holiday of Passover)}

\fmtLemma{see\-bee\-ka}
\fmtCitation{(oku)\-see\-bee\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{see\-bee\-ke\-re}
\fmtSense{to steam (e.g.\,greens) on top of food, to add on top}
\fmtExampleRL{Ona\-ka\-ti mu\-see\-bee\-ke ku ki\-na\-ga.}
\fmtExampleEN{Put nakati on the food parcel to steam it.}

\fmtLemma{see\-bee\-ke}
\fmtCitation{\textnormal{(e.g.}\,gi\-see\-bee\-ke, gu\-see\-bee\-ke, zi\-see\-bee\-ke\textnormal{)}}
\fmtPOS{adj}
\fmtSense{steamed (e.g.\,food)}
\fmtExampleRL{Nta\-ka oku\-lya omu\-ku\-bi ogu\-see\-bee\-ke.}
\fmtExampleEN{I like eating steamed sauce.}

\fmtLemma{see\-ga}
\fmtCitation{(oku)\-see\-ga}
\fmtOne{1}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{see\-ge\-re}
\fmtSubentry(b)
\fmtSense{to become poor, to become miserable, to be stricken with disease, to endure a prolonged illness}
\fmtExampleRL{Aba\-ntu ba ku\-ni ba\-see\-ge\-re olwa isa\-na okwi\-ta ebi\-ntu bya\-bwe.}
\fmtExampleEN{The people here became poor due the drought which destroyed all their crops.}

\fmtLemma{see\-ga}%
\fmtOne{2}
\fmtCitation{(oku)\-see\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{see\-ge\-re}
\fmtSense{to shake}
\fmtExampleRL{Omwa\-na ngo\-di aku\-see\-ga ensee\-ge.}
\fmtExampleEN{The child is shaking the shakers.}

\fmtLemma{see\-ge\-rya}
\fmtCitation{(oku)\-see\-ge\-rya}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{see\-gei\-rye}
\fmtSubentry(a)
\fmtSense{to grind (using the whole surface of a grinding stone)}
\fmtExampleRL{Sya ng’o\-ku\-see\-ge\-rya eme\-ngo ere\-ke kwi\-za\-mu kii\-na.}
\fmtExampleEN{Grind using the whole surface so that no dent forms on the grinding stone.}
\fmtSubentry(b)
\fmtSense{to start a dance with slow graceful movements}
\fmtExampleRL{Ng’o\-kwa\-ba oku\-bi\-na, soo\-ka ku\-see\-ge\-rya.}
\fmtExampleEN{Whenever you are going to dance, start slowly.}

\fmtLemma{see\-gu\-la}
\fmtCitation{(oku)\-see\-gu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{see\-gwi\-re}
\fmtSense{to speak vulgarly, to speak using bad language}
\fmtExampleRL{Waa\-li\-wo omu\-sai\-za eya\-see\-gu\-la\-nga.}
\fmtExampleEN{There was a vulgar man who used to speak vulgarly.}

\fmtLemma{see\-ka}%
\fmtOne{1}
\fmtCitation{(o)see\-ka}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from Ar.}
\fmtSense{sheikh (a leader in a Muslim community or organization)}
\fmtExampleRL{Omu\-zi\-gi\-ti gwa wai\-swe te\-gu\-ta na see\-ka.}
\fmtExampleEN{Our mosque does not have a sheikh.}

\fmtLemma{see\-ka}%
\fmtOne{2}
\fmtCitation{(oku)\-see\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{see\-ke\-re}
\fmtSense{to become emaciated, to loose weight (due to illness)}
\fmtExampleRL{Omwa\-la oni alwai\-re n’a\-see\-ka.}
\fmtExampleEN{This girl fell sick and lost weight.}

\fmtLemma{see\-nee\-na}
\fmtCitation{(oku)\-see\-nee\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{see\-nee\-ne\-re}
\fmtSense{to become exhausted, to abate, to subside diminish, to disappear}
\fmtExampleRL{Osa\-bbuu\-ni wa\-nge asee\-nee\-ne\-re.}
\fmtExampleEN{My soap is used up.}

\fmtLemma{see\-nya}
\fmtCitation{(oku)\-see\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{see\-nye\-re}
\fmtSubentry(a)
\fmtSense{to burn}
\fmtExampleRL{Omuu\-ro gu\-see\-nye\-re enyu\-mba yoo\-na ewei\-re\-wo.}
\fmtExampleEN{The fire has burnt down the entire house.}
\fmtSubentry(b)
\fmtSense{to eat a lot}
\fmtExampleRL{Bai\-tu omu\-sai\-za wa\-nge asee\-nya eme\-re!}
\fmtExampleEN{My dear, my husband eats a lot of food!}
\fmtSubentry(c)
\fmtSense{to beat thoroughly, to give a good thrashing}
\fmtExampleRL{Omwi\-bbi mmu\-see\-nye\-re omwi\-go olwa kwi\-bba onte.}
\fmtExampleEN{I have beaten a thief with a stick for stealing a cow.}
\fmtSubentry(d)
\fmtSense{to destroy}
\fmtSubentry(e)
\fmtSense{to collect (firewood)}
\fmtExampleRL{Ti\-ba\-see\-nye\-re nkwi, ati ti\-ba\-ku\-so\-bo\-la ku\-su\-mba me\-re yaa\-bwe.}
\fmtExampleEN{They didn't collect firewood, and now they cannot cook their food.}
\fmtSubentry(f)
\fmtFixedexpression{see\-nya (e)nkwi}
\fmtSense{to collect firewood (idiom. to pull labia minora)}
\fmtExampleRL{Ni ba\-su\-na ama\-bee\-re, ni ba\-see\-nya enkwi.}
\fmtExampleEN{When they develop breasts, they start lengthening their labia minora. (Lit.\,When they develop breasts, they start collecting firewood.)}

\fmtLemma{see\-nyuu\-ka}
\fmtCitation{(oku)\-see\-nyuu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{see\-nyuu\-ki\-re}
\fmtSense{(of hair or beard) to change colour due to malnutrition or illness}
\fmtExampleRL{Ei\-so\-ke lya mwa\-na li\-see\-nyuu\-ki\-re.}
\fmtExampleEN{The child's hair has changed its colour.}

\fmtLemma{see\-re\-ra}
\fmtCitation{(oku)\-see\-re\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{see\-rei\-re}
\fmtSense{to be slippery}
\fmtExampleRL{Enzi\-ra eku\-see\-re\-ra.}
\fmtExampleEN{The road is slippery.}

\fmtLemma{see\-see\-tu\-ka}
\fmtCitation{(oku)\-see\-see\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{see\-see\-tu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to slip, to slide, to fall off/down, to become loose, to swerve}
\fmtSubentry(b)
\fmtSense{to attain another level, to reach another level}
\fmtSubentry(c)
\fmtSense{to change one's mind}
\fmtExampleRL{Opu\-re\-zi\-de\-nti ya\-see\-see\-tu\-ki\-re omu ndo\-woo\-za.}
\fmtExampleEN{The president has changed his mind.}

\fmtLemma{see\-tu\-la}%
\fmtOne{1}
\fmtCitation{(oku)\-see\-tu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-twi\-re}
\fmtSubentry(a)
\fmtSense{to cause to move slowly, to push along, to slide along}
\fmtExampleRL{See\-tu\-la aba\-ntu abo bai\-gee\-re\-ku edi.}
\fmtExampleEN{Move these people to that end.}
\fmtSubentry(b)
\fmtSense{to push}
\fmtExampleRL{See\-tu\-la ente\-be eyoo\-gii\-rye edi.}
\fmtExampleEN{Push that chair to that place.}
\fmtSubentry(c)
\fmtSense{to set for grazing}
\fmtExampleRL{Ya\-ba osee\-tu\-le ente.}
\fmtExampleEN{Go and set the cows for grazing.}
\fmtSubentry(d)
\fmtSense{to dance}
\fmtExampleRL{Bo\-na omu\-ka\-li n’a\-ku\-see\-tu\-la enke\-nde.}
\fmtExampleEN{Look at the woman dancing.}

\fmtLemma{see\-yee\-ya}
\fmtCitation{(oku)\-see\-yee\-ya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{see\-yee\-ye\-re}
\fmtSubentry(a)
\fmtSense{to float, to sail}
\fmtExampleRL{Eki\-do\-mo\-la ki\-ku\-see\-yee\-ya ku nya\-nja.}
\fmtExampleEN{A jerry can is floating on the lake.}
\fmtSubentry(b)
\fmtSense{to glide, to hover}
\fmtExampleRL{Ebi\-nyo\-nyi bi\-ku\-see\-yee\-ya mu ibba\-nga.}
\fmtExampleEN{The birds are gliding in the air.}
\fmtSubentry(c)
\fmtSense{to drive slowly}
\fmtExampleRL{Emo\-to\-ka eku\-see\-yee\-ya ng’e\-ku\-twa\-la aba\-go\-le.}
\fmtExampleEN{The car is moving slowly carrying the newlyweds.}

\fmtLemma{se\-fu\-lya}
\fmtCitation{(e)se\-fu\-lya}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{saucepan}
\fmtExampleRL{Lee\-ta ese\-fu\-lya egi\-koo\-to mu\-no.}
\fmtExampleEN{Bring the biggest saucepan.}

\fmtLemma{se\-ge\-nya}
\fmtCitation{(oku)\-se\-ge\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-ge\-nye\-re}
\fmtSense{to winnow}
\fmtExampleRL{Se\-ge\-nya obwi\-ta tu\-bu\-sye.}
\fmtExampleEN{Winnow the millet so that we can grind it.}

\fmtLemma{se\-ju}
\fmtCitation{(o)se\-ju}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{spear grass (\textit{Heteropogon contortus}) or \textit{Imperata cylindrica}}
\fmtExampleRL{Ose\-ju asu\-mi\-ta mu\-to. N'a\-ku\-la ayo\-mbo\-ka ma\-nyu\-mba.}
\fmtExampleEN{Do whatever you are doing while still young. When you grow up, it will be too late. (a proverb, lit.\,The spear grass pierces while young, when it grows old, it thatches houses.)}

\fmtLemma{se\-ka}
\fmtCitation{(oku)\-se\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-ke\-re}
\fmtSubentry(a)
\fmtSense{to laugh}
\fmtExampleRL{Nku\-wuu\-ra aba\-ntu ba\-ku\-se\-ka.}
\fmtExampleEN{I hear people laughing.}
\fmtSubentry(b)
\fmtFixedexpression{se\-ka bi\-mo\-do}
\fmtSense{to laugh mockingly}
\fmtExampleRL{Oyo aku\-se\-ka bi\-mo\-do.}
\fmtExampleEN{That one is just laughing mockingly at you.}
\fmtSubentry(c)
\fmtSense{(of millet flower) to sprout}
\fmtExampleRL{Obwi\-ta bu\-se\-ke\-re.}
\fmtExampleEN{The millet has sprouted.}

\fmtLemma{se\-ka\-pu\-laa\-li}
\fmtCitation{(e)se\-ka\-pu\-laa\-li}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{scapular (a short monastic cloak covering the shoulders)}

\fmtLemma{se\-kee\-ra}
\fmtCitation{(oku)\-se\-kee\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-ke\-rei\-re}
\fmtSense{to laugh at}
\fmtExampleRL{Nku\-wuu\-ra aba\-ntu ba\-ku\-nse\-kee\-ra.}
\fmtExampleEN{I hear people laughing at me.}

\fmtLemma{se\-kee\-rya}
\fmtCitation{(oku)\-se\-kee\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-ke\-rei\-rye}
\fmtSense{to smile}
\fmtExampleRL{Oku\-se\-kee\-rya ku\-kuu\-ma ba\-ntu obu\-ta\-kai\-ri\-ka.}
\fmtExampleEN{Smiling keeps people young.}

\fmtLemma{se\-ke\-sa}
\fmtCitation{(o)se\-ke\-sa}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{type of hairy poisonous caterpillar}
\fmtExampleRL{Ose\-ke\-sa anku\-bbi\-re obu\-lu\-ngu.}
\fmtExampleEN{The sekesa caterpillar irritated me with its short stinging hair.}

\fmtLemma{se\-ke\-su}
\fmtCitation{(o)se\-ke\-su}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{porcupine}
\fmtExampleRL{Ose\-ke\-su abba n’a\-ma\-su\-mu oku lu\-su\-su lwa\-mwe.}
\fmtExampleEN{The porcupine has thorny body.}

\fmtLemma{se\-ko\-ko}
\fmtCitation{(o)se\-ko\-ko}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{turkey}
\fmtExampleRL{Aba\-na\-yu\-nda ti\-ba\-te\-ra ku\-lii\-sya se\-ko\-ko.}
\fmtExampleEN{Ugandans rarely rear turkeys.}

\fmtLemma{Se\-ku\-ku\-lu}
\fmtCitation{(e)Se\-ku\-ku\-lu}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{Christmas}
\fmtExampleRL{Ese\-ku\-ku\-lu ekwi\-ka ma\-mbya.}
\fmtExampleEN{Tomorrow is the Christmas day.}

\fmtLemma{se\-ku\-la}
\fmtCitation{(oku)\-se\-ku\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-kwi\-re}
\fmtSense{to pound softly (white ants, groundnuts, cassava, maize to remove the husks)}
\fmtExampleRL{Se\-ku\-la amai\-do tu\-fu\-ne omu\-ku\-bi.}
\fmtExampleEN{Pound the groundnuts so that we can get sauce.}

\fmtLemma{Se\-la}
\fmtCitation{(e)Se\-la}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{Selah (a word used 74 times in the Hebrew Bible, it encourages a reader to have a moment of silence and reflect on the text)}

\fmtLemma{se\-lee\-bu}
\fmtCitation{(o)se\-lee\-bu}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from En.}
\fmtSense{celebrity}
\fmtExampleRL{OKa\-ye afuu\-ki\-re se\-lee\-bu.}
\fmtExampleEN{Kaye has become a celebrity.}

\fmtLemma{se\-mba}%
\fmtOne{1}
\fmtCitation{(oku)\-se\-mba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-mbe\-re}
\fmtSubentry(a)
\fmtSense{to support}
\fmtSubentry(b)
\fmtSense{to fill (e.g.\,a crack)}
\fmtExampleRL{Nkwa\-ba ku\-se\-mba bi\-wu\-lu ku lwi\-ge.}
\fmtExampleEN{I am going to fill the door cracks.}
\fmtSubentry(c)
\fmtSense{to recommend, to propose, to nominate}
\fmtExampleRL{Naa\-ni eya\-ku\-se\-mbe\-re oku\-tu\-nga omu\-lu\-mu gu\-ni.}
\fmtExampleEN{Who recommended you to get this job?}

\fmtLemma{se\-mba}%
\fmtOne{2}
\fmtCitation{(oku)\-se\-mba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-mbe\-re}
\fmtSense{to come last, to arrive last, to be last}
\fmtExampleRL{Omwe\-gi eya\-se\-mba te\-ya\-fu\-na ki\-two.}
\fmtExampleEN{A student who comes last will not get a gift.}

\fmtLemma{se\-mba\-yo}
\fmtCitation{(oku)\-se\-mba\-yo}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-mbe\-re\-yo}
\fmtSense{to come last}
\fmtExampleRL{Otee\-kee\-ra, bai\-tu se\-mba\-yo.}
\fmtExampleEN{Don’t come first, instead come last.}

\fmtLemma{se\-mbe\-ra}
\fmtCitation{(oku)\-se\-mbe\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-mbei\-re}
\fmtSubentry(a)
\fmtSense{to come near, to come close, to approach}
\fmtExampleRL{Se\-mbe\-ra ani nku\-wee\-wee\-te\-re.}
\fmtExampleEN{Come close and I will whisper to you.}
\fmtSubentry(b)
\fmtSense{to receive holy communion}
\fmtExampleRL{Aba\-ka\-tu\-lu\-ki ba\-kwa\-ba ku\-se\-mbe\-ra.}
\fmtExampleEN{The Catholics are going to receive holy communion.}

\fmtLemma{se\-mee\-rwa}
\fmtCitation{(oku)\-se\-mee\-rwa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-me\-rei\-rwe}
\fmtSense{to be happy, to be glad, to be pleased}
\fmtExampleRL{Na\-se\-me\-rei\-rwe oku\-bo\-na omwa\-na wa\-nge nga amaa\-re emi\-so\-mo gywa\-mwe eya di\-gu\-li.}
\fmtExampleEN{I was pleased to see my daughter completing her graduate studies.}

\fmtLemma{se\-mee\-rya}
\fmtCitation{(oku)\-se\-mee\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-mei\-rye}
\fmtSense{to please, to make happy}
\fmtExampleRL{Ki\-sai oku\-se\-mee\-rya abee\-gi baa\-mu.}
\fmtExampleEN{It is good to please your students.}

\fmtLemma{se\-me\-ra}%
\fmtOne{1}
\fmtCitation{(oku)\-se\-me\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-mei\-re}
\fmtSubentry(a)
\fmtSense{to be tasty, to be delicious}
\fmtExampleRL{Eme\-re eni ese\-mei\-re.}
\fmtExampleEN{This food is tasty.}
\fmtSubentry(b)
\fmtSense{to be interesting}
\fmtExampleRL{Olu\-nya\-la lu\-se\-me\-ra.}
\fmtExampleEN{The Lunyala language is interesting.}
\fmtSubentry(c)
\fmtSense{to look smart, to be neat}
\fmtExampleRL{Mboi\-ne omu\-ka\-li ase\-mei\-re.}
\fmtExampleEN{I have seen a neatly dressed woman.}

\fmtLemma{se\-me\-za}%
\fmtOne{1}
\fmtCitation{(oku)\-se\-me\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-mei\-ze}
\fmtSense{to please, to make happy}
\fmtExampleRL{Olwe\-mbo olwo lu\-nse\-mei\-ze mu\-no.}
\fmtExampleEN{That song has really pleased me.}

\fmtLemma{se\-me\-za}%
\fmtOne{2}
\fmtCitation{(oku)\-se\-me\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-mei\-ze}
\fmtSense{to operate on, to perform a surgical operation on}
\fmtExampleRL{Odo\-ki\-ta aku\-se\-me\-za omu\-lwai\-re.}
\fmtExampleEN{The doctor is operating on the patient.}

\fmtLemma{se\-mi\-na}
\fmtCitation{(e)se\-mi\-na}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{seminar, workshop}
\fmtExampleRL{Tu\-kwa\-ba omu se\-mi\-na ya bya\-bwo\-mi.}
\fmtExampleEN{We are going to attend a seminar on health matters.}

\fmtLemma{se\-mi\-nti}
\fmtCitation{(o)se\-mi\-nti}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSource{from En.}
\fmtSense{cement}
\fmtExampleRL{Oso\-bo\-la okwa\-ba oku iduu\-ka n’o\-gu\-la\-yo ensa\-wo isaa\-tu eza se\-mi\-nti?}
\fmtExampleEN{Can you please go to the shop and buy three bags of cement?}

\fmtLemma{se\-mu\-tu\-ndu}
\fmtCrossreference{\textnormal{see also} mboli}
\fmtCitation{(o)se\-mu\-tu\-ndu}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{\textit{Bagrus docmak} (semutundu, Sudan catfish)}
\fmtExampleRL{Aba\-ntu bai\-ngi ba\-ta\-ka oku\-lya ose\-mu\-tu\-ndu.}
\fmtExampleEN{Many people like to eat semutundu fish.}

\fmtLemma{se\-na}
\fmtCitation{(oku)\-se\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-ne\-re}
\fmtSense{to scoop, to ladle out, to draw out}
\fmtExampleRL{Nse\-ne\-ra amai\-zi omu pi\-pa.}
\fmtExampleEN{Draw water for me from the drum.}

\fmtLemma{se\-nda}%
\fmtOne{1}
\fmtCitation{(oku)\-se\-nda}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-nde\-re}
\fmtSubentry(a)
\fmtSense{to push aside, to shove away}
\fmtExampleRL{Se\-nda ebi\-su\-bi bii\-re ku nta\-ndo.}
\fmtExampleEN{Push the rubbish aside.}
\fmtSubentry(b)
\fmtSense{to demolish, to take down, to tear down}
\fmtExampleRL{Ba\-se\-nde\-re enyu\-mba za ba\-ntu ba Bu\-go\-loo\-bi.}
\fmtExampleEN{They demolished people's houses in Bugoloobi.}

\fmtLemma{se\-nda}%
\fmtOne{2}
\fmtCitation{(oku)\-se\-nda}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-nde\-re}
\fmtSense{to grade (land)}
\fmtExampleRL{Ei\-ta\-ka\-li lya Na\-nta\-ba ba\-li\-se\-nde\-re.}
\fmtExampleEN{Nantaba's land was graded.}

\fmtLemma{se\-nda\-se\-nda}
\fmtCitation{(oku)\-se\-nda\-se\-nda}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-nde\-re\-se\-nde\-re}
\fmtSense{to entice, to persuade, to convince}
\fmtExampleRL{Ba\-mu\-se\-nde\-re\-se\-nde\-re n’a\-yi\-ngi\-ra omu ki\-saa\-we.}
\fmtExampleEN{He was persuaded to go to the field.}

\fmtLemma{se\-nde\-kee\-rya}%
\fmtOne{1}
\fmtCitation{(oku)\-se\-nde\-kee\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-nde\-ke\-rei\-rye}
\fmtSense{to escort, to accompany}
\fmtExampleRL{Nku\-se\-nde\-kee\-rya mu\-ge\-ni.}
\fmtExampleEN{I am escorting a visitor.}

\fmtLemma{se\-nde\-kee\-rya}%
\fmtOne{2}
\fmtCitation{(oku)\-se\-nde\-kee\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-nde\-ke\-rei\-rye}
\fmtSense{to throw away, to cast away}
\fmtExampleRL{Baa\-se\-nde\-ke\-rei\-rye ebi\-si\-ngo omu mu\-ge\-ra.}
\fmtExampleEN{They threw away the charms into the swamp.}

\fmtLemma{se\-nga}
\fmtCitation{(oku)\-se\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-nge\-re}
\fmtSubentry(a)
\fmtSense{to settle, to resettle, to occupy}
\fmtExampleRL{Aba\-ntu bai\-ngi ba\-se\-nge\-re omu bi\-tu\-ndu bya iya\-nga byo\-na\-byo\-na.}
\fmtExampleEN{Many people have settled almost in every region of Uganda.}
\fmtSubentry(b)
\fmtSense{(of a man) to move into a woman's house to start a marriage}
\fmtExampleRL{Omu\-sai\-za ya\-le\-ke\-re\-wo ama\-ka gaa\-mwe n’a\-se\-nga ewa mu\-ka\-li.}
\fmtExampleEN{The man abandoned his family and moved into another's woman's home.}
\fmtSubentry(c)
\fmtFixedexpression{se\-nga (o)mu\-ka\-ma}
\fmtSense{to turn to God}
\fmtExampleRL{Aba\-ntu bai\-ngi ne ba\-se\-nga oKa\-nca.}
\fmtExampleEN{Many people turned to God.}

\fmtLemma{se\-ngee\-re\-sya}
\fmtCitation{(oku)\-se\-ngee\-re\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-ngee\-re\-se\-rye}
\fmtSense{to take, to bring, to carry, to bear, to transport}
\fmtExampleRL{Se\-ngee\-re\-sya omwa\-na omaa\-ma waa\-mwe.}
\fmtExampleEN{Take the child to his mother.}

\fmtLemma{se\-ngee\-rya}
\fmtCitation{(oku)\-se\-ngee\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-nge\-rei\-rye}
\fmtSubentry(a)
\fmtSense{to follow, to come behind, to come after, to go behind, to go after, to walk behind}
\fmtExampleRL{Ombwe\-ne aku\-se\-ngee\-rya omu\-ka\-ma waa\-mwe.}
\fmtExampleEN{The dog is following its master.}
\fmtSubentry(b)
\fmtSense{to follow up, to keep track of, to pursue further, to investigate further}
\fmtExampleRL{Se\-ngee\-rya enso\-nga ezo.}
\fmtExampleEN{Follow up on those matters.}
\fmtSubentry(c)
\fmtSense{to go and find out, to go see}
\fmtExampleRL{Se\-ngee\-rya omwa\-na nka alwi\-re\-wo okwi\-ra.}
\fmtExampleEN{Go and find out why the child has delayed.}

\fmtLemma{se\-ngei\-za}
\fmtCitation{(oku)\-se\-ngei\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-ngei\-ze\-re}
\fmtSense{to strain, to filter, to sieve}
\fmtExampleRL{Ya\-se\-ngei\-ze\-re obu\-wu\-nga.}
\fmtExampleEN{She sieved the flour.}

\fmtLemma{se\-nge\-ka}
\fmtCitation{(oku)\-se\-nge\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-nge\-ke\-re}
\fmtSense{to align, to arrange, to put in order, to lay down, to present, to organise}
\fmtExampleRL{Se\-nge\-ka enso\-nga emwei ku emwei.}
\fmtExampleEN{Organise the points one by one.}

\fmtLemma{se\-nge\-nge}
\fmtCitation{(e)se\-nge\-nge}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{barbed wi\-re}
\fmtExampleRL{Bbwa ose\-nge\-nge oku ndu\-mi\-ro ebi\-so\-lo bi\-re\-me ku\-sii\-sa me\-re yaa\-mu.}
\fmtExampleEN{Put up barbed wire around the garden to prevent animals from eating your crops.}

\fmtLemma{se\-nge\-ree\-rya}
\fmtCitation{(oku)\-se\-nge\-ree\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-nge\-rei\-rye}
\fmtSense{to apologise to, to say sorry to, to make an apology to, to ask forgiveness}
\fmtExampleRL{Ya\-ya\-ti\-rye enda\-bi\-rwa\-mu ya\-nge bai\-tu ti\-ya\-nee\-se\-nge\-rei\-rye.}
\fmtExampleEN{He broke my mirror but did not apologise to me.}

\fmtLemma{se\-ngu\-ka}
\fmtCitation{(oku)\-se\-ngu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-ngu\-ki\-re}
\fmtSense{to migrate from, to relocate from, to resettle from, to move from}
\fmtExampleRL{Aba\-nya\-la baa\-se\-ngu\-ki\-re Bu\-nyo\-ro ne baa\-ba omu Bu\-nya\-la.}
\fmtExampleEN{The Banyala migrated from Bunyoro and went to Bunyala.}

\fmtLemma{se\-ngu\-la}%
\fmtOne{1}
\fmtCitation{(oku)\-se\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-ngwi\-re}
\fmtSense{to displace, to oust}
\fmtExampleRL{Aba\-guu\-da ti\-ba\-ta na nso\-ni oku\-se\-ngu\-la aboo\-ro.}
\fmtExampleEN{The rich don’t have remorse for displacing the poor.}

\fmtLemma{se\-ngu\-la}%
\fmtOne{2}
\fmtCitation{(oku)\-se\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-ngwi\-re}
\fmtSense{to limp}
\fmtExampleRL{Omu\-ka\-li oyo aku\-se\-ngu\-la ku\-bba mu\-lwai\-re wa ku\-gu\-lu.}
\fmtExampleEN{That woman is limping because her leg hurts.}

\fmtLemma{se\-nsa}
\fmtCitation{(oku)\-se\-nsa}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{se\-nse\-re}
\fmtSubentry(a)
\fmtSense{to pour}
\fmtExampleRL{Se\-nsa amai\-zi omu nzi\-ra le\-ke tu\-wo\-ne ocuu\-cu.}
\fmtExampleEN{Pour water onto the road to stop the dust. (lit.\,Pour water onto the road so that we are spared from the dust.)}
\fmtSubentry(b)
\fmtSense{to multiply, to increase , to spead}
\fmtExampleRL{Encwi zi\-se\-nsa Kwai\-se\-nsa\-ncwi.}
\fmtExampleEN{Fish multiplies in May.}

\fmtLemma{se\-nsaa\-si}
\fmtCitation{(e)se\-nsaa\-si}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{census}
\fmtExampleRL{Se\-nsaa\-si eya\-wei\-re ti\-ba\-mbaa\-re.}
\fmtExampleEN{I was not counted during the last census.}

\fmtLemma{se\-nse\-bu\-ka}
\fmtCitation{(oku)\-se\-nse\-bu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-nse\-bu\-ki\-re}
\fmtSense{to fall apart, to tear into pieces, to break into pieces}
\fmtExampleRL{Olu\-ga\-li lu\-se\-nse\-bu\-ki\-re.}
\fmtExampleEN{The winnower has broken into pieces.}

\fmtLemma{se\-nse\-bu\-la}
\fmtCitation{(oku)\-se\-nse\-bu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-nse\-bwi\-re}
\fmtSense{to tear into pieces, to break into pieces}
\fmtExampleRL{Omwo\-jo ase\-nse\-bwi\-re olu\-ga\-li.}
\fmtExampleEN{A boy has torn the winnower into pieces.}

\fmtLemma{se\-nse\-ka}
\fmtCitation{(oku)\-se\-nse\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-nse\-ke\-re}
\fmtSubentry(a)
\fmtSense{to be destroyed}
\fmtExampleRL{Ebbo\-mu ya\-ku\-bbi\-re ekyo\-mbo\-ko eki\-koo\-to ne ki\-se\-nse\-ka.}
\fmtExampleEN{A bomb exploded in a big building, and the building was destroyed.}
\fmtSubentry(b)
\fmtSense{to accept quickly, to agree quickly}
\fmtExampleRL{Omu\-ka\-li ni na\-mu\-ko\-be\-re\-ku n’a\-se\-nse\-ka.}
\fmtExampleEN{When I proposed to the woman, she accepts quickly.}
\fmtSubentry(c)
\fmtSense{to pour out, to be poured out, to spill, to overflow, to fall down, to collapse, to pour in (as a group of people), to fail, (of a plan) to come to naught}
\fmtExampleRL{Ama\-ca\-nde ga\-nge ga\-se\-nse\-ke\-re.}
\fmtExampleEN{My simsim has poured out.}

\fmtLemma{se\-nse\-ra}
\fmtCitation{(oku)\-se\-nse\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-nsei\-re}
\fmtSense{to slip in, to penetrate, to enter, to infiltrate, to take hold}
\fmtExampleRL{Aba\-te\-mu ba\-se\-nsei\-re omu ki\-tu\-ndu kya Ka\-mpa\-la.}
\fmtExampleEN{Murderers have penetrated Kampala area.}

\fmtLemma{se\-nsu\-ka}
\fmtCitation{(oku)\-se\-nsu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-nsu\-ki\-re}
\fmtSubentry(a)
\fmtSense{(of clothes, mats) to be tattered, to be old and torn, to be in poor condition}
\fmtExampleRL{Aba\-te\-gi ba\-zwai\-re engo\-ye ezi\-se\-nsu\-ki\-re.}
\fmtExampleEN{The fishermen are wearing tattered clothes.}
\fmtSubentry(b)
\fmtSense{(of a thatched roof) to lose thatching grass}
\fmtExampleRL{enyu\-mba yaa\-mu ense\-nsu\-ki\-re.}
\fmtExampleEN{The thatching grass on her house has fallen off.}

\fmtLemma{se\-nte}
\fmtCitation{(e)se\-nte}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{money}
\fmtExampleRL{Mbii\-ki\-ra ese\-nte za\-nge.}
\fmtExampleEN{Keep this money for me.}

\fmtLemma{se\-nte\-nsi}
\fmtCitation{(e)se\-nte\-nsi}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{sentence (a set of words)}
\fmtExampleRL{Ese\-nte\-nsi eni ti\-nje\-te\-ge\-rye ku\-sai.}
\fmtExampleEN{I have not understood this sentence well.}

\fmtLemma{se\-nya}
\fmtCitation{(oku)\-se\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-nye\-re}
\fmtSense{to brush (teeth)}
\fmtExampleRL{Oba lwa\-ki aba\-ntu aba\-mwei ti\-be\-ndya ku\-se\-nya mai\-no?}
\fmtExampleEN{I wonder why some people do not want to brush their teeth.}

\fmtLemma{se\-nyi\-ga}
\fmtCitation{(o)se\-nyi\-ga}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from Lug.} 
\fmtSubentry(a)
\fmtSense{flu}
\fmtSubentry(b)
\fmtFixedexpression{se\-nyiga (o)mwandi}
\fmtSense{COVID-19}
\fmtExampleRL{Ose\-nyi\-ga omwa\-ndi ayi\-ti\-re aba\-ntu bai\-ngi.}
\fmtExampleEN{COVID-19 has killed many people.}

\fmtLemma{se\-pee\-wo}
\fmtCitation{(e)se\-pee\-wo}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{hat}
\fmtExampleRL{Omu\-kai\-re atai\-re ese\-pee\-wo oku mu\-twe.}
\fmtExampleEN{The old man is wearing a hat.}

\fmtLemma{se\-pi\-ki}
\fmtCitation{(e)se\-pi\-ki}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{cooking pan, saucepan}
\fmtExampleRL{Ese\-pi\-ki eni gi\-kai\-re mu\-no.}
\fmtExampleEN{This saucepan is very old.}

\fmtLemma{Se\-ptua\-gi\-nta}
\fmtCitation{(e)Se\-ptua\-gi\-nta}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{Septuagint (the earliest extant Greek translation of the Hebrew scriptures from the original Hebrew, known as ``The Translation of the Seventy" as it is recorded that the Septuagint was translated by 70 or 72 Jewish scholars)}

\fmtLemma{se\-pu\-li\-ngi}
\fmtCitation{(e)se\-pu\-li\-ngi}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{spring, stiff suspension}
\fmtExampleRL{Ese\-pu\-li\-ngi ya mo\-to\-ka ya\-tii\-ki\-re.}
\fmtExampleEN{The car spring is broken.}

\fmtLemma{se\-ra}
\fmtCitation{(oku)\-se\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{see\-re}
\fmtSense{to practice cannibalism}
\fmtExampleRL{Omu\-sai\-za baa\-mu\-pe\-ke\-re oku kya\-lo lwa ku\-se\-ra.}
\fmtExampleEN{A man was chased from the village because of practicing cannibalism.}

\fmtLemma{se\-re\-be\-ra}
\fmtCitation{(oku)\-se\-re\-be\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-re\-bei\-re}
\fmtSubentry(a)
\fmtSense{(of paste, food) to be watery, to be thin}
\fmtExampleRL{Ya\-su\-mbi\-re omwo\-go ni gu\-se\-re\-be\-ra.}
\fmtExampleEN{She cooked cassava and it became watery.}
\fmtSubentry(b)
\fmtSense{to weaken, to be frail}
\fmtExampleRL{Omu\-lwai\-re ase\-re\-bei\-re.}
\fmtExampleEN{The patient has become frail.}

\fmtLemma{se\-re\-be\-ri}
\fmtCitation{\textnormal{(e.g.}\,gu\-se\-re\-be\-ri, mi\-se\-ri\-be\-ri, ka\-se\-re\-be\-ri\textnormal{)}}
\fmtPOS{adj}
\fmtSense{watery}
\fmtExampleRL{Ya\-su\-mbi\-re omwo\-go gu\-se\-re\-be\-ri.}
\fmtExampleEN{She cooked watery cassava.}

\fmtLemma{se\-re\-ge\-nya}
\fmtCitation{(oku)\-se\-re\-ge\-nya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-re\-ge\-nye\-re}
\fmtSense{to grow wings}
\fmtExampleRL{Obu\-ko\-ko bu\-ku\-se\-re\-ge\-nya.}
\fmtExampleEN{The chicks are growing wings.}

\fmtLemma{se\-re\-ge\-nyi}
\fmtPOS{adv}
\fmtSense{forever and ever}
\fmtExampleRL{OKa\-nca ali\-bbaa\-wo mi\-re\-mbe se\-re\-ge\-nyi.}
\fmtExampleEN{God will live forever and ever.}

\fmtLemma{se\-re\-ka}
\fmtCitation{(oku)\-se\-re\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-re\-ke\-re}
\fmtSubentry(a)
\fmtSense{to keep secretly, to hide, to tuck way}
\fmtExampleRL{Nse\-re\-ke\-ra ese\-nte za\-nge.}
\fmtExampleEN{Keep this money secretly for me.}
\fmtSubentry(b)
\fmtSense{(of a hen) to protect (chicks)}
\fmtExampleRL{Onko\-ko ase\-re\-ka obwa\-na.}
\fmtExampleEN{The hen sits on the chicks.}
\fmtSubentry(c)
\fmtSense{to thatch, to roof}
\fmtExampleRL{Aku\-se\-re\-ka nyu\-mba yaa\-mwe.}
\fmtExampleEN{He is roofing his house.}

\fmtLemma{se\-re\-nge\-ta}
\fmtCitation{(oku)\-se\-re\-nge\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-re\-nge\-te\-re}
\fmtSense{to descend, to go down, to come down, to slope downwards}
\fmtExampleRL{Omwo\-jo ngo\-yo aku\-se\-re\-nge\-ta ku izu\-ba.}
\fmtExampleEN{That boy is sloping downwards to the well.}

\fmtLemma{se\-ru\-ka}
\fmtCitation{(oku)\-se\-ru\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-ru\-ki\-re}
\fmtSense{to suffer from measles}
\fmtExampleRL{Omwa\-na ya\-se\-ru\-ki\-re.}
\fmtExampleEN{The child is suffering from measles.}

\fmtLemma{se\-ru\-mbee\-te}
\fmtCitation{(e)se\-ru\-mbee\-te}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSubentry(a)
\fmtSense{bass drum}
\fmtExampleRL{OJi\-mmy amai\-te oku\-ku\-bba ese\-ru\-mbee\-te.}
\fmtExampleEN{Jimmy knows how to play the bass drum.}
\fmtSubentry(b)
\fmtSense{brass band}
\fmtExampleRL{Abaa\-na ba iso\-me\-ro lya Ki\-se\-nyi PS baa\-ku\-bbi\-re ku\-sai ese\-ru\-mbee\-te.}
\fmtExampleEN{The pupils of Kisenyi Primary School played in the brass band well.}

\fmtLemma{se\-rya}
\fmtCitation{(oku)\-se\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{see\-rye}
\fmtSense{(of an eye) to twitch}
\fmtExampleRL{Ki\-ki eki\-ku\-nse\-rya ai\-gu\-ru wa rii\-so.}
\fmtExampleEN{What is twitching above my eye?}

\fmtLemma{se\-rya\-se\-rya}%
\fmtOne{1}
\fmtCitation{(oku)\-se\-rya\-se\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{see\-rye\-see\-rye}
\fmtSubentry(a)
\fmtSense{to flash}
\fmtExampleRL{Ei\-gu\-ru li\-ku\-se\-rya\-se\-rya olwa nku\-bba.}
\fmtExampleEN{The sky is flashing because of lightning.}
\fmtSubentry(b)
\fmtSense{to glitter, to gleam, to glisten}
\fmtExampleRL{Mboi\-ne onjo\-ka omu\-koo\-to ng’a\-ku\-se\-rya\-se\-rya.}
\fmtExampleEN{I have seen a big snake glittering.}

\fmtLemma{se\-se\-ma}
\fmtCitation{(oku)\-se\-se\-ma}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{se\-se\-me\-re}
\fmtSense{to vomit, to bring up, to trow up, to puke}
\fmtExampleRL{Omu\-lwai\-re wa mu\-su\-ja aku\-se\-se\-ma omu\-ba\-zi.}
\fmtExampleEN{The malaria patient is vomiting the medicine.}

\fmtLemma{se\-se\-ma}%
\fmtOne{2}
\fmtCitation{(oku)\-se\-se\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{se\-se\-me\-re}
\fmtSense{to return by applying force}
\fmtExampleRL{Ya\-se\-se\-me\-re ese\-nte ze ya\-nya\-gwi\-re.}
\fmtExampleEN{He returned the money he has stolen.}

\fmtLemma{se\-taa\-ni}
\fmtCitation{(o)se\-taa\-ni}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{Satan}
\fmtExampleRL{OSe\-taa\-ni enko\-la yaa\-mwe gi\-bbi.}
\fmtExampleEN{Satan’s ways are bad.}

\fmtLemma{se\-yei\-tu\-ka}%
\fmtOne{1}
\fmtCitation{(oku)\-se\-yei\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-yei\-tu\-ki\-re}
\fmtSense{to be unstable, to be disturbed}
\fmtExampleRL{Omwa\-la oku\-zwa ni ya\-zwi\-re omu bu\-fu\-mbo bwa\-mwe ase\-yei\-tu\-ka\-mbe.}
\fmtExampleEN{Ever since that girl left her marriage, she has been unstable.}

\fmtLemma{se\-yei\-tu\-ka}%
\fmtOne{2}
\fmtCitation{(oku)\-se\-yei\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{se\-yei\-tu\-ki\-re}
\fmtSense{to walk a long way}
\fmtExampleRL{Tu\-se\-yei\-tu\-ki\-re olu\-ge\-ndo lu\-koo\-to.}
\fmtExampleEN{We have walked a long way.}

\fmtLemma{Sha\-ri\-ya}
\fmtCitation{(e)Sha\-ri\-ya}
\fmtPOS{n}%
\fmtNC{9}
\fmtSource{from En.}
\fmtSense{sharia (Islamic canonical law)}
\fmtExampleRL{Aba\-si\-raa\-mu aba\-mwei ba\-ta\-mbuu\-ra ku Sha\-ri\-ya.}
\fmtExampleEN{Some Muslims follow the sharia.}

\fmtLemma{She\-ma}
\fmtCitation{(e)She\-ma}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{Shema (a prayer that serves as a centrepiece of the morning and evening Jewish prayer services)}
\fmtExampleRL{Aba\-yu\-daa\-ya ba\-so\-ma mu\-no eshe\-ma.}
\fmtExampleEN{The Jews say the Shema.}

\fmtLemma{Shi\-nto}
\fmtCitation{(e)Shi\-nto}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{Shinto (the traditional religion of Japan)}
\fmtExampleRL{Aba\-ja\-paa\-ni ba\-ta n’e\-dii\-ni ya shi\-nto.}
\fmtExampleEN{The Japanese have the Shinto religion.}

\fmtLemma{si\-ba}
\fmtCitation{(oku)\-si\-ba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-bi\-re}
\fmtSubentry(a)
\fmtSense{to tie, to bind}
\fmtExampleRL{Ensa\-wo ya su\-kaa\-li gi\-si\-be oku gaa\-li.}
\fmtExampleEN{Tie a bag of sugar on the bicycle.}
\fmtSubentry(b)
\fmtSense{to tether, to tie up (an animal)}
\fmtExampleRL{Si\-ba ombu\-li.}
\fmtExampleEN{Tether that goat.}

\fmtLemma{si\-bu\-ka}
\fmtCrossreference{\textnormal{see} subuka}

\fmtLemma{si\-ga}%
\fmtOne{1}
\fmtCitation{(oku)\-si\-ga}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{si\-gi\-re}
\fmtSubentry(a)
\fmtSense{to plant, to grow, to sow}
\fmtExampleRL{Aya\-bi\-re ku\-si\-ga bwi\-ta.}
\fmtExampleEN{He has gone to plant millet.}
\fmtSubentry(b)
\fmtSense{to invest, to put money into, to provide capital for, to fund, to back, to finance}
\fmtExampleRL{Aba mu\-si\-ga\-mpii\-ya ba\-si\-gi\-re obu\-kai\-re bwa doo\-la omu Uga\-nda.}
\fmtExampleEN{The investors have invested millions of dollars in Uganda.}
\fmtSubentry(c)
\fmtSense{to give an offering}
\fmtExampleRL{N’o\-si\-ga omu ka\-ni\-sa oKa\-nca aku\-wa mi\-ki\-sa.}
\fmtExampleEN{If you make church offerings, God will bless you.}

\fmtLemma{si\-ga}%
\fmtOne{2}
\fmtCitation{(oku)\-si\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-gi\-re}
\fmtSense{to push (a boat into deep water)}
\fmtExampleRL{Nku\-si\-ga lya\-to.}
\fmtExampleEN{I am pushing the boat into water.}

\fmtLemma{si\-ga\-la}%
\fmtOne{1}
\fmtCitation{(o)si\-ga\-la}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{cigarette}
\fmtExampleRL{Omu\-kuu\-mi w’a\-ni afuu\-wa osi\-ga\-la.}
\fmtExampleEN{The guard of this place smokes cigarette.}

\fmtLemma{si\-ga\-la}%
\fmtOne{2}
\fmtCitation{(oku)\-si\-ga\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-gai\-re}
\fmtSense{to remain, to stay, to stay behind, to be left}
\fmtExampleRL{Enkai\-na zoo\-na\-zoo\-na zaa\-bi\-re oni ye\-nkai ni\-ye asi\-gai\-re.}
\fmtExampleEN{All the camels have left and only this one remains.}

\fmtLemma{si\-ga\-lya}
\fmtCitation{(oku)\-si\-ga\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-gai\-rye}
\fmtSense{to keep, to retain, to keep possession of, to have}
\fmtExampleRL{Na\-tu\-nda embe\-re\-ge zoo\-naa, nsi\-ga\-lye\-ku omwei.}
\fmtExampleEN{I will sell all the pigs and will keep only one.}

\fmtLemma{si\-gi\-ka}
\fmtCitation{(oku)\-si\-gi\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-gi\-ki\-re}
\fmtSubentry(a)
\fmtSense{to fix (cooking stone), to arrange, to put up}
\fmtExampleRL{Omu\-ka\-li aku\-si\-gi\-ka amai\-ga.}
\fmtExampleEN{A woman is arranging the cooking stones.}
\fmtSubentry(b)
\fmtSense{to use (stones) to prevent (e.g.\,a pot) from tilting down, to prop up (with stones)}
\fmtExampleRL{Eki\-na\-ga ki\-si\-gi\-ke ei\-bbaa\-le.}
\fmtExampleEN{Use the stone to stop the pot from tilting.}
\fmtSubentry(c)
\fmtSense{to rely on, to fall back on}
\fmtExampleRL{Nsi\-gi\-kii\-re ku bi\-ga\-mbo bye wa\-nko\-be\-re ne nda\-mba abaa\-na ba\-ni.}
\fmtExampleEN{I have relied on what you told me to counsel these children.}

\fmtLemma{si\-gi\-ri}
\fmtCitation{(e)si\-gi\-ri}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{charcoal stove}
\fmtExampleRL{Esi\-gi\-ri eni gi\-koo\-to.}
\fmtExampleEN{This charcoal stove is big.}

\fmtLemma{si\-gu\-ka}
\fmtCitation{(oku)\-si\-gu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-gu\-ki\-re}
\fmtSense{(of a tree) to fall, to go down, to topple}
\fmtExampleRL{Omu\-saa\-le gu\-si\-gu\-ki\-re.}
\fmtExampleEN{The tree has fallen.}

\fmtLemma{si\-gu\-la}
\fmtCitation{(oku)\-si\-gu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-gwi\-re}
\fmtSense{to uproot, to root out}
\fmtExampleRL{Ompu\-ngo asi\-gwi\-re omu\-saa\-le.}
\fmtExampleEN{The wind has uprooted a tree.}

\fmtLemma{sii\-ba}
\fmtCitation{(oku)\-sii\-ba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sii\-bi\-re}
\fmtSubentry(a)
\fmtSense{to spend a day}
\fmtExampleRL{Ei\-zo ya\-sii\-bi\-re mpa\-ni.}
\fmtExampleEN{He spent the whole of yesterday here.}
\fmtSubentry(b)
\fmtSense{to go without food a whole day, to observe fasting during Ramadan}
\fmtExampleRL{OKa\-ju\-ra asii\-bi\-re nza\-la, eme\-re emu\-buu\-re.}
\fmtExampleEN{Kajura went a whole day without food because he failed to get food.}

\fmtLemma{sii\-bii\-ra}
\fmtCitation{(oku)\-sii\-bii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sii\-bii\-re}
\fmtSense{to eat for a day (one type of food)}
\fmtExampleRL{Mwo\-jo, ewai\-swe tu\-sii\-bii\-ra mu\-wo\-go.}
\fmtExampleEN{My dear, we have eaten only cassava for a day.}

\fmtLemma{sii\-bu\-la}
\fmtCitation{(oku)\-sii\-bu\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sii\-bwi\-re}
\fmtSense{to say goodbye, to say farewell, to send off}
\fmtExampleRL{Aya\-bi\-re ta\-sii\-bwi\-re.}
\fmtExampleEN{He has gone without saying goodbye.}

\fmtLemma{sii\-bu\-lu\-ku\-ka}
\fmtCitation{(oku)\-sii\-bu\-lu\-ku\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sii\-bu\-lu\-ku\-ki\-re}
\fmtSubentry(a)
\fmtSense{to break the fast, to eat after fasting}
\fmtExampleRL{Aba\-si\-raa\-mu ba\-sii\-bu\-lu\-ku\-ka saa\-wa emwei eya ka\-swe\-ko.}
\fmtExampleEN{The Muslims break the fast at 7:00 p.m.}
\fmtSubentry(b)
\fmtSense{to celebrate Eid}
\fmtExampleRL{Aba\-si\-raa\-mu ba\-kwa\-ba ku\-sii\-bu\-lu\-ku\-ka ma\-mbya.}
\fmtExampleEN{The Muslims will celebrate Eid tomorrow.}

\fmtLemma{sii\-ga}
\fmtCitation{(oku)\-sii\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sii\-gi\-re}
\fmtSubentry(a)
\fmtSense{to smear, to paint, to apply (e.g. makeup)}
\fmtExampleRL{Ei\-rai enyu\-mba nga ba\-zi\-sii\-ga\-nga bi\-sai.}
\fmtExampleEN{In the past houses were smeared with cow dung.}
\fmtSubentry(b)
\fmtSense{to transmit (a desease)}
\fmtExampleRL{Emba\-ra\-ba\-ra zi\-sii\-ga ebi\-so\-lo endwai\-re.}
\fmtExampleEN{Ticks transmit diseases in animals.}

\fmtLemma{sii\-ka}
\fmtCitation{(oku)\-sii\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sii\-ki\-re}
\fmtSense{to fry, to roast}
\fmtExampleRL{Omaa\-ma aku\-sii\-ka encwi.}
\fmtExampleEN{Mammy is frying fish.}

\fmtLemma{sii\-ma}
\fmtCitation{(oku)\-sii\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sii\-mi\-re}
\fmtSubentry(a)
\fmtSense{to appreciate, to thank, to be grateful for, to be pleased with}
\fmtExampleRL{Ya\-sii\-mi\-re bye\-na\-mu\-lee\-tei\-re.}
\fmtExampleEN{He appreciated what I gave him.}
\fmtSubentry(b)
\fmtSense{to fall in love with}
\fmtExampleRL{Omwo\-jo mmu\-sii\-mi\-re.}
\fmtExampleEN{I have fallen in love with the boy.}

\fmtLemma{sii\-moo\-za}
\fmtCitation{(oku)\-sii\-moo\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sii\-moo\-ze\-re}
\fmtSense{to make a sucking noise with one's lips as an expression of contempt towards, to suck one's teeth as expression of contempt towards, to practice teeth-sucking at}
\fmtExampleRL{Omwa\-la oni aku\-sii\-moo\-za abai\-ra baa\-mwe.}
\fmtExampleEN{This girl insults her pals making a sucking noise with the lips as an expression of contempt.}

\fmtLemma{sii\-muu\-ka}
\fmtCitation{(oku)\-sii\-muu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sii\-muu\-ki\-re}
\fmtSense{to be erased, to be deleted}
\fmtExampleRL{Ebi\-ga\-mbo bye ya\-wa\-ndii\-ki\-re omu ki\-ta\-bo bya\-sii\-mu\-ki\-re.}
\fmtExampleEN{The words that she wrote in the book have been erased.}

\fmtLemma{sii\-muu\-la}
\fmtCitation{(oku)\-sii\-muu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sii\-mwi\-re}
\fmtSubentry(a)
\fmtSense{to mop, to wipe, to clean}
\fmtExampleRL{Enyu\-mba abaa\-la ba\-ku\-gi\-sii\-muu\-la.}
\fmtExampleEN{The girls are mopping the house.}
\fmtSubentry(b)
\fmtSense{to erase, to delete}
\fmtExampleRL{Sii\-muu\-la eki\-wa\-ndii\-ko ekyo.}
\fmtExampleEN{Erase that document.}
\fmtSubentry(c)
\fmtSense{to clean (a baby), to wipe (a baby)}
\fmtExampleRL{Omwa\-na oyo mu\-sii\-muu\-le ebi\-mi\-ra.}
\fmtExampleEN{Clean the mucus from the child’s nose.}

\fmtLemma{sii\-na}
\fmtCitation{(o)sii\-na}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{dark brown cow}
\fmtExampleRL{Onte osii\-na abba n’e\-ra\-ngi eya ki\-ta\-ka omwi\-ti\-ki\-ri.}
\fmtExampleEN{Siina is a dark brown cow.}

\fmtLemma{sii\-nai}
\fmtCitation{(o)sii\-nai}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{\textit{Grewia mollis} (species of flowering plant)}
\fmtExampleRL{Osii\-nai ba\-mu\-lu\-ka\-mu ebi\-dee\-ro.}
\fmtExampleEN{\textit{Grewia mollis} is used for making baskets.}

\fmtLemma{sii\-nyoo\-la}
\fmtCitation{(oku)\-sii\-nyoo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sii\-nyoi\-re}
\fmtSense{to make a sucking noise with one's lips as an expression of contempt, to suck one's teeth as expression of contempt, to practice teeth-sucking}
\fmtExampleRL{Toi\-raa\-nga\-mu ku\-sii\-nyoo\-la ba\-ntu ba\-ku\-lu.}
\fmtExampleEN{Never again suck your teeth at elders.}

\fmtLemma{sii\-nyoo\-za}
\fmtCrossreference{\textnormal{see} siinyoola}

\fmtLemma{sii\-rii\-ra}
\fmtCitation{(oku)\-sii\-rii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sii\-rii\-re}
\fmtSense{to be burnt}
\fmtExampleRL{Eme\-re eku\-sii\-rii\-ra.}
\fmtExampleEN{The food is burnt.}

\fmtLemma{sii\-rii\-rya}
\fmtCitation{(oku)\-sii\-rii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sii\-rii\-rye}
\fmtSense{to burn, to cause to burn}
\fmtExampleRL{Omu\-su\-mbi asii\-rii\-rye eki\-na\-ga.}
\fmtExampleEN{The cook burnt the saucepan.}

\fmtLemma{sii\-sa}
\fmtCitation{(oku)\-sii\-sa}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{sii\-si\-re}
\fmtSubentry(a)
\fmtSense{to destroy, to spoil, to damage, to ruin}
\fmtExampleRL{Obu\-bu\-nguu\-ra bu\-sii\-sa eme\-re.}
\fmtExampleEN{Caterpillars destroy crops.}
\fmtSubentry(b)
\fmtSense{to defile, to rape}
\fmtExampleRL{Omwa\-la waa\-mwe ba\-mu\-sii\-si\-re.}
\fmtExampleEN{His girl was defiled.}
\fmtSubentry(c)
\fmtSense{(of a child) to defecate}
\fmtExampleRL{Omwa\-na asii\-si\-re ani.}
\fmtExampleEN{The child has defecated here.}
\fmtSubentry(d)
\fmtFixedexpression{sii\-sa (o)bwi\-re}
\fmtSense{to waste time}
\fmtExampleRL{Lwa\-ki oku\-sii\-sa obwi\-re?}
\fmtExampleEN{Why are you wasting your time?}

\fmtLemma{sii\-sii\-rya}%
\fmtOne{1}
\fmtCitation{(oku)\-sii\-sii\-rya}
\fmtPOS{v}%
\fmtTrans{ditr}
\fmtPFV{sii\-si\-rii\-rye}
\fmtSense{to apply (ointment), to smear with}
\fmtExampleRL{Sii\-sii\-rya omwa\-na oyo ama\-gi\-ta.}
\fmtExampleEN{Smear the child with vaseline.}

\fmtLemma{sii\-sii\-rya}%
\fmtOne{2}
\fmtCitation{(oku)\-sii\-sii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sii\-si\-rii\-rye}
\fmtSense{to prevent, to stop, to hinder, to impede}
\fmtExampleRL{Na\-bai\-re nkwa\-ba oku\-kwa\-ta omwi\-bbi bai\-tu oTo\-m n’a\-nsii\-sii\-rya.}
\fmtExampleEN{I was going to catch the thief, but Tom prevented me.}

\fmtLemma{sii\-si\-ka\-la}
\fmtCitation{(oku)\-sii\-si\-ka\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sii\-si\-kai\-re}
\fmtSubentry(a)
\fmtSense{to fall apart, to fall to pieces, to disintegrate, to decay, to be broken}
\fmtExampleRL{Obu\-wa\-ngwa bwa Bai\-ra\-ga\-zu bwa\-sii\-si\-kai\-re aba\-fu\-gi ba ma\-twa\-le ni bai\-zi\-re.}
\fmtExampleEN{The culture of Africans fell apart when Europeans colonised Africa.}
\fmtSubentry(b)
\fmtSense{(of food) to go bad, to spoil}
\fmtExampleRL{Eme\-re esii\-si\-kai\-re.}
\fmtExampleEN{The food has gone bad.}
\fmtSubentry(c)
\fmtSense{(of a vehicle) to break down}
\fmtExampleRL{Emo\-to\-ka yaa\-mwe esii\-si\-kai\-re.}
\fmtExampleEN{His car has broken down.}
\fmtSubentry(d)
\fmtSense{to be disorganised, to be disordered}
\fmtExampleRL{Ebi\-ntu bya\-nge bi\-sii\-si\-kai\-re.}
\fmtExampleEN{My things are disorganised.}

\fmtLemma{sii\-si\-ra}
\fmtCitation{(oku)\-sii\-si\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sii\-sii\-re}
\fmtSense{to stay for a long time, to stay for too long}
\fmtExampleRL{Mu\-ta\-sii\-si\-ra mu ki\-bu\-ga tu\-li\-na ebyo\-ku\-ko\-la bi\-ngi mu kya\-lo.}
\fmtExampleEN{Do not stay in the city for too long, we have many things to be done in the village.}

\fmtLemma{sii\-woo\-lya}
\fmtCitation{(oku)\-sii\-woo\-lya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sii\-woi\-rye}
\fmtSense{to break wind, to pass gas, to fart}
\fmtExampleRL{Omu\-ge\-ni asii\-woi\-rye omu ba\-ntu.}
\fmtExampleEN{A visitor has farted in public.}

\fmtLemma{sii\-wuu\-la}
\fmtCitation{(oku)\-sii\-wuu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sii\-wi\-re}
\fmtSense{to winnow, to separate (out), to sift out}
\fmtExampleRL{Ngo\-di aku\-sii\-wuu\-la mu\-we\-mba.}
\fmtExampleEN{He is there winnowing sorghum.}

\fmtLemma{sii\-ya}
\fmtCitation{(oku)\-sii\-ya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sii\-yi\-re}
\fmtSubentry(a)
\fmtSense{to hiss}
\fmtExampleRL{Onjo\-ka n’a\-sii\-ya oli\-na kwi\-ru\-ka.}
\fmtExampleEN{When the snake hisses, you must run.}
\fmtSubentry(b)
\fmtSense{to cause itching, to cause irritation}
\fmtExampleRL{Amai\-zi ago ga\-sii\-ya oku lu\-su\-su.}
\fmtExampleEN{That water irritates the skin.}

\fmtLemma{sii\-ya\-ni}
\fmtCitation{\textnormal{(e.g.}\,mu\-sii\-ya\-ni, ba\-sii\-yaa\-ni\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{horny, sexually aroused, lustful}
\fmtExampleRL{Omu\-sai\-za n’o\-mu\-ka\-li waa\-mwe ba\-sii\-ya\-ni.}
\fmtExampleEN{The man and his wife are horny.}
\fmtSubentry(b)
\fmtSense{adulterous, unfaithful, cheating}
\fmtExampleRL{Omu\-sai\-za oyo mu\-sii\-ya\-ni.}
\fmtExampleEN{That man is adulterous.}

\fmtLemma{sii\-ya\-sii\-ya\-na}
\fmtCitation{(oku)\-sii\-ya\-sii\-ya\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sii\-ya\-sii\-yai\-ne}
\fmtSense{(of a woman) to be horny, to be sexually aroused}
\fmtExampleRL{Omu\-ka\-li oyo aku\-bo\-ne\-ka aku\-sii\-ya\-sii\-ya\-na.}
\fmtExampleEN{That woman seems to be sexually aroused.}

\fmtLemma{si\-ka}%
\fmtOne{1}
\fmtCitation{(oku)\-si\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-ki\-re}
\fmtSense{to pull, to drag, to draw}
\fmtExampleRL{Si\-ka ente\-be edi ogi\-ree\-te ani.}
\fmtExampleEN{Pull that chair and bring it here.}

\fmtLemma{si\-ka}%
\fmtOne{2}
\fmtCitation{(oku)\-si\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-ki\-re}
\fmtSense{to inherit, to become heir to}
\fmtExampleRL{Oyo ni\-ye ya\-si\-kii\-re omaa\-ma.}
\fmtExampleEN{That one became an heir to my mother.}

\fmtLemma{si\-kaa\-ti}
\fmtCitation{(e)si\-kaa\-ti}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{skirt}
\fmtExampleRL{Omwa\-la azwai\-re esi\-kaa\-ti.}
\fmtExampleEN{The girl has worn a skirt.}

\fmtLemma{si\-kaa\-vu}
\fmtCitation{(e)si\-kaa\-vu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{scarf}
\fmtExampleRL{Aba\-ka\-li ba\-ta\-ka okwe\-bbwa esi\-kaa\-vu.}
\fmtExampleEN{Women like wearing a scarf.}

\fmtLemma{si\-ki\-na}
\fmtCitation{(oku)\-si\-ki\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-ki\-ni\-re}
\fmtSource{from taboo}
\fmtSense{to respond to sexual stimulation}

\fmtLemma{si\-kwe\-ya}
\fmtCitation{(e)si\-kwe\-ya}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{square}
\fmtExampleRL{Emee\-za eni ya si\-kwe\-ya.}
\fmtExampleEN{The table is square.}

\fmtLemma{si\-ma\-ntii\-rya}
\fmtCitation{(oku)\-si\-ma\-ntii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-ma\-nti\-rii\-rye}
\fmtSense{to plaster the wall (with soil)}
\fmtExampleRL{Nku\-si\-ma\-ntii\-rya nyu\-mba ya\-nge na ita\-ka\-li.}
\fmtExampleEN{I am plastering the walls of my house using soil.}

\fmtLemma{si\-mba}%
\fmtOne{1}
\fmtCitation{(oku)\-si\-mba}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{si\-mbi\-re}
\fmtSubentry(a)
\fmtSense{to go straight}
\fmtExampleRL{N’o\-bba okwa\-ba si\-mba\-mbe ota\-kya\-mu\-ka.}
\fmtExampleEN{Go straight, don’t branch off.}
\fmtSubentry(b)
\fmtSense{to put upright}
\fmtExampleRL{Lee\-ta engwa\-nyu nsi\-mbe eki\-ti\-mba.}
\fmtExampleEN{Bring the supporting sticks and I put upright the net.}

\fmtLemma{si\-mba}%
\fmtOne{2}
\fmtCitation{(oku)\-si\-mba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-mbi\-re}
\fmtSense{to plant, to sow}
\fmtExampleRL{Ya\-si\-mbi\-re ebi\-duu\-ma bai\-tu bya\-fii\-re ei\-sa\-na.}
\fmtExampleEN{She planted maize, but it didn’t grow because of the drought.}

\fmtLemma{si\-mba\-wo}
\fmtCitation{(oku)\-si\-mba\-wo}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-mbi\-re\-wo}
\fmtSense{to nominate, to designate, to recommend, to suggest}
\fmtExampleRL{Aba\-ntu aba\-mwe ba\-kwe\-buu\-lya oku ki\-ge\-ro kya\-mwe oku bya\-kwe\-si\-mba\-wo.}
\fmtExampleEN{Some people are questioning the appropriateness of his candidature.}

\fmtLemma{si\-mbu\-ka}
\fmtCitation{(oku)\-si\-mbu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-mbu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to begin a journey, to set off, to leave}
\fmtExampleRL{Si\-mbu\-ka ati\-bu\-ni oya\-be.}
\fmtExampleEN{Set off right now and go.}
\fmtSubentry(b)
\fmtSense{to start moving, to drive off (e.g.\,a vehicle)}
\fmtExampleRL{Emo\-to\-ka zi\-si\-mbu\-ki\-re.}
\fmtExampleEN{The cars have started moving.}

\fmtLemma{si\-mbu\-la}
\fmtCitation{(oku)\-si\-mbu\-la}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{si\-mbwi\-re}
\fmtSubentry(a)
\fmtSense{to begin a journey, to set off, to leave}
\fmtExampleRL{Si\-mbu\-la oya\-be.}
\fmtExampleEN{Set off and go.}
\fmtSubentry(b)
\fmtSense{to start moving, to drive off (e.g.\,a vehicle)}
\fmtExampleRL{Ebbaa\-si ni ya\-bbai\-re eri ampi oku\-si\-mbu\-la omu\-ntu owo\-ndi ya\-yi\-zi\-re ni ye\-ku\-kuu\-tii\-rya\-mu.}
\fmtExampleEN{Just as the bus was about to leave, another person squeezed in.}
\fmtSubentry(c)
\fmtSense{to uproot, to pull out of the ground}
\fmtExampleRL{Ombu\-li na\-mu\-bbwe\-re oku nko\-ndo n’a\-gi\-si\-mbu\-la.}
\fmtExampleEN{I tethered the goat on a stump, and it uprooted it.}
\fmtSubentry(d)
\fmtSense{to toddle, to totter}
\fmtExampleRL{Omwa\-na ata\-ndi\-ki\-re oku\-si\-mbu\-la.}
\fmtExampleEN{The baby has started to toddle.}

\fmtLemma{si\-mbu\-li\-za}
\fmtCitation{(oku)\-si\-mbu\-li\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-mbu\-lii\-ze}
\fmtSubentry(a)
\fmtSense{to transfer, to move, to shift, to move a tethered goat to browse in another sport}
\fmtExampleRL{Si\-mbu\-li\-za embu\-li oku saa\-wa iku\-mi.}
\fmtExampleEN{Move the tethered goat to browse in another sport at four o'clock.}
\fmtSubentry(b)
\fmtSense{to transplant, to replant, to repot}
\fmtExampleRL{Aku\-si\-mbu\-li\-za mwa\-ni zaa\-mwe.}
\fmtExampleEN{He is transplanting his coffee.}

\fmtLemma{si\-mi\-rya}
\fmtCitation{(oku)\-si\-mi\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-mii\-rye}
\fmtSense{to beckon to, to wave at, to gesture at}
\fmtExampleRL{Na\-boi\-ne omwo\-jo ng’a\-ku\-si\-mi\-rya omwa\-la.}
\fmtExampleEN{I saw a boy beckoning to a girl.}

\fmtLemma{si\-moo\-la}
\fmtCitation{(oku)\-si\-moo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-moi\-re}
\fmtSense{to say, to speak, to pronounce}
\fmtExampleRL{Si\-moo\-la nku\-ku\-wu\-lii\-rya.}
\fmtExampleEN{Speak, I am listening to you.}

\fmtLemma{si\-moo\-ni}
\fmtCitation{(e)si\-moo\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{coin of fifty cents}
\fmtExampleRL{Ti\-nsi\-gai\-rye go\-nze esi\-moo\-ni.}
\fmtExampleEN{I don’t have even a cent.}

\fmtLemma{si\-mu}
\fmtCitation{(e)si\-mu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{telephone, smartphone}
\fmtExampleRL{Nku\-ta\-ka oku\-gu\-la esi\-mu eya mu\-le\-mbe okwe\-zi ku\-ni.}
\fmtExampleEN{I need to buy a trendy smartphone this month.}

\fmtLemma{si\-mu\-kaa\-di}
\fmtCitation{(e)si\-mu\-kaa\-di}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{simcard}
\fmtExampleRL{Nkwa\-ba ku igo\-mbo\-lo\-la ku\-wa\-ndii\-ki\-sya si\-mu\-kaa\-di ya\-nge.}
\fmtExampleEN{I am going the sub-county headquarters to register my sim card.}

\fmtLemma{si\-mu\-la}
\fmtCitation{(oku)\-si\-mu\-la}
\fmtTrans{tr}
\fmtPFV{si\-mwi\-re}
\fmtSubentry(a)
\fmtSense{to shoot (a ball), to kick}
\fmtExampleRL{Ba\-ku\-si\-mu\-la\-nga\-na pe\-na\-ti.}
\fmtExampleEN{They are shooting a penalty kick.}
\fmtSubentry(b)
\fmtSense{to sneeze}

\fmtLemma{si\-na}
\fmtCitation{(oku)\-si\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-ni\-re}
\fmtSense{to nauseate, to disgust}
\fmtExampleRL{Mwa\-na we onsi\-ni\-re.}
\fmtExampleEN{Child, you disgust me.}

\fmtLemma{Si\-naa\-goo\-ga}
\fmtCitation{(e)Si\-naa\-goo\-ga}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{synagogue}
\fmtExampleRL{Aba\-yu\-daa\-ya ba\-ra\-mi\-rya mu si\-naa\-goo\-ga.}
\fmtExampleEN{The Jews pray in the synagogue.}

\fmtLemma{si\-nda\-bu\-ka}
\fmtCitation{(oku)\-si\-nda\-bu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-nda\-bu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to get loose, to become untied, to become unfastened}
\fmtExampleRL{Omu\-gu\-gu gwa\-nge gu\-si\-nda\-bu\-ki\-re.}
\fmtExampleEN{My luggage has become unfastened.}
\fmtSubentry(b)
\fmtSense{(of a bag) to open}
\fmtExampleRL{Kwa\-ta ku\-sai ensa\-wo ya\-nge eta\-si\-nda\-bu\-ka.}
\fmtExampleEN{Hold my bag carefully so that it does not open.}
\fmtSubentry(c)
\fmtSense{(of a stomach) to stop being constipated}
\fmtExampleRL{Eki\-da kya\-nge ki\-sa\-nda\-bu\-ki\-re.}
\fmtExampleEN{My stomach has stopped being constipated.}
\fmtSubentry(d)
\fmtSense{to give birth}
\fmtExampleRL{Omu\-ka\-li wa Mba\-ngi\-re akya\-li ku\-si\-nda\-bu\-ka.}
\fmtExampleEN{Mbangire's wife has not yet given birth.}

\fmtLemma{si\-nda\-bu\-la}
\fmtCitation{(oku)\-si\-nda\-bu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-nda\-bwi\-re}
\fmtSubentry(a)
\fmtSense{to remove, to detach, to take off}
\fmtExampleRL{Si\-nda\-bu\-la omu\-pii\-ra gwa mo\-to\-ka ba\-gu\-ta\-mbe.}
\fmtExampleEN{Remove the tyre from the car to get it mended.}
\fmtSubentry(b)
\fmtSense{to untie, to unleash}
\fmtExampleRL{Si\-nda\-bu\-la ombu\-li wa\-nge.}
\fmtExampleEN{Untie my goat.}
\fmtSubentry(c)
\fmtSense{to unlock}
\fmtExampleRL{Si\-nda\-bu\-la eku\-fu\-lu oku lwi\-ge.}
\fmtExampleEN{Unlock the padlock on the door.}
\fmtSubentry(d)
\fmtSense{to unpack, to open (a parcel, a bag)}
\fmtExampleRL{Si\-nda\-bu\-la ensa\-wu yaa\-mu tu\-bo\-ne ebi\-ri\-mu.}
\fmtExampleEN{Open your bag and let's see what is in it.}

\fmtLemma{si\-ndi\-ka}
\fmtCitation{(oku)\-si\-ndi\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-ndi\-ki\-re}
\fmtSubentry(a)
\fmtSense{to push}
\fmtExampleRL{Si\-ndi\-ka olwi\-ge oyi\-ngi\-re.}
\fmtExampleEN{Push the door and enter.}
\fmtSubentry(b)
\fmtSense{to send}
\fmtExampleRL{Omu\-ka\-ma ba\-mu\-si\-ndi\-ki\-re obu\-kwe\-nda.}
\fmtExampleEN{They have sent a message to the king.}
\fmtSubentry(c)
\fmtSense{to deploy, to post, to position}
\fmtExampleRL{Si\-ndi\-ka aba\-si\-ri\-ka\-le ba\-mu\-kwa\-te.}
\fmtExampleEN{Deploy soldiers to arrest her.}
\fmtSubentry(d)
\fmtSense{to transfer}
\fmtExampleRL{Omu\-lwai\-re abbai\-re ani ba\-mu\-si\-ndi\-ki\-re Ki\-ru\-ddu.}
\fmtExampleEN{The patient who was here has been transferred to Kiruddu.}
\fmtSubentry(e)
\fmtSense{to notify, to inform}
\fmtExampleRL{Mu\-si\-ndi\-ki\-re obu\-kwe\-nda ye\-te\-gye.}
\fmtExampleEN{Notify him so that he gets prepared.}

\fmtLemma{si\-ndyo\-ka}
\fmtCitation{(oku)\-si\-ndyo\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-ndyo\-ke\-re}
\fmtSense{to spin}
\fmtExampleRL{Ekyo\-ma ekyo\-za engo\-ye ti\-ki\-si\-ndyo\-ka ku\-sai.}
\fmtExampleEN{The washing machine does not spin properly.}

\fmtLemma{si\-nga}%
\fmtOne{1}
\fmtPOS{conj}
\fmtSense{if, when, while}
\fmtExampleRL{Nkwi\-za ku\-ku\-bya\-lu\-ku\-ka si\-nga ogaa\-na oku\-so\-ma.}
\fmtExampleEN{I will disown you if you refuse to study.}

\fmtLemma{si\-nga}%
\fmtOne{2}
\fmtCitation{(oku)\-si\-nga}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{si\-ngi\-re}
\fmtSubentry(a)
\fmtSense{to win (e.g.\,a war, a game, a prize)}
\fmtExampleRL{Ai\-da\-h asi\-ngi\-re omu\-sa\-ngo.}
\fmtExampleEN{Aidah has won the case.}
\fmtSubentry(b)
\fmtSense{to be better than, to be more than, to defeat}
\fmtExampleRL{OKa\-to ni\-ye asi\-nga bai\-ra baa\-mwe.}
\fmtExampleEN{Kato is better than his pals.}
\fmtSubentry(c)
\fmtSense{to be more, to be exceedingly, to be mostly}
\fmtExampleRL{Abee\-ka egyo aba\-ku\-si\-nga ba\-ta n’a\-ka\-wa\-ngo.}
\fmtExampleEN{Most members of that family have a mental disorder.}
\fmtSubentry(d)
\fmtSense{to excel, to be excellent, to be outstanding}
\fmtExampleRL{Mu\-taa\-ni wa\-nge ni\-ye ya\-si\-ngi\-re mu ke\-mi\-si\-tu\-le.}
\fmtExampleEN{My son excelled in chemistry.}

\fmtLemma{si\-nga}%
\fmtOne{3}
\fmtCitation{(oku)\-si\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-ngi\-re}
\fmtSense{to pawn, to put in pawn, to give as security}
\fmtExampleRL{OMu\-ka\-sa ya\-koo\-re ku\-bbi oku\-si\-nga enyu\-mba yaa\-mwe.}
\fmtExampleEN{Mukasa did bad to pawn his house.}

\fmtLemma{si\-ngaa\-ni\-sya}
\fmtCitation{(oku)\-si\-ngaa\-ni\-sya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-ngaa\-ni\-si\-rye}
\fmtSense{to compete, to participate, to take part (in a contest)}
\fmtExampleRL{Ba\-ku\-si\-ngaa\-ni\-sya mu mi\-si\-nde.}
\fmtExampleEN{They are competing in athletics.}

\fmtLemma{si\-nga\-nga\-na}
\fmtCitation{(oku)\-si\-nga\-nga\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-nga\-ngai\-ne}
\fmtSense{to compete}
\fmtExampleRL{ONa\-ka\-to n’o\-Ba\-bi\-rye ba\-si\-nga\-nga\-na mu kwe\-mba.}
\fmtExampleEN{Nakato and Babirye are competing in singing.}

\fmtLemma{si\-ngi\-sya}
\fmtCitation{(oku)\-si\-ngi\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-ngi\-si\-rye}
\fmtSense{to rule against, to convict}
\fmtExampleRL{Omu\-la\-mu\-li amu\-si\-ngi\-si\-rye omu\-sa\-ngo.}
\fmtExampleEN{The judge has ruled against him.}

\fmtLemma{si\-ngu\-la}
\fmtCitation{(oku)\-si\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-ngwi\-re}
\fmtSense{to win (in a lottery)}
\fmtExampleRL{Omu\-sai\-za ya\-zee\-nyi\-re aka\-lu\-lu n’a\-nsi\-ngu\-la emo\-to\-ka.}
\fmtExampleEN{A man played a lottery game and won a car.}

\fmtLemma{si\-ni\-ka}
\fmtCitation{(oku)\-si\-ni\-ka amai\-no}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-ni\-ki\-re}
\fmtSense{to bare (one's teeth as an insult)}
\fmtExampleRL{Le\-ka ku\-si\-ni\-ka amai\-no.}
\fmtExampleEN{Don't bare your teeth.}

\fmtLemma{si\-ni\-nu\-kwa}
\fmtCitation{(oku)\-si\-ni\-nu\-kwa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-ni\-nu\-ki\-rwe}
\fmtSense{to grieve, to be in mourning, to lament}
\fmtExampleRL{OPe\-te\-r aku\-si\-ni\-nu\-kwa olwa ku\-fa kwa ita waa\-mwe.}
\fmtExampleEN{Peter is grieving because of his father's death.}

\fmtLemma{si\-no\-di}
\fmtCitation{(e)si\-no\-di}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{synod}
\fmtExampleRL{OBa\-tu\-mbya ba\-mu\-ko\-me\-rye oku\-bba omu\-kwe\-nda oku si\-no\-di.}
\fmtExampleEN{Batumbya was chosen to be a representative on the synod.}

\fmtLemma{si\-nsi\-mu\-ka}
\fmtCitation{(oku)\-si\-nsi\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-nsi\-mu\-ki\-re}
\fmtSense{to be shredded, to be torn}
\fmtExampleRL{Olu\-go\-ye lwa\-nge lu\-si\-nsi\-mu\-ki\-re.}
\fmtExampleEN{My dress has become shredded.}

\fmtLemma{si\-nsi\-mu\-la}
\fmtCitation{(oku)\-si\-nsi\-mu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-nsi\-mwi\-re}
\fmtSubentry(a)
\fmtSense{to beat, to slap (severely)}
\fmtExampleRL{Nkwa\-ba oku\-ku\-si\-nsi\-mu\-la olu\-yi.}
\fmtExampleEN{I will slap you!}
\fmtSubentry(b)
\fmtSense{to shake out (a piece of clothing after rinsing)}
\fmtExampleRL{Si\-nsi\-mu\-la olu\-go\-ye lwa\-nge olwo lu\-ka\-le ma\-ngu.}
\fmtExampleEN{Shake out my dress so that it dries quickly.}

\fmtLemma{si\-paa\-ka}
\fmtCitation{(e)si\-paa\-ka}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{(spark) magnetic fishing hook}
\fmtExampleRL{Empuu\-ta ba\-zi\-te\-ge\-sya si\-paa\-ka.}
\fmtExampleEN{They use the magnetic fishing hook to fish the Nile perch.}

\fmtLemma{si\-pa\-nda}
\fmtCitation{(e)si\-pa\-nda}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSubentry(a)
\fmtSense{catapult, slingshot}
\fmtExampleRL{Omu\-sai\-za aku\-bbi\-re onko\-wu n’e\-si\-pa\-na\-da.}
\fmtExampleEN{The man has hit a guinea fowl using a catapult.}
\fmtSubentry(b)
\fmtSense{spanner, wrench}
\fmtExampleRL{Mpee\-re\-za esi\-pa\-nda le\-ke ngu\-mye empe\-ta oku gaa\-li ya\-nge.}
\fmtExampleEN{Give me a spanner so that I can tighten my bicycle's nuts.}

\fmtLemma{si\-pee\-ya}
\fmtCitation{(o)si\-pee\-ya}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{spare parts}
\fmtExampleRL{Esi\-mu ya\-nge te\-ta na si\-pee\-ya.}
\fmtExampleEN{My phone has no spare parts.}

\fmtLemma{si\-ra}
\fmtCitation{(oku)\-si\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sii\-re}
\fmtSense{to dissolve by rubbing (a medicinal clay stick)}
\fmtExampleRL{Si\-ra emu\-mbwa onywe\-ku.}
\fmtExampleEN{Dissolve by rubbing a medicinal clay stick and drink it.}

\fmtLemma{si\-raa\-na}
\fmtCitation{(oku)\-si\-raa\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-rai\-ne}
\fmtSense{to be cursed, to be bewitched, to be unlucky}
\fmtExampleRL{Omu\-ka\-li oyo ya\-si\-rai\-ne.}
\fmtExampleEN{That woman was cursed.}

\fmtLemma{si\-ra\-mu\-ka}
\fmtCitation{(oku)\-si\-ra\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-ra\-mu\-ki\-re}
\fmtSense{to adopt Islam as a religious faith, to convert to Islam}
\fmtExampleRL{Omu\-sai\-za asi\-ra\-mu\-ki\-re.}
\fmtExampleEN{A man has converted to Islam.}

\fmtLemma{si\-ra\-mu\-la}
\fmtCitation{(oku)\-si\-ra\-mu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-ra\-mwi\-re}
\fmtSubentry(a)
\fmtSense{to circumcise (as a sign of converting to Islam)}
\fmtExampleRL{Omwo\-jo oyo baa\-mu\-si\-ra\-mwi\-re.}
\fmtExampleEN{That boy was circumcised.}
\fmtSubentry(b)
\fmtSense{to convert to Islam, to persuade to change one's faith to Islam}
\fmtExampleRL{Omwo\-jo oni baa\-mu\-si\-ra\-mwi\-re.}
\fmtExampleEN{This young man was converted to Islam.}

\fmtLemma{si\-re}
\fmtCitation{\textnormal{(e.g.}\,mu\-si\-re, ki\-se\-re, gi\-si\-re\textnormal{)}}
\fmtPOS{adj}
\fmtSense{dirty}
\fmtExampleRL{Bu\-lii\-zo nta\-nga\-la\-na omu\-sai\-za omu\-si\-re oku ka\-ta\-le.}
\fmtExampleEN{I always meet a dirty man at the market.}

\fmtLemma{si\-ri\-bbu\-ka}
\fmtCitation{(oku)\-si\-ri\-bbu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-ri\-bbu\-ki\-re}
\fmtSense{to run very fast, to rush, to sprint, to bolt}
\fmtExampleRL{Omwo\-jo oyo asi\-ri\-bbu\-ka ki\-ta\-lo.}
\fmtExampleEN{That boy runs very fast.}

\fmtLemma{si\-rii\-kii\-ra}
\fmtCitation{(oku)\-si\-rii\-kii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-rii\-ki\-rii\-re}
\fmtSubentry(a)
\fmtSense{to be silent, to be quiet}
\fmtExampleRL{Ali\-mu omu nyu\-mba asi\-rii\-ki\-rii\-re.}
\fmtExampleEN{He is quiet in the house.}
\fmtSubentry(b)
\fmtSense{to calm down}
\fmtExampleRL{Bai\-ra\-ba, mu\-si\-rii\-kii\-re!}
\fmtExampleEN{My dear friends, calm down!}
\fmtSubentry(c)
\fmtSense{to contemplate, to reflect on, to meditate on}
\fmtExampleRL{Omu\-ka\-li asi\-rii\-ki\-rii\-re.}
\fmtExampleEN{A woman is contemplating.}

\fmtLemma{si\-rii\-kii\-rya}
\fmtCitation{(oku)\-si\-rii\-kii\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-rii\-ki\-rii\-rye}
\fmtSense{to cease (fire, fighting, noise), to pause, to stop for a while}
\fmtExampleRL{Ba\-si\-rii\-ki\-rii\-rye\-mu embu\-ndu.}
\fmtExampleEN{They have ceased fighting.}

\fmtLemma{si\-rii\-mu}
\fmtCitation{(o)si\-rii\-mu}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{HIV/AIDS}
\fmtExampleRL{Ki\-sai okwe\-ke\-be\-re\-sya osi\-rii\-mu.}
\fmtExampleEN{It is good to get tested for HIV/AIDS.}

\fmtLemma{si\-ri\-ka}
\fmtCitation{(oku)\-si\-ri\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-ri\-ki\-re}
\fmtSense{to keep quite}
\fmtExampleRL{Si\-ri\-ka ota\-ba\-za ki\-ga\-mbo.}
\fmtExampleEN{Keep quiet and don’t say a word.}

\fmtLemma{si\-ri\-ki}%
\fmtOne{1}
\fmtCitation{\textnormal{(e.g.}\,mu\-si\-ri\-ki, ba\-si\-ri\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{quiet}
\fmtExampleRL{Eki\-tu\-ndu ki\-ni ki\-si\-ri\-ki.}
\fmtExampleEN{This place is quiet.}
\fmtSubentry(b)
\fmtSense{calm}
\fmtExampleRL{Omu\-ka\-li wa\-nge ata mu\-si\-ri\-ki.}
\fmtExampleEN{My wife is calm.}

\fmtLemma{si\-ri\-ki}%
\fmtOne{2}
\fmtCitation{(o)si\-ri\-ki}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{silk, type of silk cloth}
\fmtExampleRL{Omu\-ka\-li azwai\-re osi\-ri\-ki.}
\fmtExampleEN{The woman is wearing a silk cloth.}

\fmtLemma{si\-ri\-ki}%
\fmtOne{3}
\fmtCitation{(o)si\-ri\-ki}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{shirk (the sin of idolatry or polytheism in Islam)}
\fmtExampleRL{Obu\-si\-raa\-mu ti\-bwi\-ki\-ri\-za si\-ri\-ki.}
\fmtExampleEN{Islam forbids shirk.}

\fmtLemma{si\-ri\-ki\-tu}
\fmtCitation{\textnormal{(e.g.}\,ka\-si\-ri\-ki\-tu, bu\-si\-ki\-ki\-tu\textnormal{)}}
\fmtPOS{adj}
\fmtSense{tiny, smal, minute}
\fmtExampleRL{Obu\-sii\-sa obwo bu\-bba bu\-si\-ri\-ki\-tu.}
\fmtExampleEN{Those bacteria are tiny.}

\fmtLemma{si\-ri\-kya}
\fmtCitation{(oku)\-si\-ri\-kya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-ri\-ki\-rye}
\fmtSubentry(a)
\fmtSense{to hush, to silence, to quiet down}
\fmtExampleRL{Si\-ri\-kya omwa\-na oyo!}
\fmtExampleEN{Hush that child!}
\fmtSubentry(b)
\fmtSense{to stop playing (a musical instrument)}
\fmtSubentry(c)
\fmtSense{to trap (a ball in footbal), to bring (the ball) under control}
\fmtExampleRL{OMu\-ka\-sa amai\-te mu\-no oku\-si\-ri\-kya omu\-pii\-ra.}
\fmtExampleEN{Mukasa knows well how to trap a ball.}

\fmtLemma{si\-ri\-mu\-ka}
\fmtCitation{(oku)\-si\-ri\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-ri\-mu\-ki\-re}
\fmtSense{to descend, to go down, to come down, to slope downwards}
\fmtExampleRL{Ende\-ge eta\-ndi\-ki\-re oku\-si\-ri\-mu\-ka.}
\fmtExampleEN{The plane has started descending.}

\fmtLemma{si\-ri\-ngi}
\fmtCitation{(e)si\-ri\-ngi}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{shilling}
\fmtExampleRL{Na\-guu\-re ama\-too\-ke iku\-mi ku si\-ri\-ngi lu\-ku\-mi.}
\fmtExampleEN{I bought ten bananas for one thousand shilling.}

\fmtLemma{si\-ru}
\fmtCitation{\textnormal{(e.g.}\,mu\-si\-ru, ba\-si\-ru\textnormal{)}}
\fmtPOS{adj}
\fmtSense{stupid, foolish}
\fmtExampleRL{Omu\-ntu omu\-si\-ru ta\-wu\-lii\-rya aba\-ku\-mu\-la\-mba.}
\fmtExampleEN{A stupid person does not listen to the ones who give him advice.}

\fmtLemma{si\-sii\-tu\-ka}
\fmtCitation{(oku)\-si\-sii\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-sii\-tu\-ki\-re}
\fmtSense{to stumble, to trip}
\fmtExampleRL{Omu\-sai\-za asi\-sii\-tu\-ki\-re n’a\-gwa.}
\fmtExampleEN{A man stumbled and fell.}

\fmtLemma{si\-si\-mu\-ka}
\fmtCitation{(oku)\-si\-si\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-si\-mu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to awaken, to wake up}
\fmtSubentry(b)
\fmtSense{fig. to realize}

\fmtLemma{si\-si\-mu\-la}
\fmtCitation{(oku)\-si\-si\-mu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-si\-mwi\-re}
\fmtSense{to awaken, to rouse from sleep}
\fmtExampleRL{Akai\-gu\-lu ka\-nsi\-si\-mwi\-re ni ntii\-na mu\-no.}
\fmtExampleEN{The thunder awakened me from sleep and felt very scared.}

\fmtLemma{si\-si\-na\-la}
\fmtCitation{(oku)\-si\-si\-na\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-si\-nai\-re}
\fmtSubentry(a)
\fmtSense{to have protruding teeth}
\fmtExampleRL{Eki\-sai\-za ki\-ta n’a\-gai\-no agaa\-si\-si\-nai\-re.}
\fmtExampleEN{That man has protruding teeth.}
\fmtSubentry(b)
\fmtSense{to bare (one's teeth), to show (one's teeth)}
\fmtSubentry(c)
\fmtSense{to grimace (in disgust), to show revulsion}
\fmtExampleRL{Omu\-ka\-li aboi\-ne eki\-si\-re n’a\-si\-si\-na\-la.}
\fmtExampleEN{A woman saw a dirty thing and grimaced.}

\fmtLemma{si\-si\-na\-lya}
\fmtCitation{(oku)\-si\-si\-na\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-si\-nai\-rye}
\fmtSource{from abusive}
\fmtSense{to mock by imitating a protrusion of teeth}
\fmtExampleRL{Omwo\-jo asi\-si\-na\-lya mwi\-raa\-mwe.}
\fmtExampleEN{The boy mocks his pal for having protruding teeth.}

\fmtLemma{si\-si\-ra}
\fmtCitation{(oku)\-si\-si\-ra}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{si\-sii\-re}
\fmtSubentry(a)
\fmtSense{to tremble all over, to shiver}
\fmtExampleRL{Lwa\-ki omwa\-na oyo aku\-si\-si\-ra?}
\fmtExampleEN{Why is that child trembling all over?}
\fmtSubentry(b)
\fmtSense{to have a creepy feeling, to have a tingling sensation, to be startled, to be shocked, to be taken aback, to experience revulsion}
\fmtExampleRL{Mboi\-ne ama\-bbi ni nsi\-si\-ra.}
\fmtExampleEN{When I saw the faeces, I got a creepy feeling.}
\fmtSubentry(c)
\fmtSense{to apply (powder), to powder, to cover (with powder)}
\fmtExampleRL{Ni wai\-ka eka osi\-si\-re omu\-ba\-zi oku kyo\-ya.}
\fmtExampleEN{When you get home, powder the medicine on the wound.}

\fmtLemma{si\-si\-rya}
\fmtCitation{(oku)\-si\-si\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-si\-rii\-rye}
\fmtSense{to repulse, to revolt, to disgust, to sicken, to nauseate}
\fmtExampleRL{Ekyo\-loo\-ni eki\-si\-re ki\-si\-si\-rya.}
\fmtExampleEN{A dirty latrine is disgusting.}

\fmtLemma{si\-ta}
\fmtCitation{(oku)\-si\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-ti\-re}
\fmtSense{to make (a traditional bed), to spread (a mat)}
\fmtExampleRL{Mmaa\-re oku\-si\-ta eki\-ta\-bu kya\-nge.}
\fmtExampleEN{I have finished making my bed.}

\fmtLemma{Si\-taa\-ni}
\fmtCitation{(o)si\-taa\-ni}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{Satan}
\fmtExampleRL{OSi\-taa\-ni ni\-ye ya\-lee\-te\-re eki\-bbi omu nsi mu\-ni.}
\fmtExampleEN{It is Satan who brought sin in this world.}

\fmtLemma{si\-ta\-ka}
\fmtCitation{(oku)\-si\-ta\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-ta\-ki\-re}
\fmtSense{to squat (unlawfully occupy a plot of land)}
\fmtExampleRL{Aba\-nya\-la ba\-li mu bu\-le\-mee\-zi bwa ku\-si\-ta\-ka oku ita\-ka\-li lya\-bwe.}
\fmtExampleEN{The Banyala are in trouble of becoming squatters on their own land.}

\fmtLemma{si\-ta\-mpu}
\fmtCitation{(e)si\-ta\-mpu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{stamp, seal}
\fmtExampleRL{Tu\-koo\-re esi\-ta\-mpu ya iso\-me\-ro lyai\-swe.}
\fmtExampleEN{Make for us a school stamp.}

\fmtLemma{si\-ta\-pu\-li}
\fmtCitation{(e)si\-ta\-pu\-li}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{door latch}
\fmtExampleRL{Esi\-ta\-pu\-li ni\-zo zi\-gu\-mya olwi\-ge.}
\fmtExampleEN{Door latches make a door strong.}

\fmtLemma{si\-ta\-yi\-ro}
\fmtCitation{(e)si\-ta\-yi\-ro}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{style, way}
\fmtExampleRL{Esi\-ta\-yi\-ro ya iso\-ke lya\-mwe gi\-sai.}
\fmtExampleEN{Her hairstyle is good.}

\fmtLemma{si\-tee\-gi}
\fmtCitation{(e)si\-tee\-gi}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSubentry(a)
\fmtSense{stage}
\fmtExampleRL{Abe\-mbi ba\-nii\-ni\-re ku si\-tee\-gi.}
\fmtExampleEN{The singers have gone on stage.}
\fmtSubentry(b)
\fmtSense{station, stopping place, stop}
\fmtExampleRL{Ebbaa\-si ngi\-nii\-ni\-ra ku si\-tee\-gi bu\-li nkya.}
\fmtExampleEN{I board a bus at the station every morning.}

\fmtLemma{si\-tee\-ri\-ngi}
\fmtCitation{(e)si\-tee\-ri\-ngi}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{steering wheel}
\fmtExampleRL{Etu\-ra\-ki\-ta yoo\-na ebba n’e\-si\-tee\-ri\-ngi.}
\fmtExampleEN{A tractor also has a steering wheel.}

\fmtLemma{si\-te\-nse\-ni}
\fmtCitation{(e)si\-te\-nse\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSubentry(a)
\fmtSense{junction, intersection}
\fmtExampleRL{Na\-mu\-ta\-nga\-rai\-ne mu si\-te\-nse\-ni.}
\fmtExampleEN{I met him at the junction.}
\fmtSubentry(b)
\fmtFixedexpression{si\-te\-nse\-ni ya gaa\-li\-moo\-shi}
\fmtSense{railway station}
\fmtExampleRL{Na\-mwa\-jii\-rye oku si\-te\-nse\-ni ya gaa\-li\-moo\-shi.}
\fmtExampleEN{I met her at the railway station.}

\fmtLemma{si\-ti\-gi\-ma\-ta}
\fmtCitation{(e)si\-ti\-gi\-ma\-ta}
\fmtPOS{n}%
\fmtNC{10}
\fmtSource{from En.}
\fmtSense{stigmata (marks corresponding to those left on Jesus’ body by the Crucifixion)}
\fmtExampleRL{Abo\-ne\-kei\-bwe ekya\-maa\-no ekya si\-ti\-gi\-ma\-ta.}
\fmtExampleEN{She had a vision of stigmata.}

\fmtLemma{si\-ti\-ra}
\fmtCitation{(oku)\-si\-ti\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-tii\-re}
\fmtSense{to erect (a fence), to barricade, to construct (a fence)}
\fmtExampleRL{Mmaa\-re oku\-si\-ti\-ra olu\-ko\-ro\-go.}
\fmtExampleEN{I have finished constructing a fence.}

\fmtLemma{si\-too\-ki\-si}
\fmtCitation{(e)si\-too\-ki\-si}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{stocking}
\fmtExampleRL{Ba\-zwa\-la si\-too\-ki\-si za ki\-paa\-paa\-li.}
\fmtExampleEN{They wear yellow stockings.}

\fmtLemma{si\-too\-wa}
\fmtCitation{(e)si\-too\-wa}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSubentry(a)
\fmtSense{store}
\fmtExampleRL{Aya\-bi\-re oku si\-too\-wa oku\-gu\-la emi\-kii\-ki\-ryo gya nyu\-mba yaa\-mwe.}
\fmtExampleEN{He has gone to the store to buy rafters for his house.}
\fmtSubentry(b)
\fmtSense{storage room, storeroom, warehouse}
\fmtExampleRL{Emwa\-ni nzi\-tai\-re mu si\-too\-wa.}
\fmtExampleEN{I have put the coffee into the storage room.}

\fmtLemma{si\-wa}
\fmtCitation{(oku)\-si\-wa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-wi\-re}
\fmtSense{to add (sorghum to banana juice during brewing)}
\fmtExampleRL{Ni ba\-maa\-re oku\-nyuu\-ka ensa\-nde ni ba\-si\-wa\-ku omu\-we\-mba.}
\fmtExampleEN{After squeezing banana juice, they added sorghum.}

\fmtLemma{si\-ye\-ta}
\fmtCitation{(e)si\-ye\-ta}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSubentry(a)
\fmtSense{theatre (as a venue)}
\fmtExampleRL{Esi\-ye\-ta ni\-mwo ba\-zee\-nye\-ra emi\-zee\-nyo.}
\fmtExampleEN{They perform dramas at the theatre.}
\fmtSubentry(b)
\fmtSense{theatre (operation room)}
\fmtExampleRL{Omu\-lwai\-re ali mu si\-ye\-ta ba\-ku\-mu\-lo\-ngoo\-sa.}
\fmtExampleEN{The patient is in the theatre undergoing a surgical operation.}

\fmtLemma{si\-zo\-ni}
\fmtCitation{(e)si\-zo\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{season}
\fmtExampleRL{Emi\-ye\-mbe gya\-ka\-li\-rii\-rye esi\-zo\-ni eni.}
\fmtExampleEN{Mangoes are in excess this season.}

\fmtLemma{so\-ba}
\fmtCitation{(oku)\-so\-ba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-be\-re}
\fmtSubentry(a)
\fmtSense{(of a shape or pattern e.g.\,of a mat) to be wrong, to go wrong, to be improper, to be not in accordance with accepted rules or standards}
\fmtExampleRL{Omu\-kee\-ka gu\-so\-be\-re.}
\fmtExampleEN{The mat has lost pattern.}
\fmtSubentry(b)
\fmtSense{(of a woman) to become pregnant while still living with one's parents}
\fmtExampleRL{Omwa\-la aso\-be\-re.}
\fmtExampleEN{The girl became pregnant while still leaving with her parents.}

\fmtLemma{so\-baa\-so\-ba\-na}
\fmtCrossreference{\textnormal{see} soba}

\fmtLemma{so\-bba}
\fmtCitation{(oku)\-so\-bba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-bbe\-re}
\fmtSense{to move slowly}
\fmtExampleRL{Obu\-lwai\-re bu\-mu\-lu\-mi\-re, ata\-ndi\-ki\-re oku\-so\-bba.}
\fmtExampleEN{He has been sick, and thereafter he started to move very slowly.}

\fmtLemma{so\-be\-ra}
\fmtCitation{(oku)\-so\-be\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-bei\-re}
\fmtSense{to confuse, to baffle, to perplex}
\fmtExampleRL{Omwa\-na oyo by’a\-ko\-la bi\-nso\-be\-ra.}
\fmtExampleEN{It confuses me what the girl does.}

\fmtLemma{so\-be\-rwa}
\fmtCitation{(oku)\-so\-be\-rwa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-bei\-rwe}
\fmtSense{to be confused, to be baffled, to be perplexed}
\fmtExampleRL{Omwa\-na gwe baa\-na\-gi\-re omu nzi\-ra ya\-so\-bei\-rwe.}
\fmtExampleEN{The child who was abandoned on the way was confused.}

\fmtLemma{so\-bo\-ka}
\fmtCitation{(oku)\-so\-bo\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-bo\-ke\-re}
\fmtSense{to be doable, to be possible}
\fmtExampleRL{Oka\-nca w’a\-ta\-bba, ebi\-ntu te\-bi\-so\-bo\-ka.}
\fmtExampleEN{Things are impossible where there is no God.}

\fmtLemma{so\-bo\-la}
\fmtCitation{(oku)\-so\-bo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-boi\-re}
\fmtSubentry(a)
\fmtSense{to be able, to can}
\fmtExampleRL{Nso\-bo\-la okwi\-za ewaa\-mu.}
\fmtExampleEN{I can come to your place.}
\fmtSubentry(b)
\fmtSense{to manage, to accomplish, to cope with}
\fmtExampleRL{Ge\-ze\-sya omwa\-na oyo tu\-ma\-nye oba\-nga ya\-so\-bo\-la ekya mu\-sa\-nju.}
\fmtExampleEN{Interview that pupil and find out if he can manage the seventh form.}

\fmtLemma{so\-bo\-lo\-la}
\fmtCrossreference{\textnormal{see} soboora}
\fmtDialect{Lunyala}

\fmtLemma{so\-boo\-ra}
\fmtDialect{Ruruuli}
\fmtCitation{(oku)\-so\-boo\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-boi\-re}
\fmtSubentry(a)
\fmtSense{to arrange, to organise}
\fmtExampleRL{So\-boo\-ra edii\-ro lya\-mu.}
\fmtExampleEN{Organise your sitting room.}
\fmtSubentry(b)
\fmtSense{to explain, to interpret}
\fmtExampleRL{So\-boo\-ra oru\-ba\-zo lwa\-mwe.}
\fmtExampleEN{Interpret his speech.}
\fmtSubentry(c)
\fmtSense{to settle (a dispute)}
\fmtExampleRL{Ya\-ba oso\-boo\-re aboo\-jo baa\-mu.}
\fmtExampleEN{Go and settle your sons’ disputes.}

\fmtLemma{so\-bo\-ro\-ka}
\fmtCitation{(oku)\-so\-bo\-ro\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-bo\-ro\-ke\-re}
\fmtSense{to become a better person, to reform}
\fmtExampleRL{Omu\-si\-ga\-zi omu\-yaa\-ye aso\-bo\-ro\-ke\-re.}
\fmtExampleEN{The villain has become a better person.}

\fmtLemma{so\-bo\-ro\-ki}
\fmtCitation{\textnormal{(e.g.}\,mu\-so\-bo\-ro\-ki, ba\-so\-bo\-ro\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{honest, decent}
\fmtExampleRL{Omwo\-jo oni aba\-ntu ba\-mwe\-si\-ga ku\-bba ata mu\-so\-bo\-ro\-ki.}
\fmtExampleEN{People trust this boy because he is honest.}
\fmtSubentry(b)
\fmtSense{straightforward}
\fmtExampleRL{Bba mu\-so\-bo\-ro\-ki omu byo\-ku\-ba\-za.}
\fmtExampleEN{Be straightforward in your speech.}
\fmtSubentry(c)
\fmtSense{humble, modest}
\fmtExampleRL{Omu\-ntu omu\-so\-bo\-ro\-ki anywa\-na bai\-ngi.}
\fmtExampleEN{A humble person has many friends.}

\fmtLemma{so\-bwoi\-to\-ka}
\fmtCitation{(oku)\-so\-bwoi\-to\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-bwoi\-to\-ke\-re}
\fmtSubentry(a)
\fmtSense{to grow thin, to lose weight}
\fmtExampleRL{Omu\-sai\-za oyo ya\-lwai\-re n’a\-so\-bwoi\-to\-ka.}
\fmtExampleEN{That man was ill, and he grew thin.}
\fmtSubentry(b)
\fmtSense{to slip down, to slip off, to slip through, to go down}
\fmtExampleRL{Empa\-le yaa\-mwe eso\-bwoi\-to\-ke\-re.}
\fmtExampleEN{His trousers have slipped down.}

\fmtLemma{so\-bwoi\-to\-la}
\fmtCitation{(oku)\-so\-bwoi\-to\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-bwoi\-toi\-re}
\fmtSense{to cause to grow thin}
\fmtExampleRL{Omwa\-na oyo obu\-lwai\-re bu\-mu\-so\-bwoi\-toi\-re.}
\fmtExampleEN{That child grew thin because of the illness. (lit.\,The illness has caused the child to be thin).}

\fmtLemma{so\-bya}
\fmtCitation{(oku)\-so\-bya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-be\-rye}
\fmtSubentry(a)
\fmtSense{to err, to make a mistake, to be wrong}
\fmtExampleRL{Bu\-li mu\-ntu aso\-bya.}
\fmtExampleEN{Every person makes mistakes.}
\fmtSubentry(b)
\fmtSense{to fail, to fall short of, to fail to reach (a goal)}
\fmtExampleRL{Na\-so\-be\-rye\-mu ena\-mba ibi\-ri omu ki\-buu\-lyo.}
\fmtExampleEN{I failed two points in the examination.}
\fmtSubentry(c)
\fmtSense{to defile, to rape}
\fmtExampleRL{Ogu\-sai\-za ogu\-mbu\-la\-nso\-ni gwa\-so\-be\-rye oku mwa\-na.}
\fmtExampleEN{A rapist man defiled a child.}

\fmtLemma{so\-do\-ka}
\fmtCitation{(oku)\-so\-do\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-do\-ke\-re}
\fmtSubentry(a)
\fmtSense{to jump, to leap, to spring, to frog jump, to hop}
\fmtExampleRL{Aba\-sai\-za ba\-ku\-so\-do\-ka.}
\fmtExampleEN{The men are jumping.}
\fmtSubentry(b)
\fmtFixedexpression{so\-doka (o)mu\-gu\-wa}
\fmtSense{to skip rope, to jump over a rope}
\fmtExampleRL{Abaa\-la ba\-ta\-ka mu\-no oku\-so\-do\-ka omu\-gu\-wa.}
\fmtExampleEN{Girls like to skip rope very much.}

\fmtLemma{so\-doo\-ka}
\fmtCitation{(oku)\-so\-doo\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-doo\-ke\-re}
\fmtSense{to walk sluggishly, to walk slowly, to drag oneself}
\fmtExampleRL{Enko\-ko zi\-ku\-so\-doo\-ka ku\-bba zi\-rwai\-re.}
\fmtExampleEN{The hens are walking sluggishly because they are ill.}

\fmtLemma{so\-go\-nyo\-ka}
\fmtCitation{(oku)\-so\-go\-nyo\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-go\-nyo\-ke\-re}
\fmtSubentry(a)
\fmtSense{to be overripe, to be too soft, to fall to pieces (e.g.\,of overcooked food)}
\fmtExampleRL{Ame\-nju ga\-so\-go\-nyo\-ke\-re.}
\fmtExampleEN{The yellow bananas have become too soft.}
\fmtSubentry(b)
\fmtSense{to be exhausted}

\fmtLemma{so\-go\-nyo\-la}
\fmtCitation{(oku)\-so\-go\-nyo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-go\-nyoi\-re}
\fmtSense{to squash, to crush, to squeeze}
\fmtExampleRL{So\-go\-nyo\-la ame\-nju ago.}
\fmtExampleEN{Squash those yellow bananas.}

\fmtLemma{so\-ke\-ra}
\fmtCitation{(oku)\-so\-ke\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-kei\-re}
\fmtSense{to load (bullets)}
\fmtExampleRL{So\-ke\-ra ama\-sa\-si omu mbu\-ndu.}
\fmtExampleEN{Load bullets into the gun.}

\fmtLemma{so\-ko\-mbo\-la}
\fmtCitation{(oku)\-so\-ko\-mbo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-ko\-mboi\-re}
\fmtSense{to unblock, to remove (an obstruction from a pipe or drain), to pick out, to pluck out, to pick at, to dig out}
\fmtExampleRL{Oku\-so\-ko\-mbo\-la ebi\-ta\-ka\-li omu mu\-fu\-le\-je?}
\fmtExampleEN{Are you digging out soil from the drainage?}

\fmtLemma{so\-ko\-ndo\-lya}
\fmtCitation{(oku)\-so\-ko\-ndo\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-ko\-ndoi\-rye}
\fmtSubentry(a)
\fmtSense{to peck at, to make (a hole) by striking with the beak}
\fmtExampleRL{Eki\-nyo\-nyi ki\-ku\-so\-ko\-ndo\-lya omu\-saa\-le.}
\fmtExampleEN{The bird is making a hole in the tree by striking with the beak.}
\fmtSubentry(b)
\fmtSense{to remove (through a small opening or a narrow passage)}
\fmtExampleRL{So\-ko\-ndo\-lya eme\-se mu ki\-do\-mo\-la.}
\fmtExampleEN{Remove the rat from the jerry can.}

\fmtLemma{so\-la}
\fmtCrossreference{\textnormal{see} sora}
\fmtDialect{Lunyala}

\fmtLemma{so\-le\-be\-za}
\fmtCitation{(oku)\-so\-lo\-be\-za}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-lo\-bei\-ze, so\-lo\-beze\-re}
\fmtSense{to slurp (e.g.\,food), to eat noisily, to suck noisily}
\fmtExampleRL{Omwa\-na ni alya eki\-kai\-zo, aso\-lo\-be\-za.}
\fmtExampleEN{When the child chews sugarcane, he slurps.}

\fmtLemma{so\-li\-ma}
\fmtCitation{(oku)\-so\-li\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-li\-mi\-re}
\fmtSubentry(a)
\fmtSense{(of steam) to escape, to be released}
\fmtExampleRL{Ama\-ngu nga omwi\-ka gu\-ta\-ndi\-ki\-re oku\-so\-li\-ma, na\-swe\-kwi\-re eki\-na\-ga.}
\fmtExampleEN{As soon as the steam started to be released, I uncovered the pot.}
\fmtSubentry(b)
\fmtSense{to walk stealthily, to move stealthily, to sneak}
\fmtExampleRL{Omu\-sai\-za ya\-so\-li\-mi\-re mpo\-la n’a\-ya\-ba.}
\fmtExampleEN{The man walked stealthily away.}
\fmtSubentry(c)
\fmtSense{to steal little by little}
\fmtExampleRL{Naa\-ni aku\-so\-li\-ma empii\-ya zaa\-nge?}
\fmtExampleEN{Who is stealing my money little by little?}

\fmtLemma{so\-lo\-bya}
\fmtCitation{(oku)\-so\-lo\-bya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-lo\-be\-rye}
\fmtSense{to prune with a pruning pole}
\fmtExampleRL{Ako\-le\-se\-rye olu\-so\-lo\-byo oku\-so\-lo\-bya enda\-ga\-la.}
\fmtExampleEN{She used the pruning pole to prune the banana leaves.}

\fmtLemma{so\-ma}%
\fmtOne{1}
\fmtCitation{(oku)\-so\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-me\-re}
\fmtSubentry(a)
\fmtSense{to peck at, to pick up (e.g.\,termites)}
\fmtExampleRL{Onko\-ko anso\-me\-re omu ki\-rwai\-re.}
\fmtExampleEN{A hen has pecked me in the wound.}
\fmtSubentry(b)
\fmtSense{to nag, to henpeck, to find foult with}
\fmtSubentry(c)
\fmtSense{to eat (soup), to sip soup, to sup}
\fmtExampleRL{Omwa\-na abbai\-re aku\-so\-ma omu\-cwe ne gu\-mwo\-ca.}
\fmtExampleEN{The child was eating soup, and it burnt her.}
\fmtSubentry(c)
\fmtSense{(of a snake) to bite}
\fmtExampleRL{Onjo\-ka ya\-mu\-so\-me\-re nga aku\-zwa omu ki\-bi\-ra.}
\fmtExampleEN{He was bitten by a snake while coming from the forest.}

\fmtLemma{so\-ma}%
\fmtOne{2}
\fmtCitation{(oku)\-so\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-me\-re}
\fmtSubentry(a)
\fmtSense{to read}
\fmtExampleRL{Mu\-wo\-nge ya\-so\-me\-re eBbai\-bbu\-li oku Sa\-nde ku\-sai mu\-no.}
\fmtExampleEN{Muwonge read the Bible on Sunday quite well.}
\fmtSubentry(b)
\fmtSense{to study}
\fmtExampleRL{Omwa\-na aku\-so\-ma ku\-sai.}
\fmtExampleEN{The child is studying well.}

\fmtLemma{so\-mba}
\fmtCitation{(oku)\-so\-mba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-mbe\-re}
\fmtSense{to collect (in several trips), to to gather, to to bring together}
\fmtExampleRL{So\-mba ama\-bbaa\-le ago ogai\-ngi\-rye omu nyu\-mba.}
\fmtExampleEN{Collect those stones and put them into the house.}

\fmtLemma{so\-mboo\-ka}
\fmtCitation{(oku)\-so\-mboo\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-mboo\-ke\-re}
\fmtSense{to congregate, to to gather in a crowd, to to come in large numbers}
\fmtExampleRL{Aba\-ntu baa\-so\-mboo\-ke\-re oku\-ra\-mya Oka\-nca.}
\fmtExampleEN{Many people congregated to pray to God.}

\fmtLemma{so\-mboo\-la}
\fmtCitation{(oku)\-so\-mboo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-mboi\-re}
\fmtSubentry(a)
\fmtSense{to collect, to to gather (in large quantities)}
\fmtExampleRL{Emi\-saa\-le ba\-ku\-so\-mboo\-la gya kwo\-mbo\-ka ki\-ga\-ngo.}
\fmtExampleEN{They are collecting many poles for constructing a hall.}
\fmtSubentry(b)
\fmtSense{to gather together, to bring together in a large group, to bring in crowds, to assemble}
\fmtExampleRL{Omu\-ka\-li ya\-so\-mboi\-re aba\-ntu okwa\-ba oku\-mu\-bo\-na.}
\fmtExampleEN{She brought together very many people.}

\fmtLemma{so\-me\-ra}
\fmtCitation{(oku)\-so\-me\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-me\-re}
\fmtSense{to study (a field, discipline)}
\fmtExampleRL{Ya\-so\-me\-rei\-re bya\-bu\-lu\-mi.}
\fmtExampleEN{He studied agriculture.}

\fmtLemma{so\-me\-sya}
\fmtCitation{(oku)\-so\-me\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-me\-se\-rye}
\fmtSense{to teach}
\fmtExampleRL{Amai\-te oku\-tu\-so\-me\-sya Olu\-ju\-ngu ni\-tu\-cu\-raa\-nu\-ka ei\-so\-mo.}
\fmtExampleEN{He knows how to teach us English, and we fully understand.}

\fmtLemma{so\-mo\-ka}
\fmtCitation{(oku)\-so\-mo\-ka}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{so\-mo\-ke\-re}
\fmtSubentry(a)
\fmtSense{to cross (e.g.\,a lake, a river), to cross over (a body of water)}
\fmtExampleRL{Aba\-ntu ba\-ku\-so\-mo\-ka enya\-nja.}
\fmtExampleEN{People are crossing the lake.}
\fmtSubentry(b)
\fmtSense{(euphemism) to die, to pass away}
\fmtExampleRL{Ya\-so\-me\-ke\-re mu mwa\-ka ogu\-wei\-re.}
\fmtExampleEN{He passed away last year.}

\fmtLemma{so\-mo\-la}
\fmtCitation{(oku)\-so\-mo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-moi\-re}
\fmtSubentry(a)
\fmtSense{(of a boat) to transport, to cause to cross over, to ferry over, to cross over (a body of water)}
\fmtExampleRL{Naa\-ni eya\-so\-mo\-la aba\-ntu omwi\-ga?}
\fmtExampleEN{Who will transport the people across the river?}
\fmtSubentry(b)
\fmtSense{to take away slowly or gradually, to steal one by one}
\fmtExampleRL{Lwa\-ki oku\-so\-mo\-la ebya\-ma bya Bu\-ka\-ma?}
\fmtExampleEN{Why are you stealing the kingdom’s secrets?}

\fmtLemma{so\-mo\-lya}
\fmtCitation{(oku)\-so\-mo\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-moi\-rye}
\fmtSense{to take away slowly or gradually, to steal one by one}
\fmtExampleRL{Omwa\-na aso\-moi\-rye omu\-ni\-ki omu ka\-na\-ga.}
\fmtExampleEN{A child has snatched piece of meat from the cooking pot.}

\fmtLemma{so\-na}
\fmtCitation{(oku)\-so\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-ne\-re}
\fmtSense{to weave, to weave, to sew, to stitch}
\fmtExampleRL{Nkya\-so\-na mu\-kee\-ka gwa\-nge.}
\fmtExampleEN{I am still weaving my mat.}

\fmtLemma{so\-nda}
\fmtCitation{(oku)\-so\-nda}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-nde\-re}
\fmtSense{to contribute money (for a wedding or funeral), to gather together little by little}
\fmtExampleRL{Tu\-ku\-so\-nda se\-nte za mba\-ga.}
\fmtExampleEN{We are gathering little by little money for the construction of a church.}

\fmtLemma{so\-nga}%
\fmtOne{1}
\fmtPOS{conj}
\fmtSense{yet, but nevertheless, whereas}
\fmtExampleRL{Na\-ya\-kai\-re Mu\-go\-ngo so\-nga ye aya\-kai\-re Gwe\-ro.}
\fmtExampleEN{I went to Mugongo whereas she went to Gwero.}

\fmtLemma{so\-nga}%
\fmtOne{2}
\fmtCitation{(oku)\-so\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-nge\-re}
\fmtSubentry(a)
\fmtSense{to slaughter (a dying animal)}
\fmtExampleRL{Onte ali ampi oku\-fa ba\-mu\-so\-nga.}
\fmtExampleEN{A cow that is about to die is slaughtered.}
\fmtSubentry(b)
\fmtSense{to spear (an animal) for the second time}
\fmtExampleRL{So\-nga oki\-so\-lo omwi\-te.}
\fmtExampleEN{Spear the animal again to kill it.}

\fmtLemma{so\-ngo\-ba}
\fmtCitation{(oku)\-so\-ngo\-ba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-ngo\-be\-re}
\fmtSense{(of a boy) to break voice}
\fmtExampleRL{Omu\-taa\-ni wa\-nge aso\-ngo\-be\-re.}
\fmtExampleEN{My son's voice has broken.}

\fmtLemma{so\-ngo\-ka}
\fmtCitation{(oku)\-so\-ngo\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-ngo\-ke\-re}
\fmtSense{(of a pencil or stick) to be sharpened, to be pointed}
\fmtExampleRL{Eka\-laa\-mu eso\-ngo\-ke\-re ku\-sai.}
\fmtExampleEN{The pencil is well sharpened.}

\fmtLemma{so\-ngo\-la}
\fmtCrossreference{\textnormal{see also} songolya}
\fmtCitation{(oku)\-so\-ngo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-ngoi\-re}
\fmtSense{to make pointed, to sharpen (e.g.\,a stick, a pencil)}
\fmtExampleRL{So\-ngo\-la eka\-laa\-mu za baa\-na.}
\fmtExampleEN{Sharpen the children's pencils.}

\fmtLemma{so\-ngo\-li}
\fmtCitation{\textnormal{(e.g.}\,mu\-so\-ngo\-li, ki\-so\-nso\-ngo\-li, lu\-so\-ngo\-li\textnormal{)}}
\fmtPOS{adj}
\fmtSense{pointed, elongated}
\fmtExampleRL{Eka\-laa\-mu ya\-nge gi\-so\-go\-li.}
\fmtExampleEN{My pencil is sharp.}

\fmtLemma{so\-ngo\-lya}
\fmtCrossreference{\textnormal{see also} songola}
\fmtCitation{(oku)\-so\-ngo\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-ngoi\-rye}
\fmtSense{to make pointed, to sharpen (e.g.\,a stick, a pencil)}
\fmtExampleRL{So\-ngo\-lya eka\-laa\-mu ya\-nge!}
\fmtExampleEN{Sharpen my pencil!}

\fmtLemma{so\-nse\-ka}
\fmtDialect{Lunyala}
\fmtCitation{(oku)\-so\-nse\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-nse\-ke\-re}
\fmtSense{to insert, to push (into a hole or opening), to slip in}
\fmtExampleRL{So\-nse\-ka oku\-gu\-lu omu nkai\-to.}
\fmtExampleEN{Push the foot into the shoe.}

\fmtLemma{so\-nso\-da}
\fmtCitation{(oku)\-so\-nso\-da}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-nso\-de\-re}
\fmtSense{to limp}
\fmtExampleRL{Lwa\-ki omwa\-na oyo aku\-so\-nso\-da?}
\fmtExampleEN{Why is that child limping?}

\fmtLemma{so\-nso\-ka}
\fmtDialect{Ruruuli}
\fmtCitation{(oku)\-so\-nso\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-nso\-ke\-re}
\fmtSense{to insert, to push in, to slip in}
\fmtExampleRL{Ya\-so\-nso\-ke\-re omu\-saa\-le omu ki\-wu\-lu.}
\fmtExampleEN{He pushed a stick into the hole.}

\fmtLemma{so\-nya}
\fmtCitation{(oku)\-so\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-nye\-re}
\fmtSense{to instigate, to prompt, to incite, to encourage, to provoke}
\fmtExampleRL{Omwo\-jo oni aku\-so\-nya abai\-ra baa\-mwe oku\-lwa\-na.}
\fmtExampleEN{This boy is provoking his pals to fight.}

\fmtLemma{so\-nyi\-wa}
\fmtCitation{(oku)\-so\-nyi\-wa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-nyi\-wi\-re}
\fmtSense{to forgive, to pardon, to excuse}
\fmtExampleRL{So\-nyi\-wa aku\-koo\-re oku\-bbi.}
\fmtExampleEN{Forgive one who does you wrong.}

\fmtLemma{so\-nyi\-wa\-la}
\fmtCitation{(oku)\-so\-nyi\-wa\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-nyi\-wai\-re}
\fmtSense{to be ashamed, to feel guilty}
\fmtExampleRL{Eki\-sai\-za ba\-ki\-kwai\-te ki\-ku\-syo\-la, ki\-ri awo ki\-so\-nyi\-wai\-re.}
\fmtExampleEN{A man was arrested practicing night-dancing, and he is ashamed.}

\fmtLemma{soo\-ba}
\fmtCitation{(oku)\-soo\-ba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{soo\-be\-re}
\fmtSense{to walk slowly, to go slowly, to move stealthily, to move carefully}
\fmtExampleRL{Soo\-ba okwa\-te onko\-ko.}
\fmtExampleEN{Move slowly and catch the hen.}

\fmtLemma{soo\-da}
\fmtCitation{(o)soo\-da}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{soda}
\fmtExampleRL{Tu\-ku\-nya osoo\-da.}
\fmtExampleEN{We are drinking soda.}

\fmtLemma{soo\-ka}
\fmtCitation{(oku)\-soo\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{soo\-ke\-re}
\fmtSense{to do first, to be first, to be the first one}
\fmtExampleRL{Ni\-ye ya\-tu\-soo\-ke\-re oku\-naa\-ba omu nga\-lo.}
\fmtExampleEN{He was the first to wash the hands.}

\fmtLemma{soo\-ki\-si}
\fmtCitation{(e)soo\-ki\-si}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{sock}
\fmtExampleRL{Nku\-zwa\-la soo\-ki\-si.}
\fmtExampleEN{I am wearing socks.}

\fmtLemma{soo\-le}
\fmtCitation{(e)soo\-le}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{sole (of a shoe or footwear)}
\fmtExampleRL{Esoo\-le ya nkai\-to yaa\-mu gi\-gu\-mu.}
\fmtExampleEN{The sole of your shoe is strong.}

\fmtLemma{soo\-loo\-ba}
\fmtCitation{(oku)\-soo\-loo\-ba}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{soo\-loo\-be\-re}
\fmtSubentry(a)
\fmtSense{to be superior to, to be senior to}
\fmtExampleRL{Opu\-re\-zi\-de\-nti abba ya\-soo\-loo\-be\-re omu ba\-ntu.}
\fmtExampleEN{A president is superior to other people.}
\fmtSubentry(b)
\fmtSense{to be outstanding}
\fmtExampleRL{Omwe\-gi oyo ya\-soo\-lo\-be\-re ku bai\-raa\-mwe.}
\fmtExampleEN{That student was outstanding among others.}

\fmtLemma{soo\-na}
\fmtCitation{(oku)\-soo\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{soo\-ne\-re}
\fmtSubentry(a)
\fmtSense{to steal an opportunity from, to snatch an opportunity from}
\fmtExampleRL{Emo\-to\-ka yaa\-ku\-bbai\-re ya\-nge bai\-tu oBye\-rwa\-nju ya\-nsoo\-ne\-re.}
\fmtExampleEN{That car could have been mine, but Bye\-rwa\-nju snatched the opportunity from me.}
\fmtSubentry(b)
\fmtSense{to trick in order to get information}
\fmtExampleRL{Omu\-ka\-li oyo mu\-soo\-ne aku\-ko\-be\-re eki\-yi\-nza oku\-bba nga ni\-kyo ekyai\-ti\-re oi\-ba waa\-mwe.}
\fmtExampleEN{Trick that woman so that she can tell you what could have caused her husband's death.}

\fmtLemma{soo\-noo\-ka}
\fmtCitation{(oku)\-soo\-noo\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{soo\-noo\-ke\-re}
\fmtSense{to walk slowly, to act slowly, to move slowly and cautiously, to walk stealthily}
\fmtExampleRL{Lwa\-ki oku\-soo\-noo\-ka, wa\-bba mu\-lwai\-re?}
\fmtExampleEN{Why are you walking slowly? Are you sick?}

\fmtLemma{soo\-nya}
\fmtCitation{(oku)\-soo\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{soo\-nye\-re}
\fmtSense{to peck}
\fmtExampleRL{Onko\-ko ansoo\-nye\-re omu kyo\-ya.}
\fmtExampleEN{A hen has pecked me in the wound.}

\fmtLemma{soo\-si}%
\fmtOne{1}
\fmtCitation{(o)soo\-si}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{dung of a tender calf (before it starts grazing)}
\fmtExampleRL{Onya\-na eya\-ka\-bya\-li\-bwa awu\-lu\-ka soo\-si.}
\fmtExampleEN{A newly born calf defecates `soosi'.}

\fmtLemma{soo\-si}%
\fmtOne{2}
\fmtCitation{(e)soo\-si}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{saucer}
\fmtExampleRL{Amai\-zi ga mu\-ge\-nyi ga\-twa\-re ku soo\-si.}
\fmtExampleEN{Carry the water for the visitor in the saucer.}

\fmtLemma{soo\-za}
\fmtCitation{(oku)\-soo\-za}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{soo\-ze\-re}
\fmtSense{to practise teeth-sucking, to make a sucking noise with the lips as an expression of contempt, to suck one's teeth as an expression of contempt}

\fmtLemma{so\-ra}
\fmtCitation{(oku)\-so\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{soi\-re}
\fmtSubentry(a)
\fmtSense{to dig up, to harvest (e.g.\,groundnuts)}
\fmtExampleRL{Amai\-do ba\-ga\-so\-ra oku\-zwa ku ki\-ti\-na.}
\fmtExampleEN{The groundnuts are removed from the stem.}
\fmtSubentry(b)
\fmtFixedexpression{so\-ra mwe\-so}
\fmtSense{to play a board game}

\fmtLemma{so\-ro\-ma}
\fmtCitation{(oku)\-so\-ro\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-ro\-me\-re}
\fmtSense{to gather, to compile, to assemble, to collect}
\fmtExampleRL{So\-ro\-ma emwa\-ni ezo zi\-wee\-wo.}
\fmtExampleEN{Gather all that coffee.}

\fmtLemma{so\-roo\-lya}
\fmtCitation{(oku)\-so\-roo\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-roi\-rye}
\fmtSense{to gather, to collect (taxes), to requisition}
\fmtExampleRL{So\-roo\-lya omu\-so\-lo oku\-zwa omu ka\-ta\-le.}
\fmtExampleEN{Collect the tax from the market.}

\fmtLemma{so\-roo\-ra}
\fmtCrossreference{\textnormal{see also} sosola}
\fmtCitation{(oku)\-so\-roo\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-roi\-re}
\fmtSubentry(a)
\fmtSense{to pick, to pick out (bones of fish, one's nails, etc.)}
\fmtExampleRL{Soo\-ka oso\-roo\-re oncwi nga okya\-la ku\-mu\-wee\-rya baa\-na.}
\fmtExampleEN{First pick out the bones from the fish before you give to the children.}
\fmtSubentry(b)
\fmtSense{to sort, to organise}
\fmtExampleRL{So\-roo\-ra\-mu ebi\-ko\-po ebi\-sai.}
\fmtExampleEN{Sort out the fine cups.}
\fmtSubentry(c)
\fmtSense{to segregate, to marginalise, to discriminate against (e.g.\,a person of another religion)}
\fmtExampleRL{Aba\-gai\-ga ba\-te\-ra oku\-so\-roo\-ra aba\-see\-ge.}
\fmtExampleEN{The rich tend to discriminate against the poor.}

\fmtLemma{so\-sa}
\fmtCitation{(oku)\-so\-sa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-se\-re}
\fmtSense{to check, to examine}
\fmtExampleRL{Ate\-ge\-re omu\-te\-go, aso\-sa.}
\fmtExampleEN{The one who sets a trap, checks it.}

\fmtLemma{so\-soi\-dya}
\fmtCitation{(oku)\-so\-soi\-dya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-soi\-de\-rye}
\fmtSense{to provoke, to instigate, to incite, to encourage, to pick a quarrel with}
\fmtExampleRL{Ka nya\-be mmu\-so\-soi\-dye alwa\-ne.}
\fmtExampleEN{Let me go and provoke him to fight.}

\fmtLemma{so\-so\-la}
\fmtCrossreference{\textnormal{see also} soroora}
\fmtCitation{(oku)\-so\-so\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-soi\-re}
\fmtSubentry(a)
\fmtSense{to pick, to to pick out (bones of fish, one's nails, to etc.)}
\fmtExampleRL{Soo\-ka oso\-so\-le oncwi nga okya\-la ku\-mu\-wee\-rya baa\-na.}
\fmtExampleEN{First pick out the bones from the fish before you give to the children.}
\fmtSubentry(b)
\fmtSense{to sort, to organise}
\fmtSubentry(c)
\fmtSense{to segregate, to discriminate}
\fmtExampleRL{Aba\-gai\-ga ba\-te\-ra oku\-so\-roo\-ra aba\-na\-ku.}
\fmtExampleEN{The rich tend to discriminate against the needy.}

\fmtLemma{so\-soo\-to\-la}
\fmtCitation{(oku)\-so\-soo\-to\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-soo\-toi\-re}
\fmtSense{to unwrap (a bundle of cooked food before serving)}
\fmtExampleRL{So\-soo\-to\-la eme\-re tu\-lye.}
\fmtExampleEN{Unwrap the food so that we can eat.}

\fmtLemma{so\-tya}
\fmtCitation{(oku)\-so\-tya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{soi\-te\-rye}
\fmtSense{to move slowly (due to fatigue or illness), to drag oneself}
\fmtExampleRL{Na\-mwa\-kai\-re\-ku obwi\-re ng’a\-ku\-so\-tya.}
\fmtExampleEN{I passed by her while she was walking very slowly.}

\fmtLemma{so\-waa\-ni}
\fmtCrossreference{\textnormal{see} saani}

\fmtLemma{so\-wee\-ra}
\fmtCitation{(oku)\-so\-wee\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{so\-wei\-re, so\-we\-rei\-re}
\fmtSense{to compensate, to make up (to)}
\fmtExampleRL{Nii\-we wa\-mu\-ku\-bbi\-re olu\-bba\-le, mu\-so\-wee\-re!}
\fmtExampleEN{It is you who beat her, and she got a cut, so compensate her.}

\fmtLemma{so\-wo\-nka\-na}
\fmtCitation{(oku)\-so\-wo\-nka\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{so\-wo\-nkai\-ne}
\fmtSense{to be elongated}
\fmtExampleRL{Eki\-go\-go ki\-ni kya\-so\-wo\-nkai\-ne.}
\fmtExampleEN{This banana plant is elongated.}

\fmtLemma{so\-ya\-bi\-ni}
\fmtCitation{(o)so\-ya\-bi\-ni}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{soya bean}
\fmtExampleRL{Oso\-ya\-bi\-ni n'a\-ka\-la apaa\-lu\-nka.}
\fmtExampleEN{When soya bean seeds get dry, they pop.}

\fmtLemma{Ssa\-be\-pi\-si\-koo\-pi}
\fmtCitation{(o)Ssa\-be\-pi\-si\-koo\-pi}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from En.}
\fmtSense{archbishop (of the Catholic church)}
\fmtExampleRL{OSsaa\-be\-pi\-skoo\-pi ni\-ye akuu\-ra ei\-sa\-za eri\-ku\-lu.}
\fmtExampleEN{The archbishop is the head of the archdiocese.}

\fmtLemma{su\-bii\-ra}
\fmtCitation{(oku)\-su\-bii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-bi\-rii\-re}
\fmtSense{to tell (a story, a legend), to narrate}
\fmtExampleRL{Ka nku\-su\-bii\-re eki\-su\-bii\-ro kya Wa\-ka\-me n’Onta\-le.}
\fmtExampleEN{Let me tell you a story about the Hair and the Lion.}

\fmtLemma{su\-bu\-ka}%
\fmtOne{1}
\fmtCitation{(oku)\-su\-bu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{su\-bu\-ki\-re}
\fmtSense{to hail from, to originate}
\fmtExampleRL{Osu\-bu\-ka ayi?}
\fmtExampleEN{Where do you hail from?}

\fmtLemma{su\-bu\-ka}%
\fmtOne{2}
\fmtCitation{(oku)\-su\-bu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-bu\-ki\-re}
\fmtSense{to tie, to bind}
\fmtExampleRL{Su\-bu\-ka ombu\-li oyo.}
\fmtExampleEN{Tie that goat.}

\fmtLemma{su\-bu\-la\-ku}
\fmtCitation{(oku)\-su\-bu\-la\-ku}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{su\-bwi\-re\-ku}
\fmtSense{to give up}
\fmtExampleRL{Na\-li\-ndi\-re odu\-ree\-va nga ta\-kwi\-za, ni nsu\-bu\-la\-ku, ni nya\-ba.}
\fmtExampleEN{I waited for the driver until I gave up and left.}

\fmtLemma{su\-bwa}
\fmtCitation{(oku)\-su\-bwa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-bi\-rwe}
\fmtSense{to miss (an opportunity)}
\fmtExampleRL{Wa\-su\-bi\-rwe oku\-bo\-na oIsa\-ba\-ruu\-li.}
\fmtExampleEN{You missed the opportunity to see the Isabaruuli.}

\fmtLemma{su\-ka}
\fmtCitation{(oku)\-su\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-ki\-re}
\fmtSense{to pour}
\fmtExampleRL{Ya\-su\-ki\-re amai\-zi omu ki\-ko\-po n’a\-wee\-rya omwa\-na waa\-mwe.}
\fmtExampleEN{She poured water into the cup and gave it to her child.}

\fmtLemma{su\-kaa\-li}
\fmtCitation{(o)su\-kaa\-li}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSubentry(a)
\fmtSense{sugar}
\fmtExampleRL{Osu\-kaa\-li ya\-nii\-ni\-re ebbee\-yi.}
\fmtExampleEN{The price of sugar has risen.}
\fmtSubentry(b)
\fmtSense{diabetes}
\fmtExampleRL{E Mu\-la\-go wa\-li\-yo aba\-lwai\-re ba su\-kaa\-li bai\-ngi.}
\fmtExampleEN{There are many patients with diabetes in Mulago hospital.}
\fmtSubentry(c)
\fmtSense{baby-girl}
\fmtExampleRL{Na\-byai\-re su\-kaa\-li.}
\fmtExampleEN{I gave birth to a baby-girl.}

\fmtLemma{su\-kaa\-li\-gu\-lu}
\fmtCitation{(o)su\-kaa\-li\-gu\-lu}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{brown sticky sugar commonly used to crude waragi}
\fmtExampleRL{Osu\-kaa\-li\-gu\-lu ba\-mu\-ko\-la\-mu wa\-ra\-gi.}
\fmtExampleEN{The brown sticky sugar is used to make crude waragi.}

\fmtLemma{su\-kii\-ra}
\fmtCitation{(oku)\-su\-kii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-ki\-rii\-re}
\fmtSubentry(a)
\fmtSense{to water, to irrigate}
\fmtExampleRL{Mu\-su\-kii\-re ebi\-ru\-me bya\-nywe le\-ke mu\-go\-bo\-lo\-le.}
\fmtExampleEN{Irrigate your plants for maximum profits.}
\fmtSubentry(b)
\fmtSense{to assist by pouring water (for a person) to wash hands}
\fmtExampleRL{Su\-kii\-ra aba\-ge\-ni omu nga\-lo.}
\fmtExampleEN{Help pouring water for the visitors to wash hands.}

\fmtLemma{su\-ku\-ka}
\fmtCitation{(oku)\-su\-ku\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{su\-ku\-ki\-re}
\fmtSubentry(a)
\fmtSense{to flow rapidly, to pour, to be shaken out (e.g.\,a liquid from a container)}
\fmtExampleRL{Amai\-zi ga\-ku\-su\-ku\-ka oku\-li\-ga omu ki\-do\-mo\-la.}
\fmtExampleEN{The water is pouring from the jerry can.}
\fmtSubentry(b)
\fmtSense{to rain}
\fmtExampleRL{Oi\-ke\-ndi asu\-ku\-ki\-re mwi\-ngi olwa\-lee\-ro.}
\fmtExampleEN{It has rained heavily today.}

\fmtLemma{su\-ku\-la}
\fmtCitation{(oku)\-su\-ku\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-kwi\-re}
\fmtSubentry(a)
\fmtSense{to shake}
\fmtExampleRL{Su\-ku\-la na maa\-ni eki\-do\-mo\-la kye\-re.}
\fmtExampleEN{Shake the jerry can so that it cleans properly.}
\fmtSubentry(b)
\fmtSense{to position (e.g.\,a child on one's back)}
\fmtExampleRL{Omwa\-na waa\-mu mu\-su\-ku\-le ata\-gwa.}
\fmtExampleEN{Push forward the child so that she does not fall down.}
\fmtSubentry(c)
\fmtSense{to pull up (e.g.\,one's pants)}
\fmtSubentry(d)
\fmtSense{(of a calf) to hit (the udder of a cow)}
\fmtExampleRL{Onya\-na asu\-ku\-la eki\-bee\-re le\-ka ama\-tai gai\-ze.}
\fmtExampleEN{The calf hits the udder to make the milk flow.}

\fmtLemma{suu\-ku\-lu\-du\-lee\-va}
\fmtCitation{(e)suu\-ku\-lu\-du\-lee\-va}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{screwdriver}
\fmtExampleRL{Aba\-ma\-ka\-ni\-ka ba\-ko\-le\-sya esuu\-ku\-lu\-du\-lee\-va oku\-bbwa ena\-ti.}
\fmtExampleEN{The mechanics use screwdrivers to tighten the nuts.}

\fmtLemma{su\-la}
\fmtCitation{(oku)\-su\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{suu\-re}
\fmtSense{to sleep, to spend a night, to pass a night}
\fmtExampleRL{Su\-la ani oya\-be ema\-mbya.}
\fmtExampleEN{Spend the night here and go tomorrow.}

\fmtLemma{su\-lii\-ra}
\fmtCitation{(oku)\-su\-lii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{su\-li\-rii\-re}
\fmtSubentry(a)
\fmtSense{to be on the verge of, to be about to}
\fmtExampleRL{Omu\-ka\-li wa\-nge aku\-su\-lii\-ra ku\-bya\-la.}
\fmtExampleEN{My wife is about to give birth.}
\fmtSubentry(b)
\fmtSense{to plan (especially for the following day)}
\fmtExampleRL{Nku\-su\-lii\-ra ku\-kee\-ra Ka\-mpa\-la.}
\fmtExampleEN{I plan to go very early to Kampala.}
\fmtSubentry(c)
\fmtSense{to spend the night with, to stay overnight with (often implying inconvenience to the usual in habitants), to spend the night preoccupied with (a present concern or future plans)}

\fmtLemma{su\-li\-ka}
\fmtCitation{(oku)\-su\-li\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-li\-ki\-re}
\fmtSubentry(a)
\fmtSense{to turn upside down, to turn over, to reverse}
\fmtSubentry(b)
\fmtSense{to put at an angle, to incline, to slant, to tilt}
\fmtExampleRL{Su\-li\-ka efu\-la\-si\-ka amai\-zi ga\-zwe\-mu goo\-na.}
\fmtExampleEN{Slant the flask so that all the water pours out.}
\fmtSubentry(c)
\fmtSense{to bow down (e.g.\,one's head)}
\fmtExampleRL{Ki\-sai oku\-su\-li\-ka omu\-twe nga obwoi\-ne omu\-ntu omu\-ku\-lu.}
\fmtExampleEN{It is good to bow your head if you see an elder.}

\fmtLemma{su\-li\-sya}
\fmtCitation{(oku)\-su\-li\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-li\-si\-rye}
\fmtSense{to accommodate for a night, to give company at night}
\fmtExampleRL{Amu\-su\-li\-si\-rye ebi\-ro bi\-bi\-ri.}
\fmtExampleEN{He accommodated him for two nights.}

\fmtLemma{su\-li\-za}
\fmtCitation{(oku)\-su\-li\-za}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{su\-lii\-ze, su\-li\-zi\-re}
\fmtSense{to sniffle, to sniff, to draw air into the nose, to snuffle, to snort}
\fmtExampleRL{Abaa\-na aba\-mwe ba\-ta\-ka oku\-su\-li\-za mu ki\-fo kya ku\-fu\-lya.}
\fmtExampleEN{Some children prefer sniffling to blowing their noses.}

\fmtLemma{su\-lu\-ka}
\fmtCitation{(oku)\-su\-lu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-lu\-ki\-re}
\fmtSense{(of a roof) to incline, to make slope, to tilt}
\fmtExampleRL{Enyu\-mba baa\-gi\-su\-lu\-ki\-re ku\-sai.}
\fmtExampleEN{The roof of the house was inclined well.}

\fmtLemma{su\-lye\-nke}
\fmtCitation{(o)su\-lye\-nke}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{raven}
\fmtExampleRL{Osu\-lye\-nke alya ame\-nju.}
\fmtExampleEN{The raven eats the yellow bananas.}

\fmtLemma{su\-ma\-ntya}
\fmtCitation{(oku)\-su\-ma\-ntya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-ma\-nti\-rye}
\fmtSense{to make smooth (the surface of pottery)}
\fmtExampleRL{Nku\-su\-ma\-ntya mi\-ru\-gwa gya bi\-na\-ga.}
\fmtExampleEN{I am smoothening the frames of my pots.}

\fmtLemma{su\-mba}
\fmtCitation{(oku)\-su\-mba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-mbi\-re}
\fmtSense{to boil, to cook}
\fmtExampleRL{Su\-mba amai\-zi ago tu\-ga\-nywe.}
\fmtExampleEN{Boil that water and we'll take it.}

\fmtLemma{su\-mba\-ta}
\fmtCitation{(oku)\-su\-mba\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-mba\-ti\-re}
\fmtSense{to hug, to embrace}
\fmtExampleRL{Omu\-sai\-za asu\-mba\-ti\-re omwi\-raa\-mwe mu nge\-ri ya ki\-nya\-nko\-re.}
\fmtExampleEN{A man has hugged his fellow man in Ankore style.}

\fmtLemma{su\-mbe}
\fmtCitation{\textnormal{(e.g.}\,mu\-su\-mbe, ki\-su\-mbe, mi\-su\-mbe\textnormal{)}}
\fmtPOS{adj}
\fmtSubentry(a)
\fmtSense{cooked}
\fmtExampleRL{Eme\-re eni gi\-su\-mbe.}
\fmtExampleEN{This food is cooked.}
\fmtSubentry(b)
\fmtSense{boiled}
\fmtExampleRL{Nywa amai\-zi aga\-su\-mbe go\-nkai.}
\fmtExampleEN{Drink only boiled water.}

\fmtLemma{su\-mbee\-ta}
\fmtCitation{(oku)\-su\-mbee\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-mbee\-te\-re}
\fmtSense{to comfort, to console}
\fmtExampleRL{Su\-mbee\-ta omwa\-na ata\-ku\-nga.}
\fmtExampleEN{Comfort that child so that it does not cry.}

\fmtLemma{su\-mbii\-ra}
\fmtCitation{(oku)\-su\-mbii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-mbi\-rii\-re}
\fmtSense{to prepare (a girl) for marital life (a series of functions traditionally performed by the paternal aunt and sister-in-law of the bride, including instructions)}
\fmtExampleRL{Omwa\-la ba\-mu\-su\-mbii\-ra nga asi\-gai\-rye\-yo ebi\-ro bi\-doo\-li aswe\-re\-bwe.}
\fmtExampleEN{A girl is prepared for marital life shortly before marriage.}

\fmtLemma{su\-mbi\-rwa}
\fmtCitation{(oku)\-su\-mbi\-rwa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-mbii\-rwe}
\fmtSense{(of a woman) to marry}
\fmtExampleRL{OJa\-ne asu\-mbii\-rwe omu\-sai\-za omu\-sai.}
\fmtExampleEN{Jane has married a good man.}

\fmtLemma{su\-mbi\-sya}
\fmtCitation{(oku)\-su\-mbi\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-mbi\-si\-rye}
\fmtSense{to compliment (the cook for preparing food after eating)}
\fmtExampleRL{Omu\-sai\-za alii\-re bai\-tu ta\-su\-mbi\-si\-rye mu\-ka\-li waa\-mwe.}
\fmtExampleEN{The man has eaten but has not complimented his wife for cooking.}

\fmtLemma{su\-mbu\-ka\-na}
\fmtCitation{(oku)\-su\-mbu\-ka\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{su\-mbu\-kai\-ne}
\fmtSense{to be bothered}
\fmtExampleRL{Ti\-nta\-ka ku\-su\-mbu\-ka\-na n’a\-bi\-ntu bi\-ta\-li\-na ma\-ku\-ru.}
\fmtExampleEN{I don’t want to be bothered with things that are useless.}

\fmtLemma{su\-mbu\-lu\-ku\-ka}
\fmtCitation{(oku)\-su\-mbu\-lu\-ku\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{su\-mbu\-lu\-ku\-ki\-re}
\fmtSense{to break (down), to become inoperative}
\fmtExampleRL{Ekyo\-to kya gaa\-li ya\-nge ki\-su\-mbu\-lu\-ku\-ki\-re.}
\fmtExampleEN{The front derailleur of my bicycle has broken.}

\fmtLemma{su\-mbuu\-sa}
\fmtCitation{(e)su\-mbuu\-sa}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{samosa (triangular fried savory pastry filled with vegetables or meat)}
\fmtExampleRL{Mpee\-rya\-yo esu\-mbuu\-sa ibi\-ri.}
\fmtExampleEN{Give me two samosas, please.}

\fmtLemma{su\-mi\-ka}
\fmtCitation{(oku)\-su\-mi\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-mi\-ki\-re}
\fmtSubentry(a)
\fmtSense{to knot on, to fasten on (an item of barkcloth clothing), to wear (a dress)}
\fmtSubentry(b)
\fmtSense{to install (a heir)}
\fmtExampleRL{Tu\-ku\-su\-mi\-ki\-ra mu\-si\-ka wa ngu\-li Ka\-to\-ngo\-le.}
\fmtExampleEN{We are installing an heir for the late Katongole.}

\fmtLemma{su\-mi\-rya}
\fmtCitation{(oku)\-su\-mi\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-mii\-rye}
\fmtSense{to motion, to signal with a hand}
\fmtExampleRL{Mu\-su\-mi\-rye\-ku le\-ke ize ani atu\-ya\-mbe\-ku.}
\fmtExampleEN{Signal to him to come and help us.}

\fmtLemma{su\-mi\-ta}
\fmtCitation{(oku)\-su\-mi\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-mi\-ti\-re}
\fmtSubentry(a)
\fmtSense{to pierce, to spear, to prick, to stab}
\fmtExampleRL{Su\-mi\-ta oki\-so\-lo oyo atu\-ma\-lii\-re embu\-li.}
\fmtExampleEN{Spear that animal, it has eaten all our goats.}
\fmtSubentry(b)
\fmtFixedexpression{su\-mi\-ta lya mpaa\-li}
\fmtSense{to spear an animal which is not in a trap}
\fmtExampleRL{Omu\-yii\-gi asu\-mi\-ti\-re lya mpaa\-li.}
\fmtExampleEN{The hunter has speared the animal before it fell into the net.}

\fmtLemma{su\-mu\-ka}
\fmtCitation{(oku)\-su\-mu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{su\-mu\-ki\-re}
\fmtSense{to get a hole, to be pierced, to be bored}
\fmtExampleRL{Ese\-fu\-li\-ya ya\-nge esu\-mu\-ki\-re.}
\fmtExampleEN{My saucepan got a hole.}

\fmtLemma{su\-mu\-la}
\fmtCitation{(oku)\-su\-mu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-mwi\-re}
\fmtSense{to punch, to pierce, to make a hole in}
\fmtExampleRL{Omu\-su\-maa\-li gu\-su\-mu\-la omu\-pii\-ra gwa gaa\-li.}
\fmtExampleEN{A nail punches the bicycle tyre.}

\fmtLemma{su\-na}
\fmtCitation{(oku)\-su\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-ni\-re}
\fmtSubentry(a)
\fmtSense{to pinch}
\fmtExampleRL{Lwa\-ki osu\-na abai\-raa\-wo?}
\fmtExampleEN{Why are you pinching your friends?}
\fmtSubentry(b)
\fmtSense{to play a string musical instrument (e.g.\,guitar)}
\fmtExampleRL{Omu Ugu\-nda mu\-li\-mu aba\-ntu ba\-doo\-li aba\-mai\-te oku\-su\-na egi\-ta.}
\fmtExampleEN{There are a few people in Uganda who know how to play egita.}
\fmtSubentry(c)
\fmtFixedexpression{su\-na (a)ma\-bee\-re}
\fmtSense{(of a girl) to start growing breasts}
\fmtExampleRL{Omwa\-la asu\-ni\-re ama\-bee\-re.}
\fmtExampleEN{The girl has started to grow breasts.}

\fmtLemma{su\-nge\-rya}
\fmtCitation{(oku)\-su\-nge\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-nge\-rei\-rye}
\fmtSense{to peep at, to glance at}
\fmtExampleRL{Nii\-we naa\-ni aku\-nsu\-ngee\-rya?}
\fmtExampleEN{Who is that peeping at me?}

\fmtLemma{su\-nna}
\fmtCitation{(e)su\-nna}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from Ar.}
\fmtSense{sunnah (optional good deed)}
\fmtExampleRL{Aba\-si\-raa\-mu ba\-ta\-ka mu\-no oku\-ko\-la esu\-nna.}
\fmtExampleEN{Muslims are fond of doing sunnahs.}

\fmtLemma{su\-nsa}
\fmtCitation{(o)su\-nsa}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{pumpkin leaf used as food}
\fmtExampleRL{Osu\-nsa ba\-mu\-ko\-la\-mu omu\-ku\-bi ogu\-ru\-ngi\-re.}
\fmtExampleEN{Pumpkin leaves are used to make a sauce mixed with groundnut paste.}

\fmtLemma{su\-nsu\-la}%
\fmtOne{1}
\fmtCitation{(oku)\-su\-nsu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-nswi\-re}
\fmtSubentry(a)
\fmtSense{to comb}
\fmtExampleRL{Ya\-ba osu\-nsu\-le ei\-so\-ke.}
\fmtExampleEN{Go and comb your hair.}
\fmtSubentry(b)
\fmtSense{to select (e.g.\,an employee), to pick out (e.g.\,a winner)}
\fmtExampleRL{Olwa\-lee\-ro oma\-ne\-gya wa Bu\-ruu\-li FM akwa\-ba ku\-nsu\-la\-mu aba\-ko\-li.}
\fmtExampleEN{Today, the manager of Buruuli FM is going to select employees.}
\fmtSubentry(c)
\fmtFixedexpression{su\-nsu\-la pa\-mpa}
\fmtSense{to gin (cotton), separate (cotton) from its seeds}
\fmtExampleRL{Ekyo\-ma ki\-ku\-su\-nsu\-la opa\-mpa oku\-ko\-la\-mu engo\-ye.}
\fmtExampleEN{The ginnery is ginning cotton for making garments.}

\fmtLemma{su\-nsu\-la}%
\fmtOne{2}
\fmtCitation{(oku)\-su\-nsu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-nswi\-re}
\fmtSubentry(a)
\fmtSense{to carry on one's shoulders}
\fmtExampleRL{Asu\-nswi\-re omwa\-na.}
\fmtExampleEN{He is carrying a child on his shoulders.}
\fmtSubentry(b)
\fmtSense{to drive or serve as chauffeur for (the Isabanyala or Isabaruuli)}

\fmtLemma{su\-nsuu\-na}
\fmtCitation{(oku)\-su\-nsuu\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{su\-nsuu\-ni\-re}
\fmtSense{to slurp}
\fmtExampleRL{Twa\-nsu\-nsuu\-ni\-re nga tu\-ku\-lya encwi.}
\fmtExampleEN{We slurp while eating the fish.}

\fmtLemma{su\-pa}
\fmtCitation{(o)su\-pa}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{concentrated local gin (war gin)}
\fmtExampleRL{Osu\-pa omwi\-ngi ai\-ta ng’o\-mu\-nywe\-re.}
\fmtExampleEN{The concentrated local gin can kill if you consume it excessively.}

\fmtLemma{su\-pu}
\fmtCitation{(o)su\-pu}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSource{from En.}
\fmtSense{soup}
\fmtExampleRL{Bu\-li lu\-na\-ku ya\-su\-mbi\-ra\-nga onyi\-na waa\-mwe osu\-pu.}
\fmtExampleEN{Every day, she used to cook soup for her mother.}

\fmtLemma{su\-ree\-rya}
\fmtCitation{(oku)\-su\-ree\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-rei\-rye}
\fmtSubentry(a)
\fmtSense{to swing, to ride a swing, to wave}
\fmtSubentry(b)
\fmtSense{to keep on the go, to cause to move about, to bother}
\fmtExampleRL{Abaa\-na ba\-bbai\-re ba\-ku\-su\-ree\-rya omu\-ka\-li oku\-zwa ma\-mbya.}
\fmtExampleEN{The children have been keeping the woman on the go since morning.}

\fmtLemma{su\-ru\-ka}
\fmtCitation{(oku)\-su\-ru\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{su\-ru\-ki\-re}
\fmtSubentry(a)
\fmtSense{to fly}
\fmtExampleRL{Aka\-nyo\-nyi ka\-su\-ru\-ki\-re\-wo ani.}
\fmtExampleEN{The bird has flown away.}
\fmtSubentry(b)
\fmtSense{to change location, to migrate, to move}
\fmtExampleRL{Omwi\-ka\-li waa\-wo ya\-su\-ru\-ki\-re.}
\fmtExampleEN{The resident of that place migrated.}

\fmtLemma{su\-ru\-kya}
\fmtCitation{(oku)\-su\-ru\-kya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-ru\-ki\-rye}
\fmtSubentry(a)
\fmtSense{to transfer, to transport, to deliver}
\fmtExampleRL{Embu\-li baa\-zi\-su\-ru\-ki\-rye ne ba\-zi\-twa\-la omu faa\-mu.}
\fmtExampleEN{The goats were transferred to the farm.}
\fmtSubentry(b)
\fmtSense{to get rid (of termites by using chemicals)}
\fmtExampleRL{Ya\-su\-ri\-ki\-rye eki\-swa.}
\fmtExampleEN{He got rid of termites from the termite mound.}

\fmtLemma{su\-sa}%
\fmtOne{1}
\fmtCitation{(oku)\-su\-sa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{si\-si\-re}
\fmtSense{(of a baby) to gain weight, to get plump}
\fmtExampleRL{Omwa\-na wa\-nge asu\-si\-re.}
\fmtExampleEN{My baby has gained weight.}

\fmtLemma{su\-sa}%
\fmtOne{2}
\fmtCitation{(oku)\-su\-sa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{si\-si\-re}
\fmtSubentry(a)
\fmtSense{to peel, to shell, to husk}
\fmtExampleRL{Su\-sa ama\-too\-ke ago.}
\fmtExampleEN{Peel those matoke please.}
\fmtSubentry(b)
\fmtSense{to rebuke, to admonish, to reprove, to chastise, to critisise}

\fmtLemma{su\-su\-bu\-ka}
\fmtCitation{(oku)\-su\-su\-bu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{su\-su\-bu\-ki\-re}
\fmtSense{(of skin) to peel off}
\fmtExampleRL{Omu\-ko\-no gwa\-mwe gu\-zi\-mbi\-re era ati gu\-ku\-su\-su\-bu\-ka.}
\fmtExampleEN{His arm is swollen, and the skin is peeling off.}

\fmtLemma{su\-su\-bu\-la}
\fmtCitation{(oku)\-su\-su\-bu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-su\-bwi\-re}
\fmtSense{to remove (tree bark)}
\fmtExampleRL{Su\-su\-bu\-la omu\-saa\-le ogwo tu\-too\-le\-ku olu\-bu\-go.}
\fmtExampleEN{Skin off the outer layer of the tree and we will make bark cloth.}

\fmtLemma{su\-su\-ka}
\fmtCitation{(oku)\-su\-su\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{su\-su\-ki\-re}
\fmtSubentry(a)
\fmtSense{to spill, to overflow, to boil over}
\fmtExampleRL{Ama\-tai ga\-ku\-su\-su\-ka la\-bu\-ka oga\-yi\-ye ku muu\-ro.}
\fmtExampleEN{The milk is spilling, hurry and remove it from the fire.}
\fmtSubentry(b)
\fmtSense{(of skin) to peel off}
\fmtExampleRL{Olu\-su\-su lwa\-mwe lu\-su\-su\-ki\-re.}
\fmtExampleEN{Her skin has peeled off.}

\fmtLemma{su\-su\-la}
\fmtCitation{(oku)\-su\-su\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-swi\-re}
\fmtSubentry(a)
\fmtSense{to peel}
\fmtExampleRL{Su\-su\-la ame\-nju nga okya\-li ku\-ga\-lya.}
\fmtExampleEN{Peel the yellow banana before you eat it.}
\fmtSubentry(b)
\fmtSense{to wipe, to rub, to rub dry, to clean}
\fmtExampleRL{Su\-su\-la entuu\-yo omu bu\-syo bwa\-mu.}
\fmtExampleEN{Wipe the sweat from your forehead.}

\fmtLemma{su\-suu\-la}
\fmtCitation{(oku)\-su\-suu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-swi\-re}
\fmtSense{to clean (esp. a baby)}
\fmtExampleRL{Su\-suu\-la omwa\-na ku\-sai n’o\-kya\-li ku\-mu\-twa\-la mu irwa\-ro.}
\fmtExampleEN{Clean the baby properly before you take her to hospital.}

\fmtLemma{su\-suu\-na}
\fmtCitation{(oku)\-su\-suu\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-suu\-ni\-re}
\fmtSense{to suck}
\fmtExampleRL{Lwa\-ki oku\-su\-suu\-na omu\-so\-koo\-lo ogwo otyo?}
\fmtExampleEN{Why are you sucking the bone marrow like that?}

\fmtLemma{su\-suu\-rya}
\fmtCitation{(oku)\-su\-suu\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-swi\-rye}
\fmtSense{to remove by wiping, to wipe off}
\fmtExampleRL{Su\-suu\-rya ei\-ta\-ka\-li oku ki\-su\-suu\-ryo.}
\fmtExampleEN{Wipe off the soil from the doormat.}

\fmtLemma{su\-ta\-ma}
\fmtCitation{(oku)\-su\-ta\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{su\-ta\-mi\-re}
\fmtSense{to squat, to sit on one's haunches, to sit on one's heels}
\fmtExampleRL{Ki\-bbi omu\-ka\-li oku\-su\-ta\-ma mu bu\-wa\-ngwa bwa Ba\-ga\-nda.}
\fmtExampleEN{According to the customs of the Baganda, it is bad for a woman to squat.}

\fmtLemma{suu\-bi}
\fmtCitation{(e)suu\-bi}
\fmtPOS{n}%
\fmtNC{5}
\fmtSubentry(a)
\fmtSense{chance, possibility}
\fmtExampleRL{Nko\-be\-ra oba nga wa\-kya\-li\-wo esuu\-bi eri\-ndi.}
\fmtExampleEN{Tell me whether there is still another chance.}
\fmtSubentry(b)
\fmtSense{hope, optimism}
\fmtExampleRL{Nkya\-li\-na esuu\-bi li\-ngi okwi\-ra\-yo ku iso\-me\-ro.}
\fmtExampleEN{I still have much hope to continue with my studies.}

\fmtLemma{suu\-bi\-ra}
\fmtCitation{(oku)\-suu\-bi\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{suu\-bi\-rii\-re}
\fmtSubentry(a)
\fmtSense{to hope, to intend}
\fmtExampleRL{Nku\-suu\-bi\-ra okwi\-za ema\-mbya.}
\fmtExampleEN{I am hoping to come tomorrow.}
\fmtSubentry(b)
\fmtSense{to anticipate, to sense, to expect}
\fmtExampleRL{Nku\-suu\-bi\-ra\-yo aka\-bbi omu nyu\-mba.}
\fmtExampleEN{I sense danger in the house.}

\fmtLemma{suu\-bi\-rii\-ra}
\fmtCitation{(oku)\-suu\-bi\-rii\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{suu\-bi\-rii\-re}
\fmtSense{to suspect, to guess, to estimate}
\fmtExampleRL{Naa\-ni gw’o\-ku\-suu\-bi\-rii\-ra okwi\-bba ente zaa\-mu?}
\fmtExampleEN{Whom do you suspect of stealing your cows?}

\fmtLemma{suu\-bi\-rya}
\fmtCitation{(oku)\-suu\-bi\-rya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{suu\-bii\-rye}
\fmtSense{to promise}
\fmtExampleRL{Aba\-na bya\-bu\-fu\-zi ba\-suu\-bi\-rya ebi\-ntu bi\-ngi ni ba\-ta\-bi\-kii\-rya.}
\fmtExampleEN{Politicians promise many things, which they never fulfil.}

\fmtLemma{suu\-bi\-za}
\fmtCitation{(oku)\-suu\-bi\-za}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{suu\-bii\-ze}
\fmtSense{to make a promise}
\fmtExampleRL{Lwa\-ki osuu\-bi\-za\-nga n’o\-ta\-yi\-kii\-rya.}
\fmtExampleEN{Why do you make promises, which you never fulfil?}

\fmtLemma{suu\-bu\-la}
\fmtCitation{(oku)\-suu\-bu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{suu\-bwi\-re}
\fmtSense{to trade, to deal (in)}
\fmtExampleRL{Oi\-te wa\-nge asuu\-bu\-la nte.}
\fmtExampleEN{My father trades in cattle.}

\fmtLemma{suu\-kii\-ra}
\fmtCitation{(oku)\-suu\-kii\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{suu\-ki\-rii\-re}
\fmtSense{to rage, to be angry, to be furious}
\fmtExampleRL{Ti ki\-sai omwe\-be\-mbe\-zi oku\-suu\-kii\-ra awa\-ta\-li nso\-nga.}
\fmtExampleEN{It is not good for a leader to become angry without a good reason.}

\fmtLemma{suu\-la}
\fmtCitation{(e)suu\-la}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{chapter}
\fmtExampleRL{Swe\-ku\-la esuu\-la ya ka\-sa\-tu omu Ma\-ta\-yo.}
\fmtExampleEN{Open the third chapter in Matthew.}

\fmtLemma{suu\-muu\-ka}
\fmtCitation{(oku)\-suu\-muu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}/\fmtTrans{tr}
\fmtPFV{suu\-muu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to grow, to get bigger, to grow up, to mature}
\fmtExampleRL{Enya\-na za\-nge zi\-suu\-muu\-ki\-re\-mu nku\-so\-bo\-la ati oku\-zi\-tu\-nda.}
\fmtExampleEN{My calves have grown up I can now sell them.}
\fmtSubentry(b)
\fmtSense{to emit (gas, to air), to discharge, to release}
\fmtExampleRL{mo\-to\-ka ezi\-kai\-re zi\-ku\-suu\-muu\-ka ekii\-ka eki\-bbi.}
\fmtExampleEN{Old cars are emitting bad gas.}

\fmtLemma{suu\-nya}
\fmtCitation{(oku)\-suu\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{suu\-nyi\-re}
\fmtSense{to whip, to cane}
\fmtExampleRL{Omwa\-na na\-mu\-suu\-nyi\-re obu\-suu\-nyo bu\-sa\-tu bwo\-nkai.}
\fmtExampleEN{I only whipped the child three times.}

\fmtLemma{suu\-ta}%
\fmtOne{1}
\fmtCitation{(oku)\-suu\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{su\-lii\-re}
\fmtSense{to compliment, to praise, to express approval of}
\fmtExampleRL{Bu\-li mu\-byai\-re asuu\-ta mu\-no omwa\-na waa\-mwe.}
\fmtExampleEN{Every parent praises his child.}

\fmtLemma{suu\-ta}%
\fmtOne{2}
\fmtCitation{(oku)\-suu\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{suu\-ti\-re}
\fmtSense{to score a goal (e.g.\,in netball, volley ball)}
\fmtExampleRL{We nii\-we okwa\-ba oku\-tu\-suu\-ti\-ra omu\-pii\-ra.}
\fmtExampleEN{You are the one who is going to score a goal for us.}

\fmtLemma{su\-waa\-ni}
\fmtCrossreference{\textnormal{see} saani}

\fmtLemma{swa}
\fmtCitation{(oku)\-swa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{swi\-re}
\fmtSource{from archaic}
\fmtSense{to die, to pass away, to perish}
\fmtExampleRL{Bu\-li mwa\-ka aba\-ntu ba\-swe\-ra omu bu\-ta\-ndwa.}
\fmtExampleEN{Every year many people die in accidents.}

\fmtLemma{swa\-bu\-la}
\fmtCitation{(oku)\-swa\-bu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{swa\-bwi\-re}
\fmtSense{to beat, to cane, to strike}
\fmtExampleRL{Ya\-swa\-bwi\-re omwa\-na obwi\-go ku\-bba ya\-so\-me\-re ku\-bbi.}
\fmtExampleEN{He caned the child because he did not perform well in his studies.}

\fmtLemma{swa\-ga}
\fmtCitation{(oku)\-swa\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{swa\-gi\-re}
\fmtSense{to look for, to search for}
\fmtExampleRL{Nkwa\-ba ku\-swa\-ga saa\-wa ya\-nge ya\-gwi\-re omu mai\-zi.}
\fmtExampleEN{I am looking for my watch, which fell into water.}

\fmtLemma{swa\-ga\-na}
\fmtCitation{(oku)\-swa\-ga\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{swa\-gai\-ne}
\fmtSense{to look for, to search for}
\fmtExampleRL{Ka nku\-swa\-ga\-ni\-re omu\-sai\-za eya\-ku\-swe\-ra.}
\fmtExampleEN{Let me search for a man to marry you.}

\fmtLemma{swa\-ga\-nii\-rya}
\fmtCitation{(oku)\-swa\-ga\-nii\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{swa\-ga\-nii\-rye, swa\-ga\-ni\-rii\-rye}
\fmtSense{to do research, to investigate}
\fmtExampleRL{Aku\-swa\-ga\-ni\-rii\-rya oku omu\-ba\-zi gwa ku\-ge\-ma mu\-su\-ja.}
\fmtExampleEN{He is doing research on malaria vaccination.}

\fmtLemma{swa\-gi}
\fmtCitation{(o)swa\-gi}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{device used to search for things underwater}
\fmtExampleRL{Ba\-guu\-re oswa\-gi ei\-zo.}
\fmtExampleEN{They bought a device used to search for drowned things yesterday.}

\fmtLemma{swa\-gi\-rya}
\fmtCitation{(oku)\-swa\-gi\-rya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{swa\-gii\-rye}
\fmtSubentry(a)
\fmtSense{to make sound of footsteps, to make noise (by people or animals moving)}
\fmtExampleRL{Nku\-wuu\-ra aba\-ntu nga ba\-ku\-swa\-gi\-rya.}
\fmtExampleEN{I hear people walking.}
\fmtSubentry(b)
\fmtSense{to perform a ritual which involves going to the bush to invoke spirits (mainly performed by traditional priests)}

\fmtLemma{swa\-ha\-bba}
\fmtCitation{(o)swa\-ha\-ba}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from Ar.}
\fmtSubentry(a)
\fmtSense{friend}
\fmtExampleRL{Oswa\-ha\-bba alwai\-re omu\-su\-ja.}
\fmtExampleEN{My friend is suffering from malaria.}
\fmtSubentry(b)
\fmtSense{person who saw and lived with Mohammed}

\fmtLemma{swai\-tu\-lya}
\fmtCitation{(oku)\-swai\-tu\-lya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{swai\-twi\-rye}
\fmtSense{to husk, to remove husks from}
\fmtExampleRL{Swai\-tu\-lya obwi\-ta tu\-bu\-twa\-le oku kyo\-ma.}
\fmtExampleEN{Husk the millet so that we can take it to the grinding mill.}

\fmtLemma{swa\-kaa\-nia}
\fmtCitation{(oku)\-swa\-kaa\-nia}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{swa\-kai\-nie}
\fmtSense{to rustle}
\fmtExampleRL{Nkuu\-ra ebi\-ku\-swa\-kaa\-nia awo.}
\fmtExampleEN{I hear something rustling.}

\fmtLemma{swa\-ki\-ra}
\fmtCitation{(oku)\-swa\-ki\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{swa\-kii\-re}
\fmtSubentry(a)
\fmtSense{(of a cobra) to spread the hood}
\fmtExampleRL{Oncwe\-ra ya\-swa\-kii\-re ni ya\-bwoi\-ne abaa\-na.}
\fmtExampleEN{The cobra spread its hood when it saw the children.}
\fmtSubentry(b)
\fmtSense{to be angry, to rage, to be furious}
\fmtExampleRL{Omu\-ka\-li aswa\-kii\-re nka ompi\-ri.}
\fmtExampleEN{The woman is raging like a snake.}

\fmtLemma{swa\-la}
\fmtCitation{(oku)\-swa\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{swai\-re}
\fmtSense{to be embarrassed, to be ashamed}
\fmtExampleRL{Oi\-za ku\-swa\-la nga to\-su\-nswi\-re iso\-ke.}
\fmtExampleEN{You will get embarrassed if you do not comb your hair.}

\fmtLemma{swa\-li\-sya}
\fmtCitation{(oku)\-swa\-li\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{swa\-li\-si\-rye}
\fmtSense{to bring shame to, to embarrass, to disgrace}
\fmtExampleRL{Omwi\-raa\-mu omu\-swa\-li\-si\-rye bu\-ndi.}
\fmtExampleEN{You have brought shame to your colleague a lot.}

\fmtLemma{swa\-lu\-ka}
\fmtCitation{(oku)\-swa\-lu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{swa\-lu\-ki\-re}
\fmtSense{to be very ashamed, to be extremely embarrassed}
\fmtExampleRL{Oswai\-re mu\-no n’o\-swa\-lu\-ka olwa ku\-ku\-bba mu\-ka\-li waa\-mu.}
\fmtExampleEN{You should be extremely embarrassed because of beating your wife.}

\fmtLemma{swa\-nku\-la}
\fmtCitation{(oku)\-swa\-nku\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{swa\-nkwi\-re}
\fmtSense{to munch, to eat noisily, to chew noisily}
\fmtExampleRL{Lwa\-ki olya oku\-swa\-nku\-la nk’o\-mbe\-re\-ge?}
\fmtExampleEN{Why do you munch like a pig?}

\fmtLemma{swa\-nyuu\-la}
\fmtCitation{(oku)\-swa\-nyuu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{swa\-nywi\-re}
\fmtSense{to beat, to cane, to flog, to whip}
\fmtExampleRL{Swa\-nyuu\-la omwa\-na waa\-mu ale\-ke omu\-ze ogwo.}
\fmtExampleEN{Cane your child to stop that bad habit.}

\fmtLemma{swa\-swa\-na}
\fmtCitation{(oku)\-swa\-swa\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{swa\-wai\-ne}
\fmtSubentry(a)
\fmtSense{to be estranged, to be alienated}
\fmtExampleRL{Ya\-swa\-swai\-ne n’a\-le\-ka\-wo ama\-ka olwa nkaa\-ya\-na.}
\fmtExampleEN{He became estranged from his family after the argument.}
\fmtSubentry(b)
\fmtSense{to divorce, to get divorced}
\fmtExampleRL{Omu\-ka\-li aswa\-swai\-ne ku\-bba oi\-baa\-mwe mwe\-nzi.}
\fmtExampleEN{The woman has divorced her husband because he is lustful.}

\fmtLemma{swa\-ya}
\fmtCitation{(e)swa\-ya}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{way of expressing bliss (e.g.\,by performing a dance, ululation)}

\fmtLemma{swa\-za}
\fmtCitation{(oku)\-swa\-za}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{swa\-zi\-re, swai\-ze}
\fmtSense{to embarrass, to humiliate}
\fmtExampleRL{Omu\-koi ya\-nswa\-zi\-re ni ya\-gai\-ne oku\-lya e bu\-koi.}
\fmtExampleEN{The brother-in-law embarrassed me when he refused to dine with the in-laws.}

\fmtLemma{swe\-ka}
\fmtCitation{(oku)\-swe\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{swe\-ke\-re}
\fmtSubentry(a)
\fmtSense{to cover, to coat}
\fmtExampleRL{Wee\-swe\-ke ebbu\-la\-ngi\-ti.}
\fmtExampleEN{Cover yourself with the blanket.}
\fmtSubentry(b)
\fmtSense{to roof}
\fmtExampleRL{Tu\-swe\-ke enyu\-mba n’ei\-su\-bi.}
\fmtExampleEN{Let's roof the house with grass.}
\fmtSubentry(c)
\fmtSense{to give a dress (to a woman who has lost her husband)}
\fmtExampleRL{Omu\-ka\-li ba\-mu\-jwe\-ke\-re ni ya\-fee\-rei\-rwe oi\-baa\-mwe.}
\fmtExampleEN{The woman was given a dress when she lost her husband.}

\fmtLemma{swe\-ku\-la}
\fmtCitation{(oku)\-swe\-ku\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{swe\-kwi\-re}
\fmtSubentry(a)
\fmtSense{to uncover}
\fmtExampleRL{Swe\-ku\-la omwa\-na empii\-ta gi\-ngi.}
\fmtExampleEN{Uncover the child, it is too hot.}
\fmtSubentry(b)
\fmtSense{to open}
\fmtExampleRL{Swe\-ku\-la esuu\-la ya ka\-sa\-tu omu Ma\-ta\-yo.}
\fmtExampleEN{Open the third chapter in Matthew.}
\fmtSubentry(c)
\fmtSense{to undress}

\fmtLemma{swe\-ra}
\fmtCitation{(oku)\-swe\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{swei\-re}
\fmtSense{(of a man) to wed, to marry}
\fmtExampleRL{Nswei\-re omu\-ka\-li owo\-ndi.}
\fmtExampleEN{I have married another wife.}

\fmtLemma{swe\-rwa}
\fmtCitation{(oku)\-swe\-rwa}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{swei\-rwe}
\fmtSense{(of a woman) to marry}
\fmtExampleRL{Na\-nsi\-ko\-mbi ya\-swei\-rwe omu\-sai\-za omu\-sai.}
\fmtExampleEN{Nansikombi married a nice man.}

\fmtLemma{swe\-ta}%
\fmtOne{1}
\fmtCrossreference{\textnormal{see also} siyeta}
\fmtCitation{(e)swe\-ta}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSubentry(a)
\fmtSense{theatre (as a venue)}
\fmtSubentry(b)
\fmtSense{operating theatre, operating room}
\fmtExampleRL{Omu\-ka\-li ba\-mu\-swe\-ke\-re ni ya\-fee\-rei\-rwe oi\-baa\-mwe.}
\fmtExampleEN{The patient has been taken to the operating theatre for a surgical operation.}

\fmtLemma{swe\-ta}%
\fmtOne{2}
\fmtCitation{(e)swe\-ta}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{sweater}
\fmtExampleRL{Ta\-ndi\-ka oku\-ka\-mba eswe\-ta.}
\fmtExampleEN{Start knitting the sweater.}

\fmtLemma{swi\-ci}
\fmtCitation{(e)swi\-ci}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{switch, button}
\fmtExampleRL{Taa\-ku eswi\-ci eri ema\-be\-ga wa ka\-ba\-da.}
\fmtExampleEN{Turn on the switch behind the cupboard.}

\fmtLemma{swii}
\fmtPOS{ideo}
\fmtSense{all}
\fmtExampleRL{Eme\-re aje\-ye\-re swii!}
\fmtExampleEN{He has swept out all the food.}

\fmtLemma{swoi\-ta\-na}
\fmtCitation{(oku)\-swoi\-ta\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{swoi\-tai\-ne}
\fmtSense{to quarrel incessantly}
\fmtExampleRL{Abaa\-la abo ba\-ku\-swoi\-ta\-na\-mbe.}
\fmtExampleEN{Those girls are just quarrelling incessantly.}

\fmtLemma{sya}%
\fmtOne{1}
\fmtCitation{(oku)\-sya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{syei\-re}
\fmtSense{to grind}
\fmtExampleRL{Nku\-sya bwi\-ta bwa ku\-tu\-nda.}
\fmtExampleEN{I am grinding millet for sale.}

\fmtLemma{sya}%
\fmtOne{2}
\fmtPOS{ideo}
\fmtSense{totally}
\fmtExampleRL{Sya! Nje ngai\-ne by’o\-ku\-nko\-ba.}
\fmtExampleEN{I have totally objected to what you are telling me.}

\fmtLemma{sya\-ta\-sya\-ta}%
\fmtOne{1}
\fmtCitation{(oku)\-sya\-ta\-sya\-ta}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sya\-ti\-re\-sya\-ti\-re}
\fmtSense{to wander aimlelsly, to walk aimlessly or blindly, to walk in confusion}
\fmtExampleRL{Omu\-si\-ri\-ka\-le ya\-ku\-bbi\-re ama\-sa\-si ni ba\-sya\-ta\-sya\-ta.}
\fmtExampleEN{The soldier fired bullets and they ran in confusion.}

\fmtLemma{sya\-ta\-sya\-ta}%
\fmtOne{2}
\fmtCitation{(e)sya\-ta\-sya\-ta}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{very hard times}
\fmtExampleRL{Mu\-sai\-za wai\-tu ali mu sya\-ta\-sya\-ta.}
\fmtExampleEN{The man is having a very hard time.}

\fmtLemma{sya\-tuu\-ka}
\fmtCitation{(oku)\-sya\-tuu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sya\-tuu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to walk hurriedly, to rush}
\fmtExampleRL{Omu ki\-fo kya ku\-ta\-mbu\-la, osya\-tuu\-ka\-mbe.}
\fmtExampleEN{Instead of walking normally you are just rushing.}
\fmtSubentry(b)
\fmtSense{to wander at random, to walk aimlessly, to roam}
\fmtSubentry(c)
\fmtSense{to trespass}
\fmtExampleRL{Mu\-ta\-sya\-tuu\-ki\-ra mu ndu\-mi\-ro ya\-nge.}
\fmtExampleEN{Don't trespass in my garden.}

\fmtLemma{sye\-ra}
\fmtCitation{(oku)\-sye\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{syei\-re}
\fmtSense{to sell (expensively), to hike (price)}
\fmtExampleRL{Aku\-sye\-ra omu\-ba\-zi gwa\-mu.}
\fmtExampleEN{He sells his medicine very expensively.}

\fmtLemma{sye\-sye\-tu\-ka}
\fmtCitation{(oku)\-sye\-sye\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sye\-sye\-tu\-ki\-re}
\fmtSubentry(a)
\fmtSense{to move over, to make room for someone to sit}
\fmtExampleRL{Sa\-ba aba\-ntu abo ba\-sye\-sye\-tu\-ke.}
\fmtExampleEN{Ask those people to move over.}
\fmtSubentry(b)
\fmtSense{to slip, to slide, to fall off, to fall down, to become loose}
\fmtSubentry(c)
\fmtSense{to make a slow progress}
\fmtExampleRL{Omwa\-na asye\-sye\-tu\-ki\-re mu bya\-ba\-nye\-ga yaa\-mwe.}
\fmtExampleEN{The child has made some slow progress in his studies.}

\fmtLemma{sye\-tu\-ka}%
\fmtOne{1}
\fmtCitation{(oku)\-sye\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sye\-tu\-ki\-re}
\fmtSense{to fight out}
\fmtExampleRL{Ni\-wa\-le\-me\-ra\-ku tu\-kwi\-za ku\-sye\-tu\-ka.}
\fmtExampleEN{If you insist, we will fight it out the hard way.}

\fmtLemma{sye\-tu\-ka}%
\fmtOne{2}
\fmtCitation{(oku)\-sye\-tu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{sye\-tu\-ki\-re}
\fmtSense{to move along slowly, to move aside (in sitting), to slide over, to slip over}
\fmtExampleRL{Tu\-sye\-tu\-ke mpo\-la\-mpo\-la nga ni\-mwo tu\-kwi\-ge\-ree\-ra eme\-re.}
\fmtExampleEN{Let us move along slowly towards the food.}

\fmtLemma{sye\-tu\-la}
\fmtCitation{(oku)\-sye\-tu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{sye\-twi\-re}
\fmtSense{to shift, to move over (a heavy object)}
\fmtExampleRL{Twa\-be tu\-sye\-tu\-le ensu\-wa ya mai\-zi.}
\fmtExampleEN{Let us go and move over that pot of water.}

\fmtLemma{syo\-la}
\fmtCitation{(oku)\-syo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{syoi\-re}
\fmtSense{to bewitch (by performing night dancing), to put a spell on, to curse}
\fmtExampleRL{Ei\-zo twa\-kwai\-te omu\-syo\-li ng’a\-ku\-syo\-la endu\-mi\-ro ya mu\-li\-ri\-rwa\-na waa\-mwe.}
\fmtExampleEN{Yesterday we arrested a night dancer bewitching his neighbour’s garden.}

\fmtLemma{syo\-ma}
\fmtCitation{(oku)\-syo\-ma}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{syo\-me\-re}
\fmtSubentry(a)
\fmtSense{to come for, to collect, to pick up}
\fmtExampleRL{Syo\-ma omwa\-na abbai\-re ani ebi\-ro bi\-ngi.}
\fmtExampleEN{Pick up that child, he's been here already for too long.}
\fmtSubentry(b)
\fmtSense{to fetch (e.g.\,water)}
\fmtExampleRL{Baa\-bi\-re ku izu\-ba ku\-syo\-ma mai\-zi.}
\fmtExampleEN{They have gone to the well to fetch water.}

\fmtLemma{syo\-too\-ki}
\fmtCitation{\textnormal{(e.g.}\,ki\-syo\-too\-ki, mu\-syo\-too\-ki, ba\-syo\-too\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{flat}
\fmtExampleRL{Lwa\-ki oku\-ta\-ka omu\-ntu omu\-syo\-too\-ki wa ma\-ta\-ko?}
\fmtExampleEN{Why do you like someone with a flat bottom?}
\end{letter}