\begin{letter}{Y}

\fmtLemma{ya}%
\fmtOne{1}
\fmtCitation{(oku)\-ya}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yii\-re}
\fmtSubentry(a)
\fmtSense{to be burnt}
\fmtExampleRL{Enyu\-mba ya Bya\-mpo\-ra eyii\-re, ta\-si\-gai\-rye ka\-ntu koo\-na.}
\fmtExampleEN{Byampora’s house got burnt up. He was left with nothing.}
\fmtSubentry(b)
\fmtSense{to boil}
\fmtExampleRL{Ama\-zi ga\-yii\-re.}
\fmtExampleEN{The water has boiled.}
\fmtSubentry(c)
\fmtSense{(of food) to be ready}
\fmtExampleRL{Eme\-re eyii\-re.}
\fmtExampleEN{The food is ready.}
\fmtSubentry(d)
\fmtSense{to be exhausted}
\fmtExampleRL{Wu\-mu\-la\-mu okwi\-za ku\-ya mu\-no.}
\fmtExampleEN{Rest please, you will get exhausted.}
\fmtSubentry(e)
\fmtSense{to be about to give birth}
\fmtExampleRL{Omu\-ka\-li wa Mba\-ngi\-re ayii\-re, akwa\-ba ku\-bya\-la.}
\fmtExampleEN{Mbangire’s wife is about to give birth.}
\fmtSubentry(f)
\fmtSense{to engage in (\eg a shady deal)}
\fmtExampleRL{OYa\-kaa\-na aku\-ya abai\-ra baa\-mwe.}
\fmtExampleEN{John falsely engages his friends in shady deals.}

\fmtLemma{ya}%
\fmtOne{2}
\fmtCitation{\textnormal{(e.g.}\,mu\-ya, ka\-ya\textnormal{)}}
\fmtPOS{adj}
\fmtSense{new}
\fmtExampleRL{OSe\-mba\-tya ya\-guu\-re aka\-ru\-ndi\-ru\-ndi aka\-ya.}
\fmtExampleEN{Sembatya bought a new bicycle peddle-rod.}

\fmtLemma{yaa\-bwe}
\fmtCrossreference{\textnormal{see} abwe}

\fmtLemma{yaa\-jwa}
\fmtCitation{\textnormal{(e.g.}\,ki\-yaa\-ja, mu\-yaa\-ja, zi\-yaa\-ja\textnormal{)}}
\fmtPOS{adj}
\fmtSense{half-dry (\eg fish, wood, coffee),}
\fmtExampleRL{To\-ta mwa\-ni zi\-yaa\-jwa mu nsa\-wo.}
\fmtExampleEN{Don't put half-dry coffee into bags.}

\fmtLemma{yaa\-ka}
\fmtCitation{\textnormal{(e.g.}\,mu\-yaa\-ka, gi\-yaa\-ka\textnormal{)}}
\fmtPOS{adj}
\fmtSense{new}
\fmtExampleRL{Omu\-so\-me\-sya ya\-guu\-re ata\-laa\-si gi\-yaa\-ka.}
\fmtExampleEN{The teacher bought a new atlas.}

\fmtLemma{yaa\-mu}
\fmtCrossreference{\textnormal{see} amu}

\fmtLemma{yaa\-nge}
\fmtCrossreference{\textnormal{see} ange}
\fmtCrossreference{\textnormal{see} amwe\textnormal{{\fmtOne{1}}}}

\fmtLemma{yaa\-nywe}
\fmtCrossreference{\textnormal{see} nywe}

\fmtLemma{yaa\-si\-yaa\-si}
\fmtCitation{(o)yaa\-si\-yaa\-si}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{tension, anxiety, pressure, worry}
\fmtExampleRL{Tu\-li ku yaa\-si\-yaa\-si wa kwi\-rya\-yo abaa\-na ku ma\-so\-me\-ro.}
\fmtExampleEN{We are anxious of sending children to school.}

\fmtLemma{Yaa\-we}
\fmtCitation{(o)Yaa\-we}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{Yahweh}
\fmtExampleRL{OYaa\-we ni\-ye Ka\-nca.}
\fmtExampleEN{Yahweh is God.}

\fmtLemma{yaa\-ya}
\fmtCitation{(o)yaa\-ya}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{babysitter, nanny}
\fmtExampleRL{ONa\-ndi\-ra yaa\-ya wa baa\-na ba Jo\-hn.}
\fmtExampleEN{Nandira is John’s babysitter.}

\fmtLemma{yaa\-ye}
\fmtPOS{interj}
\fmtSense{cry of pain, grief or terror, cry for help}
\fmtExampleRL{Yaa\-ye! Ba\-ku\-nyi\-ta!}
\fmtExampleEN{Help! I am dying!}

\fmtLemma{yaa\-yu\-ka}
\fmtCitation{(oku)\-yaa\-yu\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yaa\-yu\-ki\-re}
\fmtSense{to become a rogue, to become a villain}
\fmtExampleRL{Omwo\-jo wa Bya\-mpo\-ra ayaa\-yu\-ki\-re.}
\fmtExampleEN{Byampora’s boy has become a rogue.}

\fmtLemma{ya\-mba}
\fmtCitation{(oku)\-ya\-mba}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ya\-mbi\-re}
\fmtSense{to help, to assist, to lend a hand to}
\fmtExampleRL{Ama\-lu\-su ga\-bbaa\-mu eki\-ru\-ngo eki\-ya\-mba oku\-sya eme\-re.}
\fmtExampleEN{The saliva contains enzymes which help to digest food.}

\fmtLemma{ya\-na}
\fmtCitation{(oku)\-ya\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{ya\-ni\-re}
\fmtSense{to blame}

\fmtLemma{ya\-nge}
\fmtCrossreference{\textnormal{see} -ange}

\fmtLemma{ya\-zi\-rwai\-ne\-ku}
\fmtCitation{(e)ya\-zi\-rwai\-ne\-ku}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSense{war veteran}
\fmtExampleRL{Eya\-zi\-rwai\-ne\-ku omu Lu\-wee\-ro afii\-re ta\-fu\-ni\-re ka\-sii\-mo.}
\fmtExampleEN{The Luweero war veteran has passed on before being paid his gratuity.}

\fmtLemma{ye}
\fmtPOS{pro}
\fmtSense{he/she (class 1 independent pronoun)}
\fmtExampleRL{Omu\-sa\-ma\-li\-ya omu\-sai ye mu\-ntu aju\-na aba\-ndi.}
\fmtExampleEN{A good Samaritan is a person who helps others.}

\fmtLemma{yee\-kaa\-lu}
\fmtCitation{(e)yee\-kaa\-lu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{temple, church}
\fmtExampleRL{Tu\-kwa\-ba mu yee\-kaa\-lu ku\-ra\-mya.}
\fmtExampleEN{We are going to the temple to pray.}

\fmtLemma{yee\-na}
\fmtCrossreference{\textnormal{see} ena}
\fmtPOS{pro}
\fmtSense{also she/he, she/he too, he/she as well}
\fmtExampleRL{Yee\-na ali mu Ka\-mpa\-la.}
\fmtExampleEN{She is also in Kampala.}

\fmtLemma{Ye\-ru\-saa\-le\-mi}
\fmtCitation{(e)Ye\-ru\-saa\-le\-mi}
\fmtPOS{n}%
\fmtNC{9}
\fmtSense{Jerusalem}
\fmtExampleRL{OYe\-su ya\-ku\-lii\-re Ye\-ru\-saa\-le\-mi.}
\fmtExampleEN{Jesus grew up in Jerusalem.}

\fmtLemma{Ye\-su}
\fmtCitation{(o)Ye\-su}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{Jesus}
\fmtExampleRL{OYe\-su mwa\-na wa Ka\-nca.}
\fmtExampleEN{Jesus is a son of God.}

\fmtLemma{ye\-te}
\fmtPOS{conj}
\fmtSense{how about}
\fmtExampleRL{Ye\-te oyo abbai\-re ki?}
\fmtExampleEN{How about that one, what is happening to him?}

\fmtLemma{yii}
\fmtPOS{interj}
\fmtSense{expression of surprise}
\fmtExampleRL{Yii. Obwai\-rai ti\-mai\-te.}
\fmtExampleEN{My goodness, I didn’t know.}

\fmtLemma{yii\-ya}
\fmtCitation{(oku)\-yii\-ya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yii\-yi\-re}
\fmtSense{to innovate, to come up with a new idea}
\fmtExampleRL{Abee\-be\-mbe\-zi ba\-li oku\-yii\-ya ama\-zi\-ra aga\-zwa\-mu empii\-ya.}
\fmtExampleEN{Leaders should come up with new ways of making money.}

\fmtLemma{yi\-kwe\-ta}
\fmtCitation{(e)yi\-kwe\-ta}
\fmtPOS{n}%
\fmtNC{5}
\fmtSource{from En.}
\fmtSense{equator}
\fmtExampleRL{Eyi\-kwe\-ta eya\-kaa\-ra mu ta\-wu\-ni ya Ka\-ya\-bwe.}
\fmtExampleEN{The equator passes through Kayabwe town in Uganda.}

\fmtLemma{yi\-mba\-ga\-ta\-nya}
\fmtCitation{(o\-ku\-yi\-mba\-ga\-ta\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yi\-mba\-ga\-ta\-ni\-rye}
\fmtSense{to multitask, to do two or more things at once, to carry many different things at one ago}
\fmtExampleRL{To\-yi\-mba\-ga\-ta\-nya mi\-ri\-mu gi\-ngi, gi\-kwa\-ba ku\-ku\-le\-mee\-ra.}
\fmtExampleEN{Don’t try to do many things at once, you won’t manage.}

\fmtLemma{yi\-mbe}
\fmtCitation{(e)yi\-mbe}
\fmtPOS{n}%
\fmtNC{5/6}
\fmtSense{African civet cat}
\fmtExampleRL{Oma\-wa\-nge yei\-zi\-ra yi\-mbe.}
\fmtExampleEN{My mother belongs to the African civet cat clan.}

\fmtLemma{yi\-mboo\-ka}
\fmtCitation{(oku)\-yi\-mboo\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yi\-mboo\-ke\-re}
\fmtSubentry(a)
\fmtSense{to germinate, to sprout, to start growing}
\fmtExampleRL{Obu\-si\-go bu\-yi\-mboo\-ke\-re.}
\fmtExampleEN{The seeds have germinated.}
\fmtSubentry(b)
\fmtSense{to regain consciousness}
\fmtExampleRL{OBya\-lu\-mbe aka\-bi\-re\-ku, n’a\-yi\-mboo\-ka.}
\fmtExampleEN{Byalumbe had fainted, but now he has regained consciousness.}
\fmtSubentry(c)
\fmtSense{to rise from the dead, to be resurrected}
\fmtExampleRL{OYe\-su ya\-yi\-mboo\-ke\-re oku\-zwa omu ba\-fu.}
\fmtExampleEN{Jesus rose from the dead.}

\fmtLemma{yi\-mboo\-la}
\fmtCitation{(oku)\-yi\-mboo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yi\-mboi\-re}
\fmtSense{to raise from the dead, to resurrect, to bring back to life}
\fmtExampleRL{OYe\-su ya\-yi\-mboi\-re aba\-fu.}
\fmtExampleEN{Jesus resurrected the dead.}

\fmtLemma{yi\-mbya}
\fmtCitation{(oku)\-yi\-mbya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yi\-mbi\-rye}
\fmtSubentry(a)
\fmtSense{to add, to mix}
\fmtExampleRL{Yi\-mbya amai\-zi n’o\-bu\-sye le\-ke oko\-le omwe\-nge.}
\fmtExampleEN{Mix water with millet flour to make beer.}
\fmtSubentry(b)
\fmtSense{to join}
\fmtExampleRL{We\-yi\-mbye oku bai\-raa\-mu.}
\fmtExampleEN{Join your friends.}
\fmtSubentry(c)
\fmtSense{to bring together, to gather}
\fmtExampleRL{Oku\-ra\-mya kwa Sa\-nde ku\-yi\-mbya aba\-ntu bai\-ngi.}
\fmtExampleEN{Sunday prayers bring together many people.}
\fmtSubentry(d)
\fmtSense{to mediate for peace}
\fmtExampleRL{Ya\-ba oyi\-mbye aba\-ntu aba\-kwa\-ra\-ta\-na.}
\fmtExampleEN{Go and mediate between those people who are in conflict.}

\fmtLemma{yi\-mbya\-nga\-ta\-na}
\fmtCitation{(oku)\-yi\-mbya\-nga\-ta\-na}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yi\-mbya\-nga\-tai\-ne}
\fmtSense{to carry many different things at once}
\fmtExampleRL{Ayi\-mbya\-nga\-tai\-ne na bi\-ngi.}
\fmtExampleEN{She has carried all sorts of things at once.}

\fmtLemma{yi\-mpa\-wa\-la}
\fmtCitation{(oku)\-yi\-mpa\-wa\-la}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yi\-mpa\-wai\-re}
\fmtSense{to become shorter}
\fmtExampleRL{Nku\-bo\-na ota\-ndi\-ki\-re oku\-yi\-mpa\-wa\-la.}
\fmtExampleEN{I can see you have started becoming shorter.}

\fmtLemma{yi\-nci}
\fmtCitation{(e)yi\-nci}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{inch}
\fmtExampleRL{Aka\-bu\-nguu\-ra ako ka\-li\-ku eyi\-nci ibi\-ri.}
\fmtExampleEN{That caterpillar is two inches long.}

\fmtLemma{yi\-nda}
\fmtCitation{(oku)\-yi\-nda}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yi\-ndi\-re}
\fmtSense{(of water) to rumble}
\fmtExampleRL{Nga oi\-ke\-ndi ya\-ka\-kya amai\-zi ga\-yi\-nda mu\-no.}
\fmtExampleEN{When it stops raining, water rumbles a lot.}

\fmtLemma{yi\-ndi\-kwa}
\fmtCitation{(oku)\-yi\-ndi\-kwa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yi\-ndi\-ku\-rwe}
\fmtSense{to be in labour}
\fmtExampleRL{Omu\-ka\-li oyo ayi\-ndu\-ku\-rwe.}
\fmtExampleEN{That woman is in labour.}

\fmtLemma{yi\-ndu\-ki\-ra\-nga\-na}
\fmtCitation{(oku)\-yi\-ndu\-ki\-ra\-nga\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yi\-ndu\-ki\-ra\-ngai\-ne}
\fmtSense{(of two parties) to resume a negotiation or a quarrel}
\fmtExampleRL{ONa\-bi\-ga\-mbo n’o\-Wa\-ne\-nge ba\-yi\-ndu\-ki\-ra\-ngai\-ne olwa mwa\-na.}
\fmtExampleEN{Nabigambo and Wanenge have resumed the quarrel over the child.}

\fmtLemma{yi\-ndu\-la}
\fmtCrossreference{\textnormal{see} indula}

\fmtLemma{yi\-nga}
\fmtCitation{(oku)\-yi\-nga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yi\-ngi\-re}
\fmtSubentry(a)
\fmtSense{to be better, to be above others}
\fmtExampleRL{Omu\-ka\-li oni ayi\-ngi\-re oku ba\-ndi.}
\fmtExampleEN{This woman is better than others.}
\fmtSubentry(b)
\fmtSense{to exceed, to surpass, to be too much}
\fmtSubentry(c)
\fmtSense{(of a situation) to be unbearable to}
\fmtExampleRL{Embee\-ra enyi\-ngi\-re.}
\fmtExampleEN{The situation is unbearable to me.}

\fmtLemma{yi\-nga\-yi\-nga\-nwa}
\fmtCitation{(oku)\-yi\-nga\-yi\-nga\-nwa}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yi\-nga\-yi\-ngai\-nwe}
\fmtSense{(of an unborn baby) not to be well positioned}
\fmtExampleRL{Omwa\-na wa mu\-ka Mu\-bi\-ru aku\-yi\-nga\-yi\-nga\-nwa enda ya\-ku\-zwa\-mu.}
\fmtExampleEN{The foetus of Mrs. Mubiru is not in a normal position, she is likely to abort.}

\fmtLemma{yi\-ngi\-ni}
\fmtCitation{(e)yi\-ngi\-ni}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{engine}
\fmtExampleRL{Eyi\-ngi\-ni ya mo\-to\-ka ya\-nge esii\-si\-kai\-re.}
\fmtExampleEN{My engine has broken down.}

\fmtLemma{yi\-ngi\-ni\-ya}
\fmtCitation{(o)yi\-ngi\-ni\-ya}
\fmtPOS{n}%
\fmtNC{1a/2}
\fmtSource{from En.}
\fmtSense{engineer}
\fmtExampleRL{Oyi\-ngi\-ni\-ya wa di\-si\-tu\-li\-ki\-ti ta\-li\-na bu\-ku\-gu bu\-ku\-ma\-la.}
\fmtExampleEN{The district engineer is a novice.}

\fmtLemma{yi\-ngu\-la}
\fmtCitation{(oku)\-yi\-ngu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yi\-ngwi\-re, yi\-ngu\-lii\-re}
\fmtSense{to cut (meat)}
\fmtExampleRL{Omu\-te\-mi wa nya\-ma anyi\-ngu\-lii\-re mu\-ni\-ki gwe\-ree\-re.}
\fmtExampleEN{The butcher cut for me only lean meat.}

\fmtLemma{yi\-nsu\-wa}
\fmtCitation{(e)yi\-nsu\-wa}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{insurance}
\fmtExampleRL{Enyu\-mba ya\-nge ka ngi\-te omu yi\-nsu\-wa.}
\fmtExampleEN{Let me put my house in insurance.}

\fmtLemma{yi\-nta\-vi\-yu}
\fmtCitation{(e)yi\-nta\-vi\-yu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{interview}
\fmtExampleRL{Baa\-bi\-re Nai\-ro\-bi ku\-ko\-la yi\-nta\-vi\-yu ya bu\-so\-me\-sa.}
\fmtExampleEN{They have gone to Nairobi to be interviewed for a teaching job.}

\fmtLemma{yi\-nza}
\fmtCitation{(oku)\-yi\-nza}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{(none)}
\fmtSense{can, may}
\fmtExampleRL{Mwa\-la we oyi\-nza oku\-di\-ba olwa nge\-so ezi\-bbi!}
\fmtExampleEN{You may fail to get a man to marry you because of your bad manners!}

\fmtLemma{yi\-rii\-ki}
\fmtCitation{\textnormal{(e.g.}\,mu\-yi\-rii\-ki, ba\-yi\-rii\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{quarrelsome}
\fmtExampleRL{Omu\-sai\-za omu\-yi\-rii\-ki aswei\-re omu\-ka\-li omu\-si\-ri\-ki.}
\fmtExampleEN{A quarrelsome man has married a humble woman.}

\fmtLemma{yi\-ri\-zi}
\fmtCitation{(e)yi\-ri\-zi}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{amulet, charm}
\fmtExampleRL{Omwi\-bbi abbai\-re n’e\-yi\-ri\-zi omu nke\-nde.}
\fmtExampleEN{The thief had an amulet on the waist.}

\fmtLemma{yo}%
\fmtOne{1}
\fmtCitation{(e)yo}
\fmtPOS{adv}
\fmtSense{the\-re}
\fmtExampleRL{Ba\-li eyo ba\-ku\-zee\-nya.}
\fmtExampleEN{They are there playing.}

\fmtLemma{yo}%
\fmtOne{2}
\fmtPOS{pro}
\fmtSense{it (class 9 independent pronoun)}

\fmtLemma{yo}%
\fmtOne{3}
\fmtCitation{\textnormal{(e.g.}\,ga\-yo, gu\-yo, gi\-yo\textnormal{)}}
\fmtPOS{adj}
\fmtSense{hot}
\fmtExampleRL{Ama\-tai ga\-yo.}
\fmtExampleEN{The milk is hot.}

\fmtLemma{yo\-byo}
\fmtCitation{(o)yo\-byo}
\fmtPOS{n}%
\fmtNC{1a/10}
\fmtSense{spider weed (\textit{Cleome gynandra}, its leaves and shoots are eaten boiled or in sauces)}
\fmtExampleRL{Ti\-naa\-lye kai\-ta lwa\-lee\-ro okwi\-ca nga oyo\-byo ayii\-re.}
\fmtExampleEN{I won’t eat millet until the spider weed sauce is ready.}

\fmtLemma{yo\-ga\-ti}
\fmtCitation{(o)yo\-ga\-ti}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSource{from En.}
\fmtSense{yoghurt}
\fmtExampleRL{Mpee\-rya\-ku oyo\-ga\-ti ndye\-ku.}
\fmtExampleEN{Give me some yoghurt so that I eat a bit of it.}

\fmtLemma{yoi\-to\-ka}
\fmtCitation{(oku)\-yoi\-to\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yoi\-to\-ke\-re}
\fmtSense{to struggle with something}
\fmtExampleRL{Ba\-ku\-yoi\-to\-ka gwa bwi\-bbi.}
\fmtExampleEN{They are struggling with a case of theft.}

\fmtLemma{yo\-ko}
\fmtCitation{(o)yo\-ko}
\fmtPOS{n}%
\fmtNC{1a}
\fmtSense{endod (\textit{Phytolacca dodecandra})}
\fmtExampleRL{Oyo\-ko ata\-mba endwai\-re zi\-ngi.}
\fmtExampleEN{\textit{Phytolacca dodecandra} cures many diseases.}

\fmtLemma{yo\-nge\-ra}
\fmtCitation{(okw)yo\-nge\-ra}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yo\-ngei\-re}
\fmtSense{to go ahead, to continue, to proceed}
\fmtExampleRL{Yo\-nge\-ra oku\-ko\-la emi\-ru\-mo gya\-mu.}
\fmtExampleEN{Go ahead and do your work.}

\fmtLemma{yo\-nja}
\fmtCitation{(oku)\-yo\-nja}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yo\-nje\-rye}
\fmtSense{to clean, to make clean}
\fmtExampleRL{Yo\-nja eko\-mpyu\-ta yaa\-mu.}
\fmtExampleEN{Clean your computer.}

\fmtLemma{yo\-njo}
\fmtCitation{\textnormal{(e.g.}\,mu\-yo\-njo, ba\-yo\-njo, ki\-yo\-njo\textnormal{)}}
\fmtPOS{adj}
\fmtSense{clean, neat, smart}
\fmtExampleRL{Emee\-za eni gi\-yo\-njo.}
\fmtExampleEN{This table is clean.}

\fmtLemma{yoo\-ka}
\fmtCitation{(oku)\-yoo\-ka}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yoo\-ke\-re}
\fmtSense{(of eyes) to redden}
\fmtExampleRL{Atei\-re oka\-la\-li omu mai\-so ati ga\-yoo\-ke\-re.}
\fmtExampleEN{He has put red pepper in the eyes and now they have turned read.}

\fmtLemma{yoo\-ki}
\fmtCitation{\textnormal{(e.g.}\,ga\-yoo\-ki, lu\-yoo\-ki, mi\-yoo\-ki\textnormal{)}}
\fmtPOS{adj}
\fmtSense{reddish}
\fmtExampleRL{Amai\-so gaa\-mwe ga\-yoo\-ki.}
\fmtExampleEN{His eyes are reddish.}

\fmtLemma{yoo\-la}
\fmtCitation{(oku)\-yoo\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yoi\-re}
\fmtSubentry(a)
\fmtSense{to collect, to pick, to fetch, to pick up}
\fmtExampleRL{Yoo\-la ka\-sa\-si\-ro oyo.}
\fmtExampleEN{You collect that garbage.}
\fmtSubentry(b)
\fmtSense{to catch, to arrest}
\fmtExampleRL{Ba\-mu\-yoi\-re ni ba\-mu\-twa\-la ku po\-lii\-si.}
\fmtExampleEN{They arrested him and took him to the police.}
\fmtSubentry(c)
\fmtSense{to fall into (a net), to get trapped in}
\fmtExampleRL{Eki\-so\-lo ki\-yoi\-re eki\-ti\-mba kya\-nge.}
\fmtExampleEN{The animal has fallen into my net.}

\fmtLemma{yoo\-na}
\fmtCrossreference{\textnormal{see} ona}

\fmtLemma{yoo\-ta\-na}
\fmtCitation{(oku)\-yoo\-ta\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yoo\-tai\-ne}
\fmtSense{to struggle, to have difficulty coping}
\fmtExampleRL{Aba\-guu\-da ba\-yoo\-ta\-na mu\-no.}
\fmtExampleEN{The rich struggle a lot.}

\fmtLemma{yo\-roo\-ka}
\fmtCitation{(oku)\-yo\-roo\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yo\-roo\-ke\-re}
\fmtSense{to detect movement (within the surroundings)}
\fmtExampleRL{Mu\-wu\-lii\-re eki\-ku\-yo\-roo\-ka enzai?}
\fmtExampleEN{Have you heard something moving about outside?}

\fmtLemma{yo\-yoo\-ta}
\fmtCitation{(oku)\-yo\-yoo\-ta}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yo\-yoo\-te\-re}
\fmtSense{to perfect, to beautify, to put in good condition}
\fmtExampleRL{Omu\-si\-bi w’a\-ba\-go\-le ni\-ye ayo\-yoo\-ta ei\-so\-ke lya ba\-go\-le.}
\fmtExampleEN{The bridal dresser is the one who treats the brides' hair.}

\fmtLemma{yo\-yo\-ta\-nya}
\fmtCitation{(oku)\-yo\-yo\-ta\-nya}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yo\-yo\-ta\-ni\-rye}
\fmtSense{to struggle to cool (food or drink) in the mouth}
\fmtExampleRL{Alii\-re eme\-re egi\-yo atuu\-ni aku\-gi\-yo\-yo\-ta\-nya\-mbe.}
\fmtExampleEN{He has eaten hot food and he is now struggling to cool it in the mouth.}

\fmtLemma{yu\-ga}
\fmtCitation{(oku)\-yu\-ga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yu\-gi\-re}
\fmtSense{to stir, to mingle (\eg millet), to mix, to mash}
\fmtExampleRL{Aku\-yu\-gi\-re obu\-se\-ra ku\-sai mu\-no.}
\fmtExampleEN{She stirred the millet porridge very well.}

\fmtLemma{yu\-nga}
\fmtCitation{(oku)\-yu\-nga}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yu\-ngi\-re}
\fmtSubentry(a)
\fmtSense{to join, to connect, to splice (a rope)}
\fmtExampleRL{Naa\-ni ayi\-nza oku\-yu\-nga omu\-gu\-wa gwa\-nge ogwa\-tii\-ki\-re.}
\fmtExampleEN{Who can splice my torn rope?}
\fmtSubentry(b)
\fmtSense{to unify, to bring together, to unite}
\fmtExampleRL{Omwo\-yo gwa iya\-nga gu\-yu\-nga aba\-ntu.}
\fmtExampleEN{Patriotism can unify people.}
\fmtSubentry(c)
\fmtSense{to set (a fractured bone)}
\fmtExampleRL{Ei\-gu\-mba ni lye\-nde\-ka, omu\-yu\-ngi ali\-yu\-nga.}
\fmtExampleEN{When a bone fractures an orthopaedic surgeon puts it together.}

\fmtLemma{yu\-nguu\-ka}
\fmtCitation{(oku)\-yu\-nguu\-ka}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yu\-nguu\-ki\-re}
\fmtSense{to taste impurities in a drink or unready food (esp. maize paste)}
\fmtExampleRL{Noo\-lya aka\-wu\-nga aka\-ta\-yii\-re ku\-sai oyu\-nguu\-ka nka enka\-nja omu mu\-nwa.}
\fmtExampleEN{When you eat half-cooked posho, you feel the taste of impurities in the mouth.}

\fmtLemma{yu\-nguu\-la}
\fmtCitation{(oku)\-yu\-nguu\-la}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yu\-ngwi\-re}
\fmtSense{to sieve away husks}
\fmtExampleRL{Omu\-ka\-li ayu\-ngwi\-re ebi\-su\-su by’a\-mai\-do.}
\fmtExampleEN{The woman has sieved away groundnut pods.}

\fmtLemma{yu\-nguu\-ra}
\fmtCitation{(oku)\-yu\-nguu\-ra}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yu\-ngwi\-re}
\fmtSense{to separate, to isolate, to analyse}
\fmtExampleRL{Yu\-nguu\-ra amai\-do oku\-zwa omu bi\-yi\-mba.}
\fmtExampleEN{Separate the groundnuts from the beans.}

\fmtLemma{yu\-ni\-foo\-mu}
\fmtCitation{(e)yu\-ni\-foo\-mu}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{uniform}
\fmtExampleRL{Abee\-gi ba iso\-me\-ro eryo ba\-zwa\-la eyu\-ni\-foo\-mu gi\-tu\-ku\-li\-ki.}
\fmtExampleEN{Pupils of that school wear a red uniform.}

\fmtLemma{yu\-ni\-vaa\-si\-te}
\fmtCitation{(e)Yu\-ni\-vaa\-si\-te}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSource{from En.}
\fmtSense{university}
\fmtExampleRL{Omwa\-na wa\-nge ai\-ngii\-re eyu\-ni\-vaa\-si\-te.}
\fmtExampleEN{My child has joined the university.}

\fmtLemma{yu\-nja}
\fmtCitation{(oku)\-yu\-nja}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yu\-nji\-re}
\fmtSense{to harvest (esp. banana), to cut down, to cut}
\fmtExampleRL{Ya\-ba oyu\-nje ama\-too\-ke gaa\-mu.}
\fmtExampleEN{Go and harvest your bananas.}

\fmtLemma{yuu\-ga}
\fmtCitation{(oku)\-yuu\-ga}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yuu\-gi\-re}
\fmtSense{to be unstable, to sway, to tremor, to move about, to shake}
\fmtExampleRL{Omu\-saa\-le gu\-ni gu\-ku\-yuu\-ga mu\-no lwa mpu\-ngo.}
\fmtExampleEN{This tree is shaking a lot because of the strong.}

\fmtLemma{yuu\-ga\-na}
\fmtCitation{(oku)\-yuu\-ga\-na}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yuu\-gai\-ne, yuu\-ga\-ni\-re}
\fmtSense{to be unstable, to sway, to tremor, to move about, to shake}
\fmtExampleRL{Ama\-ka ga Mu\-si\-si ga\-ku\-yuu\-ga\-na.}
\fmtExampleEN{Musisi’s family is unstable.}

\fmtLemma{yuu\-guu\-ma}
\fmtCitation{(oku)\-yuu\-guu\-ma}
\fmtPOS{v}%
\fmtTrans{intr}
\fmtPFV{yuu\-guu\-mi\-re}
\fmtSense{to tremor, to shake}
\fmtExampleRL{Omu\-te\-nga n’a\-ya\-ka\-la ensi eyuu\-guu\-ma.}
\fmtExampleEN{When there is an earthquake, the earth tremors.}

\fmtLemma{yuu\-ja}
\fmtCrossreference{\textnormal{see also} zongooza}
\fmtCitation{(oku)\-yuu\-ja}
\fmtPOS{v}%
\fmtTrans{tr}
\fmtPFV{yuu\-ji\-re}
\fmtSubentry(a)
\fmtSense{to shake gently}
\fmtExampleRL{Aku\-yuu\-ja ama\-tai.}
\fmtExampleEN{He is gently shaking milk to make ghee.}
\fmtSubentry(b)
\fmtSense{to do a gentle massage of unborn baby}
\fmtExampleRL{ONa\-mu\-so\-ke aya\-ba owa mu\-le\-rwa oku\-mu\-yuu\-ja.}
\fmtExampleEN{Namusoke visits the midwife to have her pregnancy massaged.}

\fmtLemma{yuu\-ya}
\fmtCitation{(e)yuu\-ya}
\fmtPOS{n}%
\fmtNC{9/10}
\fmtSense{\textit{Acalypha bipartita} (type of plant)}
\fmtExampleRL{Tu\-su\-mbi\-re oku mu\-ku\-bi gwa yuu\-ya.}
\fmtExampleEN{Cook for us some \textit{Acalypha bipartita} for the sauce.}
\end{letter}