\documentclass[output=paper,colorlinks,citecolor=brown,newtxmath]{langsci/langscibook}
\ChapterDOI{10.5281/zenodo.3764859}
%\bibliography{localbibliography}
%\input{localpackages.tex}
%\input{localcommands.tex}
\IfFileExists{../localcommands.tex}{
  \input{../localpackages}
  \input{../localcommands}
  \input{../localhyphenation}
  \togglepaper[9]%%chapternumber
}{}
%\togglepaper[9]

\author{Ilya Naumov\affiliation{National Research University Higher School of Economics}}

\title{Constraining the distribution of the perdurative in Russian}

\abstract{In this paper, I explore the constraints on the distribution of the perdurative prefix \textit{pro-} in Russian. Applying several diagnostics proposed by \cite{tatevosov2009mnozestvennaja, tatevosov2013mnozestvennaja}, I show that the perdurative \textit{pro-} is a ``selectionally restricted" prefix associated with an additional restriction: it can combine with a predicate built on a \isi{secondary imperfective} stem only under a pluractional interpretation. I argue that this restriction is an instantiation of a more general semantic requirement imposed by the perdurative: it can be formed from a predicate if there are no subevents making up the activity component of this predicate that are in \citeposst{landman1992progressive} ``stage-of relation''.

\keywords{aspect, (a)telicity, subeventual structure, verbal prefixes, perdurative, Russian}}




\begin{document}%
\maketitle

\section{Introduction} \label{sec:naumov:1}
This chapter focuses on the distribution of the \isi{perdurative} in \ili{Russian}. So-called \isi{perdurative} verbs in \ili{Russian} are verbs marked by the so-called \isi{perdurative prefix} \textit{pro-}. Perdurative-prefixed verbs convey the meaning of temporal duration. The addition of the prefix to a \isi{verb} stem makes a co-occurring measure expression obligatory, see \REF{ex:naumov:1} and \REF{ex:naumov:1b}.\footnote{There is no consensus on the syntactic status of measure expressions occuring with \isi{perdurative} verbs. A common view is that they are adjuncts \citep[a.o.]{borik2006aspect,ramchand2008verb,gehrke2008ps}. Some researchers argue for a quasi-argument status for these phrases \citep{fowler1993argument}. \citet{schoorlemmer1995participial} distinguishes different types of \isi{accusative} phrases: some of them are treated as arguments, while others as adjuncts. The most recent and at the same time the most radical analysis is presented in \citet{Žaucer2009vp,Žaucer2012syntax}, where all \isi{accusative} phrases are claimed to be unselected direct objects introduced by the prefix. I will not go into this problem here and refer the interested reader to the cited works.}

\ea \ea \label{ex:naumov:1}
\gll Lena	govori-la	(\hspace{-2pt} ves’	večer)		po	telefonu.\\
     L.	talk-\textsc{pst} {} all	evening	on	phone\\
\glt `Lena talked on the phone for the whole evening.'
\ex\gll Lena pro-govori-la *(\hspace{-2pt} ves' večer)		po telefonu.\\
     L. \textsc{pro}$_{\textsc{per}}$-talk-\textsc{pst} {} all	evening	on	phone\\
\glt `Lena spent the whole evening talking on the phone.'\label{ex:naumov:1b}
\z \z

\noindent While the syntax (\citealt{babko1999zero,ramchand2005time,svenonius2004slavic,romanova2007constructing,gehrke2008ps,Žaucer2009vp,tolskaya2015verbal}, a.o.) and semantics (\citealt{isacenko1960grammaticeskij,bondarko1967,forsyth1970grammar,flier1985scope,zalizniak2000,gehrke2003aspectual,filip2005measures}, a.o.) of the \isi{perdurative} in \ili{Russian} has been extensively discussed, the constraints on its distribution, to the best of my knowledge, have not been the subject of a separate research. Nevertheless, data suggest that these constraints are intricate. For example, the \isi{perdurative} can be combined with some predicates describing atelic processes, \REF{ex:naumov:2} (recall also \REF{ex:naumov:1b}), but not with others, \REF{ex:naumov:3}.

\ea[*]{\label{ex:naumov:2}
\gll Kolja		pro-pisa-l		pismo	vsjo	utro.\\
     K.		\textsc{pro}$_\textsc{per}$-write-\textsc{pst}	letter	all	morning\\
\glt Intended: `Kolja spent the whole morning writing the letter.'}
\z

\ea[]{\label{ex:naumov:3}
\gll Kolja	pro-taska-l		čemodan	neskol’ko	časov.\\
     K.		\textsc{pro}$_\textsc{per}$-carry-\textsc{pst}	suitcase	several		hours\\
\glt `Kolja spent several hours carrying the suitcase.'}
\z

\noindent As \citet{pazelskaya2006} convincingly argue, so-called simplex \isi{imperfective} stems, like \textit{pisa-} and \textit{taska-}, project an activity \isi{event structure}. With respect to common diagnostics on subeventual complexity (such as the interpretation of \textit{opjat’} `again' and \isi{negation}) both \textit{pisat’ pis’mo} `write a letter' and \textit{taskat’ čemodan} `carry a suitcase'
 demonstrate identical properties: they are subeventually simplex denoting events that do not lead to the attainment of the \isi{result state}. If nothing else is added here, it remains a mystery why the \isi{perdurative} is bad from the former and fine from the latter.

Moreover, the \isi{perdurative} cannot be derived from predicates describing some goal-oriented activity that stops before the corresponding \isi{result state} is attained. An example of such a predicate is \textit{otkryvat' zamok} `open a lock'. The unavailability of the \isi{perdurative} from this predicate is demonstrated in \REF{ex:naumov:4}.

\ea \textit{Situation:} The lock in the door is broken. After spending half an hour trying to open it, Vasja gave up.\medskip\\
\gll *\hspace{-2pt} Vasja		pro-[[otkr]-yva]-l		zamok		polčasa.\\
     {} V.
     \textsc{pro}$_\textsc{per}$-open-\textsc{yva}-\textsc{pst}	lock		half.an.hour\\
\glt Intended: `Vasja spent half an hour opening the lock.' \label{ex:naumov:4}
\z

\noindent Unlike in the examples above, in \REF{ex:naumov:4} the \isi{perdurative} attaches to the so-called \isi{secondary imperfective} \isi{verb} stem containing the so-called \isi{secondary imperfective suffix} \textit{-yva}. It has been claimed by \citet{altshuler2013,altshuler2014} and \citet{tatevosov2017temporal} that in this type of predicates \textit{-yva} functions as a partitive operator. It extracts (proper) non-final parts (more precisely, stages) from the extension of the original telic \isi{event predicate} \textit{otrkyt’ zamok} `open a lock'. Why cannot the \isi{perdurative} successfully operate on this \isi{eventuality} description?

In principle, the \isi{perdurative} can be derived from a prediсate based on the \isi{secondary imperfective} \isi{verb} stem but only if this predicate has a \isi{pluractional interpretation}. Consider the examples in \REF{ex:naumov:5}. While the sentence in \REF{ex:naumov:5a} has two possible readings, the sentence in \REF{ex:naumov:5b}, which contains a perdurative-prefixed \isi{verb}, is only compatible with a scenario in which the agent opened his mouth repeatedly.\footnote{I would like to thank an anonymous reviewer for suggesting scenario №1 for examples in \REF{ex:naumov:5}.} That is, the restriction on the possible interpretation arises at the stage when the \isi{perdurative} attaches to the stem.

\ea \label{ex:naumov:5} \ea \label{ex:naumov:5a}
\gll Kolja 		otkr-yva-l 		rot 		minutu.\\
     K. 		open-\textsc{yva}-\textsc{pst} 	mouth 		minute\\
\glt `Kolja was opening the mouth for a minute.' \\
\ea[\ding{51}]{Scenario №1: Kolja was opening the mouth wider and wider for a minute until he got cramps in his cheekbones.}
\ex[\ding{51}]{Scenario №2: Kolja was opening the mouth several times for a minute until he realized that he was not producing any sounds.}
\z

\ex{ \label{ex:naumov:5b}
\gll Kolja 	pro-[[otkr]-yva]-l 		rot 		minutu.\\
     K. \textsc{pro}$_\textsc{per}$-open-\textsc{yva}-\textsc{pst} mouth 		minute\\
\glt `Kolja spent a minute opening the mouth.' \\
\ea[\ding{55}]{ Scenario №1: Kolja spent a minute opening the mouth wider and wider until he got cramps in his cheekbones.}
\ex[\ding{51}]{Scenario №2: Kolja spent a minute opening the mouth several times until he realized that he was not producing any sounds.}\z
}
\z\z


\noindent\sloppy Although predicates describing non-culminating events, as in \REF{ex:naumov:4}, and plural events, as in \REF{ex:naumov:5b}, are made up from the same morphological pieces that come in the same order (the \isi{secondary imperfective suffix} \textit{-yva} attached before the \isi{perdurative} \textit{pro-}), they still differ with respect to the availability of the \isi{perdurative}. What is the underlying property distinguishing these classes of predicates that the \isi{perdurative} is sensitive to?

In the remainder of the paper I will explore restrictions on the distribution of the \isi{perdurative} in greater detail. In \sectref{sec:naumov:2}, I will show that the distribution of the \isi{perdurative} is subject to (morpho)syntactic constraints. Applying the diagnostics that were proposed by \citet{tatevosov2009mnozestvennaja,tatevosov2013mnozestvennaja}, I will argue that the \isi{perdurative} \textit{pro-} is a \isi{selectionally restricted} prefix. Taking into account the interaction of the \isi{perdurative} \textit{pro-} with some other affixes, I will claim that it originates in a \isi{functional projection} between \textit{v}P and \isi{AspP}. In \sectref{sec:naumov:3}, I will present empirical evidence indicating that the distribution of the \isi{perdurative} is also subject to semantic constraints. Namely, I will argue that predicates allowing the derivation of the \isi{perdurative} form a natural class with respect to one semantic property: subevents making up the \isi{activity component} of these predicates are not in \citeposst{landman1992progressive} ``\isi{stage-of relation}''. The main results and several open questions are listed in \sectref{sec:naumov:4}.


\section{Determing the position of the prefix} \label{sec:naumov:2}
\subsection{The perdurative \textit{pro-} in the big picture} \label{sec:naumov:2.1}
One of the most important observations concerning prefixes in \ili{Russian} and, more broadly, in \ili{Slavic} languages that has been made so far is that they form a heterogeneous class and fall at least into two types: \textsc{lexical prefixes} (LPs) and \textsc{superlexical prefixes} (SLPs), or internal and external in other terminology. For the first time this dichotomy was argued for by \citet{babko1999zero} and subsequently became the subject of intense discussion \citep[see, e.\,g.,][]{svenonius2004slavic, ramchand2005time, romanova2005superlexical, romanova2007constructing, tolskaya2015verbal}. SLPs have several characteristics that distinguish them from lexical ones:
\begin{itemize}
    \item External prefixes merge outside VP.
\item External prefixes are attached over the internal ones.
\item External prefixes do not affect the argument 	structure of the \isi{verb} stem, or predictably decrease its transitivity.
\item External prefixes are semantically compositional.
\item External prefixes express temporal or quantifying meanings.
\end{itemize}

\noindent Among the works listed above, there is no agreement on the position of the \isi{perdurative} \textit{pro-} within this classification. Some authors include it in the list of SLPs, while others do not.


\begin{table}[h!]
\caption{The perdurative \textit{pro-} as an SLP}
\small
\label{tab1}
\begin{tabularx}{\textwidth}{cCCCCC}
  \lsptoprule
            \citeauthor{babko1999zero}  &	\citeauthor{ramchand2005time} &	\citeauthor{svenonius2004slavic}  &	\citeauthor{romanova2007constructing}  &	\citeauthor{gehrke2008ps}  &	\citeauthor{tolskaya2015verbal}\\
            (\citeyear{babko1999zero}) & (\citeyear{ramchand2005time}) & (\citeyear{svenonius2004slavic}) & (\citeyear{romanova2007constructing}) & (\citeyear{gehrke2008ps}) & (\citeyear{tolskaya2015verbal}) \\
  \midrule
 $+$ &    $+$  &    $-$ & $-$ & $+$ & $-$\\
  \lspbottomrule
 \end{tabularx}
\end{table}

This binary opposition has been recently refined. Focusing on \ili{Russian} data, \citet{tatevosov2009mnozestvennaja,tatevosov2013mnozestvennaja} claims that prefixes commonly subsumed under the label SLPs fall, in fact, into at least two distinct groups with respect to constraints that regulate their distribution. Namely, there are prefixes that demonstrate selectional restrictions, \REF{ex:naumov:6a}, and prefixes that demonstrate positional restrictions, \REF{ex:naumov:6b}.\footnote{ The property of being ``formally (im)\isi{perfective}'' does not imply carrying any \isi{aspectual} semantics. This is a morphological notion.}


\ea \label{ex:naumov:6} Possible restrictions on the distribution of SLPs:
\ea \label{ex:naumov:6a} Selectional restriction. The possibility of attaching a prefix to a stem can be constrained by the stem's formal (im)\isi{perfectivity}.
\ex \label{ex:naumov:6b} Positional restriction. The possibility of attaching a prefix to a stem can be constrained by the positional relationship between this prefix and the \isi{secondary imperfective suffix} \textit{-yva}.
\z \z

\noindent With respect to these two restrictions there emerge \textsc{selectionally restricted} (\isi{SR}) and \textsc{positionally restricted} (\isi{PR}) prefixes. If the distribution of a given prefix can be described through \REF{ex:naumov:6a}, this prefix is said to be an SR-prefix. If the distribution of a given prefix can be described through \REF{ex:naumov:6b}, this prefix is said to be a PR-prefix. The list of SR-prefixes includes: \isi{delimitative} \textit{po-}, cumulative \mbox{\textit{na-},} distributive \textit{pere-}, inchoative \textit{za-}. The list of PR-prefixes includes: completive \mbox{\textit{do-},} \isi{repetitive} \textit{pere-}, attenuative \textit{pod-}, attenuative \textit{pri-}. The \isi{perdurative} \textit{pro-} is not considered separately by \citet{tatevosov2009mnozestvennaja, tatevosov2013mnozestvennaja}.

There is one more alternative view on the lexical/superlexical distinction found in  \citet{Žaucer2009vp,Žaucer2012syntax}. The author of these works consistently argues that at least some SLPs merge within the same resultative projection as LPs. The \isi{perdurative} \textit{pro-} is claimed to be one of these SLPs. It should be pointed out, however, that \citeauthor{Žaucer2009vp}'s (\citeyear{Žaucer2009vp,Žaucer2012syntax}) proposal is based mainly on data from \ili{Slovenian}. The present work does not set as its goal to revise it. What I aim to do is to try to determine the position of the \isi{perdurative} \textit{pro-} in \ili{Russian}. For this I will use diagnostics proposed by \citet{tatevosov2009mnozestvennaja, tatevosov2013mnozestvennaja}.

\subsection{The position of the perdurative \textit{pro-}} \label{2.2}
In this part of the paper, I will follow \cite{tatevosov2009mnozestvennaja,tatevosov2013mnozestvennaja} and assume that SLPs fall into at least two separate classes: SR-prefixes and PR-prefixes.
The class membership of a given prefix is determined via the restrictions from \REF{ex:naumov:6}. Below I will show that the distribution of the \isi{perdurative} \textit{pro-} is subject to selectional restrictions and argue that the \isi{perdurative} \textit{pro-} is an SR-prefix.

First, the \isi{perdurative} \textit{pro-} selects for formally \isi{imperfective} stems. Its distribution falls under the generalization in \REF{ex:naumov:7}.

\ea \label{ex:naumov:7}
The \isi{perdurative} \textit{pro-} merges with formally \isi{imperfective} stems.
\z

\noindent It has already been demonstrated in \sectref{sec:naumov:1} that the \isi{perdurative} \textit{pro-} can combine both with simplex \isi{imperfective} stems and \isi{imperfective} stems derived through applying the \isi{secondary imperfective suffix} \textit{-yva}. Here, I repeat the relevant examples.

    \ea Perdurative \textit{pro-} with a non-derived (non-prefixed) \isi{imperfective} stem:\label{ex:naumov:8}\smallskip\\
    \gll Lena	pro-govori-la		ves’	večer		po	telefonu. \\
L. \textsc{pro}$_\textsc{per}$-talk-\textsc{pst}	all	evening	on	phone\\
\glt `Lena spent the whole evening talking on the phone.'
\z


    \ea Perdurative \textit{pro-} with a stem imperfectivized by         \textit{-yva:}\label{ex:naumov:9}\smallskip\\
    \gll Kolja 	pro-[[otkr]-yva]-l 		rot 		minutu.\\
     K. \textsc{pro}$_\textsc{per}$-open-\textsc{yva}-\textsc{pst} mouth 		minute\\
    \glt `Kolja spent a minute opening the mouth.'
\ea[\ding{55}]{\ Scenario №1: Kolja spent a minute opening the mouth wider and wider until he got cramps in his cheekbones.}
\ex[\ding{51}]{Scenario №2: Kolja spent a minute opening the mouth several times until he realized that he was not producing any sounds.}
\z\z

\noindent What is the source of the restriction on possible interpretations in \REF{ex:naumov:9}?

As was shown in \sectref{sec:naumov:1} (recall \REF{ex:naumov:5a}), when the \isi{perdurative} \textit{pro-} is not attached above  \textit{-yva}, the sentence is also compatible with a progressive interpretation. \citet{tatevosov2015severing} claims that when \textit{-yva} induces a progressive interpretation, it functions as a partitive operator and merges in a position between VP and \textit{v}. One can naturally assume that the second (pluractional) interpretation of the sentence in \REF{ex:naumov:5a} is also induced by the \textit{-yva} suffix and that in this case it functions as a pluractional operator. It has been proposed that pluractional operators apply very low in the syntactic structure, namely, at the level of V (see, e.g, \citealt{lasersohn1995plurality,van2004adverbials}). If this view is correct, the restriction in \REF{ex:naumov:9} could be treated as (morpho)syntactic. Namely, despite no surface difference, it could be stated that \textit{-yva} has the possibility to enter the derivation in two distinct hierarchical positions: within VP and above it. When the suffix induces a progressive interpretation, it merges, as \citet{tatevosov2015severing} proposes, between VP and \textit{v}. In contrast, when this suffix induces a \isi{pluractional interpretation}, it adjoins to V. The incompatibility of the \isi{perdurative} \textit{pro-} and the ``partitive'' \textit{-yva} could be explained by claiming that they compete for the same position and, thus, block the derivation. Such a configuration would look like in \REF{ex:naumov:10}.

\ea \label{ex:naumov:10}
{[\textsubscript{\textit{v}P} {\ldots} [\textsubscript{\textit{v}$'$} \ldots [\textsubscript{F} \textit{pro}\textsubscript{\textsc{per}}- \textit{-yva}\textsubscript{\textsc{part}} {\ldots} [\textsubscript{VP} {\ldots}]]]].}
\z

\noindent The explanation for the compatibility of the \isi{perdurative} \textit{pro-} and the ``pluractional''
\textit{-yva} would be that the latter occupies a position lower in the tree and does not prevent the former from merging with a stem, \REF{ex:naumov:11}.

\ea \label{ex:naumov:11}
{[\textsubscript{\textit{v}P} {\ldots} [\textsubscript{\textit{v}$'$} {\ldots} [\textsubscript{F} \textit{pro}\textsubscript{\textsc{per}}- {\ldots} [\textsubscript{VP} {\ldots} [\textsubscript{V$'$} \textit{-yva}\textsubscript{\textsc{iter}}  {\ldots}]]]]].}
\z

\noindent While this line of reasoning may be true, I cannot come up with any empirical evidence in favor of it.\footnote{In fact, below I will argue, relying on the fact that the \isi{perdurative} \textit{pro-} stacks above the \isi{repetitive} \textit{pere-}, that the \isi{perdurative} \textit{pro-} merges above \textit{v}P. This, if true, can be taken as an indirect argument against such a reasoning.} Moreover, as \citet{gianina2015pluractionality} show, there are a number of empirical and theoretical challenges for the view that pluractional operators are V-level operators. They argue for high aspect-level pluractionality. I will pursue a different path and try to show in \sectref{3} that the restriction we observe in examples like \REF{ex:naumov:5b} and \REF{ex:naumov:9} occurs due to semantic reasons and can be explained without any specific assumptions about the syntax of pluractional operators.

On the flip side, the \isi{perdurative} \textit{pro-} does not select for formally \isi{perfective} stems. Its distribution falls under the generalization in \REF{ex:naumov:12}.

\ea \label{ex:naumov:12}
The \isi{perdurative} \textit{pro-} does not merge with formally \isi{perfective} stems.
\z

    \ea Perdurative \textit{pro-} with a non-derived (non-prefixed) \isi{perfective} stem:\label{ex:naumov:13}\smallskip\\
    \gll * Vasja		pro-reši-l			zadanije	desjat’		minut.\\
    {} V.		\textsc{pro}$_\textsc{per}$-solve-\textsc{pst}		task		ten		minutes
    \\
    \glt \ \ Intended: `Vasja spent ten minutes solving the task.'
    \z

    \ea Perdurative \textit{pro-} with a \isi{perfective} stem derived by prefixation:\label{ex:naumov:14}\smallskip\\
    \gll * Maša	pro-[na-[pisa]]-la		pis’mo		dva	časa.\\
    {} M.	\textsc{pro}$_\textsc{per}$-\textsc{na}-write-\textsc{pst}	letter		two	hours\\
    \glt \ \ Intended: `Maša spent two hours writing the letter.'
    \z


\noindent As can be seen from the examples, the \isi{perdurative} \textit{pro-} cannot combine either with simplex \isi{perfective} stems, \REF{ex:naumov:13}, or with \isi{perfective} stems derived by prefixation, \REF{ex:naumov:14}.


The intermediate conclusion that can be drawn at this stage is the following. The distribution of the \isi{perdurative} \textit{pro-} is constrained by the (im)\isi{perfectivity} of the stem. Specifically, the \isi{perdurative} \textit{pro-} merges with formally \isi{imperfective} stems and does not merge with formally \isi{perfective} ones. Therefore, the distribution of the \isi{perdurative} \textit{pro-} is subject to selectional restrictions and hence the \isi{perdurative} \textit{pro-} is an SR-prefix. Next, I will consider constraints that regulate the distribution of PR-prefixes and argue that (i) the \isi{perdurative} \textit{pro-} is not subject to these constraints; (ii) the \isi{syntactic position} in which the \isi{perdurative} \textit{pro-} merges is above \textit{v}P.

PR-prefixes do not impose restrictions on the (im)\isi{perfectivity} of the stem with which they combine \citep{tatevosov2009mnozestvennaja,tatevosov2013mnozestvennaja}. As was shown in \REF{ex:naumov:13} and \REF{ex:naumov:14}, this does not hold for the \isi{perdurative} \textit{pro-} because the \isi{perdurative} \textit{pro-} requires the stem to which it attaches not to be formally \isi{perfective}.

The distribution of PR-prefixes falls under the generalization in \REF{ex:naumov:15}.

\ea \label{ex:naumov:15}
PR-prefixes do not merge above the \isi{secondary imperfective suffix} \textit{-yva}.
\z

\noindent As was shown in \REF{ex:naumov:9}, repeated here as \REF{ex:naumov:16}, this generalization also does not hold for the \isi{perdurative} \textit{pro-}.


 \ea Perdurative \textit{pro-} above the \isi{secondary imperfective suffix} \textit{-yva:}\label{ex:naumov:16}\smallskip\\
     \gll Kolja	pro-[[otkr]-yva]-l		rot		minutu.\\
K.	\textsc{pro}$_\textsc{per}$-open-\textsc{yva}-\textsc{pst}	mouth		minute\\
     \glt `Kolja spent a minute opening the mouth.'
     \z


\noindent Another configuration, when it attaches under the \isi{secondary imperfective}, is also present, \REF{ex:naumov:17}.

\ea The \isi{perdurative} \textit{pro-} under the \isi{secondary imperfective suffix} \textit{-yva:}\label{ex:naumov:17}\smallskip\\
    \gll Prošloj	zimoj	ja	[\hspace{-2pt} pro-[lež]]-iva-l	na	divane	po	desjat’ časov	v	den’.\\
    Last	winter	I	{} \textsc{pro}$_\textsc{per}$-lay-\textsc{yva}-\textsc{pst}	on	sofa	by	ten hours	in	day \\
    \glt `Last winter I spent ten hours a day laying on the sofa.'
    \z


\noindent Unlike PR-prefixes, SR-prefixes do not have restrictions relative to the position of the \isi{secondary imperfective} \textit{-yva}. The fact that the \isi{perdurative} \textit{pro-} can merge both under and above the suffix also unites it with SR-prefixes.


There is one more observation concerning SR- vs. PR-prefixes dichotomy: SR-prefixes attach above PR-prefixes and they cannot merge as adjacent heads, \textit{-yva} must merge between them \citep{tatevosov2009mnozestvennaja,tatevosov2013mnozestvennaja}. As the example in \REF{ex:naumov:18} demonstrates this is true for the \isi{perdurative} \textit{pro-}. In this example, the \isi{perdurative} \textit{pro-} attaches to the stem that already contains the \isi{repetitive} suffix \textit{pere-}, which is assumed to be a PR-prefix.\footnote{Note that \REF{ex:naumov:18} is compatible only with a \isi{pluractional interpretation}, as is reflected in the translation.}

\ea \label{ex:naumov:18}
\gll Vasja	\minsp{\{} pro-[[pere-[čit]]-yva]-l / \minsp{*} pro-[pere-[čita]]-l	\} etot	abzac polčasa,	no 	ničego		ne	ponja-l. \\
V. {} \textsc{pro}$_\textsc{per}$-\textsc{pere}$_\textsc{rep}$-read-\textsc{yva}-\textsc{pst} {} {} \textsc{pro}$_\textsc{per}$-\textsc{pere}$_\textsc{rep}$-read-\textsc{pst}		{} 	this	paragraph	half.an.hour	but	nothing	not	understand-\textsc{pst} \\
\glt `Vasja spent half an hour reading this paragraph over and over again, but did not understand anything.'
\z

\noindent PR-prefixes have the possibility to enter the derivation in two distinct syntactic positions: between VP and \textit{v}P, and above \textit{v}P \citep{tatevosov2008intermediate}. The position of the prefix affects the interpretation of the predicate. If the prefix merges before \textit{v}P is projected and takes in its scope only the \isi{result state}, the restitutive reading obtains. In contrast, if the prefix merges after \textit{v}P is projected and takes in its scope the whole \isi{event}, the \isi{repetitive} reading obtains. In \REF{ex:naumov:18}, we observe the second possibility: \textit{pere-} enters the derivation above \textit{v}P, and, thus, yields the \isi{repetitive} reading `Vasja read the paragraph, and that had happened before'. Therefore, as the \isi{perdurative} \textit{pro-} attaches above \textit{pere-}, it must merge in a position above \textit{v}P. We are open to two different possibilities: (a) the \isi{perdurative} \textit{pro-} merges in \isi{AspP} and (b) the \isi{perdurative} \textit{pro-} merges lower, between \textit{v}P and \isi{AspP}. Although (a) seems to be more straightforward, below I will speak in favor of (b).\footnote{Note that in principle both of them are consistent with the semantic proposal made in \sectref{3}.}

In formal Slavic literature, there have been various proposals regarding the position in which SLPs merge. They can be divided into two groups with respect to how they treat the hierarchical relationship between prefixes and \isi{perfectivity}: (i) \isi{perfectivity} is introduced as high as prefixes; (ii) \isi{perfectivity} is introduced higher than prefixes.

The first view is the mainstream and presented in a large number of works on the syntax of prefixation. For example, \cite{babko1999zero} treats SL-prefixes as left-adjuncts to Asp and \cite{svenonius2004slavic} treats SL-prefixes as PPs occupying SpecAspP (see also \citealt{ramchand2005time} for a similar proposal). These authors assume that \isi{perfectivity} is directly encoded in SL-prefixes or introduced sufficiently local to them.

The second view is less popular and presented in \citet{pinon1994} and \citet{filip2000quantization, filip2005measures,filip2008events}. These authors focus on the \isi{delimitative} prefix \textit{po-} and claim that it is an \isi{event} modifier with the semantics of a measure \isi{adverbial}. The result of its application is an \isi{event predicate}, not a property of times. According to these works, \isi{perfectivity} is not part of the meaning of the \isi{delimitative} \textit{po-}. It is introduced by a phonologically silent operator located higher in the syntactic structure.

The crucial thing is that these approaches have different predictions about the possibility of multiple pieces of \isi{aspectual} morphology within a word form. Specifically, if we assume that the \isi{perdurative} \textit{pro-} merges in the projection of \isi{aspectual} operators, we do not expect any other \isi{aspectual} morphology after its application. Data, however, suggest just the opposite. It has already been shown in \REF{ex:naumov:17}, repeated here as \REF{ex:naumov:19}, that the \isi{secondary imperfective suffix} \textit{-yva} can attach above the \isi{perdurative} \textit{pro-}.

\ea \label{ex:naumov:19}
\gll Prošloj	zimoj	ja	[\hspace{-2pt} pro-[lež]]-iva-l	   na	divane	   po	desjat’ časov	v	den’. \\
Last	winter	I   {}	\textsc{pro}$_\textsc{per}$-lay-\textsc{yva}-\textsc{pst}	   on	sofa	   by	ten hours	in	day \\
\glt `Last winter I spent ten hours a day laying on the sofa.'
\z

\noindent If we take \isi{aspectual} morphology to be merged in the projection of \isi{aspectual} operators, then in the case when there are multiple pieces of this morphology within a word form, as in \REF{ex:naumov:19}, we have to postulate several adjacent \isi{AspP} projections. However, there exists a well-established ban on consecutive identical projections. \citet{de2018negation} have recently proposed the following formulation of this constraint:

\ea[*] {\label{ex:naumov:20}
$\langle$X, X$\rangle$ \\
The functional sequence must not contain two immediately consecutive identical projections.}
\z

\noindent That is, the syntactic structure with several adjacent \isi{AspP} projections would contradict the constraint on admissible functional sequence and would be undesirable from a theoretical point of view. This is the reason why I opt for (b): the \isi{perdurative} \textit{pro-} merges in a \isi{functional projection} between \textit{v}P and \isi{AspP}.\footnote{A reviewer poses the following question: ``Which projection is positioned there? Are there more projections present? In fact, it seems to me that there could be only different types of \isi{aspectual} projections like in \citet{cinque1999}''. While I have no basis to make more concrete claims regarding the nature of this projection, I assume that an approach in the spirit of \citet{cinque1999} can be potentially undertaken. For example, \citet{markova2011} proposes that some prefixes in \ili{Bulgarian} are derived in dedicated \isi{aspectual} projections a la \citet{cinque1999}.}

Summarizing all that has been discussed so far, I make the following conclusions:

\begin{itemize}
    \item The distribution of the \isi{perdurative} \textit{pro-} is subject to constraints that regulate the distribution of SR-prefixes.
\item The \isi{syntactic position} in which the \isi{perdurative} \textit{pro-} merges is between \textit{v}P and \isi{AspP}.
\item The \isi{perdurative} \textit{pro-} felicitously combines with predicates based on \isi{secondary imperfective} \isi{verb} stems only when they receive a \isi{pluractional interpretation}.
\end{itemize}

\noindent In the next section, I will try to argue for a unified semantic constraint that regulates the distribution of the \isi{perdurative} in \ili{Russian}. The proposal will be based on the observation about the specific behaviour of pluractional predicates with respect to \citeposst{landman1992progressive} \isi{stage-of relation}.

\section{A semantic constraint on the distribution of the perdurative} \label{sec:naumov:3}
\subsection{Theoretical background} \label{sec:naumov:3.1}

I adopt a neo-Davidsonian version of \isi{event} semantics where verbs are represented as one-place predicates over sets of eventualities. I assume that the denotation of \textit{v}P is a predicate of events which is mapped to predicates of times at Asp (see, e.g., \citealt{wolfgang1994time} and much further literature). In addition, adhering to the common view in predicate decomposition, I presuppose that while accomplishment predicates are subeventually complex and consist of, at least, two separate subevents—activity (or process \isi{subevent}) and \isi{result state} (or become \isi{subevent})—that are connected by a finite set of causal relations (\citealt{dowty1979word}; \citealt{rothstein2004structuring}; a.o.), activity predicates have simple structures consisting of a single activity \isi{subevent}. I assume, following \citet{pazelskaya2006}, that predicates based on ``simplex \isi{imperfective}'' \isi{verb} stems are associated with an activity \isi{event structure}. Following \citet{altshuler2013,altshuler2014} and \citet{gronn2003semantics, gronn2013}, I take both simplex imperfectives and secondary imperfectives as denoting (not necessarily proper) parts (or stages) of complete eventualities.

\subsection{Previous studies} \label{sec:naumov:3.2}
To the best of my knowledge, the semantics and the distribution of the \isi{perdurative} in \ili{Russian} have not been the subject of a separate study so far. Traditional Slavic aspectology is rather consistent in giving a more or less uniform treatment of the meaning of the \isi{perdurative}. For example, \cite[243--244]{isacenko1960grammaticeskij} proposes that \isi{perdurative} verbs denote ``the completion of a process that lasted for a specific period of time''. \cite[16]{bondarko1967} postulate a ``long-term Aktionsart''; its nucleus includes ``verbs with the prefix \textit{pro-} denoting an action that covers a specific period of time''. \cite[23]{forsyth1970grammar} notes that verbs marked by the \isi{perdurative} \textit{pro-}, as opposed to verbs marked by the \isi{delimitative} \textit{po-}, ``suggest a longer period of time'' during which an action is performed. A ``long-term'' or ``\isi{perdurative}'' Aktionsart is postulated by \cite[112--113]{zalizniak2000}. The authors claim that this Aktionsart includes ``verbs denoting an action that took place within a closed specific period of time''. In formal literature, we find similar opinions on the semantics of the \isi{perdurative}. For example, \cite[26]{gehrke2003aspectual} indicates that the \isi{perdurative} ``refers to an unexpectedly long duration of a situation, where it always has to be made explicit that this duration is specific''. \cite[32]{filip2005measures} notes that the \isi{perdurative} \textit{pro-} ``indicates a relatively long temporal extent of the \isi{event} (with connotations of wasted time, boredom, and the like).''

Although the inference that the \isi{eventuality} from the extension of the verbal predicate marked by the \isi{perdurative prefix} \textit{pro-} lasts long is indeed frequently present, it is, in fact, optional and can be canceled by some linguistic means. Consider, e.g., the sentence in \REF{ex:naumov:21}, where the \isi{perdurative} \isi{verb} is followed by the adverb \textit{vsego liš’} `just, only'.


\ea \label{ex:naumov:21}
\gll  V	očeredi		Kolja	pro-stoja-l		vsego	liš’	sutki.\\
in	queue		K.	\textsc{pro}$_\textsc{per}$-stay-\textsc{pst} total	only	day \\
\glt `Kolja stood in the queue only for a day.'
\z

\noindent Here, the use of the adverb explicitly indicates that an \isi{event} of staying described by the \isi{perdurative} \isi{verb} \textit{pro-stojat’} `stay' did not last long \isi{relative} to expectations. The fact that the inference of long duration arises in the presence of the \isi{perdurative} \textit{pro-} and can be absent in the appropriate context, suggests that this inference is an \isi{implicature}. The generation of implicatures is generally taken to be due to the non-use of a non-weaker alternative: the choice of one alternative implies the \isi{negation} of the other. The only candidate for the position of the alternative item for the \isi{perdurative} \textit{pro-} is the so-called \isi{delimitative} prefix \textit{po-}, which also conveys the meaning of temporal duration. Indeed, the \isi{perdurative} and the \isi{delimitative} are usually considered together and sometimes assumed to impose similar restrictions on the predicates with which they combine (see, e.g., the references above). Constraints on the distribution of the \isi{delimitative} have been recently explored by \citet{tatevosov2017temporal}. In the next subsection, I will give an outline of the theory developed in these works and take it as a starting point in identifying constraints on the distribution of the \isi{perdurative}.

\subsection{The delimitative and unique temporal arrangement} \label{sec:naumov:3.3}

\citet{tatevosov2017temporal} claims that in \ili{Russian} the derivation of non-culminating accom\-plish\-ments—predicates that appear in \isi{perfective} clauses and describe some goal-oriented activity that stops before the corresponding \isi{result state} is attained—proceeds in two steps: the \isi{secondary imperfective suffix} \textit{-yva} first merges with a formally \isi{perfective} stem and then the \isi{delimitative} prefix \textit{po-} attaches to the resulting complex.\footnote{\citet{tatevosov2017temporal} proposes that non-culminating accomplishments exist in \ili{Russian}, and the \isi{delimitative} \textit{po-} is a means to derive them. There are, however, alternative views on this problem. For example, \citet{martin2017non} claims that \ili{Russian} does not possess non-culminating accomplishments, and the delimitative-prefixed verbs do not exemplify this class of predicates. I believe that this theoretical debate is orthogonal to the current purposes.} Each step is subject to certain semantic restrictions. The ones that we are interested in here are those that emerge when the \isi{delimitative} comes into play. \citet{tatevosov2017temporal} proposes that the \isi{delimitative} can be derived from an \isi{event predicate} iff the \isi{activity component} of this predicate does not contain in its extension subevents that are temporally ordered in a unique way. The notion of unique temporal orderedness is formalized through the property of \textsc{unique temporal arrangement} (\isi{UTA}). An informal definition of \isi{UTA} is given in \REF{ex:naumov:22}.

\eanoraggedright \textit{Definition:} Unique \isi{temporal arrangement}\smallskip\\
Whenever an \isi{event} $e$ falls under $P$, there is exactly one way for $e$ to start, there is exactly one way for $e$ to finish, and for any non-final part of $e$ there is exactly one follow up.\label{ex:naumov:22}
\z

\noindent This property is best understood by looking at concrete examples. Let us compare predicates \textit{zapivat’ tabletku} `wash down a pill' and \textit{zapolnjat’ anketu} `fill in a form', which are both based on \isi{secondary imperfective} \isi{verb} stems and within our assumption denote non-final parts of complete eventualities. The predicate \textit{zapivat’ tabletku} `wash down a pill' does not allow the formation of the \isi{delimitative}, while \textit{zapolnjat’ anketu} `fill in a form' does. These predicates differ with respect to the \isi{UTA} property. The \isi{activity component} of the predicate \textit{zapivat’ tabletku} `wash down a pill' consists of very specific subevents that have to come in a very specific order to represent an activity of washing down a pill. In contrast, the predicate \textit{zapolnjat’ anketu} `fill in a form' denotes events in which subevents making up the activity of filling in a form do not have to come in any specific order. Even if some of them are skipped or occur more than once, their sum still represents an activity of filling in a form. According to \citet{tatevosov2017temporal}, the \isi{delimitative} \textit{po-} is sensitive to this difference. Crucially, the \isi{delimitative} \textit{po-} can also combine with predicates based on simplex \isi{imperfective} \isi{verb} stems, like \textit{govorit’ po telefonu} `talk on a phone', which are associated with an acitivity \isi{event structure} and also do not possess \isi{UTA}. Putting these facts together, \citet{tatevosov2017temporal} concludes that the complement of the \isi{delimitative}, whether it is derived or non-derived, must be an Activity.

The important fact for the purposes of this work is that, despite superficial similarity, the \isi{perdurative} and the \isi{delimitative} impose different requirements on predicates with which they combine. As was shown in \sectref{sec:naumov:1}, the \isi{perdurative}, unlike the \isi{delimitative}, can be derived from some activity predicates but cannot be derived from others. Consider the contrast between \REF{ex:naumov:23} and \REF{ex:naumov:24}.\footnote{It must be pointed out that the \isi{verb} \textit{taskat’} in \REF{ex:naumov:23b} and \REF{ex:naumov:24b} is an ``indeterminate'' motion \isi{verb}. Indeterminate motion verbs are used to describe undirected motion, motion back and forth, motion that is not associated with any particular path being covered during the unfolding of an \isi{event}. Neither the \isi{delimitative} nor the \isi{perdurative} can be derived from a predicate when it describes motion in a single direction.}

\ea \label{ex:naumov:23} \ea [*]{\label{ex:naumov:23a}
\gll  Kolja		pro-pisa-l		pismo		neskol’ko	minut.\\
K. \textsc{pro}$_\textsc{per}$-write-\textsc{pst}   letter		several		minutes    \\
\glt Intended: `Kolja spent several minutes writing the letter.'}
\ex[]{\label{ex:naumov:23b}
\gll Kolja	pro-taska-l		čemodan	neskol’ko	časov. \\
K. \textsc{pro}$_\textsc{per}$-carry-\textsc{pst} suitcase	several		hours \\
\glt `Kolja spent several hours carrying the suitcase.'}
\z \z

\ea \label{ex:naumov:24} \ea \label{ex:naumov:24a}
\gll  Kolja	po-pisa-l		pismo		neskol’ko	minut.\\
K. \textsc{po}$_\textsc{del}$-write-\textsc{pst}   letter		several		minutes    \\
\glt `Kolja spent several minutes writing the letter.'
\ex \label{ex:naumov:24b}
\gll Kolja	po-taska-l		čemodan	neskol’ko	časov. \\
K. \textsc{po}$_\textsc{del}$-carry-\textsc{pst} suitcase	several		hours \\
\glt `Kolja spent several hours carrying the suitcase.'
\z \z

\noindent Since the boundary separating predicates that allow the \isi{perdurative} from predicates that do not allow it runs inside the class of activities, more should be said about their internal structure. A more specific question that I would like to address is: are there any properties that differentiate predicates like \textit{pisat’ pismo} `write a letter’ from predicates like \textit{taskat’ čemodan} `carry a suitcase’? The next subsection is devoted to seeking an answer to this question.

\subsection{Theories of atelicity} \label{sec:naumov:3.4}

In the literature on \isi{aspectual} composition, it is consistently argued that activity predicates are atelic. What lies under this notion varies across different approaches. In their influential work, carried out within the framework of temporal semantics, \citet{bennett1978toward} propose that atelic predicates must satisfy the \textsc{subinterval property}.

\ea \label{ex:naumov:25}
\textit{Definition:} Subinterval property\smallskip\\
$\cnst{sub}(P) \leftrightarrow \forall i [\cnst{at}(P, i) \shortrightarrow \forall j[j \subset i \shortrightarrow \cnst{at}(P, j)]]$ \\
\textit{P} possesses the \isi{subinterval property} iff, if \textit{P} is true at $i$, it is true at every subinterval of $i$.
\z

\noindent This definition, however, is too strong and gives rise to the well-known ``minimal-parts problem'' \citep{dowty1979word}. The problem is that it is not the case that all activity predicates are true at every subinterval during which the events from their extension take place. Dynamic eventualities never hold at points, they take time to establish themselves. Therefore, it is reasonable to judge whether a given predicate possesses the \isi{subinterval property} only \isi{relative} to an interval that is ``sufficiently large" for the \isi{event} from its extension to unfold.

Now that we have equipped ourselves with the notion of \isi{subinterval property}, we can return to the question formulated at the end of the previous subsection and consider the predicates \textit{pisat’ pismo} `write a letter’ and \textit{taskat’ čemodan} `carry a suitcase’ in detail. These predicates are based on underived \isi{imperfective} stems and, within the assumption introduced in \sectref{sec:naumov:3.1}, denote non-final parts of complete events and belong to the class of activities. The \isi{subinterval property} is not what distinguishes them: both \textit{pisat’ pismo} `write a letter’ and \textit{taskat’ čemodan} `carry a suitcase’ possess this property. If a nonfinal part of an \isi{event} of writing a letter holds at an interval $i$, it also holds at every sufficiently large subinterval of $i$. Similarly, if a non-final part of an \isi{event} of carrying a suitcase holds at an interval $i$, it also holds at every sufficiently large subinterval of $i$. The \isi{perdurative}, however, can be derived only from the latter but not from the former.

In the mereological approach to \isi{aspectual} composition \citep{krifka1989nominal,krifka1989nominal, krifka1992nominal,krifka1998origins}, atelic predicates are defined as being cumulative. A predicate is \textsc{cumulative} if it satisfies the properties of \textsc{additivity}, \REF{ex:naumov:26}, and \textsc{divisivity}, \REF{ex:naumov:27}.

\ea \label{ex:naumov:26}
\textit{Definition:} Additivity\smallskip\\
$\forall P[\cnst{cum}(P) \leftrightarrow \forall x \forall y[P(x) \wedge P(y) \shortrightarrow P(x \oplus y)]]$ \\
$P$ is additive iff whenever it applies to the entities $x$ and $y$, it also applies to the sum $x \oplus y$
\z

\ea \label{ex:naumov:27}
\textit{Definition:} Divisivity\smallskip\\
$\forall P[\cnst{div}(P) \leftrightarrow \forall x \forall x' [P(x) \wedge x' \subset x \shortrightarrow P(x')]]$ \\
$P$ is divisive iff whenever $P$ applies to $x$, then it must also apply to any $x'$ that is properly included in $x$.
\z

\noindent The notion of cumulativity also cannot help us to distinguish \textit{pisat’ pismo} `write a letter’ from \textit{taskat’ čemodan} `carry a suitcase’. Both these predicates are additive and divisive down to minimal parts.\footnote{Due to space limitations, I leave the verification of this statement to the reader.}

\citet{landman2010incremental,landman2012felicity} claim that neither the \isi{subinterval property} nor the property of cumulativity is adequate for distinguishing activities from other classes of predicates. They argue that activity predicates are lexically constrained as being incrementally homogeneous.

\textsc{Incremental homogeneity} is based on two essential components: \textsc{cross-temporal identity} and \textsc{\isi{event} onsets}. Cross-temporal identity is a semantic primitive that is used to compare events with different running times. Events with different running times count as the same \isi{event} if they are in an equivalence relation of \isi{cross-temporal identity}, \REF{es28}.

\eanoraggedright \label{es28}
\textit{Definition:} Equivalence relation of \isi{cross-temporal identity}\smallskip\\
$e_1$ is cross-temporally identical to $e_2$, $e_1 \sim e_2$ iff $e_1$ and $e_2$ count as one and the same \isi{event}, i.e. for counting purposes $e_1$ and $e_2$ count as one \isi{event}.
\z

\noindent The second component is the notion of onset of an \isi{event}. For $e$, the onset of $e$ is the smallest initial part of $e$ that is big enough to count both as $e$ and as cross-temporally identical to $e$, \REF{ex:naumov:29}.

\ea \label{ex:naumov:29}
\textit{Definition:} Onset of an \isi{event}\smallskip\\
Let $e$ be an \isi{eventuality} of \isi{verb} type $V$. \\
The onset of $e$, \isi{relative} to $V$, $O(e,V)$ is the smallest \isi{eventuality} of type $V$ such that: $O(e,V) \sim e$ and $\tau(O(e,V) \subseteq_{\cnst{in}} e$.
\z

\noindent Taking these, the notion of \isi{incremental homogeneity} is defined, \REF{ex:naumov:30}.
\ea \label{ex:naumov:30}
\textit{Definition:} Incremental homogeneity\smallskip\\
Let $\alpha$ be a VP with \isi{event type} $\alpha$ and verbal \isi{event type} $V_\alpha$.

Let $e\in V_\alpha$ and $e \in \alpha$. \\
$e$ is incrementally homogeneous w.r.t. $\alpha$ and $V_\alpha$ iff for every interval $i$: if $\tau(O(e,V)) \subseteq_{\cnst{in}} i \subseteq_{\cnst{in}} \tau(e)$ then there is an \isi{eventuality} $e'$ of \isi{event type} $\alpha$ such that: $e' \sim e$ and $\tau(e') = i$.
\z

\noindent For an \isi{event} $e \in \alpha$ to be incrementally homogeneous, the onset of $e$ must count both as \isi{event type} V$_\alpha$ and as \isi{event type} $\alpha$. If $\alpha$ does not hold at the onset of $e$, neither $e$ is incrementally homogeneous nor the predicate that is true of $e$. This is exactly what happens to predicates like \textit{pisat’ pismo} `write a letter’. Let the predicate \textit{pisat’ pismo} `write a letter’ be an \isi{event type} $\alpha$, with verbal \isi{event type} V$_\alpha$, which is \textit{pisat’} `write’.

\ea \label{ex:naumov:31} \ea
 $\alpha = \lambda e. \textsc{write}(e) \wedge \cnst{theme}(\textsc{letter})(e)$ \\
    Event type of writing letter events
\ex $V_\alpha = \lambda e. \textsc{write}(e)$ \\
    Event type of writing events
\z \z

\noindent Suppose that $e$ is a non-final part of an \isi{event} of writing a letter. According to the definition, the onset of $e$ is its most initial proper part that counts as a writing activity. It is not the case, however, that this onset must count as a non-final part of an \isi{event} of writing a letter. An \isi{event} of someone diligently tracing out the first letter of the letter undoubtedly counts as writing but very unlikely counts as writing a letter. Thus, the predicate \textit{pisat’ pismo} `write a letter’ is not incrementally homogeneous, and the \isi{perdurative} cannot be derived from it.

In contrast, \textit{taskat’ čemodan} `carry a suitcase’ is an incrementally homogeneous predicate. Suppose that $e$ is a non-final part of an \isi{event} of carrying a suitcase. According to the definition, the onset of $e$ is its most initial proper part that counts as a carrying activity. What is crucial is that this onset counts also as a non-final part of an \isi{event} of carrying a suitcase. An \isi{event} of someone taking a suitcase and making a few steps in different directions counts both as an \isi{event} of carrying and as an \isi{event} of carrying a suitcase. Thus, the predicate\textit{ taskat’ čemodan} `carry a suitcase’ is incrementally homogeneous, and the \isi{perdurative} can be derived from it.

Can we stop here and say that the restriction on \isi{incremental homogeneity} is what regulates the distribution of the \isi{perdurative}? The answer is negative. There are incrementally homogeneous predicates that are marginal with the \isi{perdurative}. These are the predicates in which, as \citet[106]{tatevosov2009event} claim, ``the activity that up to its final point does not contribute to the development of the become \isi{subevent} at all''. An example of such a predicate is \textit{otkryvat’ zamok} `open a lock’. This is an atelic predicate obtained from its telic counterpart \textit{otkryt’ zamok} `open a lock’ through applying the \isi{secondary imperfective suffix} \textit{-yva}. Within our assumption, this predicate denotes non-final parts of a complete \isi{event} of opening a lock. These non-final parts do not induce any change of state. The \isi{result state} is brought about by the very final \isi{subevent}. This \isi{subevent}, however, is not in the denotation of the \isi{secondary imperfective} \textit{otkryvat’ zamok} `open a lock’.

Consider the example in \REF{ex:naumov:32}. In the provided situation, the sentence is judged as ungrammatical.

\ea \label{ex:naumov:32} \textit{Situation:} The lock in the door is broken. After spending half an hour trying to open it, Vasja gave up.\smallskip\\
\gll * Vasja		pro-[[otkr]-yva]-l		zamok		polčasa.\\
    {} V.		\textsc{pro}$_\textsc{per}$-open-\textsc{yva}-\textsc{pst}	lock		half.an.hour\\
\glt \ \ Intended: `Vasja spent half an hour opening the lock.'
\z

\noindent It is easy to show that this predicate is indeed incrementally homogeneous. Let $e$ be a non-final part of a complete \isi{event} of opening a lock. According to the definition of onset given above, the onset of $e$ is its most initial proper part that counts as an opening activity. Crucially, this onset counts as a non-final part of an \isi{event} of opening a lock, too. The very first manipulation with a lock aimed at unlocking it counts both as opening and as opening a lock. Thus, the predicate \textit{otkryvat’ zamok} `open a lock’ is incrementally homogeneous. However, the \isi{perdurative} is blocked.

In the next subsection, I will try to identify the property that distinguishes predicates that allow the \isi{perdurative} from predicates that do not and try to formulate a unified semantic constraint that regulates the distribution of the \isi{perdurative} \textit{pro-} in \ili{Russian}.

\subsection{The perdurative and arrangement by stages} \label{sec:naumov:3.5}

I start this subsection from the observation, previously presented in \sectref{sec:naumov:1}, that the \isi{perdurative} can be derived from predicates based on the \isi{secondary imperfective} \isi{verb} stems only if they receive a \isi{pluractional interpretation}, \REF{ex:naumov:33}.

\ea{ \label{ex:naumov:33}
\gll Kolja 	pro-[[otkr]-yva]-l 		rot 		minutu.\\
     K. \textsc{pro}$_\textsc{per}$-open-\textsc{yva}-\textsc{pst} mouth 		minute\\
\glt `Kolja spent a minute opening the mouth.'
\ea[\ding{55}]{\ Scenario №1: Kolja spent a minute opening the mouth wider and wider until he got cramps in his cheekbones.}
\ex[\ding{51}]{Scenario №2: Kolja spent a minute opening the mouth several times until he realized that he was not producing any sounds.}\z}
\z

\noindent The above contrast suggests that the distinguishing property of pluractional predicates might serve as a clue to our understanding what the constraint on the distribution of the \isi{perdurative} is. There have been proposed various analyses of \isi{event} pluractionality in formal semantic literature (\citealt{lasersohn1995plurality,van2004adverbials,tovena2010pluractionality,henderson2012ways}, a.o.). Space prevents me from reviewing them in detail here. To put it briefly, the basic idea shared by these works is that a pluractional marker takes an underlying predicate and ensures that there is a multiplicity of atomic events of the same type. Importantly, these atomic events are self-sufficient, they do not stand in any special temporal (like \isi{UTA}) or causal relation to each other. I capture this property by utilizing \citeposst{landman1992progressive} \isi{stage-of relation}.

The notion of a \textsc{stage} has been proposed by \citet{landman1992progressive} to define the semantics of the progressive in \ili{English}. He claims that the progressive is a function from a set of events denoted by VP to a set of stages of those events. A progressive sentence is true if a VP-\isi{event} stage develops into an \isi{event} of the same kind denoted by VP. Stage-of relation is a partial ordering of the set of events. For an \isi{event} $e$ to be a stage of another \isi{event} $e′$, $e$ must share enough characteristics with $e′$ and must develop into $e′$ in some possible world that is near enough to the world of evaluation.

Pluractional predicates are organized in a special way with respect to the \isi{stage-of relation}: although every atomic \isi{event} from the extension of the pluractional predicate is a stage of the plural \isi{event} denoted by this predicate, no atomic \isi{event} is a stage of any other atomic \isi{event}. I claim that other predicates allowing the derivation of the \isi{perdurative} are organized exactly as pluractional predicates with respect to the \isi{stage-of relation}. I propose that for a predicate to be able to form the \isi{perdurative}, subevents making up the \isi{activity component} of this predicate must not be arranged by the \isi{stage-of relation}. Below I will argue that if a predicate blocks the formation of the \isi{perdurative}, it possesses the property of \textsc{arrangement by stages} (\isi{AbS}). A formal definition of \isi{AbS} is given in \REF{ex:naumov:34}.

\ea \label{ex:naumov:34}
Let $e$ be a \isi{partial eventuality} from the extension of an \isi{event predicate} P, and \\
Let $e'$ and $e''$ be stages of $e$ such that
\ea $e' \subseteq e$, and
\ex $e'' \subseteq e$, and\smallskip\\
P($e$) possesses \isi{AbS} iff \\
$\forall e' \forall e''\,[\,e' \ll_{\cnst{t}} e'' \rightarrow e'$ is a stage of $e'']$\smallskip\\
where $\ll_{\cnst{t}}$ is a temporal precedence relation on events.
\z \z

\noindent \REF{ex:naumov:34} says that an \isi{event predicate} P is an \isi{AbS} predicate iff for all contextually salient subevents $e'$ and $e''$ in its denotation such that both $e'$ and $e''$ are stages of $e$, if $e'$ temporally precedes $e''$, then $e'$ is a stage of $e''$. \isi{AbS} is stronger than \isi{incremental homogeneity}. I claim that this property is what distinguishes predicates that allow the \isi{perdurative} from predicates that do not.

Let us consider the predicate \textit{otkryvat’ zamok} `open a lock' again. Imagine a scenario in which one has a bunch of numbered keys but does not know which one opens the lock. Suppose that a stage of a \isi{partial eventuality} $e$ from the extension of the predicate \textit{otkryvat’ zamok} `open a lock' is an \isi{event} of using key number one. Let it be $e'$. Suppose that another stage of $e$ is an \isi{event} of using key number two. Let it be $e''$. In addition, assume that $e'$ temporally precedes $e''$. Clearly, both $e'$ and $e''$ are stages of the bigger \isi{event} $e$. The crucial fact is that $e'$ is a stage of $e''$, too. According to the original definition proposed in \citet{landman1992progressive}, ``an \isi{event} is a stage of another \isi{event} if the second can be regarded as a more developed version of the first, that is, if we can point at it and say, ``It's the same \isi{event} in a further stage of development.'' \citep[23]{landman1992progressive}. Although \citet{landman1992progressive} does not provide an explanation for what it is to be ``a more developed version'', intuitively, a more developed \isi{event} is nothing more than a next step a rational agent takes to achieve the desired goal. Returning to the discussed scenario, if the agent loses the hope of opening the lock with the first two keys, it is very unlikely that she uses key number one again. In case she picks it and performs the same kind of activity as before (without changing the turning direction of the key, the applied force, etc.), this can hardly count as a more developed version of the previous \isi{event}, which is a sum of using key number one and using key number two, because this activity does not bring the agent closer to her goal—have the lock opened. For an \isi{event} to fall under the denotation of \textit{otkryvat’ zamok} `open a lock', this \isi{event} has to consist of subevents such that each subsequent \isi{subevent} is a more developed version of the previous one. That is, \textit{otkryvat’ zamok} `open a lock' is an \isi{AbS} predicate. This is the reason why the \isi{perdurative} is not allowed from it.

As was shown in the previous sections, predicates like \textit{govorit’ po telefonu} `talk on the phone' and \textit{taskat’ čemodan} `carry a suitcase' allow the formation of the \isi{perdurative}. My proposal correctly accounts for this fact. Let us consider the predicate \textit{taskat’ čemodan} `carry a suitcase'. Indeterminate verbs of motion in \ili{Russian} can express a wide range of meanings \citep{forsyth1970grammar}. In episodic contexts, the \isi{perdurative} is licensed when a predicate has the so-called multiple directions reading. This reading arises in a scenario with a single \isi{event} of motion ``in various unspecified directions'' (\citealt{forsyth1970grammar}: 321). It is easy to show that when \textit{taskat’ čemodan} `carry a suitcase' receives the multiple direction reading, it does not possess \isi{AbS}. Suppose that a stage of a \isi{partial eventuality} $e$ from the extension of the predicate \textit{taskat’ čemodan} `carry a suitcase' is an \isi{event} of moving in one particular direction. Let it be $e'$. Suppose that another stage of $e$ is an \isi{event} of moving in the other direction. Let it be $e''$. Assume that $e'$ temporally precedes $e''$. Both $e'$ and $e''$ are stages of the bigger \isi{event} $e$: an \isi{event} of moving in multiple directions is a more developed version of an \isi{event} of moving in a single direction. However, an \isi{event} of moving in one direction can hardly count as a more developed version of an \isi{event} of moving in the other direction. Moreover, after the agent tried to go in two different directions, she can return to the starting point and take the first direction again. This action will still be in the extension of the predicate \textit{taskat’ čemodan} `carry a suitcase'. For an \isi{event} to fall under its denotation, it does not need to consist of subevents such that every subsequent \isi{subevent} is a more developed version of the previous one. The same is true for the predicate \textit{govorit’ po telefonu} `talk on a phone'.

\largerpage
The \isi{perdurative} also cannot be derived from predicates associated with an incremental relation between the activity and change of state components. Predicates of this type denote events such that for every activity \isi{subevent} there is a change of state that it induces \citep{rothstein2004structuring}. An instance of such a predicate is \textit{pisat’ pismo} `write a letter'. When one writes a letter, every part of the writing activity corresponds to some part of the process of being written. The \isi{perdurative} cannot be derived from this predicate, \REF{ex:naumov:35}.

\ea[*] {\label{ex:naumov:35}
\gll Kolja		pro-pisa-l		pismo	vsjo	utro.\\
     K.		\textsc{pro}$_\textsc{per}$-write-\textsc{pst}	letter	all	morning\\
\glt Intended: `Kolja spent the whole morning writing the letter.'}
\z

\noindent My proposal predicts it because this predicate possesses \isi{AbS}: an activity of writing a letter consists of subevents such that every subsequent \isi{subevent} is a more developed version of the previous one. Let me show why. Suppose that a stage of a \isi{partial eventuality} $e$ from the extension of the predicate \textit{pisat’ pismo} `write a letter' is an \isi{event} of writing an address. Let it be $e'$. Suppose that another stage of $e$ is an \isi{event} of writing a salutation. Let it be $e''$. Assume that $e'$ temporally precedes $e''$. Since each of these subevents makes a contribution to the state of being written, at the moment when the agent writes the salutation the letter is already finished to the extent of the address. The \isi{subevent} of writing the salutation extends the degree to which the letter is finished. Therefore, it counts as a more developed version of the \isi{subevent} of writing the address. After the salutation is done, the agent can return to the address-part of the letter and, e.g., correct the name of the street or re-write it anew if she realizes that she has made a mistake, and this action will count as a more developed \isi{subevent} and will be in the extension of the predicate \textit{pisat’ pismo} `write a letter'. What she is very unlikely to do is to perform the same kind of activity and write the address again without any changes. The action of writing the address again will not count as a more developed \isi{subevent} and will not be in the extension of the predicate \textit{pisat’ pismo} `write a letter'.

\section{Conclusion} \label{sec:naumov:4}

In this study, I have analyzed (morpho)syntactic and semantic constraints on the derivation of the \isi{perdurative} in \ili{Russian}. Considering the restrictions that the \isi{perdurative prefix} \textit{pro-} demonstrates when combined with a number of other affixes, I have argued that this prefix is a \isi{selectionally restricted} prefix that merges in the functional domain between \textit{v}P and \isi{AspP}. I have shown that the \isi{perdurative} \textit{pro-} selects for grammatically \isi{imperfective} stems and combines with predicates based on \isi{secondary imperfective} \isi{verb} stems only when they receive a \isi{pluractional interpretation}. I have demonstrated that the distribution of the \isi{perdurative} is also constrained semantically. Having started from the observation that the \isi{perdurative} can be derived from predicates associated with an activity \isi{event structure} and having discussed the existing proposals for what it is to be an activity predicate, I concluded that none of them is able to account for its distribution. I have argued that predicates allowing the formation of the \isi{perdurative} do not possess the \isi{AbS} property. In other words, activity subevents from their extension do not stand in \citeposst{landman1992progressive} \isi{stage-of relation} to each other.

Several questions remain. How does the analysis developed here account for stative predicates like \textit{žit’ v Moskve} `live in Moscow', which allow the \isi{perdurative}? In principle, the proposed analysis predicts the availability of the \isi{perdurative} from statives because statives do not have stages \citep{landman1992progressive}, and hence there is simply nothing that can be arranged by the \isi{stage-of relation}. This idea, if correct, needs to be spelled out in more detail. If the \isi{perdurative} is not responsible for introducing \isi{perfectivity}, what is its semantic contribution? A potential line to follow, as I see it, is to make use of the notion of maximality, recently much discussed in the literature (see \citealt{altshuler2014,filip2017semantics}, a.o.). The \isi{perdurative} can be treated as an \isi{event} modifier extracting a maximal stage (or part) of a \isi{partial eventuality}. The elaboration of this hypothesis is left for future research.

\section*{Abbreviations}
\begin{tabularx}{.45\textwidth}{lX}
{AbS} & arrangement by stages\\
\textsc{del}&{delimitative}\\
{LP}&{lexical prefixes}\\
\textsc{per}&{perdurative}\\
{PR} & {positionally restricted}\\
&\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lX}
\textsc{pst}&past\\
\textsc{rep}&{repetitive}\\
{SLP}&{super lexical prefixes}\\
{SR} & {selectionally restricted}\\
{UTA}& {unique temporal arrangement}\\
\end{tabularx}

\section*{Acknowledgments}
The present paper has benefited a lot from comments and questions that I received from two anonymous reviewers as well as from the audience at FDSL 12.5. Many thanks to all these people. I am especially grateful to Sergei Tatevosov whose support and whose own work have influenced this project enormously. This work would have not been completed without Lena Pasalskaya's expertise in \LaTeX. Needless to say that all errors and shortcomings are my own responsibility.



\sloppy
\printbibliography[heading=subbibliography,notkeyword=this]

\end{document}
