\addchap{\lsAcknowledgementTitle} 

This book is a modified version of my Ph.D. dissertation, which was defended in May 2017 at University of New Mexico. While the book features an updated language sample, the findings and interpretations are largely unchanged from the original study. Neither work would have been possible without the mentorship and support of many people along the way.

I am particularly grateful to the members of my dissertation committee. My advisor, Caroline Smith, provided helpful and constructive feedback on all aspects of the dissertation and my progress through the Ph.D. Joan Bybee and Ian Maddieson introduced me to language change and phonological typology, respectively, and I have greatly benefited from the inspiring work and mentorship of both. I thank Ioana Chitoran for many thought-provoking discussions of the findings and the next steps in this line of research. Thanks also to Bill Croft.

While writing the original dissertation I was fortunate to have the financial support of the Joseph H.\ Greenberg Endowed Fellowship (2011--2015) and the Russell J.\ and Dorothy S.\ Bilinski Fellowship (2016--2017).

Since October 2017, I have been employed as a postdoctoral researcher at Laboratoire Dynamique du Langage (UMR5596, CNRS \& Université Lyon 2) in Lyon, France. While revising this book, I have been funded by LabEx ASLAN (ANR-10-LABX-0081) of the Université de Lyon within the \ili{French} program Investissements d’Avenir. DDL has been a warm, welcoming, and intellectually exciting environment in which to work, and for this I am grateful to my supervisor François Pellegrino, laboratory director Antoine Guillaume, and Anetta Kopecka, the head of the Description, Typologie, et Terrain (DTT) research axis. Special thanks also go to my fellow ``postdoc ladies'' -- Natalia Chousou-Polydouri, Marine Vuillermet, Laetitia de Almeida, Elisa Demuru, and Natasha Aralova -- for their friendship and moral support during this stage of our lives and careers, and to DDL alumna Kasia Janic for the use of her magical apartment.

Many specialists kindly answered my inquiries about details of the languages and language families discussed in this book, including Doug Marmion (\ili{Wutung}), Matthew Carter (\ili{Bashkir}), Netra P.\ Paudyal (\ili{Darai}), Jonathan Bobaljik (\ili{Itelmen}), Andrey Filchenko (\ili{Eastern Khanty}), Bonny Sands (\ili{Hadza}), Kirk Miller (\ili{Hadza}), Logan Sutton (Kiowa-Tanoan), Rosa Vallejos (\ili{Cocama-Cocamilla}), Sylvia Tufvesson (\ili{Semai}), Kasia Wojtylak (\ili{Murui Huitoto}), and Albert Alvarez Gonzalez (Uto-Az\-tecan). Tim Zingler, Ricardo Napoleão de Souza, Lexie Adams, and Marc\linebreak Adams helped me with the translation of reference material from the original \ili{German}, Portuguese, and \ili{Spanish}. George Moroz helped construct and format the maps.

Enormous thanks to the team at Language Science Press for all their assistance with improving this work and moving it towards publication, particularly Martine Grice, Sebastian Nordhoff, Felix Kopecky, an anonymous reviewer, and the community proofreading volunteers.

I am grateful to many dear friends for their support over the years that this work was in progress, including Caron Treloar, Lexie Adams, Erin Debenport, Cora McKenna, Francesca Di Garbo, Giorgio Iemmolo, Logan Sutton, Jason\linebreak Timm, Ricardo Napoleão de Souza, Fabian Armijo, Nick Mathews, Amanda Delgado Galvan, Maurice Pico, Sylvia Tufvesson, Geny Gonzalez Castaño, and Esteban Díaz Montenegro.

Finally, to my family: thank you for everything.
