\addchap{\lsAcknowledgementTitle} 


\section*{Sources of illustrations}

Origins of the photographs used as illustrations to the text are
provided below. Where possible, attribution and/or permission for use
are provided. Wherever possible, I have attempted to contact the
sources of photos to obtain permission for their use. Where only a URL
is given, the photo used was found on that site, but no information
was provided there as to the source.

\begin{description}
\item[Figure~\ref{fig:ch.saussure_life.young_saussure}] ``Ferdinand de
  Saussure as a young man'': Photo in de Saussure family collection,
  used with permission. 
\item[Figure~\ref{fig:ch.saussure_life.saussure_family}] ``Ferdinand
  de Saussure with his wife Marie and sons Jacques [l.]  and Raymond
  [r.]'' Photo in de Saussure family collection, used with permission.
\item[Figure~\ref{fig:ch.saussure_sound.saussure}] ``Ferdinand de
  Saussure'': Photo by ``F. H. Jullien Genève'', Frank-Henri Jullien
  (1882–1938), in de Saussure family collection, used with permission.
\item[Figure~\ref{fig:ch.kazan_baudouin}] ``Jan Niecisław Baudouin de
  Courtenay''.
  \url{https://en.wikipedia.org/wiki/Jan_Baudouin_de_Courtenay}. Public
  Domain.
\item[Figure\ref{fig:ch.kazan_kruszewski}] ``\name{Mikołaj}{Kruszewski}''.
  Wikipedia File:Крушевский, Николай Вячеславович
  (1851—1887).jpg. Public Domain.
\item[Figure~\ref{fig:ch.kazan_scerba}] ``\name{Lev Vladimiovič}{Ščerba}''.
  \url{https://ru.wikipedia.org/wiki/Щерба,_Лев_Владимирович}. Origin
  unknown.
\item[Figure~\ref{fig:ch.prague_jakobson_1916}] ``\name{Roman}{Jakobson} in
  1916''.
  \url{https://upload.wikimedia.org/wikipedia/commons/b/b5/Roman_Yakobson.jpg}.
  Public Domain.
\item[Figure~\ref{fig:ch.prague_trubetzkoy_younger}] ``Prince Nikolai
  Sergeievič Trubetzkoy (1927)''. 
  Archive of the University of Vienna, picture archive Originator:
  Theo Bauer, Wien XIV. Signatur: 106.I.1278. 
\item[Figure~\ref{fig:ch.prague_trubetzkoy_prague_circle}] ``Detail
  from the 1930 International Phonology Meeting''. Detail from group
  photo at \url{http://www.philologoz.ru/trubetzkoy/plk1930.jpg}.
\item[Figure~\ref{fig:ch.prague_trubetzkoy_1939}] ``Prince Nikolai
  Sergeievič Trubetzkoy''. Frontispiece to
  \citealt{trubetzkoy39:grundzuge}.
\item[Figure~\ref{fig:ch.prague_trubetzkoy}] ``Prince Nikolai
  Sergeievič Trubetzkoy''.
  \url{https://commons.wikimedia.org/wiki/File:Trubetzkoy.jpg}. Public Domain.
\item[Figure~\ref{fig:ch.jakobson_jakobson_1920}]  "\name{Roman}{Jakobson} in
  1920".  \url{https://monoskop.org/Roman_Jakobson}.
\item[Figure~\ref{fig:ch.jakobson_jakobson_levi-strauss}] ``Roman
  Jakobson and \name{Claude}{Levi-Strauss} (1972)''. 
  \url{https://journals.openedition.org/lettre-cdf/docannexe/image/210/img-2.jpg}. 
\item[Figure~\ref{fig:ch.jakobson_jakobson_lsa}] ``\name{Roman}{Jakobson} in
later years''. Used with permission of the family of Rosamond
T. Halle.
\item[Figure~\ref{fig:ch.hjelmslev.young_hjelmslev}] ``Louis Trolle
  Hjelmslev as a young MA''
  \url{https://cc.au.dk/infrastrukturalisme/om-louis-hjelmslev-og-hans-kreds/billeder-hvordan-saa-de-ud/}
  Used with permission.
\item[Figure~\ref{fig:ch.hjelmslev.uldall-hjelmslev}] ``Hans Jørgen
  Uldall and Louis Hjelmslev''
  \url{https://cc.au.dk/en/infrastructuralism/archives/louis-hjelmslevs-archive-at-the-royal-library/louis-hjelmslev-and-hans-joergen-uldall/. Used
  with Permission.}
\item[Figure~\ref{fig:ch.hjelmslev.hjelmslev2}] ``Louis
  Hjelmslev''. \url{https://cc.au.dk/en/infrastructuralism/archives/louis-hjelmslevs-archive-at-the-royal-library/louis-hjelmslev-and-jens-holt/}. Used
  with permission.
\item[Figure~\ref{fig:ch.hjelmslev.hjelmslev3}] ``Louis Hjelmslev''
  \url{https://literariness.org/2018/03/19/key-theories-of-louis-hjelmslev/}. 
\item[Figure~\ref{fig:ch.hjelmslev.hjelmslev}] ``Louis Hjelmslev''
  \url{https://www.carlsbergfondet.dk/da/Forskningsaktiviteter/Forskningsprojekter/Andre-forskningsprojekter/Hans-Basboell-Four-Great-Danish-Linguists}.
\item[Figure~\ref{fig:ch.hjelmslev.efj-young}] ``\name{Eli}{Fischer-Jørgensen}
  (1949)'' Photo provided by Prof. \name{Gunver}{Skytte}; used with
  permission.
\item[Figure~\ref{fig:ch.hjelmslev.efj}] ``\name{Eli}{Fischer-Jørgensen}
  (1968)'' By courtesy of the Royal Danish Academy of Sciences and
  Letters.
\item[Figure~\ref{fig:ch.hjelmslev.efj_lab}] ``\name{Eli}{Fischer-Jørgensen}
  in the Phonetics Lab (1981)'' Photo from (defunct) ``Nordisk
  Pressefoto''. Used with permission of Prof. \name{Gunver}{Skytte}.
\item[Figure~\ref{fig:ch.martinet.martinet0}] ``André
  Martinet''. Unattributed photo from cover of volumes of Martinet's
  \textsl{Œuvres} as published by E.M.E.
\item[Figure~\ref{fig:ch.martinet.martinet2}] ``\name{André}{Martinet}''.
  \url{http://learlinguistic.blogspot.com/2011/09/functional-linguistcs-prague-school.html}
\item[Figure~\ref{fig:ch.martinet.martinet3}] ``\name{André}{Martinet}''.
  \url{https://nadezda-fundamentosdelinguistica.blogspot.com/2020/08/andre-martinet.html} 
\item[Figure~\ref{fig:ch.firth.sweet}] ``Henry
  Sweet''. \url{https://en.wikipedia.org/wiki/Henry_Sweet#/media/File:Sweet_Henry.jpg}
  (Public domain)
\item[Figure~\ref{fig:ch.firth.jones1}] ``\name{Daniel}{Jones}''. 
\item[Figure~\ref{fig:ch.firth.jones2}] ``\name{Daniel}{Jones}''.
\item[Figure~\ref{fig:ch.firth.icphs_1935}] ``From the 1935 London
  ICPhS''. Detail of group picture of participants in the 2nd
  International Congress of Phonetic Sciences, London, 1935. Photo in
  possession of Dept. of Linguistics and English Language, University
  of Edinburgh. Bequest of \name{Elizabeth}{Uldall}.
\item[Figure~\ref{fig:ch.firth:firth-abercrombie-jones} "J. R. Firth, \name{David}{Abercrombie}, \name{Daniel}{Jones}".] 
\item[Figure~\ref{fig:ch.firth.firth_soas}] ``John Rupert
  Firth''. Special Collections, SOAS National Research Library, SOAS
  University of London. Used with permission.
\item[Figure~\ref{fig:ch.firth.firth_wiki}] ``John Rupert
  Firth''. \url{https://commons.wikimedia.org/wiki/File:John_Rupert_Firth.png}. (Public
  Domain) 
\item[Figure~\ref{fig:ch.boas.young_whitney}] ``\name{William Dwight}{Whitney}
  (1858)'': Engraving ``William D. Whitney Professor of Sanskrit and
  Instructor in Modern Languages in Yale College'' by J. C. Buttre, in
  author's Possession.
\item[Figure~\ref{fig:ch.boas.whitney_painting}] ``William Dwight
  Whitney'': Painting by Huc-Mazelet Luquiens, in Yale University Art
  Gallery Collection. Public Domain.
\item[Figure~\ref{fig:ch.boas.powell}] ``\name{John Wesley}{Powell}''. National
  Portait Gallery, Washington, DC. Painter: Edmund Clarence Messer
  (1842 - 1919). Public Domain.
\item[Figure~\ref{fig:ch.boas.boas_student}] ``\name{Franz}{Boas} as a Student
  (1881)''. American Philosophical Society Library, Graphics \#
  U4-1-24. Used with permission.
\item[Figure~\ref{fig:ch.boas.hunt_family}] ``Boas with \name{George}{Hunt}
  Family 1894''. American Philosophical Society Library, Graphics \#
  U5-1-28. Used with permission.
\item[Figure~\ref{fig:ch.boas.boas_1924}] ``\name{Franz}{Boas} with his wife
  and daughter at the 21st International Congress of Americanists, The
  Hague, 1924''. Detail from American Philosophical society Library
  Archive photo, Legacy Identifier APSimg2460, Graphics Number F8-2-1.
\item[Figure~\ref{fig:ch.boas.hunt_tsukwani}] ``\name{George}{Hunt} and
  Tsukwani (Tsax̠is, 1930)''. Photo by J. B. Scott. American Museum of
Natural History.
\item[Figure~\ref{fig:ch.boas.boas}] ``\name{Franz}{Boas} (1912)''. American
  Philosophical Society Library, Graphics \# U5-1-25. Used with permission.
\item[Figure~\ref{fig:ch.sapir.tillohash}] ``\name{Tony}{Tillohash}, Kaibab
  Paiute Indian, in his Carlisle School Uniform''. 1910 photograph
  taken by \name{Edward}{Sapir}. \citealt{sapir30:ute.texts}, insert following
  p. 309.
\item[Figure~\ref{fig:ch.sapir.sapir_1913}] ``\name{Edward}{Sapir} (1913)''.
  Photo in Sapir family collection, used with permission.
\item[Figure~\ref{fig:ch.sapir.sapir_1925}] ``\name{Edward}{Sapir} (1925)''.
  Photo in Sapir family collection, used with permission.
\item[Figure~\ref{fig:ch.sapir.haas_cat}] ``Mary Haas and
  Cat''. American Philosophical Society Library, \name{Mary Rosamond}{Haas}
  Collection. 
\item[Figure~\ref{fig:ch.sapir.swadesh_army}] ``\name{Morris}{Swadesh}''.
  John Simon Guggenheim Foundation
  \url{https://www.gf.org/fellows/all-fellows/morris-h-swadesh/}.
\item[Figure~\ref{fig:ch.sapir.sapir_stock}] ``\name{Edward}{Sapir}
  (ca. 1930)''. From Wikipedia, Public Domain.
\item[Figure~\ref{fig:ch.sapir.sapir_1937}] ``\name{Edward}{Sapir}
  (1937)''. Photo in Sapir family collection, used with permission.
\item[Figure~\ref{fig:ch.sapir.sapir_yale}] ``\name{Edward}{Sapir} (ca. 1938)'':
  Photo in Sapir family collection, used with permission.
\item[Figure~\ref{fig:ch.sapir.sapir_utes}] ``\name{Edward}{Sapir} (1909)''
  {[Group photo with Utes and J. \name{Alden}{Mason}]}. Photo in Sapir family
  collection, used with permission.
\item[Figure~\ref{fig:ch.bloomfield.bloomfield_standing}] ``Leonard
  Bloomfield in his 30s''. University of Chicago Photographic Archive,
  [apf1-00756], Special Collections Research Center, University of
  Chicago Library
\item[Figure~\ref{fig:ch.bloomfield.bloomfield_1944}] ``Leonard
  Bloomfield (1944)''. Courtesy of Department of Linguistics, Yale University.
\item[Figure~\ref{fig:ch.structuralists.smith}] ``Henry Lee
  Smith''. 
  \url{https://www.buffalo.edu/ubreporter/archive/2011_10_27/flashback.html}.
\item[Figure~\ref{fig:ch.structuralists.joos}] ``Martin
  Joos''. Copyright, Board of
  Regents, University of Wisconsin System. Used with permission.
\item[Figure~\ref{fig:ch.structuralists.joos-phonetics}] ``\name{Martin}{Joos}
  engaged in early acoustic phonetic research''. Copyright, Board of
  Regents, University of Wisconsin System. Used with permission.
\item[Figure~\ref{fig:ch.structuralists.trager}] ``George
  Trager''. \url{https://www.gf.org/fellows/all-fellows/george-l-trager/}.
\item[Figure~\ref{fig:ch.structuralists.swadesh}] ``Morris
  Swadesh''.  \url{https://alchetron.com/Morris-Swadesh}.
\item[Figure~\ref{fig:ch.structuralists.twaddell}] ``W. Freeman
  Twaddell''. Photo supplied by William Twaddell, used with permission.
\item[Figure~\ref{fig:ch.structuralists.bloch}] ``Bernard
  Bloch''. \url{https://ling.yale.edu/about/history/people/bernard-bloch}. 
\item[Figure~\ref{fig:ch.structuralists.hockett}] ``\name{Charles}{Hockett}''.
  \url{https://en.wikipedia.org/wiki/Charles_F._Hockett}.
\item[Figure~\ref{fig:ch.structuralists.pike}] ``Kenneth
  Pike''. 
  \url{http://www.sil.org/sites/default/files/styles/history_colorbox/public/history/43_-_ken_pike_m.jpg}.
\item[Figure~\ref{fig:ch.structuralists.harris}] ``Zellig
  Harris''. \url{https://en.wikipedia.org/wiki/Zellig_Harris}.
\item[Figure~\ref{fig:ch.structuralists.sapir_harris}] ``\name{Edward}{Sapir} (Left)
  and \name{Zellig}{Harris} (New Haven, ca. 1937)''. Photo in Sapir family
  collection, used with permission.
\item[Figure~\ref{fig:ch.genphon.chomsky_1950s}] ``\name{Noam}{Chomsky}
  (1950s)''
\item[Figure~\ref{fig:ch.genphon.halle_1960s}] ``\name{Morris}{Halle}
  (1960s)''
\item[Figure~\ref{fig:ch.genphon.jakobson_fant_halle}] ``Jakobson,
  Fant and Halle (ca. 1970)'' Photo supplied by \name{Tim}{Halle}, used with
  permission.
\item[Figure~\ref{fig:ch.genphon.kittery}] ``\name{Morris}{Halle} and Noam
  Chomsky (1953)'' Photo supplied by \name{John}{Halle}, used
  with permission.
\item[Figure~\ref{fig:ch.genphon.morris-roman}] ``\name{Morris}{Halle} and
  \name{Roman}{Jakobson}'' Photo supplied by \name{Tim}{Halle}, used with
  permission.
\item[Figure~\ref{fig:ch.spe.morrisnoam50}] ``\name{Morris}{Halle} and Noam
  Chomsky (2011 (1988 (1953)))''. Photo supplied by \name{John}{Halle}, used
  with permission.
\item[Figure~\ref{fig:ch.spe.kiparsky}] ``\name{Paul}{Kiparsky}'' Photo
  supplied by \name{Paul}{Kiparsky}, used with permission.
\item[Figure~\ref{fig:ch.spe.bybee}] ``\name{Joan}{Bybee} [Hooper]''. Photo
  by John Sumrow, used with permission from \name{Joan}{Bybee}.
\item[Figure~\ref{fig:ch.spe.donegan-stampe}] ``\name{Patricia}{Donegan} and
  \name{David}{Stampe}'': Photo supplied by \name{Patricia}{Donegan}, used with permission.
\item[Figure~\ref{fig:ch.spe:hallechomsky}] ``\name{Morris}{Halle} and Noam
  Chomsky over the years (2016 (2011 (1988 (1955))))'': Photo by Max
  Strong, used with permission.
\item[Figure~\ref{fig:ch.otlabphon.liberman}] ``\name{Mark}{Liberman}'' Photo
supplied by \name{Mark}{Liberman}, used with permission.
\item[Figure~\ref{fig:ch.otlabphon.prince}] ``\name{Alan}{Prince}''
\item[Figure~\ref{fig:ch.otlabphon.goldsmith}] ``\name{John}{Goldsmith}
  (1978)''. Photo supplied by \name{John}{Goldsmith}, used with permission.
\item[Figure~\ref{fig:ch.otlabphon.smolensky}] ``\name{Paul}{Smolensky}''
  Photo supplied by \name{Paul}{Smolensky}, used with permission.
\end{description}


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "/Users/sra/Dropbox/Docs/Books/P20C_2/LSP/main.tex"
%%% End: 
