\chapter{Phonologically conditioned metathesis and associated processes}\label{ch:PhoMet}

\section{Introduction}\label{sec:Int ch:PhoMet}
In this chapter I discuss the use of M\=/forms (metathesised forms)
before vowel-initial enclitics,
as well as the other phonological process with which it occurs.
Such M\=/forms occur only before vowel-initial enclitics
and it is possible to analyse them as conditioned by the this environment.\footnote{
		The processes described in this chapter only occur before vowel-initial enclitics.
		Comparable environments do not exist in Amarasi.
		Vowel-initial suffixes -- of which there are only two
		(see \srf{sec:PeoGroSuf} and \srf{sec:Suf-aq}) --
		only occur on VVC{\#} final stems
		which remain unchanged before vowel-initial enclitics.}
M\=/forms before an enclitic boundary are glossed {\Mvv}
(M with an equals sign for clitic above it).
Three phonological processes, including metathesis,
are triggered before vowel-initial enclitics.
These processes are summarised in \qf{ex:ProEncBou} below.

\begin{exe}
	\ex{Processes at Enclitic Boundaries}\label{ex:ProEncBou}
	\begin{xlist}
		\ex{Metathesis}
		\ex{Consonant Insertion}
		\ex{Vowel Assimilation}
	\end{xlist}
\end{exe}

In this chapter I present an analysis of these processes.
Metathesis is triggered by \tsc{Crisp-Edge};
the need to keep prosodic words phonologically distinct (\srf{sec:Met ch:PhoMet}).
Consonant insertion is triggered by the need for feet to have an onset consonant (\srf{sec:ConIns}).
Vowel assimilation is triggered by metathesis occurring after consonant insertion (\srf{sec:VowAss ch:PhoMet}).

The behaviour of stems before vowel-initial enclitics
is quite diverse between different varieties of Meto
and it must be emphasised that the discussion here
only describes Kotos Amarasi as spoken in the hamlet of Koro{\Q}oto.
Occasional notes on other varieties of Meto occur at some points as relevant
and more extensive comparative data is presented in \srf{sec:HisDev}.

The Amarasi vowel-initial enclitics, all of which trigger these processes,
are given in \trf{tab2:AmaVowIniEnc}.\footnote{
		In \cite{ed16b} I analysed many of these
		vowel-initial enclitics as containing only a single vowel,
		in line with their usual phonetic realisation.
		Since then, I have collected comparative data
		from other varieties of Meto which is best analysed
		by positing that \emph{all} these enclitics have two underlying vowels.
		This evidence is presented in \srf{sec:HisDev}.}
The enclitics \ve{=ii}, \ve{=ein} and \ve{=ee} have
different uses when attached to verbs than they do with nouns.
The enclitic \ve{=ein} displays some complex allomorphy
and is associated with unusual consonant insertion (\srf{sec:PluEnc}).
The function and syntactic behaviour of most of these enclitics
is discussed in Chapter \ref{ch:SynMet}.

\begin{table}[h]
	\caption{Amarasi vowel-initial enclitics}\label{tab2:AmaVowIniEnc}
	\centering
		\begin{tabular}{lll}\lsptoprule
			Form						&Gloss	& Use\\ \midrule
			\ve{=ii}				&{\ii}	& definite referent near/relevant to speaker\\
											&{\ii}	& raises discourse prominence\\
			\ve{=aan/=ana}	&{\aan}	&	definite referent near/relevant to addressee\\
			\ve{=ee}				&{\ee}	& definite referent near/relevant to a third person\\
											&{\eeV}	&	third person P argument (object) of verb\\
			\ve{=aa}				&{\aa}	&	definite referent near/relevant to no one (≈obviative)\\
			\ve{=ein/=eni}	&{\ein}	&	definite plural\\
											&{\einV}&	third person plural verbal argument (S/A/P)\\
			\ve{=ees/=esa}	&{\es}	&	the numeral one (1); indefinite singular\\
			\ve{=een/=ena}	&{\een}	&	inceptive, beginning of state/event \\
			\ve{=aah/=aha}	&just		&	restrictive\\
			\ve{=oo-n}				&{\oo}	& reflexive	(takes genitive suffix \srf{sec:GenSuf}) \\\lspbottomrule
		\end{tabular}
\end{table}

In addition to these vowel-initial enclitics,
the vowel-initial forms of the sentence enclitics \ve{=ma/=ama} `and'
and \ve{=te/=ate} {\te} occasionally, though not obligatorily,
trigger metathesis on their host. The vowel-initial allomorphs
of these enclitics only occur after consonant-final stems in my data.

The structure of the M\=/form of stems
before these vowel-initial enclitics
is summarised in \trf{tab:AmaMfoEnc}
according to the nine unique surface phonotactic
shapes of U\=/forms which can be identified.
The different M\=/forms in \trf{tab:AmaMfoEnc}
are completely predictable based on the corresponding U\=/form,
while the M\=/forms are not fully predictive of the U\=/forms.

\begin{table}[ht]
	\centering
	\caption{Amarasi M-forms before enclitics}\label{tab:AmaMfoEnc}
		\stl{0.5em}\begin{tabular}{lrclrcll}\lsptoprule
							&	U\=/form																	&		&M\=/form																&	U\=/form		&		&M\=/form				&gloss		\\ \midrule
						1.&V\sub{1}{\sA}C\sub{1}V\sub{2}{\sB}				&\ra&V\sub{1}{\sA}V{\sA}C\sub{1}C{\sB}		&\ve{fafi}	&\ra&\ve{faaf\j=}	&`pig'		\\
						2.&V\sub{1}C\sub{1}V\sub{2}C\sub{2}					&\ra&V\sub{1}V\sub{2}C\sub{1}C\sub{2}			&\ve{muʔit}	&\ra&\ve{muiʔt=}	&`animal'	\\
						3.&V\sub{1}\sub{\A\tsc{hi}}C\sub{1}V\sub{2}\sub{\B\tsc{mid}}
																												&\ra&V\sub{1}{\sA}V{\sA}C\sub{1}C{\sB}		&\ve{ume}		&\ra&\ve{uum\j=}	&`house'	\\
						4.&V\sub{1}{\sA}C\sub{1}a\sub{2}(C\sub{2})	&\ra&V\sub{1}{\sA}V{\sA}C\sub{1}(C\sub{2})&\ve{n-sosa}&\ra&\ve{n-soos=}	&`buy'	\\
						5.&V\sub{1}V\sub{2}C\sub{1}V\sub{3}{\sA}		&\ra&V\sub{1}V\sub{2}C\sub{1}C{\sA}				&\ve{n-aiti}	&\ra&\ve{n-ait\j=}	&`pick up'	\\
						6.&V\sub{1}V\sub{2}C\sub{1}V\sub{3}C\sub{2}	&\ra&V\sub{1}V\sub{2}C\sub{1}C\sub{2}			&\ve{nautus}&\ra&\ve{nauts=}	&`beetle'	\\
						7.&V\sub{1}V\sub{2}C\sub{1}									&\ra&V\sub{1}V\sub{2}C\sub{1}							&\ve{kaut}	&\ra&\ve{kaut=}		&`papaya'	\\
						8.&V\sub{1}{\sA}V\sub{2}{\sB}								&\ra&V\sub{1}{\sA}V{\sA}C{\sB}						&\ve{ai}		&\ra&\ve{aa\j=}		&`fire'		\\
%						9.&V\sub{1}C\sub{1}V\sub{2}-ʔ								&\ra&V\sub{1}V\sub{2}C\sub{1}							&\ve{mabe-ʔ}&\ra&\ve{maeb=}		&`time'		\\
						\lspbottomrule
		\end{tabular}
\end{table}

In this chapter I describe each of these M\=/forms in detail
and analyse the ways in which these M\=/forms are derived from the U\=/form.
This analysis has two fundamental elements:
feet should begin with a consonant
and prosodic words should
be phonologically separate from one another.

\input{chapters/PhonologicalMetathesis/Met}
\input{chapters/PhonologicalMetathesis/ConIns}
	\input{chapters/PhonologicalMetathesis/UforSuff}
\input{chapters/PhonologicalMetathesis/VowAss}
\input{chapters/PhonologicalMetathesis/CliHosFinA}
\input{chapters/PhonologicalMetathesis/PluEnc}
	\input{chapters/PhonologicalMetathesis/AnaGwInsVVn}
\input{chapters/PhonologicalMetathesis/MulVowIniCli}
\input{chapters/PhonologicalMetathesis/Dia}
\input{chapters/PhonologicalMetathesis/Sum}
