\documentclass[output=paper]{langsci/langscibook} 
\title{STAMP morphs in the Macro-Sudan Belt} 
\author{%
Gregory D. S. Anderson
\affiliation{Living Tongues Institute for Endangered Languages
\newline
\& University of South Africa (UNISA)} 
}
\chapterDOI{10.17169/langsci.b121.497} %will be filled in at production


\abstract{
STAMP morphs $-$portmanteau subject-tense-aspect-mood-polarity morphs exhibiting functional and  formal properties of both pronominals and auxiliary verbs$-$ are characteristic of many of the genetic units across the Macro-Sudan Belt. STAMP morphs typically occur within a constructional frame, which may include a verb in a fully unmarked form or various construction-specific finite or non-finite forms. STAMP morphs often originate in, and synchronically may appear in, auxiliary verb constructions. Further, univerbation of STAMP morphs with following verbs has yielded various series of inflectional prefixes in a range of families and individual languages. After introducing the functional and formal/constructional properties of STAMP morphs, the paper discusses the origin of STAMP morph constructions and their subsequent developments into bound inflectional or conjugational prefix series. The paper closes by presenting a typology of STAMP morphs in three important Macro-Sudan Belt macro-groupings, viz. Chadic, Central Sudanic, and Niger-Congo languages; and for the last mentioned, particularly the Benue-Congo taxon. 
}

\maketitle
\begin{document}


\section{Introduction}\label{sec:Anderson:1}

In this article I introduce and exemplify a curious and characteristic feature of a number of languages found across different genetic units spanning the large areal complex of equatorial Africa called the \ili{Macro-Sudan} Belt \citep{Güldemann2008}. This feature has challenged analysts pursuing descriptions of various western and central African languages, exhibiting functional properties typically associated with both subject pronouns and auxiliary verbs in many other languages; it has recently been called a STAMP morph \citep{Anderson2012, Anderson2015}. This is mnemonic for what these elements largely are, portmanteau morphs that encode the referent properties of semantic arguments that typically play the syntactic role of `S[ubject]'–that is, the person, number and gender properties of such an actant–in combination with categories of T[ense], A[spect], M[ood] and P[olarity]. Such elements have also been previously called the \textit{tense-person complex} \citep{Creissels2005}, and \textit{pronominal predicative markers} or \textit{pronominal auxiliaries} \citep{Vydrine2011,Erman2002} in the Africanist literature. 

  Portmanteau STAMP morphs are found throughout languages representing different genetic units of the \ili{Macro-Sudan} Belt. In their most basic form, they are the sole means of encoding referent properties of the syntactic subject, in addition to most of the types of categories encoded by verbal Tense, Aspect and \textit{Aktisonsart}, Mood and Polarity morphology. An example of the simplest STAMP morph construction can be seen in \REF{ex:anderson:1} and \REF{ex:anderson:2} from \ili{Tarok}, a language of Nigeria belonging to the \ili{Tarokoid} Plateau genetic unit (\ili{Benue}-Congo stock, \ili{Niger-Congo} phylum).
  
\ea\label{ex:anderson:1}
\ili{Tarok} \citep[238]{Sibomana1981}        [\textsc{\ili{Tarokoid}/Plateau}]\\
\gll \textbf{n}        wá     ù-dɨŋ.    \\
\textsc{1.pfv}  drink \textsc{clsfr}-water\\
\glt 'I have drunk the water.'      
\z

\ea\label{ex:anderson:2}
\ili{Tarok}   \\
\gll \textbf{mi}    wá  a-tí      ipín.  \\
1\textsc{.irr}  drink  \textsc{clsfr}-tea tomorrow \\
\glt 'I will drink tea tomorrow.'
\z

In this study, I discuss some of the characteristics of STAMP morphs and STAMP morph constructions in a set of languages and genetic units of the \ili{Macro-Sudan} Belt. \sectref{sec:Anderson:2} discusses the genetic units that constitute the typological sampling of African languages used in this study. \sectref{sec:Anderson:3} discusses the range of functions associated with STAMP morph constructions, and \sectref{sec:Anderson:4} explores their formal constructional features. \sectref{sec:Anderson:5} offers some thoughts on the origins of STAMP morphs.  \sectref{sec:Anderson:6} discusses the subsequent role such formations have played in the development of prefixed conjugation series that are likewise found across the different genetic units of this linguistic area. Sections 7-9 briefly examine the role of STAMP morph constructions and prefix conjugations that developed out of erstwhile STAMP morphs in three macro-level groupings of genetic units (stocks) in the \ili{Macro-Sudan} Belt. This includes the \ili{Chadic} stock (\sectref{sec:Anderson:7}), the \ili{Central Sudanic} stock of \ili{Nilo-Saharan} (\sectref{sec:Anderson:8}), and the various genetic units found within the \ili{Benue}-Congo stock of \ili{Niger-Congo} (\sectref{sec:Anderson:9}).
 
\section{Typological Sampling and the {Macro-Sudan} Belt}\label{sec:Anderson:2}

In his seminal work on the \ili{Macro-Sudan} Belt [MSB], \cite{Güldemann2008} remained agnostic about the usefulness of higher order units in Africa, at least with respect to typological sampling. Instead he suggested the fourteen genera in \tabref{tab:Anderson:1}, with Ijoid, Dogon, \ili{Chadic} and \ili{Songhay} being peripheral to the proposed area insofar as they show fewer of the characteristic core-features identified as definitional of the MSB, than do the majority of other identified genera.

\begin{table}
\caption{Genera in the \ili{Macro-Sudan} Belt according to \cite{Güldemann2008}} 
\label{tab:Anderson:1} 
\begin{tabularx}{\textwidth}{XllX}
\lsptoprule
 Adamawa &  non-\ili{Bantu} \ili{Benue} Congo &  \ili{Bongo}-Bagirmi &  Dogon\\
\tablevspace
 \ili{Kwa} &  Kru &  \ili{Gur} &  \ili{Songhay}\\
\tablevspace
 \ili{Mande} &  Moru-\ili{Mangbetu} &  Ubangian &  Ijoid\\
\tablevspace
 \ili{Chadic} &  Atlantic &  & \\
\lspbottomrule
\end{tabularx}
\end{table} 

\newpage 
Since 2008, a number of revisions to standard classifications of African languages have been offered, including statements by \cite{Dimmendaal2008, Dimmendaal2011} and \citet{Sands2009}. In previous work on African linguistic typology, I have suggested a larger range of sampling units as appropriate for cross-linguistic work, which have internal diversification on the level of \ili{Germanic} or \ili{Romance}; I call these `genetic units for typological sampling' \citep{Anderson2011}, and write them in small-caps to distinguish them from names of both languages and larger taxa. Many genera in \tabref{tab:Anderson:1} would thus be split into multiple different genetic units; in this system, the number of such genetic units in the \ili{Macro-Sudan} Belt to use in typological sampling totals over fifty. Not all these genetic units have attested STAMP morph constructions, but the vast majority does. In my corpus the genetic units that include STAMP morph constructions (or prefix conjugation series that historically derive from such constructions, see section 5) include at least those in \tabref{tab:Anderson:2}.

\begin{table} 
\caption{Genetic units w/STAMP morphs in MSB \citep{Anderson2011, Dimmendaal2008, Dimmendaal2011, Sands2009}}
\label{tab:Anderson:2}
  
\begin{tabularx}{\textwidth}{XXXX}
\lsptoprule
\multicolumn{4}{l}{\ili{Atlantic-Congo} families, isolates and stocks}  \\ 
\midrule
  \ili{Bantoid} &  \ili{Gbe} &  \ili{Kulango} &  \ili{Potou-Tano} \\
  \ili{Bendi} &  \ili{Gur} &  \ili{Leko-Nimbari} &  \ili{Senufic} \\
  \ili{Cangin} &  \ili{Igboid} &  \ili{Mbum-Day} &  \ili{Tarokoid} \\
 \ili{Cross River} & \ili{Jen Bambukic} &  \ili{Na-Togo} &  \ili{Tenda} \\
  \ili{Edoid} &  \ili{Ka-Togo} &  \ili{Nupoid} &  \ili{Ukaan} \\
  \ili{Ega} &  \ili{Kainji} &  \ili{Okoid} &  \ili{Waja-Kam} \\
  \ili{Fali} &  \ili{Kru} &  \ili{Plateau} &  \ili{Wolof} \\
  \ili{Ga-Dangme} \\
\end{tabularx} 
 
\bigskip

\begin{tabularx}{\textwidth}{XXX}
\ili{Chadic} sub-families             && \ili{Mande} families \\ 
\cmidrule{1-1}\cmidrule{3-3}
 \ili{Biu-Mandara} (Central) && \il{Mande!East Mande}{East Mande} \\
 \il{Chadic!East Chadic}East Chadic && \il{Mande!Northwest Mande}Northwest Mande \\
 \ili{Masa} && \il{Mande!Southeast Mande}Southeast Mande \\
 \il{Chadic!West Chadic}West Chadic && \il{Mande!Southwest Mande}Southwest Mande \\ 
\end{tabularx} 

\bigskip

\begin{tabularx}{\textwidth}{lXX}
 \ili{Central Sudanic} families && \ili{Ubangian} families \\ 
\cmidrule{1-1}\cmidrule{3-3}
 \ili{Kresh} && \ili{Gbaya} \\
 \ili{Lendu} && \ili{Ngbandi} \\
 \ili{Mangbetu} && \ili{Mba} \\
 \ili{Mangbutu-Lese/Efe} && \ili{Zande} \\
 \ili{Moru-Ma’di} && +traces in:\\
 \ili{Sara-\ili{Bongo}-Bagirmi} &&  ~~~~~~\ili{Ngbaka} \\
                   &&  ~~~~~~\ili{Banda} \\
\lspbottomrule                                          
\end{tabularx} 
\end{table}





                  
 


\section{Functions of STAMP morphs}\label{sec:Anderson:3}

STAMP morphs usually combine with a following verb to encode the TAM categories of the event, and the person, number and, where relevant, gender categories of the subject. Such forms can be found in as diverse an array of languages of the \ili{Macro-Sudan} Belt as the  {Biu-Mandara (Central) Chadic} language \ili{Merey} \REF{ex:anderson:3} of Cameroon, the \ili{Kru} language \ili{Wobé} \REF{ex:anderson:4} of Côte d'Ivoire, and \ili{Bongo} of the \ili{Sara-\ili{Bongo}-Bagirmi} family of \ili{Central Sudanic} \REF{ex:anderson:5} spoken in South Sudan.

\ea\label{ex:anderson:3}
\ili{Merey}  \citep[8]{Gravina2007}                 [\ili{Biu-Mandara} Chadic]\\
\gll na  zal.    \\
\textsc{1.pst}  call   \\
\glt `I called.'  
\z

 \newpage
\ea\label{ex:anderson:4}
\ea\label{ex:anderson:4a}
\ili{Wobé} \citep[\ili{Wobé} 3]{HoferLink1973}\footnote{Hofer \& Link's volume does not number the chapters consecutively, but rather each chapter begins anew at page 1. Thus this form comes from page 3 of the \ili{Wobé} chapter. Their page-numbering convention is followed here.}    [\ili{Kru}]\\
\gll ẽ\textsuperscript{ 2}    gyi\textsuperscript{32}.            \\
1.\textsc{pst}  come  \\
\glt `I have come.'    
 

\ex \label{ex:anderson:4b}
\ili{Wobé}\\
\gll  ma\textsuperscript{2}  gyi\textsuperscript{32}.\\
1.\textsc{npst}  come\\
\glt `I am coming.'  
\z
\z

\ea\label{ex:anderson:5}
\ili{Bongo}   \citep[75]{TuckerBryan1966}       [\ili{Sara-\ili{Bongo}-Bagirmi}]\\
\gll ma    bi.\\
1.\textsc{indef}  give  \\
\glt `I am giving.'
\z

One of the curious and characteristic features of STAMP morph constructions is the possible encoding of negative polarity without any distinct negative polarity scope operator, the negative polarity being encoded together with TAM categories and referent properties of the subject in the STAMP morph itself. Formations of this type can be found in languages of the \ili{Macro-Sudan} Belt such as \ili{Duka} (\ili{Kainji}) \REF{ex:anderson:6} or \ili{Mano} (\il{Mande!Southeast Mande}Southeast Mande) \REF{ex:anderson:7}. 

\ea\label{ex:anderson:6}
\ili{Duka}  \citep[13]{BendorSamuelEtAl1973}      [\ili{Kainji}]\\
\gll mân  hé  ò-k\'ɔt.  \\
\textsc{I.fut.neg}  go  to-bush\\
\glt `I won't go to the bush.'      
\z

\ea\label{ex:anderson:7}
\ili{Mano} \citep[226]{Vydrine2009}           [\il{Mande!Southeast Mande}Southeast Mande]\\
\gll l\`{ɛ}\'{ɛ}      máá    wè    gè\={e}.\\
\textsc{3sg.neg[hab]}  \ili{Mano}    language  speak\\
\glt `S/he doesn't speak \ili{Mano}.'
\z

As noted earlier, the referent categories expressed in STAMP morphs are typically restricted to subjects, but in a small number of instances one also finds portmanteau subject > object forms within a STAMP morph, as in the following sentence from \ili{Kohumono} (\ili{Cross River}).

\ea\label{ex:anderson:8}
 \ili{Kohumono} \citep[355]{Cook1972}         [\ili{Cross River}]\\
\gll {β\'ɔ}    fà.\\
1>2.\textsc{npst}  bite\\
\glt `I bite you.'
\z

In \ili{Guro} (\il{Mande!Southeast Mande}Southeast Mande), one finds portmanteau subject > object-encoding STAMP morphs that also reference polarity, as in \textit{ɓ}\textit{e} vs. \textit{yaa} in the following sentences (note the tone differences on the verbs in these constructions as well as the presence of the postverbal negator \textit{ɗo} in \REF{ex:anderson:9b}). 

\ea\label{ex:anderson:9}
\ea\label{ex:anderson:9a}
\ili{Guro} \citep[239]{Vydrine2009}       [\il{Mande!Southeast Mande}Southeast Mande]\\
\gll  ɓe      zuru-o.      \\
  \textsc{2sg>3sg.ipfv}    wash-\textsc{ipfv} \\
\glt `[You] wash him/her/it.'       

\ex \label{ex:anderson:9b}
\gll  yaa      zùrù-ò    ɗo.\\
  \textsc{2sg>3sg.ipfv.neg}  wash-\textsc{ipfv}  \textsc{neg}\\
\glt `[You] don't wash him/her/it.'
\z
\z

The next section (\sectref{sec:Anderson:4}) turns to the range of constructional types in which STAMP morphs participate.


\section{Formal subtypes of STAMP morph constructions}\label{sec:Anderson:4}


STAMP morph constructions exhibit a wide-range of formal sub-types. As mentioned previously, the simplest construction consists of a STAMP morph and an unmarked or bare stem form of the verb \REF{ex:anderson:10}.

\ea\label{ex:anderson:10}
Bare Stem Construction:  STAMP  Verb
\z

Some examples of this constructional type of STAMP morph may be found in \ili{Fyem} \REF{ex:anderson:11}. \ili{Wolof} also has an extensive system of such STAMP morph constructions, as do a number of \ili{Mande} languages. Upwards of ten separate paradigms may be found.

\ea\label{ex:anderson:11}
\ea\label{ex:anderson:11a}
\ili{Fyem} \citep[32, 35]{Nettle1998}          [\ili{Plateau}]\\
\gll   náá  soo  Gindiríŋ.    \\
\textsc{I.prf}  go  Gindiri  \\
\glt `I went to Gindiri.'   
\ex \label{ex:anderson:11b}
\gll ín  soo  dirámméka.\\
\textsc{I.ipfv}  go  farm.your.\textsc{oblq}\\
\glt `I will go to your farm.'
\z
\z

In a second STAMP morph construction the verb appears in a construction-specific tonal form, which suggests that there is an associated floating tone projecting from the STAMP morph onto the verb stem.

\ea\label{ex:anderson:12}
 STAMP + Floating Tone Construction:   STAMP Verb\textsubscript{<construction-specific tonal form>}  
 \z

West \ili{Chadic} Guus (also known as \ili{Sigidi}) also has a highly elaborated system of STAMP morphs (\tabref{tab:Anderson:3}), some of which have associated floating tones that they project onto the verb. The Recent Past (\textsc{rec)} form projects a high tone onto the following verb. The Future (\textsc{fut)} does the same, and differs from the otherwise identical-looking subjunctive [\textsc{sbjn]} in whether it assigns this floating high tone (\textsc{fut}), or does not (\textsc{sbjn}); the two are thus \textit{constructionally} distinct.

%%please move \begin{table} just above \begin{tabular
\begin{table}
\caption{STAMP morphs in Guus [\ili{Sigidi}], West \ili{Chadic} \citep[8-9]{Caron2001}}
\label{tab:Anderson:3}

\begin{tabularx}{\textwidth}{XXXXXXXXX} & \scshape aor & \scshape imm & \scshape sbjn & \scshape fut & \scshape hab & \scshape pfv & \scshape rec & \scshape irr\\
\lsptoprule
1 & \itshape ma & \itshape maa & \textit{m}ə & \textit{m}ə\textit{\textsuperscript{+H}} & \itshape mak & \itshape map & \textit{mam}\textit{\textsuperscript{+}}\textsuperscript{H} & \textit{m}ə\textit{s}\\
2 & \itshape ka & \itshape kaa & \textit{k}ə & \textit{k}ə\textit{\textsuperscript{+H}} & \itshape kak & \itshape kap & \textit{kam}\textit{\textsuperscript{+}}\textsuperscript{H} & \textit{k}ə\textit{s}\\
\lspbottomrule
\end{tabularx}
\end{table}

\ea\label{ex:anderson:13}
\ili{Guus} [\ili{Sigidi}]  \citep[11]{Caron2001}          [\il{Chadic!West Chadic}West Chadic]\\
\gll ən  ka    ɗuu  karáŋ    tʃí    məʃi.\\
if  you\textsc{.irr}  beat  dog    s/he\textsc{.fut}  die\\
\glt `If you beat the dog, it will die.'   
\z

The divergent \ili{Atlantic-Congo} (or \ili{Gur}) isolate language \ili{Kulango} is another language of the \ili{Macro-Sudan} Belt that reflects floating tones associated with STAMP morphs. In \ili{Kulango} \REF{ex:anderson:14} the first singular habitual STAMP morph projects a low tone onto the first stem syllable of the verb, while the corresponding subjunctive form projects a high tone. Both STAMP morphs bear high tones themselves, but one projects a floating low tone and the other a floating high tone.

\ea\label{ex:anderson:14}
\ili{Kulango} \citep[193]{Elders2007}          [\ili{Kulango}]\\
\ea\label{ex:anderson:14a}
\gll má     dɔlɪ.  \\
  \textsc{I.hab}\textsubscript{<+L>}  sell\\        
\glt `I sell.'            

\ex \label{ex:anderson:14b}
\gll mɪ    dɔlɪ.\\
  1.\textsc{sbjn}\textsubscript{<+H>}  sell\\
\glt 'May I sell.'
\z
\z

The \ili{Waja-Kam} language \ili{Dadiya} shows similar constructions.

\ea\label{ex:anderson:15}
\ili{Dadiya}   \citep[196]{Jungraithmayr1968}       [\ili{Waja-Kam}]  \\
\gll \'{n}    já.    \\
you\textsc{.prf}  eat:\textsc{prf}    \\
\glt `You have eaten.'  
\z

\newpage 
In a third sub-type of STAMP morph construction, the verb appears in a construction-specific co-grammaticalized aspectual form \REF{ex:anderson:16}. 

\ea\label{ex:anderson:16}
Aspect Construction:    STAMP  Verb\textsubscript{<\textsc{asp}>}    
\z

Such STAMP morph constructions are found in \ili{Tiv} (\ili{Bantoid}) \REF{ex:anderson:17}, and Ndut-\ili{Falor} (\ili{Cangin}) \REF{ex:anderson:18}.

\ea\label{ex:anderson:17}
\ea\label{ex:anderson:17a}
\ili{Tiv} \citep[114]{Arnott1958}          [\ili{Bantoid/Tivoid}]\\
\gll vé     pìnè.        \\
  they\textsc{.pst}   ask          \textsc{}\\
\glt `They asked.'          

\ex \label{ex:anderson:17b}
\gll mbá       pi\textsuperscript{{\textbar}}ne-n.  \\
  they.\textsc{prs}    ask-\textsc{asp}\\
\glt `They are asking.'      
\z
\z

\ea\label{ex:anderson:18}
\ili{Ndut-Falor} \citep[Ndut-\ili{Falor} 4]{Pichl1973}      [\ili{Cangin}]\\
\gll mi    acɛ.          \\
I.\textsc{rls}  come:\textsc{prf}\\
\glt `I have come.'  
\z

A sub-type of this construction is found with the verb stem having a morphological aspect marker as well being reduplicated \REF{ex:anderson:19}. 

\ea\label{ex:anderson:19}
   STAMP  \textsc{redpl}-Verb-\textsc{asp} 
\z

This is found in the following formation from \ili{Dadiya} of the \ili{Waja-Kam} genetic unit \REF{ex:anderson:20}:

\ea\label{ex:anderson:20}
\ili{Dadiya}   \citep[197]{Jungraithmayr1968}        [\ili{Waja-Kam}]\\
\ea\label{ex:anderson:20a}
\gll m\`ən  nò-lɛ.\\
  I.\textsc{npst}  drink-\textsc{prog}        \\
\glt `I am drinking.'         

\ex \label{ex:anderson:20b}
\gll  mon    jà-jà-l. \\
  you.\textsc{npst}  \textsc{rdpl}-eat-\textsc{prog}\\
\glt `You are eating.'
\z
\z 

In some African languages this construction-determined aspectual form is itself realized tonally \REF{ex:anderson:21}.

\ea\label{ex:anderson:21}
  Tonal Morphology  STAMP   Verb\textsc{\textsubscript{<Asp=Tone>}}
\z

An example of such a realization comes from \il{Bozom@`Bozom}`Bozom, a \ili{Gbaya} Ubangi language \REF{ex:anderson:22}, in which the \textsc{ipfv} and \textsc{pfv} verb forms are distinguished tonally.

\ea\label{ex:anderson:22}
\il{Bozom@`Bozom}`Bozom   \citep[159]{Monino1995}        [\ili{Gbaya} Ubangi]\\
\ea\label{ex:anderson:22a}
\gll ʔà̰    ré.  \\
  he\textsc{.rls}     enter.\textsc{ipfv}\\
\glt `He enters.' 

\ex \label{ex:anderson:22b}
\gll  má  rè. \\
  he\textsc{.irr}  enter\textsc{.pfv}\\
\glt `He will enter.'
\z
\z

Verbs in STAMP morph constructions can also appear in specific construction-de\-pen\-dent modal forms \REF{ex:anderson:23}. 


\ea\label{ex:anderson:23}
Modal Construction  STAMP  Verb\textsubscript{<}\textsc{\textsubscript{modal}}\textsubscript{>}\\
\z

Examples of this type can be found in Ndut-\ili{Falor} \REF{ex:anderson:24} and \ili{Duka} \REF{ex:anderson:25}.

\ea\label{ex:anderson:24}
\ili{Ndut-Falor} \citep[Ndut-\ili{Falor} 4]{Pichl1973}      [\ili{Cangin}]\\
\gll ma[y]  ayɛ.\\
I.\textsc{fut}  come:\textsc{mod}  \\
\glt `I will come.' 
\z

\ea\label{ex:anderson:25}
 \ili{Duka}  \citep[96-98; 105]{BendorSamuelEtAl1973}    [\ili{Kainji}]\\
 \gll m\'{ɛ}    he\'{ɛ}.   \\
I.\textsc{irr}  go:\textsc{cond}  \\
\glt `If I go...'  
\z

Split negative marking is relatively common in AVCs (Anderson 2006; 2011) where the negative is encoded on a lexical verb but other obligatory categories are on the auxiliary. In Africa, this pattern is a feature of auxiliary verb constructions in the languages of the \ili{Rashad Kordofanian} genetic unit, e.g. in \ili{Rashad} proper \REF{ex:anderson:26}.

\ea\label{ex:anderson:26}
\ili{Rashad} \citep[297]{TuckerBryan1966}      [\ili{Rashad Kordofanian}]\\
\gll ŋi   fas    k-eyɛ    y-ɛn.\\
I  meat    \textsc{neg}-eat  1-\textsc{aux}\\
\glt `I am not eating meat.'
\z

Thus, it is not overly surprising that negative can also be encoded in STAMP morph constructions as well \REF{ex:anderson:27}, albeit reflecting different syntactic configurations in the languages of the \ili{Macro-Sudan} Belt (the lexical verb follows the functional element) than in the Nuba Hills languages (where the reverse tends to be true).

\ea\label{ex:anderson:27}
Split Negative Construction:  \textsc{STAMP  V}erb\textsc{\textsubscript{<neg>}}
\z

Such a formation in the negative future in \ili{Gã} of the \ili{Ga-Dangme} Kwa genetic unit \REF{ex:anderson:28}.\footnote{Negative can be encoded in the STAMP morph itself as well. When so, the verb is thus in a `co-negative' form.}

\ea\label{ex:anderson:28}
\ili{Ga} \citep[105]{Kropp1988}        [\ili{Ga-Dangme} Kwa]\\
\gll e\textsuperscript{{\textbar}}    bá\textsuperscript{!}-\'{ŋ}.\\
he\textsc{.fut}  come-\textsc{neg}\\
\glt `He will not come.'
\z

The other common split inflectional pattern found cross-linguistically, but one that is not overly common in the languages of Africa and particularly not in those of the \ili{Macro-Sudan} Belt \citep{Anderson2011}, is a subject-object split pattern \citep{Anderson2006}. In the split subject-object pattern, the object is encoded on the lexical verb that subcategorizes for it, and the other obligatory inflectional categories including subject information appear on the auxiliary as in the auxiliary verb construction in \REF{ex:anderson:29} from \ili{Gidar}.

\ea\label{ex:anderson:29}
\ili{Gidar} \citep[263]{Frajzyngier2008}        [\ili{Biu-Mandara} Chadic]\\
\gll wá-n\`{ə}    mp\`{ə}r-kó.\\
\textsc{fut}-1    chew-2\\
\glt `I will eat you.'
\z

This pattern also appears to be reflected in STAMP morph constructions as well \REF{ex:anderson:30}. 

\ea\label{ex:anderson:30}
Split Object Construction:   STAMP  Verb\textsc{\textsubscript{<obj>}}
\z

Such a formation may be found in a small number of \ili{Chadic} languages, for example, West \ili{Chadic} \ili{Polci} \REF{ex:anderson:31} or the Biu-Mandara (Central) \ili{Chadic} language \ili{Mofu-Gudur} \REF{ex:anderson:31}. 

\ea\label{ex:anderson:31}
\ili{Polci} \citep[153]{Caron2008}          [\il{Chadic!West Chadic}West Chadic]\\
\gll Gǎrbà   kən  ndʒaŋ  sloː  wú  ɗe  kə   fǔː-m. \\
Garba  \textsc{cop}  cut  meat  \textsc{acc}  \textsc{inj}  \textsc{2:aor}  say-1  \\
\glt `If Garba slaughters a beast, tell me'
\z

\ea\label{ex:anderson:32}
\ili{Mofu-Gudur} \citep[4]{Pohlig1992}          [\ili{Biu-Mandara} Chadic]\\
\gll fá    tá-ka    ɗáf.      \\
\textsc{prog.3}  prepare-2\textsc{.io}  food    \\
\glt `She is preparing you food.'
\z

\section{Origin of STAMP morphs from auxiliary constructions}\label{sec:Anderson:5}

Where might STAMP morphs have come from? In other words, how did they develop? In known instances, they result from the fusing of a pronominal marker or pronoun with a following auxiliary verb, fused into a single portmanteau complex \citep{Anderson2006, Anderson2011}.\footnote{There are also instances where STAMP morphs have no obvious etymology, as in many \ili{Mande} languages.} An example showing synchronic variation between a STAMP morph and its source auxiliary verb construction comes from \ili{Limbum} \REF{ex:anderson:33}.

\ea\label{ex:anderson:33}
\ili{Limbum} \citep[\ili{Limbum} 3-4]{FiorePeck1973/1980}      [\ili{Bantoid}]\\
\gll mɛ:\textsuperscript{32}  wɨ\textsuperscript{3}   {\textasciitilde}   mɛ\textsuperscript{3}  ʃe\textsuperscript{2}  wɨ\textsuperscript{3 }   mɛ:\textsuperscript{32} < mɛ\textsuperscript{3}=ʃe\textsuperscript{2}.\\
\textsc{1.prog}  come  ~  1  \textsc{prog}  come\\
\glt `I am coming.'
\z

Researchers on individual languages and families have recognized this connection. Thus \citet[101]{Shimizu1983}, when describing STAMP morphs in the \ili{Leko-Nimbari} language \ili{Zing} \ili{Mumuye} states that ``the surface differences in subject pronouns are in fact due to the TAM markers, which are realized on them or contracted with them.'' Further, \citet[35]{Babaev2010}, in discussing issues in the reconstruction of \ili{Benue}-Congo cautions the reader that ``… various phonological processes of merging person markers with predicative markers of tense, aspect, modality and polarity have made the situation in many languages obscure.''

  Language-specific internal evidence sheds light on this most likely source for the development of STAMP morphs. In Sara of the \ili{Sara-\ili{Bongo}-Bagirmi} family of the \ili{Central Sudanic} stock, the so-called indefinite series subject pronouns are not used with finite, inflected forms of a following vowel-initial verb, but rather with an infinitive form.
  
  \ea\label{ex:anderson:34}
  Sara \citep[75]{TuckerBryan1966}      [\ili{Sara-\ili{Bongo}-Bagirmi}]\\
\gll ma    k-usa.\\
1:\textsc{def}  \textsc{inf}-eat\\
\glt `I am eating.', `I eat (\textsc{hab}).' 
\z

Though a finite construction that seemingly consists of a combination of a subject pronoun  plus an infinitive verb form is odd typologically, such a formation is entirely consistent with typological norms if the subject pronoun actually has an auxiliary verb `hidden' within it, or in other words is actually a STAMP morph. Sara and its sister languages (see \sectref{sec:Anderson:8} below) are far from the only languages that show constructions with infinitive verb forms in combination with specific STAMP morphs. The entirely unrelated \ili{Bantoid} language \ili{Nomaande} shows similar structures. 

\ea\label{ex:anderson:35}
\ili{Nomaande} \citep[22]{Wilkendorf2001}          [\ili{Bantoid}]\\
\gll y\'{ɔ}    ɔcɔba  ná-áyá.\\
3.\textsc{prs}  \textsc{inf}:go  to-\textsc{3sg}\\
\glt `He is going to him (i.e., his home).' 
\z

Similarly, in Dott (also known as \ili{Dass} or \ili{Zoɗi}) of the \il{Chadic!West Chadic}West Chadic family \REF{ex:anderson:36}, the continuous or progressive STAMP morph series requires a verbal noun form of the lexical verb in the predicate. 

\ea\label{ex:anderson:36}
\ili{Dott}/\ili{Zoɗi} \citep[165]{Caron2002}            [\il{Chadic!West Chadic}West Chadic]\\
\gll taa    tʃét-ti.\\
\textsc{3pl.cont}  come-\textsc{vn}\\
\glt `They are coming.'
\z

Yet other languages of the \ili{Macro-Sudan} Belt require dependent-marked forms of verbs to appear in combination with STAMP morphs in specific constructional subtypes. Such languages include the \ili{Kainji} language \ili{Duka} \REF{ex:anderson:37} and \il{Bozom@`Bozom}`Bozom \REF{ex:anderson:38} of the \ili{Gbaya} Ubangi family. 

\ea\label{ex:anderson:37}
\ili{Duka}  \citep[96-98; 105]{BendorSamuelEtAl1973}     [\ili{Kainji}]\\
\gll m\'{ɛ}     \`{ə}m-hà     á.\\
I.\textsc{irr}  \textsc{dep}-go   \textsc{neg}  \\
\glt `I am not going.' 
\z

\ea\label{ex:anderson:38}
\il{Bozom@`Bozom}`Bozom \citep[159]{Monino1995}            [\ili{Gbaya} Ubangi]\\
\gll ʔà̰    rè-á.      \\
he.\textsc{rls}  enter.\textsc{prf}-\textsc{dep} \\
\glt `He has entered.'
\z

\section{STAMP morphs developing into prefixal conjugations}\label{sec:Anderson:6}

%%start here

Having established that STAMP morph constructions often derive from AVCs, the question arises as to where these formations ``go'' in their subsequent historical developments. In ``normal'' AVCs, an auxiliary is often drawn into the verbal complex, and loses its phonological integrity as a freestanding element. Similar developments are seen with STAMP morph constructions. In the \ili{Macro-Sudan} Belt one typically finds that old STAMP morphs have gotten drawn into the verb to become prefixal conjugation markers. This can be seen in cognate formations in Nchumuru varieties. Bejamso-Grubi Nchumuru \REF{ex:anderson:39} has freestanding STAMP morphs, while in closely related \ili{Banda} Nchumuru \REF{ex:anderson:40}, these have become a prefixed conjugation set.

\ea\label{ex:anderson:39}
\ili{Bejamso-Grubi Nchumuru}  \citep[Nchumuru 5]{Price1975}   [\ili{Potou-Tano} Kwa]\\
\gll màá  bà.\\
I.\textsc{prf}  come\\
\glt `I have come.'
\z

\ea\label{ex:anderson:40}
\ili{Banda Nchumuru} \citep[Nuchumuru 4]{Cleal1973}      [\textsc{Potou-Tano} Kwa]\\
\gll mà-ba.\\
\textsc{1pst}:\textsc{prf}-come\\
\glt `I have come.'
\z

\largerpage
Such developments have been known in African linguistics for more than a century.\footnote{\citet[211]{Seidel1898} pointed out that in \ili{Kwa} languages of \ili{Togo} ``die verbalen Präfixe, wahrscheinlich Reste ehemahliger Hilfsverben, verschmelzen nicht selten mit den Pronominalpräfixen zu einer Silbe''. What the status of STAMP morphs might be prosodically in a given language in this region bears close examination, as they may turn out to be clitics or prefixes. To be sure, \citet[93]{CreisselsEtAl2008} caution that ``many descriptions of African languages do not identify pronominal markers appropriately, treating them as independent words''. Such is definitely the case in the East \ili{Mande} \ili{Boko/Busa} cluster \citep{Jones1998}, where the orthography treats STAMP morphs as freestanding elements but phonologically they are prefixes. As a whole, one can find certain variation between Anglophone vs. Francophone meta-analytic traditions in the Africanist literature, with Anglophone analysts preferring more prefixes in their descriptions while Francophone scholars rather often favor analyses with pre-verbal free-standing elements. This is not an absolute.} 
I now turn to brief examination of STAMP morph constructions and historically-related prefixal conjugations in \ili{Chadic}, \ili{Central Sudanic}, and certain \ili{Niger-Congo} families.

\section{STAMP morphs and prefix conjugations in {Chadic} languages}\label{sec:Anderson:7}
 
The next three sections briefly present STAMP morphs in three of the representative stocks found in the \ili{Macro-Sudan} Belt (each consisting of several genetic units for typological sampling). STAMP morphs are a characteristic of the entire \ili{Chadic} macro-family.\footnote{It is important to not confuse STAMP morphs with a series of intransitive copy pronouns that are also characteristic of \ili{Chadic} \citep{Frajzyngier1977, Burquest1986}. They are unrelated phenomena.} Simplex formations of the most basic type \REF{ex:anderson:10} are characteristic of \il{Chadic!West Chadic}West Chadic languages like \ili{Angas} \REF{ex:anderson:41} and Gerka \REF{ex:anderson:42}.

\largerpage[-2] %long distance
\ea\label{ex:anderson:41}
\ili{Angas}  \citep[38]{Burquest1973}         [\il{Chadic!West Chadic}West Chadic]\\
\gll ŋâː    jì.\\
1.\textsc{compl}  come\\
\glt  `I have come.'
\z

\ea\label{ex:anderson:42}
\ili{Gerka}  \citep[173]{Jungraithmayr1968}\\
\gll kà  tà  ƒàm.\\
\textsc{2.prf}  drink  water\\
\glt `You have drunk water.'
\z

On the other end of the spectrum, in some \ili{Biu-Mandara} Chadic languages, e.g., Vamé \REF{ex:anderson:43}, one now finds only bound prefix conjugations which have derived from STAMP morph constructions. 

\ea\label{ex:anderson:43}
Vamé  \citep[11]{Kinnaird2006}          [\ili{Biu-Mandara} Chadic]\\
\gll \`{ə}ŋ-lɛ.\\
\textsc{fut}.1-go\\
\glt `I will go.'
\z

\largerpage[2]
A common pattern seen throughout the \ili{Macro-Sudan} Belt is for a perfective series to be bound prefixal conjugations, but an imperfective series to remain in freestanding STAMP morph constructions. An example of such a system is found in the \ili{Biu-Mandara} Chadic language \ili{Mbuko} \REF{ex:anderson:44}.

\ea\label{ex:anderson:44}
\ili{Mbuko}  \citep[7]{Gravina2001}           [\ili{Biu-Mandara} Chadic]\\
\ea\label{ex:anderson:44a}
\gll n\`{ə}-zlàmbál.    \\
  1\textsc{pfv/ant}-throw:\textsc{ant} \\
\glt `I threw.'    

\ex \label{ex:anderson:44b}
\gll n\={i}  zl\={a}mb\={a}l.\\
\textsc{1.ipfv}  throw\\
\glt `I am throwing.'  
\z
\z

\ili{Merey} (also \ili{Biu-Mandara} Chadic) shows an unusual intra-pardigmatic split between free-standing forms in the first singular past but bound prefix elements in the corresponding third  person \REF{ex:anderson:45} (masculine singular) past form. 

 \ea\label{ex:anderson:45}
\ili{Merey}  \citep[8]{Gravina2007}          [\ili{Biu-Mandara} Chadic]\\
\ea\label{ex:anderson:45a}
\gll na    zal.      \\
  \textsc{1.pst}    call\\
\glt `I called.'

\ex \label{ex:anderson:45b}
\gll a-zal.\\
  3\textsc{.pst}-call\\
\glt `He called.'
\z
\z

Once bound, such prefix conjugational elements can appear with new auxiliary verbs, as in the following \ili{Buduma} formation \REF{ex:anderson:46}.

\ea\label{ex:anderson:46}
\ili{Buduma} \citep[376]{Pawlak2001}, \citep[55]{Lukas1939}    [\ili{Biu-Mandara} Chadic]\\
\gll a-kol  a  jai-ni.\\
\textsc{3.prs}-be  at  sit\textsc{-vn}\\
\glt `He is/was sitting.' 
\z

While prefix conjugations are archaic features of Afroasiatic languages \citep{Hodge1971, Schuh1976, Mukarovsky1983, Voigt1987}, the ones attested in \ili{Chadic} languages do not reflect inherited structures \citep{Schuh1976, Voigt1989, Jungraithmayr2005, Jungraithmayr2006}, but rather reflect secondary developments typically derived from the univerbation of STAMP morphs with following verbs \citep{Caron2006, Shay2008, Anderson2011, Anderson2012}.

\section{STAMP morphs and conjugational prefixes in {Central Sudanic} languages}\label{sec:Anderson:8}

STAMP morphs and prefixed conjugation series derived from STAMP morphs are also found in the various families of the \ili{Central Sudanic} stock \citep{Anderson2015}. Within the analytic tradition of \ili{Ma'di} \REF{ex:anderson:47}, a member of the \ili{Moru-Ma’di} genetic unit \citet{BlackingsFabb2003} analyze the freestanding STAMP morph \textit{ka}/\textit{k\'ɔ} as a pronominal, but \citet{TuckerBryan1966} analyze it as an auxiliary. Such disagreement over what traditional part-of-speech has to be assigned to these complexes underscores the unusual but characteristic features of STAMP morphs. 

\newpage 
\ea\label{ex:anderson:47}
\ili{Ma'di} \citep[13]{BlackingsFabb2003}      [\ili{Moru-Ma’di}]\\
\ea\label{ex:anderson:47a}
\gll ká    gbándà  \`{}ɲa.  \\
  \textsc{3sg.indef}  cassava  \textsc{npst}:eat\\
\glt `He eats/is eating cassava.'      

\ex\label{ex:anderson:47b}
\gll kɔ    \`{}ɲa-ʔa.\\
  3\textsc{sg.indef}  \textsc{npst}:eat-\textsc{obj}\\
\glt `He eats/is eating it.'
\z
\z

Various \ili{Sara-\ili{Bongo}-Bagirmi} languages require infinitive forms of vowel-initial verbs when used with freestanding STAMP morphs, for example, Ngambay \REF{ex:anderson:48}, Kabba \REF{ex:anderson:49} and \ili{Kenga} \REF{ex:anderson:50}.

\ea\label{ex:anderson:48}
\ili{Ngambay} \citep[118]{Vandame1963}      [\ili{Sara-\ili{Bongo}-Bagirmi}]\\
\ea\label{ex:anderson:48a}
\={m}    k-ào   àl   ngà.\\
   1.\textsc{fut}   \textsc{inf}-go  \textsc{neg}   \textsc{adv}\\
\glt `I will not go again.'

\ex \label{ex:anderson:48b}
\gll á   k-ùs\={a}    né   ngà   uà.\\
 2.\textsc{fut}   \textsc{inf}-eat thing   \textsc{adv}   \textsc{q}\\
\glt `You are already going to eat?'
\z
\z

\ea\label{ex:anderson:49}
\ili{Kabba}  \citep[220]{Moser2004}         [\ili{Sara-\ili{Bongo}-Bagirmi}]\\
\gll má  k-àw  lò  t\'{ə}  àáng.    \\
1:\textsc{fut}  \textsc{inf}-go  place  \textsc{loc}  \textsc{neg}    \\
\glt `I shall go nowhere.'
\z

\ea\label{ex:anderson:50}
\ili{Kenga}  \citep[15]{Neukom2010}        [\ili{Sara-\ili{Bongo}-Bagirmi}]\\
\gll \={m}    k-\`{ɔ}s\={ɔ}.\\
1.\textsc{fut}  \textsc{inf}-eat\\
\glt `I will eat.'
\z

As with \ili{Chadic}, many languages in different sub-families of \ili{Central Sudanic} have a bound `definite' perfective STAMP series that contrasts with an unbound `indefinite' imperfective series. The pattern is found in \ili{Bongo} \REF{ex:anderson:51}, \ili{Lugbara} \REF{ex:anderson:52}, Lendu \REF{ex:anderson:53} and \ili{Oke'bu} \REF{ex:anderson:54}, from four different genetic units of \ili{Central Sudanic}. 

\ea\label{ex:anderson:51}
\ili{Bongo}   \textit{*mi- < m-i} \citep[75]{TuckerBryan1966}    [\ili{Sara-\ili{Bongo}-Bagirmi}]\\
\ea\label{ex:anderson:51a}
\gll mi-bi.\\
1.\textsc{def}-give\\
\glt `I give, I gave.'

\ex \label{ex:anderson:51b}
\gll ma    bi.\\
 1.\textsc{indef}  give\\
\glt  `I am giving.'    
\z
\z

\newpage 
\ea\label{ex:anderson:52}
\ili{Lugbara} \citep[47]{TuckerBryan1966}         [\ili{Moru-Ma’di}] \\
\ea\label{ex:anderson:52a}
\gll á-tsɔ    mvá.\\
  1.\textsc{def}-beat  child\\
  \glt `I beat the child.'
  
  \ex \label{ex:anderson:52b}
\gll ma    mvá  tsɔ.\\
  1.\textsc{indef}  child  beat\\
\glt `I am beating the child.'
\z
\z

\ea\label{ex:anderson:53}
\ili{Lendu}  \citep[46]{TuckerBryan1966}         [\ili{Lendu}]\\
\ea\label{ex:anderson:53a}
\gll má-drr    mbí.\\
  1.\textsc{def}-pull  rope\\
  \glt `I pull the rope.' 

\ex\label{ex:anderson:53b}
\gll má    mbi  dr\'{r}.\\
  \textsc{1.indef}  rope  pull\\
\glt `I am pulling the rope.'
\z
\z

\ea\label{ex:anderson:54}
\ili{Oke'bu} \citep[48]{TuckerBryan1966}        [\ili{Mangbutu}]\\
\ea\label{ex:anderson:54a}
\gll l-ómá     ùnzu.\\
  \textsc{2.def}-beat  child\\
\glt `You beat the child.'


\ex \label{ex:anderson:54b}
\gll láà    unzú  òma.\\
  \textsc{2.indef}  child  beat\\
\glt `You are beating the child.'
\z
\z

As the reader may have noticed, there is also a difference in the basic clausal syntax of the two series in these \ili{Central Sudanic} languages, with VO order in the definite/perfective series and OV in the indefinite/imperfective one. The VO order in the definite series vs. OV order in the indefinite is, however, not universal in \ili{Central Sudanic} and is not found, for example, in \ili{Baka} \REF{ex:anderson:55} of the \ili{Sara-\ili{Bongo}-Bagirmi} family or Kresh \REF{ex:anderson:56}.

\ea\label{ex:anderson:55}
\ili{Baka} \citep[75]{TuckerBryan1966}        [\textsc{Sara-\ili{Bongo}-Bagirmi}]\\
\ea\label{ex:anderson:55a}
\gll m-áne    yí.\\
  \textsc{1.def}-eat  thing\\
\glt `I ate (something).'

\ex \label{ex:anderson:55b}
\gll má    y-ane    yi.\\
 1.\textsc{indef}  \textsc{inf}-eat    thing\\
\glt `I am eating (something).'
\z
\z


\newpage 
\ea\label{ex:anderson:56}
\ili{Kresh} \citep[75]{TuckerBryan1966}     [\ili{Kresh}]\\
\ea\label{ex:anderson:56a}
\gll m-omò    nòmò.\\
  1-drink   drink    \\
\glt I drink a drink.'     


\ex \label{ex:anderson:56b}
\gll ma     y-òmò    nòmò.\\
  \textsc{1.indef}  \textsc{inf}-drink   drink\\
\glt `I am drinking.'
\z
\z

Note that these two languages however also have infinitive forms of the verb with vowel initial stems with the indefinite series STAMP morph, like those in \REF{ex:anderson:48}-\REF{ex:anderson:50}. This feature further underscores the likely origin of STAMP morph in an auxiliary verb construction in \ili{Central Sudanic} languages.

Of course, once a bound STAMP morph prefix exists in a given \ili{Central Sudanic} language, it is free to attach to auxiliary verbs in new AVCs, as seen in the following \ili{Lugbara} sentence \REF{ex:anderson:57}.

\ea\label{ex:anderson:57}
\ili{Lugbara} \citep[46, 47]{TuckerBryan1966}      [\ili{Moru-Ma’di}]\\
\gll ma-ŋga  mvá  tsɔ.\\
\textsc{1.def-aux}  child  beat\\
\glt `I shall beat the child.'    
\z

Variation in closely related varieties between bound and unbound formations can be seen in \ili{Sara-\ili{Bongo}-Bagirmi} languages. Compare in this regard \ili{Furu} \REF{ex:anderson:58}, where a freestanding STAMP morph is found, with the corresponding sentence in its close sister language \ili{Bagiro} \REF{ex:anderson:59}, where it is a prefix.

\ea\label{ex:anderson:58}
\ili{Furu} \citep[91]{Boyeldieu1990}           [\ili{Sara-\ili{Bongo}-Bagirmi}]\\
\gll mí  gáli  gɔ.          \\
I   know  \textsc{neg}      \\
\glt `I don't know.'
\z

\ea\label{ex:anderson:59}
\ili{Bagiro}  \citep[91]{Boyeldieu1990}           [\ili{Sara-\ili{Bongo}-Bagirmi}]\\
\gll mú-gá{\downstep}li  g\`{ɔ}.\\
1-know  \textsc{neg}\\
\glt `I don't know.'  
\z

\ili{Mangbetu} languages only have bound prefixes, but nevertheless reflect a possible trace of two originally distinct STAMP sets, albeit with the opposition having become phonologized in such languages as \ili{Mangbetu} and Meje. Thus vowel initial stems take the second singular prefix \textit{ni-} \REF{ex:anderson:60}-\REF{ex:anderson:61} while consonant-initial stems take \textit{(m)ú-} \REF{ex:anderson:62}-\REF{ex:anderson:63}:

% \newcommand{\andersonE}{$\text{\textgreek{'e}}$} 

\newcommand{\andersonE}{\'{ɛ}}
\ea\label{ex:anderson:60}
\ili{Mangbetu} \citep[106]{Larochette1958}        [\ili{Mangbetu}]\\
\gll ni-{\andersonE}si-(a).      \\
2.\textsc{prs/fut}-do-\textsc{tam}        \\
\glt `You (will) do.' 
\z

\ea\label{ex:anderson:61}
\ili{Meje} \citep[106]{Larochette1958}           [\ili{Mangbetu}]\\
\gll ni-{\andersonE}sí-a.\\
\textsc{prs/fut}-do-\textsc{tam}\\
\glt `You (will) do.' 
\z

\ea\label{ex:anderson:62}
\ili{Mangbetu} \citep[106-7]{Larochette1958}        [\ili{Mangbetu}]\\
\gll mú-ta.\\
2.\textsc{prs/fut}-carry    \\
\glt `You (will) carry.'     
\z

\ea\label{ex:anderson:63}
\ili{Meje} \citep[106-7]{Larochette1958}        [\ili{Mangbetu}]\\
\gll ú-ta.\\
\textsc{prs/fut}-carry\\
\glt `You (will) carry.' 
\z

The \ili{Mangbutu} language Lese also has only bound prefix conjugations. Again these appear to reflect two originally distinct series, such as the present progressive in the first singular in \textit{má}- \REF{ex:anderson:64} and the future progressive first singular in \textit{m\'ʊ-} \REF{ex:anderson:65}.

\ea\label{ex:anderson:64}
\ili{Lese} \citep[51]{TuckerBryan1966}        [\ili{Mangbutu}]\\
\gll má-dz\`ɔ    k\`ɔɗí  à-n\'ʊ.   \\
1:\textsc{tam-aux}  meat  \textsc{dep}-eat   \\
\glt `I am eating meat.'   
\z

\ea\label{ex:anderson:65}
\ili{Lese} \citep[51]{TuckerBryan1966}        [\ili{Mangbutu}]\\
\gll m\'ʊ-dzá    k\`ɔɗí  à-n\`ʊ.\\
1:\textsc{tam-aux}  meat  \textsc{dep}-eat\\
\glt `I shall be eating meat.'  
\z

While widespread and hence tempting to reconstruct the bound perfective series and unbound imperfective series pattern all the way back to \ili{Proto-Central Sudanic}, the overall frequency of such patterning in the \ili{Macro-Sudan} Belt suggests that we should do so only with caution. Nevertheless, I tentatively reconstructed this to \ili{Proto-Central Sudanic} \citep{Anderson2015}. This is because the sub-families that lack this pattern have two bound series, and never the opposite patterning, such that bound imperfective series appear to be attested only if bound perfective series exist; this distribution is also found across the \ili{Benue}-Congo and \ili{Chadic} languages of the \ili{Macro-Sudan} Belt.

\section{STAMP morphs and conjugational prefixes in {Niger-Congo} languages}\label{sec:Anderson:9}

STAMP morphs and conjugational prefixes derived from STAMP morphs are found in a wide range of \ili{Niger-Congo} languages. These include \ili{Klao} of the \ili{Kru} family \REF{ex:anderson:66}, \ili{Dadiya} of the \ili{Waja-Kam} family \REF{ex:anderson:67}, \ili{Gã} of the \ili{Ga-Dangme} Kwa genetic unit \REF{ex:anderson:68}, and \ili{Kulango} \REF{ex:anderson:69}.

\ea\label{ex:anderson:66}
\ili{Klao} \citep[3, 18]{Marchese1982}            [\ili{Kru}]\\
\gll \={ɔ}\={ɔ}    bl\={e}.  \\
he\textsc{:ipfv}   sing\\
\glt `He is singing.'
\z

  
\ea\label{ex:anderson:67}
\ili{Dadiya}   \citep[196]{Jungraithmayr1968}        [\ili{Waja-Kam}]\\
\gll \'{n}    já.   \\
you\textsc{.pfv}  eat.\textsc{pfv} \\
\glt `You have eaten.'
\z

\ea\label{ex:anderson:68}
\ili{Ga} \citep[105]{Kropp1988}          [\ili{Ga-Dangme} Kwa]  \\
\gll èe     n\`{\~u}  n\`{\~u}.  \\
3\textsc{:prs}  water  drink\\
\glt `He is drinking water.'
\z

\ea\label{ex:anderson:69}
\ili{Kulango} \citep[193]{Elders2007}            [\ili{Kulango}]\\
\ea\label{ex:anderson:69a}
\gll m\'ɪ   d\'ɔl\`ɪ.\\
  \textsc{1.sbjn}  sell\\
\glt `May I sell.'    

\ex \label{ex:anderson:69b}
\gll   m\'ɪ\`ɪ   d\`ɔl\`ɪ.\\
  1.\textsc{prog}  sell\\
  \glt `I am selling.'  
\z
\z

Such formations are also widespread in \ili{Mande} languages, such as the following example from \ili{Jo}(wulu) of the \il{Mande!Northwest Mande}Northwest Mande genetic unit \REF{ex:anderson:70}.

\ea\label{ex:anderson:70}
\ili{Jo} (\ili{Jowulu}) \citep[11]{Kim2002}          [\il{Mande!Northwest Mande}Northwest Mande]  \\
\gll kaa    ku-ki  kulu  n\~u.  \\
\textsc{3pl.fut}  \textit{to}-\textsc{def}  hot  eat    \\
\glt `They will eat hot \textit{to.}'        
\z

Within \ili{Benue}-Congo there is evidence for two (possibly three) sets of contrastive STAMP morphs series.\footnote{The reconstructions for Proto-\ili{Benue}-Congo are preliminary and impressionistic. Space limitations prevent further demonstration of the details of the complex \ili{Benue}-Congo situation.} One is realized as a bound prefix series, while the other(s) remain(s) freestanding STAMP morphs in most but not all relevant languages. These series consist of a bound realis/perfect(ive) series marked by \textit{*}\textit{ma-} (possibly with an associated low tone *\textit{mà}-) in the first singular, which contrasts with one (or two) other \textit{m}-initial series with front vowels (or no vowel) marking irrealis/imperfective action. Thus, within \ili{Benue}-Congo itself, there is some evidence for the two non-perfect(ive) series. However, the broader comparative data from \ili{Atlantic-Congo} (or \ili{Volta-Congo}) languages suggest that there were originally two sets of forms, and one may well have split into two contrasting series during the development of \ili{Benue}-Congo, or, alternatively, in the history of individual sub-groups within \ili{Benue}-Congo. However, they likely originated in a single imperfective/irrealis series from an earlier state of the language. One of these sets generally has a high-tone in the first singular and marks irrealis or future semantics. The first singular marker in the other set appears optionally without a vowel and with a non-high (mid or low) tone, and encodes present/non-past/imperfective semantics \REF{ex:anderson:71}.

\ea\label{ex:anderson:71}
 \ili{Proto-\ili{Benue}-Congo}\\
Pattern 1  \textit{*ma-} I.\textsc{prf/pst/rls}  :  \textit{*mé}  I.\textsc{irr/fut}\\  
Pattern 2  \textit{*ma-} I.\textsc{prf/pst/rls}  :  \textit{*mé} I.\textsc{irr/fut}  :   \textit{*mi, *m(V)} I.\textsc{prs/npst}\\
\z

Evidence for the *\textit{ma-} ({\textasciitilde}\textbf{*}\textit{mà}-) perfective/realis/past series in \ili{Benue}-Congo \REF{ex:anderson:71} comes from a range of different sub-groups of the stock. Thus one finds formally and functionally cognate elements in such \ili{Benue}-Congo languages as \ili{Bantoid} \ili{Ndemli} \REF{ex:anderson:72}, Eleme of the \ili{Cross River} family \REF{ex:anderson:73}, and \ili{Berom} of the \ili{Plateau} sub-stock \REF{ex:anderson:74}.

\ea\label{ex:anderson:72}
\ili{Ndemli}   \citep[72]{Ngoran1999}            [\ili{Bantoid}]\\
\gll mà-tóm. \\
\textsc{I.pst-}send\\
\glt  `I sent.'   
\z

\ea\label{ex:anderson:73}
\ili{Eleme}  \citep[1482]{Bond2008}, \citep{Bond2010}        [\ili{Cross River}]\\
\gll ma-ʔà.\\
1.\textsc{ant.prf}-leave\\
\glt `I left.'
\z

\ea\label{ex:anderson:74}
\ili{Berom}  \citep[299, 301]{Bouquiaux1970}          [\ili{Plateau}]\\
\gll mà-ciŋ.\\
1.\textsc{pst/prf}-dig\\
\glt `I (have) dug.'
\z

Evidence for the high-toned irrealis or future series marked in the first singular by \textit{*mé} can be found in a range of \ili{Benue}-Congo languages, e.g., the \ili{Bantoid} language \ili{Tiv} \REF{ex:anderson:75}, the \ili{Kainji} language \ili{Duka} \REF{ex:anderson:76} or \ili{Berom} of the \ili{Plateau} stock \REF{ex:anderson:77}.

\ea\label{ex:anderson:75}
 \ili{Tiv} \citep[\ili{Tiv} 4]{Arnott19671980}        [\ili{Bantoid}]\\
 \gll mé    \textsuperscript{!}va.\\
\textsc{1fut}  come\\
\glt `I will come.'
\z

\ea\label{ex:anderson:76}
\ili{Duka} \citep[17]{BendorSamuelEtAl1973}      [\ili{Kainji}]\\
\gll  m\'ɛ    róà    sə  á.\\
I.\textsc{irr}  \textsc{rem.fut}  drink  \textsc{neg}\\
\glt `I would not drink it.'
\z

\ea\label{ex:anderson:77}
\ili{Berom}  \citep[300]{Bouquiaux1970}        [\ili{Plateau}]\\
\gll mé    ciŋ.\\
1.\textsc{sbjn}  dig\\
\glt `I must dig.'
\z

Evidence for a third series of STAMP morphs in Proto-\ili{Benue}-Congo is more tenuous and must remain beyond of the scope of the present study.

\section{Summary}\label{sec:Anderson:10}
\largerpage
This study has been a preliminary survey of STAMP morph constructions in a range of genetic linguistic units spoken across the \ili{Macro-Sudan} Belt. A comparative analysis of STAMP morphs across the languages of the \ili{Macro-Sudan} Belt, and the prefixal conjugations that historically derived from them, yields significant insights into various layers in the history of verbal conjugation in many genetic units. Perhaps the most noteworthy insight is that there is an absolute pattern whereby, if present, a bound series encodes perfective/realis categories and an unbound series rather encodes imperfective/irrealis categories. This situation is found in \ili{Chadic}, \ili{Central Sudanic} and \ili{Benue}-Congo languages alike. This pattern has been reconstructed to \ili{Proto-Central Sudanic} \citep{Anderson2015}. One can also reconstruct a probably bound perfective/realis conjugation series (e.g., with first singular \textit{*ma-/=)} for Proto-\ili{Benue}-Congo, but its immediate ancestral formation may still have been an unbound STAMP morph. The corresponding imperfective/irrealis series (e.g., with first singular \textit{*mé {\textasciitilde} *mI {\textasciitilde} *mÍ}) appears to have certainly been an unbound STAMP morph in the proto-language. As in \ili{Central Sudanic}, \ili{Benue}-Congo languages only have bound imperfective series STAMP morphs if perfective series elements are also bound. The precise modeling of the \ili{Chadic} developments is the object of current ongoing research, but the same distributional trend appears to be manifested across the \ili{Chadic} languages as well. 

While insights into the inflectional history of the genetic units in this macro-region are gained by such an analysis (e.g. how secondary prefix conjugations arose in \ili{Central Sudanic} languages), others remain unanswered to date. The most salient of such questions is the following: Why is it that if bound/prefixal conjugations are found which appear to derive from STAMP morph constructions, in \ili{Chadic}, various \ili{Niger-Congo} families and \ili{Central Sudanic} languages across the \ili{Macro-Sudan} Belt the pattern is always one where the bound series encodes perfective/realis and the unbound series imperfective/irrealis categories, but never the reverse situation? Resolving this important and intriguing question remains a primary goal of future research. 

\section*{Abbreviations and symbols}

 

\begin{tabularx}{.45\textwidth}{ll}
1,2,3 & 1\textsuperscript{st}, 2\textsuperscript{nd}, 3\textsuperscript{rd} person\\
>  & acting on  \\
\textsc{acc}  & accusative\\  
\textsc{adv}   & adverb\\
\textsc{ant}   & anterior \\
\textsc{aor}   & aorist\\  
\textsc{asp}   & aspect\\
\textsc{aux}   & auxiliary  \\
\textsc{clsfr}   & classifier  \\
\textsc{compl}   & completive\\
\textsc{cond}   & conditional  \\ 
\textsc{cont}   & continuative  \\
\textsc{cop}   & copula\\
\textsc{def}   & definite\\  
\textsc{dep}   & dependent \\
\textsc{fut}   & future\\
\textsc{+H}   & high tone\\ 
\textsc{io}  & indirect object  \\
\textsc{hab}   & habitual  \\
\textsc{imm}  & immediate\\
\textsc{indef}  & indefinite\\  
\textsc{inf}  & infinitive  \\
\textsc{inj}  & injunctive  \\
\textsc{ipfv}  & imperfective  \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{ll}
\textsc{irr}   & irrealis  \\
\textsc{+L}  & low tone  \\
\textsc{loc}  & locative  \\
\textsc{mod}  & modal  \\
\textsc{neg}  & negative  \\
\textsc{npst}  & non-past \\
\textsc{obj}  & object  \\
\textsc{oblq}  & oblique  \\
\textsc{pfv}  & perfective  \\
\textsc{pl}  & plural\\
\textsc{prf}  & perfect  \\ 
\textsc{prog}  & progressive \\ 
\textsc{prs}  & present\\
\textsc{pst}  & past  \\
\textsc{q}  & interrogative\\  
\textsc{rec}  & recent\\
\textsc{rdpl}  &reduplication  \\
\textsc{rem}  & remote  \\
\textsc{rls}  & realis\\
\textsc{sbjn}  & subjunctive\\  
\textsc{sg}  & singular  \\
\textsc{tam}  & tense-aspect-mood \textsc{}\\
\textsc{vn    }   & verbal noun\\
\\
\end{tabularx}
\largerpage[2]
 

{\sloppy
\printbibliography[heading=subbibliography,notkeyword=this]
}
\end{document}