\documentclass[output=paper]{langsci/langscibook} 
\title{Valency changing processes in {Akan}}
\author{%
E. Kweku Osam\affiliation{University of Ghana} 
}
\chapterDOI{10.17169/langsci.b121.478} %will be filled in at production




\abstract{Valency has been considered as both a semantic and syntactic notion. Semantically it is used to refer to the participants in an event; as a syntactic notion it is used to indicate the number of arguments in a construction. In \ili{Akan}, a Kwa language spoken in Ghana, we can identify various transitivity classes of verbs: verbs that are strictly intransitive; those that are strictly transitive; and verbs that are used ditransitively. Apart from these, there are verbs that can be used both intransitively and transitively. Critical to the study of the notion of valency in Akan is the fact that there is clear evidence for grammatical relations in the language. As is the case in many languages, Akan possesses morphosyntactic means through which the valency of verbs can be adjusted. The application of these morphosyntactic processes reduces or increases the valency of verbs. This paper examines these processes in Akan. The critical valency-reducing processes in Akan are reflexivization, reciprocals, anticausative/inchoative constructions, impersonal constructions, object omission constructions, and unspecified object constructions. Valency-increasing processes include causativization and agentivization through serialization.}

\maketitle
\begin{document}

\section{Introduction}\label{§1:introduction.osam}

The linguistics literature is replete with studies on valency -- those that are theoretical, cross-linguistic, and others that focus on the study of valency in specific languages. The interest in understanding the notion of valency has resulted in a number of volumes dedicated to the subject. Notable among these are \citet{comriepolinsky1993}, \citet{dixonaikhenvald2000a}, \citet{malchukovcomrieinpress}. The papers in \citet{malchukovcomrieinpress} came out of the Leipzig Valency Classes Project and the Conference on Valency Classes in the World's Languages held in 2011. Worthy of mention is also \citet{nicholsetal2004} that put forward the idea of the basic valency orientation of languages. As in the case of the Leipzig Project, a number of scholars adopting the methodology of Nichols and associates, have examined the basic valency orientation of various languages (see, for example, \citealt{narogg2009,planklahiri2009,vangelderen2011,luraghi2012}).  

The goal of this paper is to examine the valency of \ili{Akan} verbs and explore the morphosyntactic processes that apply to modify the valency of verbs. It is expected that the paper will help expand further our understanding of the behavior of verbs in \ili{Akan}, building on what has been reported in \citet{osam2008a}. 

This paper is based on the \ili{Akan} Verbs Database project which was implemented in the Department of Linguistics, University of Ghana, between 2009 and 2011, and for which I was the Principal Investigator. The project was funded by the University of Ghana Research Fund. By the end of the project, a database of over 3,500 verb stems and other verb forms had been created.

The data come from various sources, including published narratives, translated works and, in some cases, constructed examples based on my knowledge as a native speaker.

The paper is structured as follows: \sectref{§2:morphosyntactic.osam} provides an overview of the verbal morphology, focusing on the tense-aspects distinctions of the language and clause structure/gram\-mat\-i\-cal relations. This section also covers some aspects of serial verb constructions in the language since they have a bearing on valency adjustment in the language. \sectref{§3:akan.osam} deals with \ili{Akan} verb classes. In \sectref{§4:decreasing.osam} the morphosyntactic processes that reduce the valency of verbs are discussed; \sectref{§5:increasing.osam} and \sectref{§6:valence.osam} focus on the processes that increase valency. \sectref{§7:summary.osam} is the summary and conclusion.

The label \ili{Akan} is used to refer to a group of closely related dialects spoken in Ghana, and partially in the south eastern parts of Cote d´Ivoire. It belongs to the \ili{Kwa} sub-family of \ili{Niger-Congo}. The name also refers to the people who speak the language. The dialects of \ili{Akan} include: \ili{Agona}, \ili{Ahafo}, \ili{Akuapem}, \ili{Akwamu}, \ili{Akyem}, \ili{Asante}, \ili{Assin}, \ili{Bono}, \ili{Denkyira}, \ili{Fante}, \ili{Kwahu}, and \ili{Wassa}. Generally, except \ili{Fante}, all the other dialects tend to be classified as \ili{Twi} in terms of \ili{Akan} dialectology. Current speaker population is estimated at nearly ten million.

\section{Some morphosyntactic features of Akan}\label{§2:morphosyntactic.osam}

In this section I provide an overview of tense/aspect and clause structure in the language. I should point out that \ili{Akan} is a two tone language and one of the outstanding phonological features of the language is the presence of Tongue Root Harmony in all the dialects and rounding harmony in the \ili{Fante} dialect. The discussion throughout the paper will draw examples from the \ili{Fante} (Fa) dialect and in some cases from the \ili{Asante} (As) and \ili{Akuapem} (Ak) dialects as well. Examples drawn from \ili{Fante} will generally not be identified as such; those from \ili{Asante} and \ili{Akuapem} will be indicated as As and Ak, respectively, in parentheses generally at the end of a translation.

\subsection{Verbal affixes}\label{§2.1:verbal.osam}

In \ili{Akan} verbal constructions are generally formed with the verb and its person, tense, aspect, mood, and polarity affixes. There are also verbal prefixes to mark motion towards or away from a deictic centre (Table \ref{tab:1.osam}).

\begin{table}
\begin{tabular}{ll}
\lsptoprule
     \textsc{1sg} & \textit{me-}\\
     \textsc{2sg} & \textit{wo-, i-} (Fa)\\
     \textsc{3sg} & \textit{ɔ-}\\
     \textsc{3sg} (inanimate) & \textit{ɛ-} (in Ak, As and some Fa subdialects)\\
     \textsc{1pl} & \textit{yɛ-}\\
     \textsc{2pl} & \textit{mo-/wɔ-} (Fa)\\
     \textsc{3pl} & \textit{wɔ-/yɛ-}\\
\lspbottomrule
\end{tabular}
\caption {Subject Prefixes}
\label{tab:1.osam}
\end{table}

The pre-verbal affixes in \ili{Akan} include the tense/aspect markers. As I have argued elsewhere \citep{osam1994a,osam2008b}, \ili{Akan} can be said to have a Future Tense and the following aspectual forms: Completive (\textsc{compl}), Perfect (\textsc{perf}), Progressive (\textsc{prog}), Habitual (\textsc{hab}), Continuative (\textsc{cont}), and Consecutive (\textsc{consec}).

The Future tense is coded by the prefix \textit{bɛ-}. The realization of the vowel is determined by vowel harmony, especially in the \ili{Fante} dialect (see \citealt{dolphyne1988}).

\ea\label{ex:2.osam}
\ea\label{ex:2a.osam}
\gll   \'{I}yí  b\'{ɛ}-mà    kóŕ-y\'{ɛ}    à-bà    hóḿ    ńtámú.\\
      this  \textsc{fut}-make  one-be    \textsc{cons}-come  \textsc{2pl.poss}  middle\\
\glt `This will bring unity among you.' \citep[79]{krampah1970}
\ex\label{ex:2b.osam}
\gll \ Dàákyé yè-bé-nyá    bì  á-ká    á-kyèr\'{ɛ}    à-fófóró.  \\
       future   \textsc{1pl.subj}-\textsc{fut}-get  some  \textsc{cons}-say \textsc{cons}-show  \textsc{pl}-new\\
\glt    `In future we will have something to tell others.' (\citealt[34]{adi1973}; Ak)
\z
\z


What I consider the Completive is what in the general literature on \ili{Akan} is referred to as the Past tense. However, I have shown in various places \citep{osam1994a,osam2004,osam2008a} that this form is more of an aspect than tense. The Completive is a suffix in the affirmative; but a prefix in the negative.  

\ea\label{ex:3.osam}
\ea\label{ex:3a.osam}
\gll  Né  ǹ-sèẁ-nóḿ    bèènú   nyìná  yɛ́-ɛ̀ èdzìbàń.\\
     \textsc{3sg}.\textsc{poss} \textsc{pl}-in.law-\textsc{col}  two   all  make-\textsc{compl}  food\\
\glt `Each of his two in-laws cooked.' \citep[57]{krampah1970}
\ex\label{ex:3b.osam}
\gll Né  ǹ-sèẁ-nóḿ    bèènú   nyìná  à-ǹ-yɛ́ èdzìbàń.\\
     \textsc{  3sg}.\textsc{poss} \textsc{pl}-in.law-\textsc{col}  two  all  \textsc{compl-neg}-make  food\\
\glt `His two in-laws did not cook.'
\z 
\z 

The Perfect aspect is realized by the prefix \textit{a-}, and generally agrees with the vowel of the verb root in ATR harmony.

\ea\label{ex:4.osam}
\ea\label{ex:4a.osam}
\gll  Bànyíń nó  á-tɔ̀     àsàásé.\\
       man  \textsc{def}  \textsc{perf}-buy  land\\
\glt `The man has bought a piece of land.'
\ex\label{ex:4b.osam}
\gll \`{M}-bòfrá  nó  á-bà    fíé.\\
     \textsc{pl}-child  \textsc{def}  \textsc{perf}-come  home\\
\glt `The children have come home.'
\z 
\z 

The Progressive is a prefix, \textit{re-}.

\ea\label{ex:5.osam}
\ea\label{ex:5a.osam}
\gll Mààmé   nó  rè-hwè-hwɛ́     né    bá   nó.\\
       woman    \textsc{def}  \textsc{prog-redu}-look  \textsc{3sg.poss}  child  \textsc{def}\\
\glt `The woman is looking for her child.'
\ex\label{ex:5b.osam}
\gll   Àbòfrá  nó  rù-tù-tú    ǹ-wúrá    nó.\\
       child  \textsc{def}  \textsc{prog-redu}-uproot  \textsc{pl}-weed  \textsc{def}\\
\glt `The child is removing the weeds.'
\z 
\z 

\ili{Akan} also has a Habitual aspect which is realized by tone (see \citealt{dolphyne1988} for more discussion on this).

\ea\label{ex:6.osam}
\ea\label{ex:6a.osam}
\gll Òkùàfó  nó   kɔ̀    hàbáń  mù  ànɔ̀pá    bíárá.\\
       farmer  \textsc{def}  go;\textsc{hab}    farm  in  morning  every\\
\glt `The farmer goes to the farm every morning.'
\ex\label{ex:6b.osam}
\gll Akosua    tɔ̀ǹ     ǹdzɛ́ḿbá  wɔ̀  gúá-mú.\\
       Akosua  sell;\textsc{hab}  things    in  market-in\\
\glt `Akosua sells in the market.'
\z 
\z 


Two of the aspects, the Continuative and Consecutive, are treated as derived aspects. The Continuative is used for stative verbs where the Progressive is used for dynamic verbs. The Consecutive aspect applies to non-initial verbs in a serial construction in which the initial verb is in either the Progressive aspect or the Future tense. Generally, the Consecutive is realized as a low tone \textit{à-}.

\ea\label{ex:7.osam}
\ea\label{ex:7a.osam}
\gll   Àbòfrá  nó  tsè    dùá  nó  ásé.\\
       child  \textsc{def}  sit/\textsc{cont}  tree  \textsc{def}  under\\
\glt `The child is sitting under the tree.'
\ex\label{ex:7b.osam}
\gll Kwesi  gyìnà    fíé  nó  ényíḿ.\\
       Kwesi  stand/\textsc{cont}  house  \textsc{def}  front\\
\glt `Kwesi is standing in front of the house.'
\z 
\z 


\ea\label{ex:8.osam}
\ea\label{ex:8a.osam}
\gll  Bànyíń  nó  bɛ́-sáẃ   à-kyèrɛ́    dɔ́ḿ  nó.\\
       man  \textsc{def}  \textsc{fut}-dance  \textsc{cons}-show  crowd  \textsc{def}\\
\glt   `The man will dance for the crowd.'
\ex\label{ex:8b.osam}
\gll    Kofi  rè-náńtsèẁ  à-kɔ̀    skúùl.\\
       Kofi  \textsc{prog}-walk  \textsc{cons}-go  school\\
\glt   `Kofi is walking to school.'
\z 
\z 


There are verbal constructions in \ili{Akan} which sometimes have two prefixes, \textit{bɛ-} and \textit{kɔ-}. These prefixes arise from the verbs for `come' and `go', respectively. I have referred to these as Motional prefixes \citep{osam2002}. They reflect the movement towards or away from a deictic centre where an event takes place. Movement towards the deictic centre is marked by the `come' verb; and away from the centre is marked by the `go' verb. 

\ea\label{ex:9.osam}
\ea\label{ex:9a.osam}
\gll Ama  bɔ̀-sɔ́-ɔ̀      gyá  wɔ̀  dáń    nó  ékyì\`{r}.\\
       Ama  come-light-\textsc{compl}  fire  at  building  \textsc{def}  back\\
\glt   `Ama came and lit a fire behind the building.'
\ex\label{ex:9b.osam}
\gll  Esi  kɛ̀-fá-à      èkùtú    nó  bá-à    fíé.\\
       Esi  go-take-\textsc{compl}  orange    \textsc{def}  come-\textsc{compl}  home\\
\glt   `Esi went and brought the oranges home.'
\z 
\z 

Even though it is too early to say that \ili{Akan} has a prospective aspect, the language has the means to express prospective meaning. This is done through a combination of the Progressive and the `come' Motional prefix.

\ea\label{ex:10.osam}
\ea\label{ex:10a.osam}
\gll   Ɔ̀màǹpànyíń  rè-bɔ́-sɔ́ŕ.\\
       president  \textsc{prog}-come-stand\\
\glt   `The president is about to rise.'
\ex\label{ex:10b.osam}
\gll    Hyɛ́ń    nó  rè-bé-gyíná.\\
       vehicle    \textsc{def}  \textsc{prog}-come-stop\\
\glt   `The vehicle is about to stop.'
\z 
\z 

The language makes a two way distinction in terms of mood -- the indicative and the Imperative. The Imperative has two manifestations; what we may call Imperative proper and the Optative. Whereas the Imperative has no segmental representation, the Optative is realized through the use of a homorganic nasal with a high tone, \textit{ń-}. 

\ea\label{ex:11.osam}
\ea\label{ex:11a.osam}
\gll Gyàè!\\
       stop/\textsc{imp}\\
\glt   `Stop it.'
\ex\label{ex:11b.osam}
\gll  Yɛ́-ń-kɔ́.\\
       \textsc{1pl}-\textsc{opt}-go\\
\glt   `Let's go.'
\z 
\z 

The language uses a homorganic nasal prefix to express negation.

\ea\label{ex:12.osam}
\ea\label{ex:12a.osam}
\gll  Mì-ǹ-nyí\`{m}     èdwúmá  â   ɔ́-yɛ́.\\
       \textsc{1sg.subj}-\textsc{neg}-know  work    \textsc{rel}  \textsc{3sg.subj}-do\\
\glt   `I don't know what work she does.'
\ex\label{ex:12b.osam}
\gll   Mààmé    nó  à-ǹ-ká       àsɛ́ḿ  nó  à-ǹ-kyèrɛ́ né    ḿ-bá    nó.\\
       woman    \textsc{def}  \textsc{compl}-\textsc{neg}-say  matter  \textsc{def}  \textsc{compl}-\textsc{neg}-show    \textsc{3sg.poss}  \textsc{pl}-child  \textsc{def}\\
\glt   `The woman didn't tell her children about the case.'  
\z 
\z 

\subsection{Verbal reduplication}\label{§2.2:verbalreduplicaion.osam}

Many verbs in \ili{Akan} are subject to the morphological process of reduplication. Generally, verbal reduplication is required where the event is iterative and/or where either the Subject or Direct Object NPs or both are plural.

\ea\label{ex:13.osam}
\ea\label{ex:13a.osam}
\gll   Áféí  ǹ-nà\`{m}fò   yí   kyìǹ-kyíń-ń     à-dɔ́fó     kàkrá â  wó-níḿ     wɔ́ń     nó   sò krá-krà-à       wɔ̀ǹ.\\
       now  \textsc{pl}-friend  these  \textsc{redu}-roam-\textsc{compl}  \textsc{pl}-lover  few \textsc{rel}  \textsc{3pl.subj}-know  \textsc{3pl.obj}  \textsc{dcm}  on \textsc{redu}-bid.farewell-\textsc{compl}  \textsc{3pl.obj}\\
\glt   `Now these two friends visited some of their friends to bid them farewell.' (\citealt[41]{adi1973}; Ak)
\ex\label{ex:13b.osam}
\gll   Né    nà    nà  n-égyà      é-wú-wú.\\
       \textsc{3sg.poss}  mother    and  \textsc{3sg.poss}-father  \textsc{perf-redu}-die\\
\glt   `Both his mother and father are dead.' \citep[24]{martin1936}
\z 
\z 

\subsection{\ili{Akan} clause structure}\label{§2.3:akan.osam}

The syntax of \ili{Akan} distinguishes grammatical relations, as has been discussed in various studies \citep{osam1994a,osam1996,osam1997,osam2000,osam2004}. The language has Subject and Direct Object with Nominative Accusative marking that is realized through word order. Word Order in the language is strictly SVO, with very little variation.

\subsubsection{Transitive constructions}\label{§2.3.1:transitive.osam}

The prototypical \ili{Akan} transitive clause has A (as Subject) and O (as Direct Object) as core arguments. The A argument precedes the O; and each can be realized as full NP or as a pronominal element.

\ea\label{ex:14.osam}
\ea\label{ex:14a.osam}
\gll  Akosua    kyè-è    àkókɔ́    nó.\\
       Akosua  catch-\textsc{compl}  chicken  \textsc{def}\\
\glt   `Akosua caught the chicken.' (As)
\ex\label{ex:14b.osam}
\gll    Ò-kùà-fó    nó  á-dáá-dàà    ɔ̀-bɔ̀-fó    nó.\\
       \textsc{sg}-farm-\textsc{idm}    \textsc{def}  \textsc{perf-redu}-deceive  \textsc{sg}-hit-\textsc{idm}  \textsc{def}\\
\glt   `The farmer has deceived the hunter.'
\z 
\z 

If the O argument is inanimate, it is not represented by a pronominal form unless some other clausal element comes after it.

\ea\label{ex:15.osam}
\ea\label{ex:15a.osam}
\gll   Ò-wíà-à    náḿ  nó.\\
       \textsc{3sg.subj}-steal-\textsc{compl}  fish  \textsc{def}\\
\glt   `He stole the fish.'
\ex\label{ex:15b.osam}
\gll   Ò-wíà-è      Ø.\\
       \textsc{3sg.subj}-steal-\textsc{compl}\\
\glt   `He stole it.'
\ex\label{ex:15c.osam}
\gll   Ò-wíà-à      nò    ńdédà.\\
       \textsc{3sg.subj}-steal-\textsc{ compl}  \textsc{3sg.obj}  yesterday\\
\glt   `He stole it yesterday.'
\z 
\z 

A transitive clause could also have optional Oblique elements expressed by locational (both spatial and temporal) phrases. It could also be a postpositional phrase. Generally, obliques would occur in clause-initial or clause-final positions. 

\ea\label{ex:16.osam}
\ea\label{ex:16a.osam}
\gll   Mààmé    nó  sí-ì    dáń  ḿfédà.\\
       woman    \textsc{def}  build-\textsc{compl}  house  last.year\\
\glt   `The woman built a house last year.'
\ex\label{ex:16b.osam}
\gll   \'{M}fédà    mààmé    nó  síì    dáń.\\
       last.year  woman    \textsc{def}  build-\textsc{compl}  house\\
\glt   `Last year the woman built a house.'
\z 
\z 


\subsubsection{Intransitive constructions}\label{§2.3.2:intransitive.osam}

The single argument of an intransitive clause, the S argument, always precedes the predicate. 

\ea\label{ex:17.osam}
\ea\label{ex:17a.osam}
\gll  Èdzìbàń  nó  á-bèǹ.  \\
       food    \textsc{def}  \textsc{perf}-be.cooked\\
\glt   `The food is cooked.'  
\ex\label{ex:17b.osam}
\gll   Ɔ̀-sɔ́-fó      nó  wú-ù    ńdédà.\\
       \textsc{sg}-pray-\textsc{idm}    \textsc{def}  die-\textsc{compl}  yesterday\\
\glt   `The priest died yesterday.'
\z 
\z 

\subsubsection{Ditransitive constructions}\label{§2.3.3:ditransitive.osam}

\ili{Akan} has ditransitive constructions in which there are three core arguments, \textsc{agent, benefactive} and \textsc{theme}. In ditransitive constructions, the NP in the immediate post-verbal position is grammatically the Direct Object and semantically the \textsc{benefactive}. The entity that is transferred, the \textsc{theme}, occurs after the \textsc{benefactive} NP. I have referred to this as the \textit{Asymmetrical Object} \citep{osam2000}.

\ea\label{ex:18.osam}
\ea\label{ex:18a.osam}
\gll  Mààmé    nó  má-à    \`{m}-bòfrá  nó  èdzìbáń.\\
       woman    \textsc{def}  give-\textsc{compl}  \textsc{pl}-child  \textsc{def}  food\\
\glt   `The woman gave the children food.'
\ex\label{ex:18b.osam}
\gll   Pàpá  nó  kyɛ́-ɛ̀    hɔ̀ǹ    sìká.\\
       man  \textsc{def}  gift-\textsc{compl}  3\textsc{pl}.\textsc{obj}  money\\
\glt   `The man gave them money.'  
\z 
\z 


\subsubsection{Serial verb constructions}\label{§2.3.4:serial.osam}

One feature of \ili{Akan} syntax crucial to valency discussions is serial verb constructions (\textsc{SVC}). \ili{Akan} serialization has been studied extensively (see, for example, \citealt{lord1973,schachter1974,essilfie1977,forson1990,osam1994a,osam1994b,osam1997,osam2004,osam2014,agyeman2002,hellanetal2003,kambon2012}). Without going into the details of \ili{Akan} SVCs, it is important to identify some salient features.

\textit{Subject marking:} the subject may be a NP or a pronominal form that occurs on only the initial verb. 

\ea\label{ex:19.osam}
\ea\label{ex:19a.osam}
\gll   Yaakwa  n-ábákáń     gyínà-è  yɛ́-ɛ̀ kyèámé bɔ́-ɔ̀      fíé   àmàǹdzɛ̀ɛ́  kyérɛ̀-ɛ̀    à-hɔ́hó    nó.\\
       Yaakwa  \textsc{3sg.poss}-first.born  stand-\textsc{compl}  be-\textsc{compl}   spokesperson tell-\textsc{compl}  home  news    show-\textsc{compl}  \textsc{pl}-visitor  \textsc{def}\\
\glt   `Yaakwa's eldest son stood as the spokesperson and briefed the visitors.' \citep[83]{krampah1970}
\ex\label{ex:19b.osam}
\gll   Wɔ̀-twé-è    nó  gyìná-à  ǹkyɛ́ń.\\
       \textsc{3pl.subj}-pull\textsc{-compl}  \textsc{def}  stand-\textsc{compl}  aside\\
\glt   `They pulled him aside.' (\citealt[31]{adi1973}; Ak)
\z 
\z 


Generally, there is uniformity in tense/aspect coding, as shown in the examples above. However, mixed tense/aspect is possible in some SVCs. 


\textit{Negation Marking:} across all dialects of \ili{Akan}, each verb in the series takes the negation prefix when the sentence is negated, as in \REF{ex:20.osam}. 


\ea
\label{ex:20.osam}
\gll Mààmé  nó  à-ǹ-tɔ́      èdzìbàń  à-\`{m}-má  né    ḿ-bá    nó. \\
     woman  \textsc{def}  \textsc{compl}-\textsc{neg}-buy  food    \textsc{compl}-\textsc{neg}-give \textsc{3.sg.poss}  \textsc{pl}-child  \textsc{def}\\
\glt `The woman did not buy food for her children.'
\z

In some serial constructions, the initial verb is \textit{de/dze}. This is a form with reduced verbal properties. The \textit{de} is used in the \ili{Twi} dialects and \textit{dze} is used in \ili{Fante}. 

\ea\label{ex:21.osam}
\ea\label{ex:21a.osam}
\gll Wɔ̀-dè    àsɛ́ḿ  nó  kɔ́-ɔ̀    àhè\`{m}fíé.\\
       \textsc{3pl.subj}-take  matter  \textsc{def}  go-\textsc{compl}  palace\\
\glt   `They took the case to the palace.' (Ak)
\ex\label{ex:21b.osam}
\gll   Ɔ̀-dzè    n-áhòɔ̀dzéń    nyìná  yɛ́-ɛ̀      èdwúmá nyá-à    sìká.\\
       \textsc{3sg.subj}-take  \textsc{3sg.poss}-strength  all  do-\textsc{compl}  work get-\textsc{compl}  money\\
\glt   `She worked very hard and made money.'
\z 
\z 


The de serialization is also relevant in the expression of agentive arguments in the case of verbs of spatial location. This will be further discussed in \sectref{§5.2:agent.osam}.

\section{\ili{Akan} verb classes}\label{§3:akan.osam}

Based on their argument structure, various transitivity classes of verbs can be identified in \ili{Akan}. Some are strictly intransitive (\sectref{§3.1:strictly.osam}), some are strictly transitive (\sectref{§3.2:strictly.osam}), some are ditransitive (\sectref{§3.3:ditransitive.osam}), and some have varying expressions of arguments. Verbs of the last type are addressed in \sectref{§4:decreasing.osam} and \sectref{§5:increasing.osam}.

\subsection{Strictly intransitive/monovalent}\label{§3.1:strictly.osam}

Monovalent verbs occur with a single argument; the S argument only (Table \ref{tab:2.osam}).

\begin{table}
\begin{tabular}{llll}
\lsptoprule
\textit{béń} & ‘be cooked’ & \textit{bèr̀} & ‘be ripe, be fair in complexion’ \\
\textit{dwó} & `cool' & \textit{fé} & `become soft/smooth'  \\
\textit{hóń} & `swell' & \textit{hwéḿ} & `blow one's nose' \\
\textit{hwèǹtsì} & `sneeze' & \textit{péŕ} & `struggle, roll around during sleep'\\
\textit{prɔ̀} & `rot' & \textit{húŕ} & `boil' \\
\lspbottomrule
\end{tabular}
\caption {Monovalent verbs}
\label{tab:2.osam}
\end{table}

\ea\label{ex:23.osam}
\ea\label{ex:23a.osam}
\gll   Èdzìbàń  nó  á-bèǹ.\\
       food    \textsc{def}  \textsc{perf}-cook\\
\glt   `The food is cooked.'
\ex\label{ex:23b.osam}
\gll  Àǹkàá    nó  á-prɔ́.\\
       oranges  \textsc{def}  \textsc{perf}-rot\\
\glt   `The oranges are rotten.'
\z 
\z 


\subsection{Strictly transitive/bivalent}\label{§3.2:strictly.osam}

There are verbs that require two arguments, A and O (Table \ref{tab:3.osam}).

\begin{table}
\begin{tabular}{llllll}
\lsptoprule
 \textit{bò\`{r}} & `beat' & \textit{dààdàà} & `deceive' & \textit{hyìrà} & `bless' \\
 \textit{ká} & `bite' & \textit{kú} & `kill' & \textit{kyé} & `catch'\\
\textit{ny\'{ɛ}ń} & `rear' & \textit{pàà} & `curse' & \textit{sànè} & `infect'    \\
\textit{sìè} & `bury' & \textit{táń} & `hate'  &  & \\
\lspbottomrule
\end{tabular}
\caption{Bivalent verbs}
\label{tab:3.osam}
\end{table}

\ea\label{ex:25.osam}
\ea\label{ex:25a.osam}
\gll  Ɔ̀-bɔ̀-fóɔ́  nó  kú-ù     ɔ̀sónó.\\
       \textsc{sg}-hit-\textsc{idm}  \textsc{def}  kill-\textsc{compl}  elephant\\
\glt   `The hunter killed an elephant.'  
\ex\label{ex:25b.osam}
\gll  Abam  síè-è     né    mààmé.\\
       Abam  bury-\textsc{compl}  \textsc{3sg.poss}  mother\\
\glt   `Abam buried his mother.'
\z 
\z 


\subsection{Ditransitive/trivalent}\label{§3.3:ditransitive.osam}

There are verbs that are ditransitive or trivalent. These require three core arguments (Table \ref{tab:4.osam}).

\begin{table}
\begin{tabular}{llll}
\lsptoprule
\textit{ky\'{ɛ}} & ‘to gift, give as a gift’ & \textit{gyé} & ‘charge’ \\
\textit{má} & `give' & \textit{kyèr\`{ɛ}} & `teach, show'  \\
\lspbottomrule
\end{tabular}
\caption{Trivalent verbs}
\label{tab:4.osam}
\end{table}

\ea\label{ex:27.osam}
\ea\label{ex:27a.osam}
\gll \'{N}tsí  ɔ̀-má-à      nò    à-hɛ́ń    èbìásá.  \\
       so  \textsc{3sg.subj}-give-\textsc{compl}  \textsc{3sg.obj}  \textsc{pl}-boat    three\\
\glt `So she gave him three boats.' \citep[8]{martin1936}
\ex\label{ex:27b.osam}
\gll   Esi  kyɛ́-ɛ̀    pàpá  nó   sìká.\\
       Esi  gift-\textsc{compl}  man  \textsc{def}  money\\
\glt   `Esi gave the man money.'
\z 
\z 


\section{Decreasing valence}\label{§4:decreasing.osam}

There are morphosyntactic processes that reduce verb valence. In various languages, morphological derivations are utilized in reducing the valence of a verb. But as has been pointed out in the literature, verbs can also manifest a change in the valence structure without the application of any morphological processes: ``Alternations in a verb's valency pattern are not necessarily the result of a morphological derivational process. Verbs or whole classes of verbs may have alternate valency patterns without any change in their formal makeup ...'' \citep[1131]{haspelmathmuellerbardey2004}.

\subsection{Reflexivization}\label{§4.1:reflexivization.osam}

Reflexivization in \ili{Akan} is marked by the use of a possessive (\textsc{poss}) pronoun and the morpheme \textit{ho} `self'.

\ea\label{ex:28.osam}
\ea\label{ex:28a.osam}
\gll   Kofi  é-kù    \ili{Yaw}.\\
       Kofi  \textsc{perf}-kill  \ili{Yaw}\\
\glt `Kofi has killed \ili{Yaw}.'
\ex\label{ex:28b.osam}
\gll   Kofi    é-kù    nó    hó.\\
       Kofi    \textsc{perf}-kill  \textsc{3sg.poss}  self\\
\glt `Kofi has killed himself.'
\z 
\z 


In reflexivization, the notion of coreference is crucial. \citet[44]{kemmer1993} notes that ``Coreference ...  means that two participants in a single event frame designate the same entity in the described situation.''  It requires that the A and O arguments have the same referent. Reflexive constructions in \ili{Akan} involving bivalent verbs have the coreferential O argument replaced by the POSS+\textit{ho}.  The reduction in the valency of the verb lies in the fact that there is no semantic differentiation between the A and O arguments. 

Apart from the direct reflexive, certain verbs, specifically grooming, or body care actions \citep{kemmer1993} are used reflexively and, consequently, demonstrate (semantic) valency decreasing properties. Examples of such verbs in \ili{Akan} are: \textit{pepa} `wipe', \textit{sera} `smear/use lotion or oil on the body', \textit{twutwuw} `wash (as with a sponge/washcloth)', \textit{siesie} `dress up'. The actions coded by these verbs can be carried out by an \textsc{agent} on a \textsc{patient} entity or the \textsc{agent} can carry it out on themselves. 

\ea\label{ex:29.osam}
\ea\label{ex:29a.osam}
\gll   Aba  twùtwúẁ-ẁ    Ekua  nó    hó.\\
       Aba  wash.wash-\textsc{compl}  Ekua  \textsc{3sg.poss}  self\\
\glt   `Aba washed Ekua.'
\ex\label{ex:29b.osam}
\gll   Aba  twùtwúẁ-ẁ    nó    hó.\\
       Aba  wash.wash-\textsc{compl}  \textsc{3sg.poss}  self\\
\glt   `Aba washed someone/herself.'
\z 
\z 


\ea\label{ex:30.osam}
\ea\label{ex:30a.osam}
\gll   Efua  sérà-à    Kwesi    nó    hó.\\
       Efua  smear-\textsc{compl}  Kwesi    \textsc{3sg.poss}  self\\
\glt   `Efua used lotion on Kwesi.'
\ex\label{ex:30b.osam}
\gll    Efua  sérà-à    nó    hó.\\
       Efua  smear-\textsc{compl}  \textsc{3sg.poss}  self\\
\glt   `Efua used lotion on someone/herself.'
\z 
\z 


In \REF{ex:29a.osam} and \REF{ex:30a.osam}, the entities in subject position, as the \textsc{agents}, carry out the activities on certain other individuals, the entities in the \textsc{patient} role. In \REF{ex:29b.osam} and \REF{ex:30b.osam}, the replacement of the full NP with the reflexive pronoun in the post-verbal position creates ambiguities. In each sentence, the referent of the reflexive pronoun could be the entity in the subject position, the \textsc{agent}; or it could be an individual already mentioned in the context of the discourse; that is Ekua in \REF{ex:29a.osam} and Kwesi in \REF{ex:30a.osam}.

Where the referents of the reflexive pronouns in \REF{ex:29b.osam} and \REF{ex:30b.osam} are the \textsc{agents} in the subject positions in the two sentences, we can argue for a reduction in the valency of the verbs \textit{sera} `smear/use lotion or oil on the body' and \textit{twutwuw} `wash (as with a sponge/washcloth)', on the basis of coreferentiality and non-individuation of the participants involved in the situation. 

In addition to the reduction in the valency of the verb \textit{sera} through reflexivization, its valency can also be reduced through its use in an intransitive construction as \REF{ex:31b.osam} shows.

\ea\label{ex:31.osam}
\ea\label{ex:31a.osam}
\gll  Efua  sérà-à    nó    hó.\\
       Efua  smear-\textsc{compl}  \textsc{3sg.poss}  self\\
\glt   `Efua used lotion on herself.'
\ex\label{ex:31b.osam}
\gll Efua  sérà-è.\\
       Efua  smear-\textsc{compl}\\
\glt   `Efua used lotion (on herself).'
\z 
\z 


There is another feature of the verb \textit{sera} pertaining to valency adjustment that will be discussed in \sectref{§4.5.2:unspecified.osam}

\subsection{Reciprocals}\label{§4.2:reciprocals.osam}

The reciprocal in \ili{Akan} is formed in ways similar to the reflexive. It also uses the morpheme \textit{ho}. But unlike the reflexive, the possessive pronoun that combines with \textit{ho} is in the plural \REF{ex:32.osam}. Similar to the reflexive, the reciprocal creates reduced valency due to coreferentiality and lack of individuation.

\ea\label{ex:32.osam}
\ea\label{ex:32a.osam}
\gll   Hwɛ́,  émí     nà   wó-égyà     yè-húń-ǹ hɛ́ń    hó   béŕ   â  mí-dzí-ì     \`{m}-fé   èdùònú   ànáń nó. \\
       look  \textsc{1sg.emph}  and  \textsc{2sg.poss}-father  \textsc{1pl.subj}-see-\textsc{compl} \textsc{1pl.poss}  self  time  \textsc{rel}  \textsc{1sg.subj}-eat-\textsc{compl}  \textsc{pl}-year twenty    four    \textsc{dcm}\\
\glt `Your father and I got to know each other when I was twenty-four years.' \citep[9]{martin1936}
\ex\label{ex:32b.osam}
\gll  \`{M}-bèráńtsé  nà  ǹ-kàtàásíá  nó   nyé     Araba Akɔm  hám-èè      yè-yáẁ-ẁ    hɔ̀ń    hó  mà ɔ̀-yɛ́-ɛ̀        àsɛ́ḿ  wɔ̀  skúùl. \\
       \textsc{pl}-man    and  \textsc{pl}-woman  \textsc{def}  accompany  Araba Akɔm quarrel-\textsc{compl}    \textsc{redu}-insult-\textsc{compl}  \textsc{3pl.poss}  self  that \textsc{3sg.subj}-make-\textsc{compl}  issue  in  school\\
\glt `The young men and young women quarreled with Araba Akɔm and insulted each other such that it became an issue in the school.' \citep[21]{martin1936}  
\z 
\z 


\subsection{Anticausative/Inchoative}\label{§4.3:anticausative.osam}

The anticausative (or decausative, inchoative, spontaneous, pseudopassive, being the various ways in which this type of construction has been labelled) works by removing the \textsc{agent} argument in the construction. In his characterization of the causative/inchoative alternation, \citeauthor{haspelmath1993} states that:

\begin{quote}
An inchoative/causative verb pair is defined semantically: it is a pair of verbs which express the same basic situation (generally a change in state, more rarely a going-on) and differ only in that the causative verb meaning includes an Agent participant who causes the situation, whereas the inchoative verb meaning excludes a causing Agent and presents the situation as occurring spontaneously ... Inchoative verbs are generally intransitive and causative verbs are transitive ... \citep[90]{haspelmath1993}
\end{quote}

As pointed out by \citet[1132]{haspelmathmuellerbardey2004}: 

\begin{quote}
In many languages there is a strong requirement for all sentences to have subjects. When in such languages a valency-changing category removes the agent argument from the subject position, the patient argument must take up the subject position instead.
\end{quote}

\ili{Akan} is a typical example of such languages. When a sentence has a single argument, that argument is always in the subject position. In \ili{Akan} anticausatives the \textsc{theme} argument is the subject of the sentence. 

Unlike some other languages where there is a derivational process to indicate either the demotion of the \textsc{agent} argument or its introduction, in \ili{Akan} there is no change in the morphology of the verb stem to reflect the process of anticausative. This puts \ili{Akan} into what \citet[91]{haspelmath1993} describes as non-directed alternation, that is, where ``... neither the inchoative nor the causative verb is derived from the other.''

The verbs in the language that can be used in the anticausative construction include: \textit{bɔ} `break', \textit{bu} `break', \textit{bue} `open', \textit{butu} `overturn', \textit{dum} `put out', \textit{hyew} `burn', \textit{koa} `bend', \textit{kyea} `bend', \textit{nane} `melt', \textit{sɛe} `destroy', \textit{te} `tear, pluck', \textit{tuei} `puncture', \textit{woso} `shake, vibrate'. 

Example \REF{ex:33a.osam} and \REF{ex:34a.osam} are causative and \REF{ex:33b.osam} and \REF{ex:34b.osam} are anticausative/inchoative.

\ea\label{ex:33.osam}
\ea\label{ex:33a.osam}
\gll nà  wɔ̀-bò-bɔ́-ɔ̀  ǹ-kùràbá   nó â ɔ́-dzé-dzè    hɔ́ń    ńsá-mú    nó.\\
       and  \textsc{3pl.subj}-\textsc{redu}-break\textsc{-compl}  \textsc{pl}-jar    \textsc{def}  \textsc{rel}    \textsc{3pl.subj}-\textsc{redu}-hold  \textsc{3pl.poss}  hand-in  \textsc{dcm}\\
\glt `...  and they broke the jars that were in their hands.' (Judges 7:19 \ili{Fante} Bible; \citealt{bible1974})
\ex\label{ex:33b.osam}
\gll   ǹ-kùràbá   nó  â ɔ́-dzé-dzè    hɔ́ń ńsá-mú    nó    bò-bɔ́-è. \\
       \textsc{pl}-jar    \textsc{def}  \textsc{rel}  \textsc{3pl.subj}-\textsc{redu}-hold  \textsc{3pl.poss}  hand    \textsc{dcm}    \textsc{redu}-break-\textsc{compl}\\
\glt `...  and the jars in their hands broke.'
\z 
\z 

\ea\label{ex:34.osam}
\ea\label{ex:34a.osam}
\gll nà  Moses    hyéẁ-ẁ   tsíŕ  nó.\\
       and   Moses     burn-\textsc{compl}  head  \textsc{def}\\
\glt `...  and Moses burned the head.' (Leviticus 8:20 \ili{Fante} Bible; \citealt{bible1974})
\ex\label{ex:34b.osam}
\gll  nà  tsíŕ   nó  hyéẁ-èè.\\
       and  head  \textsc{def}  burn-\textsc{compl}  \\
\glt `. .  . and the head burned.'
\z 
\z 


\subsection{Impersonal constructions}\label{§4.4:impersonal.osam}

Another means of valency decrease is through impersonal constructions. The notion of impersonal construction adopted here follows \citet{siewierska2008,siewierska2011} and \citet{malchukovogawa2011}. These are constructions that do not have a referential subject. \citet{malchukovogawa2011}, following \citet{keenan1976}, argue that the subject in an impersonal construction deviates from the prototype subject. In \citeauthor{keenan1976}'s (\citeyear{keenan1976}) approach, the canonical subject is expected to have the following properties \citep[23]{malchukovogawa2011}:

\begin{itemize}[noitemsep]
 \item  a referential argument
 \item a definite NP
 \item topical
 \item animate
 \item  agentive
\end{itemize}


Based on cross-linguistic studies, various coding strategies of impersonal constructions have been isolated. One of these is the pronominal impersonal \citep{siewierska2011}. In some languages this involves the use of a regular personal pronoun as the subject of the construction. \ili{Akan} does this by using the regular 3 person plural subject pronoun, as illustrated in \REF{ex:35.osam}.

\ea\label{ex:35.osam}
\ea\label{ex:35a.osam}
\gll  Wɔ̀-á-kyè\`{r}    èwì-fó    nó.\\
       \textsc{3pl.subj}-\textsc{perf}-catch  thief-\textsc{pl}  \textsc{def}\\
\glt `They have arrested the thieves.'/`The thieves have been arrested.'
\ex\label{ex:35b.osam}
\gll   Wɔ̀-á-tò    ésíkyìré  nó    bó  mú.\\
       \textsc{3pl.subj}-\textsc{perf}-raise  sugar    \textsc{3sg.poss}  price  in\\
\glt   `They have increased the price of sugar.'/`The price of sugar has been increased.'
\z 
\z 


The subject pronouns in \REF{ex:35.osam} are non-referential and non-individuated \citep{hopperthompson1980}. Example \REF{ex:36.osam} below is taken from the Apostles' Creed of the Christian faith. 

\ea
\label{ex:36.osam}
\gll Wɔ̀-bɔ́-ɔ̀       Nò    \`{m}bèàmúdúá  mù, Ò-wú-ì,     wò-síé-è     Nò.\\
     \textsc{3pl.subj}-hit\textsc{-compl}  \textsc{3sg.obj}  cross in \textsc{3sg.subj}-die  \textsc{3pl.subj}-bury\textsc{-compl}  \textsc{3sg.obj}\\
\glt `He was crucified, dead and buried.' (Source: Christian Asɔr Ndwom; \citealt{methodist1937})
\z

Even though in the constructions in \REF{ex:35.osam} and \REF{ex:36.osam} there are two arguments, A and O, functionally, there is reduced valency because the subjects are not prototypical subjects. The sentences have \textsc{agent} subjects. Nonetheless, the \textsc{agents} involved are not distinct because the pronominal form used is non-referential. This makes it impossible to identify the referent, an obvious way of downgrading the \textsc{agent} argument.

\subsection{Object omission/suppression}\label{§4.5:object.osam}

Decreasing valency can also involve verbs that potentially have A and O arguments. However, the O argument remains suppressed either because it is understood or it is known that any one of a range of entities can fill that argument position.

\subsubsection{Understood object}\label{§4.5.1:understood.osam}

Verbs that allow understood objects, falling in the category of Inherent Complement Verbs (\textit{dwanse} `urinate', \textit{bow} `be drunk', \textit{nye} `defecate', \textit{dɔr} `be fatty', and sa `dance') can occur with both overt A and O arguments. However, in many instances, the O argument is not expressed because speakers know what it is, thereby reducing the grammatical valence of the verb. The verbs and their inherent complements are listed below:

\ea\label{ex:37.osam}
\ea\label{ex:37a.osam}
\gll  dwàǹsè   dwáńsé\\
   `urinate'  `urine' \\
\ex\label{ex:37b.osam}
\gll   bòẁ    ǹsá\\ 
  {`be drunk'}  `drink/alcohol'\\
\ex\label{ex:37c.osam}
\gll   nyè    bíń\\ 
  `defecate'  `faeces' \\
\ex\label{ex:37d.osam}
\gll    dɔ̀rè    sèràdé\'{ɛ}\\
  {`be fatty'}  `fat'  \\
\ex\label{ex:37e.osam}
\gll    sà    àsá\\
   `dance'    {`a dance'}  \\
\z 
\z 

\ea\label{ex:38.osam}
\ea\label{ex:38a.osam}
\gll   Bànyíń  nó  á-bòẁ    ǹsá.\\
       man  \textsc{def}  \textsc{perf}-be.drunk  alcohol\\
\glt `The man is drunk (with alcohol).'
\ex\label{ex:38b.osam}
\gll  Bànyíń  nó  á-bòẁ.\\
       man  \textsc{def}  \textsc{perf}-be.drunk\\
\glt `The man is drunk.'
\z 
\z 


\ea\label{ex:39.osam}
\ea\label{ex:39a.osam}
\gll   Àpɔ̀ǹkyé  nó  á-dɔ̀rè      sèràdé\'{ɛ}.\\
       goat    \textsc{def}  \textsc{perf}-be.fatty    fat\\
\glt   `The goat is fatty.' (Lit. `The goat is fatty with fat.') (As)
\ex\label{ex:39b.osam}
\gll   Àpɔ̀ǹkyé  nó  á-dɔ́ré.      \\
       goat    \textsc{def}  \textsc{perf}-be.fatty  \\
\glt   `The goat is fatty.' (As)
\z 
\z 


In \REF{ex:38a.osam} and \REF{ex:39a.osam} the verbs are used with their inherent complements. However, in \REF{ex:38b.osam} and \REF{ex:39b.osam}, there are no overt complements, reflecting a reduction in the grammatical valence of the verbs \textit{bow} `be drunk' and \textit{dɔre} `be fatty'.

\subsubsection{Unspecified Object}\label{§4.5.2:unspecified.osam}

\ili{Akan} has a limited number of verbs that demonstrate varying valency: monovalent, bivalent, and trivalent. The reduction in valence of these verbs revolves around the non-expression of the O argument in all instances. This means that the single argument of the intransitive construction is an \textsc{agent}. It is more appropriate then to talk about the suppression of the O argument. Another feature of these verbs is that they tend to take various items as the O argument. Examples of the verbs are: \textit{soa} `carry', \textit{hyɛ} `wear, dress up', \textit{sera} `apply body lotion, smear', \textit{sa} `administer enema', \textit{son} `apply herbal nasal drop', \textit{tua} `douche'.

\ea\label{ex:40.osam}
\ea\label{ex:40a.osam}
\gll   Araba  sérà-à    ǹkú.\\
       Araba  smear-\textsc{compl}  shea.butter\\
\glt `Araba applied shea butter on her body.'
\ex\label{ex:40b.osam}
\gll   Araba  sérà-è.\\
       Araba  smear-\textsc{compl}\\
\glt `Araba applied (some substance) to her body.'
\z 
\z 

In the first example above, the A and O arguments are expressed. However in the second example the O argument remains unexpressed; the single argument, S, is the \textsc{agent}. The ditransitive use of these verbs is illustrated in \REF{ex:41.osam}:

\ea\label{ex:41.osam}
\ea\label{ex:41a.osam}
\gll  Araba  sérà-à    àbòfrá  nó  ǹkú.\\
       Araba  smear-\textsc{compl}  child  \textsc{def}  shea.butter\\
\glt `Araba applied shea butter on the child.'
\ex\label{ex:41b.osam}
\gll   Araba  sérà-à    àbòfrá  nó.\\
       Araba  smear-\textsc{compl}  child  \textsc{def}\\
\glt `Araba applied (some substance) on the child.'
\z 
\z 


In \REF{ex:41a.osam}, the verb has three arguments: A, E, and O, in that order. In \REF{ex:41b.osam}, the O argument is unspecified, leaving the A and E arguments.

\section{Increasing valency}\label{§5:increasing.osam}

There are processes that lead to an increase in the arguments of a verb. There are two main ways in \ili{Akan} that this happens: the introduction of an \textsc{agent} through serialization and causativization.

\subsection{Causativization}\label{§5.1:causativization.osam}

Causative constructions increase valence by adding a causing \textsc{agent} to an event. As seen in \sectref{§4.3:anticausative.osam}, \ili{Akan} has verbs that permit the anticausative construction. The causative variant would have an \textsc{agent} present, as in \REF{ex:41a.osam} and \REF{ex:43a.osam}. In \REF{ex:42b.osam} and \REF{ex:43b.osam}, the change in state of the \textsc{patient} entities ` chains and stick, respectively ` is captured without the specification of the responsible \textsc{agent}. In \REF{ex:42a.osam} and \REF{ex:43a.osam}, on the other hand, the events are presented with the causing \textsc{agents} overtly stated. Effectively, in \REF{ex:42a.osam} and \REF{ex:43a.osam} where we have the causative constructions, the valency of the verbs has been increased by the addition of the causing \textsc{agents}.

\ea\label{ex:42.osam}
\ea\label{ex:42a.osam} Causative \\
\gll  Nà  ò-bù-bú-ù      hɔ́ń    \`{m}pòkyèr\'{ɛ}  mù   èsìǹ-ésíń.\\
     and  \textsc{3sg.subj}-\textsc{redu}-break-\textsc{compl}  \textsc{3pl.poss}  chains    in  \textsc{redu}-piece\\
\glt `... and he broke their chains in pieces.' (\ili{Fante} Bible Psalm 107:14;  \citealt{bible1974})
\ex\label{ex:42b.osam} Anticausative\\
\gll Hɔ́ń  \`{m}pòkyèr\'{ɛ}   mù  bù-bú-ù     èsìǹ-ésíń.\\
     \textsc{3pl.poss}  chains    in  \textsc{redu}-break-\textsc{compl}  \textsc{redu}-piece\\
\glt `Their chains broke into pieces.'
\z 
\z 


\ea\label{ex:43.osam}
\ea\label{ex:43a.osam} Causative \\
\gll Kofi  á-kyéá    àbàá  nó.\\
     Kofi  \textsc{perf}-bend  stick  \textsc{def}\\
\glt `Kofi has bent the stick.'
\ex\label{ex:43b.osam} Anticausative \\
\gll Àbàá  nó  á-kyéá.\\
     stick  \textsc{def}  \textsc{perf}-bend\\
\glt `The stick is bent.'
\z 
\z 


\subsection{Agent introduction through serialization}\label{§5.2:agent.osam}

There are verbs that code the location or spatial configuration of an entity. Examples  include: \textit{da} `lie, be at', \textit{twer} `lean', \textit{bea} `lie', \textit{sɛn} `hang', \textit{hyɛ} ` be in', \textit{gu} `be in', \textit{tar} `paste, stick', \textit{fam} `stick', and \textit{si} `stand'. For verbs like these, the introduction of an agentive NP requires the use of a serial construction. Even though the resulting construction cannot be said to increase the valency of the verbs, it shows how an agentive argument can be introduced through the syntactic strategy of serialization.

\ea\label{ex:44.osam}
\ea\label{ex:44a.osam}
\gll  Nà  ɔ̀-dzè     \`{m}-pòmá   nó  hyè-hy\'{ɛ}-\`{ɛ}  àdáká     nó    hó. \\
             and  \textsc{3sg.subj}-take  \textsc{pl}-pole    \textsc{def}  \textsc{redu}-put-\textsc{compl} box    \textsc{3sg.poss}  self  \\
\glt  `And he took the poles and put them on the ark.'/`And he put the poles on the ark.' (\ili{Fante} Bible, Exodus 40: 20; \citealt{bible1974})
\ex\label{ex:44b.osam}
\gll    \`{M}-pòmá   nó   hyè-hy\`{ɛ}   àdáká  nó    hó.\\
       \textsc{pl}-pole    \textsc{def}  \textsc{redu-cont}  box  \textsc{3sg.poss}  self  \\
\glt   `The poles are on the ark.'
\z 
\z 


\ea\label{ex:45.osam}
\ea\label{ex:45a.osam}
\gll   Wɔ̀-dè    wɔ̀ń    á-ky\'{ɛ}ḿ   sèǹ-s\'{ɛ}nè w-à-fásúó    hó   hyìá. \\
       \textsc{3pl.subj}-take  \textsc{3pl.poss}  \textsc{pl}-shield  \textsc{redu}-hang/\textsc{hab} \textsc{3sg.poss-pl}-wall  self  meet/\textsc{hab}  \\
\glt `They hang their shields around your walls.' (Asante Bible, Ezekiel 27: 11;  \citealt{bible1964})
\ex\label{ex:45b.osam}
\gll   Wɔ̀ń    á-ky\'{ɛ}ḿ    sèǹ-s\'{ɛ}nè     w-à-fásúó hó  hyìá.\\
       \textsc{3pl.poss}  \textsc{pl}-shield  \textsc{redu}-hang/\textsc{hab}  \textsc{3sg.poss-pl}-wall self  meet/\textsc{hab}\\
\glt `Their shields hang around your walls.' (As)
\z 
\z 

\ea\label{ex:46.osam}
\ea\label{ex:46a.osam}
\gll Kàǹdzéá  nó  sì    póń  nó  dó.\\
       lantern    \textsc{def}  stand/\textsc{cont}  table  \textsc{def}  on\\
\glt `The lantern is on the table.'
\ex\label{ex:46b.osam}
\gll   Ato  dzè  kàǹdzéá  nó  sí-ì    póń  nó  dó.\\
       Ato  take  lantern    \textsc{def}  stand-\textsc{compl}  table  \textsc{def}  on\\
\glt `Ato placed the lantern on the table.'
\z 
\z 


\section{Valence adjustment through reduplication}\label{§6:valence.osam}

One valency adjusting morphological process in \ili{Akan} with limited application is verbal reduplication. So far, only two verbs have been identified in the language that change their valence when reduplicated. The verbs are \textit{da} `sleep' and \textit{di} `eat'. Their reduplicated forms are \textit{deda} `put to sleep' and \textit{didi} `eat'.

\ea\label{ex:47.osam}
\ea\label{ex:47a.osam}
\gll  Ama  dà-à    àwìá.\\
       Ama  sleep-\textsc{compl}  afternoon\\
\glt   `Ama slept in the afternoon.' (As)
\ex\label{ex:47b.osam}
\gll   Ama  dè-dá-à    àbòfrá    nó  àwìá.\\
       Ama  \textsc{redu}-sleep\textsc{-compl}  child    \textsc{def}  afternoon\\
\glt   `Ama put the child to sleep in the afternoon.' (As)
\z 
\z 


As shown in \REF{ex:47b.osam}, the reduplication of the verb \textit{da} `sleep' is a means by which an \textsc{agent} argument is introduced.

The behavior of \textit{di} when reduplicated is the reverse of the reduplication of \textit{da}. When \textit{di} is reduplicated, it loses the capacity to have a \textsc{patient} argument.

\ea\label{ex:48.osam}
\ea\label{ex:48a.osam}
\gll  Kofi  dí-ì    àdùáń    nó.\\
       Kofi  eat-\textsc{compl}  food    \textsc{def}\\
\glt   `Kofi ate the food.' (Ak)
\ex\label{ex:48b.osam}
\gll Kofi  dì-dí-ì.\\
       Kofi  \textsc{redu}-eat-\textsc{compl}\\
\glt   `Kofi ate.' (Ak)
\ex\label{ex:48c.osam}
\gll   Kofi  dì-dí-ì      àdùáń    nó.\\
       Kofi  \textsc{redu}-eat-\textsc{compl}  food    \textsc{def}\\
\glt   `Kofi ate the food.'
\z 
\z 


There is a polysemous use of \textit{didi} where it takes a postpositional phrase in the post-verbal position \REF{ex:49.osam}. But the use of the reduplicated form of the verb \textit{di} in this situation is metaphorical and does not contradict the case made about the verb.

\ea\label{ex:49.osam}
\ea\label{ex:49a.osam}
\gll  Òbíárá    dì-dí    n-àdwúmá    hó.\\
       everybody  \textsc{redu}-eat  \textsc{3sg.poss}-work  self\\
\glt   `Everybody benefits from their work.' (Ak)
\ex\label{ex:49b.osam}
\gll    \`{N}krɔ̀fó  nó  dì-dí-ì      ɔ̀héń   nó  ásé.\\
       people  \textsc{def}  \textsc{redu}-eat-\textsc{compl}  chief  \textsc{def}  under\\
\glt   `The people sabotaged the chief.' (Ak)
\z 
\z 


\section{Summary and conclusion}\label{§7:summary.osam}

The notion of valency has received extensive treatment in the linguistics literature. I set out in this paper to examine the valency of \ili{Akan} verbs and to investigate the morphosyntactic ways in which the valency of verbs can be modified. 

I have shown in the preceding discussion that we can identify verbs in the language that are invariably monovalent ` that is, verbs that take only one core argument, the S argument; those that are bivalent ` requiring two core arguments, A and O; and those that are trivalent, needing three core arguments, that is, A, O, and E.

Apart from the verbs with invariant argument structure, there are many verbs that exhibit variations in the expression of their arguments. I have shown in the paper that overall, the morphosyntactic mechanisms by which the valency of verbs is modified in \ili{Akan} fit into various cross-linguistic patterns. \ili{Akan} is not known as a language with complex morphology. Consequently, the valency adjusting processes tend to be more syntactic than morphological. Verbs in the language that can undergo a reduction in the expression of their arguments do so through reflexivization, the use of reciprocals, anticausatives, impersonal constructions, and various forms of object suppression. It has been shown in the paper that where there is valency decrease resulting in only one argument being expressed, the single argument is always the S argument.  Increase in verb valency is achieved through causativization, and agentivization through serialization. It has also been demonstrated that the language uses reduplication in a very limited way to adjust verb valency. As stated in the paper, the use of reduplication applies to only two verbs in the language.

\citet[25-27]{dixonaikhenvald2000b}, in ending their paper, identify some topics that need to be investigated regarding the notion of valency cross-linguistically:

\begin{quote}
Our preliminary impression is that, across the languages of the world, there tend to be more valency-increasing derivations (comitative and applicative) than valency-reducing derivations (passive, antipassive, reflexive, reciprocal, etc.). This needs to be verified, through study of a large representative sample of languages; if it is true, linguists should seek an explanation. \citep[26]{dixonaikhenvald2000b}  
\end{quote}

From what has been presented in this paper, it is obvious that in \ili{Akan} there are more valency decreasing morphosyntactic strategies than those used to increase valency. Based on my knowledge of the languages that are genetically related to \ili{Akan}, for example, Ga and \ili{Ewe}, one would expect a similar tendency. However, this is an issue that needs to be investigated. More broadly, the tools that have been developed by the Leipzig Project and by Nichols and associates need to be applied to \ili{Akan} and related languages in order to contribute to our further understanding of the notion valency.

\section*{Abbreviations}


\begin{tabularx}{.55\textwidth}{lQ}
{\sc{col}} &   collective \\     
{\sc{compl}} & completive       \\
{\sc{cons}} &   consecutive      \\    
{\sc{cont}} & continuative      \\   
{\sc{dcm}} & dependent    clause marker \\
{\sc{def}} &   definite      \\
{\sc{emph}} &   emphatic      \\
{\sc{fut}} &   future      \\
{\sc{hab}} &   habitual      \\
{\sc{idm}} &   identity marker      \\
{\sc{imp}} &   imperative      \\
\end{tabularx}
\begin{tabularx}{.40\textwidth}{lX}
{\sc{neg}} & negation      \\
{\sc{obj}} &   objective     \\
{\sc{opt}}  &   optative     \\  
{\sc{perf}} &   perfect     \\
{\sc{pl}}  &   plural      \\    
{\sc{poss}} & possessive       \\
{\sc{prog}} &  progressive     \\ 
{\sc{redu}} &   reduplication      \\
{\sc{rel}} &   relativizer      \\
{\sc{sg}} &   singular      \\
{\sc{subj}} &   subject      \\
\end{tabularx}

{\sloppy
\printbibliography[heading=subbibliography,notkeyword=this]
}
\end{document}