\addchap{Acknowledgments}
Many thanks

To the Moloko men and women who shared their stories and fables with me.  These are the people whose stories we have used for this analysis: Abelden, Ali Gaston, Baba Abba, Dungaya, Dungaya Daniel, Dugujé, Kama Joseph, Majay Moïze, Mala, Malatina Moïze, Mana Samuel, Njida, Sali Anouldéo Justin, Tajay Suzanne, Tajike, and Tsokom.

To the Moloko men who transcribed and translated the texts, entered them into the computer, and helped me understand what they mean: Ali Gaston, Holmaka Marcel, Mana Djeme Isaac, Mana Samuel, Oumar Abraham, Sali Anouldéo Justin, and Sambo Joël.

To colleagues who also worked among the Moloko:  Megan Mamalis, Alan and DeEtte Starr, Ginger Boyd, and Catherine Bow.  

To Jenni Beadle, for smoothly taking the verb tone files from shoebox to the chart in the appendix.

To Dr. Aaron Shryock, Rhonda Thwing, and Richard Gravina, for tireless interest in the intricacies of Moloko, and miles and miles of red ink in the early drafts.

To Sean Allison, for gracious, detailed comments and challenges on one of the later drafts.

To Dr. Doris Payne, for incredible insights, encouragement, and perseverance. 

To Barb Penner and Felix Kopecky, for expert typesetting. 

To the editors, staff and many reviewers from Language Science Press for their wise and discerning oversight and guidance in all aspects of bringing this document to publication and making it available to others. 

To the Moloko people who welcomed me to their land and into their homes, and for whose sakes we strive to understand more about this language.  

\textit{Malan manjan ana Hərmbəlom!} (Glory to God)