\documentclass[output=paper]{langsci/langscibook} 
\ChapterDOI{10.5281/zenodo.3744503}
\author{Ahmad Al-Jallad\affiliation{The Ohio State University}}
\title{Pre-Islamic Arabic}
% \keywords{} 
\abstract{This chapter provides an overview of Arabic in contact in the pre-Islamic period, from the early first millennium BCE to the rise of Islam. Contact languages include Akkadian, Aramaic, Ancient South Arabian, Canaanite, Dadanitic, and Greek. The chapter concludes with two case studies on contact-induced development: the emergence of the definite article and the realization of the feminine ending.}

\begin{document}
\maketitle

\section{Preliminaries}  
\subsection{Language contact in the pre-Islamic period}

\begin{quote}
[I]n the Djāhiliyya, ``the Age of Ignorance'' […], the Arabs lived to a great extent in almost complete isolation from the outer world… [t]his accounts for the prima facie astonishing fact that \ili{Arabic}, though appearing on the stage of history hundreds of years after the Canaanites and Aramaeans, nevertheless in many respects has a more archaic character than these old \ili{Semitic} languages. The Arabs, being almost completely isolated from outer influences and living under the same primitive conditions of their ancestors preserved the archaic structure of their language. (\citealt{Blau1981}: 18).
\end{quote}
This is the image of \ili{Arabic}’s pre-Islamic past that emerges from \ili{Classical} \ili{Arabic} sources. For writers such as Ibn Khaldūn,\ia{Ibn Khaldūn@Ibn Khaldūn} contact-induced change in \ili{Arabic} was a by-product of the Arab conquests, and served to explain the differences between the colloquial(s) of his time and the literary language. More than a century and a half of epigraphic and archaeological research in Arabia and adjacent areas has rendered this view of \ili{Arabic}’s past untenable. \ili{Arabic} first appears in the epigraphic record in the early first millennium BCE, and for most of its pre-Islamic history, the language interacted in diverse ways with a number of related \ili{Semitic} languages and \ili{Greek}. This chapter will outline the various foci of contact between \ili{Arabic} and other languages in the pre-Islamic period based on documentary evidence. Following this, I offer two short case studies showing how contact-induced change in the pre-Islamic period may explain some of the key features of \ili{Arabic} today.   

\subsection{Old Arabic}
\ili{Old} \ili{Arabic} is an umbrella term for the diverse forms of the language attested in documentary and literary sources from the pre-Islamic period, including inscriptions, papyri, and transcriptions in \ili{Greek}, \ili{Latin}, and cuneiform texts. The present usage does not refer to \ili{Classical} \ili{Arabic} or the linguistic material attributed to the pre-Islamic period collected in the eighth and ninth centuries CE, such as poetry and proverbs, as we cannot be sure about their authenticity, especially with regard to their linguistic features. \citet{Al-Jallad2017early} defines the corpus of \ili{Old} \ili{Arabic} as follows: \ili{Safaitic}, an Ancient North Arabian script concentrated in the Syro-Jordanian Ḥarrah (end of the 1st millennium BCE to 4th c. CE), \ili{Hismaic}, an Ancient North Arabian script spanning from central Jordan to northwest Arabia (chronology unclear, but overlapping with \ili{Nabataean}), the substratum of \ili{Nabataean Aramaic}, along with a few \ili{Arabic}-language texts carved in this script (2nd c. BCE to 4th c. CE), the Nabataeo-\ili{Arabic} inscriptions (3rd c. CE to 5th c. CE), pre-Islamic \ili{Arabic} script inscriptions (5th c. CE to early 7th c. CE) and isolated inscriptions in the \ili{Greek}, \ili{Dadanitic} (the oasis of Dadān, modern-day al-ʕUlā, northwest Ḥiǧāz), and Ancient \ili{South Arabian} alphabets (varied chronology). 

In geographic terms, \ili{Old} \ili{Arabic} is attested mainly in the southern Levant, the Sinai, and northwestern Arabia, as far south as Ḥegrā (Madāʔin Ṣāleḥ). Within this area a variety of non-\ili{Arabic} languages were spoken and written, with which \ili{Old} \ili{Arabic} interacted. The main contact language was \ili{Imperial Aramaic}, which served as a literary language across North Arabia in the latter half of the first millennium BCE until, perhaps, the rise of Islam. Since contact must be viewed through the lens of writing, it is in most cases difficult to determine how extensive multilingualism was outside of literate circles.

\section{Contact languages}
\subsection{Arabic and Akkadian}
The first attestations of \ili{Arabic} are preserved in cuneiform documents. While no \ili{Arabic} texts written in cuneiform have yet been discovered, isolated lexical items survive in this medium. Livingstone (\citeyear{Livingstone1997}) identified an example of the \ili{Old} \ili{Arabic} word for `camel' with the {definite} {article} in the inscriptions of Tiglathpileser III (744--727 BCE): \textit{a-na-qa-a-te} = (\textit{h/ʔ})\textit{an-nāq-āte} ‘the she-camels’. Aside from this, almost all other \ili{Arabic} material consists of personal and divine names. There are reports of “Arabs” in Mesopotamia – inhabiting walled towns in western Babylonia – as early as the eighth century BCE (\citealt{Ephal1974}: 112). While we cannot be sure that the people whom the Babylonians called Arabs were in fact \ili{Arabic} speakers, a few texts in dispersed Ancient North Arabian scripts hail from this region. So far, all seem to contain only personal names with \ili{Arabic} or Arabian etymologies.\footnote{``Dispersed Ancient North Arabian'' is a temporary term given to the Ancient North Arabian inscriptions on seals, pottery, bricks, etc. which have been found in various parts of Mesopotamia and elsewhere \citep[33]{Macdonald2000}.}  These facts can only suggest the possibility of contact between speakers of \ili{Arabic} and \ili{Akkadian} in the early first millennium BCE.

\subsection{Arabic and Canaanite}\label{Canaanite}
Contact between \ili{Arabic} speakers and speakers of \ili{Canaanite} languages is documented in the \ili{Hebrew} Bible (\citealt{Ephal1982}: ch.2; \citealt{Retsö2003}: ch.8), and there is one inscription directly attesting to contact between both groups. An Ancient North Arabian inscription from Bāyir, Jordan contains a prayer in \ili{Old} \ili{Arabic} to three gods of the Iron Age \ili{Canaanite} kingdoms of Moab, Ammon, and Edom (\citealt{Hayajnehetal2015}). The text is accompanied by a \ili{Canaanite} inscription, which remains undeciphered. The reading of the \ili{Arabic} according to the edition is as follows:

\ea Bāyir inscription (\citealt{Hayajnehetal2015})\\
\gll h 	mlkm 	w-kms 	w-qws 	b-km 		   ʕwðn h-ʔsḥy 		m-mdwst	\\
     \textsc{voc} \textsc{pn} \textsc{conj-pn} \textsc{conj-pn} \textsc{prep-2pl.m} protect.\textsc{prf.1pl} \textsc{dem}-well.\textsc{pl} \textsc{prep}-ruin \\
\glt ‘O Malkom, Kemosh, and Qaws, we place under your protection these wells against ruin.’
\z

\subsection{Arabic and Aramaic}
Evidence for contact between \ili{Arabic} and \ili{Aramaic} spans from the middle of the first millennium BCE to the late sixth century CE, and is concentrated in the southern Levant and northwest Arabia.\footnote{See \citet{Stein2018} on the role of \ili{Aramaic} in the Arabian Peninsula in the pre-Islamic period.} Perhaps one of the earliest examples of \ili{Arabic} speakers using \ili{Aramaic} as a written language comes from the fifth-century-BCE Nile Delta. A king of Qedar, Qayno \ia{Qayno@Qayno} son of Gośam,\ia{Gośam@Gośam}\footnote{The symbol \textit{ś} denotes the \ili{Old} \ili{Arabic} reflex of \ili{Classical} \ili{Arabic} 〈\kern 1pt{\arabscript{ش}}〉, which is usually transcribed \textit{š}. /ś/ was likely realized as a voiceless lateral fricative [ɬ].} commissioned an \ili{Aramaic} votive inscription dedicated to \textit{hn-ʔlt} ‘the goddess’ (\citealt{Rabinowitz1956}). \ili{Arabic} names can be found in transcription across the Levant in \ili{Aramaic} inscriptions (\citealt{Israel1995}), and in most cases names with an \ili{Arabic} etymology terminating in the characteristic final \textit{-w}, reflecting an original nominative case (\citealt{Al-Jalladforthcoming}). 

\ili{Arabic} and \ili{Aramaic} language contact reaches a climax in the written record at the end of the first millennium BCE with the arrival of inscriptions in the Nabataean script. The Nabataeans established a kingdom in the region of Edom in the fourth century BCE, which at its greatest extent spanned from the Ḥawrān to the northern Ḥiǧāz. While they, like their contemporaries across the Near East, wrote in a form of \ili{Imperial} \ili{Aramaic}, the spoken language of the royal house and large segments of the population was \ili{Arabic}. Unlike other examples of \ili{Aramaic} written by \ili{Arabic} speakers so far, Nabataean incorporated \ili{Arabic} elements into its writing school, such as the optative use of the perfect, the negator \textit{ɣayr}, and a significant number of lexical items relating to daily life (\citealt{Gzella2015}: 242--243).

Perhaps one of the most interesting examples of contact between the two languages is found in Nabataean legal papyri from the Judaean desert (1st--2nd c. CE). These \ili{Aramaic}-language legal documents contain a number of glosses in \ili{Arabic}, for example: \textit{ʕqd} /ʕaqd/ ‘contract’; \textit{mʕnm} /maɣnam/ ‘profit’; \textit{prʕ} /faraʕ/ ‘to branch out’; \textit{ṣnʕh} /ṣanʕah/ ‘handiwork’, etc. (\citealt{Yardeni2014}). Macdonald (\citeyear{Macdonald2010}: 20) has suggested, based on this evidence, that {Nabataean} legal proceedings would have taken place in \ili{Arabic}, while all written records were made in \ili{Aramaic}.

In addition to the use of \ili{Arabic} within \ili{Aramaic}, a unique votive inscription from ʕEn ʕAvdat (Negev, Israel) contains three verses of an \ili{Arabic} hymn to the deified Nabataean king \ia{Obodat@ʕObodat}ʕObodat embedded within an \ili{Aramaic} text. While undated (but likely earlier than 150 CE), the text is certainly the earliest example of continuous \ili{Arabic} language written in the Nabataean script, as before this almost all examples are isolated words and personal names.\largerpage[-3]\pagebreak

\ea ʕEn ʕAvdat inscription\footnote{This is my translation; the \textit{editio princeps} is Negev, Naveh \& Shaked (\citeyear{Negevetal1986}); it is discussed most recently in Fiema et al. (\citeyear{Fiemaetal2015}: 399–402) and Kropp (\citeyear{Kropp2017}).} \\
\ea{Aramaic} \\ 
\gll    dkyr b-ṭb q[r]ʔ qdm ʕbdt ʔlhʔ w-dkyr mn ktb grmʔlhy br tymʔlhy šlm lqbl ʕbdt ʔlhʔ \\
remember.\textsc{ptcp.pass} \textsc{prep}-good read.\textsc{ptcp.act} \textsc{prep} \textsc{pn} god.\textsc{def} \textsc{conj}-remember.\textsc{ptcp.pass} \textsc{rel} write.\textsc{prf.3sg.m} \textsc{pn} son.\textsc{cs} \textsc{pn} be\_secure.\textsc{prf.3sg.m} \textsc{prep} \textsc{pn} god.\textsc{def} \\ 
\glt `May he who reads this aloud be remembered for good before ʕObodat the god, and may he who wrote be remembered. May Garmallāhi son of Taymallāhi be secure in the presence of ʕObodat (the god).'\\

\ex{Arabic} \\
\gll   p-ypʕl lʔ pdʔ w-lʔ ʔtrʔ p-kn hnʔ ybʕ-nʔ ʔl-mwtw lʔ ʔbʕ-h p-kn hnʔ ʔrd grḥw lʔ yrd-nʔ \\
       \textsc{conj}-act.\textsc{impf.3sg.m} \textsc{neg} ransom.\textsc{acc} \textsc{conj}-\textsc{neg} scar.\textsc{acc} \textsc{conj}-be.\textsc{inf} here seek.\textsc{impf.3sg.m-1pl} \textsc{def}-death.\textsc{nom} \textsc{neg} make.obtain.\textsc{inf-3sg.m} \textsc{conj}-be.\textsc{inf} here want.\textsc{prf.3sg.m} wound.\textsc{nom} \textsc{neg} want.\textsc{impf.3sg.m-1pl} \\
\glt `May he act that there be neither ransom nor scar; so be it that death would seek us, may he not aid its seeking, and so be it that a wound would desire (a victim), let it not desire us!'\\

\ex{Aramaic} \\
\gll   grmʔlhy ktb yd-h \\
\textsc{pn} writing.\textsc{cs} hand.\textsc{3sg.m} \\
\glt `Garmallāhi, the writing of his hand.’\\
\z
\z

The presence of \ili{Aramaic} is much more lightly felt in the desert hinterland to the east and north of Nabataea. A small handful of \ili{Safaitic}--\ili{Aramaic} bilingual inscriptions are known (\citealt{Hayajneh2009}: 214--215). In one \ili{Safaitic} text, produced by a Nabataean, the author gives his name and affiliation to social groups in a type of \ili{Aramaic}, but then writes the remainder of the inscription in \ili{Old} \ili{Arabic}, suggesting that this individual may have been bilingual.\pagebreak

\ea Nabataean \ili{Safaitic} (\citealt{Al-Jallad2015Safaitic}: 19; C 2820) \\ 
\gll l ʔʔsd bn rbʔl bn ʔʔsd bn rbʔl nbṭwy slmy w brḥ ḫlqt śty h-dr w tð̣r h-smy \\
\textsc{prep} \textsc{pn} son.\textsc{cs} \textsc{pn} son.\textsc{cs} \textsc{pn} son.\textsc{cs} \textsc{pn} Nabataean Salamite \textsc{conj} depart.\textsc{prf.3sg.m} period.\textsc{cs} winter \textsc{def}-region \textsc{conj} keep\_watch.\textsc{prf.3sg.m} \textsc{def}-sky \\
\glt `By ʔAʔsad son of Rabbʔel son of  ʔAʔsad son of Rabbʔel, the Nabataean Salamite, and he set off from this place for the period of winter and kept watch for the rains.' \\
\z

A handful of \ili{Aramaic} loans are found in the \ili{Safaitic} inscriptions: \textit{sfr} ‘writing’; \textit{ʔsyt} ‘hide, trap’; \textit{lṣṭ} ‘thief’, ultimately from \ili{Greek} \textit{lēistḗs}. Other words, such as \textit{mdbr} /madbar/ ‘the Hamad, wilderness’ and \textit{nḫl} /naḫl/ ‘valley’, are absent in \ili{Classical} \ili{Arabic} yet appear in the Northwest \ili{Semitic} languages. These do not appear to be loans, however, as their meanings and phonologies are local and \ili{Arabic}, respectively. They should instead be regarded as genuine cognates that did not make it into the Islamic-period lexica. 

\subsection{\textit{Provincia Arabia} and the Nabataeo-Arabic script}\label{Provincia}
In 106 CE, under circumstances that remain poorly understood, the Romans annexed the Nabataean Kingdom and established their Province of Arabia. While Nabataean political independence ended, their script, writing tradition and language continued to thrive and evolve. This is exemplified by the famous tomb inscription of Raqōś bint ʕAbd-Manōto from Madāʔin Ṣāliḥ.\ia{Raqōś bint ʕAbd-Manōto@Raqōś bint ʕAbd-Manōto} Dated to 267 CE, the text is a legal inscription associated with the grave of a woman who died in al-Ḥegr. Unlike other grave inscriptions at this site, the Raqōś\ia{Raqōś bint ʕAbd-Manōto@Raqōś bint ʕAbd-Manōto} inscription is composed almost entirely in \ili{Arabic}, with the \ili{Aramaic} components restricted to the introductory demonstrative \textit{dnh} ‘this’, the words for ‘son’ and ‘daughter’, the dating formula, and the name of the deity. The \ili{Aramaic} components are bolded below:


\ea Madāʔin Ṣāliḥ inscription (JSNab 17)\footnote{For the latest discussion of this text, see Macdonald’s contribution to Fiema et al. (\citeyear{Fiemaetal2015}: 402--405).}\label{Raqosh}\\ 
\gll   \textbf{dnh} qbrw ṣnʕ-h kʕbw \textbf{br} ḥrtt l-rqwš \textbf{brt} ʕbdmnwtw ʔm-h w-h hlkt fy ʔl-ḥgrwy \textbf{šnt} \textbf{mʔh} \textbf{w-štyn} \textbf{w-tryn} \textbf{b-yrḥ} \textbf{tmwz} w-lʕn \textbf{mryʕlmʔ} mn yšnʔ ʔl-qbrw d[ʔ] 	w-mn 		yftḥ-h ḥšy w wld-h w-lʕn mn yqbr w-{y}ʕly mn-h \\
       \textsc{dem} grave build.\textsc{prf.3sg.m-3sg.m} \textsc{pn} son \textsc{pn} \textsc{prep-pn} daughter \textsc{pn} mother-\textsc{3sg.m} \textsc{conj-3sg.f} die.\textsc{prf.3sg.f} \textsc{prep} \textsc{def-pn} year hundred \textsc{conj}-sixty \textsc{conj-two} \textsc{prep}-month Tammūz \textsc{conj}-curse.\textsc{prf.3sg.m} \textsc{pn} \textsc{rel} desecrate.\textsc{impf.3sg.m} \textsc{def}-grave \textsc{dem} \textsc{conj-rel} open.\textsc{impf.3sg.m-3sg.m} except w children-\textsc{3sg.m} \textsc{conj}-curse.\textsc{prf.3sg.m} \textsc{rel} bury.\textsc{impf.3sg.m} \textsc{conj}-remove.\textsc{impf.3sg.m} \textsc{prep-3sg.m} \\
\glt   `\textbf{This} is a grave that Kaʕbo \textbf{son} of Ḥāreθat constructed for Raqōś \textbf{daughter} of ʕAbd-Manōto,\ia{Raqōś bint ʕAbd-Manōto@Raqōś bint ʕAbd-Manōto} his mother, and she perished in al-Ḥegro \textbf{year one hundred and sixty two in the month of Tammūz}. May \textbf{the Lord of the World}  curse anyone who desecrates this grave and anyone who would open it, with the exception of his children, and may he curse anyone who would bury or remove from it (a body).'\\
\z

During the same period, the classical Nabataean script continues to evolve towards what we consider the \ili{Arabic} script (\citealt{Nehmé2010}). Its letter forms take on a more cursive character, and the connecting element of each letter goes across the bottom of the text. Nehmé considers the letter forms typical of the \ili{Arabic} script to have evolved from Nabataean between the third and fifth centuries CE. In inscriptions from this period, the \ili{Arabic} component begins to increase at the expense of \ili{Aramaic} (\citealt{Nehmé2017}). This trend may suggest that knowledge of \ili{Aramaic} was waning in these centuries, or that the writing tradition itself was transforming – \ili{Aramaic} was slowly being replaced by \ili{Arabic}. If we think in terms of writing schools, there may not have been much \ili{Arabic}--\ili{Aramaic} {bilingualism} in Arabia outside of the scribal class – indeed, scholars have continued to debate whether \ili{Nabataean Aramaic} was ever a colloquial, and there are good arguments to doubt that it was (\citealt{Gzella2015}: 240). The remnants of \ili{Aramaic} in the latest phases of the Nabataeo-\ili{Arabic} inscriptions, however, most certainly functioned as a code, grams for \ili{Arabic} words, a situation comparable to the Aramaeograms of Pahlavi (cf. \citealt{Nyberg1974}).

\subsection{The Arabic inscriptions of the sixth century CE}
In \ili{Arabic} inscriptions of the sixth century, written \ili{Arabic} and \ili{Aramaic} continue the stable situation of contact witnessed in the Nabataeo-\ili{Arabic} period. \ili{Aramaic} fossils are employed in dating formulae and the word for ‘son’, and possibly the first person pronoun. But otherwise, the language of these texts is entirely \ili{Arabic}. Perhaps the most famous among these is the inscription of Jebel Usays, given in \REF{Usays}, in which the \ili{Aramaic} components are bolded.

\ea\label{Usays} Jebel Usays inscription\footnote{For the latest discussion of this text, see Macdonald’s contribution to Fiema et al. (\citeyear{Fiemaetal2015}: 405).}\\
\gll   \textbf{ʔnh}\footnotemark{} rqym \textbf{br} mʕrf ʔl-ʔwsy ʔrsl-ny ʔlḥrt ʔl-mlk ʕly ʔsys mslḥh snt 423 \\
       \textsc{1sg} \textsc{pn} son \textsc{pn} \textsc{def}-Awsite send.\textsc{prf.3sg.m-1sg} \textsc{pn} \textsc{def}-king \textsc{prep} Usays outpost year 423  \\
\glt   `\textbf{I}, Ruqaym \textbf{son} of Muʕarrif the Awsite, al-Ḥāriθ the king sent me to Usays as an outpost, year 423.' [= 528/9 CE]
\z

\footnotetext{While it has been suggested that the spelling \textit{ʔnh} reflects a pausal form \citep{Larcher2010}, it seems more likely in light of the Thaʕlabah Nabataeo-\ili{Arabic} inscription \citep{AvnerNehmeRobin2013}, which spells `I' as \textit{ʔnh}, that this form reflects the \ili{Aramaic} spelling of the pronoun rather than an \ili{Arabic} variant.}

\subsection{Arabic, Greek and Aramaic in sixth-century Petra}
In 1993, a corpus of carbonized \ili{Greek} papyri – some 140 rolls – was discovered at the Byzantine church of Petra.\footnote{These papyri are edited in a five-volume series: the \textit{Petra Papyri I--V} (2002--2018), various editors, Amman: American Center of Oriental
Research. See \citet{Arjavaetal2018} for the last volume.} These documents attest to a trilingual situation at the city: \ili{Greek} served as the official administrative language, while \ili{Arabic} and \ili{Aramaic} appear to have been spoken languages. The microtoponyms (names of small plots of lands and vineyards) are in both \ili{Arabic} and \ili{Aramaic}, and oftentimes the same word is expressed in both languages, as in Table \ref{tab:Petra}. 

\begin{table}
\caption{Arabic--Aramaic equivalents in the Petra Papyri (\citealt{Al-Jallad2018Petra}: 41)}
\label{tab:Petra}
\begin{tabular}{lllll}
\lsptoprule
Translation & \ili{Arabic} & \ili{Aramaic} \\
\midrule
`land markers' & \multicolumn{1}{l}{Αραμ /ārām/} & \multicolumn{1}{l}{Εραμαεια /eramayyā/}  \\
`farm' & \multicolumn{1}{l}{αλ-Ναϲβα /al-naṣbah/} & \multicolumn{1}{l}{Ναϲβαθα /naṣbatā/}  \\
`canal' & \multicolumn{1}{l}{αλ-Κεϲεβ /al-qeṣeb/} & \multicolumn{1}{l}{Κιϲβα/Κειϲβα /qiṣbā/}  \\
\lspbottomrule
\end{tabular}
\end{table}

This naturally suggests that, alongside literacy in \ili{Greek}, there was spoken bilingualism in \ili{Arabic} and \ili{Aramaic}, perhaps a stable situation extending back to Nabataean times. 

\subsection{Arabic and Ancient South Arabian}
\ili{Classical} \ili{Arabic} sources note a situation of close contact between \ili{Arabic} and “Ḥimyaritic”, a term used for a language they associated with the pre-Islamic kingdom of Ḥimyar in what is today {Yemen}. The pre-Islamic inscriptions from the northern Yemeni Jawf, the so-called Haram region, attest to a similar situation. These texts are composed in Sabaic, but contain a significant admixture of non-Sabaic linguistic material. Some scholars (e.g. \citealt{Robin1991}) have considered \ili{Arabic} to be the contributing source, but in most cases the non-Sabaic linguistic features are not specific to \ili{Arabic}, such as the use of the {causative} verb ʔaCCaC, which is attested in \ili{Aramaic} and \ili{Gəʕəz} for example, rather than haCCaC as in Sabaic. As Macdonald (\citeyear{Macdonald2000}: 55) rightly puts it, these inscriptions are basically Sabaic, with a small admixture from North Arabian languages, but not necessarily \ili{Arabic}. Four texts from this region, however, exhibit the \ili{Arabic} isogloss of \textit{lam} for past-{tense} {negation}, suggesting that some form of \ili{Arabic} may have contributed to their mixed character.\footnote{For a list of the Haram inscriptions, see Macdonald (\citeyear{Macdonald2000}: 61), who labels these texts Sabaeo-North-Arabian.}   

Mixed North/South Arabian texts can be found further to the north, in Naǧrān and Qaryat al-Fāw. The most famous is perhaps the grave inscription of \textit{Rbbl bn Hfʕm}. This unique text attests features that can be attributed to both non-Sabaic and Sabaic sources. On the non-Sabaic side, it uses the {definite} {article} \textit{ʔl}, the {causative} morpheme \textit{ʔ-} rather than \textit{h-}, and occasionally the 3rd person pronoun \textit{h} rather than \textit{hw}. At the same time, the text employs mimation, {clitic} pronouns with long vowels, e.g. \textit{-hw}, and {prepositions} not known in \ili{Arabic} (\citealt{Al-Jallad2018ANA}: 30). At Naǧrān, one occasionally encounters \ili{Arabic} lexical items, such as \textit{ldy} ‘at’ and \textit{ʕnd} ‘with’ in otherwise perfectly good \ili{South Arabian} texts. So then, how are we to interpret the mixed character of these texts? For Qaryat al-Fāw, Durand (\citeyear{Durand2017}: 95, fn.32) has suggested, based on the significant amount of Petraean pottery, that a sizable Nabataean colony existed at the oasis. It could be the case that Nabataean colonists introduced \ili{Arabic} to the oasis, where it naturally gained {prestige} as a trade language given its links with the north. The mixed nature of some of the inscriptions of this site could therefore be interpreted in two ways. If they reflect a spoken variety, then perhaps they are the result of {convergence} between the \ili{Arabic} introduced by the Nabataeans and Sabaic, similar to the modern dialects of \ili{Yemeni} \ili{Arabic} today, which are essentially \ili{Arabic} with a significant \ili{South Arabian} admixture.\footnote{On these varieties, see \citet{Watson2018}.} If we are dealing with an artificial scribal {register}, then the language may be the result of a scribe attempting to produce a text in \ili{Arabic}, for an \ili{Arabic}-speaking customer, but inadvertently introducing Sabaicisms from the language he is more used to writing. A similar phenomenon might be at play in the \ili{Aramaic}–\ili{Hasaitic} tomb inscription from Mleiha.\footnote{\ili{Hasaitic} is the name given to the pre-Islamic script and language of East Arabia.} There, the scribe – seemingly unintentionally – uses the \ili{Aramaic} word for son, \textit{br}, in the \ili{Hasaitic} portion of the text, suggesting perhaps that he was bilingual and more used to writing in \ili{Aramaic} (\citealt{Overlaetetal2016}).

\subsection{Arabic in the Ḥiǧāz}
Before the arrival of the Nabataeans, the written language of the oasis of al-ʕUlā and associated environs in the northern Ḥiǧāz was \ili{Dadanitic}, a non-\ili{Arabic} \ili{Central Semitic} language. A few texts, however, display features that are unambiguously \ili{Arabic}. The best known of these is JSLih 384. This short text is written in the \ili{Dadanitic} script but seems to be, in other respects, produced in a dialect of \ili{Old} \ili{Arabic}, notably making use of the {relative} pronoun \textit{ʔlt} /ʔallatī/. Two other \ili{Dadanitic} texts make use of the \ili{Arabic} construction \textit{ʔn yfʕl}, that is, the use of the subordinator \textit{ʔan} with a {modal} verb. In addition to this, one occasionally finds the \textit{ʔ(l)} {definite} {article} employed in these inscriptions. The interpretation of this contact situation, like that in South Arabia, is unclear. Do these few texts represent the writings of travelers or immigrants from the north, whose spoken language influenced the dictation of text to the scribe? Or do they reflect unique points on a dialect continuum? The complex linguistic situation at ancient Dadan is the subject of a fascinating study by \citet{Kootstra2019}.

\subsection{Arabic and the languages of the Thamudic inscriptions}\largerpage
Even more difficult to distill is the possible contact situation between \ili{Arabic} and the more shadowy pre-\ili{Arabic} \ili{Semitic} languages of north and central Arabia. We are afforded a small glimpse of these languages by the laconic \ili{Thamudic} inscriptions, mainly those classified in the C, D, and F scripts.\footnote{\ili{Thamudic} B, C, and D are discussed in Macdonald (\citeyear{Macdonald2000}) and Al-Jallad (\citeyear{Al-Jallad2017early}; \citeyear{Al-Jallad2018ANA}); \ili{Thamudic} F is outlined in Prioletta \& Robin (\citeyear{PriolettaRobin2018}).}  While it is difficult to say much about the languages these scripts express, they are clearly distinct from \ili{Arabic} (\citealt{Al-Jallad2017early}: 321--322). The only evidence for contact between \ili{Arabic} and any of these languages is found in the tomb inscription of Raqōś\ia{Raqōś bint ʕAbd-Manōto@Raqōś bint ʕAbd-Manōto} at Madāʔin Ṣāliḥ, illustrated in \REF{Raqosh}. This text, as discussed in §\ref{Provincia}, is written mainly in \ili{Arabic}, with a few fossilized \ili{Aramaic} components. Alongside the main inscription, there is a short text inscribed in the \ili{Thamudic} D script stating: \textit{ʔn rqś bnt ʕbdmnt} ‘This is Raqōś, daughter of ʕAbdo-Manōto’.\ia{Raqōś bint ʕAbd-Manōto@Raqōś bint ʕAbd-Manōto} The use of the introductory element \textit{ʔn} `this' or perhaps `for', rather than the \ili{Arabic} demonstrative \textit{dʔ} /ðā/ or perhaps its feminine equivalent \textit{dy} /ðī/, employed in the Nabataean text, indicates that we are dealing with a third language.\footnote{While it is tempting to interpret \textit{ʔn} as the first-person singular pronoun \textit{ʔanā}, such a formula would indeed be strange in a grave epitaph. Perhaps \textit{ʔn} is {cognate} with the demonstrative/presentative element *han, or perhaps it should be construed as a dative `to, for' {cognate} with East \ili{Semitic} \textit{ana}.} Did Raqōś\ia{Raqōś bint ʕAbd-Manōto@Raqōś bint ʕAbd-Manōto} originally hail from a nomadic community who spoke a non-\ili{Arabic} \ili{Semitic} language expressed in the \ili{Thamudic} D script? And did she later come to live in \ili{Arabic}-speaking Ḥegrā? Was the use of this script on her grave a tribute to her heritage? These {questions} are impossible to answer with the data available to us now, but they widen the scope of investigation when examining \ili{Arabic}’s history. The available fragments of evidence support the suggestion put forth recently by Souag (\citeyear{Souag2018blog}): we must consider the possibility of unknown \ili{Semitic} {substrate}(s) in the development of early \ili{Arabic}.

\subsection{Arabic and Greek}
The nexus of \ili{Arabic}--\ili{Greek} contact, based on the inscriptions known so far, is the Syro-Jordanian Harrah, the basalt desert that stretches from the Hawrān to northern Arabia. \ili{Greek} inscriptions are occasionally found throughout this region, interacting with the local \ili{Arabic} dialects in diverse ways. The commonest type of bilingual text consists of simple signatures in \ili{Safaitic} and \ili{Greek}. These texts, illustrated in \REF{Taym}, only prove that the author knew how to write his name in \ili{Greek}, and do not constitute evidence for genuine {bilingualism}. 

\ea Graeco-\ili{Arabic} inscription A1 (\citealt{Al-Jalladal-Manaser2016}: 56)\label{Taym}\\
\ea{Greek} \\ 
\gll    Θαιμος Γαφαλου \\
        Taym Gaḥfal\\
\glt `Taym, son of Gaḥfal'\\

\ex{Arabic} \\
\gll   l-tm bn gḥfl \\
       \textsc{prep}-Taym son Gaḥfal \\
\glt `for/by Taym, son of Gaḥfal'\\
\z
\z

 The second inscription discussed by \citet{Al-Jalladal-Manaser2016}, illustrated in \REF{Ghawth}, provides more insight into the different degrees of \ili{Arabic}--\ili{Greek} {bilingualism}. The author carves a short text in both \ili{Greek} and \ili{Old} \ili{Arabic}, indicating that he knew both languages but that his command of \ili{Arabic} was obviously better.

\ea Graeco-\ili{Arabic} inscription 2 (\citealt{Al-Jalladal-Manaser2016}: 58)\label{Ghawth}\\
\ea{Arabic} \\
\gll   l-ɣθ w tḥll ʔfwh ʕql sr \\
       \textsc{prep}-Ghawth \textsc{conj} go.\textsc{prf.3sg.m} \textsc{prep} protected\_area Sayr\\
\glt `By Ghawth and he went into the protected area of Sayr.'\\
\ex{Greek} \\ 
\gll    Γαυτος ἀπῆλθεν [ε]ἰς τόν Ακελον Σαιρου\\
Ghawth.\textsc{nom} depart.\textsc{aor.3sg} \textsc{prep} \textsc{def.m.acc.sg} \textit{ʕaql}.\textsc{acc.sg} Sayr.\textsc{gen}\\ 
\glt `Ghawth, he went away to the \textit{ʕaql} of Sayr.'\\
\z
\z

The author translates the \ili{Arabic} into \ili{Greek} effectively, but seems not to have known the \ili{Greek} word for the culturally specific term \textit{ʕaql}, ‘a protected area of pasturage’. In this case, he simply wrote the word out in \ili{Greek}: \textit{Ακελον}.

There is evidence that some nomadic \ili{Arabic} speakers did master the \ili{Greek} language, as one sometimes comes across very well-composed texts in \ili{Greek}, attesting to full-scale {bilingualism}, at least in writing (for example A2 in \citet{Al-Jalladal-Manaser2015}. This level of {bilingualism}, however, must have been rare. There is no appreciable influence from \ili{Greek} on the \ili{Arabic} of the \ili{Safaitic} inscriptions. A few {loanwords} are known, e.g. \textit{qṣr} ‘Caesar’, \textit{lṣṭ} ‘thief', but these more likely come through \ili{Aramaic}.

\subsection{Arabic in eastern Arabia}
The inscriptional record of eastern Arabia is relatively poor when compared to the western two-thirds of the Peninsula. Nevertheless, the extant texts point towards contact between \ili{Aramaic} and the local Arabian language, called \ili{Hasaitic} by scholars. This language, however, cannot be regarded as a form of \ili{Arabic}, and there are no pre-Islamic attestations of \ili{Arabic} from eastern Arabia yet (\citealt{Al-Jallad2018ANA}: 260--261). 

\section{Grammatical features arising from contact}
This section offers a contact-based explanation for two linguistic features found in \ili{Old} \ili{Arabic}: the {definite} {article}, and the realization of the feminine ending.

\subsection{Definite article}
It has long been established that the overt marking of {definiteness} in the \ili{Semitic} languages is a relatively late innovation (\citealt{HuehnergardRubin2011}: 260--261). All varieties of \ili{Arabic} today attest some form of the {definite} {article} -- most commonly variants of \textit{ʔal} but other forms exist as well, mainly in southwest Arabia, including \textit{am}, \textit{an}, and \textit{a-}, with gemination of the following consonant. In light of the {comparative} evidence, did \ili{Arabic} innovate this feature independently or was contact with other \ili{Semitic} languages involved?

The evidence suggests that the prefixed {article} *han- emerged in the central Levant sometime in the late second millennium BCE, after the diversification of Northwest \ili{Semitic} (\citealt{Tropper2001,Gzella2006,Pat-El2006}). It seems clear that by the early first millennium BCE, the {article} had spread across the southern Levant and to North Arabia, as it is found in Taymanitic, \ili{Thamudic} B, and \ili{Dadanitic}, as well as in the \ili{Old} \ili{Arabic} of the \ili{Safaitic} inscriptions. In the latter case, contact with \ili{Canaanite} is substantiated in the inscriptional record in the form of the Bāyir inscription (see §\ref{Canaanite} above). 

All of these languages, including the earliest \ili{Old} \ili{Arabic}, took over the form of the {article} unchanged; that is \textit{h-} with the assimilation of the /n/ before a consonant, the exception being \ili{Dadanitic}, which preserves the /n/ before laryngeal consonants, e.g. \textit{h-mlk} /ham-malk/ ‘the king’ vs. \textit{hn-ʔʕly} ‘the upper’ /han-ʔaʕlay/. We cannot, however, argue for the spread of the {definite} {article} to Proto-\ili{Arabic}. The original, article-less situation is attested in the inscriptions of Central Jordan stretching down to the Hismā, known as \ili{Hismaic} (\citealt{GrafZwettler2004}). These texts are in unambiguously \ili{Arabic} language, but they lack the {definite} {article}. The \textit{h-}morpheme exists, but it has a strong demonstrative force. Indeed, in a few \ili{Nabataean}--\ili{Hismaic} bilingual inscriptions, the {definite} {article} \textit{ʔl} of the \ili{Nabataean} component is rendered as zero in the \ili{Hismaic} text (\citealt{Hayajneh2009}). A minority of \ili{Safaitic} inscriptions also lack the {definite} {article} (\citealt{Al-Jallad2018ANA}), showing that it had not spread to all varieties of \ili{Arabic} even as late as the turn of the Era. Thus, like \ili{Hebrew} and \ili{Aramaic}, the earliest linguistic stages of \ili{Arabic} – and indeed Proto-\ili{Arabic} – lacked a fully grammaticalized {definite} {article}. Contact with \ili{Canaanite} then seems to be the likeliest explanation for the appearance of the \textit{h-}{article} in \ili{Old} \ili{Arabic}.

While the \textit{h-} {article} is the commonest form in \ili{Old} \ili{Arabic}, whence the \textit{ʔal} form? The \textit{ʔal} {article} appears to be a later development from the original \textit{han} {article}, through two irregular sound changes: \textit{h} > \textit{ʔ} and \textit{n} > \textit{l}.\footnote{The origins of the \textit{al}-{article} are discussed in detail in \citet{Al-Jalladforthcoming}.} The former is well attested in \ili{Arabic} (e.g. the {causative} ʔaCCaCa from haCCaCa), while the latter is not uncommon in loans (e.g. \textit{finǧān} vs. \textit{finǧāl} ‘cup’). The \textit{ʔal} {article} appears to have developed in the western dialects of \ili{Old} \ili{Arabic}, attested first in the Nile Delta (cf. the famous \textit{αλιλατ} \textit{al-ʔilat} ‘the goddess’ mentioned in \iai{Herodotus}, \textit{Histories} I: 131), and is the regular form of the {article} in the dialect of the Nabataeans, who were situated in ancient Edom, stretching south to the Ḥiǧāz. The \textit{ʔal}-{article} is attested sporadically at Dadān in the western Ḥiǧāz as well. Based on the inscriptional record, the \textit{ʔal}-{article} was a typical linguistic feature of settled, rather than nomadic groups, being attested most frequently in the \ili{Nabataean} dialect, and in cities and oases like Petra and Ḥegrā. The nomads used a variety of {definite} {article} forms. It was perhaps not until the rise of Islam, and the resulting {prestige} given to official \ili{Arabic} of the Umayyad state, that the \textit{ʔal} {article} began to dominate at the expense of other forms.

\subsection{The feminine ending}
In most modern \ili{Arabic} dialects, the feminine ending *-at is realized as \textit{-a(h)} in all contexts except the construct state, where it retains its original form \textit{-at}. In \ili{Classical} \ili{Arabic}, it is \textit{-at} in all situations, except for in utterance-final position, where it is realized as \textit{-ah}. The Quranic Consonantal Text resembles the situation in the modern dialects, as do the transitional Nabataeo-\ili{Arabic} and sixth-century \ili{Arabic} script inscriptions (\citealt{Nehmé2017}). Yet, if we go back further to the first century CE, it seems that varieties of \ili{Arabic} written in the \ili{Hismaic} and \ili{Safaitic} script never experienced the sound change \textit{-at} > \textit{-ah} in any position – the feminine ending is always written as 〈t〉. In the \ili{Arabic} of the Nabataeans, however, the sound change of \textit{-at} to \textit{-ah} seems to have operated as early as the third century BCE (\citealt{Al-Jallad2017early}: §5.2.1). 

The sound change \textit{-at} > \textit{-ah} is common in the \ili{Central Semitic} languages, but the distribution can vary. In \ili{Phoenician}, it applies to verbs but not nouns, while in \ili{Hebrew} it applies equally to nouns and verbs (\citealt{HuehnergardRubin2011}: 265--266). The most common \ili{Arabic} distribution matches \ili{Aramaic}: it applies to nouns but not verbs. I would suggest that, since this sound change is first attested in a dialect of \ili{Arabic} for which we have abundant evidence of heavy contact with \ili{Aramaic}, it is likely a contact-induced change (see also van Putten, this volume).\ia{van Putten, Marijn@van Putten, Marijn} Contact, or the lack thereof, may explain its absence in the ancient nomadic dialects, where, as we have seen above, there is little evidence for contact with \ili{Aramaic}. Thus, like the \textit{ʔal} {article}, the \textit{-at} to \textit{-ah} change would have been a typical feature of \ili{Arabic} dialects of settled groups in the pre-Islamic period. In later forms of \ili{Arabic}, the change spreads even to nomadic dialects, as we find it operational today across the Arabian Peninsula. Yet, the chronology of this {diffusion} is not quite clear. In an important study by \citet{vanPutten2017}, the Dosiri dialect of Kuwait appears to preserve the archaic situation where the feminine ending is realized as \textit{-at} in all positions.

\section{Conclusion}\largerpage
Contact must be factored into our understanding of {language change} for \ili{Arabic} at every attested stage. A summary of the facts above show that \ili{Arabic} was in most intense contact with \ili{Aramaic}, a situation that persisted for over a millennium prior to the rise of Islam, which may explain the high number of \ili{Aramaic} {loanwords} into \ili{Arabic}, and indeed some striking structural parallels, such as the distribution of the sound change \textit{-at} > \textit{-ah}. At the same time, there is very little evidence for contact with Sabaic (\ili{Old South Arabian}), a contact situation only represented by a small number of mixed texts.  This nicely matches the absence of \ili{South Arabian} influence on \ili{Old} \ili{Arabic} and later forms of the language, with the exception of those dialects spoken in southwest Arabia.

\section*{Further reading}
\begin{furtherreading}
\item \citet{Al-Jallad2018ANA} provides a comprehensive outline of the languages and scripts of pre-Islamic North Arabia.
\item \citet{Macdonald2003} gives a description of the multilingual environment of ancient Nabataea.
\item \citet{Nehmé2010} outlines the development of the \ili{Arabic} script based on the newest Nabataeo-\ili{Arabic} inscriptions from Northwest Arabia.
\item \citet{Stein2018} gives an outline of the use of \ili{Aramaic} in pre-Islamic Arabia.
\end{furtherreading}

\section*{Abbreviations}
\begin{tabularx}{.55\textwidth}{@{}lQ@{}}
\textsc{1, 2, 3} & 1st, 2nd, 3rd person \\
\textsc{acc} & accusative \\
\textsc{act} & active \\
\textsc{aor} & aorist \\
BCE & before Common Era \\
c. & century \\
CE & Common Era \\
\textsc{conj} & conjunction \\
\textsc{def} & {definite} \\
\textsc{dem} & demonstrative \\
\textsc{f} & feminine \\
\textsc{impf} & imperfect (prefix conjugation) \\
\end{tabularx}%
\begin{tabularx}{.5\textwidth}{@{}lQ@{}}
\textsc{inf} & {infinitive} \\
\textsc{m} & masculine \\
\textsc{nom} & nominative \\
\textsc{pass} & {passive} \\
\textsc{pl} & plural \\
\textsc{pn} & proper noun \\
\textsc{prep} & {preposition} \\
\textsc{prf} & perfect (suffix conjugation) \\
\textsc{ptcp} & {participle} \\
\textsc{rel} & {relative} \\
\textsc{sg} & singular \\
\textsc{voc} & {vocative} \\
\end{tabularx}%


\sloppy
\printbibliography[heading=subbibliography,notkeyword=this] 
\end{document}
