\documentclass[output=paper]{langsci/langscibook}
\ChapterDOI{10.5281/zenodo.3744517}
\author{Adam Benkato\affiliation{University of California, Berkeley}}
\title{Maghrebi Arabic}
\abstract{This chapter gives an overview of contact-induced changes in the Maghrebi dialect group in North Africa. It includes both a general summary of relevant research on the topic and a selection of case studies which exemplify contact-induced changes in the areas of phonology, morphology, syntax, and lexicon.}
\maketitle

\begin{document}

\section{The Maghrebi Arabic varieties}

In \ili{Arabic} dialectology, \textsc{Maghrebi} is generally considered to be one of the main dialect groups of \ili{Arabic}, denoting the dialects spoken in a region stretching from the Nile delta to Africa’s Atlantic coast -- in other words, the dialects of Mauritania, Morocco, Algeria, Tunisia, Libya, parts of western Egypt, and Malta. The main isogloss distinguishing Maghrebi dialects from non-Maghrebi dialects is the first person of the imperfect, as shown in Table \ref{tab:1:niktib} (cf. Lucas \& Čéplö, this volume).\footnote{More about the exact distribution of this isogloss can be found in \citet{Behnstedt2016niktib}.}\ia{Lucas, Christopher@Lucas, Christopher}\ia{Čéplö, Slavomír@Čéplö, Slavomír}


\begin{table}
\caption{First-person imperfect `write' in Maghrebi and non-Maghrebi Arabic}
\label{tab:1:niktib}
 \fittable{
 \begin{tabular}{lllll}
  \lsptoprule
         & \multicolumn{2}{c}{Non-Maghrebi}       & \multicolumn{2}{c}{Maghrebi} \\\cmidrule(lr){2-3}\cmidrule(lr){4-5}
         & \ili{Classical Arabic} & \ili{Baghdad} \ili{Arabic} & \ili{Casablanca} \ili{Arabic} & \ili{Maltese} \\
           \midrule
Singular & \textit{aktub}            & \textit{aktib}          & \textit{nəktəb}            & \textit{nikteb}  \\
Plural   & \textit{naktub}           & \textit{niktib}         & \textit{nkətbu}            & \textit{niktbu}  \\
\lspbottomrule
\end{tabular}}
\end{table}

This Maghrebi group of dialects is in turn traditionally held to consist of two subtypes: those spoken by sedentary populations in the old urban centers of North Africa, and those spoken by nomadic populations. The former of these, usually referred to as “pre-Hilali” (better: “first-layer”) would have originated with the earliest Arab communities established across North Africa ({\textasciitilde}7th--8th centuries CE) up to the Iberian Peninsula. The latter of these, usually referred to as “Hilali” (better: “second-layer”), is held to have originated with the westward migration of a large group of Bedouin tribes ({\textasciitilde}11th century CE) out of the Arabian Peninsula and into North Africa via Egypt. Their distribution is roughly as follows.\footnote{More will not be said about the subgroups of Maghrebi dialects that have been proposed. For more details about the features and distribution of Maghrebi dialects see \citet{Pereira2011}; for more detail on the complex distribution of varieties in Morocco see \citet{Heath2002}.} First-layer dialects exist in cities such as \ili{Tunis}, Kairouan, Mahdia, \ili{Sousse}, \ili{Sfax} (Tunisia), Jijel, \ili{Algiers}, Cherchell, \ili{Tlemcen} (Algeria), Tangier, Tetuan, southern Rif villages, Rabat, Fez, Taza, so-called “northern” dialects (Morocco), \ili{Maltese}, and formerly \ili{Andalusi} and {Sicilian}\il{Sicilian Arabic} dialects; most Judeo-\ili{Arabic} dialects formerly spoken in parts of North Africa are also part of this group. Second-layer dialects are spoken by populations of nearly all other regions, from western Egypt, through all urban and rural parts of Libya, to the remaining urban and rural parts of Algeria and Morocco. Though some differences between these two subtypes are clear (such as [q, ʔ, k] vs. [g] for *q), there have probably been varying levels of interdialectal mixture and contact since the eleventh century CE. In many cases, first-layer varieties of urban centers have been influenced by neighboring second-layer ones, leading to new dialects formed on the basis of inter-dialectal contact. It is important to note that North Africa is becoming increasingly urbanized and so not only is the traditional sedentary/nomadic distinction anachronistic (if it was ever completely accurate), but also that intensifying \isi{dialect contact} accompanying urbanization means that new ways of thinking about Maghrebi dialects are necessary. It is also possible to speak of the recent but ongoing koinéization of multiple local varieties into supralocal or even roughly national varieties—thus one can speak, in a general way, of “\ili{Libyan} \ili{Arabic}” or “\ili{Moroccan} \ili{Arabic}”. This chapter will not deal with contact between mutually intelligible varieties of a language although this is equally important for the understanding of both the history and present of Maghrebi dialects.\footnote{The emergence of new Maghrebi varieties resulting from migration and mixture is discussed in \citet{Pereira2007} and \citet{Gibson2002}, for example. The oft-cited distinction between urban and nomadic dialects is also problematized by the existence of the so-called rural or village dialects (though this is also a problematic ecolinguistic term), on which see \citet{Mion2015}. Dialect contact outside of the Maghreb is discussed by Cotter (this volume).}\ia{Cotter, William M.@Cotter, William M.}

\section{ Languages in contact}

Contact between \ili{Arabic} and other languages in North Africa began in the late seventh century CE, when Arab armies began to spread westward through North Africa, reaching the Iberian Peninsula by the early eighth century CE and founding or occupying settlements along the way. Their dialects would have come into contact with the languages spoken in coastal regions at that time, including varieties of \ili{Berber} and \ili{Late Latin}, and possibly even late forms of \ili{Punic} and \ili{Greek}. The numbers of \ili{Arabic} speakers moving into North Africa at the time of initial conquests were likely to have been quite small.\footnote{See Heath (this volume) for discussion of \ili{Late} \ili{Latin} influence in \ili{Moroccan} \ili{Arabic} dialects.}\ia{Heath, Jeffrey@Heath, Jeffrey} By the time of the migration of Bedouin groups beginning in the eleventh century, it is doubtful that languages other than \ili{Berber} and \ili{Arabic} survived in the Maghreb. The \isi{Arabization} of coastal hinterlands and the Sahara increased in pace after the eleventh century. \ili{Berber} varieties continue to be spoken natively by millions in Morocco and Algeria, and by smaller communities in Libya, Tunisia, Mauritania, and Egypt. Any changes in an \ili{Arabic} variety due to \ili{Berber} are almost certainly the result of \ili{Berber} speakers adopting \ili{Arabic} rather than \ili{Arabic} speakers adopting \ili{Berber} -- the sociolinguistic situation in North Africa is such that L1 \ili{Arabic} speakers rarely acquire \ili{Berber}.

  Beginning in the sixteenth century, most of North Africa came under the control of the Ottoman Empire and thus into contact with varieties of \ili{Turkish}, although the effect of \ili{Turkish} is essentially limited to cultural borrowings (see §\ref{lexicon}). The sociolinguistic conditions in which \ili{Turkish} was spoken in North Africa are poorly understood.

  The advent of \isi{colonialism} imposed different European languages on the region, most prominently \ili{French} (in Mauritania, Morocco, Algeria, and Tunisia), \ili{Italian} (in Libya), and \ili{Spanish} (in Morocco). \ili{Romance} words in dialects outside of Morocco may also derive from forms of \ili{Spanish} (via Andalusi refugees to North Africa in the 16th--17th centuries) or from the Mediterranean \ili{Lingua Franca}.\footnote{On the \ili{Lingua Franca} see Nolan (this volume).}\ia{Nolan, Joanna@Nolan, Joanna}

  The effects on Maghrebi \ili{Arabic} of contact with \ili{Chadic} (e.g. \ili{Hausa}) or \ili{Nilo-Saharan} (e.g. \ili{Songhay}, Tebu) languages is largely unstudied since in most cases data from the relevant \ili{Arabic} varieties is lacking. Yet some borrowings from these languages can be found in \ili{Arabic} and \ili{Berber} varieties throughout the region \citep{Souag2013lexical}.\footnote{See also \citet{Souag2016sahara} for an overview of contact in the Sahara region not limited to \ili{Arabic}.} Lastly, \ili{Hebrew} loans are present in most Jewish \ili{Arabic} dialects of North Africa \citep{Yoda2013}, though unfortunately these dialects hardly exist anymore.


  To restate these facts in Van Coetsem’s (\citeyear{VanCoetsem1988,VanCoetsem2000}) terms, there are two major contact situations at work in Maghrebi \ili{Arabic} in general, though the specifics will of course differ from variety to variety. The first is change in \ili{Arabic} driven by \isi{source-language} (\ili{Berber}) dominant speakers; this \isi{transfer} type is \textsc{imposition}. The second is change in \ili{Arabic} driven by \isi{recipient-language} (\ili{Arabic}) dominant speakers where the \isi{source language} is a European \isi{colonial} language; this \isi{transfer} type is called \textsc{borrowing}.\footnote{Another good illustration of the two \isi{transfer} types in the Van Coetsemian framework can be found in Winford (\citeyear{Winford2005}: 378–381).} So far, ``dominance'' describes linguistic dominance, that is, the fact that a speaker is more proficient in one of the languages involved in the contact situation. However, social dominance, referring to the social and political status of a language (\citealt{VanCoetsem1988}: 13), is also important, especially in North Africa.

\section{ Contact-induced changes in Maghrebi dialects}\largerpage

\subsection{Phonology}

Changes in Maghrebi \ili{Arabic} phonology due to contact with \ili{Berber} are difficult to prove. There are several cases, for example, where historical changes in \ili{Arabic} phonology may be argued to be the result of contact with \ili{Berber} \textit{or} the result of internal developments. These include the change of *ǧ to /ž/ in many varieties, or the emergence of phonemic /ẓ/ \citep{Souag2016sahara}. Another example, the pronunciation /ṭ/ in some first-layer varieties where most \ili{Arabic} varieties have /ð̣/, has also been explained as a result of \ili{Berber} influence, or as unclear directionality \citep[187]{Kossmann2013book}, while \citet{Al-Jallad2015Maghreb} argues that it is actually an archaism within \ili{Arabic}.

  The \isi{merger} in \ili{Arabic} of the vowels *{a} and *{i} (and even *{u}) to a single \isi{phoneme} /ǝ/ in some, especially first-layer, varieties, is often attributed to \ili{Berber} influence, as many \ili{Berber} varieties have only a single short vowel \isi{phoneme} /ǝ/. However Kossmann (\citeyear{Kossmann2013book}: 171–174) points out that \ili{Berber} also \isi{merged} older *ă and *ǝ to a single \isi{phoneme} /ǝ/ and that it cannot be proven that the reduction happened in \ili{Berber} before it happened in \ili{Arabic}. Hence, again the directionality of influence is difficult to show.

  Related to this development is also that many Maghrebi varieties disallow vowels in light syllables (often described as the deletion of short vowels in open syllables), such that *katab ‘he wrote’ > Tripoli \textit{kt\kern 0.75ptǝ\kern -0.75ptb} or *kitāb ‘book’ > Algerian \textit{ktāb}.\footnote{Since the short vowels \isi{merge} to schwa in many \ili{Moroccan} and Algerian varieties, \isi{vowel length} is no longer contrastive and it is common to transcribe e.g. \textit{ktab} rather than \textit{ktāb}.} Meanwhile, second-layer varieties often do allow vowels in light syllables (e.g. Benghazi \textit{kitab} ‘he wrote’, Douz \textit{m\textsuperscript{i}}\textit{šē} ‘he went’). While proto-\ili{Berber} and some modern varieties allow vowels in light syllables, most \ili{Berber} varieties of Algeria and Morocco do not. This is another example of a similar development wherein the directionality of influence is unclear (see \citealt{Souag2017syllable}: 62–65 for further discussion).

  In the \ili{Arabic} variety of \ili{Ghomara}, northwest Morocco, *d and *t are spirantized to /ð/ and /θ/ initially (*d only), postvocalically and finally \citep{Naciri-Azzouz2016}: e.g. \textit{māθ\kern 0.75ptǝ\kern -0.75ptθ} `she died' (*mātat), \textit{warθ} `inheritance' (although etymologically *warθ, dialects of the wider Jbala region of Morocco have no interdentals so *wart), \textit{ðāba} `now' (*dāba), \textit{ḫ\kern 0.75ptǝ\kern -0.75ptðma} `work' (*ḫidma), \textit{wāḥ\kern 0.75ptǝ\kern -0.75ptð} `one' (*wāḥid). Naciri-Azzouz points out that the distribution of spirantization is the same as in \ili{Ghomara} \ili{Berber}, a variety spoken by groups in the same region.\footnote{The \ili{Berber} variety of \ili{Ghomara} exhibits an extreme amount of influence from dialectal \ili{Arabic}, see Mourigh (\citeyear{Mourigh2015}). Kossmann (\citeyear[431]{Kossmann2013book}) writes that given the existence of parallel morphological systems for virtually all grammatical categories (nominal, adjectival, pronominal and verbal morphology) and a high \isi{loanword} count (more than 30\% of basic lexicon is \ili{Arabic}) it would be possible to call \ili{Ghomara} \ili{Berber} a mixed language.}

  New phonemes have been borrowed into Maghrebi varieties through contact with European languages: for example, /p/ and nasalized vowels in more recent \ili{French} loans in \ili{Tunisian} \ili{Arabic}, or /v, č, ǧ/ in \ili{Italian} loans in \ili{Libyan} \ili{Arabic} (\textit{grīǧū} ‘gray’ < \textit{grigio}).

  \subsection{Morphology}\label{Morphology}\largerpage

In the realm of morphology, changes in \ili{Arabic} varieties due to contact vary depending on whether the relationship between \ili{Arabic} and the contact language is substratal, adstratal, or superstratal.

  Morphological influence from \ili{Berber} on the \ili{Arabic} varieties of the northern Maghreb is not overly common.\footnote{Documentation of the varieties where such influence would be more expected, such as \ili{Arabic}-speaking towns in the otherwise \ili{Berber}-speaking Nafusa Mountains in Libya, is lacking.} In some places where \ili{Berber}--\ili{Arabic} \isi{bilingualism} is or was more common, contact has led to the borrowing of \ili{Berber} nouns into \ili{Arabic} together with their morphology, a phenomenon known as “\isi{parallel system borrowing}”.\footnote{For a closer look at \isi{parallel system borrowing} in the context of \ili{Arabic} and \ili{Berber} contact, see \citet{Kossmann2010}, mostly discussing the borrowing of \ili{Arabic} paradigms into \ili{Berber}.} In \ili{Ḥassāniyya}, for example, many nouns have been transferred together with their \isi{gender} and number marking.\footnote{See Taine-Cheikh (this volume).}\ia{Taine-Cheikh, Catherine@Taine-Cheikh, Catherine} In the dialect of Jijel, \ili{Berber} singular nouns are transferred together with their prefixes (\textit{āwtūl} ‘hare’, cf. \ili{Kabyle} \textit{āwtūl}); plurals are then formed in a way which resembles \ili{Berber} but is not identical (Jijel \textit{āsr\kern 0.75ptǝ\kern -0.75ptf}, \textit{ās\kern 0.75ptǝ\kern -0.75ptrf\kern 0.75ptǝ\kern -0.75ptn} ‘bush(es)’, cf. \ili{Kabyle} \ili{Berber} \textit{āsr\kern 0.75ptǝ\kern -0.75ptf}, \textit{īs\kern 0.75ptǝ\kern -0.75ptrf\kern 0.75ptǝ\kern -0.75ptn}); moreover, the prefix \textit{ā}{}- is also used with nouns of \ili{Arabic} origin (\textit{āfḫ\kern 0.75ptǝ\kern -0.75ptd} ‘thigh’, \ili{Arabic} *faḫað) \citep[302–318]{Marçais1956}.

  In Algeria and Morocco the circumfix \textit{tā-...-t}, which occurs on feminine nouns in \ili{Berber}, can derive abstract nouns (e.g. Jijel \textit{tāk\kern 0.75ptǝ\kern -0.75ptbūrt} ‘boasting’, \textit{tāw\kern 0.75ptǝ\kern -0.75ptḥḥūnt} ‘having labor pains’) and in \ili{Moroccan} \ili{Arabic} \textit{tā-...-t} is the regular way of forming nouns of professions and traits (e.g. \textit{tān\kern 0.75ptǝ\kern -0.75ptžžā{\R}t} ‘carpentry’) \citep{Kossmann2013chapter}.

  The verbal morphology of \ili{Arabic} dialects is much less affected by \ili{Berber}, though \ili{Ḥassāniyya} again provides an interesting example. It has a \isi{causative} prefix \textit{sä}- used with both inherited \ili{Arabic} verbs and borrowed \ili{Berber} verbs, and most likely to be borrowed from \ili{Berber} \isi{causative} forms in \textit{s-/š-} \citep{Taine-Cheikh2008chapter}.

  \ili{Turkish} influence on morphology is restricted to the suffix -\textit{ği}/\textit{{}-ži} (< \textit{{}-ci}) used to indicate professions and borrowed widely into \ili{Arabic} dialects in general. In Tunisia, its use has been extended to derive adjectives of quality from nouns (\textit{sukkārži} ‘drunkard’) and has also even been added to borrowed \ili{French} nouns (\textit{bankāži} ‘banker’ < \ili{French} \textit{banque}). As Manfredi (\citeyear[410]{Manfredi2018}) points out, the productivity of this borrowed \isi{derivational} morpheme constitutes one example of how \isi{recipient-language} agentivity can introduce morphological innovations via borrowing.

  \ili{French} (and other \ili{Romance}) verbs are also routinely borrowed into Maghrebi varieties. \citet{Talmoudi1986} discusses their integration into different forms of the verbal system of \ili{Tunisian} \ili{Arabic}, e.g. \textit{mannak} `to be absent' < \ili{French} \textit{manquer} (\citeyear[81--82]{Talmoudi1986}) or (\textit{t)rānā} `to train' < \ili{French} \textit{entrainer} (\citeyear[21--24]{Talmoudi1986}).

\subsection{ Syntax}

Syntax is often the least documented aspect of the grammar of Maghrebi \ili{Arabic} varieties and research on contact-induced changes in syntax is still in its infancy. Much attention has been devoted recently to explaining the rise of bipartite \isi{negation} in \ili{Arabic} and \ili{Berber}; in varieties of both languages the word for ‘thing’ (\ili{Arabic} \textit{šayʔ}, \ili{Berber} *ḱăra) has been grammaticalized postverbally in a marker of \isi{negation}:

\begin{multicols}{2}\raggedcolumns
\begin{exe}
\ex
\ili{Arabic} (Benghazi)\\
\gll \textbf{mā}-šift-hā-\textbf{š}\\
     \textbf{\textsc{neg}}-see.\textsc{prf}.\textsc{1sg}-\textsc{3sg.f}-\textbf{\textsc{neg}}\\
\glt `I didn’t see her'.\columnbreak
\ex  \ili{Berber} (Tarifit)\\
\gll \textbf{wā}  t-ẓṛiɣ  \textbf{ša}\\
     \textbf{\textsc{neg}}  \textsc{3sg.f}-see.\textsc{prf}.\textsc{1sg}  \textbf{\textsc{neg}}\\
\glt `I didn’t see her'.
\end{exe}
\end{multicols}

Although some accounts give no attention to \ili{Berber}, while others attribute the \ili{Arabic} development solely to \ili{Berber}, the development in both languages in the same contexts is probably not a coincidence, though there is no current consensus on the direction of \isi{transfer} \textendash \,see Lucas (this volume) for discussion.\footnote{See \citet{Lucas2007,Lucas2010,Lucas2018} and \citet{Souag2018thing} for further discussion of the \isi{grammaticalization} of `thing’ for indefinite quantification and polar question marking in \ili{Arabic} and \ili{Berber}. \citet[324–334]{Kossmann2013book} surveys the situation in the \ili{Berber} languages. See \citet{Lafkioui2013reinventing} for an overview of \isi{negation} in especially \ili{Moroccan} \ili{Arabic}, as well as discussion of a variety of \ili{Moroccan} \ili{Arabic} which features the discontinuous morpheme \textit{mā}{}- ... -\textit{bū}, where the latter part has been borrowed from Tarifit.}\ia{Lucas, Christopher@Lucas, Christopher} However, it must be noted that not all \ili{Berber} varieties have double \isi{negation} (e.g. \ili{Tashelhiyt} \textit{ur} \textit{nniγ} \textit{ak} ‘I didn’t tell you’ where the only negator is \textit{ur}).

  In another area, recent work on the variety of \ili{Tunis} has yielded interesting conclusions: while possessives with \ili{French} nouns are overwhemingly analytic (\textit{l-prononciation} \textit{mtēʕ-ha} ‘her pronunciation’) and those with \ili{Arabic} nouns are almost as overwhelmingly synthetic (\textit{nuṭq-u} ‘his pronunciation’), the frequent occurence of \ili{French} loan nouns may be triggering an increase in the overall \isi{frequency} of analytical possessives over syntactic ones, including those with \ili{Arabic} nouns \citep{Sayahi2015}.

  The remainder of this section will discuss one particularly interesting case: the first-layer dialect of Jijel, a city in eastern Algeria. At the time of its description \citep{Marçais1956}, it showed little influence from second-layer varieties, but displayed wide-ranging influence from \ili{Berber} in multiple domains. In a recent \isi{article}, \citet{Kossmann2014} has demonstrated how a \ili{Berber} marker of non-verbal predication was adopted into the \ili{Arabic} dialect of Jijel as a focus marker. Here I will briefly summarize Kossmann’s arguments with a few examples. In the Jijel dialect, as described by Marçais and reanalyzed by Kossmann, a morpheme \textit{d} occurs in the following syntactic contexts (examples \xxref{nonverbal}{left focalizations} are all from \citealt[129\textendash131]{Kossmann2014}, who retranscribes from Marçais’ texts): before non-verbal predicates (\ref{nonverbal}), in clefts with a noun/pronoun in the cleft (\ref{clefts}), in secondary predication with a specific noun (\ref{secondary pred}), as a marker of subject (or object) focus (\ref{focus}), and in left-moved focalizations (\ref{left focalizations}).

\begin{exe}
\ex\label{nonverbal}

\gll l-lila  d-ǝl-ʕid\\
     \textsc{def}-night  \textsc{d}-\textsc{def}-feast\\
\glt ‘Tonight is the feast.’
\ex\label{clefts}

\gll d-hum  ǝddǝ  šraw-ǝh   qbǝl-ma  nǝzdad\\
     \textsc{d}-\textsc{3pl.m} \textsc{rel}  buy.\textsc{prf.3pl.m}-\textsc{3sg.m}  before-\textsc{comp} be.born.\textsc{impf.1sg}\\
\glt ‘It is them who bought it before I was born.’
\ex\label{secondary pred}

\gll ṛa-na  nqǝṭṭʕu-č  d  ǝṭ-ṭraf\\
     \textsc{prst}-\textsc{1pl}  cut.\textsc{impf.1pl-sg}  \textsc{d}  \textsc{def}-pieces\\
\glt ‘We will cut you (into) pieces.’
\ex\label{focus}

\gll tkǝṣṣṛǝt  d  l-idura\\
     break.\textsc{prf.3sg.f}  \textsc{d}  \textsc{def}-bowl\\
\glt ‘The bowl has broken.’
\ex\label{left focalizations}

\gll qalu  d  ǝṛ-ṛbiʕ  dǝḫlǝt\\
     say.\textsc{prf.3pl}  \textsc{d}  \textsc{def}-spring  enter.\textsc{prf.3sg.f}\\
\glt ‘They say spring has come.’
\end{exe}

Although previous analyses attempted to explain \textit{d} within \ili{Arabic}, Kossmann notes that an \ili{Arabic}-internal \isi{derivation} of \textit{d} is impossible. However, \ili{Kabyle}, the \ili{Berber} language neighboring the Jijel area has an element \textit{d} (realized [ð] due to spirantization in \ili{Kabyle}) which is used in (pro)nominal predicates (\ref{Berber pred}), cleft constructions (\ref{Berber cleft}), and secondary predication when non-verbal (\ref{Berber secondary}). Examples \xxref{Berber pred}{Berber secondary} are all \ili{Kabyle} \ili{Berber}, taken from Kossmann (\citeyear[135\textendash136]{Kossmann2014}). This element \textit{d} is attested in \ili{Berber} more widely, too, and is likely reconstructible to older stages of the language.\largerpage[2]

\begin{exe}
\ex\label{Berber pred}
\gll d-yǝlli-m\\
     \textsc{d}-daughter-\textsc{2sg.f}\\
\glt ‘Is it your daughter?’
\ex\label{Berber cleft}
\gll d-ay-ǝn  i  d-tǝnna  abrid  amǝnzu\\
     \textsc{d}-this-\textsc{deict}  \textsc{rel}  hither-say.\textsc{3sg.f}  road  first\\
\glt ‘This is what she said the first time.’
\ex\label{Berber secondary}

\gll ad  nǝǧʕǝl  iman  nn-ǝɣ  d-inǝbgiwǝn  n  ṛǝbbi\\
     \textsc{mod} make.\textsc{1pl} self \textsc{gen-1pl}  \textsc{d}-guests \textsc{gen}  lord\\
\glt ‘We shall pretend to be beggars (lit. guests of God).’
\end{exe}

Thus \ili{Berber} \textit{d} is the best candidate for the origin of Jijel \ili{Arabic} \textit{d}, though its usage in (\ili{Kabyle}) \ili{Berber} (where it is primarily a marker of syntactic organization) differs from that of Jijel \ili{Arabic} (where it is mainly a marker of information structure). In a simplified scenario with a \ili{Berber} variety as \isi{source language} and Jijel \ili{Arabic} as recipient, \textit{d} would likely have been imposed into Jijel \ili{Arabic} with its exact \ili{Berber} functions. As Kossmann notes, though, speech communities are full of variation and language contact is a “negotiation between the \isi{frequency} of non-native speech and the \isi{prestige} of the native way of speaking” \citep[138]{Kossmann2014}. Kossmann thus proposes a scenario in which larger groups of \ili{Berber} speakers switched to a variety of Jijel \ili{Arabic} and began imposing their own \textit{d}; the native Jijel \ili{Arabic} speakers, fewer in number, began adopting \textit{d} but understood it differently and interpreted it as a focus marker, introducing it into new contexts; eventually the variety of Jijel \ili{Arabic} with \textit{d} in all these functions became nativized. Per Kossmann (\citeyear[138\textendash139]{Kossmann2014}), two processes would have taken place: the \isi{transfer} of a \isi{source-language} feature by speakers dominant in the \isi{source language} (\ili{Berber}), followed by the borrowing of this feature by speakers dominant in the \isi{recipient language} (\ili{Arabic}), and its eventual regularization in that variety. Jijel \ili{Arabic} is an excellent example of what may happen when large numbers of \ili{Berber} speakers switch to \ili{Arabic}.

\subsection{Lexicon}\label{lexicon}\largerpage[2]

Much work on contact and Maghrebi \ili{Arabic} has focused on \isi{loanwords}, the most salient effects of borrowing, with secondary attention to their phonological or morphological adaptation. The concept of social dominance has particular relevance for borrowing: in the North African context, the \isi{colonial} languages, and especially \ili{French}, have high social status for both \ili{Arabic} and \ili{Berber} native speakers. One also must modify the idea of linguistic dominance to include those who acquire two languages natively (2L1 speakers; see \citealt[525]{Lucas2015}), definitely the case for certain speakers of \ili{Berber} and \ili{Arabic} in North Africa.

  Unsurprisingly, we see firstly that the majority of words borrowed into \ili{Arabic} varieties are nouns, and secondly that the lexical domains into which these borrowings fall are often restricted. Social dominance seems to play a role in the nature of the nouns borrowed.

  \ili{Berber} loans are found in most Maghrebi \ili{Arabic} varieties, though their number ranges from only a handful of words in the east to many more in the west (cf. §\ref{Morphology} above). Almost all Maghrebi varieties have borrowed the words \textit{ž(i){\R}āna} ‘frog’ and \textit{fakrūna} ‘turtle’, while in some oases \ili{Berber} influence in agricultural terminology can be seen. Again, the documentation of the relevant varieties is often insufficient.

Several studies on contact between Maghrebi \ili{Arabic} varieties and European languages exist. For \ili{French} in Morocco, \citet{Heath1989} argues that \isi{code-switching} and borrowing are essentially the same in a bilingual community which has established borrowing routines.\footnote{\citet[87]{VanCoetsem1988} notes that for bilingual speakers who have a balance in linguistic dominance between the two languages, the separation between the two \isi{transfer} types (borrowing and \isi{imposition}) will be weaker. Hence, either of the two dominant languages can serve as the \isi{recipient language} in \isi{code-switching} behavior. \citeauthor{Winford2005} (2005, esp. 394–396), expanding on Van Coetsem’s framework, points out that \isi{code-switching} is inherently linked to the borrowing \isi{transfer} type. In the Maghreb, this scenario is possible for \ili{Berber}\textendash \ili{Arabic} bilinguals as well as for some \ili{French}\textendash \ili{Arabic} bilinguals. See \citet{Ziamari2008} for an insightful and more recent analysis of \ili{Moroccan} \ili{Arabic} in contact with \ili{French} using a ``matrix language frame" analysis.} For \ili{French} in Tunisia, \citet{Talmoudi1986} analyzes the phonological and morphological adaptation of \ili{French} verbs into \ili{Arabic}. \citet[127–151]{Sayahi2014} gives a broader view of lexical borrowing in diglossic or bilingual communities, \isi{focusing} on \ili{French} in Tunisia and \ili{Spanish} in Morocco. \citet{Vicente2005} studies \ili{Arabic}-\ili{Spanish} \isi{code-switching} in Ceuta, a \ili{Spanish} enclave in northern Morocco. \ili{Italian} in Tunisia is studied briefly by \citet{Cifoletti1994}. Studies of contact with \ili{Turkish} are limited to discussion of lexical borrowing: on Morocco see \citet{Procházka2012}; on Algeria, see \citet{BenCheneb1922}, to be read with the review by G. S. Colin \citep[21–30]{Colin1999}.

The remainder of this section will consider the influence of \ili{Turkish} and \ili{Italian} on Libyan \ili{Arabic} (henceforth \ili{LA}), a hitherto under-researched topic. Uniquely in the Maghreb region there is at present no superstratum language spoken widely by \ili{Arabic} speakers in Libya, while there are also fewer \ili{Berber} speakers than in Algeria or Morocco. As far as documented varieties of \ili{LA} (Tripoli and Benghazi) go, contact situations are historical and not active.

  There seems to be an impression among dialectologists that \ili{LA} varieties have the largest number of \ili{Turkish} loans, though there is not a published basis for this. Procházka (\citeyear[191]{Procházka2005}) suggests that the number of (\ili{Ottoman}) \ili{Turkish} loans in a given \ili{Arabic} dialect is proportional to the length and intensity of Ottoman rule. By this criterion Libya should have quite a few, as the regions now constituting Libya were under control of the Ottoman Empire from 1551 to 1911, but Procházka estimates that the dialect would show 200 to 500 surviving loans, fewer than in other dialects. Another important factor is likely to be that Libya’s population was very small during the period of Ottoman rule so that the long-term presence of even a few thousand \ili{Turkish} speakers could have had a significant effect. However, I cannot yet offer a statistical analysis of \ili{Turkish} words in \ili{LA}.\footnote{The only study dedicated to \ili{Turkish} loans in \ili{LA} is \citet{Türkmen1988}, who lists 90 words. However, the basis for his wordlist seems unclear and  several items are either spurious or incorrect (e.g. there is no word \textit{kabak} ‘pumpkin’ in Benghazi \ili{Arabic} but there is \textit{bkaywa} ‘pumpkin’, identified by \citet{Souag2013lexical} as a loan from \ili{Hausa}). \ili{Turkish} words in \ili{LA} cited here are from the Benghazi variety, author’s data.} It is clear so far, though, that the effects of \ili{Turkish} on \ili{LA} can mainly be seen in the lexicon and, in my data, almost entirely in nouns. In terms of their semantic domains, Procházka (\citeyear[192]{Procházka2005}) points out that the majority of \ili{Turkish} loans in \ili{Arabic} dialects in general fall into three categories, roughly described as: private life; law, government, social classes; and army, war. By far the majority of surviving loans would belong to the first of these classes (such as \textit{šīšma} ‘tap’ < \textit{çeşme}, \textit{dizdān} ‘wallet’ < \textit{cüzdan}), or the second (such as \textit{fayramān} ‘order’ < \textit{ferman}, \textit{ḥafð̣a} ‘week’ < \textit{hafte}) while I suspect that words from the third class are increasingly rarer. Outside of these, only a few words other than nouns seem to be present, such as \textit{du\.gri} ‘straight ahead’ and \textit{balki} ‘maybe’. The length of time since \ili{Turkish} was last actively spoken in Libya no doubt means that the number of \ili{Turkish} loans actively used by speakers has been decreasing.

  \ili{LA} is unique among Maghrebi varieties in having had \ili{Italian} as the main European contact language. \ili{Italian} had a presence in what is now Libya from the 1800s, but this was mainly limited to the Tripolitanian Jewish community and wealthy merchant families. The Italian colonization of Libya officially began in 1911; though the majority of the region was not brought under Italian control until the early 1930s, large numbers of Italian colonists had begun to settle in Libya in the 1920s. From that period until 1970, when the remaining Italian citizens were expelled from the country, Italians made up 15\% or more of the population and the language was in widespread use. From the 1970s on, \ili{Italian} was scarcely used in Libya, and the teaching of foreign languages was banned in 1984, not to return again until 2005.\footnote{For more information on the return of \ili{Italian} instruction to Libya, see \citet{Danna2018phonetic}.} Many of the postwar generation spoke (and still speak) \ili{Italian}, though they rarely use it anymore, but few Libyans of younger generations do. The 1920s to the 1970s can thus be regarded as the main period of contact between \ili{LA} and \ili{Italian}.\footnote{The \ili{Italian} words in Yoda’s (\citeyear{Yoda2005}) study of Tripoli Judeo-\ili{Arabic} need to be seen slightly differently than \ili{Italian} words in non-Jewish dialects, owing to a different history of contact between the Tripolitanian Jewish community and Italy.} However, the concentration of Italians differed from region to region and thus may have influenced local varieties differently. The primary study devoted to analyzing \ili{Italian} loans in \ili{LA} is that of \citet{Abdu1988} who, \isi{focusing} on the variety of Tripoli, draws up a list of nearly 700 items (a few are misidentified), of which about 50\% were recognized by a majority of those surveyed. Some 93\% of these are nouns and the remainder are practically all derived from nouns or adjectives, such as \textit{bwōno} ‘well done!’ < \textit{buono} ‘good’ or \textit{faryaz} ‘to go out of order’ < \ili{Italian} \textit{fuori} \textit{uso}.\footnote{See Abdu (\citeyear[271]{Abdu1988}) and \citet{Danna2018phonetic}. Some denominal verbs are cited by Abdu, but more extensive data might reveal several more in use: for example in the variety of Benghazi, I identified \textit{fu{\R}an}  ‘to brake (intransitive)’ < \textit{frayno} ‘brake’ < \ili{Italian} \textit{freno}, not listed by Abdu.} Abdu’s study (\citeyear[248–268]{Abdu1988}) groups \ili{Italian} loans into some 22 semantic categories, the vast majority of which relate to material culture. Examples of these from the Benghazi variety are \textit{byāmbu} ‘lead’ < \textit{piombo}, \textit{bōskō} ‘zoo’ < \textit{bosco} `wood', \textit{furkayta} ‘fork’ < \textit{forchetta}, \textit{maršabīdi} ‘sidewalk’ < \textit{marciapiede} (author’s data).

  As \citet{Danna2018phonetic} points out, the adaptation of \ili{Italian} words to \ili{LA} phonology varies: new phonemes, particularly  [v] and [č], sometimes occur but are sometimes adapted to the dialects’ pre-existing phonologies, an indication of “subsidiary phonological borrowing” (\citealt{VanCoetsem1988}: 98). Of course, the maintenance of new phonemes often depends on speakers continuing to have access to the \isi{source language}; as this is no longer the case in Libya, \ili{Italian} borrowings in \ili{LA} are traversing a different trajectory than \ili{French} borrowings in other Maghrebi varieties, where only the oldest borrowings have been phonologically integrated.

  The overwhelming majority of surviving \ili{Turkish} and \ili{Italian} loans in \ili{LA} are nouns, widely acknowledged to be the most easily-borrowed word class due to their being the least disruptive of the \isi{recipient language}’s argument structure \citep{Myers-Scotton2002}, though a few verbs derived dialect-internally do exist. Furthermore, almost all the nouns are cultural borrowings — “lexical content-words that denote an object or concept hitherto unfamiliar to the receiving society, terminology related to institutions that are the property of the neighboring [or colonizing] culture, and so on” \citep[210]{Matras2011universals}. Cultural borrowings are to be differentiated from core borrowings, the latter being words that more or less duplicate already existing words and which originate in a bilingual \isi{code-switching} context. These facts lead us to conclude that \ili{Turkish} and \ili{Italian} borrowings in \ili{Libyan} varieties would be from (1) to (2) on the borrowing scale proposed by Thomason \& Kaufman (\citeyear[78–83]{ThomasonKaufman1988}). While (1) of the scale involves lexical borrowing of non-basic vocabulary only, (2) includes some function words as well as new phones appearing in those \isi{loanwords}. Colonial language contact situations are typically ones of \isi{recipient-language} agentivity, as the number of indigenous people learning the \isi{colonial} language is many times more than the number of colonizers learning indigenous languages. Without a longer period of sustained \isi{bilingualism} or language education motivated by continued contact with the metropole, \ili{Italian} has affected \ili{LA} to a much smaller degree than \ili{French} has Libya's Maghrebi neighbours.

\section{Conclusion}

The general parameters of the Maghrebi linguistic landscape and contact situations are relatively well understood. However, more documentation of Maghrebi varieties is needed, and more specifically, of those where contact situations -- especially with \ili{Berber} -- may have existed. Additionally, further research into the sociolinguistic factors affecting \isi{bilingualism} in \ili{Berber} and \ili{Arabic}, or regarding the intersection of \isi{diglossia} with \isi{bilingualism}, will no doubt add to our knowledge of the parameters of contact-induced change more generally. Finally, inter-dialectal contact as well as the gradual rise of national or at least supra-local varieties certainly merits continuing attention.

\section*{Further reading}
\begin{furtherreading}
\item \citet{Kossmann2013book} is the most extensive study so far of \ili{Berber}\textendash \ili{Arabic} contact, written from a Berberological point of view but important for Arabists.
\item \citet{Sayahi2014} studies the intersection of dialects, \ili{Standard} \ili{Arabic}, \ili{French} and \ili{Spanish} in Tunisia and Morocco.
\item \citet{Souag2016sahara} summarizes contact in the Saharan region among \ili{Arabic}, \ili{Berber}, \ili{Hausa}, \ili{Songhay}, \ili{Chadic}, etc.
\item \citet{Ziamari2008} is the most up-to-date work discussing \isi{code-switching} and borrowing strategies between \ili{Moroccan} \ili{Arabic} and \ili{French}.
\end{furtherreading}

\section*{Acknowledgements}
The research for this article was supported by a grant from the Alexander von Humboldt-Stiftung.


\section*{Abbreviations}
\setlength{\columnsep}{30pt}
\begin{multicols}{2}
\begin{tabbing}
\textsc{ipfv} \hspace{1em} \= before common era\kill
\textsc{1, 2, 3} \> 1st, 2nd, 3rd person \\
\textsc{comp} \> {complementizer} \\
\textsc{def} \> {definite} \\
\textsc{deict} \> deictic \\
\textsc{f} \> feminine \\
\textsc{gen} \> genitive \\
\textsc{impf} \> imperfect (prefix conjugation) \\
LA \> Libyan Arabic \\
\textsc{m} \> masculine \\
\textsc{mod} \> {modal} \\
\textsc{neg} \> negative \\
\textsc{pl} \> plural \\
\textsc{prf} \> perfect (suffix conjugation) \\
\textsc{prst} \> presentative \\
\textsc{rel} \> {relative} \\
\textsc{sg} \> singular
\end{tabbing}
\end{multicols}


\sloppy
\printbibliography[heading=subbibliography,notkeyword=this]
\end{document}
