\documentclass[output=paper]{langsci/langscibook} 
\ChapterDOI{10.5281/zenodo.3744557}
\author{Stefano Manfredi\affiliation{CNRS, SeDyL}}
\title{Contact and calquing}
\abstract{The notion of calquing refers to the transfer of semantic and syntactic patterns deprived of morphophonological matter. By providing examples of lexical and grammatical calques in a number of Arabic dialects and Arabic-based contact languages, this chapter identifies ways to relate the process of calquing to Van Coetsem’s psycholinguistic principle of language dominance.}
\IfFileExists{../localcommands.tex}{
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 
  

\section{Introduction}

In its simplest definition, \textsc{calquing} is a type of contact-induced change in which a word or sentence structure is transferred without actual morphemes \citep[260]{Thomason2001}. Calques are sometimes called loan translations, as they typically represent a word-by-word (or morpheme-by-morpheme) translation of a lexeme or a sentence from another language. Heath (\citeyear[367]{Heath1984}) labels this process “pattern \isi{transfer}” and distinguishes it from “\isi{matter borrowing}” which is instead linked to the integration of morphophonological material. \cite{Ross2007}, for his part, points out that that \isi{calquing} can also have important grammatical effects, and he considers it a necessary precondition for contact-induced morphosyntactic restructuring (what Ross calls “metatypy”).  

Broadly speaking, we can distinguish two types of \isi{calquing}: lexical \isi{calquing}, which entails the \isi{transfer} of semantic properties of lexical items, and grammatical \isi{calquing}, which instead implies the \isi{transfer} of the functional properties of morphemes and syntactic constructions. Using Ross’s words (\citeyear[126]{Ross2007}), lexical \isi{calquing} consists of remodelling lexical “ways of saying things”, whereas grammatical \isi{calquing} consists of remodelling grammatical “ways of saying things”. Despite this fundamental difference, lexical and grammatical \isi{calquing} share a single cause: bilingual speakers’ need to express the same meaning in two languages (\citealt[32]{Sasse1992}). This also means that everything that expresses meaning (i.e. morphemes, lexemes, and constructions) can, in principle, be a source of \isi{calquing}. 

Focusing mainly on the \isi{transfer} of linguistic matter, Van Coetsem (\citeyear{VanCoetsem1988}) does not overtly mention the possibility of \isi{transfer} of lexical and grammatical meanings through \isi{calquing}. The present chapter thus aims at relating contact-induced changes produced by \isi{calquing} to the principle of language dominance as postulated by Van Coetsem. 


 \section{Contact-induced changes and calquing}


 \subsection{Lexical calquing}\label{sec:lex}

According to Haspelmath (\citeyear[39]{Haspelmath2009}), a lexical \isi{calque} is a lexical unit that was created by an item-by-item translation of the source unit. This type of contact-induced change occurs as bilingual speakers reorganise the lexicon of one of their languages to match the semantic organisation of the other (\citealt[132]{Ross2007}). Adopting the psycholinguistic standpoint of language dominance, Winford (\citeyear[345]{Winford2003}) regards lexical \isi{calquing} as a subtype of lexical borrowing, which is a combination of \isi{recipient language} (\isi{RL}) lexemes in imitation of \isi{source language} (\isi{SL}) semantic patterns. In contrast, I will show that, though lexical \isi{calquing} can easily be triggered by RL-dominant speakers, it can also be a product of \isi{imposition} via \isi{SL} agentivity. In order to do this, I will mainly focus on \isi{calquing} of \isi{compound nouns}. A compound noun is here defined as a series of two or more lexemes, which is semantically conceived as a single unit. Each component of the compound can function as a lexeme independent from the other(s), and may show some phonological and/or morphological constraints within the compound when compared to its isolated syntactic usage \citep{Bauer2001}. Against this backdrop, I will specifically discuss noun–noun \isi{compounds}, as they represent the more uniform phenomenon of nominal compounding in the world’s languages \cite{Pepperforthcoming}. As we will see, the \isi{transfer} of the semantics of \isi{compound nouns} does not imply any morphosyntactic change in \ili{Arabic}, as calqued \isi{compounds} are typically adjusted to fit \isi{RL} morphosyntactic patterns.  

Generally speaking, lexical \isi{calquing} through borrowing can occur in indirect contact situations characterized by a very low degree of \isi{bilingualism}. This is because \isi{RL} monolinguals can also be agents of lexical borrowing (\citealt[10]{VanCoetsem1988}). Typical instances of lexical \isi{calquing} via \isi{RL} agentivity are related to the \isi{transfer} of the semantic patterns of \ili{English} \isi{compound nouns} in modern \ili{Arabic} dialects. This kind of \isi{transfer} is linked to the expansion of the non-core \ili{Arabic} lexicon for expressing previously unknown concepts. A prime example is the \ili{English} \isi{calque} \textit{lōḥit} \textit{il-mafatīḥ} `keyboard' (lit. ‘the board of keys’) in \ili{Egyptian Arabic} \citep[9]{WilmsenWoidich2011}. Here, it can be clearly seen that the \isi{transfer} of the semantic organization of the \isi{SL} compound noun does not affect the morphosyntax of the \isi{RL}, as the \isi{word order} of the \ili{English} nominal juxtaposition is reversed to fit the \ili{Arabic} construct state.  

Lexical \isi{calquing} can also take place in prolonged contact situations, as testified by numerous \ili{Italian} \isi{compounds} in \ili{Maltese}. A singular case of mixed \isi{calquing} is that of \textit{wi\.c\.c} \textit{tost} ‘shameless person’ (lit. ‘tough face’) deriving from the \ili{Italian} compound \textit{faccia} \textit{tosta} ‘shameless person’ (lit. ‘tough face’) \citep{Aquilina1987}. On the one hand, the first lexical item of the compound presents an \ili{Arabic} phono\-logical form while expressing semantic properties associated with the lexeme ‘face’ in \ili{Italian}. On the other hand, the second lexical item clearly results from the borrowing of the adjective \textit{tosto} ‘hard, tough’ retaining both the \ili{Italian} phono\-logical matter and semantic properties. The mixed nature of this compound brings to the fore the complementary relationship between \isi{RL} and \isi{SL} agentivity and shows that it is not always a trivial matter to distinguish between \isi{imposition} and borrowing. However, \ili{Maltese} also gives evidence of genitive \isi{compounds} in which both lexical components have an \ili{Arabic} phonological form coupled with \ili{Italian} semantic properties. This is the case of the \isi{compound nouns} \textit{saba'} \textit{ta'} \textit{sieq} ‘toe’ calqued on the \ili{Italian} \textit{dito} \textit{del} \textit{piede} ‘toe’ \citep{Pepperforthcoming}. Such instances of lexical \isi{calquing} clearly mirror semantic properties of \isi{SL} lexemes and they most plausibly result from borrowing via \isi{RL} agentivity (cf. Lucas \& Ćeplö, this volume).\ia{Lucas, Christopher@Lucas, Christopher}\ia{Čéplö, Slavomír@Čéplö, Slavomír} 

\ili{Ḥassāniyya} \ili{Arabic}, for its part, presents many \isi{compound nouns} that are traditionally analysed in terms of substratum interference from \ili{Zenaga} \ili{Berber} \citep{Taine-Cheikh2008chapter,Taine-Cheikh2012}. Also in this case, the \isi{transfer} of the semantic properties of the \isi{SL} does not produce any morphosyntactic change in \ili{Arabic}, as we can see in the pairs of examples in \REF{ex:crow} and \REF{ex:ankle}.

\ea\label{ex:crow}
\ea{\ili{Ḥassāniyya} \ili{Arabic} \citep[126]{Taine-Cheikh2008chapter}}\\
\gll   k{\R}aʕ lә-ɣ{\R}ab  \\
       foot \textsc{def}-crow\\
\glt   `aquatic herbaceous plant' (lit.  ‘crow’s foot’)

\ex{\ili{Zenaga} \ili{Berber} \citep[126]{Taine-Cheikh2008chapter}}\\
\gll   að̣aʔ{\R} әn tayyaḷ \\
       foot \textsc{gen} crow \\
\glt   `aquatic herbaceous plant' (lit. ‘crow’s foot’)
\z
\ex \label{ex:ankle}
\ea{\ili{Ḥassāniyya} \ili{Arabic} \citep[126]{Taine-Cheikh2008chapter}}\\
\gll   sayllāl lә-ʕrāgib \\
       ripper \textsc{def}-ankle.\textsc{pl}\\
\glt   `honey badger' (lit. ‘ripper of ankles’)

\ex{\ili{Zenaga} \ili{Berber} \citep[126]{Taine-Cheikh2008chapter}}\\
\gll   amәssäf әn ūržan \\
       ripper \textsc{gen} ankle.\textsc{pl} \\
\glt   `honey badger' (lit. ‘ripper of ankles’)
\z
\z

Taine-Cheikh (\citeyear[126]{Taine-Cheikh2008chapter}) stresses that it is somewhat difficult to trace back the origin of these \isi{compounds}. Accordingly, she speaks of a process of \isi{convergence} between the two languages, rather than determining the direction of the semantic \isi{transfer}. However, it should be observed that these \isi{compound nouns} are not attested in other spoken varieties of \ili{Arabic}. Furthermore, since at least the mid-twentieth century, Berbers in Mauritania have been gradually loosing competence in \ili{Zenaga}, in favour of \ili{Arabic} \citep[100]{Taine-Cheikh2012}, while \ili{Zenaga} is rarely acquired as second language by \ili{Ḥassāniyya} \ili{Arabic} speakers. In such a context, the most probable agents of contact-induced change were former \ili{Berber}-dominant speakers who gradually shifted to \ili{Arabic}. Thus, it seems plausible that the \isi{transfer} of the semantic properties of \ili{Zenaga} \isi{compounds} has been achieved through \isi{imposition}, rather than through borrowing.

\ili{Nigerian} \ili{Arabic} also shows interesting instances of lexical \isi{calquing} as a consequence of a longstanding contact with \ili{Kanuri}, a \ili{Nilo-Saharan} language widely spoken in the \isi{Lake Chad} area. Owens (\citeyear{Owens2015,Owens2016idioms}) gives evidence of the \isi{transfer} of the semantic properties of numerous \isi{compound nouns} including the lexeme \textit{{\R}ās} ‘head’. Similar to the previous instances of compound \isi{calquing}, the integration of \ili{Kanuri} semantic patterns does not affect the \ili{Arabic} morphosyntax, as we can see in the  pairs of examples in \REF{ex:roof} and \REF{ex:tassel}. 

\ea\label{ex:roof}
\ea{\ili{Nigerian} \ili{Arabic} \citep[69]{Owens2016idioms}}\\
\gll   {\R}ās al-bēt \\
       head \textsc{def}-house\\
\glt   `roof' (lit. ‘head of house’)

\ex{\ili{Kanuri} \citep[69]{Owens2016idioms}}\\
\gll   kǝla fato-be   \\
       head house-\textsc{gen}\\
\glt   `roof' (lit. ‘head of house’)
\z\pagebreak\largerpage
\ex\label{ex:tassel}
\ea{\ili{Nigerian} \ili{Arabic} \citep[65]{Owens2016idioms}}\\
\gll   {\R}ās al-qalla  \\
       head \textsc{def}-corn\\
\glt   `tassel' (lit. ‘head of corn')

\ex{\ili{Kanuri} \citep[65]{Owens2016idioms}}\\
\gll   kǝla argǝm-be \\
       head corn-\textsc{gen} \\
\glt   `tassel' (lit. ‘head of corn')
\z
\z

According to Owens (\citeyear[65]{Owens2016idioms}), \ili{Kanuri}–\ili{Arabic} \isi{bilingualism}, with \ili{Arabic} being a minority language, would have been the foremost factor underlying the \isi{transfer} of these \isi{compound nouns} into \ili{Nigerian} \ili{Arabic}. He further stresses that \ili{Kanuri} is the main source of \isi{compound nouns} in a number of other minority languages in the area (e.g. \ili{Kotoko}, Glayda, and \ili{Fulfulde}) and that there is little evidence of \ili{Kanuri} to \ili{Arabic} shift in the region \citep[147]{Owens2014}. However, the fact that \ili{Kanuri} represents the majority language of northeastern Nigeria, does not shed light on the \isi{transfer} mechanism lying behind lexical \isi{calquing} in \ili{Nigerian} \ili{Arabic}. This is because speakers can be linguistically dominant in a socially subordinate language \citep[376]{Winford2005}. In fact, such contact settings are closely tied to \isi{SL} agentivity, as the youngest bilingual generations tend to impose semantic features from their dominant language (i.e. \ili{Kanuri}) onto the ancestral language (i.e. \ili{Arabic}). It is only at a later stage that these innovations are borrowed by older bilingual speakers who are still dominant in \ili{Arabic}. 

The fact that \ili{Nigerian} \ili{Arabic} speakers have gradually developed a high bilingual proficiency in \ili{Kanuri} is also testified by the \isi{transfer} of a number of \isi{idiomatic} expressions. In this regard, Ross (\citeyear[122]{Ross2007}) observes that \isi{calquing} of meaning is not only reflected in word compounding, but also in lexical collocations of \isi{idiomatic} expressions. These are combinations of lexical items that are semantically idiosyncratic as they have a pairing of form and meaning that cannot be predicted from the rest of the grammar. The pair of examples in \REF{distract} provide evidence of an \isi{idiomatic} \ili{Kanuri} \isi{calque} in \ili{Nigerian} \ili{Arabic}. 

\ea\label{distract}
\ea{\ili{Nigerian} \ili{Arabic} \citep[77]{Ritt-Benmimounetc2017}}\\
\gll   šuqul šāl {\R}ās-i\\
       something carry.\textsc{prf.3sg.m} head-\textsc{obl.1sg}\\
\glt   `Something distracted me.' (Lit. ‘Something carried my head.’)

\ex{\ili{Kanuri} \citep[77]{Ritt-Benmimounetc2017}}\\
\gll   awode kǝla   gō-zǝ-na \\
       something head carry-\textsc{3sg-prf}\\
\glt   `Something distracted me.' (Lit. ‘Something carried head.’)
\z
\z

Given that \isi{idiomatic} expressions are syntactically compositional (i.e. their lexical components behave syntactically as they do in non-\isi{idiomatic} expressions), it is not only the meanings expressed by the lexeme ‘head’ which correspond between \ili{Nigerian} \ili{Arabic} and \ili{Kanuri}, but also their \isi{idiomatic} collocations, which align between the two languages \citep[157]{Owens2014}. Besides, it is worth noting that also \isi{idiomatic} expressions are adjusted to fit \isi{RL} morphosyntactic patterns. This is evidenced by the inalienable possession of body parts in \ili{Nigerian} \ili{Arabic} (\textit{{\R}ās-i} ‘my head’), which is instead unattested in the \isi{SL} (\textit{k\kern .75ptǝ\kern -1ptla} ‘head’). Even if we cannot exclude the possibility that these kinds of calques are a product of borrowing, it is evident that their integration needs a high proficiency in the \isi{SL} for individuating the single \isi{idiomatic} collocations of lexical items. Furthermore, differently from borrowed calques, imposed \isi{idiomatic} expressions can significantly affect the lexical semantics of the \isi{RL} created by SL-dominant bilinguals, and thus produce grammatical changes in the long run. 

Finally, lexical \isi{calquing} via \isi{SL} agentivity can also take place in extreme contact situations such as creolization. For instance, \ili{Juba} \ili{Arabic}, the \ili{Arabic}-based pidgincreole spoken in South Sudan, shows numerous calques in which \ili{Arabic}-derived lexemes are compounded according to the semantic patterns of \ili{Bari}, the main \isi{substrate} language of \ili{Juba} \ili{Arabic} (\citealt{Nakao2012}; \citealt[50]{Manfredi2017}; Avram, this volume)\ia{Avram, Andrei A.@Avram, Andrei A.} As we can see in \REF{tree} and \REF{eleph}, the \isi{word order} in \ili{Juba} \ili{Arabic} \isi{compounds} follows the order of \ili{Bari} \isi{compounds}. However, this cannot be seen as an innovative morphosyntactic development, as the possessed–possessor order matches also with the \ili{Arabic} \isi{lexifier}. 

\ea\label{tree}
\begin{multicols}{2}\raggedcolumns
\ea{\ili{Juba} \ili{Arabic} \citep[136]{Nakao2012}} \\
\gll   éna ta séjera  \\
       eye \textsc{gen} tree \\
\glt   `fruit’ (lit. ‘eye of tree’)
\columnbreak
\ex{\ili{Bari} \citep[136]{Nakao2012}}\\
\gll   koŋe lo-ködini\\
       eye \textsc{gen}-tree\\
\glt   `fruit’ (lit. ‘eye of tree’)
\z\end{multicols}\begin{multicols}{2}\raggedcolumns
\ex\label{eleph}
\ea{\ili{Juba} \ili{Arabic} \citep[137]{Nakao2012}} \\
\gll   ída ta fil\\
       hand \textsc{gen} elephant \\
\glt   `trunk’ (lit. ‘hand of elephant’)
\columnbreak
\ex{\ili{Bari} \citep[137]{Nakao2012}}\\
\gll   könin lo-tome\\
       hand \textsc{gen}-elephant \\
\glt   `trunk’ (lit. ‘hand of elephant’)
\z\end{multicols}
\z

Given that the asymmetric contact situation leading to \isi{creole} \isi{formation} limits access to the \isi{superstrate} language (i.e. \ili{Sudanese} \ili{Arabic}), the semantic patterns of \isi{substrate} languages (i.e. \ili{Bari}) can be easily carried over into the \isi{creole} in ways peculiar to \isi{imposition} via \isi{SL} agentivity. 

All things considered, unlike lexical borrowing, lexical \isi{calquing} allows for a semantic overlapping of \isi{RL} and \isi{SL} lexical entries and it can also produce important structural changes. 


 
 \subsection{Grammatical calquing}\label{sec:gra}

Grammatical \isi{calquing} brings about a match between the grammatical categories of two languages and the memberships of these categories \citep[132]{Ross2007}. \citet{HeineKuteva2005} suggest that the grammatical changes induced by \isi{calquing} can be better analysed in terms of \isi{contact-induced grammaticalization} (cf. Leddy-Cecere, this volume).\ia{Leddy-Cecere, Thomas@Leddy-Cecere, Thomas} In fact, the \isi{calquing} of the semantic properties of lexical and grammatical items may lead to the \isi{grammaticalization} of innovative syntactic structures in the \isi{RL} matching with those of the \isi{SL}. From the traditional sociohistorical perspective of contact-induced change (\citealt{ThomasonKaufman1988}), grammatical \isi{calquing} is basically seen as a product of \isi{language shift}. In contrast, Ross (\citeyear[131]{Ross2007}) argues that grammatical calques can widely occur in situations of language maintenance. Actually, the different grammatical outputs of \isi{calquing} mainly depend on the way in which they are transferred from the \isi{SL} into the \isi{RL} and, by \isi{extension}, on different kinds and degrees of \isi{bilingualism}.   

For the purposes of this chapter, I distinguish between three different types of grammatical \isi{calquing}: 

\begin{enumerate} 
\item Calquing of polyfunctionality of lexical items without syntactic change;
\item Calquing of polyfunctionality of grammatical items leading to syntactic change; 
\item Narrow syntactic \isi{calquing} (without \isi{calquing} of polyfunctionality of lexical/grammatical items).
\end{enumerate}

Being lexical in nature, the first of these three types of grammatical \isi{calquing} can be triggered by both \isi{imposition} via \isi{SL} agentivity and borrowing via \isi{RL} agentivity, whereas the two latter types are likely to result only from \isi{imposition} via \isi{SL} agentivity. 

Calquing of polyfunctionality patterns of lexical items is by far the most common type of grammatical \isi{calquing}, and it can be exemplified by the comparison of reflexive anaphors in different \ili{Arabic} dialects. As is well known, \ili{Classical} and \ili{Standard} \ili{Arabic} express a reflexive meaning either by means of agent-oriented derived verbs lacking an overtly expressed patient (e.g. \textit{istaḥamma} ‘he washed himself’) or by anaphoric constructions in which the syntagm \textit{nafs-}\textsc{pro.poss} ‘soul-\textsc{pro.poss’} marks coreferentiality between the agent and the patient of the predicate (e.g. \textit{qatala} \textit{nafsa-hu} ‘he killed himself’). Nevertheless, as a result of contact with different languages, a number of modern \ili{Arabic} dialects have grammaticalized other lexical sources for expressing a reflexive meaning. Western \ili{Maghrebi} dialects are a case in point. As we can see in \REF{15}--\REF{16}, both \ili{Moroccan} and \ili{Ḥassāniyya} \ili{Arabic} have grammaticalized the nominal syntagm \textit{{\R}āṣ=}\textsc{pro.poss} ‘head-\textsc{pro.poss’} as default reflexive anaphor.

\ea\label{15}
{\ili{Moroccan} \ili{Arabic} (D. Caubet, personal communication)} \\
\gll   qtәl {\R}ās-o\\
       kill.\textsc{prf.3sg.m} head-\textsc{3sg.m}\\
\glt   `He killed himself.' (Lit. ‘He killed his head.’)
\ex \label{16}
{\ili{Ḥassāniyya} \ili{Arabic} \citep[16]{Taine-Cheikh2008chapter}}\\
\gll   ktәl {\R}āṣ-u\\
       kill.\textsc{prf.3sg.m} head-\textsc{3sg.m}\\
\glt   `He killed himself.' (Lit. ‘He killed his head.’)
\z

This reflexive use of the lexeme ‘head’ has generally been interpreted as \isi{substrate} interference from \ili{Berber} languages \citep[197]{ElAissati2011}, in which the same \isi{grammaticalization} path is attested, as shown in the following examples from \ili{Tarifiyt} and \ili{Zenaga}: 

\ea
{\ili{Tarifiyt} \ili{Berber} (\citealt[95]{Kossmann2000})}\\
\gll   yәtšaθ iḫәf nnәs\\
       beat.\textsc{prf.3sg.m} head \textsc{poss.3sg.m}\\
\glt   `He beats himself.' (Lit. ‘He beats his head.’)
\ex
{\ili{Zenaga} \ili{Berber} \citep[126]{Taine-Cheikh2008chapter}}\\
\gll   yәʔna iʔf-әn-š\\
       kill.\textsc{prf.3sg.m} head-\textsc{gen-3sg.m}\\
\glt   `He killed himself.' (Lit. ‘He killed his head.’)
\z

Lexemes for ‘head’ are the second most common source of \isi{grammaticalization} of reflexive anaphors worldwide \citep{KönigTöpper2013} and its occurrence is particularly common in West Africa (\citealt[50]{Heine2011}). In this scenario, it should be stressed that the reflexive function of the lexeme ‘head’ is an innovative feature of both \ili{Arabic} and \ili{Berber} varieties of northwestern Africa. Other \ili{Berber} languages typically use the reflexive anaphor \textit{iman-}\textsc{poss} ‘soul-\textsc{poss}’, as we can see in the following example from \ili{Kabyle}.

\ea
{\ili{Kabyle} \citep{Mettouchi2012}}\\
\gll   n-səlk-dd       iman-ntəɣ\\
       \textsc{1pl}-spare.\textsc{prf-prox} soul.\textsc{abs.sg.m-poss.1pl.f}\\
\glt   `We saved ourselves.'\\
\z

In addition, the known \ili{Arabic}–\ili{Berber} contact situation, in which second language acquirers of \ili{Berber} only played a marginal role in triggering contact-induced change in \ili{Arabic}, suggests that the contact induced \isi{grammaticalization} of ‘head’ in westernmost \ili{Arabic} dialects resulted from an \isi{imposition} enacted by former \ili{Berber}-dominant speakers.  

A similar instance of \isi{calquing} in the domain of anaphoric reflexive constructions is found in Kordofanian \ili{Baggara} \ili{Arabic}, a \ili{Western Sudanic} dialect spoken in the Nuba Mountains area, in central Sudan. In this case, the source of the reflexive anaphor is the lexeme for ‘neck’, as we can see in \REF{20}.  

\ea\label{20}
{Kordofanian \ili{Baggara} \ili{Arabic} \citep[176]{Manfredi2010}}\\
\gll  abrahīm gaṣṣa ragabt-a\\
       Ibrahim cut.\textsc{prf.3sg.m} neck-\textsc{3sg.m}\\
\glt   `Ibrahim cut himself.' (Lit. `Ibrahim cut his neck.')
\z

Different from ‘head’, the \isi{grammaticalization} of ‘neck’ as a reflexive anaphor is quite rare in Africa (\citealt[50]{Heine2011}), but it is attested in a number of Niger-Kordofanian languages spoken in the same region. Such is the case of \ili{Tagoi} \REF{21} and \ili{Koalib} \REF{22}. 

\ea\label{21}
{\ili{Tagoi} \citep[26]{Alamin2015}}\\
\gll   t-áɡám t-ùrúŋ ínní\\
       \textsc{nc}-neck \textsc{nc}-\textsc{poss.3} kill.\textsc{prf.3}\\
\glt   `He killed himself.' (Lit. ‘He killed his neck.’)
\ex\label{22}
{\ili{Koalib} (N. Quint, personal communication)} \\
\gll   ɛ̀ɽnyɛ́ r-ɔ́kwɽɔ̀ r-ùŋwún\\
       kill \textsc{nc}-neck \textsc{nc-poss.3}\\
\glt   `to kill oneself' (lit. ‘to kill one's neck’)
\z

Similarly to the situation described with reference to western \ili{Maghrebi} dialects, \ili{Arabic}-speaking groups in the Nuba Mountains have hardly developed any bilingual competence in local Niger-Kordofanian languages. Therefore, it seems likely that the \isi{calquing} of the polyfunctionality patterns of ‘neck’ has been imposed by Arabized populations who were dominant in the \isi{SL}. 

\ili{Maltese} also provides remarkable examples of \isi{calquing} of polyfunctionality of lexical items. This is particularly evident in the domain of \isi{auxiliary verbs} (\citealt{Vanhove1993}; \citealt{VanhoveCaubet2009}). A well-known example is that of the lexical verb \textit{\.gie} ‘come’ used as an auxiliary for expressing a dynamic \isi{passive} \REF{23} in the same way as \ili{Italian} \REF{24}.

\ea\label{23}
{\ili{Maltese} \citep[214]{BorgAzzopardi-Alexander1997}}\\
\gll   it-tabib \.gie afdat bi-l-każ\\
       \textsc{def}-doctor come.\textsc{prf.3sg.m} trust.\textsc{ptcp.pass} with-\textsc{def}-case\\
\glt   `The doctor was entrusted with the case.' (Lit. ‘The doctor came entrusted with the case.’)
\ex \label{24}
{\ili{Italian} (own knowledge)} \\
\gll   non venne creduto\\
       \textsc{neg} come.\textsc{pst.3sg.m} trust.\textsc{ptcp.pst}\\
\glt   `He was not trusted.' (Lit. ‘He did not come trusted.’)
\z

Even if \isi{imposition} played a role in the emergence of \ili{Maltese} (Lucas \& Ćeplö, this volume),\ia{Lucas, Christopher@Lucas, Christopher}\ia{Čéplö, Slavomír@Čéplö, Slavomír} it is generally accepted that intertwined languages emerge mainly from a widespread process of borrowing in Van Coetsem’s terminology (\citealt[397]{Winford2005}; \citealt{Manfredi2018}). This suggests that, unlike the aforementioned \isi{grammaticalization} of reflexive anaphors in \ili{Arabic} dialects, the \isi{calquing} of polyfunctionality of lexical verb ‘come’ in \ili{Maltese} was most likely triggered by agentivity of \isi{RL} dominant speakers. 

Regardless of the different contact situations, what unites all the previous instances of grammatical \isi{calquing} is the fact that the \isi{transfer} of patterns of \isi{grammaticalization} did not produce any syntactic change in \ili{Arabic}. In contrast to the above, the \isi{calquing} of polyfunctionality of grammatical items can be accompanied by important typological changes. This is the case of the \isi{grammaticalization} of prototypical \isi{passive} constructions in \ili{Juba} \ili{Arabic} (\citealt[92]{Manfredi2017}; \citeyear[415]{Manfredi2018}). As we can see in \REF{passive}, the South Sudanese pidgincreole presents an innovative \isi{passive} construction in which the patient occupies the syntactic slot of a preverbal subject, whereas the oblique-marked agent is introduced by the comitative \isi{preposition} \textit{ma-} ‘with’.  

\ea\label{passive}
{\ili{Juba} \ili{Arabic} (\citealt[86]{Manfredi2017})}\\
\gll   bab de kasurú ma-jón \\
       door \textsc{prox.sg} break.\textsc{pass} with-John\\
\glt   `This door has been broken by John.' (Lit. ‘This door has been broken with John.’)
\z
 
Interestingly, this prototypical \isi{passive} construction is not attested in the \isi{lexifier} language of \ili{Juba} \ili{Arabic} (i.e. \ili{Sudanese} \ili{Arabic}), which instead makes use of impersonal \isi{passive} constructions with a default \textsc{3pl.m} subject.

\ea
{\ili{Sudanese} \ili{Arabic} (own knowledge)}\\
\gll   kassaru-hu\\
       break.\textsc{prf.3pl.m-3sg.m}\\
\glt   `It got broken.' (Lit. ‘They have broken it.’) 
\z

Indeed, the \isi{grammaticalization} of this complex syntactic structure is the result of the \isi{calquing} of the functional properties associated with the comitative \isi{preposition} of the main \isi{substrate} language, \ili{Bari}. \ili{Bari} presents the same kind of prototypical \isi{passive} construction in which an oblique-marked agent is introduced by the \isi{preposition} \textit{ko-} ‘with’.

\ea
{\ili{Bari} (\citealt[65]{Owen1909})}\\
\gll   niena  wuret   a-wur-ö ko-nan \\
       \textsc{prox.sg} book \textsc{3sg.pst}-write-\textsc{pass} with-\textsc{1sg}\\
\glt   `This book has been written by me.' (Lit. ‘This book has been written with me.’)
\z

If we assume that the emergence of \isi{creole} languages is always induced by the disruption of the \isi{transmission} of the \isi{lexifier} language \citep{Comrie2011}, we can conclude that \ili{Bari} speakers have imposed the semantics of their dominant language on a grammatical item derived from \ili{Arabic}, and thus induced profound changes in the \isi{word order} of the \isi{creole} when compared to its \isi{lexifier} language.\footnote{This kind of syntactic change accompanied by the \isi{calquing} of semantic properties of \isi{substrate} items in \isi{creole} languages is traditionally labelled ``relexification'' \citep{Lefebvre1998}.}  In light of the above, the contact dynamics lying behind the \isi{calquing} of polyfunctionality of grammatical items are quite restrictive as they are most likely a product of \isi{imposition} via \isi{SL} agentivity.\largerpage[2]

The third kind of grammatical \isi{calquing} is linked to the \isi{transfer} of syntactic patterns without \isi{transfer} of polyfunctionality of either lexical or grammatical items. This narrow type of syntactic \isi{calquing} can be exemplified by possessor doubling in \ili{Central Asian} \ili{Arabic} \citep{Ratcliffe2005}. Clitic doubling is a construction in which a \isi{clitic} co-occurs with a full nominal phrase in argument position, forming a discontinuous constituent with it. Various forms of \isi{clitic doubling} have arisen in a number of \ili{Arabic} varieties as a result of contact with different \isi{substrate}/\isi{adstrate} languages \citep{Souag2017clitic}. In regard to possessive constructions, \ili{Arabic} typically presents a possessed–possessor order. In contrast, \ili{Central Asian} \ili{Arabic} \REF{28} gives evidence of the opposite order with obligatory possessor doubling in the same way as \ili{Tajik} \REF{29}. 

\ea\label{28}
{\ili{Central Asian} \ili{Arabic} (\citealt{Ratcliffe2005}; \citealt[56]{Souag2017clitic})}\\
\gll   amīr wald-u\\
        prince son-\textsc{3sg.m}\\
\glt   `the prince’s son'
\ex \label{29}
{\ili{Tajik} (\citealt[56]{Souag2017clitic})}\\
\gll   buḫoro universitet-aš\\
       Bukhara university-\textsc{3sg}\\
\glt   `Bukhara University'
\z

Souag (\citeyear[157]{Souag2017clitic}) states that double possessor constructions in \ili{Central Asian} \ili{Arabic} are instances of grammatical \isi{calquing}, accommodated through the reinterpretation of pre-existing topicalized constructions. This means that, unlike the syntactic changes induced by the \isi{calquing} of polyfunctionality of mor-\linebreak phemes, the emergence of double possessor constructions in Bukhara \ili{Arabic} would have been favoured by a formal congruence between \isi{SL} and \isi{RL} syntactic structures. As such, this instance of contact-induced morphosyntactic restructuring (metatypy) does not derive from a direct copying of a double possessor construction. Rather, it consists in speakers expressing a possessive meaning in \ili{Arabic} by using a construction which they equate with the construction in adstratal languages (\citealt[128]{Ross2007}). If we consider that the youngest speakers of \ili{Central Asian} \ili{Arabic} are gradually losing competence in their ancestral language in favour of socially dominant languages (\citealt[128]{Chikovani2005}), it is plausible to assume that this kind of syntactic restructuring can only be a result of \isi{imposition} via \isi{SL} agentivity. Still, given our limited diachronic knowledge, we cannot exclude the hypothesis of an early process of borrowing enacted by former \ili{Arabic}-dominant speakers.   

\section{Conclusion}\largerpage

Van Coetsem (\citeyear[20]{VanCoetsem1988}) suggests that the variable outcomes of language contact are primarily a reflex of the ``\isi{stability gradient}'' of language, which induces speakers to preserve the domains of their dominant language that are less affected by change. As lexicon is the most unstable linguistic domain, it is likely to be transferred via \isi{RL} agentivity. In contrast, morphosyntax and phonology are considered to be relatively stable domains and they are expected to be transferred only via \isi{SL} agentivity. Against this background, it is unclear how the \isi{transfer} of semantic features deprived of morphophonological matter should be understood in relation to the linguistic dominance of the agents of contact-induced change. 

If we look at the previously analysed instances of lexical \isi{calquing} (§\ref{sec:lex}), it is evident that the \isi{transfer} of the semantic features of nominal \isi{compounds} can take place within speech communities with a very low degree of \isi{bilingualism}, as in the case of {Egyptian}-\ili{Arabic}-dominant speakers borrowing the semantics of \ili{English} \isi{compounds}. But it is also true that compound \isi{calquing} can be a product of \isi{imposition} resulting from ongoing \isi{language shift} or pidginization, and the \isi{transfer} of semantic features of single lexical items within \isi{idiomatic} expressions always requires a widespread proficiency in the \isi{SL}, as in the case of \ili{Arabic}–\ili{Kanuri} \isi{bilingualism} in northern Nigeria.\largerpage

As far as grammatical \isi{calquing} is concerned (§\ref{sec:gra}), I have shown that \isi{calquing} of the polyfunctionality of lexical items can be triggered either by \isi{imposition}, as in the case of \isi{substrate} interference in Ḥass\-āniyya\il{Ḥassāniyya} and \ili{Baggara} dialects, or by borrowing in the emergence of intertwined languages such as \ili{Maltese}. Calquing of polyfunctionality of grammatical items, for its part, requires a higher degree of linguistic abstraction for the identification of a functional overlap between morphemes. Accordingly, this type of \isi{transfer} will typically occur via \isi{imposition} by SL-dominant speakers in deep contact situations such as creolization. In the same manner, narrow syntactic \isi{calquing} requires high bilingual proficiency, as it necessitates the recognition of some formal congruence between the \isi{SL} and the \isi{RL}, as shown by the emergence of possessor doubling in \ili{Central Asian} \ili{Arabic}.  

To stay somewhat in line with the \isi{stability gradient} principle, we could argue that, in absence of the \isi{transfer} of linguistic matter, the semantic properties of morphemes and syntactic constructions are more stable than those of lexical items. However, such a generalization would be misleading without an in-depth knowledge of the sociolinguistic circumstances underlying a specific instance of second language acquisition (i.e. symmetric \isi{bilingualism}, asymmetric \isi{bilingualism}, multilingualism, pidginization/creolization). Thus, it becomes evident that the recognition of different patterns of \isi{bilingualism} within the same community remains the only way to identify the \isi{transfer} type at play in a given contact situation, regardless of its different structural outputs.  

Drawing on the available literature, this chapter has surveyed only a few instances of  \isi{calquing} induced by contact between \ili{Arabic} and other languages. This is mainly because we lack information about \isi{calquing} in \isi{dialect contact} situations. Indeed, it is regrettable that studies dealing with \isi{dialect contact} and \isi{new dialect} \isi{formation} are still exclusively focused on the \isi{diffusion} of few lexical and morphophonological features, while disregarding the \isi{transfer} of semantic and syntactic patterns. Fine-grained analyses of semantic changes induced by \isi{dialect contact} thus remain a major desideratum for the development an aggregate variationist \ili{Arabic} dialectology.  

\section*{Further reading}
\begin{furtherreading}
\item \citet{Keesing1988} adopts the notion of \isi{calquing} and describes the \isi{transfer} of semantic properties of Oceanic morphemes in Melanesian Pidgin.
\item \citet{Meyerhoff2009} focuses on the notions of \isi{replication}, \isi{transfer}, and \isi{calquing}, thereby strengthening connections between variationist sociolinguistics and contact linguistics.
\item \citet{Zuckermann2009} provides numerous instances of \isi{calquing} in Modern \ili{Hebrew} and analyses them in the light of the Congruence Principle.
\end{furtherreading}

\section*{Abbreviations}

\begin{tabularx}{.5\textwidth}[t]{@{}lQ@{}}
\textsc{1, 2, 3} & 1st, 2nd, 3rd person \\
\textsc{def} & {definite} \isi{article} \\
\textsc{f} & feminine \\
\textsc{gen} & genitive \\
\textsc{m} & masculine \\
\textsc{nc} & noun class  \\
\textsc{obl} & oblique \\
\textsc{pass} & {passive} \\
\textsc{pst} & past \\
\textsc{pl} & plural \\
\end{tabularx}%
\begin{tabularx}{.5\textwidth}[t]{@{}lQ@{}}
\textsc{poss} & possessive pronoun \\
\textsc{prg} & pragmatic marker \\
\textsc{prf} & perfect (prefix conjugation) \\
\textsc{pro} & pronoun \\
\textsc{prox} & proximal \\
\isi{RL} & {recipient language} \\
\textsc{refl} & reflexive \\
\textsc{sg} & singular \\
\isi{SL} & {source language} \\
\textsc{abs} & absolute state \\
\end{tabularx}%

{\sloppy\printbibliography[heading=subbibliography,notkeyword=this]}
\end{document}
