\chapter*{Acknowledgement}
\label{sec:acknow}

%jungle
%sumo naa
%teacha wusa
%nana
%mansa bele

The completion of this project was dependent upon a multitude of factors, the most important being  the  knowledge and generosity of Chakali-speaking individuals. I am  indebted to  Daniel Kanganu Karija,   Fuseini Mba Zien, Afia Kala Tangu,  Awie Bakuri Ahmed,  Seidu Kassim Tangu, Henry Seidu Daannaa, Godfrey Bayon Tangu, Kpersi-Naa\dag, Kotia Nwabipe\dag, Mangwe-Naa\dag, Mark Zoon-Naa\dag, Amoa Bari-Naa, Siibu Jakalia, Kala Osman,  and Adam Sakara Baduong, among many others,  for their instructions and enlightenments. I owe special gratitude to Daniel Kanganu Karija for his loyal contribution since the very beginning of my learning journey. A special thank you goes  to Jonas Kpierekoh, a specialist in agroforestery and principal programme officer at the  Environmental Protection Agency, who spent time with me and local experts in \isi{Ducie} working on the scientific name of many tree species.  With immense gratitude, I  thank everyone in the villages where I stayed -- especially Ducie and \isi{Gurumbele} --  for their warm welcome, hospitality, and cooperation. 

This book has undergone several stages and has benefited from the suggestions  and criticisms of many people.  I would like to express my deep thanks to  Mary Esther Kropp Dakubu (who is sadly no more),  for her guidance, support,  advice,  and  linguistic insight throughout the years.  I wish to thank my wife Léonie, and friends and colleagues, Benjamin Waldron,  Felix K. Ameka,  Kaja Borthen, Albert Awedoba,  Tyson Farrell, Lars Hellan, Jolanta Bachan, Rachel Selbach, Kenneth Mango, and Assibi Apatewon Amidu, who kindly suggested corrections, and commented on  earlier versions.  For his assistance with the carthography, thanks to Per Wirehn. I  gratefully  acknowledge  the  generous assistance John Rennison and   Tony Naden have provided at different stages.  Thanks as well to ALGaD Series' Editors and to  Sebastian Nordhoff at Language Science Press for his editorial aid in preparing this book for publication.