\addchap{Acknowledgments} 

I wish to express my gratitude to the people in Equatorial Guinea whose support is the foundation of this work. I am deeply indebted to Françoise Tatchouop for her hospitality and generosity during my stays in Equatorial Guinea. I am most grateful to my language teachers, the late Natalia “Abuela” Toichoa Borico, María Rosa “Mami Rose” Buesule Toichoa, Rudolfo “Djunais” Beaka Chale, Ildefonso “Boye” Ntutumu, Sonia “Lage” Belobe Toichoa, Sandra Eyang Ncoha Belobe, Miguel Ángel Ñat Buesule, Fermín Beaka Chale, Juan Antonio Tonca Toichoa, Agustín Gaspar Nguema Eñeso, Rosalía Ndjoku, Maura Toichoa Lopete, María Fernanda Aboki Sami, Hilda Méndez Sami, Eduardo Mejía, Ursus Megua Kofi, Miguel Fernández Ndonjo, as well as María, Beatriz, Lindo, and Charlie of Ela Nguema. I also wish to thank Trinidad Morgades Besari, the late Samuel Ebuka, the late Edward Jones, and Ruth Jones for sharing their wealth of knowledge on the history and the peoples of Bioko with me.


A special note of thanks goes to Danae Perez for copy editing and the correction of Spanish examples, as well as to Hing-Yuet Fung for a final round of editing and problem fixing in LaTeX. I am grateful to Sebastian Nordhoff and Redmer Kronemeijer for assistance with the conversion to LaTeX and typesetting, and to the members of the Language Science Press community for proofreading. I also wish to thank three anonymous reviewers for their detailed comments and suggestions.


I am obliged to Pieter Muysken, whose support secured funding from the Centre for Language Studies at Radboud University Nijmegen for field research between 2003 and 2007. I equally wish to acknowledge the support of grant no. 27613316 of the Research Grants Council of the Government of Hong Kong, which provided me with funding for complementary research in 2016.

{\raggedleft
Kofi Yakpo\\
Hong Kong, {December 2018} \par
}

