\documentclass[output=paper]{langscibook}
\ChapterDOI{10.5281/zenodo.5792959}
\author{Mikael Kalm\affiliation{Stockholm University}}
\title{The emergence of adverbial infinitives in Swedish}
\abstract{Prepositional infinitives are commonly used in present-day Swedish to express a range of adverbial notions, including purposive, abessive, substitutive, temporal, and instrumental meanings. However, only purposive infinitives are attested in Old Swedish. The aim of this article is to give an account of the emergence of adverbial infinitives in the history of Swedish. Other adverbial infinitives emerged from the late 17\textsuperscript{th} century onwards, possibly as a result of contact-induced (replica) grammaticalization. It is furthermore argued that the emergence of adverbial infinitives should be seen as a result of the increasing importance and consequent demands for precision of the written language, and as part of the establishing of a written norm, separate from the spoken language. This assumption is supported by data from the traditional dialect of Övdalian, where adverbial infinitives are not used.

\keywords{adverbial infinitives, prepositional infinitives, Swedish, Övdalian, language Ausbau, grammatical replication, contact-induced change, Verschriftlichung}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\SetupAffiliations{mark style=none}
\maketitle 
%\shorttitlerunninghead{}%%use this for an abridged title in the page headers



\section{Introduction}\label{sec:kalm:1}\largerpage
The development of Swedish into a literary language is to a large extent characterized by intensive contact with other languages: \ili{Latin} and \ili{Low German} in the Middle Ages and \ili{High German} and \ili{French} in the Early Modern era. The first steps towards a literary and somewhat standardized written language were already being taken during the late Middle Ages in the monasteries and convents, most importantly at the abbey of \isi{Vadstena}. The \ili{Latin} influence on these texts is considerable, not least at a syntactic level, causing a split between written \ili{Old Swedish} and the spoken varieties. This has been described as a process of language \isi{Ausbau} (see \citealt{Hoder2009, Hoder2010}). A more homogeneous written language took shape after the Reformation, with the translation of the \isi{Bible} into Swedish during the first half of the 16\textsuperscript{th} century as an important landmark. It was not until the end of the 18\textsuperscript{th} century, however, that the \isi{standardization} process was completed (more or less). As mentioned above, this post-Reformation period is characterized by contact between Swedish and \ili{High German} and \ili{French} (as well as \ili{Latin}, but to a much lesser extent than before). Traces of this influence are clearly visible in the vocabulary even today (see \citealt{Teleman2003Swedish,Teleman2003Tradis}), but we know considerably less about the syntactic influence from these languages. 



This paper investigates a syntactic \isi{innovation} from the dynamic post-Ref\-or\-ma\-tion period in the history of written Swedish, namely the emergence of \isi{adverbial} infinitives, i.e. \isi{prepositional infinitive} clauses functioning as \isi{adverbial} adjuncts. While present-day Swedish has a range of such prepositional clauses, they are not attested in \ili{Old Swedish}. A few examples of the \isi{adverbial} infinitives discussed in this article are given in \REF{ex:kalm:1} below, all from present-day Swedish.\largerpage  



\ea
\label{ex:kalm:1}
\ea \is{purposive (infinitive)}Purposive\label{ex:kalm:1a}\\
\gll Han gick ut för att få sig lite luft.\\
he went out for \textsc{im} get.\textsc{inf} \textsc{refl} some air\\
\glt ‘He went out to get some air.’

\ex \is{abessive (infinitive)}Abessive\label{ex:kalm:1b}\\
\gll Hon sålde företaget utan att fråga de anställda.\\
she sold company.\textsc{def} without \textsc{im} ask.\textsc{inf} the employees\\
\glt ‘She sold the company without asking the employees.’

\ex \is{standardization}Standardization\label{ex:kalm:1c}\\
\gll Jag satt hemma istället för att gå ut.\\
I sat home instead for \textsc{im} go.\textsc{inf} out\\
\glt‘I stayed home instead of going out.’

\ex Temporal\label{ex:kalm:1d}\\
\gll Hon läste medicin efter att ha kommit tillbaka från Berlin.\\
she read medicine after \textsc{im} have.\textsc{inf} come back from Berlin\\
\glt ‘She studied medicine after having returned from Berlin.’

\ex Instrumental\label{ex:kalm:1e}\\
\gll Han gjorde sig oumbärlig genom att alltid vara steget före.\\
he made \textsc{refl} indispensable through \textsc{im} always be.\textsc{inf} step.\textsc{def} before\\
\glt ‘He made himself indispensable by always being one step ahead.’
\z
\z

\is{adverbial}Adverbial infinitives (as well as \isi{adverbial} constructions more generally) have not attracted much attention from Scandinavian scholars. One exception is \citet{Holm1967}, who makes several claims about \isi{adverbial} infinitives in his influential survey of prose styles in the history of Swedish. Firstly, focusing on present-day North \ili{Germanic}, Holm suggests that \isi{adverbial} infinitives belong to a written register and are not used in “uneducated” (Sw. \textit{obokligt}) speech, nor in the traditional dialects. He also claims that \isi{adverbial} infinitives emerged with inspiration from other languages, such as \ili{High German} and \ili{French}, and that they are not attested in \ili{Old Swedish} \citep[27]{Holm1967}. Holm does not present any evidence to support his claims. 



The aim of this article is to give a more detailed account of \isi{adverbial} infinitives in the history of Swedish, and to compare standard Swedish with the \isi{non-standardized variety} of \ili{Övdalian} (see \sectref{sec:kalm:4.3}) with regard to the use of prepositional infinitives to express \isi{adverbial} notions. I will also discuss whether the emergence of \isi{adverbial} infinitives in Swedish could be an example of contact-induced \isi{grammaticalization} (see \citealt{HeineKuteva2003, HeineKuteva2005}) and if this, in turn, can be understood as part of an ongoing language \isi{Ausbau} process (in the sense of \citealt{Kloss1967}). If so, \isi{adverbial} infinitives should originally have been part of a written register rather than a spoken register. Important questions are therefore: (1) when \isi{adverbial} infinitives emerged in Swedish, and (2) whether there is support for the claim that they first appeared as part of a written register. In order to answer the first \isi{question}, I have investigated Early Modern and \ili{Late Modern Swedish} texts (ca. 1500–2000) from the digital corpora the \ili{Old Swedish} Text Bank (Sw. \textit{Fornsvenska textbanken}) and the Swedish Literature Bank (Sw. \textit{Litteraturbanken}), both available through the corpus infrastructure \isi{Korp} \citep{BorinEtAl2012}.\footnote{\url{https://spraakbanken.gu.se/korp/}}  In addition to this, the corpus of Swedish drama dialogue (Sw. \textit{Svensk dramadialog}; \citealt{MarttalaStromquist2001}), covering the period 1725–2000, has also been investigated in its entirety. To shed light on the second \isi{question} (i.e. whether \isi{adverbial} infinitives first appeared as part of a written register), I have conducted a contrastive study of Swedish and the \isi{traditional dialect} of \ili{Övdalian}, using translations from Swedish into \ili{Övdalian}.



The article is organized as follows. In \sectref{sec:kalm:2}, I give a brief introduction to the concept of language \isi{Ausbau} in relation to the emergence of standardized written Swedish. \sectref{sec:kalm:3} provides a background to the empirical study, with a survey of the history of the \ili{Germanic} infinitive and a typological perspective on \isi{adverbial} infinitives in European languages. In \sectref{sec:kalm:4}, I present my empirical findings, and \sectref{sec:kalm:5} provides a theoretical discussion. Finally, \sectref{sec:kalm:6} contains some concluding remarks. 


\section{Functional expansion and formal elaboration of written Swedish}\label{sec:kalm:2}



In the process of codifying a traditionally spoken language or \isi{dialect}, i.e. transforming it into a written language, the written language will soon differ in certain regards from the spoken varieties. Over time, the use of the written language is normally expanded into new domains. This expansion itself leads to an elaboration of the written language, as it is shaped in more or less deliberate ways to meet certain (communicative or aesthetic) needs. This should be understood as an ongoing process, and is identified by \citet{Kloss1967} as a diachronic process of language \isi{Ausbau}. It can also be understood as a process of \isi{Verschriftlichung} (see \citealt{Fischer2007}: 37–38), i.e. the elaboration of the written language will lead to register-specific developments, such as more complex syntactic constructions. The functional expansion and subsequent formal elaboration of the written language lead to a split between the spoken and written varieties of a language (see \citealt{Hoder2009}). In societies in which a large part of the population has access to the written language (as in most modern Western societies), phenomena originally considered only as part of the written register can spread to the spoken language \parencites[37]{Fischer2007}[]{Weis2004}. 



In Swedish, the language \isi{Ausbau} process was characterized by influence from other languages. The earliest longer texts in \ili{Old Swedish} written with the \ili{Latin} alphabet are the \is{provincial law}provincial laws (from the 13\textsuperscript{th} century onwards), but it was mainly in the context of \isi{Vadstena} Abbey that the shaping of \ili{Old Swedish} into a literary language took place. Since \ili{Latin} was the predominant language in this environment, the Swedish texts produced there were often heavily influenced by \ili{Latin}, for instance regarding \isi{relativization} and other syntactic patterns (see further \citealt{Hoder2009, Hoder2010, Wollin1981, Wollin1983}). The bulk of the texts were also translations or paraphrases of \ili{Latin} originals (i.e. not original works in Swedish). Many of these \ili{Latin}-inspired constructions survived into \ili{Early Modern Swedish} through the Reformation \isi{Bible} translations, which were influenced by the \isi{Vadstena} tradition (see \citealt{Stahle1970, Teleman2002, Teleman2003Swedish}). Outside of the religious context, another language was in more direct contact with Swedish, namely \ili{Low German}. During the Middle Ages, the \isi{Hanseatic League} had a strong impact on the Scandinavian countries, in particular in urban settings, where a large part of the population were \ili{Low German} merchants. Unlike \ili{Latin}, the \ili{Low German} influence was not limited to the written language, but also affected the spoken language of the cities (see \citealt{Braunmuller1997,Braunmuller2005,}). The influence can be observed on all linguistic levels, including numerous grammatical loans, for example conjunctions such as \textit{men} ‘but’, \textit{samt} ‘and’, auxiliaries such as \textit{måste} ‘must’, \textit{bliva} ‘become’, and derivational morphology such as \textit{be}{}-, \textit{för}{}-. (See for example \citealt{Braunmuller2005} for an overview of \ili{Low German} (and \ili{Latin}) influence on Swedish.)


During the Early Modern era (i.e. post-Reformation), written Swedish was used in more contexts (for instance in private letters, diaries, science reports, novels, etc.) and no longer mainly in religious or administrational contexts. The Reformation in the first half of the 16\textsuperscript{th} century marked the end of the text production within the monasteries, since they were closed down. Instead, original production in Swedish increased, and the \ili{Latin} influence decreased. Simultaneously, the importance of the \isi{Hanseatic League} diminished, thus leading to less contact between Swedish and \ili{Low German}. Instead, \ili{High German} became an important language in Sweden, affecting the emerging written standard, not least with regard to the \isi{Bible} translation in 1541. The \ili{High German} influence was not only orthographic and lexical, but also syntactic. For example, it has been suggested that the increasing OV frequency of the time \citep{Petzell2011} and so-called finite \textit{ha}{}-drop in subordinate clauses (see \citealt{Johannisson1945}, who first made the claim, and \citealt{Backstrom2019}, who puts it to the test) are both due to contact with \ili{High German} (see also \citealt{Braunmuller2005} for an overview). \ili{High German} continued to be an important language in Sweden throughout the Early and \ili{Late Modern Swedish} period, but from the 17\textsuperscript{th} century onwards, Swedish was also affected by \ili{French}. The \ili{French} influence, however, was not as far reaching as the \ili{High German} one, affecting mainly the lexicon (see \citealt{Gellerstam2005}). By the end of the 18\textsuperscript{th} century, a more or less standardized written Swedish was in place (\citealt{Teleman2002}, see also \citetv{chapters/01}).


\section{Background}\label{sec:kalm:3}


In this section, I give a short introduction to the historical origins of the infinitive and the \isi{infinitive marker} with special focus on North \ili{Germanic}. Thereafter, I discuss \isi{adverbial} infinitives from a typological perspective, comparing present-day Swedish with some other European languages, primarily \ili{Romance} languages and \ili{German}. I also summarize the historical development of \isi{adverbial} infinitives in \ili{Romance} languages. 


\subsection{The Germanic infinitive: A short introduction}\label{sec:kalm:3.1}



In Swedish, as in many other European languages, the infinitive is considered the basic form of the verb, used for example as the dictionary entry. Unlike in \ili{English}, where the infinitive is identical to the verb stem, the Swedish infinitive is an inflected form, at least when the verb stem is consonantal.\footnote{There are North \ili{Germanic} dialects in which the verb stem is identical to the infinitive, as in \ili{English} (see \citealt{Delsing2014Verbsystemet}).}  The Swedish infinitive is always vocalic; if the stem ends with a consonant, an \textit{a}{}-suffix is added. However, if the verb stem is vocalic, there is no formal difference between the stem and the infinitive, as illustrated in \REF{ex:kalm:2} below. 


\ea\label{ex:kalm:2}
\begin{tabular}[t]{llll}
{} & stem & infinitive & {}\\
a. & \textit{läs} & \textit{läs-a} & ‘read’\\
{} & \textit{köp} & \textit{köp-a} & ‘buy’\\	
b. & \textit{sy} & \textit{sy}{}-${\emptyset}$ & ‘sew’\\
{} & \textit{hoppa} & \textit{hoppa}{}-${\emptyset}$ & ‘jump’\\
\end{tabular}
\z

From a historical perspective, the verbal status of the infinitive is less obvious. The characteristic \textit{a}{}-ending of the Swedish infinitive is the remnant of an Old \ili{Germanic} derivational suffix -\textit{an(a)} used in the formation of verbal nouns (\citealt[193]{FalkTorp1900}). Originally, the infinitive was thus nominal rather than verbal, and as such it took case endings like any other nominal element. Examples of this can be found in West \ili{Germanic}, but not in \ili{Ancient Nordic} \citep[205]{Prokosch1939}. Over time, the deverbal noun was reanalysed as a verb, as the derivational affix was grammaticalized into an inflectional ending. The -\textit{n} was lost at an early stage in North \ili{Germanic} languages but is preserved in, for example, \ili{German} and \ili{Dutch} (see \citealt[636]{Noreen1898}). The infinitive is thus common to all the \ili{Germanic} languages. There are infinitives in other \ili{Indo-European} languages as well, for example in \ili{Romance}, but these are etymologically distinct from their \ili{Germanic} counterparts (see \citealt[205]{Prokosch1939}, \citealt[193]{FalkTorp1900}).



In \ili{Germanic} (and also in \ili{Romance}), the infinitive came to be associated with prepositions that were later grammaticalized into infinitive markers. While there is a common \ili{Germanic} infinitive, the \isi{infinitival preposition} is different in the North \ili{Germanic} languages than in the \ili{Continental Germanic} languages (including \ili{English}). The Nordic preposition is \textit{at} (‘at’, ‘by’), identical to the present-day Swedish preposition \textit{åt}, and with the same semantics (see \citealt{Hellquist1948}), and the Proto-\ili{Germanic} is *\textit{tō} (‘at’, ‘by’), thus corresponding to \ili{Dutch} \textit{te}, Eng. \textit{to}, Ger. \textit{zu}, etc. Note that the infinitive markers in \ili{Romance} languages have similar semantics, for example \ili{French} \textit{de} and \textit{à} (see \citealt[15]{Beckman1934}).\footnote{In \ili{Romance}, the infinitive markers seem to retain more of their prepositional status than in \ili{Germanic}. See \sectref{sec:kalm:3.2} below.}  With the deverbal noun as a \isi{complement}, the prepositional phrase was used to express allative or \isi{locative} meanings. As the nominal status of the infinitive was gradually lost, and it was reanalysed as a verb, the prepositional infinitives came to express \is{purposive (infinitive)}purposive meanings instead. This can be understood as a \isi{semantic change} in which the prepositional phrase first expressed a concrete, spatial \isi{goal} and later a more abstract \isi{goal} (see \citealt{Los2005}). In \ili{Gothic}, for example, there is a regular system in which the \isi{infinitival preposition} \textit{du} (optionally) can be used in \is{purposive (infinitive)}purposive infinitives, while non-\is{purposive (infinitive)}purposive infinitives are always bare, i.e. appear without the \isi{infinitival preposition}. A few examples are given in \REF{ex:kalm:3} below. The distribution of the Old \ili{High German} \textit{ze} seems to be similar to that of \ili{Gothic} \textit{du} \citep{Haspelmath1989}. 

\ea
\label{ex:kalm:3}
\ea \is{purposive (infinitive)}Purposive\\\label{ex:kalm:3a}
\gll sat du aíthron (\ili{Gothic})\\
sit.\textsc{pst.3sg} to pray.\textsc{inf}\\
\glt ‘he sat down to pray’ (from \citealt{Wright1954}: 193–194) 

\ex Object\\\label{ex:kalm:3b}
\gll othedun faríhnan ina (\ili{Gothic})\\
fear\textsc{.pst.3pl} ask.\textsc{inf} him.\textsc{acc}\\
\glt ‘they feared to ask him’ (from \citealt{Wright1954}: 193–194)

\ex Subject\\ \label{ex:kalm:3c}
\gll ni godh ist niman hláif barne (\ili{Gothic})\\
not good be.\textsc{prs.3sg} take.\textsc{inf} bread.\textsc{acc} child.\textsc{pl.gen}\\
\glt‘it is not good to take the bread of the children’ (from \citealt{Wright1954}: 193–194)
\z
\z

Over time, the \isi{infinitival preposition} spread from \is{purposive (infinitive)}purposive to non-\is{purposive (infinitive)}purposive contexts in all of the \ili{Germanic} languages. The results of this process are similar across \ili{Germanic} languages, but there is also variation between the languages, even within the North \ili{Germanic} branch (see \citealt{Haspelmath1989,Los2005,Kalm2016Prepositioner, Kalm2016Satsekvivalenta,Kalm2019}). The final stages of the spread of infinitival prepositions to non-\is{purposive (infinitive)}purposive contexts can partly be traced in \ili{Early Old Swedish} (EOS, ca. 1225–1375). In many of the earliest Swedish texts (i.e. the \is{provincial law}provincial laws from the 13\textsuperscript{th} century), \textit{at} is regularly used in \is{purposive (infinitive)}purposive infinitives, while its use outside of these contexts is highly irregular (\citealt{Kalm2016Satsekvivalenta}: 186; see also \citealt{Kalm2019} for a comparison with Old \ili{Danish} and Old Gutnish). A few examples are given in \REF{ex:kalm:4}. Note that present-day Swedish requires the \isi{infinitive marker} in all of the contexts exemplified.\pagebreak


\ea
\label{ex:kalm:4}
\ea \is{purposive (infinitive)}Purposive\\\label{ex:kalm:4a}
\gll Combær han til at köpæ iorþ\\ 
comes he there \textsc{im} buy.\textsc{inf} soil\\
\glt‘He comes there to buy land’ (\isi{EWL}, EOS) 

\ex \is{complement}Complement to \isi{adjective}\\\label{ex:kalm:4b}
\gll siþæn ær han skyldughær kunungær wæræ\\
 then is he obliged king be.\textsc{inf}\\
\glt ‘then he is obliged to be king’ (UL, EOS)
 
\ex Subject\\\label{ex:kalm:4c}
\gll Nu ær þæt klokkarans skuld. bæra bok ok stol i sokn\\
now is it clerc.\textsc{def.gen} obligation carry.\textsc{inf} book and stole in parish.\textsc{def}\\
\glt ‘Now it is the liability of the clerk to bring the \isi{Bible} and the stole to the parish’ (ÖgL, EOS)
\z 
\z 



The use of the \isi{infinitival preposition}/\isi{infinitive marker} stabilized during the \ili{Old Swedish} period, and by the beginning of the Early Modern period (the 16\textsuperscript{th} century onwards), its distribution was generally the same as in present-day Swedish (see however \citealt{Hellquist1902}: 194–195 for some exceptions to this). The prepositional status of the \isi{infinitive marker} was gradually lost during the \ili{Old Swedish} period (see further \citealt{Kalm2016Satsekvivalenta}: 195–199). 


\subsection{Typological and diachronic perspectives on adverbial infinitives}\label{sec:kalm:3.2}

As mentioned earlier, the historical development of \isi{adverbial} constructions has not been a major focus in Scandinavian linguistics. In part, this is probably due to the difficulties associated with defining the \isi{adverbial} category as such; \isi{adverbial} notions can be expressed in various ways (see \citealt[3]{Van_der_auwera1998}). From a typological point of view, it can be noted that \isi{adverbial} clauses (finite and non-finite) are common in the languages of Europe, although there is substantial variation with regard to the exact syntactic expression (see \citealt{Hengenveld1998} for a comprehensive overview). Focusing on \isi{adverbial} infinitives, we can note that many of the \isi{adverbial} notions that are expressed using prepositional infinitives in Swedish correspond to prepositional participles in \ili{English} (see the examples in \REF{ex:kalm:1} above and their respective translations into \ili{English}). Prepositional infinitives can be used in, for instance, \ili{German} and \ili{French} to express some of these notions. See \REF{tab:kalm:1} for examples. 



\ea Examples of prepositional infinitives in \ili{French} and \ili{German}.\label{tab:kalm:1}
\ea \is{purposive (infinitive)}Purposive 
\ea \ili{German}\\
   \gll Sie   isst   viel,   um   Gewicht   zu   gewinnen.\\
        She  eats  a.lot   for   weight     \textsc{im}   win.\textsc{inf}\\
   \glt ‘She eats a lot in order to gain weight’\\
\ex \ili{French}\\
    \gll Elle   mange   beaucoup pour   prendre   du   poids.\\
    she   eats   a.lot for   gain.\textsc{inf}   \textsc{prep.art}   weight \\
\glt ‘She eats a lot in order to gain weight.’\\
\z
\ex \is{abessive (infinitive)}Abessive 
\ea \ili{German}\\
\gll Er   ging,   ohne     etwas     zu   sagen.\\
     he   left   without   something   \textsc{im}   say.\textsc{inf}\\
\glt ‘He left without saying anything.’
\ex \ili{French}\\
\gll Il   est   sorti   sans     rien     dire.\\
     he   is   left   without   nothing   say.\textsc{inf}\\
\glt ‘He left without saying anything.’
\z
\ex \is{standardization}Standardization
\ea \ili{German}\\
\gll Er   blieb   vor     dem   Fernseher,   anstatt mit   seinen   Freunden   aus-zu-gehen.\\
     he   stayed   in.front.of   \textsc{art}   TV     instead.of with   his   friends   out-\textsc{im}{}-go.\textsc{inf}\\
\glt ‘He stayed in front of the TV instead of going out with his friends.’
\ex \ili{French}\\
\gll Il   est   resté   devant     la   télé au   lieu   de   sortir     avec   ses   amis.\\
     he   is   stayed   in.front.of   \textsc{art}   TV in   stead   of   going.out.\textsc{inf}    with   his   friends\\
\glt ‘He stayed in front of the TV instead of going out with his friends.’
\z
\ex Temporal (anteriority)
\ea \ili{German}\\
    –
\ex \ili{French}\\
\gll Après   avoir     réfléchi,   elle   lui   répondit.\\
after   have.\textsc{inf}   reflected   she   him   answered\\
\glt ‘After having reflected, she answered him.’\\
\z
\z
\z


Note that the prepositional infinitives in \ili{French} normally do not contain the equivalent of an \isi{infinitive marker} (e.g. \textit{att} in Swedish). In \ili{German}, however, \textit{zu} is used in a similar way to its Swedish counterpart. This is probably due to the fact that the infinitive markers of \ili{Romance} have retained their prepositional status to a greater extent than in the \ili{Germanic} languages. As is evident from the examples above, there is no \isi{perfect} overlap between the languages, i.e. there are differences in the range of prepositional infinitives available (see \citealt{Hengenveld1998}). Unlike in Swedish, temporal posteriority may be expressed using a \isi{prepositional infinitive} in \ili{French} (see \ref{ex:kalm:5a} below). In Swedish, this would correspond to a finite clause. In some \ili{Romance} languages, such as \ili{Spanish} and \ili{Portuguese}, there is also a \is{concessive (infinitive)}concessive infinitive. This would also normally correspond to a finite clause in Swedish. Examples of \ili{Romance} prepositional infinitives expressing temporal posteriority and \is{concessive (infinitive)}concessive meaning are given in \REF{ex:kalm:5a} and \REF{ex:kalm:5b} respectively. 


\ea
\label{ex:kalm:5}
\ea \ili{French}\label{ex:kalm:5a}\\ 
\gll Il faut réfléchir avant de parler.\\
\textsc{indf} must reflect.\textsc{inf} before \textsc{prep} speak.\textsc{inf}\\
\glt ‘One must reflect before speaking.’ (from \citealt{Grevisse1993}: 1298) 

\ex \ili{Spanish}\label{ex:kalm:5b}\\ 
\gll Ambos procesos requieren de las mismas técnicas a pesar de ser totalmente diferentes.\\
both processes require \textsc{prep} \textsc{art} same techniques \textsc{prep} despite \textsc{prep} be.\textsc{inf} totally different\\
\glt ‘Both processes require the same techniques even though they are totally different.’ (from \citealt{Schulte2007What}: 533)
\z 
\z 



Conversely, the Swedish \is{instrumental (infinitive)}instrumental infinitive corresponds to other constructions in both \ili{Romance} languages and \ili{German}. In \ili{French}, a contrastive study shows that \textit{genom} ‘through, by’\,+\,\textit{att}\,+\,infinitive (i.e. what I have called \is{instrumental (infinitive)}instrumental infinitive; see \REF{ex:kalm:1e} above) generally corresponds to the use of the \textit{gérondif} \citep[128]{Hellqvist2015}. It is thus quite evident that the system of using a \isi{prepositional infinitive} to express certain \isi{adverbial} notions is common in many languages, but the exact range of prepositional infinitives available varies quite a lot between different languages. If \citet{Holm1967} is correct in assuming that \isi{adverbial} infinitives in Swedish developed on the basis of \ili{French} and/or \ili{High German} models, it would be the construction as such (i.e., the use of prepositional infinitives as \isi{adverbial} adjuncts) that was replicated, rather than the individual components. 



\is{adverbial}Adverbial infinitives are not attested in \ili{Latin}, but are innovations in the vernaculars. In his investigation of \isi{adverbial} infinitives in the history of \ili{Spanish}, \ili{Portuguese}, and \ili{Romanian}, \citet{Schulte2007Prepositional,Schulte2007What} shows that prepositional infinitives emerged independently in each of the three languages. In spite of this, the developments show striking similarities, as the relative chronological order in which the different prepositional infinitives appeared is identical. The order is shown in \REF{ex:kalm:6} below. 


\ea \label{ex:kalm:6}
\is{purposive (infinitive)}purposive > \is{abessive (infinitive)}abessive > temporal > \is{substitutive (infinitive)}substitutive > \is{concessive (infinitive)}concessive
\z

The development of \isi{adverbial} infinitives in \ili{Romance} can be understood as a gradual process in which one prepositional construction is attested after the other. It should be noted that \ili{Romanian} has not yet developed a \is{concessive (infinitive)}concessive infinitive, in contrast with \ili{Portuguese} and \ili{Spanish}. \citet{Schulte2007Prepositional} argues that his results show that the emergence of \isi{adverbial} infinitives is in line with general tendencies in how \isi{adverbial} categories evolve in languages (cf. \citealt{Cristofaro2005}; see also \sectref{sec:kalm:5.1}). With regard to Swedish, it can be noted that all of the \isi{adverbial} infinitives in \REF{ex:kalm:6} are attested, with the exception of \is{concessive (infinitive)}concessive infinitives. At least, this is the impression given by the Swedish Academy grammar (\citealt{TelemanEtAl1999}/3: 589–591), in which no mention of \is{concessive (infinitive)}concessive infinitives is made. Nevertheless, it is possible to find examples of what appear to be \is{concessive (infinitive)}concessive infinitives, mainly in informal Swedish. The following examples are excerpted using the present-day Swedish corpora in the corpus infrastructure \isi{Korp} \citep{BorinEtAl2012}. 


\ea
\label{ex:kalm:7}
\ea  \label{ex:kalm:7a}
\gll jag blir (sur) på nära och kära som trots att ha verkat vara okej med {instruktionen …} kräver att jag ska vara anträffbar\\
I become sulky at close and dear that despite \textsc{im} have.\textsc{inf} seemed be.\textsc{inf} okay with instruction.\textsc{def} demand that I shall be available\\ 
\glt ‘I get (sulky) at my loved ones who even though they seem to be okay with the instruction … demand that I be available’ (Blogg 2007)

\ex  \label{ex:kalm:7b}
\gll Men, det känns ändå kul att belönas med drygt 65\% trots att ha ridit en ritt jag inte var nöjd med.\\
but it feels still fun to reward.\textsc{inf.pass} with fully 65\% despite \textsc{im} have.\textsc{inf} ridden a ride I not was pleased with\\ 
\glt ‘But, it is still fun to be rewarded with slightly more than 65\% even though I had ridden a ride I was not satisfied with.’ (Blogg 2016)
\z 
\z 



To sum up, \isi{adverbial} infinitives of different sorts exist in many European languages, including \ili{French} and \ili{German}. It has been suggested that the emergence of \isi{adverbial} infinitives in \ili{Romance} follows a general pattern of how \isi{adverbial} categories evolve in languages. An interesting \isi{question} is therefore whether this analysis can also be extended to Swedish. In the next section, the empirical investigation of \isi{adverbial} infinitives in Swedish and \ili{Övdalian} will be presented. 


\section{Adverbial infinitives in Swedish}\label{sec:kalm:4}
In this section, I present the results of the empirical investigation of \isi{adverbial} infinitives in the history of Swedish and in the \isi{traditional dialect} \ili{Övdalian}. Since \is{purposive (infinitive)}purposive infinitives have a somewhat separate history from the other \isi{adverbial} infinitives, as they are attested even in \ili{Old Swedish}, they are discussed separately, in \sectref{sec:kalm:4.1}.


\subsection{Purposive infinitives}\label{sec:kalm:4.1}
\is{purposive (infinitive)}Purposive infinitives are common throughout the history of Swedish, but their form has varied substantially over time. In present-day Swedish, \is{purposive (infinitive)}purposive infinitives are normally introduced with the preposition \textit{för} ‘for’. In specific contexts, where the \is{purposive (infinitive)}purposive reading of the infinitive is implied, it is possible to omit the preposition. An example of such a context is given in \REF{ex:kalm:8a} below, where the matrix verb is \textit{skicka} ‘send’. In \REF{ex:kalm:8b}, the preposition is obligatory since it is not a “\is{purposive (infinitive)}purposive” context (see \citealt{Kalm2016Satsekvivalenta}: 137–138). Note that both constructions are \is{purposive (infinitive)}purposive, the difference being that the preposition \textit{för} is obligatory only in \REF{ex:kalm:8b}.


\ea \ili{Present-day Swedish}\label{ex:kalm:8}
\ea \label{ex:kalm:8a}
\gll Vi skickade honom (för) att köpa mjölk.\\
we sent him \textsc{prep} \textsc{im} buy.\textsc{inf} milk\\
\glt ‘We sent him to buy milk.’

\ex  \label{ex:kalm:8b}
\gll Jag satte mig på andra sidan gången för att inte vara i vägen.\\
 I sat \textsc{refl} at other side.\textsc{def} aisle.\textsc{def} \textsc{prep} \textsc{im} not be.\textsc{inf} in way.\textsc{def}\\
\glt ‘I sat down on the other side of the aisle in order not to be in the way.’
\z 
\z 


In earlier stages of Swedish, the form of the \is{purposive (infinitive)}purposive infinitive varied both diachronically and synchronically (depending on the semantic context of the infinitive clause). In the earliest \ili{Old Swedish} texts (i.e. the \is{provincial law}provincial laws from the 13\textsuperscript{th} century), \is{purposive (infinitive)}purposive infinitives were normally not prepositional (see \citealt{Kalm2016Satsekvivalenta}: 120). As mentioned in \sectref{sec:kalm:3.1} above, the \isi{infinitive marker} \textit{at} spread from \is{purposive (infinitive)}purposive to non-\is{purposive (infinitive)}purposive contexts in the 13\textsuperscript{th} century. A possible explanation as to why \is{purposive (infinitive)}purposive infinitives were not prepositional in the earliest or most archaic time might be that \textit{at} itself was considered a marker of the \is{purposive (infinitive)}purposive reading and therefore a preposition was not needed. Given the etymological status of \textit{at} as a preposition, it is also possible that it maintained its prepositional status at the time, and that it could not be introduced by yet another preposition. 



Leaving the categorical status of the \isi{infinitive marker} aside, the use of \textit{at} in \is{purposive (infinitive)}purposive infinitives was very common in the earliest stages of \ili{Old Swedish}. It was not completely obligatory, however, and it is possible to find examples of bare infinitives with a \is{purposive (infinitive)}purposive reading.\footnote{It can be noted that in earlier stages of \ili{Germanic} languages, the infinitival ending was a case suffix, representing the \isi{accusative}. Since the \isi{accusative} itself could denote a \isi{goal}, neither the \isi{infinitival preposition} (the \isi{infinitive marker}) nor a preposition was necessary (see \citealt{Haspelmath1989,Jeffers1975}).} Interestingly, this use of the bare infinitive seems to be restricted to contexts in which the \is{purposive (infinitive)}purposive reading of the infinitive clause is implied, typically in combination with verbs of motion, i.e. similar to the contexts in which the preposition \textit{för} ‘for’ can be omitted in present-day Swedish.\footnote{The bare \is{purposive (infinitive)}purposive infinitive in combination with verbs of motion is also attested in Estonian Swedish (\citealt{Lagman1958}: 88–89; see also \citealt{Jorgensen1970}: 38).}  Below, \REF{ex:kalm:9a} is a typical example of a \is{purposive (infinitive)}purposive \textit{at}-infinitive from the 13\textsuperscript{th} century, and \REF{ex:kalm:9b} shows a bare infinitive with a \is{purposive (infinitive)}purposive reading. Note that the verb \textit{sænda} ‘send’ in \REF{ex:kalm:9b} is semantically identical to the present-day Swedish verb in \REF{ex:kalm:8a} above.\largerpage

\ea
\label{ex:kalm:9}
\ea \label{ex:kalm:9a}
\gll han ær i. sokn farin siukum. at hiælpæ.\\
 he is in parish gone sick \textsc{im} help.\textsc{inf}\\ 
\glt ‘he went to the city to help the sick’ (\isi{EWL}, EOS)

\ex \label{ex:kalm:9b}
\gll han sænde sina dicipulos viþa vm væruldena pradica.\\ 
 he sent his disciples wide around world.\textsc{def} preach.\textsc{inf}\\
\glt ‘he sent his disciples around the world to preach’ (Leg, EOS)
\z 
\z 

In the \ili{Late Old Swedish} period (LOS, ca. 1375–1526), it became increasingly common to explicitly mark the \is{purposive (infinitive)}purposive semantics of the infinitive using the preposition \textit{til} ‘to’, possibly with inspiration from \ili{Low German} (see also \sectref{sec:kalm:5.1}). The increasing use of prepositional \is{purposive (infinitive)}purposive infinitives seems to correlate with the distributional expansion of \textit{at} through which the connection between \textit{at} and the \is{purposive (infinitive)}purposive meaning is weakened. Two examples of \textit{til}{}-governed \is{purposive (infinitive)}purposive \textit{at}{}-infinitives from the \ili{Late Old Swedish} period are given in \REF{ex:kalm:10} below. 

\ea
\label{ex:kalm:10}
\ea  \label{ex:kalm:10a}
\gll Til mykla oglädhi war thu här komin aff franz til at göra nakan wanhedher konungx döttrom\\
for much unhappiness were you here come of France to \textsc{im} make.\textsc{inf} some dishonor king.\textsc{gen} daughters\\
\glt ‘to our great displeasure, you came here from France to dishonor the daughters of the king’ (KM, LOS)

\ex\label{ex:kalm:10b}
\gll jak nidherfoor til Colne til ath thär faa reliquias\\
 I down.traveled to Cologne to \textsc{im} there get.\textsc{inf} relics\\
\glt ‘I travelled to Cologne to get relics there’ (Måns, LOS, p. 332)
\z 
\z 


During the Early Modern period (ca. 1526 onward), there is considerable variation in the form of the \is{purposive (infinitive)}purposive infinitive.\footnote{There is also substantial variation in the form of \is{purposive (infinitive)}purposive infinitives in \ili{German}, both synchronically and diachronically (see \citealt{Demske2011}).}  In addition to the \ili{Old Swedish} patterns, some \is{purposive (infinitive)}purposive infinitives are also introduced by \textit{til} ‘to’ alone, as in \REF{ex:kalm:11a} below, or by \textit{til} ‘to’ in combination with the preposition \textit{för} ‘for’, as in \REF{ex:kalm:11b}. This is a consequence of the \isi{grammaticalization} of \textit{til} (\textit{at}) as an \isi{infinitive marker} during the 16\textsuperscript{th} and 17\textsuperscript{th} centuries (see \cites[]{Kalm2014}[]{Kalm2016Prepositioner}[203–221]{Kalm2016Satsekvivalenta}). In \REF{ex:kalm:11c} there is an example of \textit{til at} as a complex \isi{infinitive marker} in a \is{purposive (infinitive)}purposive infinitive introduced by the preposition \textit{för} ‘for’. Not until the late 18\textsuperscript{th} century was the present-day system with a \textit{för}{}-introduced \textit{at}{}-infinitive established. An early example is given in \REF{ex:kalm:11d}. 


\ea
\label{ex:kalm:11}
\ea  \label{ex:kalm:11a}
\gll Men nu ha vi komme hit te gratulera\\
but now have we come here \textsc{prep}/\textsc{im} congratulate.\textsc{inf}\\
\glt ‘But now we have come to congratulate’ (2cKUSINE, 1791) 
\ex  \label{ex:kalm:11b}
\gll Ä de inte Ni, som kom te mej för te berätta hur Ni har hört\\
is it not you that came to me for \textsc{im} tell.\textsc{inf} how you have heard\\
\glt ‘Is it not you, who came to me, to tell me how you had heard’ (2cKUSINE, 1791)

\ex  \label{ex:kalm:11c}
\gll Häldre ger jag ut min plåt för sådant, än jag går på Operan och knuffas, för til at få se et par illa uphängda Gudar träta ur Dis dure med hwarandra.\\
rather give I out my ticket for such than I go to Opera.\textsc{def} and jostle for \textsc{im} \textsc{im} get.\textsc{inf} see.\textsc{inf} a pair poorly hanged gods argue from Dis dure with each.other\\
\glt ‘I would rather give away my ticket than go to the Opera and hustle in order to see a pair of poorly hanging gods arguing in D-sharp major with each other.’ (2aSTERBH, 1776)

\ex \label{ex:kalm:11d}
\gll Sedan Danske skeep flotten had liggat söder i Callmars sundh och blockuerat i 2 veckor så gick han åth Gottland för att forsorga sigh medh wedh och proviant.\\
since \ili{Danish} ship fleet.\textsc{def} had layed south in Kalmar.\textsc{gen} channel and blocked in 2 weeks so went he to Gotland for \textsc{im} support.\textsc{inf} \textsc{refl} with firewood and provisions\\ 
\glt ‘Since the \ili{Danish} fleet had been in the south, blocking the channel of Kalmar for two weeks, he went to Gotland to get supplies of firewood and provisions.’ (Bol, ca. 1697, p. 78)
\z 
\z 


\begin{sloppypar}
The form of the \is{purposive (infinitive)}purposive infinitive has evidently varied substantially throughout the history of Swedish: it has been bare, and introduced by different infinitive markers (\textit{at}, \textit{til at}, \textit{til}) and prepositions (\textit{til} ‘to’\textit{, för} ‘for’). In \tabref{tab:kalm:2}, I give an overview of the proportions of the \is{purposive (infinitive)}purposive constructions in the history of Swedish. 
\end{sloppypar}


\begin{table}
\caption{Distribution of purposive infinitives in the history of Swedish, based on data from \citet{Kalm2016Satsekvivalenta} and the corpus of Swedish drama dialogue ($N = 1230$).\label{tab:kalm:2}}
\begin{tabular}{lrrrrrr} 
\lsptoprule
& bare inf. & \textit{at}  & \textit{til at}  & \textit{til}  & \textit{för til at}  & \textit{för at} \\\midrule
1225–1300 & 22\% & 78\% & – & – & – & –\\
1301–1375 & 8\% & 85\% & 8\% & – & – & –\\
1376–1450 & 2\% & 84\% & 14\% & – & – & –\\
1451–1526 & 5\% & 70\% & 23\% & 2\% & – & –\\
1527–1600 & – & 25\% & 74\% & 1\% & – & –\\
1601–1675 & 3\% & 66\% & 30\% & 1\% & – & 1\%\\
1725–1750 & – & 43\% & 55\% & – & – & 2\%\\
1775–1800 & – & 15\% & 1\% & 8\% & 2\% & 74\%\\
1825–1850 & – & 6\% & – & – & – & 94\%\\
1875–1900 & – & 2\% & – & – & – & 98\%\\
1925–1950 & – & 1\% & – & – & – & 99\%\\
1975–2000 & – & 2\% & – & – & – & 98\%\\
\lspbottomrule
\end{tabular}
\end{table}


The data in \tabref{tab:kalm:2} show that the \is{purposive (infinitive)}purposive infinitive most commonly takes the form of a non-prepositional \textit{at}{}-infinitive in \ili{Early Old Swedish} (i.e. 1225–1300 and 1301–1375). There are also examples of bare infinitives and, from the 14\textsuperscript{th} century onwards, \textit{til}{}-introduced \textit{at-}infinitives. The latter became more common over time, and in the 16\textsuperscript{th} century, this was the dominant pattern.\footnote{This is partly due to the fact that \textit{til at} was grammaticalized into a complex \isi{infinitive marker} during the 16\textsuperscript{th} century. The use of \textit{til at} was not restricted to \is{purposive (infinitive)}purposive contexts, but used more generally by certain writers in the 16\textsuperscript{th}–18\textsuperscript{th} centuries. See \textcites[]{Kalm2016Prepositioner}[203–221]{Kalm2016Satsekvivalenta} for further details.} The variation in form increased during the 17\textsuperscript{th} and early 18\textsuperscript{th} centuries, but decreased during the second half of the 18\textsuperscript{th} century, with the rather sudden shift from \textit{til} ‘to’ to \textit{för} ‘for’ as the general \is{purposive (infinitive)}purposive preposition. In the early 18\textsuperscript{th} century, \textit{för} ‘for’ was quite rare in \is{purposive (infinitive)}purposive contexts, while it dominated the \is{purposive (infinitive)}purposive infinitives in the latter part of the century. It is not quite clear what motivation there was for this shift in \is{purposive (infinitive)}purposive preposition.\footnote{As pointed out by an anonymous reviewer, one possibility could be influence from the \is{purposive (infinitive)}purposive \textit{pour} ‘for’\,+\,infinitive construction in \ili{French}.}  Other constructions (such as \textit{för te}\,+\,infinitive or \textit{för til at}\,+\,infinitive) must be considered quite marginal in comparison with these more frequent patterns. From the 19\textsuperscript{th} century onwards, only \textit{för}{}-introduced \textit{at}{}-infinitives and bare \textit{at}{}-infinitives are attested with \is{purposive (infinitive)}purposive meaning. 


\subsection{Other adverbial infinitives}\label{sec:kalm:4.2}

\is{adverbial}Adverbial infinitives are prepositional in present-day Swedish, with the above-mentioned exceptions of certain \is{purposive (infinitive)}purposive infinitives (see example \REF{ex:kalm:8} above), and the \isi{infinitive marker} \textit{att} is obligatory. Prepositional infinitives were already common in \ili{Old Swedish}, but the prepositions were then part of the lexical construction of a governing verb, noun, or \isi{adjective}, i.e. they did not introduce \isi{adverbial} adjuncts. The earliest examples of \isi{adverbial} (\isi{adjunct}) infinitives appear in texts from the second half of the 17\textsuperscript{th} century. First attested are \is{abessive (infinitive)}abessive infinitives in a text from 1657, and \is{substitutive (infinitive)}substitutive infinitives in a text from 1675; see \REF{ex:kalm:12} below. 


\ea
\label{ex:kalm:12}
\ea \is{abessive (infinitive)}Abessive\\\label{ex:kalm:12a}
\gll[Då] kom wijd Minans springande en fransos flygandess och ståendes på föttren       mitt ibland Officerarne utan at wara på något sätt skadder.\\
then came at mine.\textsc{def.gen} exploding a Frenchman flying and standing at foot.\textsc{pl.def} midst among officer.\textsc{pl.def} without \textsc{im} be.\textsc{inf} at any way hurt \\ 
\glt ‘[Then] came when the mine exploded a Frenchman flying and standing on his feet in the midst of the officers, without being in any way hurt.’ (Rålamb, 1657)

\ex \is{standardization}Standardization\\\label{ex:kalm:12b}
\gll i stället för at fruckta dem, utbrast han i desse ord: Ju flere Fijender ju meer ähra.\\
 in place for to fear.\textsc{inf} them exclaimed he in these words the more enemies the more honor\\
\glt ‘instead of fearing them, he exclaimed these words: The more enemies, the more glory.’ (Mål-roo, 1675)
\z 
\z 


Both \is{abessive (infinitive)}abessive and \is{substitutive (infinitive)}substitutive infinitives seem to have been used only sporadically during the late 17\textsuperscript{th} century, but they rapidly become common in texts from the 18\textsuperscript{th} century, especially the \is{abessive (infinitive)}abessive infinitives (see \citealt[129–130]{Kalm2016Satsekvivalenta}). The first examples of temporal and instrumental infinitives date from much later. The earliest instance of a \isi{temporal infinitive} is from 1779, and the first \is{instrumental (infinitive)}instrumental infinitive appears in 1829; see \REF{ex:kalm:13} below.  

\ea
\label{ex:kalm:13}
\ea Temporal\label{ex:kalm:13a}\\
\gll Då jag efter att ha likafullt gått {ut …} åter tillbaka kom holt jag före att man borde i så critiqva omständigheter ej våga dröja längre utan strax gjöra Revolten.\\
as I after \textsc{im} have.\textsc{inf} nevertheless gone.\textsc{sup} out again back came held I for that one should in such critical circumstances not dare wait longer but soon make revolt.\textsc{def}\\
\glt ‘As I, after nevertheless having gone out, came back, I meant that one should, in such critical conditions, not dare to wait any longer but immediately begin the revolt.’ (Ehrensvärd, 1779, p. 6)

\ex Instrumental\label{ex:kalm:13b}\\
\gll {konsten} segrar icke genom att trotsa, utan genom att följa naturen\\
art.\textsc{def} prevails not through \textsc{im} defy.\textsc{inf} but through \textsc{im} follow.\textsc{inf} nature.\textsc{def}\\
\glt ‘Art will not prevail by defying, but by following nature’ (von Unge, 1829)
\z 
\z 


In a similar way to the situation in \ili{Romance} (\citealt{Schulte2007Prepositional,Schulte2007What}), the development of \isi{adverbial} infinitives in Swedish thus seems to have been a gradual process. Over time, more \isi{adverbial} notions came to be expressed with prepositional infinitives. In \REF{ex:kalm:14} below, I have summarized the order of first appearance for the \isi{adverbial} infinitives investigated in this article. A comparison with the corresponding process in \ili{Romance} (see \REF{ex:kalm:6} above) shows that the processes are similar, but not identical. As is the case in \ili{Romance}, \is{purposive (infinitive)}purposive infinitives are the first to be attested, followed by \is{abessive (infinitive)}abessive infinitives. While \is{substitutive (infinitive)}substitutive infinitives are attested earlier than temporal infinitives in Swedish, it is the opposite order in \ili{Romance}. The \is{instrumental (infinitive)}instrumental infinitive has no equivalent in \ili{Romance}, and there are no examples of \is{concessive (infinitive)}concessive infinitives in the Swedish corpora investigated for this study.\footnote{See, however, the examples from present-day Swedish in \REF{ex:kalm:7} above.}     

\ea  \label{ex:kalm:14}
\gll \is{purposive (infinitive)}purposive >  \is{abessive (infinitive)}abessive  > \is{substitutive (infinitive)}substitutive > temporal > instrumental\\
(1225) {} (1657) {} (1675) {} (1779) {} (1829)\\
\z 



In conclusion, we have seen that it became possible to express certain \isi{adverbial} notions with prepositional infinitives from the second half of the 17\textsuperscript{th} century onwards. The emergence of such \isi{adverbial} infinitives seems to have been a gradual process in the sense that \is{abessive (infinitive)}abessive and \is{substitutive (infinitive)}substitutive infinitives are attested about a hundred years before the first example of temporal infinitives. Instrumental infinitives are attested during the first half of the 19\textsuperscript{th} century. The process shows similarities with the \isi{grammaticalization} of \isi{adverbial} infinitives in \ili{Romance} (see \citealt{Schulte2007Prepositional,Schulte2007What}). 


\subsection{The case of Övdalian}\label{sec:kalm:4.3}


As mentioned above, \citet[27]{Holm1967} claims that \isi{adverbial} (prepositional) infinitives are not attested in traditional dialects. If this is correct, it means that other linguistic resources need to be used to express these \isi{adverbial} notions. As already mentioned, Holm does not present any data to support his claim, and it can also be difficult to find the relevant data. When comparing standardized languages with each other, it can be fruitful to work contrastively with translations between the two languages. This method allows us to find specific constructions in a language and see how they are translated into another language. Dialects and other non-standard varieties are normally spoken, and rarely written down, and the comparison between the variety and the \isi{standard language} is therefore more difficult to conduct, at least when you are interested in the usage of a specific construction. From a Swedish point of view, there is one important exception to this, namely \ili{Övdalian}. 

\ili{Övdalian} (Sw. \textit{älvdalska}) is traditionally considered a Swedish \isi{dialect}, but in recent times there have been attempts to get it acknowledged as a \isi{minority language} in Sweden because of its distance, linguistically speaking, from standard Swedish. \ili{Övdalian} is spoken in the northern parts of the province of Dalarna by around 2,500 people \citep[27]{Garbacz2009}. The variety is characterized both by many archaic traits (such as the persistence of a partial \isi{case system} and subject-verb \isi{inflection}) and by innovations (for example, secondary diphthongs). It has received considerable attention from linguists in the last decades, with the result that it is fairly well described (see for instance \citealt{Garbacz2009,BentzenEtAl2015}). There is also an interest among the speakers of \ili{Övdalian} in preserving the variety.\footnote{There is, for example, an association for the preservation of \ili{Övdalian}, Ulum Dalska, founded in 1985.} As a consequence, \ili{Övdalian} has been codified in dictionaries and grammars in recent years, unlike most other North \ili{Germanic} dialects or non-standard varieties. An orthographic norm is in the process of being established, and there are at the moment a fair number of texts available in \ili{Övdalian}, both original works and translations. 

It is possible that \ili{Övdalian} is going through a process of both \isi{standardization} and \isi{Verschriftlichung} at the moment, but given its relative isolation historically, it has not been influenced by other languages, and not even by standard Swedish to any great extent, it seems. At present, however, it is likely that \ili{Övdalian} is being affected by Swedish, not least since all of its speakers are bilingual. 

For this investigation, I have chosen to use two \ili{Övdalian} translations from Swedish: the novel \textit{Hunden/Rattsjin} ‘The dog’ and the Gospel of John (in Sw. Johannesevangeliet, Joh.; in Övd. Juanneswaundsjilą, Jua.). I have excerpted all of the \isi{adverbial} infinitives in the Swedish texts and then compared them with their respective translations. The aim is to establish whether \ili{Övdalian} uses prepositional infinitives or other linguistic resources to express the notions that are expressed with prepositional infinitives in Swedish. My hope is that this contrastive study will give us clues as to how \isi{adverbial} notions might be expressed in varieties that have not gone through a process of \isi{Verschriftlichung} or language \isi{Ausbau}, including earlier stages of Swedish. That being said, it is important to remember that the \ili{Övdalian} texts might be affected by the Swedish originals. 

There are in total 79 \isi{adverbial} infinitives in the two Swedish texts, evenly distributed between them. An overview of the Swedish data is given in \tabref{tab:kalm:3}. 

\begin{table}
\caption{Adverbial infinitives in the Swedish texts.\label{tab:kalm:3}}
\begin{tabular}{lrrr} 
\lsptoprule
& \textit{Hunden} & \textit{Johannesevangeliet} & Total\\\midrule
\is{purposive (infinitive)}Purposive & 28 & 35 & 63\\
\is{abessive (infinitive)}Abessive & 10 & 1 & 11\\
Instrumental & 1 & 3 & 4\\
Temporal & 0 & 1 & 1\\
Total & 39 & 40 & 79\\
\lspbottomrule
\end{tabular}
\end{table}


The \is{purposive (infinitive)}purposive infinitives are by far the most common. There are only 16 examples of other \isi{adverbial} infinitives in the two texts; \is{substitutive (infinitive)}substitutive infinitives are not attested. All of the \is{purposive (infinitive)}purposive infinitives are introduced by the preposition \textit{för} ‘for’ in the Swedish texts. These are translated using several different constructions in \ili{Övdalian}, including prepositional infinitives. An overview is given in \tabref{tab:kalm:4}. 



\begin{table}
\caption{Övdalian translations of purposive infinitives.\label{tab:kalm:4}}
\begin{tabular}{lrrr} 
\lsptoprule
& \textit{Rattsjin} & \textit{Juanneswaundsjilą} & Total\\\midrule
\textit{fer te}{}-inf. (‘for to’) & 7 & 19 & 26\\
\textit{og} coordination (‘and’) & 11 & 12 & 23\\
\textit{so} clause (‘so’) & 5 & 2 & 7\\
\textit{fer} clause (‘for’) & 5 & 0 & 5\\
bare inf. & 0 & 1 & 1\\
other\footnote{In this one instance, the translation is far from the Swedish original for some reason, and it is difficult to say what part of the translation would correspond to the \is{purposive (infinitive)}purposive infinitive.} & 0 & 1 & 1\\
\lspbottomrule
\end{tabular}
\end{table}

In the reference grammar of \ili{Övdalian} (\citealt{AkerbergNystrom2012}), the authors note that the \is{purposive (infinitive)}purposive \textit{för att}{}-infinitive (‘for to’) is somewhat difficult to translate to \ili{Övdalian}. Apparently, \ili{Övdalian} would prefer to use a construction with the conjunction \textit{og} ‘and’\,+\,infinitive instead (\citealt{AkerbergNystrom2012}: 539). However, in the translations I have investigated, prepositional (\is{purposive (infinitive)}purposive) infinitives are common (26 instances; see \tabref{tab:kalm:4}), and they are used much like in the Swedish original. Note that the \ili{Övdalian} \isi{infinitive marker} is \textit{te} (see \sectref{sec:kalm:4.1} above). In the \is{purposive (infinitive)}purposive infinitives in \REF{ex:kalm:15} and \REF{ex:kalm:16} below, both Swedish and \ili{Övdalian} have infinitives introduced with the preposition \textit{för/fer} ‘for’ and the \isi{infinitive marker} (\textit{att/te}).


\ea
\label{ex:kalm:15}
\ea \ili{Present-day Swedish}\\\label{ex:kalm:15a}
\gll Nu fick han lyfta benen högt för att komma fram i snösörjan.\\
now got he lift.\textsc{inf} leg.\textsc{pl.def} high for \textsc{im} come.\textsc{inf} forward in slush.\textsc{def}\\ 
\glt ‘He now had to raise his leg up high in order to make his way through the slush.’ (Hunden, p. 26)

\ex \ili{Övdalian}\\\label{ex:kalm:15b}
\gll Nu wart an lypt fuätum og fer te tågå sig framm i wåtsn\k{i}uäm.\\
now was he lift feet and for \textsc{im} come.\textsc{inf} \textsc{refl} forward in slush.\textsc{def}\\
\glt ‘He now had to raise his leg up high in order to make his way through the slush.’ (Rattsjin, p. 26)
\z 
\ex
\label{ex:kalm:16}
\ea \ili{Present-day Swedish}\\\label{ex:kalm:16a}
\gll Detta sade de för att sätta honom på prov och få något att anklaga honom för.\\
this said they for \textsc{im} put.\textsc{inf} him at test and get something \textsc{im} accuse.\textsc{inf} him for\\
\glt ‘They were using this \isi{question} as a trap,{~}in order to have a basis for accusing him.’ (Joh. 8:6)

\ex \ili{Övdalian}\\\label{ex:kalm:16b}
\gll Edar sagd dier bar fer tä frest an og fer tä fǫ nod tä klågå ǫ an fer.\\
this said they just for \textsc{im} tempt.\textsc{inf} him and for \textsc{im} get.\textsc{inf} something \textsc{im} complain.\textsc{inf} at him for\\
\glt ‘They were using this \isi{question} as a trap,{~}in order to have a basis for accusing him.’ (Jua. 8:6)
\z 
\z 

\textit{Og}-introduced structures are also common, but they are far from always being non-finite. Rather, \textit{og} commonly introduces a finite clause from which the subject has been omitted, as in example \REF{ex:kalm:17} below: \textit{slätsjer} ‘lick’ is the singular form of the present \isi{tense} of the verb. Instead of a \is{purposive (infinitive)}purposive reading, the construction emphasizes the simultaneity of the two verb actions. In \REF{ex:kalm:18}, however, the \isi{prepositional infinitive} in Swedish is translated using \textit{og}\,+\,the infinitive: \textit{dsjärå} ‘do, make, create’ is an unambiguous infinitive. Because of the \ili{Övdalian} tendency to drop vocalic endings (see \citealt{AkerbergNystrom2012}: 532), it is sometimes difficult to establish whether the verb is finite or non-finite. An example of this can be found in \REF{ex:kalm:19}. There is no way of knowing whether it is the infinitival \textit{a}{}-ending or the past \isi{tense} third person plural \textit{e}{}-ending that is omitted from the form \textit{myöt} ‘meet’. 


\ea
\label{ex:kalm:17}
\ea \ili{Present-day Swedish}\label{ex:kalm:17a}\\
\gll Ibland stannar han för att slicka på skaren.\\
sometimes stops he for \textsc{im} lick.\textsc{inf} at snow.crust.\textsc{def}\\
\glt ‘Sometimes he stops to lick the snow crust.’ (Hunden, p. 13)

\ex \ili{Övdalian}\label{ex:kalm:17b}\\ 
\gll Millumað stanner an og slätsjer skårån liteð.\\
Sometimes stops he and licks snow.crust.\textsc{def} little\\
\glt ‘Sometimes he stops to lick the snow crust a little bit.’ (Rattsjin, p. 13)
\z 
\ex
\label{ex:kalm:18}
\ea \ili{Present-day Swedish}\label{ex:kalm:18a}\\
\gll Skulle jag annars säga att jag går bort för att bereda plats för er?\\
would I otherwise say.\textsc{inf} that I go away for \textsc{im} prepare.\textsc{inf} place for you\\
\glt ‘If that were not so, would I have told you that I am going there{~}to prepare a place for you?’ (Joh. 14:2)

\ex \ili{Övdalian}\label{ex:kalm:18b}\\
\gll Edd ig sagt ellest, ig far dait og dsjärå ruom ad id?\\
had I said otherwise I go there and make.\textsc{inf} room at you\\
 \glt ‘If that were not so, would I have told you that I am going there{~}to prepare a place for you?’ (Jua. 14:2)
\z 
\ex
\label{ex:kalm:19}
\ea \ili{Present-day Swedish}\label{ex:kalm:19a}\\
\gll När folk hörde att han hade gjort detta tecken drog de ut för att möta honom.\\
when people heard that he had made this sign went they out for \textsc{im} meet.\textsc{inf} him\\
\glt ‘Many people, because they had heard that he had performed this sign,{~}went out to meet him.’ (Joh. 12:18)

\ex \ili{Övdalian}\label{ex:kalm:19b}\\
\gll So mes fuotsjed fingg är, an add gart ed teckned, fuor dier aut og myöt onum.\\
so when people got hear he had done that sign went they out and meet.\textsc{inf/pres} him\\
\glt ‘Many people, because they had heard that he had performed this sign,{~}went out to meet him.’ (Jua. 12:18)
\z 
\z 


There are also twelve instances in which \ili{Övdalian} has a \isi{subordinate clause} instead of a \isi{prepositional infinitive}; ten of these examples are found in Rattsjin. According to \citet[491]{AkerbergNystrom2012}, both \textit{fer} and \textit{so} are \is{purposive (infinitive)}purposive subjunctions (but \textit{so} can also be used to introduce resultative clauses). Two examples are given in \REF{ex:kalm:20} and \REF{ex:kalm:21}.  

\ea
\label{ex:kalm:20}
\ea \ili{Present-day Swedish}\label{ex:kalm:20a}\\
\gll En gång hade han plumsat i för att försöka nå dem.\\
one time had he splashed in for \textsc{im} try.\textsc{inf} reach them\\ 
\glt ‘Once, he had splashed in in an attempt to reach them.’ (Hunden, p. 51)

\ex \ili{Övdalian}\label{ex:kalm:20b}\\
\gll Iesn add an pulsað åv auti fer an willd biuäð til kum að diem.\\
once had he splashed off into for he wanted try to come at them\\
\glt ‘Once, he had splashed in in an attempt to reach them.’ (Rattsjin, p. 51)
\z 
\ex
\label{ex:kalm:21}
\ea \ili{Present-day Swedish}\label{ex:kalm:21a}\\ 
\gll Själva stannade de utanför, för att inte bli orena\\
self stayed they outside for \textsc{im} not become.\textsc{inf} unclean\\
\glt ‘To avoid ceremonial uncleanness they did not enter the palace’ (Joh. 18:28)

\ex \ili{Övdalian}\label{ex:kalm:21b}\\
\gll og siuover dsjingg dier it in i hlotted, so dier uld it werd \k{u}oriener.\\
and self went they not in in palace.\textsc{def} so they would not become.\textsc{inf} unclean\\
\glt ‘and to avoid ceremonial uncleanness they did not enter the palace,{~}because they wanted to be able to eat the Passover.’ (Jua. 18:28)
\z 
\z\largerpage[2]


Finally, there is one example of a bare infinitive with a \is{purposive (infinitive)}purposive meaning, reproduced as \REF{ex:kalm:22} below. Interestingly, the bare infinitive appears in what I referred to above as a \is{purposive (infinitive)}purposive context, i.e. in combination with a verb phrase expressing motion. In these contexts, we also find bare infinitives in \ili{Old Swedish}.\footnote{This also seems to be the case in Estonian Swedish (\citealt{Lagman1958}: 88–89).} 


\ea
\label{ex:kalm:22}
\ea \ili{Present-day Swedish}\label{ex:kalm:22a}\\
\gll När Maria hörde det, steg hon strax upp och gick för att möta honom.\\
when Mary heard that went she soon up and went for \textsc{im} meet.\textsc{inf} him\\
\glt ‘When Mary heard this, she got up quickly and went to meet him.’ (Joh. 11:29)

\ex \ili{Övdalian}\label{ex:kalm:22b}\\ 
\gll Snjäst Mari fick är ed, raitt ǫ upp sig og fuor stad myöt onum.\\
when Mary got hear that raised she up \textsc{refl} and went along meet.\textsc{inf} him\\
\glt ‘When Mary heard this, she got up quickly and went to meet him.’ (Jua. 11:29)
\z 
\z 


\is{abessive (infinitive)}Abessive infinitives are relatively common in the Swedish texts, especially in the novel \textit{Hunden}. They are generally not translated with a prepositional construction in \ili{Övdalian}. Instead of using embedding, as in Swedish, the \ili{Övdalian} translators have normally chosen to use a coordinate negated structure instead, as in (\ref{ex:kalm:23}--\ref{ex:kalm:24}) below. On one occasion, the translator used a finite subordinate structure instead; see \REF{ex:kalm:25} below.\largerpage[-1]


\ea
\label{ex:kalm:23}
\ea \ili{Present-day Swedish}\label{ex:kalm:23a}\\
\gll Han följde dem alltid med blicken men låg kvar utan att känna upphetsning.\\
he followed them always with eye.\textsc{def} but stayed put without \textsc{im} feel.\textsc{inf} excitement\\
\glt ‘He followed them always with his eye, but stayed put without feeling excitement.’ (Hunden, p. 52)\\

\ex \ili{Övdalian}\label{ex:kalm:23b}\\ 
\gll An fygd ǫ ðiem min ogum olltiett og låg kwer og wart it ekster.\\
he followed at them with eyes always and stayed put and became not excited\\
\glt ‘He followed them always with his eye, but stayed put without feeling excitement.’ (Rattsjin, p. 52)
\z 
\ex
\label{ex:kalm:24}
\ea \ili{Present-day Swedish}\label{ex:kalm:24a}\\
\gll Han som låg under vindfället lyssnade utan att förstå.\\
he who laid under windfall.\textsc{def} listened without \textsc{im} understand.\textsc{inf}\\
\glt ‘He who was under the windfall listened without understanding.’ (Hunden, p. 82)\\
\ex \ili{Övdalian}\label{ex:kalm:24b}\\ 
\gll An so låg under windfellę lydd men bigript it noð.\\
he who laid under windfall.\textsc{def} listened but understood not anything\\
\glt ‘He who was under the windfall listened without understanding.’ (Rattsjin, p. 82)
\z
\ex
\label{ex:kalm:25}
\ea \ili{Present-day Swedish}\label{ex:kalm:25a}\\
\gll Hunden gick i skogskanten, långa sträckor utan att synas alls.\\
dog.\textsc{def} went in forest.edge.\textsc{def} long distances without \textsc{im} see.\textsc{inf.pass} at.all\\
\glt ‘The dog wandered by the edge of the forest, long distances without revealing himself at all’ (Hunden, p. 109)\\
\ex \ili{Övdalian}\label{ex:kalm:25b}\\ 
\gll Rattsjin dsjikk laungg strettsjur i skuägkantem, so int an syndes noð.\\
dog.\textsc{def} went long distances in forest.edge.\textsc{def} so not he see.\textsc{pst.pass} at.all\\
\glt ‘The dog wandered by the edge of the forest, long distances without revealing himself at all.’ (Rattsjin, p. 109)
\z 
\z 


There are two examples (one from each text) in which the translators use a \isi{prepositional infinitive} just like in the Swedish original. This may be the result of a slavish translation from Swedish. The example from the Gospel of John is given in \REF{ex:kalm:26} below.\pagebreak


\ea
\label{ex:kalm:26}
\ea \ili{Present-day Swedish}\label{ex:kalm:26a}\\
\gll Ty den som Gud har sänt talar Guds ord; Gud ger Anden utan att mäta.\\
for it who God has sent speaks God.\textsc{gen} word God gives spirit.\textsc{def} without \textsc{im} measure.\textsc{inf}\\
\glt ‘For the one whom God has sent speaks the words of God, for God gives the Spirit without limit.’ (Joh. 3:34)\\

\ex \ili{Övdalian}\label{ex:kalm:26b}\\
\gll Fer an so Gud ar stsjickad, an glemer Gudes uord, fer Gud dsjäv Andan autǫ tä mela.\\
for he who God has sent he speaks God.\textsc{gen} word for God gives spirit.\textsc{def} without \textsc{im} measure.\textsc{inf}\\
\glt ‘For the one whom God has sent speaks the words of God, for God gives the Spirit without limit.’ (Jua. 3:34)
\z
\z 


We can therefore conclude that \is{abessive (infinitive)}abessive meanings are normally not expressed through prepositional infinitives in \ili{Övdalian}, but by other means. The most frequent pattern used in the texts investigated is a coordinated finite structure containing a negation, instead of a non-finite embedded structure as in the Swedish original. Even though this is clearly an attempt to convey the same meaning in the translation, it must be noted that the translation has a slightly different meaning than the Swedish original. 



Let us turn next to the four instrumental infinitives found in the Swedish texts. Three of them have been translated using a finite subordinate structure, as in (\ref{ex:kalm:27}–\ref{ex:kalm:28}). In \REF{ex:kalm:27}, the Swedish \is{instrumental (infinitive)}instrumental infinitive has been translated with a temporal \isi{subordinate clause} introduced by the subordinator \textit{dar}, and in \REF{ex:kalm:28} the infinitive is translated with a \textit{bar}-introduced (‘only’, ‘just’) \isi{subordinate clause}. In \REF{ex:kalm:29}, the infinitive has been translated with a \isi{relative clause} instead. These examples suggest that instrumental infinitives are not used in \ili{Övdalian} and that there might not be any specialized means to express instrumental meaning. 


\ea
\label{ex:kalm:27}
\ea \ili{Present-day Swedish}\label{ex:kalm:27a}\\
\gll Mycket folk följde efter, därför att de såg de tecken han gjorde genom att bota de sjuka.\\
 much people followed after because that they saw the signs he did through \textsc{im} cure.\textsc{inf} the sick\\
\glt ‘And a large crowd was following him, because they saw the signs he performed on the sick.’ (Joh. 6:2)\\

\ex \ili{Övdalian}\label{ex:kalm:27b}\\
\gll Ed war mitsjid fuok so fygde, fer dier såg tecknę an garde, dar an buoted diem so war kliener.\\
 it was much people who followed for they say signs.\textsc{def} he did when he cured those who were sick\\
\glt ‘And a large crowd was following him, because they saw the signs that he was doing by curing the sick.’ (Jua. 6:2)
\z
\ex
\label{ex:kalm:28}
\ea \ili{Present-day Swedish}\label{ex:kalm:28a}\\
\gll Det hade hänt mellan några valpar och en grovröstad karl som kunde bli hälften så stor genom att ställa sig på knä.\\
 it had happened between some puppies and a gruff.voiced man who could become half so big through \textsc{im} stand.\textsc{inf} \textsc{refl} on knee\\
\glt ‘It had happened between a couple of puppies and a man with a gruff voice who could halve his size by kneeling.’ (Hunden, p. 86)\\

\ex \ili{Övdalian}\label{ex:kalm:28b}\\ 
\gll Eð add ennt millǫ nogum wepum og ienum gruävröstaðum kalle so wart elptn so stur bar an stelld sig ǫ kninę.\\
it had happened between some puppies and a gruff.voiced man who became half so big only he stood \textsc{refl} on knees\\
\glt ‘It had happened between a couple of puppies and a man with a gruff voice who could halve his size by kneeling.’ (Rattsjin, p. 86)
\z 
\ex
\label{ex:kalm:29}
\ea \ili{Present-day Swedish} \label{ex:kalm:29a}\\
\gll Men dessa har upptecknats för att ni skall tro att Jesus är Messias, Guds son, och för att ni genom att tro skall ha liv i hans namn.\\
but these have record.\textsc{sup.pass} for that you shall believe that Jesus is Messiah God.\textsc{gen} son and for that you through \textsc{im} believe.\textsc{inf} shall have life in his name \\
\glt ‘But these are written so that you may believe that Jesus is the Christ, the Son of God, and that by believing you may have life in his name.’ (Joh. 20:31)

\ex \ili{Övdalian}\label{ex:kalm:29b}\\ 
\gll Men ittad ar uort skrievt, so ulid truo att Iesus ir Messias, Gudes Sun, og so ulid åvå laived i namnę onumes, dar truoid.\\
 but this has become written so you believe that Jesus is Messiah God.\textsc{gen} son and so you have life.\textsc{def} in name his who believes\\
\glt ‘But these are written so that you may believe that Jesus is the Christ, the Son of God, and that by believing you may have life in his name.’ (Jua. 20:31)
\z
\z 

Finally, the only example of a \isi{temporal infinitive} has been translated with a temporal \isi{subordinate clause}: 


\ea
\label{ex:kalm:30}
\ea \ili{Present-day Swedish}\label{ex:kalm:30a}\\ 
\gll Efter att ha sagt detta fortsatte han: …\\
 after \textsc{im} have.\textsc{inf} said this continued he\\
\glt ‘After having said this, he continued: …’ (Joh. 11:11)

\ex \ili{Övdalian}\label{ex:kalm:30b}\\
\gll Mes an add sagt edar, fuortsett an: …\\
 as he had said this continued he\\
\glt ‘As he had said this, he continued: …’ (Jua. 11:11)
\z
\z 


To sum up, the contrastive study shows that Swedish \isi{adverbial} infinitives are normally translated using other constructions in \ili{Övdalian}. \is{purposive (infinitive)}Purposive infinitives, by far the most common construction in the Swedish originals, are translated in several different ways. Often, a \isi{prepositional infinitive} is used, much like in Swedish, but another common strategy is to use constructions with the coordinator \textit{og} instead. To a lesser degree, finite subordinate clauses are used, and in one instance a bare infinitive (in combination with a verb phrase expressing motion). \is{abessive (infinitive)}Abessive infinitives are most often translated with a finite coordinate clause containing a negation. This means that the translators are trying to convey the \is{abessive (infinitive)}abessive meaning using another linguistic construction (coordination instead of embedding). In the case of instrumental infinitives, the translation uses temporal or other subordinate clauses, which leads to a slightly different (and less specific) meaning. The only example of a \isi{temporal infinitive} has been translated using a temporal finite clause in \ili{Övdalian}. 

The results partly support \citegen[27]{Holm1967} claim that prepositional infinitives are not found in traditional dialects. In any case, they are not as frequent as in Swedish, and not all types seem to occur. Since \isi{adverbial} infinitives are not attested in \ili{Old Swedish}, it is possible that similar (or the same) constructions to those that we find in \ili{Övdalian} might have been used in earlier stages of Swedish to express the various \isi{adverbial} notions. 


\subsection{Summary of empirical findings}\label{sec:kalm:4.4}


This investigation of \isi{adverbial} infinitives in the history of Swedish takes \is{purposive (infinitive)}purposive infinitives as its point of departure since they are the only type attested as early as \ili{Old Swedish}. In the earliest stages of Swedish, \is{purposive (infinitive)}purposive infinitives are not introduced by a preposition, like their present-day counterparts. Over time there is an increasing tendency for \is{purposive (infinitive)}purposive infinitives to be prepositional, and by the second half of the 18\textsuperscript{th} century, almost all of them are. The connection between a specific form and a specific meaning is thus strengthened over time. Furthermore, there is a rather sudden change of the \is{purposive (infinitive)}purposive preposition from \textit{til} ‘to’ to \textit{för} ‘for’ during the second half of the 18\textsuperscript{th} century. The reasons for this shift remain unclear. 



Other \isi{adverbial} infinitives are attested from the second half of the 17\textsuperscript{th} century onwards in a gradual process whereby more \isi{adverbial} notions are expressed with prepositional infinitives over time. \is{abessive (infinitive)}Abessive infinitives are attested first in a text from 1657, and slightly later (in a text from 1675) we find the first example of a \is{substitutive (infinitive)}substitutive infinitive. Temporal infinitives are not attested until the late 18\textsuperscript{th} century (first example from 1779), and the first instance of an \is{instrumental (infinitive)}instrumental infinitive is found in a text from 1829. This order of appearance is similar (although not identical) to the corresponding emergence of \isi{adverbial} infinitives in \ili{Romance}. 



In addition to the historical survey of \isi{adverbial} infinitives, I have presented a contrastive investigation of \ili{Övdalian} translations of Swedish texts containing \isi{adverbial} infinitives. It has been claimed that traditional dialects, such as \ili{Övdalian}, do not use prepositional infinitives as \isi{adverbial} adjuncts (see \citealt{Holm1967}: 27). This investigation gives some support to these claims, since the \isi{adverbial} infinitives in the Swedish originals are often translated with other constructions in the \ili{Övdalian} versions. This is especially evident in the case of \is{abessive (infinitive)}abessive infinitives, as the translators use a coordinated negated clause (with a few exceptions). It is possible that \ili{Övdalian} can give us clues as to how these \isi{adverbial} notions were expressed in earlier stages of Swedish, i.e. before the emergence of \isi{adverbial} infinitives. 


\section{Discussion}\label{sec:kalm:5}

In this section, the empirical results from the investigation are discussed. I first discuss the role of \isi{language contact} with regard to \isi{adverbial} infinitives in Swedish. Thereafter, I turn to the \isi{question} of whether \isi{adverbial} infinitives should be seen as part of a process of specialization and elaboration of the written language (i.e. language \isi{Ausbau}). 


\subsection{The role of language contact}\label{sec:kalm:5.1}

Whereas \is{purposive (infinitive)}purposive infinitives are already attested in \ili{Early Old Swedish}, other \isi{adverbial} infinitives did not appear until the Early and \ili{Late Modern Swedish} period. \is{purposive (infinitive)}Purposive infinitives were normally introduced by the \isi{infinitive marker} \textit{at} in \ili{Early Old Swedish}, and since \textit{at} etymologically is a preposition (see \sectref{sec:kalm:3.1} above), it could be argued that the (\is{purposive (infinitive)}purposive) \textit{at}{}-infinitive should be regarded as a prepositional phrase at this time (see \citealt{Kalm2016Satsekvivalenta}). Non-\is{purposive (infinitive)}purposive infinitives were often bare (i.e. not introduced by \textit{at}) in \ili{Early Old Swedish}, but later on, the use of the \textit{at}{}-infinitive spread at the expense of the bare infinitive. By the \ili{Late Old Swedish} period, \textit{at} thus seems to have lost its \is{purposive (infinitive)}purposive semantics and its prepositional status, and the \textit{at}{}-infinitive was used in \is{purposive (infinitive)}purposive as well as non-\is{purposive (infinitive)}purposive contexts. As a consequence of the lost relationship between the \textit{at}{}-infinitive and the \is{purposive (infinitive)}purposive meaning, it became increasingly common for \is{purposive (infinitive)}purposive \textit{at}{}-infinitives to be introduced by the preposition \textit{til}, possibly as a way of explicitly marking the \is{purposive (infinitive)}purposive reading of the infinitive. There is, however, also a general tendency (especially during the \ili{Early Modern Swedish} period) for infinitive clauses to be introduced by \textit{til}, and it has been argued that \textit{til at} was grammaticalized as a complex \isi{infinitive marker} during this time (see \citealt{Kalm2014,Kalm2016Satsekvivalenta}). It is not unlikely that the promotion of \textit{til} during this time was facilitated by \ili{Low German} influence in much the same way as \citet{Nesse2002} suggests that the \isi{infinitive marker} \textit{te} of the \ili{Norwegian} Bergen \isi{dialect} emerged as a result of \isi{language contact} with \ili{Low German}. The \ili{Low German} \textit{te} had a double function as both \isi{infinitive marker} and preposition (just like in present-day \ili{German} and \ili{English}).

The first signs of the modern system of prepositional \isi{adverbial} infinitives appeared by the end of the \ili{Early Modern Swedish} period. By this time, the \ili{Low German} influence on Swedish had already come to an end and it is thus not likely that the appearance of the first \isi{adverbial} infinitives was due to contact with \ili{Low German}. Instead, \citet{Holm1967} suggests that the constructions emerged as a result of \isi{language contact} with \ili{High German} and/or \ili{French}. Of the two languages, at least \ili{High German} seems like a possible candidate for such an influence on Swedish, since we know of other syntactic constructions that arose due to influence from \ili{High German} (see \sectref{sec:kalm:2} above). (As noted, the \ili{French} influence on Swedish that we know of today is lexical rather than syntactic.) If the emergence of \isi{adverbial} infinitives in Swedish was due to \isi{language contact}, this could be understood as a case of contact-induced (or replica) \isi{grammaticalization} (see \citealt{HeineKuteva2003, HeineKuteva2005}) in much the same way as \citet{Nesse2002} accounts for the appearance of the \isi{infinitive marker} \textit{te} in the \ili{Norwegian} Bergen \isi{dialect}. This would mean that the (foreign) system of using prepositional infinitives to express \isi{adverbial} notions was replicated in Swedish. The fact that the \is{instrumental (infinitive)}instrumental infinitive is specifically Swedish (i.e. not attested in \ili{High German}) is not an argument against this analysis, since it is the initiation of the \isi{grammaticalization} process that is contact-dependent, not the entire process. It is thus expected that you would find language-specific development even in cases of contact-induced \isi{grammaticalization}. 



Even though it cannot be ruled out that the emergence of \isi{adverbial} infinitives in Swedish should be understood as a contact phenomenon, it is equally possible that they emerged independently in Swedish in much the same way as has been argued for the \ili{Romance} languages (\citealt{Schulte2007Prepositional, Schulte2007What}). \citet{Schulte2007Prepositional, Schulte2007What} shows that new \isi{adverbial} infinitives appeared in the same relative order in \ili{Spanish}, \ili{Portuguese}, and \ili{Romanian}. The relative order is very similar (albeit not identical) to the corresponding development in Swedish. Compare the \ili{Romance} cline in \REF{ex:kalm:6} and the Swedish in \REF{ex:kalm:14}, repeated as \REF{ex:kalm:31} and \REF{ex:kalm:32} below. 

\ea
\label{ex:kalm:31}
\is{purposive (infinitive)}purposive > \is{abessive (infinitive)}abessive > temporal > \is{substitutive (infinitive)}substitutive > \is{concessive (infinitive)}concessive
\ex
\label{ex:kalm:32}
\is{purposive (infinitive)}purposive > \is{abessive (infinitive)}abessive > \is{substitutive (infinitive)}substitutive > temporal > instrumental
\z

In \ili{Romance}, as in Swedish, \is{purposive (infinitive)}purposive infinitives are attested first, followed by \is{abessive (infinitive)}abessive infinitives. Thereafter, the order differs with regard to temporal and \is{substitutive (infinitive)}substitutive infinitives, and while Swedish developed an \is{instrumental (infinitive)}instrumental infinitive, \ili{Romance} has a \is{concessive (infinitive)}concessive infinitive. Interestingly, there are examples of \is{concessive (infinitive)}concessive infinitives in (informal) present-day Swedish as well (see the examples in \REF{ex:kalm:7} above). Since it is not likely that the appearance of \is{concessive (infinitive)}concessive infinitives in present-day Swedish is due to contact with \ili{Romance} languages, it rather seems like the development follows some general pattern of how \isi{adverbial} constructions evolve in a language. This is also what \citet{Schulte2007Prepositional, Schulte2007What} suggests with regard to \ili{Romance}, and he relates the order to \citegen{Cristofaro2005} typological hierarchy of “deranked” (roughly: non-finite) \isi{adverbial} clauses. According to Cristoforo, there is a general tendency for deranked constructions to be associated first with \is{purposive (infinitive)}purposive meaning (see also \citealt{Haspelmath1989}) and then later with other semantics. The overall result of the study is that there are cross-linguistic patterns as to how \isi{adverbial} (non-finite) clauses evolve in languages, and \citet{Schulte2007Prepositional, Schulte2007What} shows that \ili{Romance} languages fit into this description. This paper suggests that the development of \isi{adverbial} infinitives in Swedish has followed the same general pattern, starting with \is{purposive (infinitive)}purposive infinitives and ending with \is{concessive (infinitive)}concessive infinitives in a similar way to the \ili{Romance} languages.



To sum up, it is possible that \isi{language contact} has affected the emergence of \isi{adverbial} infinitives in Swedish, but it cannot be ruled out that we are dealing with a “natural” evolution of \isi{adverbial} categories. It is likely that the use of \textit{til} in combination with \is{purposive (infinitive)}purposive (and other) infinitival clauses is due to contact with \ili{Low German}, but when it comes to other \isi{adverbial} infinitives, it is less clear that they should be understood as contact phenomena. Since the two explanations are not mutually exclusive, it is possible that the \isi{grammaticalization} was initiated by \isi{language contact}, but that the continuation of the process was independent, following general patterns of how \isi{adverbial} non-finite constructions evolve. In the next section, I will discuss the \isi{adverbial} infinitives in relation to the ongoing specialization of the written language during the Late Old and \ili{Early Modern Swedish} period. 


\subsection{Adverbial infinitives as an Ausbau phenomenon}\label{sec:kalm:5.2}


Simultaneously with the expansion into new domains, a written language normally undergoes formal elaborations of different kinds, visible for instance in syntactic complexity (see \cites[38--39]{Fischer2007}[]{Kloss1967}). During the Middle Ages, when the bulk of the text production in Swedish consisted of translations from other languages, the elaboration of \ili{Old Swedish} into a literary language was conducted under more or less direct influence from the source language, predominantly \ili{Latin} (see \citealt{Wollin1981,Wollin1983,Hoder2009, Hoder2010}). This caused a split between the spoken and written varieties of \ili{Old Swedish}, since the spoken language was not affected by \ili{Latin}. In the emerging cities, it was instead influences from \ili{Low German} that shaped (the spoken and written) Swedish (see \sectref{sec:kalm:2} and \sectref{sec:kalm:5.1}). The increasing use of \isi{participle} constructions in \ili{Late Old Swedish}, for example, is considered to be the result of such a \ili{Latin} influence \citep{Ahlberg1942,Hoder2010}. Some of these \isi{participle} constructions seem to be used to express \isi{adverbial} notions that today could be expressed using prepositional infinitives. In \REF{ex:kalm:33} below, the negated \isi{participle} has an \is{abessive (infinitive)}abessive meaning and would correspond to \textit{utan} ‘without’\,+\,\textit{att}\,+\,infinitive in present-day Swedish. (Note that \ili{English} uses a construction similar to the one found in \ili{Old Swedish}.)


\ea 
\label{ex:kalm:33}
 \gll far {iak …} til iherusalem eyg vitande hwat mik skal ouir koma\\ 
travel I to Jerusalem not knowing what me shall over come.\textsc{inf}\\
 \glt ‘I travel … to Jerusalem not knowing what shall come over me’ (ApG, ca. 1385, p. 164; from \citealt{Ahlberg1942}: 163)
\z


Although these \ili{Latin}-inspired constructions were exclusively part of a written register, i.e. they were not used in the spoken language of the time, many of the \isi{participle} constructions survived in the 16\textsuperscript{th} century Reformation \isi{Bible} translations and remained part of a religious register for centuries \citep[17–19]{Stahle1970}. For the most part, they have not survived into present-day Swedish (neither spoken nor written). 

After the Reformation, text production in Swedish increased rapidly, but not predominantly in religious contexts, as opposed to the situation in the Middle Ages. Texts were also original works in Swedish to a much greater extent than before, i.e. not translations or paraphrases. This means that the elaboration of Swedish during this period was somewhat less directly influenced by other languages than was the case during the late Middle Ages, when the first steps towards a more homogeneous, standardized Swedish were taken. As we have seen, a new \is{abessive (infinitive)}abessive construction appeared during the late 17\textsuperscript{th} century. Instead of participles like the one in \REF{ex:kalm:33}, prepositional infinitives were used to express the same \isi{adverbial} notion, as in \REF{ex:kalm:34} below. The fact that specifically \is{abessive (infinitive)}abessive constructions appeared during periods (and in contexts) in which the written language was being elaborated might suggest that they are part of an ongoing process of \isi{Verschriftlichung}. While the first of these processes (during the late Middle Ages) was interrupted by the Reformation, the second process has continued. 


\ea
\label{ex:kalm:34}
 \gll Skulle det wäl wara möjeligit, at jag fått en stiufmor, utan at weta deraf.\\
should it well be.\textsc{inf} possible that I get.\textsc{sup} a stepmother without \textsc{im} know.\textsc{inf} thereof\\
\glt ‘Would it be possible that I had got a stepmother without knowing anything of it?’ (1dSMUL, 1738)
\z


The emergence of constructions such as the one in \REF{ex:kalm:33} above is an example of how the demands of new genres and text types lead to a specialization of the language. In this case, it is quite obvious that the construction in \isi{question} has emerged with inspiration from \ili{Latin}, but as we have seen in \sectref{sec:kalm:5.1}, this is not necessarily the case with \isi{adverbial} infinitives. In spite of this, both the \ili{Latin}-inspired constructions and the \isi{adverbial} infinitives can be understood as part of the same process of formal elaboration and specialization that a language undergoes when it is being used in new functional contexts. An important piece of evidence supporting the assumption that the development of \isi{adverbial} infinitives is part of a written-language specific development is the fact that prepositional \isi{adverbial} infinitives (with the exception of \is{purposive (infinitive)}purposive infinitives) do not seem to be used in the traditional, spoken \isi{dialect} of \ili{Övdalian}. Since the linguistic distance between Swedish and \ili{Övdalian} is substantial, and has been for a long time, \ili{Övdalian} has not been influenced by written Swedish in the same way as other regional and local varieties of spoken Swedish have been. Its isolated location also means that contact with other languages has been minimal. Until recently, the \isi{dialect} has not been used in writing to any large extent, which means that it has not been through processes of \isi{Verschriftlichung} or language \isi{Ausbau}. The empirical investigation showed that the \ili{Övdalian} translators very often used other constructions to convey the meaning expressed by \isi{adverbial} infinitives in the Swedish originals. Many of the \is{purposive (infinitive)}purposive infinitives were expressed with finite clauses, both coordinated and subordinated. The \is{abessive (infinitive)}abessive infinitives, on the other hand, were almost consistently translated with a negated coordinated structure, like the one in \REF{ex:kalm:24} above, here repeated as \REF{ex:kalm:35}.


\ea
\label{ex:kalm:35}
\ea \ili{Present-day Swedish} \\
\gll Han som låg under vindfället lyssnade utan att förstå.\\
 he who laid under windfall.\textsc{def} listened without \textsc{im} understand.\textsc{inf}\\
\glt ‘He who was under the windfall listened without understanding.’ (Hunden, p. 82)

\ex \ili{Övdalian}\\
\gll An so låg under windfellę lydd men bigript it noð.\\
he who laid under windfall.\textsc{def} listened but understood not anything\\
\glt ‘He who was under the windfall listened without understanding.’ (Rattsjin, p. 82)
\z
\z


Even though both constructions convey roughly the same meaning, the Swedish \isi{prepositional infinitive} can be understood as a grammaticalized construction expressing \is{abessive (infinitive)}abessive meaning, while the coordinated structure in \ili{Övdalian} is not exclusively used with this meaning. Prepositional infinitives thus offer a more precise way of expressing \isi{adverbial} notions, which in earlier stages of the language might have been expressed as in \ili{Övdalian}. This can be understood as a kind of \isi{grammaticalization} where a specific meaning (in this case \is{abessive (infinitive)}abessive) is connected to a specific construction (in this case a \isi{prepositional infinitive}). When it comes to the \is{abessive (infinitive)}abessive meaning, there thus exists a more or less one-to-one relationship between form and function in Swedish (in the sense that the construction \textit{utan} ‘without’\,+\,\textit{att}-infinitive always has \is{abessive (infinitive)}abessive meaning), but this is not the case in \ili{Övdalian}. This becomes even clearer when we compare instrumental infinitives in Swedish with their \ili{Övdalian} translations. In these instances, it rather seems like \ili{Övdalian} lacks the ability to express this precise meaning. Instead, a construction with a temporal meaning is used, as in example \REF{ex:kalm:27b} above. 



There are thus reasons to assume that \isi{adverbial} constructions (for example expressing \is{abessive (infinitive)}abessive meaning) are part of a process of elaboration and specialization of the written language when it is being used in new domains. \is{abessive (infinitive)}Abessive constructions emerged in the late Middle Ages, then clearly under the influence of \ili{Latin}, as well as in the Early Modern period. Both periods are characterized by an increasing use of the written language. The comparison with \ili{Övdalian} shows that specialized \isi{adverbial} constructions are not used in the same way as in present-day Swedish. Since the variety has remained mostly spoken (at least until recently), this supports the assumption that \isi{adverbial} infinitives emerged as part of a written register.  


\section{Summary and conclusions}\label{sec:kalm:6}


This article traces the development of prepositional infinitives functioning as \isi{adverbial} adjuncts throughout the history of Swedish, and investigates how present-day Swedish \isi{adverbial} infinitives are translated into the \isi{traditional dialect} of \ili{Övdalian}. The results show that the emergence of \isi{adverbial} infinitives is a gradual process in many ways similar to the corresponding development in \ili{Romance} languages. While \is{purposive (infinitive)}purposive infinitives were already attested in \ili{Old Swedish}, other \isi{adverbial} infinitives emerged during the post-Reformation era, starting with \is{abessive (infinitive)}abessive and \is{substitutive (infinitive)}substitutive infinitives in the 17\textsuperscript{th} century, followed by temporal and instrumental infinitives in the late 18\textsuperscript{th} and early 19\textsuperscript{th} centuries, respectively. This was a dynamic period in the history of Swedish, characterized by a substantial functional expansion of the written language and the subsequent formal elaborations of the written form (i.e. language \isi{Ausbau}). The contrastive investigation of \ili{Övdalian} translations of Swedish texts suggests that \isi{adverbial} infinitives are not used in the \isi{dialect}, with the exception of \is{purposive (infinitive)}purposive infinitives. Instead, other strategies are used in order to express the same \isi{adverbial} notions. As opposed to \ili{Standard Swedish}, \ili{Övdalian} has not existed as a literary (written) language until recently. I have taken the absence of \isi{adverbial} infinitives in \ili{Övdalian} as an indication that they originally were part of a written rather than spoken register. I take the emergence of \isi{adverbial} infinitives in Swedish to be part of a process of \isi{Verschriftlichung} and language \isi{Ausbau}. 



The article also explores the possibility that \isi{adverbial} infinitives are the result of \isi{language contact} with \ili{High German} and \ili{French}, as suggested by \citet{Holm1967}. Since we know of other instances of syntactic influence from \ili{High German} on Swedish, but of mainly lexical influence from \ili{French}, \ili{High German} stands out as the more probable candidate in this regard. Moreover, it is quite clear that some of the \ili{Old Swedish} \isi{participle} constructions which are used in much the same way as present-day Swedish \isi{adverbial} infinitives emerged under the influence of \ili{Latin}. It is, however, not evident that the same is true for the later emergence of infinitival constructions. \is{adverbial}Adverbial infinitives are common in European languages, and even though it cannot be ruled out that the Swedish constructions are (partly) due to \isi{language contact}, there is also the possibility that they emerged independently as a result of the elaboration and specialization of the written language during the Early Modern period. The empirical findings of this investigation show that the relative order in which the \isi{adverbial} infinitives appeared appears to convey a more general pattern of how \isi{adverbial} constructions tend to evolve cross-linguistically (especially with regard to the initial and final stages of the process). Even if the emergence of the first \isi{adverbial} infinitives is the result of \isi{language contact} (in that case an example of \isi{replica grammaticalization}), the subsequent \isi{grammaticalization} of other \isi{adverbial} infinitives could still be independent. In other words, the possibility of \isi{language contact} does not exclude that the subsequent \isi{grammaticalization} process was independent and in line with cross-linguistic patterns. 


\section*{Abbreviations}
\begin{tabbing}
EOS\hspace{1ex}\= \ili{Early Old Swedish}\kill
EOS \> \ili{Early Old Swedish}\\
IM  \> \isi{infinitive marker}\\
LOS \> \ili{Late Old Swedish}\\
\end{tabbing}


\section*{Quoted texts}
\begin{description}[font=\normalfont]\sloppy
\item[1dSMUL:] Modée, Reinhold Gustaf (b. 1698). \textit{Håkan Smulgråt} [Håkan Cheapskate]. Stockholm, 1739. See  \citet{MarttalaStromquist2001}. Available through LB.
\item[2aSTERBH:] Kexél, Olof (b. 1748). \textit{Sterbhus-kammereraren Mulpus eller Caffe-huset i Stora Kyrkobrinken} [The chief accountant of the estate Mulpus or the coffee house in the main church hill]. Stockholm, 1776. See \citet{MarttalaStromquist2001}. Available through LB.
\item[2cKUSINE:] Envallsson, Carl (b. 1756). \textit{Kusinerna, eller: Fruntimmers-sqvallret} [The cousins or the gossip of the women]. Stockholm, 1807. See \citet{MarttalaStromquist2001}. Available through LB.
\item[ApG:] Apostla Gerningar [Acts of the Apostles]. 1385. In \textit{Klosterläsning}. Edited by G. E. Klemming. (Samlingar utgivna av Svenska fornskriftssällskapet 22.) Stockholm, 1877–1878. Available through FTB\slash \isi{Korp}.
\item[Blogg:] Bloggmix [A selection of Swedish blogs]. 1998–2017. Available through \isi{Korp}.
\item[Bol:] Bolinus, Andreas (b. 1642). \textit{En dagbok från 1600-talet} [A diary from the 17\textsuperscript{th} century]. Edited by E. Brunnström. Stockholm, 1913. 
\item[Leg:] Stephens, George (ed.). 1847. \textit{Svenska medeltidens kloster- och helgonabok} […] \textit{Ett forn-svenskt legendarium} […] [The Swedish medieval book of monasteries and saints … An \ili{Old Swedish} collection of legends]. Stockholm: Norstedts. Originally written some time between 1276 and 1307. Earliest manuscript (Codex Bureanus) from ca. 1350. Available through FTB\slash \isi{Korp}.
\item[Ehrensvärd:] \textit{C. A. Ehrensvärds brev} [The letters of C. A. Ehrensvärd]. Auth. b. 1745. Edited by Gunhild Bergh. Stockholm, 1916.
\item[Hunden:] Ekman, Kerstin. 1986. \textit{Hunden} [The dog]. Stockholm: Bonnier. 
\item[Joh.:] \textit{Johannesevangeliet} [The gospel of John]. In \textit{Bibel 2000}. Gothenburg: Bokförlaget Cordia AB, 1999. 
\item[Jua.:] \textit{Juanneswaundsjilą: Johannesevangeliet på älvdalska} [The gospel of John in \ili{Övdalian}]. Knivsta: Lars Steensland, 1989.
\item[KM:] Karl Magnus saga [The story of Karl Magnus]. End of 14\textsuperscript{th} century. In \textit{Prosadikter från Sveriges medeltid}. Edited by G. E. Klemming. (Samlingar utgivna av Svenska fornskriftsällskapet 28.) Stockholm, 1887–1889. Available through FTB\slash \isi{Korp}.
\item[Mål-roo:] Columbus, Samuel (b. 1642). \textit{Mål-Roo eller roo-mål} [Language amusement or amusing language]. Ca. 1675. Edited by Bengt Hesselman. (Nordiska texter och undersökningar 6.) Stockholm: Hugo Geebers förlag, 1935. Available through FTB\slash \isi{Korp}.
\item[Måns:] \textit{Peder Månssons brev på svenska från Rom till \isi{Vadstena} kloster 1508–1519} [Peder Månsson’s letters in Swedish from Rome to \isi{Vadstena} monastery 1508–1519]. Auth. b. ca. 1460. Edited by Robert Geete. (Småstycken på forn-svenska. Andra serien.) Stockholm, 1915.
\item[Rattsjin:] Ekman, Kerstin. 2000. \textit{Rattsjin} [The dog]. Älvdalen: Juts böcker. 
\item[Rålamb:] Callmer, Christian (ed.). 1963. \textit{Diarium under resa till Konstantinopel 1657–1658} [`Diary during a journey to Constantinople 1657–1658’, undertaken by Claes Rålamb (b. 1622)]. (Historiska handlingar 37:3.) Stockholm: Norstedts.
\item[UL:] Schlyter, Carl Johan (ed.). 1834. \emph{Samling af Sweriges gamla lagar. Tredje bandet.} \textit{Uplands-Lagen} [Collection of the old laws of Sweden. Volume three. The law of Uppland]. Stockholm: Norstedts. Originally written in 1296. Available through FTB\slash \isi{Korp}.
\item[von Unge:] von Unge, Otto Sebastian (b. 1797). \textit{Vandring genom Dalarne, jemte Författarens Resa söderut} [Walk through Dalecarlia, as well as the author’s journey to the south]. Stockholm: Z. Hæggström, 1829. Available through LB.
\item[\isi{EWL}:] Collin, Hans Samuel \& Carl Johan Schlyter (eds.). 1827. \emph{Samling af Sweriges gamla lagar. Första bandet. Westgötalagen} [Collection of the old laws of Sweden. Volume one. The Westrogothic law]. Stockholm: Z. Haeggström. Originally written in the 1220s. Available through FTB\slash \isi{Korp}.
\item[ÖgL:] Collin, Hans Samuel \& Carl Johan Schlyter (eds.). 1830. \emph{Samling af Sweriges gamla lagar. Andra bandet.} \textit{Östgöta-Lagen} [Collection of the old laws of Sweden. Volume two. The Ostrogothic law]. Stockholm: Norstedts. Originally written in the 1280s. Available through FTB\slash \isi{Korp}.
\end{description}


\section*{Electronic corpora}
\begin{description}[font=\normalfont]
\item[FTB:] \isi{Fornsvenska textbanken} [The text bank of \ili{Old Swedish}]:  \url{https://project2.sol.lu.se/fornsvenska} 
\item[\isi{Korp}:] \url{https://spraakbanken.gu.se/korp/?mode=all_hist}
\item[LB:] The Swedish literature bank: \url{http://www.litteraturbanken.se}
\end{description}

{\sloppy\printbibliography[heading=subbibliography,notkeyword=this]}
\end{document}
