\addchap{Acknowledgments}
An academic monograph, such as this one, may have a single author, but is really a project dependent on the support of many more people.

First of all, I wish to thank my three PhD supervisors:\footnote{The work presented in this book is based on my doctoral dissertation, which was accepted by the University of Cologne's Faculty of Arts and Humanities in 2016.} Gerrit Dimmendaal for all the time he has dedicated to mentoring my work, for the cordial reception at the Institute for African Studies and Egyptology and for believing in my abilities from the very first moment on; Martin Becker for hours of discussion and answering tons of e-mails on the mind-twisting issues of tense, aspect and modality; and Silvia Kutscher for motivating me to pursue a doctorate and for channelling my work.

I am further indebted to Dan King and Helen Eaton of SIL International for access to their data and for lengthy discussions of Nyakyusa, to Dörte Borchers for encouraging me to continue working on this language and to Robert Botne, Thera Crane, Tiffany Kershner, Axel Fleisch and Frank Seidel for debating all things tense and aspect.

My trips to Lwangwa would not have been possible without the accommodation offered by Kai and Susanne Hoffmann. Further thanks go to Marcelo and Melanie Reimer and to Heinke Schimanowski for their help in times of need.

Nancy Winters and Martin Werner helped me a great deal with my first steps in the village and have provided great companionship far from home. My first contacts in the Nyakyusa community were established with the help of Astol Benson and Anthon Mwangake. Martin Mwakaje has been a great help in finding further speakers for text collecting.

Descriptive linguistic research greatly relies on the cooperation of native speakers. I am thankful to all my language assistants, especially Elisha Mwakyoma, who has been so patient from my first steps in empirical linguistics on, and Herbert Zabron \lq\lq The Professor'' Mwaikema, who has been a dedicated language teacher.

\largerpage
The Institute for African Studies and Egyptology with all of its members has been a welcoming and supportive environment. Further thanks go to Monica Feinen, who has provided the map used in this study.

I also wish to express my gratitude towards a.r.t.e.s. Graduate School for the Humanities Cologne for the great working conditions and for awarding me a generous scholarship that made this research possible.

Further thanks go to Lee Bickmore and Sebastian Nordhoff for providing helpful feedback and for guiding me through the publication process. This book has also benefited greatly from the comments of the two anonymous reviewers. I wish to thank Mary Chambers and the Language Science Press community proofreaders for improving my non-native English.

Last but not least, I wish to thank all my friends, who lifted my spirits when writing or fieldwork strained my nerves, especially Alexandra, Anne, Benjamin, Eka, Hares, Jens and Willi.


